### 2016-12-28 pivot/paper.tex
Add Ecoop draft

\documentclass[a4paper,UKenglish]{lipics-v2016}
%This is a template for producing LIPIcs articles. 
%See lipics-manual.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
% for section-numbered lemmas etc., use "numberwithinsect"

\usepackage{microtype}%if unwanted, comment out or use option "draft"
\usepackage{lipsum}
\usepackage{semantic}

\bibliographystyle{plainurl}% the recommended bibstyle

% ==================================================================================================

\title{Data exploration through dot-driven development\footnote{This 
  work was supported by Google Digital News Initiative.}}
\titlerunning{Data exploration through dot-driven development}

\author[1]{Tomas Petricek}
\affil[1]{The Alan Turing Institute, London, UK\\
  \texttt{tomas@tomasp.net}}
\authorrunning{T. Petricek}
\Copyright{Tomas Petricek}
\subjclass{D.3.2 Very high-level languages}
\keywords{data science, type providers, pivot, aggregation}

\EventEditors{John Q. Open and Joan R. Acces}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}

% ==================================================================================================

% Formatting for source code & types
\definecolor{cmtclr}{rgb}{0.0,0.6,0.0}
\definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
\definecolor{numclr}{rgb}{0.0,0.4,0.0}
\definecolor{strclr}{rgb}{0.4,0.3,0.0}
\definecolor{prepclr}{rgb}{0.6,0.0,0.2}

\newcommand{\langl}{\begin{picture}(4.5,7)
\put(1.1,2.5){\rotatebox{60}{\line(1,0){5.5}}}
\put(1.1,2.5){\rotatebox{300}{\line(1,0){5.5}}}
\end{picture}}
\newcommand{\rangl}{\begin{picture}(4.5,7)
\put(.9,2.5){\rotatebox{120}{\line(1,0){5.5}}}
\put(.9,2.5){\rotatebox{240}{\line(1,0){5.5}}}
\end{picture}}

\newcommand{\lsep}{~|~}
\newcommand{\num}[1]{\textcolor{numclr}{#1}}
\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
\newcommand{\ident}[1]{\textnormal{\sffamily #1}}
\newcommand{\qident}[1]{\textnormal{\sffamily \guillemotleft #1\guillemotright}}
% \newcommand{\qident}[1]{\textnormal{\sffamily ‹#1›}}

\newcommand{\dom}{\ident{dom}}

% ==================================================================================================

% TODO: Introduce pivot type provider somewhere early

\begin{document}
\maketitle

\begin{abstract}
Data literacy is becoming increasingly important in the modern world. While spreadsheets make 
simple data analytics accessible to a large number of people, creating transparent scripts that 
can be checked, modified, reproduced and formally analyzed requires expert programming skills. 
In this paper, we describe the design of a data exploration language that makes the task more 
accessible by embedding advanced programming concepts in a simple core language.

The core language uses type providers, but we employ them in a novel way -- rather than providing 
types with members for accessing data, we provide types with members that allow the user to access 
data, but also compose complex queries using only member access (``dot''). This lets us recreate 
functionality that usually requires complex type systems (row polymporphism, dependent typing) 
in an extremely simple object-based language.

We formalize our approach using a minimal object-based calculus and prove that
programs constructed using the provided types represent valid data transformations. We then discuss 
a prototype implementation of the language, together with a simple editor that bridges
some of the gaps between programming and spreadsheets. We believe that this provides a pathway towards
democratizing data science -- our use of type providers significantly reduce the complexity of 
languages that one needs to understand in order to write scripts for exploring data.
\end{abstract}

% ==================================================================================================

\section{Introduction -- Simplifying data exploration}
\label{sec:intro}

The rise of big data and open data initiatives means that there is an increasing amount of raw data 
available. At the same time, the fact that ``post-truth'' was chosen as the word of 2016 suggests 
that there has never been greater need for increasing data literacy and building tools that let 
anyone -- including journalists and interested citizens -- explore such data and use it to support facts.

Spreadsheet tools made data exploration accessible to a large number of people, but operations 
performed in a spreadsheet application cannot be reproduced or replicated with different 
input parameters -- the manual mode of interaction is not repeatable and it breaks the link with 
the original data source. The answer to this problem is to explore data programmatically. A program 
can be run repeatedly and its parameters can be modified.

However, even with the programming tools generally accepted as simple, exploring data is 
surprisingly difficult. For example, consider the following Python program (using the pandas
library), which reads a list of all Olympic medals ever awarded (see Appendix~A) and finds top 8 
athletes by the number of gold medals they won in Rio 2016:

\noindent
\begin{equation*}
\begin{array}{l}
\ident{olympics} = \ident{pd.read\_csv}(\str{olympics.csv})\\[0.5em]
\ident{olympics}[\ident{olympics}[\str{Games}]==\str{Rio (2016)}]\\
\qquad .\ident{groupby}(\str{Athlete})\\
\qquad .\ident{agg}(\left\{\str{Gold}:\ident{sum}\right\})\\
\qquad .\ident{sort\_values}(\ident{by}=\str{Gold}, \ident{ascending}=\kvd{False})\\
\qquad .\ident{head}(\num{8})
\end{array}
\end{equation*}

\noindent
The code is short and easy to understand, but writing or modifying it requires the user to 
understand intricate details of Python and be well aware of the structure of the data source.

The short example specifies operation parameters in three different ways -- indexing $[\ldots]$ 
is used for filtering; aggregation takes a dictionary $\left\{\ldots\right\}$ and sorting uses 
optional parameters. The dynamic nature of Python makes the code simple, but it also means that
auto-completion on member names (after typing dot) is not commonplace and so fining the operation
names (\ident{groupby}, \ident{sort\_values}, \ident{head}, ...) often requires using a search 
engine. Furthermore, column names are specified as strings and so the user often needs to refer back
to the structure of the data source and be careful to avoid typos.

The language presented in this paper reduces the number of language features by making member access
the primary mechanism. For example, finding top 8 athletes by the number of gold medals in Rio 2016 
can be written as (in code $\qident{\ldots}$ is written as \textquotesingle$\ldots$\textquotesingle):
%
\begin{equation*}
\begin{array}{l}
\ident{olympics}\\
\quad.\qident{filter data}.\qident{Games is}.\qident{Rio (2016)}.\ident{then}\\
\quad.\qident{group data}.\qident{by Athlete}.\qident{sum Gold}.\ident{then}\\
\quad.\qident{sort data}.\qident{by Gold descending}.\ident{then}\\
\quad.\qident{paging}.\ident{take}(\num{8})\\
\end{array}
\end{equation*}

\noindent
The language is object-based with nominal typing. This means that auto-completion can 
provide list of available members when writing and modifying code. The members (such as 
\qident{by Gold descending}) are provided by a type provier based on the knowledge of
the data source and transformations applied so far -- thus only valid and meaningful 
operations are provided. The rest of the paper gives a detailed description of the mechanism.

\subparagraph{Contributions.} This paper explores an interesting corner in the programming
language design space. We support it by a practical implementation and a formalization with
a correctness proof. Our specific contributions are:

\begin{itemize}
\item We use type providers in a new way (Section~\ref{sec:tps}). Previous work focused on providing 
  members for direct data access. In contrast, our pivot type provider (Section~\ref{sec:pivot}) lazily 
  provides types with members that can be used for composing queries, making it possible to perform
  entire date exploration through single programming mechanism (Section~\ref{sec:analysis-dot}).

\item Our mechanism shows how to embed fancy types into a simple nominally-typed programming  
  language (Section~\ref{sec:analysis-col}). Our proof-of-concept language tracks names and types of 
  available columns in the data set (inspired by row types), but the same mechanism can be used to 
  implement other features in any Java-like language.

\item We implement the language outlined in this paper (\url{github.com/the-gamma}) and make 
  it available as a JavaScript component that can be used to build open and transparent 
  data-driven visualizations (\url{thegamma.net}) and discuss the implementation (Section~\ref{sec:impl}).

\item We formalize the language (Section~\ref{sec:foo}) and the type provider for data exploration
  (Section~\ref{sec:pivot}) and show that queries constructed using the type provider are valid
  (Section~\ref{sec:proofs}). Our formalization also covers the laziness of type providers, which
  is an important aspect not covered in the existing literature.
\end{itemize}

% ==================================================================================================

\section{Background -- Type providers}
\label{sec:tps}

The work presented in this paper consists of a simple nominally-typed host language and the pivot
type provider, which provides types with members that can be used to construct and execute queries
against an external data source. This section briefly reviews the existing work on type providers
and explains what is new about the pivot type provider.

\subparagraph{Information-rich programming.} Type providers were first presented as a mechanism
for providing type-safe access to rich information sources. According to Syme et al. [X],
a type provider is a compile-time component that imports external information source into a
programming language. It provides two things to the compiler (or editor) hosting it. First, it 
provides a type signature that models the external source using structures understood by the
host language (e.g.~types with members). Second, it provides an implementation for the signatures
which accesses data from the external source.

For example, the World Bank type provider [X] provides a fine-grained access to development 
indicators about countries of the world. The following accesses CO2 emissions in 2010 for all 
available countries:
%
\begin{equation*}
\begin{array}{l}
\ident{world}.\ident{byYear}.\qident{2010}.\qident{Climate Change}.\qident{CO2 emissions (kt)}
\end{array}
\end{equation*}

\noindent
The provided schema consists of types with members such as \qident{CO2 emissions (kt)} and \qident{2010}.
The members are generated by the type provider based on the meta-data provided by the World Bank.
The second part provided by the type provider is code that is executed when the above code is run.
For the example above:
%
\begin{equation*}
\begin{array}{l}
\ident{series.create}(\str{CO2 emissions (kt)},\str{Year},\str{Value},\\
\qquad \ident{world.getByYear}(\num{2010},\str{EN.ATM.CO2E.KT}))
\end{array}
\end{equation*}

\noindent
Here, the provided code relies on runtime library that provides a representation of a series 
(mapping from keys to values) and the \ident{getByYear} function that downloads data for a specified
indicator represented by a code. The type provider provides a type-safe access to known indicators,
which exist only as strings in the compiled code, increasing safety and making data access easier
thanks to auto-completion (which offers a list of available indicators).

\subparagraph{Types from data.} Recent work on the F\# Data library [X] uses type providers for 
accessing data in structured formats such as XML, CSV and JSON. This is done by inferring the 
structure of the data from a sample document, provided as a static parameter to a type provider.
In the following example (adapted from [X]), a sample URL is passed to \ident{JsonProvider}:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Weather} = \ident{JsonProvider}\langl\str{http://api.owm.org/?q=London}\rangl \\[0.5em]
 \kvd{let}~\ident{ldn} = \ident{Weather.GetSample}()\\
 \ident{printfn}~\str{The temperature in London is \%f}~\ident{ldn.Main.Temp}
\end{array}
\end{equation*}

\noindent
As in the World Bank example, the JSON type provider generates types with members that let us access
data in the external data source -- here, we access the temperature using \ident{ldn.Main.Temp}. 
The provided code attempts to access the corresponding nested field and convert it to a number.
The relative safety property guarantees that this will not fail if the sample is representative
of the actual data loaded at runtime.

\subparagraph{Pivot type provider.} The pivot type provider presented in this paper follows the 
same general mechanism as the F\# type providers discussed above, although it is embedded in a 
simple language that runs in a web browser.

The main difference between our work and the type providers discussed above is that we do not use
type providers for mapping external data sources into the type system (by providing members that
correspond to parts of the data). Instead, we use type providers for (lazily) generating types 
with members that let users compose type-safe queries over the data source. As discussed in 
Section~\ref{sec:foo}, we model the provided code by a relational algebra.

This means that our use of type providers is more akin to meta-programming or code generation with
one important difference -- the schema provided by the pivot type provider is potentially infinite
(as there are always more operations that can be applied). To support this, we use the fact that
type providers are integrated into the type system and types can be provided lazilly. This is 
also a new aspect of our formalization in Section~\ref{sec:foo}.

% ==================================================================================================

\section{Analysis -- What makes data exploration scripts complex}
\label{sec:analysis}

We started by contrasting a data exploration script using a popular Python library pandas 
with a script that can be written using the pivot type provider (Section~\ref{sec:intro}). In this
section, we analyze what makes the Python code complex and we discuss how our alternative design
aims to simplify it. We use Python as a concrete example, but other commonly used tools (mentioned
throughout) share the same properties:

\begin{enumerate}
\item The filtering operation is written using indexing $[\ldots]$ while all other operations are
  written using member invocation with (optionally named) parameters. In the first case, we write
  an expression $\ident{olympics}[\str{Games}]==\str{Rio (2016)}$ returning a vector of Booleans
  while in the other, we specify a parameter value using $\ident{by}=\str{Gold}$. In other languges,
  a parameter can also be a lambda function specifying a predicate or a transformation.

\item The aggregation operation takes a dictionary $\left\{\ldots\right\}$, which is yet another
  concept the user needs to understand. Here, it lets us specify one or more aggregations to
  be applied over a group. A way of specifying multiple operations or results is common in other 
  languages. For example, anonymous types in LINQ play the same role.

\item The editor tooling available for Python is limited -- editors that provide auto-completion
  rely on a mix of advanced static analysis and simple (not always correct) hints and often fail
  for chained operations such as the one in our example\footnote{For an anecdotal evidence,
  see for example: \url{stackoverflow.com/questions/25801246/}}. Statically-typed languages
  provide better tooling, but at the cost of higher complexity\footnote{Again,
  See \url{fslab.org/Deedle}
  and \url{extremeoptimization.com/Documentation/Data_Frame/Data_Frames.aspx} }.

\item In the Python example (as well as in most other data manipulation 
  libraries\footnote{deedle, etc.}), column names are specified as strings. This makes static 
  checking of column names and auto-completion help difficult. For example, \str{Gold} is a valid 
  column name when calling \ident{sort\_values} -- we know that because it is a key of the 
  dictionary passed to \ident{agg} before.
\end{enumerate}

\noindent
We use the above four points as guiding principles for the design discussed in the rest of this
section. We unify as many distinct languages constructs as possible by making member
access the primary operation (Section~\ref{sec:analysis-dot}); we use simple nominal typing to
enable auto-completion (Section~\ref{sec:analysis-auto}); we track column names using the pivot
type provider (Section~\ref{sec:analysis-col}) and we use operation-chaining via member access
for constructing dictionaries (Section~\ref{sec:analysis-and}).

% --------------------------------------------------------------------------------------------------

\subsection{Unifying language constructs with member access}
\label{sec:analysis-dot}

We unify this via dot

Could this be done without type provider?

\subsection{Dot-driven development}
\label{sec:analysis-auto}

  ○ We use simple nominal typing
\subsection{Column names}
\label{sec:analysis-col}

The last two points are more interesting -- proponents of static typing would correctly
point out that both member names and column names can be tracked by a more sophisticated type
system. However, powerful types usually increase the complexity of the programming language.

  ○ We use type providers to provide available options
  (e.g. something)
\subsection{Dictionary syntax}
\label{sec:analysis-and}
  ○ We create the "and blah. and blah. then" pattern
  (e.g. grouping and sorting)

~
\begin{equation*}
\begin{array}{l}
\ident{olympics}\\
\quad.\qident{filter data}.\qident{Games is}.\qident{Rio (2016)}.\ident{then}\\
\quad.\qident{group data}.\qident{by Athlete}.\qident{sum Gold}.\ident{then}\\
\quad.\qident{sort data}.\qident{by Gold descending}.\ident{then}\\
\quad.\qident{paging}.\ident{take}(8)\\
\quad.\qident{get series}.\qident{with key Athlete}.\qident{and value Gold}
\end{array}
\end{equation*}

Introduction

\begin{equation*}
\begin{array}{l}
\ident{olympics}\\
\quad.\qident{filter data}.\qident{Games is}.\qident{Rio (2016)}.\ident{then}\\
\quad.\qident{group data}.\qident{by Athlete}.\qident{sum Gold}.\ident{then}\\
\quad.\qident{sort data}.\qident{by Gold descending}.\ident{then}\\
\quad.\qident{paging}.\ident{take}(8)\\
\quad.\qident{get series}.\qident{with key Athlete}.\qident{and value Gold}
\end{array}
\end{equation*}

Details
 - Grouping
 - Filtering
 - Joining

\section{Formalising The Gamma script}
\label{sec:foo}

Target language is super simple OO language, but we can use it to encode very fancy types!!

\subsection{Relational algebra}
\begin{equation*}
\begin{array}{rclll}
  R & = & \Pi_{f_1, \ldots, f_n} (R) &&\textnormal{get columns}\\
   & \lsep & \sigma_\varphi (R) &&\textnormal{filter by predicate}\\
   & \lsep & \tau_{f_1, \ldots, f_n}(R) &&\textnormal{sort by fields}\\
   & \lsep & \Phi_{f, \rho_1, \ldots, \rho_n} (R) &&\textnormal{group by $f$ and aggregate}\\
   \\
  \rho & = & \ident{count}  \\
   & \lsep & \ident{sum}~f    \\       
   & \lsep & \ident{dist}~f\\
   & \lsep & \ident{conc}~f\\
\end{array}
\end{equation*}
\subsection{Foo calculus}
\begin{equation*}
\begin{array}{rcl}
 \tau &=& \ident{num} \lsep \ident{string} \lsep \ident{series}\langl \tau_1, \tau_2 \rangl \lsep \ident{Query}\\
 l &=& \kvd{type}~C(\overline{x:\tau}) = \overline{m} \\[0.0em]
 m &=& \kvd{member}~N:\tau= e
\end{array}
\end{equation*}

\begin{equation*}
\inference[(foo)]
  {L_1; \Gamma \vdash e : C; L_2 \qquad ((\kvd{type}~C(\overline{x:\tau}) = ..\;\kvd{member}~N_i : \tau_i = e_i\;..), L) \in L_2}
  {L_1; \Gamma \vdash e.N_i:\tau_i; L_2 \cup L}
\end{equation*}

\newpage

\begin{figure}
\begin{equation*}
\begin{array}{ll}
\ident{main}(F) = C \mapsto (l, L_1 \cup \ldots \cup L_4)\\[0.5em]
\quad l = \kvd{type}~C(x:\ident{Query}) = \\
\qquad \qquad \kvd{member}~\qident{drop fields} : C_1 = C_1(x) &\textnormal{where}~C_1, L_1 = \ident{drop}(F)\\
\qquad \qquad \kvd{member}~\qident{sort data} : C_2 = C_2(x) &\textnormal{where}~C_2, L_2 = \ident{sort}(F)\\
\qquad \qquad \kvd{member}~\qident{group data} : C_3 = C_3(x) &\textnormal{where}~C_3, L_3 = \ident{group}(F)\\
\qquad \qquad \kvd{member}~\qident{get series} : C_4 = C_4(x) &\textnormal{where}~C_4, L_4 = \ident{get-key}(F)\\
\\
\ident{drop}(F) = C, \{l\} \cup \bigcup L_f\\[0.5em]
\quad l = \kvd{type}~C(x:\ident{Query}) = \\
\qquad \qquad \kvd{member}~\ident{then} : C = C(x)                 &\textnormal{where}~C, L = \ident{main}(F)\\
\qquad \qquad \kvd{member}~\qident{drop $f$} : C_f = &\forall f \in \dom(F)~\textnormal{where}~C_f, L_f = \ident{drop}(F')\\
\qquad \qquad \qquad  C_f(\Pi_{\ident{dom}(F')}(x))& \quad\textnormal{and}~ F' = \{ f'\mapsto\tau'\in F, f' \neq f\}\\
\end{array}
\end{equation*}

\caption{Foo}
\label{fig:foo}
\end{figure}

\begin{figure}
\begin{equation*}
\begin{array}{ll}
\ident{get-key}(F) = C, \{l\} \cup \bigcup L_f \\[0.25em]
\quad l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)~\textnormal{where} \\
\qquad \qquad \kvd{member}~\qident{with key $f$} : C_f = C_f(x) &\quad C_f, L_f = \ident{get-val}(F, f)\\
\\
\ident{get-val}(F, f_k) = C, \{l\} \cup \bigcup L_f \\[0.25em]
\quad l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)\setminus\{f\}~\textnormal{where} \\
\qquad \qquad \kvd{member}~\qident{and value $f$} : C_f = C_f(\Pi_{f_k, f_v}(x)) &\quad  C_f~\textnormal{defined as below} \\
\quad l_f = \kvd{type}~C_f(x:\ident{Query}) = &\quad\textnormal{and}~\tau_k, \tau_v~\textnormal{such that} \\
\qquad \qquad \kvd{member}~\ident{series} : \ident{series}\langl\tau_k, \tau_v\rangl = x &\quad \tau_k = F(f_k), \tau_v = F(f) &\\
\end{array}
\end{equation*}

\caption{Foo}
\label{fig:foo}
\end{figure}

\section{Pivot type provider}
\label{sec:pivot}

Omit paging, because it's boring

Our formalization of the pivot type provider follows the style used by Petricek et al. \cite{fsdata}
in their formalization of F\# Data. In our case, a type provider is a function that takes a schema
$F$ (mapping from field names to field types) and produces a class $C$ together with collection 
of other class defintins $L$ that are used by $C$:
%
\begin{equation*}
\ident{provider}(F) = C, L    ~~\textnormal{where}~F=\{ f_1 \mapsto \tau_1, \ldots, f_n \mapsto \tau_n \}
\end{equation*}
%
The provided class $C$ has a constructor taking \ident{Query}. It may be a class with further 
members that allow refining the query or a class with a single member \ident{series} that returns
a value of type $\ident{series}\langl \tau_1, \tau_2 \rangl$. In the second case, we can evaluate
the provided transformation on input $R$ using an expression $(\kvd{new}~C(R)).\ident{series}$.

\subparagraph{Top-level type.} The top-level type allows the user to select a transformation. The
members return objects with members that allow specifying further properties of each transformation.

\subparagraph{Sorting.} yadda

\begin{figure}
\begin{equation*}
\begin{array}{ll}
\ident{sort}(F) = C, \{l\} \cup \bigcup L_f \cup \bigcup L'_f\\[0.25em]
\quad l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)~\textnormal{where} \\
\qquad \qquad \kvd{member}~\qident{by $f$ descending} : C_f = C_f(x) &\quad C_f, L_f = \ident{sort-and}(F, \{ f \mapsto \ident{desc} \})\\
\qquad \qquad \kvd{member}~\qident{by $f$ ascending} : C'_f = C'_f(x) &\quad C'_f, L'_f = \ident{sort-and}(F, \{ f \mapsto \ident{asc} \})\\
\\[0.5em]
\ident{sort-and}(F,S) = C, \{l\} \cup \bigcup L_f \cup \bigcup L'_f\\[0.25em]
\quad l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)\setminus\dom(S)~\textnormal{where} \\
\qquad \qquad \kvd{member}~\qident{and $f$ descending} : C_f = C_f(x) &\quad C_f, L_f = \ident{sort-and}(F, S\cup\{ f \mapsto \ident{desc} \})\\
\qquad \qquad \kvd{member}~\qident{and $f$ ascending} : C'_f = C'_f(x) &\quad C'_f, L'_f = \ident{sort-and}(F, S\cup\{ f \mapsto \ident{asc} \})\\
\qquad \qquad \kvd{member}~\ident{then} : C = C(\tau_S(x))                 &\textnormal{where}~C, L = \ident{main}(F)\\
\end{array}
\end{equation*}
\caption{Foo bar}
\label{fig:foobar}
\end{figure}

\begin{figure}
\begin{equation*}
\begin{array}{ll}
\ident{group}(F) = C, \{l\} \cup \bigcup L_f \\[0.25em]
\quad l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)~\textnormal{where} \\
\qquad \qquad \kvd{member}~\qident{by $f$} : C_f = C_f(x) &\quad C_f, L_f = \ident{agg}(F, \{ f \mapsto F(f) \}, \emptyset)\\
\\[0.5em]
\ident{agg}(F,S,G) = C, \{l\} \cup \bigcup L_f \cup \bigcup L'_f\\[0.25em]
\quad l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)\setminus\dom(S)~\textnormal{where} \\
\qquad \qquad \kvd{member}~\qident{distinct $f$} : C_f = C_f(x) &\quad C_f, L_f = \ident{agg}(F, S\cup\{ f \mapsto \ident{num}, G\cup\{\ident{dist}~f\} \})\\
\qquad \qquad \kvd{member}~\qident{sum $f$} : C'_f = C'_f(x) &\quad C'_f, L'_f = \ident{agg}(F, S\cup\{ f \mapsto \ident{num} \}, G\cup\{\ident{sum}~f\})\\
\qquad \qquad \kvd{member}~\qident{concat $f$} : C'_f = C'_f(x) &\quad C'_f, L'_f = \ident{agg}(F, S\cup\{ f \mapsto \ident{string} \}, G\cup\{\ident{conc}~f\})\\
\qquad \qquad \kvd{member}~\qident{count all} : C = C(\tau_S(x))                 &\textnormal{where}~C, L = \ident{main}(F)\\
\qquad \qquad \kvd{member}~\ident{then} : C = C(\tau_S(x))                 &\textnormal{where}~C, L = \ident{main}(F)\\
\end{array}
\end{equation*}
\caption{Foo bar 2}
\label{fig:foobar2}
\end{figure}

!!

TODO: Somehow delay $L$

!!

We only generate sorting methods for fields not used yet

Helper \ident{sort-and} also takes sorting keys already used

\section{Properties}
\label{sec:proofs}

\section{Implementation}
\label{sec:impl}

~
\newpage

\subparagraph*{Acknowledgements.}

I want to thank \dots

\appendix
\section{Morbi eros magna}

Morbi eros magna, vestibulum non posuere non, porta eu quam. Maecenas vitae orci risus, eget imperdiet mauris. Donec massa mauris, pellentesque vel lobortis eu, molestie ac turpis. Sed condimentum convallis dolor, a dignissim est ultrices eu. Donec consectetur volutpat eros, et ornare dui ultricies id. Vivamus eu augue eget dolor euismod ultrices et sit amet nisi. Vivamus malesuada leo ac leo ullamcorper tempor. Donec justo mi, tempor vitae aliquet non, faucibus eu lacus. Donec dictum gravida neque, non porta turpis imperdiet eget. Curabitur quis euismod ligula.

%%
%% Bibliography
%%

%% Either use bibtex (recommended),

\bibliography{lipics-v2016-sample-article}

%% .. or use the thebibliography environment explicitely

\end{document}

### 2016-12-29 pivot/paper.tex
More work on draft

@@
 \usepackage{microtype}%if unwanted, comment out or use option "draft"
-\usepackage{lipsum}
+\usepackage{pifont}
+\usepackage{fp}
+\usepackage{graphicx}
 \usepackage{semantic}
@@
 \end{picture}}
+
+\newcommand{\ball}[1]{\FPeval{\result}{clip(201+#1)}\textnormal{\ding{\result}}}
 
@@
 
-We unify this via dot
+LISP is the best example of a language that unifies many language constructs using a single
+form. In LISP, everything is an s-experssion. That is, either a list or a symbol. In contrast, 
+a typical data processing language uses a number of distinct constructs including indexers (for 
+range selection and filtering), method calls (for transformations) and named parameters (for further 
+configuration). Consider filtering and sorting:
+%
+\begin{equation*}
+\begin{array}{l}
+\ident{data}[\ident{data}[\str{Games}]==\str{Rio (2016)}] \qquad\ball{1}\\
+\ident{data}.\ident{filter}(\kvd{fun}~\ident{row}\rightarrow\ident{row}.\ident{Games}==\str{Rio (2016)})\qquad\ball{2}\\
+\ident{data}.\ident{sort\_values}(\ident{by}=\str{Gold}, \ident{ascending}=\kvd{False})\qquad\ball{3}\\
+\end{array}
+\end{equation*}
 
-Could this be done without type provider?
+\noindent
+Python and pandas uses indexers for filtering \ball{1} which can alternatively be written (e.g.~in 
+LINQ) using a method taking a predicate as a lambda function \ball{2}. Operations that are 
+parameterized only by column name, such as sorting in pandas \ball{3} are often methods with 
+named parameters.
+
+We aim to unify the above examples using a single language construct that offers a high-level programming
+model\footnote{In contrast, s-expressions in LISP are typically used as a lower-level encoding of 
+higher-level constructs. For example \ident{lambda} symbol encodes a lambda function as an s-expression.}
+and can be supported by modern tooling (as discussed in Section~\ref{sec:analysis-auto}).
+Member access provides an extremely simple programming construct that is, in conjunction with
+the type provider mechanism capable of expressing the above data transformations in a uniform way: 
+%
+\begin{equation*}
+\begin{array}{l}
+\ident{data}.\qident{sort data}.\qident{by Gold descending}\qquad\ball{1}\\
+\ident{data}.\qident{filter data}.\qident{Games is}.\qident{Rio (2016)}\qquad\ball{2}\\
+\end{array}
+\end{equation*}
+
+\noindent
+The member names tend to be longer and descriptive and we write them using \qident{...}.
+The names are not usually typed by the user (see~Section~\ref{sec:analysis-auto}) and so the
+length is not an issue when writing code. The above two examples illustrate two interesting
+aspects of our approach.
+
+\subparagraph{Members, type providers, discoverability.}
+When sorting \ball{1} the member that specifies how sorting is done includes the name of the
+column. This is possible because the pivot type provider tracks the column names (see 
+Section~\ref{sec:analysis-col}) and provides members based on the available columns suitable
+for use as sort keys. When filtering \ball{2}, the member \qident{Rio (2016)} is provided based
+on the values in the data source (we discuss this furter in Section~\ref{sec:analysis-auto}).
+
+These two examples illustrate that member access can be expressive, but it requires huge number
+of types with huge number of members. Type providers address this by integration with the type
+system (formalized in Section~\ref{sec:foo}) that discovers members lazily. This is why 
+approaches based on code generation or pre-processors would not be viable.
+
+Using descriptive member names is only viable when the names are discoverable. The above
+code could be executed in a dynamically-typed languge that allows custom message-not-understood
+handlers, but it would be impossible to get the name right when writing it. Our approach 
+relies on discovering names through auto-completion as discussed in Section~\ref{sec:analysis-auto}.
+
+\subparagraph{Are members expressive enough?}
+
+no... but it works fine! See Section~\ref{sec:impl}.
+   there are ways to make it more expressive too.
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.35,trim=0mm 0mm 0mm 0mm,clip]{filter.png} % left bottom right top
+\end{center}
+\vspace{-0.5em}
+\caption{Auto-completion offering the available values of the athlete name column}
+\label{fig:subtyping-diagram}
+\end{figure}
 
@@
 
-  ○ We use simple nominal typing
+Autocompletion - see figure.
+
+\subparagraph{Abstraction and data.}
+
+- Needs concrete data set or a sample, but that's fine
+
+\subparagraph{Values vs. types.}
+
+- provide values? Could be $\qident{Games is}(\str{Rio (2016)})$
+
 \subsection{Column names}
@@
   (e.g. something)
+
+Considerations
+
+ - renaming columns
+
 \subsection{Dictionary syntax}
@@
 
+Rio case study
+
 ~

### 2016-12-31 pivot/paper.tex
Finishing analysis

@@
 \subjclass{D.3.2 Very high-level languages}
-\keywords{data science, type providers, pivot, aggregation}
+\keywords{Data science, type providers, pivot tables, aggregation}
 
@@
 data, but also compose complex queries using only member access (``dot''). This lets us recreate 
-functionality that usually requires complex type systems (row polymporphism, dependent typing) 
-in an extremely simple object-based language.
+functionality that usually requires complex type systems (row polymporphism, phantom types, 
+dependent typing) in an extremely simple object-based language.
 
@@
 available. At the same time, the fact that ``post-truth'' was chosen as the word of 2016 suggests 
-that there has never been greater need for increasing data literacy and building tools that let 
-anyone -- including journalists and interested citizens -- explore such data and use it to support facts.
+that there has never been a greater need for increasing data literacy and building tools that let 
+anyone -- including journalists and interested citizens -- explore such data and use it to support 
+facts in a transparent way.
 
-Spreadsheet tools made data exploration accessible to a large number of people, but operations 
-performed in a spreadsheet application cannot be reproduced or replicated with different 
-input parameters -- the manual mode of interaction is not repeatable and it breaks the link with 
-the original data source. The answer to this problem is to explore data programmatically. A program 
-can be run repeatedly and its parameters can be modified.
+Spreadsheets made data exploration accessible to a large number of people, but operations 
+performed on spreadsheets cannot be reproduced or replicated with different input parameters.
+The manual mode of interaction is not repeatable and it breaks the link with the original data 
+source, making spreadsheets error-prone [X]. The answer to this problem is to explore data 
+programmatically. A program can be run repeatedly and its parameters can be modified.
 
@@
 surprisingly difficult. For example, consider the following Python program (using the pandas
-library), which reads a list of all Olympic medals ever awarded (see Appendix~A) and finds top 8 
-athletes by the number of gold medals they won in Rio 2016:
+library), which reads a list of all Olympic medals ever awarded (see Appendix~\ref{app:olympics-csv}) 
+and finds top 8 athletes by the number of gold medals they won in Rio 2016:
 
@@
 understand intricate details of Python and be well aware of the structure of the data source. 
-
 The short example specifies operation parameters in three different ways -- indexing $[\ldots]$ 
@@
 optional parameters. The dynamic nature of Python makes the code simple, but it also means that
-auto-completion on member names (after typing dot) is not commonplace and so fining the operation
-names (\ident{groupby}, \ident{sort\_values}, \ident{head}, ...) often requires using a search 
-engine. Furthermore, column names are specified as strings and so the user often needs to refer back
+auto-completion on member names (after typing dot) is not commonplace and so finding the operation
+names (\ident{groupby}, \ident{sort\_values}, \ident{head}, ...) often requires using internet 
+search. Furthermore, column names are specified as strings and so the user often needs to refer back
 to the structure of the data source and be careful to avoid typos.
@@
 The language presented in this paper reduces the number of language features by making member access
-the primary mechanism. For example, finding top 8 athletes by the number of gold medals in Rio 2016 
-can be written as (in code $\qident{\ldots}$ is written as \textquotesingle$\ldots$\textquotesingle):
+the primary programming mechanism. Finding top 8 athletes by the number of gold medals from Rio 2016 
+can be written as (in source code $\qident{\ldots}$ is written as \textquotesingle$\ldots$\textquotesingle):
 %
@@
 \noindent
-The language is object-based with nominal typing. This means that auto-completion can 
-provide list of available members when writing and modifying code. The members (such as 
-\qident{by Gold descending}) are provided by a type provier based on the knowledge of
+The language is object-based with nominal typing. This enables auto-completion that  
+provides a list of available members when writing and modifying code. The members (such as 
+\qident{by Gold descending}) are generated by a type provier based on the knowledge of
 the data source and transformations applied so far -- thus only valid and meaningful 
-operations are provided. The rest of the paper gives a detailed description of the mechanism.
+operations are offered. The rest of the paper gives a detailed analysis and description of the mechanism.
 
-\subparagraph{Contributions.} This paper explores an interesting corner in the programming
-language design space. We support it by a practical implementation and a formalization with
-a correctness proof. Our specific contributions are:
+\subparagraph{Contributions.} This paper explores an interesting particular corner of the 
+programming language design space. We support it by a detailed analysis (Section~\ref{sec:analysis}), 
+formal treatment (Section~\ref{sec:pivot}) and an implementation with 
+a case study (Section~\ref{sec:impl}). Our contributions are:
 
@@
 
-\item Our mechanism shows how to embed fancy types into a simple nominally-typed programming  
-  language (Section~\ref{sec:analysis-col}). Our proof-of-concept language tracks names and types of 
-  available columns in the data set (inspired by row types), but the same mechanism can be used to 
-  implement other features in any Java-like language.
+\item Our mechanism illustrates how to embed fancy types [X] into a simple nominally-typed programming  
+  language (Section~\ref{sec:columns}). We track names and types of available columns of the 
+  manipulated data set (using a mechanism akin to row types), but the same mechanism can be used for 
+  embedding other advanced typing schemes into any Java-like language.
 
-\item We implement the language outlined in this paper (\url{github.com/the-gamma}) and make 
-  it available as a JavaScript component that can be used to build open and transparent 
-  data-driven visualizations (\url{thegamma.net}) and discuss the implementation (Section~\ref{sec:impl}).
+\item We implement the language (\url{github.com/the-gamma}), make it available as a JavaScript 
+  component that can be used to build transparent data-driven visualizations (\url{thegamma.net}) 
+  and discuss a case study visualizing fun facts about Olympic medals (Section~\ref{sec:impl}).
 
@@
 The work presented in this paper consists of a simple nominally-typed host language and the pivot
-type provider, which provides types with members that can be used to construct and execute queries
+type provider, which generates types with members that can be used to construct and execute queries
 against an external data source. This section briefly reviews the existing work on type providers
@@
 \subparagraph{Information-rich programming.} Type providers were first presented as a mechanism
-for providing type-safe access to rich information sources. According to Syme et al. [X],
-a type provider is a compile-time component that imports external information source into a
-programming language. It provides two things to the compiler (or editor) hosting it. First, it 
-provides a type signature that models the external source using structures understood by the
-host language (e.g.~types with members). Second, it provides an implementation for the signatures
-which accesses data from the external source.
+for providing type-safe access to rich information sources. A type provider is a compile-time 
+component that imports external information source into a programming language [X]. It provides two 
+things to the compiler or editor hosting it: a type signature that models the external source 
+using structures understood by the host language (e.g.~types) and an implementation for the 
+signatures which accesses data from the external source.
 
 For example, the World Bank type provider [X] provides a fine-grained access to development 
-indicators about countries of the world. The following accesses CO2 emissions in 2010 for all 
-available countries:
+indicators about countries. The following accesses CO2 emissions by~country~in~2010:
 %
@@
 The provided schema consists of types with members such as \qident{CO2 emissions (kt)} and \qident{2010}.
-The members are generated by the type provider based on the meta-data provided by the World Bank.
+The members are generated by the type provider based on the meta-data obtained from the World Bank.
 The second part provided by the type provider is code that is executed when the above code is run.
-For the example above:
+For the example above, the code looks as follows:
 %
@@
 \noindent
-Here, the provided code relies on runtime library that provides a representation of a series 
-(mapping from keys to values) and the \ident{getByYear} function that downloads data for a specified
-indicator represented by a code. The type provider provides a type-safe access to known indicators,
-which exist only as strings in the compiled code, increasing safety and making data access easier
-thanks to auto-completion (which offers a list of available indicators).
+Here, a runtime library consists of a data series type (mapping from keys to values) and the 
+\ident{getByYear} function that downloads data for a specified indicator represented by an ID. 
+The type provider provides a type-safe access to known indicators, which exist only as strings in 
+the compiled code, increasing safety and making data access easier thanks to auto-completion 
+(which offers a list of available indicators).
 
@@
 structure of the data from a sample document, provided as a static parameter to a type provider.
-In the following example (adapted from [X]), a sample URL is passed to \ident{JsonProvider}:
+In the following example, adapted from [X], a sample URL is passed to \ident{JsonProvider}:
 %
@@
 The provided code attempts to access the corresponding nested field and convert it to a number.
-The relative safety property guarantees that this will not fail if the sample is representative
-of the actual data loaded at runtime.
+The relative safety property of the type provider guarantees that this will not fail if the sample 
+is representative of the actual data loaded at runtime.
 
@@
 The main difference between our work and the type providers discussed above is that we do not use
-type providers for mapping external data sources into the type system (by providing members that
-correspond to parts of the data). Instead, we use type providers for (lazily) generating types 
-with members that let users compose type-safe queries over the data source. As discussed in 
-Section~\ref{sec:foo}, we model the provided code by a relational algebra.
+type providers for importing external data sources (by providing members that correspond to parts 
+of the data). Instead, we use type providers for lazily generating types with members that let 
+users compose type-safe queries over the data source. As discussed in Section~\ref{sec:foo}, we 
+model the provided code by a relational algebra.
 
@@
 
-\section{Analysis -- What makes data exploration scripts complex}
+\section{Analysis -- Simplifying data scripting}
 \label{sec:analysis}
 
-We started by contrasting a data exploration script using a popular Python library pandas 
-with a script that can be written using the pivot type provider (Section~\ref{sec:intro}). In this
-section, we analyze what makes the Python code complex and we discuss how our alternative design
-aims to simplify it. We use Python as a concrete example, but other commonly used tools (mentioned
-throughout) share the same properties:
+In Section~\ref{sec:intro}, we contrasted a data exploration script written using a popular Python 
+library pandas with a script written using the pivot type provider. In this section, we analyze 
+what makes the Python code complex (Section~\ref{sec:analysis-why}) and how our design simplifies it.
 
-\begin{enumerate}
+\subsection{What makes data exploration scripts complex}
+\label{sec:analysis-why}
+
+We consider the Python example from Section~\ref{sec:intro} for concreteness, but the following 
+four points are shared with other commonly used libraries and languages. We use them to guide our
+alternative design discussed in the rest of this section.
+
+\begin{itemize}
 \item The filtering operation is written using indexing $[\ldots]$ while all other operations are
@@
   concept the user needs to understand. Here, it lets us specify one or more aggregations to
-  be applied over a group. A way of specifying multiple operations or results is common in other 
-  languages. For example, anonymous types in LINQ play the same role.
+  be applied over a group. A similar way of specifying multiple operations or results is common 
+  in other languages. For example, anonymous types in LINQ play the same role.
 
@@
   libraries\footnote{deedle, etc.}), column names are specified as strings. This makes static 
-  checking of column names and auto-completion help difficult. For example, \str{Gold} is a valid 
-  column name when calling \ident{sort\_values} -- we know that because it is a key of the 
+  checking of column names and auto-completion difficult. For example, \str{Gold} is a valid 
+  column name when calling \ident{sort\_values}, but we only know that because it is a key of the 
   dictionary passed to \ident{agg} before.
-\end{enumerate}
+\end{itemize}
 
 \noindent
-We use the above four points as guiding principles for the design discussed in the rest of this
-section. We unify as many distinct languages constructs as possible by making member
+In our design, we unify many distinct languages constructs by making member
 access the primary operation (Section~\ref{sec:analysis-dot}); we use simple nominal typing to
-enable auto-completion (Section~\ref{sec:analysis-auto}); we track column names using the pivot
-type provider (Section~\ref{sec:analysis-col}) and we use operation-chaining via member access
-for constructing dictionaries (Section~\ref{sec:analysis-and}).
+enable auto-completion (Section~\ref{sec:analysis-auto}); we use operation-chaining via member access
+for constructing dictionaries (Section~\ref{sec:analysis-and}) and we track column names using the 
+pivot type provider (Section~\ref{sec:columns}).
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.35,trim=0mm 0mm 0mm 0mm,clip]{filter.png} % left bottom right top
+\end{center}
+\caption{Auto-completion offering the available values of the athlete name column}
+\label{fig:dot-driven}
+\end{figure}
 
@@
 
-LISP is the best example of a language that unifies many language constructs using a single
-form. In LISP, everything is an s-experssion. That is, either a list or a symbol. In contrast, 
+LISP is perhaps the best example of a language that unifies many distinct constructs using a single
+form. In LISP, everything is an s-experssion, that is, either a list or a symbol. In contrast, 
 a typical data processing language uses a number of distinct constructs including indexers (for 
@@
 \noindent
-Python and pandas uses indexers for filtering \ball{1} which can alternatively be written (e.g.~in 
+Pandas uses indexers for filtering \ball{1} which can alternatively be written (e.g.~in 
 LINQ) using a method taking a predicate as a lambda function \ball{2}. Operations that are 
@@
 \begin{array}{l}
-\ident{data}.\qident{sort data}.\qident{by Gold descending}\qquad\ball{1}\\
-\ident{data}.\qident{filter data}.\qident{Games is}.\qident{Rio (2016)}\qquad\ball{2}\\
+\ident{data}.\qident{sort data}.\qident{by Gold descending}.\ident{then}\qquad\ball{1}\\
+\ident{data}.\qident{filter data}.\qident{Games is}.\qident{Rio (2016)}.\ident{then}\qquad\ball{2}\\
 \end{array}
@@
 column. This is possible because the pivot type provider tracks the column names (see 
-Section~\ref{sec:analysis-col}) and provides members based on the available columns suitable
+Section~\ref{sec:columns}) and provides members based on the available columns suitable
 for use as sort keys. When filtering \ball{2}, the member \qident{Rio (2016)} is provided based
@@
 
-\subparagraph{Are members expressive enough?}
+\subparagraph{Expressivity of members.} 
+Using member access as the primary mechanism for programming reduces the expressivity of the 
+language -- our aim is to create a domain-specific language for data exploration, rather than
+a general purpose language\footnote{Designing a general purpose language based on member access
+is a separate interesting problem.}. For this purpose, the sequential nature of member accesses 
+matches well with the sequential nature of data transformations.
 
-no... but it works fine! See Section~\ref{sec:impl}.
-   there are ways to make it more expressive too.
+The members provided, for example, for filtering limit the number of conditions that can be 
+written, because the user is limited to choosing one of the provided members.
+As discussed in the case study based on our implementation (Section~\ref{sec:impl}),
+this appears sufficient for many common data exploration tasks. The mechanism could be made
+more expressive, but we leave this for future work -- for example, the type provider could
+accept or reject member names written by the user (as in internet search) rather than providing 
+names from which the user can choose (as in web directories).
+
+% --------------------------------------------------------------------------------------------------
 
 \begin{figure}
-\begin{center}
-\includegraphics[scale=0.35,trim=0mm 0mm 0mm 0mm,clip]{filter.png} % left bottom right top
-\end{center}
-\vspace{-0.5em}
-\caption{Auto-completion offering the available values of the athlete name column}
-\label{fig:subtyping-diagram}
+\begin{equation*}
+\quad \begin{array}{l}
+\qident{drop columns}\\
+\quad\rightarrow~\qident{drop Athlete}\\
+\quad\rightarrow~\qident{drop Discipline}\\
+\quad\rightarrow~\qident{drop Year}\\[0.5em]
+\qident{sort data}\\
+\quad\rightarrow~\qident{by Athlete}\\
+\quad\rightarrow~\qident{by Athlete descending}\\
+\quad\rightarrow~\qident{by Discipline}\\
+\end{array}\qquad
+\begin{array}{l}
+\qident{group data}\\
+\quad\rightarrow~\qident{by Athlete}\\
+\qquad\quad\rightarrow~\qident{distinct Discipline}\\
+\qquad\quad\rightarrow~\qident{average Year}\\
+\qquad\quad\rightarrow~\qident{sum Year}\\[0.5em] 
+\quad\rightarrow~\qident{by Year}\\
+\qquad\quad\rightarrow~\qident{distinct Athlete}\\
+\qquad\quad\rightarrow~\qident{distinct Discipline}\\
+\end{array}
+\end{equation*}
+\caption{Subset of members provided by the pivot type provider}
+\label{fig:pivot-members}
 \end{figure}
 
-\subsection{Dot-driven development}
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Tooling and dot-driven development}
 \label{sec:analysis-auto}
 
-Autocompletion - see figure.
+Source code editors for object-based languages with nominal type sytems often provide 
+auto-completion for members of objects. This combination appear to work extremely well; the
+member list is a complete list of what might follow after typing ``dot'' and it can be easily
+obtained for an instance of known type. The fact that developers can often rely on just typing
+``dot'' and choosing an appropriate member led to a semi-serious phrase dot-driven development,
+that we (equally semi-seriously) adopt in this paper.
 
-\subparagraph{Abstraction and data.}
+Type providers in F\# rely on dot-driven development when navigating through data. When writing code 
+to access current temperature $\ident{ldn.Main.Temp}$ in Section~\ref{sec:tps}, the auto-completion
+offers various available properties, such as \ident{Wind} and \ident{Clouds} once ``dot'' is typed
+after $\ident{ldn.Main}$. Other type providers [X] follow a similar pattern. It is worth noting
+that despite the use of nominal typing, the names of types rarely explicitly appear in code -- we
+do not need to know the name of the type of $\ident{ldn.Main}$, but we need to know its members.
+Thus the type name can be arbitrary [X] and is used more as a lookup key.
 
-- Needs concrete data set or a sample, but that's fine
+The pivot type provider presented in this paper uses dot-driven development for suggesting 
+transformations as well as possible values of parameters. This is illustrated in Figure~\ref{fig:dot-driven}
+where the user wants to obtain medals of a specific athlete and is offered a list of possible 
+names. The editor filters the list as the user starts typing the required name.
+
+In Figure~\ref{fig:pivot-members}, we list a subset of the provided members that were used in the
+example in Section~\ref{sec:intro}. After choosing \qident{sort data}, the user is offered 
+the possible sorting keys and directions. After choosing \qident{group data}, the user first 
+selects the grouping key and then can choose one or more aggregations that can be applied on
+other columns of the group. Thus an enitre data transformation (such as choosing top 8 athletes
+by the number of gold medals) can be constructed using dot-driven development.
 
 \subparagraph{Values vs. types.}
+As the Figure~\ref{fig:dot-driven} illustrates, the pivot type provider sometimes blurs the 
+distinction between values and (members of) types. In the example in Section~\ref{sec:intro},
+\str{Rio (2016)} is a string value in Python, but a statically-typed member \qident{Rio (2016)}
+when using pivot type provider. This is a recurring theme in type provider development\footnote{The
+\ident{Individuals} property in the Freebase type provider [X] imports values into types in a similar way.}.
 
-- provide values? Could be $\qident{Games is}(\str{Rio (2016)})$
+Our language supports method calls such as $\ident{head}(\num{8})$ and an alternative design for
+filtering would be to provide a method such as $\qident{Games is}(\str{Rio (2016)})$. However,
+the fact that we can offer possible values largely simplifies writing of the script for the most 
+common case when the user is interested in one of the known values.
 
-\subsection{Column names}
-\label{sec:analysis-col}
+Unlike in traditional development, a data scientist doing data exploration often has the entire 
+data set available. The pivot type provider uses this when offering possible values for filtering
+(Section~X), but all other operations (Section~Y) require only meta-data (names and types of 
+columns). Following the example of type providers for structured data formats [X], the schema could
+be inferred from a representative sample.
 
-The last two points are more interesting -- proponents of static typing would correctly
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Expressing structured logic using members}
+\label{sec:analysis-and}
+
+In the motivating example, the \ident{agg} method takes a dictionary that specifies one or more 
+aggregates to be calculated over a group. We sum the number of gold medals, but we could also sum
+the number of silver and bronze medals, concatenate names of teams for the athlete and perform other 
+aggregations. In this case, we provide a nested structure (list of aggregations) as a parameter of 
+a single operation (grouping).
+
+This is an interesting case, because when encoding program as a sequence of member accesses, 
+there is no built-in support for nesting. In the pivot type provider, we use ``then'' design
+pattern to provide operations that require nesting. The following example specifies multiple 
+aggregations and then sorts data by multiple keys:
+%
+\begin{equation*}
+\begin{array}{l}
+\ident{olympics}.\\
+\quad\qident{group data}.\qident{by Athlete}.\\
+\qquad.\qident{sum Gold}.\qident{sum Silver}.\qident{concat Team}.\ident{then} \qquad\ball{1}\\
+\quad.\qident{sort data}.\\
+\qquad.\qident{by Gold descending}.\qident{and Silver descending}.\ident{then} \qquad\ball{2}\\
+\end{array}
+\end{equation*}
+
+\noindent
+When grouping, we sum the number of gold and silver medals and concatenates team names~\ball{1}. 
+Then we sorts the grouped data using multiple sorting keys \ball{2} -- first by number of gold 
+medals and then by silver medals (within a group with the same number of gold medals).
+
+\subparagraph{The ``then'' pattern.}
+Nesting is no doubt essential programming construct and it may very well be desirable to support it 
+directly in the language, but the ``then'' pattern lets us express it without explicit langauge
+support. In both cases, the nested structure is specified by selecting one or more members and then 
+ending the nested structure using the \ident{then} member.
+
+In case of grouping, we choose aggregations (\qident{sum Gold}, \qident{concat Team}, etc.) after 
+we specify grouping key using \qident{by Athlete}. In case of sorting, we specify the first key
+using \qident{by Gold descending} and then add more nested keys using \qident{and Silver descending}.
+Thanks to the dot-driven development and the ``then'' pattern, the user is offered possible 
+parameter values (aggregations or sorting keys) even when creating a nested structure. This is
+harder to do with general-purpose nesting as ordinary languages (even domain-specific languages)
+often provide many different ways for creating a nested structure.
+
+\subparagraph{Renaming columns.}
+The pivot type provider automatically chooses names for the columns obtained as the result of
+aggregation. In the above example \ball{1}, the resulting data set will have columns Athlete
+(the grouping key) together with Gold, Silver and Team (based on the aggregated columns).
+The user cannot currently rename the columns.
+
+% https://github.com/fsharp/fslang-design/blob/master/FSharp-4.0/StaticMethodArgumentsDesignAndSpec.md
+
+In F\# type providers, this could be done using methods with static parameters [X] by writing, 
+for example, $\ident{g}.\qident{sum Gold as}\langl\str{Total Gold}\rangl()$. In F\#, the value of 
+the static parameter (here, \str{Total Gold}) is passed to the type provider, which can use it to
+generate the type signature of the method and the return type with member name according to the
+value of the static parameter.
+
+% ==================================================================================================
+
+\section{Tracking column names}
+\label{sec:columns}
+
+proponents of static typing would correctly
 point out that both member names and column names can be tracked by a more sophisticated type
@@
 
-  ○ We use type providers to provide available options
-  (e.g. something)
+% --------------------------------------------------------------------------------------------------
 
-Considerations
+\subsection{Using row types and phantom types}
+\label{sec:columns-row}
 
- - renaming columns
+Phantom types + row types
 
-\subsection{Dictionary syntax}
-\label{sec:analysis-and}
-  ○ We create the "and blah. and blah. then" pattern
-  (e.g. grouping and sorting)
-
-~
 \begin{equation*}
-\begin{array}{l}
-\ident{olympics}\\
-\quad.\qident{filter data}.\qident{Games is}.\qident{Rio (2016)}.\ident{then}\\
-\quad.\qident{group data}.\qident{by Athlete}.\qident{sum Gold}.\ident{then}\\
-\quad.\qident{sort data}.\qident{by Gold descending}.\ident{then}\\
-\quad.\qident{paging}.\ident{take}(8)\\
-\quad.\qident{get series}.\qident{with key Athlete}.\qident{and value Gold}
-\end{array}
+  \tau = [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n] \lsep \tau_\ident{lbl}
 \end{equation*}
 
-Introduction
+\begin{equation*}
+\inference
+  {\Gamma \vdash e : [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]}
+  {\Gamma \vdash e.\qident{drop columns}: [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]_{\ident{sort}}}
+\end{equation*}
 
 \begin{equation*}
-\begin{array}{l}
-\ident{olympics}\\
-\quad.\qident{filter data}.\qident{Games is}.\qident{Rio (2016)}.\ident{then}\\
-\quad.\qident{group data}.\qident{by Athlete}.\qident{sum Gold}.\ident{then}\\
-\quad.\qident{sort data}.\qident{by Gold descending}.\ident{then}\\
-\quad.\qident{paging}.\ident{take}(8)\\
-\quad.\qident{get series}.\qident{with key Athlete}.\qident{and value Gold}
-\end{array}
+\inference
+  {\Gamma \vdash e : [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]_{\ident{sort}}}
+  {\Gamma \vdash e.\qident{then}: [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]}
 \end{equation*}
 
-Details
- - Grouping
- - Filtering
- - Joining
+\begin{equation*}
+\inference
+  {\Gamma \vdash e : [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]_{\ident{sort}}}
+  {\Gamma \vdash e.\qident{drop~$f_i$}: [f_1\!:\!\tau_1, \ldots, f_{i-1}\!:\!\tau_{i-1}, f_{i+1}\!:\!\tau_{i+1}, \ldots, f_n\!:\!\tau_n ]_{\ident{sort}}}
+\end{equation*}
+
+\subsection{Using type providers}
+\label{sec:columns-tp}
+
+% --------------------------------------------------------------------------------------------------
 
@@
 \quad l = \kvd{type}~C(x:\ident{Query}) = \\
-\qquad \qquad \kvd{member}~\qident{drop fields} : C_1 = C_1(x) &\textnormal{where}~C_1, L_1 = \ident{drop}(F)\\
+\qquad \qquad \kvd{member}~\qident{drop columns} : C_1 = C_1(x) &\textnormal{where}~C_1, L_1 = \ident{drop}(F)\\
 \qquad \qquad \kvd{member}~\qident{sort data} : C_2 = C_2(x) &\textnormal{where}~C_2, L_2 = \ident{sort}(F)\\
@@
 
+\begin{equation*}
+\begin{array}{l}
+\ident{olympics}\\
+\quad.\qident{filter data}.\qident{Games is}.\qident{Rio (2016)}.\ident{then}\\
+\quad.\qident{group data}.\qident{by Athlete}.\qident{sum Gold}.\ident{then}\\
+\quad.\qident{sort data}.\qident{by Gold descending}.\ident{then}\\
+\quad.\qident{paging}.\ident{take}(8)\\
+\quad.\qident{get series}.\qident{with key Athlete}.\qident{and value Gold}
+\end{array}
+\end{equation*}
+
 ~
@@
 \appendix
-\section{Morbi eros magna}
+\section{Sample of the Olympic medals data set}
+\label{app:olympics-csv}
 
-Morbi eros magna, vestibulum non posuere non, porta eu quam.
-Maecenas vitae orci risus, eget imperdiet mauris.
-Donec massa mauris, pellentesque vel lobortis eu, molestie ac turpis.
-Sed condimentum convallis dolor, a dignissim est ultrices eu.
-Donec consectetur volutpat eros, et ornare dui ultricies id.
-Vivamus eu augue eget dolor euismod ultrices et sit amet nisi.
-Vivamus malesuada leo ac leo ullamcorper tempor.
-Donec justo mi, tempor vitae aliquet non, faucibus eu lacus.
-Donec dictum gravida neque, non porta turpis imperdiet eget.
-Curabitur quis euismod ligula.
+{\small
+\begin{equation*}
+\begin{array}{l}
+\textbf{\sffamily {Games},{Year},{Discipline},{Athlete},{Team},{Gender},{Event},{Medal},{Gold},{Silver},{Bronze}}\\
+\ident{Athens (1896)},\num{1896},\ident{Swimming},\ident{Alfred Hajos},\kvd{HUN},\ident{Men},\ident{100m freestyle men},\ident{Gold},\num{1},\num{0},\num{0}\\
+\ident{Athens (1896)},\num{1896},\ident{Swimming},\ident{Otto Herschmann},\kvd{AUT},\ident{Men},\ident{100m freestyle men},\ident{Silver},\num{0},\num{1},\num{0}\\
+\ident{Athens (1896)},\num{1896},\ident{Swimming},\ident{Dimitrios Drivas},\kvd{GRE},\ident{Men},\ident{100m freestyle for sailors men},\ident{Bronze},\num{0},\num{0},\num{1}\\
+\ident{Athens (1896)},\num{1896},\ident{Swimming},\ident{Ioannis Malokinis},\kvd{GRE},\ident{Men},\ident{100m freestyle for sailors men},\ident{Gold},\num{1},\num{0},\num{0}\\
+\ident{Athens (1896)},\num{1896},\ident{Swimming},\ident{Spiridon Chasapis},\kvd{GRE},\ident{Men},\ident{100m freestyle for sailors men},\ident{Silver},\num{0},\num{1},\num{0}\\
+\end{array}
+\end{equation*} }
 

### 2016-12-31 pivot/paper.tex
Tracking columns, part one

@@
 data, but also compose complex queries using only member access (``dot''). This lets us recreate 
-functionality that usually requires complex type systems (row polymporphism, phantom types, 
+functionality that usually requires complex type systems (row polymporphism, type state, 
 dependent typing) in an extremely simple object-based language.
@@
 
-\section{Introduction -- Simplifying data exploration}
+\section{Introduction}
 \label{sec:intro}
@@
 
-\section{Background -- Type providers}
+\section{Using type providers in a novel way}
 \label{sec:tps}
@@
 
-\section{Analysis -- Simplifying data scripting}
+\section{Simplifying data scripting languages}
 \label{sec:analysis}
@@
 
-proponents of static typing would correctly
-point out that both member names and column names can be tracked by a more sophisticated type
-system. However, powerful types usually increase the complexity of the programming language.
+The last difficulty with data scripting discussed in Section~\ref{sec:analysis-why} is that pandas 
+(and other data exploration libraries, even for statically-typed languages) track column names as 
+strings at runtime, making code error-prone and auto-complete on column names difficult to provie.
+Proponents of static typing would correctly point out that column names and their types can be
+tracked by a more sophisticated type system.
+
+In this section, we discuss our approach -- we track column names statically using a mechanism
+that is inspired by row types and type state (Section~\ref{sec:columns-row}), however we 
+embed the mechanism using type providers into a simple nominal type system (Section~\ref{sec:columns-tp}).
+This way, the host language for the pivot type provider can be extremely simple -- and indeed, the
+mechanism could be added to languages such as Java or TypeScript with minimal effort.
 
@@
 
-\subsection{Using row types and phantom types}
-\label{sec:columns-row}
-
-Phantom types + row types
+\begin{figure}
 
 \begin{equation*}
-  \tau = [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n] \lsep \tau_\ident{lbl}
+\inference[(\emph{drop-start})~]
+  {\Gamma \vdash e : [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]}
+  {\Gamma \vdash e.\qident{drop columns}: [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]_{\ident{drop}}}
 \end{equation*}
@@
 \begin{equation*}
-\inference
-  {\Gamma \vdash e : [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]}
-  {\Gamma \vdash e.\qident{drop columns}: [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]_{\ident{sort}}}
+\inference[(\emph{drop-col})~]
+  {\Gamma \vdash e : [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]_{\ident{drop}}}
+  {\Gamma \vdash e.\qident{drop~$f_i$}: [f_1\!:\!\tau_1, \ldots, f_{i-1}\!:\!\tau_{i-1}, f_{i+1}\!:\!\tau_{i+1}, \ldots, f_n\!:\!\tau_n ]_{\ident{drop}}}
 \end{equation*}
@@
 \begin{equation*}
-\inference
-  {\Gamma \vdash e : [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]_{\ident{sort}}}
+\inference[(\emph{drop-then})~]
+  {\Gamma \vdash e : [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]_{\ident{drop}}}
   {\Gamma \vdash e.\qident{then}: [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]}
@@
 
+\caption{Tracking available column names with row types and type state}
+\label{fig:fancy-types}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Using row types and type state}
+\label{sec:columns-row}
+
+There are several common data transformations that modify the structure of the data set and affect
+what columns (and of what types) are available. When grouping and aggregating data, the resulting
+data set has columns depending on the aggregates calculated. Another simpler operation is adding
+or removing a column from the data set. For example, given the Olympic medals data set, we can 
+drop games and year as follows:
+%
 \begin{equation*}
-\inference
-  {\Gamma \vdash e : [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]_{\ident{sort}}}
-  {\Gamma \vdash e.\qident{drop~$f_i$}: [f_1\!:\!\tau_1, \ldots, f_{i-1}\!:\!\tau_{i-1}, f_{i+1}\!:\!\tau_{i+1}, \ldots, f_n\!:\!\tau_n ]_{\ident{sort}}}
+  \begin{array}{l}
+    \ident{olympics}.\qident{drop columns}.\qident{drop Games}.\qident{drop Year}.\ident{then}
+  \end{array}
 \end{equation*}
 
-\subsection{Using type providers}
+\noindent
+Operations on the type of rows in the data set can be captured using row types [X]. In addition, 
+we need to annotate type with a form of type state [Y] to restrict what operations are available. 
+When dropping columns, we first access the \qident{drop columns} member, which sets the state to
+a state where we can drop individual columns using \qident{drop $f$}. The \ident{then} member can 
+then be used to complete the operation and choose another transformation.
+
+To illustrate tracking of columns using row types and type state, consider a simple language with
+variables (representing external data sources) and member access. Types can be either primitive 
+types $\alpha$, types annotated with a type state \ident{lbl} or row type with fields $f$:
+%
+\begin{equation*}
+\begin{array}{rcl}  
+e&=&v \lsep e.N\\
+\tau&=&\alpha \lsep \tau_\ident{lbl} \lsep [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n] 
+\end{array}  
+\end{equation*}
+
+\noindent
+Typing rules for members that are used to drop columns are shown in Figure~\ref{fig:fancy-types}.
+When \qident{drop columns} is invoked on a record, the type is annotated with a state \ident{drop}
+(\emph{drop-start}) indicating that individual columns may be dropped. The \ident{then} 
+operation (\emph{drop-then}) removes the state label. Individual members can be removed using
+\qident{drop $f_i$} and the (\emph{drop-col}) rule ensures the dropped column is available in the
+input row type and removes it.
+
+Other data transformations could be type checked in a similar way, but there are two drawbacks.
+First, row types and type state (although relatively straightforward) make the host language more
+complex. Second, rules such as (\emph{drop-col}) make auto-completion more difficult, because the
+editor needs to understand the rules and calculate what members may be invoked. This is a distinct
+operation from type checking and type infernece (which operate on complete programs) that needs 
+to be formalized and implemented.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Using the pivot type provider}
 \label{sec:columns-tp}
+
+If $\Gamma\vdash e:[f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]$ using row types and
+$\Gamma\vdash e:C$ then $\ident{fields}(C) = \{ f_1\mapsto\tau_1 \}$
+\begin{equation*}
+\end{equation*}
+
+\newpage
 

### 2017-01-01 pivot/paper.tex
Relational algebra

@@
 % ==================================================================================================
+
+\theoremstyle{plain}
+\theoremstyle{definition}
+\newtheorem{rmrk}{Remark}
 
@@
 
-If $\Gamma\vdash e:[f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]$ using row types and
-$\Gamma\vdash e:C$ then $\ident{fields}(C) = \{ f_1\mapsto\tau_1 \}$
+In our approach, the information about available fields is used by a type provider to provide
+types with appropriate members. This is hidden from the host language, which only sees class types
+with members. Provided class definitions consist of a constructor and members: 
+%
 \begin{equation*}
+\begin{array}{rcl}
+ l &=& \kvd{type}~C(x:\tau) = \overline{m} \\
+ m &=& \kvd{member}~N:\tau= e
+\end{array}
 \end{equation*}
 
-\newpage
+\noindent
+During type-checking, the type system keeps track of a list of provided class definitions $L$. 
+Checking member access is then just a matter of finding the corresponding class definition and
+finding the member type:
+%
+\begin{equation*}
+\inference[(\emph{member})~]
+  {L; \Gamma \vdash e : C & \kvd{type}~C(x:\tau) = ..\;\kvd{member}~N_i : \tau_i = e_i\;.. \in L}
+  {L; \Gamma \vdash e.N_i:\tau_i}
+\end{equation*}
+
+\noindent
+The rule, adapted from [X], does not capture laziness of type providers that is 
+important for the pivot type provider (where the number of provided classes is potentially 
+infinite). We discuss this aspect in Section~\ref{sec:foo}.
+
+Using type providers and nominal type system seemingly hides knowledge about fields available
+in the data set. However, for types constructed by the pivot type provider, we can define a 
+mapping \ident{fields} that returns the fields available in the data set represented by the 
+class. The type provider encodes the logic expressed in Section~\ref{sec:columns-row} in the
+following sense:
+
+\begin{rmrk}[Encoding of fancy types]
+\label{thm:encoding-fancy}
+If $\Gamma\vdash e:[f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]$ using a type system defined in 
+Figure~\ref{fig:fancy-types} and $\Gamma\vdash e:C$ using nominal typing and $C$ is a type 
+provided by the pivot type provider then $\ident{fields}(C) = \{ f_1\mapsto\tau_1, \ldots, f_n\mapsto\tau_n \}$.
+\end{rmrk}
+
+\noindent
+In the following two sections, we focus on formalizing the pivot type provider and the nominally 
+typed host language. We define the \ident{fields} predicate in Section~\ref{sec:proofs} and use it 
+to prove properties of the pivot type provider.
+
+We do not fully develop the type system based on fancy types sketched in Section~\ref{sec:columns-row}.
+However, the remark illustrates one interesting aspect of our work -- the type provider 
+mechanism makes it possible to express safety guarantees that would normally require row types and 
+type state in a simple nominally typed language. In a similar way, type providers have been used
+to encode session types [X], suggesting that this is a generally useful approach.
 
@@
    & \lsep & \ident{conc}~f\\
+\\
 \end{array} 
 \end{equation*}
+
+\newcommand{\vect}[1]{\langl #1 \rangl}
+
+$v=\vect{\{f_1\mapsto v_{1,1}, \ldots, f_n\mapsto v_{1,n}\},\ldots,\{f_1\mapsto v_{m,1}, \ldots, f_n\mapsto v_{m,n}\}}$
+
+\begin{figure}
+\begin{equation*}
+\begin{array}{l}
+\Pi_{f_{p(1)}, \ldots, f_{p(m)}} \{f_1\mapsto \vect{v_{1,1},\ldots,v_{1,r}}, \ldots, f_n\mapsto \vect{v_{n,1},\ldots,v_{n,r}}\} \leadsto \\
+  \quad \{f_{p(1)}\mapsto \vect{v_{p(1),1},\ldots,v_{p(1),r}}, \ldots, f_{p(m)}\mapsto \vect{v_{p(m),1},\ldots,v_{p(m),r}}\}  
+\\
+\\
+\sigma_\varphi \{f_1\mapsto \vect{v_{1,1},\ldots,v_{1,r}}, \ldots, f_n\mapsto \vect{v_{n,1},\ldots,v_{n,r}}\} \leadsto\\
+  \quad \{f_1\mapsto \vect{\ldots, v_{1,j}, \ldots}, \ldots, f_n\mapsto \vect{\ldots, v_{n,j},\ldots}\}
+  \qquad (\forall j.\;\varphi\, \{f_1\mapsto v_{1,j}, \ldots, f_n\mapsto v_{n,j}\})
+\\
+\\
+\tau_{f_{p(1)}, \ldots, f_{p(m)}} \{f_1\mapsto \vect{v_{1,1},\ldots,v_{1,r}}, \ldots, f_n\mapsto \vect{v_{n,1},\ldots,v_{n,r}}\} \leadsto\\
+  \quad \{f_1\mapsto \vect{v_{1,q(1)},\ldots,v_{1,q(r)}}, \ldots, f_n\mapsto \vect{v_{n,q(1)},\ldots,v_{n,q(r)}}\} 
+  \quad \textnormal{where $q$ permutation}\\
+  \qquad \textnormal{such that}~i\leq j\implies(v_{p(1),q(i)}, \ldots, v_{p(m),q(i)}) \leq (v_{p(1),q(j)}, \ldots, v_{p(m),q(j)})
+  \\
+  \\
+\Phi_{f_i, \rho_1, \ldots, \rho_m} \{f_1\mapsto \vect{v_{1,1},\ldots,v_{1,r}}, \ldots, f_n\mapsto \vect{v_{n,1},\ldots,v_{n,r}}\} \leadsto\\
+  \quad \{\ident{nm}(\rho_1)\mapsto a_1, \ldots, \ident{nm}(\rho_m)\mapsto a_m \} 
+  \quad \textnormal{where}\\
+  \qquad \{g_1,\ldots,g_s\}=\{\{l~|~k\in 1\ldots r, ~v_{i,l}=v_{i,k}\}, ~l\in 1\ldots r\} \\
+  ~~\begin{array}{ll}    
+    \qquad a_i = \vect{|g_1|, \ldots, |g_s|}  & \textnormal{when}~\rho_i = \ident{count} \\
+    \qquad a_i = \vect{\Sigma_{k\in g_1}v_{j,k}, \ldots, \Sigma_{k\in g_s}v_{j,k}} & \textnormal{when}~\rho_i = \ident{sum}~f_j \\
+    \qquad a_i = \vect{\Pi_{k\in g_1}v_{j,k}, \ldots, \Sigma_{k\in g_s}v_{j,k}}   & \textnormal{when}~\rho_i = \ident{conc}~f_j \\
+    \qquad a_i = \vect{|\{v_{j,k}, k\in g_1\}|, \ldots, |\{v_{j,k}, k\in g_s\}|}  & \textnormal{when}~\rho_i = \ident{dist}~f_j
+  \end{array}
+\end{array}
+\end{equation*}
+
+\caption{foo}
+\label{fig:zzz}
+\end{figure}
+
+%\noindent
+\begin{equation*}
+\Pi_{f_{p(1)}, \ldots, f_{p(n)}} \{f_1\mapsto \overline{v_1}, \ldots, f_n\mapsto \overline{v_m}\}
+  \leadsto \{f_{p(1)}\mapsto \overline{v_{p(1)}}, \ldots, f_{p(n)}\mapsto \overline{v_{p(n)}}\}
+\end{equation*}
+
 \subsection{Foo calculus}
@@
 \begin{array}{rcl}
- \tau &=& \ident{num} \lsep \ident{string} \lsep \ident{series}\langl \tau_1, \tau_2 \rangl \lsep \ident{Query}\\
- l &=& \kvd{type}~C(\overline{x:\tau}) = \overline{m} \\[0.0em]
+ \tau &=& C \lsep \ident{num} \lsep \ident{string} \lsep \ident{series}\langl \tau_1, \tau_2 \rangl \lsep \ident{Query}\\
+ l &=& \kvd{type}~C(x:\tau) = \overline{m} \\[0.0em]
  m &=& \kvd{member}~N:\tau= e
@@
 \begin{equation*}
+\begin{array}{rcl}
+ v &\narrow{=}& d \lsep \ident{None} \lsep \ident{Some}(v) \lsep \kvd{new}~C(\overline{v}) \lsep v_1::v_2 \\[0.0em]
+ e &\narrow{=}& d \lsep op \lsep e_1~e_2 \lsep \lambda x.e \lsep e.N \lsep \kvd{new}~C(\overline{e}) \\
+   &\narrow{|}& \ident{None} \lsep\kvd{match}~e~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\
+   &\narrow{|}& \ident{Some}(e) \lsep e_1 = e_2 \lsep {\kvd{if}~e_1~\kvd{then}~e_2~\kvd{else}~e_3} \lsep \kvd{nil} \\
+   &\narrow{|}& e_1 :: e_2 \lsep \kvd{match}~e~\kvd{with}~x_1::x_2 \rightarrow e_1 \,|\, \kvd{nil} \rightarrow e_2
+\end{array}
+\end{equation*}
+
+\begin{equation*}
 \inference[(foo)]
-  {L_1; \Gamma \vdash e : C; L_2 \qquad ((\kvd{type}~C(\overline{x:\tau}) = ..\;\kvd{member}~N_i : \tau_i = e_i\;..), L) \in L_2}
+  {L_1; \Gamma \vdash e : C; L_2 \qquad ((\kvd{type}~C(x:\tau) = ..\;\kvd{member}~N_i : \tau_i = e_i\;..), L) \in L_2}
   {L_1; \Gamma \vdash e.N_i:\tau_i; L_2 \cup L}

### 2017-01-02 pivot/paper.tex
Relational algebra, take 2

@@
 
+\newcommand{\vect}[1]{\langl #1 \rangl}
 \newcommand{\langl}{\begin{picture}(4.5,7)
@@
 
-% --------------------------------------------------------------------------------------------------
+% ==================================================================================================
 
-\section{Formalising The Gamma script}
-\label{sec:foo}
+\begin{figure}
 
-Target language is super simple OO language, but we can use it to encode very fancy types!!
-
-\subsection{Relational algebra}
 \begin{equation*}
 \begin{array}{rclll}
-  R & = & \Pi_{f_1, \ldots, f_n} (R) &&\textnormal{get columns}\\
-   & \lsep & \sigma_\varphi (R) &&\textnormal{filter by predicate}\\
-   & \lsep & \tau_{f_1, \ldots, f_n}(R) &&\textnormal{sort by fields}\\
-   & \lsep & \Phi_{f, \rho_1, \ldots, \rho_n} (R) &&\textnormal{group by $f$ and aggregate}\\
-   \\
-  \rho & = & \ident{count}  \\
-   & \lsep & \ident{sum}~f    \\       
-   & \lsep & \ident{dist}~f\\
-   & \lsep & \ident{conc}~f\\
+  D & = & \multicolumn{3}{l}{ 
+    \{f_1\mapsto \vect{v_{1,1},\ldots,v_{1,r}}, \ldots, f_n\mapsto \vect{v_{n,1},\ldots,v_{n,r}}\} 
+  }\\[0.5em]
+  R & = & \Pi_{f_1, \ldots, f_n} (R) &&\textnormal{Projection -- select specified column names}\\
+   & \lsep & \sigma_\varphi (R) &&\textnormal{Selection -- filter rows by given predicate}\\
+   & \lsep & \tau_{f_1, \ldots, f_n}(R) &&\textnormal{Sorting -- sort by specified columns}\\
+   & \lsep & \Phi_{f, \rho_1/f_1, \ldots, \rho_n/f_n} (R) &&\textnormal{Grouping -- group by and calculate aggregates}
+   \\[0.5em]
+  \rho & = & \ident{count}  &&\textnormal{Count number of rows in the group}\\
+   & \lsep & \ident{sum}~f  &&\textnormal{Sum numerical values of the column $f$}\\
+   & \lsep & \ident{dist}~f &&\textnormal{Count number of distinct values of the column $f$}\\
+   & \lsep & \ident{conc}~f &&\textnormal{Concatenat string values of the column $f$}\\
 \\
@@
 
-\newcommand{\vect}[1]{\langl #1 \rangl}
+\vspace{-1.5em}
+\caption{Relational algebra with values, sorting and aggregation}
+\label{fig:foo-rel}
+\end{figure}
 
-$v=\vect{\{f_1\mapsto v_{1,1}, \ldots, f_n\mapsto v_{1,n}\},\ldots,\{f_1\mapsto v_{m,1}, \ldots, f_n\mapsto v_{m,n}\}}$
+% --------------------------------------------------------------------------------------------------
 
@@
   \\
-\Phi_{f_i, \rho_1, \ldots, \rho_m} \{f_1\mapsto \vect{v_{1,1},\ldots,v_{1,r}}, \ldots, f_n\mapsto \vect{v_{n,1},\ldots,v_{n,r}}\} \leadsto\\
-  \quad \{\ident{nm}(\rho_1)\mapsto a_1, \ldots, \ident{nm}(\rho_m)\mapsto a_m \} 
+\Phi_{f_i, \rho_1/f_1', \ldots, \rho_m/f_m'} \{f_1\mapsto \vect{v_{1,1},\ldots,v_{1,r}}, \ldots, f_n\mapsto \vect{v_{n,1},\ldots,v_{n,r}}\} \leadsto\\
+  \quad \{f_1' \mapsto a_1, \ldots, f_m'\mapsto a_m \} 
   \quad \textnormal{where}\\
@@
     \qquad a_i = \vect{\Pi_{k\in g_1}v_{j,k}, \ldots, \Sigma_{k\in g_s}v_{j,k}}   & \textnormal{when}~\rho_i = \ident{conc}~f_j \\
-    \qquad a_i = \vect{|\{v_{j,k}, k\in g_1\}|, \ldots, |\{v_{j,k}, k\in g_s\}|}  & \textnormal{when}~\rho_i = \ident{dist}~f_j
-  \end{array}
+    \qquad a_i = \vect{|\{v_{j,k}, k\in g_1\}|, \ldots, |\{v_{j,k}, k\in g_s\}|}  & \textnormal{when}~\rho_i = \ident{dist}~f_j \\
+  \end{array}\\
 \end{array}
@@
 
-\caption{foo}
-\label{fig:zzz}
+\caption{Vector-based semantics for operations of the extended relational algebra}
+\label{fig:foo-relsem}
 \end{figure}
 
-%\noindent
-\begin{equation*}
-\Pi_{f_{p(1)}, \ldots, f_{p(n)}} \{f_1\mapsto \overline{v_1}, \ldots, f_n\mapsto \overline{v_m}\}
-  \leadsto \{f_{p(1)}\mapsto \overline{v_{p(1)}}, \ldots, f_{p(n)}\mapsto \overline{v_{p(n)}}\}
-\end{equation*}
+% --------------------------------------------------------------------------------------------------
+
+\section{Formalising the host language}
+\label{sec:foo}
+
+Type providers often provide a thin type-safe layer over a richer untyped runtime component. In case 
+of providers for data access (Section~\ref{sec:tps}), the untyped runtime component performs operations
+over an external data source. In case of the pivot type provider, the untyped runtime component 
+is a relational algebra, which we use for modelling data transformations. We formalize the 
+relational algebra in Section~\ref{sec:foo-rel}, together with the object-based host language
+in Section~\ref{sec:foo-foo}.
+
+\subsection{Relational algebra with vector semantics}
+\label{sec:foo-rel}
+
+The focus of our work is on data aggregation and so we use a form of relational algebra with 
+extensions for grouping and sorting [X,Y,Z]. The syntax is defined in Figure~\ref{fig:foo-rel}.
+We write $f$ for column (field) names and we include definition of a data value $D$, which maps
+columns to vectors of length $r$ storing the data (values $v$ are defined below).
+Aside from standard projection $\Pi$ and selection $\sigma$, our algebra includes sorting 
+$\tau$ which takes one or more column names forming the sort key and aggregation $\Phi$, which
+requires a single grouping key and several aggregations together with names of the
+new columns to be returned.
+
+The semantics of the algbra is given in Figure~\ref{fig:foo-relsem}. We use vector-based semantics
+to support sorting and duplicate entries, but otherwise the formalization captures the usual
+behaviour. In projection and sorting, we write $f_{p(1)}, \ldots, f_{p(m)}$ to refer to a selection
+of fields from $f_1, \ldots, f_n$. Assuming $m\leq n$, $p$ can be seen as a mapping from $\{1\ldots m\}$
+to a subset of $\{1\ldots n\}$. In selection, $\varphi$ is a predicate applied to a mapping from 
+column names to values. In sorting, we assume that there is a permutation on row indices $q$ such
+that the tuples obtained by selecting values according to the sort key are ordered.
+
+The most complex operation is grouping. We need to group data by the value of the column $f_i$ and
+then apply aggregations $\rho_1,\ldots,\rho_m$. To do this, we first obtain a set of groups $g_1,\ldots, g_s$
+where each group represents a set of indicies of rows belonging to each group. For a given group 
+$g_i$ we can then obtain values of column $j$ for rows in the group as $\{v_{j,k}, k\in g_i\}$.
+This is used when performing the aggregations -- $|g_i|$ gives the number of rows in the group,
+$\Sigma$ sums numerical values and $\Pi$ (a slight notation abuse) concatenates string values.
 
 \subsection{Foo calculus}
+\label{sec:foo-foo}
+
 \begin{equation*}

### 2017-01-05 pivot/paper.tex
Pivot type provider

@@
 \usepackage{semantic}
+\usepackage{lipsum}
 
@@
 
-\newcommand{\lsep}{~|~}
+\newcommand{\lsep}{~\,|\,~}
 \newcommand{\num}[1]{\textcolor{numclr}{#1}}
@@
   }\\[0.5em]
-  R & = & \Pi_{f_1, \ldots, f_n} (R) &&\textnormal{Projection -- select specified column names}\\
-   & \lsep & \sigma_\varphi (R) &&\textnormal{Selection -- filter rows by given predicate}\\
-   & \lsep & \tau_{f_1, \ldots, f_n}(R) &&\textnormal{Sorting -- sort by specified columns}\\
-   & \lsep & \Phi_{f, \rho_1/f_1, \ldots, \rho_n/f_n} (R) &&\textnormal{Grouping -- group by and calculate aggregates}
+  e & = & \Pi_{f_1, \ldots, f_n} (e) &&\textnormal{Projection -- select specified column names}\\
+   & \lsep & \sigma_\varphi (e) &&\textnormal{Selection -- filter rows by given predicate}\\
+   & \lsep & \tau_{f_1\mapsto\omega_1, \ldots, f_n\mapsto\omega_n}(e) &&\textnormal{Sorting -- sort by specified columns}\\
+   & \lsep & \Phi_{f, \rho_1/f_1, \ldots, \rho_n/f_n} (e) &&\textnormal{Grouping -- group by and calculate aggregates}
    \\[0.5em]
+  \omega & = & \ident{desc}\lsep\ident{asc}  &&\textnormal{Sort order -- descending or ascending}\\
   \rho & = & \ident{count}  &&\textnormal{Count number of rows in the group}\\
@@
 \\
-\tau_{f_{p(1)}, \ldots, f_{p(m)}} \{f_1\mapsto \vect{v_{1,1},\ldots,v_{1,r}}, \ldots, f_n\mapsto \vect{v_{n,1},\ldots,v_{n,r}}\} \leadsto\\
+\tau_{f_{p(1)}\mapsto\omega_1, \ldots, f_{p(m)}\mapsto\omega_m} \{f_1\mapsto \vect{v_{1,1},\ldots,v_{1,r}}, \ldots, f_n\mapsto \vect{v_{n,1},\ldots,v_{n,r}}\} \leadsto\\
   \quad \{f_1\mapsto \vect{v_{1,q(1)},\ldots,v_{1,q(r)}}, \ldots, f_n\mapsto \vect{v_{n,q(1)},\ldots,v_{n,q(r)}}\} 
   \quad \textnormal{where $q$ permutation}\\
-  \qquad \textnormal{such that}~i\leq j\implies(v_{p(1),q(i)}, \ldots, v_{p(m),q(i)}) \leq (v_{p(1),q(j)}, \ldots, v_{p(m),q(j)})
+  \qquad \textnormal{such that}~\forall i, j.~i\leq j\implies(u_{1,i}, \ldots, v_{m,i}) \leq (v_{1,j}, \ldots, v_{m,j})~ \textnormal{where}\\
+  \qquad \quad u_{k,l}=v_{p(k),q(l)} \quad \,~~(\textnormal{when}~\omega_k=\ident{asc}) \\
+  \qquad \quad u_{k,l}=-v_{p(k),q(l)} \quad (\textnormal{when}~\omega_k=\ident{desc})
   \\
@@
 \Phi_{f_i, \rho_1/f_1', \ldots, \rho_m/f_m'} \{f_1\mapsto \vect{v_{1,1},\ldots,v_{1,r}}, \ldots, f_n\mapsto \vect{v_{n,1},\ldots,v_{n,r}}\} \leadsto\\
-  \quad \{f_1' \mapsto a_1, \ldots, f_m'\mapsto a_m \} 
+  \quad \{f_1' \mapsto a_1, \ldots, f_m'\mapsto a_m, f_i\mapsto b \} 
   \quad \textnormal{where}\\
@@
   ~~\begin{array}{ll}    
+    \qquad b = \vect{v_{i,k_1},\ldots,v_{i, k_s}} & \textnormal{where}~k_j \in g_j \\
     \qquad a_i = \vect{|g_1|, \ldots, |g_s|}  & \textnormal{when}~\rho_i = \ident{count} \\
@@
     \qquad a_i = \vect{\Pi_{k\in g_1}v_{j,k}, \ldots, \Sigma_{k\in g_s}v_{j,k}}   & \textnormal{when}~\rho_i = \ident{conc}~f_j \\
-    \qquad a_i = \vect{|\{v_{j,k}, k\in g_1\}|, \ldots, |\{v_{j,k}, k\in g_s\}|}  & \textnormal{when}~\rho_i = \ident{dist}~f_j \\
+    \qquad a_i = \vect{|\{v_{j,k}\,|\, k\in g_1\}|, \ldots, |\{v_{j,k}\,|\, k\in g_s\}|}  & \textnormal{when}~\rho_i = \ident{dist}~f_j \\
   \end{array}\\
@@
 
-\section{Formalising the host language}
+\section{Formalising the host language and runtime}
 \label{sec:foo}
 
-Type providers often provide a thin type-safe layer over a richer untyped runtime component. In case 
+Type providers often provide a thin type-safe layer over richer untyped runtime components. In case 
 of providers for data access (Section~\ref{sec:tps}), the untyped runtime component performs operations
-over an external data source. In case of the pivot type provider, the untyped runtime component 
-is a relational algebra, which we use for modelling data transformations. We formalize the 
-relational algebra in Section~\ref{sec:foo-rel}, together with the object-based host language
-in Section~\ref{sec:foo-foo}.
+over external data sources. In case of the pivot type provider, the untyped runtime component 
+is a relational algebra modelling data transformations. We formalize the relational algebra in 
+Section~\ref{sec:foo-rel}, together with the object-based host language in Section~\ref{sec:foo-foo}.
 
@@
 Aside from standard projection $\Pi$ and selection $\sigma$, our algebra includes sorting 
-$\tau$ which takes one or more column names forming the sort key and aggregation $\Phi$, which
-requires a single grouping key and several aggregations together with names of the
-new columns to be returned.
+$\tau$ which takes one or more columns forming the sort key (with sort order $\omega$)
+and aggregation $\Phi$, which requires a single grouping key and several aggregations together with 
+names of the new columns to be returned.
 
@@
 column names to values. In sorting, we assume that there is a permutation on row indices $q$ such
-that the tuples obtained by selecting values according to the sort key are ordered.
+that the tuples obtained by selecting values according to the given sort key are ordered. The 
+auxiliary definition $u_{k,l}$ negates the number to reverse the sort order when descending order
+is required.
 
@@
 where each group represents a set of indicies of rows belonging to each group. For a given group 
-$g_i$ we can then obtain values of column $j$ for rows in the group as $\{v_{j,k}, k\in g_i\}$.
-This is used when performing the aggregations -- $|g_i|$ gives the number of rows in the group,
+$g_i$ we can then obtain values of column $j$ for rows in the group as $\{v_{j,k}\,|\, k\in g_i\}$.
+This is used to calculate the resulting data set -- the field $f_i$ becomes a new column formed
+by the group keys (obtained by picking one of the indices from $g_j$ for each group); other fields
+are calculated by aggregating data in various ways -- $|g_i|$ gives the number of rows in the group,
 $\Sigma$ sums numerical values and $\Pi$ (a slight notation abuse) concatenates string values.
 
-\subsection{Foo calculus}
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}
+\begin{equation*}
+\begin{array}{rcl}
+  v &=& C(v)\,\, \lsep \ident{series}\langl \tau_1, \tau_2 \rangl(v)\; \lsep n \lsep s \lsep D \\
+  e &=& C(e)\,\, \lsep \ident{series}\langl \tau_1, \tau_2 \rangl(e)\;\, \lsep x \lsep v \lsep e.N \lsep \ldots\\
+  E &=& C(E) \lsep \ident{series}\langl \tau_1, \tau_2 \rangl(E)\, \lsep E.N \\
+    &|& \Pi_{f_1, \ldots, f_n}(E) \lsep \sigma_\varphi(E) \lsep \tau_{f_1, \ldots, f_n}(E) \lsep \Phi_{f_i, \rho_1/f_1', \ldots, \rho_m/f_m'} (E) \\
+    \\
+ \tau &=& C \lsep \kvd{num} \lsep \kvd{string} \lsep \ident{series}\langl \tau_1, \tau_2 \rangl \lsep \ident{Query}\\
+ l &=& \kvd{type}~C(x:\tau) = \overline{m} \\
+ m &=& \kvd{member}~N:\tau= e
+\end{array}
+\end{equation*}
+\vspace{-1em}
+
+\[
+\inference[(\emph{member})~]
+{ L(C) = (\kvd{type}~C(x:\tau)= \ldots~\kvd{member}~N_i : \tau_i = e_i~\ldots), L'}
+{ (C(v)).N_i \leadsto_L e_i[x \leftarrow v] }\\
+\]
+\vspace{-2.5em}
+
+\[
+\inference[(\emph{context})~]
+{e \leadsto_L e'}
+{E[e] \leadsto_L E[e']}
+\]
+
+\caption{Syntax and remaining reduction rules of the Foo calculus}
+\label{fig:foo-syntax}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Foo calculus with lazy context}
 \label{sec:foo-foo}
 
-\begin{equation*}
-\begin{array}{rcl}
- \tau &=& C \lsep \ident{num} \lsep \ident{string} \lsep \ident{series}\langl \tau_1, \tau_2 \rangl \lsep \ident{Query}\\
- l &=& \kvd{type}~C(x:\tau) = \overline{m} \\[0.0em]
- m &=& \kvd{member}~N:\tau= e
-\end{array}
-\end{equation*}
-
-\begin{equation*}
-\begin{array}{rcl}
- v &\narrow{=}& d \lsep \ident{None} \lsep \ident{Some}(v) \lsep \kvd{new}~C(\overline{v}) \lsep v_1::v_2 \\[0.0em]
- e &\narrow{=}& d \lsep op \lsep e_1~e_2 \lsep \lambda x.e \lsep e.N \lsep \kvd{new}~C(\overline{e}) \\
-   &\narrow{|}& \ident{None} \lsep\kvd{match}~e~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\
-   &\narrow{|}& \ident{Some}(e) \lsep e_1 = e_2 \lsep {\kvd{if}~e_1~\kvd{then}~e_2~\kvd{else}~e_3} \lsep \kvd{nil} \\
-   &\narrow{|}& e_1 :: e_2 \lsep \kvd{match}~e~\kvd{with}~x_1::x_2 \rightarrow e_1 \,|\, \kvd{nil} \rightarrow e_2
-\end{array}
-\end{equation*}
-
-\begin{equation*}
-\inference[(foo)]
-  {L_1; \Gamma \vdash e : C; L_2 \qquad ((\kvd{type}~C(x:\tau) = ..\;\kvd{member}~N_i : \tau_i = e_i\;..), L) \in L_2}
+We model the host language using a variant of the Foo calculus [X]. The core of the calculus models
+a simple object-based language with objects and members. The syntax of the language is shown in
+Figure~\ref{fig:foo-syntax}. The relational algebra defined in Figure~\ref{fig:foo-rel} is included
+in the Foo calculus as a model of the runtime components of the pivot type provider -- the values
+include the data value $D$ and the expressions include all the operations of the relational algebra.
+
+The Foo calculus includes two special types. \ident{Query} is a type of data and queries 
+constructed using the relational algebra. The type $\ident{series}\langl \tau_1, \tau_2 \rangl$ 
+models a data series mapping keys of type $\tau_1$ to values of type $\tau_2$. A series is created 
+as a result of a query written using the pivot type provider. In the actual implementation, a series 
+can be passed to a charting library. A series is a statically-typed wrapper over a \ident{Query} 
+value and the proofs in Section~\ref{sec:proofs} show that a series obtained from the pivot type 
+provider contains keys and values of matching types.
+
+\subparagraph{Reduction rules.} The reduction relation $\leadsto_L$ is prameterized by a function
+$L$ that maps class names to class definitions, together with nested classes associated with the
+class definition (used during type-checking as discussed below). The map is not used in the 
+reduction rules for the relational algebra, given in Figure~\ref{fig:foo-relsem} and so it was 
+omitted there.
+
+The remaining reduction rules are given in Figure~\ref{fig:foo-syntax}. The (\emph{member}) rule 
+performs lookup using $L(C)$ to find the definition of the member that is being accessed and then
+it reduces member access by substituting the evaluated constructor argument $v$ for a variable $x$.
+We assume standard capture-avoiding substitution $[x\leftarrow v]$. The rule ignores the nested 
+class definitions $L'$. The (\emph{context}) rule performs reduction in an evaluation context $E$.
+
+\subparagraph{Type checking.} One interesting aspect of type-checking with type providers is that
+type providers can provide potentially infinite number of types. In F\#, the types are provided 
+lazily as the type checker explores parts of the type space used by the program [X]. Consider:
+%
+\begin{equation*}
+\ident{olympics}.\qident{group data}.\qident{by Athlete}.\qident{sum Gold}.\ident{then}
+\end{equation*}
+
+\noindent
+The type checker initially knows the type of $\ident{olympics}$ is a class $C_1$ with member 
+$\qident{group data}$ and it knows that the type of this member is $C_2$. However, it only needs to 
+obtain full definition of $C_2$ when checking the member $\qident{by Athlete}$. Types of other 
+member of $C_1$ remain unevaluated when they do not appear in the source code. This aspect of type
+providers have been omitted in previous work [X,Y], but it is necessary for the pivot type provider.
+The typing rules given are written using the following judgement:
+%
+\begin{equation*}
+L_1; \Gamma \vdash e : \tau; L_2
+\end{equation*}
+
+\noindent
+The judgement states that given class definitions $L_1$ and a variable context $\Gamma$, the type
+of expression $e$ is $\tau$ and the type checking evaluated further class definitions that are
+now included in $L_2$. The structure of class definitions $L_1$ is a function mapping a class name 
+$C$ to a pair consisting of the definition and a function that provides definitions of 
+further classes: 
+%
+\begin{equation*}
+L_1(C) = \kvd{type}~C(x:\tau) = \overline{m}, L'
+\end{equation*}
+
+\noindent
+The types of members of the class $C$ may be already known classes defined by $L_1$, but also classes 
+defined by $L'$. This models laziness as $L'$ is a function that may never be evaluated.
+
+The rules that define type-checking are shown in Figure~\ref{fig:foo-typecheck}. The two rules
+that force the discovery of new classes are (\emph{new}) and (\emph{member}). In both cases,
+we perform a lookup using $L_2(C)$ which returns a class definition together with further classes
+$L$. We treat functions as sets and join $L_2$ with further classes defined by $L$ using $L_2 \cup L$.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+
+\[
+\inference[(\emph{num})~]{}{L; \Gamma \vdash n : \kvd{num}; L }
+\quad
+\inference[(\emph{string})~]{}{L; \Gamma \vdash s : \kvd{string}; L}
+\quad
+\inference[(\emph{var})~]{}{L; \Gamma, x:\tau\vdash x:\tau; L}
+\]
+\vspace{-2em}
+
+\[
+\inference[(\emph{data})~]
+  {L; \Gamma \vdash v_{i,j} : \tau;L \quad \tau \in \{ \kvd{num}, \kvd{string} \}}
+  {L; \Gamma \vdash \{f_1\mapsto \vect{v_{1,1},\ldots,v_{1,r}}, \ldots, f_n\mapsto \vect{v_{n,1},\ldots,v_{n,r}}\} : \ident{Query};L }
+\]
+\vspace{-2em}
+
+\[
+\inference[(\emph{proj})~]
+  {L_1; \Gamma\vdash e : \ident{Query}; L_2}
+  {L_1; \Gamma\vdash \Pi_{f_1, \ldots, f_n} (e) : \ident{Query}; L_2}
+\qquad\inference[(\emph{sort})~]
+  {L_1; \Gamma\vdash e : \ident{Query}; L_2}
+  {L_1; \Gamma\vdash \tau_{f_1, \ldots, f_n}(e): \ident{Query}; L_2}
+\]
+
+\[
+\inference[(\emph{sel})~]
+  {L_1; \Gamma\vdash e : \ident{Query}; L_2}
+  {L_1; \Gamma\vdash \sigma_\varphi (e) : \ident{Query}; L_2}
+\qquad
+\inference[(\emph{group})~]
+  {L_1; \Gamma\vdash e : \ident{Query}; L_2}
+  {L_1; \Gamma\vdash \Phi_{f, \rho_1/f_1, \ldots, \rho_n/f_n}(e): \ident{Query}; L_2}
+\]
+\vspace{-2em}
+
+\[
+\inference[(\emph{series})~]
+  {L_1; \Gamma\vdash e : \ident{Query}; L_2}
+  {L_1; \Gamma\vdash \ident{series}\langl \tau_1, \tau_2 \rangl(e) : \ident{series}\langl \tau_1, \tau_2 \rangl; L_2}
+\]
+\vspace{-2em}
+
+\[
+\inference[(\emph{new})~]
+  {L_1; \Gamma \vdash e : \tau, L_2 & L_2(C) = (\kvd{type}~C(x:\tau) = \ldots), L}
+  {L_1; \Gamma \vdash C(e) : C; L_2 \cup L}
+\]
+\vspace{-2em}
+
+\[
+\inference[(\emph{member})~]
+  {L_1; \Gamma \vdash e : C; L_2 \qquad L_2(C)=(\kvd{type}~C(x:\tau) = ..\;\kvd{member}~N_i : \tau_i = e_i\;..), L}
   {L_1; \Gamma \vdash e.N_i:\tau_i; L_2 \cup L}
-\end{equation*}
-
-\newpage
+\]
+\vspace{-2em}
+
+\[
+\inference[(\emph{class})~]
+  { L; \Gamma,x:\tau \vdash e_i : \tau_i; L_{i} & m_i = \kvd{member}~N_i:\tau_i= e_i }
+  { L \vdash \kvd{type}~C(x:\tau) = m_1\ldots m_n; L_1 \cup \ldots \cup L_n }
+\]
+\caption{Type-checking of Foo expressions with lazy context}
+\label{fig:foo-typecheck}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+The rules for primitive types and variables are standard.
+Input data (\emph{data}) is of type \ident{Query} and all the operations of relational algebra 
+take a value of type \ident{Query} and produce a result of type \ident{Query}. This is the 
+untyped (or one-typed) subset of the Foo calculus. A \ident{Query} value can be converted into a
+series type (\emph{series}) of any type\footnote{This can be seen as a boundary between static and
+dynamic typing in gradually typed languages [X,Y,Z].} -- in the pivot type provider, the 
+\ident{Query} type is always hidden and users can only work with type-safe series type.
+
+Finally, we also define a judgement $L_1 \vdash l; L_2$ which states that given known classes $L_1$,
+a class definition $l$ is well typed and produces known classes $L_2$. This is defined by the
+(\emph{class}) rule that checks the types of bodies of all members of the class.
+
+% --------------------------------------------------------------------------------------------------
 
@@
 \begin{array}{ll}
-\ident{main}(F) = C \mapsto (l, L_1 \cup \ldots \cup L_4)\\[0.5em]
+\ident{pivot}(F) = C, \{ C \mapsto (l, L_1 \cup \ldots \cup L_4) \}\qquad\ball{1}\\[0.5em]
 \quad l = \kvd{type}~C(x:\ident{Query}) = \\
@@
 \\
-\ident{drop}(F) = C, \{l\} \cup \bigcup L_f\\[0.5em]
-\quad l = \kvd{type}~C(x:\ident{Query}) = \\
-\qquad \qquad \kvd{member}~\ident{then} : C = C(x)                 &\textnormal{where}~C, L = \ident{main}(F)\\
-\qquad \qquad \kvd{member}~\qident{drop $f$} : C_f = &\forall f \in \dom(F)~\textnormal{where}~C_f, L_f = \ident{drop}(F')\\
-\qquad \qquad \qquad  C_f(\Pi_{\ident{dom}(F')}(x))& \quad\textnormal{and}~ F' = \{ f'\mapsto\tau'\in F, f' \neq f\}\\
-\end{array}
-\end{equation*}
-
-\caption{Foo}
-\label{fig:foo}
-\end{figure}
-
-\begin{figure}
-\begin{equation*}
-\begin{array}{ll}
-\ident{get-key}(F) = C, \{l\} \cup \bigcup L_f \\[0.25em]
+\ident{get-key}(F) = C, \{ C \mapsto (l, \bigcup L_f)\}\qquad\ball{2} \\[0.25em]
 \quad l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)~\textnormal{where} \\
@@
 \\
-\ident{get-val}(F, f_k) = C, \{l\} \cup \bigcup L_f \\[0.25em]
-\quad l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)\setminus\{f\}~\textnormal{where} \\
-\qquad \qquad \kvd{member}~\qident{and value $f$} : C_f = C_f(\Pi_{f_k, f_v}(x)) &\quad  C_f~\textnormal{defined as below} \\
-\quad l_f = \kvd{type}~C_f(x:\ident{Query}) = &\quad\textnormal{and}~\tau_k, \tau_v~\textnormal{such that} \\
-\qquad \qquad \kvd{member}~\ident{series} : \ident{series}\langl\tau_k, \tau_v\rangl = x &\quad \tau_k = F(f_k), \tau_v = F(f) &\\
-\end{array}
-\end{equation*}
-
-\caption{Foo}
-\label{fig:foo}
+\ident{get-val}(F, f_k) = \{ C \mapsto (l, \{\}) \}\qquad\ball{3} \\[0.25em]
+\quad l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)\setminus\{f_k\}~\textnormal{where} \\
+\qquad \qquad \kvd{member}~\qident{and value $f$} : \ident{series}\langl\tau_k, \tau_v\rangl = &\quad  \tau_k = F(f_k),~ \tau_v = F(f) \\
+\qquad \qquad \quad \ident{series}\langl\tau_k, \tau_v\rangl(\Pi_{f_k, f}(x))\qquad\ball{4}
+\end{array}
+\end{equation*}
+\vspace{-1em}
+
+\caption{Pivot type provider -- providing entry-point type and accessing transformed data}
+\label{fig:tp-main}
 \end{figure}
+
+% ==================================================================================================
 
@@
 
-Omit paging, because it's boring
-
-Our formalization of the pivot type provider follows the style used by Petricek et al. \cite{fsdata}
-in their formalization of F\# Data. In our case, a type provider is a function that takes a schema
-$F$ (mapping from field names to field types) and produces a class $C$ together with collection 
-of other class defintins $L$ that are used by $C$:
-%
-\begin{equation*}
-\ident{provider}(F) = C, L    ~~\textnormal{where}~F=\{ f_1 \mapsto \tau_1, \ldots, f_n \mapsto \tau_n \}
-\end{equation*}
-%
-The provided class $C$ has a constructor taking \ident{Query}. It may be a class with further 
-members that allow refining the query or a class with a single member \ident{series} that returns
-a value of type $\ident{series}\langl \tau_1, \tau_2 \rangl$. In the second case, we can evaluate
-the provided transformation on input $R$ using an expression $(\kvd{new}~C(R)).\ident{series}$.
-
-\subparagraph{Top-level type.} The top-level type allows the user to select a transformation. The
-members return objects with members that allow specifying further properties of each transformation.
-
-\subparagraph{Sorting.} yadda
+In an implementation, a type provider is an executable component that is called by the compiler
+and the editor to provide information about types on demand. In our formalization, we follow the 
+style used by Petricek et al. [X], but we add laziness as discussed in Section~\ref{sec:foo-foo}.
+We model the core operations (dropping columns, grouping and sorting) in Section~\ref{sec:pivot-core}
+and then refine the model to include filtering based on values Section~\ref{sec:pivot-filter}.
+We omit paging, as it does not affect the shape of data, but modelling it would require richer 
+host language.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Pivot type provider}
+\label{sec:pivot-core}
+
+A type provider is a function that takes static parameters, such as schema of the input data set, 
+and returns a class name $C$ together with a mapping that defines the body of the class, together 
+with definitions of associated classes $L$ that may be used by the members of the class $C$. 
+In our case, the schema $F$ is a mapping from field names to field types:
+%
+\begin{equation*}
+\begin{array}{l}
+\ident{pivot}(F) = C, \{ C \mapsto (\kvd{type}~C(x:\ident{Query}) = \ldots,L) \} \quad \textnormal{where}~F=\{ f_1 \mapsto \tau_1, \ldots, f_n \mapsto \tau_n \}
+\end{array}
+\end{equation*}
+%
+The class $C$ provided by the pivot tpe provider has a constructor taking \ident{Query}, which 
+represents the, possibly already partly transformed, input data set. It generates members that
+allow the user to refine the query and access the data. The type provider is defined using several
+helper functions discussed in the rest of this section.
+
+\subparagraph{Entry-point and data access.} Figure~\ref{fig:tp-main} shows three of the functions
+defining the pivot type provider. The \ident{pivot} function \ball{1} defines the entry-point type, 
+which lets the user choose which operation to perform before specifying parameters of the operation. 
+This is the type of \ident{olympics} in the examples throughout this paper. The definition generates 
+a new class $C$ with members that wrap the input data in further classes generated by other parts
+of the type provider. The result of \ident{pivot} is the class name $C$ together with definition of 
+the class and further generated types. The definition is a function that only needs to be evaluated 
+when a program accesses a member of the class $C$, modelling the laziness of the type provider.
+In the implementation, we return the name $C$, but only need to compute the definition when the 
+type checker needs to access the body.
+
+The \ident{get-key} \ball{2} and \ident{get-val} \ball{3} functions provide members that can be used 
+to choose two columns from the data set as keys and values and obtain the resulting data set as a 
+value of type $\ident{series}\langl \tau_1, \tau_2 \rangl$. For example, the following expression 
+has a type $\ident{series}\langl \kvd{string}, \kvd{num} \rangl$:
+%
+\begin{equation*}
+\begin{array}{l}
+\ident{olympics}.\qident{get series}.\qident{with key Athlete}.\qident{and value Year}
+\end{array}
+\end{equation*}
+
+\noindent
+The \ident{get-key} function generates a class with one member for each field in the data set.
+The returned class $C_f$ is generated by \ident{get-val} and lets the user choose any of the 
+remaining fields as the value. The key and value columns are then selected using $\Pi_{f_k,f}$ \ball{4}.
+The series is then created with a data set containing only the key and value columns (we assume 
+the order of columns is preserved). Creating a series does not statically enforce that the 
+data set has the right structure, but the properties discussed in Section~\ref{sec:proofs} show 
+that series obtained from the pivot type provider is constructed correctly.
+
+% --------------------------------------------------------------------------------------------------
 
@@
 \begin{array}{ll}
-\ident{sort}(F) = C, \{l\} \cup \bigcup L_f \cup \bigcup L'_f\\[0.25em]
-\quad l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)~\textnormal{where} \\
-\qquad \qquad \kvd{member}~\qident{by $f$ descending} : C_f = C_f(x) &\quad C_f, L_f = \ident{sort-and}(F, \{ f \mapsto \ident{desc} \})\\
-\qquad \qquad \kvd{member}~\qident{by $f$ ascending} : C'_f = C'_f(x) &\quad C'_f, L'_f = \ident{sort-and}(F, \{ f \mapsto \ident{asc} \})\\
+\ident{drop}(F) = C, \{ C \mapsto (l, L' \cup \bigcup L_f) \}\qquad\ball{1}\\[0.5em]
+~ l = \kvd{type}~C(x:\ident{Query}) = &\forall f \in \dom(F)~\textnormal{where}~C_f, L_f = \ident{drop}(F')\\
+\quad \qquad \kvd{member}~\qident{drop $f$} : C_f = C_f(\Pi_{\ident{dom}(F')}(x)) & \quad\textnormal{and}~ F' = \{ f'\mapsto\tau'\in F, f' \neq f\}\\
+\quad \qquad \kvd{member}~\ident{then} : C' = C'(x)\qquad\ball{2}                 &\textnormal{where}~C', L' = \ident{pivot}(F)\\
+\\
+\ident{sort}(F) = C, \{ C \mapsto (l, \bigcup L_f \cup \bigcup L'_f) \}\qquad\ball{3}\\[0.25em]
+~ l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\{f~|~f\mapsto\kvd{num}\in F\},~\textnormal{where} \\
+\quad \qquad \kvd{member}~\qident{by $f$ descending} : C_f = C_f(x) &\quad C_f, L_f = \ident{sort-and}(F, \vect{f \mapsto \ident{desc}})\\
+\quad \qquad \kvd{member}~\qident{by $f$ ascending} : C'_f = C'_f(x) &\quad C'_f, L'_f = \ident{sort-and}(F, \vect{f \mapsto \ident{asc}})\\
 \\[0.5em]
-\ident{sort-and}(F,S) = C, \{l\} \cup \bigcup L_f \cup \bigcup L'_f\\[0.25em]
-\quad l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)\setminus\dom(S)~\textnormal{where} \\
-\qquad \qquad \kvd{member}~\qident{and $f$ descending} : C_f = C_f(x) &\quad C_f, L_f = \ident{sort-and}(F, S\cup\{ f \mapsto \ident{desc} \})\\
-\qquad \qquad \kvd{member}~\qident{and $f$ ascending} : C'_f = C'_f(x) &\quad C'_f, L'_f = \ident{sort-and}(F, S\cup\{ f \mapsto \ident{asc} \})\\
-\qquad \qquad \kvd{member}~\ident{then} : C = C(\tau_S(x))                 &\textnormal{where}~C, L = \ident{main}(F)\\
-\end{array}
-\end{equation*}
-\caption{Foo bar}
-\label{fig:foobar}
+\multicolumn{2}{l}{ \ident{sort-and}(F,\vect{s_1,\ldots,s_n}) = C, \{ C \mapsto (l, \bigcup L_f \cup \bigcup L'_f \cup L') \}\qquad\ball{4} }\\[0.25em]
+~ l = \kvd{type}~C(x:\ident{Query}) = &\hspace{-1em}\forall f\in\{f~|~f\mapsto\kvd{num}\in F, \,\nexists i.s_i=f'\mapsto\omega\wedge f'=f\}\\
+\quad \qquad \kvd{member}~\qident{and $f$ descending} : C_f = C_f(x) &\quad C_f, L_f = \ident{sort-and}(F, \vect{s_1,..,s_n, f \mapsto \ident{desc}})\\
+\quad \qquad \kvd{member}~\qident{and $f$ ascending} : C'_f = C'_f(x) &\quad C'_f, L'_f = \ident{sort-and}(F, \vect{s_1,..,s_n, f \mapsto \ident{asc}})\\
+\quad \qquad \kvd{member}~\ident{then} : C' = C'(\tau_{s_1,\ldots,s_n}(x))\quad~~\ball{5}           &\textnormal{where}~C', L' = \ident{pivot}(F)\\
+\end{array}
+\end{equation*}
+
+\caption{Pivot type provider -- dropping columns and sorting data}
+\label{fig:tp-ds}
 \end{figure}
 
+% --------------------------------------------------------------------------------------------------
+
+\subparagraph{Dropping columns and sorting.} Functions that provide types for the 
+\qident{drop columns} and \qident{sort data} members are defined in Figure~\ref{fig:tp-ds}.
+The \ident{drop} function \ball{1} builds a new type that lets the user drop any of the available columns.
+The resulting type $C_f$ is recursively generated by \ident{drop} so that multiple columns can 
+be dropped before completing the transformation using the \ident{then} operation \ball{2}, whose 
+return type is generated using the main \ident{pivot} function. Note that columns removed from the 
+schema $F'$ as the types are generated match the columns removed from the data set at runtime  
+using $\Pi_{\dom(F')}$.
+
+Types for defining the sorting transformation are split between two functions; \ident{sort}~\ball{3} 
+generates type for choosing the first sorting key and \ident{sort-and}~\ball{4} lets the user add more
+keys. The members are restricted to numerical columns (by checking $f\mapsto\kvd{num}\in F$).
+The sort keys are kept as a vector. The \ident{sort} operation creates a singleton vector;
+\ident{sort-and} appends a new key to the end and the \ident{then} member \ball{5} generates code 
+that passes the collected sort keys to the $\tau$ operation of the relational algebra.
+When generating members for adding further sort keys, we exclude the columns that are used
+already (by checking that the column $f$ does not match column name of any of the existing 
+keys $s_i = f'\mapsto\omega$).
+
+% --------------------------------------------------------------------------------------------------
+
 \begin{figure}
+\vspace{-0.75em}
 \begin{equation*}
 \begin{array}{ll}
-\ident{group}(F) = C, \{l\} \cup \bigcup L_f \\[0.25em]
-\quad l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)~\textnormal{where} \\
-\qquad \qquad \kvd{member}~\qident{by $f$} : C_f = C_f(x) &\quad C_f, L_f = \ident{agg}(F, \{ f \mapsto F(f) \}, \emptyset)\\
-\\[0.5em]
-\ident{agg}(F,S,G) = C, \{l\} \cup \bigcup L_f \cup \bigcup L'_f\\[0.25em]
-\quad l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)\setminus\dom(S)~\textnormal{where} \\
-\qquad \qquad \kvd{member}~\qident{distinct $f$} : C_f = C_f(x) &\quad C_f, L_f = \ident{agg}(F, S\cup\{ f \mapsto \ident{num}, G\cup\{\ident{dist}~f\} \})\\
-\qquad \qquad \kvd{member}~\qident{sum $f$} : C'_f = C'_f(x) &\quad C'_f, L'_f = \ident{agg}(F, S\cup\{ f \mapsto \ident{num} \}, G\cup\{\ident{sum}~f\})\\
-\qquad \qquad \kvd{member}~\qident{concat $f$} : C'_f = C'_f(x) &\quad C'_f, L'_f = \ident{agg}(F, S\cup\{ f \mapsto \ident{string} \}, G\cup\{\ident{conc}~f\})\\
-\qquad \qquad \kvd{member}~\qident{count all} : C = C(\tau_S(x))                 &\textnormal{where}~C, L = \ident{main}(F)\\
-\qquad \qquad \kvd{member}~\ident{then} : C = C(\tau_S(x))                 &\textnormal{where}~C, L = \ident{main}(F)\\
-\end{array}
-\end{equation*}
-\caption{Foo bar 2}
-\label{fig:foobar2}
+\ident{group}(F) = C, \{ C\mapsto (l, \bigcup L_f) \}\quad\ball{1} \\[0.25em]
+~ l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)~\textnormal{where} \\
+\quad \qquad \kvd{member}~\qident{by $f$} : C_f = C_f(x) &\quad C_f, L_f = \ident{agg}(F, f, \{ f \mapsto F(f) \}, \emptyset)~~\ball{2}\\
+\\
+\multicolumn{2}{l}{ \ident{agg}(F,f,G,S) = C, \{C\mapsto (l, \bigcup L_f \cup \bigcup L'_f \cup\bigcup L''_f\cup L'\cup L'') \}\quad\ball{3}}\\[0.25em]
+~ l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)\setminus\dom(S)\\
+\quad \qquad \kvd{member}~\qident{distinct $f$} : C_f = C_f(x)  \\
+\quad \qquad \kvd{member}~\qident{sum $f$} : C'_f = C'_f(x) & \quad \textnormal{when}~F(f)=\kvd{num}\quad\ball{4} \\
+\quad \qquad \kvd{member}~\qident{concat $f$} : C''_f = C''_f(x) & \quad \textnormal{when}~F(f)=\kvd{string}\quad\ball{5} \\
+\quad \qquad \kvd{member}~\qident{count all} : C' = C'(x) & \quad \textnormal{when}~\ident{Count}\notin G\quad\ball{6} \\
+\quad \qquad \kvd{member}~\ident{then} : C'' = C''(\Phi_{f, \rho_1/f_1, \ldots, \rho_n/f_n}(x)) &\quad\textnormal{where}~\{\rho_1/f_1,\ldots,\rho_n/f_n\}=S\quad\ball{7} \\[0.5em]
+\qquad  \textnormal{where}\\[0.5em]
+\multicolumn{2}{l}{
+~\,\qquad \begin{array}{l}
+ C_f, L_f = \ident{agg}(F,f, G\cup\{ f \mapsto \kvd{num}\}, S\cup\{\ident{dist}~f/f\} )\\
+ C'_f, L'_f = \ident{agg}(F,f, G\cup\{ f \mapsto \kvd{num} \}, S\cup\{\ident{sum}~f/f\})\\
+ C''_f, L''_f = \ident{agg}(F,f, G\cup\{ f \mapsto \kvd{string} \}, S\cup\{\ident{conc}~f/f\})\\
+ C', L' = \ident{agg}(F,f, G\cup\{ \ident{Count}\mapsto\kvd{int} \}, S\cup{\ident{count}/\ident{Count}})\\
+ C'', L'' = \ident{pivot}(G)
+\end{array} 
+}
+\end{array}
+\end{equation*}
+\vspace{-1em}
+
+\caption{Pivot type provider -- grouping and aggregation}
+\label{fig:tp-group}
 \end{figure}
 
-!!
-
-TODO: Somehow delay $L$
-
-!!
-
-We only generate sorting methods for fields not used yet
-
-Helper \ident{sort-and} also takes sorting keys already used
+% --------------------------------------------------------------------------------------------------
+
+\subparagraph{Grouping and aggregation.} The final part of the pivot type provider is defined in
+Figure~\ref{fig:tp-group}. The \ident{group} function \ball{1} generates a class that lets the user 
+select a column to use as the grouping key and \ident{agg} is used to provide aggregates that can be 
+calculated over grouped data. The \ident{agg} function \ball{3} takes the schema of the input data 
+set $F$, column $f$ to be used as the group key, a schema of the data set that will be produced as 
+the result $G$ and a set of aggregation operations collected so far $S$. Initially~\ball{2}, the 
+resulting schema contains only the column used as the key with its original type (which is always 
+generated by $\Phi$) and the set of aggregations to be calculated is empty.
+
+The \ident{agg} function is invoked recursively (similarly to \ident{drop} and \ident{sort-and})
+to add further aggregation operations, or until the user selects the \ident{then} member \ball{7}, 
+which applies the grouping using $\Phi$ and returns a class generated by the entry-point \ident{pivot}  
+function.
+
+When calculating an aggregate over a specific column, the type provider reuses the column name from
+the input data set in the resulting data set. Consequently, the \ident{agg} function offers 
+aggregation operations only using columns that have not been already used. This somewhat limits
+the expressivity, but it simplifies the programming model. Furthermore, \qident{sum $f$} \ball{4} 
+is only provided for columns of type \kvd{num} and \qident{concat $f$} \ball{5} is only provided 
+for strings. Finally, the \qident{count all} aggregation \ball{6} is not related to a specific 
+field and is exposed once, adding a column \ident{Count} to the schema of the resulting data set.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Filtering}
+\label{sec:pivot-filter}
+\newpage
+
+% ==================================================================================================
 
@@
 \label{sec:proofs}
+
+% ==================================================================================================
 

### 2017-01-10 pivot/paper.tex
Properties of pivot

@@
   (Section~\ref{sec:pivot}) and show that queries constructed using the type provider are valid
-  (Section~\ref{sec:proofs}). Our formalization also covers the laziness of type providers, which
+  (Section~\ref{sec:pivot-prop}). Our formalization also covers the laziness of type providers, which
   is an important aspect not covered in the existing literature.
@@
 \noindent
-During type-checking, the type system keeps track of a list of provided class definitions $L$. 
+During type checking, the type system keeps track of a list of provided class definitions $L$. 
 Checking member access is then just a matter of finding the corresponding class definition and
@@
 In the following two sections, we focus on formalizing the pivot type provider and the nominally 
-typed host language. We define the \ident{fields} predicate in Section~\ref{sec:proofs} and use it 
+typed host language. We define the \ident{fields} predicate in Section~\ref{sec:pivot-prop} and use it 
 to prove properties of the pivot type provider.
@@
 \begin{figure}
+\vspace{-1em}
 \begin{equation*}
@@
   E &=& C(E) \lsep \ident{series}\langl \tau_1, \tau_2 \rangl(E)\, \lsep E.N \\
-    &|& \Pi_{f_1, \ldots, f_n}(E) \lsep \sigma_\varphi(E) \lsep \tau_{f_1, \ldots, f_n}(E) \lsep \Phi_{f_i, \rho_1/f_1', \ldots, \rho_m/f_m'} (E) \\
+    &|& \Pi_{f_1, \ldots, f_n}(E) \lsep \sigma_\varphi(E) \lsep \tau_{f_1, \ldots, f_n}(E) \lsep \Phi_{f, \rho_1/f_1, \ldots, \rho_m/f_m} (E) \\
     \\
@@
 \end{equation*}
-\vspace{-1em}
+\vspace{-1.5em}
 
@@
 \]
-\vspace{-2.5em}
+\vspace{-3em}
 
@@
 \]
-
 \caption{Syntax and remaining reduction rules of the Foo calculus}
@@
 can be passed to a charting library. A series is a statically-typed wrapper over a \ident{Query} 
-value and the proofs in Section~\ref{sec:proofs} show that a series obtained from the pivot type 
-provider contains keys and values of matching types.
+value and the proofs in Section~\ref{sec:pivot-prop} show that a series obtained from the pivot type 
+provider contains keys and values of matching~types.
 
@@
 $L$ that maps class names to class definitions, together with nested classes associated with the
-class definition (used during type-checking as discussed below). The map is not used in the 
+class definition (used during type checking as discussed below). The map is not used in the 
 reduction rules for the relational algebra, given in Figure~\ref{fig:foo-relsem} and so it was 
@@
 
-\subparagraph{Type checking.} One interesting aspect of type-checking with type providers is that
+\subparagraph{Type checking.} One interesting aspect of type checking with type providers is that
 type providers can provide potentially infinite number of types. In F\#, the types are provided 
@@
 The judgement states that given class definitions $L_1$ and a variable context $\Gamma$, the type
-of expression $e$ is $\tau$ and the type checking evaluated further class definitions that are
-now included in $L_2$. The structure of class definitions $L_1$ is a function mapping a class name 
+of expression $e$ is $\tau$ and the type checking evaluated class definitions that are
+now included in~$L_2$. The structure of class definitions $L_1$ is a function mapping a class name 
 $C$ to a pair consisting of the definition and a function that provides definitions of 
-further classes: 
-%
-\begin{equation*}
-L_1(C) = \kvd{type}~C(x:\tau) = \overline{m}, L'
-\end{equation*}
-
-\noindent
-The types of members of the class $C$ may be already known classes defined by $L_1$, but also classes 
-defined by $L'$. This models laziness as $L'$ is a function that may never be evaluated.
-
-The rules that define type-checking are shown in Figure~\ref{fig:foo-typecheck}. The two rules
-that force the discovery of new classes are (\emph{new}) and (\emph{member}). In both cases,
-we perform a lookup using $L_2(C)$ which returns a class definition together with further classes
-$L$. We treat functions as sets and join $L_2$ with further classes defined by $L$ using $L_2 \cup L$.
+delayed classes: 
+%
+\begin{equation*}
+L_1(C) = \kvd{type}~C(x:\tau) = \overline{m}, L
+\end{equation*}
+
+\noindent
+The class $C$ may use classes defined in $L_1$, but also delayed classes from $L$. This models 
+laziness as $L$ is a function that may never be evaluated. Since $L$ is potentially infinite, we 
+cannot check class definitions upfront as in typical object calculi [X]. Instead, we check that 
+that members are well typed as they appear in the source code, which matches the behaviour of F\# 
+type providers. In general, this means that $L$ may contain classes with incorrectly typed 
+members. We prove that this is not the case for the pivot type provider (Section~\ref{sec:pivot-prop}).
+
+The rules that define type checking are shown in Figure~\ref{fig:foo-typecheck}. The two rules
+that force the discovery of new classes are (\emph{new}) and (\emph{member}). In (\emph{new}), we
+find the class definition and delayed classes using $L_2(C)$. We treat functions as sets and 
+join $L_2$ with delayed classes defined by $L$ using $L_2 \cup L$. In (\emph{member}), we obtain
+the class definition and discover delayed classes in the same way, but we also check that the body
+of the member is well-typed.
 
@@
 \begin{figure}[t]
-
+\vspace{-1em}
 \[
@@
 \inference[(\emph{new})~]
-  {L_1; \Gamma \vdash e : \tau, L_2 & L_2(C) = (\kvd{type}~C(x:\tau) = \ldots), L}
+  {L_1; \Gamma \vdash e : \tau, L_2 & L_2(C) = (\kvd{type}~C(x:\tau) = \ldots),L}
   {L_1; \Gamma \vdash C(e) : C; L_2 \cup L}
@@
 \inference[(\emph{member})~]
-  {L_1; \Gamma \vdash e : C; L_2 \qquad L_2(C)=(\kvd{type}~C(x:\tau) = ..\;\kvd{member}~N_i : \tau_i = e_i\;..), L}
-  {L_1; \Gamma \vdash e.N_i:\tau_i; L_2 \cup L}
+  {L_1; \Gamma \vdash e : C; L_2 \qquad L_2\cup L; \Gamma,x:\tau \vdash e_i : \tau_i; L_3 \\
+   L_2(C)=(\kvd{type}~C(x:\tau) = ..\;\kvd{member}~N_i : \tau_i = e_i\;..), L }
+  {L_1; \Gamma \vdash e.N_i:\tau_i; L_3}
 \]
-\vspace{-2em}
-
-\[
-\inference[(\emph{class})~]
-  { L; \Gamma,x:\tau \vdash e_i : \tau_i; L_{i} & m_i = \kvd{member}~N_i:\tau_i= e_i }
-  { L \vdash \kvd{type}~C(x:\tau) = m_1\ldots m_n; L_1 \cup \ldots \cup L_n }
-\]
+\vspace{-1em}
 \caption{Type-checking of Foo expressions with lazy context}
@@
 
-The rules for primitive types and variables are standard.
-Input data (\emph{data}) is of type \ident{Query} and all the operations of relational algebra 
-take a value of type \ident{Query} and produce a result of type \ident{Query}. This is the 
-untyped (or one-typed) subset of the Foo calculus. A \ident{Query} value can be converted into a
-series type (\emph{series}) of any type\footnote{This can be seen as a boundary between static and
-dynamic typing in gradually typed languages [X,Y,Z].} -- in the pivot type provider, the 
-\ident{Query} type is always hidden and users can only work with type-safe series type.
-
-Finally, we also define a judgement $L_1 \vdash l; L_2$ which states that given known classes $L_1$,
-a class definition $l$ is well typed and produces known classes $L_2$. This is defined by the
-(\emph{class}) rule that checks the types of bodies of all members of the class.
+The rules for primitive types and variables are standard. Input data (\emph{data}) is of type 
+\ident{Query} and all the operations of relational algebra take a value of type \ident{Query} and 
+produce a result of type \ident{Query}. An untyped \ident{Query} value can be converted into a 
+series type (\emph{series}) of any type -- in the pivot type provider, the \ident{Query} type is 
+always hidden and users can only work with type-safe series type. The bridging of typed and 
+untyped parts of the language resembles the boundary between static and dynamic typing in 
+gradually typed languages.% [X,Y,Z].
 
@@
 \\
-\ident{get-val}(F, f_k) = \{ C \mapsto (l, \{\}) \}\qquad\ball{3} \\[0.25em]
+\ident{get-val}(F, f_k) = C, \{ C \mapsto (l, \{\}) \}\qquad\ball{3} \\[0.25em]
 \quad l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)\setminus\{f_k\}~\textnormal{where} \\
@@
 \end{equation*}
-\vspace{-1em}
-
-\caption{Pivot type provider -- providing entry-point type and accessing transformed data}
+
+\caption{Pivot type provider -- entry-point type and accessing transformed data}
 \label{fig:tp-main}
@@
 
-\section{Pivot type provider}
+\section{Formalising the pivot type provider}
 \label{sec:pivot}
@@
 This is the type of \ident{olympics} in the examples throughout this paper. The definition generates 
-a new class $C$ with members that wrap the input data in further classes generated by other parts
+a new class $C$ with members that wrap the input data in delayed classes generated by other parts
 of the type provider. The result of \ident{pivot} is the class name $C$ together with definition of 
-the class and further generated types. The definition is a function that only needs to be evaluated 
+the class and delayed generated types. The definition is a function that only needs to be evaluated 
 when a program accesses a member of the class $C$, modelling the laziness of the type provider.
@@
 the order of columns is preserved). Creating a series does not statically enforce that the 
-data set has the right structure, but the properties discussed in Section~\ref{sec:proofs} show 
+data set has the right structure, but the properties discussed in Section~\ref{sec:pivot-prop} show 
 that series obtained from the pivot type provider is constructed correctly.
@@
 \begin{figure}
-\vspace{-0.75em}
 \begin{equation*}
@@
 ~ l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)\setminus\dom(S)\\
-\quad \qquad \kvd{member}~\qident{distinct $f$} : C_f = C_f(x)  \\
 \quad \qquad \kvd{member}~\qident{sum $f$} : C'_f = C'_f(x) & \quad \textnormal{when}~F(f)=\kvd{num}\quad\ball{4} \\
@@
 \quad \qquad \kvd{member}~\qident{count all} : C' = C'(x) & \quad \textnormal{when}~\ident{Count}\notin G\quad\ball{6} \\
-\quad \qquad \kvd{member}~\ident{then} : C'' = C''(\Phi_{f, \rho_1/f_1, \ldots, \rho_n/f_n}(x)) &\quad\textnormal{where}~\{\rho_1/f_1,\ldots,\rho_n/f_n\}=S\quad\ball{7} \\[0.5em]
+\quad \qquad \kvd{member}~\qident{distinct $f$} : C_f = C_f(x)  \\
+\quad \qquad \kvd{member}~\ident{then} : C'' = C''(\Phi_{f, \rho_1/f_1, \ldots, \rho_n/f_n}(x)) &\textnormal{where}~\{\rho_1/f_1,\ldots,\rho_n/f_n\}=S\quad\ball{7} \\[0.5em]
 \qquad  \textnormal{where}\\[0.5em]
@@
 \end{equation*}
-\vspace{-1em}
 
@@
 
-\subsection{Filtering}
+\newpage
+
+\subsection{Properties of the pivot type provider}
+\label{sec:pivot-prop}
+
+If we were using the relational algebra formalised in Section~\ref{sec:foo-rel} to construct 
+queries, we can obtain an invalid program, e.g.~by attempting to select
+a column $f$ using $\Pi_{f}$ from a data set that does not contain the column. This is not an issue 
+when using the pivot type provider, because the provided types allow the user to construct only 
+correct data transformations.
+
+To formalise this, we prove partial soundness of the Foo calculus (Theorem~\ref{thm:foo-sound}),
+which characterises the invalid programs that can be written using the \ident{Query}-typed 
+expressions and then prove safety of the pivot type provider (Theorem~\ref{thm:pivot-safe}),
+which shows that such errors do not occur when using the provided types.
+
+\subparagraph{Foo calculus.} The Foo calculus consists of the relational algebra and simple object
+calculus where objects can be constructed and their members accessed. It permits recursion as a
+member can invoke itself on a new object instance. To accommodate this, we formalise soundness 
+using progress (Lemma~\ref{thm:foo-progress}) and preservation (Lemma~\ref{thm:foo-pres}).
+
+The soundness is partial because the evaluation can get stuck when an operation of the relational 
+algebra on a given data set is undefined.
+
+\begin{theorem}[Partial soundness]
+\label{thm:foo-sound}
+For all $L_0, e, e'$, if $L_0, \emptyset \vdash e : \tau, L_1$ and $e\leadsto_{L_1}^{*} e'$ 
+then either $e'$ is a value, or there exists $e''$ such that $e'\leadsto_{L_1}^{*} e''$, or
+$e'$ has one of the following forms:
+$E[\Pi_{f_1, \ldots, f_n}(D)]$, $E[\sigma_\varphi(D)]$, $\tau_{f_1, \ldots, f_n}(D)]$ or 
+$E[\Phi_{f, \rho_1/f_1, \ldots, \rho_m/f_m} (D)]$ for some $E, D$.
+\end{theorem}
+\begin{proof}
+Direct consequence of Lemma~\ref{thm:foo-progress} and Lemma~\ref{thm:foo-pres}.
+\end{proof}
+
+\begin{lemma}[Partial progress]
+\label{thm:foo-progress}
+For all $L_0, e$ such that $L_0, \emptyset \vdash e : \tau, L_1$ then either, $e$ is a value,
+there exists $e'$ such that $e\leadsto_{L_1} e'$ or $e$ has one of
+the following forms: $E[\Pi_{f_1, \ldots, f_n}(D)]$, $E[\sigma_\varphi(D)]$, $\tau_{f_1, \ldots, f_n}(D)]$ 
+or $E[\Phi_{f, \rho_1/f_1, \ldots, \rho_m/f_m} (D)]$ for some $E$ and $D$.
+\end{lemma}
+\begin{proof}
+By induction over $\vdash$. For data, strings and numbers, the expression is always a value. 
+For relational algebra operations, the expression can either be reduced or has one of the required 
+forms. For (\emph{member}) the expression can always be reduced.
+\end{proof}
+
+\begin{lemma}[Type preservation]
+\label{thm:foo-pres}
+For all $L_0, e, e'$ such that $L_0, \emptyset \vdash e : \tau, L_1$ and $e\leadsto_{L_1} e'$ then
+$L_1, \emptyset \vdash e' : \tau, L_2$ for some $L_2$.
+\end{lemma}
+\begin{proof}
+By induction over $\leadsto_{L_1}$. Cases for relational algebra operations and for (\emph{context})
+are straightforward. The (\emph{member}) case follows from a standard substitution lemma and the 
+fact that type checking of member access also type checks the body of the member.
+\end{proof}
+
+\subparagraph{Correctness of the pivot provider.} 
+The pivot type provider defined by \ident{pivot} defines an entry-point class and a context $L$ 
+containing delayed classes. The type system cannot check type definitions in $L$ upfront, but 
+we prove that the body of all provided members is well-typed. Furthermore, type checking could fail
+if a delayed class was not discovered before it is needed in the (\emph{new}) and (\emph{member}) 
+typing rules (Figure~\ref{fig:foo-typecheck}). We show that this cannot happen for the context
+constructed by the \ident{pivot} function. To avoid operating over potentially infinite contexts,
+we first define an expansion operation $\downarrow_n L$ that evaluates the first $n$ levels of the 
+nested context $L$ and flattens it.
+
+\begin{definition}[Expansion]
+\label{thm:pivot-well}
+Given a context $L$, we define $n^{\textnormal{th}}$ expansion of $L$, written 
+$\downarrow_n L$:
+\begin{itemize}
+\item $\downarrow_n L = \downarrow_{n-1} L \cup \bigcup L_n$ assuming that $\downarrow_{n-1} L = \{ C_0\mapsto(l_0, L_0),\ldots,C_n\mapsto(l_n, L_n) \}$
+\item $\downarrow_0 L = L$  
+\end{itemize}
+\end{definition}
+
+\begin{theorem}[Correctness of lazy contexts]
+\label{thm:pivot-lazy}
+Given $C,L=\ident{pivot}(F)$ then for any $e$ if there exists $i, \tau$ such that 
+$\downarrow_i L; \emptyset \vdash e : \tau; L'$ then also
+$L; \emptyset \vdash e : \tau; L''$.
+\end{theorem}
+\begin{proof}
+Assume there exists $F,e,i$ such that $\downarrow_i L; \emptyset \vdash e : \tau; L'$ but
+not $L; \emptyset \vdash e : \tau; L''$. This is a contradiction as (\emph{new}) and (\emph{member})
+typing rules expand $L$ defined by \ident{pivot} sufficiently to discover all types that may have
+been used in the type-checking of $e$ using $\downarrow_i L$.
+\end{proof}
+
+\begin{theorem}[Correctness of provided types]
+\label{thm:pivot-well}
+For all $F,n$ let $C_0,L_0=\ident{pivot}(F)$ and assume that $C \in \dom(\downarrow_{n} L)$ 
+where $\downarrow_{n} L(C) = (\kvd{type}~C(x:\tau) = ..\;\kvd{member}~N_i:\tau_i= e_i\;..), L'$.
+It holds that for all $i$ the body of $N_i$ is well-typed, i.e.~
+$L \cup L'; x:\tau \vdash e_i : \tau_i; L''$.
+\end{theorem}
+\begin{proof}
+By examination of the functions defining the type provider; the expressions $e_i$ are well-typed
+and use only types defined in $L\cup L'$.
+\end{proof}
+
+\subparagraph{Safety of provided transformations.}
+The two properties discussed above ensure that the types provided by the pivot type provider can 
+be used to type check expressions constructed by the users of the type provider in the expected
+way. An expression will not fail to type check because of an error in the provided types.
+
+Now we can turn to the key theorem of the paper, which states that any expression constructed 
+using (just) the provided types can be evaluated to a value of correct type. For simplicity, we
+only assume expressions that access a series using the \qident{get series} member. However, this
+covers all data transformations that can be constructed using the type provider.
+
+\begin{theorem}[Safety of pivot type provider]
+\label{thm:pivot-safe}
+Given a schema $F=\{f_1\mapsto\tau_1, \ldots, f_n\mapsto\tau_n \}$, let $C,L=\ident{pivot}(F)$ then for any 
+expression $e$ that does not contain relational algebra operations or $\ident{Query}$-typed values as sub-expression, 
+if $L;x\!:\!C \vdash e : \ident{series}\langl\tau_1,\tau_2\rangl; L'$ then for all $D = 
+\{ f_1\mapsto\vect{v_{1,1},\ldots,v_{1,m}}, \ldots, f_n\mapsto\vect{v_{n,1},\ldots,v_{n,m}} \}$
+such that $\vdash v_{i, j} : \tau_i$ it holds that $e[x\leftarrow C(D)]\leadsto_{L'}^{*} 
+  \ident{series}\langl\tau_k,\tau_v\rangl(\{ f_k\mapsto{k_1,\ldots,k_r}, f_v\mapsto{v_1,\ldots,v_r} \})$
+  such that for all $j$ $\vdash k_j : \tau_k$ and $\vdash v_j : \tau_v$.
+\end{theorem}
+\begin{proof}
+Define a mapping $\ident{fields}(C)$ that returns the fields expected in the data set passed to a 
+class $C$ provided by the pivot type provider. Let $\ident{fields}(C) = F$ for $C$ provided using:
+\begin{equation*}
+\begin{array}{l}
+\begin{array}{lcl}
+\ident{pivot}(F) = C, L &&
+\ident{get-key}(F) = C, L \\
+\ident{drop}(F) = C, L &&
+\ident{get-val}(F, f_k) = C, L\\
+\end{array}
+\quad\begin{array}{l}
+\ident{sort}(F) = C, L\\
+\ident{sort-and}(F,\vect{s_1,\ldots,s_n}) = C, L\\
+\end{array}\\[-0.1em]
+\;\, \ident{group}(F) = C, L \;\;\,\quad \ident{agg}(F,f,G,S) = C, L
+\end{array}
+\end{equation*}
+
+\noindent
+By induction over $\leadsto_{L'}$, show that when $C(v).N_i$ is reduced using (\emph{member}) then 
+$v$ is a value $\{ f_1\mapsto\vect{v_{1,1},\ldots,v_{1,m}}, \ldots, f_n\mapsto\vect{v_{n,1},\ldots,v_{n,m}} \}$
+s.t. $\ident{fields}(C)=\{f_1\mapsto\tau_1, \ldots, f_n\mapsto\tau_n \}$ and 
+$\vdash v_{i, j} : \tau_i$. Thus the class provided by \ident{get-val} is constructed with 
+a data set containing the required columns of corresponding types.
+\end{proof}
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\begin{equation*}
+\begin{array}{ll}
+  \ident{pivot}(F, D) = C, \{ C \mapsto (l, L_1 \cup L_2 \cup \ldots  \}\quad \ball{1}\\[0.5em]
+  ~ l = \kvd{type}~C(x:\ident{Query}) = \\
+  \quad \qquad \kvd{member}~\qident{drop columns} : C_1 = C_1(x) &\textnormal{where}~C_1, L_1 = \ident{drop}(F, D)\\
+  \quad \qquad \kvd{member}~\qident{filter data} : C_2 = C_2(x) &\textnormal{where}~C_2, L_2 = \ident{filter}(F, D)\\
+  \quad \qquad (\ldots)\\
+\\
+\ident{drop}(F, D) = C, \{ C \mapsto (l, L' \cup \bigcup L_f) \}\\[0.5em]
+~ l = \kvd{type}~C(x:\ident{Query}) = & \forall f \in \dom(F)~\\
+\quad \qquad \kvd{member}~\qident{drop $f$} : C_f =  & \quad\textnormal{where}~ F' = \{ f'\mapsto\tau'\in F, f' \neq f\}\\
+\quad \qquad \quad C_f(\Pi_{\ident{dom}(F')}(x)) &\quad\textnormal{and}~C_f, L_f = \ident{drop}(F', \Pi_{\ident{dom}(F')}(D))\quad\ball{2}\\
+\quad \qquad \kvd{member}~\ident{then} : C' = C'(x)                 &\textnormal{where}~C', L' = \ident{pivot}(F, D)\\
+\\
+\ident{filter}(F, D) = C, \{ C \mapsto (l, L' \cup \bigcup L_f) \}\\[0.5em]
+~ l = \kvd{type}~C(x:\ident{Query}) = & \forall f \in \dom(F)~\\
+\quad \qquad \kvd{member}~\qident{$f$ is} : C_f = C_f(x) & \quad\textnormal{where}~C_f, L_f = \ident{filter-val}(F, f, D)\quad\ball{3}\\
+\quad \qquad \kvd{member}~\ident{then} : C' = C'(x)                 &\textnormal{where}~C', L' = \ident{pivot}(F, D)\\
+\\
+\ident{filter-val}(F, f, D) = C, \{ C \mapsto (l, \cup \bigcup L_v) \}&\textnormal{where}~D=\{ f\mapsto\vect{v_1, \ldots, v_n}, \ldots \}\quad\ball{4} \\[0.5em]
+~ l = \kvd{type}~C(x:\ident{Query}) = & \forall v \in \{v_1, \ldots, v_n \} \\
+\quad \qquad \kvd{member}~\qident{$\,v\,$} : C_v =  & \quad\textnormal{where}~C_v, L_v = \ident{filter}(F, \sigma_{\varphi_v}(D))\\
+\quad \qquad \quad C_v(\sigma_{\varphi_v}(x)) &\quad\textnormal{and}~ \varphi_v (r) = r(f) = v \quad \ball{5}
+\end{array}
+\end{equation*}
+
+\caption{Pivot type provider -- grouping and aggregation}
+\label{fig:tp-group}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Adding the filtering operation}
 \label{sec:pivot-filter}
+
+Add $D$ representing input data (in reality, we do not need all of it; just a function)
+
+zz
+
+% ==================================================================================================
+
 \newpage
-
-% ==================================================================================================
-
-\section{Properties}
-\label{sec:proofs}
-
-% ==================================================================================================
 

### 2017-01-10 pivot/paper.tex
Filtering section

@@
 \\
-\ident{drop}(F, D) = C, \{ C \mapsto (l, L' \cup \bigcup L_f) \}\\[0.5em]
+\ident{drop}(F, D) = C, \{ C \mapsto (l, L' \cup \bigcup L_f) \}\quad\ball{2}\\[0.5em]
 ~ l = \kvd{type}~C(x:\ident{Query}) = & \forall f \in \dom(F)~\\
 \quad \qquad \kvd{member}~\qident{drop $f$} : C_f =  & \quad\textnormal{where}~ F' = \{ f'\mapsto\tau'\in F, f' \neq f\}\\
-\quad \qquad \quad C_f(\Pi_{\ident{dom}(F')}(x)) &\quad\textnormal{and}~C_f, L_f = \ident{drop}(F', \Pi_{\ident{dom}(F')}(D))\quad\ball{2}\\
+\quad \qquad \quad C_f(\Pi_{\ident{dom}(F')}(x)) &\quad\textnormal{and}~C_f, L_f = \ident{drop}(F', \Pi_{\ident{dom}(F')}(D))\quad\ball{3}\\
 \quad \qquad \kvd{member}~\ident{then} : C' = C'(x)                 &\textnormal{where}~C', L' = \ident{pivot}(F, D)\\
@@
 ~ l = \kvd{type}~C(x:\ident{Query}) = & \forall f \in \dom(F)~\\
-\quad \qquad \kvd{member}~\qident{$f$ is} : C_f = C_f(x) & \quad\textnormal{where}~C_f, L_f = \ident{filter-val}(F, f, D)\quad\ball{3}\\
+\quad \qquad \kvd{member}~\qident{$f$ is} : C_f = C_f(x) & \quad\textnormal{where}~C_f, L_f = \ident{filter-val}(F, f, D)\quad\ball{4}\\
 \quad \qquad \kvd{member}~\ident{then} : C' = C'(x)                 &\textnormal{where}~C', L' = \ident{pivot}(F, D)\\
 \\
-\ident{filter-val}(F, f, D) = C, \{ C \mapsto (l, \cup \bigcup L_v) \}&\textnormal{where}~D=\{ f\mapsto\vect{v_1, \ldots, v_n}, \ldots \}\quad\ball{4} \\[0.5em]
+\ident{filter-val}(F, f, D) = C, \{ C \mapsto (l, \cup \bigcup L_v) \}&\textnormal{where}~D=\{ f\mapsto\vect{v_1, \ldots, v_n}, \ldots \}\quad\ball{5} \\[0.5em]
 ~ l = \kvd{type}~C(x:\ident{Query}) = & \forall v \in \{v_1, \ldots, v_n \} \\
 \quad \qquad \kvd{member}~\qident{$\,v\,$} : C_v =  & \quad\textnormal{where}~C_v, L_v = \ident{filter}(F, \sigma_{\varphi_v}(D))\\
-\quad \qquad \quad C_v(\sigma_{\varphi_v}(x)) &\quad\textnormal{and}~ \varphi_v (r) = r(f) = v \quad \ball{5}
+\quad \qquad \quad C_v(\sigma_{\varphi_v}(x)) &\quad\textnormal{and}~ \varphi_v (r) = r(f) = v \quad \ball{6}
 \end{array}
@@
 \caption{Pivot type provider -- grouping and aggregation}
-\label{fig:tp-group}
+\label{fig:tp-filter}
 \end{figure}
@@
 
-Add $D$ representing input data (in reality, we do not need all of it; just a function)
-
-zz
+The example given in Section~\ref{sec:intro} obtained top 8 athletes based on the number
+of gold medals from Rio 2016. It used two operations that were omitted in the formalisation in
+Section~\ref{sec:pivot-core}. We omitted paging to keep the host language simple, but we also
+omitted filtering, which lets us write $\qident{filter data}.\qident{Games is}.\qident{Rio (2016)}$.
+This operation is worth further discussion. To support it, the type provider needs not only the
+schema of the data set, but also sample data set that is used to offer the available values such 
+as $\qident{Rio (2016)}$.
+
+\noindent
+In the revised model, the \ident{pivot} function which models the type provider takes the schema
+$F$ together with sample data $D$ and provides the type with class context:
+%
+\begin{equation*}
+\begin{array}{l}
+\ident{pivot}(F, D) = C, L \quad \textnormal{where}\\
+\qquad F=\{ f_1 \mapsto \tau_1, \ldots, f_n \mapsto \tau_n \}\\
+\qquad D=\{ f_1\mapsto \vect{v_{1,1},\ldots,v_{1,r}}, \ldots, f_n\mapsto \vect{v_{n,1},\ldots,v_{n,r}}\} 
+\end{array}
+\end{equation*}
+
+\noindent
+In prior work on type providers [X], the actual input value is not available when writing the code
+and so the schema is inferred from a representative sample value. In exploratory data analysis, the 
+data set is often available at the time of writing the code and so $D$ can be the actual data set.
+However, it could equally be a sample or representative subset of the data.
+
+The Figure~\ref{fig:tp-filter} shows a revised version of the \ident{pivot} function \ball{1} 
+together with one of the operations discussed before and the newly added \ident{filter} function.
+
+As members members performing data transformations are generated, the provider applies the same
+transformation on the sample data. For example, the revised \ident{drop} function \ball{2} takes
+the sample data set $D$; when calling \ident{drop} recursively to generate nested class after
+dropping a column \ball{3}, it removes the column from the schema (as before), but it also removes
+the column from the sample dataset. This means that as nested types are provided, the sample data
+used is always representative of data what will be passed to the class at runtime.
+
+After choosing the \qident{filter data} member, the class provided by \ident{filter} lets the 
+user select one of the columns \ball{4} based on the schema. The \ident{filter-val} function then 
+generates a class with members based on the available values for the specified column in the data 
+$D$ \ball{5}. The predicate that filters data based on the value \ball{6} is used both in the 
+provided runtime code and also when restricting the sample data set using $\sigma_{\varphi_v}(D)$ 
+before providing further filtering by recursively calling \ident{filter}.
+
+The fact that we transform the sample data when providing types is important for two reasons.
+It makes it possible to apply filtering after aggregation (which changes the format of data)
+and it means that more appropriate values are provided for faceted data. For example,
+Figure~\ref{fig:pivot-filter} shows some of the provided members when filtering data by medal,
+team and individual athlete. Once we refine the team using $\qident{Team is}.\ident{Mongolia}$
+and attempt to filter by athlete using \qident{Athlete is}, the type provider offers only the
+names of Mongolian athletes.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}
+\begin{equation*}
+\hspace{-1em}
+\begin{array}{l}
+\ident{olympics}.\qident{filter data}.\qident{Medal is}.\ident{Gold}.\qident{Team is}\\[0.25em]
+\quad\begin{array}{l}
+\rightarrow \qident{Czech Republic}.\qident{Athlete is}\\[0.25em]
+\qquad \rightarrow \qident{Barbora Spotakova}\\
+\qquad \rightarrow \qident{David Kostelecky}\\
+\qquad \rightarrow \qident{David Svoboda}\\
+\end{array}\qquad
+\begin{array}{l}
+\rightarrow \ident{Mongolia}.\qident{Athlete is}\\[0.25em]
+\qquad \rightarrow \qident{Badar-Uugan Enkhbat}\\
+\qquad \rightarrow \qident{Tuvshinbayar Naidan}\\
+\\
+\end{array}
+\end{array}
+\end{equation*}
+\caption{Subset of members provided by the filtering operation}
+\label{fig:pivot-filter}
+\end{figure}
 
 % ==================================================================================================
-
-\newpage
 

### 2017-01-11 pivot/paper.tex
Case study

@@
 \begin{center}
-\includegraphics[scale=0.35,trim=0mm 0mm 0mm 0mm,clip]{filter.png} % left bottom right top
+\includegraphics[scale=0.35,trim=0mm 0mm 0mm 0mm,clip]{images/filter.png} % left bottom right top
 \end{center}
@@
 
-\section{Implementation}
+\begin{figure}[b]
+\begin{center}
+\includegraphics[scale=0.30,trim=0mm 0mm 0mm 0mm,clip]{images/table.png} % left bottom right top
+\end{center}
+\caption{Athletes by the number of medals over the enitre history of Olympic games}
+\label{fig:case-table}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\section{Case study: Visualizing Olympic medalists}
 \label{sec:impl}
 
-Rio case study
-
-\begin{equation*}
-\begin{array}{l}
-\ident{olympics}\\
-\quad.\qident{filter data}.\qident{Games is}.\qident{Rio (2016)}.\ident{then}\\
-\quad.\qident{group data}.\qident{by Athlete}.\qident{sum Gold}.\ident{then}\\
-\quad.\qident{sort data}.\qident{by Gold descending}.\ident{then}\\
-\quad.\qident{paging}.\ident{take}(8)\\
-\quad.\qident{get series}.\qident{with key Athlete}.\qident{and value Gold}
-\end{array}
-\end{equation*}
-
-~
-\newpage
+We used The Gamma script with the pivot type provider to build an interactive web site 
+(\url{rio2016.thegamma.net}) that visualizes a number of facts about Olympic medalists 
+using the data set discussed in Appendix~\ref{app:olympics-csv} and used throughout this paper.
+The web site lets the readers view and modify the source code and we also developed a number of
+tools that make working with the source code easier, going beyond the basic auto-completion tooling
+to enable dot-driven development as discussed in Section~\ref{sec:analysis-auto}. In this section,
+we reivew our experience and outline some of the additional tools (available at 
+\url{github.com/the-gamma}).
+
+\subparagraph{Building tables and charts.} 
+Building rich visualizations with the simplicity of the pivot type provider was left for future
+work -- we provided functions for building basic kinds of visualizations (table, column chart, 
+pie chart and timeline) and we extended the host language with more advanced features that can
+be used to customize the displays. Figure~\ref{fig:case-table} shows a sample table, listing
+top athletes over the entire history of Olympic games.
+
+The data transformation used to construct the table uses the operations discussed in this paper
+together with paging functionality and \qident{get the data} which returns the entire data set
+of type \ident{Query}, as opposed to extracting a series with keys and values:
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~\ident{data}~=~\ident{olympics}\\
+\quad .\qident{group data}.\qident{by Athlete}\\
+\qquad .\qident{sum Gold}.\qident{sum Silver}.\qident{sum Bronze}.\qident{concat Team}.\ident{then}\\
+\quad .\qident{sort data}.\qident{by Gold descending}\\
+\qquad .\qident{and by Silver descending}.\qident{and by Bronze descending}.\ident{then}\\
+\quad .\ident{paging}.\ident{take}(10).\qident{get the data}\\[0.5em]
+\ident{table.create}(\ident{data})
+\end{array}
+\end{equation*}
+
+\noindent
+The $\ident{table.create}$ operation on the last line generates a table based on the columns 
+available in the data set. We omit the additional customization which specifies that medals should
+be rendered as images. For most visualizations we built, the pivot type provider was expressive
+enough to capture the core logic of the operation, but further joining of data was sometimes
+needed. Possible extensions that would allow capturing those are discussed in 
+Section~\ref{sec:further}.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\begin{center}
+\includegraphics[scale=0.32,trim=0mm 0mm 0mm 0mm,clip]{images/options.png} % left bottom right top
+\end{center}
+\caption{User interface with automatically provided grouping and sorting options}
+\label{fig:case-opts}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\subparagraph{Generating interactive user interfaces.}
+Although the pivot type provider simplifies code needed for data exploration, not every reader will 
+be able to write or modify it. The simplicity of the host language makes it possible to easily 
+generate user interface that allows changing of some of the parameters of the program. 
+Figure~\ref{fig:case-opts} shows an example for the above code snippet that we implemented as 
+part of the visualization.
+
+The user interface lets the user choose aggregations to be caluclated over a group and select 
+columns used for sorting. It is generated automatically by looking for specific pattern in the 
+chain of member accesses -- we annotate members with annotations denoting whether a member is 
+start of a list, list item or end of a list. The editor then looks for parts of the chain of the
+form $\qident{list start}.\qident{list item 1}.\qident{list item 2}.\qident{list end}$ and generates
+a control that lets the user remove or add list items. An item cannot be removed if the operation
+caused the code to break (e.g.~when it adds a member that is needed later) and items to be added
+are chosen using available members (as in the standard auto-complete). The header
+shown in Figure~\ref{fig:case-opts} are provided as additional annotations attached to 
+\qident{list start}.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\begin{center}
+\includegraphics[scale=0.31,trim=0mm 0mm 0mm 0mm,clip]{images/pivot.png} % left bottom right top
+\end{center}
+\caption{Auto-completion offering the available values of the athlete name column}
+\label{fig:case-ed}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\subparagraph{Spreadsheet-inspired live editor.}
+The third editor extension that we developed for the pivot type provider aims to bridge the gap 
+between code and user interface. This is done through a direct manipulation interface [X] inspired
+by spreadsheet applications. When exploring data in a spreadsheet, the user can always see the 
+data they work with and the results of an action will be immediately visible. For queries written
+using the pivot type provider, the results can also be calculated immediately using the sample 
+data set provided when using the refined version of the type provider as discussed in 
+Section~\ref{sec:pivot-filter}.
+
+The Figure~\ref{fig:case-ed} shows the sample expression (discussed above) in the live 
+editor\footnote{The live editor can be tested live as part of the documentation for the 
+JavaScript package at \url{thegamma.net}}.
+Note that the selected part of code is the \qident{by Gold descending} identifier and so the 
+preview shows data available at that point of the query evaluation. Athletes with largest number
+of gold medals appear first, but silver or bronze medals are not used as secondary sorting keys yet.
+As the user moves through the code, or writes the code, the live preview is updated accordingly.
+Finally, the editor also makes it possible to modify the code through the user interface -- the
+``x'' buttons can be used to remove sort keys or transformations and ``+'' buttons (on the right)
+can be used to add more transformations or parameters.
+
+Unlike the user interface for specifying lists, the live editor has been implemented specifically
+for the pivot type provider, but it still heavily relies on the simple structure provided by the 
+fact that entire transformation can be written as a single chain of member accesses. In particular,
+we identify individual transformations (\qident{group by}, \qident{sort by}, etc.) and then 
+generate different user interface for specifying parameters. For sorting, as shown in 
+Figure~\ref{fig:case-ed}, the second level lets the user add or remove members. For grouping or
+paging, the second level looks different and lets the user choose the key and the number of 
+elements to take, respectively.
+
+% ==================================================================================================
+
+\section{Discussion and related work}
+
+a
+
+b
+
+c
+
+\subsection{Further work}
+\label{sec:further}
+
+\subsection{Related work}
+
+visual query systems
+
+https://people.csail.mit.edu/ebakke/research/sieuferd_sigmod2016.pdf
+
+\subsection{Conclusions}
 

### 2017-01-12 pivot/paper.tex
References and related work

@@
 \EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-\EventShortTitle{CVIT 2016}
-\EventAcronym{CVIT}
+\EventShortTitle{DRAFT}
+\EventAcronym{DRAFT}
 \EventYear{2016}
@@
 The manual mode of interaction is not repeatable and it breaks the link with the original data 
-source, making spreadsheets error-prone [X]. The answer to this problem is to explore data 
+source, making spreadsheets error-prone \cite{exceldep}. The answer to this problem is to explore data 
 programmatically. A program can be run repeatedly and its parameters can be modified.
@@
 
-\item Our mechanism illustrates how to embed fancy types [X] into a simple nominally-typed programming  
+\item Our mechanism illustrates how to embed fancy types \cite{fancytypes} into a simple nominally-typed programming  
   language (Section~\ref{sec:columns}). We track names and types of available columns of the 
@@
 for providing type-safe access to rich information sources. A type provider is a compile-time 
-component that imports external information source into a programming language [X]. It provides two 
+component that imports external information source into a programming language \cite{inforich}. It provides two 
 things to the compiler or editor hosting it: a type signature that models the external source 
@@
 
-For example, the World Bank type provider [X] provides a fine-grained access to development 
+For example, the World Bank type provider \cite{ageweb} provides a fine-grained access to development 
 indicators about countries. The following accesses CO2 emissions by~country~in~2010:
@@
 
-\subparagraph{Types from data.} Recent work on the F\# Data library [X] uses type providers for 
+\subparagraph{Types from data.} Recent work on the F\# Data library \cite{fsdata} uses type providers for 
 accessing data in structured formats such as XML, CSV and JSON. This is done by inferring the 
 structure of the data from a sample document, provided as a static parameter to a type provider.
-In the following example, adapted from [X], a sample URL is passed to \ident{JsonProvider}:
+In the following example, adapted from \cite{fsdata}, a sample URL is passed to \ident{JsonProvider}:
 %
@@
 offers various available properties, such as \ident{Wind} and \ident{Clouds} once ``dot'' is typed
-after $\ident{ldn.Main}$. Other type providers [X] follow a similar pattern. It is worth noting
+after $\ident{ldn.Main}$. Other type providers \cite{inforich} follow a similar pattern. It is worth noting
 that despite the use of nominal typing, the names of types rarely explicitly appear in code -- we
 do not need to know the name of the type of $\ident{ldn.Main}$, but we need to know its members.
-Thus the type name can be arbitrary [X] and is used more as a lookup key.
+Thus the type name can be arbitrary \cite{fsdata} and is used more as a lookup key.
 
@@
 when using pivot type provider. This is a recurring theme in type provider development\footnote{The
-\ident{Individuals} property in the Freebase type provider [X] imports values into types in a similar way.}.
+\ident{Individuals} property in the Freebase type provider \cite{inforich} imports values into types in a similar way.}.
 
@@
 (Section~X), but all other operations (Section~Y) require only meta-data (names and types of 
-columns). Following the example of type providers for structured data formats [X], the schema could
+columns). Following the example of type providers for structured data formats \cite{fsdata}, the schema could
 be inferred from a representative sample.
@@
 
-% https://github.com/fsharp/fslang-design/blob/master/FSharp-4.0/StaticMethodArgumentsDesignAndSpec.md
-
-In F\# type providers, this could be done using methods with static parameters [X] by writing, 
+In F\# type providers, this could be done using methods with static parameters \cite{staticpar} by writing, 
 for example, $\ident{g}.\qident{sum Gold as}\langl\str{Total Gold}\rangl()$. In F\#, the value of 
@@
 \noindent
-Operations on the type of rows in the data set can be captured using row types [X]. In addition, 
-we need to annotate type with a form of type state [Y] to restrict what operations are available. 
+Operations on the type of rows in the data set can be captured using row types \cite{rowtypes}. 
+In addition, we need to annotate type with a form of type state \cite{typestate} to restrict what operations are available. 
 When dropping columns, we first access the \qident{drop columns} member, which sets the state to
@@
 \noindent
-The rule, adapted from [X], does not capture laziness of type providers that is 
+The rule, adapted from \cite{fsdata}, does not capture laziness of type providers that is 
 important for the pivot type provider (where the number of provided classes is potentially 
@@
 type state in a simple nominally typed language. In a similar way, type providers have been used
-to encode session types [X], suggesting that this is a generally useful approach.
+to encode session types \cite{sessiontp}, suggesting that this is a generally useful approach.
 
@@
 The focus of our work is on data aggregation and so we use a form of relational algebra with 
-extensions for grouping and sorting [X,Y,Z]. The syntax is defined in Figure~\ref{fig:foo-rel}.
+extensions for grouping and sorting \cite{relalg,dbsys}. The syntax is defined in Figure~\ref{fig:foo-rel}.
 We write $f$ for column (field) names and we include definition of a data value $D$, which maps
@@
 
-We model the host language using a variant of the Foo calculus [X]. The core of the calculus models
+We model the host language using a variant of the Foo calculus \cite{fsdata}. The core of the calculus models
 a simple object-based language with objects and members. The syntax of the language is shown in
@@
 type providers can provide potentially infinite number of types. In F\#, the types are provided 
-lazily as the type checker explores parts of the type space used by the program [X]. Consider:
+lazily as the type checker explores parts of the type space used by the program \cite{inforich}. 
+Consider:
 %
@@
 member of $C_1$ remain unevaluated when they do not appear in the source code. This aspect of type
-providers have been omitted in previous work [X,Y], but it is necessary for the pivot type provider.
+providers have been omitted in previous work \cite{fsdata,liteq}, but it is necessary for the pivot type provider.
 The typing rules given are written using the following judgement:
@@
 laziness as $L$ is a function that may never be evaluated. Since $L$ is potentially infinite, we 
-cannot check class definitions upfront as in typical object calculi [X]. Instead, we check that 
+cannot check class definitions upfront as in typical object calculi \cite{objects}. Instead, we check that 
 that members are well typed as they appear in the source code, which matches the behaviour of F\# 
@@
 untyped parts of the language resembles the boundary between static and dynamic typing in 
-gradually typed languages.% [X,Y,Z].
+gradually typed languages~\cite{gradual}.
 
@@
 and the editor to provide information about types on demand. In our formalization, we follow the 
-style used by Petricek et al. [X], but we add laziness as discussed in Section~\ref{sec:foo-foo}.
+style used by Petricek et al. \cite{fsdata}, but we add laziness as discussed in Section~\ref{sec:foo-foo}.
 We model the core operations (dropping columns, grouping and sorting) in Section~\ref{sec:pivot-core}
@@
 \noindent
-In prior work on type providers [X], the actual input value is not available when writing the code
+In prior work on type providers \cite{fsdata}, the actual input value is not available when writing the code
 and so the schema is inferred from a representative sample value. In exploratory data analysis, the 
@@
 The third editor extension that we developed for the pivot type provider aims to bridge the gap 
-between code and user interface. This is done through a direct manipulation interface [X] inspired
-by spreadsheet applications. When exploring data in a spreadsheet, the user can always see the 
+between code and user interface. This is done through a direct manipulation interface \cite{directman} 
+inspired by spreadsheet applications. When exploring data in a spreadsheet, the user can always see the 
 data they work with and the results of an action will be immediately visible. For queries written
@@
 
-Unlike the user interface for specifying lists, the live editor has been implemented specifically
-for the pivot type provider, but it still heavily relies on the simple structure provided by the 
-fact that entire transformation can be written as a single chain of member accesses. In particular,
-we identify individual transformations (\qident{group by}, \qident{sort by}, etc.) and then 
-generate different user interface for specifying parameters. For sorting, as shown in 
-Figure~\ref{fig:case-ed}, the second level lets the user add or remove members. For grouping or
-paging, the second level looks different and lets the user choose the key and the number of 
+Unlike the user interface for modifying lists, the live editor works specifically with the pivot 
+type provider. It still relies on the simple structure provided by the fact that entire 
+transformation can be written as a single chain of member accesses. In particular,
+we identify individual transformations (\qident{group by}, \qident{sort by}, etc.) and 
+generate different user interface for specifying parameters of each transformation. For sorting, as 
+shown in Figure~\ref{fig:case-ed}, the user can add or remove sort keys. For grouping 
+or paging, the user interface lets the user choose the grouping key and the number of 
 elements to take, respectively.
@@
 
-a
-
-b
-
-c
-
-\subsection{Further work}
+Although the technical focus of this paper is on the programming language theory behind the pivot type
+provider (Section~\ref{sec:pivot}), the paper also outlines interesting human-computer interaction 
+aspects (Section~\ref{sec:impl}). We discuss further related directions in this
+section before~concluding.
+
+\subsection{Related and further work}
 \label{sec:further}
 
-\subsection{Related work}
-
-visual query systems
-
-https://people.csail.mit.edu/ebakke/research/sieuferd_sigmod2016.pdf
+\subparagraph{Type providers.}
+Type providers have been pioneered in F\# \cite{inforich}. They can be seen as a form of dependent 
+typing \cite{idris-tp}; we take the opposite perspective and use type providers as a mechanism for 
+implementing other type system features. Our focus on using type providers for describing 
+computations is different from other type provider work \cite{fsdata,liteq,ageweb}, which focuses 
+on mapping of external data into types.
+
+\subparagraph{Fancy types.}
+The pivot type provider makes data exploration safer as it does not allow construction of invalid
+queries. Alternative approach would be to use fancy types \cite{fancytypes}. This can be done
+using using row types and typestate or phantom types \cite{rowtypes,typestate,phanty} as sketched
+in in Section~\ref{sec:columns-row}. The idea of using type providers in this way has also been
+explored for session types \cite{sessionty,sessiontp} and it would be interesting to see whether
+our approach can be applied in other areas such as web development \cite{urweb}.
+
+\subparagraph{Human-computer intercation.}
+We discussed how the pivot type provider simplifies the programming model (Section~\ref{sec:analysis}),
+but it would be interesting to explore this aspect empirically through the perspective of HCI. 
+The live editor shown in Section~\ref{sec:impl} offers a form of direct manipulation 
+\cite{directman,directman2,dynamicq}. Unlike spreadsheets, we construct a transformation rather than
+actually transforming data, which makes it more related to systems for query construction
+\cite{spreadsheetalgebra,querydirect}. We also see code as equally important to the direct 
+manipulation interface.
+
+\subparagraph{Relational algebra.}
+Our operational semantics used to model data transformations (Section~\ref{sec:foo}) was based on 
+relational algebra \cite{relalg,dbsys}, although our focus was on aggregation, which has been 
+added to the core algebra in a number of different ways \cite{sumtables,datacube,relalg-alpha,relalg-sparql}.
+The pivot type provider does not provide operations for joining data sets, which is an interesting
+problem for further work as it requires extensions to the type provider mechanism -- the join
+operation is parameterized by two data sets that are being combined.
 
 \subsection{Conclusions}
-
-\subparagraph*{Acknowledgements.}
-
-I want to thank \dots
+There is lots of data (big data), but using them is hard (post-truth).
+We want to make it easier to explore data without sacrificing reproducibility 
+and other good things we get from having source code.
+
+In this paper, we focus on making the language as simple as possible. We unify many different
+PL concepts using member access as the primary programming mechanism thanks to type providers.
+This makes programming easier through auto-completed, but we also explored other tooling ranging
+from simple interactive UIs to direct manipulation tools.
+
+The pivot type provider provides a safe and easy to use layer over an underlying 
+relational algebra. We formalize the host language and the type provider and we prove
+that queries constructed using the type provider are correct.
+
+TL;DR - better toools to democratize data stuff.
+
+\bibliography{paper}
 
@@
 
-\bibliography{lipics-v2016-sample-article}
-
 %% .. or use the thebibliography environment explicitely

### 2017-01-13 pivot/paper.tex
Final edits, add appendix

@@
 In this paper, we describe the design of a data exploration language that makes the task more 
-accessible by embedding advanced programming concepts in a simple core language.
+accessible by embedding advanced programming concepts into a simple core language.
 
 The core language uses type providers, but we employ them in a novel way -- rather than providing 
-types with members for accessing data, we provide types with members that allow the user to access 
-data, but also compose complex queries using only member access (``dot''). This lets us recreate 
-functionality that usually requires complex type systems (row polymporphism, type state, 
-dependent typing) in an extremely simple object-based language.
-
-We formalize our approach using a minimal object-based calculus and prove that
-programs constructed using the provided types represent valid data transformations. We then discuss 
-a prototype implementation of the language, together with a simple editor that bridges
-some of the gaps between programming and spreadsheets. We believe that this provides a pathway towards
-democratizing data science -- our use of type providers significantly reduce the complexity of 
-languages that one needs to understand in order to write scripts for exploring data.
+types with members for accessing data, we provide types with members that allow the user to also 
+compose rich and correct queries using just member access (``dot''). This way, we recreate 
+functionality that usually requires complex type systems (row polymporphism, type state and dependent 
+typing) in an extremely simple object-based language.
+
+We formalize our approach using an object-based calculus and prove that programs constructed using 
+the provided types represent valid data transformations. We discuss a case study developed using the 
+language, together with additional editor tooling that bridges some of the gaps between programming 
+and spreadsheets. We believe that this work provides a pathway towards democratizing data science 
+-- our use of type providers significantly reduce the complexity of languages that one needs to 
+understand in order to write scripts for exploring data.
 \end{abstract}
@@
 The rise of big data and open data initiatives means that there is an increasing amount of raw data 
-available. At the same time, the fact that ``post-truth'' was chosen as the word of 2016 suggests 
-that there has never been a greater need for increasing data literacy and building tools that let 
-anyone -- including journalists and interested citizens -- explore such data and use it to support 
-facts in a transparent way.
+available. At the same time, the fact that ``post-truth'' was chosen as the word of 2016 \cite{posttruth} 
+suggests that there has never been a greater need for increasing data literacy and building tools that let 
+anyone -- including journalists and interested citizens -- explore such data and use it to make
+transparent factual claims.
 
@@
 The manual mode of interaction is not repeatable and it breaks the link with the original data 
-source, making spreadsheets error-prone \cite{exceldep}. The answer to this problem is to explore data 
-programmatically. A program can be run repeatedly and its parameters can be modified.
+source, making spreadsheets error-prone \cite{exceldep,sprerrors}. One solution is to explore data 
+programmatically, as programs can be run repeatedly and their parameters can be modified.
 
@@
 surprisingly difficult. For example, consider the following Python program (using the pandas
-library), which reads a list of all Olympic medals ever awarded (see Appendix~\ref{app:olympics-csv}) 
+library), which reads a list of all Olympic medals awarded (see Appendix~\ref{app:olympics-csv}) 
 and finds top 8 athletes by the number of gold medals they won in Rio 2016:
@@
 the primary programming mechanism. Finding top 8 athletes by the number of gold medals from Rio 2016 
-can be written as (in source code $\qident{\ldots}$ is written as \textquotesingle$\ldots$\textquotesingle):
+can be written as:
 %
@@
 provides a list of available members when writing and modifying code. The members (such as 
-\qident{by Gold descending}) are generated by a type provier based on the knowledge of
-the data source and transformations applied so far -- thus only valid and meaningful 
+\qident{by Gold descending}) are generated by the pivot type provider based on the knowledge of
+the data source and transformations applied so far -- only valid and meaningful 
 operations are offered. The rest of the paper gives a detailed analysis and description of the mechanism.
 
-\subparagraph{Contributions.} This paper explores an interesting particular corner of the 
-programming language design space. We support it by a detailed analysis (Section~\ref{sec:analysis}), 
+\subparagraph{Contributions.} This paper explores an interesting new area of the 
+programming language design space. We support our design by a detailed analysis (Section~\ref{sec:analysis}), 
 formal treatment (Section~\ref{sec:pivot}) and an implementation with 
@@
 
-\item Our mechanism illustrates how to embed fancy types \cite{fancytypes} into a simple nominally-typed programming  
+\item Our mechanism illustrates how to embed ``fancy types'' \cite{fancytypes} into a simple nominally-typed programming  
   language (Section~\ref{sec:columns}). We track names and types of available columns of the 
-  manipulated data set (using a mechanism akin to row types), but the same mechanism can be used for 
+  manipulated data set (using a mechanism akin to row types), but our mechanism can be used for 
   embedding other advanced typing schemes into any Java-like language.
 
-\item We implement the language (\url{github.com/the-gamma}), make it available as a JavaScript 
-  component that can be used to build transparent data-driven visualizations (\url{thegamma.net}) 
-  and discuss a case study visualizing fun facts about Olympic medals (Section~\ref{sec:impl}).
-
-\item We formalize the language (Section~\ref{sec:foo}) and the type provider for data exploration
-  (Section~\ref{sec:pivot}) and show that queries constructed using the type provider are valid
+\item We formalize the language (Section~\ref{sec:foo}) and the pivot type provider (Section~\ref{sec:pivot}) 
+  and show that queries for exploring data constructed using the type provider are correct
   (Section~\ref{sec:pivot-prop}). Our formalization also covers the laziness of type providers, which
   is an important aspect not covered in the existing literature.
+
+\item We implement the language (\url{github.com/the-gamma}), make it available as a JavaScript 
+  component (\url{thegamma.net}) that can be used to build transparent data-driven visualizations 
+  and discuss a case study visualizing facts about Olympic medalists (Section~\ref{sec:impl}).
 \end{itemize}
@@
 \ident{getByYear} function that downloads data for a specified indicator represented by an ID. 
-The type provider provides a type-safe access to known indicators, which exist only as strings in 
-the compiled code, increasing safety and making data access easier thanks to auto-completion 
-(which offers a list of available indicators).
+The indicators exist only as strings in compiled code, but the type provider provides a 
+type-safe access to known indicators, increasing safety and making data access easier thanks 
+to auto-completion (which offers a list of available indicators).
 
@@
 data in the external data source -- here, we access the temperature using \ident{ldn.Main.Temp}. 
-The provided code attempts to access the corresponding nested field and convert it to a number.
+The provided code attempts to access the corresponding nested field and converts it to a number.
 The relative safety property of the type provider guarantees that this will not fail if the sample 
@@
 same general mechanism as the F\# type providers discussed above, although it is embedded in a 
-simple language that runs in a web browser.
+simple host language that runs in a web browser.
 
@@
 type providers for importing external data sources (by providing members that correspond to parts 
-of the data). Instead, we use type providers for lazily generating types with members that let 
-users compose type-safe queries over the data source. As discussed in Section~\ref{sec:foo}, we 
-model the provided code by a relational algebra.
+of the data). Instead, we use type providers to lazily generate types with members that let 
+users compose type-safe queries over the data source.
 
@@
 one important difference -- the schema provided by the pivot type provider is potentially infinite
-(as there are always more operations that can be applied). To support this, we use the fact that
-type providers are integrated into the type system and types can be provided lazilly. This is 
+(as there are always more operations that can be applied). The implementation relies on the fact 
+that type providers are integrated into the type system and types can be provided lazilly. This is 
 also a new aspect of our formalization in Section~\ref{sec:foo}.
@@
 
-In Section~\ref{sec:intro}, we contrasted a data exploration script written using a popular Python 
-library pandas with a script written using the pivot type provider. In this section, we analyze 
-what makes the Python code complex (Section~\ref{sec:analysis-why}) and how our design simplifies it.
+In Section~\ref{sec:intro}, we contrasted a data exploration script written using the popular Python 
+library pandas \cite{pandas} with a script written using the pivot type provider. In this section, we 
+analyze what makes the Python code complex (Section~\ref{sec:analysis-why}) and how our design simplifies it.
 
@@
 We consider the Python example from Section~\ref{sec:intro} for concreteness, but the following 
-four points are shared with other commonly used libraries and languages. We use them to guide our
-alternative design discussed in the rest of this section.
+four points are shared with other commonly used libraries and languages. We use the four points to 
+inform our alternative design as discussed in the rest of this section.
 
@@
   an expression $\ident{olympics}[\str{Games}]==\str{Rio (2016)}$ returning a vector of Booleans
-  while in the other, we specify a parameter value using $\ident{by}=\str{Gold}$. In other languges,
+  while in the other, we specify a column name using $\ident{by}=\str{Gold}$. In other languges,
   a parameter can also be a lambda function specifying a predicate or a transformation.
@@
   be applied over a group. A similar way of specifying multiple operations or results is common 
-  in other languages. For example, anonymous types in LINQ play the same role.
+  in other languages. For example, anonymous types in LINQ \cite{linq} play the same role.
 
@@
   for chained operations such as the one in our example\footnote{For an anecdotal evidence,
-  see for example: \url{stackoverflow.com/questions/25801246/}}. Statically-typed languages
-  provide better tooling, but at the cost of higher complexity\footnote{Again,
-  See \url{fslab.org/Deedle}
-  and \url{extremeoptimization.com/Documentation/Data_Frame/Data_Frames.aspx} }.
-
-\item In the Python example (as well as in most other data manipulation 
-  libraries\footnote{deedle, etc.}), column names are specified as strings. This makes static 
-  checking of column names and auto-completion difficult. For example, \str{Gold} is a valid 
-  column name when calling \ident{sort\_values}, but we only know that because it is a key of the 
-  dictionary passed to \ident{agg} before.
+  see for example: \url{stackoverflow.com/questions/25801246}}. Statically-typed languages
+  provide better tooling, but at the cost of higher complexity\footnote{A detailed evaluation 
+  is out of the scope of this paper, but the reader can compare the Python example with 
+  F\# code using Deedle (\url{fslab.org/Deedle}), Haskell Frames library (\url{acowley.github.io/Frames})
+  and similar C\# project (\url{extremeoptimization.com/Documentation/Data_Frame})}.
+
+\item In the Python example (as well as in most other data manipulation libraries), column names are 
+  specified as strings\footnote{This is the case for Deedle and the aforementioned C\# library. 
+  Haskell Frames \cite{frames} tracks column names statically, arguably at the cost of higher code 
+  complexity when compared with Python.}. This makes static checking of column names and auto-completion 
+  difficult. For example, \str{Gold} is a valid column name when calling \ident{sort\_values}, but 
+  we only know that because it is a key of the dictionary passed to \ident{agg} before.
 \end{itemize}
@@
 enable auto-completion (Section~\ref{sec:analysis-auto}); we use operation-chaining via member access
-for constructing dictionaries (Section~\ref{sec:analysis-and}) and we track column names using the 
-pivot type provider (Section~\ref{sec:columns}).
+for constructing dictionaries (Section~\ref{sec:analysis-and}) and we track column names statically
+in the pivot type provider (Section~\ref{sec:columns}).
 
@@
 \ident{data}[\ident{data}[\str{Games}]==\str{Rio (2016)}] \qquad\ball{1}\\
-\ident{data}.\ident{filter}(\kvd{fun}~\ident{row}\rightarrow\ident{row}.\ident{Games}==\str{Rio (2016)})\qquad\ball{2}\\
+\ident{data}.\ident{filter}(\kvd{fun}~\ident{row}\rightarrow\ident{row}.\ident{Games}=\str{Rio (2016)})\qquad\ball{2}\\
 \ident{data}.\ident{sort\_values}(\ident{by}=\str{Gold}, \ident{ascending}=\kvd{False})\qquad\ball{3}\\
@@
 We aim to unify the above examples using a single language construct that offers a high-level programming
-model\footnote{In contrast, s-expressions in LISP are typically used as a lower-level encoding of 
-higher-level constructs. For example \ident{lambda} symbol encodes a lambda function as an s-expression.}
-and can be supported by modern tooling (as discussed in Section~\ref{sec:analysis-auto}).
+model and can be supported by modern tooling (as discussed in Section~\ref{sec:analysis-auto}).
 Member access provides an extremely simple programming construct that is, in conjunction with
-the type provider mechanism capable of expressing the above data transformations in a uniform way: 
+the type provider mechanism, capable of expressing the above data transformations in a uniform way: 
 %
@@
 \noindent
-The member names tend to be longer and descriptive and we write them using \qident{...}.
+The member names tend to be longer and descriptive. Quoted names appear as \textquotesingle$\ldots$\textquotesingle~in
+code, but we typeset them using \qident{...} for readability.
 The names are not usually typed by the user (see~Section~\ref{sec:analysis-auto}) and so the
@@
 for use as sort keys. When filtering \ball{2}, the member \qident{Rio (2016)} is provided based
-on the values in the data source (we discuss this furter in Section~\ref{sec:analysis-auto}).
+on the values in the data source (we discuss this furter in Section~\ref{sec:pivot-filter}).
 
@@
 
-Using descriptive member names is only viable when the names are discoverable. The above
+Using descriptive member names is only possible when the names are discoverable. The above
 code could be executed in a dynamically-typed languge that allows custom message-not-understood
@@
 The members provided, for example, for filtering limit the number of conditions that can be 
-written, because the user is limited to choosing one of the provided members.
-As discussed in the case study based on our implementation (Section~\ref{sec:impl}),
+written, because the user is restricted to choosing one of the provided members.
+As illustrated by the case study based on our implementation (Section~\ref{sec:impl}),
 this appears sufficient for many common data exploration tasks. The mechanism could be made
@@
 \quad\rightarrow~\qident{by Discipline}\\
+\quad\rightarrow~\qident{by Discipline descending}\\
 \end{array}\qquad
@@
 \quad\rightarrow~\qident{by Athlete}\\
-\qquad\quad\rightarrow~\qident{distinct Discipline}\\
 \qquad\quad\rightarrow~\qident{average Year}\\
@@
 \qquad\quad\rightarrow~\qident{distinct Athlete}\\
+\qquad\quad\rightarrow~\qident{concat Athlete}\\
 \qquad\quad\rightarrow~\qident{distinct Discipline}\\
+\qquad\quad\rightarrow~\qident{concat Discipline}\\
 \end{array}
@@
 Source code editors for object-based languages with nominal type sytems often provide 
-auto-completion for members of objects. This combination appear to work extremely well; the
+auto-completion for members of objects. This combination works extremely well in practice; the
 member list is a complete list of what might follow after typing ``dot'' and it can be easily
@@
 do not need to know the name of the type of $\ident{ldn.Main}$, but we need to know its members.
-Thus the type name can be arbitrary \cite{fsdata} and is used more as a lookup key.
+Thus the type name can be arbitrary \cite{fsdata} and is used merely as a lookup key.
 
@@
 \subparagraph{Values vs. types.}
-As the Figure~\ref{fig:dot-driven} illustrates, the pivot type provider sometimes blurs the 
-distinction between values and (members of) types. In the example in Section~\ref{sec:intro},
+As Figure~\ref{fig:dot-driven} illustrates, the pivot type provider sometimes blurs the 
+distinction between values and types. In the example in Section~\ref{sec:intro},
 \str{Rio (2016)} is a string value in Python, but a statically-typed member \qident{Rio (2016)}
-when using pivot type provider. This is a recurring theme in type provider development\footnote{The
+when using the pivot type provider. This is a recurring theme in type provider development\footnote{The
 \ident{Individuals} property in the Freebase type provider \cite{inforich} imports values into types in a similar way.}.
@@
 data set available. The pivot type provider uses this when offering possible values for filtering
-(Section~X), but all other operations (Section~Y) require only meta-data (names and types of 
-columns). Following the example of type providers for structured data formats \cite{fsdata}, the schema could
-be inferred from a representative sample.
+(Section~\ref{sec:pivot-filter}), but all other operations (Section~\ref{sec:pivot-core}) require 
+only meta-data (names and types of columns). Following the example of type providers for structured 
+data formats~\cite{fsdata}, the schema could be inferred from a representative sample.
 
@@
 This is an interesting case, because when encoding program as a sequence of member accesses, 
-there is no built-in support for nesting. In the pivot type provider, we use ``then'' design
+there is no built-in support for nesting. In the pivot type provider, we use the ``then'' design
 pattern to provide operations that require nesting. The following example specifies multiple 
@@
 When grouping, we sum the number of gold and silver medals and concatenates team names~\ball{1}. 
-Then we sorts the grouped data using multiple sorting keys \ball{2} -- first by number of gold 
-medals and then by silver medals (within a group with the same number of gold medals).
+Then we sort the grouped data using multiple sorting keys \ball{2} -- first by the number of gold 
+medals and then silver medals (within a group with the same number of gold medals).
 
 \subparagraph{The ``then'' pattern.}
-Nesting is no doubt essential programming construct and it may very well be desirable to support it 
-directly in the language, but the ``then'' pattern lets us express it without explicit langauge
-support. In both cases, the nested structure is specified by selecting one or more members and then 
+Nesting is an essential programming construct and it may be desirable to support it directly in 
+the language, but the ``then'' pattern lets us express nesting without langauge support. In both 
+of the cases above, the nested structure is specified by selecting one or more members and then 
 ending the nested structure using the \ident{then} member.
@@
 Thanks to the dot-driven development and the ``then'' pattern, the user is offered possible 
-parameter values (aggregations or sorting keys) even when creating a nested structure. This is
-harder to do with general-purpose nesting as ordinary languages (even domain-specific languages)
-often provide many different ways for creating a nested structure.
+parameter values (aggregations or sorting keys) even when creating a nested structure. We also use
+the simple structure of the ``then'' pattern to automatically generate interactive user
+interfaces for specifying aggregation and sorting parameters (Section~\ref{sec:impl}).
 
@@
 
-In F\# type providers, this could be done using methods with static parameters \cite{staticpar} by writing, 
+In type providers for F\#, this could be done using methods with static parameters \cite{staticpar} by writing, 
 for example, $\ident{g}.\qident{sum Gold as}\langl\str{Total Gold}\rangl()$. In F\#, the value of 
@@
 The last difficulty with data scripting discussed in Section~\ref{sec:analysis-why} is that pandas 
-(and other data exploration libraries, even for statically-typed languages) track column names as 
-strings at runtime, making code error-prone and auto-complete on column names difficult to provie.
+(and most other data exploration libraries, even for statically-typed languages) track column names as 
+strings at runtime, making code error-prone and auto-complete on column names difficult to support.
 Proponents of static typing would correctly point out that column names and their types can be
@@
 
-\begin{figure}
-
-\begin{equation*}
-\inference[(\emph{drop-start})~]
-  {\Gamma \vdash e : [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]}
-  {\Gamma \vdash e.\qident{drop columns}: [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]_{\ident{drop}}}
-\end{equation*}
-
-\begin{equation*}
-\inference[(\emph{drop-col})~]
-  {\Gamma \vdash e : [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]_{\ident{drop}}}
-  {\Gamma \vdash e.\qident{drop~$f_i$}: [f_1\!:\!\tau_1, \ldots, f_{i-1}\!:\!\tau_{i-1}, f_{i+1}\!:\!\tau_{i+1}, \ldots, f_n\!:\!\tau_n ]_{\ident{drop}}}
-\end{equation*}
-
-\begin{equation*}
-\inference[(\emph{drop-then})~]
-  {\Gamma \vdash e : [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]_{\ident{drop}}}
-  {\Gamma \vdash e.\qident{then}: [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]}
-\end{equation*}
-
-\caption{Tracking available column names with row types and type state}
-\label{fig:fancy-types}
-\end{figure}
-
-% --------------------------------------------------------------------------------------------------
-
 \subsection{Using row types and type state}
@@
 what columns (and of what types) are available. When grouping and aggregating data, the resulting
-data set has columns depending on the aggregates calculated. Another simpler operation is adding
-or removing a column from the data set. For example, given the Olympic medals data set, we can 
-drop games and year as follows:
+data set has columns depending on the aggregates calculated. For simplicity, we consider another 
+operation -- removing column from the data set. For example, given the Olympic medals data set, we 
+can drop \ident{Games} and \ident{Year} columns as follows:
 %
@@
 \noindent
-Operations on the type of rows in the data set can be captured using row types \cite{rowtypes}. 
-In addition, we need to annotate type with a form of type state \cite{typestate} to restrict what operations are available. 
+Operations that change the type of rows in the data set can be captured using row types~\cite{rowtypes}. 
+In addition, we need to annotate type with a form of typestate \cite{typestate} to restrict what operations are available. 
 When dropping columns, we first access the \qident{drop columns} member, which sets the state to
@@
 types $\alpha$, types annotated with a type state \ident{lbl} or row type with fields $f$:
-%
+
+\noindent
 \begin{equation*}
@@
 Other data transformations could be type checked in a similar way, but there are two drawbacks.
-First, row types and type state (although relatively straightforward) make the host language more
+First, row types and typestate (although relatively straightforward) make the host language more
 complex. Second, rules such as (\emph{drop-col}) make auto-completion more difficult, because the
@@
 
+\begin{figure}
+
+\begin{equation*}
+\inference[(\emph{drop-start})~]
+  {\Gamma \vdash e : [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]}
+  {\Gamma \vdash e.\qident{drop columns}: [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]_{\ident{drop}}}
+\end{equation*}
+
+\begin{equation*}
+\inference[(\emph{drop-col})~]
+  {\Gamma \vdash e : [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]_{\ident{drop}}}
+  {\Gamma \vdash e.\qident{drop~$f_i$}: [f_1\!:\!\tau_1, \ldots, f_{i-1}\!:\!\tau_{i-1}, f_{i+1}\!:\!\tau_{i+1}, \ldots, f_n\!:\!\tau_n ]_{\ident{drop}}}
+\end{equation*}
+
+\begin{equation*}
+\inference[(\emph{drop-then})~]
+  {\Gamma \vdash e : [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]_{\ident{drop}}}
+  {\Gamma \vdash e.\qident{then}: [f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]}
+\end{equation*}
+
+\caption{Tracking available column names with row types and type state}
+\label{fig:fancy-types}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
 \subsection{Using the pivot type provider}
@@
 
-In our approach, the information about available fields is used by a type provider to provide
-types with appropriate members. This is hidden from the host language, which only sees class types
-with members. Provided class definitions consist of a constructor and members: 
+In our approach, the information about available fields is used by the pivot type provider to provide
+types with appropriate members. This is hidden from the host language, which only sees class types.
+Provided class definitions consist of a constructor and members: 
 %
@@
 \noindent
-During type checking, the type system keeps track of a list of provided class definitions $L$. 
+During type checking, the type system keeps track of a lookup of provided class definitions $L$. 
 Checking member access is then just a matter of finding the corresponding class definition and
@@
 \inference[(\emph{member})~]
-  {L; \Gamma \vdash e : C & \kvd{type}~C(x:\tau) = ..\;\kvd{member}~N_i : \tau_i = e_i\;.. \in L}
+  {L; \Gamma \vdash e : C & L(C)=\kvd{type}~C(x:\tau) = ..\;\kvd{member}~N_i : \tau_i = e_i\;..}
   {L; \Gamma \vdash e.N_i:\tau_i}
@@
 
-Using type providers and nominal type system seemingly hides knowledge about fields available
+Using type providers and nominal type system hides knowledge about fields available
 in the data set. However, for types constructed by the pivot type provider, we can define a 
@@
 mechanism makes it possible to express safety guarantees that would normally require row types and 
-type state in a simple nominally typed language. In a similar way, type providers have been used
+typestate in a simple nominally typed language. In a similar way, type providers have been used
 to encode session types \cite{sessiontp}, suggesting that this is a generally useful approach.
@@
     \qquad a_i = \vect{\Sigma_{k\in g_1}v_{j,k}, \ldots, \Sigma_{k\in g_s}v_{j,k}} & \textnormal{when}~\rho_i = \ident{sum}~f_j \\
-    \qquad a_i = \vect{\Pi_{k\in g_1}v_{j,k}, \ldots, \Sigma_{k\in g_s}v_{j,k}}   & \textnormal{when}~\rho_i = \ident{conc}~f_j \\
+    \qquad a_i = \vect{\Pi_{k\in g_1}v_{j,k}, \ldots, \Pi_{k\in g_s}v_{j,k}}   & \textnormal{when}~\rho_i = \ident{conc}~f_j \\
     \qquad a_i = \vect{|\{v_{j,k}\,|\, k\in g_1\}|, \ldots, |\{v_{j,k}\,|\, k\in g_s\}|}  & \textnormal{when}~\rho_i = \ident{dist}~f_j \\
@@
 Type providers often provide a thin type-safe layer over richer untyped runtime components. In case 
-of providers for data access (Section~\ref{sec:tps}), the untyped runtime component performs operations
-over external data sources. In case of the pivot type provider, the untyped runtime component 
+of providers for data access (Section~\ref{sec:tps}), the untyped runtime component performs lookups
+into external data sources. In case of the pivot type provider, the untyped runtime component 
 is a relational algebra modelling data transformations. We formalize the relational algebra in 
-Section~\ref{sec:foo-rel}, together with the object-based host language in Section~\ref{sec:foo-foo}.
+Section~\ref{sec:foo-rel}, followed by the object-based host language in Section~\ref{sec:foo-foo}.
 
@@
 We write $f$ for column (field) names and we include definition of a data value $D$, which maps
-columns to vectors of length $r$ storing the data (values $v$ are defined below).
+column names to vectors of length $r$ storing the data (values $v$ are defined below).
 Aside from standard projection $\Pi$ and selection $\sigma$, our algebra includes sorting 
@@
 constructed using the relational algebra. The type $\ident{series}\langl \tau_1, \tau_2 \rangl$ 
-models a data series mapping keys of type $\tau_1$ to values of type $\tau_2$. A series is created 
-as a result of a query written using the pivot type provider. In the actual implementation, a series 
-can be passed to a charting library. A series is a statically-typed wrapper over a \ident{Query} 
-value and the proofs in Section~\ref{sec:pivot-prop} show that a series obtained from the pivot type 
-provider contains keys and values of matching~types.
+models a type-safe data series mapping keys of type $\tau_1$ to values of type $\tau_2$ that
+can be used, for example, as input for a charting library. A series is a typed wrapper over a 
+\indet{Query} value and the proofs in Section~\ref{sec:pivot-prop} show that a series obtained 
+from the pivot type provider contains keys and values of matching~types.
 
@@
 \subparagraph{Type checking.} One interesting aspect of type checking with type providers is that
-type providers can provide potentially infinite number of types. In F\#, the types are provided 
+type providers can provide potentially infinite number of types. The types are provided 
 lazily as the type checker explores parts of the type space used by the program \cite{inforich}. 
@@
 obtain full definition of $C_2$ when checking the member $\qident{by Athlete}$. Types of other 
-member of $C_1$ remain unevaluated when they do not appear in the source code. This aspect of type
+members of $C_1$ remain unevaluated when they do not appear in the source code. This aspect of type
 providers have been omitted in previous work \cite{fsdata,liteq}, but it is necessary for the pivot type provider.
@@
 of expression $e$ is $\tau$ and the type checking evaluated class definitions that are
-now included in~$L_2$. The structure of class definitions $L_1$ is a function mapping a class name 
+now included in~$L_2$. The resulting context obtained by type checking contains all definitions
+that may be needed when running the program and is passed to the reduction operation $\leadsto_{L}$.
+
+The structure of class definitions $L$ is a function mapping a class name 
 $C$ to a pair consisting of the definition and a function that provides definitions of 
@@
 \begin{equation*}
-L_1(C) = \kvd{type}~C(x:\tau) = \overline{m}, L
-\end{equation*}
-
-\noindent
-The class $C$ may use classes defined in $L_1$, but also delayed classes from $L$. This models 
-laziness as $L$ is a function that may never be evaluated. Since $L$ is potentially infinite, we 
+L(C) = \kvd{type}~C(x:\tau) = \overline{m}, L'
+\end{equation*}
+
+\noindent
+The class $C$ may use classes defined in $L$, but also delayed classes from $L'$. This models 
+laziness as $L'$ is a function that may never be evaluated. Since $L$ is potentially infinite, we 
 cannot check class definitions upfront as in typical object calculi \cite{objects}. Instead, we check that 
@@
 \begin{figure}[t]
-\vspace{-1em}
+\vspace{-1.5em}
 \[
@@
 \]
-\vspace{-2em}
+\vspace{-2.25em}
 
@@
 \]
-\vspace{-1em}
+\vspace{-0.5em}
 \caption{Type-checking of Foo expressions with lazy context}
+\vspace{-0.5em}
 \label{fig:foo-typecheck}
@@
 The rules for primitive types and variables are standard. Input data (\emph{data}) is of type 
-\ident{Query} and all the operations of relational algebra take a value of type \ident{Query} and 
-produce a result of type \ident{Query}. An untyped \ident{Query} value can be converted into a 
-series type (\emph{series}) of any type -- in the pivot type provider, the \ident{Query} type is 
-always hidden and users can only work with type-safe series type. The bridging of typed and 
-untyped parts of the language resembles the boundary between static and dynamic typing in 
-gradually typed languages~\cite{gradual}.
+\ident{Query} and all the operations of relational algebra take \ident{Query} input and 
+produce \ident{Query} results. An untyped \ident{Query} value can be converted into a 
+series (\emph{series}) of any type, akin to the boundary between static and dynamic typing 
+in gradually typed languages~\cite{gradual}. When provided by the pivot type provider, 
+the operation produces series with values of correct~types.
 
@@
 
-In an implementation, a type provider is an executable component that is called by the compiler
-and the editor to provide information about types on demand. In our formalization, we follow the 
-style used by Petricek et al. \cite{fsdata}, but we add laziness as discussed in Section~\ref{sec:foo-foo}.
+A type provider is an executable component called by the compiler and the editor to 
+provide information about types on demand. In our formalisation, we follow the 
+style of Petricek et al. \cite{fsdata}, but we add laziness as discussed in Section~\ref{sec:foo-foo}.
 We model the core operations (dropping columns, grouping and sorting) in Section~\ref{sec:pivot-core}
-and then refine the model to include filtering based on values Section~\ref{sec:pivot-filter}.
-We omit paging, as it does not affect the shape of data, but modelling it would require richer 
-host language.
+and refine the model to include filtering Section~\ref{sec:pivot-filter}.
+For simplicity we omit paging, which does not affect the shape of data.
 
@@
 A type provider is a function that takes static parameters, such as schema of the input data set, 
-and returns a class name $C$ together with a mapping that defines the body of the class, together 
-with definitions of associated classes $L$ that may be used by the members of the class $C$. 
+and returns a class name $C$ together with a mapping that defines the body of the class and 
+definitions of delayed classes $L$ that may be used by the members of the class $C$. 
 In our case, the schema $F$ is a mapping from field names to field types:
@@
 when a program accesses a member of the class $C$, modelling the laziness of the type provider.
-In the implementation, we return the name $C$, but only need to compute the definition when the 
-type checker needs to access the body.
+In the implementation, we return the name $C$ together with a function that computes the definition 
+of the class when the type checker needs to inspect the body.
 
@@
 \ident{sort}(F) = C, \{ C \mapsto (l, \bigcup L_f \cup \bigcup L'_f) \}\qquad\ball{3}\\[0.25em]
-~ l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\{f~|~f\mapsto\kvd{num}\in F\},~\textnormal{where} \\
+~ l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\{f~|~F(f)=\kvd{num} \},~\textnormal{where} \\
 \quad \qquad \kvd{member}~\qident{by $f$ descending} : C_f = C_f(x) &\quad C_f, L_f = \ident{sort-and}(F, \vect{f \mapsto \ident{desc}})\\
@@
 \multicolumn{2}{l}{ \ident{sort-and}(F,\vect{s_1,\ldots,s_n}) = C, \{ C \mapsto (l, \bigcup L_f \cup \bigcup L'_f \cup L') \}\qquad\ball{4} }\\[0.25em]
-~ l = \kvd{type}~C(x:\ident{Query}) = &\hspace{-1em}\forall f\in\{f~|~f\mapsto\kvd{num}\in F, \,\nexists i.s_i=f'\mapsto\omega\wedge f'=f\}\\
+~ l = \kvd{type}~C(x:\ident{Query}) = &\hspace{-1em}\forall f\in\{f~|~F(f)=\kvd{num}, \,\nexists i.s_i=f'\mapsto\omega\wedge f'=f\}\\
 \quad \qquad \kvd{member}~\qident{and $f$ descending} : C_f = C_f(x) &\quad C_f, L_f = \ident{sort-and}(F, \vect{s_1,..,s_n, f \mapsto \ident{desc}})\\
@@
 return type is generated using the main \ident{pivot} function. Note that columns removed from the 
-schema $F'$ as the types are generated match the columns removed from the data set at runtime  
+schema $F'$ match the columns removed from the data set at runtime  
 using $\Pi_{\dom(F')}$.
@@
 generates type for choosing the first sorting key and \ident{sort-and}~\ball{4} lets the user add more
-keys. The members are restricted to numerical columns (by checking $f\mapsto\kvd{num}\in F$).
+keys. The members are restricted to numerical columns (by checking $F(f)=\kvd{num}$).
 The sort keys are kept as a vector. The \ident{sort} operation creates a singleton vector;
@@
 already (by checking that the column $f$ does not match column name of any of the existing 
-keys $s_i = f'\mapsto\omega$).
+keys $\nexists i.\,s_i = f'\mapsto\omega$).
 
@@
 resulting schema contains only the column used as the key with its original type (which is always 
-generated by $\Phi$) and the set of aggregations to be calculated is empty.
+implicitly added by $\Phi$) and the set of aggregations to be calculated is empty.
 
@@
 
-\newpage
-
 \subsection{Properties of the pivot type provider}
@@
 If we were using the relational algebra formalised in Section~\ref{sec:foo-rel} to construct 
-queries, we can obtain an invalid program, e.g.~by attempting to select
+queries, we can write an invalid program, e.g.~by attempting to select
 a column $f$ using $\Pi_{f}$ from a data set that does not contain the column. This is not an issue 
@@
 \label{thm:foo-sound}
-For all $L_0, e, e'$, if $L_0, \emptyset \vdash e : \tau, L_1$ and $e\leadsto_{L_1}^{*} e'$ 
-then either $e'$ is a value, or there exists $e''$ such that $e'\leadsto_{L_1}^{*} e''$, or
+For all $L_0, e, e'$, if $L_0, \emptyset \vdash e : \tau, L_1$ and $e\leadsto_{L_1} e'$ 
+then either $e'$ is a value, or there exists $e''$ such that $e'\leadsto_{L_1} e''$, or
 $e'$ has one of the following forms:
@@
 For relational algebra operations, the expression can either be reduced or has one of the required 
-forms. For (\emph{member}) the expression can always be reduced.
+forms. For (\emph{member}) typing guarantees reduction is possible.
 \end{proof}
@@
 The pivot type provider defined by \ident{pivot} defines an entry-point class and a context $L$ 
-containing delayed classes. The type system cannot check type definitions in $L$ upfront, but 
-we prove that the body of all provided members is well-typed. Furthermore, type checking could fail
-if a delayed class was not discovered before it is needed in the (\emph{new}) and (\emph{member}) 
+containing delayed classes. Our type system does not check type definitions in $L$ upfront
+(although this is possible in dependently-typed languages~\cite{idris-tp}), but 
+we prove that the body of all provided members is well-typed.
+
+Type checking can also fail if a delayed class was not discovered before it is needed in the (\emph{new}) and (\emph{member}) 
 typing rules (Figure~\ref{fig:foo-typecheck}). We show that this cannot happen for the context
@@
 Given a context $L$, we define $n^{\textnormal{th}}$ expansion of $L$, written 
-$\downarrow_n L$:
-\begin{itemize}
-\item $\downarrow_n L = \downarrow_{n-1} L \cup \bigcup L_n$ assuming that $\downarrow_{n-1} L = \{ C_0\mapsto(l_0, L_0),\ldots,C_n\mapsto(l_n, L_n) \}$
-\item $\downarrow_0 L = L$  
-\end{itemize}
+$\downarrow_n L$ such that
+$\downarrow_{n+1} L = \downarrow_{n} L \cup \bigcup L_n$ where $\downarrow_{n} L = \{ C_0\mapsto(l_0, L_0),\ldots,C_n\mapsto(l_n, L_n) \}$
+and $\downarrow_0 L = L$.
 \end{definition}
@@
 
-\noindent
-In the revised model, the \ident{pivot} function which models the type provider takes the schema
+In the revised formalisation, the \ident{pivot} function which models the type provider takes the schema
 $F$ together with sample data $D$ and provides the type with class context:
@@
 \noindent
-In prior work on type providers \cite{fsdata}, the actual input value is not available when writing the code
-and so the schema is inferred from a representative sample value. In exploratory data analysis, the 
+In prior work \cite{fsdata}, the input value is not available when writing 
+the code and so the schema is inferred from a representative sample. In exploratory data analysis, the 
 data set is often available at the time of writing the code and so $D$ can be the actual data set.
-However, it could equally be a sample or representative subset of the data.
 
@@
 together with one of the operations discussed before and the newly added \ident{filter} function. 
-
 As members members performing data transformations are generated, the provider applies the same
@@
 After choosing the \qident{filter data} member, the class provided by \ident{filter} lets the 
-user select one of the columns \ball{4} based on the schema. The \ident{filter-val} function then 
+user select one of the columns \ball{4} based on the schema; \ident{filter-val} then 
 generates a class with members based on the available values for the specified column in the data 
 $D$ \ball{5}. The predicate that filters data based on the value \ball{6} is used both in the 
-provided runtime code and also when restricting the sample data set using $\sigma_{\varphi_v}(D)$ 
-before providing further filtering by recursively calling \ident{filter}.
+runtime code and when restricting the sample data set using $\sigma_{\varphi_v}(D)$ 
+in the type provider when recursively calling \ident{filter}.
 
@@
 \subparagraph{Building tables and charts.} 
-Building rich visualizations with the simplicity of the pivot type provider was left for future
-work -- we provided functions for building basic kinds of visualizations (table, column chart, 
-pie chart and timeline) and we extended the host language with more advanced features that can
-be used to customize the displays. Figure~\ref{fig:case-table} shows a sample table, listing
-top athletes over the entire history of Olympic games.
-
-The data transformation used to construct the table uses the operations discussed in this paper
+As part of the case study, we implement functions for building basic visualizations 
+(table, column chart, pie chart and timeline) and we extended the host language with more advanced 
+features that can be used to customize the displays. Building rich visualizations with the 
+simplicity of the pivot type provider is an interesting future work. Figure~\ref{fig:case-table} 
+shows a sample table, listing top athletes over the entire history of Olympic games.
+
+The data transformation used to construct the table include the operations discussed in this paper
 together with paging functionality and \qident{get the data} which returns the entire data set
@@
 \subparagraph{Generating interactive user interfaces.}
-Although the pivot type provider simplifies code needed for data exploration, not every reader will 
-be able to write or modify it. The simplicity of the host language makes it possible to easily 
-generate user interface that allows changing of some of the parameters of the program. 
+Although the pivot type provider simplifies code needed for data exploration, not everyone will 
+be able to write or modify source code. The simplicity of the host language makes it possible to  
+automatically generate user interface that allows changing of some of the parameters of the program. 
 Figure~\ref{fig:case-opts} shows an example for the above code snippet that we implemented as 
@@
 chain of member accesses -- we annotate members with annotations denoting whether a member is 
-start of a list, list item or end of a list. The editor then looks for parts of the chain of the
+start of a list, list item or an end of a list. The editor then looks for parts of the chain of the
 form $\qident{list start}.\qident{list item 1}.\qident{list item 2}.\qident{list end}$ and generates
-a control that lets the user remove or add list items. An item cannot be removed if the operation
-caused the code to break (e.g.~when it adds a member that is needed later) and items to be added
-are chosen using available members (as in the standard auto-complete). The header
+a component that lets the user remove or add list items. An item cannot be removed if the operation
+would break the code (e.g.~when it adds a member that is needed later) and items to be added
+are chosen using available members (as in the standard auto-complete). The headers
 shown in Figure~\ref{fig:case-opts} are provided as additional annotations attached to 
@@
 
-\begin{figure}[t]
-\begin{center}
-\includegraphics[scale=0.31,trim=0mm 0mm 0mm 0mm,clip]{images/pivot.png} % left bottom right top
-\end{center}
-\caption{Auto-completion offering the available values of the athlete name column}
-\label{fig:case-ed}
-\end{figure}
-
-% --------------------------------------------------------------------------------------------------
-
 \subparagraph{Spreadsheet-inspired live editor.}
 The third editor extension that we developed for the pivot type provider aims to bridge the gap 
-between code and user interface. This is done through a direct manipulation interface \cite{directman} 
+between code and user interfaces. This is done through a direct manipulation editor \cite{directman} 
 inspired by spreadsheet applications. When exploring data in a spreadsheet, the user can always see the 
-data they work with and the results of an action will be immediately visible. For queries written
-using the pivot type provider, the results can also be calculated immediately using the sample 
-data set provided when using the refined version of the type provider as discussed in 
-Section~\ref{sec:pivot-filter}.
+data they work with and the results of an action will be immediately visible. When exploring data
+using the pivot type provider, the intermediate results can be calculated immediately using the 
+sample data set provided when instantiating the refined version of the type provider with filtering
+support (Section~\ref{sec:pivot-filter}).
 
@@
 Note that the selected part of code is the \qident{by Gold descending} identifier and so the 
-preview shows data available at that point of the query evaluation. Athletes with largest number
-of gold medals appear first, but silver or bronze medals are not used as secondary sorting keys yet.
+preview shows results as computed at that point of the query evaluation. Athletes with largest number
+of gold medals appear first, but silver or bronze medals are not yed used as secondary sorting keys.
 As the user moves through the code, or writes the code, the live preview is updated accordingly.
-Finally, the editor also makes it possible to modify the code through the user interface -- the
+
+Finally, the editor also makes it possible to modify the code through the user interface. The
 ``x'' buttons can be used to remove sort keys or transformations and ``+'' buttons (on the right)
@@
 Unlike the user interface for modifying lists, the live editor works specifically with the pivot 
-type provider. It still relies on the simple structure provided by the fact that entire 
+type provider. However, it still relies on the simple structure provided by the fact that entire 
 transformation can be written as a single chain of member accesses. In particular,
@@
 
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\begin{center}
+\includegraphics[scale=0.31,trim=0mm 0mm 0mm 0mm,clip]{images/pivot.png} % left bottom right top
+\end{center}
+\caption{Spreadsheet-inspired live editor for the pivot type provider}
+\label{fig:case-ed}
+\end{figure}
+
 % ==================================================================================================
@@
 The pivot type provider makes data exploration safer as it does not allow construction of invalid
-queries. Alternative approach would be to use fancy types \cite{fancytypes}. This can be done
-using using row types and typestate or phantom types \cite{rowtypes,typestate,phanty} as sketched
-in in Section~\ref{sec:columns-row}. The idea of using type providers in this way has also been
+queries. Alternative approach would be to use fancy types, such as those available in Haskell \cite{frames,fancytypes}. 
+The approach sketched in Section~\ref{sec:columns-row} used row types and typestate or phantom types 
+\cite{rowtypes,typestate,phanty}. The idea of using type providers to encode fancy types has also been
 explored for session types \cite{sessionty,sessiontp} and it would be interesting to see whether
@@
 actually transforming data, which makes it more related to systems for query construction
-\cite{spreadsheetalgebra,querydirect}. We also see code as equally important to the direct 
-manipulation interface.
+\cite{spreadsheetalgebra,querydirect}. Our approach is somewhat different in that we see code 
+as equally important to the direct manipulation interface.
 
@@
 \subsection{Conclusions}
-There is lots of data (big data), but using them is hard (post-truth).
-We want to make it easier to explore data without sacrificing reproducibility 
-and other good things we get from having source code.
-
-In this paper, we focus on making the language as simple as possible. We unify many different
-PL concepts using member access as the primary programming mechanism thanks to type providers.
-This makes programming easier through auto-completed, but we also explored other tooling ranging
-from simple interactive UIs to direct manipulation tools.
-
-The pivot type provider provides a safe and easy to use layer over an underlying 
-relational algebra. We formalize the host language and the type provider and we prove
-that queries constructed using the type provider are correct.
-
-TL;DR - better toools to democratize data stuff.
+In this paper, we presented a simple programming language for data exploration. The language 
+addresses two problems with the current tooling for data science. One one hand, spreadsheets are 
+easy to use, but are error-prone and do not lead to reproducible scripts that could be modified
+or checked for correctness. On the other hand, even simple data exploration libraries require 
+the user to understand non-trivial programming concepts and offer only little help when writing 
+data exploration code.
+
+We reduce the number of concepts in the language by making member access (``dot'') the primary
+programming mechanism and we implement type provider for data exploration, which offers 
+available transformations and their parameters as members of a provided type. This leads to a 
+simple language that can be well supported by standard tooling such as auto-completion. 
+We also explore other possibilities for tooling enabled by this model ranging from 
+simple interactive user interfaces to direct manipulation~tools.
+
+The pivot type provider offers a safe and easy to use layer over an underlying relational algebra 
+that we use to model data transformations. As a key technical contribution of this paper, we 
+formalise the type provider and prove that queries constructed using the types it provides are 
+correct. Achieving this property by ohter means would require a language with complex
+type system features such as typestate and row types.
+
+We believe that the simple programming model for data exploration presented in this paper can
+contribute to democratization of data exploration -- you should not need to be an experienced 
+programmer to build a transparent visualization using facts that matter to you!
 
@@
 
+The data set used as an example in the case study discussed in Section~\ref{sec:impl} as well as
+in the examples disussed throughout the paper is a single CSV file listing the entire history of
+Olympic medals awarded since 1896. The data set can be found at \url{https://github.com/the-gamma/workyard}
+together with scripts used to obtain it. The following is a representative example listing the
+first 5 rows:
+%
 {\small
@@
 \textbf{\sffamily {Games},{Year},{Discipline},{Athlete},{Team},{Gender},{Event},{Medal},{Gold},{Silver},{Bronze}}\\
-\ident{Athens (1896)},\num{1896},\ident{Swimming},\ident{Alfred Hajos},\kvd{HUN},\ident{Men},\ident{100m freestyle men},\ident{Gold},\num{1},\num{0},\num{0}\\
-\ident{Athens (1896)},\num{1896},\ident{Swimming},\ident{Otto Herschmann},\kvd{AUT},\ident{Men},\ident{100m freestyle men},\ident{Silver},\num{0},\num{1},\num{0}\\
-\ident{Athens (1896)},\num{1896},\ident{Swimming},\ident{Dimitrios Drivas},\kvd{GRE},\ident{Men},\ident{100m freestyle for sailors men},\ident{Bronze},\num{0},\num{0},\num{1}\\
-\ident{Athens (1896)},\num{1896},\ident{Swimming},\ident{Ioannis Malokinis},\kvd{GRE},\ident{Men},\ident{100m freestyle for sailors men},\ident{Gold},\num{1},\num{0},\num{0}\\
-\ident{Athens (1896)},\num{1896},\ident{Swimming},\ident{Spiridon Chasapis},\kvd{GRE},\ident{Men},\ident{100m freestyle for sailors men},\ident{Silver},\num{0},\num{1},\num{0}\\
+\ident{Athens (1896)},\num{1896},\ident{Swimming},\ident{Alfred Hajos},\kvd{HUN},\ident{Men},\ident{100m freestyle},\ident{Gold},\num{1},\num{0},\num{0}\\
+\ident{Athens (1896)},\num{1896},\ident{Swimming},\ident{Otto Herschmann},\kvd{AUT},\ident{Men},\ident{100m freestyle},\ident{Silver},\num{0},\num{1},\num{0}\\
+\ident{Athens (1896)},\num{1896},\ident{Swimming},\ident{Dimitrios Drivas},\kvd{GRE},\ident{Men},\ident{100m freestyle for sailors},\ident{Bronze},\num{0},\num{0},\num{1}\\
+\ident{Athens (1896)},\num{1896},\ident{Swimming},\ident{Ioannis Malokinis},\kvd{GRE},\ident{Men},\ident{100m freestyle for sailors},\ident{Gold},\num{1},\num{0},\num{0}\\
+\ident{Athens (1896)},\num{1896},\ident{Swimming},\ident{Spiridon Chasapis},\kvd{GRE},\ident{Men},\ident{100m freestyle for sailors},\ident{Silver},\num{0},\num{1},\num{0}\\
 \end{array}
@@
 
-%%
-%% Bibliography
-%%
-
-%% Either use bibtex (recommended),
-
-%% .. or use the thebibliography environment explicitely
+\noindent
+The column names are the same as the column names used to generate the \ident{olympics} value using
+the pivot type provider. The script to generate the file de-normalizes the \ident{Medal} column
+and adds \ident{Gold}, \ident{Silver} and \ident{Bronze} columns which are numerical and can thus
+be easily summed. When loading the data, we also transform country codes such as \kvd{GRE} to
+full country names.
 

### 2017-01-13 pivot/paper.tex
Typos

@@
 (as there are always more operations that can be applied). The implementation relies on the fact 
-that type providers are integrated into the type system and types can be provided lazilly. This is 
+that type providers are integrated into the type system and types can be provided lazily. This is 
 also a new aspect of our formalization in Section~\ref{sec:foo}.
@@
   an expression $\ident{olympics}[\str{Games}]==\str{Rio (2016)}$ returning a vector of Booleans
-  while in the other, we specify a column name using $\ident{by}=\str{Gold}$. In other languges,
+  while in the other, we specify a column name using $\ident{by}=\str{Gold}$. In other languages,
   a parameter can also be a lambda function specifying a predicate or a transformation.
@@
 LISP is perhaps the best example of a language that unifies many distinct constructs using a single
-form. In LISP, everything is an s-experssion, that is, either a list or a symbol. In contrast, 
+form. In LISP, everything is an s-expression, that is, either a list or a symbol. In contrast, 
 a typical data processing language uses a number of distinct constructs including indexers (for 
@@
 for use as sort keys. When filtering \ball{2}, the member \qident{Rio (2016)} is provided based
-on the values in the data source (we discuss this furter in Section~\ref{sec:pivot-filter}).
+on the values in the data source (we discuss this further in Section~\ref{sec:pivot-filter}).
 
@@
 Using descriptive member names is only possible when the names are discoverable. The above
-code could be executed in a dynamically-typed languge that allows custom message-not-understood
+code could be executed in a dynamically-typed language that allows custom message-not-understood
 handlers, but it would be impossible to get the name right when writing it. Our approach 
@@
 
-Source code editors for object-based languages with nominal type sytems often provide 
+Source code editors for object-based languages with nominal type systems often provide 
 auto-completion for members of objects. This combination works extremely well in practice; the
@@
 selects the grouping key and then can choose one or more aggregations that can be applied on
-other columns of the group. Thus an enitre data transformation (such as choosing top 8 athletes
+other columns of the group. Thus an entire data transformation (such as choosing top 8 athletes
 by the number of gold medals) can be constructed using dot-driven development.
@@
 Nesting is an essential programming construct and it may be desirable to support it directly in 
-the language, but the ``then'' pattern lets us express nesting without langauge support. In both 
+the language, but the ``then'' pattern lets us express nesting without language support. In both 
 of the cases above, the nested structure is specified by selecting one or more members and then 
@@
 editor needs to understand the rules and calculate what members may be invoked. This is a distinct
-operation from type checking and type infernece (which operate on complete programs) that needs 
+operation from type checking and type inference (which operate on complete programs) that needs 
 to be formalized and implemented.
@@
    & \lsep & \ident{dist}~f &&\textnormal{Count number of distinct values of the column $f$}\\
-   & \lsep & \ident{conc}~f &&\textnormal{Concatenat string values of the column $f$}\\
+   & \lsep & \ident{conc}~f &&\textnormal{Concatenate string values of the column $f$}\\
 \\
@@
 
-The semantics of the algbra is given in Figure~\ref{fig:foo-relsem}. We use vector-based semantics
+The semantics of the algebra is given in Figure~\ref{fig:foo-relsem}. We use vector-based semantics
 to support sorting and duplicate entries, but otherwise the formalization captures the usual
@@
 then apply aggregations $\rho_1,\ldots,\rho_m$. To do this, we first obtain a set of groups $g_1,\ldots, g_s$
-where each group represents a set of indicies of rows belonging to each group. For a given group 
+where each group represents a set of indices of rows belonging to each group. For a given group 
 $g_i$ we can then obtain values of column $j$ for rows in the group as $\{v_{j,k}\,|\, k\in g_i\}$.
@@
 
-\subparagraph{Reduction rules.} The reduction relation $\leadsto_L$ is prameterized by a function
+\subparagraph{Reduction rules.} The reduction relation $\leadsto_L$ is parameterized by a function
 $L$ that maps class names to class definitions, together with nested classes associated with the
@@
 A type provider is an executable component called by the compiler and the editor to 
-provide information about types on demand. In our formalisation, we follow the 
+provide information about types on demand. In our formalization, we follow the 
 style of Petricek et al. \cite{fsdata}, but we add laziness as discussed in Section~\ref{sec:foo-foo}.
@@
 %
-The class $C$ provided by the pivot tpe provider has a constructor taking \ident{Query}, which 
+The class $C$ provided by the pivot type provider has a constructor taking \ident{Query}, which 
 represents the, possibly already partly transformed, input data set. It generates members that
@@
 
-If we were using the relational algebra formalised in Section~\ref{sec:foo-rel} to construct 
+If we were using the relational algebra formalized in Section~\ref{sec:foo-rel} to construct 
 queries, we can write an invalid program, e.g.~by attempting to select
@@
 
-To formalise this, we prove partial soundness of the Foo calculus (Theorem~\ref{thm:foo-sound}),
-which characterises the invalid programs that can be written using the \ident{Query}-typed 
+To formalize this, we prove partial soundness of the Foo calculus (Theorem~\ref{thm:foo-sound}),
+which characterizes the invalid programs that can be written using the \ident{Query}-typed 
 expressions and then prove safety of the pivot type provider (Theorem~\ref{thm:pivot-safe}),
@@
 calculus where objects can be constructed and their members accessed. It permits recursion as a
-member can invoke itself on a new object instance. To accommodate this, we formalise soundness 
+member can invoke itself on a new object instance. To accommodate this, we formalize soundness 
 using progress (Lemma~\ref{thm:foo-progress}) and preservation (Lemma~\ref{thm:foo-pres}).
@@
 The example given in Section~\ref{sec:intro} obtained top 8 athletes based on the number
-of gold medals from Rio 2016. It used two operations that were omitted in the formalisation in
+of gold medals from Rio 2016. It used two operations that were omitted in the formalization in
 Section~\ref{sec:pivot-core}. We omitted paging to keep the host language simple, but we also
@@
 
-In the revised formalisation, the \ident{pivot} function which models the type provider takes the schema
+In the revised formalization, the \ident{pivot} function which models the type provider takes the schema
 $F$ together with sample data $D$ and provides the type with class context:
@@
 \end{center}
-\caption{Athletes by the number of medals over the enitre history of Olympic games}
+\caption{Athletes by the number of medals over the entire history of Olympic games}
 \label{fig:case-table}
@@
 to enable dot-driven development as discussed in Section~\ref{sec:analysis-auto}. In this section,
-we reivew our experience and outline some of the additional tools (available at 
+we review our experience and outline some of the additional tools (available at 
 \url{github.com/the-gamma}).
@@
 
-The user interface lets the user choose aggregations to be caluclated over a group and select 
+The user interface lets the user choose aggregations to be calculate over a group and select 
 columns used for sorting. It is generated automatically by looking for specific pattern in the 
@@
 preview shows results as computed at that point of the query evaluation. Athletes with largest number
-of gold medals appear first, but silver or bronze medals are not yed used as secondary sorting keys.
+of gold medals appear first, but silver or bronze medals are not yet used as secondary sorting keys.
 As the user moves through the code, or writes the code, the live preview is updated accordingly.
@@
 
-\subparagraph{Human-computer intercation.}
+\subparagraph{Human-computer interaction.}
 We discussed how the pivot type provider simplifies the programming model (Section~\ref{sec:analysis}),
@@
 that we use to model data transformations. As a key technical contribution of this paper, we 
-formalise the type provider and prove that queries constructed using the types it provides are 
-correct. Achieving this property by ohter means would require a language with complex
+formalize the type provider and prove that queries constructed using the types it provides are 
+correct. Achieving this property by other means would require a language with complex
 type system features such as typestate and row types.
@@
 The data set used as an example in the case study discussed in Section~\ref{sec:impl} as well as
-in the examples disussed throughout the paper is a single CSV file listing the entire history of
+in the examples discussed throughout the paper is a single CSV file listing the entire history of
 Olympic medals awarded since 1896. The data set can be found at \url{https://github.com/the-gamma/workyard}

### 2017-01-13 pivot/paper.tex
One more typo

@@
 compose rich and correct queries using just member access (``dot''). This way, we recreate 
-functionality that usually requires complex type systems (row polymporphism, type state and dependent 
+functionality that usually requires complex type systems (row polymorphism, type state and dependent 
 typing) in an extremely simple object-based language.

### 2017-04-12 ecoop-2017/paper.tex
Organize papers

--- 2017-01-13 pivot/paper.tex
@@
 \title{Data exploration through dot-driven development\footnote{This 
-  work was supported by Google Digital News Initiative.}}
+  work was supported by The Alan Turing Institute under the EPSRC grant EP/N510129/1
+  and by the Google Digital News Initiative.}}
 \titlerunning{Data exploration through dot-driven development}

### 2017-04-19 ecoop-2017/artifact.tex
Add artifact source, description and Dockerfile

\documentclass[a4paper,UKenglish]{lipics-v2016}

\usepackage{upquote}
\usepackage{microtype}
\usepackage{pifont}
\usepackage{fp}
\usepackage{graphicx}
\usepackage{semantic}
\usepackage{lipsum}

\bibliographystyle{plainurl}

% ==================================================================================================

\title{Data exploration through dot-driven development (artifacat evaluation)}
\titlerunning{Data exploration through dot-driven development}

\author[1]{Tomas Petricek}
\affil[1]{The Alan Turing Institute, London, UK\\
  \texttt{tomas@tomasp.net}}
\authorrunning{T. Petricek}
\Copyright{Tomas Petricek}
\subjclass{D.3.2 Very high-level languages}
\keywords{Data science, type providers, pivot tables, aggregation}

\EventEditors{John Q. Open and Joan R. Acces}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{DRAFT}
\EventAcronym{DRAFT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}

% ==================================================================================================

\theoremstyle{plain}
\theoremstyle{definition}
\newtheorem{rmrk}{Remark}

% Formatting for source code & types
\definecolor{cmtclr}{rgb}{0.0,0.6,0.0}
\definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
\definecolor{numclr}{rgb}{0.0,0.4,0.0}
\definecolor{strclr}{rgb}{0.4,0.3,0.0}
\definecolor{prepclr}{rgb}{0.6,0.0,0.2}

\newcommand{\vect}[1]{\langl #1 \rangl}
\newcommand{\langl}{\begin{picture}(4.5,7)
\put(1.1,2.5){\rotatebox{60}{\line(1,0){5.5}}}
\put(1.1,2.5){\rotatebox{300}{\line(1,0){5.5}}}
\end{picture}}
\newcommand{\rangl}{\begin{picture}(4.5,7)
\put(.9,2.5){\rotatebox{120}{\line(1,0){5.5}}}
\put(.9,2.5){\rotatebox{240}{\line(1,0){5.5}}}
\end{picture}}

\newcommand{\ball}[1]{\FPeval{\result}{clip(201+#1)}\textnormal{\ding{\result}}}

\newcommand{\lsep}{~\,|\,~}
\newcommand{\num}[1]{\textcolor{numclr}{#1}}
\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
\newcommand{\ident}[1]{\textnormal{\sffamily #1}}
\newcommand{\qident}[1]{\textnormal{\sffamily \guillemotleft #1\guillemotright}}
% \newcommand{\qident}[1]{\textnormal{\sffamily ‹#1›}}

\newcommand{\dom}{\ident{dom}}

% ==================================================================================================

\begin{document}
\maketitle

% ==================================================================================================

~

~

\section{Installation instructions}

The tool implemented as part of the submitted paper is available as a JavaScript library that
parses, type-checks and runs sample scripts and creates an editor that can be used for writing
code in the browser. The library calls a server to evaluate queries constructed using the pivot
type provider.

The presented artifact consists of a simple web server packaged as a Docker container. When 
started, the web server hosts the service that provides data to the web browser, as well as
a sample web page with a number of demos illustrating the language.

To run the artifact, you will need to install Docker (\url{https://www.docker.com}) and the
Google Chrome browser (\url{http://www.google.co.uk/chrome}) and also download the provided
docker image \texttt{thegamma-ecoop17.tar}. Next, use the \texttt{docker load} command to 
unpack the image and \texttt{docker run} to start the web server:

\begin{verbatim}
    docker load --input thegamma-ecoop17.tar
    docker run -p 8889:80 thegamma-ecoop17
\end{verbatim}

\noindent
Alternatively, the \texttt{thegamma-ecoop17} Docker image is also hosted on the Docker repository
and can be downloaded and executed directly from the Docker repository:

\begin{verbatim}
    docker run -p 80:8889 tomasp/thegamma-ecoop17
\end{verbatim}

\noindent
The above commands specify the \texttt{-p 8889:80} parameter, which instructs Docker to map the
server hosted at port 80 inside the Docker container to a port 8889 of the host machine. Once
the Docker container starts, you will be able to access the sample web site by opening 
\url{http://localhost:8889} in the Google Chrome browser.

\begin{figure}[t]
\begin{center}
\includegraphics[scale=0.35,trim=0mm 0mm 0mm 0mm,clip]{images/af-preview.png}
\end{center}
\caption{Visualization showing top medalist from Rio 2016 with source code}
\label{fig:chart}
\end{figure}

\section{Exploring the examples}

The sample page contains a number of examples that illustrate the pivot type provider. When the
page loads, the examples are executed and the results appear as illustrated in Figure~\ref{fig:chart}.
For each example, the web page shows the full source code using the pivot type provider together
with several additional libraries not presented in the paper (such as \texttt{chart} and \texttt{table}),
which are used for creating visualizations.

Clicking on the ``Edit source code'' button opens an editor where the code can be modified. The
editor supports auto-completion, which is the key feature made possible by the type system of the
language and the pivot type provider. The following three sections briefly describe three scenarios
that illustrate interesting aspects of the artifact.

\begin{figure}
\begin{center}
\includegraphics[scale=0.35,trim=0mm 0mm 0mm 0mm,clip]{images/af-editor.png}
\end{center}
\caption{Visualization showing top medalist from Rio 2016 with source code.}
\label{fig:code}
\end{figure}

~

\subsection{Demo: Obtain top medallists from London 2012}

In the first demo, we make a simple change to an existing source code to create a bar chart
showing top medallists from London 2012 rather than top medallists from Rio 2016.

\begin{enumerate}
\item Go to the first visualization ``Top medalists from Rio 2016'' and click the ``Edit source code''
  button to load the source code in the editor.
\item Locate the \texttt{Rio (2016)} member on the third line and delete it in the text editor.
  Additionally, delete the ``dot'' after \textt{Games is}.
\item Type ``.'' immediately after the quoted \texttt{Games is} member. An auto-complete list
  should appear showing the available Olympic games together with their years.
  Choose \texttt{London (2012)} and make sure the identifier is separated from the following
  \texttt{then} member with another ``dot''.
\item Click ``Update page''. The visualization on the original demo page should change to show
  the new chart.
\end{enumerate}

\subsection{Demo: Obtain Mongolian gold medalists in code}

This demo illustrates the experience of writing code using the pivot type provider from scratch.
We start with an empty code editor, start writing code and use auto-complete to create a simple
query.

\begin{enumerate}
\item Click on the ``Edit source code'' button for any of the examples and delete all source code
  in the editor, so that we can start from empty file.

\item Type \texttt{olympics.} and note that auto-complete offers different data transformations 
  as documented in the paper. Start by choosing \textt{filter data}, followed by 
  \textt{Team is} and \texttt{Mongolia}. Finally, complete the filtering by choosing \texttt{then}.
  Typing ``dot'' again offers the list of all available transformations.

\item Complete the query by typing ``dot'' and selecting items from auto-complete. When auto-complete
  is displayed, you can navigate using arrow keys and filter items by typing substrings of the
  member names. As you type, you can see the data calculated so far in the preview below the
  current line.

\begin{verbatim}
  olympics.'filter data'.'Team is'.Mongolia.then
    .'filter data'.'Medal is'.Gold.then.'get the data'
\end{verbatim}

\item Now that the data query is written, complete the sample by creating a table to show the results.
  Wrap the code in the following (note that the code written in previous step needs to be indented):

\begin{verbatim}
  let altan = (...)
  table.create(altan)   
\end{verbatim}

\item When you place the cursor inside the \texttt{table.create} call, you should see table as a
  preview. Click ``Update page'' and you should see the new table appear in the original demo page.
\end{enumerate}

\subsection{Demo: Count medals per discipline using the pivot user interface}

This demo introduces the user interface built on top of the pivot type provider as mentioned in the
case study presented in the paper. We will use the user interface to construct a simple query that
finds the number of medals awarded per discipline and finds the disciplines with the largest number
of medals. To do this, we make small edits to the code and complete the rest of the work using
a simple user interface.

\begin{enumerate}
\item Go to the ``All time medalists table'' demo and click ``Edit source code''.
\item In the displayed source code, click anywhere inside the query that performs the data aggregatoion
  so that the user interface displayed in Figure~\ref{fig:code} appears.
\item Remove all the operations except for \texttt{group by} by clicking on the ``x'' button on 
  the right. Start from the last one and remove \texttt{get the data}, \texttt{paging},
  \texttt{get the data} and \texttt{sort by}. Now, remove all the calculated aggregations 
  by clicking on ``x'' buttons in the yellow toolbar. 
\item Change the grouping key by selecting \texttt{by Discipline} in the drop down
  showing \texttt{by Athlete}. Next, add aggregation \texttt{count all} by clicking on the 
  ``+'' button in the yellow toolbar.
\item Add \texttt{sort by} transformation by clicking on the yellow ``+'' in the tab panel showing
  a list of operations. Then add sorting key \texttt{by count descending} by clicking on the ``+''
  in the yellow toolbar.    
\item Finally, add the \texttt{get the data} transformation to obtain the data. The final source
  code should look as follows:
\begin{verbatim}
  let data =
    olympics
      .'group data'.'by Discipline'.'count all'.then
      .'sort data'.'by count descending'.then
      .'get the data'
  table.create(data)  
\end{verbatim}
\item Finally, click ``Update page'' and review the newly produced table in the main page. You should
  see athletics at the top with 3856 medals and Jeu de Paume at the bottom with just 3 medals.
\end{enumerate}

\end{document}

### 2017-04-19 ecoop-2017-artifact/web/lib/thegamma/embed.js

@@ -1,0 +1,94 @@
+if (!thegammaInit) { var thegammaInit = false; }
+
+function openDialog(id) {
+  document.getElementById("thegamma-" + id + "-dialog").style.display="block";
+  setTimeout(function() { 
+    document.getElementById("thegamma-" + id + "-dialog").style.opacity=1;
+    document.getElementById("thegamma-" + id + "-dialog-window").style.top="0px";
+  },1);
+}
+function closeDialog(id) {
+  document.getElementById("thegamma-" + id + "-dialog").style.opacity=0;
+  document.getElementById("thegamma-" + id + "-dialog-window").style.top="-500px";
+  setTimeout(function() { 
+    document.getElementById("thegamma-" + id + "-dialog").style.display="none";
+  },400)
+}
+
+function loadTheGamma() {
+  require.config({
+    paths:{'vs':'https://thegamma.net/lib/thegamma/vs'},
+    map:{ "*":{"monaco":"vs/editor/editor.main"}}
+  });
+  require(["monaco", "https://thegamma.net/lib/thegamma/thegamma.js"], function (_, g) {      
+    thegamma.forEach(function (id) {
+      var services = "https://thegamma-services.azurewebsites.net/";      
+      var providers = 
+        g.providers.createProviders({ 
+          "worldbank": g.providers.rest(services + "worldbank"),
+          "libraries": g.providers.library("https://thegamma.net/lib/thegamma/libraries.json"),
+          "olympics": g.providers.pivot(services + "pdata/olympics") });
+        
+      // Create context and setup error handler
+      var ctx = g.gamma.createContext(providers);
+      ctx.errorsReported(function (errs) { 
+        var lis = errs.slice(0, 5).map(function (e) { 
+          return "<li><span class='err'>error " + e.number + "</span>" +
+            "<span class='loc'>at line " + e.startLine + " col " + e.startColumn + "</span>: " +
+            e.message;
+        });        
+        var ul = "<ul>" + lis + "</ul>";
+        document.getElementById("thegamma-" + id + "-errors").innerHTML = ul;
+      });
+      
+      // Get and run default code, setup update handler
+      var editor;
+      var code = document.getElementById(id + "-code").innerHTML;
+      ctx.evaluate(code, "thegamma-" + id + "-out");
+      document.getElementById("thegamma-" + id + "-update").onclick = function() {
[…]

### 2017-05-05 programming-2017-demo/final/abstract.tex
ACM version of Programming abstract

\documentclass[sigconf,english]{acmart}
\usepackage{demo}
\begin{document}

\setcopyright{acmlicensed}

\copyrightyear{2017} 
\acmYear{2017} 
\setcopyright{rightsretained} 
\acmConference{Programming '17}{April 03-06, 2017}{Brussels, Belgium}
\acmDOI{http://dx.doi.org/10.1145/3079368.3079382}
\acmISBN{978-1-4503-4836-2/17/04}

\title{Tools for open, transparent and engaging storytelling}

\author{Tomas Petricek}
\affiliation{%
  \institution{The Alan Turing Institute}
  \city{London}
  \country{UK}}
\email{tomas@tomasp.net}

\date{5 May 2017}

\begin{abstract}
The rise of Big Data and Open Government Data initiatives means that there is an increasing amount
of raw data about the world available. At the same time, ``post-truth'' has been chosen as the word 
of 2016 \cite{bbc} and the general public increasingly distrusts statistics \cite{guardian}. In 
other words, data science has more capabilities to help us understand the world than ever before, 
yet it is becoming less relevant in public discussion.

This should perhaps not be a surprise as data science is often opaque, non-experts find results 
difficult to interpret and verify, and creating data-driven reports requires advanced skills and 
is limited to a small number of specialists.

The purpose of the proposed demo is to present The Gamma project (\url{http://thegamma.net}, \cite{dot}) 
which aims to democratize data science. The Gamma encourages everyone -- including journalists and 
interested citizens -- to understand how presented claims are justified, explore data on their own 
and make their own transparent factual claims. If the society is to benefit from the possibilities 
available through data science, it is essential to make data-driven storytelling widely accessible, 
open and engaging.
\end{abstract}

\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10003120.10003145.10003151.10011771</concept_id>
<concept_desc>Human-centered computing~Visualization toolkits</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10011007.10011006.10011008.10011009.10011022</concept_id>
<concept_desc>Software and its engineering~Very high level languages</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}
\ccsdesc[500]{Human-centered computing~Visualization toolkits}
\ccsdesc[300]{Software and its engineering~Very high level languages}

% Keywords: PLEASE REPLACE
\keywords{Programming experience; example articles}

\maketitle
\section{Open data-driven storytelling}
On one hand, spreadsheets made data exploration accessible to a large number of people, but 
operations performed on spreadsheets are error-prone and cannot be easily reproduced or 
replicated with different data source. On the other hand, data analyses written as programs
can be modified and run repeatedly, but even with the simplest programming tools available, 
building an end-to-end analysis that reads data from a government data source, performs analysis 
and produces an interactive visualization requires expert programming and data science skills.

The Gamma aims to build programming tools that let anyone explore data from a wide range of data 
sources, including open government data, and publish data-driven reports that are:

\begin{itemize}
\item	\textbf{Transparent and accountable}. Readers can review how data is used and discover misleading uses of data.
\item	\textbf{Reproducible and connected}. Readers can run the analysis themselves using the original data source.
\item	\textbf{Open and engaging}. Readers can modify parameters and share reports on different aspects of the data.
\end{itemize}

\noindent
To achieve this, we treat data-driven reports as reproducible programs written in a simple web-based 
scripting language that is integrated with primary data sources (using type providers \cite{inforich,fsdata,idris}) 
and we develop editor tooling that bridges the gap between program-ming and spreadsheets.

\begin{figure*}
\begin{center}
\includegraphics[scale=0.50,trim=0mm 60mm 0mm 0mm,clip]{fig1.png}
\end{center}
\vspace{-2em}
\caption{\textnormal{\small The Gamma project uses a type provider to generate types with members based on the structure 
of the processed data. When writing script to work with data, the auto-completion offers help based 
on the types. In the scripting language behind The Gamma, almost all data processing work is done by 
typing ``.'' and choosing one of the available members, leading to an extremely simple programming model 
that can be well supported by editor tooling.}}
\label{fig1}
\end{figure*}

\begin{figure*}
\begin{center}
\includegraphics[scale=0.2,trim=0mm 0mm 0mm 0mm,clip]{fig2.png}
\end{center}
\caption{\textnormal{\small When writing data transformations, users can directly edit code that represents the data 
  transformation, but they can also see the preview of the result of the data transformation written
  so far and edit the data transformation in a spreadsheet-inspired manner. Here, the user decided to 
  aggregate data by team (country) and is now choosing aggregations to perform over the group.}}
\label{fig2}
\end{figure*}

\section{Demo outline}
The work presented in the demo session has been focused on building a simple web-based 
library that could be used by data journalists to present visualizations obtained by aggregating and 
summarizing tabular data. Figure~\ref{fig1} and Figure~\ref{fig2} illustrate some of the steps 
performed during a sample task -- given a data table recording individual medals awarded over the 
entire history of the Olympic games, we want to calculate the number of medals per country.

\section{Innovative aspects of the project}
The project is innovative in two ways. It creates a new simple scripting language for working 
with data and it complements the language with spreadsheet-inspired tooling:

\begin{itemize}
\item \textbf{Simple data-aware language}. When writing code, the programming language understands 
  the data source and data transformations performed so far and offers all available operations when 
  ``.'' is typed. For example, when ``Games is'' is typed in Figure~\ref{fig1}, the editor understands 
  what values are available and offers the user ``Rio (2016)'' in the completion list. This means 
  that the user can construct the whole program just by choosing one of the available operations.

\item \textbf{Spreadsheet-inspired editing}. One of the reasons why spreadsheets are easy to use is 
  that the user can always see the data they are working with and manipulate it directly. We adapt 
  this paradigm to programming, taking inspiration from direct-manipulation user interfaces
  \cite{spreadsheetalgebra,querydirect,directman}. In our live editor, the user can always see preview of the 
  aggregation constructed so far, making data exploration easier. As demonstrated in Figure~\ref{fig2}, 
  many transformations can be created using the user interface without writing code directly. 
  Yet, the final result is still an open and reproducible script.
\end{itemize}

\noindent
These two innovations make it possible to create web-based data-driven reports that are transparent 
(anyone can see how they are created), open (readers can modify them and share their results) and 
engaging (reader can explore other fun aspects of the data).

The Gamma project is the first step of an increasingly important research that aims to democratize 
data science and encourage every citizen to make factual claims backed by data -- be it for fun or to hold the government accountable.

\begin{acks}
This work was supported by The Alan Turing Institute under the EPSRC grant EP/N510129/1
and by a the Google Digital News Initiative Prototype Fund. 
\end{acks}

\newpage
\nocite{*}

\bibliographystyle{ACM-Reference-Format}
\bibliography{sigproc}
\end{document}

### 2017-05-16 ecoop-2017/artifact.tex
Update artifact

@@
-\documentclass[a4paper,UKenglish]{lipics-v2016}
+\documentclass[a4paper,UKenglish]{darts}
+\usepackage{microtype}
+\bibliographystyle{plainurl}
+\input{artifactcmds}
 
-\usepackage{upquote}
-\usepackage{microtype}
-\usepackage{pifont}
-\usepackage{fp}
-\usepackage{graphicx}
-\usepackage{semantic}
-\usepackage{lipsum}
-
-\bibliographystyle{plainurl}
-
-% ==================================================================================================
-
-\title{Data exploration through dot-driven development (artifacat evaluation)}
-\titlerunning{Data exploration through dot-driven development}
+\title{Data exploration through dot-driven development (Artifact)\footnote{This artifact is a companion of the paper:  Tomas Petricek, ``Data exploration through dot-driven development'', Proceedings of the 31st European Conference on Object-Oriented Programming (ECOOP 2017), June 18-23, 2017, Barcelona, Spain.
+This work was supported in part by The Alan Turing Institute under the EPSRC grant EP/N510129/1 and by the Google Digital News Initiative.}}
+\titlerunning{Data exploration through dot-driven development (Artifact)}
 
@@
 \affil[1]{The Alan Turing Institute, London, UK\\
+ and Microsoft Research, Cambridge, UK\\
   \texttt{tomas@tomasp.net}}
@@
 \Copyright{Tomas Petricek}
+
 \subjclass{D.3.2 Very high-level languages}
@@
 
-\EventEditors{John Q. Open and Joan R. Acces}
-\EventNoEds{2}
-\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-\EventShortTitle{DRAFT}
-\EventAcronym{DRAFT}
-\EventYear{2016}
-\EventDate{December 24--27, 2016}
-\EventLocation{Little Whinging, United Kingdom}
-\EventLogo{}
-\SeriesVolume{42}
-\ArticleNo{23}
-
-% ==================================================================================================
-
-\theoremstyle{plain}
-\theoremstyle{definition}
-\newtheorem{rmrk}{Remark}
-
-% Formatting for source code & types
-\definecolor{cmtclr}{rgb}{0.0,0.6,0.0}
-\definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
-\definecolor{numclr}{rgb}{0.0,0.4,0.0}
-\definecolor{strclr}{rgb}{0.4,0.3,0.0}
-\definecolor{prepclr}{rgb}{0.6,0.0,0.2}
-
-\newcommand{\vect}[1]{\langl #1 \rangl}
-\newcommand{\langl}{\begin{picture}(4.5,7)
-\put(1.1,2.5){\rotatebox{60}{\line(1,0){5.5}}}
-\put(1.1,2.5){\rotatebox{300}{\line(1,0){5.5}}}
-\end{picture}}
-\newcommand{\rangl}{\begin{picture}(4.5,7)
-\put(.9,2.5){\rotatebox{120}{\line(1,0){5.5}}}
-\put(.9,2.5){\rotatebox{240}{\line(1,0){5.5}}}
-\end{picture}}
-
-\newcommand{\ball}[1]{\FPeval{\result}{clip(201+#1)}\textnormal{\ding{\result}}}
-
-\newcommand{\lsep}{~\,|\,~}
-\newcommand{\num}[1]{\textcolor{numclr}{#1}}
-\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
-\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
-\newcommand{\ident}[1]{\textnormal{\sffamily #1}}
-\newcommand{\qident}[1]{\textnormal{\sffamily \guillemotleft #1\guillemotright}}
-% \newcommand{\qident}[1]{\textnormal{\sffamily ‹#1›}}
-
-\newcommand{\dom}{\ident{dom}}
-
-% ==================================================================================================
+%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\Volume{3}
+\Issue{2}
+\Article{1}
+\RelatedConference{European Conference on Object-Oriented Programming (ECOOP 2017), June 18-23, 2017, Barcelona, Spain}
+% Editor-only macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
 \begin{document}
+
 \maketitle
 
-% ==================================================================================================
+\begin{abstract}
+This artifact presents The Gamma, a simple programming environment for data exploration
+that uses member access as the primary mechanism for constructing queries. The artifact 
+consists of two parts. The user facing web-based component allows users to explore a simple
+dataset of Olympic medal winners while a back-end service provides the data and evaluates
+queries executed by the user.
 
-~
+The purpose of the artifact is to illustrate the pivot type provider, which provides a simple
+way for constructing queries in a object-based programming language equipped with member 
+access. As illustrated by a number of walkthroughs, the pivot type provider can be use to
+construct new queries from code or using the user interface, but it also encourages the user
+to modify existing code.  
+\end{abstract}
 
-~
+\begin{scope}
+The artifact is designed to provide a repeatble way of evaluating the user experience that
+is provided by the pivot type provider. As discussed in the paper, the purpose of the project
+is to simplify data exploration (by using a simple programming language), make it open and 
+transparent (by revealing the source code) and encourage further data exploration (by making
+it easy to change the source code). The presented artifact supports this by providing a number
+of guided walkthroughs that illustrate the above features.
 
-\section{Installation instructions}
-
-The tool implemented as part of the submitted paper is available as a JavaScript library that
-parses, type-checks and runs sample scripts and creates an editor that can be used for writing
-code in the browser. The library calls a server to evaluate queries constructed using the pivot
-type provider.
-
-The presented artifact consists of a simple web server packaged as a Docker container. When 
-started, the web server hosts the service that provides data to the web browser, as well as
-a sample web page with a number of demos illustrating the language.
-
-To run the artifact, you will need to install Docker (\url{https://www.docker.com}) and the
-Google Chrome browser (\url{http://www.google.co.uk/chrome}) and also download the provided
-docker image \texttt{thegamma-ecoop17.tar}. Next, use the \texttt{docker load} command to 
-unpack the image and \texttt{docker run} to start the web server:
-
-\begin{verbatim}
-    docker load --input thegamma-ecoop17.tar
-    docker run -p 8889:80 thegamma-ecoop17
-\end{verbatim}
-
-\noindent
-Alternatively, the \texttt{thegamma-ecoop17} Docker image is also hosted on the Docker repository
-and can be downloaded and executed directly from the Docker repository:
-
-\begin{verbatim}
-    docker run -p 80:8889 tomasp/thegamma-ecoop17
-\end{verbatim}
-
-\noindent
-The above commands specify the \texttt{-p 8889:80} parameter, which instructs Docker to map the
-server hosted at port 80 inside the Docker container to a port 8889 of the host machine. Once
-the Docker container starts, you will be able to access the sample web site by opening 
-\url{http://localhost:8889} in the Google Chrome browser.
-
-\begin{figure}[t]
-\begin{center}
-\includegraphics[scale=0.35,trim=0mm 0mm 0mm 0mm,clip]{images/af-preview.png}
-\end{center}
-\caption{Visualization showing top medalist from Rio 2016 with source code}
-\label{fig:chart}
-\end{figure}
-
-\section{Exploring the examples}
-
-The sample page contains a number of examples that illustrate the pivot type provider. When the
+The artifact web page contains a number of examples that illustrate the pivot type provider. When the
 page loads, the examples are executed and the results appear as illustrated in Figure~\ref{fig:chart}.
@@
 
-\begin{figure}
+\begin{figure}[t]
 \begin{center}
-\includegraphics[scale=0.35,trim=0mm 0mm 0mm 0mm,clip]{images/af-editor.png}
+\includegraphics[scale=0.35,trim=0mm 0mm 0mm 0mm,clip]{images/af-preview.png}
 \end{center}
-\caption{Visualization showing top medalist from Rio 2016 with source code.}
-\label{fig:code}
+\caption{Visualization showing top medalist from Rio 2016 with source code}
+\label{fig:chart}
 \end{figure}
 
-~
-
-\subsection{Demo: Obtain top medallists from London 2012}
+\subsection*{Walkthrough 1: Obtain top medallists from London 2012}
 
@@
 \item Locate the \texttt{Rio (2016)} member on the third line and delete it in the text editor.
-  Additionally, delete the ``dot'' after \textt{Games is}.
+  Additionally, delete the ``dot'' after \texttt{Games is}.
 \item Type ``.'' immediately after the quoted \texttt{Games is} member. An auto-complete list
@@
 
-\subsection{Demo: Obtain Mongolian gold medalists in code}
+\subsection*{Walkthrough 2: Obtain Mongolian gold medalists in code}
 
@@
 \item Type \texttt{olympics.} and note that auto-complete offers different data transformations 
-  as documented in the paper. Start by choosing \textt{filter data}, followed by 
-  \textt{Team is} and \texttt{Mongolia}. Finally, complete the filtering by choosing \texttt{then}.
+  as documented in the paper. Start by choosing \texttt{filter data}, followed by 
+  \texttt{Team is} and \texttt{Mongolia}. Finally, complete the filtering by choosing \texttt{then}.
   Typing ``dot'' again offers the list of all available transformations.
@@
 
-\subsection{Demo: Count medals per discipline using the pivot user interface}
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.35,trim=0mm 0mm 0mm 0mm,clip]{images/af-editor.png}
+\end{center}
+\caption{Visualization showing top medalist from Rio 2016 with source code.}
+\label{fig:code}
+\end{figure}
+
+\subsection*{Walkthrough 3: Count medals per discipline using the pivot user interface}
 
@@
 \end{enumerate}
+\end{scope}
+
+\begin{content}
+The artifact package includes:
+\begin{itemize}
+\item a Docker image ({\tt thegamma-ecoop17.tar}), which contains the necessary client-side
+  and server-side components for running the artifact
+\item an F\# source code ({\tt thegamma-source.zip}) which contains the full source code for 
+  the client-side and server-side components behind the artifact
+\end{itemize}
+To make evaluating of the user interface easier, we provide a Docker container that
+can be used to host the system. The Docker container is based on the standard {\tt fsharp} 
+image supplied by the F\# Software Foundation.
+
+The included source code provides complete source code
+for the server-side components (running as F\# code on the server) and client-side components
+(translated to JavaScript using the Fable compiler). When compiling code from scratch, several
+packages are downloaded prior to the compilation from standard .NET and JavaScript package
+repositories NuGet and npm.
+\end{content}
+
+\begin{getting}
+
+\subsection*{Running the artifact}
+The presented artifact consists of a simple web server packaged as a Docker container. When 
+started, the web server hosts the service that provides data to the web browser, as well as
+a sample web page with a number of demos illustrating the language.
+
+Running the artifact requires Docker (\url{https://www.docker.com}) and the Google Chrome 
+browser (\url{http://www.google.co.uk/chrome}) and also download the provided
+docker image \texttt{thegamma-ecoop17.tar}. To run the artifact, use the \texttt{docker load} command to 
+unpack the image and \texttt{docker run} to start the web server:
+
+\begin{verbatim}
+    docker load --input thegamma-ecoop17.tar
+    docker run -p 8889:80 thegamma-ecoop17
+\end{verbatim}
+
+\noindent
+Alternatively, the \texttt{thegamma-ecoop17} Docker image is also hosted on the Docker repository
+and can be downloaded and executed directly from the Docker repository:
+
+\begin{verbatim}
+    docker run -p 8889:80 tomasp/thegamma-ecoop17
+\end{verbatim}
+
+\noindent
+The above commands specify the \texttt{-p 8889:80} parameter, which instructs Docker to map the
+server hosted at port 80 inside the Docker container to a port 8889 of the host machine. Once
+the Docker container starts, you will be able to access the sample web site by opening 
+\url{http://localhost:8889} in the Google Chrome browser.
+
+\subsection*{Latest version}  
+The tool implemented as part of the submitted paper is also available as a JavaScript library
+that is made available through the standard JavaScript package manager npm. The full
+project soruce code is also hosted on GitHub. The following links can be used to obtain the latest
+version:
+\begin{itemize}
+  \item \url{https://github.com/the-gamma/thegamma-script}
+  \item \url{https://npmjs.com/package/thegamma-script}
+\end{itemize}
+
+\end{getting}
+
+\begin{platforms}
+  The artifact is known to work on any platform running Docker version 17. It has been 
+  specifically tested using Docker version 17.03.1-ce-win12 on the Windows operating system.
+  5~GB of free space on disk and at least 2~GB of free space in RAM are sufficient for running
+  the artifact.
+\end{platforms}
+
+\license{MIT ({\url https://opensource.org/licenses/MIT})}
+
+% ARTIFACT: section specifying the md5 sum of the artifact master file
+% uploaded to the Dagstuhl Research Online Publication Server, enabling 
+% downloaders to check that the file is the expected version and suffered 
+% no damage during download.
+\mdsum{0e0e2873157bb6d69cbff5822523e41a}
+
+% ARTIFACT: section specifying the size of the artifact master file uploaded
+% to the Dagstuhl Research Online Publication Server
+\artifactsize{335 MB}
+
+% ARTIFACT: optional appendix
+%\appendix
+
+%\section{My Appendix}
+
+% Add here any further material you would like to include. For instance, if the artifact is itself a PDF document, add it here.
+
+% ARTIFACT: include here any additional references, if needed...
+
+%% Either use bibtex (recommended), but commented out in this sample
+
+%\bibliography{dummybib}
+
+%% .. or use the thebibliography environment explicitely
 

### 2017-05-16 ecoop-2017/artifact.tex
Final artifact version

@@
 way for constructing queries in a object-based programming language equipped with member 
-access. As illustrated by a number of walkthroughs, the pivot type provider can be use to
+access. The pivot type provider can be use to
 construct new queries from code or using the user interface, but it also encourages the user
@@
 transparent (by revealing the source code) and encourage further data exploration (by making
-it easy to change the source code). The presented artifact supports this by providing a number
-of guided walkthroughs that illustrate the above features.
+it easy to change the source code). The presented artifact supports this claim.
 
 The artifact web page contains a number of examples that illustrate the pivot type provider. When the
-page loads, the examples are executed and the results appear as illustrated in Figure~\ref{fig:chart}.
+page loads, the examples are executed and the results appear.
 For each example, the web page shows the full source code using the pivot type provider together
@@
 that illustrate interesting aspects of the artifact.
-
-\begin{figure}[t]
-\begin{center}
-\includegraphics[scale=0.35,trim=0mm 0mm 0mm 0mm,clip]{images/af-preview.png}
-\end{center}
-\caption{Visualization showing top medalist from Rio 2016 with source code}
-\label{fig:chart}
-\end{figure}
-
-\subsection*{Walkthrough 1: Obtain top medallists from London 2012}
-
-In the first demo, we make a simple change to an existing source code to create a bar chart
-showing top medallists from London 2012 rather than top medallists from Rio 2016.
-
-\begin{enumerate}
-\item Go to the first visualization ``Top medalists from Rio 2016'' and click the ``Edit source code''
-  button to load the source code in the editor.
-\item Locate the \texttt{Rio (2016)} member on the third line and delete it in the text editor.
-  Additionally, delete the ``dot'' after \texttt{Games is}.
-\item Type ``.'' immediately after the quoted \texttt{Games is} member. An auto-complete list
-  should appear showing the available Olympic games together with their years.
-  Choose \texttt{London (2012)} and make sure the identifier is separated from the following
-  \texttt{then} member with another ``dot''.
-\item Click ``Update page''. The visualization on the original demo page should change to show
-  the new chart.
-\end{enumerate}
-
-\subsection*{Walkthrough 2: Obtain Mongolian gold medalists in code}
-
-This demo illustrates the experience of writing code using the pivot type provider from scratch.
-We start with an empty code editor, start writing code and use auto-complete to create a simple
-query.
-
-\begin{enumerate}
-\item Click on the ``Edit source code'' button for any of the examples and delete all source code
-  in the editor, so that we can start from empty file.
-
-\item Type \texttt{olympics.} and note that auto-complete offers different data transformations 
-  as documented in the paper. Start by choosing \texttt{filter data}, followed by 
-  \texttt{Team is} and \texttt{Mongolia}. Finally, complete the filtering by choosing \texttt{then}.
-  Typing ``dot'' again offers the list of all available transformations.
-
-\item Complete the query by typing ``dot'' and selecting items from auto-complete. When auto-complete
-  is displayed, you can navigate using arrow keys and filter items by typing substrings of the
-  member names. As you type, you can see the data calculated so far in the preview below the
-  current line.
-
-\begin{verbatim}
-  olympics.'filter data'.'Team is'.Mongolia.then
-    .'filter data'.'Medal is'.Gold.then.'get the data'
-\end{verbatim}
-
-\item Now that the data query is written, complete the sample by creating a table to show the results.
-  Wrap the code in the following (note that the code written in previous step needs to be indented):
-
-\begin{verbatim}
-  let altan = (...)
-  table.create(altan)   
-\end{verbatim}
-
-\item When you place the cursor inside the \texttt{table.create} call, you should see table as a
-  preview. Click ``Update page'' and you should see the new table appear in the original demo page.
-\end{enumerate}
-
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.35,trim=0mm 0mm 0mm 0mm,clip]{images/af-editor.png}
-\end{center}
-\caption{Visualization showing top medalist from Rio 2016 with source code.}
-\label{fig:code}
-\end{figure}
-
-\subsection*{Walkthrough 3: Count medals per discipline using the pivot user interface}
-
-This demo introduces the user interface built on top of the pivot type provider as mentioned in the
-case study presented in the paper. We will use the user interface to construct a simple query that
-finds the number of medals awarded per discipline and finds the disciplines with the largest number
-of medals. To do this, we make small edits to the code and complete the rest of the work using
-a simple user interface.
-
-\begin{enumerate}
-\item Go to the ``All time medalists table'' demo and click ``Edit source code''.
-\item In the displayed source code, click anywhere inside the query that performs the data aggregatoion
-  so that the user interface displayed in Figure~\ref{fig:code} appears.
-\item Remove all the operations except for \texttt{group by} by clicking on the ``x'' button on 
-  the right. Start from the last one and remove \texttt{get the data}, \texttt{paging},
-  \texttt{get the data} and \texttt{sort by}. Now, remove all the calculated aggregations 
-  by clicking on ``x'' buttons in the yellow toolbar. 
-\item Change the grouping key by selecting \texttt{by Discipline} in the drop down
-  showing \texttt{by Athlete}. Next, add aggregation \texttt{count all} by clicking on the 
-  ``+'' button in the yellow toolbar.
-\item Add \texttt{sort by} transformation by clicking on the yellow ``+'' in the tab panel showing
-  a list of operations. Then add sorting key \texttt{by count descending} by clicking on the ``+''
-  in the yellow toolbar.    
-\item Finally, add the \texttt{get the data} transformation to obtain the data. The final source
-  code should look as follows:
-\begin{verbatim}
-  let data =
-    olympics
-      .'group data'.'by Discipline'.'count all'.then
-      .'sort data'.'by count descending'.then
-      .'get the data'
-  table.create(data)  
-\end{verbatim}
-\item Finally, click ``Update page'' and review the newly produced table in the main page. You should
-  see athletics at the top with 3856 medals and Jeu de Paume at the bottom with just 3 medals.
-\end{enumerate}
 \end{scope}
@@
 
-\subsection*{Running the artifact}
 The presented artifact consists of a simple web server packaged as a Docker container. When
-started, the web server hosts the service that provides data to the web browser, as well as
-a sample web page with a number of demos illustrating the language.
-
-Running the artifact requires Docker (\url{https://www.docker.com}) and the Google Chrome 
-browser (\url{http://www.google.co.uk/chrome}) and also download the provided
-docker image \texttt{thegamma-ecoop17.tar}. To run the artifact, use the \texttt{docker load} command to 
-unpack the image and \texttt{docker run} to start the web server:
-
-\begin{verbatim}
-    docker load --input thegamma-ecoop17.tar
-    docker run -p 8889:80 thegamma-ecoop17
-\end{verbatim}
-
-\noindent
-Alternatively, the \texttt{thegamma-ecoop17} Docker image is also hosted on the Docker repository
-and can be downloaded and executed directly from the Docker repository:
+executed the artifact can be viewed by navigating to \url{http://localhost:8889} as described
+in the artifact documentation. The artifact can be also obtained and started from
+Docker hub using:
 
@@
 \noindent
-The above commands specify the \texttt{-p 8889:80} parameter, which instructs Docker to map the
-server hosted at port 80 inside the Docker container to a port 8889 of the host machine. Once
-the Docker container starts, you will be able to access the sample web site by opening 
-\url{http://localhost:8889} in the Google Chrome browser.
-
-\subsection*{Latest version}  
 The tool implemented as part of the submitted paper is also available as a JavaScript library
@@
 \begin{platforms}
-  The artifact is known to work on any platform running Docker version 17. It has been 
+  The artifact is known to work on any platform running Docker version 17 using the Google Chrome
+  browser. It has been 
   specifically tested using Docker version 17.03.1-ce-win12 on the Windows operating system.

### 2017-05-16 ecoop-2017/paper.tex
Final version edits
Final artifact version
Final version

@@
 
-\EventEditors{John Q. Open and Joan R. Acces}
-\EventNoEds{2}
-\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-\EventShortTitle{DRAFT}
-\EventAcronym{DRAFT}
-\EventYear{2016}
-\EventDate{December 24--27, 2016}
-\EventLocation{Little Whinging, United Kingdom}
+\EventEditors{Peter M\"uller}
+\EventNoEds{1}
+\EventNoEds{1}
+\EventLongTitle{31st European Conference on Object-Oriented Programming (ECOOP 2017)} 
+\EventShortTitle{ECOOP 2017} 
+\EventAcronym{ECOOP} 
+\EventYear{2017} 
+\EventDate{June 18--23, 2017} 
+\EventLocation{Barcelona, Spain} 
 \EventLogo{} 
-\SeriesVolume{42}
-\ArticleNo{23}
+\SeriesVolume{74} 
+\ArticleNo{55}
 
@@
 available. At the same time, the fact that ``post-truth'' was chosen as the word of 2016 \cite{posttruth} 
-suggests that there has never been a greater need for increasing data literacy and building tools that let 
-anyone -- including journalists and interested citizens -- explore such data and use it to make
-transparent factual claims.
+suggests that there has never been a greater need for increasing data literacy and tools that let 
+anyone explore such data and use it to make transparent factual claims.
 
@@
 
-In Figure~\ref{fig:pivot-members}, we list a subset of the provided members that were used in the
+Figure~\ref{fig:pivot-members} lists a subset of the members from the
 example in Section~\ref{sec:intro}. After choosing \qident{sort data}, the user is offered 
-the possible sorting keys and directions. After choosing \qident{group data}, the user first 
+the possible sorting keys. After choosing \qident{group data}, the user first 
 selects the grouping key and then can choose one or more aggregations that can be applied on
@@
 
-Our language supports method calls such as $\ident{head}(\num{8})$ and an alternative design for
-filtering would be to provide a method such as $\qident{Games is}(\str{Rio (2016)})$. However,
-the fact that we can offer possible values largely simplifies writing of the script for the most 
+Our language supports method calls and so some of the opertaions that are currently exposed as
+member access could equally be provided as methods. For example, filtering could be written as 
+$\qident{Games is}(\str{Rio (2016)})$. However, the fact that we can offer possible values when
+filtering largely simplifies writing of the script for the most 
 common case when the user is interested in one of the known values.
@@
 \noindent
-When grouping, we sum the number of gold and silver medals and concatenates team names~\ball{1}. 
-Then we sort the grouped data using multiple sorting keys \ball{2} -- first by the number of gold 
+When grouping, we sum the number of gold and silver medals and concatenates distinct team names~\ball{1}. 
+Then we sort the grouped data using two sorting keys \ball{2} -- first by the number of gold 
 medals and then silver medals (within a group with the same number of gold medals).
@@
 
-In type providers for F\#, this could be done using methods with static parameters \cite{staticpar} by writing, 
+In type providers for F\#, renaming of columns could be encoded using methods with static parameters \cite{staticpar} by writing, 
 for example, $\ident{g}.\qident{sum Gold as}\langl\str{Total Gold}\rangl()$. In F\#, the value of 
@@
 Operations that change the type of rows in the data set can be captured using row types~\cite{rowtypes}. 
+Row types make it possible to statically track operations on records that add or remove fields
+and so they can be used for the typing of operations such as $\qident{drop Year}$.
 In addition, we need to annotate type with a form of typestate \cite{typestate} to restrict what operations are available. 
@@
 \end{equation*}
-\vspace{-1.5em}
+\vspace{-1.85em}
 
@@
 \]
-\vspace{-3em}
+\vspace{-3.15em}
 
@@
 can be used, for example, as input for a charting library. A series is a typed wrapper over a 
-\indet{Query} value and the proofs in Section~\ref{sec:pivot-prop} show that a series obtained 
+\ident{Query} value and the proofs in Section~\ref{sec:pivot-prop} show that a series obtained 
 from the pivot type provider contains keys and values of matching~types.
@@
 obtain full definition of $C_2$ when checking the member $\qident{by Athlete}$. Types of other 
-members of $C_1$ remain unevaluated when they do not appear in the source code. This aspect of type
+members of $C_1$ remain unevaluated. This aspect of type
 providers have been omitted in previous work \cite{fsdata,liteq}, but it is necessary for the pivot type provider.
-The typing rules given are written using the following judgement:
+The typing rules given are written as:
 %
@@
 \begin{figure}[t]
-\vspace{-1.5em}
+\vspace{-0.5em}
 \[
@@
 \]
-\vspace{-2.25em}
+\vspace{-1.75em}
 
@@
 \caption{Type-checking of Foo expressions with lazy context}
-\vspace{-0.5em}
 \label{fig:foo-typecheck}
@@
 \begin{figure}
-\vspace{-1em}
 \begin{equation*}
@@
 \quad \qquad \kvd{member}~\ident{then} : C' = C'(x)\qquad\ball{2}                 &\textnormal{where}~C', L' = \ident{pivot}(F)\\
-\\
+\\[0.5em]
 \ident{sort}(F) = C, \{ C \mapsto (l, \bigcup L_f \cup \bigcup L'_f) \}\qquad\ball{3}\\[0.25em]
@@
 \quad \qquad \kvd{member}~\qident{by $f$ ascending} : C'_f = C'_f(x) &\quad C'_f, L'_f = \ident{sort-and}(F, \vect{f \mapsto \ident{asc}})\\
-\\[0.5em]
+\\[1em]
 \multicolumn{2}{l}{ \ident{sort-and}(F,\vect{s_1,\ldots,s_n}) = C, \{ C \mapsto (l, \bigcup L_f \cup \bigcup L'_f \cup L') \}\qquad\ball{4} }\\[0.25em]
@@
 \quad \qquad \kvd{member}~\qident{by $f$} : C_f = C_f(x) &\quad C_f, L_f = \ident{agg}(F, f, \{ f \mapsto F(f) \}, \emptyset)~~\ball{2}\\
-\\
+\\[0.5em]
 \multicolumn{2}{l}{ \ident{agg}(F,f,G,S) = C, \{C\mapsto (l, \bigcup L_f \cup \bigcup L'_f \cup\bigcup L''_f\cup L'\cup L'') \}\quad\ball{3}}\\[0.25em]
@@
 
+\newpage
+
 % --------------------------------------------------------------------------------------------------
@@
 \begin{figure}[t]
+\vspace{-1em}
 \begin{equation*}
@@
 \begin{equation*}
-\hspace{-1em}
 \begin{array}{l}
@@
 \begin{figure}[b]
+\vspace{1em}
 \begin{center}
-\includegraphics[scale=0.30,trim=0mm 0mm 0mm 0mm,clip]{images/table.png} % left bottom right top
+\includegraphics[scale=0.42,trim=0mm 0mm 0mm 0mm,clip]{images/table2.png} % left bottom right top
 \end{center}
@@
 
-The user interface lets the user choose aggregations to be calculate over a group and select 
-columns used for sorting. It is generated automatically by looking for specific pattern in the 
+The user interface lets the user choose aggregations to be calculated over a group and select 
+columns used for sorting. It is generated automatically by looking for a specific pattern in the 
 chain of member accesses -- we annotate members with annotations denoting whether a member is 
@@
 inspired by spreadsheet applications. When exploring data in a spreadsheet, the user can always see the 
-data they work with and the results of an action will be immediately visible. When exploring data
-using the pivot type provider, the intermediate results can be calculated immediately using the 
-sample data set provided when instantiating the refined version of the type provider with filtering
-support (Section~\ref{sec:pivot-filter}).
+data they work with and the results of an action will be immediately visible. This is not usually 
+the case when writing code in text editor. However, when exploring data using the pivot type 
+provider, the intermediate results can be calculated immediately using the sample data set provided 
+when instantiating the refined version of the type provider with filtering support (Section~\ref{sec:pivot-filter}).
 
@@
 preview shows results as computed at that point of the query evaluation. Athletes with largest number
-of gold medals appear first, but silver or bronze medals are not yet used as secondary sorting keys.
+of gold medals appear first, but silver or bronze medals are not yet used as secondary sorting keys
+and so the secondary ordering is arbitrary.
 As the user moves through the code, or writes the code, the live preview is updated accordingly.
@@
 ``x'' buttons can be used to remove sort keys or transformations and ``+'' buttons (on the right)
-can be used to add more transformations or parameters.
+can be used to add more transformations or to specify additional parameters within the ``then'' 
+pattern. In case of sorting, this allows adding further sorting keys.
 
@@
 
+~
+
 % --------------------------------------------------------------------------------------------------
@@
 \begin{center}
-\includegraphics[scale=0.31,trim=0mm 0mm 0mm 0mm,clip]{images/pivot.png} % left bottom right top
+\includegraphics[scale=0.35,trim=0mm 0mm 0mm 0mm,clip]{images/pivot.png} % left bottom right top
 \end{center}
@@
 
-\section{Discussion and related work}
-
-Although the technical focus of this paper is on the programming language theory behind the pivot type
-provider (Section~\ref{sec:pivot}), the paper also outlines interesting human-computer interaction 
+\section{Related and further work}
+
+The technical focus of this paper is on the programming language theory behind the pivot type
+provider (Section~\ref{sec:pivot}), but the paper also outlines interesting human-computer interaction 
 aspects (Section~\ref{sec:impl}). We discuss further related directions in this
-section before~concluding.
-
-\subsection{Related and further work}
+section before concluding.
+
+\subsection{Further  work}
 \label{sec:further}
 
+The pivot type provider shows the feasibility of using dot-driven development as a mechanism behind 
+simple programming tools for data exploration. Extending the mechanism to handle large and dirty 
+datasets poses a number of interesting challenges.
+
+\vspace{-0.25em}
+\subparagraph{Scalability.} A benefit of our approach based on relational algebra, is that the 
+query constructed by the pivot type provider can be translated to SQL and executed by a database 
+engine. This means that evaluating the query over large data sets does not pose a problem. However,
+the completion lists generated from data when filtering may require further consideration.
+
+We plan to explore a number of possibilities such as grouping the values by a prefix (e.g.
+\qident{starting with LO}.\ident{London} and \qident{starting with CA}.\ident{Cambridge})
+or grouping the values by their frequency (for example, \qident{occurring less than 100 
+times}.\ident{Grantchester} and \qident{occurring more than 10000 times}.\ident{London}). 
+Such encoding makes it possible to scale to an arbitrary data size, provided that the backing 
+data storage is equipped with an appropriate index.
+
+\vspace{-0.25em}
+\subparagraph{Expressivity.} 
+The case studies presented in the paper show that the pivot type provider is practically useful in 
+its current form, but we acknowledge that its expressivity is limited to simple queries.
+Making the tool more expressive to allow tasks such as denormalisation, handling of missing values 
+and dirty data is an important problem. 
+Unlike data querying (which is captured by the relational algebra), there is no generally accepted 
+``algebra of data cleaning'' and so more foundational work is needed, possibly building on 
+from tools such as Wrangler \cite{wrangler} and PADS \cite{pads}. We believe that the 
+``dot-driven development'' methodology can support richer languages and we intend to explore this
+direction in the future.
+
+\subsection{Related work}
+\label{sec:further}
+
+Our work builds on type providers, which have been pioneered in F\# \cite{inforich}.
+The technical contributions are related to several works on type systems. This section
+also gives an  overview of related work on human-computer interaction and commercial tools
+for data visualization.
+
 \subparagraph{Type providers.}
-Type providers have been pioneered in F\# \cite{inforich}. They can be seen as a form of dependent 
+Type providers first appeared in F\# \cite{inforich} and can also be seen as a form of dependent 
 typing \cite{idris-tp}; we take the opposite perspective and use type providers as a mechanism for 
@@
 computations is different from other type provider work \cite{fsdata,liteq,ageweb}, which focuses 
-on mapping of external data into types.
+on mapping of external data into types. To our best knowledge, the Azure type provider \cite{azureprov} 
+is the first type provider that provides members for specifying a restricted form of queries.
 
@@
 
+\vspace{-0.25em}
 \subparagraph{Relational algebra.}
@@
 
-\subsection{Conclusions}
+\vspace{-0.25em}
+\subparagraph{Commercial tools.}
+There is a wide range of commercial tools for building dashboards and data visualizations such as
+Microsoft Power BI \cite{powerbi}, Tableau \cite{tableau} and Qlik \cite{qlik}. Those allow users
+to build data visualizations through a user interface and embedded scripting capabilities. 
+The main difference from the pivot type provider is that none of these tools treats source code
+as primary and so they do not provide the same level of reproducibility as scripts written using
+the pivot type provider.
+
+\section{Conclusions}
 In this paper, we presented a simple programming language for data exploration. The language 
-addresses two problems with the current tooling for data science. One one hand, spreadsheets are 
+addresses two problems with the current tooling for data science. On one hand, spreadsheets are 
 easy to use, but are error-prone and do not lead to reproducible scripts that could be modified
@@
 
+\vspace{-0.25em}
+\subparagraph{Acknowledgements.}
+The author is grateful to Don Syme for numerous discussions about type providers, James
+Geddes and Kenji Takeda for suggestions and useful references and to Mariana Marasoiu and
+Alan Blackwell for ideas on human-computer interaction aspects of the work. Finally, thanks
+to the anonymous reviewers for useful suggestions and corrections.
+
 \bibliography{paper}

### 2017-05-22 ecoop-2017/paper.tex
Proceedings tweaks, mostly negative hspace, lolz

@@
 ~ l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\{f~|~F(f)=\kvd{num} \},~\textnormal{where} \\
-\quad \qquad \kvd{member}~\qident{by $f$ descending} : C_f = C_f(x) &\quad C_f, L_f = \ident{sort-and}(F, \vect{f \mapsto \ident{desc}})\\
-\quad \qquad \kvd{member}~\qident{by $f$ ascending} : C'_f = C'_f(x) &\quad C'_f, L'_f = \ident{sort-and}(F, \vect{f \mapsto \ident{asc}})\\
+\quad \qquad \kvd{member}~\qident{by $f$ desc} : C_f = C_f(x) &\quad C_f, L_f = \ident{sort-and}(F, \vect{f \mapsto \ident{desc}})\\
+\quad \qquad \kvd{member}~\qident{by $f$ asc} : C'_f = C'_f(x) &\quad C'_f, L'_f = \ident{sort-and}(F, \vect{f \mapsto \ident{asc}})\\
 \\[1em]
 \multicolumn{2}{l}{ \ident{sort-and}(F,\vect{s_1,\ldots,s_n}) = C, \{ C \mapsto (l, \bigcup L_f \cup \bigcup L'_f \cup L') \}\qquad\ball{4} }\\[0.25em]
-~ l = \kvd{type}~C(x:\ident{Query}) = &\hspace{-1em}\forall f\in\{f~|~F(f)=\kvd{num}, \,\nexists i.s_i=f'\mapsto\omega\wedge f'=f\}\\
-\quad \qquad \kvd{member}~\qident{and $f$ descending} : C_f = C_f(x) &\quad C_f, L_f = \ident{sort-and}(F, \vect{s_1,..,s_n, f \mapsto \ident{desc}})\\
-\quad \qquad \kvd{member}~\qident{and $f$ ascending} : C'_f = C'_f(x) &\quad C'_f, L'_f = \ident{sort-and}(F, \vect{s_1,..,s_n, f \mapsto \ident{asc}})\\
-\quad \qquad \kvd{member}~\ident{then} : C' = C'(\tau_{s_1,\ldots,s_n}(x))\quad~~\ball{5}           &\textnormal{where}~C', L' = \ident{pivot}(F)\\
+~ l = \kvd{type}~C(x:\ident{Query}) = &\hspace{-4em}\forall f\in\{f~|~F(f)=\kvd{num}, \,\nexists i.s_i=f'\mapsto\omega\wedge f'=f\}\\
+\quad \qquad \kvd{member}~\qident{$f$ desc} : C_f = C_f(x) &\hspace{-2em} C_f, L_f = \ident{sort-and}(F, \vect{s_1,..,s_n, f \mapsto \ident{desc}})\\
+\quad \qquad \kvd{member}~\qident{$f$ asc} : C'_f = C'_f(x) &\hspace{-2em} C'_f, L'_f = \ident{sort-and}(F, \vect{s_1,..,s_n, f \mapsto \ident{asc}})\\
+\quad \qquad \kvd{member}~\ident{then} : C' = C'(\tau_{s_1,\ldots,s_n}(x))  &\hspace{-2em}\textnormal{where}~C', L' = \ident{pivot}(F)\qquad ~~\ball{5}\\
 \end{array}
@@
 generates type for choosing the first sorting key and \ident{sort-and}~\ball{4} lets the user add more
-keys. The members are restricted to numerical columns (by checking $F(f)=\kvd{num}$).
+keys. For space reasons, we abbreviate \ident{ascending} and \ident{descending} as \ident{asc} and
+\ident{desc} in the generated member names and we omit \ident{and} in name of further keys such as 
+\qident{and Gold descending}.
+
+The members are restricted to numerical columns (by checking $F(f)=\kvd{num}$).
 The sort keys are kept as a vector. The \ident{sort} operation creates a singleton vector;
@@
 \begin{equation*}
+\hspace{-1.5em}
 \begin{array}{ll}
 \ident{group}(F) = C, \{ C\mapsto (l, \bigcup L_f) \}\quad\ball{1} \\[0.25em]
-~ l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)~\textnormal{where} \\
-\quad \qquad \kvd{member}~\qident{by $f$} : C_f = C_f(x) &\quad C_f, L_f = \ident{agg}(F, f, \{ f \mapsto F(f) \}, \emptyset)~~\ball{2}\\
+~ l = \kvd{type}~C(x:\ident{Query}) = &\hspace{-2.5em}\forall f\in\dom(F)~\textnormal{where} \\
+\quad \qquad \kvd{member}~\qident{by $f$} : C_f = C_f(x) &\hspace{-2.5em}\quad C_f, L_f = \ident{agg}(F, f, \{ f \mapsto F(f) \}, \emptyset)~~\ball{2}\\
 \\[0.5em]
@@
 \begin{definition}[Expansion]
-\label{thm:pivot-well}
+\label{thm:pivot-expand}
 Given a context $L$, we define $n^{\textnormal{th}}$ expansion of $L$, written 
@@
 \subsection{Related work}
-\label{sec:further}
+\label{sec:related}
 

### 2017-09-16 live/paper.tex
Working on live programming environment paper

\documentclass[sigplan,10pt]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}

\usepackage{lineno,hyperref,xcolor}
\usepackage{stmaryrd}
\usepackage{amssymb}
\usepackage{xypic}
\usepackage{semantic}
\usepackage{booktabs} 
\usepackage{subcaption}

\acmConference[PL'17]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
\acmYear{2017}
\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}
\setcopyright{none}
\bibliographystyle{ACM-Reference-Format}

\title{Design and implementaiton of a live coding environment}

\author{Tomas Petricek}
\affiliation{
  \institution{The Alan Turing Institute}
  \country{London, United Kingdom}
}
\email{tomas@tomasp.net}

\definecolor{cmtclr}{rgb}{0.0,0.6,0.0}
\definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
\definecolor{numclr}{rgb}{0.0,0.4,0.0}
\definecolor{strclr}{rgb}{0.4,0.3,0.0}
\definecolor{prepclr}{rgb}{0.6,0.0,0.2}
\newcommand{\vect}[1]{\langl #1 \rangl}
\newcommand{\langl}{\begin{picture}(4.5,7)
\put(1.1,2.5){\rotatebox{60}{\line(1,0){5.5}}}
\put(1.1,2.5){\rotatebox{300}{\line(1,0){5.5}}}
\end{picture}}
\newcommand{\rangl}{\begin{picture}(4.5,7)
\put(.9,2.5){\rotatebox{120}{\line(1,0){5.5}}}
\put(.9,2.5){\rotatebox{240}{\line(1,0){5.5}}}
\end{picture}}
\newcommand{\ball}[1]{\FPeval{\result}{clip(201+#1)}\textnormal{\ding{\result}}}
\newcommand{\lsep}{~\,|\,~}
\newcommand{\num}[1]{\textcolor{numclr}{#1}}
\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
\newcommand{\ident}[1]{\textnormal{\sffamily #1}}
\newcommand{\qident}[1]{\textnormal{\sffamily \guillemotleft #1\guillemotright}}
\newcommand{\dom}{\ident{dom}}
\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}

\newcommand{\bndclr}[1]{\textcolor{kvdclr}{#1}}
\newcommand{\bkndclr}[1]{\textcolor{prepclr}{#1}}
\newcommand{\blblclr}[1]{\textcolor{numclr}{#1}}
\newcommand{\bnd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
\newcommand{\bknd}[1]{\textnormal{\textcolor{prepclr}{\sffamily #1}}}
\newcommand{\blbl}[1]{\textnormal{\textcolor{numclr}{\sffamily #1}}}

\begin{document}
\begin{abstract}
Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
\end{abstract}
\maketitle

% ==================================================================================================

\section{Introduction}
One of the aspects that make spreadsheet tools such as Excel more accessible than programming 
environments is their liveness. When you change a value in a cell in Excel, the whole spreadsheet
is updated instantly and you see the new results immediately.

Increasing number of programming environments aim to provide the same live experience for more
standard programming languages, but doing this is not easy. Fully recomputing the whole program after
every single change is inefficient and calculating how a change in source code changes the result
is extremely hard when the editor allows arbitrary manipulation of program text. For example, 
consider the following simple program that gets the release years of 10 most expensive movies 
in a data set \ident{movies}:
%
\begin{equation*}
\begin{array}{l}  
\kvd{let}~\ident{top} =\ident{movies}\\
\quad .\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\ident{.take}(\num{10})\\
\quad .\ident{map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{yyyy}))}\\
\end{array}
\end{equation*}
%
A live coding environment shows a preview of the list of dates \ident{top} and the programmer then
changes the code by making $\num{10}$ a variable and changing the date format to see the full date:

\begin{equation*}
\begin{array}{l}  
\kvd{let}~\ident{count} = \num{10}\\
\kvd{let}~\ident{top} = \ident{movies}\\
\quad .\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\ident{.take}(\ident{count})\\
\quad \ident{.map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{dd-mm-yyyy}))}\\
\end{array}
\end{equation*}
%
Ideally, the live coding environment should understand the change, reuse a cached result of the
first two transformations (sorting and taking 10 elements) and only evaluate the last \ident{map}
and format the release dates of already computed top 10 movies.

This is not difficult to achieve if we represent the program in a structured way and allow the user
to edit code only via well-understood primitive operations such as ``extract variable'' (which has
no effect on the result) or ``change constant value'' (which forces recomputation of subsequent
transformations). However, many programmers still prefer to edit programs as free form text.
In this paper, we present the design and implementation of a live coding system that is capable
of reusing previously evaluated expressions as in the example above, yet, is integrated into an
ordinary text editor.

The motivation for this paper is twofold. First, implementing a live programming system requires
a different way of thinking about compilers and interpreters than the one presented in classic
programming language literature. Second, an increasing number of programming environments 
incorporate aspects of live programming [X,Y], yet relatively little has been written on their
architecture and implementation.

% --------------------------------------------------------------------------------------------------

\begin{figure*}
\includegraphics[scale=0.45]{scrape.png}
\caption{xxx}  
\end{figure*}

\newpage

% ==================================================================================================

\section{Live programming for data exploration}

The Gamma introduction and what this tries to do. See pretty screenshot.

This is what we care about - it's not live programming for your typical haskell,
but for R or Python.

\newpage

% ==================================================================================================

\section{Formalising live coding environment}

In this section, we present a formalisation of a live coding environment for a small,
expression-based programming language that supports \kvd{let} binding, member invocations
and $\lambda$ abstractions. This is the necessary minimum for data exploration as described
in the previous section and it excludes constructs such as a mechanism for defining new objects.
%
\begin{equation*}
\begin{array}{lcl}
e &=& \kvd{let}~x = e~\kvd{in}~e ~|~ \lambda x\rightarrow e ~|~ e.m(e, \ldots, e) ~|~ x ~|~ n
\end{array}
\end{equation*}
%
Here, $m$ ranges over member names, $x$ over variables and $n$ over primitive values such as 
numbers. Function values can be passed as arguments to methods (provided by a type provider), but 
for the purpose of this paper, we do not need to be able to invoke them directly.

\paragraph{The problem with functions.}
In the context of live programming, \kvd{let} binding and member access are unproblematic.
We can evaluate them and provide live preview for both of them, including all their sub-expressions.
Function values are more problematic, because their sub-expressions cannot be evaluated. For example:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{page} = \lambda x \rightarrow \ident{movies}.\ident{skip}(x*\num{10}).\ident{take}(\num{10})
\end{array}
\end{equation*}
%
We can provide live preview for the \ident{movies} sub-expression, but not for 
$\ident{movies}.\ident{skip}(x*\num{10})$ because we cannot obtain the value of $x$ without
running the rest of the program and analysing how the function is called later.

The method described in this paper does not provide live preview for sub-expressions that
contain free variables (which are not global objects provided by a type provider), but we
describe possible ways of doing so in Section~X and more speculative design of live coding 
friendly fucntions in Section~Y.

% --------------------------------------------------------------------------------------------------

\begin{figure*}
\begin{equation*}
\begin{array}{l}
\xymatrix{
& \bnd{val}(\num{10}) &\qquad\quad& \bnd{val}(\num{15})\\
\bnd{var}(\ident{data}) & 
  \bnd{mem}(\ident{skip},s_0)\ar[l]_{\blbl{arg}(0)} \ar[u]^{\blbl{arg}(1)} && 
  \bnd{mem}(\ident{take},s_1)\ar[ll]_{\blbl{arg}(0)} \ar[u]_{\blbl{arg}(1)}
}\\[5em]
\xymatrix{
&& \bnd{val}(\num{10})\\
\bnd{var}(\ident{data}) & 
  \bnd{mem}(\ident{skip}, s_0)\ar[l]_{\blbl{arg}(0)} \ar@/^/[ru]^{\blbl{arg}(1)} && 
  \bnd{mem}(\ident{take}, s_2)\ar[ll]_{\blbl{arg}(0)} \ar@/_/[lu]_{\blbl{arg}(1)}
}
\end{array}\qquad
\begin{array}{l}
\textnormal{a.) The first graph is constructed from}\\
\textnormal{the following initial expression:}\\[0.5em]
\quad\kvd{let}~x = \num{15}~\kvd{in}\\
\quad\ident{data.skip}(\num{10}).\ident{take}(x)\\
\\
\textnormal{b.) The second diagram shows the updated graph}\\
\textnormal{after the programmer changes $x$ to $10$:}\\[0.5em]
\quad\kvd{let}~x = \num{10}~\kvd{in}\\
\quad\ident{data.skip}(\num{10}).\ident{take}(x)
\end{array}
\end{equation*}
\caption{Dependency graphs formed by two steps of the live programming process. }
\label{fig:dep-graph}
\end{figure*}

% --------------------------------------------------------------------------------------------------

\subsection{Maintaining dependency graph}

The key idea behind our implementation is to maintain a dependency graph with nodes 
representing individual operations of the computation that can be partially evaluated
to obtain a preview. Each time the program text is modified, we parse it afresh (using an
error-recovering parser) and bind the abstract syntax tree to the dependency graph.

We remember the previously created nodes of the graph. When binding a new expression to
the graph, we reuse previously created nodes that have the same dependencies. For
expressions that have a new structure, we create new nodes (using a fresh symbol to 
identify them).

The nodes of the graph serve as unique keys into a lookup table with previously
evaluated operations of the computation. When a preview is requested, we use the node
bound to the expression to find a preview, or evaluate it by first forcing the evaluation 
of all parents in the dependency graph.

\paragraph{Elements of the graph.} The nodes of the graph represent individual operations
to be computed. In our design, the nodes themselves are used as keys, so we attach a unique 
\emph{symbol} to some of the nodes. That way, we can create two unique nodes representing, 
for example, access to a member named \ident{take} which differ in their dependencies.

Furthermore, the graph edges are labelled with labels indicating the kind of dependency. For
a method call, the labels are ``first argument'', ``second argument'' and so on. Formally:
%
\begin{equation*}
\begin{array}{rcll}
s&\hspace{-0.25em}\in\hspace{-0.25em}& \textit{Symbol}\\
i&\hspace{-0.25em}\in\hspace{-0.25em}& \textit{Integer}\\
n&\hspace{-0.25em}\in\hspace{-0.25em}& \textit{Primitive values}\\
x&\hspace{-0.25em}\in\hspace{-0.25em}& \textit{Variable names}\\
m&\hspace{-0.25em}\in\hspace{-0.25em}& \textit{Member names}\\[0.5em]
\bndclr{v}&\hspace{-0.25em}\in\hspace{-0.25em}&\bnd{val}(n)~|~\bnd{var}(x)~|~\bnd{mem}(m, s)~|~\bnd{fun}(x, s)&(\textit{Vertices})\\
\blblclr{l}&\hspace{-0.25em}\in\hspace{-0.25em}&\blbl{body}~|~\blbl{arg}(i)&(\textit{Edge labels})\\
\end{array}
\end{equation*}
%
The \bnd{val} node represents a primitive value and contains the value itself. Two occurences
of $\num{10}$ in the source code will be represented by the same node. Member access \bnd{mem}
contains the member name, together with a unique symbol -- two member access nodes with different 
dependencies will contain a different symbol. Dependencies of member access are labelled with 
\blbl{arg} indicating the index of the argument (the instance has index $0$ and arguments 
start with $1$).

Finally, nodes \bnd{fun} and \bnd{var} represent function values and variables bound by $\lambda$ 
abstraction. For simplicity, we use variable names rather than de Bruijn indices and so 
renaming a bound variable foces recomputation.

\paragraph{Example graph.} Figure~\ref{fig:dep-graph} illustrates how we construct and update the 
dependency graph. Node representing $\ident{take}(x)$ depends on the argument -- the
number $\num{15}$ -- and the instance, which is a node representing $\ident{skip}(\num{10})$.
This, in turn, depends on the instance \ident{data} and the number $\num{10}$. Note that variables
bound via \kvd{let} binding such as $x$ do not appear as $\bnd{var}$ nodes. The node using it
depends directly on the node representing the result of the expression that is assigned to $x$.

After changing the value of $x$, we create a new graph. The dependencies of the node 
$\bnd{mem}(\ident{skip}, s_0)$ are unchanged and so the node is reused. This means that this
part of the program is not recomputed. The $\blbl{arg}(1)$ dependency of the \ident{take} call 
changed and so we create a node $\bnd{mem}(\ident{skip}, s_2)$ with a new fresh symbol $s_2$.
The preview for this node is then recomputed as needed using the already known values of its
dependencies.

% --------------------------------------------------------------------------------------------------

\begin{figure*}[t]
\begin{equation*}
\begin{array}{l}
\ident{bind}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n)) = \\ 
\qquad \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n)\\
\quad \textnormal{when}~\bndclr{v_i}, (V_i, E_i) = \ident{bind}_{\Gamma, \Delta}(e_i)\\
\quad \textnormal{and}~(\bknd{mem}(m),[(\bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v_n}, \blbl{arg}(n))]) \notin \ident{dom}(\Delta)\\
\quad \textnormal{let}~\bndclr{v} = \bnd{mem}(m, s), s~\textnormal{fresh}\\
\quad \textnormal{let}~E = \{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v},\bndclr{v_n}, \blbl{arg}(n)) \}
\\[0.75em]
\ident{bind}_{\Gamma, \Delta}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V, \{e\} \cup E)\\
\quad \textnormal{when}~\bndclr{v_0}, (V, E) = \ident{bind}_{\Gamma, \Delta}(e)\\
\quad \textnormal{and}~\bndclr{v} = X(\bknd{fun}(x),[(\bndclr{v_0}, \blbl{body})])\\
\quad \textnormal{let}~e = (\bndclr{v}, \bndclr{v_0}, \blbl{body}) 
\\[0.75em]
\ident{bind}_{\Gamma, \Delta}(n) = \bnd{val}(n), (\{ \bnd{val}(n) \}, \emptyset) }
\\[0.5em]
\ident{bind}_{\Gamma, \Delta}(x) = \bnd{var}(x), (\{ \bnd{var}(x) \}, \emptyset) }\quad \textnormal{when}~x \notin \ident{dom}(\Gamma)
\\[0.5em]
\ident{bind}_{\Gamma, \Delta}(x) = \bndclr{v}, (\{ \bndclr{v} \}, \emptyset)\quad \textnormal{when}~\bndclr{v} = \Gamma(x)
\\[0.5em]
\end{array}
\begin{array}{l}
\ident{bind}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n)) = \\ 
\qquad \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n)\\
\quad \textnormal{when}~\bndclr{v_i}, (V_i, E_i) = \ident{bind}_{\Gamma, \Delta}(e_i)\\
\quad \textnormal{and}~(\bknd{mem}(m),[(\bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v_n}, \blbl{arg}(n))]) \notin \ident{dom}(\Delta)\\
\quad \textnormal{let}~\bndclr{v} = \bnd{mem}(m, s), s~\textnormal{fresh}\\
\quad \textnormal{let}~E = \{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v},\bndclr{v_n}, \blbl{arg}(n)) \}
\\[0.75em]
\ident{bind}_{\Gamma, \Delta}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V, \{e\} \cup E)\\
\quad \textnormal{when}~\bndclr{v_0}, (V, E) = \ident{bind}_{\Gamma, \Delta}(e)\\
\quad \textnormal{and}~\bndclr{v} = X(\bknd{fun}(x),[(\bndclr{v_0}, \blbl{body})])\\
\quad \textnormal{let}~e = (\bndclr{v}, \bndclr{v_0}, \blbl{body}) 
\\[0.75em]
\ident{bind}_{\Gamma, \Delta}(\kvd{let}~x=e_1~\kvd{in}~e_2) = \bndclr{v}, (\{\bndclr{v}\} \cup V \cup V_1, E \cup E_1)\\
\quad \textnormal{let}~\bndclr{v_1}, (V_1,E_1) = \ident{bind}_{\Gamma,\Delta}(e_1)\\
\quad \textnormal{let}~\Gamma_1 = \Gamma \cup \{(x,\bndclr{v_1})\} \\
\quad \textnormal{let}~\bndclr{v}, (V, E) = \ident{bind}_{\Gamma, \Delta}(e_2)
\end{array}
\end{equation*}
\caption{Binding}
\end{figure*}

% --------------------------------------------------------------------------------------------------

\paragraph{Reusing graph nodes.} The binding process takes an expression and constructs a 
dependency graph, reusing existing nodes when possible. For this, we keep a lookup table 
of member access and function value nodes. The key is formed by a node kind (for 
disambiguation) together with a list of dependencies. A node kind is a member access or a function:
%
\begin{equation*}
\begin{array}{lcll}
\bkndclr{k}&\in&\bknd{fun}(x)~|~\bknd{mem}(m)&\qquad(\textit{Node  kinds})
\end{array}
\end{equation*}
%
Given a lookup table $X$, we write $X(\bkndclr{k}, [(\bndclr{n_1},\blblclr{l_1}), \ldots,
(\bndclr{v_n},\blblclr{l_n})])$ to perform a lookup for a node of a kind $\bkndclr{k}$ that has dependencies
$\bndclr{v_1}, \ldots, \bndclr{v_n}$ labelled with labels $\blblclr{l_1}, \ldots, \blblclr{l_n}$.

For example, when creating the graph in Figure~\ref{fig:dep-graph} (b), we perform the following
lookup for the \ident{skip} member access:
%
\[ X(\bknd{mem}(\ident{skip}), [(\bnd{var}(\ident{data}),\blbl{arg}(0)), (\bnd{val}(\num{10}), \blbl{arg}(1))]) \]
%
The lookup returns the node $\bnd{mem}(\ident{skip}, s_0)$ known from the previous step. We then perform
the following lookup for the \ident{take} member access:
%
\[ X(\bknd{mem}(\ident{take}), [(\bnd{mem}(\ident{skip}, s_0),\blbl{arg}(0)), (\bnd{val}(\num{10}), \blbl{arg}(1))]) \]
%
In the previous graph, the argument of \ident{take} was $\num{15}$ rather than $\num{10}$ and so
this lookup fails. We then construct a new node $\bnd{mem}(\ident{take}, s_2)$ using a fresh
symbol $s_2$.

% --------------------------------------------------------------------------------------------------

\subsection{Binding expressions to a graph}
Some simple definitions -- $s$ is a unique symbol, the rest is as before

We represent graph as (V, E) where E = (v1, v2, lbl), rather than having a function on the side

\vspace{1em}
\noindent
Then we define binding function such that

\begin{equation*}
\ident{bind}_{\Gamma, \Delta}(e) = (V, E), v
\end{equation*}

\vspace{1em}
\noindent
Here, $X$ is a mapping from a pair of vertex kind and a list of vertices 
to a vertext with a set of edges leading from it i.e. 
$(K, [v_1, \ldots, v_n]) = v, E$. Then, the $\ident{bind}_{\Gamma, \Delta}$ function
takes an expression and returns a dependency graph for the expression
together with the vertex representing the expression. The vertices of the 
graph are formed by $V$ and the edges are labelled by labels $L$.

\newpage

% --------------------------------------------------------------------------------------------------

\newpage

\newpage

\section*{References}

\bibliography{paper}

\end{document}

### 2017-09-18 live/paper.tex
More writing on binding process

@@
 \section{Formalising live coding environment}
+\label{sec:formal}
 
@@
 \subsection{Maintaining dependency graph}
+\label{sec:formal-deps}
 
@@
 \begin{figure*}[t]
+\vspace{-0.5em}
 \begin{equation*}
-\begin{array}{l}
-\ident{bind}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n)) = \\ 
+\begin{array}{ll}
+\ident{bind}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n)) = &(1)~~\\ 
 \qquad \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n)\\
 \quad \textnormal{when}~\bndclr{v_i}, (V_i, E_i) = \ident{bind}_{\Gamma, \Delta}(e_i)\\
-\quad \textnormal{and}~(\bknd{mem}(m),[(\bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v_n}, \blbl{arg}(n))]) \notin \ident{dom}(\Delta)\\
+\quad \textnormal{and}~\bndclr{v} = \Delta(\bknd{mem}(m),[(\bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v_n}, \blbl{arg}(n))])\\
+\quad \textnormal{let}~E = \{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v}, \bndclr{v_n}, \blbl{arg}(n))\}
+\\[0.75em]
+\ident{bind}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n)) = &(2)\\ 
+\qquad \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n)\\
+\quad \textnormal{when}~\bndclr{v_i}, (V_i, E_i) = \ident{bind}_{\Gamma, \Delta}(e_i)\\
+\quad \textnormal{and}~(\bknd{mem}(m),[(\bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v_n}, \blbl{arg}(n))]) \notin \ident{dom}(\Delta)\hspace{-200em}.\\
 \quad \textnormal{let}~\bndclr{v} = \bnd{mem}(m, s), s~\textnormal{fresh}\\
@@
 \\[0.75em]
-\ident{bind}_{\Gamma, \Delta}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V, \{e\} \cup E)\\
-\quad \textnormal{when}~\bndclr{v_0}, (V, E) = \ident{bind}_{\Gamma, \Delta}(e)\\
-\quad \textnormal{and}~\bndclr{v} = X(\bknd{fun}(x),[(\bndclr{v_0}, \blbl{body})])\\
+\ident{bind}_{\Gamma, \Delta}(\kvd{let}~x=e_1~\kvd{in}~e_2) = \bndclr{v}, (\{\bndclr{v}\} \cup V \cup V_1, E \cup E_1)&(3)\\
+\quad \textnormal{let}~\bndclr{v_1}, (V_1,E_1) = \ident{bind}_{\Gamma,\Delta}(e_1)\\
+\quad \textnormal{let}~\Gamma_1 = \Gamma \cup \{(x,\bndclr{v_1})\} \\
+\quad \textnormal{let}~\bndclr{v}, (V, E) = \ident{bind}_{\Gamma_1, \Delta}(e_2)
+\end{array}
+\begin{array}{ll}
+\ident{bind}_{\Gamma, \Delta}(n) = \bnd{val}(n), (\{ \bnd{val}(n) \}, \emptyset) }&(4)
+\\[0.75em]
+\ident{bind}_{\Gamma, \Delta}(x) = \bndclr{v}, (\{ \bndclr{v} \}, \emptyset)\quad \textnormal{when}~\bndclr{v} = \Gamma(x)&(6)
+\\[1em]
+\ident{bind}_{\Gamma, \Delta}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V, \{e\} \cup E)&(7)\\
+\quad \textnormal{when}~\Gamma_1 = \Gamma \cup \{ x, \bnd{var}(x) \}\\
+\quad \textnormal{and}~\bndclr{v_0}, (V, E) = \ident{bind}_{\Gamma_1, \Delta}(e)\\
+\quad \textnormal{and}~\bndclr{v} = \Delta(\bknd{fun}(x),[(\bndclr{v_0}, \blbl{body})])\\
 \quad \textnormal{let}~e = (\bndclr{v}, \bndclr{v_0}, \blbl{body}) 
-\\[0.75em]
-\ident{bind}_{\Gamma, \Delta}(n) = \bnd{val}(n), (\{ \bnd{val}(n) \}, \emptyset) }
-\\[0.5em]
-\ident{bind}_{\Gamma, \Delta}(x) = \bnd{var}(x), (\{ \bnd{var}(x) \}, \emptyset) }\quad \textnormal{when}~x \notin \ident{dom}(\Gamma)
-\\[0.5em]
-\ident{bind}_{\Gamma, \Delta}(x) = \bndclr{v}, (\{ \bndclr{v} \}, \emptyset)\quad \textnormal{when}~\bndclr{v} = \Gamma(x)
+\\[1em]
+\ident{bind}_{\Gamma, \Delta}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V, \{e\} \cup E)&(8)\\
+\quad \textnormal{when}~\Gamma_1 = \Gamma \cup \{ x, \bnd{var}(x) \}\\
+\quad \textnormal{and}~\bndclr{v_0}, (V, E) = \ident{bind}_{\Gamma_1, \Delta}(e)\\
+\quad \textnormal{and}~(\bknd{fun}(x),[(\bndclr{v_0}, \blbl{body})])\notin \ident{dom}(\Delta)\\
+\quad \textnormal{let}~\bndclr{v} = \bnd{fun}(s, x), s~\textnormal{fresh}\\
+\quad \textnormal{let}~e = (\bndclr{v},\bndclr{v_0},\blbl{body}) 
 \\[0.5em]
 \end{array}
-\begin{array}{l}
-\ident{bind}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n)) = \\ 
-\qquad \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n)\\
-\quad \textnormal{when}~\bndclr{v_i}, (V_i, E_i) = \ident{bind}_{\Gamma, \Delta}(e_i)\\
-\quad \textnormal{and}~(\bknd{mem}(m),[(\bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v_n}, \blbl{arg}(n))]) \notin \ident{dom}(\Delta)\\
-\quad \textnormal{let}~\bndclr{v} = \bnd{mem}(m, s), s~\textnormal{fresh}\\
-\quad \textnormal{let}~E = \{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v},\bndclr{v_n}, \blbl{arg}(n)) \}
-\\[0.75em]
-\ident{bind}_{\Gamma, \Delta}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V, \{e\} \cup E)\\
-\quad \textnormal{when}~\bndclr{v_0}, (V, E) = \ident{bind}_{\Gamma, \Delta}(e)\\
-\quad \textnormal{and}~\bndclr{v} = X(\bknd{fun}(x),[(\bndclr{v_0}, \blbl{body})])\\
-\quad \textnormal{let}~e = (\bndclr{v}, \bndclr{v_0}, \blbl{body}) 
-\\[0.75em]
-\ident{bind}_{\Gamma, \Delta}(\kvd{let}~x=e_1~\kvd{in}~e_2) = \bndclr{v}, (\{\bndclr{v}\} \cup V \cup V_1, E \cup E_1)\\
-\quad \textnormal{let}~\bndclr{v_1}, (V_1,E_1) = \ident{bind}_{\Gamma,\Delta}(e_1)\\
-\quad \textnormal{let}~\Gamma_1 = \Gamma \cup \{(x,\bndclr{v_1})\} \\
-\quad \textnormal{let}~\bndclr{v}, (V, E) = \ident{bind}_{\Gamma, \Delta}(e_2)
-\end{array}
 \end{equation*}
-\caption{Binding}
+\vspace{-0.5em}
+\caption{Rules of the binding process, whic constructs a dependency graph for an expression.}
+\label{fig:binding-rules}
 \end{figure*}
@@
 %
-Given a lookup table $X$, we write $X(\bkndclr{k}, [(\bndclr{n_1},\blblclr{l_1}), \ldots,
+Given a lookup table $\Delta$, we write $\Delta(\bkndclr{k}, [(\bndclr{n_1},\blblclr{l_1}), \ldots,
 (\bndclr{v_n},\blblclr{l_n})])$ to perform a lookup for a node of a kind $\bkndclr{k}$ that has dependencies
@@
 %
-\[ X(\bknd{mem}(\ident{skip}), [(\bnd{var}(\ident{data}),\blbl{arg}(0)), (\bnd{val}(\num{10}), \blbl{arg}(1))]) \]
+\[ \Delta(\bknd{mem}(\ident{skip}), [(\bnd{var}(\ident{data}),\blbl{arg}(0)), (\bnd{val}(\num{10}), \blbl{arg}(1))]) \]
 %
@@
 %
-\[ X(\bknd{mem}(\ident{take}), [(\bnd{mem}(\ident{skip}, s_0),\blbl{arg}(0)), (\bnd{val}(\num{10}), \blbl{arg}(1))]) \]
+\[ \Delta(\bknd{mem}(\ident{take}), [(\bnd{mem}(\ident{skip}, s_0),\blbl{arg}(0)), (\bnd{val}(\num{10}), \blbl{arg}(1))]) \]
 %
@@
 
-\subsection{Binding expressions to a graph}
-Some simple definitions -- $s$ is a unique symbol, the rest is as before
+\subsection{Binding an expressions to a graph}
+\label{sec:formal-bind}
 
-We represent graph as (V, E) where E = (v1, v2, lbl), rather than having a function on the side
+When constructing the dependency graph, our implementation annotates the nodes of the 
+abstract syntax tree with the nodes of the dependency graph, forming a mapping
+$e \rightarrow \bndclr{v}$. For this reason, we call the process \emph{binding}.
 
-\vspace{1em}
-\noindent
-Then we define binding function such that
+The process of binding is defined by the rules in Figure~\ref{fig:binding-rules}.
+The \ident{bind} function is annotated with a lookup table $\Delta$ discussed in 
+Section~\ref{sec:formal-deps} and a variable context $\Gamma$. The variable context 
+is a map from variable names to dependency graph nodes and is used for variables bound
+using \kvd{let} binding.
 
+When applied on an expression $e$, binding $\ident{bind}_{\Gamma,\Delta}(e)$ returns a
+dependency graph $(V, E)$ paired with a node $\bndclr{v}$ corresponding to the expression $e$.
+In the graph, $V$ is a set of nodes $\bndclr{v}$ and $E$ is a set of labelled edges
+$(\bndclr{v_1}, \bndclr{v_2}, \blblclr{l})$. We attach the label directly to the edge rather than
+keeping a separate colouring function as this makes the formalisation simpler.
+
+\paragraph{Binding member access.} In all rules, we recursively bind sub-expressions to get
+a dependency graph for each sub-expression and a graph node that represents it. The nodes 
+representing sub-expressions are then used as dependencies for lookup into $\Delta$, together
+with their labels. When binding a member access, we reuse an existing node if it is defined by 
+$\Delta$ (1) or we create a new node containing a fresh symbol when the domain of $\Delta$ does 
+not contain a key describing the current member access~(2).
+
+\paragraph{Binding let binding.} For \kvd{let} binding (3), we first bind the expression $e_1$ assigned
+to the variable to obtain a graph node $\bndclr{v_1}$. We then bind the body expression $e_2$,
+but using a variable context $\Gamma_1$ that maps the value of the variable to the graph node
+$\bndclr{v_1}$. The variable context is used when binding a variable (6) and so all variables 
+declared using \kvd{let} binding will be bound to a graph node representing the value assinged 
+to the variable. The node bound to the overall \kvd{let} expression is then the graph node bound
+to the body expression.
+
+\paragraph{Binding function values.} If a function value uses its argument, we will not be able
+to evaluate its body. In this case, the graph node bound to a function will depend on a synthetic
+node $\bnd{var}(x)$ that represents the variable with no value. When binding a function, we 
+create the synthetic variable and add it to the variable context $\Gamma_1$ before binding the
+body. As with member access, the node representing a function may (7) or may not (8) be already 
+present in the lookup table.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[!b]
 \begin{equation*}
-\ident{bind}_{\Gamma, \Delta}(e) = (V, E), v
+\begin{array}{l}
+\Delta_{i}(\bknd{mem}(m), [(\bndclr{v_0},\blbl{arg}(0)),\ldots, (\bndclr{v_n},\blbl{arg}(n))]) = \bndclr{v}\\
+\quad \textnormal{for all}~\bnd{mem}(m, s) \in V\\
+\quad \textnormal{such that}~(\bnd{mem}(m, s), \bndclr{v_i}, \blbl{arg}(i)) \in E~\textnormal{for}~i\in 0,..,n
+\\[0.75em]
+\Delta_{i}(\bknd{fun}(x), [(\bndclr{v_1},\blbl{body})]) = \bndclr{v}\\
+\quad \textnormal{for all}~\bnd{fun}(x, s) \in V\\
+\quad \textnormal{such that}~(\bnd{fun}(x, s), \bndclr{v_1}, \blbl{body}) \in E
+\\[0.75em]
+\Delta_{i}(\bkndclr{v}) = \Delta_{i-1}(\bkndclr{v})\quad \textnormal{(otherwise)}
+\end{array}
 \end{equation*}
+\caption{Updating the node cache after binding a new graph}
+\label{fig:loop}
+\end{figure}
 
-\vspace{1em}
-\noindent
-Here, $X$ is a mapping from a pair of vertex kind and a list of vertices 
-to a vertext with a set of edges leading from it i.e. 
-$(K, [v_1, \ldots, v_n]) = v, E$. Then, the $\ident{bind}_{\Gamma, \Delta}$ function
-takes an expression and returns a dependency graph for the expression
-together with the vertex representing the expression. The vertices of the 
-graph are formed by $V$ and the edges are labelled by labels $L$.
+\subsection{Edit and rebind loop}
 
+The binding process formalised in Section~\ref{sec:formal-bind} specifies how to update the 
+dependency graph after updated program text is parsed. During live coding, this is done 
+repeatedly as the programmer edits code. Throughout the process, we maintain a series of 
+lookup table states $\Delta_0, \Delta_1, \Delta_2, \ldots$. Initially, the lookup table is 
+empty, i.e.~$\Delta_0 = \emptyset$.
+
+At a step $i$, we parse an expression $e_i$ and calculate the new dependency graph and a node
+bound to the top-level expression using the previous $\Delta$:
+%
+\begin{equation*}
+\bndclr{v}, (V, E) = \ident{bind}_{\emptyset, \Delta_{i-1}}(e_i)
+\end{equation*}
+%
+The new state of the node cache is then computed by adding newly created nodes from the graph
+$(V, E)$ to the previous cache $\Delta_{i-1}$. This is done for function and member nodes that
+contain unique symbols as defined in Figure~\ref{fig:loop}. We do not need to cache nodes
+representing primitive values and variables as those do not contain symbols and will remain the
+same due to the way they are constructed.
+
+% ==================================================================================================
+
+\section{Evaluating previews}
+\newpage
+
+\section{Properties}
 \newpage

### 2017-10-01 live/paper.tex
Evaluating previews section

@@
 % ==================================================================================================
+% TODO: Change values from 'n' to 'o' to indicate they are objects and can have members
+% ==================================================================================================
 
@@
 \newpage
+~
 
 % ==================================================================================================
-
-\section{Formalising live coding environment}
-\label{sec:formal}
-
-In this section, we present a formalisation of a live coding environment for a small,
-expression-based programming language that supports \kvd{let} binding, member invocations
-and $\lambda$ abstractions. This is the necessary minimum for data exploration as described
-in the previous section and it excludes constructs such as a mechanism for defining new objects.
-%
-\begin{equation*}
-\begin{array}{lcl}
-e &=& \kvd{let}~x = e~\kvd{in}~e ~|~ \lambda x\rightarrow e ~|~ e.m(e, \ldots, e) ~|~ x ~|~ n
-\end{array}
-\end{equation*}
-%
-Here, $m$ ranges over member names, $x$ over variables and $n$ over primitive values such as 
-numbers. Function values can be passed as arguments to methods (provided by a type provider), but 
-for the purpose of this paper, we do not need to be able to invoke them directly.
-
-\paragraph{The problem with functions.}
-In the context of live programming, \kvd{let} binding and member access are unproblematic.
-We can evaluate them and provide live preview for both of them, including all their sub-expressions.
-Function values are more problematic, because their sub-expressions cannot be evaluated. For example:
-%
-\begin{equation*}
-\begin{array}{l}
-\kvd{let}~\ident{page} = \lambda x \rightarrow \ident{movies}.\ident{skip}(x*\num{10}).\ident{take}(\num{10})
-\end{array}
-\end{equation*}
-%
-We can provide live preview for the \ident{movies} sub-expression, but not for 
-$\ident{movies}.\ident{skip}(x*\num{10})$ because we cannot obtain the value of $x$ without
-running the rest of the program and analysing how the function is called later.
-
-The method described in this paper does not provide live preview for sub-expressions that
-contain free variables (which are not global objects provided by a type provider), but we
-describe possible ways of doing so in Section~X and more speculative design of live coding 
-friendly fucntions in Section~Y.
-
-% --------------------------------------------------------------------------------------------------
 
@@
 
+\newpage
+
+\section{Formalising live coding environment}
+\label{sec:formal}
+
+In this section, we present a formalisation of a live coding environment for a small,
+expression-based programming language that supports \kvd{let} binding, member invocations
+and $\lambda$ abstractions. This is the necessary minimum for data exploration as described
+in the previous section.
+
+It excludes constructs such as a mechanism for defining new objects as we assume that those
+are imported from the context through a mechanims such as type providers.
+%
+\begin{equation*}
+\begin{array}{lcl}
+e &=& \kvd{let}~x = e~\kvd{in}~e ~|~ \lambda x\rightarrow e ~|~ e.m(e, \ldots, e) ~|~ x ~|~ n
+\end{array}
+\end{equation*}
+%
+Here, $m$ ranges over member names, $x$ over variables and $n$ over primitive values such as 
+numbers. Function values can be passed as arguments to methods (provided by a type provider), but 
+for the purpose of this paper, we do not need to be able to invoke them directly.
+
+\paragraph{The problem with functions.}
+In the context of live programming, \kvd{let} binding and member access are unproblematic.
+We can evaluate them and provide live preview for both of them, including all their sub-expressions.
+Function values are more problematic, because their sub-expressions cannot be evaluated. For example:
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~\ident{page} = \lambda x \rightarrow \ident{movies}.\ident{skip}(x*\num{10}).\ident{take}(\num{10})
+\end{array}
+\end{equation*}
+%
+We can provide live preview for the \ident{movies} sub-expression, but not for 
+$\ident{movies}.\ident{skip}(x*\num{10})$ because we cannot obtain the value of $x$ without
+running the rest of the program and analysing how the function is called later.
+
+The method described in this paper does not provide live preview for sub-expressions that
+contain free variables (which are not global objects provided by a type provider), but we
+describe possible ways of doing so in Section~X and more speculative design of live coding 
+friendly fucntions in Section~Y.
+
+% --------------------------------------------------------------------------------------------------
+
 \subsection{Maintaining dependency graph}
@@
 of all parents in the dependency graph.
-
-\paragraph{Elements of the graph.} The nodes of the graph represent individual operations
-to be computed. In our design, the nodes themselves are used as keys, so we attach a unique 
-\emph{symbol} to some of the nodes. That way, we can create two unique nodes representing, 
-for example, access to a member named \ident{take} which differ in their dependencies.
-
-Furthermore, the graph edges are labelled with labels indicating the kind of dependency. For
-a method call, the labels are ``first argument'', ``second argument'' and so on. Formally:
-%
-\begin{equation*}
-\begin{array}{rcll}
-s&\hspace{-0.25em}\in\hspace{-0.25em}& \textit{Symbol}\\
-i&\hspace{-0.25em}\in\hspace{-0.25em}& \textit{Integer}\\
-n&\hspace{-0.25em}\in\hspace{-0.25em}& \textit{Primitive values}\\
-x&\hspace{-0.25em}\in\hspace{-0.25em}& \textit{Variable names}\\
-m&\hspace{-0.25em}\in\hspace{-0.25em}& \textit{Member names}\\[0.5em]
-\bndclr{v}&\hspace{-0.25em}\in\hspace{-0.25em}&\bnd{val}(n)~|~\bnd{var}(x)~|~\bnd{mem}(m, s)~|~\bnd{fun}(x, s)&(\textit{Vertices})\\
-\blblclr{l}&\hspace{-0.25em}\in\hspace{-0.25em}&\blbl{body}~|~\blbl{arg}(i)&(\textit{Edge labels})\\
-\end{array}
-\end{equation*}
-%
-The \bnd{val} node represents a primitive value and contains the value itself. Two occurences
-of $\num{10}$ in the source code will be represented by the same node. Member access \bnd{mem}
-contains the member name, together with a unique symbol -- two member access nodes with different 
-dependencies will contain a different symbol. Dependencies of member access are labelled with 
-\blbl{arg} indicating the index of the argument (the instance has index $0$ and arguments 
-start with $1$).
-
-Finally, nodes \bnd{fun} and \bnd{var} represent function values and variables bound by $\lambda$ 
-abstraction. For simplicity, we use variable names rather than de Bruijn indices and so 
-renaming a bound variable foces recomputation.
-
-\paragraph{Example graph.} Figure~\ref{fig:dep-graph} illustrates how we construct and update the 
-dependency graph. Node representing $\ident{take}(x)$ depends on the argument -- the
-number $\num{15}$ -- and the instance, which is a node representing $\ident{skip}(\num{10})$.
-This, in turn, depends on the instance \ident{data} and the number $\num{10}$. Note that variables
-bound via \kvd{let} binding such as $x$ do not appear as $\bnd{var}$ nodes. The node using it
-depends directly on the node representing the result of the expression that is assigned to $x$.
-
-After changing the value of $x$, we create a new graph. The dependencies of the node 
-$\bnd{mem}(\ident{skip}, s_0)$ are unchanged and so the node is reused. This means that this
-part of the program is not recomputed. The $\blbl{arg}(1)$ dependency of the \ident{take} call 
-changed and so we create a node $\bnd{mem}(\ident{skip}, s_2)$ with a new fresh symbol $s_2$.
-The preview for this node is then recomputed as needed using the already known values of its
-dependencies.
 
@@
 \vspace{-0.5em}
-\caption{Rules of the binding process, whic constructs a dependency graph for an expression.}
+\caption{Rules of the binding process, which constructs a dependency graph for an expression.}
 \label{fig:binding-rules}
 \end{figure*}
+
+% --------------------------------------------------------------------------------------------------
+\paragraph{Elements of the graph.} The nodes of the graph represent individual operations
+to be computed. In our design, the nodes themselves are used as keys, so we attach a unique 
+\emph{symbol} to some of the nodes. That way, we can create two unique nodes representing, 
+for example, access to a member named \ident{take} which differ in their dependencies.
+
+Furthermore, the graph edges are labelled with labels indicating the kind of dependency. For
+a method call, the labels are ``first argument'', ``second argument'' and so on. Formally:
+%
+\begin{equation*}
+\begin{array}{rcll}
+s&\hspace{-0.25em}\in\hspace{-0.25em}& \textit{Symbol}\\
+i&\hspace{-0.25em}\in\hspace{-0.25em}& \textit{Integer}\\
+n&\hspace{-0.25em}\in\hspace{-0.25em}& \textit{Primitive values}\\
+x&\hspace{-0.25em}\in\hspace{-0.25em}& \textit{Variable names}\\
+m&\hspace{-0.25em}\in\hspace{-0.25em}& \textit{Member names}\\[0.5em]
+\bndclr{v}&\hspace{-0.25em}\in\hspace{-0.25em}&\bnd{val}(n)~|~\bnd{var}(x)~|~\bnd{mem}(m, s)~|~\bnd{fun}(x, s)&(\textit{Vertices})\\
+\blblclr{l}&\hspace{-0.25em}\in\hspace{-0.25em}&\blbl{body}~|~\blbl{arg}(i)&(\textit{Edge labels})\\
+\end{array}
+\end{equation*}
+%
+The \bnd{val} node represents a primitive value and contains the value itself. Two occurences
+of $\num{10}$ in the source code will be represented by the same node. Member access \bnd{mem}
+contains the member name, together with a unique symbol -- two member access nodes with different 
+dependencies will contain a different symbol. Dependencies of member access are labelled with 
+\blbl{arg} indicating the index of the argument (the instance has index $0$ and arguments 
+start with $1$).
+
+Finally, nodes \bnd{fun} and \bnd{var} represent function values and variables bound by $\lambda$ 
+abstraction. For simplicity, we use variable names rather than de Bruijn indices and so 
+renaming a bound variable foces recomputation.
+
+\paragraph{Example graph.} Figure~\ref{fig:dep-graph} illustrates how we construct and update the 
+dependency graph. Node representing $\ident{take}(x)$ depends on the argument -- the
+number $\num{15}$ -- and the instance, which is a node representing $\ident{skip}(\num{10})$.
+This, in turn, depends on the instance \ident{data} and the number $\num{10}$. Note that variables
+bound via \kvd{let} binding such as $x$ do not appear as $\bnd{var}$ nodes. The node using it
+depends directly on the node representing the result of the expression that is assigned to $x$.
+
+After changing the value of $x$, we create a new graph. The dependencies of the node 
+$\bnd{mem}(\ident{skip}, s_0)$ are unchanged and so the node is reused. This means that this
+part of the program is not recomputed. The $\blbl{arg}(1)$ dependency of the \ident{take} call 
+changed and so we create a node $\bnd{mem}(\ident{skip}, s_2)$ with a new fresh symbol $s_2$.
+The preview for this node is then recomputed as needed using the already known values of its
+dependencies.
 
@@
 \begin{figure}[!b]
+\vspace{-0.5em}
 \begin{equation*}
@@
 \end{equation*}
+\vspace{-0.5em}
 \caption{Updating the node cache after binding a new graph}
@@
 \end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure*}
+\begin{equation*}
+\qquad\qquad\quad\begin{array}{l}
+\inference[(lift-expr)~]
+  {\bndclr{v} \Downarrow \llbracket e \rrbracket_\Gamma}
+  {\bndclr{v} \Downarrow_{\ident{lift}} \llbracket e \rrbracket_\Gamma}
+\\
+\\
+\inference[(lift-prev)~]
+  {\bndclr{v} \Downarrow p}
+  {\bndclr{v} \Downarrow_{\ident{lift}} \llbracket p \rrbracket_\emptyset}
+\\
+\\
+\inference[(val)~]
+  {~}
+  {\bnd{val}(n) \Downarrow n }
+\\
+\\
+\inference[(var)~]
+  {~}
+  {\bnd{var}(x) \Downarrow \llbracket x \rrbracket_x}
+  \\[0.5em]~
+\end{array}
+\qquad
+\begin{array}{l}
+\inference[(fun-val)~]
+  { (\bnd{fun}(x, s), \bndclr{v}, \blbl{body}) \in E & \bndclr{v} \Downarrow p }
+  { \bnd{fun}(x, s) \Downarrow \lambda x\rightarrow p }
+\\
+\\
+\inference[(fun-bind)~]
+  { (\bnd{fun}(x, s), \bndclr{v}, \blbl{body}) \in E & \bndclr{v} \Downarrow \llbracket e \rrbracket_{x} }
+  { \bnd{fun}(x, s) \Downarrow \lambda x\rightarrow e }
+\\
+\\
+\inference[(fun-expr)~]
+  { (\bnd{fun}(x, s), \bndclr{v}, \blbl{body}) \in E & \bndclr{v} \Downarrow \llbracket e \rrbracket_{x,\Gamma} }
+  { \bnd{fun}(x, s) \Downarrow \llbracket \lambda x\rightarrow e \rrbracket_\Gamma }
+\\[0.5em]~
+\end{array}
+\end{equation*}
+\begin{equation*}  
+\qquad\qquad\begin{array}{l}
+\inference[(mem-val)~]
+  { \forall i\in\{0\ldots k\}.(\bnd{mem}(m, s), \bndclr{v_i}, \blbl{arg}(i)) \in E & \bndclr{v_i} \Downarrow p_i 
+    & p_0.m(p_1, \ldots, p_k) \rightsquigarrow p }
+  { \bnd{mem}(m, s) \Downarrow p }
+\\  
+\\
+\inference[(mem-expr)~]
+  { \forall i\in\{0\ldots k\}.(\bnd{mem}(m, s), \bndclr{v_i}, \blbl{arg}(i)) \in E & \exists j\in\{0\ldots k\}.\bndclr{v_j}\diagup\hspace{-1em}\Downarrow p_j
+   & \bndclr{v_i} \Downarrow_{\ident{lift}} \llbracket e_i \rrbracket_{\Gamma_i}  }
+  { \bnd{mem}(m, s) \Downarrow \llbracket e_0.m(e_1, \ldots, e_k) \rrbracket_{\Gamma_0,\ldots,\Gamma k} }
+\\
+\end{array}
+\end{equation*}
+
+\caption{Rules that define evaluation of previews over a dependency graph for an expression}  
+\label{fig:eval}
+\end{figure*}
+
+% --------------------------------------------------------------------------------------------------
 
@@
 \section{Evaluating previews}
-\newpage
+\label{sec:previews}
+
+The mechanism for constructing dependency graphs defined in Section~\ref{sec:formal} makes it
+possible to provide live previews when editing code without recomputing the whole program 
+each time the source code changes.
+
+The nodes in the dependency graph correspond to individual operations that will be performed
+when running the program. When the dependencies of an operation do not change while editing 
+code, the subsequent dependency graph will reuse a node used to represent the operation.
+
+Our live editor keeps a map from graph nodes to live previews, so a new preview only needs to be
+computed when a new node appears in the dependency graph (and the user moves the cursor to a 
+code location that corresponds to the node). This section describes how previews are evaluated.
+
+\paragraph{Previews and delayed previews.}
+As discussed in Section~\ref{sec:formal}, the body of a function cannot be easily evaluated to
+a value if it uses the bound variable. We do not attempt to ``guess'' possible arguments and,
+instead, provide a full preview only for sub-expressions with free variables bound by a let binding.
+For a function body that uses the bound variable, we obtain a \emph{delayed preview}, which is
+an expression annotated with a list of variables that need to be provided before the expression
+can be evaluated. We use the following notation:
+%
+\begin{equation*}
+\begin{array}{rcll}
+p&\hspace{-0.25em}\in\hspace{-0.25em}&n~|~\lambda x\rightarrow e&\quad(\textit{Fully evaluated previews})\\
+d&\hspace{-0.25em}\in\hspace{-0.25em}&p~|~\llbracket e \rrbracket_\Gamma&\quad(\textit{Evaluated and delayed previews})\\
+\end{array}
+\end{equation*}
+%
+Here, $p$ ranges over fully evaluated values. It can be either a primitive value (such as number,
+string or an object) or a function value with no free variables. A possibly delayed preview $d$ 
+can then be either evaluated preview $p$ or an expression $e$ that requires variables $\Gamma$.
+For simplicity, we use an untyped language and so $\Gamma$ is a list of variables $x_1, \ldots, x_n$.
+
+\paragraph{Evaluation and splicing.}
+In this paper, we omit the specifics of the underlying programming langauge and we focus on the
+live coding mechanism. However, we assume that the language is equipped with an evaluation 
+reduction $e \rightsquigarrow p$ that reduces a closed expression $e$ into a value $p$.
+
+For delayed previews, we construct a delayed expression using splicing. For example, assuming
+we have a delayed previews $\llbracket e_0 \rrbracket_x$ and $\llbracket e_1 \rrbracket_y$. 
+If we need to invoke a member $m$ on $e_0$ using $e_1$ as an argument, we construct a new 
+delayed preview $\llbracket e_0.m(e_1) \rrbracket_{x, y}$. This operation is akin to expression
+splicing from meta-programming [X,Y] and can be more formally captured by Contextual Modal Type 
+Theory (CMTT) as outlined below.
+
+\paragraph{Evaluation of previews.}
+The evaluation of previews is defined in Figure~\ref{fig:eval}. Given a dependency graph $(V, E)$,
+we define a relation $\bndclr{v}\Downarrow d$ that evaluates a sub-expression corresponding to 
+the node $\bndclr{v}$ to a (possibly delayed) preview $d$.
+
+The auxiliary relation $\bndclr{v}\Downarrow_{\ident{lift}} d$ always evaluates
+to a delayed preview. If the ordinary evaluation returns a delayed preview, so does the auxiliary
+relation (\emph{lift-expr}). If the ordinary evaluation returns a value, the value is wrapped
+into a delayed preview requiring no variables (\emph{lift-prev}).
+
+Graph node representing a value is evaluated to a value (\emph{val}) and a graph node representing
+an unbound variable is reduced to a delayed preview that requires the variable and returns its
+value (\emph{var}).
+
+For member access, we distinguish two cases. If all arguments evaluate to values (\emph{member-val}),
+then we use the evaluation relation $\rightsquigarrow$, immediately evaluate the member access and 
+produce a value. If one of the arguments is delayed (\emph{member-expr}), because the member access 
+is in the body of a lambda function, then we produce a delayed member access experession that
+requires the union of the variables required by the individual arguments.
+
+Finally, the evaluation of function values is similar, but requires three cases. If the body can 
+be reduced to a value with no unbound variables (\emph{fun-val}), we return a lambda function that
+returns the value. If the body requires only the bound variable (\emph{fun-bind}), we return a 
+lambda function with the delayed preview as the body. If the body requires further variables,
+the result is a delayed preview containing a function.
+
+\paragraph{Caching previews.}
+For simplicity, the relation $\Downarrow$ in Figure~\ref{fig:eval} does not specify how previews 
+are cached and linked to graph nodes. In practice, this is done by maintaining a lookup table
+$\bndclr{v} \rightarrow p$ from graph nodes $\bdnclr{v}$ to (possibly delayed) previews $p$.
+Whenever the $\Downarrow$ relation is used to obtain a preview for a graph node, we first 
+attempt to find an already evaluated preview using the lookup table. If the preview has not
+been previously evaluated, we evaluate it and add it to the lookup table.
+
+This means that evaluated previews can be reused in two ways. First, multiple nodes can 
+depend on one sub-graph in a single dependency graph (e.g.~if the same sub-expression appears
+twice in the program). Second, the keys of the lookup table are graph nodes and nodes are 
+reused when a new dependency graph is constructed after the user edits the source code.
+
+\paragraph{Semantics of delayed previews.}
+The focus of this paper is on the design and implementation of a live coding environment, 
+but it is worth noting that the structure of delayed previews is closely linked to the work
+on Contextual Modal Type Theory (CMTT) [X] and comonads [Y].
+
+In CMTT, $\lbrack \Psi \rbrack A$ denotes that a proposition $A$ is valid in context $\Psi$,
+which is closely related to our delayed preivews written as $\llbracket A \rrbracket_\Psi$.
+CMTT defines rules for composing context-dependent propositions that would allow us to express
+the splicing operation used in (\emph{mem-expr}). In categorical terms, the context-dependent
+proposition can be modelled as an indexed comonad [X]. The evaluation of a preview with no 
+context dependencies (built implicitly into our evaluation rules) corresponds to the counit
+operation of a comonad and would be explicitly written as $\llbracket A \rrbracket_\emptyset \rightarrow A$.
+
+% ==================================================================================================
 
@@
 \newpage
+
+Something about how dependency graph bits are reused correctly, i.e. if there is a source code
+change that can change the result, the graph will change.
+
+Something about various code edit operations that will not cause recomputation.
 

### 2017-10-03 live/paper.tex
Properties of previews

@@
 \usepackage{subcaption}
+
+\newcounter{thc}
+\newcounter{dfc}
+
+\theoremstyle{plain}
+\newtheorem{theorem}[thc]{Theorem}
+\newtheorem{lemma}[thc]{Lemma}
+
+\theoremstyle{definition}
+\newtheorem{definition}[dfc]{Definition}
 
@@
 
+\newpage
 \section{Properties}
-\newpage
+
+\subsection{Host language}
+
+We require that the host language allows eliminating let bindings without changing the meaning
+of the program. That is
+
+\begin{definition}[Host language reduction]
+The reduction $\rightsquigarrow$ reduces host language expression such that:
+\begin{itemize}
+\item  
+Given an expression $e$ and $e'$ such that $e'$ is obtained from $e$
+by replacing any sub-expression $\kvd{let}~x=e_1~\kvd{in}~e_2$ with the expression 
+$e_2[x\leftarrow e_1]$, written as $e\rightsquigarrow_\ident{let}e'$, then if $e \rightsquigarrow e''$ then also $e' \rightsquigarrow e''$.
+
+\item 
+Compositionality, i.e. if $e \rightsquigarrow e'$ and $C[e] \rightsquigarrow e''$ then also
+  $C[e'] \rightsquigarrow e''$.
+\end{itemize}
+\end{definition}
+
+We don't care about non-termination here.
+(so, does $\rightsquigarrow$ do small step or big step reduction?)
+
+\subsection{Correctness of previews}
+Eliminating let bindings does not change the result in the host language. It also yields the same
+dependency graph.
+
+\begin{lemma}[Let elimintion]
+Given an expression $e_1$ such that $e_1 \rightsquigarrow_\ident{let} e_2$ and a lookup table $\Delta$
+then if $\bndclr{v_1}, (V_1, E_1) = \ident{bind}_{\emptyset, \Delta}(e_1)$ and  
+$\bndclr{v_2}, (V_2, E_2) = \ident{bind}_{\emptyset, \Delta}(e_2)$ then it holds that
+$\bndclr{v_1} = \bndclr{v_2}$ and also $(V_1, E_1) = (V_2, E_2)$.
+\end{lemma}
+
+This simplifies the correctness proof as we only need to consider expressions not containing
+let bindings (if there were let bindings, both the result and the preview would be equivalent
+to one that would be obtained if the let bindings were elimninated)
+
+Now, if we have any let-free expression, we show that the preview obtained via the dependency graph
+is a value that is the same as the one we'd get by directly evaluating the expression.
+
+\begin{lemma}[Let-free correctness]
+Given an expression $e$ that has no free variables and does not contain let bindnigs, together with
+a lookup table $\Delta$ obtained from any sequence of expressions according to Figure~\ref{fig:loop}, 
+let $\bndclr{v}, (V, E) = \ident{bind}_{\emptyset, \Delta}(e)$. If $\bndclr{v}\Downarrow d$ 
+over a graph $(V, E)$ then $d = p$ for some $p$ and $e \rightsquigarrow p$.
+\end{lemma}
+\begin{proof}
+Every dependency graph node $\bndclr{v}$ corresponds to an expression $e$ with free variables
+$FV(e)$. We obtain preview $\bndclr{v}\Downarrow d$ such that if $FV(e) = \emptyset$ then
+$d = p$ for some $p$, otherwise, $d = \llbracket e' \rrbracket_{FV(e)}$.  
+We show that either $e \rightsquigarrow p$ or that for any $C$ such that $FV(C[e]) = \emptyset$
+it holds that $C[e] \rightsquigarrow e''$ and $C[e'] \rightsquigarrow e''$.
+By induction over the structure of the evaluation:
+
+\vspace{0.5em}\noindent\emph{(val)} $\bnd{val}(n) \Downarrow n$ and so $p=n$. The only
+  \ident{bind} case that generates $\bnd{val}(n)$ in the absence of let bindings is~(4) and
+  so $e = n$.
+
+\vspace{0.5em}\noindent\emph{(mem-val)} $\bnd{mem}(m, s) \Downarrow p$. By induction, there
+  are nodes $\bndclr{v_i}$ representing expressions $e_i$ and previews $\bndclr{v_i}\Downarrow p_i$
+  such that $e_i \rightsquigarrow p_i$. The only \ident{bind} case that generates the $\bnd{mem}(m, s)$
+  node in the absence of let binding is~(2) and so $e=e_0.m(e_1, \ldots, e_n)$. From 
+  (\emph{mem-val}), $p_0.m(p_1, \ldots, p_n) \rightsquigarrow p$ and from the compositionality 
+  of $\rightsquigarrow$, it is the case that $e \rightsquigarrow p$.
+
+\end{proof}
+
+\begin{theorem}[Correctness]
+(Same as let-free version, but using let elimination to say that this works for any expression)
+\end{theorem}
+
+\subsection{Reuse of previews}
+
+* Inlining let leads to an equivalent dependency graph
+
+* Given two expressions, if we inline lets and they are structurally different, then 
+  their dependency graphs are structurally different
 

### 2017-10-25 live/paper.tex
Abstract and introduction

@@
 \theoremstyle{plain}
+\newtheorem{lem}[thc]{Lemma}
 \newtheorem{theorem}[thc]{Theorem}
-\newtheorem{lemma}[thc]{Lemma}
 
@@
 
-\title{Design and implementaiton of a live coding environment}
+\title{Design and implementaiton of a live coding environment for data science}
 
@@
 \begin{abstract}
-Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+Data science can be done by directly manipulating data using spreadsheets, or by writing 
+data manipulation scripts using a programming language. The former is error-prone and
+does not scale, while the latter requires expert skills. Live coding has the potential to 
+bridge this gap and make writing of transparent, reproducible scripts more accessible.
 
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+In this paper, we describe a live programming environment for data science that provides instant 
+previews and contextual hints, while allowing the user to edit code in an unrestricted way in
+a text editor. This is challenging as the user can make arbitrary edits and we cannot fully 
+recompute previews after every change. The presented interpreter partially evalautes programs as 
+they are written and reuses results of previous evaluations, making it possible to provide instant 
+feedback during development.
+
+We formalise how programs are interpreted and how previews are computed, prove correctness
+of the previews and formally specify when can previews be reused. We hope this work will provide
+solid and easy to reuse foundations for the emerging trend of live programming environments.
 \end{abstract}
@@
 A live coding environment shows a preview of the list of dates \ident{top} and the programmer then
-changes the code by making $\num{10}$ a variable and changing the date format to see the full date:
-
+modifies the code by making the constant $\num{10}$ a variable and changing the date format to see 
+the full date:
+%
 \begin{equation*}
@@
 first two transformations (sorting and taking 10 elements) and only evaluate the last \ident{map}
-and format the release dates of already computed top 10 movies.
+to differently format the release dates of already computed top 10 movies.
 
@@
 no effect on the result) or ``change constant value'' (which forces recomputation of subsequent
-transformations). However, many programmers still prefer to edit programs as free form text.
-In this paper, we present the design and implementation of a live coding system that is capable
+transformations). However, many programmers prefer to edit programs as free form text.
+We present the design and implementation of a live coding system that is capable
 of reusing previously evaluated expressions as in the example above, yet, is integrated into an
-ordinary text editor.
+ordinary text editor. The main contributions of this paper are:
 
-The motivation for this paper is twofold. First, implementing a live programming system requires
-a different way of thinking about compilers and interpreters than the one presented in classic
-programming language literature. Second, an increasing number of programming environments 
-incorporate aspects of live programming [X,Y], yet relatively little has been written on their
-architecture and implementation.
+\begin{itemize}
+\item We introduce The Gamma (Section~\ref{sec:live}), a simple live coding environment for
+  data science. We review its design and implementation and explain how it 
+  bridges the gap between programming and spreadsheets.
+
+\item Implementing a live programming system requires different way of thinking about compilers
+  and interpreters than the one presented in classic programming language literature. Our 
+  formalisation (Section~\ref{sec:formal}) captures the essence of the new perspective.
+
+\item We formalise the evaluation of previews (Section~\ref{sec:previews}) and prove that our
+  evaluation and caching mechanism is produces correct previews (Section~\ref{sec:properties-correct}) 
+  and can effectively reuse partial results (Section~\ref{sec:properties-reuse}).
+
+\item In the final more speculative part (Section~\ref{sec:design}), we consider alternative
+  language designs that would enable further live coding experiences, which are difficult to
+  build using our current system.
+\end{itemize}
+
+\noindent
+We hope the architecture and its formal presentation in this paper can contribute important 
+foundations to the growing and important trend of text-based live coding environments.
 
@@
 \section{Live programming for data exploration}
+\label{sec:live}
 
@@
 \section{Properties}
+\label{sec:properties}
 
@@
 \subsection{Correctness of previews}
+\label{sec:properties-correct}
+
 Eliminating let bindings does not change the result in the host language. It also yields the same
@@
 
-\begin{lemma}[Let elimintion]
+\begin{lem}[Let elimintion]
 Given an expression $e_1$ such that $e_1 \rightsquigarrow_\ident{let} e_2$ and a lookup table $\Delta$
@@
 $\bndclr{v_1} = \bndclr{v_2}$ and also $(V_1, E_1) = (V_2, E_2)$.
-\end{lemma}
+\end{lem}
 
@@
 
-\begin{lemma}[Let-free correctness]
+\begin{lem}[Let-free correctness]
 Given an expression $e$ that has no free variables and does not contain let bindnigs, together with
@@
 over a graph $(V, E)$ then $d = p$ for some $p$ and $e \rightsquigarrow p$.
-\end{lemma}
+\end{lem}
 \begin{proof}
@@
 \subsection{Reuse of previews}
+\label{sec:properties-reuse}
 
@@
 
+\section{Takeaways}
+\label{sec:design}
+
 \newpage

### 2017-10-26 live/paper.tex
Updating informal gamma intro

@@
 \usepackage{subcaption}
+\usepackage{enumitem}
+\setlist{leftmargin=6mm}
 
@@
 previews and contextual hints, while allowing the user to edit code in an unrestricted way in
-a text editor. This is challenging as the user can make arbitrary edits and we cannot fully 
-recompute previews after every change. The presented interpreter partially evalautes programs as 
-they are written and reuses results of previous evaluations, making it possible to provide instant 
-feedback during development.
+a text editor.
+
+Using a text editor is challenging as any edit can significantly change the structure of code and
+fully recomputing previews after every change is too expensive. The presented interpreter partially 
+evalautes programs as they are written and reuses results of previous evaluations, making it 
+possible to provide instant feedback during development.
 
@@
 %
-A live coding environment shows a preview of the list of dates \ident{top} and the programmer then
+A live coding environment shows a preview of the list of dates. Next, assume that the programmer
 modifies the code by making the constant $\num{10}$ a variable and changing the date format to see 
@@
 
-This is not difficult to achieve if we represent the program in a structured way and allow the user
-to edit code only via well-understood primitive operations such as ``extract variable'' (which has
+This is not difficult if we represent the program in a structured way and allow the user
+to edit code only via primitive operations such as ``extract variable'' (which has
 no effect on the result) or ``change constant value'' (which forces recomputation of subsequent
 transformations). However, many programmers prefer to edit programs as free form text.
+
 We present the design and implementation of a live coding system that is capable
 of reusing previously evaluated expressions as in the example above, yet, is integrated into an
-ordinary text editor. The main contributions of this paper are:
+ordinary text editor. Our main contributions are:
 
-\begin{itemize}
+\begin{itemize}[itemsep=3pt]
 \item We introduce The Gamma (Section~\ref{sec:live}), a simple live coding environment for
@@
 
-\item In the final more speculative part (Section~\ref{sec:design}), we consider alternative
+\item In more speculative conclusions (Section~\ref{sec:design}), we consider alternative
   language designs that would enable further live coding experiences, which are difficult to
@@
 \begin{figure*}
-\includegraphics[scale=0.45]{scrape.png}
-\caption{xxx}  
+\includegraphics[scale=0.21]{wiki.png}
+\includegraphics[scale=0.21]{drop.png}
+\caption{Scraping 2016 science events. Preview of the Wikipedia source page (left) and live filtering (right).}
+\label{fig:thegamma}
 \end{figure*}
-
-\newpage
 
@@
 
-The Gamma introduction and what this tries to do. See pretty screenshot.
+% https://cacm.acm.org/blogs/blog-cacm/169199-data-science-workflow-overview-and-challenges/fulltext
 
-This is what we care about - it's not live programming for your typical haskell,
-but for R or Python.
+The Gamma aims to make basic data exploration accessible to non-experts, such as data journalists [X], who
+need to use data transparently. As such, we focus on simple scripts that can be written using
+tools such as Jupyter notebooks~[X]. Those scripts follow a typical data science workflow [X];
+they acquire and reformat data, run a number of analyses using the clean data and then create 
+several visualizations.
 
-\newpage
-~
+An important property of data science workflow is that we work with readily available concrete data.
+Data scientists load inputs into memory and refer to it in subsequent REPL interactions. They might 
+later wrap completed working code into a function (and run on multiple datasets), but 
+do not start with functions. We reflect this pattern in our language.
+
+In this section, we provide a brief overview of The Gamma, a simple live coding environment for
+data science. We discuss the language, type providers and the user interface,
+before focusing on the algorithms behind live previews in Section~\ref{sec:formal}.
+
+\subsection{The Gamma scripting language}
+Figure~\ref{fig:thegamma} shows The Gamma script that scrapes items from a Wikipedia page, 
+collects those marked as ``Events'' and filters them. The script illustrates two aspects of the
+scripting language used by The Gamma -- its structure and its use of type providers for
+dot-driven data exploration.
+
+The scripting language is not intended to be as expressive as, say, R or Python and so it
+has a very simple structure -- scripts are a sequence of let-bindings (that obtain or transform 
+data) or statements (that produce visualizations). This reflects the fact that we always work
+with concrete data and allows us to provide previews.
+
+The most notable limitation of The Gamma is that the scripting language does not support top-level 
+functions. This is not a problem for the simple scripts we consider, but it would be an issue for
+a general-purpose language. We discuss potential design for functions that would still be based
+on working with concrete data in Section~\ref{sec:design}.
+
+\subsection{Dot-driven data exploration}
+
+As ilustrated by the second let binding in Figure~\ref{fig:thegamma}, many opertaions
+in The Gamma can be expressed using member access via dot. The underlying mechanism is based
+on type providers [X,Y] and the specific type provider used in this example has been described 
+elsewhere [X].
+
+Given a data source (scraped Wikipedia page), the type provider generates type with members that allow
+a range of transformations of the data such as grouping, sorting and filtering. Some of the members
+are based only on the schema (e.g.~\ident{\textquotesingle Item Contains\textquotesingle} or 
+\ident{\textquotesingle Drop Heading 2\textquotesingle}), but some may also be generated based
+on (a sample of) the dataset (e.g.~the second member in \ident{\textquotesingle Heading 2 
+is\textquotesingle.\textquotesingle Events[edit]\textquotesingle}).
+
+What matters for the purpose of this paper is 
+the fact that most operations are expressed via member access matters. First, it means 
+that we need to provide live previews for sub-expressions formed by a chain of member accesses.
+Second, it means that the result of any member access expression depends on the instance and,
+possibly, on the parameters provided for a method call.
+
+\subsection{Direct manipulation and live previews}
+The screenshot in Figure~\ref{fig:thegamma} show the editor as implemented in The Gamma. This
+includes live previews (discussed in this paper), but also an editor that provides spreadsheet-like
+interface for editing the data transformation script.
+
+In our implementation, previews appears below the currently selected let binding.
+The preview can be of several types such as web page (left) or data table (right). In this paper,
+we describe how we evaluate scripts to obtain the resulting object and we ignore how such objects
+are rendered.
+
+The Figure~\ref{fig:thegamma} also shows a special handling of expressions constructed using the
+pivot type provider. The editor recognises individual data transformations and provides a simple
+user interface for adding and removing transformations and changing their parameters that changes
+the source code accordingly. This aspect is not discussed in the present paper.
 
@@
 
+\section{Related and future}
+\label{sec:future}
+
+Integrate with Python/R via Jupyter
+
 \newpage

### 2017-11-06 live/paper.tex
Properties section

@@
 \usepackage{enumitem}
+\usepackage[shortlabels]{enumerate}
 \setlist{leftmargin=6mm}
@@
 \\[0.75em]
-\ident{bind}_{\Gamma, \Delta}(x) = \bndclr{v}, (\{ \bndclr{v} \}, \emptyset)\quad \textnormal{when}~\bndclr{v} = \Gamma(x)&(6)
+\ident{bind}_{\Gamma, \Delta}(x) = \bndclr{v}, (\{ \bndclr{v} \}, \emptyset)\quad \textnormal{when}~\bndclr{v} = \Gamma(x)&(5)
 \\[1em]
-\ident{bind}_{\Gamma, \Delta}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V, \{e\} \cup E)&(7)\\
+\ident{bind}_{\Gamma, \Delta}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V, \{e\} \cup E)&(6)\\
 \quad \textnormal{when}~\Gamma_1 = \Gamma \cup \{ x, \bnd{var}(x) \}\\
@@
 \\[1em]
-\ident{bind}_{\Gamma, \Delta}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V, \{e\} \cup E)&(8)\\
+\ident{bind}_{\Gamma, \Delta}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V, \{e\} \cup E)&(7)\\
 \quad \textnormal{when}~\Gamma_1 = \Gamma \cup \{ x, \bnd{var}(x) \}\\
@@
 \begin{array}{l}
-\Delta_{i}(\bknd{mem}(m), [(\bndclr{v_0},\blbl{arg}(0)),\ldots, (\bndclr{v_n},\blbl{arg}(n))]) = \bndclr{v}\\
+\Delta_{i}(\bknd{mem}(m), [(\bndclr{v_0},\blbl{arg}(0)),\ldots, (\bndclr{v_n},\blbl{arg}(n))]) = \bnd{mem}(m, s)\\
 \quad \textnormal{for all}~\bnd{mem}(m, s) \in V\\
@@
 \\[0.75em]
-\Delta_{i}(\bknd{fun}(x), [(\bndclr{v_1},\blbl{body})]) = \bndclr{v}\\
+\Delta_{i}(\bknd{fun}(x), [(\bndclr{v_1},\blbl{body})]) = \bnd{fun}(x, s)\\
 \quad \textnormal{for all}~\bnd{fun}(x, s) \in V\\
@@
 \newpage
-\section{Properties}
+\section{Properties of preview evaluation}
 \label{sec:properties}
 
+The dependency graph makes it possible to cache partial results when evaluating previews. 
+The construction of dependency graphs and evaluation of previews needs to have two main properties.
+First, if we evaluate a preview using dependency graph with caching, it should be the same as the
+value we would obtain by evaluating the expression directly. Second, the evaluation of previews 
+using dependency graphs should -- in some cases -- reuse previously evaluated partial results.
+In other words, this section shows that our mechanism is both correct and implements a useful 
+optimization.
+
 \subsection{Host language}
+The method in this paper does not rely on the particulars of how the underlying programming 
+language works. For this reason, we do not give a concrete definition of how member access
+$e.m(e_1,\ldots, e_n)$ reduces to a value.
 
@@
 of the program. That is
+
+[Rightsquigarrow allows beta anywhere? then it works...]
 
@@
   $C[e'] \rightsquigarrow e''$.
+
+\item
+Extensionality -- if $C1[e1] \rightsquigarrow e$ and $C1[e2] \rightsquigarrow e$ then $C2[fn e1] \rightsquigarrow ee$ and $C2[fn e2] \rightsquigarrow ee$
+
 \end{itemize}
@@
 \label{sec:properties-correct}
+
+First is corrctness (Theorem~\ref{thm:correcntess}) -- no matter how a graph is constructed, when 
+we use it to evaluate a preview for an expression, the preview is the same as the value we would 
+obtain by evaluating the expression directly. Second is determinacy (Theorem~\ref{thm:determinacy}) --
+if we cache a preview for a graph node and update the graph, the preview would still be the same
+when evaluated using an updated graph.
+
+This is the correctness and determinacy property discussed in Section~\ref{sec:properties-correct}.
 
@@
 
-\begin{lem}[Let-free correctness]
-Given an expression $e$ that has no free variables and does not contain let bindnigs, together with
-a lookup table $\Delta$ obtained from any sequence of expressions according to Figure~\ref{fig:loop}, 
+xxx
+
+\begin{theorem}[Determinacy]
+\label{thm:determinacy}
+Downarrow will always evaluate the same, if we add things to a graph in the defined way.
+(this means that we can cache)
+\end{theorem}
+
+~
+
+~
+
+~
+
+[Only dependency graph matters, not the history and symbols]
+[Symbols are effectively hashes of dependencies]
+
+\begin{lemma}
+\label{thm:lemma-lookup}
+Lemma X - lookup with red mem(m) returns blue mem(m, s)
+\end{lemma}
+
+\begin{theorem}[Let-free correctness]
+Given an expression $e$ that has no free variables and does not contain let bindings, together
+with a lookup table $\Delta$ obtained from any sequence of expressions according to Figure~\ref{fig:loop}
 let $\bndclr{v}, (V, E) = \ident{bind}_{\emptyset, \Delta}(e)$. If $\bndclr{v}\Downarrow d$ 
 over a graph $(V, E)$ then $d = p$ for some $p$ and $e \rightsquigarrow p$.
-\end{lem}
+\end{theorem}
 \begin{proof}
-Every dependency graph node $\bndclr{v}$ corresponds to an expression $e$ with free variables
-$FV(e)$. We obtain preview $\bndclr{v}\Downarrow d$ such that if $FV(e) = \emptyset$ then
-$d = p$ for some $p$, otherwise, $d = \llbracket e' \rrbracket_{FV(e)}$.  
-We show that either $e \rightsquigarrow p$ or that for any $C$ such that $FV(C[e]) = \emptyset$
-it holds that $C[e] \rightsquigarrow e''$ and $C[e'] \rightsquigarrow e''$.
-By induction over the structure of the evaluation:
+First note that, when combining recursively constructed sub-graphs, the \ident{bind} operation
+adds new nodes and edges leading from those new nodes. Therefore, an evaluation using $\Downarrow$
+over a sub-graph will also be valid over the new graph.
 
-\vspace{0.5em}\noindent\emph{(val)} $\bnd{val}(n) \Downarrow n$ and so $p=n$. The only
-  \ident{bind} case that generates $\bnd{val}(n)$ in the absence of let bindings is~(4) and
-  so $e = n$.
-
-\vspace{0.5em}\noindent\emph{(mem-val)} $\bnd{mem}(m, s) \Downarrow p$. By induction, there
-  are nodes $\bndclr{v_i}$ representing expressions $e_i$ and previews $\bndclr{v_i}\Downarrow p_i$
-  such that $e_i \rightsquigarrow p_i$. The only \ident{bind} case that generates the $\bnd{mem}(m, s)$
-  node in the absence of let binding is~(2) and so $e=e_0.m(e_1, \ldots, e_n)$. From 
-  (\emph{mem-val}), $p_0.m(p_1, \ldots, p_n) \rightsquigarrow p$ and from the compositionality 
-  of $\rightsquigarrow$, it is the case that $e \rightsquigarrow p$.
-
+Next, we prove a more general property using induction showing that for $e$ such 
+that $\bndclr{v}, (V, E) = \ident{bind}_{\emptyset, \Delta}(e)$:
+\begin{enumerate}
+\item[a.] If $FV(e)=\emptyset$ then $\bndclr{v} \Downarrow p$ for some $p$ and $e \rightsquigarrow p$
+\item[b.] Otherwise, $\bndclr{v} \Downarrow \llbracket e_p \rrbracket_{FV(e)}$ for some $e_p$ and
+  for any evaluation context $C[-]$ such that $FV(C[e_p])=\emptyset$ it holds that if
+  $C[e_p] \rightsquigarrow v$ for some $v$ then $C[e] \rightsquigarrow v$.  
+\end{enumerate}
+The proof is done by induction over the binding process, which follows the structure 
+of the expression $e$ and can be found in Appendix~\ref{sec:app-correctness}.
 \end{proof}
@@
 \begin{theorem}[Correctness]
+\label{thm:correctness}
 (Same as let-free version, but using let elimination to say that this works for any expression)
@@
 
+\appendix 
+\section{Appendix}
+\label{sec:app-correctness}
+\begin{theorem}[Let-free correctness]
+Given an expression $e$ that has no free variables and does not contain let bindings, together
+with a lookup table $\Delta$ obtained from any sequence of expressions according to Figure~\ref{fig:loop}
+let $\bndclr{v}, (V, E) = \ident{bind}_{\emptyset, \Delta}(e)$. If $\bndclr{v}\Downarrow d$ 
+over a graph $(V, E)$ then $d = p$ for some $p$ and $e \rightsquigarrow p$.
+\end{theorem}
+\begin{proof}
+First note that, when combining recursively constructed sub-graphs, the \ident{bind} operation
+adds new nodes and edges leading from those new nodes. Therefore, an evaluation using $\Downarrow$
+over a sub-graph will also be valid over the new graph.
+
+Next, we prove a more general property using induction showing that for $e$ such 
+that $\bndclr{v}, (V, E) = \ident{bind}_{\emptyset, \Delta}(e)$:
+\begin{enumerate}
+\item[a.] If $FV(e)=\emptyset$ then $\bndclr{v} \Downarrow p$ for some $p$ and $e \rightsquigarrow p$
+\item[b.] Otherwise, $\bndclr{v} \Downarrow \llbracket e_p \rrbracket_{FV(e)}$ for some $e_p$ and
+  for any evaluation context $C[-]$ such that $FV(C[e_p])=\emptyset$ it holds that if
+  $C[e_p] \rightsquigarrow v$ for some $v$ then $C[e] \rightsquigarrow v$.  
+\end{enumerate}
+
+The proof is done by induction over the binding process, which follows the structure 
+of the expression $e$:
+
+\vspace{0.75em}\noindent(1) $\ident{bind}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n))$ -- 
+  Here $e = e_0.m(e_1, \ldots, e_n)$, $\bndclr{v_i}$ are graph nodes obtained by induction for
+  expressions $e_i$ and $\{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v}, \bndclr{v_n}, \blbl{arg}(n))\} \subseteq E$
+  From Lemma~\ref{thm:lemma-lookup}, $\bndclr{v} = \bnd{mem}(m, s)$ for some $s$.
+
+  If $FV(e)=\emptyset$, then $\bndclr{v_i} \Downarrow p_i$ for $i\in 0\ldots n$ and   
+  $\bndclr{v}\Downarrow p$ using (\emph{mem-val}) such that $p_0.m(p_1, \ldots, p_n) \rightsquigarrow p$. 
+  From induction hypothesis, $e_i \rightsquigarrow p_i$ and so, using compositionality of $\rightsquigarrow$, 
+  $e_0.m(e_1, \ldots, e_n) \rightsquigarrow p$.
+
+  If $FV(e)\neq\emptyset$, then $\bnclr{v_i} \Downarrow_\ident{lift} \llbracket e'_i \rrbracket$ for $i\in 0\ldots n$ and 
+  $\bndclr{v}\Downarrow \llbracket e'_0.m(e'_1, \ldots, e'_n) \rrbracket_{FV(e)}$ using (\emph{mem-expr}).
+  From induction hypothesis, for any $C[-]$, if $C[e'_i] \rightsquigarrow v_i$ then $C[e_i] \rightsquigarrow v_i$.
+  Using compositionality, it also holds that for any $C[-]$, if $C[e'_0.m(e'_1, \ldots, e'_n)]\rightsquigarrow v$
+  then also $C[e_0.m(e_1, \ldots, e_n)] \rightsquigarrow v$.
+
+\vspace{0.75em}\noindent(2) $\ident{bind}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n))$ --
+  This case is similar to (1), except that the fact that $\bndclr{v} = \bnd{mem}(m, s)$
+  holds by construction, rather than using Lemma~\ref{thm:lemma-lookup}.
+
+\vspace{0.75em}\noindent(3) We assume that the expression $e$ does not include let bindings and
+  so this case never happens.
+
+\vspace{0.75em}\noindent(4) $\ident{bind}_{\Gamma, \Delta}(n)$ -- In this case $e=n$ and $\bndclr{v} = \bnd{val}(n)$. The preview of
+  $\bnd{val}(n)$ is evaluated to $n$ using the (\emph{val}) case.
+
+\vspace{0.75em}\noindent(5) $\ident{bind}_{\Gamma, \Delta}(x)$ -- The initial $\Gamma$ is empty and
+  there are no let bindings, so $x$ must have been added to $\Gamma$ by case (6) or (7). Hence,
+  $\bndclr{v}=\bnd{var}(x)$. Using (\emph{var}) $\bndclr{v}\Downarrow \llbracket x \rrbracket_{x}$
+  and so $e_p = e = x$ and the second case (b.) trivially holds.
+
+\vspace{0.75em}\noindent(6) $\ident{bind}_{\Gamma, \Delta}(\lambda x\rightarrow e)$ -- Assume
+  $\bndclr{v_b}$ is a graph node representing the body. The evaluation can use one of three rules:
+
+  If $FV(e)=\emptyset$ then $\bndclr{v_b}\Downarrow p_b$ for some $p_b$ and $\bndclr{v}\Downarrow \lambda x.p_b$ 
+  using (\emph{fun-val}). From induction $e_b \rightsquigarrow p_b$
+  and so, using extensionality of $\rightsquigarrow$ also $\lambda x.e_b \rightsquigarrow \lambda x.p_b$.
+
+  If $FV(e)=\{x\}$ then $\bndclr{v_b}\Downarrow \llbracket e_b \rrbracket_x$ for some $e_b$ and
+  $\bndclr{v}\Downarrow \lambda x.e_b$ using (\emph{fun-bind}). From 
+  induction, for any $C[-]$, if $C[e_b] \rightsquigarrow v$ then also $C[e] \rightsquigarrow v$.
+  By extensionality, using an empty context $C[-]=-$ it holds that if $\lambda x.e \rightsquigarrow \lambda x.e_b$.
+
+  Otherwise, $\bndclr{v_b}\Downarrow \llbracket e_b \rrbracket_{x,\Gamma}$ for some $e_b$ and
+  $\bndclr{v}\Downarrow \llbracket\lambda x.e_b\rrbracket_\Gamma$ using (\emph{fun-expr}). From
+  induction, for any $C[-]$, if $C[e_b] \rightsquigarrow v$ then also $C[e] \rightsquigarrow v$.
+  By extensionality, if $C'[\lambda x.e_b] \rightsquigarrow v$ then also
+  $C'[\lambda x.e] \rightsquigarrow v$.
+\end{proof}
+
 \end{document}

### 2017-11-07 live/paper.tex
Tweaking proofs section

@@
 
-This means that evaluated previews can be reused in two ways. First, multiple nodes can 
-depend on one sub-graph in a single dependency graph (e.g.~if the same sub-expression appears
+The evaluated previews can be reused in two ways. First, multiple nodes can 
+depend on one sub-graph in a single dependency graph (if the same sub-expression appears
 twice in the program). Second, the keys of the lookup table are graph nodes and nodes are 
@@
 
-\newpage
 \section{Properties of preview evaluation}
@@
 The dependency graph makes it possible to cache partial results when evaluating previews. 
-The construction of dependency graphs and evaluation of previews needs to have two main properties.
+The mechanism needs to satisfy two properties.
 First, if we evaluate a preview using dependency graph with caching, it should be the same as the
@@
 using dependency graphs should -- in some cases -- reuse previously evaluated partial results.
-In other words, this section shows that our mechanism is both correct and implements a useful 
-optimization.
+In other words, we show that the mechanism is correct and implements a useful optimization.
 
-\subsection{Host language}
-The method in this paper does not rely on the particulars of how the underlying programming 
-language works. For this reason, we do not give a concrete definition of how member access
-$e.m(e_1,\ldots, e_n)$ reduces to a value.
+\subsection{Modelling expression evaluation}
+In The Gamma language, computations are expressed using
+member access, written as $e.m(e_1,\ldots, e_n)$. In this paper, we do not define how member
+access evaluates. This has been done elsewhere [X], but more importantly, the evaluation of
+previews does not rely on the exact specifics of the evaluation, provided that the language
+satisfies certan basic conditions. The following definitions provides the necessary structure 
+for discussing correctness of previews.
 
-We require that the host language allows eliminating let bindings without changing the meaning
-of the program. That is
+Partial evaluation may reduce an expression under $\lambda$-abstraction. We do not require that
+the reduction of the host language allows this. Instead, we define an extended reduction relation
+and use that in the proofs. The host language only needs to compose well with such extended reduction
+as captured by the \emph{compositionality} property below. We also require that the language allows
+elimination of let bindings.
 
-[Rightsquigarrow allows beta anywhere? then it works...]
+\begin{definition}[Host language]
+Given a relation on expressions $e_1 \rightsquigarrow e_2$ that models evaluation, we define:
 
-\begin{definition}[Host language reduction]
-The reduction $\rightsquigarrow$ reduces host language expression such that:
 \begin{itemize}
-\item  
-Given an expression $e$ and $e'$ such that $e'$ is obtained from $e$
-by replacing any sub-expression $\kvd{let}~x=e_1~\kvd{in}~e_2$ with the expression 
-$e_2[x\leftarrow e_1]$, written as $e\rightsquigarrow_\ident{let}e'$, then if $e \rightsquigarrow e''$ then also $e' \rightsquigarrow e''$.
+\item[--]  
+ A \emph{preview evaluation context} (also referred to as \emph{context}):
 
-\item 
-Compositionality, i.e. if $e \rightsquigarrow e'$ and $C[e] \rightsquigarrow e''$ then also
-  $C[e'] \rightsquigarrow e''$.
+\vspace{-1.25em}
+\begin{equation*}
+\begin{array}{lcl}
+C[-] &\hspace{-1em}=\hspace{-1em}& \kvd{let}~x = -~\kvd{in}~e ~|~ \kvd{let}~x = e~\kvd{in}~- ~|~ \lambda x\rightarrow - \\
+     &&  e_0.m(e_1, \ldots, e_{k-1}, -, e_{k+1}, \ldots e_n)
+\end{array}
+\end{equation*}
 
-\item
-Extensionality -- if $C1[e1] \rightsquigarrow e$ and $C1[e2] \rightsquigarrow e$ then $C2[fn e1] \rightsquigarrow ee$ and $C2[fn e2] \rightsquigarrow ee$
+\item[--]  
+An extended reduction relation $\rightsquigarrow_\beta$ such that, for any context $C$, 
+$C[e_1] \rightsquigarrow_\beta C[e_2]$ whenever $e_1 \rightsquigarrow e_2$.
 
+\item[--]  
+Let elimination $\rightsquigarrow_\ident{let}$ such that, using  capture-avoiding substitution,
+  $C[\kvd{let}~x=e_1~\kvd{in}~e_2] \rightsquigarrow_\ident{let} C[e_2[x\leftarrow e_1]]$
+\end{itemize}
+
+\noindent
+We say that $\rightsquigarrow$ is a suitable \emph{host language reduction} if:
+\begin{itemize}
+\item[--] It satisfies the \emph{compositionality} property, that is if $e \rightsquigarrow e'$ and 
+$C[e]\rightsquigarrow_\beta e''$ then also $C[e']\rightsquigarrow_\beta e''$.
+
+\item[--] Let elimination does not affect the result, i.e.~if $e \rightsquigarrow_{\ident{let}} e'$ 
+and $e'\rightsquigarrow_\beta e''$ then also $e\rightsquigarrow_\beta e''$ 
 \end{itemize}
@@
 
-We don't care about non-termination here.
-(so, does $\rightsquigarrow$ do small step or big step reduction?)
+\noindent
+The host language in The Gamma is a simple call-by-value functional language without side-effects,
+and so it satisfies both compositionality and allows let bindings to be eliminated, although the
+latter affects the performance. The mechanism for preview evaluation presented here would also 
+work for call-by-name languages, but it would suffer from the expected difficulties in the 
+presence of side-effects or non-determinism.
 
@@
 let $\bndclr{v}, (V, E) = \ident{bind}_{\emptyset, \Delta}(e)$. If $\bndclr{v}\Downarrow d$ 
-over a graph $(V, E)$ then $d = p$ for some $p$ and $e \rightsquigarrow p$.
+over a graph $(V, E)$ then $d = p$ for some $p$ and $e \rightsquigarrow_\beta p$.
 \end{theorem}
@@
 \begin{enumerate}
-\item[a.] If $FV(e)=\emptyset$ then $\bndclr{v} \Downarrow p$ for some $p$ and $e \rightsquigarrow p$
-\item[b.] Otherwise, $\bndclr{v} \Downarrow \llbracket e_p \rrbracket_{FV(e)}$ for some $e_p$ and
+\item[a.] If $FV(e)\,=\,\emptyset$ then $\bndclr{v} \Downarrow p$ for some $p$ and $e \rightsquigarrow_\beta p$
+\item[b.] If $FV(e)\neq\emptyset$ then $\bndclr{v} \Downarrow \llbracket e_p \rrbracket_{FV(e)}$ for some $e_p$ and
   for any evaluation context $C[-]$ such that $FV(C[e_p])=\emptyset$ it holds that if
-  $C[e_p] \rightsquigarrow v$ for some $v$ then $C[e] \rightsquigarrow v$.  
+  $C[e] \rightsquigarrow_\beta C[e_p]$.  
 \end{enumerate}
+
 The proof is done by induction over the binding process, which follows the structure 
@@
 \appendix
+
 \section{Appendix}
 \label{sec:app-correctness}
+
+\setcounter{thc}{-1}
 \begin{theorem}[Let-free correctness]
@@
 \begin{enumerate}
-\item[a.] If $FV(e)=\emptyset$ then $\bndclr{v} \Downarrow p$ for some $p$ and $e \rightsquigarrow p$
-\item[b.] Otherwise, $\bndclr{v} \Downarrow \llbracket e_p \rrbracket_{FV(e)}$ for some $e_p$ and
+\item[a.] If $FV(e)\,=\,\emptyset$ then $\bndclr{v} \Downarrow p$ for some $p$ and $e \rightsquigarrow_\beta p$
+\item[b.] If $FV(e)\neq\emptyset$ then $\bndclr{v} \Downarrow \llbracket e_p \rrbracket_{FV(e)}$ for some $e_p$ and
   for any evaluation context $C[-]$ such that $FV(C[e_p])=\emptyset$ it holds that if
-  $C[e_p] \rightsquigarrow v$ for some $v$ then $C[e] \rightsquigarrow v$.  
+  $C[e] \rightsquigarrow_\beta C[e_p]$.  
 \end{enumerate}
@@
   $\bndclr{v}\Downarrow p$ using (\emph{mem-val}) such that $p_0.m(p_1, \ldots, p_n) \rightsquigarrow p$. 
-  From induction hypothesis, $e_i \rightsquigarrow p_i$ and so, using compositionality of $\rightsquigarrow$, 
-  $e_0.m(e_1, \ldots, e_n) \rightsquigarrow p$.
+  From induction hypothesis, $e_i \rightsquigarrow_\beta p_i$ and so, using compositionality of $\rightsquigarrow$, 
+  $e_0.m(e_1, \ldots, e_n) \rightsquigarrow_\beta p$.
 
   If $FV(e)\neq\emptyset$, then $\bnclr{v_i} \Downarrow_\ident{lift} \llbracket e'_i \rrbracket$ for $i\in 0\ldots n$ and 
-  $\bndclr{v}\Downarrow \llbracket e'_0.m(e'_1, \ldots, e'_n) \rrbracket_{FV(e)}$ using (\emph{mem-expr}).
-  From induction hypothesis, for any $C[-]$, if $C[e'_i] \rightsquigarrow v_i$ then $C[e_i] \rightsquigarrow v_i$.
-  Using compositionality, it also holds that for any $C[-]$, if $C[e'_0.m(e'_1, \ldots, e'_n)]\rightsquigarrow v$
-  then also $C[e_0.m(e_1, \ldots, e_n)] \rightsquigarrow v$.
+  using (\emph{mem-expr}), $\bndclr{v}\Downarrow \llbracket e'_0.m(e'_1, \ldots, e'_n) \rrbracket_{FV(e)}$.
+  From induction hypothesis, for any $C[-]$, it holds that $C[e_i] \rightsquigarrow_\beta C[e'_i]$.
+  Using compositionality, it also holds that for any $C[-]$, it is the case that
+  $C[e_0.m(e_1, \ldots, e_n)] \rightsquigarrow_\beta C[e'_0.m(e'_1, \ldots, e'_n)]$.
 
@@
   If $FV(e)=\emptyset$ then $\bndclr{v_b}\Downarrow p_b$ for some $p_b$ and $\bndclr{v}\Downarrow \lambda x.p_b$ 
-  using (\emph{fun-val}). From induction $e_b \rightsquigarrow p_b$
-  and so, using extensionality of $\rightsquigarrow$ also $\lambda x.e_b \rightsquigarrow \lambda x.p_b$.
+  using (\emph{fun-val}). From induction $e_b \rightsquigarrow_\beta p_b$
+  and so by definition also $\lambda x.e_b \rightsquigarrow_\beta \lambda x.p_b$.
 
@@
   $\bndclr{v}\Downarrow \lambda x.e_b$ using (\emph{fun-bind}). From 
-  induction, for any $C[-]$, if $C[e_b] \rightsquigarrow v$ then also $C[e] \rightsquigarrow v$.
-  By extensionality, using an empty context $C[-]=-$ it holds that if $\lambda x.e \rightsquigarrow \lambda x.e_b$.
+  induction, for any context $C[-]$, it holds that $C[e] \rightsquigarrow_\beta C[e_b]$.
+  Using a context $C[-]=\lambda x.-$ it holds that $\lambda x.e \rightsquigarrow_\beta \lambda x.e_b$.
 
@@
   $\bndclr{v}\Downarrow \llbracket\lambda x.e_b\rrbracket_\Gamma$ using (\emph{fun-expr}). From
-  induction, for any $C[-]$, if $C[e_b] \rightsquigarrow v$ then also $C[e] \rightsquigarrow v$.
-  By extensionality, if $C'[\lambda x.e_b] \rightsquigarrow v$ then also
-  $C'[\lambda x.e] \rightsquigarrow v$.
+  induction, for any context $C[-]$, it holds that $C[e] \rightsquigarrow_\beta C[e_b]$ 
+  and for any context $C'[-]$, by definition of $\rightsquigarrow_\beta$ also 
+  $C'[\lambda x.e] \rightsquigarrow C'[\lambda x.e_b]$.
 \end{proof}

### 2017-11-14 live/paper.tex
Finish correctness of previews

@@
 \begin{figure*}[t]
-\vspace{-0.5em}
 \begin{equation*}
@@
 \quad \textnormal{let}~E = \{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v}, \bndclr{v_n}, \blbl{arg}(n))\}
-\\[0.75em]
+\\[1em]
 \ident{bind}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n)) = &(2)\\ 
@@
 \quad \textnormal{let}~E = \{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v},\bndclr{v_n}, \blbl{arg}(n)) \}
-\\[0.75em]
+\\[1em]
 \ident{bind}_{\Gamma, \Delta}(\kvd{let}~x=e_1~\kvd{in}~e_2) = \bndclr{v}, (\{\bndclr{v}\} \cup V \cup V_1, E \cup E_1)&(3)\\
@@
 \ident{bind}_{\Gamma, \Delta}(n) = \bnd{val}(n), (\{ \bnd{val}(n) \}, \emptyset) }&(4)
-\\[0.75em]
+\\[1em]
 \ident{bind}_{\Gamma, \Delta}(x) = \bndclr{v}, (\{ \bndclr{v} \}, \emptyset)\quad \textnormal{when}~\bndclr{v} = \Gamma(x)&(5)
-\\[1em]
+\\[1.25em]
 \ident{bind}_{\Gamma, \Delta}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V, \{e\} \cup E)&(6)\\
@@
 \quad \textnormal{let}~e = (\bndclr{v}, \bndclr{v_0}, \blbl{body}) 
-\\[1em]
+\\[1.25em]
 \ident{bind}_{\Gamma, \Delta}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V, \{e\} \cup E)&(7)\\
@@
 \quad \textnormal{and}~(\bknd{fun}(x),[(\bndclr{v_0}, \blbl{body})])\notin \ident{dom}(\Delta)\\
-\quad \textnormal{let}~\bndclr{v} = \bnd{fun}(s, x), s~\textnormal{fresh}\\
+\quad \textnormal{let}~\bndclr{v} = \bnd{fun}(x, s), s~\textnormal{fresh}\\
 \quad \textnormal{let}~e = (\bndclr{v},\bndclr{v_0},\blbl{body}) 
-\\[0.5em]
-\end{array}
-\end{equation*}
-\vspace{-0.5em}
+\\[0.75em]
+\end{array}
+\end{equation*}
 \caption{Rules of the binding process, which constructs a dependency graph for an expression.}
@@
 \begin{figure}[!b]  
-\vspace{-0.5em}
 \begin{equation*}
 \begin{array}{l}
-\Delta_{i}(\bknd{mem}(m), [(\bndclr{v_0},\blbl{arg}(0)),\ldots, (\bndclr{v_n},\blbl{arg}(n))]) = \bnd{mem}(m, s)\\
-\quad \textnormal{for all}~\bnd{mem}(m, s) \in V\\
-\quad \textnormal{such that}~(\bnd{mem}(m, s), \bndclr{v_i}, \blbl{arg}(i)) \in E~\textnormal{for}~i\in 0,..,n
+\ident{update}_{V,E}(\Delta_{i-1}) = \Delta_i~\textnormal{such that:}
 \\[0.75em]
-\Delta_{i}(\bknd{fun}(x), [(\bndclr{v_1},\blbl{body})]) = \bnd{fun}(x, s)\\
-\quad \textnormal{for all}~\bnd{fun}(x, s) \in V\\
-\quad \textnormal{such that}~(\bnd{fun}(x, s), \bndclr{v_1}, \blbl{body}) \in E
+\quad\Delta_{i}(\bknd{mem}(m), [(\bndclr{v_0},\blbl{arg}(0)),\ldots, (\bndclr{v_n},\blbl{arg}(n))]) = \bnd{mem}(m, s)\\
+\quad\quad \textnormal{for all}~\bnd{mem}(m, s) \in V\\
+\quad\quad \textnormal{such that}~(\bnd{mem}(m, s), \bndclr{v_i}, \blbl{arg}(i)) \in E~\textnormal{for}~i\in 0,..,n
 \\[0.75em]
-\Delta_{i}(\bkndclr{v}) = \Delta_{i-1}(\bkndclr{v})\quad \textnormal{(otherwise)}
-\end{array}
-\end{equation*}
-\vspace{-0.5em}
+\quad\Delta_{i}(\bknd{fun}(x), [(\bndclr{v_1},\blbl{body})]) = \bnd{fun}(x, s)\\
+\quad\quad \textnormal{for all}~\bnd{fun}(x, s) \in V\\
+\quad\quad \textnormal{such that}~(\bnd{fun}(x, s), \bndclr{v_1}, \blbl{body}) \in E
+\\[0.75em]
+\quad\Delta_{i}(\bkndclr{v}) = \Delta_{i-1}(\bkndclr{v})\quad \textnormal{(otherwise)}
+\end{array}
+\end{equation*}
 \caption{Updating the node cache after binding a new graph}
 \label{fig:loop}
+\vspace{0.5em}
 \end{figure}
@@
 
+\section{Type checking}
+\label{sec:types}
+
+We need to cache type checking results too.
+
+The key trick is to separate the process into faster binding which constructs dependency
+graph and slower type checking and evaluation. (This assumes that binding can be done, perhaps
+as an over-approximation without full type checking.)
+
+\subsection{Provided types}
+
+a
+
+\subsection{Type checking over dependency graphs}
+
+a
+
+\newpage
+~
+\newpage
+
+% ==================================================================================================
+
 \section{Properties of preview evaluation}
@@
 Partial evaluation may reduce an expression under $\lambda$-abstraction. We do not require that
-the reduction of the host language allows this. Instead, we define an extended reduction relation
+the reduction of the host language does this. Instead, we define an extended reduction relation
 and use that in the proofs. The host language only needs to compose well with such extended reduction
@@
 \begin{definition}[Host language]
-Given a relation on expressions $e_1 \rightsquigarrow e_2$ that models evaluation, we define:
+\label{def:host}
+Given a relation on expressions $e_1 \rightsquigarrow e_2$ that models small-step evaluation, we define:
 
@@
 
+% --------------------------------------------------------------------------------------------------
+
 \subsection{Correctness of previews}
@@
 
-First is corrctness (Theorem~\ref{thm:correcntess}) -- no matter how a graph is constructed, when 
-we use it to evaluate a preview for an expression, the preview is the same as the value we would 
-obtain by evaluating the expression directly. Second is determinacy (Theorem~\ref{thm:determinacy}) --
-if we cache a preview for a graph node and update the graph, the preview would still be the same
-when evaluated using an updated graph.
-
-This is the correctness and determinacy property discussed in Section~\ref{sec:properties-correct}.
-
-Eliminating let bindings does not change the result in the host language. It also yields the same
-dependency graph.
+To show that the evalauted previews are correct, we prove two properties. Correctness 
+(Theorem~\ref{thm:correcntess}) shows that, no matter how a graph is constructed, when we use
+it to evaluate a preview for an expression, the preview is the same as the value we would obtain
+by evaluating the expression directly. Determinacy (Theorem~\ref{thm:determinacy}) shows that if
+we cache a preview for a graph node and update the graph, the preview we would evaluate using the
+updated graph would be the same as the cached preview.
+
+To simplify the proofs, we consider expressions without let bindings. This is possible, because
+eliminating let bindings does not change the result in the host language (Definition~\ref{def:host})
+and it also does not change the constructed dependency graph as shown in Lemma~\ref{thm:let-elimination}.
 
 \begin{lem}[Let elimintion]
-Given an expression $e_1$ such that $e_1 \rightsquigarrow_\ident{let} e_2$ and a lookup table $\Delta$
-then if $\bndclr{v_1}, (V_1, E_1) = \ident{bind}_{\emptyset, \Delta}(e_1)$ and  
-$\bndclr{v_2}, (V_2, E_2) = \ident{bind}_{\emptyset, \Delta}(e_2)$ then it holds that
-$\bndclr{v_1} = \bndclr{v_2}$ and also $(V_1, E_1) = (V_2, E_2)$.
+\label{thm:let-elimination}  
+Given an expression $e_1$ such that $e_1 \rightsquigarrow_\ident{let} e_2$ and a lookup table $\Delta_0$
+then if $\bndclr{v_1}, (V_1, E_1) = \ident{bind}_{\emptyset, \Delta_0}(e_1)$ and  
+$\bndclr{v_2}, (V_2, E_2) = \ident{bind}_{\emptyset, \Delta_1}(e_2)$ such that $\Delta_1 = \ident{update}_{V_1, E_1}(\Delta_0)$
+then it holds that $\bndclr{v_1} = \bndclr{v_2}$ and also $(V_1, E_1) = (V_2, E_2)$.
 \end{lem}
-
-This simplifies the correctness proof as we only need to consider expressions not containing
-let bindings (if there were let bindings, both the result and the preview would be equivalent
-to one that would be obtained if the let bindings were elimninated)
-
-Now, if we have any let-free expression, we show that the preview obtained via the dependency graph
-is a value that is the same as the one we'd get by directly evaluating the expression.
-
-xxx
-
-\begin{theorem}[Determinacy]
-\label{thm:determinacy}
-Downarrow will always evaluate the same, if we add things to a graph in the defined way.
-(this means that we can cache)
-\end{theorem}
-
-~
-
-~
-
-~
-
-[Only dependency graph matters, not the history and symbols]
-[Symbols are effectively hashes of dependencies]
-
-\begin{lemma}
-\label{thm:lemma-lookup}
-Lemma X - lookup with red mem(m) returns blue mem(m, s)
-\end{lemma}
+\begin{proof}
+Assume $e_1 = C[\kvd{let}~x=e'~\kvd{in}~e'']$ and the resulting $e_2 = C[e''[x\leftarrow e']]$. The case 
+$\ident{bind}_{\Gamma,\Delta}(\kvd{let}~x=e'~\kvd{in}~e'')$ when binding $e_1$ is handled using
+(3).
+
+When binding $e_1$, the node resulting from binding $e'$ is added to the graph $V_1, E_1$ and is 
+referenced each time $x$ is used. When binding $e_2$, the node representing $e'$ is a primitive value,
+or already present in $\Delta_1$ (added by $\ident{update}_{V_1, E_1}$) and is reused each time 
+$\ident{bind}_{\Gamma,\Delta_1}(e')$ is called.
+\end{proof}
+
+The Lemma~\ref{thm:let-elimination} provides a way of removing let bindings from an expression,
+such that the resulting dependency graph remains the same. Here, we bind the original expression
+first, which adds the node for $e'$ to $\Delta$. In our implementation, this is not needed
+because $\Delta$ is updated while the graph is being constructed using $\ident{bind}$. 
+To keep the formalisation simpler, we separate the process of building the dependency graph 
+and updating $\Delta$.
+
+Now, we can show that, given a let-free expression, the preview obtained using a correctly
+constructed dependency graph is the same as the one we would obtain by directly evaluating the
+expression. This requires a simple auxiliary lemma and the full proof is shown in 
+Appendix~\ref{sec:app-correctness}.
+
+\begin{lem}
+\label{thm:lemma-lookup}[Lookup inversion]
+Given $\Delta$ obtained using \ident{update} as defined in Figure~\ref{fig:loop} then: 
+\begin{itemize}
+\raggedright
+\item[--] If $\bndclr{v}=\Delta(\bknd{fun}(x),[(\bndclr{v_0}, \blblclr{l_0})])$
+then $\bndclr{v}=\bnd{fun}(x, s)$ for some $s$. 
+\item[--] If $\bndclr{v}=\Delta(\bknd{mem}(m),[(\bndclr{v_0}, \blblclr{l_0}), \ldots, (\bndclr{v_n}, \blblclr{l_n})])$
+then $\bndclr{v}=\bnd{mem}(m, s)$ for some $s$. 
+\end{itemize}
+\end{lem}
+\begin{proof}
+By construction of $\Delta$ in Figure~\ref{fig:loop}.  
+\end{proof}
 
 \begin{theorem}[Let-free correctness]
+\label{thm:let-free-correct}
 Given an expression $e$ that has no free variables and does not contain let bindings, together
@@
 \end{enumerate}
-
+%
 The proof is done by induction over the binding process, which follows the structure 
@@
 
+\noindent
+The correctness theorem combines the previous two results.
+
 \begin{theorem}[Correctness]
-\label{thm:correctness}
-(Same as let-free version, but using let elimination to say that this works for any expression)
+\label{thm:correcntess}
+Consider an expression $e_1$ that has no free variables together with a lookup table $\Delta_1$ obtained 
+from any sequence of expressions according to Figure~\ref{fig:loop} and $e_2$ such that
+$e_1\rightsquigarrow_\beta e_2$ and let $\bndclr{v_1}, (V_1, E_1) = \ident{bind}_{\emptyset, \Delta_1}(e_1)$.
+
+Let $\Delta_2 = \ident{update}_{V_1, E_1}(\Delta_1)$ and $\bndclr{v_2}, (V_2, E_2) = \ident{bind}_{\emptyset, \Delta_2}(e_2)$.
+If $\bndclr{v_2}\Downarrow d$ over a graph $(V_2, E_2)$ then $d = p$ for some $p$ and $e \rightsquigarrow_\beta p$.
 \end{theorem}
+\begin{proof}
+Direct consequence of Lemma~\ref{thm:let-elimination} and Theorem~\ref{thm:let-free-correct}.
+\end{proof}
+
+As discussed above when introducing Lemma~\ref{thm:let-elimination}, in our implementation,
+$\Delta$ is updated during the recursive binding process and so a stronger version of the property
+holds -- namely, $e \rightsquigarrow_\beta p$ for a $p$ that is obtained by calculating preview 
+over a graph obtained directly for the original expression $e$. We note that this is the case, but
+do not show it formally to keep aid the clarity of our formalisation.
+
+The second important property that guarantees the correctness of previews shown by the user in 
+our implementation is determinacy. This makes it possible to cache the previews evaluated using
+$\Downarrow$ using the corresponding graph node as a lookup key.
+
+\begin{theorem}[Determinacy]
+\label{thm:determinacy}
+Let $\Delta_1 = \emptyset$, for any $e_1, e_2$, assume that the first expression is bound,
+i.e.~$\bndclr{v_1}, (V_1, E_1) = \ident{bind}_{\emptyset, \Delta_1}(e_1)$, the graph node cache
+is updated $\Delta_2 = \ident{update}_{V_1, E_1}(\Delta_1)$ and a new expression is bound, i.e.~
+$\bndclr{v_2}, (V_2, E_2) = \ident{bind}_{\emptyset, \Delta_2}(e_2)$. Now, for any $\bndclr{v}$, 
+if $\bndclr{v} \Downarrow p$ over $(V_1, E_1)$ then also
+ $\bndclr{v} \Downarrow p$ over $(V_2, E_2)$.
+\end{theorem}
+\begin{proof}
+By induction over $\Downarrow$ over $(V_1, E_1)$, we show that the same evaluation rules also
+apply over $(V_2, E_2)$.
+
+This is the case, because new graph nodes added to $\Delta_2$ by $\ident{update}_{V_1, E_2}$
+are only ever added as new nodes in $\ident{bind}_{\emptyset, \Delta_2}$ and so the existing 
+nodes and edges of $(V_1, E_1)$ used during the evaluation are unaffected.
+\end{proof}
+
+The mechanism used for caching previews, as discussed at the end of Section~\ref{sec:previews},
+keeps a preview or a partial preview $d$ in a lookup table indexed by nodes $\bndclr{v}$. The
+Throrem~\ref{thm:determinacy} guarantees that this is a valid strategy. As we update dependency
+graph during code editing, previous nodes will continue representing the same sub-expressions.
 
@@
 
-* Inlining let leads to an equivalent dependency graph
-
-* Given two expressions, if we inline lets and they are structurally different, then 
-  their dependency graphs are structurally different
-
-Something about how dependency graph bits are reused correctly, i.e. if there is a source code
-change that can change the result, the graph will change.
-
-Something about various code edit operations that will not cause recomputation.
+\subparagraph{1. Introducing let}
+
+$C_1[C_2[e]]$ to
+$C_1[\kvd{let}~x=e~\kvd{in}~C_2[x]]$
+
+Also holds for multiple replaces.
+
+\subparagraph{2. Eliminating let}
+
+$C_1[\kvd{let}~x=e~\kvd{in}~C_2[x]]$ to
+$C_1[\kvd{let}~x=e~\kvd{in}~C_2[e]]$ to
+$C_1[C_2[e]]$
+
+or
+
+$C_1[\kvd{let}~x=e~\kvd{in}~C_2[x]]$ to
+$C_1[C_2[x]]$ to
+$C_1[C_2[e]]$
+
+Also holds for multiple variables...
+
+\subparagraph{3. Editing non-dependency in let}
+
+$C[\kvd{let}~x=e_1~\kvd{in}~e_2]$ such that $x\notin FV(e_2)$ to
+$C[\kvd{let}~x=e_1'~\kvd{in}~e_2]$
+
+\subparagraph{3. Editing non-dependency in chain} a
+
+$C[e.m(e_1, \ldots, e_n).m'(e'_1, \ldots, e'_k)]$ to
+
+$C[e.m(e_1, \ldots, e_n).m''(e''_1, \ldots, e''_k)]$
+
+preview of $e.m(e_1, \ldots, e_n)$ remains
 
@@
   Here $e = e_0.m(e_1, \ldots, e_n)$, $\bndclr{v_i}$ are graph nodes obtained by induction for
-  expressions $e_i$ and $\{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v}, \bndclr{v_n}, \blbl{arg}(n))\} \subseteq E$
+  expressions $e_i$ and $\{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v}, \bndclr{v_n}, \blbl{arg}(n))\} \subseteq E$.
   From Lemma~\ref{thm:lemma-lookup}, $\bndclr{v} = \bnd{mem}(m, s)$ for some $s$.

### 2017-11-15 live/paper.tex
Reuse of previews and type checking

@@
 \section{Introduction}
+\label{sec:intro}
+
 One of the aspects that make spreadsheet tools such as Excel more accessible than programming 
@@
 
+\begin{figure*}[t]
+\begin{equation*}
+\begin{array}{ll}
+\ident{bind}_{\Gamma, \Delta, c}(e_0.m(e_1, \ldots, e_n)) = &(2b)\\ 
+\qquad \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n)\\
+\quad \textnormal{when}~\bndclr{v_0}, (V_0, E_0) = \ident{bind}_{\Gamma, \Delta, \bot}(e_0)\\
+\quad \textnormal{and}~c_i = (\bndclr{v_0}, \blbl{callsite}(m,i))\hspace{4.8em} (i \in 1\ldots n)\\
+\quad \textnormal{and}~\bndclr{v_i}, (V_i, E_i) = \ident{bind}_{\Gamma, \Delta, c_i}(e_i)\qquad (i\in 1\ldots n)\\
+\quad \textnormal{and}~(\bknd{mem}(m),[(\bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v_n}, \blbl{arg}(n))]) \notin \ident{dom}(\Delta)\hspace{-200em}.\\
+\quad \textnormal{let}~\bndclr{v} = \bnd{mem}(m, s), s~\textnormal{fresh}\\
+\quad \textnormal{let}~E = \{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v},\bndclr{v_n}, \blbl{arg}(n)) \}
+\end{array}
+\qquad\qquad
+\begin{array}{ll}
+\ident{bind}_{\Gamma, \Delta, (\bndclr{v_c}, \blblclr{l_c} )}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V_0, E \cup E_0)&(7b)\\
+\quad \textnormal{when}~(\bknd{var}(x),[(\bndclr{v_c}, \blblclr{l_c} )])\notin \ident{dom}(\Delta)\\
+\quad \textnormal{let}~\bndclr{v_x} = \bnd{var}(x, s_x), s_x~\textnormal{fresh}\\
+\quad \textnormal{and}~\Gamma_1 = \Gamma \cup \{ x, \bndclr{v_x} \}\\
+\quad \textnormal{and}~\bndclr{v_0}, (V_0, E_0) = \ident{bind}_{\Gamma_1, \Delta, \bot}(e)\\
+\quad \textnormal{when}~(\bknd{fun}(x),[(\bndclr{v_0}, \blbl{body}), (\bndclr{v_c}, \blblclr{l_c} )])\notin \ident{dom}(\Delta)\\
+\quad \textnormal{let}~\bndclr{v} = \bnd{fun}(x, s_f), s_f~\textnormal{fresh}\\
+\quad \textnormal{let}~E = \{ (\bndclr{v},\bndclr{v_0},\blbl{body}), (\bndclr{v},\bndclr{v_c},\blblclr{l_c}), (\bndclr{v_x},\bndclr{v_c},\blblclr{l_c}) \}
+\\[0.75em]
+\end{array}
+\end{equation*}
+\caption{Revised binding rules, tracking call sites of function values.} 
+\label{fig:binding-rules-callsite}
+\end{figure*}
+
+\begin{figure*}
+\begin{equation*}
+\begin{array}{l}
+\inference[(val)~]
+  {\Sigma(n) = \tau}
+  {\bnd{val}(n) \vdash \tau }
+\\
+\\
+\inference[(var)~]
+  { (\bnd{var}(x,s), \bndclr{v}, \blbl{callsite}(m, i)) \in E \\
+    \bndclr{v} \vdash (.., m:(\tau_1, \ldots, \tau_k)\rightarrow \tau,..) & \tau_i=\tau'\rightarrow\tau'' }
+  { \bnd{var}(x, s) \vdash \tau' }
+\\
+\\
+\inference[(fun-val)~]
+  { (\bnd{fun}(x,s), \bndclr{v_b}, \blbl{body}) \in E \\
+    (\bnd{var}(x,s), \bndclr{v_c}, \blbl{callsite}(m, i)) \in E \\
+    \bndclr{v_c} \vdash (.., m:(\tau_1, \ldots, \tau_k)\rightarrow \tau,..) \\
+    \tau_i=\tau'\rightarrow\tau'' & \bndclr{v_b} \vdash \tau'' }
+  { \bnd{fun}(x, s) \vdash \tau' \rightarrow \tau'' }
+\\
+\\
+\inference[(mem)~]
+  { \forall i\in\{0\ldots k\}.(\bnd{mem}(m, s), \bndclr{v_i}, \blbl{arg}(i)) \in E \\
+    \bndclr{v_0} \vdash (.., m:(\tau_1, \ldots, \tau_k) \rightarrow \tau, ..)& \bndclr{v_i} \vdash \tau_i }
+  { \bnd{mem}(m, s) \vdash \tau }
+\\  
+\\
+\end{array}
+\end{equation*}
+
+\caption{Rules that define evaluation of previews over a dependency graph for an expression}  
+\label{fig:tc}
+\end{figure*}
+
+% --------------------------------------------------------------------------------------------------
+
 \section{Type checking}
@@
 
+\subsection{Type checking functions}
+
+%
+\begin{equation*}
+\begin{array}{lcll}
+\bkndclr{k}&\in&\bknd{fun}(x)~|~\bknd{mem}(m)~|~\bknd{var}(x)&\quad(\textit{Node kinds})\\
+c&\in&\bot ~|~(\bndclr{v}, \blblclr{l})&\quad(\textit{Call sites})\\
+\end{array}
+\end{equation*}
+%
+
+Binding of
+
+$o.m(\lambda x\rightarrow x)$
+
+\begin{equation*}
+\xymatrix{
+&\bnd{val}(o)\\
+&\bnd{mem}(m, s_0) \ar[u]_{\blbl{arg}(0)} \ar[dl]^{\blbl{arg}(1)} \\
+\bnd{fun}(x, s_2) \ar@/^/[uur]^{\blbl{callsite}(m, 1)} \ar[rr]^{\blbl{body}} && \bnd{var}(x, s_1) \ar@/_/[uul]_{\blbl{callsite}(m, 1)}
+}  
+\end{equation*}
+
+Update Figure~\ref{fig:loop} accordingly
+
+We show cases (2) and (7); cases (1) and (6) that reuse nodes are analogous;
+other cases just call bind recursively with no node
+
 \subsection{Type checking over dependency graphs}
 
-a
+\begin{equation*}
+\begin{array}{lcll}
+\tau&\in&\alpha ~|~ \tau \rightarrow \tau ~|~ (m_1\!:\!\sigma_1, \ldots, m_n\!:\!\sigma_n) &\quad(\textit{Types})\\
+\sigma &\in& (\tau_1, \ldots, \tau_n) \rightarrow \tau &\quad(\textit{Members})\\
+\end{array}
+\end{equation*}
 
@@
 
-\section{Properties of preview evaluation}
+\section{Properties of live coding environment}
 \label{sec:properties}
@@
 To show that the evalauted previews are correct, we prove two properties. Correctness 
-(Theorem~\ref{thm:correcntess}) shows that, no matter how a graph is constructed, when we use
+(Theorem~\ref{thm:correcntess}) guarantees that, no matter how a graph is constructed, when we use
 it to evaluate a preview for an expression, the preview is the same as the value we would obtain
-by evaluating the expression directly. Determinacy (Theorem~\ref{thm:determinacy}) shows that if
+by evaluating the expression directly. Determinacy (Theorem~\ref{thm:determinacy}) guarantees that if
 we cache a preview for a graph node and update the graph, the preview we would evaluate using the
@@
 
+% --------------------------------------------------------------------------------------------------
+
 \subsection{Reuse of previews}
@@
 
-\subparagraph{1. Introducing let}
-
-$C_1[C_2[e]]$ to
-$C_1[\kvd{let}~x=e~\kvd{in}~C_2[x]]$
-
-Also holds for multiple replaces.
-
-\subparagraph{2. Eliminating let}
-
-$C_1[\kvd{let}~x=e~\kvd{in}~C_2[x]]$ to
-$C_1[\kvd{let}~x=e~\kvd{in}~C_2[e]]$ to
-$C_1[C_2[e]]$
-
-or
-
-$C_1[\kvd{let}~x=e~\kvd{in}~C_2[x]]$ to
-$C_1[C_2[x]]$ to
-$C_1[C_2[e]]$
-
-Also holds for multiple variables...
-
-\subparagraph{3. Editing non-dependency in let}
-
-$C[\kvd{let}~x=e_1~\kvd{in}~e_2]$ such that $x\notin FV(e_2)$ to
-$C[\kvd{let}~x=e_1'~\kvd{in}~e_2]$
-
-\subparagraph{3. Editing non-dependency in chain} a
-
-$C[e.m(e_1, \ldots, e_n).m'(e'_1, \ldots, e'_k)]$ to
-
-$C[e.m(e_1, \ldots, e_n).m''(e''_1, \ldots, e''_k)]$
-
-preview of $e.m(e_1, \ldots, e_n)$ remains
-
-% --------------------------------------------------------------------------------------------------
-
-\newpage
+In the motivating example in Section~\ref{sec:intro}, the programmer first extracted a constant
+value into a let binding and then modified a parameter of the last method call in a call chain.
+We argued that the live coding environment should reuse partially evaluated previews for these
+two cases. In this section, we prove that this is, indeed, the case in our system.
+
+Figure~\ref{fig:operations} shows a list of six code edit operations where a previeww of the 
+expression (cases 1-4), or a sub-expression (cases 5-6), can be reused. This is the case, because 
+the graph nodes that are bound to the sub-expression before and after the code is changed are the 
+same and hence, a cached preview (stored using the graph node as the key) can be reused.
+
+In some of the operations (cases 1 and 3), the code is changed via an intermediate expression
+that is semantically different and has only partial preview. This illustrates a typical way of
+working with code in a text editor using cut and paste operations. Cases 1 and 3 illustrate how
+our approach allows this way of editing code.
+
+Finally, it is worth noting that our list is not exhaustive. In particular, cases 1-4 only cover
+let bindings where the bound variable is used once. However, previews can also be reused if the
+variable appears multiple times.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]  
+\raggedright
+\subparagraph{1. Let introduction A.} 
+The expression $C_1[C_2[e]]$ is changed to $C_1[\kvd{let}~x=e~\kvd{in}~C_2[x]]$
+via semantically non-equivalent expression $C_1[C_2[x]]$ where $x$ is unbound variable.
+
+\subparagraph{2. Let introduction B.} 
+The expression $C_1[C_2[e]]$ is changed to $C_1[\kvd{let}~x=e~\kvd{in}~C_2[x]]$
+via $C_1[\kvd{let}~x=e~\kvd{in}~C_2[e]]$ where $x$ is unused variable.
+
+\subparagraph{3. Let elimination A.} The expression
+$C_1[\kvd{let}~x=e~\kvd{in}~C_2[x]]$ is changed to
+$C_1[C_2[e]]$ via semantically non-equivalent expression
+$C_1[C_2[x]]$ where $x$ is unbound variable.
+
+\subparagraph{4. Let elimination B.} The expression
+$C_1[\kvd{let}~x=e~\kvd{in}~C_2[x]]$ is changed to
+$C_1[C_2[e]]$ via $C_1[\kvd{let}~x=e~\kvd{in}~C_2[e]]$ 
+where $x$ is unused variable.
+
+\subparagraph{5. Editing a non-dependency in let.} Assuming $x\notin FV(e_2)$, the expression
+$C_1[\kvd{let}~x=e_1~\kvd{in}~C_2[e_2]]$ changes to an expression
+$C_1[\kvd{let}~x=e_1'~\kvd{in}~C_2[e_2]]$. The preview of a sub-expression $e_2$ is not recomputed.
+
+\subparagraph{6. Editing a non-dependency in a chain.} The expression 
+$C[e.m(e_1, \ldots, e_n).m'(e'_1, \ldots, e'_k)]$ is changed to an expression
+$C[e.m(e_1, \ldots, e_n).m''(e''_1, \ldots, e''_k)]$. The preview of a sub-expression
+$e.m(e_1, \ldots, e_n)$ is not recomputed.
+
+\caption{Code edit operations that enable preview reuse}
+\label{fig:operations}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{lem}[Binding sub-expressions]
+\label{thm:sub-expr}
+Given any $\Delta_1$ together with $e_1 = C[C_1[e]]$ and $e_2 C[C_2[e]]$, such that all free 
+variables of $e$ are bound in $C$, assume that the first expression is bound,
+i.e.~$\bndclr{v_1}, (V_1, E_1) = \ident{bind}_{\emptyset, \Delta_1}(e_1)$, the graph node cache
+is updated $\Delta_2 = \ident{update}_{V_1, E_1}(\Delta_1)$ and the second expression is bound, i.e.~
+$\bndclr{v_2}, (V_2, E_2) = \ident{bind}_{\emptyset, \Delta_2}(e_2)$.
+
+Now, assume $\bndclr{v}, G = \ident{bind}_{\Gamma_1, \Delta_1}(e)$ and 
+$\bndclr{v'}, G' = \ident{bind}_{\Gamma_2, \Delta_2}(e)$ are the recursive calls to bind
+$e$ during the first and the second binding, respectively. Then, the graph nodes assigned to the
+sub-expression $e$ are the same, i.e.~$\bndclr{v} = \bndclr{v'}$.
+\end{lem}
+\begin{proof}
+First, assuming that $\forall x\in FV(e). \Gamma_1(x) = \Gamma_2(x)$, we show by induction over the binding process of $e$
+when binding $C[C_1[e]]$ that the result is the same. In cases (1) and (6), the updated $\Delta_2$
+contains the required key and so the second binding proceeds using the same case. In cases 
+(2) and (7), the second binding reuses the node created by the first binding using case (1) and
+(6), respectively. Cases (4) and (5) are the same and case (3) follows directly via induction.
+
+Second, when binding let bindings in $C[-]$, the initial $\Gamma = \emptyset$ during both bindings
+and so the nodes added to $\Gamma_1$ and $\Gamma_2$ are the same. $C_1$ and $C_2$ do not add
+any new nodes used in $e$ to $\Gamma_1$ and $\Gamma_2$ and so $\bndclr{v} = \bndclr{v'}$
+using the above.
+\end{proof}
+
+\begin{theorem}[Preview reuse]
+Given the sequence of expressions as specified in Figure~\ref{fig:operations}, if the expressions
+are bound in sequence and graph node cache updated as specified in Figure~\ref{fig:loop}, then 
+the graph nodes assigned to the specifeid sub-expressions are the same.
+\end{theorem}
+\begin{proof}
+Consequence of Lemma~\ref{thm:sub-expr}, using appropriate contexts. In cases with 
+intermediate expressions (1)-(4), binding the intermediate expression introduces additional
+nodes to $\Delta$, but those are ignored when binding the final expression.
+\end{proof}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Properties of type checking}
+
+a
+
+% --------------------------------------------------------------------------------------------------
 

### 2017-11-16 live/paper.tex
Type checking text

@@
 \definecolor{numclr}{rgb}{0.0,0.4,0.0}
-\definecolor{strclr}{rgb}{0.4,0.3,0.0}
+\definecolor{strclr}{rgb}{0.4,0.4,0.0}
+\definecolor{rstrclr}{rgb}{0.5,0.1,0.0}
 \definecolor{prepclr}{rgb}{0.6,0.0,0.2}
@@
 \newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
+\newcommand{\rstr}[1]{\textnormal{\textcolor{rstrclr}{\sffamily "#1"}}}
 \newcommand{\ident}[1]{\textnormal{\sffamily #1}}
@@
 
-\section{Formalising live coding environment}
+\section{Formalising live coding infrastructure}
 \label{sec:formal}
 
-In this section, we present a formalisation of a live coding environment for a small,
+In this section, we present a formalisation of a live coding infrastructure for a small,
 expression-based programming language that supports \kvd{let} binding, member invocations
@@
 
-Finally, the evaluation of function values is similar, but requires three cases. If the body can 
+The evaluation of function values is similar, but requires three cases. If the body can 
 be reduced to a value with no unbound variables (\emph{fun-val}), we return a lambda function that
@@
 lambda function with the delayed preview as the body. If the body requires further variables,
-the result is a delayed preview containing a function.
+the result is a delayed preview.
 
@@
 are cached and linked to graph nodes. In practice, this is done by maintaining a lookup table
-$\bndclr{v} \rightarrow p$ from graph nodes $\bdnclr{v}$ to (possibly delayed) previews $p$.
-Whenever the $\Downarrow$ relation is used to obtain a preview for a graph node, we first 
+from graph nodes $\bdnclr{v}$ to (possibly delayed) previews $p$.
+Whenever $\Downarrow$ is used to obtain a preview for a graph node, we first 
 attempt to find an already evaluated preview using the lookup table. If the preview has not
@@
 \begin{figure*}[t]
+\vspace{-0.5em}
 \begin{equation*}
@@
 \end{equation*}
+\vspace{-0.5em}
 \caption{Revised binding rules, tracking call sites of function values.} 
 \label{fig:binding-rules-callsite}
+\vspace{-0.5em}
 \end{figure*}
 
-\begin{figure*}
-\begin{equation*}
-\begin{array}{l}
-\inference[(val)~]
-  {\Sigma(n) = \tau}
-  {\bnd{val}(n) \vdash \tau }
-\\
-\\
-\inference[(var)~]
-  { (\bnd{var}(x,s), \bndclr{v}, \blbl{callsite}(m, i)) \in E \\
-    \bndclr{v} \vdash (.., m:(\tau_1, \ldots, \tau_k)\rightarrow \tau,..) & \tau_i=\tau'\rightarrow\tau'' }
-  { \bnd{var}(x, s) \vdash \tau' }
-\\
-\\
-\inference[(fun-val)~]
-  { (\bnd{fun}(x,s), \bndclr{v_b}, \blbl{body}) \in E \\
-    (\bnd{var}(x,s), \bndclr{v_c}, \blbl{callsite}(m, i)) \in E \\
-    \bndclr{v_c} \vdash (.., m:(\tau_1, \ldots, \tau_k)\rightarrow \tau,..) \\
-    \tau_i=\tau'\rightarrow\tau'' & \bndclr{v_b} \vdash \tau'' }
-  { \bnd{fun}(x, s) \vdash \tau' \rightarrow \tau'' }
-\\
-\\
-\inference[(mem)~]
-  { \forall i\in\{0\ldots k\}.(\bnd{mem}(m, s), \bndclr{v_i}, \blbl{arg}(i)) \in E \\
-    \bndclr{v_0} \vdash (.., m:(\tau_1, \ldots, \tau_k) \rightarrow \tau, ..)& \bndclr{v_i} \vdash \tau_i }
-  { \bnd{mem}(m, s) \vdash \tau }
-\\  
-\\
-\end{array}
-\end{equation*}
-
-\caption{Rules that define evaluation of previews over a dependency graph for an expression}  
-\label{fig:tc}
-\end{figure*}
-
-% --------------------------------------------------------------------------------------------------
+% ==================================================================================================
 
@@
 
-We need to cache type checking results too.
-
-The key trick is to separate the process into faster binding which constructs dependency
-graph and slower type checking and evaluation. (This assumes that binding can be done, perhaps
-as an over-approximation without full type checking.)
-
-\subsection{Provided types}
-
-a
-
-\subsection{Type checking functions}
-
-%
-\begin{equation*}
-\begin{array}{lcll}
-\bkndclr{k}&\in&\bknd{fun}(x)~|~\bknd{mem}(m)~|~\bknd{var}(x)&\quad(\textit{Node kinds})\\
-c&\in&\bot ~|~(\bndclr{v}, \blblclr{l})&\quad(\textit{Call sites})\\
-\end{array}
-\end{equation*}
-%
-
-Binding of
-
-$o.m(\lambda x\rightarrow x)$
-
+Evaluating live previews can be an expensive operations, so being able to cache partial previews
+is a must for a live coding environment. Type checking is typically fast, so the main focus of
+this paper is on live previews. However, asynchronous type providers in The Gamma (Section~\ref{sec:types-providers}) 
+can make type checking time consuming, and so we use the dependency graph also for type checking
+(Section~\ref{sec:types-graph}). Type checking lambda functions (Section~\ref{sec:types-funs}) 
+requires a slight extension of the model discussed in Section~\ref{sec:formal}.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Asynchronously provided types}
+\label{sec:types-providers}
+
+Data available in The Gamma can be defined using several kinds of type providers. The type provider
+used in Figure~\ref{fig:thegamma} as asynchronous [X]. It downloads the sample URL and generates 
+types based on the contents of the web page. The parameter to \ident{web.scrape} is a static 
+parameter and is evaluated during type-checking. We omit details in this paper, but we note this
+works similarly to F\# type providers [X].
+
+Type providers can also be implemented as REST services [X] to allow anyone implement a data source
+in the language of their choice. In this case, each member of a call-chain returns a type that is
+generated based on the result of an HTTP request. For example, when the user types \ident{worldbank}
+(to access information about countries), the type provider makes a request to   
+\url{http://thegamma-services.azurewebsites.net/worldbank}, which returns two members:
+%
+\begin{equation*}
+\begin{array}{l}
+[~\{\rstr{name}\!:\str{byYear},\rstr{returns}\!:\\
+~\quad \{\rstr{kind}\!:\str{nested},\rstr{endpoint}\!:\str{/pickYear}\}\}\\
+\hspace{0.55em}\{\rstr{name}\!:\str{byCountry},\rstr{returns}\!:\\
+~\quad \{\rstr{kind}\!:\str{nested},\rstr{endpoint}\!:\str{/pickCountry}\}\}~]
+\end{array}
+\end{equation*}
+%
+This indicates that \ident{worldbank} has members \ident{byYear} and \ident{byCountry}.
+If the user types \ident{worldbank.byCountry}, a request is made to the specified URL 
+\url{http://thegamma-services.azurewebsites.net/worldbank/pickCountry}:
+%
+\begin{equation*}
+\begin{array}{l}
+[~\{\rstr{name}\!:\str{Andorra},\rstr{trace}\!:[\str{"country=AR"}], \\
+\quad \rstr{returns}\!:\{\rstr{kind}\!:\str{nested},\rstr{endpoint}\!:\str{/pickTopic}\}\}\\
+\hspace{0.55em}\{\rstr{name}\!:\str{Afghanistan},\rstr{trace}\!:[\str{"country=AF"}],\\
+\quad \rstr{returns}\!:\{\rstr{kind}\!:\str{nested},\rstr{endpoint}\!:\str{/pickTopic}\}\},\,..]
+\end{array}
+\end{equation*}
+%
+This returns a list of countries which can then be accessed as members via 
+\ident{worldbank.byCountry.Andorra}, etc.
+
+This is one reason for why type checking in The Gamma can be time consuming. Other
+type providers may perform other more computationally intensive work to provide types and so
+it is desirable to reuse type-checking results during live coding. The rest of this section
+shows how this is done using the dependency graph discussed in Section~\ref{sec:formal}.
+
+\begin{figure}[!b]
+\vspace{-0.5em}
 \begin{equation*}
@@
 \end{equation*}
-
-Update Figure~\ref{fig:loop} accordingly
-
-We show cases (2) and (7); cases (1) and (6) that reuse nodes are analogous;
-other cases just call bind recursively with no node
+\caption{Dependency graph for $o.m(\lambda x\rightarrow x)$}
+\label{fig:graph-func}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Revised binding of functions}
+\label{sec:types-funs}
+
+The Gamma script supports lambda functions, but only in a limited way. A function can be passed
+as a parameter to a method, which makes type checking of functions easier. For example, consider:
+%
+\begin{equation*}
+\begin{array}{l}  
+\ident{movies}.\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())
+\end{array}
+\end{equation*}
+%
+If \ident{movies} is a collection of \ident{Movie} objects, the type of the lambda function must be
+$\ident{Movie}\rightarrow\ident{bool}$ and so the type of $x$ is \ident{Movie}. This is similar to
+type checking of lambda functions in C\# [X], where type is also inferred from the context (or has
+to be specified explicitly).
+
+We do not currently allow lambda functions as stand-alone let-bound 
+values. This could be done by requiring explicit types, or introducing polymorphism, but it was not
+necessary for the limited domain of non-expert data exploration.
+
+\paragraph{Dependency graph for functions.}
+In the binding process specified in Section~\ref{sec:formal}, a variable is a leaf of the dependency
+graph. In the revised model, it depends on the context in which it appears. A new edge labelled
+$\blbl{callsite}(m, i)$ indicates that the source node is the input variable of a function 
+passed as the $i^\textnormal{th}$ argument to the $m$ member of the expression represented by the
+target node. A node representing function is linked to the call site using the same edge.
+
+Figure~\ref{fig:graph-func} shows the result of binding $o.m(\lambda x\rightarrow x)$. Both
+$\bnd{fun}(x, s_2)$ and $\bnd{var}(x, s_1)$ now depend on the node representing $o$. The new 
+$\blbl{callsite}$ edge makes it possible to type-check function and variable nodes just using their 
+dependencies. As before, the member invocation $\bnd{mem}(m, s_0)$ 
+depends on the instance using $\blbl{arg}(0)$ and on its argument using $\blbl{arg}(1)$.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure*}
+\vspace{0.5em}
+\begin{equation*}
+\begin{array}{l}
+\qquad\inference[(val)~]
+  {\Sigma(n) = \alpha}
+  {\bnd{val}(n) \vdash \alpha }
+\quad\qquad
+\inference[(var)~]
+  { (\bnd{var}(x,s), \bndclr{v}, \blbl{callsite}(m, i)) \in E &
+    \bndclr{v} \vdash (.., m:(\tau_1, \ldots, \tau_k)\rightarrow \tau,..) & \tau_i=\tau'\rightarrow\tau'' }
+  { \bnd{var}(x, s) \vdash \tau' }
+\\[2.5em]
+\inference[(fun)~]
+  { \{(\bnd{fun}(x,s), \bndclr{v_b}, \blbl{body}), (\bnd{var}(x,s), \bndclr{v_c}, \blbl{callsite}(m, i))\} \subseteq E &
+    \bndclr{v_c} \vdash (.., m:(\tau_1, \ldots, \tau_k)\rightarrow \tau,..) &
+    \tau_i=\tau'\rightarrow\tau'' & \bndclr{v_b} \vdash \tau'' }
+  { \bnd{fun}(x, s) \vdash \tau' \rightarrow \tau'' }
+\\[2.5em]
+\qquad\qquad\quad\inference[(mem)~]
+  { \forall i\in\{0\ldots k\}.(\bnd{mem}(m, s), \bndclr{v_i}, \blbl{arg}(i)) \in E &
+    \bndclr{v_0} \vdash (.., m:(\tau_1, \ldots, \tau_k) \rightarrow \tau, ..)& \bndclr{v_i} \vdash \tau_i }
+  { \bnd{mem}(m, s) \vdash \tau }
+\\[1em]
+\end{array}
+\end{equation*}
+\caption{Rules that define evaluation of previews over a dependency graph for an expression}  
+\vspace{0.5em}
+\label{fig:tc}
+\end{figure*}
+
+% --------------------------------------------------------------------------------------------------
+
+\paragraph{Revised binding process.}
+For the revised binding process, we introduce a new edge label $\blbl{callsite}$. Variable nodes
+now have dependencies and so we cache them and attach a symbol $s$ to \bnd{var}, so we also 
+introduce a node kind $\bknd{var}$ as part of a lookup key for $\Delta$.
+
+The \ident{bind} function now has a parameter $c$, in addition to $\Gamma$ and $\Delta$, which 
+represents the context in which the binding happens. This is either a member invocation (labelled 
+with instance node $\bndclr{v}$ and \blbl{callsite} label $\blblclr{l}$), or not a member 
+invocation writted as $\bot$. The updated definitions are:
+%
+\begin{equation*}
+\begin{array}{lcll}
+\bndclr{v} &\hspace{-0.5em}\in\hspace{-0.5em}&\bnd{val}(n)~|~\bnd{var}(x, s)~|~\bnd{mem}(m, s)~|~\bnd{fun}(x, s)&(\textit{Vertices})\\
+\blblclr{l}&\hspace{-0.5em}\in\hspace{-0.5em}&\blbl{body}~|~\blbl{arg}(i)~|~\blbl{callsite}(m,i)&(\textit{Edge labels})\\
+\bkndclr{k}&\hspace{-0.5em}\in\hspace{-0.5em}&\bknd{fun}(x)~|~\bknd{mem}(m)~|~\bknd{var}(x)&(\textit{Node kinds})\\
+          c&\hspace{-0.5em}\in\hspace{-0.5em}&\bot ~|~(\bndclr{v}, \blblclr{l})&(\textit{Call sites})\\
+\end{array}
+\end{equation*}
+%
+The key parts of the revised definiton of the \ident{bind} function are shown in Figure~\ref{fig:binding-rules-callsite}.
+We now write $\ident{bind}_{\Gamma,\Delta,c}$ where $c$ represents the context in which the
+binding occurs. This is set to $\bot$ in all cases, except when binding arguments of a member call.
+In (2b), we first recursively bind the instance node using $\bot$ as the context and then 
+bind all the arguments using $(\bndclr{v_0}, \blbl{callsite}(m, i))$ as the context for $i^{\textnormal{th}}$ 
+argument. The rest is as in case (2) before. The case (1) is updated similarly and is not shown 
+here for brevity.
+
+When binding a function (7b), we now also store variable nodes in $\Delta$ and so we check if a 
+variable with the given call site exists. If no, we create a fresh node $\bnd{var}(x, s_x)$.
+The node is added to $\Gamma$ as before. At the end, we now also include a call site edge from the 
+variable node and from the function node in $E$. We omit a similar variant of the case (6).
+The remaining cases (3)-(5) are the same, except that \ident{bind} has the additional $c$ parameter
+and recursive calls always set it to $\bot$.
+
+Finally, the \ident{update} function in Figure~\ref{fig:loop} also needs to be updated to store the
+newly created \bnd{var} nodes. This is done by adding the following single case:
+
+\begin{equation*}
+\begin{array}{l}
+\Delta_{i}(\bknd{var}(x), [(\bndclr{v},\blbl{callsite}(m,i))]) = \bnd{var}(x, s)\\
+\quad\quad \textnormal{for all}~\bnd{var}(x, s) \in V\\
+\quad\quad \textnormal{such that}~(\bnd{var}(x, s), \bndclr{v}, \blbl{callsite}(m,i)) \in E
+\end{array}
+\end{equation*}
+
+% --------------------------------------------------------------------------------------------------
 
 \subsection{Type checking over dependency graphs}
-
+\label{sec:types-graph}
+
+The type system for The Gamma supports a number of primitive types (such as integers and strings) 
+written as $\alpha$. Composed types include functions and objects with members. Objects are be 
+provided by type providers, but we omit the details here. Types of object members are written as
+$\sigma$ and can have multiple arguments:
+%
 \begin{equation*}
 \begin{array}{lcll}
-\tau&\in&\alpha ~|~ \tau \rightarrow \tau ~|~ (m_1\!:\!\sigma_1, \ldots, m_n\!:\!\sigma_n) &\quad(\textit{Types})\\
+\tau&\in&\alpha ~|~ \tau \rightarrow \tau ~|~ \{m_1\!:\!\sigma_1, \ldots, m_n\!:\!\sigma_n\} &\quad(\textit{Types})\\
 \sigma &\in& (\tau_1, \ldots, \tau_n) \rightarrow \tau &\quad(\textit{Members})\\
@@
 \end{equation*}
-
-\newpage
-~
-\newpage
+%
+The typing judgements are written in the form $\bnd{v}\vdash \tau$. They are parameterised by the
+dependency graph $(V, E)$, but this is not modified during type checking so we keep it implicit
+rather than writing e.g.~$\bnd{v}\vdash_{(V,E)} \tau$.
+
+\paragraph{Type checking.}
+The typing rules are shown in Figure~\ref{fig:tc}. Types of primitive values $n$ are obtained
+using a global lookup table $\Sigma$ (\emph{val}). When type checking a member call (\emph{mem}),
+we find its dependencies $\bndclr{v_i}$ and check that the first one (instance) is an object with 
+the required member $m$. The types of input arguments of the member then need to match the types 
+of the remaining (non-instance) nodes.
+
+Type checking a function (\emph{fun}) and a variable (\emph{var}) is similar. In both cases,
+we follow the \blbl{callsite} edge to find the member that accepts the function as an argument.
+We obtain the type of the function from the type of the $i^\textnormal{th}$ argument of the 
+member. We use the input type as the type of variable (\emph{var}). For functions, we also check 
+that the resulting type matches the type of the body (\emph{fun}).
+
+\paragraph{Caching results.}
+Performing type checking over the dependency graph, rather than over the abstract syntax tree,
+enables us to reuse the results of previously type checked parts of a program. As when 
+caching evaluated previews (Section~\ref{sec:previews}), we build a lookup table mapping graph
+nodes to types. When type checking a node, we first check the cache and, only if it is new, 
+follow the $\vdash$ relation to obtain the type.
+
+As a result, code can be type checked on-the-fly during editing, even when asynchronous type 
+providers are used, and the programmer gets instant feedback without delays.
 
@@
 
+\section{Summary}
+The key trick is to separate the process into faster binding which constructs dependency
+graph and slower type checking and evaluation. (This assumes that binding can be done, perhaps
+as an over-approximation without full type checking.)
+
 \section*{References}

### 2017-11-17 live/paper.tex
Finish paper

@@
-\documentclass[sigplan,10pt]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+\documentclass[sigplan,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
 
@@
 
-Using a text editor is challenging as any edit can significantly change the structure of code and
-fully recomputing previews after every change is too expensive. The presented interpreter partially 
-evalautes programs as they are written and reuses results of previous evaluations, making it 
-possible to provide instant feedback during development.
+Supporting a text editor is challenging as any edit can significantly change the structure of code and
+fully recomputing previews after every change is too expensive. We present a technique that allows the type
+checker and the interpreter to run on the fly, while code is written, and reuse results of previous 
+runs. This makes it possible to efficiently provide instant feedback and live previews during development.
 
 We formalise how programs are interpreted and how previews are computed, prove correctness
-of the previews and formally specify when can previews be reused. We hope this work will provide
+of the previews and formally specify when can previews be reused. We believe this work provides
 solid and easy to reuse foundations for the emerging trend of live programming environments.
@@
 One of the aspects that make spreadsheet tools such as Excel more accessible than programming 
-environments is their liveness. When you change a value in a cell in Excel, the whole spreadsheet
-is updated instantly and you see the new results immediately.
+environments is their liveness. When you change a value in a cell in Excel \cite{spreadsheet}, the  
+whole spreadsheet is updated instantly and you see the new results immediately.
 
@@
 
-This is not difficult if we represent the program in a structured way and allow the user
-to edit code only via primitive operations such as ``extract variable'' (which has
+This is not difficult if we represent the program in a structured way \cite{structure-based,hazelnut} 
+and allow the user to edit code via known operations such as ``extract variable'' (which has
 no effect on the result) or ``change constant value'' (which forces recomputation of subsequent
@@
 
+\item We follow the same method to implement a type checker (Section~\ref{sec:types}) for our
+  language that is capable of reusing previous results. This makes it possible to efficiently support
+  asynchronously provided types (Section~\ref{sec:types-providers}).
+
 \item In more speculative conclusions (Section~\ref{sec:design}), we consider alternative
@@
 \noindent
-We hope the architecture and its formal presentation in this paper can contribute important 
-foundations to the growing and important trend of text-based live coding environments.
+We hope the architecture and its formal presentation in this paper can contribute crucial 
+foundations for the growing and important trend of text-based live coding environments.
 
@@
 
-% https://cacm.acm.org/blogs/blog-cacm/169199-data-science-workflow-overview-and-challenges/fulltext
-
-The Gamma aims to make basic data exploration accessible to non-experts, such as data journalists [X], who
-need to use data transparently. As such, we focus on simple scripts that can be written using
-tools such as Jupyter notebooks~[X]. Those scripts follow a typical data science workflow [X];
+The Gamma aims to make basic data exploration accessible to non-experts, such as data journalists \cite{ddj}, 
+who need to use data transparently. As such, we focus on simple scripts that can be written using
+tools such as Jupyter notebooks \cite{jupyter}. Those scripts follow a typical data science workflow \cite{workflow};
 they acquire and reformat data, run a number of analyses using the clean data and then create 
@@
 in The Gamma can be expressed using member access via dot. The underlying mechanism is based
-on type providers [X,Y] and the specific type provider used in this example has been described 
-elsewhere [X].
+on type providers \cite{providers-fsharp,providers-idris}. The specific type provider used in 
+the above example has been described elsewhere \cite{gamma}.
 
@@
 
-The method described in this paper does not provide live preview for sub-expressions that
-contain free variables (which are not global objects provided by a type provider), but we
-describe possible ways of doing so in Section~X and more speculative design of live coding 
-friendly fucntions in Section~Y.
+The method described in this paper obtains delayed previews for sub-expressions that contain
+free varaibles (which could be fully evaluated if values for free variables were provided), but
+we describe a more speculative design of live coding friendly functions in Section~\ref{sec:design}.
 
@@
 
-The key idea behind our implementation is to maintain a dependency graph with nodes 
-representing individual operations of the computation that can be partially evaluated
+The key idea behind our implementation is to maintain a dependency graph \cite{dependencies} with 
+nodes representing individual operations of the computation that can be partially evaluated
 to obtain a preview. Each time the program text is modified, we parse it afresh (using an
@@
 delayed preview $\llbracket e_0.m(e_1) \rrbracket_{x, y}$. This operation is akin to expression
-splicing from meta-programming [X,Y] and can be more formally captured by Contextual Modal Type 
-Theory (CMTT) as outlined below.
+splicing from meta-programming \cite{metaml,quotations} and can be more formally captured by 
+Contextual Modal Type Theory (CMTT) as outlined below.
 
@@
 but it is worth noting that the structure of delayed previews is closely linked to the work
-on Contextual Modal Type Theory (CMTT) [X] and comonads [Y].
+on Contextual Modal Type Theory (CMTT) \cite{cmtt} and comonads \cite{cmtt-denotation}.
 
@@
 the splicing operation used in (\emph{mem-expr}). In categorical terms, the context-dependent
-proposition can be modelled as an indexed comonad [X]. The evaluation of a preview with no 
+proposition can be modelled as an indexed comonad \cite{effectrev,graded}. The evaluation of a preview with no 
 context dependencies (built implicitly into our evaluation rules) corresponds to the counit
@@
 Data available in The Gamma can be defined using several kinds of type providers. The type provider
-used in Figure~\ref{fig:thegamma} as asynchronous [X]. It downloads the sample URL and generates 
+used in Figure~\ref{fig:thegamma} as asynchronous \cite{async}. It downloads the sample URL and generates 
 types based on the contents of the web page. The parameter to \ident{web.scrape} is a static 
 parameter and is evaluated during type-checking. We omit details in this paper, but we note this
-works similarly to F\# type providers [X].
-
-Type providers can also be implemented as REST services [X] to allow anyone implement a data source
+works similarly to F\# type providers \cite{providers-fsharp}.
+
+Type providers can also be implemented as REST services \cite{rest} to allow anyone implement a data source
 in the language of their choice. In this case, each member of a call-chain returns a type that is
@@
 $\ident{Movie}\rightarrow\ident{bool}$ and so the type of $x$ is \ident{Movie}. This is similar to
-type checking of lambda functions in C\# [X], where type is also inferred from the context (or has
+type checking of lambda functions in C\# \cite{csharp}, where type is also inferred from the context (or has
 to be specified explicitly).
@@
 member access, written as $e.m(e_1,\ldots, e_n)$. In this paper, we do not define how member
-access evaluates. This has been done elsewhere [X], but more importantly, the evaluation of
+access evaluates. This has been done elsewhere \cite{gamma}, but more importantly, the evaluation of
 previews does not rely on the exact specifics of the evaluation, provided that the language
@@
 
-a
-
-% --------------------------------------------------------------------------------------------------
-
-\section{Takeaways}
+As noted in Section~\ref{sec:types}, the focus of this paper is on live previews, but we also
+use the method based on reusing nodes in a dependency graph for type checking. We do not discuss 
+properties of type checking in detail, but we briefly note how the different properties of 
+live previews extend to corresponding properties of type checking.
+
+\paragraph{Type checking result reuse.} In Section~\ref{sec:properties-reuse}, we show that 
+certain source code edits do not cause the recomputation of previews for the whole expression
+or a sub-expression. The edits are given in Figure~\ref{fig:operations}. The proof uses the fact
+that the newly bound graph (after code edit) reuses nodes of the previous graph. This implies that
+type checking results can be reused in exactly the same way as live previews -- they are also 
+stored in a lookup table with graph nodes as keys.
+
+\paragraph{Correctness.} The correctness property (Theorem~\ref{thm:correctness}) shows that 
+graph-based preview evaluation matches direct evaluation of expressions. To show a corresponding
+property for type checking, we would need to provide ordinary type system based on the structure
+of the expression and prove that the two are equivalent. In our implementation, we only use the
+presented graph-based type checking method, so we do not provide an alternate account in this paper.
+
+\paragraph{Determinacy.} The determinacy property (Theorem~\ref{thm:determinacy}) guarantees that
+previews can be cached, because evaluating them again, using $\Downarrow$ over an updated graph, 
+would yield the same result. The same property holds for $\vdash$, meaning that type checking 
+results can be cached. Although the Theorem~\ref{thm:determinacy} talks explicitly about 
+$\Downarrow$, it can be easily extended for $\vdash$, because the proof depends on how the graph
+is updated using $\ident{update}_{V,E}$ and the binding process.
+
+% ==================================================================================================
+
+\section{Design lessons}
 \label{sec:design}
 
-\section{Related and future}
+The motivation for the presented work, briefly outlined in Section~\ref{sec:live}, is to build a 
+simple data exploration environment that would allow non-experts, like data journalists, 
+transparently work with data. In this paper, we focused on providing live coding experience, 
+which is one important step towards the goal. However, the language we use is a mix of established 
+object-oriented (member access) and functional (function values) features with type providers.
+
+If we were to design a new programming language, there are lessons we can learn from the 
+cases that make type checking and preview evaluation in this paper difficult. This section 
+briefly considers those.
+
+\paragraph{Functions and type providers.} When using type providers in a nominally-typed language,
+the provided types are named, but the names are typically not easy to type \cite{fsdata}. This is
+not a problem in typical usage where provided members are accessed via dot. Using the \ident{worldbank}
+example from Section~\ref{sec:types-providers}, we can access population of two countries using:
+%
+\begin{equation*}
+\begin{array}{l}
+\ident{worldbank.byCountry.\textquotesingle United Kingdom\textquotesingle}.\\
+\quad \ident{Indicators.\textquotesingle Population (total)\textquotesingle}\\[0.5em]
+\ident{worldbank.byCountry.\textquotesingle Czech Republic\textquotesingle}.\\
+\quad \ident{Indicators.\textquotesingle Population (total)\textquotesingle}\hspace{9em}
+\end{array}
+\end{equation*}
+%
+However, the fact that the provided types do not have nice names becomes a problem when want to 
+extract code to access population into a function:
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~\ident{getPopulation}~c=\\
+\quad c.\ident{Indicators.\textquotesingle Population (total)\textquotesingle}\hspace{8.5em}
+\end{array}
+\end{equation*}
+%
+Here, the compiler cannot infer the type of $c$ from usage and so we are required to provide a
+type annotation using an automatically generated name.
+
+\paragraph{Functions and live previews.} 
+Providing live previews in a language with ordinary functions suffers from the same problem as 
+type checking of functions.
+
+Our live preview evaluation, discussed in Section~\ref{sec:previews}, can obtain only a delayed 
+preview for the body of \ident{getPopulation}. The delayed preview we would obtain in this case
+is $\llbracket c.\ident{Indicators}.\ident{\textquotesingle Population (total)\textquotesingle} \rrbracket_{c}$.
+
+If we know the type of $c$, we can provide a user interface that lets the user specify a value 
+for $c$ (or, more generally, free variables of the preview) and then evaluate the preview, but 
+it is difficult to provide a meaningful preview automatically.
+
+\paragraph{Reusable code blocks.}
+In data science scripting, we start with a concrete example and then turn code into a reusable
+function. This pattern could be supported by the language in a way that makes type checking
+and preview evaluation easier. Using an imaginary notation, we could write:
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~\ident{uk}~=~\ident{worldbank.byCountry.\textquotesingle United Kingdom\textquotesingle}\hspace{1.5em}\\[0.5em]
+\kvd{def}~\ident{getPopulation}~=\\
+\quad\lbrack\ident{country}\!:\!\ident{uk}\rbrack.\ident{Indicators.\textquotesingle Population (total)\textquotesingle}\\[0.5em]
+\ident{getPopulation}~\ident{worldbank.byCountry.China}\\
+\ident{getPopulation}~\ident{worldbank.byCountry.India}
+\end{array}
+\end{equation*}
+%
+This notation is speculative and we intend to implement it in future prototypes of The Gamma.
+The second line is an expression that accesses the population of the UK, using a concrete
+data source as the input, but it also defines a named function \ident{getPopulation} that has a
+parameter \ident{country}. In a way, we are providing \emph{type annotation} by example, together
+with a \emph{value annotation} that can be used for live previews.
+
+This way of constructing abstractions is perhaps more akin to how spreadsheets are used -- we often
+write a formula using a concrete cell and then use the ``drag down'' operation to extend it to 
+other inputs.
+
+% ==================================================================================================
+
+\section{Related and future work}
 \label{sec:future}
 
-Integrate with Python/R via Jupyter
-
-\newpage
+This paper approaches the problem of live coding environments from a theoretical programming
+language perspective with a special focus on tooling for data science. Hence, the related and
+future work spans numerous areas.
+
+\paragraph{Design and human-computer interaction.}
+From a design perspective, the idea of live programming environments has been popularised by 
+Bret Victor \cite{learnable}. Active research on novel forms of interaction happens in areas
+such as live coded music \cite{beyond,sonic}.
+The idea of live previews can be extended to direct manipulation \cite{direct}. The Gamma
+provides limited support for directly manipulating data (Section~\ref{sec:live}), but we 
+intend to explore this direction further.
+
+\paragraph{Data science tooling.}
+An essential tool in data science is REPL (read-eval-print-loop) \cite{drscheme}, which is now
+widely available. This has been integrated with rich graphical outputs in tools such as Jupyter
+notebooks \cite{jupyter,ipython}, but such previews are updated using an explicit command.
+Integrating our work with Jupyter to provide instant live previews for R or Python would be
+an interesting extension of the presented work.
+
+\paragraph{Live coding and live previews.}
+Live previews have been implemented in LightTable \cite{lighttable} and, more recently, in
+editors such as Chrom Developer Tools, but neither presents a simple description of their
+inner workings. An issue that attracts much attention is keeping state during code edits
+\cite{alive,livingit}. This would be an interesting problem if we extended our work to
+event-based reactive programming.
+
+\paragraph{Structured editing.}
+An alternative approach to ours is to avoid using text editors. Structured editors 
+\cite{structure-based} allow the user to edit the AST and could, in principle, recompute previews
+based on the performed operations, or preview evaluation as in interactive functional programming
+\cite{interactive}. A promising direction is using bi-directional lambda calculus~\cite{hazelnut}. 
+Finally, abandoning text also enables building richer, more human-centric 
+abstractions as illustrated by Subtext~\cite{subtext}. Our current focus, however, remains
+on text-based editors.
+
+\paragraph{Dependency analysis.}
+Our use of dependency graphs \cite{dependencies} is first-order. Building dependency graphs 
+involving function calls using modern compiler methods \cite{optimizing} or program slicing 
+\cite{slicing} would allow us to deduce possible inputs for functions and use those for 
+previews rather than changing the language as suggested in Section~\ref{sec:design}. This direction
+is worth considering, but it requires more empirical usability testing.
+
+\paragraph{Semantics and partial evaluation.}
+The evaluation of previews can be seen as a form of partial evaluation \cite{partial}, done in a
+way that allows reuse of results. This can be done implicitly or explicitly in the form of 
+multi-stage programming \cite{metaml}. Both can provide useful perspective for formally analysing
+how previews are evaluated. Semantically, the evaluation of previews can be seen as a modality
+\cite{modal} and delayed previews are linked to contextual modal type thoery \cite{cmtt}, which, 
+in turn, can be understood in terms of comonads \cite{cmtt-denotation}. This provides an intriguing
+direction for rigorous analysis of the presented system.
+
+% ==================================================================================================
 
 \section{Summary}
-The key trick is to separate the process into faster binding which constructs dependency
-graph and slower type checking and evaluation. (This assumes that binding can be done, perhaps
-as an over-approximation without full type checking.)
-
-\section*{References}
+We present The Gamma, a live coding environment for data exploration. The environment
+bridges the gap between spreadsheets and scripting -- live previews give users rapid feedback, 
+while the final result is a fully reproducible script.
+
+In this paper, we focus on the challenge of efficiently providing live previews and type checking 
+code during editing in a free-form text editor. This is a challenge, because users can perform 
+arbitrary text transformations and we cannot recompute previews after each edit.
+
+The key trick is to separate the process into fast \emph{binding phaes}, 
+which constructs a dependency graph and slower \emph{evaluation phase} and \emph{type checking phase} 
+that can cache results, using the nodes from the dependency graph created during binding as keys.
+This makes it possible to quickly parse updated code, reconstruct dependency graph and compute
+preview using previous, partially evaluated, results.
+
+We describe our approach formally, which serves two purposes. First, we aim to provide easy to
+use foundations for the growing and important trend of text-based live coding environments. 
+Second, we explore the properties of our system and prove that our method does not recompute 
+previews in a number of common cases and, at the same time, the optimisation still produces 
+correct previews.
 

### 2017-11-17 live/paper.tex
Spellcheck, tweaks

@@
 \usepackage{lineno,hyperref,xcolor}
+\usepackage{flushend}
 \usepackage{stmaryrd}
@@
 
-\title{Design and implementaiton of a live coding environment for data science}
+\title{Design and implementation of a live coding environment for data science}
 
@@
 
-As ilustrated by the second let binding in Figure~\ref{fig:thegamma}, many opertaions
+As illustrated by the second let binding in Figure~\ref{fig:thegamma}, many operations
 in The Gamma can be expressed using member access via dot. The underlying mechanism is based
@@
 It excludes constructs such as a mechanism for defining new objects as we assume that those
-are imported from the context through a mechanims such as type providers.
+are imported from the context through a mechanism such as type providers.
 %
@@
 The method described in this paper obtains delayed previews for sub-expressions that contain
-free varaibles (which could be fully evaluated if values for free variables were provided), but
+free variables (which could be fully evaluated if values for free variables were provided), but
 we describe a more speculative design of live coding friendly functions in Section~\ref{sec:design}.
@@
 
-Furthermore, the graph edges are labelled with labels indicating the kind of dependency. For
+Furthermore, the graph edges are labeled with labels indicating the kind of dependency. For
 a method call, the labels are ``first argument'', ``second argument'' and so on. Formally:
@@
 %
-The \bnd{val} node represents a primitive value and contains the value itself. Two occurences
+The \bnd{val} node represents a primitive value and contains the value itself. Two occurrences
 of $\num{10}$ in the source code will be represented by the same node. Member access \bnd{mem}
 contains the member name, together with a unique symbol -- two member access nodes with different 
-dependencies will contain a different symbol. Dependencies of member access are labelled with 
+dependencies will contain a different symbol. Dependencies of member access are labeled with 
 \blbl{arg} indicating the index of the argument (the instance has index $0$ and arguments 
@@
 abstraction. For simplicity, we use variable names rather than de Bruijn indices and so 
-renaming a bound variable foces recomputation.
+renaming a bound variable forces recomputation.
 
@@
 (\bndclr{v_n},\blblclr{l_n})])$ to perform a lookup for a node of a kind $\bkndclr{k}$ that has dependencies
-$\bndclr{v_1}, \ldots, \bndclr{v_n}$ labelled with labels $\blblclr{l_1}, \ldots, \blblclr{l_n}$.
+$\bndclr{v_1}, \ldots, \bndclr{v_n}$ labeled with labels $\blblclr{l_1}, \ldots, \blblclr{l_n}$.
 
@@
 dependency graph $(V, E)$ paired with a node $\bndclr{v}$ corresponding to the expression $e$.
-In the graph, $V$ is a set of nodes $\bndclr{v}$ and $E$ is a set of labelled edges
+In the graph, $V$ is a set of nodes $\bndclr{v}$ and $E$ is a set of labeled edges
 $(\bndclr{v_1}, \bndclr{v_2}, \blblclr{l})$. We attach the label directly to the edge rather than
@@
 $\bndclr{v_1}$. The variable context is used when binding a variable (6) and so all variables 
-declared using \kvd{let} binding will be bound to a graph node representing the value assinged 
+declared using \kvd{let} binding will be bound to a graph node representing the value assigned 
 to the variable. The node bound to the overall \kvd{let} expression is then the graph node bound
@@
 \paragraph{Evaluation and splicing.}
-In this paper, we omit the specifics of the underlying programming langauge and we focus on the
+In this paper, we omit the specifics of the underlying programming language and we focus on the
 live coding mechanism. However, we assume that the language is equipped with an evaluation 
@@
 produce a value. If one of the arguments is delayed (\emph{member-expr}), because the member access 
-is in the body of a lambda function, then we produce a delayed member access experession that
+is in the body of a lambda function, then we produce a delayed member access expression that
 requires the union of the variables required by the individual arguments.
@@
 In CMTT, $\lbrack \Psi \rbrack A$ denotes that a proposition $A$ is valid in context $\Psi$,
-which is closely related to our delayed preivews written as $\llbracket A \rrbracket_\Psi$.
+which is closely related to our delayed previews written as $\llbracket A \rrbracket_\Psi$.
 CMTT defines rules for composing context-dependent propositions that would allow us to express
 the splicing operation used in (\emph{mem-expr}). In categorical terms, the context-dependent
-proposition can be modelled as an indexed comonad \cite{effectrev,graded}. The evaluation of a preview with no 
+proposition can be modeled as an indexed comonad \cite{effectrev,graded}. The evaluation of a preview with no 
 context dependencies (built implicitly into our evaluation rules) corresponds to the counit
@@
 In the binding process specified in Section~\ref{sec:formal}, a variable is a leaf of the dependency
-graph. In the revised model, it depends on the context in which it appears. A new edge labelled
+graph. In the revised model, it depends on the context in which it appears. A new edge labeled
 $\blbl{callsite}(m, i)$ indicates that the source node is the input variable of a function 
@@
 The \ident{bind} function now has a parameter $c$, in addition to $\Gamma$ and $\Delta$, which 
-represents the context in which the binding happens. This is either a member invocation (labelled 
+represents the context in which the binding happens. This is either a member invocation (labeled 
 with instance node $\bndclr{v}$ and \blbl{callsite} label $\blblclr{l}$), or not a member 
-invocation writted as $\bot$. The updated definitions are:
+invocation written as $\bot$. The updated definitions are:
 %
@@
 %
-The key parts of the revised definiton of the \ident{bind} function are shown in Figure~\ref{fig:binding-rules-callsite}.
+The key parts of the revised definition of the \ident{bind} function are shown in Figure~\ref{fig:binding-rules-callsite}.
 We now write $\ident{bind}_{\Gamma,\Delta,c}$ where $c$ represents the context in which the
@@
 
-\subsection{Modelling expression evaluation}
+\subsection{Modeling expression evaluation}
 In The Gamma language, computations are expressed using
@@
 previews does not rely on the exact specifics of the evaluation, provided that the language
-satisfies certan basic conditions. The following definitions provides the necessary structure 
+satisfies certain basic conditions. The following definitions provides the necessary structure 
 for discussing correctness of previews.
@@
 
-To show that the evalauted previews are correct, we prove two properties. Correctness 
+To show that the evaluated previews are correct, we prove two properties. Correctness 
 (Theorem~\ref{thm:correcntess}) guarantees that, no matter how a graph is constructed, when we use
@@
 keeps a preview or a partial preview $d$ in a lookup table indexed by nodes $\bndclr{v}$. The
-Throrem~\ref{thm:determinacy} guarantees that this is a valid strategy. As we update dependency
+Theorem~\ref{thm:determinacy} guarantees that this is a valid strategy. As we update dependency
 graph during code editing, previous nodes will continue representing the same sub-expressions.
@@
 
-Figure~\ref{fig:operations} shows a list of six code edit operations where a previeww of the 
+Figure~\ref{fig:operations} shows a list of six code edit operations where a preview of the 
 expression (cases 1-4), or a sub-expression (cases 5-6), can be reused. This is the case, because 
@@
 are bound in sequence and graph node cache updated as specified in Figure~\ref{fig:loop}, then 
-the graph nodes assigned to the specifeid sub-expressions are the same.
+the graph nodes assigned to the specified sub-expressions are the same.
 \end{theorem}
@@
 transparently work with data. In this paper, we focused on providing live coding experience, 
-which is one important step towards the goal. However, the language we use is a mix of established 
+which is one important step toward the goal. However, the language we use is a mix of established 
 object-oriented (member access) and functional (function values) features with type providers.
@@
 Live previews have been implemented in LightTable \cite{lighttable} and, more recently, in
-editors such as Chrom Developer Tools, but neither presents a simple description of their
+editors such as Chrome Developer Tools, but neither presents a simple description of their
 inner workings. An issue that attracts much attention is keeping state during code edits
@@
 how previews are evaluated. Semantically, the evaluation of previews can be seen as a modality
-\cite{modal} and delayed previews are linked to contextual modal type thoery \cite{cmtt}, which, 
+\cite{modal} and delayed previews are linked to contextual modal type theory \cite{cmtt}, which, 
 in turn, can be understood in terms of comonads \cite{cmtt-denotation}. This provides an intriguing
@@
 
-The key trick is to separate the process into fast \emph{binding phaes}, 
+The key trick is to separate the process into fast \emph{binding phase}, 
 which constructs a dependency graph and slower \emph{evaluation phase} and \emph{type checking phase} 

### 2018-03-31 wip-live-programming/paper.tex
Cleanup, start revising

--- 2017-11-17 live/paper.tex
@@
-\documentclass[sigplan,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+\documentclass[sigplan,10pt]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
 
@@
 \author{Tomas Petricek}
-\affiliation{
-  \institution{The Alan Turing Institute}
-  \country{London, United Kingdom}
-}
-\email{tomas@tomasp.net}
+%\affiliation{
+%  \institution{The Alan Turing Institute}
+%  \country{London, United Kingdom}
+%}
+%\email{tomas@tomasp.net}
 
@@
 
-\paragraph{Reusable code blocks.}
+\paragraph{Wormhole abstractions.}
 In data science scripting, we start with a concrete example and then turn code into a reusable
@@
 %
-This notation is speculative and we intend to implement it in future prototypes of The Gamma.
+We tentatively call this notation \emph{wormhole} abstraction and we intend to implement it in future prototypes of The Gamma.
 The second line is an expression that accesses the population of the UK, using a concrete
@@
 
+\newpage
 \appendix
@@
 
+\newpage  
 \vspace{0.75em}\noindent(2) $\ident{bind}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n))$ --

### 2018-10-20 wip-live-programming/paper.tex
Uncommitted changes

@@
 \usepackage{lineno,hyperref,xcolor}
+\usepackage{ragged2e}
 \usepackage{flushend}
@@
 
-\title{Design and implementation of a live coding environment for data science}
+\title{What you see is what you write}
+\subtitle{On the fly evaluation of live previews for data manipulation scripts}
 
 \author{Tomas Petricek}
-%\affiliation{
-%  \institution{The Alan Turing Institute}
-%  \country{London, United Kingdom}
-%}
-%\email{tomas@tomasp.net}
+\affiliation{
+  \institution{The Alan Turing Institute}
+  \country{London, United Kingdom}
+}
+\email{tomas@tomasp.net}
 
@@
 \begin{abstract}
-Data science can be done by directly manipulating data using spreadsheets, or by writing 
-data manipulation scripts using a programming language. The former is error-prone and
-does not scale, while the latter requires expert skills. Live coding has the potential to 
-bridge this gap and make writing of transparent, reproducible scripts more accessible.
-
-In this paper, we describe a live programming environment for data science that provides instant 
-previews and contextual hints, while allowing the user to edit code in an unrestricted way in
-a text editor.
-
-Supporting a text editor is challenging as any edit can significantly change the structure of code and
-fully recomputing previews after every change is too expensive. We present a technique that allows the type
-checker and the interpreter to run on the fly, while code is written, and reuse results of previous 
-runs. This makes it possible to efficiently provide instant feedback and live previews during development.
-
-We formalise how programs are interpreted and how previews are computed, prove correctness
-of the previews and formally specify when can previews be reused. We believe this work provides
-solid and easy to reuse foundations for the emerging trend of live programming environments.
+Liveness is one of the aspects that make spreadsheets easy to use. When you edit a formula in 
+Excel, the whole spreadsheet is automatically updated and you see the result immediately. 
+Providing the same behaviour for data manipulation scripts would make simple data scripting 
+more accessible, but it is difficult to do efficiently while allowing users to edit code in 
+an unrestricted way in a text editor.
+
+In this paper, we describe an efficient method for evaluating live previews for simple data 
+manipulation scripts. This is challenging as any edit in a text editor can significantly 
+change the structure of code and fully recomputing previews after every change is too expensive. 
+Our method allows the interpreter to run while code is written and reuse results of previous 
+runs. This makes it possible to efficiently provide live previews on the fly during development.
+
+We formalise how live previews are computed, prove their correctness and formally specify when 
+a live preview can be reused. We also demonstrate the efficiency of our method using an empirical 
+study. Although our focus is on a simple data scripting language, the work provides easy to use 
+foundations that can be shared by a wide range of live programming environments.
 \end{abstract}
@@
 
-One of the aspects that make spreadsheet tools such as Excel more accessible than programming 
-environments is their liveness. When you change a value in a cell in Excel \cite{spreadsheet}, the  
-whole spreadsheet is updated instantly and you see the new results immediately.
-
-Increasing number of programming environments aim to provide the same live experience for more
-standard programming languages, but doing this is not easy. Fully recomputing the whole program after
-every single change is inefficient and calculating how a change in source code changes the result
-is extremely hard when the editor allows arbitrary manipulation of program text. For example, 
-consider the following simple program that gets the release years of 10 most expensive movies 
-in a data set \ident{movies}:
-%
-\begin{equation*}
-\begin{array}{l}  
-\kvd{let}~\ident{top} =\ident{movies}\\
-\quad .\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\ident{.take}(\num{10})\\
-\quad .\ident{map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{yyyy}))}\\
-\end{array}
-\end{equation*}
-%
-A live coding environment shows a preview of the list of dates. Next, assume that the programmer
-modifies the code by making the constant $\num{10}$ a variable and changing the date format to see 
-the full date:
-%
-\begin{equation*}
-\begin{array}{l}  
-\kvd{let}~\ident{count} = \num{10}\\
-\kvd{let}~\ident{top} = \ident{movies}\\
-\quad .\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\ident{.take}(\ident{count})\\
-\quad \ident{.map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{dd-mm-yyyy}))}\\
-\end{array}
-\end{equation*}
-%
-Ideally, the live coding environment should understand the change, reuse a cached result of the
-first two transformations (sorting and taking 10 elements) and only evaluate the last \ident{map}
-to differently format the release dates of already computed top 10 movies.
+Data science tools such as Jupyter notebooks \cite{jupyter} and programming environments
+with REPLs \cite{rmarkdown,drscheme} embed program outputs directly into the environment, 
+but the outputs are only updated after the user finishes writing code and explicitly runs it.
+Doing better is not easy. Fully recomputing the whole program after every single key stroke is 
+inefficient and calculating how a change in source code changes the result is extremely hard 
+when the editor allows arbitrary manipulation of program text.
+
+For example, consider two steps of a data manipulation process that are documented in 
+Figure~\ref{fig:intro-live}. The user first writes code to load and aggregate broadband speed data 
+(described in Appendix~\ref{sec:app-broadband}). Using a live preview, she spots a problem with 
+the data and decides to fix it by ignoring invalid rows. She then changes the code by adding
+a call to filter the data and also extracts a constant into a named variable.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[b]
+\caption{\small{Exploring broadband speed data. We group reports by whether they are from urban
+  or rural area and calculate average download speeds. Resulting table is displayed as we type.}}
+\label{fig:intro-live}
+
+\vspace{0.5em}
+\noindent
+\includegraphics[scale=0.38]{bb1.png}
+
+\vspace{0.5em}
+\justify
+\small{We notice that Urban/rural may be empty and filter out empty values and we
+  extract the URL into a variable. To update live preview, the system needs to evaluate
+  only the last filtering operation.}
+
+\vspace{0.75em}
+\noindent
+\includegraphics[scale=0.38]{bb2.png}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+A system that provides live previews should understand the code change, reuse cached result of data 
+loading and aggregation and only evaluate the filtering operation. Regardless of the size of the 
+original input, this will be fast as it filters an aggregated series with just three rows.
 
@@
 and allow the user to edit code via known operations such as ``extract variable'' (which has
-no effect on the result) or ``change constant value'' (which forces recomputation of subsequent
-transformations). However, many programmers prefer to edit programs as free form text.
+no effect on the result) or ``add operation to a chain'' (which requires evaluation of the new
+transformation). However, many programmers prefer to edit programs as free form text.
 
@@
 \section{Appendix}
+\label{sec:app-broadband}
+
+\section{Appendix}
 \label{sec:app-correctness}

### 2018-11-14 wip-live-programming/paper.tex
Rewriting the background section

--- 2018-03-31 wip-live-programming/paper.tex
@@
 \newcommand{\ident}[1]{\textnormal{\sffamily #1}}
-\newcommand{\qident}[1]{\textnormal{\sffamily \guillemotleft #1\guillemotright}}
+\newcommand{\qident}[1]{\textnormal{\sffamily \textquotesingle #1\textquotesingle}}
 \newcommand{\dom}{\ident{dom}}
@@
 
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure*}
-\includegraphics[scale=0.21]{wiki.png}
-\includegraphics[scale=0.21]{drop.png}
-\caption{Scraping 2016 science events. Preview of the Wikipedia source page (left) and live filtering (right).}
-\label{fig:thegamma}
-\end{figure*}
+\newpage
 
@@
 
-\section{Live programming for data exploration}
-\label{sec:live}
-
-The Gamma aims to make basic data exploration accessible to non-experts, such as data journalists \cite{ddj}, 
-who need to use data transparently. As such, we focus on simple scripts that can be written using
-tools such as Jupyter notebooks \cite{jupyter}. Those scripts follow a typical data science workflow \cite{workflow};
-they acquire and reformat data, run a number of analyses using the clean data and then create 
-several visualizations.
-
-An important property of data science workflow is that we work with readily available concrete data.
-Data scientists load inputs into memory and refer to it in subsequent REPL interactions. They might 
-later wrap completed working code into a function (and run on multiple datasets), but 
-do not start with functions. We reflect this pattern in our language.
-
-In this section, we provide a brief overview of The Gamma, a simple live coding environment for
-data science. We discuss the language, type providers and the user interface,
-before focusing on the algorithms behind live previews in Section~\ref{sec:formal}.
-
-\subsection{The Gamma scripting language}
-Figure~\ref{fig:thegamma} shows The Gamma script that scrapes items from a Wikipedia page, 
-collects those marked as ``Events'' and filters them. The script illustrates two aspects of the
-scripting language used by The Gamma -- its structure and its use of type providers for
-dot-driven data exploration.
-
-The scripting language is not intended to be as expressive as, say, R or Python and so it
-has a very simple structure -- scripts are a sequence of let-bindings (that obtain or transform 
-data) or statements (that produce visualizations). This reflects the fact that we always work
-with concrete data and allows us to provide previews.
-
-The most notable limitation of The Gamma is that the scripting language does not support top-level 
-functions. This is not a problem for the simple scripts we consider, but it would be an issue for
-a general-purpose language. We discuss potential design for functions that would still be based
-on working with concrete data in Section~\ref{sec:design}.
-
-\subsection{Dot-driven data exploration}
-
-As illustrated by the second let binding in Figure~\ref{fig:thegamma}, many operations
-in The Gamma can be expressed using member access via dot. The underlying mechanism is based
-on type providers \cite{providers-fsharp,providers-idris}. The specific type provider used in 
-the above example has been described elsewhere \cite{gamma}.
-
-Given a data source (scraped Wikipedia page), the type provider generates type with members that allow
-a range of transformations of the data such as grouping, sorting and filtering. Some of the members
-are based only on the schema (e.g.~\ident{\textquotesingle Item Contains\textquotesingle} or 
-\ident{\textquotesingle Drop Heading 2\textquotesingle}), but some may also be generated based
-on (a sample of) the dataset (e.g.~the second member in \ident{\textquotesingle Heading 2 
-is\textquotesingle.\textquotesingle Events[edit]\textquotesingle}).
-
-What matters for the purpose of this paper is 
-the fact that most operations are expressed via member access matters. First, it means 
-that we need to provide live previews for sub-expressions formed by a chain of member accesses.
-Second, it means that the result of any member access expression depends on the instance and,
-possibly, on the parameters provided for a method call.
-
-\subsection{Direct manipulation and live previews}
-The screenshot in Figure~\ref{fig:thegamma} show the editor as implemented in The Gamma. This
-includes live previews (discussed in this paper), but also an editor that provides spreadsheet-like
-interface for editing the data transformation script.
-
-In our implementation, previews appears below the currently selected let binding.
-The preview can be of several types such as web page (left) or data table (right). In this paper,
-we describe how we evaluate scripts to obtain the resulting object and we ignore how such objects
-are rendered.
-
-The Figure~\ref{fig:thegamma} also shows a special handling of expressions constructed using the
-pivot type provider. The editor recognises individual data transformations and provides a simple
-user interface for adding and removing transformations and changing their parameters that changes
-the source code accordingly. This aspect is not discussed in the present paper.
+\section{Understanding how data scientists work}
+
+Data scientists often use general-purpose programming languages such as Python, but the kind of 
+code they write and the way they interact with the system is very different from how software 
+engineers work.
+
+This paper focuses on simple data wrangling and data exploration as done, for example, by 
+journalists analysing government datasets. This section illustrates how such data analyses 
+look and provides a justification for design decisions that we make when designing the data 
+exploration calculus.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Data wrangling in Jupyter}
+\label{sec:background-jupyter}
+
+Increasing numbers of data analysts use notebook systems such as Jupyter or R Markdown, which 
+make it possible to combine text, equations and code with results of running the code such as tables 
+or visualizations. Notebooks blur the traditional distinction between development and execution. 
+Data analysts write small snippets of code, run them immediately to see results and then revise 
+them interactively based on the results. We aim to take this trend further and build foundations
+for seamlessly integrating development and execution in a notebook-like environment.
+
+Jupyter notebooks are used by a variety of users ranging from scientists who implement complex
+models to journalists who load data, perform simple aggregations and create visualizations. Our
+focus is on simple use cases, such as those of journalists, because we believe that this is where
+better tooling can provide the most value.
+
+As an example, consider the data anlysis done by the Financial Times for an article on plastic 
+waste. The analysis joins datasets from Eurostat, UN Comtrade and more, aggregates the data and 
+builds a visualization comparing the plastic waste flows in 2017 and 2018.
+Figure~\ref{fig:ft-uncomtrade} shows an extract from one notebook from the data analysis. 
+The code of the analysis has a number of remarkable properties:
+%
+\begin{itemize}
+\item There is no abstraction. The analysis uses lambda functions as arguments to library calls,
+  but it does not define any functions. Parameterization is done either by defining a list of 
+  inputs and then having a for loop, or by defining a top-level variable and setting its value.
+  In our example, \ident{material} is set to \str{plastics} and a comment lists other options. 
+  This lets the analyst easily see and check the results of intermediate steps of the analysis.
+
+\item The code is structured as a sequence of commands. Some commands define a variable, either by
+  loading data, or by transforming data loaded previously. Even in Python, data is often treated
+  as immutable. Other commands produce an output that is displayed in the notebook.
+
+\item There are many corner cases, such as the fact that the \ident{keep\_default\_na} parameter 
+  needs to be set to handle Namibia correctly. These are discovered interactively by writing and 
+  running a code snippet and examining the output, so providing a rapid feedback is essential.
+\end{itemize}
+
+Many Jupyter notebooks are more complex than the example discussed here. They might define
+helper functions or even structure code using object-oriented programming. However, simple 
+data analyses such as the one discussed here are frequent enough to pose an interesting and
+important problem for programming tools research.
+
+\begin{figure}
+\includegraphics[scale=0.6]{notebook.png}
+\caption{An excerpt from a Jupyter notebook by Financial Times reporters, which loads trade
+  data from the UN trade database and adds ISO country codes.}
+\label{fig:ft-uncomtrade}
+\end{figure}
+
+\begin{figure}
+\includegraphics[scale=0.52]{gamma.png}
+\caption{The Gamma script to aggregate waste data. We select three countries and group
+data by country. We then count the number of distinct reporting periods per country and type `.'
+to choose further aggregation operations.}
+\label{fig:gamma-waste}
+\vspace{-1em}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Dot-driven data exploration in The Gamma}
+\label{sec:background-gamma}
+
+Data exploration of the kind discussed in the previous section has been the motivation
+for a recent programming language research project The Gamma. The scripting language in The Gamma
+captures the structure discussed above.
+
+A script is a sequence of commands that can either define a 
+variable or produce an output. The language does not support top-level function declarations, 
+although we propose an extension that provides a simple abstraction mechanism and is compatible with
+providing live previews in Section~X. Lambda functions can be used only as method 
+arguments.
+
+Given the expressiveness of The Gamma, libraries for it are implemented in other languages, such as
+JavaScript. More importantly, The Gamma supports type providers, which can be used to generate types,
+on the fly, for accessing and exploring external data sources. Type providers provide object types 
+with members and The Gamma makes using those convenient by providing auto-complete when the user 
+types the dot symbol (`.') to access a member.
+
+The combination of type providers and auto-complete on~``.'' makes it possible to complete a 
+large number of data exploration tasks through a very simple interaction of selecting operations from
+a list. An example in Figure~\ref{fig:gamma-waste} summarizes waste data using a type provider.
+The names of identifiers such as \qident{average Waste (kg)} do not have
+any special meaning -- they are merely members generated by the type provider, based on information
+about the data source. The type provider used in this example generates an object with members for
+different data transformations, such as \qident{group data}, which return furhter objects with 
+members for specifying transformation parameters, such as selecting the grouping key using
+\qident{by Country}.
+
+The Gamma is more complex than shown here -- it supports methods that can take
+functions as arguments -- but the above example illustrates the fact that non-trivial data
+exploration can be done using a very simple language.
+
+We complement our theoretical work on foundations of live programming environments for data
+exploration with an implementation for The Gamma. As discussed in the previous
+section, the assumptions about structure of code that are explicit in The Gamma are also
+implicitly present in Python and R data analyses produced by journalists, economists and
+other users with other than  programming background. Consequently, our theoretical work 
+applies to simple data analyses using Python and readers not familiar with The Gamma
+can assume that we use a small subset of Python.
+
+% ==================================================================================================
+
+\section{Calculus}
+
+x
+
+\begin{equation*}
+\begin{array}{lcl}
+p &=& c_1; \ldots; c_n\\
+c &=& \kvd{let}~x = e ~|~ e\\
+e &=& n ~|~ x ~|~ e.m(a, \ldots, a)\\
+a &=& e ~|~ \lambda x\rightarrow e
+\end{array}
+\end{equation*}
+
+\newpage
 

### 2018-12-06 wip-live-programming/paper.tex
Add section on live programming, working on calculus

@@
-\documentclass[sigplan,10pt]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
-
+\documentclass[acmsmall,anonymous,fleqn]{acmart}\settopmatter{printfolios=false,printccs=false,printacmref=false}
+
+%% For double-blind review submission, w/ CCS and ACM Reference
+%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
+%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
+%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+%% For single-blind review submission, w/ CCS and ACM Reference
+%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
+%% For final camera-ready submission, w/ required CCS and ACM Reference
+%\documentclass[acmsmall]{acmart}\settopmatter{}
+
+\acmJournal{PACMPL}
+\acmVolume{1}
+\acmNumber{OOPSLA}
+\acmArticle{1}
+\acmYear{2019}
+\acmMonth{1}
+\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
+\startPage{1}
+
+%% Copyright information
+%% Supplied to authors (based on authors' rights management selection;
+%% see authors.acm.org) by publisher for camera-ready submission;
+%% use 'none' for review submission.
+\setcopyright{none}
+%\setcopyright{acmcopyright}
+%\setcopyright{acmlicensed}
+%\setcopyright{rightsretained}
+%\copyrightyear{2018}           %% If different from \acmYear
+
+%% Bibliography style
+\bibliographystyle{ACM-Reference-Format}
+%% Citation style
+%% Note: author/year citations are required for papers published as an
+%% issue of PACMPL.
+\citestyle{acmauthoryear}   %% For author/year citations
+
+\usepackage{booktabs}
+\usepackage{subcaption}
 \usepackage{lineno,hyperref,xcolor}
@@
 \usepackage{enumitem}
-\usepackage[shortlabels]{enumerate}
+\usepackage{enumerate}
+\usepackage{amsmath}
+
 \setlist{leftmargin=6mm}
-
 \newcounter{thc}
@@
 
-\acmConference[PL'17]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
-\acmYear{2017}
-\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
-\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
-\startPage{1}
-\setcopyright{none}
-\bibliographystyle{ACM-Reference-Format}
-
-\title{Design and implementation of a live coding environment for data science}
-
+\title{Live coding}
 \author{Tomas Petricek}
-%\affiliation{
-%  \institution{The Alan Turing Institute}
-%  \country{London, United Kingdom}
-%}
-%\email{tomas@tomasp.net}
+
+\affiliation{
+  \institution{University of Kent}
+  \country{United Kingdom}
+}
+\email{tomas@tomasp.net}
+
+\begin{CCSXML}
+<ccs2012>
+<concept>
+<concept_id>10011007.10011006.10011008</concept_id>
+<concept_desc>Software and its engineering~General programming languages</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+<concept>
+<concept_id>10003456.10003457.10003521.10003525</concept_id>
+<concept_desc>Social and professional topics~History of programming languages</concept_desc>
+<concept_significance>300</concept_significance>
+</concept>
+</ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Software and its engineering~General programming languages}
+\ccsdesc[300]{Social and professional topics~History of programming languages}
 
@@
 \newcommand{\ball}[1]{\FPeval{\result}{clip(201+#1)}\textnormal{\ding{\result}}}
-\newcommand{\lsep}{~\,|\,~}
+\newcommand{\lsep}{\;\;|\;\;}
 \newcommand{\num}[1]{\textcolor{numclr}{#1}}
@@
 \section{Understanding how data scientists work}
+\label{sec:background}
 
@@
 
-Increasing numbers of data analysts use notebook systems such as Jupyter or R Markdown, which 
+Data analysts increasingly use notebook systems such as Jupyter or R Markdown, which 
 make it possible to combine text, equations and code with results of running the code such as tables 
-or visualizations. Notebooks blur the traditional distinction between development and execution. 
-Data analysts write small snippets of code, run them immediately to see results and then revise 
-them interactively based on the results. We aim to take this trend further and build foundations
-for seamlessly integrating development and execution in a notebook-like environment.
+or visualizations. Notebooks blur the distinction between development and execution. 
+Data analysts write small snippets of code, run them to see results immediately and then revise 
+them interactively based on the results. We aim to build foundations for such seamless integration 
+of development and execution in a notebook environment.
 
@@
 models to journalists who load data, perform simple aggregations and create visualizations. Our
-focus is on simple use cases, such as those of journalists, because we believe that this is where
-better tooling can provide the most value.
+initial focus is on simple use cases, such as those of journalists, because we believe that this 
+is where better tooling can provide the most value.
 
@@
 \begin{itemize}
-\item There is no abstraction. The analysis uses lambda functions as arguments to library calls,
-  but it does not define any functions. Parameterization is done either by defining a list of 
+\item There is no abstraction. The full analysis uses lambda functions as arguments to library calls,
+  but it does not define any top-level functions. Parameterization is done by defining a list of 
   inputs and then having a for loop, or by defining a top-level variable and setting its value.
@@
 
+\vspace{0.5em}
 \item The code is structured as a sequence of commands. Some commands define a variable, either by
@@
 
+\vspace{0.5em}
 \item There are many corner cases, such as the fact that the \ident{keep\_default\_na} parameter 
@@
 data analyses such as the one discussed here are frequent enough to pose an interesting and
-important problem for programming tools research.
+important domain for programming tools research.
 
@@
 
+% --------------------------------------------------------------------------------------------------
+
 \begin{figure}
-\includegraphics[scale=0.52]{gamma.png}
-\caption{The Gamma script to aggregate waste data. We select three countries and group
-data by country. We then count the number of distinct reporting periods per country and type `.'
-to choose further aggregation operations.}
-\label{fig:gamma-waste}
+\includegraphics[scale=0.28]{gamma1.png}
+\caption{\small The Gamma script to aggregate Olympic medals. We select Rio 2016 games and
+count the number of distinct events per athlete. We then type `.' to choose 
+further aggregation operations.}
+\label{fig:gamma-dot}
 \vspace{-1em}
@@
 
-% --------------------------------------------------------------------------------------------------
-
 \subsection{Dot-driven data exploration in The Gamma}
@@
 Data exploration of the kind discussed in the previous section has been the motivation
-for a recent programming language research project The Gamma. The scripting language in The Gamma
+for a recent programming environment called The Gamma. The scripting language in The Gamma
 captures the structure discussed above.
@@
 
-Given the expressiveness of The Gamma, libraries for it are implemented in other languages, such as
+Given the limited expressiveness of The Gamma, libraries for it are implemented in other languages, such as
 JavaScript. More importantly, The Gamma supports type providers, which can be used to generate types,
@@
 large number of data exploration tasks through a very simple interaction of selecting operations from
-a list. An example in Figure~\ref{fig:gamma-waste} summarizes waste data using a type provider.
-The names of identifiers such as \qident{average Waste (kg)} do not have
+a list. An example in Figure~\ref{fig:gamma-dot} summarizes data on Olympic medals using a type provider.
+The names of identifiers such as \qident{sum Bronze} do not have
 any special meaning -- they are merely members generated by the type provider, based on information
@@
 members for specifying transformation parameters, such as selecting the grouping key using
-\qident{by Country}.
+\qident{by Athlete}.
 
@@
 
-\section{Calculus}
-
-x
-
-\begin{equation*}
+% https://www.ft.com/content/360e2524-d71a-11e8-a854-33d6f82e62f8
+% https://github.com/ft-interactive/recycling-is-broken-notebooks
+
+\begin{figure}
+\includegraphics[scale=0.28 ]{gamma2.png}
+\caption{\small Live preview in The Gamma programming environment. The table is produced as
+  the data anlyst types and updates on-the-fly to show result at the current cursor position.}
+\label{fig:gamma-preview}
+\end{figure}
+
+\section{Live coding for data exploration}
+
+This paper presents the foundations of a live coding environment for data exploration that 
+follows the analysis in the previous section. We consider simple data exploration tasks, such as 
+those done by journalists and we implement our ideas in the editor for TheGamma.
+
+Figure~\ref{fig:gamma-preview} shows a TheGamma script exporing Olympic medal statistics in 
+our live coding environment. Before discussing the technical details, it is worth highlighting
+a number of important points about our approach.
+
+\paragraph{Real-world use cases.}
+Our work has been motivated by the needs outlined in the previous section. In particular, we 
+have been focusing on a langauge that is simple, yet sufficient for completing basic data 
+exploration tasks that a journalist might want to do. We have used our live coding environment
+to perform a number of simple data analyses looking at topics such as public government spending, 
+crises in financial markets and activities of a research institute. Qualitative observations about
+our experience are reported in Section~X. Our tools can be used in any modern web browser and are 
+available as an open-source package.
+
+\paragraph{Text editor integration.}
+Our approach is based on an ordinary text editor that allows unrestricted modification of the
+code. Our parser supports error-recovery and we attempt to give previews whenever the code can be
+correctly parsed and type-checked up to the current location. Notably, we do not, in general, know
+how the user modified the code. This is in contrast with approaches based on structured editing
+where the editor tracks what changes to code have been made. Using an ordinary text editor makes
+providing live previews harder, but we believe that many users still favor the flexibility of 
+plain text editors.
+
+\paragraph{Data stay, code changes.}
+Our focus has been on the scenario when input data is a static file that does not change, but the
+data analyst modifies the source code. This is in contrast with work on incremental computation,
+which focuses on scenarios where code stays the same, but input data change.
+
+That said, many of the methods that we used for evaluating live previews such as maintaining 
+the dependency graph with nodes for individual syntactic elements are closely related to methods
+used in incremental computation research. We do not attempt to combine the two, but note that
+this is an interesting future direction.
+
+\newpage
+\begin{figure}
+\raggedright
+{\small\sffamily Programs, commands, expressions, arguments and values}
+\begin{equation*}
+\hspace{-2em}
 \begin{array}{lcl}
 p &=& c_1; \ldots; c_n\\
-c &=& \kvd{let}~x = e ~|~ e\\
-e &=& n ~|~ x ~|~ e.m(a, \ldots, a)\\
-a &=& e ~|~ \lambda x\rightarrow e
-\end{array}
-\end{equation*}
+c &=& \kvd{let}~x = e \lsep e\\
+\end{array}
+\quad
+\begin{array}{lcl}
+e &=& o \lsep x \\
+  &|& e.m(a, \ldots, a)
+\end{array}
+\quad
+\begin{array}{lcl}
+a &=& e \lsep \lambda x\rightarrow e\\
+v &=& o \lsep \lambda x\rightarrow e\\
+\end{array}
+\end{equation*}
+
+\vspace{0.5em}
+{\small\sffamily Evaluation contexts of expressions}
+\begin{equation*}
+\hspace{-2em}  
+\begin{array}{lcl}
+C_e[-] &=& -.m(a_1, \ldots, a_n) ~|~ -.m(v_1, \ldots, v_{m}, -, a_1,\ldots , a_n) ~|~ -\\
+C_c[-] &=& \kvd{let}~x = C_e[-] ~|~ C_e[-]\\
+C_p[-] &=& c_1;\; \ldots;\; c_{k-1};\; C_c[-];\; c_{k+1};\; \ldots;\; c_n\\
+\end{array}
+\end{equation*}
+
+\vspace{0.5em}
+{\small\sffamily Let elimination and member reduction}
+\begin{equation*}
+\hspace{-2em}  
+\begin{array}{ll}
+\hspace{-0.5em}\begin{array}{l}  
+c_1;\; \ldots;\; c_{k-1};\; \kvd{let}~x=o;\; c_{k+1};\; \ldots; c_n \rightsquigarrow\\
+\qquad  c_1;\; \ldots;\; c_{k-1};\; c_{k+1}[x\leftarrow o];\; \ldots; c_n[x\leftarrow o]\end{array} &
+\quad\text{\small\sffamily(let)}\\[1.2em]
+{o.m(v_1, \ldots, v_n) \rightsquigarrow_\epsilon o'}\implies
+  {C_p[o.m(v_1, \ldots, v_n)] \rightsquigarrow C_p[o']} & 
+\quad\text{\small\sffamily (external)}    
+\end{array}  
+\end{equation*}
+
+\vspace{-0.5em}
+\caption{Syntax, contexts and reduction rules of the data exploration calculus}
+\label{fig:dec-calculus}
+\end{figure}
+
+\section{Data exploration calculus}
+\label{sec:calculus}
+
+The \emph{data exploration calculus} is a small, formally tractable language for data exploration. 
+The calculus is not Turing-complete and it can only be used together with external libraries that
+define what objects are available and what the behaviour of their members is. However, this is 
+sufficient to capture the simple data analyses discussed in Section~\ref{sec:background}. We 
+define the caluclus in this section and then use it as the basis for discussing our live preview 
+mechanism in Section~\ref{sec:formal}.
+
+\subsection{Language syntax}
+The calculus combines object-oriented features such as member access with functional features
+such as lambda functions. The syntax is defined in Figure~\ref{fig:dec-calculus}. We write $x$ for 
+variables and $o$ for object values. A program $p$ in the data exploration calculus consists of a 
+sequence of commands $c$. A command can be either let binding or an expression. Lambda functions 
+can only appear directly as arguments in method calls. An argument $a$ can thus be a lambda 
+function or an expression.
+
+\subsection{Operational semantics}
+The data exploration calculus is a call-by-value language. We model the evaluation as a small-step
+reduction $\rightsquigarrow$. Fully evaluating a program results in an irreducible sequence of 
+objects $o_1;\; \ldots;\; o_n$ which can be displayed as results of the data analysis. The 
+operational semantics is parameterized by a relation $\rightsquigarrow_\epsilon$ that models
+the functionality of the external libraries used with the calculus and defines reduction rules
+for member accesses. The relation has the following form:
+%
+\begin{equation*}
+o.m(v_1, \ldots, v_n) \rightsquigarrow_\epsilon o'
+\end{equation*}
+%
+foo
+
+bar
+
+\newpage
+
+\subsection{Equational theory}
+In The Gamma language, computations are expressed using
+member access, written as $e.m(e_1,\ldots, e_n)$. In this paper, we do not define how member
+access evaluates. This has been done elsewhere \cite{gamma}, but more importantly, the evaluation of
+previews does not rely on the exact specifics of the evaluation, provided that the language
+satisfies certain basic conditions. The following definitions provides the necessary structure 
+for discussing correctness of previews.
+
+Partial evaluation may reduce an expression under $\lambda$-abstraction. We do not require that
+the reduction of the host language does this. Instead, we define an extended reduction relation
+and use that in the proofs. The host language only needs to compose well with such extended reduction
+as captured by the \emph{compositionality} property below. We also require that the language allows
+elimination of let bindings.
+
+\begin{definition}[Host language]
+\label{def:host}
+Given a relation on expressions $e_1 \rightsquigarrow e_2$ that models small-step evaluation, we define:
+
+\begin{itemize}
+\item[--]  
+ A \emph{preview evaluation context} (also referred to as \emph{context}):
+
+\vspace{-1.25em}
+\begin{equation*}
+\begin{array}{lcl}
+C[-] &\hspace{-1em}=\hspace{-1em}& \kvd{let}~x = -~\kvd{in}~e ~|~ \kvd{let}~x = e~\kvd{in}~- ~|~ \lambda x\rightarrow - \\
+     &&  e_0.m(e_1, \ldots, e_{k-1}, -, e_{k+1}, \ldots e_n)
+\end{array}
+\end{equation*}
+
+\item[--]  
+An extended reduction relation $\rightsquigarrow_\beta$ such that, for any context $C$, 
+$C[e_1] \rightsquigarrow_\beta C[e_2]$ whenever $e_1 \rightsquigarrow e_2$.
+
+\item[--]  
+Let elimination $\rightsquigarrow_\ident{let}$ such that, using  capture-avoiding substitution,
+  $C[\kvd{let}~x=e_1~\kvd{in}~e_2] \rightsquigarrow_\ident{let} C[e_2[x\leftarrow e_1]]$
+\end{itemize}
+
+\noindent
+We say that $\rightsquigarrow$ is a suitable \emph{host language reduction} if:
+\begin{itemize}
+\item[--] It satisfies the \emph{compositionality} property, that is if $e \rightsquigarrow e'$ and 
+$C[e]\rightsquigarrow_\beta e''$ then also $C[e']\rightsquigarrow_\beta e''$.
+
+\item[--] Let elimination does not affect the result, i.e.~if $e \rightsquigarrow_{\ident{let}} e'$ 
+and $e'\rightsquigarrow_\beta e''$ then also $e\rightsquigarrow_\beta e''$ 
+\end{itemize}
+\end{definition}
+
+\noindent
+The host language in The Gamma is a simple call-by-value functional language without side-effects,
+and so it satisfies both compositionality and allows let bindings to be eliminated, although the
+latter affects the performance. The mechanism for preview evaluation presented here would also 
+work for call-by-name languages, but it would suffer from the expected difficulties in the 
+presence of side-effects or non-determinism.
 

### 2019-01-17 wip-live-programming/paper.tex
Updating formalisation, formatting tweaks

@@
 \usepackage{booktabs}
-\usepackage{subcaption}
 \usepackage{enumitem}
@@
 \definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
+\definecolor{idclr}{rgb}{0.3,0.3,0.3}
 \definecolor{numclr}{rgb}{0.0,0.4,0.0}
@@
 \newcommand{\rstr}[1]{\textnormal{\textcolor{rstrclr}{\sffamily "#1"}}}
-\newcommand{\ident}[1]{\textnormal{\sffamily #1}}
+\newcommand{\ident}[1]{\textnormal{\textcolor{idclr}{\sffamily #1}}}
 \newcommand{\qident}[1]{\textnormal{\sffamily \textquotesingle #1\textquotesingle}}
@@
 \newcommand{\blbl}[1]{\textnormal{\textcolor{numclr}{\sffamily #1}}}
+\newcommand{\narrow}[1]{\hspace{-0.6em}#1\hspace{-0.6em}}
 
@@
 
+\begin{figure}[b]
+\includegraphics[scale=0.5]{notebook.png}
+\caption{An excerpt from a Jupyter notebook by Financial Times reporters, which loads trade
+  data from the UN trade database and adds ISO country codes.}
+\label{fig:ft-uncomtrade}
+\end{figure}
+
 \section{Understanding how data scientists work}
@@
 code they write and the way they interact with the system is very different from how software
-engineers work.
-
-This paper focuses on simple data wrangling and data exploration as done, for example, by 
-journalists analysing government datasets. This section illustrates how such data analyses 
-look and provides a justification for design decisions that we make when designing the data 
-exploration calculus.
+engineers work. This paper focuses on simple data wrangling and data exploration as done, for
+example, by journalists analysing government datasets. This section illustrates how such data
+analyses look and provides a justification for the design of our data exploration calculus.
 
@@
 Data analysts write small snippets of code, run them to see results immediately and then revise
-them interactively based on the results. We aim to build foundations for such seamless integration 
-of development and execution in a notebook environment.
+them.
+
+% interactively based on the results. We aim to build foundations for such seamless integration
+% of development and execution in a notebook environment.
 
@@
 models to journalists who load data, perform simple aggregations and create visualizations. Our
-initial focus is on simple use cases, such as those of journalists, because we believe that this 
-is where better tooling can provide the most value.
-
-As an example, consider the data anlysis done by the Financial Times for an article on plastic 
+initial focus is on simple use cases, such as those of journalists. Tooling that makes notebook
+systems more interactive and spreadsheet-like in such cases is of crucial importance as it enables
+more people to produce reproducible and transparent data analyses.
+
+As an example, consider the anlysis done by the Financial Times for an article on plastic
 waste. The analysis joins datasets from Eurostat, UN Comtrade and more, aggregates the data and
-builds a visualization comparing the plastic waste flows in 2017 and 2018.
+builds a visualization comparing plastic waste flows in 2017 and 2018.
 Figure~\ref{fig:ft-uncomtrade} shows an extract from one notebook from the data analysis.
@@
 
-\begin{figure}
-\includegraphics[scale=0.6]{notebook.png}
-\caption{An excerpt from a Jupyter notebook by Financial Times reporters, which loads trade
-  data from the UN trade database and adds ISO country codes.}
-\label{fig:ft-uncomtrade}
-\end{figure}
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure}
-\includegraphics[scale=0.28]{gamma1.png}
-\caption{\small The Gamma script to aggregate Olympic medals. We select Rio 2016 games and
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[b]
+  \centering
+  \subcaptionbox{
+    \raggedright\footnotesize The Gamma script to aggregate Olympic medals. We select Rio 2016 games and
     count the number of distinct events per athlete. We then type `.' to choose
-further aggregation operations.}
+    further aggregation operations.
     \label{fig:gamma-dot}
-\vspace{-1em}
+  }{\includegraphics[scale=0.26]{gamma1.png}}
+  \subcaptionbox{
+    \raggedright\footnotesize Live preview in The Gamma programming environment. The table is produced as
+      the data anlyst types and updates on-the-fly to show result at the current cursor position.
+    \label{fig:gamma-preview}
+  }{\includegraphics[scale=0.26]{gamma2.png}}
+  \caption{\small TBD}
+  \label{first-figure-with-subfigures}
 \end{figure}
@@
 
-\begin{figure}
-\includegraphics[scale=0.28 ]{gamma2.png}
-\caption{\small Live preview in The Gamma programming environment. The table is produced as
-  the data anlyst types and updates on-the-fly to show result at the current cursor position.}
-\label{fig:gamma-preview}
-\end{figure}
-
 \section{Live coding for data exploration}
+
+% TODO: What is live coding? (Show preview while typing!)
 
@@
 
-\paragraph{Real-world use cases.}
+\begin{itemize}
+\item\textbf{Real-world use cases.}
 Our work has been motivated by the needs outlined in the previous section. In particular, we
@@
 
-\paragraph{Text editor integration.}
+\item\textbf{Text editor integration.}
 Our approach is based on an ordinary text editor that allows unrestricted modification of the
@@
 
-\paragraph{Data stay, code changes.}
+\item\textbf{Data stay, code changes.}
 Our focus has been on the scenario when input data is a static file that does not change, but the
@@
 this is an interesting future direction.
+\end{itemize}
 
@@
 \begin{array}{lcl}
-p &=& c_1; \ldots; c_n\\
-c &=& \kvd{let}~x = e \lsep e\\
-\end{array}
-\quad
+p &\narrow{=}& c_1; \ldots; c_n\\
+c &\narrow{=}& \kvd{let}~x = e \lsep e\\
+\end{array}
+\qquad
 \begin{array}{lcl}
-e &=& o \lsep x \\
-  &|& e.m(a, \ldots, a)
-\end{array}
-\quad
+e &\narrow{=}& o \lsep x \\
+  &\narrow{|}& e.m(a, \ldots, a)
+\end{array}
+\qquad
 \begin{array}{lcl}
-a &=& e \lsep \lambda x\rightarrow e\\
-v &=& o \lsep \lambda x\rightarrow e\\
+a &\narrow{=}& e \lsep \lambda x\rightarrow e\\
+v &\narrow{=}& o \lsep \lambda x\rightarrow e\\
 \end{array}
@@
 \begin{array}{lcl}
-C_e[-] &=& -.m(a_1, \ldots, a_n) ~|~ -.m(v_1, \ldots, v_{m}, -, a_1,\ldots , a_n) ~|~ -\\
-C_c[-] &=& \kvd{let}~x = C_e[-] ~|~ C_e[-]\\
-C_p[-] &=& c_1;\; \ldots;\; c_{k-1};\; C_c[-];\; c_{k+1};\; \ldots;\; c_n\\
+C_e[-] &=& -.m(a_1, \ldots, a_n) ~\lsep~ -.m(v_1, \ldots, v_{m}, -, a_1,\ldots , a_n) ~\lsep~ -\\
+C_c[-] &=& \kvd{let}~x = C_e[-] \hspace{1.25em}\lsep~ C_e[-]\\
+C_p[-] &=& o_1;\; \ldots;\; o_{k};\; C_c[-];\; c_{1};\; \ldots;\; c_n\\
 \end{array}
@@
 \hspace{-0.5em}\begin{array}{l}
-c_1;\; \ldots;\; c_{k-1};\; \kvd{let}~x=o;\; c_{k+1};\; \ldots; c_n \rightsquigarrow\\
-\qquad  c_1;\; \ldots;\; c_{k-1};\; c_{k+1}[x\leftarrow o];\; \ldots; c_n[x\leftarrow o]\end{array} &
+o_1;\; \ldots;\; o_{k};\; \kvd{let}~x=o;\; c_{1};\; \ldots; c_n \rightsquigarrow\\
+\qquad  o_1;\; \ldots;\; o_{k};\; c_{1}[x\leftarrow o];\; \ldots; c_n[x\leftarrow o]\end{array} &
 \quad\text{\small\sffamily(let)}\\[1.2em]
@@
 The calculus combines object-oriented features such as member access with functional features
-such as lambda functions. The syntax is defined in Figure~\ref{fig:dec-calculus}. We write $x$ for 
-variables and $o$ for object values. A program $p$ in the data exploration calculus consists of a 
-sequence of commands $c$. A command can be either let binding or an expression. Lambda functions 
-can only appear directly as arguments in method calls. An argument $a$ can thus be a lambda 
-function or an expression.
+such as lambda functions. The syntax is defined in Figure~\ref{fig:dec-calculus}. Object values
+$o$ are defined by external libraries that are used in conjunction with the core calculus.
+
+A program $p$ in the data exploration calculus consists of a sequence of commands $c$. A command
+can be either let binding or an expression. Let bindings define variables $x$ that can be used in
+subsequent commands. The calculus supports lambda functions, but they
+can only appear as arguments in method calls. An argument $a$ can thus be a lambda
+function or an expression.\footnote{Similar but weaker restrictions on the use of lambda functions
+exist in other languages. To guide type inference, lambda functions in C\# can appear as either
+method arguments or assigned to an explicitly typed variable.}
 
@@
 reduction $\rightsquigarrow$. Fully evaluating a program results in an irreducible sequence of
-objects $o_1;\; \ldots;\; o_n$ which can be displayed as results of the data analysis. The 
+objects $o_1;\; \ldots;\; o_n$ which can be displayed as intermediate results of the data analysis. The
 operational semantics is parameterized by a relation $\rightsquigarrow_\epsilon$ that models
@@
 \begin{equation*}
-o.m(v_1, \ldots, v_n) \rightsquigarrow_\epsilon o'
-\end{equation*}
-%
-foo
-
-bar
+o_1.m(v_1, \ldots, v_n) \rightsquigarrow_\epsilon o_2
+\end{equation*}
+%
+Figure~\ref{fig:dec-calculus} defines the reduction rules in terms of $\rightsquigarrow_\epsilon$
+and evaluation contexts; $C_e$ specifies left-to-right evaluation of arguments of a method call,
+$C_c$ specifies evluation of a command and $C_p$ defines top-to-bottom evaluation of a program.
+The rule {\small\sffamily (external)} applies reductions provided by an external library in a
+call-by-value order and {\small\sffamily (let)} substitutes a value of an evaluated variable in
+all subsequent commands.
+
+\subsection{Properties}
+
+TBD: Some adaptation of the stuff below, as needed for the proofs.
 
@@
 
-\subsection{Equational theory}
-In The Gamma language, computations are expressed using
-member access, written as $e.m(e_1,\ldots, e_n)$. In this paper, we do not define how member
-access evaluates. This has been done elsewhere \cite{gamma}, but more importantly, the evaluation of
-previews does not rely on the exact specifics of the evaluation, provided that the language
-satisfies certain basic conditions. The following definitions provides the necessary structure 
-for discussing correctness of previews.
-
-Partial evaluation may reduce an expression under $\lambda$-abstraction. We do not require that
-the reduction of the host language does this. Instead, we define an extended reduction relation
-and use that in the proofs. The host language only needs to compose well with such extended reduction
-as captured by the \emph{compositionality} property below. We also require that the language allows
-elimination of let bindings.
-
-\begin{definition}[Host language]
-\label{def:host}
-Given a relation on expressions $e_1 \rightsquigarrow e_2$ that models small-step evaluation, we define:
-
-\begin{itemize}
-\item[--]  
- A \emph{preview evaluation context} (also referred to as \emph{context}):
-
-\vspace{-1.25em}
-\begin{equation*}
-\begin{array}{lcl}
-C[-] &\hspace{-1em}=\hspace{-1em}& \kvd{let}~x = -~\kvd{in}~e ~|~ \kvd{let}~x = e~\kvd{in}~- ~|~ \lambda x\rightarrow - \\
-     &&  e_0.m(e_1, \ldots, e_{k-1}, -, e_{k+1}, \ldots e_n)
-\end{array}
-\end{equation*}
-
-\item[--]  
-An extended reduction relation $\rightsquigarrow_\beta$ such that, for any context $C$, 
-$C[e_1] \rightsquigarrow_\beta C[e_2]$ whenever $e_1 \rightsquigarrow e_2$.
-
-\item[--]  
-Let elimination $\rightsquigarrow_\ident{let}$ such that, using  capture-avoiding substitution,
-  $C[\kvd{let}~x=e_1~\kvd{in}~e_2] \rightsquigarrow_\ident{let} C[e_2[x\leftarrow e_1]]$
-\end{itemize}
-
-\noindent
-We say that $\rightsquigarrow$ is a suitable \emph{host language reduction} if:
-\begin{itemize}
-\item[--] It satisfies the \emph{compositionality} property, that is if $e \rightsquigarrow e'$ and 
-$C[e]\rightsquigarrow_\beta e''$ then also $C[e']\rightsquigarrow_\beta e''$.
-
-\item[--] Let elimination does not affect the result, i.e.~if $e \rightsquigarrow_{\ident{let}} e'$ 
-and $e'\rightsquigarrow_\beta e''$ then also $e\rightsquigarrow_\beta e''$ 
-\end{itemize}
-\end{definition}
-
-\noindent
-The host language in The Gamma is a simple call-by-value functional language without side-effects,
-and so it satisfies both compositionality and allows let bindings to be eliminated, although the
-latter affects the performance. The mechanism for preview evaluation presented here would also 
-work for call-by-name languages, but it would suffer from the expected difficulties in the 
-presence of side-effects or non-determinism.
-
-\newpage
+% TODO: Adapt as needed
+%
+% \subsection{Equational theory}
+% In The Gamma language, computations are expressed using
+% member access, written as $e.m(e_1,\ldots, e_n)$. In this paper, we do not define how member
+% access evaluates. This has been done elsewhere \cite{gamma}, but more importantly, the evaluation of
+% previews does not rely on the exact specifics of the evaluation, provided that the language
+% satisfies certain basic conditions. The following definitions provides the necessary structure
+% for discussing correctness of previews.
+%
+% Partial evaluation may reduce an expression under $\lambda$-abstraction. We do not require that
+% the reduction of the host language does this. Instead, we define an extended reduction relation
+% and use that in the proofs. The host language only needs to compose well with such extended reduction
+% as captured by the \emph{compositionality} property below. We also require that the language allows
+% elimination of let bindings.
+%
+% \begin{definition}[Host language]
+% \label{def:host}
+% Given a relation on expressions $e_1 \rightsquigarrow e_2$ that models small-step evaluation, we define:
+%
+% \begin{itemize}
+% \item[--]
+%  A \emph{preview evaluation context} (also referred to as \emph{context}):
+%
+% \vspace{-1.25em}
+% \begin{equation*}
+% \begin{array}{lcl}
+% C[-] &\hspace{-1em}=\hspace{-1em}& \kvd{let}~x = -~\kvd{in}~e ~|~ \kvd{let}~x = e~\kvd{in}~- ~|~ \lambda x\rightarrow - \\
+%      &&  e_0.m(e_1, \ldots, e_{k-1}, -, e_{k+1}, \ldots e_n)
+% \end{array}
+% \end{equation*}
+%
+% \item[--]
+% An extended reduction relation $\rightsquigarrow_\beta$ such that, for any context $C$,
+% $C[e_1] \rightsquigarrow_\beta C[e_2]$ whenever $e_1 \rightsquigarrow e_2$.
+%
+% \item[--]
+% Let elimination $\rightsquigarrow_\ident{let}$ such that, using  capture-avoiding substitution,
+%   $C[\kvd{let}~x=e_1~\kvd{in}~e_2] \rightsquigarrow_\ident{let} C[e_2[x\leftarrow e_1]]$
+% \end{itemize}
+%
+% \noindent
+% We say that $\rightsquigarrow$ is a suitable \emph{host language reduction} if:
+% \begin{itemize}
+% \item[--] It satisfies the \emph{compositionality} property, that is if $e \rightsquigarrow e'$ and
+% $C[e]\rightsquigarrow_\beta e''$ then also $C[e']\rightsquigarrow_\beta e''$.
+%
+% \item[--] Let elimination does not affect the result, i.e.~if $e \rightsquigarrow_{\ident{let}} e'$
+% and $e'\rightsquigarrow_\beta e''$ then also $e\rightsquigarrow_\beta e''$
+% \end{itemize}
+% \end{definition}
+%
+% \noindent
+% The host language in The Gamma is a simple call-by-value functional language without side-effects,
+% and so it satisfies both compositionality and allows let bindings to be eliminated, although the
+% latter affects the performance. The mechanism for preview evaluation presented here would also
+% work for call-by-name languages, but it would suffer from the expected difficulties in the
+% presence of side-effects or non-determinism.
+%
+% \newpage
 
@@
 
-\begin{figure*}
-\begin{equation*}
-\begin{array}{l}
-\xymatrix{
-& \bnd{val}(\num{10}) &\qquad\quad& \bnd{val}(\num{15})\\
-\bnd{var}(\ident{data}) & 
-  \bnd{mem}(\ident{skip},s_0)\ar[l]_{\blbl{arg}(0)} \ar[u]^{\blbl{arg}(1)} && 
-  \bnd{mem}(\ident{take},s_1)\ar[ll]_{\blbl{arg}(0)} \ar[u]_{\blbl{arg}(1)}
-}\\[5em]
-\xymatrix{
-&& \bnd{val}(\num{10})\\
-\bnd{var}(\ident{data}) & 
-  \bnd{mem}(\ident{skip}, s_0)\ar[l]_{\blbl{arg}(0)} \ar@/^/[ru]^{\blbl{arg}(1)} && 
-  \bnd{mem}(\ident{take}, s_2)\ar[ll]_{\blbl{arg}(0)} \ar@/_/[lu]_{\blbl{arg}(1)}
+\section{Formalising live coding infrastructure}
+\label{sec:formal}
+
+A naive way of providing live previews during code editing would be to re-evaluate the code
+after each change. This would be wasteful -- for example, when writing
+code, most changes are additive and the preview can be updated by evaluating just the newly added
+code. In this section, we develop foundations for an efficient way of displaying live previews
+for the data exploration calculus.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Maintaining dependency graph}
+\label{sec:formal-deps}
+
+The key idea behind our method is to maintain a dependency graph \cite{dependencies} with
+nodes representing individual operations of the computation that can be (partially) evaluated
+to obtain a preview. Each time the program text is modified, we parse it afresh (using an
+error-recovering parser) and bind the abstract syntax tree to the dependency graph.
+When binding a new expression to the graph, we reuse previously created nodes as long as
+they have the same dependencies. For expressions that have a new structure, we create new nodes.
+
+The nodes of the graph serve as unique keys into a lookup table with previously
+evaluated parts of the program. When a preview is requested for an expression, we use the graph
+node bound to the expression to find a preview. If a preview has not been evaluated, we force
+the evaluation of all dependencies in the graph and then evaluate the operation represented by
+the current node.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[b]
+\centering
+\subcaptionbox{
+  \raggedright\small
+  Graph constructed from the following initial expression:
+  $\kvd{let}~x = \num{15}~\kvd{in}~\ident{data.skip}(\num{10}).\ident{take}(x)$
+  \label{fig:dep-graph-a}
+}{
+  $\xymatrix{
+  \bnd{val}(\num{10}) & \bnd{val}(\num{15})\\
+  \bnd{mem}(\ident{skip},s_0)\ar[d]_{\blbl{arg}(0)} \ar[u]^{\blbl{arg}(1)} &
+  \bnd{mem}(\ident{take},s_1)\ar[l]_{\blbl{arg}(0)} \ar[u]_{\blbl{arg}(1)}\\
+  \bnd{var}(\ident{data})
+  }\qquad$
 }
-\end{array}\qquad
-\begin{array}{l}
-\textnormal{a.) The first graph is constructed from}\\
-\textnormal{the following initial expression:}\\[0.5em]
-\quad\kvd{let}~x = \num{15}~\kvd{in}\\
-\quad\ident{data.skip}(\num{10}).\ident{take}(x)\\
-\\
-\textnormal{b.) The second diagram shows the updated graph}\\
-\textnormal{after the programmer changes $x$ to $10$:}\\[0.5em]
-\quad\kvd{let}~x = \num{10}~\kvd{in}\\
-\quad\ident{data.skip}(\num{10}).\ident{take}(x)
-\end{array}
-\end{equation*}
+\hspace{1em}
+\subcaptionbox{
+  \raggedright\small
+  Updated graph after changing $x$ to $10$:
+  $\kvd{let}~x = \num{10}~\kvd{in}~\ident{data.skip}(\num{10}).\ident{take}(x)$
+  \label{fig:dep-graph-b}
+}{
+  $\xymatrix{
+  \bnd{val}(\num{10})&\\
+  \bnd{mem}(\ident{skip}, s_0)\ar[d]_{\blbl{arg}(0)} \ar[u]^{\blbl{arg}(1)} &
+  \bnd{mem}(\ident{take}, s_2)\ar[l]_{\blbl{arg}(0)} \ar@/_/[lu]_{\blbl{arg}(1)}\\
+  \bnd{var}(\ident{data})
+}\qquad$
+}
 \caption{Dependency graphs formed by two steps of the live programming process. }
 \label{fig:dep-graph}
-\end{figure*}
-
-% --------------------------------------------------------------------------------------------------
-
-\newpage
-
-\section{Formalising live coding infrastructure}
-\label{sec:formal}
-
-In this section, we present a formalisation of a live coding infrastructure for a small,
-expression-based programming language that supports \kvd{let} binding, member invocations
-and $\lambda$ abstractions. This is the necessary minimum for data exploration as described
-in the previous section.
-
-It excludes constructs such as a mechanism for defining new objects as we assume that those
-are imported from the context through a mechanism such as type providers.
-%
-\begin{equation*}
-\begin{array}{lcl}
-e &=& \kvd{let}~x = e~\kvd{in}~e ~|~ \lambda x\rightarrow e ~|~ e.m(e, \ldots, e) ~|~ x ~|~ n
-\end{array}
-\end{equation*}
-%
-Here, $m$ ranges over member names, $x$ over variables and $n$ over primitive values such as 
-numbers. Function values can be passed as arguments to methods (provided by a type provider), but 
-for the purpose of this paper, we do not need to be able to invoke them directly.
-
-\paragraph{The problem with functions.}
-In the context of live programming, \kvd{let} binding and member access are unproblematic.
-We can evaluate them and provide live preview for both of them, including all their sub-expressions.
-Function values are more problematic, because their sub-expressions cannot be evaluated. For example:
-%
-\begin{equation*}
-\begin{array}{l}
-\kvd{let}~\ident{page} = \lambda x \rightarrow \ident{movies}.\ident{skip}(x*\num{10}).\ident{take}(\num{10})
-\end{array}
-\end{equation*}
-%
-We can provide live preview for the \ident{movies} sub-expression, but not for 
-$\ident{movies}.\ident{skip}(x*\num{10})$ because we cannot obtain the value of $x$ without
-running the rest of the program and analysing how the function is called later.
-
-The method described in this paper obtains delayed previews for sub-expressions that contain
-free variables (which could be fully evaluated if values for free variables were provided), but
-we describe a more speculative design of live coding friendly functions in Section~\ref{sec:design}.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Maintaining dependency graph}
-\label{sec:formal-deps}
-
-The key idea behind our implementation is to maintain a dependency graph \cite{dependencies} with 
-nodes representing individual operations of the computation that can be partially evaluated
-to obtain a preview. Each time the program text is modified, we parse it afresh (using an
-error-recovering parser) and bind the abstract syntax tree to the dependency graph.
-
-We remember the previously created nodes of the graph. When binding a new expression to
-the graph, we reuse previously created nodes that have the same dependencies. For
-expressions that have a new structure, we create new nodes (using a fresh symbol to 
-identify them).
-
-The nodes of the graph serve as unique keys into a lookup table with previously
-evaluated operations of the computation. When a preview is requested, we use the node
-bound to the expression to find a preview, or evaluate it by first forcing the evaluation 
-of all parents in the dependency graph.
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure*}[t]
-\begin{equation*}
-\begin{array}{ll}
-\ident{bind}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n)) = &(1)~~\\ 
-\qquad \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n)\\
-\quad \textnormal{when}~\bndclr{v_i}, (V_i, E_i) = \ident{bind}_{\Gamma, \Delta}(e_i)\\
-\quad \textnormal{and}~\bndclr{v} = \Delta(\bknd{mem}(m),[(\bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v_n}, \blbl{arg}(n))])\\
-\quad \textnormal{let}~E = \{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v}, \bndclr{v_n}, \blbl{arg}(n))\}
-\\[1em]
-\ident{bind}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n)) = &(2)\\ 
-\qquad \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n)\\
-\quad \textnormal{when}~\bndclr{v_i}, (V_i, E_i) = \ident{bind}_{\Gamma, \Delta}(e_i)\\
-\quad \textnormal{and}~(\bknd{mem}(m),[(\bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v_n}, \blbl{arg}(n))]) \notin \ident{dom}(\Delta)\hspace{-200em}.\\
-\quad \textnormal{let}~\bndclr{v} = \bnd{mem}(m, s), s~\textnormal{fresh}\\
-\quad \textnormal{let}~E = \{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v},\bndclr{v_n}, \blbl{arg}(n)) \}
-\\[1em]
-\ident{bind}_{\Gamma, \Delta}(\kvd{let}~x=e_1~\kvd{in}~e_2) = \bndclr{v}, (\{\bndclr{v}\} \cup V \cup V_1, E \cup E_1)&(3)\\
-\quad \textnormal{let}~\bndclr{v_1}, (V_1,E_1) = \ident{bind}_{\Gamma,\Delta}(e_1)\\
-\quad \textnormal{let}~\Gamma_1 = \Gamma \cup \{(x,\bndclr{v_1})\} \\
-\quad \textnormal{let}~\bndclr{v}, (V, E) = \ident{bind}_{\Gamma_1, \Delta}(e_2)
-\end{array}
-\begin{array}{ll}
-\ident{bind}_{\Gamma, \Delta}(n) = \bnd{val}(n), (\{ \bnd{val}(n) \}, \emptyset) }&(4)
-\\[1em]
-\ident{bind}_{\Gamma, \Delta}(x) = \bndclr{v}, (\{ \bndclr{v} \}, \emptyset)\quad \textnormal{when}~\bndclr{v} = \Gamma(x)&(5)
-\\[1.25em]
-\ident{bind}_{\Gamma, \Delta}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V, \{e\} \cup E)&(6)\\
-\quad \textnormal{when}~\Gamma_1 = \Gamma \cup \{ x, \bnd{var}(x) \}\\
-\quad \textnormal{and}~\bndclr{v_0}, (V, E) = \ident{bind}_{\Gamma_1, \Delta}(e)\\
-\quad \textnormal{and}~\bndclr{v} = \Delta(\bknd{fun}(x),[(\bndclr{v_0}, \blbl{body})])\\
-\quad \textnormal{let}~e = (\bndclr{v}, \bndclr{v_0}, \blbl{body}) 
-\\[1.25em]
-\ident{bind}_{\Gamma, \Delta}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V, \{e\} \cup E)&(7)\\
-\quad \textnormal{when}~\Gamma_1 = \Gamma \cup \{ x, \bnd{var}(x) \}\\
-\quad \textnormal{and}~\bndclr{v_0}, (V, E) = \ident{bind}_{\Gamma_1, \Delta}(e)\\
-\quad \textnormal{and}~(\bknd{fun}(x),[(\bndclr{v_0}, \blbl{body})])\notin \ident{dom}(\Delta)\\
-\quad \textnormal{let}~\bndclr{v} = \bnd{fun}(x, s), s~\textnormal{fresh}\\
-\quad \textnormal{let}~e = (\bndclr{v},\bndclr{v_0},\blbl{body}) 
-\\[0.75em]
-\end{array}
-\end{equation*}
-\caption{Rules of the binding process, which constructs a dependency graph for an expression.}
-\label{fig:binding-rules}
-\end{figure*}
-
-% --------------------------------------------------------------------------------------------------
-\paragraph{Elements of the graph.} The nodes of the graph represent individual operations
-to be computed. In our design, the nodes themselves are used as keys, so we attach a unique 
-\emph{symbol} to some of the nodes. That way, we can create two unique nodes representing, 
-for example, access to a member named \ident{take} which differ in their dependencies.
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsubsection{Elements of the graph}
+The nodes of the graph represent individual operations of the computation. In our design, the nodes
+are used as cache keys, so we attach a unique symbol $s$ to some of the nodes. That way, we can
+create two unique nodes representing, for example, access to a member named \ident{take} which
+differ in their dependencies.
 
 Furthermore, the graph edges are labeled with labels indicating the kind of dependency. For
-a method call, the labels are ``first argument'', ``second argument'' and so on. Formally:
+a method call, the labels are ``first argument'', ``second argument'' and so on. Writing
+$s$ for symbols and $i$ for integers, nodes (vertices) $\bndclr{v}$ and edge labels $\blblclr{l}$
+are defined as:
 %
@@
 \begin{array}{rcll}
-s&\hspace{-0.25em}\in\hspace{-0.25em}& \textit{Symbol}\\
-i&\hspace{-0.25em}\in\hspace{-0.25em}& \textit{Integer}\\
-n&\hspace{-0.25em}\in\hspace{-0.25em}& \textit{Primitive values}\\
-x&\hspace{-0.25em}\in\hspace{-0.25em}& \textit{Variable names}\\
-m&\hspace{-0.25em}\in\hspace{-0.25em}& \textit{Member names}\\[0.5em]
-\bndclr{v}&\hspace{-0.25em}\in\hspace{-0.25em}&\bnd{val}(n)~|~\bnd{var}(x)~|~\bnd{mem}(m, s)~|~\bnd{fun}(x, s)&(\textit{Vertices})\\
-\blblclr{l}&\hspace{-0.25em}\in\hspace{-0.25em}&\blbl{body}~|~\blbl{arg}(i)&(\textit{Edge labels})\\
-\end{array}
-\end{equation*}
-%
-The \bnd{val} node represents a primitive value and contains the value itself. Two occurrences
+\bndclr{v}&=&\bnd{val}(o) \lsep \bnd{var}(x)\lsep \bnd{mem}(m, s)\lsep \bnd{fun}(x, s)&(\textit{Vertices})\\
+\blblclr{l}&=&\blbl{body} \lsep \blbl{arg}(i)&(\textit{Edge labels})\\
+\end{array}
+\end{equation*}
+%
+The \bnd{val} node represents a primitive value and contains the object itself. Two occurrences
 of $\num{10}$ in the source code will be represented by the same node. Member access \bnd{mem}
-contains the member name, together with a unique symbol -- two member access nodes with different 
-dependencies will contain a different symbol. Dependencies of member access are labeled with 
-\blbl{arg} indicating the index of the argument (the instance has index $0$ and arguments 
-start with $1$).
-
-Finally, nodes \bnd{fun} and \bnd{var} represent function values and variables bound by $\lambda$ 
-abstraction. For simplicity, we use variable names rather than de Bruijn indices and so 
-renaming a bound variable forces recomputation.
-
-\paragraph{Example graph.} Figure~\ref{fig:dep-graph} illustrates how we construct and update the 
+contains the member name, together with a unique symbol $s$ -- two member access nodes with
+different dependencies will contain a different symbol. Dependencies of member access are labeled
+with \blbl{arg} indicating the index of the argument (the instance has index $0$ and arguments
+start with $1$). Finally, nodes \bnd{fun} and \bnd{var} represent function values and variables
+bound by $\lambda$ abstraction. For simplicity, we use variable names rather than de Bruijn
+indices and so renaming a bound variable forces recomputation.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsubsection{Example graph}
+Figure~\ref{fig:dep-graph} illustrates how we construct and update the
 dependency graph. Node representing $\ident{take}(x)$ depends on the argument -- the
@@
 bound via \kvd{let} binding such as $x$ do not appear as $\bnd{var}$ nodes. The node using it
-depends directly on the node representing the result of the expression that is assigned to $x$.
+depends directly on the node representing the expression that is assigned to $x$.
 
 After changing the value of $x$, we create a new graph. The dependencies of the node
-$\bnd{mem}(\ident{skip}, s_0)$ are unchanged and so the node is reused. This means that this
-part of the program is not recomputed. The $\blbl{arg}(1)$ dependency of the \ident{take} call 
-changed and so we create a node $\bnd{mem}(\ident{skip}, s_2)$ with a new fresh symbol $s_2$.
-The preview for this node is then recomputed as needed using the already known values of its
+$\bnd{mem}(\ident{skip}, s_0)$ are unchanged and so the symbol $s_0$ attached to the node remains
+the same and previously computed previews can be reused. This part of the program is not recomputed.
+The $\blbl{arg}(1)$ dependency of the \ident{take} call
+changed and so we create a new node $\bnd{mem}(\ident{skip}, s_2)$ with a fresh symbol $s_2$.
+The preview for this node is then computed as needed using the already known values of its
 dependencies.
@@
 
-\paragraph{Reusing graph nodes.} The binding process takes an expression and constructs a 
-dependency graph, reusing existing nodes when possible. For this, we keep a lookup table 
-of member access and function value nodes. The key is formed by a node kind (for 
-disambiguation) together with a list of dependencies. A node kind is a member access or a function:
-%
-\begin{equation*}
-\begin{array}{lcll}
-\bkndclr{k}&\in&\bknd{fun}(x)~|~\bknd{mem}(m)&\qquad(\textit{Node  kinds})
-\end{array}
-\end{equation*}
-%
-Given a lookup table $\Delta$, we write $\Delta(\bkndclr{k}, [(\bndclr{n_1},\blblclr{l_1}), \ldots,
-(\bndclr{v_n},\blblclr{l_n})])$ to perform a lookup for a node of a kind $\bkndclr{k}$ that has dependencies
+\subsubsection{Reusing graph nodes}
+The binding process takes an expression and constructs a dependency graph reusing existing nodes
+when possible. For this, we use a lookup table of member access and function value nodes. The key
+in the lookup table is formed by a node kind (for disambiguation) together with a list of
+dependencies. A node kind is a member access containing the member name or a function containing
+the bound variable name; a lookup table $\Delta$ then maps a node kind with a list of dependencies
+to a cached node:
+%
+\begin{equation*}
+\begin{array}{ll}
+\bkndclr{k}~=~\bknd{fun}(x)~|~\bknd{mem}(m)&(\textit{Node kinds})\\[0.2em]
+\Delta(\bkndclr{k}, [(\bndclr{v_1},\blblclr{l_1}), \ldots, (\bndclr{v_n},\blblclr{l_n})])  & (\textit{Lookup for a node})
+\end{array}
+\end{equation*}
+%
+The example on the second line looks for a node of a kind $\bkndclr{k}$ that has dependencies
 $\bndclr{v_1}, \ldots, \bndclr{v_n}$ labeled with labels $\blblclr{l_1}, \ldots, \blblclr{l_n}$.
-
-For example, when creating the graph in Figure~\ref{fig:dep-graph} (b), we perform the following
-lookup for the \ident{skip} member access:
+We write $\Delta(\bndclr{k}, l)\!\downarrow$ when a value for a given key is not defined.
+For example, when creating the graph in Figure~\ref{fig:dep-graph-b},
+we perform the following lookup for the \ident{skip} member access:
 %
@@
 
-\subsection{Binding an expressions to a graph}
+\begin{figure*}[t]
+\begin{equation*}
+\hspace{-3.5em}
+\begin{array}{ll}
+\ident{bind-expr}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n)) = \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n) &(1)~~\\
+\quad \textnormal{when}~\bndclr{v_i}, (V_i, E_i) = \ident{bind-expr}_{\Gamma, \Delta}(e_i)
+~\textnormal{and}~\bndclr{v} = \Delta(\bknd{mem}(m),[(\bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v_n}, \blbl{arg}(n))])\\
+\quad \textnormal{let}~E = \{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v}, \bndclr{v_n}, \blbl{arg}(n))\}
+\\[1em]
+\ident{bind-expr}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n)) = \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n)&(2)\\
+\quad \textnormal{when}~\bndclr{v_i}, (V_i, E_i) = \ident{bind-expr}_{\Gamma, \Delta}(e_i)
+~\textnormal{and}~\Delta(\bknd{mem}(m),[(\bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v_n}, \blbl{arg}(n))])\!\downarrow\\
+\quad \textnormal{let}~\bndclr{v} = \bnd{mem}(m, s), s~\textnormal{fresh}
+~\textnormal{and}~E = \{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v},\bndclr{v_n}, \blbl{arg}(n)) \}
+\\[1em]
+\ident{bind-expr}_{\Gamma, \Delta}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V, \{e\} \cup E)&(3)\\
+\quad \textnormal{when}~\Gamma_1 = \Gamma \cup \{ x, \bnd{var}(x) \}
+~\textnormal{and}~\bndclr{v_0}, (V, E) = \ident{bind-expr}_{\Gamma_1, \Delta}(e)
+~\textnormal{and}~\bndclr{v} = \Delta(\bknd{fun}(x),[(\bndclr{v_0}, \blbl{body})])\\
+\quad \textnormal{let}~e = (\bndclr{v}, \bndclr{v_0}, \blbl{body})
+\\[1em]
+\ident{bind-expr}_{\Gamma, \Delta}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V, \{e\} \cup E)&(4)\\
+\quad \textnormal{when}~\Gamma_1 = \Gamma \cup \{ x, \bnd{var}(x) \}
+~\textnormal{and}~\bndclr{v_0}, (V, E) = \ident{bind-expr}_{\Gamma_1, \Delta}(e)
+~\textnormal{and}~\Delta(\bknd{fun}(x),[(\bndclr{v_0}, \blbl{body})])\!\downarrow\\
+\quad \textnormal{let}~\bndclr{v} = \bnd{fun}(x, s), s~\textnormal{fresh}
+~\textnormal{and}~e = (\bndclr{v},\bndclr{v_0},\blbl{body})
+\\[1em]
+\ident{bind-expr}_{\Gamma, \Delta}(o) = \bnd{val}(o), (\{ \bnd{val}(o) \}, \emptyset) }&(5)
+\\[1em]
+\ident{bind-expr}_{\Gamma, \Delta}(x) = \bndclr{v}, (\{ \bndclr{v} \}, \emptyset)~\textnormal{when}~\bndclr{v} = \Gamma(x)&(6)
+\end{array}
+\end{equation*}
+\caption{Binding rules that define a construction of a dependency graph for an expression.}
+\label{fig:binding-rules-expr}
+\end{figure*}
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure*}[t]
+\begin{equation*}
+\hspace{-3.5em}
+\begin{array}{ll}
+\ident{bind-prog}_{\Gamma, \Delta}(\kvd{let}~x=e; c_2; \ldots; c_n) = \bndclr{v_1};\ldots;\bndclr{v_n}, (\{\bndclr{v}\} \cup V \cup V_1, E \cup E_1)
+  \hspace{9.35em}&(7)\\
+\quad \textnormal{let}~\bndclr{v_1}, (V_1,E_1) = \ident{bind-expr}_{\Gamma,\Delta}(e_1)~\textnormal{and}~\Gamma_1 = \Gamma \cup \{(x,\bndclr{v_1})\} \\
+\quad \textnormal{and}~\bndclr{v_2}; \ldots; \bndclr{v_n}, (V, E) = \ident{bind-prog}_{\Gamma_1, \Delta}(c_2; \ldots; c_n)
+\\[1em]
+\ident{bind-prog}_{\Gamma, \Delta}(e; c_2; \ldots; c_n) = \bndclr{v_1};\ldots;\bndclr{v_n}, (\{\bndclr{v}\} \cup V \cup V_1, E \cup E_1)&(8)\\
+\quad \textnormal{let}~\bndclr{v_1}, (V_1,E_1) = \ident{bind-expr}_{\Gamma,\Delta}(e)
+~\textnormal{and}~\bndclr{v_2}; \ldots; \bndclr{v_n}, (V, E) = \ident{bind-prog}_{\Gamma_1, \Delta}(c_2; \ldots; c_n)
+\\[1em]
+\ident{bind-prog}_{\Gamma, \Delta}([]) = [], (\emptyset, \emptyset)&(9)
+\end{array}
+\end{equation*}
+\caption{Binding rules that define a construction of a dependency graph for a program.}
+\label{fig:binding-rules-prog}
+\end{figure*}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Binding expressions to a graph}
 \label{sec:formal-bind}
 
-When constructing the dependency graph, our implementation annotates the nodes of the 
-abstract syntax tree with the nodes of the dependency graph, forming a mapping
-$e \rightarrow \bndclr{v}$. For this reason, we call the process \emph{binding}.
-
-The process of binding is defined by the rules in Figure~\ref{fig:binding-rules}.
-The \ident{bind} function is annotated with a lookup table $\Delta$ discussed in 
-Section~\ref{sec:formal-deps} and a variable context $\Gamma$. The variable context 
-is a map from variable names to dependency graph nodes and is used for variables bound
-using \kvd{let} binding.
-
-When applied on an expression $e$, binding $\ident{bind}_{\Gamma,\Delta}(e)$ returns a
-dependency graph $(V, E)$ paired with a node $\bndclr{v}$ corresponding to the expression $e$.
+After parsing updated code, we update the dependency graph and link each node of the abstract
+syntax tree to a node of the dependency graph. This process is called
+binding and is defined by the \ident{bind-expr} function (Figure~\ref{fig:binding-rules-expr})
+and \ident{bind-prog} function (Figure~\ref{fig:binding-rules-prog}). Both functions are
+annotated with a lookup table $\Delta$ discussed in Section~\ref{sec:formal-deps} and a variable
+context $\Gamma$. The variable context is a map from variable names to dependency graph nodes and
+is used for variables bound using \kvd{let} binding.
+
+When applied on an expression $e$, binding $\ident{bind-expr}_{\Gamma,\Delta}(e)$ returns
+a node $\bndclr{v}$ corresponding to the expression $e$ paired with a dependency graph $(V, E)$.
 In the graph, $V$ is a set of nodes $\bndclr{v}$ and $E$ is a set of labeled edges
@@
 keeping a separate colouring function as this makes the formalisation simpler.
-
-\paragraph{Binding member access.} In all rules, we recursively bind sub-expressions to get
-a dependency graph for each sub-expression and a graph node that represents it. The nodes 
-representing sub-expressions are then used as dependencies for lookup into $\Delta$, together
-with their labels. When binding a member access, we reuse an existing node if it is defined by 
-$\Delta$ (1) or we create a new node containing a fresh symbol when the domain of $\Delta$ does 
-not contain a key describing the current member access~(2).
-
-\paragraph{Binding let binding.} For \kvd{let} binding (3), we first bind the expression $e_1$ assigned
-to the variable to obtain a graph node $\bndclr{v_1}$. We then bind the body expression $e_2$,
-but using a variable context $\Gamma_1$ that maps the value of the variable to the graph node
-$\bndclr{v_1}$. The variable context is used when binding a variable (6) and so all variables 
-declared using \kvd{let} binding will be bound to a graph node representing the value assigned 
-to the variable. The node bound to the overall \kvd{let} expression is then the graph node bound
-to the body expression.
-
-\paragraph{Binding function values.} If a function value uses its argument, we will not be able
-to evaluate its body. In this case, the graph node bound to a function will depend on a synthetic
-node $\bnd{var}(x)$ that represents the variable with no value. When binding a function, we 
-create the synthetic variable and add it to the variable context $\Gamma_1$ before binding the
-body. As with member access, the node representing a function may (7) or may not (8) be already 
-present in the lookup table.
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure}[!b]  
+The $\ident{bind-prog}_{\Gamma,\Delta}$ function works similarly, but takes a sequence of
+commands and returns a sequence of nodes.
+
+\subsubsection{Binding expressions.} When binding member access, we use \ident{bind-expr} recursively
+to get a node and a dependency graph for each sub-expression. The nodes representing sub-expressions are
+then used as dependencies for lookup into $\Delta$, together with their labels. If a node already
+exists in $\Delta$ it is reused (1). Alternativelym, we create a new node containing a fresh symbol~(2).
+
+If a lambda function uses its argument, we will not be able to evaluate its body. In this case, the
+graph node bound to a function will depend on a synthetic node $\bnd{var}(x)$ that represents the
+variable with unknown value. When binding a function, we create the synthetic variable and add it
+to the variable context $\Gamma_1$ before binding the body. As with member access, the node
+representing a function may (3) or may not (4) be already present in the lookup table.
+
+\subsubsection{Binding programs.} When binding a program, we bind the first command and then
+recursively process remaining commands until we reach an empty list of commands (9).
+For \kvd{let} binding (7), we bind the expression $e$ assigned to the variable to obtain a
+graph node $\bndclr{v_1}$. We then bind the remaining commands using a variable context $\Gamma_1$
+that maps the value of the variable to the graph node $\bndclr{v_1}$. The variable context is used
+when binding a variable (6) and so all variables declared using \kvd{let} binding will be bound to
+a graph node representing the value assigned to the variable. When the command is just an
+expression (8), we bind the expression using \ident{bind-expr}.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}
 \begin{equation*}
@@
 \quad\Delta_{i}(\bknd{mem}(m), [(\bndclr{v_0},\blbl{arg}(0)),\ldots, (\bndclr{v_n},\blbl{arg}(n))]) = \bnd{mem}(m, s)\\
-\quad\quad \textnormal{for all}~\bnd{mem}(m, s) \in V\\
-\quad\quad \textnormal{such that}~(\bnd{mem}(m, s), \bndclr{v_i}, \blbl{arg}(i)) \in E~\textnormal{for}~i\in 0,..,n
+\quad\quad \textnormal{when}~\bnd{mem}(m, s) \in V
+~\textnormal{and}~(\bnd{mem}(m, s), \bndclr{v_i}, \blbl{arg}(i)) \in E~\textnormal{for}~i\in 0,..,n
 \\[0.75em]
 \quad\Delta_{i}(\bknd{fun}(x), [(\bndclr{v_1},\blbl{body})]) = \bnd{fun}(x, s)\\
-\quad\quad \textnormal{for all}~\bnd{fun}(x, s) \in V\\
-\quad\quad \textnormal{such that}~(\bnd{fun}(x, s), \bndclr{v_1}, \blbl{body}) \in E
+\quad\quad \textnormal{when}~\bnd{fun}(x, s) \in V
+~\textnormal{and}~(\bnd{fun}(x, s), \bndclr{v_1}, \blbl{body}) \in E
 \\[0.75em]
@@
 \end{equation*}
+\vspace{-1em}
 \caption{Updating the node cache after binding a new graph}
 \label{fig:loop}
-\vspace{0.5em}
+\vspace{-0.5em}
 \end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Edit and rebind loop}
+
+The binding process formalised in Section~\ref{sec:formal-bind} specifies how to update the
+dependency graph after updated program text is parsed. During live coding, this is done
+repeatedly as the programmer edits code. Throughout the process, we maintain a series of
+lookup table states $\Delta_0, \Delta_1, \Delta_2, \ldots$ The initial lookup table is
+empty, i.e.~$\Delta_0 = \emptyset$. At a step $i$, we parse a program $p_i$ (consisting of
+several commands) and obtain a new dependency graph using the previous $\Delta$. The result is
+a sequence of nodes corresponding to commands of the program and a graph $(V, E)$:
+%
+\begin{equation*}
+\bndclr{v_1}; \ldots; \bndclr{v_n}, (V, E) = \ident{bind-prog}_{\emptyset, \Delta_{i-1}}(p_i)
+\end{equation*}
+%
+The new state of the cache is computed by calling the $\ident{update}_{V, E}(\Delta_{i-1})$ function
+defined in Figure~\ref{fig:loop}. The function adds newly created nodes from the graph
+$(V, E)$ to the previous cache $\Delta_{i-1}$ and returns a new cache $\Delta_{i}$. We only cache
+nodes for function and member accesses -- nodes for variables and primitive values will remain
+the same thanks to the way they are constructed.
+
+% ==================================================================================================
+
+\section{Evaluating previews}
+\label{sec:previews}
+
+The mechanism for constructing dependency graphs defined in Section~\ref{sec:formal} makes it
+possible to provide live previews when editing code without recomputing the whole program
+each time the source code changes.
+
+The nodes in the dependency graph correspond to individual operations that will be performed
+when running the program. When the dependencies of an operation do not change while editing
+code, the subsequent dependency graph will reuse a node used to represent the operation.
+
+Our live editor keeps a map from graph nodes to live previews, so a new preview only needs to be
+computed when a new node appears in the dependency graph (and the user moves the cursor to a
+code location that corresponds to the node). This section describes how previews are evaluated.
 
@@
 
-\subsection{Edit and rebind loop}
-
-The binding process formalised in Section~\ref{sec:formal-bind} specifies how to update the 
-dependency graph after updated program text is parsed. During live coding, this is done 
-repeatedly as the programmer edits code. Throughout the process, we maintain a series of 
-lookup table states $\Delta_0, \Delta_1, \Delta_2, \ldots$. Initially, the lookup table is 
-empty, i.e.~$\Delta_0 = \emptyset$.
-
-At a step $i$, we parse an expression $e_i$ and calculate the new dependency graph and a node
-bound to the top-level expression using the previous $\Delta$:
-%
-\begin{equation*}
-\bndclr{v}, (V, E) = \ident{bind}_{\emptyset, \Delta_{i-1}}(e_i)
-\end{equation*}
-%
-The new state of the node cache is then computed by adding newly created nodes from the graph
-$(V, E)$ to the previous cache $\Delta_{i-1}$. This is done for function and member nodes that
-contain unique symbols as defined in Figure~\ref{fig:loop}. We do not need to cache nodes
-representing primitive values and variables as those do not contain symbols and will remain the
-same due to the way they are constructed.
-
-% ==================================================================================================
-
-\section{Evaluating previews}
-\label{sec:previews}
-
-The mechanism for constructing dependency graphs defined in Section~\ref{sec:formal} makes it
-possible to provide live previews when editing code without recomputing the whole program 
-each time the source code changes.
-
-The nodes in the dependency graph correspond to individual operations that will be performed
-when running the program. When the dependencies of an operation do not change while editing 
-code, the subsequent dependency graph will reuse a node used to represent the operation.
-
-Our live editor keeps a map from graph nodes to live previews, so a new preview only needs to be
-computed when a new node appears in the dependency graph (and the user moves the cursor to a 
-code location that corresponds to the node). This section describes how previews are evaluated.
+\subsection{The problem with functions.}
+In the context of live programming, \kvd{let} binding and member access are unproblematic.
+We can evaluate them and provide live preview for both of them, including all their sub-expressions.
+Function values are more problematic, because their sub-expressions cannot be evaluated. For example:
+%
+\begin{equation*}
+\begin{array}{l}
+pd.read\_csv(\strf{demo.csv}).map)
+\kvd{let}~\ident{page} = \lambda x \rightarrow \ident{movies}.\ident{skip}(x*\num{10}).\ident{take}(\num{10})
+\end{array}
+\end{equation*}
+%
+We can provide live preview for the \ident{movies} sub-expression, but not for
+$\ident{movies}.\ident{skip}(x*\num{10})$ because we cannot obtain the value of $x$ without
+running the rest of the program and analysing how the function is called later.
+
+The method described in this paper obtains delayed previews for sub-expressions that contain
+free variables (which could be fully evaluated if values for free variables were provided), but
+we describe a more speculative design of live coding friendly functions in Section~\ref{sec:design}.
 
@@
 \begin{array}{ll}
-\ident{bind}_{\Gamma, \Delta, c}(e_0.m(e_1, \ldots, e_n)) = &(2b)\\ 
+\ident{bind-expr}_{\Gamma, \Delta, c}(e_0.m(e_1, \ldots, e_n)) = &(2b)\\
 \qquad \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n)\\
-\quad \textnormal{when}~\bndclr{v_0}, (V_0, E_0) = \ident{bind}_{\Gamma, \Delta, \bot}(e_0)\\
+\quad \textnormal{when}~\bndclr{v_0}, (V_0, E_0) = \ident{bind-expr}_{\Gamma, \Delta, \bot}(e_0)\\
 \quad \textnormal{and}~c_i = (\bndclr{v_0}, \blbl{callsite}(m,i))\hspace{4.8em} (i \in 1\ldots n)\\
-\quad \textnormal{and}~\bndclr{v_i}, (V_i, E_i) = \ident{bind}_{\Gamma, \Delta, c_i}(e_i)\qquad (i\in 1\ldots n)\\
+\quad \textnormal{and}~\bndclr{v_i}, (V_i, E_i) = \ident{bind-expr}_{\Gamma, \Delta, c_i}(e_i)\qquad (i\in 1\ldots n)\\
 \quad \textnormal{and}~(\bknd{mem}(m),[(\bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v_n}, \blbl{arg}(n))]) \notin \ident{dom}(\Delta)\hspace{-200em}.\\
@@
 \begin{array}{ll}
-\ident{bind}_{\Gamma, \Delta, (\bndclr{v_c}, \blblclr{l_c} )}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V_0, E \cup E_0)&(7b)\\
+\ident{bind-expr}_{\Gamma, \Delta, (\bndclr{v_c}, \blblclr{l_c} )}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V_0, E \cup E_0)&(7b)\\
 \quad \textnormal{when}~(\bknd{var}(x),[(\bndclr{v_c}, \blblclr{l_c} )])\notin \ident{dom}(\Delta)\\
@@
 \quad \textnormal{and}~\Gamma_1 = \Gamma \cup \{ x, \bndclr{v_x} \}\\
-\quad \textnormal{and}~\bndclr{v_0}, (V_0, E_0) = \ident{bind}_{\Gamma_1, \Delta, \bot}(e)\\
+\quad \textnormal{and}~\bndclr{v_0}, (V_0, E_0) = \ident{bind-expr}_{\Gamma_1, \Delta, \bot}(e)\\
 \quad \textnormal{when}~(\bknd{fun}(x),[(\bndclr{v_0}, \blbl{body}), (\bndclr{v_c}, \blblclr{l_c} )])\notin \ident{dom}(\Delta)\\
@@
 
-The \ident{bind} function now has a parameter $c$, in addition to $\Gamma$ and $\Delta$, which 
+The \ident{bind-expr} function now has a parameter $c$, in addition to $\Gamma$ and $\Delta$, which
 represents the context in which the binding happens. This is either a member invocation (labeled
@@
 %
-The key parts of the revised definition of the \ident{bind} function are shown in Figure~\ref{fig:binding-rules-callsite}.
-We now write $\ident{bind}_{\Gamma,\Delta,c}$ where $c$ represents the context in which the
+The key parts of the revised definition of the \ident{bind-expr} function are shown in Figure~\ref{fig:binding-rules-callsite}.
+We now write $\ident{bind-expr}_{\Gamma,\Delta,c}$ where $c$ represents the context in which the
 binding occurs. This is set to $\bot$ in all cases, except when binding arguments of a member call.
@@
 variable node and from the function node in $E$. We omit a similar variant of the case (6).
-The remaining cases (3)-(5) are the same, except that \ident{bind} has the additional $c$ parameter
+The remaining cases (3)-(5) are the same, except that \ident{bind-expr} has the additional $c$ parameter
 and recursive calls always set it to $\bot$.
@@
 Given an expression $e_1$ such that $e_1 \rightsquigarrow_\ident{let} e_2$ and a lookup table $\Delta_0$
-then if $\bndclr{v_1}, (V_1, E_1) = \ident{bind}_{\emptyset, \Delta_0}(e_1)$ and  
-$\bndclr{v_2}, (V_2, E_2) = \ident{bind}_{\emptyset, \Delta_1}(e_2)$ such that $\Delta_1 = \ident{update}_{V_1, E_1}(\Delta_0)$
+then if $\bndclr{v_1}, (V_1, E_1) = \ident{bind-expr}_{\emptyset, \Delta_0}(e_1)$ and
+$\bndclr{v_2}, (V_2, E_2) = \ident{bind-expr}_{\emptyset, \Delta_1}(e_2)$ such that $\Delta_1 = \ident{update}_{V_1, E_1}(\Delta_0)$
 then it holds that $\bndclr{v_1} = \bndclr{v_2}$ and also $(V_1, E_1) = (V_2, E_2)$.
@@
 Assume $e_1 = C[\kvd{let}~x=e'~\kvd{in}~e'']$ and the resulting $e_2 = C[e''[x\leftarrow e']]$. The case
-$\ident{bind}_{\Gamma,\Delta}(\kvd{let}~x=e'~\kvd{in}~e'')$ when binding $e_1$ is handled using
+$\ident{bind-expr}_{\Gamma,\Delta}(\kvd{let}~x=e'~\kvd{in}~e'')$ when binding $e_1$ is handled using
 (3).
@@
 or already present in $\Delta_1$ (added by $\ident{update}_{V_1, E_1}$) and is reused each time
-$\ident{bind}_{\Gamma,\Delta_1}(e')$ is called.
+$\ident{bind-expr}_{\Gamma,\Delta_1}(e')$ is called.
 \end{proof}
@@
 first, which adds the node for $e'$ to $\Delta$. In our implementation, this is not needed
-because $\Delta$ is updated while the graph is being constructed using $\ident{bind}$. 
+because $\Delta$ is updated while the graph is being constructed using $\ident{bind-expr}$.
 To keep the formalisation simpler, we separate the process of building the dependency graph
@@
 with a lookup table $\Delta$ obtained from any sequence of expressions according to Figure~\ref{fig:loop}
-let $\bndclr{v}, (V, E) = \ident{bind}_{\emptyset, \Delta}(e)$. If $\bndclr{v}\Downarrow d$ 
+let $\bndclr{v}, (V, E) = \ident{bind-expr}_{\emptyset, \Delta}(e)$. If $\bndclr{v}\Downarrow d$
 over a graph $(V, E)$ then $d = p$ for some $p$ and $e \rightsquigarrow_\beta p$.
@@
 \begin{proof}
-First note that, when combining recursively constructed sub-graphs, the \ident{bind} operation
+First note that, when combining recursively constructed sub-graphs, the \ident{bind-expr} operation
 adds new nodes and edges leading from those new nodes. Therefore, an evaluation using $\Downarrow$
@@
 Next, we prove a more general property using induction showing that for $e$ such
-that $\bndclr{v}, (V, E) = \ident{bind}_{\emptyset, \Delta}(e)$:
+that $\bndclr{v}, (V, E) = \ident{bind-expr}_{\emptyset, \Delta}(e)$:
 \begin{enumerate}
@@
 from any sequence of expressions according to Figure~\ref{fig:loop} and $e_2$ such that
-$e_1\rightsquigarrow_\beta e_2$ and let $\bndclr{v_1}, (V_1, E_1) = \ident{bind}_{\emptyset, \Delta_1}(e_1)$.
-
-Let $\Delta_2 = \ident{update}_{V_1, E_1}(\Delta_1)$ and $\bndclr{v_2}, (V_2, E_2) = \ident{bind}_{\emptyset, \Delta_2}(e_2)$.
+$e_1\rightsquigarrow_\beta e_2$ and let $\bndclr{v_1}, (V_1, E_1) = \ident{bind-expr}_{\emptyset, \Delta_1}(e_1)$.
+
+Let $\Delta_2 = \ident{update}_{V_1, E_1}(\Delta_1)$ and $\bndclr{v_2}, (V_2, E_2) = \ident{bind-expr}_{\emptyset, \Delta_2}(e_2)$.
 If $\bndclr{v_2}\Downarrow d$ over a graph $(V_2, E_2)$ then $d = p$ for some $p$ and $e \rightsquigarrow_\beta p$.
@@
 Let $\Delta_1 = \emptyset$, for any $e_1, e_2$, assume that the first expression is bound,
-i.e.~$\bndclr{v_1}, (V_1, E_1) = \ident{bind}_{\emptyset, \Delta_1}(e_1)$, the graph node cache
+i.e.~$\bndclr{v_1}, (V_1, E_1) = \ident{bind-expr}_{\emptyset, \Delta_1}(e_1)$, the graph node cache
 is updated $\Delta_2 = \ident{update}_{V_1, E_1}(\Delta_1)$ and a new expression is bound, i.e.~
-$\bndclr{v_2}, (V_2, E_2) = \ident{bind}_{\emptyset, \Delta_2}(e_2)$. Now, for any $\bndclr{v}$, 
+$\bndclr{v_2}, (V_2, E_2) = \ident{bind-expr}_{\emptyset, \Delta_2}(e_2)$. Now, for any $\bndclr{v}$,
 if $\bndclr{v} \Downarrow p$ over $(V_1, E_1)$ then also
@@
 This is the case, because new graph nodes added to $\Delta_2$ by $\ident{update}_{V_1, E_2}$
-are only ever added as new nodes in $\ident{bind}_{\emptyset, \Delta_2}$ and so the existing 
+are only ever added as new nodes in $\ident{bind-expr}_{\emptyset, \Delta_2}$ and so the existing
 nodes and edges of $(V_1, E_1)$ used during the evaluation are unaffected.
@@
 variables of $e$ are bound in $C$, assume that the first expression is bound,
-i.e.~$\bndclr{v_1}, (V_1, E_1) = \ident{bind}_{\emptyset, \Delta_1}(e_1)$, the graph node cache
+i.e.~$\bndclr{v_1}, (V_1, E_1) = \ident{bind-expr}_{\emptyset, \Delta_1}(e_1)$, the graph node cache
 is updated $\Delta_2 = \ident{update}_{V_1, E_1}(\Delta_1)$ and the second expression is bound, i.e.~
-$\bndclr{v_2}, (V_2, E_2) = \ident{bind}_{\emptyset, \Delta_2}(e_2)$.
-
-Now, assume $\bndclr{v}, G = \ident{bind}_{\Gamma_1, \Delta_1}(e)$ and 
-$\bndclr{v'}, G' = \ident{bind}_{\Gamma_2, \Delta_2}(e)$ are the recursive calls to bind
+$\bndclr{v_2}, (V_2, E_2) = \ident{bind-expr}_{\emptyset, \Delta_2}(e_2)$.
+
+Now, assume $\bndclr{v}, G = \ident{bind-expr}_{\Gamma_1, \Delta_1}(e)$ and
+$\bndclr{v'}, G' = \ident{bind-expr}_{\Gamma_2, \Delta_2}(e)$ are the recursive calls to bind
 $e$ during the first and the second binding, respectively. Then, the graph nodes assigned to the
@@
 
-\vspace{0.75em}\noindent(1) $\ident{bind}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n))$ -- 
+\vspace{0.75em}\noindent(1) $\ident{bind-expr}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n))$ --
   Here $e = e_0.m(e_1, \ldots, e_n)$, $\bndclr{v_i}$ are graph nodes obtained by induction for
@@
 \newpage
-\vspace{0.75em}\noindent(2) $\ident{bind}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n))$ --
+\vspace{0.75em}\noindent(2) $\ident{bind-expr}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n))$ --
   This case is similar to (1), except that the fact that $\bndclr{v} = \bnd{mem}(m, s)$
@@
 
-\vspace{0.75em}\noindent(4) $\ident{bind}_{\Gamma, \Delta}(n)$ -- In this case $e=n$ and $\bndclr{v} = \bnd{val}(n)$. The preview of
+\vspace{0.75em}\noindent(4) $\ident{bind-expr}_{\Gamma, \Delta}(n)$ -- In this case $e=n$ and $\bndclr{v} = \bnd{val}(n)$. The preview of
   $\bnd{val}(n)$ is evaluated to $n$ using the (\emph{val}) case.
 
-\vspace{0.75em}\noindent(5) $\ident{bind}_{\Gamma, \Delta}(x)$ -- The initial $\Gamma$ is empty and
+\vspace{0.75em}\noindent(5) $\ident{bind-expr}_{\Gamma, \Delta}(x)$ -- The initial $\Gamma$ is empty and
   there are no let bindings, so $x$ must have been added to $\Gamma$ by case (6) or (7). Hence,
@@
 
-\vspace{0.75em}\noindent(6) $\ident{bind}_{\Gamma, \Delta}(\lambda x\rightarrow e)$ -- Assume
+\vspace{0.75em}\noindent(6) $\ident{bind-expr}_{\Gamma, \Delta}(\lambda x\rightarrow e)$ -- Assume
   $\bndclr{v_b}$ is a graph node representing the body. The evaluation can use one of three rules:

### 2019-01-23 wip-live-programming/paper.tex
Updating correctness proofs

@@
 \theoremstyle{definition}
+\newtheorem{axiom}[dfc]{Axiom}
 \newtheorem{definition}[dfc]{Definition}
@@
 \newcommand{\narrow}[1]{\hspace{-0.6em}#1\hspace{-0.6em}}
+\newcommand{\rname}[1]{{\sffamily\small(#1)}}
 
@@
 \quad .\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\ident{.take}(\num{10})\\
-\quad .\ident{map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{yyyy}))}\\
+\quad .\ident{map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{yyyy}))\\
 \end{array}
@@
 \quad .\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\ident{.take}(\ident{count})\\
-\quad \ident{.map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{dd-mm-yyyy}))}\\
+\quad \ident{.map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{dd-mm-yyyy}))\\
 \end{array}
@@
 \raggedright
-{\small\sffamily Programs, commands, expressions, arguments and values}
+{\small\sffamily Programs, commands, terms, expressions and values}
 \begin{equation*}
@@
 p &\narrow{=}& c_1; \ldots; c_n\\
-c &\narrow{=}& \kvd{let}~x = e \lsep e\\
+c &\narrow{=}& \kvd{let}~x = t \lsep t\\
 \end{array}
@@
 \begin{array}{lcl}
-e &\narrow{=}& o \lsep x \\
-  &\narrow{|}& e.m(a, \ldots, a)
+t &\narrow{=}& o \lsep x \\
+  &\narrow{|}& t.m(e, \ldots, e)
 \end{array}
@@
 \begin{array}{lcl}
-a &\narrow{=}& e \lsep \lambda x\rightarrow e\\
+e &\narrow{=}& t \lsep \lambda x\rightarrow e\\
 v &\narrow{=}& o \lsep \lambda x\rightarrow e\\
@@
 \begin{array}{lcl}
-C_e[-] &=& -.m(a_1, \ldots, a_n) ~\lsep~ -.m(v_1, \ldots, v_{m}, -, a_1,\ldots , a_n) ~\lsep~ -\\
-C_c[-] &=& \kvd{let}~x = C_e[-] \hspace{1.25em}\lsep~ C_e[-]\\
+C_e[-] &=& C_e[-].m(e_1, \ldots, e_n) ~\lsep~ o.m(v_1, \ldots, v_{m}, C_e[-], e_1,\ldots , e_n) ~\lsep~ -\\
+C_c[-] &=& \kvd{let}~x = C_e[-] ~\lsep~ C_e[-]\\
 C_p[-] &=& o_1;\; \ldots;\; o_{k};\; C_c[-];\; c_{1};\; \ldots;\; c_n\\
@@
 o_1;\; \ldots;\; o_{k};\; \kvd{let}~x=o;\; c_{1};\; \ldots; c_n \rightsquigarrow\\
-\qquad  o_1;\; \ldots;\; o_{k};\; c_{1}[x\leftarrow o];\; \ldots; c_n[x\leftarrow o]\end{array} &
-\quad\text{\small\sffamily(let)}\\[1.2em]
+\qquad  o_1;\; \ldots;\; o_{k};\; o; c_{1}[x\leftarrow o];\; \ldots; c_n[x\leftarrow o]\end{array} &
+\quad\text{\rname{let}}\\[1.2em]
 {o.m(v_1, \ldots, v_n) \rightsquigarrow_\epsilon o'}\implies
   {C_p[o.m(v_1, \ldots, v_n)] \rightsquigarrow C_p[o']} &
-\quad\text{\small\sffamily (external)}
+\quad\text{\rname{external}}
 \end{array}
@@
 A program $p$ in the data exploration calculus consists of a sequence of commands $c$. A command
-can be either let binding or an expression. Let bindings define variables $x$ that can be used in
+can be either let binding or a term. Let bindings define variables $x$ that can be used in
 subsequent commands. The calculus supports lambda functions, but they
-can only appear as arguments in method calls. An argument $a$ can thus be a lambda
-function or an expression.\footnote{Similar but weaker restrictions on the use of lambda functions
+can only appear as arguments in method calls. A term $t$ can be a value, variable or a
+member access, while an expression $e$, which can appear as an argument in member access,
+can be a lambda function or a term.\footnote{Similar but weaker restrictions on the use of lambda functions
 exist in other languages. To guide type inference, lambda functions in C\# can appear as either
@@
 \subsection{Operational semantics}
+\label{sec:calculus-semantics}
+
 The data exploration calculus is a call-by-value language. We model the evaluation as a small-step
@@
 $C_c$ specifies evluation of a command and $C_p$ defines top-to-bottom evaluation of a program.
-The rule {\small\sffamily (external)} applies reductions provided by an external library in a
-call-by-value order and {\small\sffamily (let)} substitutes a value of an evaluated variable in
-all subsequent commands.
+The rule \rname{external} applies reductions provided by an external library in a
+call-by-value order and \rname{let} substitutes a value of an evaluated variable in
+all subsequent commands. Note that \rname{let} leaves the value of the variable
+in the resulting list of commands.
 
@@
 
-TBD: Some adaptation of the stuff below, as needed for the proofs.
+DEFINE $\rightsquigarrow^{*}$ as a reflexive, transitive closure of $\rightsquigarrow$
+
+DEFINE $\rightsquigarrow_\ident{let}$ (on a program) - CBN subsitution
+
+\begin{definition}[External library]
+\label{def:external}
+
+Satsifies \emph{compositionality} such that:
+
+Given $e_0, e_1, \ldots, e_n$ and $e'_0, e'_1, \ldots, e'_n$ and $m$ such that
+$e_0.m(e_1, \ldots, e_n) \rightsquigarrow^{*} o$ and
+$e'_0.m(e'_1, \ldots, e'_n) \rightsquigarrow^{*} o'$ then
+if for any contexts
+$C_0, C_1, \ldots, C_n$ it holds that if $C_i[e_i] \rightsquigarrow^{*} o_i$ and $C_i[e'_i] \rightsquigarrow^{*} o_i$ for some $o_i$
+then $o = o'$.
+\end{definition}
+
+TBD: LET above should leave the evaluated object in the list of commmands
+
+LET ELIMINATION DOES NOT CHANGE RESULTS
+
+Let elimination $\rightsquigarrow_\ident{let}$ such that, using  capture-avoiding substitution,
+$C[\kvd{let}~x=e_1~\kvd{in}~e_2] \rightsquigarrow_\ident{let} C[e_2[x\leftarrow e_1]]$
+
+Let elimination does not affect the result, i.e.~if $e \rightsquigarrow_{\ident{let}} e'$
+and $e'\rightsquigarrow_\beta e''$ then also $e\rightsquigarrow_\beta e''$
+
+\begin{lem}[Let elimination for a program]
+\label{thm:let-lang-elimination}
+does not matter
+\end{lem}
 
@@
 
-% TODO: Adapt as needed
-%
-% \subsection{Equational theory}
-% In The Gamma language, computations are expressed using
-% member access, written as $e.m(e_1,\ldots, e_n)$. In this paper, we do not define how member
-% access evaluates. This has been done elsewhere \cite{gamma}, but more importantly, the evaluation of
-% previews does not rely on the exact specifics of the evaluation, provided that the language
-% satisfies certain basic conditions. The following definitions provides the necessary structure
-% for discussing correctness of previews.
-%
-% Partial evaluation may reduce an expression under $\lambda$-abstraction. We do not require that
-% the reduction of the host language does this. Instead, we define an extended reduction relation
-% and use that in the proofs. The host language only needs to compose well with such extended reduction
-% as captured by the \emph{compositionality} property below. We also require that the language allows
-% elimination of let bindings.
-%
-% \begin{definition}[Host language]
-% \label{def:host}
-% Given a relation on expressions $e_1 \rightsquigarrow e_2$ that models small-step evaluation, we define:
-%
-% \begin{itemize}
-% \item[--]
-%  A \emph{preview evaluation context} (also referred to as \emph{context}):
-%
-% \vspace{-1.25em}
-% \begin{equation*}
-% \begin{array}{lcl}
-% C[-] &\hspace{-1em}=\hspace{-1em}& \kvd{let}~x = -~\kvd{in}~e ~|~ \kvd{let}~x = e~\kvd{in}~- ~|~ \lambda x\rightarrow - \\
-%      &&  e_0.m(e_1, \ldots, e_{k-1}, -, e_{k+1}, \ldots e_n)
-% \end{array}
-% \end{equation*}
-%
-% \item[--]
-% An extended reduction relation $\rightsquigarrow_\beta$ such that, for any context $C$,
-% $C[e_1] \rightsquigarrow_\beta C[e_2]$ whenever $e_1 \rightsquigarrow e_2$.
-%
-% \item[--]
-% Let elimination $\rightsquigarrow_\ident{let}$ such that, using  capture-avoiding substitution,
-%   $C[\kvd{let}~x=e_1~\kvd{in}~e_2] \rightsquigarrow_\ident{let} C[e_2[x\leftarrow e_1]]$
-% \end{itemize}
-%
-% \noindent
-% We say that $\rightsquigarrow$ is a suitable \emph{host language reduction} if:
-% \begin{itemize}
-% \item[--] It satisfies the \emph{compositionality} property, that is if $e \rightsquigarrow e'$ and
-% $C[e]\rightsquigarrow_\beta e''$ then also $C[e']\rightsquigarrow_\beta e''$.
-%
-% \item[--] Let elimination does not affect the result, i.e.~if $e \rightsquigarrow_{\ident{let}} e'$
-% and $e'\rightsquigarrow_\beta e''$ then also $e\rightsquigarrow_\beta e''$
-% \end{itemize}
-% \end{definition}
-%
-% \noindent
-% The host language in The Gamma is a simple call-by-value functional language without side-effects,
-% and so it satisfies both compositionality and allows let bindings to be eliminated, although the
-% latter affects the performance. The mechanism for preview evaluation presented here would also
-% work for call-by-name languages, but it would suffer from the expected difficulties in the
-% presence of side-effects or non-determinism.
-%
-% \newpage
-
 % ==================================================================================================
 
-\section{Formalising live coding infrastructure}
+\section{Formalising live coding environment}
 \label{sec:formal}
@@
 \\[1em]
-\ident{bind-expr}_{\Gamma, \Delta}(o) = \bnd{val}(o), (\{ \bnd{val}(o) \}, \emptyset) }&(5)
+\ident{bind-expr}_{\Gamma, \Delta}(o) = \bnd{val}(o), (\{ \bnd{val}(o) \}, \emptyset) &(5)
 \\[1em]
@@
 \begin{array}{ll}
-\ident{bind-prog}_{\Gamma, \Delta}(\kvd{let}~x=e; c_2; \ldots; c_n) = \bndclr{v_1};\ldots;\bndclr{v_n}, (\{\bndclr{v}\} \cup V \cup V_1, E \cup E_1)
+\ident{bind-prog}_{\Gamma, \Delta}(\kvd{let}~x=e; c_2; \ldots; c_n) = \bndclr{v_1};\ldots;\bndclr{v_n}, (\{\bndclr{v_1}\} \cup V \cup V_1, E \cup E_1)
   \hspace{9.35em}&(7)\\
@@
 \\[1em]
-\ident{bind-prog}_{\Gamma, \Delta}(e; c_2; \ldots; c_n) = \bndclr{v_1};\ldots;\bndclr{v_n}, (\{\bndclr{v}\} \cup V \cup V_1, E \cup E_1)&(8)\\
+\ident{bind-prog}_{\Gamma, \Delta}(e; c_2; \ldots; c_n) = \bndclr{v_1};\ldots;\bndclr{v_n}, (\{\bndclr{v_1}\} \cup V \cup V_1, E \cup E_1)&(8)\\
 \quad \textnormal{let}~\bndclr{v_1}, (V_1,E_1) = \ident{bind-expr}_{\Gamma,\Delta}(e)
@@
 
-The mechanism for constructing dependency graphs defined in Section~\ref{sec:formal} makes it
-possible to provide live previews when editing code without recomputing the whole program
-each time the source code changes.
-
-The nodes in the dependency graph correspond to individual operations that will be performed
-when running the program. When the dependencies of an operation do not change while editing
-code, the subsequent dependency graph will reuse a node used to represent the operation.
-
-Our live editor keeps a map from graph nodes to live previews, so a new preview only needs to be
-computed when a new node appears in the dependency graph (and the user moves the cursor to a
-code location that corresponds to the node). This section describes how previews are evaluated.
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure*}
-\begin{equation*}
-\qquad\qquad\quad\begin{array}{l}
-\inference[(lift-expr)~]
+The binding process described in the previous section constructs a dependency graph after code
+changes. The nodes in the dependency graph correspond to individual operations that will be performed
+when running the program. When evaluating a preview, we attach (partial) results to
+nodes of the graph. Since the binding process reuses nodes when their dependencies do not change,
+previews for expressions (or sub-expressions) of a program can be reused when updating a preview.
+
+In this section, we describe how previews are evaluated. The evaluation is done over the dependency
+graph, rather than over the structure of the program expressions as in the operational semantics
+given in Section~\ref{sec:calculus-semantics}. We analyse the preview evaluation formally in
+Section~\ref{sec:properties} and show that the resulting previews are the same as the result we
+would get by directly evaluating the code and, also, that no recomputation occurs when code is
+edited in certain ways.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Previews and delayed previews}
+
+Programs in the data exploration calculus consist of sequence of commands. Those can always be
+evaluated to a value with a preview that can be displayed to the user. However, we also support
+previews for all sub-expressions. This can be problematic if the current sub-expression is inside
+the body of a function. For example:
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~\ident{top} = \ident{movies}.\ident{take}(\num{10}).\ident{map}(\lambda x \rightarrow x.\ident{getReleased}().\ident{format}(\str{dd-mm-yyyy}))
+\end{array}
+\end{equation*}
+%
+Here, we can directly evaluate sub-expressions \ident{movies} and $\ident{movies}.\ident{take}(\num{10})$,
+but not $x.\ident{getReleased}()$ and $x.\ident{getReleased}().\ident{format}(\str{dd-mm-yyyy})$
+because they contain a free variable $x$. Our preview evaluation algorithm addresses this by
+producing two kinds of previews. A \emph{fully evaluated preview} is just a value, while
+a \emph{delayed preview} is a partially evalauted expression with free variables:
+
+\begin{equation*}
+\begin{array}{rcll}
+p&=&o \lsep  \lambda x\rightarrow e&\quad(\textit{Fully evaluated previews})\\
+d&=&p \lsep  \llbracket e \rrbracket_\Gamma&\quad(\textit{Evaluated and delayed previews})\\
+\end{array}
+\end{equation*}
+%
+A fully evaluated preview $p$ can be either a primitive object or a function value with no free
+variables. A possibly delayed preview $d$ can be either an evaluated preview $p$ or an expression
+$e$ that requires variables $\Gamma$. For simplicity, we use an untyped language and so $\Gamma$
+is just a list of variables $x_1, \ldots, x_n$.
+
+A delayed preview is not necessarily the body of a lambda function as it appears in the source
+code. We partially evaluate sub-expressions of the body that do not have free variables or that
+have free variables bound by an earlier let binding.
+
+Our implementation does not currently display
+delayed previews to the user, but there is a number of possible approaches for doing that.
+Most interestingly, since lambda functions always appear as arguments of a member access, we could
+force the evaluation of the surrounding expression, capture (a number of) values passed as inputs
+to the lambda function and display a preview based on those. Finally, in Section~\ref{sec:design},
+we also consider a more speculative design for an abstraction mechanism that supports live previews
+that could, in some cases, replace lambda function.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure*}[b]
+\begin{equation*}
+\hspace{-2.5em}
+\begin{array}{l}
+\inference[\rname{lift-expr}~]
   {\bndclr{v} \Downarrow \llbracket e \rrbracket_\Gamma}
@@
 \\
-\inference[(lift-prev)~]
+\inference[\rname{lift-prev}~]
   {\bndclr{v} \Downarrow p}
@@
 \\
-\inference[(val)~]
+\inference[\rname{val}~]
   {~}
-  {\bnd{val}(n) \Downarrow n }
+  {\bnd{val}(o) \Downarrow o }
 \\
 \\
-\inference[(var)~]
+\inference[\rname{var}~]
   {~}
@@
 \end{array}
-\qquad
+\qquad\quad
 \begin{array}{l}
-\inference[(fun-val)~]
+\inference[\rname{fun-val}~]
   { (\bnd{fun}(x, s), \bndclr{v}, \blbl{body}) \in E & \bndclr{v} \Downarrow p }
@@
 \\
-\inference[(fun-bind)~]
+\inference[\rname{fun-bind}~]
   { (\bnd{fun}(x, s), \bndclr{v}, \blbl{body}) \in E & \bndclr{v} \Downarrow \llbracket e \rrbracket_{x} }
@@
 \\
-\inference[(fun-expr)~]
+\inference[\rname{fun-expr}~]
   { (\bnd{fun}(x, s), \bndclr{v}, \blbl{body}) \in E & \bndclr{v} \Downarrow \llbracket e \rrbracket_{x,\Gamma} }
@@
 \begin{equation*}
-\qquad\qquad\begin{array}{l}
-\inference[(mem-val)~]
+\hspace{-2.5em}
+\begin{array}{l}
+\inference[\rname{mem-val}~]
   { \forall i\in\{0\ldots k\}.(\bnd{mem}(m, s), \bndclr{v_i}, \blbl{arg}(i)) \in E & \bndclr{v_i} \Downarrow p_i
-    & p_0.m(p_1, \ldots, p_k) \rightsquigarrow p }
+    & p_0.m(p_1, \ldots, p_k) \rightsquigarrow_\epsilon p }
   { \bnd{mem}(m, s) \Downarrow p }
@@
 \\
-\inference[(mem-expr)~]
+\inference[\rname{mem-expr}~]
   { \forall i\in\{0\ldots k\}.(\bnd{mem}(m, s), \bndclr{v_i}, \blbl{arg}(i)) \in E & \exists j\in\{0\ldots k\}.\bndclr{v_j}\diagup\hspace{-1em}\Downarrow p_j
@@
 
-\caption{Rules that define evaluation of previews over a dependency graph for an expression}
+\caption{Rules that define evaluation of previews over a dependency graph for a program}
 \label{fig:eval}
@@
 
-\subsection{The problem with functions.}
-In the context of live programming, \kvd{let} binding and member access are unproblematic.
-We can evaluate them and provide live preview for both of them, including all their sub-expressions.
-Function values are more problematic, because their sub-expressions cannot be evaluated. For example:
-%
-\begin{equation*}
-\begin{array}{l}
-pd.read\_csv(\strf{demo.csv}).map)
-\kvd{let}~\ident{page} = \lambda x \rightarrow \ident{movies}.\ident{skip}(x*\num{10}).\ident{take}(\num{10})
-\end{array}
-\end{equation*}
-%
-We can provide live preview for the \ident{movies} sub-expression, but not for
-$\ident{movies}.\ident{skip}(x*\num{10})$ because we cannot obtain the value of $x$ without
-running the rest of the program and analysing how the function is called later.
-
-The method described in this paper obtains delayed previews for sub-expressions that contain
-free variables (which could be fully evaluated if values for free variables were provided), but
-we describe a more speculative design of live coding friendly functions in Section~\ref{sec:design}.
-
-\paragraph{Previews and delayed previews.}
-As discussed in Section~\ref{sec:formal}, the body of a function cannot be easily evaluated to
-a value if it uses the bound variable. We do not attempt to ``guess'' possible arguments and,
-instead, provide a full preview only for sub-expressions with free variables bound by a let binding.
-For a function body that uses the bound variable, we obtain a \emph{delayed preview}, which is
-an expression annotated with a list of variables that need to be provided before the expression
-can be evaluated. We use the following notation:
-%
-\begin{equation*}
-\begin{array}{rcll}
-p&\hspace{-0.25em}\in\hspace{-0.25em}&n~|~\lambda x\rightarrow e&\quad(\textit{Fully evaluated previews})\\
-d&\hspace{-0.25em}\in\hspace{-0.25em}&p~|~\llbracket e \rrbracket_\Gamma&\quad(\textit{Evaluated and delayed previews})\\
-\end{array}
-\end{equation*}
-%
-Here, $p$ ranges over fully evaluated values. It can be either a primitive value (such as number,
-string or an object) or a function value with no free variables. A possibly delayed preview $d$
-can then be either evaluated preview $p$ or an expression $e$ that requires variables $\Gamma$.
-For simplicity, we use an untyped language and so $\Gamma$ is a list of variables $x_1, \ldots, x_n$.
-
-\paragraph{Evaluation and splicing.}
-In this paper, we omit the specifics of the underlying programming language and we focus on the
-live coding mechanism. However, we assume that the language is equipped with an evaluation
-reduction $e \rightsquigarrow p$ that reduces a closed expression $e$ into a value $p$.
-
-For delayed previews, we construct a delayed expression using splicing. For example, assuming
-we have a delayed previews $\llbracket e_0 \rrbracket_x$ and $\llbracket e_1 \rrbracket_y$.
-If we need to invoke a member $m$ on $e_0$ using $e_1$ as an argument, we construct a new
-delayed preview $\llbracket e_0.m(e_1) \rrbracket_{x, y}$. This operation is akin to expression
-splicing from meta-programming \cite{metaml,quotations} and can be more formally captured by
-Contextual Modal Type Theory (CMTT) as outlined below.
-
-\paragraph{Evaluation of previews.}
+\subsection{Evaluation of previews}
 The evaluation of previews is defined in Figure~\ref{fig:eval}. Given a dependency graph $(V, E)$,
 we define a relation $\bndclr{v}\Downarrow d$ that evaluates a sub-expression corresponding to
-the node $\bndclr{v}$ to a (possibly delayed) preview $d$.
+the node $\bndclr{v}$ to a possibly delayed preview $d$. The nodes $V$ and edges $E$ of the
+dependency graph are parameters of the $\Downarrow$ relation, but they do not change during the
+evaluation and so we do not explicity write them.
 
@@
 to a delayed preview. If the ordinary evaluation returns a delayed preview, so does the auxiliary
-relation (\emph{lift-expr}). If the ordinary evaluation returns a value, the value is wrapped
-into a delayed preview requiring no variables (\emph{lift-prev}).
-
-Graph node representing a value is evaluated to a value (\emph{val}) and a graph node representing
+relation \rname{lift-expr}. If the ordinary evaluation returns a value, the value is wrapped
+into a delayed preview requiring no variables \rname{lift-prev}.
+%
+A node representing a value is evaluated to a value \rname{val} and a node representing
 an unbound variable is reduced to a delayed preview that requires the variable and returns its
-value (\emph{var}).
-
-For member access, we distinguish two cases. If all arguments evaluate to values (\emph{member-val}),
-then we use the evaluation relation $\rightsquigarrow$, immediately evaluate the member access and
-produce a value. If one of the arguments is delayed (\emph{member-expr}), because the member access
-is in the body of a lambda function, then we produce a delayed member access expression that
-requires the union of the variables required by the individual arguments.
+value~\rname{var}.
+
+For member access, we distinguish two cases. If all arguments evaluate to values \rname{member-val},
+then we use the evaluation relation defined by external libraries $\rightsquigarrow_\epsilon$ to
+immediately evaluate the member access and produce a value. If some of the arguments are
+delayed \rname{member-expr}, because the member access is inside the body of a lambda function,
+we produce a delayed member access expression that requires the union of the variables
+required by the individual arguments.
 
 The evaluation of function values is similar, but requires three cases. If the body can
-be reduced to a value with no unbound variables (\emph{fun-val}), we return a lambda function that
-returns the value. If the body requires only the bound variable (\emph{fun-bind}), we return a
+be reduced to a value with no unbound variables \rname{fun-val}, we return a lambda function that
+returns the value. If the body requires only the bound variable \rname{fun-bind}, we return a
 lambda function with the delayed preview as the body. If the body requires further variables,
@@
 
-\paragraph{Caching previews.}
+\subsection{Caching of evaluated previews}
+\label{sec:previews-cache}
+
 For simplicity, the relation $\Downarrow$ in Figure~\ref{fig:eval} does not specify how previews
 are cached and linked to graph nodes. In practice, this is done by maintaining a lookup table
-from graph nodes $\bdnclr{v}$ to (possibly delayed) previews $p$.
+from graph nodes $\bndclr{v}$ to (possibly delayed) previews $p$.
 Whenever $\Downarrow$ is used to obtain a preview for a graph node, we first
@@
 
-The evaluated previews can be reused in two ways. First, multiple nodes can
+The cached evaluated previews can be reused in two ways. First, multiple nodes can
 depend on one sub-graph in a single dependency graph (if the same sub-expression appears
@@
 
-\paragraph{Semantics of delayed previews.}
-The focus of this paper is on the design and implementation of a live coding environment,
-but it is worth noting that the structure of delayed previews is closely linked to the work
-on Contextual Modal Type Theory (CMTT) \cite{cmtt} and comonads \cite{cmtt-denotation}.
-
-In CMTT, $\lbrack \Psi \rbrack A$ denotes that a proposition $A$ is valid in context $\Psi$,
-which is closely related to our delayed previews written as $\llbracket A \rrbracket_\Psi$.
-CMTT defines rules for composing context-dependent propositions that would allow us to express
-the splicing operation used in (\emph{mem-expr}). In categorical terms, the context-dependent
-proposition can be modeled as an indexed comonad \cite{effectrev,graded}. The evaluation of a preview with no
-context dependencies (built implicitly into our evaluation rules) corresponds to the counit
-operation of a comonad and would be explicitly written as $\llbracket A \rrbracket_\emptyset \rightarrow A$.
-
-% ==================================================================================================
-
-\begin{figure*}[t]
-\vspace{-0.5em}
-\begin{equation*}
-\begin{array}{ll}
-\ident{bind-expr}_{\Gamma, \Delta, c}(e_0.m(e_1, \ldots, e_n)) = &(2b)\\
-\qquad \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n)\\
-\quad \textnormal{when}~\bndclr{v_0}, (V_0, E_0) = \ident{bind-expr}_{\Gamma, \Delta, \bot}(e_0)\\
-\quad \textnormal{and}~c_i = (\bndclr{v_0}, \blbl{callsite}(m,i))\hspace{4.8em} (i \in 1\ldots n)\\
-\quad \textnormal{and}~\bndclr{v_i}, (V_i, E_i) = \ident{bind-expr}_{\Gamma, \Delta, c_i}(e_i)\qquad (i\in 1\ldots n)\\
-\quad \textnormal{and}~(\bknd{mem}(m),[(\bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v_n}, \blbl{arg}(n))]) \notin \ident{dom}(\Delta)\hspace{-200em}.\\
-\quad \textnormal{let}~\bndclr{v} = \bnd{mem}(m, s), s~\textnormal{fresh}\\
-\quad \textnormal{let}~E = \{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v},\bndclr{v_n}, \blbl{arg}(n)) \}
-\end{array}
-\qquad\qquad
-\begin{array}{ll}
-\ident{bind-expr}_{\Gamma, \Delta, (\bndclr{v_c}, \blblclr{l_c} )}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V_0, E \cup E_0)&(7b)\\
-\quad \textnormal{when}~(\bknd{var}(x),[(\bndclr{v_c}, \blblclr{l_c} )])\notin \ident{dom}(\Delta)\\
-\quad \textnormal{let}~\bndclr{v_x} = \bnd{var}(x, s_x), s_x~\textnormal{fresh}\\
-\quad \textnormal{and}~\Gamma_1 = \Gamma \cup \{ x, \bndclr{v_x} \}\\
-\quad \textnormal{and}~\bndclr{v_0}, (V_0, E_0) = \ident{bind-expr}_{\Gamma_1, \Delta, \bot}(e)\\
-\quad \textnormal{when}~(\bknd{fun}(x),[(\bndclr{v_0}, \blbl{body}), (\bndclr{v_c}, \blblclr{l_c} )])\notin \ident{dom}(\Delta)\\
-\quad \textnormal{let}~\bndclr{v} = \bnd{fun}(x, s_f), s_f~\textnormal{fresh}\\
-\quad \textnormal{let}~E = \{ (\bndclr{v},\bndclr{v_0},\blbl{body}), (\bndclr{v},\bndclr{v_c},\blblclr{l_c}), (\bndclr{v_x},\bndclr{v_c},\blblclr{l_c}) \}
-\\[0.75em]
-\end{array}
-\end{equation*}
-\vspace{-0.5em}
-\caption{Revised binding rules, tracking call sites of function values.}
-\label{fig:binding-rules-callsite}
-\vspace{-0.5em}
-\end{figure*}
-
-% ==================================================================================================
-
-\section{Type checking}
-\label{sec:types}
-
-Evaluating live previews can be an expensive operations, so being able to cache partial previews
-is a must for a live coding environment. Type checking is typically fast, so the main focus of
-this paper is on live previews. However, asynchronous type providers in The Gamma (Section~\ref{sec:types-providers})
-can make type checking time consuming, and so we use the dependency graph also for type checking
-(Section~\ref{sec:types-graph}). Type checking lambda functions (Section~\ref{sec:types-funs})
-requires a slight extension of the model discussed in Section~\ref{sec:formal}.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Asynchronously provided types}
-\label{sec:types-providers}
-
-Data available in The Gamma can be defined using several kinds of type providers. The type provider
-used in Figure~\ref{fig:thegamma} as asynchronous \cite{async}. It downloads the sample URL and generates
-types based on the contents of the web page. The parameter to \ident{web.scrape} is a static
-parameter and is evaluated during type-checking. We omit details in this paper, but we note this
-works similarly to F\# type providers \cite{providers-fsharp}.
-
-Type providers can also be implemented as REST services \cite{rest} to allow anyone implement a data source
-in the language of their choice. In this case, each member of a call-chain returns a type that is
-generated based on the result of an HTTP request. For example, when the user types \ident{worldbank}
-(to access information about countries), the type provider makes a request to
-\url{http://thegamma-services.azurewebsites.net/worldbank}, which returns two members:
-%
-\begin{equation*}
-\begin{array}{l}
-[~\{\rstr{name}\!:\str{byYear},\rstr{returns}\!:\\
-~\quad \{\rstr{kind}\!:\str{nested},\rstr{endpoint}\!:\str{/pickYear}\}\}\\
-\hspace{0.55em}\{\rstr{name}\!:\str{byCountry},\rstr{returns}\!:\\
-~\quad \{\rstr{kind}\!:\str{nested},\rstr{endpoint}\!:\str{/pickCountry}\}\}~]
-\end{array}
-\end{equation*}
-%
-This indicates that \ident{worldbank} has members \ident{byYear} and \ident{byCountry}.
-If the user types \ident{worldbank.byCountry}, a request is made to the specified URL
-\url{http://thegamma-services.azurewebsites.net/worldbank/pickCountry}:
-%
-\begin{equation*}
-\begin{array}{l}
-[~\{\rstr{name}\!:\str{Andorra},\rstr{trace}\!:[\str{"country=AR"}], \\
-\quad \rstr{returns}\!:\{\rstr{kind}\!:\str{nested},\rstr{endpoint}\!:\str{/pickTopic}\}\}\\
-\hspace{0.55em}\{\rstr{name}\!:\str{Afghanistan},\rstr{trace}\!:[\str{"country=AF"}],\\
-\quad \rstr{returns}\!:\{\rstr{kind}\!:\str{nested},\rstr{endpoint}\!:\str{/pickTopic}\}\},\,..]
-\end{array}
-\end{equation*}
-%
-This returns a list of countries which can then be accessed as members via
-\ident{worldbank.byCountry.Andorra}, etc.
-
-This is one reason for why type checking in The Gamma can be time consuming. Other
-type providers may perform other more computationally intensive work to provide types and so
-it is desirable to reuse type-checking results during live coding. The rest of this section
-shows how this is done using the dependency graph discussed in Section~\ref{sec:formal}.
-
-\begin{figure}[!b]
-\vspace{-0.5em}
-\begin{equation*}
-\xymatrix{
-&\bnd{val}(o)\\
-&\bnd{mem}(m, s_0) \ar[u]_{\blbl{arg}(0)} \ar[dl]^{\blbl{arg}(1)} \\
-\bnd{fun}(x, s_2) \ar@/^/[uur]^{\blbl{callsite}(m, 1)} \ar[rr]^{\blbl{body}} && \bnd{var}(x, s_1) \ar@/_/[uul]_{\blbl{callsite}(m, 1)}
-}
-\end{equation*}
-\caption{Dependency graph for $o.m(\lambda x\rightarrow x)$}
-\label{fig:graph-func}
-\end{figure}
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Revised binding of functions}
-\label{sec:types-funs}
-
-The Gamma script supports lambda functions, but only in a limited way. A function can be passed
-as a parameter to a method, which makes type checking of functions easier. For example, consider:
-%
-\begin{equation*}
-\begin{array}{l}
-\ident{movies}.\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())
-\end{array}
-\end{equation*}
-%
-If \ident{movies} is a collection of \ident{Movie} objects, the type of the lambda function must be
-$\ident{Movie}\rightarrow\ident{bool}$ and so the type of $x$ is \ident{Movie}. This is similar to
-type checking of lambda functions in C\# \cite{csharp}, where type is also inferred from the context (or has
-to be specified explicitly).
-
-We do not currently allow lambda functions as stand-alone let-bound
-values. This could be done by requiring explicit types, or introducing polymorphism, but it was not
-necessary for the limited domain of non-expert data exploration.
-
-\paragraph{Dependency graph for functions.}
-In the binding process specified in Section~\ref{sec:formal}, a variable is a leaf of the dependency
-graph. In the revised model, it depends on the context in which it appears. A new edge labeled
-$\blbl{callsite}(m, i)$ indicates that the source node is the input variable of a function
-passed as the $i^\textnormal{th}$ argument to the $m$ member of the expression represented by the
-target node. A node representing function is linked to the call site using the same edge.
-
-Figure~\ref{fig:graph-func} shows the result of binding $o.m(\lambda x\rightarrow x)$. Both
-$\bnd{fun}(x, s_2)$ and $\bnd{var}(x, s_1)$ now depend on the node representing $o$. The new
-$\blbl{callsite}$ edge makes it possible to type-check function and variable nodes just using their
-dependencies. As before, the member invocation $\bnd{mem}(m, s_0)$
-depends on the instance using $\blbl{arg}(0)$ and on its argument using $\blbl{arg}(1)$.
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure*}
-\vspace{0.5em}
-\begin{equation*}
-\begin{array}{l}
-\qquad\inference[(val)~]
-  {\Sigma(n) = \alpha}
-  {\bnd{val}(n) \vdash \alpha }
-\quad\qquad
-\inference[(var)~]
-  { (\bnd{var}(x,s), \bndclr{v}, \blbl{callsite}(m, i)) \in E &
-    \bndclr{v} \vdash (.., m:(\tau_1, \ldots, \tau_k)\rightarrow \tau,..) & \tau_i=\tau'\rightarrow\tau'' }
-  { \bnd{var}(x, s) \vdash \tau' }
-\\[2.5em]
-\inference[(fun)~]
-  { \{(\bnd{fun}(x,s), \bndclr{v_b}, \blbl{body}), (\bnd{var}(x,s), \bndclr{v_c}, \blbl{callsite}(m, i))\} \subseteq E &
-    \bndclr{v_c} \vdash (.., m:(\tau_1, \ldots, \tau_k)\rightarrow \tau,..) &
-    \tau_i=\tau'\rightarrow\tau'' & \bndclr{v_b} \vdash \tau'' }
-  { \bnd{fun}(x, s) \vdash \tau' \rightarrow \tau'' }
-\\[2.5em]
-\qquad\qquad\quad\inference[(mem)~]
-  { \forall i\in\{0\ldots k\}.(\bnd{mem}(m, s), \bndclr{v_i}, \blbl{arg}(i)) \in E &
-    \bndclr{v_0} \vdash (.., m:(\tau_1, \ldots, \tau_k) \rightarrow \tau, ..)& \bndclr{v_i} \vdash \tau_i }
-  { \bnd{mem}(m, s) \vdash \tau }
-\\[1em]
-\end{array}
-\end{equation*}
-\caption{Rules that define evaluation of previews over a dependency graph for an expression}
-\vspace{0.5em}
-\label{fig:tc}
-\end{figure*}
-
-% --------------------------------------------------------------------------------------------------
-
-\paragraph{Revised binding process.}
-For the revised binding process, we introduce a new edge label $\blbl{callsite}$. Variable nodes
-now have dependencies and so we cache them and attach a symbol $s$ to \bnd{var}, so we also
-introduce a node kind $\bknd{var}$ as part of a lookup key for $\Delta$.
-
-The \ident{bind-expr} function now has a parameter $c$, in addition to $\Gamma$ and $\Delta$, which
-represents the context in which the binding happens. This is either a member invocation (labeled
-with instance node $\bndclr{v}$ and \blbl{callsite} label $\blblclr{l}$), or not a member
-invocation written as $\bot$. The updated definitions are:
-%
-\begin{equation*}
-\begin{array}{lcll}
-\bndclr{v} &\hspace{-0.5em}\in\hspace{-0.5em}&\bnd{val}(n)~|~\bnd{var}(x, s)~|~\bnd{mem}(m, s)~|~\bnd{fun}(x, s)&(\textit{Vertices})\\
-\blblclr{l}&\hspace{-0.5em}\in\hspace{-0.5em}&\blbl{body}~|~\blbl{arg}(i)~|~\blbl{callsite}(m,i)&(\textit{Edge labels})\\
-\bkndclr{k}&\hspace{-0.5em}\in\hspace{-0.5em}&\bknd{fun}(x)~|~\bknd{mem}(m)~|~\bknd{var}(x)&(\textit{Node kinds})\\
-          c&\hspace{-0.5em}\in\hspace{-0.5em}&\bot ~|~(\bndclr{v}, \blblclr{l})&(\textit{Call sites})\\
-\end{array}
-\end{equation*}
-%
-The key parts of the revised definition of the \ident{bind-expr} function are shown in Figure~\ref{fig:binding-rules-callsite}.
-We now write $\ident{bind-expr}_{\Gamma,\Delta,c}$ where $c$ represents the context in which the
-binding occurs. This is set to $\bot$ in all cases, except when binding arguments of a member call.
-In (2b), we first recursively bind the instance node using $\bot$ as the context and then
-bind all the arguments using $(\bndclr{v_0}, \blbl{callsite}(m, i))$ as the context for $i^{\textnormal{th}}$
-argument. The rest is as in case (2) before. The case (1) is updated similarly and is not shown
-here for brevity.
-
-When binding a function (7b), we now also store variable nodes in $\Delta$ and so we check if a
-variable with the given call site exists. If no, we create a fresh node $\bnd{var}(x, s_x)$.
-The node is added to $\Gamma$ as before. At the end, we now also include a call site edge from the
-variable node and from the function node in $E$. We omit a similar variant of the case (6).
-The remaining cases (3)-(5) are the same, except that \ident{bind-expr} has the additional $c$ parameter
-and recursive calls always set it to $\bot$.
-
-Finally, the \ident{update} function in Figure~\ref{fig:loop} also needs to be updated to store the
-newly created \bnd{var} nodes. This is done by adding the following single case:
-
-\begin{equation*}
-\begin{array}{l}
-\Delta_{i}(\bknd{var}(x), [(\bndclr{v},\blbl{callsite}(m,i))]) = \bnd{var}(x, s)\\
-\quad\quad \textnormal{for all}~\bnd{var}(x, s) \in V\\
-\quad\quad \textnormal{such that}~(\bnd{var}(x, s), \bndclr{v}, \blbl{callsite}(m,i)) \in E
-\end{array}
-\end{equation*}
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Type checking over dependency graphs}
-\label{sec:types-graph}
-
-The type system for The Gamma supports a number of primitive types (such as integers and strings)
-written as $\alpha$. Composed types include functions and objects with members. Objects are be
-provided by type providers, but we omit the details here. Types of object members are written as
-$\sigma$ and can have multiple arguments:
-%
-\begin{equation*}
-\begin{array}{lcll}
-\tau&\in&\alpha ~|~ \tau \rightarrow \tau ~|~ \{m_1\!:\!\sigma_1, \ldots, m_n\!:\!\sigma_n\} &\quad(\textit{Types})\\
-\sigma &\in& (\tau_1, \ldots, \tau_n) \rightarrow \tau &\quad(\textit{Members})\\
-\end{array}
-\end{equation*}
-%
-The typing judgements are written in the form $\bnd{v}\vdash \tau$. They are parameterised by the
-dependency graph $(V, E)$, but this is not modified during type checking so we keep it implicit
-rather than writing e.g.~$\bnd{v}\vdash_{(V,E)} \tau$.
-
-\paragraph{Type checking.}
-The typing rules are shown in Figure~\ref{fig:tc}. Types of primitive values $n$ are obtained
-using a global lookup table $\Sigma$ (\emph{val}). When type checking a member call (\emph{mem}),
-we find its dependencies $\bndclr{v_i}$ and check that the first one (instance) is an object with
-the required member $m$. The types of input arguments of the member then need to match the types
-of the remaining (non-instance) nodes.
-
-Type checking a function (\emph{fun}) and a variable (\emph{var}) is similar. In both cases,
-we follow the \blbl{callsite} edge to find the member that accepts the function as an argument.
-We obtain the type of the function from the type of the $i^\textnormal{th}$ argument of the
-member. We use the input type as the type of variable (\emph{var}). For functions, we also check
-that the resulting type matches the type of the body (\emph{fun}).
-
-\paragraph{Caching results.}
-Performing type checking over the dependency graph, rather than over the abstract syntax tree,
-enables us to reuse the results of previously type checked parts of a program. As when
-caching evaluated previews (Section~\ref{sec:previews}), we build a lookup table mapping graph
-nodes to types. When type checking a node, we first check the cache and, only if it is new,
-follow the $\vdash$ relation to obtain the type.
-
-As a result, code can be type checked on-the-fly during editing, even when asynchronous type
-providers are used, and the programmer gets instant feedback without delays.
+\subsection{Theories of delayed previews}
+The operational semantics presented in this paper serves two purposes. It gives a simple guide
+for implementing text-based live coding environments for data science and we use it to prove that
+our optimized way of producing live previews is correct. However, some aspects of our mechanism
+are related to important work in semantics of programming languages and deserve to be mentioned.
+
+The construction of delayed previews is related to meta-programming. Assuming
+we have delayed previews $\llbracket e_0 \rrbracket_x$ and $\llbracket e_1 \rrbracket_y$ and
+we invoke a member $m$ on $e_0$ using $e_1$ as an argument. To do this, we construct a new
+delayed preview $\llbracket e_0.m(e_1) \rrbracket_{x, y}$. This operation is akin to expression
+splicing from meta-programming \cite{metaml,quotations}.
+
+The semantics of delayed previews can be more formally captured by Contextual Modal Type Theory
+(CMTT) \cite{cmtt} and comonads \cite{cmtt-denotation}. In CMTT, $\lbrack \Psi \rbrack A$ denotes
+that a proposition $A$ is valid in context $\Psi$, which is similar to our delayed previews written
+as $\llbracket A \rrbracket_\Psi$. CMTT defines rules for composing context-dependent propositions
+that would allow us to express the splicing operation used in \rname{mem-expr}. In categorical
+terms, the context-dependent proposition can be modeled as an indexed comonad \cite{effectrev,graded}.
+The evaluation of a preview with no context dependencies (built implicitly into our evaluation rules)
+corresponds to the counit operation of a comonad and would be explicitly written as
+$\llbracket A \rrbracket_\emptyset \rightarrow A$.
 
@@
 
-The dependency graph makes it possible to cache partial results when evaluating previews.
-The mechanism needs to satisfy two properties.
-First, if we evaluate a preview using dependency graph with caching, it should be the same as the
-value we would obtain by evaluating the expression directly. Second, the evaluation of previews
-using dependency graphs should -- in some cases -- reuse previously evaluated partial results.
-In other words, we show that the mechanism is correct and implements a useful optimization.
-
-\subsection{Modeling expression evaluation}
-In The Gamma language, computations are expressed using
-member access, written as $e.m(e_1,\ldots, e_n)$. In this paper, we do not define how member
-access evaluates. This has been done elsewhere \cite{gamma}, but more importantly, the evaluation of
-previews does not rely on the exact specifics of the evaluation, provided that the language
-satisfies certain basic conditions. The following definitions provides the necessary structure
-for discussing correctness of previews.
-
-Partial evaluation may reduce an expression under $\lambda$-abstraction. We do not require that
-the reduction of the host language does this. Instead, we define an extended reduction relation
-and use that in the proofs. The host language only needs to compose well with such extended reduction
-as captured by the \emph{compositionality} property below. We also require that the language allows
-elimination of let bindings.
-
-\begin{definition}[Host language]
-\label{def:host}
-Given a relation on expressions $e_1 \rightsquigarrow e_2$ that models small-step evaluation, we define:
-
-\begin{itemize}
-\item[--]
- A \emph{preview evaluation context} (also referred to as \emph{context}):
-
-\vspace{-1.25em}
-\begin{equation*}
-\begin{array}{lcl}
-C[-] &\hspace{-1em}=\hspace{-1em}& \kvd{let}~x = -~\kvd{in}~e ~|~ \kvd{let}~x = e~\kvd{in}~- ~|~ \lambda x\rightarrow - \\
-     &&  e_0.m(e_1, \ldots, e_{k-1}, -, e_{k+1}, \ldots e_n)
-\end{array}
-\end{equation*}
-
-\item[--]
-An extended reduction relation $\rightsquigarrow_\beta$ such that, for any context $C$,
-$C[e_1] \rightsquigarrow_\beta C[e_2]$ whenever $e_1 \rightsquigarrow e_2$.
-
-\item[--]
-Let elimination $\rightsquigarrow_\ident{let}$ such that, using  capture-avoiding substitution,
-  $C[\kvd{let}~x=e_1~\kvd{in}~e_2] \rightsquigarrow_\ident{let} C[e_2[x\leftarrow e_1]]$
-\end{itemize}
-
-\noindent
-We say that $\rightsquigarrow$ is a suitable \emph{host language reduction} if:
-\begin{itemize}
-\item[--] It satisfies the \emph{compositionality} property, that is if $e \rightsquigarrow e'$ and
-$C[e]\rightsquigarrow_\beta e''$ then also $C[e']\rightsquigarrow_\beta e''$.
-
-\item[--] Let elimination does not affect the result, i.e.~if $e \rightsquigarrow_{\ident{let}} e'$
-and $e'\rightsquigarrow_\beta e''$ then also $e\rightsquigarrow_\beta e''$
-\end{itemize}
-\end{definition}
-
-\noindent
-The host language in The Gamma is a simple call-by-value functional language without side-effects,
-and so it satisfies both compositionality and allows let bindings to be eliminated, although the
-latter affects the performance. The mechanism for preview evaluation presented here would also
-work for call-by-name languages, but it would suffer from the expected difficulties in the
-presence of side-effects or non-determinism.
+The use of a dependency graph when evaluating previews makes it possible to cache partial results
+through a mechanism discussed in Section~\ref{sec:formal-deps} and Section~\ref{sec:previews-cache}.
+The mechanism satisfies two properties. First, if we evaluate a preview using dependency
+graph with caching, it is the same as the value we would obtain by evaluating the expression
+directly. Second, the evaluation of previews using dependency graphs reuses -- in some cases --
+previously evaluated partial results. In other words, we show that the mechanism is correct
+and implements a useful optimization.
 
@@
 To show that the evaluated previews are correct, we prove two properties. Correctness
-(Theorem~\ref{thm:correcntess}) guarantees that, no matter how a graph is constructed, when we use
-it to evaluate a preview for an expression, the preview is the same as the value we would obtain
-by evaluating the expression directly. Determinacy (Theorem~\ref{thm:determinacy}) guarantees that if
-we cache a preview for a graph node and update the graph, the preview we would evaluate using the
-updated graph would be the same as the cached preview.
-
-To simplify the proofs, we consider expressions without let bindings. This is possible, because
-eliminating let bindings does not change the result in the host language (Definition~\ref{def:host})
-and it also does not change the constructed dependency graph as shown in Lemma~\ref{thm:let-elimination}.
-
-\begin{lem}[Let elimintion]
-\label{thm:let-elimination}
-Given an expression $e_1$ such that $e_1 \rightsquigarrow_\ident{let} e_2$ and a lookup table $\Delta_0$
-then if $\bndclr{v_1}, (V_1, E_1) = \ident{bind-expr}_{\emptyset, \Delta_0}(e_1)$ and
-$\bndclr{v_2}, (V_2, E_2) = \ident{bind-expr}_{\emptyset, \Delta_1}(e_2)$ such that $\Delta_1 = \ident{update}_{V_1, E_1}(\Delta_0)$
-then it holds that $\bndclr{v_1} = \bndclr{v_2}$ and also $(V_1, E_1) = (V_2, E_2)$.
+(Theorem~\ref{thm:correcntess}) guarantees that, no matter how a dependency graph is constructed,
+when we use it to evaluate previews for a program, the previews are the same as the values we
+would obtain by evaluating the program commands directly. Determinacy (Theorem~\ref{thm:determinacy})
+guarantees that if we cache a preview for a graph node and update the graph, the preview we would
+evaluate using the updated graph would be the same as the cached preview.
+
+To simplify the proofs, we consider programs without let bindings. This is possible, because
+eliminating let bindings does not change the result of evaluation, as shown earlier in
+Lemma~\ref{thm:let-lang-elimination}, and it also does not change the constructed dependency graph
+as shown below in Lemma~\ref{thm:let-grp-elimination}.
+
+\begin{lem}[Let elimintion for a dependency graph]
+\label{thm:let-grp-elimination}
+Given programs $p_1, p_2$ such that $p_1 \rightsquigarrow_\ident{let} p_2$ and a lookup table
+$\Delta_0$ then if $\bndclr{v_1}; \ldots; \bndclr{v_n}, (V, E) = \ident{bind-prog}_{\emptyset, \Delta_0}(p_1)$ and
+$\bndclr{v'_1};\ldots;\bndclr{v'_n}, (V', E') = \ident{bind-prog}_{\emptyset, \Delta_1}(p_2)$ such that $\Delta_1 = \ident{update}_{V, E}(\Delta_0)$
+then for all $i$, $\bndclr{v_i} = \bndclr{v'_i}$ and also $(V, E) = (V', E')$.
 \end{lem}
 \begin{proof}
-Assume $e_1 = C[\kvd{let}~x=e'~\kvd{in}~e'']$ and the resulting $e_2 = C[e''[x\leftarrow e']]$. The case
-$\ident{bind-expr}_{\Gamma,\Delta}(\kvd{let}~x=e'~\kvd{in}~e'')$ when binding $e_1$ is handled using
-(3).
-
-When binding $e_1$, the node resulting from binding $e'$ is added to the graph $V_1, E_1$ and is
-referenced each time $x$ is used. When binding $e_2$, the node representing $e'$ is a primitive value,
-or already present in $\Delta_1$ (added by $\ident{update}_{V_1, E_1}$) and is reused each time
-$\ident{bind-expr}_{\Gamma,\Delta_1}(e')$ is called.
+Assume $p_1 = c_1; \ldots; c_{k-1}; \kvd{let}~x=e; c_{k+1};\ldots; c_n$ and the let binding is
+eliminated resulting in $p_2 = c_1; \ldots; c_{k-1}; e; c_{k+1}[x\leftarrow e];\ldots; c_n[x\leftarrow e]$.
+When binding $p_1$, the case $\ident{bind-prog}_{\Gamma,\Delta}(\kvd{let}~x=e)$ is handled using (7)
+and the node resulting from binding $e$ is added to the graph $V, E$. It is then referenced each
+time $x$ appears in subsequent commands $c_{k+1}; \ldots; c_n$.
+When binding $p_2$, the node resulting from binding $e$ is a primitive value or a node already
+present in $\Delta_1$ (added by $\ident{update}_{V, E}$) and is reused each time
+$\ident{bind-expr}_{\Gamma,\Delta_1}(e)$ is called.
 \end{proof}
 
-The Lemma~\ref{thm:let-elimination} provides a way of removing let bindings from an expression,
-such that the resulting dependency graph remains the same. Here, we bind the original expression
-first, which adds the node for $e'$ to $\Delta$. In our implementation, this is not needed
+The Lemma~\ref{thm:let-grp-elimination} provides a way of removing let bindings from a program,
+such that the resulting dependency graph remains the same. Here, we bind the original program
+first, which adds the node for $e$ to $\Delta$. In our implementation, this is not needed
 because $\Delta$ is updated while the graph is being constructed using $\ident{bind-expr}$.
 To keep the formalisation simpler, we separate the process of building the dependency graph
-and updating $\Delta$.
+and updating $\Delta$ and thus Lemma~\ref{thm:let-grp-elimination} requires an extra binding step.
 
@@
 constructed dependency graph is the same as the one we would obtain by directly evaluating the
-expression. This requires a simple auxiliary lemma and the full proof is shown in
-Appendix~\ref{sec:app-correctness}.
-
-\begin{lem}
-\label{thm:lemma-lookup}[Lookup inversion]
+expression. This requires a simple auxiliary lemma.
+
+\begin{lem}[Lookup inversion]
+\label{thm:lemma-lookup}
 Given $\Delta$ obtained using \ident{update} as defined in Figure~\ref{fig:loop} then:
@@
 
-\begin{theorem}[Let-free correctness]
+\begin{theorem}[Term preview correctness]
 \label{thm:let-free-correct}
-Given an expression $e$ that has no free variables and does not contain let bindings, together
-with a lookup table $\Delta$ obtained from any sequence of expressions according to Figure~\ref{fig:loop}
-let $\bndclr{v}, (V, E) = \ident{bind-expr}_{\emptyset, \Delta}(e)$. If $\bndclr{v}\Downarrow d$
-over a graph $(V, E)$ then $d = p$ for some $p$ and $e \rightsquigarrow_\beta p$.
+Given a term $t$ that has no free variables, together with a lookup table $\Delta$ obtained
+from any sequence of programs using $\ident{bind-prog}$ (Figure~\ref{fig:binding-rules-prog}) and
+$\ident{update}$ (Figure~\ref{fig:loop}), then
+let $\bndclr{v}, (V, E) = \ident{bind-expr}_{\emptyset, \Delta}(t)$.
+
+\vspace{0.25em}
+\noindent
+If $\bndclr{v}\Downarrow p$
+over a graph $(V, E)$ then $p = o$ for some value $o$ and $t \rightsquigarrow^{*} o$.
+\end{theorem}
+
+\begin{proof}
+First note that, when combining recursively constructed sub-graphs, the \ident{bind-expr} function
+adds new nodes and edges leading from those new nodes. Therefore, an evaluation using~$\Downarrow$
+over a sub-graph will also be valid over the new graph -- the newly added nodes and edges do not introduce
+non-determinism to the rules given in Figure~\ref{fig:eval}.
+
+We prove a more general property showing that for any $e$, its binding
+$\bndclr{v}, (V, E) = \ident{bind-expr}_{\emptyset, \Delta}(e)$ and any evaluation context $C$
+such that $C[e]\rightsquigarrow o$ for some $o$, one of the following holds:
+%
+\begin{enumerate}
+\item[a.] If $FV(e)\,=\,\emptyset$ then $\bndclr{v} \Downarrow p$ for some $p$ and $C[p] \rightsquigarrow o$
+\item[b.] If $FV(e)\neq\emptyset$ then $\bndclr{v} \Downarrow \llbracket e_p \rrbracket_{FV(e)}$ for some $e_p$ and $C[e_p] \rightsquigarrow o$
+\end{enumerate}
+%
+In the first case, $p$ is a value, but it is not always the case that $e \rightsquigarrow^{*} p$,
+because $p$ may be lambda function and preview evaluation may reduce sub-expression in the body of
+the function. Using a context $C$ in which the value reduces to an object avoids this problem.
+
+The proof of the theorem follows from the more general property. Using a context $C[-]=-$, the
+term $t$ reduces $t \rightsquigarrow^{*}t' \rightsquigarrow_\epsilon o$ for some $o$ and the
+preview $p$ is a value $o$ because $C[p] = p = o$.
+The proof is by induction over the binding process, which follows the structure of the expression.
+The full proof is shown in Appendix~\ref{sec:app-correctness}.
+\end{proof}
+
+\noindent
+The correctness theorem combines the previous two results.
+
+\begin{theorem}[Program preview correctness]
+\label{thm:correcntess}
+Consider a program $p=c_1;\ldots;c_n$ that has no free variables, together with a lookup table
+$\Delta_0$ obtained from any sequence of programs using $\ident{bind-prog}$
+(Figure~\ref{fig:binding-rules-prog}) and $\ident{update}$ (Figure~\ref{fig:loop}). Assume a
+let-free program $p'=t_1;\ldots;t_n$ such that $p\rightsquigarrow^{*}_\ident{let}p'$.
+
+\vspace{0.25em}
+\noindent
+Let $\bndclr{v_1};\ldots;\bndclr{v_n}, (V, E) = \ident{bind-prog}_{\emptyset, \Delta_0}(p)$
+and define updated lookup table $\Delta_1=\ident{update}_{V, E}(\Delta_0)$ and
+let $\bndclr{v'_1};\ldots;\bndclr{v'_n}, (V', E') = \ident{bind-prog}_{\emptyset, \Delta_1}(p)$.
+
+\vspace{0.25em}
+\noindent
+If $\bndclr{v'_i}\Downarrow p_i$ over a graph $(V', E')$ then $p_i=o_i$ for some value $o_i$ and
+$t_i \rightsquigarrow o_i$.
 \end{theorem}
 \begin{proof}
-First note that, when combining recursively constructed sub-graphs, the \ident{bind-expr} operation
-adds new nodes and edges leading from those new nodes. Therefore, an evaluation using $\Downarrow$
-over a sub-graph will also be valid over the new graph.
-
-Next, we prove a more general property using induction showing that for $e$ such
-that $\bndclr{v}, (V, E) = \ident{bind-expr}_{\emptyset, \Delta}(e)$:
-\begin{enumerate}
-\item[a.] If $FV(e)\,=\,\emptyset$ then $\bndclr{v} \Downarrow p$ for some $p$ and $e \rightsquigarrow_\beta p$
-\item[b.] If $FV(e)\neq\emptyset$ then $\bndclr{v} \Downarrow \llbracket e_p \rrbracket_{FV(e)}$ for some $e_p$ and
-  for any evaluation context $C[-]$ such that $FV(C[e_p])=\emptyset$ it holds that if
-  $C[e] \rightsquigarrow_\beta C[e_p]$.
-\end{enumerate}
-%
-The proof is done by induction over the binding process, which follows the structure
-of the expression $e$ and can be found in Appendix~\ref{sec:app-correctness}.
+Direct consequence of Lemma~\ref{thm:let-grp-elimination} and Theorem~\ref{thm:let-free-correct}.
 \end{proof}
 
-\noindent
-The correctness theorem combines the previous two results.
-
-\begin{theorem}[Correctness]
-\label{thm:correcntess}
-Consider an expression $e_1$ that has no free variables together with a lookup table $\Delta_1$ obtained
-from any sequence of expressions according to Figure~\ref{fig:loop} and $e_2$ such that
-$e_1\rightsquigarrow_\beta e_2$ and let $\bndclr{v_1}, (V_1, E_1) = \ident{bind-expr}_{\emptyset, \Delta_1}(e_1)$.
-
-Let $\Delta_2 = \ident{update}_{V_1, E_1}(\Delta_1)$ and $\bndclr{v_2}, (V_2, E_2) = \ident{bind-expr}_{\emptyset, \Delta_2}(e_2)$.
-If $\bndclr{v_2}\Downarrow d$ over a graph $(V_2, E_2)$ then $d = p$ for some $p$ and $e \rightsquigarrow_\beta p$.
+As discussed earlier, our implementation updates $\Delta$ during the recursive binding process and
+so a stronger version of the property holds -- namely, previews calculated over a graph obtained
+directly for the original program $p$ are the same as the values of the fully evaluated program.
+We note that this is the case, but do not show it formally to aid the clarity of our formalisation.
+
+The second important property that guarantees that our mechanism always displays correct previews
+is determinacy. This makes it possible to cache the previews evaluated using~$\Downarrow$ using
+the corresponding graph node as a lookup key.
+
+\begin{theorem}[Program preview determinacy]
+\label{thm:determinacy}
+For some $\Delta$ and for any programs $p, p'$, assume that the first program is bound,
+i.e.~$\bndclr{v_1};\ldots;\bndclr{v_n}, (V, E) = \ident{bind-prog}_{\emptyset, \Delta}(p)$,
+the graph node cache is updated $\Delta' = \ident{update}_{V, E}(\Delta)$ and the second program is
+bound, i.e.~$\bndclr{v'_1};\ldots;\bndclr{v'_m}, (V', E') = \ident{bind-prog}_{\emptyset, \Delta'}(p')$.
+Now, for any $\bndclr{v}$, if $\bndclr{v} \Downarrow p$ over $(V, E)$ then also
+ $\bndclr{v} \Downarrow p$ over $(V', E')$.
 \end{theorem}
 \begin{proof}
-Direct consequence of Lemma~\ref{thm:let-elimination} and Theorem~\ref{thm:let-free-correct}.
+By induction over $\Downarrow$ over $(V, E)$, we show that the same evaluation rules also
+apply over $(V', E')$.
+This is the case, because new graph nodes added to $\Delta'$ by $\ident{update}_{V, E}$
+are only ever added as new nodes in $\ident{bind-prog}_{\emptyset, \Delta'}$ and so the existing
+nodes and edges of $(V, E)$ used during the evaluation are unaffected.
 \end{proof}
 
-As discussed above when introducing Lemma~\ref{thm:let-elimination}, in our implementation,
-$\Delta$ is updated during the recursive binding process and so a stronger version of the property
-holds -- namely, $e \rightsquigarrow_\beta p$ for a $p$ that is obtained by calculating preview
-over a graph obtained directly for the original expression $e$. We note that this is the case, but
-do not show it formally to keep aid the clarity of our formalisation.
-
-The second important property that guarantees the correctness of previews shown by the user in
-our implementation is determinacy. This makes it possible to cache the previews evaluated using
-$\Downarrow$ using the corresponding graph node as a lookup key.
-
-\begin{theorem}[Determinacy]
-\label{thm:determinacy}
-Let $\Delta_1 = \emptyset$, for any $e_1, e_2$, assume that the first expression is bound,
-i.e.~$\bndclr{v_1}, (V_1, E_1) = \ident{bind-expr}_{\emptyset, \Delta_1}(e_1)$, the graph node cache
-is updated $\Delta_2 = \ident{update}_{V_1, E_1}(\Delta_1)$ and a new expression is bound, i.e.~
-$\bndclr{v_2}, (V_2, E_2) = \ident{bind-expr}_{\emptyset, \Delta_2}(e_2)$. Now, for any $\bndclr{v}$,
-if $\bndclr{v} \Downarrow p$ over $(V_1, E_1)$ then also
- $\bndclr{v} \Downarrow p$ over $(V_2, E_2)$.
-\end{theorem}
-\begin{proof}
-By induction over $\Downarrow$ over $(V_1, E_1)$, we show that the same evaluation rules also
-apply over $(V_2, E_2)$.
-
-This is the case, because new graph nodes added to $\Delta_2$ by $\ident{update}_{V_1, E_2}$
-are only ever added as new nodes in $\ident{bind-expr}_{\emptyset, \Delta_2}$ and so the existing
-nodes and edges of $(V_1, E_1)$ used during the evaluation are unaffected.
-\end{proof}
-
-The mechanism used for caching previews, as discussed at the end of Section~\ref{sec:previews},
+The mechanism used for caching previews, as discussed in Section~\ref{sec:previews-cache},
 keeps a preview or a partial preview $d$ in a lookup table indexed by nodes $\bndclr{v}$. The
@@
 
+% ==================================================================================================
+
+\section{Type checking}
+\label{sec:types}
+
+Evaluating live previews can be an expensive operations, so being able to cache partial previews
+is a must for a live coding environment. Type checking is typically fast, so the main focus of
+this paper is on live previews. However, asynchronous type providers in The Gamma (Section~\ref{sec:types-providers})
+can make type checking time consuming, and so we use the dependency graph also for type checking
+(Section~\ref{sec:types-graph}). Type checking lambda functions (Section~\ref{sec:types-funs})
+requires a slight extension of the model discussed in Section~\ref{sec:formal}.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Asynchronously provided types}
+\label{sec:types-providers}
+
+Data available in The Gamma can be defined using several kinds of type providers. The type provider
+used in Figure~\ref{fig:thegamma} as asynchronous \cite{async}. It downloads the sample URL and generates
+types based on the contents of the web page. The parameter to \ident{web.scrape} is a static
+parameter and is evaluated during type-checking. We omit details in this paper, but we note this
+works similarly to F\# type providers \cite{providers-fsharp}.
+
+Type providers can also be implemented as REST services \cite{rest} to allow anyone implement a data source
+in the language of their choice. In this case, each member of a call-chain returns a type that is
+generated based on the result of an HTTP request. For example, when the user types \ident{worldbank}
+(to access information about countries), the type provider makes a request to
+\url{http://thegamma-services.azurewebsites.net/worldbank}, which returns two members:
+%
+\begin{equation*}
+\begin{array}{l}
+[~\{\rstr{name}\!:\str{byYear},\rstr{returns}\!:\\
+~\quad \{\rstr{kind}\!:\str{nested},\rstr{endpoint}\!:\str{/pickYear}\}\}\\
+\hspace{0.55em}\{\rstr{name}\!:\str{byCountry},\rstr{returns}\!:\\
+~\quad \{\rstr{kind}\!:\str{nested},\rstr{endpoint}\!:\str{/pickCountry}\}\}~]
+\end{array}
+\end{equation*}
+%
+This indicates that \ident{worldbank} has members \ident{byYear} and \ident{byCountry}.
+If the user types \ident{worldbank.byCountry}, a request is made to the specified URL
+\url{http://thegamma-services.azurewebsites.net/worldbank/pickCountry}:
+%
+\begin{equation*}
+\begin{array}{l}
+[~\{\rstr{name}\!:\str{Andorra},\rstr{trace}\!:[\str{"country=AR"}], \\
+\quad \rstr{returns}\!:\{\rstr{kind}\!:\str{nested},\rstr{endpoint}\!:\str{/pickTopic}\}\}\\
+\hspace{0.55em}\{\rstr{name}\!:\str{Afghanistan},\rstr{trace}\!:[\str{"country=AF"}],\\
+\quad \rstr{returns}\!:\{\rstr{kind}\!:\str{nested},\rstr{endpoint}\!:\str{/pickTopic}\}\},\,..]
+\end{array}
+\end{equation*}
+%
+This returns a list of countries which can then be accessed as members via
+\ident{worldbank.byCountry.Andorra}, etc.
+
+This is one reason for why type checking in The Gamma can be time consuming. Other
+type providers may perform other more computationally intensive work to provide types and so
+it is desirable to reuse type-checking results during live coding. The rest of this section
+shows how this is done using the dependency graph discussed in Section~\ref{sec:formal}.
+
+\begin{figure}[!b]
+\vspace{-0.5em}
+\begin{equation*}
+\xymatrix{
+&\bnd{val}(o)\\
+&\bnd{mem}(m, s_0) \ar[u]_{\blbl{arg}(0)} \ar[dl]^{\blbl{arg}(1)} \\
+\bnd{fun}(x, s_2) \ar@/^/[uur]^{\blbl{callsite}(m, 1)} \ar[rr]^{\blbl{body}} && \bnd{var}(x, s_1) \ar@/_/[uul]_{\blbl{callsite}(m, 1)}
+}
+\end{equation*}
+\caption{Dependency graph for $o.m(\lambda x\rightarrow x)$}
+\label{fig:graph-func}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure*}[t]
+\vspace{-0.5em}
+\begin{equation*}
+\begin{array}{ll}
+\ident{bind-expr}_{\Gamma, \Delta, c}(e_0.m(e_1, \ldots, e_n)) = &(2b)\\
+\qquad \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n)\\
+\quad \textnormal{when}~\bndclr{v_0}, (V_0, E_0) = \ident{bind-expr}_{\Gamma, \Delta, \bot}(e_0)\\
+\quad \textnormal{and}~c_i = (\bndclr{v_0}, \blbl{callsite}(m,i))\hspace{4.8em} (i \in 1\ldots n)\\
+\quad \textnormal{and}~\bndclr{v_i}, (V_i, E_i) = \ident{bind-expr}_{\Gamma, \Delta, c_i}(e_i)\qquad (i\in 1\ldots n)\\
+\quad \textnormal{and}~(\bknd{mem}(m),[(\bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v_n}, \blbl{arg}(n))]) \notin \ident{dom}(\Delta)\hspace{-200em}.\\
+\quad \textnormal{let}~\bndclr{v} = \bnd{mem}(m, s), s~\textnormal{fresh}\\
+\quad \textnormal{let}~E = \{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v},\bndclr{v_n}, \blbl{arg}(n)) \}
+\end{array}
+\qquad\qquad
+\begin{array}{ll}
+\ident{bind-expr}_{\Gamma, \Delta, (\bndclr{v_c}, \blblclr{l_c} )}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V_0, E \cup E_0)&(7b)\\
+\quad \textnormal{when}~(\bknd{var}(x),[(\bndclr{v_c}, \blblclr{l_c} )])\notin \ident{dom}(\Delta)\\
+\quad \textnormal{let}~\bndclr{v_x} = \bnd{var}(x, s_x), s_x~\textnormal{fresh}\\
+\quad \textnormal{and}~\Gamma_1 = \Gamma \cup \{ x, \bndclr{v_x} \}\\
+\quad \textnormal{and}~\bndclr{v_0}, (V_0, E_0) = \ident{bind-expr}_{\Gamma_1, \Delta, \bot}(e)\\
+\quad \textnormal{when}~(\bknd{fun}(x),[(\bndclr{v_0}, \blbl{body}), (\bndclr{v_c}, \blblclr{l_c} )])\notin \ident{dom}(\Delta)\\
+\quad \textnormal{let}~\bndclr{v} = \bnd{fun}(x, s_f), s_f~\textnormal{fresh}\\
+\quad \textnormal{let}~E = \{ (\bndclr{v},\bndclr{v_0},\blbl{body}), (\bndclr{v},\bndclr{v_c},\blblclr{l_c}), (\bndclr{v_x},\bndclr{v_c},\blblclr{l_c}) \}
+\\[0.75em]
+\end{array}
+\end{equation*}
+\vspace{-0.5em}
+\caption{Revised binding rules, tracking call sites of function values.}
+\label{fig:binding-rules-callsite}
+\vspace{-0.5em}
+\end{figure*}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Revised binding of functions}
+\label{sec:types-funs}
+
+The Gamma script supports lambda functions, but only in a limited way. A function can be passed
+as a parameter to a method, which makes type checking of functions easier. For example, consider:
+%
+\begin{equation*}
+\begin{array}{l}
+\ident{movies}.\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())
+\end{array}
+\end{equation*}
+%
+If \ident{movies} is a collection of \ident{Movie} objects, the type of the lambda function must be
+$\ident{Movie}\rightarrow\ident{bool}$ and so the type of $x$ is \ident{Movie}. This is similar to
+type checking of lambda functions in C\# \cite{csharp}, where type is also inferred from the context (or has
+to be specified explicitly).
+
+We do not currently allow lambda functions as stand-alone let-bound
+values. This could be done by requiring explicit types, or introducing polymorphism, but it was not
+necessary for the limited domain of non-expert data exploration.
+
+\paragraph{Dependency graph for functions.}
+In the binding process specified in Section~\ref{sec:formal}, a variable is a leaf of the dependency
+graph. In the revised model, it depends on the context in which it appears. A new edge labeled
+$\blbl{callsite}(m, i)$ indicates that the source node is the input variable of a function
+passed as the $i^\textnormal{th}$ argument to the $m$ member of the expression represented by the
+target node. A node representing function is linked to the call site using the same edge.
+
+Figure~\ref{fig:graph-func} shows the result of binding $o.m(\lambda x\rightarrow x)$. Both
+$\bnd{fun}(x, s_2)$ and $\bnd{var}(x, s_1)$ now depend on the node representing $o$. The new
+$\blbl{callsite}$ edge makes it possible to type-check function and variable nodes just using their
+dependencies. As before, the member invocation $\bnd{mem}(m, s_0)$
+depends on the instance using $\blbl{arg}(0)$ and on its argument using $\blbl{arg}(1)$.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure*}
+\vspace{0.5em}
+\begin{equation*}
+\begin{array}{l}
+\qquad\inference[\rname{val}~]
+  {\Sigma(n) = \alpha}
+  {\bnd{val}(n) \vdash \alpha }
+\quad\qquad
+\inference[\rname{var}~]
+  { (\bnd{var}(x,s), \bndclr{v}, \blbl{callsite}(m, i)) \in E &
+    \bndclr{v} \vdash (.., m:(\tau_1, \ldots, \tau_k)\rightarrow \tau,..) & \tau_i=\tau'\rightarrow\tau'' }
+  { \bnd{var}(x, s) \vdash \tau' }
+\\[2.5em]
+\inference[\rname{fun}~]
+  { \{(\bnd{fun}(x,s), \bndclr{v_b}, \blbl{body}), (\bnd{var}(x,s), \bndclr{v_c}, \blbl{callsite}(m, i))\} \subseteq E &
+    \bndclr{v_c} \vdash (.., m:(\tau_1, \ldots, \tau_k)\rightarrow \tau,..) &
+    \tau_i=\tau'\rightarrow\tau'' & \bndclr{v_b} \vdash \tau'' }
+  { \bnd{fun}(x, s) \vdash \tau' \rightarrow \tau'' }
+\\[2.5em]
+\qquad\qquad\quad\inference[\rname{mem}~]
+  { \forall i\in\{0\ldots k\}.(\bnd{mem}(m, s), \bndclr{v_i}, \blbl{arg}(i)) \in E &
+    \bndclr{v_0} \vdash (.., m:(\tau_1, \ldots, \tau_k) \rightarrow \tau, ..)& \bndclr{v_i} \vdash \tau_i }
+  { \bnd{mem}(m, s) \vdash \tau }
+\\[1em]
+\end{array}
+\end{equation*}
+\caption{Rules that define evaluation of previews over a dependency graph for an expression}
+\vspace{0.5em}
+\label{fig:tc}
+\end{figure*}
+
+% --------------------------------------------------------------------------------------------------
+
+\paragraph{Revised binding process.}
+For the revised binding process, we introduce a new edge label $\blbl{callsite}$. Variable nodes
+now have dependencies and so we cache them and attach a symbol $s$ to \bnd{var}, so we also
+introduce a node kind $\bknd{var}$ as part of a lookup key for $\Delta$.
+
+The \ident{bind-expr} function now has a parameter $c$, in addition to $\Gamma$ and $\Delta$, which
+represents the context in which the binding happens. This is either a member invocation (labeled
+with instance node $\bndclr{v}$ and \blbl{callsite} label $\blblclr{l}$), or not a member
+invocation written as $\bot$. The updated definitions are:
+%
+\begin{equation*}
+\begin{array}{lcll}
+\bndclr{v} &\hspace{-0.5em}\in\hspace{-0.5em}&\bnd{val}(n)~|~\bnd{var}(x, s)~|~\bnd{mem}(m, s)~|~\bnd{fun}(x, s)&(\textit{Vertices})\\
+\blblclr{l}&\hspace{-0.5em}\in\hspace{-0.5em}&\blbl{body}~|~\blbl{arg}(i)~|~\blbl{callsite}(m,i)&(\textit{Edge labels})\\
+\bkndclr{k}&\hspace{-0.5em}\in\hspace{-0.5em}&\bknd{fun}(x)~|~\bknd{mem}(m)~|~\bknd{var}(x)&(\textit{Node kinds})\\
+          c&\hspace{-0.5em}\in\hspace{-0.5em}&\bot ~|~(\bndclr{v}, \blblclr{l})&(\textit{Call sites})\\
+\end{array}
+\end{equation*}
+%
+The key parts of the revised definition of the \ident{bind-expr} function are shown in Figure~\ref{fig:binding-rules-callsite}.
+We now write $\ident{bind-expr}_{\Gamma,\Delta,c}$ where $c$ represents the context in which the
+binding occurs. This is set to $\bot$ in all cases, except when binding arguments of a member call.
+In (2b), we first recursively bind the instance node using $\bot$ as the context and then
+bind all the arguments using $(\bndclr{v_0}, \blbl{callsite}(m, i))$ as the context for $i^{\textnormal{th}}$
+argument. The rest is as in case (2) before. The case (1) is updated similarly and is not shown
+here for brevity.
+
+When binding a function (7b), we now also store variable nodes in $\Delta$ and so we check if a
+variable with the given call site exists. If no, we create a fresh node $\bnd{var}(x, s_x)$.
+The node is added to $\Gamma$ as before. At the end, we now also include a call site edge from the
+variable node and from the function node in $E$. We omit a similar variant of the case (6).
+The remaining cases (3)-(5) are the same, except that \ident{bind-expr} has the additional $c$ parameter
+and recursive calls always set it to $\bot$.
+
+Finally, the \ident{update} function in Figure~\ref{fig:loop} also needs to be updated to store the
+newly created \bnd{var} nodes. This is done by adding the following single case:
+
+\begin{equation*}
+\begin{array}{l}
+\Delta_{i}(\bknd{var}(x), [(\bndclr{v},\blbl{callsite}(m,i))]) = \bnd{var}(x, s)\\
+\quad\quad \textnormal{for all}~\bnd{var}(x, s) \in V\\
+\quad\quad \textnormal{such that}~(\bnd{var}(x, s), \bndclr{v}, \blbl{callsite}(m,i)) \in E
+\end{array}
+\end{equation*}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Type checking over dependency graphs}
+\label{sec:types-graph}
+
+The type system for The Gamma supports a number of primitive types (such as integers and strings)
+written as $\alpha$. Composed types include functions and objects with members. Objects are be
+provided by type providers, but we omit the details here. Types of object members are written as
+$\sigma$ and can have multiple arguments:
+%
+\begin{equation*}
+\begin{array}{lcll}
+\tau&\in&\alpha ~|~ \tau \rightarrow \tau ~|~ \{m_1\!:\!\sigma_1, \ldots, m_n\!:\!\sigma_n\} &\quad(\textit{Types})\\
+\sigma &\in& (\tau_1, \ldots, \tau_n) \rightarrow \tau &\quad(\textit{Members})\\
+\end{array}
+\end{equation*}
+%
+The typing judgements are written in the form $\bnd{v}\vdash \tau$. They are parameterised by the
+dependency graph $(V, E)$, but this is not modified during type checking so we keep it implicit
+rather than writing e.g.~$\bnd{v}\vdash_{(V,E)} \tau$.
+
+\paragraph{Type checking.}
+The typing rules are shown in Figure~\ref{fig:tc}. Types of primitive values $n$ are obtained
+using a global lookup table $\Sigma$ \rname{val}. When type checking a member call \rname{mem},
+we find its dependencies $\bndclr{v_i}$ and check that the first one (instance) is an object with
+the required member $m$. The types of input arguments of the member then need to match the types
+of the remaining (non-instance) nodes.
+
+Type checking a function \rname{fun} and a variable \rname{var} is similar. In both cases,
+we follow the \blbl{callsite} edge to find the member that accepts the function as an argument.
+We obtain the type of the function from the type of the $i^\textnormal{th}$ argument of the
+member. We use the input type as the type of variable \rname{var}. For functions, we also check
+that the resulting type matches the type of the body \rname{fun}.
+
+\paragraph{Caching results.}
+Performing type checking over the dependency graph, rather than over the abstract syntax tree,
+enables us to reuse the results of previously type checked parts of a program. As when
+caching evaluated previews (Section~\ref{sec:previews}), we build a lookup table mapping graph
+nodes to types. When type checking a node, we first check the cache and, only if it is new,
+follow the $\vdash$ relation to obtain the type.
+
+As a result, code can be type checked on-the-fly during editing, even when asynchronous type
+providers are used, and the programmer gets instant feedback without delays.
+
 % --------------------------------------------------------------------------------------------------
@@
 
+\section{TODO}
+
+TBD: In section about dependency graphs, include a subsection discussing how this relates
+to incremental computation (just like "Semantics of delayed previews" links that to CMTT etc.)
+
 \section{Summary}
@@
 
-\setcounter{thc}{-1}
-\begin{theorem}[Let-free correctness]
-Given an expression $e$ that has no free variables and does not contain let bindings, together
-with a lookup table $\Delta$ obtained from any sequence of expressions according to Figure~\ref{fig:loop}
-let $\bndclr{v}, (V, E) = \ident{bind-expr}_{\emptyset, \Delta}(e)$. If $\bndclr{v}\Downarrow d$
-over a graph $(V, E)$ then $d = p$ for some $p$ and $e \rightsquigarrow_\beta p$.
+\begin{theorem}[Term preview correctness]
+Given a term $t$ that has no free variables, together with a lookup table $\Delta$ obtained
+from any sequence of programs using $\ident{bind-prog}$ (Figure~\ref{fig:binding-rules-prog}) and
+$\ident{update}$ (Figure~\ref{fig:loop}), then
+let $\bndclr{v}, (V, E) = \ident{bind-expr}_{\emptyset, \Delta}(t)$. If $\bndclr{v}\Downarrow p$
+over a graph $(V, E)$ then $p = o$ for some value $o$ and $t \rightsquigarrow^{*} o$.
 \end{theorem}
+
 \begin{proof}
-First note that, when combining recursively constructed sub-graphs, the \ident{bind-expr} operation
-adds new nodes and edges leading from those new nodes. Therefore, an evaluation using $\Downarrow$
-over a sub-graph will also be valid over the new graph.
-
-Next, we prove a more general property using induction showing that for $e$ such
-that $\bndclr{v}, (V, E) = \ident{bind-expr}_{\emptyset, \Delta}(e)$:
+First note that, when combining recursively constructed sub-graphs, the \ident{bind-expr} function
+adds new nodes and edges leading from those new nodes. Therefore, an evaluation using~$\Downarrow$
+over a sub-graph will also be valid over the new graph -- the newly added nodes and edges do not introduce
+non-determinism to the rules given in Figure~\ref{fig:eval}.
+
+We prove a more general property showing that for any $e$, its binding
+$\bndclr{v}, (V, E) = \ident{bind-expr}_{\emptyset, \Delta}(e)$ and any evaluation context $C$
+such that $C[e]\rightsquigarrow o$ for some $o$, one of the following holds:
+%
 \begin{enumerate}
-\item[a.] If $FV(e)\,=\,\emptyset$ then $\bndclr{v} \Downarrow p$ for some $p$ and $e \rightsquigarrow_\beta p$
-\item[b.] If $FV(e)\neq\emptyset$ then $\bndclr{v} \Downarrow \llbracket e_p \rrbracket_{FV(e)}$ for some $e_p$ and
-  for any evaluation context $C[-]$ such that $FV(C[e_p])=\emptyset$ it holds that if
-  $C[e] \rightsquigarrow_\beta C[e_p]$.
+\item[a.] If $FV(e)\,=\,\emptyset$ then $\bndclr{v} \Downarrow p$ for some $p$ and $C[p] \rightsquigarrow o$
+\item[b.] If $FV(e)\neq\emptyset$ then $\bndclr{v} \Downarrow \llbracket e_p \rrbracket_{FV(e)}$ for some $e_p$ and $C[e_p] \rightsquigarrow o$
 \end{enumerate}
-
-The proof is done by induction over the binding process, which follows the structure
-of the expression $e$:
+%
+In the first case, $p$ is a value, but it is not always the case that $e \rightsquigarrow^{*} p$,
+because $p$ may be lambda function and preview evaluation may reduce sub-expression in the body of
+the function. Using a context $C$ in which the value reduces to an object avoids this problem.
+
+The proof of the theorem follows from the more general property. Using a context $C[-]=-$, the
+term $t$ reduces $t \rightsquigarrow^{*}t' \rightsquigarrow_\epsilon o$ for some $o$ and the
+preview $p$ is a value $o$ because $C[p] = p = o$.
+The proof is by induction over the binding process, which follows the structure of the expression:
 
@@
   expressions $e_i$ and $\{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v}, \bndclr{v_n}, \blbl{arg}(n))\} \subseteq E$.
-  From Lemma~\ref{thm:lemma-lookup}, $\bndclr{v} = \bnd{mem}(m, s)$ for some $s$.
+  From lookup inversion Lemma~\ref{thm:lemma-lookup}, $\bndclr{v} = \bnd{mem}(m, s)$ for some $s$.
 
   If $FV(e)=\emptyset$, then $\bndclr{v_i} \Downarrow p_i$ for $i\in 0\ldots n$ and
-  $\bndclr{v}\Downarrow p$ using (\emph{mem-val}) such that $p_0.m(p_1, \ldots, p_n) \rightsquigarrow p$.
-  From induction hypothesis, $e_i \rightsquigarrow_\beta p_i$ and so, using compositionality of $\rightsquigarrow$,
-  $e_0.m(e_1, \ldots, e_n) \rightsquigarrow_\beta p$.
+  $\bndclr{v}\Downarrow p$ using \rname{mem-val} such that $p_0.m(p_1, \ldots, p_n) \rightsquigarrow p$.
+  From induction hypothesis and \emph{compositionality} of external libraries (Definition~\ref{def:external}),
+  it holds that for any $C$ such that $C[e_0.m(e_1, \ldots, e_n)] \rightsquigarrow o$ for some $o$
+  then also $C[p_0.m(p_1, \ldots, p_n)] \rightsquigarrow C[p] \rightsquigarrow o$.
 
   If $FV(e)\neq\emptyset$, then $\bnclr{v_i} \Downarrow_\ident{lift} \llbracket e'_i \rrbracket$ for $i\in 0\ldots n$ and
-  using (\emph{mem-expr}), $\bndclr{v}\Downarrow \llbracket e'_0.m(e'_1, \ldots, e'_n) \rrbracket_{FV(e)}$.
-  From induction hypothesis, for any $C[-]$, it holds that $C[e_i] \rightsquigarrow_\beta C[e'_i]$.
-  Using compositionality, it also holds that for any $C[-]$, it is the case that
-  $C[e_0.m(e_1, \ldots, e_n)] \rightsquigarrow_\beta C[e'_0.m(e'_1, \ldots, e'_n)]$.
-
-\newpage
+  $\bndclr{v}\Downarrow \llbracket e'_0.m(e'_1, \ldots, e'_n) \rrbracket_{FV(e)}$ using \rname{mem-expr}.
+  From induction hypothesis and \emph{compositionality} of external libraries (Definition~\ref{def:external}),
+  it holds that for any $C$ such that $C[e_0.m(e_1, \ldots, e_n)] \rightsquigarrow o$ for some $o$
+  then also $C[e'_0.m(e'_1, \ldots, e'_n)] \rightsquigarrow o$.
+
 \vspace{0.75em}\noindent(2) $\ident{bind-expr}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n))$ --
@@
 
-\vspace{0.75em}\noindent(3) We assume that the expression $e$ does not include let bindings and
-  so this case never happens.
-
-\vspace{0.75em}\noindent(4) $\ident{bind-expr}_{\Gamma, \Delta}(n)$ -- In this case $e=n$ and $\bndclr{v} = \bnd{val}(n)$. The preview of
-  $\bnd{val}(n)$ is evaluated to $n$ using the (\emph{val}) case.
-
-\vspace{0.75em}\noindent(5) $\ident{bind-expr}_{\Gamma, \Delta}(x)$ -- The initial $\Gamma$ is empty and
-  there are no let bindings, so $x$ must have been added to $\Gamma$ by case (6) or (7). Hence,
-  $\bndclr{v}=\bnd{var}(x)$. Using (\emph{var}) $\bndclr{v}\Downarrow \llbracket x \rrbracket_{x}$
-  and so $e_p = e = x$ and the second case (b.) trivially holds.
-
-\vspace{0.75em}\noindent(6) $\ident{bind-expr}_{\Gamma, \Delta}(\lambda x\rightarrow e)$ -- Assume
-  $\bndclr{v_b}$ is a graph node representing the body. The evaluation can use one of three rules:
-
-  If $FV(e)=\emptyset$ then $\bndclr{v_b}\Downarrow p_b$ for some $p_b$ and $\bndclr{v}\Downarrow \lambda x.p_b$
-  using (\emph{fun-val}). From induction $e_b \rightsquigarrow_\beta p_b$
-  and so by definition also $\lambda x.e_b \rightsquigarrow_\beta \lambda x.p_b$.
-
-  If $FV(e)=\{x\}$ then $\bndclr{v_b}\Downarrow \llbracket e_b \rrbracket_x$ for some $e_b$ and
-  $\bndclr{v}\Downarrow \lambda x.e_b$ using (\emph{fun-bind}). From
-  induction, for any context $C[-]$, it holds that $C[e] \rightsquigarrow_\beta C[e_b]$.
-  Using a context $C[-]=\lambda x.-$ it holds that $\lambda x.e \rightsquigarrow_\beta \lambda x.e_b$.
-
-  Otherwise, $\bndclr{v_b}\Downarrow \llbracket e_b \rrbracket_{x,\Gamma}$ for some $e_b$ and
-  $\bndclr{v}\Downarrow \llbracket\lambda x.e_b\rrbracket_\Gamma$ using (\emph{fun-expr}). From
-  induction, for any context $C[-]$, it holds that $C[e] \rightsquigarrow_\beta C[e_b]$
-  and for any context $C'[-]$, by definition of $\rightsquigarrow_\beta$ also
-  $C'[\lambda x.e] \rightsquigarrow C'[\lambda x.e_b]$.
+\vspace{0.75em}\noindent(3) $\ident{bind-expr}_{\Gamma, \Delta}(\lambda x\rightarrow e_b)$ --
+  Here $e = \lambda x\rightarrow e_b$, $\bndclr{v_b}$ is the graph node obtained by induction
+  for the expression $e_b$ and $(\bndclr{v}, \bndclr{v_b}, \blbl{body}) \in E$.
+  From lookup inversion Lemma~\ref{thm:lemma-lookup}, $\bndclr{v} = \bnd{fun}(x, s)$ for some $s$.
+  The evaluation can use one of three rules:
+
+  \begin{enumerate}
+  \item[i.]
+  If $FV(e)=\emptyset$ then $\bndclr{v_b}\Downarrow p_b$ for some $p_b$ and $\bndclr{v}\Downarrow \lambda x\rightarrow p_b$
+  using \rname{fun-val}. Let $e'_b=p_b$.
+
+  \item[ii.]
+  If $FV(e_b)=\{x\}$ then $\bndclr{v_b}\Downarrow \llbracket e'_b \rrbracket_x$ for some $e'_b$ and
+  $\bndclr{v}\Downarrow \lambda x\rightarrow e'_b$ using \rname{fun-bind}.
+
+  \item[iii.]
+  Otherwise, $\bndclr{v_b}\Downarrow \llbracket e'_b \rrbracket_{x,\Gamma}$ for some $e'_b$ and
+  $\bndclr{v}\Downarrow \llbracket\lambda x\rightarrow e'_b\rrbracket_\Gamma$ using \rname{fun-expr}.
+  \end{enumerate}
+
+\noindent
+For i.) and ii.) we show that a.) is the case; for iii.) we show that b.) is the case; that is
+for any $C$, if $C[\lambda x\rightarrow e_b] \rightsquigarrow o$ then also
+$C[\lambda x\rightarrow e'_b] \rightsquigarrow o$. For a given $C$, let $C'[-] = C[\lambda x\rightarrow -]$
+and use the induction hypothesis, i.e.~if $C'[e_b] \rightsquigarrow o$ for some $o$ then also $C'[e'_b] \rightsquigarrow o$.
+
+\vspace{0.75em}\noindent(4) $\ident{bind-expr}_{\Gamma, \Delta}(\lambda x\rightarrow e)$  --
+  This case is similar to (3), except that the fact that $\bndclr{v} = \bnd{fun}(x, s)$
+  holds by construction, rather than using Lemma~\ref{thm:lemma-lookup}.
+
+\vspace{0.75em}\noindent(5) $\ident{bind-expr}_{\Gamma, \Delta}(o)$ -- In this case $e=o$ and $\bndclr{v} = \bnd{val}(o)$
+  and $\bnd{val}(o)\Downarrow o$ using \rname{val} and so the case a.) trivially holds.
+
+\vspace{0.75em}\noindent(6) $\ident{bind-expr}_{\Gamma, \Delta}(x)$ -- The initial $\Gamma$ is empty,
+  so $x$ must have been added to $\Gamma$ by case (3) or (4). Hence,
+  $\bndclr{v}=\bnd{var}(x)$, $\bndclr{v}\Downarrow \llbracket x \rrbracket_{x}$ using \rname{var}
+  and so $e_p = e = x$ and the case b.) trivially holds.
+
 \end{proof}

### 2019-01-28 wip-live-programming/paper.tex
Properties of the data exploration calculus

@@
 
-\newpage
 \begin{figure}
 \raggedright
-{\small\sffamily Programs, commands, terms, expressions and values}
-\begin{equation*}
-\hspace{-2em}
+\hspace{1.5em}{\small\sffamily Programs, commands, terms, expressions and values}
+\begin{equation*}
 \begin{array}{lcl}
@@
 \vspace{0.5em}
-{\small\sffamily Evaluation contexts of expressions}
-\begin{equation*}
-\hspace{-2em}
+\hspace{1.5em}{\small\sffamily Evaluation contexts of expressions}
+\begin{equation*}
 \begin{array}{lcl}
@@
 \vspace{0.5em}
-{\small\sffamily Let elimination and member reduction}
-\begin{equation*}
-\hspace{-2em}
+\hspace{1.5em}{\small\sffamily Let elimination and member reduction}
+\begin{equation*}
 \begin{array}{ll}
@@
 
+\newpage
+
 \section{Data exploration calculus}
@@
 The calculus is not Turing-complete and it can only be used together with external libraries that
-define what objects are available and what the behaviour of their members is. However, this is
+define what objects are available and what the behaviour of their members is. This is
 sufficient to capture the simple data analyses discussed in Section~\ref{sec:background}. We
@@
 
+We initially present the data exploration calculus without a static type system. Our main focus
+is on the live preview mechanism which can be done equally well without types. We consider
+type checking in Section~\ref{sec:types}, because the same mechanism that is used for live previews
+can also be used for incremental type-checking.
+
 \subsection{Language syntax}
 The calculus combines object-oriented features such as member access with functional features
-such as lambda functions. The syntax is defined in Figure~\ref{fig:dec-calculus}. Object values
+including lambda functions. The syntax is defined in Figure~\ref{fig:dec-calculus}. Object values
 $o$ are defined by external libraries that are used in conjunction with the core calculus.
@@
 A program $p$ in the data exploration calculus consists of a sequence of commands $c$. A command
-can be either let binding or a term. Let bindings define variables $x$ that can be used in
+can be either a let binding or a term. Let bindings define variables $x$ that can be used in
 subsequent commands. The calculus supports lambda functions, but they
@@
 member access, while an expression $e$, which can appear as an argument in member access,
-can be a lambda function or a term.\footnote{Similar but weaker restrictions on the use of lambda functions
-exist in other languages. To guide type inference, lambda functions in C\# can appear as either
-method arguments or assigned to an explicitly typed variable.}
+can be a lambda function or a term.\footnote{Similar but weaker restrictions on the use of lambda
+functions exist in other languages. For example, in order to guide type inference, lambda functions
+in C\# can appear as either method arguments or assigned to an explicitly typed variable, but they
+cannot be assigned to an implicitly typed variable.}
 
@@
 reduction $\rightsquigarrow$. Fully evaluating a program results in an irreducible sequence of
-objects $o_1;\; \ldots;\; o_n$ which can be displayed as intermediate results of the data analysis. The
-operational semantics is parameterized by a relation $\rightsquigarrow_\epsilon$ that models
-the functionality of the external libraries used with the calculus and defines reduction rules
-for member accesses. The relation has the following form:
+objects $o_1;\; \ldots;\; o_n$ (one object for each command, including let bindings) which can be
+displayed as intermediate results of the data analysis. The operational semantics is parameterized
+by a relation $\rightsquigarrow_\epsilon$ that models the functionality of the external libraries
+used with the calculus and defines reduction rules for member accesses. The relation has the
+following form:
 %
@@
 %
+The operation is invoked on an object and takes values (objects or function values) as arguments.
+It always results in an opaque object.
 Figure~\ref{fig:dec-calculus} defines the reduction rules in terms of $\rightsquigarrow_\epsilon$
@@
 \subsection{Properties}
-
-DEFINE $\rightsquigarrow^{*}$ as a reflexive, transitive closure of $\rightsquigarrow$
-
-DEFINE $\rightsquigarrow_\ident{let}$ (on a program) - CBN subsitution
+The data exploration calculus is a very simple language and it has a number of desirable
+properties. However, some of those require that the relation $\rightsquigarrow_\epsilon$, which
+defines evaluation for external libraries, satisfies a number of conditions. Our main motivation
+is to capture properties that allow us to prove that our method of evaluating live previews,
+discussed in Section~\ref{sec:properties}, is correct.
+
+\begin{definition}[Further reductions]
+\label{def:further-red}
+We define two additional reduction relations:
+\begin{itemize}
+\item[-] We write $\rightsquigarrow^{*}$ for the reflexive, transitive closure of $\rightsquigarrow$
+\item[-] We write $\rightsquigarrow_\ident{let}$ for a call-by-name let binding elimination\\
+  $c_1;\; \ldots;\; c_{k-1};\; \kvd{let}~x=t;\; c_{k+1};\; \ldots; c_n ~\rightsquigarrow_\ident{let}~
+  c_1;\; \ldots;\; c_{k-1};\; t; c_{k+1}[x\leftarrow t];\; \ldots; c_n[x\leftarrow t]$
+\end{itemize}
+\end{definition}
+
+\noindent
+We say that two expressions
+$e$ and $e'$ are observationally equivalent if, for any context $C$, the expressions $C[e]$ and
+$C[e']$ reduce to the same value. Lambda functions $\lambda x\!\rightarrow\!2$ and
+$\lambda x\!\rightarrow\!1\!+\!1$ are not equal, but they are observationally equivalent.
+We require that external libraries satisfy two conditions. First, when a method is called with
+observationally equivalent values as arguments, it should return the same value. Second, the
+evaluation of $o.m(v_1,\ldots,v_n)$ should be defined for all $o, i$ and $v_i$. The external
+library will typically satisfy this by defining an error object and reducing all invalid calls
+to the error object, following the standard method of \citet{gowrong}.
 
@@
 \label{def:external}
-
-Satsifies \emph{compositionality} such that:
-
-Given $e_0, e_1, \ldots, e_n$ and $e'_0, e'_1, \ldots, e'_n$ and $m$ such that
-$e_0.m(e_1, \ldots, e_n) \rightsquigarrow^{*} o$ and
-$e'_0.m(e'_1, \ldots, e'_n) \rightsquigarrow^{*} o'$ then
-if for any contexts
-$C_0, C_1, \ldots, C_n$ it holds that if $C_i[e_i] \rightsquigarrow^{*} o_i$ and $C_i[e'_i] \rightsquigarrow^{*} o_i$ for some $o_i$
-then $o = o'$.
+An external library consists of a set of objects $O$ and a reduction relation
+$\rightsquigarrow_\epsilon$ that satsifies the following two properties:
+\begin{itemize}
+\item[-] \emph{Completeness.} For all $o, m, i$ and all $v_1, \ldots, v_i$, there exists $o'$ such
+  that $o.m(v_1, \ldots, v_i) \rightsquigarrow_\epsilon o'$.
+\vspace{0.5em}
+\item[-] \emph{Compositionality.} For observationally equivalent arguments, the reduction returns
+  the same object, i.e.~given $e_0, e_1, \ldots, e_n$ and $e'_0, e'_1, \ldots, e'_n$ and $m$ such that
+  $e_0.m(e_1, \ldots, e_n) \rightsquigarrow^{*} o$ and $e'_0.m(e'_1, \ldots, e'_n) \rightsquigarrow^{*} o'$ then
+  if for any contexts $C_0, C_1, \ldots, C_n$ it holds that if $C_i[e_i] \rightsquigarrow^{*} o_i$ and
+  $C_i[e'_i] \rightsquigarrow^{*} o_i$ for some $o_i$ then $o = o'$.
 \end{definition}
 
-TBD: LET above should leave the evaluated object in the list of commmands
-
-LET ELIMINATION DOES NOT CHANGE RESULTS
-
-Let elimination $\rightsquigarrow_\ident{let}$ such that, using  capture-avoiding substitution,
-$C[\kvd{let}~x=e_1~\kvd{in}~e_2] \rightsquigarrow_\ident{let} C[e_2[x\leftarrow e_1]]$
-
-Let elimination does not affect the result, i.e.~if $e \rightsquigarrow_{\ident{let}} e'$
-and $e'\rightsquigarrow_\beta e''$ then also $e\rightsquigarrow_\beta e''$
+\noindent
+The compositionality condition is essential for proving the correctness of our live preview mechanism.
+Completeness allows us to prove normalization, i.e.~all programs reduce to
+a value -- although the resulting value may be an error value provided by the external library.
+
+\begin{theorem}[Normalization]
+\label{thm:normalization}
+For all $p$, there exists $n$ and $o_1, \ldots, o_n$ such that $p\rightsquigarrow^{*} o_1;\ldots;o_n$.
+\end{lem}
+\begin{proof}
+A program that is not a sequence of values can be reduced and reduction decreases the size of the
+program. See Appendix~\ref{sec:app-normalization} for more detail.
+\end{proof}
+
+\noindent
+Although the reduction rules \rname{let} and \rname{external} of the data exploration calculus
+define an evaluation in a call-by-value order, eliminating let bindings in a call-by-name way
+using the $\rightsquigarrow_\ident{let}$ reduction does not affect the result. This
+simplifies our later proof of live preview correctness in Section~\ref{sec:properties-correct}.
 
@@
 \label{thm:let-lang-elimination}
-does not matter
+Given any program $p$ such that $p \rightsquigarrow^{*} o_1;\ldots;o_n$ for some $n$ and $o_1, \ldots, o_n$
+then if $p \rightsquigarrow_\ident{let} p'$ for some $p'$ then also $p' \rightsquigarrow^{*} o_1;\ldots;o_n$.
 \end{lem}
+\begin{proof}
+The elimination of let binding transforms a program $c_1;\; \ldots;\; c_{k-1};\; \kvd{let}~x=t;\; c_{k+1};\; \ldots; c_n$
+to a program $c_1;\; \ldots;\; c_{k-1};\; t; c_{k+1}[x\leftarrow t];\; \ldots; c_n[x\leftarrow t]$.
+The reduction steps for the new program can be constructed using the steps of $p \rightsquigarrow^{*} o_1;\ldots;o_n$.
+The new command $t$ reduces to an object $o$ using the same steps as the original term $t$
+in $\kvd{let}~x=t$ but with context $C_c = -$ rather than $C_c = \kvd{let}~x=-$; the terms $t$
+introduced by substitution also reduce using the same steps as before, but using
+contexts in which the variable $x$ originally appeared.
+\end{proof}
 
@@
 \section{Appendix}
+
+\subsection{Normalization}
+\label{sec:app-normalization}
+
+\begin{theorem}[Normalization]
+For all $p$, there exists $n$ and $o_1, \ldots, o_n$ such that $p\rightsquigarrow^{*} o_1;\ldots;o_n$.
+\end{lem}
+\begin{proof}
+We define $\ident{size}$ of a program in data exploration calculus as follows:
+\begin{equation}
+\begin{array}{rcl}
+\ident{size}(c_1; \ldots; c_n) &=& 1 + \Sigma^{n}_{i=1} \ident{size}(c_i)\\
+\ident{size}(\kvd{let}~x=t) &=& 1 + \ident{size}(t)\\
+\ident{size}(e_0.m(e_1, \ldots, e_n)) &=& 1 + \Sigma^{n}_{i=0} \ident{size}(e_i)\\
+\ident{size}(\lambda x\rightarrow e) &=& 1 + \ident{size}(e)\\
+\ident{size}(o) = \ident{size}(x) &=& 1\\
+\end{array}
+\end{equation}
+
+\noindent
+The property holds because, first, both \rname{let} and \rname{external} decrease the \ident{size}
+of the program and, second, a program is either fully evaluated, i.e.~$o_1;\ldots;o_n$ for some $n$
+or, it can be reduced using one of the reduction rules.
+\end{proof}
+
+\subsection{Term preview correctness}
 \label{sec:app-correctness}

### 2019-01-30 wip-live-programming/paper.tex
Experiments and revising introduction

@@
 
+\settopmatter{printfolios=true}
 %% Copyright information
@@
 
-\title{Live coding}
+\title{Live coding environment for a data exploration language}
 \author{Tomas Petricek}
@@
 \begin{abstract}
-Data science can be done by directly manipulating data using spreadsheets, or by writing
-data manipulation scripts using a programming language. The former is error-prone and
-does not scale, while the latter requires expert skills. Live coding has the potential to
-bridge this gap and make writing of transparent, reproducible scripts more accessible.
-
-In this paper, we describe a live programming environment for data science that provides instant
-previews and contextual hints, while allowing the user to edit code in an unrestricted way in
-a text editor.
-
-Supporting a text editor is challenging as any edit can significantly change the structure of code and
-fully recomputing previews after every change is too expensive. We present a technique that allows the type
-checker and the interpreter to run on the fly, while code is written, and reuse results of previous
-runs. This makes it possible to efficiently provide instant feedback and live previews during development.
-
-We formalise how programs are interpreted and how previews are computed, prove correctness
-of the previews and formally specify when can previews be reused. We believe this work provides
-solid and easy to reuse foundations for the emerging trend of live programming environments.
+
+A growing amount of code is written to explore and analyze data, but the way data analysts
+work with code differs from the way software engineers do. First, many data
+scientists write code that introduces few or no abstractions. Second, data exploration is an
+interactive process and it blurs the traditional distinction between development-time and run-time.
+This poses an interesting challenge for programming language research, because it allows us to
+revisit a number of traditional assumptions about programming languages and focus our attention
+on a new kind of programming tools that data analysts need.
+
+In this paper, we present \emph{data exploration calculus}, a formally tractable
+language that captures the structure of simple data analyses as done, for example, by journalists
+exploring open government data. We implement a programming environment for data exploration that
+evaluates code instantaneously during editing and shows previews of the results,
+while allowing the user to modify code in an unrestricted way in a text
+editor. Supporting a text editor is tricky as any edit can change the structure
+of code and fully recomputing program after every change would be too expensive.
+We present a new technique that allows evaluating and type-checking code while it is
+being written while reusing results of previous runs. We formalize the technique using the data
+exploration calculus, prove that it is correct and specify when previous results are reused.
+Finally, we also illustrate the practicality of our approach using empirical evaluation
+and a case study.
+
+As data analysis becomes ever more important use of programming, research on programming languages
+and tools needs to consider new kinds of code that people need to write and new kinds of tools
+that can support them. The present paper is one step in this important direction.
 \end{abstract}
@@
 % ==================================================================================================
-% TODO: Change values from 'n' to 'o' to indicate they are objects and can have members
-% ==================================================================================================
 
@@
 \label{sec:intro}
-
-One of the aspects that make spreadsheet tools such as Excel more accessible than programming
-environments is their liveness. When you change a value in a cell in Excel \cite{spreadsheet}, the
-whole spreadsheet is updated instantly and you see the new results immediately.
-
-Increasing number of programming environments aim to provide the same live experience for more
-standard programming languages, but doing this is not easy. Fully recomputing the whole program after
-every single change is inefficient and calculating how a change in source code changes the result
-is extremely hard when the editor allows arbitrary manipulation of program text. For example,
-consider the following simple program that gets the release years of 10 most expensive movies
-in a data set \ident{movies}:
+One of the aspects that make spreadsheets easier to use than programming tools is their
+liveness. When you change a value in a cell in Excel, the whole spreadsheet updates instantly
+and you immediately see new results, without having to explicitly trigger re-computation
+and without having to wait for an extensive period of time.
+
+Increasing number of programming environments aim to provide the same live development experience
+for standard programming languages, but doing this is not easy. Fully recomputing the whole program
+after every keystroke is inefficient and calculating how a change in the source code changes the
+result is extremely hard when the text editor allows arbitrary changes.
+Consider the following snippet that gets the release years of 10 most expensive movies from a
+data set \ident{movies}:
 %
@@
 \begin{array}{l}
-\kvd{let}~\ident{top} =\ident{movies}\\
-\quad .\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\ident{.take}(\num{10})\\
-\quad .\ident{map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{yyyy}))\\
-\end{array}
-\end{equation*}
-%
-A live coding environment shows a preview of the list of dates. Next, assume that the programmer
-modifies the code by making the constant $\num{10}$ a variable and changing the date format to see
-the full date:
+\kvd{let}~\ident{top} =\ident{movies}.\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\\
+\qquad .\ident{take}(\num{10}).\ident{map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{yyyy}))\\
+\end{array}
+\end{equation*}
+%
+A live coding environment computes and shows the list of years, but then the programmer modifies the code by
+making the constant $\num{10}$ a variable and changing the date format to see the full date:
 %
@@
 \kvd{let}~\ident{count} = \num{10}\\
-\kvd{let}~\ident{top} = \ident{movies}\\
-\quad .\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\ident{.take}(\ident{count})\\
-\quad \ident{.map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{dd-mm-yyyy}))\\
+\kvd{let}~\ident{top} = \ident{movies}.\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\\
+\qquad .\ident{take}(\ident{count}).\ident{map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{dd-mm-yyyy}))\\
 \end{array}
@@
 Ideally, the live coding environment should understand the change, reuse a cached result of the
-first two transformations (sorting and taking 10 elements) and only evaluate the last \ident{map}
-to differently format the release dates of already computed top 10 movies.
-
-This is not difficult if we represent the program in a structured way \cite{structure-based,hazelnut}
-and allow the user to edit code via known operations such as ``extract variable'' (which has
-no effect on the result) or ``change constant value'' (which forces recomputation of subsequent
-transformations). However, many programmers prefer to edit programs as free form text.
-
-We present the design and implementation of a live coding system that is capable
-of reusing previously evaluated expressions as in the example above, yet, is integrated into an
-ordinary text editor. Our main contributions are:
+first two transformations (sorting and taking first 10 elements) and only evaluate the
+\ident{map} operation to differently format the release dates of already computed top 10 movies.
+We present such a live coding environment, describe it formally and empirically evaluate it.
+We discuss related work in Section~\ref{sec:future}, but we review the most important directions
+next, in order to situate our contributions.
+
+\subsection{Related work}
+We focus on simple data exploration as done, for example, by journalists \cite{ddj}
+and we aim to bring the experience of using transparent and reproducible programming tools
+closer to that of using spreadsheets. When a user edits a cell in a spreadsheet, the results
+automatically update. Recomputation in spreadsheets is well-studied,~e.g. by \citet{spreadsheet},
+but it is simpler than in our case, because editing is limited to individual cells.
+
+Many visual data exploration tools support interactivity \cite{control,tableau,vizdom} and some,
+can even export the workflow as a script \cite{wrangler}, but data anlysts who prefer working with
+code typically resort to notebook systems such as Jupyter or R Markdown \cite{jupyter,rmarkdown}.
+Those are text-based, but have a limited model of recomputation. Users have to structure code in
+cells and manually reevaluate cells after a change. Those systems also do not track
+dependencies between cells, which can lead to inconsistencies; an issue that has been addressed
+by e.g.~\citet{dataflow} or \citet{wrattler}.
+
+We aim to keep the transparency and reproducibility of notebooks, but add a correct and
+efficient instantaneous evaluation mechanism. Existing text-based editors that provide instant,
+such as Lighttable \cite{lighttable} and Chrome DevTools, often work only in certain
+situations or require full recomputation. A more principled approach has been used by
+systems based on structured editing \cite{livenut,lamdu}, which benefit from the fact that code
+can only be modified via known operations with known effect on the computation graph
+(``extract variable'' has no effect on the result; ``change constant value''
+forces recomputation of subsequent code). However, we aim to cater for the many users who
+prefer to edit programs as free-form text.
+
+The technical aspects of our technique for providing instantaneous feedback
+are related to the work on incremental computation \cite{selfadjusting,incremental}. This focuses
+on cases where \emph{data} changes but program stays the same, while our focus is on cases where
+\emph{program} changes but data stays. Yet, the technical aspects are related and,
+as noted in Section~\ref{sec:future}, language that supports incremental computation could be used
+to implement the kind of system presented in this paper.
+
+\subsection{Contributions}
+We present the design and implementation of a live coding environment for a data exploration
+language that provides a correct and efficient instantaneous feedback, yet, is integrated into an
+ordinary text editor. Our most important contributions are:
 
 \begin{itemize}[itemsep=3pt]
-\item We introduce The Gamma (Section~\ref{sec:live}), a simple live coding environment for
-  data science. We review its design and implementation and explain how it
-  bridges the gap between programming and spreadsheets.
-
-\item Implementing a live programming system requires different way of thinking about compilers
-  and interpreters than the one presented in classic programming language literature. Our
-  formalisation (Section~\ref{sec:formal}) captures the essence of the new perspective.
-
-\item We formalise the evaluation of previews (Section~\ref{sec:previews}) and prove that our
-  evaluation and caching mechanism is produces correct previews (Section~\ref{sec:properties-correct})
-  and can effectively reuse partial results (Section~\ref{sec:properties-reuse}).
-
-\item We follow the same method to implement a type checker (Section~\ref{sec:types}) for our
-  language that is capable of reusing previous results. This makes it possible to efficiently support
-  asynchronously provided types (Section~\ref{sec:types-providers}).
-
-\item In more speculative conclusions (Section~\ref{sec:design}), we consider alternative
-  language designs that would enable further live coding experiences, which are difficult to
-  build using our current system.
+\item We introduce \emph{data exploration calculus} (Section~\ref{sec:calculus}), a small formally
+  tractable programming language for data exploration. The language is motivated by our review
+  of how data analysts work (Section~\ref{sec:background}) and captures the idea that scripts are
+  often lists of commands with few or no abstractions with most logic provided by external libraries.
+
+\item Implementing a live coding environment requires a different way of thinking than the one
+  presented in classic compiler literature. We capture the essence of the new perspective
+  (Section~\ref{sec:formal}) and use it to build our \emph{live preview} mechanism
+  (Section~\ref{sec:previews}) evaluates code instantaneously during editing.
+
+\item We prove that our live preview mechanism is correct (Section~\ref{sec:correctness}) and that
+  it reuses previously evaluated values when possible (Section~\ref{sec:evaluation-reuse}). We
+  illustrate the practicality of the mechanism using an empirical evaluation
+  (Section~\ref{sec:evaluation-empirical}) and a case study (Section~\ref{sec:case}).
+
+\item We extend our core environment in two ways, First, we add instantaneous type checker
+  that relies on the same core infrastructure as live previews and enables richer user experience
+  through \emph{type providers} (Section~\ref{sec:types}). Second, we add an abstraction mechanism
+  that preserves the live coding experience (Section~\ref{sec:design}).
 \end{itemize}
-
-\noindent
-We hope the architecture and its formal presentation in this paper can contribute crucial
-foundations for the growing and important trend of text-based live coding environments.
-
-\newpage
 
@@
 code they write and the way they interact with the system is very different from how software
-engineers work. This paper focuses on simple data wrangling and data exploration as done, for
+engineers work \cite{workflow}. This paper focuses on simple data wrangling and data exploration as done, for
 example, by journalists analysing government datasets. This section illustrates how such data
@@
 \end{itemize}
+
+~
+
+TODO: Most logic in external libs
+
+~
 
@@
 is to capture properties that allow us to prove that our method of evaluating live previews,
-discussed in Section~\ref{sec:properties}, is correct.
+discussed in Section~\ref{sec:correctness}, is correct.
 
@@
   $C_i[e'_i] \rightsquigarrow^{*} o_i$ for some $o_i$ then $o = o'$.
+\end{itemize}
 \end{definition}
@@
 For all $p$, there exists $n$ and $o_1, \ldots, o_n$ such that $p\rightsquigarrow^{*} o_1;\ldots;o_n$.
-\end{lem}
+\end{theorem}
 \begin{proof}
@@
 using the $\rightsquigarrow_\ident{let}$ reduction does not affect the result. This
-simplifies our later proof of live preview correctness in Section~\ref{sec:properties-correct}.
+simplifies our later proof of live preview correctness in Section~\ref{sec:correctness-correct}.
 
@@
 given in Section~\ref{sec:calculus-semantics}. We analyse the preview evaluation formally in
-Section~\ref{sec:properties} and show that the resulting previews are the same as the result we
+Section~\ref{sec:correctness} and show that the resulting previews are the same as the result we
 would get by directly evaluating the code and, also, that no recomputation occurs when code is
@@
 
-\section{Properties of live coding environment}
-\label{sec:properties}
+\section{Correctness of previews}
+\label{sec:correctness}
 
@@
 previously evaluated partial results. In other words, we show that the mechanism is correct
-and implements a useful optimization.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Correctness of previews}
-\label{sec:properties-correct}
+and implements an effective optimization. We discuss correctness in this section and evaluate
+the effectiveness of the optimization, both theoretically and empirically, in Section~\ref{sec:optimization}.
 
@@
 \item[a.] If $FV(e)\,=\,\emptyset$ then $\bndclr{v} \Downarrow p$ for some $p$ and $C[p] \rightsquigarrow o$
-\item[b.] If $FV(e)\neq\emptyset$ then $\bndclr{v} \Downarrow \llbracket e_p \rrbracket_{FV(e)}$ for some $e_p$ and $C[e_p] \rightsquigarrow o$
+\item[b.] If $FV(e)\,\neq\,\emptyset$ then $\bndclr{v} \Downarrow \llbracket e_p \rrbracket_{FV(e)}$ for some $e_p$ and $C[e_p] \rightsquigarrow o$
 \end{enumerate}
@@
 
+\section{Evaluation}
+\label{sec:evaluation}
+
 \subsection{Reuse of previews}
-\label{sec:properties-reuse}
+\label{sec:evaluation-reuse}
 
@@
 \end{proof}
+
+% ==================================================================================================
+
+\subsection{Speed}
+\label{sec:evaluation-empirical}
+
+\includegraphics[scale=0.4]{drawing.png}
+
+image.load("shadow.png").greyScale().blur(5)
+
+image.load("shadow.png").greyScale().blur(10)
+
+let s = image.load("shadow.png").greyScale().blur(5)
+s
+
+let s = image.load("shadow.png").greyScale().blur(10)
+s.combine(image.load("pope.png"), 80)
+
+let r = 80
+let s = image.load("shadow.png").greyScale().blur(10)
+s.combine(image.load("pope.png"), r)
+
+\newpage
+
+\section{Case study}
+\label{sec:case}
 
@@
 
-\paragraph{Type checking result reuse.} In Section~\ref{sec:properties-reuse}, we show that
+\paragraph{Type checking result reuse.} In Section~\ref{sec:evaluation-reuse}, we show that
 certain source code edits do not cause the recomputation of previews for the whole expression
@@
 \paragraph{Design and human-computer interaction.}
+TODO: \citet{principle}
+
 From a design perspective, the idea of live programming environments has been popularised by
@@
 
+TBD: Define live preview somewhere
+
 TBD: In section about dependency graphs, include a subsection discussing how this relates
 to incremental computation (just like "Semantics of delayed previews" links that to CMTT etc.)
+
+\cite{selfadjusting}
+\cite{incremental}
+
+Incremental computation
+
+  Self-adjusting computation
+  http://www.cs.cmu.edu/~rwh/theses/acar.pdf
+
+  Incremental Computation with Names
+  https://www.cs.tufts.edu/~jfoster/papers/oopsla15.pdf
+  -> dsl interpreter
+
+  (and other adapton stuff http://adapton.org/)
+
+Type checking
+
+  co-contextual
+  http://www.informatik.uni-marburg.de/~seba/publications/cocontextual-type-checking.pdf
+  http://drops.dagstuhl.de/opus/volltexte/2017/7262/pdf/LIPIcs-ECOOP-2017-18.pdf
+
+https://dl.acm.org/citation.cfm?id=664109
 
@@
 
-  If $FV(e)\neq\emptyset$, then $\bnclr{v_i} \Downarrow_\ident{lift} \llbracket e'_i \rrbracket$ for $i\in 0\ldots n$ and
+  If $FV(e)\neq\emptyset$, then $\bndclr{v_i} \Downarrow_\ident{lift} \llbracket e'_i \rrbracket$ for $i\in 0\ldots n$ and
   $\bndclr{v}\Downarrow \llbracket e'_0.m(e'_1, \ldots, e'_n) \rrbracket_{FV(e)}$ using \rname{mem-expr}.

### 2019-01-30 wip-live-programming/experiments/script.R

@@ -1,0 +1,13 @@
+install.packages("ggplot2")
+library(ggplot2)
+library(reshape2)
+
+dr <- read.csv("C:/Tomas/Research/TheGamma/wip-live-programming/experiments/drawing.csv")
+names(dr) <- c("full", "live")
+dr$time <- seq.int(nrow(dr))
+
+drm <- melt(dr,id = c("time"))
+names(drm) <- c("time", "method", "delay")
+
+ggplot(drm,aes(x = time,y = delay)) + 
+  geom_bar(aes(fill = method),stat = "identity",position = "dodge")

### 2019-02-02 wip-live-programming/paper.tex
Finish cleaning up intro sections

@@
 \caption{An excerpt from a Jupyter notebook by Financial Times reporters, which loads trade
-  data from the UN trade database and adds ISO country codes.}
+  data from the UN trade database (stored in a local file) and adds ISO country codes (from
+  an Excel spreadsheet).}
 \label{fig:ft-uncomtrade}
 \end{figure}
+
+% --------------------------------------------------------------------------------------------------
 
@@
 
-Data analysts increasingly use notebook systems such as Jupyter or R Markdown, which
+Data analysts increasingly use notebook systems such as Jupyter, which
 make it possible to combine text, equations and code with results of running the code such as tables
@@
 
-% interactively based on the results. We aim to build foundations for such seamless integration
-% of development and execution in a notebook environment.
-
 Jupyter notebooks are used by a variety of users ranging from scientists who implement complex
-models to journalists who load data, perform simple aggregations and create visualizations. Our
+models of physical systems to journalists who load data, perform simple aggregations and create visualizations. Our
 initial focus is on simple use cases, such as those of journalists. Tooling that makes notebook
-systems more interactive and spreadsheet-like in such cases is of crucial importance as it enables
-more people to produce reproducible and transparent data analyses.
-
-As an example, consider the anlysis done by the Financial Times for an article on plastic
-waste. The analysis joins datasets from Eurostat, UN Comtrade and more, aggregates the data and
-builds a visualization comparing plastic waste flows in 2017 and 2018.
-Figure~\ref{fig:ft-uncomtrade} shows an extract from one notebook from the data analysis.
-The code of the analysis has a number of remarkable properties:
+systems more interactive and spreadsheet-like can encourage users to choose programming tools
+over spreadsheets and produce reproducible and transparent data analyses.
+
+As an example, the anlysis done by the Financial Times for an article on plastic waste
+\cite{ftnotebooks,ftarticle} joins datasets from Eurostat, UN Comtrade and more, aggregates the data and
+builds a visualization comparing waste flows in 2017 and 2018. Figure~\ref{fig:ft-uncomtrade} shows
+an extract from one notebook from the data analysis. The code has a number of remarkable properties:
 %
 \begin{itemize}
-\item There is no abstraction. The full analysis uses lambda functions as arguments to library calls,
+\item There is no abstraction. The analysis uses lambda functions as arguments to library calls,
   but it does not define any top-level functions. Parameterization is done by defining a list of
@@
   This lets the analyst easily see and check the results of intermediate steps of the analysis.
+
+\vspace{0.5em}
+\item The code relies on external libraries. Our example uses Pandas \cite{pandas},
+  which provides opreations for data wrangling such as \ident{merge} to join datasets
+  or \ident{drop\_duplicates} to delete rows with duplicate column values. Those libraries are
+  external to the data analysis and may even be implemented in other languages -- many Python
+  libraries rely on C++.
 
@@
 
-~
-
-TODO: Most logic in external libs
-
-~
-
+\noindent
 Many Jupyter notebooks are more complex than the example discussed here. They might define
-helper functions or even structure code using object-oriented programming. However, simple
-data analyses such as the one discussed here are frequent enough to pose an interesting and
-important domain for programming tools research.
+helper functions or even structure code using object-oriented programming. However, this paper
+is motivated by our conviction that simple data analyses such as the one discussed here are frequent
+enough to pose an interesting and important domain for programming tools research.
 
@@
   \subcaptionbox{
-    \raggedright\footnotesize The Gamma script to aggregate Olympic medals. We select Rio 2016 games and
-    count the number of distinct events per athlete. We then type `.' to choose
-    further aggregation operations.
+    \raggedright\footnotesize The Gamma script to aggregate Olympic medals. The analyst selects
+      Rio 2016 games and counts the number of distinct events per athlete. After typing `.',
+      a type provider offers further aggregation operations.
     \label{fig:gamma-dot}
@@
   \subcaptionbox{
-    \raggedright\footnotesize Live preview in The Gamma programming environment. The table is produced as
+    \raggedright\footnotesize Live preview in The Gamma programming environment developed based on
+      the theory presented in this paper. The table is produced as
       the data anlyst types and updates on-the-fly to show result at the current cursor position.
@@
   }{\includegraphics[scale=0.26]{gamma2.png}}
-  \caption{\small TBD}
+  \caption{\small Data exploration in The Gamma -- using a type provider (left) and with a live
+    preview (right).}
   \label{first-figure-with-subfigures}
@@
 
-Data exploration of the kind discussed in the previous section has been the motivation
-for a recent programming environment called The Gamma. The scripting language in The Gamma
-captures the structure discussed above.
-
-A script is a sequence of commands that can either define a
-variable or produce an output. The language does not support top-level function declarations,
-although we propose an extension that provides a simple abstraction mechanism and is compatible with
-providing live previews in Section~X. Lambda functions can be used only as method
-arguments.
-
-Given the limited expressiveness of The Gamma, libraries for it are implemented in other languages, such as
-JavaScript. More importantly, The Gamma supports type providers, which can be used to generate types,
-on the fly, for accessing and exploring external data sources. Type providers provide object types
-with members and The Gamma makes using those convenient by providing auto-complete when the user
-types the dot symbol (`.') to access a member.
-
-The combination of type providers and auto-complete on~``.'' makes it possible to complete a
-large number of data exploration tasks through a very simple interaction of selecting operations from
-a list. An example in Figure~\ref{fig:gamma-dot} summarizes data on Olympic medals using a type provider.
-The names of identifiers such as \qident{sum Bronze} do not have
-any special meaning -- they are merely members generated by the type provider, based on information
+Data exploration of the kind discussed in the previous section has been the motivation for a
+recent work by \citet{gamma} that develops a small scripting language called The Gamma. The language
+follows the structure discussed above. A script is a sequence of commands that can either define a
+variable or produce an output. The language does not support top-level function declarations
+and lambda functions can be used only as method arguments.
+
+Given the limited expressiveness of The Gamma, libraries are implemented in other languages,
+such as JavaScript. The Gamma also supports type providers \cite{providers-fsharp},
+which can be used to generate types, on the fly, for accessing and exploring external data sources.
+Type providers provide object types with members and The Gamma makes using those convenient by
+providing auto-complete when the user types the dot symbol (`.') to access a member.
+
+As illustrated by \citet{gamma}, the combination of type providers and auto-complete on~`.' makes it
+possible to complete a large number of data exploration tasks through a very simple interaction of
+selecting operations from a list. An example in Figure~\ref{fig:gamma-dot} summarizes data on
+Olympic medals using a type provider. The names of identifiers such as \qident{sum Bronze} do not have
+any special meaning. They are merely members generated by the type provider, based on information
 about the data source. The type provider used in this example generates an object with members for
@@
 
-The Gamma is more complex than shown here -- it supports methods that can take
-functions as arguments -- but the above example illustrates the fact that non-trivial data
-exploration can be done using a very simple language.
-
-We complement our theoretical work on foundations of live programming environments for data
-exploration with an implementation for The Gamma. As discussed in the previous
-section, the assumptions about structure of code that are explicit in The Gamma are also
-implicitly present in Python and R data analyses produced by journalists, economists and
-other users with other than  programming background. Consequently, our theoretical work
-applies to simple data analyses using Python and readers not familiar with The Gamma
-can assume that we use a small subset of Python.
+The Gamma language is richer, but the example in Figure~\ref{fig:gamma-dot} shows that non-trivial
+data exploration can be done using a very simple language. The assumptions about strucutre of
+code that are explicit in The Gamma are implicitly present in Python and R data analyses
+produced by journalists, economists and other users with other than programming background.
+When we refer to The Gamma in this paper, readers not familiar with it can assume that we focus
+on a subset of Python.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Live coding for data exploration}
+
+The practical implementation work that complements the theory presented in this paper implements a live
+coding environment for The Gamma and is discussed in Section~\ref{sec:case} in detail. Earlier work by
+\citet{gamma} discussed The Gamma language and the type provider mechanism, but the implementation
+provided only a standard text editor with auto-completion. Our work brings the experience of using
+The Gamma closer to that of spreadsheets by developing an editor that evaluates scripts
+instantaneously in background and displays the results as \emph{live previews}.
+
+An example of a live preview is shown in Figure~\ref{fig:gamma-preview}. As noted earlier, The
+Gamma program consits of a list of commands which are either expressions or let bindings. Our
+editor displays a live preview below the command that the user is currently editing. The preview
+shows the result of evaluating the expression or the value assigned to a bound variable. When the
+user changes the code, the preview is updated automatically, without any additional interaction
+with the user.
+
+There is a number of guiding principles that our work follows. First, we allow the analyst
+to edit code in an unrestricted form in a text editor. Although structured editors provide an
+appearling alternative and make recomputation easier, we prefer the transparency
+and openness of plain text. Second, we focus on the scenario when code changes, but input
+does not. Rapid feedback allows the analyst to quickly adapt code to correctly handle
+corner cases that typical analysis involves. In contrast to work on incremental computation,
+we do not consider the case when data changes, although supporting interactive data exploration
+of streaming data is an interesting future direction.
 
@@
 
-% https://www.ft.com/content/360e2524-d71a-11e8-a854-33d6f82e62f8
-% https://github.com/ft-interactive/recycling-is-broken-notebooks
-
-\section{Live coding for data exploration}
-
-% TODO: What is live coding? (Show preview while typing!)
-
-This paper presents the foundations of a live coding environment for data exploration that
-follows the analysis in the previous section. We consider simple data exploration tasks, such as
-those done by journalists and we implement our ideas in the editor for TheGamma.
-
-Figure~\ref{fig:gamma-preview} shows a TheGamma script exporing Olympic medal statistics in
-our live coding environment. Before discussing the technical details, it is worth highlighting
-a number of important points about our approach.
-
-\begin{itemize}
-\item\textbf{Real-world use cases.}
-Our work has been motivated by the needs outlined in the previous section. In particular, we
-have been focusing on a langauge that is simple, yet sufficient for completing basic data
-exploration tasks that a journalist might want to do. We have used our live coding environment
-to perform a number of simple data analyses looking at topics such as public government spending,
-crises in financial markets and activities of a research institute. Qualitative observations about
-our experience are reported in Section~X. Our tools can be used in any modern web browser and are
-available as an open-source package.
-
-\item\textbf{Text editor integration.}
-Our approach is based on an ordinary text editor that allows unrestricted modification of the
-code. Our parser supports error-recovery and we attempt to give previews whenever the code can be
-correctly parsed and type-checked up to the current location. Notably, we do not, in general, know
-how the user modified the code. This is in contrast with approaches based on structured editing
-where the editor tracks what changes to code have been made. Using an ordinary text editor makes
-providing live previews harder, but we believe that many users still favor the flexibility of
-plain text editors.
-
-\item\textbf{Data stay, code changes.}
-Our focus has been on the scenario when input data is a static file that does not change, but the
-data analyst modifies the source code. This is in contrast with work on incremental computation,
-which focuses on scenarios where code stays the same, but input data change.
-
-That said, many of the methods that we used for evaluating live previews such as maintaining
-the dependency graph with nodes for individual syntactic elements are closely related to methods
-used in incremental computation research. We do not attempt to combine the two, but note that
-this is an interesting future direction.
-\end{itemize}
+\section{Data exploration calculus}
+\label{sec:calculus}
+
+The \emph{data exploration calculus} is a small, formally tractable language for data exploration.
+The calculus is not Turing-complete and it can only be used together with external libraries that
+define what objects are available and what the behaviour of their members is. This is
+sufficient to capture the simple data analyses discussed in Section~\ref{sec:background}. We
+define the caluclus in this section and then use it as the basis for discussing our live preview
+mechanism in Section~\ref{sec:formal}.
+
+We initially present the data exploration calculus without a static type system. The live preview
+mechanism can be discussed equally well without types. We consider type checking later in
+Section~\ref{sec:types} as the mechanism that is used for live previews can also be
+used for incremental type-checking.
+
+% --------------------------------------------------------------------------------------------------
 
@@
 
-\newpage
-
-\section{Data exploration calculus}
-\label{sec:calculus}
-
-The \emph{data exploration calculus} is a small, formally tractable language for data exploration.
-The calculus is not Turing-complete and it can only be used together with external libraries that
-define what objects are available and what the behaviour of their members is. This is
-sufficient to capture the simple data analyses discussed in Section~\ref{sec:background}. We
-define the caluclus in this section and then use it as the basis for discussing our live preview
-mechanism in Section~\ref{sec:formal}.
-
-We initially present the data exploration calculus without a static type system. Our main focus
-is on the live preview mechanism which can be done equally well without types. We consider
-type checking in Section~\ref{sec:types}, because the same mechanism that is used for live previews
-can also be used for incremental type-checking.
+% --------------------------------------------------------------------------------------------------
 
@@
 and implements an effective optimization. We discuss correctness in this section and evaluate
-the effectiveness of the optimization, both theoretically and empirically, in Section~\ref{sec:optimization}.
-
-To show that the evaluated previews are correct, we prove two properties. Correctness
+the effectiveness of the optimization, both theoretically and empirically, in Section~\ref{sec:evaluation}.
+
+To show that the previews are correct, we prove two properties. Correctness
 (Theorem~\ref{thm:correcntess}) guarantees that, no matter how a dependency graph is constructed,
@@
 
+\newpage
+
 \section{Evaluation}
@@
 variable appears multiple times.
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
-\raggedright
-\subparagraph{1. Let introduction A.}
-The expression $C_1[C_2[e]]$ is changed to $C_1[\kvd{let}~x=e~\kvd{in}~C_2[x]]$
-via semantically non-equivalent expression $C_1[C_2[x]]$ where $x$ is unbound variable.
-
-\subparagraph{2. Let introduction B.}
-The expression $C_1[C_2[e]]$ is changed to $C_1[\kvd{let}~x=e~\kvd{in}~C_2[x]]$
-via $C_1[\kvd{let}~x=e~\kvd{in}~C_2[e]]$ where $x$ is unused variable.
-
-\subparagraph{3. Let elimination A.} The expression
-$C_1[\kvd{let}~x=e~\kvd{in}~C_2[x]]$ is changed to
-$C_1[C_2[e]]$ via semantically non-equivalent expression
-$C_1[C_2[x]]$ where $x$ is unbound variable.
-
-\subparagraph{4. Let elimination B.} The expression
-$C_1[\kvd{let}~x=e~\kvd{in}~C_2[x]]$ is changed to
-$C_1[C_2[e]]$ via $C_1[\kvd{let}~x=e~\kvd{in}~C_2[e]]$
-where $x$ is unused variable.
-
-\subparagraph{5. Editing a non-dependency in let.} Assuming $x\notin FV(e_2)$, the expression
-$C_1[\kvd{let}~x=e_1~\kvd{in}~C_2[e_2]]$ changes to an expression
-$C_1[\kvd{let}~x=e_1'~\kvd{in}~C_2[e_2]]$. The preview of a sub-expression $e_2$ is not recomputed.
-
-\subparagraph{6. Editing a non-dependency in a chain.} The expression
-$C[e.m(e_1, \ldots, e_n).m'(e'_1, \ldots, e'_k)]$ is changed to an expression
-$C[e.m(e_1, \ldots, e_n).m''(e''_1, \ldots, e''_k)]$. The preview of a sub-expression
-$e.m(e_1, \ldots, e_n)$ is not recomputed.
-
-\caption{Code edit operations that enable preview reuse}
-\label{fig:operations}
-\end{figure}
 
@@
 
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\raggedright
+\subparagraph{1. Let introduction A.}
+The expression $C_1[C_2[e]]$ is changed to $C_1[\kvd{let}~x=e~\kvd{in}~C_2[x]]$
+via semantically non-equivalent expression $C_1[C_2[x]]$ where $x$ is unbound variable.
+
+\subparagraph{2. Let introduction B.}
+The expression $C_1[C_2[e]]$ is changed to $C_1[\kvd{let}~x=e~\kvd{in}~C_2[x]]$
+via $C_1[\kvd{let}~x=e~\kvd{in}~C_2[e]]$ where $x$ is unused variable.
+
+\subparagraph{3. Let elimination A.} The expression
+$C_1[\kvd{let}~x=e~\kvd{in}~C_2[x]]$ is changed to
+$C_1[C_2[e]]$ via semantically non-equivalent expression
+$C_1[C_2[x]]$ where $x$ is unbound variable.
+
+\subparagraph{4. Let elimination B.} The expression
+$C_1[\kvd{let}~x=e~\kvd{in}~C_2[x]]$ is changed to
+$C_1[C_2[e]]$ via $C_1[\kvd{let}~x=e~\kvd{in}~C_2[e]]$
+where $x$ is unused variable.
+
+\subparagraph{5. Editing a non-dependency in let.} Assuming $x\notin FV(e_2)$, the expression
+$C_1[\kvd{let}~x=e_1~\kvd{in}~C_2[e_2]]$ changes to an expression
+$C_1[\kvd{let}~x=e_1'~\kvd{in}~C_2[e_2]]$. The preview of a sub-expression $e_2$ is not recomputed.
+
+\subparagraph{6. Editing a non-dependency in a chain.} The expression
+$C[e.m(e_1, \ldots, e_n).m'(e'_1, \ldots, e'_k)]$ is changed to an expression
+$C[e.m(e_1, \ldots, e_n).m''(e''_1, \ldots, e''_k)]$. The preview of a sub-expression
+$e.m(e_1, \ldots, e_n)$ is not recomputed.
+
+\caption{Code edit operations that enable preview reuse}
+\label{fig:operations}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
 % ==================================================================================================
@@
 
-\section{Case study}
+\section{Case study / implementation}
 \label{sec:case}

### 2019-02-03 wip-live-programming/paper.tex
Reuse of previews, cbv and cbn experiments

@@
 \usepackage{amsmath}
+\usepackage[T1]{fontenc}
 
@@
 \newcommand{\rname}[1]{{\sffamily\small(#1)}}
+\newcommand{\ername}[1]{\vspace{0.75em}\rname{#1}\hspace{0.5em}}
+\newcommand{\preview}[1]{\,\textnormal{\guillemotleft} #1\textnormal{\guillemotright}\,}
 
@@
 
-\item We prove that our live preview mechanism is correct (Section~\ref{sec:correctness}) and that
+\item We prove that our live preview mechanism is correct (Section~\ref{sec:evaluation-correctness}) and that
   it reuses previously evaluated values when possible (Section~\ref{sec:evaluation-reuse}). We
@@
 is to capture properties that allow us to prove that our method of evaluating live previews,
-discussed in Section~\ref{sec:correctness}, is correct.
+discussed in Section~\ref{sec:evaluation-correctness}, is correct.
 
@@
 using the $\rightsquigarrow_\ident{let}$ reduction does not affect the result. This
-simplifies our later proof of live preview correctness in Section~\ref{sec:correctness-correct}.
+simplifies our later proof of live preview correctness in Section~\ref{sec:evaluation-correctness}.
 
@@
 
-\section{Evaluating previews}
+\section{Computing live previews}
 \label{sec:previews}
@@
 given in Section~\ref{sec:calculus-semantics}. We analyse the preview evaluation formally in
-Section~\ref{sec:correctness} and show that the resulting previews are the same as the result we
+Section~\ref{sec:evaluation} and show that the resulting previews are the same as the result we
 would get by directly evaluating the code and, also, that no recomputation occurs when code is
@@
 
-\section{Correctness of previews}
-\label{sec:correctness}
+\section{Evaluating correctness and effectiveness}
+\label{sec:evaluation}
 
@@
 previously evaluated partial results. In other words, we show that the mechanism is correct
-and implements an effective optimization. We discuss correctness in this section and evaluate
-the effectiveness of the optimization, both theoretically and empirically, in Section~\ref{sec:evaluation}.
+and implements an effective optimization. We discuss correctness first and then evaluate
+effectiveness, both theoretically (Section~\ref{sec:evaluation-reuse}) and empirically
+(Section~\ref{sec:evaluation-empirical}).
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Correctness of previews}
+\label{sec:evaluation-correctness}
 
@@
 By induction over $\Downarrow$ over $(V, E)$, we show that the same evaluation rules also
-apply over $(V', E')$.
-This is the case, because new graph nodes added to $\Delta'$ by $\ident{update}_{V, E}$
-are only ever added as new nodes in $\ident{bind-prog}_{\emptyset, \Delta'}$ and so the existing
+apply over $(V', E')$. This is the case, because graph nodes added to $\Delta'$ by $\ident{update}_{V, E}$
+are added as new nodes in $\ident{bind-prog}_{\emptyset, \Delta'}$ and
 nodes and edges of $(V, E)$ used during the evaluation are unaffected.
@@
 
-\newpage
-
-\section{Evaluation}
-\label{sec:evaluation}
-
 \subsection{Reuse of previews}
@@
 
-In the motivating example in Section~\ref{sec:intro}, the programmer first extracted a constant
-value into a let binding and then modified a parameter of the last method call in a call chain.
-We argued that the live coding environment should reuse partially evaluated previews for these
-two cases. In this section, we prove that this is, indeed, the case in our system.
-
-Figure~\ref{fig:operations} shows a list of six code edit operations where a preview of the
-expression (cases 1-4), or a sub-expression (cases 5-6), can be reused. This is the case, because
-the graph nodes that are bound to the sub-expression before and after the code is changed are the
-same and hence, a cached preview (stored using the graph node as the key) can be reused.
-
-In some of the operations (cases 1 and 3), the code is changed via an intermediate expression
-that is semantically different and has only partial preview. This illustrates a typical way of
-working with code in a text editor using cut and paste operations. Cases 1 and 3 illustrate how
-our approach allows this way of editing code.
-
-Finally, it is worth noting that our list is not exhaustive. In particular, cases 1-4 only cover
-let bindings where the bound variable is used once. However, previews can also be reused if the
-variable appears multiple times.
+In the motivating example given in Section~\ref{sec:intro}, the data analyst first extracted a
+constant value into a let binding and then modified a parameter of the last method call in a call
+chain. We argued that a live coding environment should reuse partially evaluated previews for these
+two cases. We now show that this is, indeed, the case in our system.
+
+In this section, we identify a number of code edit operations where the previously evaluated
+values for a sub-expression can be reused. The list of operations is shown in Figure~\ref{fig:operations}.
+To express the operations we define an editing context $K$ which is defined
+similarly to evaluation context $C$ from Figure~\ref{fig:dec-calculus}, but captures sub-expressions
+appearing anywhere in the program.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\raggedright
+\vspace{0.5em}
+\hspace{1.5em}{\small\sffamily Edit contexts of expressions}
+\begin{equation*}
+\begin{array}{lcl}
+K_e[-] &=& K_e[-].m(e_1, \ldots, e_n) ~\lsep~ e.m(e_1, \ldots, e_{l-1}, K_e[-], e_{l+1},\ldots , e_n) ~\lsep~ -\\
+K_c[-] &=& \kvd{let}~x = K_e[-] ~\lsep~ K_e[-]\\
+\end{array}
+\end{equation*}
+
+\vspace{0.5em}
+\hspace{1.5em}{\small\sffamily Code edit operations preserving preview for a sub-expression}
+
+\hspace{4em}\begin{minipage}[c]{0.82\textwidth}
+  \raggedright\setlength{\parindent}{-1em}
+
+  \ername{let-intro-var}
+  $\overline{c_1}; \preview{e}; \overline{c_2}$ changes to
+  $\overline{c_1}; \kvd{let}\,x = e; \preview{x}; \overline{c_2}$ where $x$ is a fresh name.
+
+  \ername{let-intro-ins}
+  $\overline{c_1}; \overline{c_2}; \preview{K_c[e]}; \overline{c_3}$ is changed to
+  $\overline{c_1}; \kvd{let}\,x=e; \overline{c_2}; \preview{K_c[x]}; \overline{c_3}$ via
+  a semantically non-equivalent expression
+  $\overline{c_1}; \overline{c_2}; K_c[x]; \overline{c_3}$ where $x$ is free.
+
+  \ername{let-intro-del}
+  $\overline{c_1}; \overline{c_2}; \preview{K_c[e]}; \overline{c_3}$ is changed to
+  $\overline{c_1}; \kvd{let}\,x=e; \overline{c_2}; \preview{K_c[x]}; \overline{c_3}$ via
+  an expression $\overline{c_1}; \kvd{let}\,x=e; \overline{c_2}; K_c[e]; \overline{c_3}$
+  with unused variable $x$.
+
+  \ername{let-elim-del}
+  $\overline{c_1}; \kvd{let}\,x=e; \overline{c_2}; \preview{K_c[x]}; \overline{c_3}$ is changed to
+  $\overline{c_1}; \overline{c_2}; \preview{K_c[e]}; \overline{c_3}$ via
+  a semantically non-equivalent expression
+  $\overline{c_1}; \overline{c_2}; K_c[x]; \overline{c_3}$ where $x$ is free.
+
+  \ername{let-elim-ins}
+  $\overline{c_1}; \kvd{let}\,x=e; \overline{c_2}; \preview{K_c[x]}; \overline{c_3}$ is changed to
+  $\overline{c_1}; \overline{c_2}; \preview{K_c[e]}; \overline{c_3}$ via
+  an expression $\overline{c_1}; \kvd{let}\,x=e; \overline{c_2}; K_c[e]; \overline{c_3}$
+  with unused variable $x$.
+
+  \ername{edit-mem}
+  $\overline{c_1}; K_c[\preview{e_0}.m(\overline{e})]; \overline{c_2}$ is changed to
+  $\overline{c_1}; K_c[\preview{e_0}.m'(\overline{e'})]; \overline{c_2}$
+
+  \ername{edit-let}
+  $\overline{c_1}; \kvd{let}\,x=e_1; \overline{c_2}; K_c[\preview{e_2}]; \overline{c_3}$ is changed to
+  $\overline{c_1}; \kvd{let}\,x=e'_1; \overline{c_2}; K_c[\preview{e_2}]; \overline{c_3}$ \hspace{2em} when
+  $x\notin FV(e_2)$.
+\end{minipage}
+
+\caption{Code edit operations that preserve previously evaluated preview}
+\label{fig:operations}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+We use the notation $\preview{e}$ to mark the part of the expressions before and after the edit
+that have the same value that will not need to be recomputed after the change and we write
+$\overline{c}$ and $\overline{e}$ for a list of commands and expressions, respectively.
+In some of the edit operations, we also specify an intermediate program that is, in some cases,
+semantically different and only has a partial preview. This illustrates a typical way of
+working with code in a text editor using cut and paste operations. For example, in
+\rname{let-intro-ins}, we assume that the analyst cuts a sub-expression $e$, replaces it with
+a variable $x$ and then adds a let binding for a variable $x$ and inserts the expression $e$
+from the clipboard. Edit with the same result is captured by \rname{let-intro-del} but here
+the analyst first inserts the let binding and then replaces the sub-expression $e$ with a
+variable $x$.
+
+The of preview-preserving code edit operations includes several ways of introducing a let
+binding, several ways of eliminating a let binding and two edit operations where the data anlyst
+modifies an unrelated part of the program. It is worth noting that the list is not exhaustive.
+Rather, it aims to illustrate some of the typical operations that the data analyst might perform
+when writing code.
+
+Theorem~\ref{thm:preview-reuse} proves that the operations given in Figure~\ref{fig:operations}
+preserve the preview for a marked sub-expression. It relies on the following lemma that
+characterizes one of the common kinds of edits more generally. Given two versions of a program
+that both contain the same sub-expression $e$, if the let bindings that define the values of
+variables used in $e$ do not change, then the graph node assigned to $e$ will be the same
+when binding the original and the updated program.
 
@@
 \label{thm:sub-expr}
-Given any $\Delta_1$ together with $e_1 = C[C_1[e]]$ and $e_2 C[C_2[e]]$, such that all free
-variables of $e$ are bound in $C$, assume that the first expression is bound,
-i.e.~$\bndclr{v_1}, (V_1, E_1) = \ident{bind-expr}_{\emptyset, \Delta_1}(e_1)$, the graph node cache
-is updated $\Delta_2 = \ident{update}_{V_1, E_1}(\Delta_1)$ and the second expression is bound, i.e.~
-$\bndclr{v_2}, (V_2, E_2) = \ident{bind-expr}_{\emptyset, \Delta_2}(e_2)$.
-
-Now, assume $\bndclr{v}, G = \ident{bind-expr}_{\Gamma_1, \Delta_1}(e)$ and
-$\bndclr{v'}, G' = \ident{bind-expr}_{\Gamma_2, \Delta_2}(e)$ are the recursive calls to bind
+Assume we have programs $p_1 = c_1; \ldots; c_k; K_c[e]; c_{k+1};\ldots; c_n$ and
+$p_2 = c'_1; \ldots; c'_k; K'_c[e]; c'_{k+1};\ldots;c'_n$ and $I\subseteq \{1\ldots k\}$ such
+that $\forall i\!\in\!I.\,c_i\!=\!c'_i$ and for each $x \in \bigcup_{i\in I}FV(c_i) \cup FV(e)$
+there exists $j\in I$ such that $c_j = \kvd{let}\,x=e$ for some $e$.
+%
+Given any $\Delta$, assume that the the first program is bound,
+i.e.~$\bndclr{v_1};\ldots;\bndclr{v_n}, (V, E) = \ident{bind-prog}_{\emptyset, \Delta}(p_1)$,
+the cache is updated $\Delta' = \ident{update}_{V, E}(\Delta)$ and the second
+program is bound,
+i.e.~$\bndclr{v'_1};\ldots;\bndclr{v'_n}, (V', E') = \ident{bind-prog}_{\emptyset, \Delta'}(p_2)$.
+
+Now, assume $\bndclr{v}, G = \ident{bind-expr}_{\Gamma, \Delta}(e)$ and
+$\bndclr{v'}, G' = \ident{bind-expr}_{\Gamma', \Delta'}(e)$ are the recursive calls to bind
 $e$ during the first and the second binding, respectively. Then, the graph nodes assigned to the
@@
 \begin{proof}
-First, assuming that $\forall x\in FV(e). \Gamma_1(x) = \Gamma_2(x)$, we show by induction over the binding process of $e$
-when binding $C[C_1[e]]$ that the result is the same. In cases (1) and (6), the updated $\Delta_2$
+First, assuming that $\forall x\in FV(e). \Gamma(x) = \Gamma'(x)$, we show by induction over the binding process of $e$
+for the first program that the result is the same. In cases (1) and (3), the updated $\Delta'$
 contains the required key and so the second binding proceeds using the same case. In cases
-(2) and (7), the second binding reuses the node created by the first binding using case (1) and
-(6), respectively. Cases (4) and (5) are the same and case (3) follows directly via induction.
-
-Second, when binding let bindings in $C[-]$, the initial $\Gamma = \emptyset$ during both bindings
-and so the nodes added to $\Gamma_1$ and $\Gamma_2$ are the same. $C_1$ and $C_2$ do not add
-any new nodes used in $e$ to $\Gamma_1$ and $\Gamma_2$ and so $\bndclr{v} = \bndclr{v'}$
-using the above.
+(2) and (4), the second binding reuses the node created by the first binding using case (1) and
+(3), respectively. Cases (5) and (6) are the same.
+
+Second, when binding let bindings in $c_1; \ldots; c_k$, the initial $\Gamma = \emptyset$ during
+both bindings. Nodes added to $\Gamma$ and $\Gamma'$ for commands $c_j$ such that $j\in I$ are
+the same and nodes added for remaining commands do not add any new nodes referenced from $e$ and
+so $\bndclr{v} = \bndclr{v'}$ using the above.
 \end{proof}
@@
 \begin{theorem}[Preview reuse]
+\label{thm:preview-reuse}
 Given the sequence of expressions as specified in Figure~\ref{fig:operations}, if the expressions
@@
 \begin{proof}
-Consequence of Lemma~\ref{thm:sub-expr}, using appropriate contexts. In cases with
-intermediate expressions (1)-(4), binding the intermediate expression introduces additional
-nodes to $\Delta$, but those are ignored when binding the final expression.
+Cases \rname{edit-let} and \rname{edit-mem} are direct consequences of Lemma~\ref{thm:sub-expr};
+for \rname{let-intro-var}, the node assigned to $x$ is the node assigned to $e$ which is the
+same as before the edit from Lemma~\ret{thm:sub-expr}.
+Cases \rname{let-intro-ins} and \rname{let-intro-del} are similar to \rname{let-intro-var}, but
+also require using induction over the binding of $K_c[e]$. Finally, cases \rname{let-elim-ins}
+and \rname{let-elim-del} are similar and also use Lemma~\ref{thm:sub-expr}
+together with induction over the binding of $K_c[x]$.
 \end{proof}
@@
 
-\begin{figure}[t]
-\raggedright
-\subparagraph{1. Let introduction A.}
-The expression $C_1[C_2[e]]$ is changed to $C_1[\kvd{let}~x=e~\kvd{in}~C_2[x]]$
-via semantically non-equivalent expression $C_1[C_2[x]]$ where $x$ is unbound variable.
-
-\subparagraph{2. Let introduction B.}
-The expression $C_1[C_2[e]]$ is changed to $C_1[\kvd{let}~x=e~\kvd{in}~C_2[x]]$
-via $C_1[\kvd{let}~x=e~\kvd{in}~C_2[e]]$ where $x$ is unused variable.
-
-\subparagraph{3. Let elimination A.} The expression
-$C_1[\kvd{let}~x=e~\kvd{in}~C_2[x]]$ is changed to
-$C_1[C_2[e]]$ via semantically non-equivalent expression
-$C_1[C_2[x]]$ where $x$ is unbound variable.
-
-\subparagraph{4. Let elimination B.} The expression
-$C_1[\kvd{let}~x=e~\kvd{in}~C_2[x]]$ is changed to
-$C_1[C_2[e]]$ via $C_1[\kvd{let}~x=e~\kvd{in}~C_2[e]]$
-where $x$ is unused variable.
-
-\subparagraph{5. Editing a non-dependency in let.} Assuming $x\notin FV(e_2)$, the expression
-$C_1[\kvd{let}~x=e_1~\kvd{in}~C_2[e_2]]$ changes to an expression
-$C_1[\kvd{let}~x=e_1'~\kvd{in}~C_2[e_2]]$. The preview of a sub-expression $e_2$ is not recomputed.
-
-\subparagraph{6. Editing a non-dependency in a chain.} The expression
-$C[e.m(e_1, \ldots, e_n).m'(e'_1, \ldots, e'_k)]$ is changed to an expression
-$C[e.m(e_1, \ldots, e_n).m''(e''_1, \ldots, e''_k)]$. The preview of a sub-expression
-$e.m(e_1, \ldots, e_n)$ is not recomputed.
-
-\caption{Code edit operations that enable preview reuse}
-\label{fig:operations}
+\subsection{Empirical evaluation of efficiency}
+\label{sec:evaluation-empirical}
+
+\begin{figure}
+\includegraphics[scale=0.5]{drawing.png}
+\caption{Zzz}
+\label{fig:drawing}
 \end{figure}
 
-% --------------------------------------------------------------------------------------------------
-
-% ==================================================================================================
-
-\subsection{Speed}
-\label{sec:evaluation-empirical}
-
-\includegraphics[scale=0.4]{drawing.png}
+\subsubsection{Methodology}
 
@@
 s.combine(image.load("pope.png"), r)
+
+\subsubsection{Conclusions}
 

### 2019-02-04 wip-live-programming/paper.tex
Adding empirical evaluation section

@@
 %
-\begin{itemize}
+\begin{itemize}[itemsep=3pt]
 \item There is no abstraction. The analysis uses lambda functions as arguments to library calls,
@@
 
-\vspace{0.5em}
 \item The code relies on external libraries. Our example uses Pandas \cite{pandas},
@@
 
-\vspace{0.5em}
 \item The code is structured as a sequence of commands. Some commands define a variable, either by
@@
 
-\vspace{0.5em}
 \item There are many corner cases, such as the fact that the \ident{keep\_default\_na} parameter
@@
 
+The key performance claim about our method of providing instantaneous feedback is that it is more
+efficient than recomputing values for the whole program (or the current command) after every
+keystroke. In the previous section, we formally proved that this is true and gave examples of
+code edit operations that do not cause recomputation. In this section, we further support this
+claim with an empirical evaluation. The purpose of this section is not to precisely evaluate
+overheads of our implementation, but to compare how much recomputation different evaluation
+strategies perform.
+
+For the purpose of the evaluation, we use a simple image manipulation library that provides
+operations for loading, greyscaling, bluring and combining images. We compare delays in
+updating the preview for three different evaluation strategies, while performing the same
+sequence of code edit operations. Using image processing as an example gives us a way to
+visualize the reuse of previously computed values. As in a typical data exploration scenario,
+the individual operations are relatively expensive compared to the overheads of building the
+dependency graph.
+
+% --------------------------------------------------------------------------------------------------
+
 \begin{figure}
-\includegraphics[scale=0.5]{drawing.png}
-\caption{Zzz}
+\raggedright
+
+\vspace{0.5em}\hspace{1.5em}{\small\sffamily (1) Enter the following code and then change parameter of blur from 4 to 8:}
+\begin{equation*}
+\begin{array}{l}
+\ident{image}.\ident{load}(\str{shadow.png}).\ident{greyScale}().\ident{blur}(\num{4})
+\end{array}
+\end{equation*}
+
+\vspace{0.5em}\hspace{1.5em}{\small\sffamily (2) Assign the result to a variable and start writing code for further operations:}
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}\,\ident{shadow}=\ident{image}.\ident{load}(\str{shadow.png}).\ident{greyScale}().\ident{blur}(\num{8})\\
+\ident{shadow}.\ident{combine}
+\end{array}
+\end{equation*}
+
+\vspace{0.5em}\hspace{1.5em}{\small\sffamily (3) Finish code to combine two images and then change parameter of combine from 20 to 80:}
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}\,\ident{shadow}=\ident{image}.\ident{load}(\str{shadow.png}).\ident{greyScale}().\ident{blur}(\num{8})\\
+\ident{shadow}.\ident{combine}(\ident{image}.\ident{load}(\str{pope.png}), \num{20})
+\end{array}
+\end{equation*}
+
+\vspace{0.5em}\hspace{1.5em}{\small\sffamily (4) Extract the parameter of combine to a let bound variable:}
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}\,\ident{ratio} = \num{80} \\
+\kvd{let}\,\ident{shadow}=\ident{image}.\ident{load}(\str{shadow.png}).\ident{greyScale}().\ident{blur}(\num{8})\\
+\ident{shadow}.\ident{combine}(\ident{image}.\ident{load}(\str{pope.png}), \ident{ratio})
+\end{array}
+\end{equation*}
+
+\caption{Code edit operations that are used in the experimental evaluation}
+\label{fig:image-edits}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[b]
+\vspace{-0.5em}
+\includegraphics[scale=0.425]{drawing.png}
+\caption{Time required to recompute the results of a sample program after individual tokens of
+  the program are added or modified for three different evaluations strategies.}
 \label{fig:drawing}
+\vspace{-0.5em}
 \end{figure}
 
-\subsubsection{Methodology}
-
-image.load("shadow.png").greyScale().blur(5)
-
-image.load("shadow.png").greyScale().blur(10)
-
-let s = image.load("shadow.png").greyScale().blur(5)
-s
-
-let s = image.load("shadow.png").greyScale().blur(10)
-s.combine(image.load("pope.png"), 80)
-
-let r = 80
-let s = image.load("shadow.png").greyScale().blur(10)
-s.combine(image.load("pope.png"), r)
-
-\subsubsection{Conclusions}
+% --------------------------------------------------------------------------------------------------
+
+\subsubsection{Code edit operations and methodology}
+
+To avoid clutter when visualizing the performance, we update the preview after complete tokens are
+added (such as identifier, number, dot or left parenthesis) rather than after individual keystrokes.
+Figure~\ref{fig:image-edits} shows the sequence of code edit operations that we use to measure
+the delays in updating a live preview. We first enter an expression to load, greyscale and blur
+an image (1) then introduce let binding (2) and add more operations (3). Finally, we extract
+one of the parameters into a variable (4). Most of the operations are simply adding code, but
+there are two cases where we modify existing code and change value of a parameter for \ident{blur}
+and \ident{combine} immediately after (1) and (3), respectively.
+
+We implement the algorithm described in Sections~\ref{sec:formal} and \ref{sec:previews} in a
+simple web-based environment that allows the user to modify code and explicitly trigger
+recomputation. It then measures time needed to recompute values for the whole program and
+displays the resulting image. If the parsing fails, e.g.~because of an unclosed parenthesis,
+previews are not updated and we record only the time taken by parsing attempt. We compare the
+delays of three different evaluation strategies:
+
+\begin{itemize}[itemsep=3pt]
+\item \emph{Call-by-value.} Following the semantics in Section~\ref{sec:calculus-semantics},
+  all sub-expressions are fully evaluated before an expression is evaluated. In this case, work may
+  be done even if the evaluation results in an error. For example, we parse
+  $\ident{image}.\ident{load}(\str{shadow.png}).\ident{blur}$ as member access with no arguments.
+  The evaluation loads the image, but then fails because blur requires one argument.
+
+\item \emph{Lazy.} To address the wastefulness of call-by-value strategy discussed above, we
+  simulate lazy evaluation by implementing a version of the image processing library where
+  operations build a delayed computation and only evaluate it when rendering an image. Using
+  this strategy, failing computations do not perform unnecessary work.
+
+\item \emph{Live.} Finally, the live strategy uses the algorithm described in
+  Sections~\ref{sec:formal} and \ref{sec:previews}. The cache is empty at the beginning of the
+  experiment and we update it after each token is added. This is the only strategy that does
+  not start afresh after reparsing code.
+\end{itemize}
+
+\subsubsection{Efficiency evaluation}
+The experimental environment is implemented in F\# and compiled to JavaScript using the Fable
+compiler. We run the experiments in Firefox (version 64.0.2, 32bit) on Windows 10
+(1809, 64bit) on a computer with Intel Core i7-7660U CPU and 16Gb RAM.
+
+Figure~\ref{fig:drawing} shows times needed to recompute previews after individual tokens
+of the program are added, deleted or modified, according to the script in
+Figure~\ref{fig:image-edits}, resulting in 38 measurements for each of the 3
+evaluation strategies. We mark a number of notable points in the chart:
+
+\begin{enumerate}[(a)]
+\item Loading image for the first time incurs small extra overhead in the live strategy.
+\item Greyscaling using the live strategy does not need to re-load the image.
+\item Accessing the blur member without arguments causes delay for call-by-value strategy.
+\item When varying the parameter of blur, the live strategy reuses previously greyscaled image.
+\item Introducing let binding does not cause recomputation when using live strategy.
+\item As in (c), accessing combine member without valud argument only affects call-by-value.
+\item Varying parameter of combine in live strategy does not recompute blur and is much faster.
+\item Introducing let binding, again, causes full recomputation for lazy can call-by-value.
+\end{enumerate}
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}
+\noindent
+\includegraphics[scale=0.48]{hist-all.png}
+\includegraphics[scale=0.48]{hist-slow.png}
+\caption{Histogram showing the distribution of delays incurred when updating previews.
+  We show a histogram computed from all delays (left) and only from delays larger
+  than 15ms (right).}
+\label{fig:drawing-hist}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\noindent
+A summarized view of the delays is provided in Figure~\ref{fig:drawing-hist}, which shows
+histograms illustrating the distribution of delays for each of the three evaluation methods.
+A large proportion of delays is very small (less than 15ms) because the parser used in our
+experimental environment often fails (e.g.~for unclosed parentheses). The histogram
+on the right summarizes only delays for edit operations where the delay for at least one of
+the strategies was over 15ms. A more robust parser would be able to recover and the delays
+for call-by-value and lazy strategies would be even more significant.
+
+\subsubsection{Summary of results}
+The purpose of our empirical is not to exactly assess the overhead of our implementation
+of the live evaluation strategy algorithm. Instead, our goal is to
+illustrate how often can previously evaluated results be reused and the impact this has when
+writing code. The experiment presented in this section is small-scale, but it is sufficient
+to illustrate the main point.
+
+When recomputing results after every edit operation
+using the \emph{call-by-value} strategy, the time needed to update the results grows
+continually. The \emph{lazy} strategy removes the overhead for programs that fail to evaluate, but
+keeps the same trend. Using our \emph{live} strategy, the computation can reuse values computed
+previously and, as a result, expensive operations (such as (d) and (g) in Figure~\ref{fig:drawing})
+are significantly faster, because they do not need to recompute operations done previously when
+writing the code. As shown in Figure~\ref{fig:drawing-hist}, there are almost no very expensive
+operations (taking over 1 second) in the \emph{live} strategy in contrast to several in the
+other two strategies.
+
+% --------------------------------------------------------------------------------------------------
 
@@
 \label{sec:case}
+
+http://gallery.thegamma.net/75/gender-pay-gap
+http://gallery.thegamma.net/73/libraries
+http://gallery.thegamma.net/47/unemployment-in-italy
+http://gallery.thegamma.net/44/classification-of-the-functions-of-government-cofog
+http://gallery.thegamma.net/37/top-10-government-departments-by-201516-expenditure
 

### 2019-02-07 wip-live-programming/paper.tex
Adding case study and language extensions

@@
   illustrate the practicality of the mechanism using an empirical evaluation
-  (Section~\ref{sec:evaluation-empirical}) and a case study (Section~\ref{sec:case}).
+  (Section~\ref{sec:evaluation-empirical}) and a case study (Section~\ref{sec:evaluation-case}).
 
@@
   that relies on the same core infrastructure as live previews and enables richer user experience
-  through \emph{type providers} (Section~\ref{sec:types}). Second, we add an abstraction mechanism
-  that preserves the live coding experience (Section~\ref{sec:design}).
+  through \emph{type providers} (Section~\ref{sec:extra-types}). Second, we add an abstraction mechanism
+  that preserves the live coding experience (Section~\ref{sec:extra-abstraction}).
 \end{itemize}
@@
     preview (right).}
-  \label{first-figure-with-subfigures}
+  \label{fig:gamma-intro}
 \end{figure}
@@
 The practical implementation work that complements the theory presented in this paper implements a live
-coding environment for The Gamma and is discussed in Section~\ref{sec:case} in detail. Earlier work by
+coding environment for The Gamma and is discussed in Section~\ref{sec:evaluation-case} in detail. Earlier work by
 \citet{gamma} discussed The Gamma language and the type provider mechanism, but the implementation
@@
 
-We initially present the data exploration calculus without a static type system. The live preview
-mechanism can be discussed equally well without types. We consider type checking later in
-Section~\ref{sec:types} as the mechanism that is used for live previews can also be
-used for incremental type-checking.
+We initially present the data exploration calculus without a static type system as the live preview
+mechanism does not rely on types. We consider type checking later in Section~\ref{sec:extra-types}
+as the mechanism that is used for live previews can also be used for instantaneous type checking.
 
@@
 force the evaluation of the surrounding expression, capture (a number of) values passed as inputs
-to the lambda function and display a preview based on those. Finally, in Section~\ref{sec:design},
+to the lambda function and display a preview based on those. Finally, in Section~\ref{sec:extra-abstraction},
 we also consider a more speculative design for an abstraction mechanism that supports live previews
@@
 
-\section{Evaluating correctness and effectiveness}
+\section{Evaluating live coding environment}
 \label{sec:evaluation}
 
-The use of a dependency graph when evaluating previews makes it possible to cache partial results
-through a mechanism discussed in Section~\ref{sec:formal-deps} and Section~\ref{sec:previews-cache}.
+Evaluating previews using a dependency graph makes it possible to cache partial results.
 The mechanism satisfies two properties. First, if we evaluate a preview using dependency
@@
 effectiveness, both theoretically (Section~\ref{sec:evaluation-reuse}) and empirically
-(Section~\ref{sec:evaluation-empirical}).
+(Section~\ref{sec:evaluation-empirical}). Finally, we describe a case study where we used the
+methods outlined in the paper to develop a larger online service for data exploration
+(Section~\ref{sec:evaluation-case}).
 
@@
 compiler. We run the experiments in Firefox (version 64.0.2, 32bit) on Windows 10
-(1809, 64bit) on a computer with Intel Core i7-7660U CPU and 16Gb RAM.
+(build 1809, 64bit) on a computer with Intel Core i7-7660U CPU and 16Gb RAM.
 
@@
 
-\newpage
-
-\section{Case study / implementation}
-\label{sec:case}
-
-http://gallery.thegamma.net/75/gender-pay-gap
-http://gallery.thegamma.net/73/libraries
-http://gallery.thegamma.net/47/unemployment-in-italy
-http://gallery.thegamma.net/44/classification-of-the-functions-of-government-cofog
-http://gallery.thegamma.net/37/top-10-government-departments-by-201516-expenditure
+\begin{figure}[t]
+\noindent
+\hspace{-1em}
+\includegraphics[scale=0.245]{gallery1.png}
+\includegraphics[scale=0.245]{gallery2.png}
+\caption{\small Data analysis counting the number of libraries per county in the UK. We load
+  and aggregate data (left) and then create a chart with a label (right).
+  The example is adapted from \url{http://gallery.thegamma.net/73/}.}
+\label{fig:gallery}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Transparent tools for data journalism}
+\label{sec:evaluation-case}
+
+In Section~\ref{sec:background-jupyter}, we motivated our work by considering how journalists
+explore open data. In addition to the theoretical and experimental work presented in this paper,
+we also implemented an online data exploration environment, equipped with live editor for The
+Gamma language that provides instananeous feedback during coding. The environment uses the
+principles presented in this paper to build a more comprehensive system that allows users,
+such as journalists, to analyse, summarize and visualize open data. In this section, we briefly
+report on our experience with the system.
+
+The environment is available at \url{http://gallery.thegamma.net}. Two screenshots shown
+in Figure~\ref{fig:gallery} illustrate a number of interesting features:
+
+\begin{itemize}[itemsep=3pt]
+\item The left screenshot uses a type provider for data aggregation introduced by \citet{gamma}
+  and briefly discussed in Secton~\ref{sec:background-gamma}. For the purpose of
+  this paper, the type provider is an external library (with objects, members and reduction relation).
+  Using the type provider relies on editor auto-complete which is based on type information. To support
+  this, we implement type checking, as discussed in Section~\ref{sec:extra-types}, reusing the mechanism
+  of live preview evaluation.
+
+\item The live preview for code written using the data agregation type provider (left) is shown
+  in a panel below the current command. This shows the preview for a sub-expression based on the
+  position of the cursor (i.e.~raw data before grouping in the screenshot). In addition to the
+  preview, our environment also generates tabs that let the user navigate to individual steps of
+  the data transformation for code using the data aggregation type provider.
+
+\item The right screenshot uses a charting library, which is another external library available in
+  the environment. Unlike for data aggregation (left), live previews for charts (right) are not
+  handled in a special way and we show the result of evaluating the whole command. The screenshot
+  shows a case where the user modifies parameters of the chart. Thanks to our live evaluation
+  strategy, this is done efficiently without reevaluating the data transformation.
+\end{itemize}
+
+\noindent
+Conducting an extensive user study that would evaluate the usability of the system from a
+human-computer ineraction perspective is left for future work. To give anecdotal evidence of
+the usability, the example shown in Figure~\ref{fig:gallery} is an analysis developed by an
+attendee of our tutorial at Mozilla Festival 2017 who had no prior programming experience.
 
@@
 
-\section{Type checking}
-\label{sec:types}
-
-Evaluating live previews can be an expensive operations, so being able to cache partial previews
-is a must for a live coding environment. Type checking is typically fast, so the main focus of
-this paper is on live previews. However, asynchronous type providers in The Gamma (Section~\ref{sec:types-providers})
-can make type checking time consuming, and so we use the dependency graph also for type checking
-(Section~\ref{sec:types-graph}). Type checking lambda functions (Section~\ref{sec:types-funs})
-requires a slight extension of the model discussed in Section~\ref{sec:formal}.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Asynchronously provided types}
-\label{sec:types-providers}
-
-Data available in The Gamma can be defined using several kinds of type providers. The type provider
-used in Figure~\ref{fig:thegamma} as asynchronous \cite{async}. It downloads the sample URL and generates
-types based on the contents of the web page. The parameter to \ident{web.scrape} is a static
-parameter and is evaluated during type-checking. We omit details in this paper, but we note this
-works similarly to F\# type providers \cite{providers-fsharp}.
-
-Type providers can also be implemented as REST services \cite{rest} to allow anyone implement a data source
-in the language of their choice. In this case, each member of a call-chain returns a type that is
-generated based on the result of an HTTP request. For example, when the user types \ident{worldbank}
-(to access information about countries), the type provider makes a request to
-\url{http://thegamma-services.azurewebsites.net/worldbank}, which returns two members:
-%
-\begin{equation*}
-\begin{array}{l}
-[~\{\rstr{name}\!:\str{byYear},\rstr{returns}\!:\\
-~\quad \{\rstr{kind}\!:\str{nested},\rstr{endpoint}\!:\str{/pickYear}\}\}\\
-\hspace{0.55em}\{\rstr{name}\!:\str{byCountry},\rstr{returns}\!:\\
-~\quad \{\rstr{kind}\!:\str{nested},\rstr{endpoint}\!:\str{/pickCountry}\}\}~]
-\end{array}
-\end{equation*}
-%
-This indicates that \ident{worldbank} has members \ident{byYear} and \ident{byCountry}.
-If the user types \ident{worldbank.byCountry}, a request is made to the specified URL
-\url{http://thegamma-services.azurewebsites.net/worldbank/pickCountry}:
-%
-\begin{equation*}
-\begin{array}{l}
-[~\{\rstr{name}\!:\str{Andorra},\rstr{trace}\!:[\str{"country=AR"}], \\
-\quad \rstr{returns}\!:\{\rstr{kind}\!:\str{nested},\rstr{endpoint}\!:\str{/pickTopic}\}\}\\
-\hspace{0.55em}\{\rstr{name}\!:\str{Afghanistan},\rstr{trace}\!:[\str{"country=AF"}],\\
-\quad \rstr{returns}\!:\{\rstr{kind}\!:\str{nested},\rstr{endpoint}\!:\str{/pickTopic}\}\},\,..]
-\end{array}
-\end{equation*}
-%
-This returns a list of countries which can then be accessed as members via
-\ident{worldbank.byCountry.Andorra}, etc.
-
-This is one reason for why type checking in The Gamma can be time consuming. Other
-type providers may perform other more computationally intensive work to provide types and so
-it is desirable to reuse type-checking results during live coding. The rest of this section
-shows how this is done using the dependency graph discussed in Section~\ref{sec:formal}.
-
-\begin{figure}[!b]
+\begin{figure}
 \vspace{-0.5em}
 \begin{equation*}
+\hspace{3em}
 \xymatrix{
-&\bnd{val}(o)\\
-&\bnd{mem}(m, s_0) \ar[u]_{\blbl{arg}(0)} \ar[dl]^{\blbl{arg}(1)} \\
-\bnd{fun}(x, s_2) \ar@/^/[uur]^{\blbl{callsite}(m, 1)} \ar[rr]^{\blbl{body}} && \bnd{var}(x, s_1) \ar@/_/[uul]_{\blbl{callsite}(m, 1)}
+\bnd{mem}(m, s_0) \ar@/_1pc/[drr]_{\blbl{arg}(1)} \ar[rr]^{\blbl{arg}(0)} && \bnd{val}(o) && \bnd{var}(x, s_1)\ar[ll]_{\blbl{callsite}(m, 1)} \\
+&&\bnd{fun}(x, s_2) \ar[u]_{\blbl{callsite}(m, 1)} \ar@/_1pc/[rru]_{\blbl{body}} &&
 }
 \end{equation*}
-\caption{Dependency graph for $o.m(\lambda x\rightarrow x)$}
+\vspace{-0.5em}
+\caption{Dependency graph for $o.m(\lambda x\rightarrow x)$ with a newly added \blbl{callsite} edges.}
 \label{fig:graph-func}
+\vspace{-0.5em}
 \end{figure}
 
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure*}[t]
-\vspace{-0.5em}
+% ==================================================================================================
+
+\section{Extending data exploration calculus}
+\label{sec:extra}
+
+The data exploration calculus presented in Section~\ref{sec:calculus} allowed us to formalise the
+live preview mechanism, but it lacks many important features. We outline two extensions.
+First, we discuss how to add instantaneous type checking by using the dependency graph construction
+described in Section~\ref{sec:formal}. Second, we consider an abstraction mechanism that preserves
+the live coding experience.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Type checking}
+\label{sec:extra-types}
+
+Instantaneous live previews give analysts quick feedback when they write incorrect code, but
+having type information is still valuable. First, it can help give better error messages. Second,
+types can be used to provide auto-complete -- when the user types `.' we can offer available
+members without having to evaluate the wait until the value of the object is available.
+
+\subsubsection{Revised dependency graph}
+Type checking of small programs is typically fast enough that no caching is necessary. However,
+The Gamma supports \emph{type providers} \cite{providers-fsharp,providers-idris}, which can
+generate types based on an external file or a REST service call, e.g.~\cite{fsdata}. For this
+reason, type checking can be relatively time consuming and can benefit from the same caching
+facilities as those available for live previews.
+
+Adding type checking requires revising the way we construct the dependency graph introduced in
+Section~\ref{sec:formal}. Previously, a variable bound by a lambda function had no dependencies.
+However, the type of the variable depends on the context in which it appears. Given an expression
+$o.m(\lambda x\rightarrow x)$, we infer the type of $x$ from the type of the first argument
+of the member $m$. A variable node for $x$ thus needs to depend on the call site of $m$.
+We capture that by adding an edge $\blbl{callsite}(m, i)$ from $x$ to $o$ which indicates that
+$x$ is the input variable of a function passes as the $i^\textnormal{th}$ argument to the $m$
+member of the expression represented by the target node. We also add $\blbl{callsite}(m, i)$
+as an edge from the node of the function. Figure~\ref{fig:graph-func} shows the
+revised dependency graph for $o.m(\lambda x\rightarrow x)$. A revised version of the algorithm
+that constructs the dependency graph is included in Appendix~\ref{sec:app-types}.
+
+\subsubsection{Type checking}
+The structure of typing rules is similar to the evaluation relation $\bndclr{v}\Downarrow d$
+defined earlyer. Given a dependency graph $(V, E)$, we define typing judgements in the form
+$\bndclr{v}\vdash \tau$. The type $\tau$ can be a primitive type, a function $\tau \rightarrow \tau$
+or an object type $\{m_1\!:\!\sigma_1, \ldots, m_n\!:\!\sigma_n\}$ with member types
+$\sigma = (\tau_1, \ldots, \tau_n) \rightarrow \tau$.
+
+The typing rules for variables, functions and member access are shown in Figure~\ref{fig:tc}.
+Remaining rules are available in Appendix~\ref{sec:app-types}.
+When type checking a member access \rname{mem}, we find its dependencies $\bndclr{v_i}$ and
+check that the instance is an object with the required member $m$. The types of arguments of the
+member then need to match the types of the remaining (non-instance) nodes.
+Type checking a function \rname{fun} and a variable \rname{var} is similar. In both cases,
+we follow the \blbl{callsite} edge to find the member that accepts the function as an argument
+and use the type of the argument to check the type of the function or infer the type of the
+variable.
+
+The results of type checking can be cached and reused in the same way as live previews, although
+we leave out the details. A property aking to correctness (Theorem~\ref{thm:correcntess})
+requires defining standard type checking over the structure of expressions, which we also
+omit for space reasons.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}
 \begin{equation*}
-\begin{array}{ll}
-\ident{bind-expr}_{\Gamma, \Delta, c}(e_0.m(e_1, \ldots, e_n)) = &(2b)\\
-\qquad \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n)\\
-\quad \textnormal{when}~\bndclr{v_0}, (V_0, E_0) = \ident{bind-expr}_{\Gamma, \Delta, \bot}(e_0)\\
-\quad \textnormal{and}~c_i = (\bndclr{v_0}, \blbl{callsite}(m,i))\hspace{4.8em} (i \in 1\ldots n)\\
-\quad \textnormal{and}~\bndclr{v_i}, (V_i, E_i) = \ident{bind-expr}_{\Gamma, \Delta, c_i}(e_i)\qquad (i\in 1\ldots n)\\
-\quad \textnormal{and}~(\bknd{mem}(m),[(\bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v_n}, \blbl{arg}(n))]) \notin \ident{dom}(\Delta)\hspace{-200em}.\\
-\quad \textnormal{let}~\bndclr{v} = \bnd{mem}(m, s), s~\textnormal{fresh}\\
-\quad \textnormal{let}~E = \{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v},\bndclr{v_n}, \blbl{arg}(n)) \}
-\end{array}
-\qquad\qquad
-\begin{array}{ll}
-\ident{bind-expr}_{\Gamma, \Delta, (\bndclr{v_c}, \blblclr{l_c} )}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V_0, E \cup E_0)&(7b)\\
-\quad \textnormal{when}~(\bknd{var}(x),[(\bndclr{v_c}, \blblclr{l_c} )])\notin \ident{dom}(\Delta)\\
-\quad \textnormal{let}~\bndclr{v_x} = \bnd{var}(x, s_x), s_x~\textnormal{fresh}\\
-\quad \textnormal{and}~\Gamma_1 = \Gamma \cup \{ x, \bndclr{v_x} \}\\
-\quad \textnormal{and}~\bndclr{v_0}, (V_0, E_0) = \ident{bind-expr}_{\Gamma_1, \Delta, \bot}(e)\\
-\quad \textnormal{when}~(\bknd{fun}(x),[(\bndclr{v_0}, \blbl{body}), (\bndclr{v_c}, \blblclr{l_c} )])\notin \ident{dom}(\Delta)\\
-\quad \textnormal{let}~\bndclr{v} = \bnd{fun}(x, s_f), s_f~\textnormal{fresh}\\
-\quad \textnormal{let}~E = \{ (\bndclr{v},\bndclr{v_0},\blbl{body}), (\bndclr{v},\bndclr{v_c},\blblclr{l_c}), (\bndclr{v_x},\bndclr{v_c},\blblclr{l_c}) \}
-\\[0.75em]
-\end{array}
-\end{equation*}
-\vspace{-0.5em}
-\caption{Revised binding rules, tracking call sites of function values.}
-\label{fig:binding-rules-callsite}
-\vspace{-0.5em}
-\end{figure*}
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Revised binding of functions}
-\label{sec:types-funs}
-
-The Gamma script supports lambda functions, but only in a limited way. A function can be passed
-as a parameter to a method, which makes type checking of functions easier. For example, consider:
-%
-\begin{equation*}
+\hspace{-3.5em}
 \begin{array}{l}
-\ident{movies}.\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())
-\end{array}
-\end{equation*}
-%
-If \ident{movies} is a collection of \ident{Movie} objects, the type of the lambda function must be
-$\ident{Movie}\rightarrow\ident{bool}$ and so the type of $x$ is \ident{Movie}. This is similar to
-type checking of lambda functions in C\# \cite{csharp}, where type is also inferred from the context (or has
-to be specified explicitly).
-
-We do not currently allow lambda functions as stand-alone let-bound
-values. This could be done by requiring explicit types, or introducing polymorphism, but it was not
-necessary for the limited domain of non-expert data exploration.
-
-\paragraph{Dependency graph for functions.}
-In the binding process specified in Section~\ref{sec:formal}, a variable is a leaf of the dependency
-graph. In the revised model, it depends on the context in which it appears. A new edge labeled
-$\blbl{callsite}(m, i)$ indicates that the source node is the input variable of a function
-passed as the $i^\textnormal{th}$ argument to the $m$ member of the expression represented by the
-target node. A node representing function is linked to the call site using the same edge.
-
-Figure~\ref{fig:graph-func} shows the result of binding $o.m(\lambda x\rightarrow x)$. Both
-$\bnd{fun}(x, s_2)$ and $\bnd{var}(x, s_1)$ now depend on the node representing $o$. The new
-$\blbl{callsite}$ edge makes it possible to type-check function and variable nodes just using their
-dependencies. As before, the member invocation $\bnd{mem}(m, s_0)$
-depends on the instance using $\blbl{arg}(0)$ and on its argument using $\blbl{arg}(1)$.
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure*}
-\vspace{0.5em}
-\begin{equation*}
-\begin{array}{l}
-\qquad\inference[\rname{val}~]
-  {\Sigma(n) = \alpha}
-  {\bnd{val}(n) \vdash \alpha }
-\quad\qquad
+\inference[\rname{mem}~]
+  { \forall i\in\{0\ldots k\}.(\bnd{mem}(m, s), \bndclr{v_i}, \blbl{arg}(i)) \in E &
+  \bndclr{v_0} \vdash \{.., m:(\tau_1, \ldots, \tau_k) \rightarrow \tau, ..\} & \bndclr{v_i} \vdash \tau_i }
+  { \bnd{mem}(m, s) \vdash \tau }
+\\[1.5em]
 \inference[\rname{var}~]
   { (\bnd{var}(x,s), \bndclr{v}, \blbl{callsite}(m, i)) \in E &
-    \bndclr{v} \vdash (.., m:(\tau_1, \ldots, \tau_k)\rightarrow \tau,..) & \tau_i=\tau'\rightarrow\tau'' }
+    \bndclr{v} \vdash \{.., m:(\tau_1, \ldots, \tau_k)\rightarrow \tau,..\} & \tau_i=\tau'\rightarrow\tau'' }
   { \bnd{var}(x, s) \vdash \tau' }
-\\[2.5em]
+\\[1.5em]
 \inference[\rname{fun}~]
-  { \{(\bnd{fun}(x,s), \bndclr{v_b}, \blbl{body}), (\bnd{var}(x,s), \bndclr{v_c}, \blbl{callsite}(m, i))\} \subseteq E &
-    \bndclr{v_c} \vdash (.., m:(\tau_1, \ldots, \tau_k)\rightarrow \tau,..) &
+  { \{(\bnd{fun}(x,s), \bndclr{v_b}, \blbl{body}), (\bnd{var}(x,s), \bndclr{v_c}, \blbl{callsite}(m, i))\} \subseteq E \\
+    \bndclr{v_c} \vdash \{.., m:(\tau_1, \ldots, \tau_k)\rightarrow \tau,..\} &
     \tau_i=\tau'\rightarrow\tau'' & \bndclr{v_b} \vdash \tau'' }
   { \bnd{fun}(x, s) \vdash \tau' \rightarrow \tau'' }
-\\[2.5em]
-\qquad\qquad\quad\inference[\rname{mem}~]
-  { \forall i\in\{0\ldots k\}.(\bnd{mem}(m, s), \bndclr{v_i}, \blbl{arg}(i)) \in E &
-    \bndclr{v_0} \vdash (.., m:(\tau_1, \ldots, \tau_k) \rightarrow \tau, ..)& \bndclr{v_i} \vdash \tau_i }
-  { \bnd{mem}(m, s) \vdash \tau }
-\\[1em]
 \end{array}
 \end{equation*}
-\caption{Rules that define evaluation of previews over a dependency graph for an expression}
-\vspace{0.5em}
+\caption{Rules that define type checking of terms and expressions over a dependency graph $(V, E)$}
 \label{fig:tc}
-\end{figure*}
-
-% --------------------------------------------------------------------------------------------------
-
-\paragraph{Revised binding process.}
-For the revised binding process, we introduce a new edge label $\blbl{callsite}$. Variable nodes
-now have dependencies and so we cache them and attach a symbol $s$ to \bnd{var}, so we also
-introduce a node kind $\bknd{var}$ as part of a lookup key for $\Delta$.
-
-The \ident{bind-expr} function now has a parameter $c$, in addition to $\Gamma$ and $\Delta$, which
-represents the context in which the binding happens. This is either a member invocation (labeled
-with instance node $\bndclr{v}$ and \blbl{callsite} label $\blblclr{l}$), or not a member
-invocation written as $\bot$. The updated definitions are:
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Feedback-friendly abstraction}
+\label{sec:extra-abstraction}
+
+The sample data analysis by Financial Times in Section~\ref{sec:background-jupyter} illustrates why
+notebook users often avoid abstraction. Wrapping code in a function would make it impossible to
+split code into cells, and see results of intermediate steps. Instead, the sample analysis used a
+global variable at the top of the notebook with possible values in a comment.
+
+Providing live previews inside ordinary functions is problematic, because we do not have readily
+available values for input parameters and our mechanism for lambda functions only provides
+delayed previews inside body of a function. We believe that extending the data exploration calculus
+with an abstraction mechanism that would support development with instantaneous feedaback is an
+interesting design problem and we briefly outline a possible solution here.
+
+Data scientists often write code interactively using a sample data set and, when it works well,
+wrap it into a function that they then call on other data. Similarly, spreadsheet users
+often write equation in the first row of a table and then use the ``drag down'' operation to
+apply it to other rows. One way of adding similar functionality to the data exploration calculus
+is to label a sequence of commands such that the sequence can be reused later with different
+inputs:
 %
 \begin{equation*}
-\begin{array}{lcll}
-\bndclr{v} &\hspace{-0.5em}\in\hspace{-0.5em}&\bnd{val}(n)~|~\bnd{var}(x, s)~|~\bnd{mem}(m, s)~|~\bnd{fun}(x, s)&(\textit{Vertices})\\
-\blblclr{l}&\hspace{-0.5em}\in\hspace{-0.5em}&\blbl{body}~|~\blbl{arg}(i)~|~\blbl{callsite}(m,i)&(\textit{Edge labels})\\
-\bkndclr{k}&\hspace{-0.5em}\in\hspace{-0.5em}&\bknd{fun}(x)~|~\bknd{mem}(m)~|~\bknd{var}(x)&(\textit{Node kinds})\\
-          c&\hspace{-0.5em}\in\hspace{-0.5em}&\bot ~|~(\bndclr{v}, \blblclr{l})&(\textit{Call sites})\\
+\begin{array}{rcl}
+p &=& c_1; \ldots; c_n\\
+c &=& \kvd{let}\,x=t \lsep t \lsep \textit{lbl}\!:\,p \lsep \textit{lbl}
 \end{array}
 \end{equation*}
-%
-The key parts of the revised definition of the \ident{bind-expr} function are shown in Figure~\ref{fig:binding-rules-callsite}.
-We now write $\ident{bind-expr}_{\Gamma,\Delta,c}$ where $c$ represents the context in which the
-binding occurs. This is set to $\bot$ in all cases, except when binding arguments of a member call.
-In (2b), we first recursively bind the instance node using $\bot$ as the context and then
-bind all the arguments using $(\bndclr{v_0}, \blbl{callsite}(m, i))$ as the context for $i^{\textnormal{th}}$
-argument. The rest is as in case (2) before. The case (1) is updated similarly and is not shown
-here for brevity.
-
-When binding a function (7b), we now also store variable nodes in $\Delta$ and so we check if a
-variable with the given call site exists. If no, we create a fresh node $\bnd{var}(x, s_x)$.
-The node is added to $\Gamma$ as before. At the end, we now also include a call site edge from the
-variable node and from the function node in $E$. We omit a similar variant of the case (6).
-The remaining cases (3)-(5) are the same, except that \ident{bind-expr} has the additional $c$ parameter
-and recursive calls always set it to $\bot$.
-
-Finally, the \ident{update} function in Figure~\ref{fig:loop} also needs to be updated to store the
-newly created \bnd{var} nodes. This is done by adding the following single case:
-
-\begin{equation*}
-\begin{array}{l}
-\Delta_{i}(\bknd{var}(x), [(\bndclr{v},\blbl{callsite}(m,i))]) = \bnd{var}(x, s)\\
-\quad\quad \textnormal{for all}~\bnd{var}(x, s) \in V\\
-\quad\quad \textnormal{such that}~(\bnd{var}(x, s), \bndclr{v}, \blbl{callsite}(m,i)) \in E
-\end{array}
-\end{equation*}
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Type checking over dependency graphs}
-\label{sec:types-graph}
-
-The type system for The Gamma supports a number of primitive types (such as integers and strings)
-written as $\alpha$. Composed types include functions and objects with members. Objects are be
-provided by type providers, but we omit the details here. Types of object members are written as
-$\sigma$ and can have multiple arguments:
-%
-\begin{equation*}
-\begin{array}{lcll}
-\tau&\in&\alpha ~|~ \tau \rightarrow \tau ~|~ \{m_1\!:\!\sigma_1, \ldots, m_n\!:\!\sigma_n\} &\quad(\textit{Types})\\
-\sigma &\in& (\tau_1, \ldots, \tau_n) \rightarrow \tau &\quad(\textit{Members})\\
-\end{array}
-\end{equation*}
-%
-The typing judgements are written in the form $\bnd{v}\vdash \tau$. They are parameterised by the
-dependency graph $(V, E)$, but this is not modified during type checking so we keep it implicit
-rather than writing e.g.~$\bnd{v}\vdash_{(V,E)} \tau$.
-
-\paragraph{Type checking.}
-The typing rules are shown in Figure~\ref{fig:tc}. Types of primitive values $n$ are obtained
-using a global lookup table $\Sigma$ \rname{val}. When type checking a member call \rname{mem},
-we find its dependencies $\bndclr{v_i}$ and check that the first one (instance) is an object with
-the required member $m$. The types of input arguments of the member then need to match the types
-of the remaining (non-instance) nodes.
-
-Type checking a function \rname{fun} and a variable \rname{var} is similar. In both cases,
-we follow the \blbl{callsite} edge to find the member that accepts the function as an argument.
-We obtain the type of the function from the type of the $i^\textnormal{th}$ argument of the
-member. We use the input type as the type of variable \rname{var}. For functions, we also check
-that the resulting type matches the type of the body \rname{fun}.
-
-\paragraph{Caching results.}
-Performing type checking over the dependency graph, rather than over the abstract syntax tree,
-enables us to reuse the results of previously type checked parts of a program. As when
-caching evaluated previews (Section~\ref{sec:previews}), we build a lookup table mapping graph
-nodes to types. When type checking a node, we first check the cache and, only if it is new,
-follow the $\vdash$ relation to obtain the type.
-
-As a result, code can be type checked on-the-fly during editing, even when asynchronous type
-providers are used, and the programmer gets instant feedback without delays.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Properties of type checking}
-
-As noted in Section~\ref{sec:types}, the focus of this paper is on live previews, but we also
-use the method based on reusing nodes in a dependency graph for type checking. We do not discuss
-properties of type checking in detail, but we briefly note how the different properties of
-live previews extend to corresponding properties of type checking.
-
-\paragraph{Type checking result reuse.} In Section~\ref{sec:evaluation-reuse}, we show that
-certain source code edits do not cause the recomputation of previews for the whole expression
-or a sub-expression. The edits are given in Figure~\ref{fig:operations}. The proof uses the fact
-that the newly bound graph (after code edit) reuses nodes of the previous graph. This implies that
-type checking results can be reused in exactly the same way as live previews -- they are also
-stored in a lookup table with graph nodes as keys.
-
-\paragraph{Correctness.} The correctness property (Theorem~\ref{thm:correctness}) shows that
-graph-based preview evaluation matches direct evaluation of expressions. To show a corresponding
-property for type checking, we would need to provide ordinary type system based on the structure
-of the expression and prove that the two are equivalent. In our implementation, we only use the
-presented graph-based type checking method, so we do not provide an alternate account in this paper.
-
-\paragraph{Determinacy.} The determinacy property (Theorem~\ref{thm:determinacy}) guarantees that
-previews can be cached, because evaluating them again, using $\Downarrow$ over an updated graph,
-would yield the same result. The same property holds for $\vdash$, meaning that type checking
-results can be cached. Although the Theorem~\ref{thm:determinacy} talks explicitly about
-$\Downarrow$, it can be easily extended for $\vdash$, because the proof depends on how the graph
-is updated using $\ident{update}_{V,E}$ and the binding process.
-
-% ==================================================================================================
-
-\section{Design lessons}
-\label{sec:design}
-
-The motivation for the presented work, briefly outlined in Section~\ref{sec:live}, is to build a
-simple data exploration environment that would allow non-experts, like data journalists,
-transparently work with data. In this paper, we focused on providing live coding experience,
-which is one important step toward the goal. However, the language we use is a mix of established
-object-oriented (member access) and functional (function values) features with type providers.
-
-If we were to design a new programming language, there are lessons we can learn from the
-cases that make type checking and preview evaluation in this paper difficult. This section
-briefly considers those.
-
-\paragraph{Functions and type providers.} When using type providers in a nominally-typed language,
-the provided types are named, but the names are typically not easy to type \cite{fsdata}. This is
-not a problem in typical usage where provided members are accessed via dot. Using the \ident{worldbank}
-example from Section~\ref{sec:types-providers}, we can access population of two countries using:
-%
-\begin{equation*}
-\begin{array}{l}
-\ident{worldbank.byCountry.\textquotesingle United Kingdom\textquotesingle}.\\
-\quad \ident{Indicators.\textquotesingle Population (total)\textquotesingle}\\[0.5em]
-\ident{worldbank.byCountry.\textquotesingle Czech Republic\textquotesingle}.\\
-\quad \ident{Indicators.\textquotesingle Population (total)\textquotesingle}\hspace{9em}
-\end{array}
-\end{equation*}
-%
-However, the fact that the provided types do not have nice names becomes a problem when want to
-extract code to access population into a function:
-%
-\begin{equation*}
-\begin{array}{l}
-\kvd{let}~\ident{getPopulation}~c=\\
-\quad c.\ident{Indicators.\textquotesingle Population (total)\textquotesingle}\hspace{8.5em}
-\end{array}
-\end{equation*}
-%
-Here, the compiler cannot infer the type of $c$ from usage and so we are required to provide a
-type annotation using an automatically generated name.
-
-\paragraph{Functions and live previews.}
-Providing live previews in a language with ordinary functions suffers from the same problem as
-type checking of functions.
-
-Our live preview evaluation, discussed in Section~\ref{sec:previews}, can obtain only a delayed
-preview for the body of \ident{getPopulation}. The delayed preview we would obtain in this case
-is $\llbracket c.\ident{Indicators}.\ident{\textquotesingle Population (total)\textquotesingle} \rrbracket_{c}$.
-
-If we know the type of $c$, we can provide a user interface that lets the user specify a value
-for $c$ (or, more generally, free variables of the preview) and then evaluate the preview, but
-it is difficult to provide a meaningful preview automatically.
-
-\paragraph{Wormhole abstractions.}
-In data science scripting, we start with a concrete example and then turn code into a reusable
-function. This pattern could be supported by the language in a way that makes type checking
-and preview evaluation easier. Using an imaginary notation, we could write:
-%
-\begin{equation*}
-\begin{array}{l}
-\kvd{let}~\ident{uk}~=~\ident{worldbank.byCountry.\textquotesingle United Kingdom\textquotesingle}\hspace{1.5em}\\[0.5em]
-\kvd{def}~\ident{getPopulation}~=\\
-\quad\lbrack\ident{country}\!:\!\ident{uk}\rbrack.\ident{Indicators.\textquotesingle Population (total)\textquotesingle}\\[0.5em]
-\ident{getPopulation}~\ident{worldbank.byCountry.China}\\
-\ident{getPopulation}~\ident{worldbank.byCountry.India}
-\end{array}
-\end{equation*}
-%
-We tentatively call this notation \emph{wormhole} abstraction and we intend to implement it in future prototypes of The Gamma.
-The second line is an expression that accesses the population of the UK, using a concrete
-data source as the input, but it also defines a named function \ident{getPopulation} that has a
-parameter \ident{country}. In a way, we are providing \emph{type annotation} by example, together
-with a \emph{value annotation} that can be used for live previews.
-
-This way of constructing abstractions is perhaps more akin to how spreadsheets are used -- we often
-write a formula using a concrete cell and then use the ``drag down'' operation to extend it to
-other inputs.
+
+\noindent
+As before, a program is a sequence of commands and a command can be either a let binding or a term.
+Additionally, a command can be a labelled sequence of commands or a reference to a label. When
+evaluating, the command $\textit{lbl}$ is replaced with the associated sequence of commands $p$
+before any other reduction occurs. Consequently, variables used in the labelled block are
+dynamically scoped and we can use let binding to redefine a value of a variable before invoking
+the block repeatedly. The correct use of dynamically scoped variables can be checked using
+coeffects \cite{coeffects}.
+
+This minimalistic abstraction mechanism supports code reuse without affecting how live previews
+are computed. The commands in the labelled block still require all let bound variables to be
+defined before the block. Those define sample data for development and can be later redefined before
+reusing the block. This functionality is not yet available in the data exploration environment
+discussed in Section~\ref{sec:evaluation-case}, but we intend to add it in a future version.
 
@@
 
-\paragraph{Design and human-computer interaction.}
+\subsubsection*{Design and human-computer interaction.}
 TODO: \citet{principle}
@@
 \cite{slicing} would allow us to deduce possible inputs for functions and use those for
-previews rather than changing the language as suggested in Section~\ref{sec:design}. This direction
+previews rather than changing the language as suggested in Section~\ref{sec:extra-abstraction}. This direction
 is worth considering, but it requires more empirical usability testing.
@@
 % ==================================================================================================
-
-\section{TODO}
-
-TBD: Define live preview somewhere
-
-TBD: In section about dependency graphs, include a subsection discussing how this relates
-to incremental computation (just like "Semantics of delayed previews" links that to CMTT etc.)
-
-\cite{selfadjusting}
-\cite{incremental}
-
-Incremental computation
-
-  Self-adjusting computation
-  http://www.cs.cmu.edu/~rwh/theses/acar.pdf
-
-  Incremental Computation with Names
-  https://www.cs.tufts.edu/~jfoster/papers/oopsla15.pdf
-  -> dsl interpreter
-
-  (and other adapton stuff http://adapton.org/)
-
-Type checking
-
-  co-contextual
-  http://www.informatik.uni-marburg.de/~seba/publications/cocontextual-type-checking.pdf
-  http://drops.dagstuhl.de/opus/volltexte/2017/7262/pdf/LIPIcs-ECOOP-2017-18.pdf
-
-https://dl.acm.org/citation.cfm?id=664109
-
+%
+% \section{TODO}
+%
+% TBD: Define live preview somewhere
+%
+% TBD: In section about dependency graphs, include a subsection discussing how this relates
+% to incremental computation (just like "Semantics of delayed previews" links that to CMTT etc.)
+%
+%
+% \cite{selfadjusting}
+% \cite{incremental}
+%
+% Incremental computation
+%
+%   Self-adjusting computation
+%   http://www.cs.cmu.edu/~rwh/theses/acar.pdf
+%
+%   Incremental Computation with Names
+%   https://www.cs.tufts.edu/~jfoster/papers/oopsla15.pdf
+%   -> dsl interpreter
+%
+%   (and other adapton stuff http://adapton.org/)
+%
+% Type checking
+%
+%   co-contextual
+%   http://www.informatik.uni-marburg.de/~seba/publications/cocontextual-type-checking.pdf
+%   http://drops.dagstuhl.de/opus/volltexte/2017/7262/pdf/LIPIcs-ECOOP-2017-18.pdf
+%
+% https://dl.acm.org/citation.cfm?id=664109
+%
 \section{Summary}
@@
 
+\section{Appenidx - Type checking}
+\label{sec:app-types}
+
+tbd
+
+% Evaluating live previews can be an expensive operations, so being able to cache partial previews
+% is a must for a live coding environment. Type checking is typically fast, so the main focus of
+% this paper is on live previews. However, asynchronous type providers in The Gamma (Section~\ref{sec:types-providers})
+% can make type checking time consuming, and so we use the dependency graph also for type checking
+% (Section~\ref{sec:types-graph}). Type checking lambda functions (Section~\ref{sec:types-funs})
+% requires a slight extension of the model discussed in Section~\ref{sec:formal}.
+%
+% % --------------------------------------------------------------------------------------------------
+%
+% \subsection{Asynchronously provided types}
+% \label{sec:types-providers}
+%
+% Data available in The Gamma can be defined using several kinds of type providers. The type provider
+% used in Figure~\ref{fig:thegamma} as asynchronous \cite{async}. It downloads the sample URL and generates
+% types based on the contents of the web page. The parameter to \ident{web.scrape} is a static
+% parameter and is evaluated during type-checking. We omit details in this paper, but we note this
+% works similarly to F\# type providers \cite{providers-fsharp}.
+%
+% Type providers can also be implemented as REST services \cite{rest} to allow anyone implement a data source
+% in the language of their choice. In this case, each member of a call-chain returns a type that is
+% generated based on the result of an HTTP request. For example, when the user types \ident{worldbank}
+% (to access information about countries), the type provider makes a request to
+% \url{http://thegamma-services.azurewebsites.net/worldbank}, which returns two members:
+% %
+% \begin{equation*}
+% \begin{array}{l}
+% [~\{\rstr{name}\!:\str{byYear},\rstr{returns}\!:\\
+% ~\quad \{\rstr{kind}\!:\str{nested},\rstr{endpoint}\!:\str{/pickYear}\}\}\\
+% \hspace{0.55em}\{\rstr{name}\!:\str{byCountry},\rstr{returns}\!:\\
+% ~\quad \{\rstr{kind}\!:\str{nested},\rstr{endpoint}\!:\str{/pickCountry}\}\}~]
+% \end{array}
+% \end{equation*}
+% %
+% This indicates that \ident{worldbank} has members \ident{byYear} and \ident{byCountry}.
+% If the user types \ident{worldbank.byCountry}, a request is made to the specified URL
+% \url{http://thegamma-services.azurewebsites.net/worldbank/pickCountry}:
+% %
+% \begin{equation*}
+% \begin{array}{l}
+% [~\{\rstr{name}\!:\str{Andorra},\rstr{trace}\!:[\str{"country=AR"}], \\
+% \quad \rstr{returns}\!:\{\rstr{kind}\!:\str{nested},\rstr{endpoint}\!:\str{/pickTopic}\}\}\\
+% \hspace{0.55em}\{\rstr{name}\!:\str{Afghanistan},\rstr{trace}\!:[\str{"country=AF"}],\\
+% \quad \rstr{returns}\!:\{\rstr{kind}\!:\str{nested},\rstr{endpoint}\!:\str{/pickTopic}\}\},\,..]
+% \end{array}
+% \end{equation*}
+% %
+% This returns a list of countries which can then be accessed as members via
+% \ident{worldbank.byCountry.Andorra}, etc.
+%
+% This is one reason for why type checking in The Gamma can be time consuming. Other
+% type providers may perform other more computationally intensive work to provide types and so
+% it is desirable to reuse type-checking results during live coding. The rest of this section
+% shows how this is done using the dependency graph discussed in Section~\ref{sec:formal}.
+%
+% \begin{figure}[!b]
+% \vspace{-0.5em}
+% \begin{equation*}
+% \xymatrix{
+% &\bnd{val}(o)\\
+% &\bnd{mem}(m, s_0) \ar[u]_{\blbl{arg}(0)} \ar[dl]^{\blbl{arg}(1)} \\
+% \bnd{fun}(x, s_2) \ar@/^/[uur]^{\blbl{callsite}(m, 1)} \ar[rr]^{\blbl{body}} && \bnd{var}(x, s_1) \ar@/_/[uul]_{\blbl{callsite}(m, 1)}
+% }
+% \end{equation*}
+% \caption{Dependency graph for $o.m(\lambda x\rightarrow x)$}
+% \label{fig:graph-func}
+% \end{figure}
+%
+% % --------------------------------------------------------------------------------------------------
+%
+% \begin{figure*}[t]
+% \vspace{-0.5em}
+% \begin{equation*}
+% \begin{array}{ll}
+% \ident{bind-expr}_{\Gamma, \Delta, c}(e_0.m(e_1, \ldots, e_n)) = &(2b)\\
+% \qquad \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n)\\
+% \quad \textnormal{when}~\bndclr{v_0}, (V_0, E_0) = \ident{bind-expr}_{\Gamma, \Delta, \bot}(e_0)\\
+% \quad \textnormal{and}~c_i = (\bndclr{v_0}, \blbl{callsite}(m,i))\hspace{4.8em} (i \in 1\ldots n)\\
+% \quad \textnormal{and}~\bndclr{v_i}, (V_i, E_i) = \ident{bind-expr}_{\Gamma, \Delta, c_i}(e_i)\qquad (i\in 1\ldots n)\\
+% \quad \textnormal{and}~(\bknd{mem}(m),[(\bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v_n}, \blbl{arg}(n))]) \notin \ident{dom}(\Delta)\hspace{-200em}.\\
+% \quad \textnormal{let}~\bndclr{v} = \bnd{mem}(m, s), s~\textnormal{fresh}\\
+% \quad \textnormal{let}~E = \{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v},\bndclr{v_n}, \blbl{arg}(n)) \}
+% \end{array}
+% \qquad\qquad
+% \begin{array}{ll}
+% \ident{bind-expr}_{\Gamma, \Delta, (\bndclr{v_c}, \blblclr{l_c} )}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V_0, E \cup E_0)&(7b)\\
+% \quad \textnormal{when}~(\bknd{var}(x),[(\bndclr{v_c}, \blblclr{l_c} )])\notin \ident{dom}(\Delta)\\
+% \quad \textnormal{let}~\bndclr{v_x} = \bnd{var}(x, s_x), s_x~\textnormal{fresh}\\
+% \quad \textnormal{and}~\Gamma_1 = \Gamma \cup \{ x, \bndclr{v_x} \}\\
+% \quad \textnormal{and}~\bndclr{v_0}, (V_0, E_0) = \ident{bind-expr}_{\Gamma_1, \Delta, \bot}(e)\\
+% \quad \textnormal{when}~(\bknd{fun}(x),[(\bndclr{v_0}, \blbl{body}), (\bndclr{v_c}, \blblclr{l_c} )])\notin \ident{dom}(\Delta)\\
+% \quad \textnormal{let}~\bndclr{v} = \bnd{fun}(x, s_f), s_f~\textnormal{fresh}\\
+% \quad \textnormal{let}~E = \{ (\bndclr{v},\bndclr{v_0},\blbl{body}), (\bndclr{v},\bndclr{v_c},\blblclr{l_c}), (\bndclr{v_x},\bndclr{v_c},\blblclr{l_c}) \}
+% \\[0.75em]
+% \end{array}
+% \end{equation*}
+% \vspace{-0.5em}
+% \caption{Revised binding rules, tracking call sites of function values.}
+% \label{fig:binding-rules-callsite}
+% \vspace{-0.5em}
+% \end{figure*}
+%
+% % --------------------------------------------------------------------------------------------------
+%
+% \subsection{Revised binding of functions}
+% \label{sec:types-funs}
+%
+% The Gamma script supports lambda functions, but only in a limited way. A function can be passed
+% as a parameter to a method, which makes type checking of functions easier. For example, consider:
+% %
+% \begin{equation*}
+% \begin{array}{l}
+% \ident{movies}.\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())
+% \end{array}
+% \end{equation*}
+% %
+% If \ident{movies} is a collection of \ident{Movie} objects, the type of the lambda function must be
+% $\ident{Movie}\rightarrow\ident{bool}$ and so the type of $x$ is \ident{Movie}. This is similar to
+% type checking of lambda functions in C\# \cite{csharp}, where type is also inferred from the context (or has
+% to be specified explicitly).
+%
+% We do not currently allow lambda functions as stand-alone let-bound
+% values. This could be done by requiring explicit types, or introducing polymorphism, but it was not
+% necessary for the limited domain of non-expert data exploration.
+%
+% \paragraph{Dependency graph for functions.}
+% In the binding process specified in Section~\ref{sec:formal}, a variable is a leaf of the dependency
+% graph. In the revised model, it depends on the context in which it appears. A new edge labeled
+% $\blbl{callsite}(m, i)$ indicates that the source node is the input variable of a function
+% passed as the $i^\textnormal{th}$ argument to the $m$ member of the expression represented by the
+% target node. A node representing function is linked to the call site using the same edge.
+%
+% Figure~\ref{fig:graph-func} shows the result of binding $o.m(\lambda x\rightarrow x)$. Both
+% $\bnd{fun}(x, s_2)$ and $\bnd{var}(x, s_1)$ now depend on the node representing $o$. The new
+% $\blbl{callsite}$ edge makes it possible to type-check function and variable nodes just using their
+% dependencies. As before, the member invocation $\bnd{mem}(m, s_0)$
+% depends on the instance using $\blbl{arg}(0)$ and on its argument using $\blbl{arg}(1)$.
+%
+% % --------------------------------------------------------------------------------------------------
+%
+% \begin{figure*}
+% \vspace{0.5em}
+% \begin{equation*}
+% \begin{array}{l}
+% \qquad\inference[\rname{val}~]
+%   {\Sigma(n) = \alpha}
+%   {\bnd{val}(n) \vdash \alpha }
+% \quad\qquad
+% \inference[\rname{var}~]
+%   { (\bnd{var}(x,s), \bndclr{v}, \blbl{callsite}(m, i)) \in E &
+%     \bndclr{v} \vdash (.., m:(\tau_1, \ldots, \tau_k)\rightarrow \tau,..) & \tau_i=\tau'\rightarrow\tau'' }
+%   { \bnd{var}(x, s) \vdash \tau' }
+% \\[2.5em]
+% \inference[\rname{fun}~]
+%   { \{(\bnd{fun}(x,s), \bndclr{v_b}, \blbl{body}), (\bnd{var}(x,s), \bndclr{v_c}, \blbl{callsite}(m, i))\} \subseteq E &
+%     \bndclr{v_c} \vdash (.., m:(\tau_1, \ldots, \tau_k)\rightarrow \tau,..) &
+%     \tau_i=\tau'\rightarrow\tau'' & \bndclr{v_b} \vdash \tau'' }
+%   { \bnd{fun}(x, s) \vdash \tau' \rightarrow \tau'' }
+% \\[2.5em]
+% \qquad\qquad\quad\inference[\rname{mem}~]
+%   { \forall i\in\{0\ldots k\}.(\bnd{mem}(m, s), \bndclr{v_i}, \blbl{arg}(i)) \in E &
+%     \bndclr{v_0} \vdash (.., m:(\tau_1, \ldots, \tau_k) \rightarrow \tau, ..)& \bndclr{v_i} \vdash \tau_i }
+%   { \bnd{mem}(m, s) \vdash \tau }
+% \\[1em]
+% \end{array}
+% \end{equation*}
+% \caption{Rules that define evaluation of previews over a dependency graph for an expression}
+% \vspace{0.5em}
+% \label{fig:tc}
+% \end{figure*}
+%
+% % --------------------------------------------------------------------------------------------------
+%
+% \paragraph{Revised binding process.}
+% For the revised binding process, we introduce a new edge label $\blbl{callsite}$. Variable nodes
+% now have dependencies and so we cache them and attach a symbol $s$ to \bnd{var}, so we also
+% introduce a node kind $\bknd{var}$ as part of a lookup key for $\Delta$.
+%
+% The \ident{bind-expr} function now has a parameter $c$, in addition to $\Gamma$ and $\Delta$, which
+% represents the context in which the binding happens. This is either a member invocation (labeled
+% with instance node $\bndclr{v}$ and \blbl{callsite} label $\blblclr{l}$), or not a member
+% invocation written as $\bot$. The updated definitions are:
+% %
+% \begin{equation*}
+% \begin{array}{lcll}
+% \bndclr{v} &\hspace{-0.5em}\in\hspace{-0.5em}&\bnd{val}(n)~|~\bnd{var}(x, s)~|~\bnd{mem}(m, s)~|~\bnd{fun}(x, s)&(\textit{Vertices})\\
+% \blblclr{l}&\hspace{-0.5em}\in\hspace{-0.5em}&\blbl{body}~|~\blbl{arg}(i)~|~\blbl{callsite}(m,i)&(\textit{Edge labels})\\
+% \bkndclr{k}&\hspace{-0.5em}\in\hspace{-0.5em}&\bknd{fun}(x)~|~\bknd{mem}(m)~|~\bknd{var}(x)&(\textit{Node kinds})\\
+%           c&\hspace{-0.5em}\in\hspace{-0.5em}&\bot ~|~(\bndclr{v}, \blblclr{l})&(\textit{Call sites})\\
+% \end{array}
+% \end{equation*}
+% %
+% The key parts of the revised definition of the \ident{bind-expr} function are shown in Figure~\ref{fig:binding-rules-callsite}.
+% We now write $\ident{bind-expr}_{\Gamma,\Delta,c}$ where $c$ represents the context in which the
+% binding occurs. This is set to $\bot$ in all cases, except when binding arguments of a member call.
+% In (2b), we first recursively bind the instance node using $\bot$ as the context and then
+% bind all the arguments using $(\bndclr{v_0}, \blbl{callsite}(m, i))$ as the context for $i^{\textnormal{th}}$
+% argument. The rest is as in case (2) before. The case (1) is updated similarly and is not shown
+% here for brevity.
+%
+% When binding a function (7b), we now also store variable nodes in $\Delta$ and so we check if a
+% variable with the given call site exists. If no, we create a fresh node $\bnd{var}(x, s_x)$.
+% The node is added to $\Gamma$ as before. At the end, we now also include a call site edge from the
+% variable node and from the function node in $E$. We omit a similar variant of the case (6).
+% The remaining cases (3)-(5) are the same, except that \ident{bind-expr} has the additional $c$ parameter
+% and recursive calls always set it to $\bot$.
+%
+% Finally, the \ident{update} function in Figure~\ref{fig:loop} also needs to be updated to store the
+% newly created \bnd{var} nodes. This is done by adding the following single case:
+%
+% \begin{equation*}
+% \begin{array}{l}
+% \Delta_{i}(\bknd{var}(x), [(\bndclr{v},\blbl{callsite}(m,i))]) = \bnd{var}(x, s)\\
+% \quad\quad \textnormal{for all}~\bnd{var}(x, s) \in V\\
+% \quad\quad \textnormal{such that}~(\bnd{var}(x, s), \bndclr{v}, \blbl{callsite}(m,i)) \in E
+% \end{array}
+% \end{equation*}
+%
+% % --------------------------------------------------------------------------------------------------
+%
+% \subsection{Type checking over dependency graphs}
+% \label{sec:types-graph}
+%
+% The type system for The Gamma supports a number of primitive types (such as integers and strings)
+% written as $\alpha$. Composed types include functions and objects with members. Objects are be
+% provided by type providers, but we omit the details here. Types of object members are written as
+% $\sigma$ and can have multiple arguments:
+% %
+% \begin{equation*}
+% \begin{array}{lcll}
+% \tau&\in&\alpha ~|~ \tau \rightarrow \tau ~|~ \{m_1\!:\!\sigma_1, \ldots, m_n\!:\!\sigma_n\} &\quad(\textit{Types})\\
+% \sigma &\in& (\tau_1, \ldots, \tau_n) \rightarrow \tau &\quad(\textit{Members})\\
+% \end{array}
+% \end{equation*}
+% %
+% The typing judgements are written in the form $\bnd{v}\vdash \tau$. They are parameterised by the
+% dependency graph $(V, E)$, but this is not modified during type checking so we keep it implicit
+% rather than writing e.g.~$\bnd{v}\vdash_{(V,E)} \tau$.
+%
+% \paragraph{Type checking.}
+% The typing rules are shown in Figure~\ref{fig:tc}. Types of primitive values $n$ are obtained
+% using a global lookup table $\Sigma$ \rname{val}. When type checking a member call \rname{mem},
+% we find its dependencies $\bndclr{v_i}$ and check that the first one (instance) is an object with
+% the required member $m$. The types of input arguments of the member then need to match the types
+% of the remaining (non-instance) nodes.
+%
+% Type checking a function \rname{fun} and a variable \rname{var} is similar. In both cases,
+% we follow the \blbl{callsite} edge to find the member that accepts the function as an argument.
+% We obtain the type of the function from the type of the $i^\textnormal{th}$ argument of the
+% member. We use the input type as the type of variable \rname{var}. For functions, we also check
+% that the resulting type matches the type of the body \rname{fun}.
+%
+% \paragraph{Caching results.}
+% Performing type checking over the dependency graph, rather than over the abstract syntax tree,
+% enables us to reuse the results of previously type checked parts of a program. As when
+% caching evaluated previews (Section~\ref{sec:previews}), we build a lookup table mapping graph
+% nodes to types. When type checking a node, we first check the cache and, only if it is new,
+% follow the $\vdash$ relation to obtain the type.
+%
+% As a result, code can be type checked on-the-fly during editing, even when asynchronous type
+% providers are used, and the programmer gets instant feedback without delays.
+%
+% % --------------------------------------------------------------------------------------------------
+%
+% \subsection{Properties of type checking}
+%
+% As noted in Section~\ref{sec:types}, the focus of this paper is on live previews, but we also
+% use the method based on reusing nodes in a dependency graph for type checking. We do not discuss
+% properties of type checking in detail, but we briefly note how the different properties of
+% live previews extend to corresponding properties of type checking.
+%
+% \paragraph{Type checking result reuse.} In Section~\ref{sec:evaluation-reuse}, we show that
+% certain source code edits do not cause the recomputation of previews for the whole expression
+% or a sub-expression. The edits are given in Figure~\ref{fig:operations}. The proof uses the fact
+% that the newly bound graph (after code edit) reuses nodes of the previous graph. This implies that
+% type checking results can be reused in exactly the same way as live previews -- they are also
+% stored in a lookup table with graph nodes as keys.
+%
+% \paragraph{Correctness.} The correctness property (Theorem~\ref{thm:correctness}) shows that
+% graph-based preview evaluation matches direct evaluation of expressions. To show a corresponding
+% property for type checking, we would need to provide ordinary type system based on the structure
+% of the expression and prove that the two are equivalent. In our implementation, we only use the
+% presented graph-based type checking method, so we do not provide an alternate account in this paper.
+%
+% \paragraph{Determinacy.} The determinacy property (Theorem~\ref{thm:determinacy}) guarantees that
+% previews can be cached, because evaluating them again, using $\Downarrow$ over an updated graph,
+% would yield the same result. The same property holds for $\vdash$, meaning that type checking
+% results can be cached. Although the Theorem~\ref{thm:determinacy} talks explicitly about
+% $\Downarrow$, it can be easily extended for $\vdash$, because the proof depends on how the graph
+% is updated using $\ident{update}_{V,E}$ and the binding process.
+%
 \end{document}

### 2019-02-07 wip-live-programming/experiments/script.R

@@ -3,11 +3,54 @@ library(ggplot2)
 library(reshape2)
 
 dr <- read.csv("C:/Tomas/Research/TheGamma/wip-live-programming/experiments/drawing.csv")
-names(dr) <- c("full", "live")
+names(dr) <- c("call-by-value", "lazy", "live")
 dr$time <- seq.int(nrow(dr))
 
 drm <- melt(dr,id = c("time"))
 names(drm) <- c("time", "method", "delay")
 
+a1 <- arrow(type = "open", angle = 90, length = unit(0.15, "inches"))
+a2 <- arrow(type = "open", angle = 90, length = unit(0.3, "inches"))
+
 ggplot(drm,aes(x = time,y = delay)) + 
-  geom_bar(aes(fill = method),stat = "identity",position = "dodge")
+  geom_bar(aes(fill = method),stat = "identity",position = "dodge") +
+  scale_fill_manual("Evaluation method", values = c("lazy" = "#aec7e8", "call-by-value" = "#1f77b4", "live" = "#ff7f0e")) +
+  xlab("Token (number)") + ylab("Delay (ms)") +
+  annotate("text", x = 37.5, y = 3250, label = c("(h) let")) +
+  geom_segment(aes(x = 37.5, y = 3100, xend = 37.5, yend = 2900), size=1, arrow = a2) +
+  annotate("text", x = 32.5, y = 3250, label = c("(g) combine")) +
+  geom_segment(aes(x = 32.5, y = 3100, xend = 32.5, yend = 2900), size=1, arrow = a2) +
+  annotate("text", x = 22, y = 2550, label = c("(f) combine")) +
+  geom_segment(aes(x = 22, y = 2400, xend = 22, yend = 1950), size=1, arrow = a1) +
+  annotate("text", x = 20, y = 2350, label = c("(e) let")) +
+  geom_segment(aes(x = 20, y = 2200, xend = 20, yend = 2050), size=1, arrow = a1) +
+  annotate("text", x = 15.5, y = 2400, label = c("(d) blur")) +
+  geom_segment(aes(x = 15.5, y = 2250, xend = 15.5, yend = 2000), size=1, arrow = a2) +
+  annotate("text", x = 12, y = 1150, label = c("(c) blur")) +
+  geom_segment(aes(x = 12, y = 1000, xend = 12, yend = 750), size=1, arrow = a1) +
[…]

### 2019-02-10 wip-live-programming/paper.tex
Related work

@@
 code typically resort to notebook systems such as Jupyter or R Markdown \cite{jupyter,rmarkdown}.
-Those are text-based, but have a limited model of recomputation. Users have to structure code in
-cells and manually reevaluate cells after a change. Those systems also do not track
-dependencies between cells, which can lead to inconsistencies; an issue that has been addressed
-by e.g.~\citet{dataflow} or \citet{wrattler}.
+Those are text-based, but have a limited model of recomputation. Users structure code in
+cells and manually reevaluate cells. Many notebook systems do not track dependencies
+between cells, which can lead to inconsistencies; an issue that has been addressed
+by e.g.~\citet{dataflow,noworkflow} or \citet{wrattler}.
 
@@
 \begin{figure}
+\vspace{-0.5em}
 \begin{equation*}
@@
 \end{equation*}
+\vspace{-0.5em}
 \caption{Rules that define type checking of terms and expressions over a dependency graph $(V, E)$}
 \label{fig:tc}
+\vspace{-0.5em}
 \end{figure}
@@
 
-The sample data analysis by Financial Times in Section~\ref{sec:background-jupyter} illustrates why
-notebook users often avoid abstraction. Wrapping code in a function would make it impossible to
-split code into cells, and see results of intermediate steps. Instead, the sample analysis used a
-global variable at the top of the notebook with possible values in a comment.
+The data analysis by Financial Times in Section~\ref{sec:background-jupyter} illustrates why
+notebook users often avoid abstraction. Wrapping code in a function makes it impossible to
+split code into cells and see results of intermediate steps. Instead, the analysis used a
+global variable with possible values in a comment.
 
@@
 \noindent
-As before, a program is a sequence of commands and a command can be either a let binding or a term.
-Additionally, a command can be a labelled sequence of commands or a reference to a label. When
-evaluating, the command $\textit{lbl}$ is replaced with the associated sequence of commands $p$
-before any other reduction occurs. Consequently, variables used in the labelled block are
+We introduce two new kinds of commands: a labelled sequence of commands and a referenceto a label.
+When evaluating, the command $\textit{lbl}$ is replaced with the associated sequence of
+commands $p$ before any other reductions. Consequently, variables used in the labelled block are
 dynamically scoped and we can use let binding to redefine a value of a variable before invoking
-the block repeatedly. The correct use of dynamically scoped variables can be checked using
+the block repeatedly. The correct use of dynamic scoping can be checked using
 coeffects \cite{coeffects}.
@@
 This minimalistic abstraction mechanism supports code reuse without affecting how live previews
-are computed. The commands in the labelled block still require all let bound variables to be
-defined before the block. Those define sample data for development and can be later redefined before
-reusing the block. This functionality is not yet available in the data exploration environment
-discussed in Section~\ref{sec:evaluation-case}, but we intend to add it in a future version.
+are computed. Commands in a labelled block require variables to be defined before the block.
+Those define sample data for development and can be redefined before reusing the block. We
+intend to implement this mechanism in a future version of our data
+exploration environment (Section~\ref{sec:evaluation-case}).
 
@@
 
-This paper approaches the problem of live coding environments from a theoretical programming
-language perspective with a special focus on tooling for data science. Hence, the related and
-future work spans numerous areas.
-
-\subsubsection*{Design and human-computer interaction.}
-TODO: \citet{principle}
-
-From a design perspective, the idea of live programming environments has been popularised by
-Bret Victor \cite{learnable}. Active research on novel forms of interaction happens in areas
-such as live coded music \cite{beyond,sonic}.
-The idea of live previews can be extended to direct manipulation \cite{direct}. The Gamma
-provides limited support for directly manipulating data (Section~\ref{sec:live}), but we
-intend to explore this direction further.
-
-\paragraph{Data science tooling.}
-An essential tool in data science is REPL (read-eval-print-loop) \cite{drscheme}, which is now
-widely available. This has been integrated with rich graphical outputs in tools such as Jupyter
-notebooks \cite{jupyter,ipython}, but such previews are updated using an explicit command.
-Integrating our work with Jupyter to provide instant live previews for R or Python would be
-an interesting extension of the presented work.
-
-\paragraph{Live coding and live previews.}
-Live previews have been implemented in LightTable \cite{lighttable} and, more recently, in
-editors such as Chrome Developer Tools, but neither presents a simple description of their
-inner workings. An issue that attracts much attention is keeping state during code edits
-\cite{alive,livingit}. This would be an interesting problem if we extended our work to
-event-based reactive programming.
-
-\paragraph{Structured editing.}
-An alternative approach to ours is to avoid using text editors. Structured editors
-\cite{structure-based} allow the user to edit the AST and could, in principle, recompute previews
-based on the performed operations, or preview evaluation as in interactive functional programming
-\cite{interactive}. A promising direction is using bi-directional lambda calculus~\cite{hazelnut}.
-Finally, abandoning text also enables building richer, more human-centric
-abstractions as illustrated by Subtext~\cite{subtext}. Our current focus, however, remains
-on text-based editors.
-
-\paragraph{Dependency analysis.}
-Our use of dependency graphs \cite{dependencies} is first-order. Building dependency graphs
-involving function calls using modern compiler methods \cite{optimizing} or program slicing
-\cite{slicing} would allow us to deduce possible inputs for functions and use those for
-previews rather than changing the language as suggested in Section~\ref{sec:extra-abstraction}. This direction
-is worth considering, but it requires more empirical usability testing.
-
-\paragraph{Semantics and partial evaluation.}
+We aim to make creating transparent data analyses, such as those done by journalists, easier and we
+present a tool that provides instantaneous feedback when writing code in a text editor. Many our
+design decisions have alternatives that have been explored in a rich body of literature.
+
+\subsubsection*{Notebooks and data science tools.}
+Basic feedback mechanism, the REPL (read-eval-print-loop), originated in Lisp
+\cite{lisp} and is now common in many languages \cite{drscheme}. Notebook systems
+such as Jupyter and R markdown \cite{rmarkdown,jupyter,ipython} structure code in cells,
+but still rely on explicitly executing individual cells using a stateful REPL environment.
+Integrating provenance tracking with notebooks \cite{dataflow,noworkflow,wrattler} addresses
+the reproducibility issues of this model, but keeps cells as units of work. Ideas such as
+dependency tracking and efficient recomputation exist in many visual data exploration tools
+\cite{control,tableau,vizdom} and scientific workflow systems \cite{taverna,kepler}. Some of the
+implementation techniques are related to our work, but we focus on text-based scripts.
+Tempe \cite{tempe} focuses on streaming data, but provides a text-based scripting environment
+with automatic code update and \citet{ripple} empirically evaluate its usability in contrast to
+REPLs. The challenges faced by Tempe are more akin to dynamic software update \cite{dsu}.
+
+\subsubsection*{Live coding and editing.}
+Live coding based on textual programs has been popularised by \citet{principle,learnable} and
+is actively developed in domains such as live coded music \cite{beyond,sonic}. \citet{liveroad}
+reviews the use of live coding in a Smalltalk derived environment. Lighttable by \citet{lighttable}
+and Chrome DevTools provide limited live previews akin to those presented in this paper,
+but without well specified recomputation model. Finally, work on keeping state during
+code edits \cite{alive,livingit} would be relevant for supporting streaming data. In the future,
+we also plan to explore how direct manipulation \cite{direct} could be integrated with our preview
+mechanism. \citet{liveroad}
+
+Using structured editors \cite{structure-based} rather than text editors is a promising alternative
+direction as editors can recompute program based on understanding of edit operations
+\cite{interactive,livenut,lamdu}. This can be elegantly implemented using bi-directional
+lambda calculus \cite{hazelnut}, but it also makes us consider more human-centric abstractions
+\cite{subtext,directprog} along the lines of the one outlined in Section~\ref{sec:extra-abstraction}.
+
+\subsubsection*{Incremental computation and dependency analysis.}
+Work on self-adjusting and incremental computation \cite{selfadjusting,adapton} handles
+recomputation when the program stays the same, but input changes. It relies on dependency tracking
+similar to ours. We rely on static code analysis whereas most incremental systems,
+e.g.~\cite{incremental,adaptivefp,deltaml} evaluate the program and use programmer-supplied
+information to build a dependency graph. \citet{adapton} implement a small adaptive interpreter
+that treats code as changing input data, suggesting an interesting implementation technique for
+live coding systems. Our use of dependency graphs \cite{dependencies} is static and first-order
+and can be seen as a form of program slicing \cite{slicing}, although our binding process is
+more directly inspired by Roslyn \cite{roslyn}, which uses it for efficient background type-checking.
+
+\subsubsection*{Semantics and partial evaluation.}
 The evaluation of previews can be seen as a form of partial evaluation \cite{partial}, done in a

### 2019-02-10 wip-live-programming/paper.tex
Spellchecking

@@
 Many visual data exploration tools support interactivity \cite{control,tableau,vizdom} and some,
-can even export the workflow as a script \cite{wrangler}, but data anlysts who prefer working with
+can even export the workflow as a script \cite{wrangler}, but data analysts who prefer working with
 code typically resort to notebook systems such as Jupyter or R Markdown \cite{jupyter,rmarkdown}.
@@
 
-As an example, the anlysis done by the Financial Times for an article on plastic waste
+As an example, the analysis done by the Financial Times for an article on plastic waste
 \cite{ftnotebooks,ftarticle} joins datasets from Eurostat, UN Comtrade and more, aggregates the data and
@@
 \item The code relies on external libraries. Our example uses Pandas \cite{pandas},
-  which provides opreations for data wrangling such as \ident{merge} to join datasets
+  which provides operations for data wrangling such as \ident{merge} to join datasets
   or \ident{drop\_duplicates} to delete rows with duplicate column values. Those libraries are
@@
       the theory presented in this paper. The table is produced as
-      the data anlyst types and updates on-the-fly to show result at the current cursor position.
+      the data analyst types and updates on-the-fly to show result at the current cursor position.
     \label{fig:gamma-preview}
@@
 about the data source. The type provider used in this example generates an object with members for
-different data transformations, such as \qident{group data}, which return furhter objects with
+different data transformations, such as \qident{group data}, which return further objects with
 members for specifying transformation parameters, such as selecting the grouping key using
@@
 The Gamma language is richer, but the example in Figure~\ref{fig:gamma-dot} shows that non-trivial
-data exploration can be done using a very simple language. The assumptions about strucutre of
+data exploration can be done using a very simple language. The assumptions about structure of
 code that are explicit in The Gamma are implicitly present in Python and R data analyses
@@
 An example of a live preview is shown in Figure~\ref{fig:gamma-preview}. As noted earlier, The
-Gamma program consits of a list of commands which are either expressions or let bindings. Our
+Gamma program consist of a list of commands which are either expressions or let bindings. Our
 editor displays a live preview below the command that the user is currently editing. The preview
@@
 to edit code in an unrestricted form in a text editor. Although structured editors provide an
-appearling alternative and make recomputation easier, we prefer the transparency
+appealing alternative and make recomputation easier, we prefer the transparency
 and openness of plain text. Second, we focus on the scenario when code changes, but input
@@
 sufficient to capture the simple data analyses discussed in Section~\ref{sec:background}. We
-define the caluclus in this section and then use it as the basis for discussing our live preview
+define the calculus in this section and then use it as the basis for discussing our live preview
 mechanism in Section~\ref{sec:formal}.
@@
 and evaluation contexts; $C_e$ specifies left-to-right evaluation of arguments of a method call,
-$C_c$ specifies evluation of a command and $C_p$ defines top-to-bottom evaluation of a program.
+$C_c$ specifies evaluation of a command and $C_p$ defines top-to-bottom evaluation of a program.
 The rule \rname{external} applies reductions provided by an external library in a
@@
 An external library consists of a set of objects $O$ and a reduction relation
-$\rightsquigarrow_\epsilon$ that satsifies the following two properties:
+$\rightsquigarrow_\epsilon$ that satisfies the following two properties:
 \begin{itemize}
@@
 
-Furthermore, the graph edges are labeled with labels indicating the kind of dependency. For
+Furthermore, the graph edges are labelled with labels indicating the kind of dependency. For
 a method call, the labels are ``first argument'', ``second argument'' and so on. Writing
@@
 contains the member name, together with a unique symbol $s$ -- two member access nodes with
-different dependencies will contain a different symbol. Dependencies of member access are labeled
+different dependencies will contain a different symbol. Dependencies of member access are labelled
 with \blbl{arg} indicating the index of the argument (the instance has index $0$ and arguments
@@
 The example on the second line looks for a node of a kind $\bkndclr{k}$ that has dependencies
-$\bndclr{v_1}, \ldots, \bndclr{v_n}$ labeled with labels $\blblclr{l_1}, \ldots, \blblclr{l_n}$.
+$\bndclr{v_1}, \ldots, \bndclr{v_n}$ labelled with labels $\blblclr{l_1}, \ldots, \blblclr{l_n}$.
 We write $\Delta(\bndclr{k}, l)\!\downarrow$ when a value for a given key is not defined.
@@
 a node $\bndclr{v}$ corresponding to the expression $e$ paired with a dependency graph $(V, E)$.
-In the graph, $V$ is a set of nodes $\bndclr{v}$ and $E$ is a set of labeled edges
+In the graph, $V$ is a set of nodes $\bndclr{v}$ and $E$ is a set of labelled edges
 $(\bndclr{v_1}, \bndclr{v_2}, \blblclr{l})$. We attach the label directly to the edge rather than
@@
 then used as dependencies for lookup into $\Delta$, together with their labels. If a node already
-exists in $\Delta$ it is reused (1). Alternativelym, we create a new node containing a fresh symbol~(2).
+exists in $\Delta$ it is reused (1). Alternatively, we create a new node containing a fresh symbol~(2).
 
@@
 producing two kinds of previews. A \emph{fully evaluated preview} is just a value, while
-a \emph{delayed preview} is a partially evalauted expression with free variables:
+a \emph{delayed preview} is a partially evaluated expression with free variables:
 
@@
 dependency graph are parameters of the $\Downarrow$ relation, but they do not change during the
-evaluation and so we do not explicity write them.
+evaluation and so we do not explicitly write them.
 
@@
 that would allow us to express the splicing operation used in \rname{mem-expr}. In categorical
-terms, the context-dependent proposition can be modeled as an indexed comonad \cite{effectrev,graded}.
+terms, the context-dependent proposition can be modelled as an indexed comonad \cite{effectrev,graded}.
 The evaluation of a preview with no context dependencies (built implicitly into our evaluation rules)
@@
 The of preview-preserving code edit operations includes several ways of introducing a let
-binding, several ways of eliminating a let binding and two edit operations where the data anlyst
+binding, several ways of eliminating a let binding and two edit operations where the data analyst
 modifies an unrelated part of the program. It is worth noting that the list is not exhaustive.
@@
 For the purpose of the evaluation, we use a simple image manipulation library that provides
-operations for loading, greyscaling, bluring and combining images. We compare delays in
+operations for loading, greyscaling, blurring and combining images. We compare delays in
 updating the preview for three different evaluation strategies, while performing the same
@@
 \item Introducing let binding does not cause recomputation when using live strategy.
-\item As in (c), accessing combine member without valud argument only affects call-by-value.
+\item As in (c), accessing combine member without valid argument only affects call-by-value.
 \item Varying parameter of combine in live strategy does not recompute blur and is much faster.
@@
 we also implemented an online data exploration environment, equipped with live editor for The
-Gamma language that provides instananeous feedback during coding. The environment uses the
+Gamma language that provides instantaneous feedback during coding. The environment uses the
 principles presented in this paper to build a more comprehensive system that allows users,
@@
 \item The left screenshot uses a type provider for data aggregation introduced by \citet{gamma}
-  and briefly discussed in Secton~\ref{sec:background-gamma}. For the purpose of
+  and briefly discussed in Section~\ref{sec:background-gamma}. For the purpose of
   this paper, the type provider is an external library (with objects, members and reduction relation).
@@
 
-\item The live preview for code written using the data agregation type provider (left) is shown
+\item The live preview for code written using the data aggregation type provider (left) is shown
   in a panel below the current command. This shows the preview for a sub-expression based on the
@@
 Conducting an extensive user study that would evaluate the usability of the system from a
-human-computer ineraction perspective is left for future work. To give anecdotal evidence of
+human-computer interaction perspective is left for future work. To give anecdotal evidence of
 the usability, the example shown in Figure~\ref{fig:gallery} is an analysis developed by an
@@
 The structure of typing rules is similar to the evaluation relation $\bndclr{v}\Downarrow d$
-defined earlyer. Given a dependency graph $(V, E)$, we define typing judgements in the form
+defined earlier. Given a dependency graph $(V, E)$, we define typing judgements in the form
 $\bndclr{v}\vdash \tau$. The type $\tau$ can be a primitive type, a function $\tau \rightarrow \tau$
@@
 The results of type checking can be cached and reused in the same way as live previews, although
-we leave out the details. A property aking to correctness (Theorem~\ref{thm:correcntess})
+we leave out the details. A property akin to correctness (Theorem~\ref{thm:correcntess})
 requires defining standard type checking over the structure of expressions, which we also
@@
 delayed previews inside body of a function. We believe that extending the data exploration calculus
-with an abstraction mechanism that would support development with instantaneous feedaback is an
+with an abstraction mechanism that would support development with instantaneous feedback is an
 interesting design problem and we briefly outline a possible solution here.
@@
 \noindent
-We introduce two new kinds of commands: a labelled sequence of commands and a referenceto a label.
+We introduce two new kinds of commands: a labelled sequence of commands and a reference to a label.
 When evaluating, the command $\textit{lbl}$ is replaced with the associated sequence of
@@
 
-\section{Appenidx - Type checking}
+\section{Appendix - Type checking}
 \label{sec:app-types}

### 2019-02-11 wip-live-programming/paper.tex
Conclusions and space tweaks

@@
 
-As discussed earlier, our implementation updates $\Delta$ during the recursive binding process and
-so a stronger version of the property holds -- namely, previews calculated over a graph obtained
-directly for the original program $p$ are the same as the values of the fully evaluated program.
-We note that this is the case, but do not show it formally to aid the clarity of our formalisation.
-
-The second important property that guarantees that our mechanism always displays correct previews
-is determinacy. This makes it possible to cache the previews evaluated using~$\Downarrow$ using
-the corresponding graph node as a lookup key.
+\noindent
+Our implementation updates $\Delta$ during the recursive binding process and so a stronger version
+of the property holds: previews calculated over a graph obtained directly for the original program
+$p$ are the same as the values of the fully evaluated program. Our formalisation omits this for
+simplicity.
+
+The second important property is determinacy, which makes it possible to cache the previews
+evaluated using~$\Downarrow$ using the corresponding graph node as a lookup key.
 
@@
 
-The mechanism used for caching previews, as discussed in Section~\ref{sec:previews-cache},
-keeps a preview or a partial preview $d$ in a lookup table indexed by nodes $\bndclr{v}$. The
-Theorem~\ref{thm:determinacy} guarantees that this is a valid strategy. As we update dependency
+\noindent
+The mechanism used for caching previews (Section~\ref{sec:previews-cache}),
+keeps a preview preview $d$ in a lookup table indexed by nodes $\bndclr{v}$.
+Theorem~\ref{thm:determinacy} guarantees that this is a valid -- as we update dependency
 graph during code editing, previous nodes will continue representing the same sub-expressions.
@@
 \raggedright
-\vspace{0.5em}
+\vspace{-0.5em}
 \hspace{1.5em}{\small\sffamily Edit contexts of expressions}
@@
 \label{fig:operations}
+\vspace{-0.5em}
 \end{figure}
@@
 We use the notation $\preview{e}$ to mark the part of the expressions before and after the edit
-that have the same value that will not need to be recomputed after the change and we write
+that have the same value and will not need to be recomputed after the change; we write
 $\overline{c}$ and $\overline{e}$ for a list of commands and expressions, respectively.
-In some of the edit operations, we also specify an intermediate program that is, in some cases,
+In some of the edit operations, we also specify an intermediate program that may be
 semantically different and only has a partial preview. This illustrates a typical way of
-working with code in a text editor using cut and paste operations. For example, in
-\rname{let-intro-ins}, we assume that the analyst cuts a sub-expression $e$, replaces it with
+working with code in a text editor using cut and paste. For example, in
+\rname{let-intro-ins}, we the analyst cuts a sub-expression $e$, replaces it with
 a variable $x$ and then adds a let binding for a variable $x$ and inserts the expression $e$
-from the clipboard. Edit with the same result is captured by \rname{let-intro-del} but here
-the analyst first inserts the let binding and then replaces the sub-expression $e$ with a
-variable $x$.
+from the clipboard. The \rname{let-intro-del} operation captures the same edit, but
+done in a different order.
 
@@
 
-\begin{figure}
+\begin{figure}[b]
 \raggedright
@@
 
-\begin{figure}[b]
+\begin{figure}[t]
 \vspace{-0.5em}
@@
 \begin{figure}
+\vspace{-0.5em}
 \noindent
@@
 \includegraphics[scale=0.48]{hist-slow.png}
+\vspace{-0.25em}
 \caption{Histogram showing the distribution of delays incurred when updating previews.
@@
 \label{fig:drawing-hist}
+\vspace{-0.75em}
 \end{figure}
@@
 \subsubsection{Summary of results}
-The purpose of our empirical is not to exactly assess the overhead of our implementation
-of the live evaluation strategy algorithm. Instead, our goal is to
-illustrate how often can previously evaluated results be reused and the impact this has when
-writing code. The experiment presented in this section is small-scale, but it is sufficient
-to illustrate the main point.
-
-When recomputing results after every edit operation
-using the \emph{call-by-value} strategy, the time needed to update the results grows
-continually. The \emph{lazy} strategy removes the overhead for programs that fail to evaluate, but
-keeps the same trend. Using our \emph{live} strategy, the computation can reuse values computed
-previously and, as a result, expensive operations (such as (d) and (g) in Figure~\ref{fig:drawing})
+The purpose of our experiment is not to exactly assess the overhead of our implementation.
+Our goal is to illustrate how often can previously evaluated results be reused and the impact
+this has when writing code. The experiment presented in this section is small-scale, but it is
+sufficient for this purpose. When recomputing results after every edit
+using the \emph{call-by-value} strategy, the time needed to update results grows
+continually. The \emph{lazy} strategy removes the overhead for programs that fail, but
+keeps the same trend. Our \emph{live} strategy reuses values computed
+previously. Consequently, expensive operations (such as (d) and (g) in Figure~\ref{fig:drawing})
 are significantly faster, because they do not need to recompute operations done previously when
@@
 \label{fig:gallery}
+\vspace{-1em}
 \end{figure}
@@
 \begin{itemize}[itemsep=3pt]
-\item The left screenshot uses a type provider for data aggregation introduced by \citet{gamma}
-  and briefly discussed in Section~\ref{sec:background-gamma}. For the purpose of
-  this paper, the type provider is an external library (with objects, members and reduction relation).
-  Using the type provider relies on editor auto-complete which is based on type information. To support
-  this, we implement type checking, as discussed in Section~\ref{sec:extra-types}, reusing the mechanism
-  of live preview evaluation.
-
-\item The live preview for code written using the data aggregation type provider (left) is shown
-  in a panel below the current command. This shows the preview for a sub-expression based on the
-  position of the cursor (i.e.~raw data before grouping in the screenshot). In addition to the
-  preview, our environment also generates tabs that let the user navigate to individual steps of
-  the data transformation for code using the data aggregation type provider.
-
-\item The right screenshot uses a charting library, which is another external library available in
-  the environment. Unlike for data aggregation (left), live previews for charts (right) are not
-  handled in a special way and we show the result of evaluating the whole command. The screenshot
-  shows a case where the user modifies parameters of the chart. Thanks to our live evaluation
+\item The left screenshot uses a type provider for data aggregation \cite{gamma}
+  (Section~\ref{sec:background-gamma}). Our system views the type provider as an external library
+  (with objects, members and reduction relation). Type providers rely on type information to
+  provide editor auto-complete, which we support by implementing type checking over a
+  dependency graph as discussed in Section~\ref{sec:extra-types}.
+
+\item When displaying live preview for code written using the data aggregation type provider (left),
+  our environment also generates tabes that show individual steps of the data transformation.
+  A tab is selected based on the cursor position and shows preview for the current sub-expression
+  (e.g.~raw data before grouping in the screenshot).
+
+\item Another external library provides support for charting (right). The environment does not
+  handle charts in a special way and displays the result of evaluating the whole command. The
+  screenshot shows a case where the user modifies parameters of the chart. Thanks to our live evaluation
   strategy, this is done efficiently without reevaluating the data transformation.
@@
 \noindent
-Conducting an extensive user study that would evaluate the usability of the system from a
-human-computer interaction perspective is left for future work. To give anecdotal evidence of
-the usability, the example shown in Figure~\ref{fig:gallery} is an analysis developed by an
-attendee of our tutorial at Mozilla Festival 2017 who had no prior programming experience.
+A user study to evaluate the usability of the system from a human-computer interaction perspective
+is left for future work. As an anecdotal evidence of the usability, the code in
+Figure~\ref{fig:gallery} was developed by an attendee of a tutorial at Mozilla Festival 2017
+who had no prior programming experience.
 
 % ==================================================================================================
+
+\section{Extending data exploration calculus}
+\label{sec:extra}
+
+The data exploration calculus presented in Section~\ref{sec:calculus} allowed us to formalise the
+live preview mechanism, but it lacks many important features. We outline two extensions.
+First, we discuss how to add instantaneous type checking by using the dependency graph construction
+described in Section~\ref{sec:formal}. Second, we consider an abstraction mechanism that preserves
+the live coding experience.
+
+% --------------------------------------------------------------------------------------------------
 
@@
 
-% ==================================================================================================
-
-\section{Extending data exploration calculus}
-\label{sec:extra}
-
-The data exploration calculus presented in Section~\ref{sec:calculus} allowed us to formalise the
-live preview mechanism, but it lacks many important features. We outline two extensions.
-First, we discuss how to add instantaneous type checking by using the dependency graph construction
-described in Section~\ref{sec:formal}. Second, we consider an abstraction mechanism that preserves
-the live coding experience.
-
 % --------------------------------------------------------------------------------------------------
@@
 
-\begin{figure}
+\begin{figure}[b]
 \vspace{-0.5em}
@@
 \subsubsection*{Semantics and partial evaluation.}
-The evaluation of previews can be seen as a form of partial evaluation \cite{partial}, done in a
-way that allows reuse of results. This can be done implicitly or explicitly in the form of
-multi-stage programming \cite{metaml}. Both can provide useful perspective for formally analysing
-how previews are evaluated. Semantically, the evaluation of previews can be seen as a modality
-\cite{modal} and delayed previews are linked to contextual modal type theory \cite{cmtt}, which,
-in turn, can be understood in terms of comonads \cite{cmtt-denotation}. This provides an intriguing
-direction for rigorous analysis of the presented system.
+The evaluation of previews is a form of partial evaluation \cite{partial}, done in a
+way that allows caching. This can be done implicitly or explicitly in the form of
+multi-stage programming \cite{metaml}. Semantically, the evaluation of previews can be seen as a
+modality \cite{modal} and delayed previews are linked to contextual modal type theory \cite{cmtt},
+formally modelled using comonads \cite{cmtt-denotation}. This suggests a direction for rigorous
+analysis of the presented system.
 
@@
 \section{Summary}
-We present The Gamma, a live coding environment for data exploration. The environment
-bridges the gap between spreadsheets and scripting -- live previews give users rapid feedback,
-while the final result is a fully reproducible script.
-
-In this paper, we focus on the challenge of efficiently providing live previews and type checking
-code during editing in a free-form text editor. This is a challenge, because users can perform
-arbitrary text transformations and we cannot recompute previews after each edit.
-
-The key trick is to separate the process into fast \emph{binding phase},
-which constructs a dependency graph and slower \emph{evaluation phase} and \emph{type checking phase}
-that can cache results, using the nodes from the dependency graph created during binding as keys.
-This makes it possible to quickly parse updated code, reconstruct dependency graph and compute
-preview using previous, partially evaluated, results.
-
-We describe our approach formally, which serves two purposes. First, we aim to provide easy to
-use foundations for the growing and important trend of text-based live coding environments.
-Second, we explore the properties of our system and prove that our method does not recompute
-previews in a number of common cases and, at the same time, the optimisation still produces
-correct previews.
+One of the reasons that make spreadsheets easier to use than programming tools is that they
+provide instant feedback. We aim to make programming tools as instantaneous as spreadsheets.
+We formalized a number of key aspects of simple data analyses such as those don by journalists
+and then used our observations to build both theory and simple practical data analytics tools.
+
+Our \emph{data exploration calculus} is a simple formally tractable language
+for data exploration. The calculus captures key observations about simple data analyses. They
+rely on logic from external libraries, implement few abstractions and are written as lists of
+commands. We later revised the calculus to add type checking and simple abstraction mechanism.
+
+Our main technical contribution is a \emph{live preview} mechanism that efficiently evaluates
+code during editing and instantly provides a preview of the result. We allow users to edit code
+in unconstrained way in an text editor, which makes this particularly challenging. The key trick
+is to separate the process into fast \emph{binding phase}, which constructs a dependency graph
+and slower \emph{evaluation phase} that can cache results. This makes it possible to quickly parse
+updated code, reconstruct dependency graph and compute preview using previous, partially
+evaluated, results.
+
+We evaluated our approach in three ways. First, we proved that our mechanism is correct and
+that it reuses evaluated values for many common code edit operations. Second, we conducted an
+experimental study that illustrates how often are previously evaluated results reused during
+typical programming scenario. Thirdly, we used our research as a basis for online data exploration
+environment, which shows the practical usability of our work.
 

### 2019-02-22 design/paper.tex
Adding wip on design principles paper

\documentclass{sigchi}

\CopyrightYear{2019}
\setcopyright{rightsretained}
\doi{http://dx.doi.org/10.475/123_4}
\isbn{123-4567-24-567/08/06}
\conferenceinfo{NA}{TBA}
\acmPrice{\$0.00}

\usepackage{xcolor}
\usepackage{balance}
\usepackage{graphics}
\usepackage[T1]{fontenc}
\usepackage{txfonts}
\usepackage{mathptmx}
\usepackage[pdflang={en-US},pdftex]{hyperref}
\usepackage{color}
\usepackage{booktabs}
\usepackage{textcomp}

\usepackage{microtype}
\usepackage{ccicons}

\def\plaintitle{The Gamma}
\def\plainauthor{First Author, Second Author, Third Author,
  Fourth Author, Fifth Author, Sixth Author}
\def\emptyauthor{}
\def\plainkeywords{Authors' choice; of terms; separated; by
  semicolons; include commas, within terms only; required.}
\def\plaingeneralterms{Documentation, Standardization}

\makeatletter
\def\url@leostyle{%
  \@ifundefined{selectfont}{
    \def\UrlFont{\sf}
  }{
    \def\UrlFont{\small\bf\ttfamily}
  }}
\makeatother
\urlstyle{leo}

\def\pprw{8.5in}
\def\pprh{11in}
\special{papersize=\pprw,\pprh}
\setlength{\paperwidth}{\pprw}
\setlength{\paperheight}{\pprh}
\setlength{\pdfpagewidth}{\pprw}
\setlength{\pdfpageheight}{\pprh}

\definecolor{todoColor}{rgb}{0.6,0.0,0.2}
\definecolor{linkColor}{RGB}{6,125,233}

\newcommand{\todo}[1]{\textcolor{todoColor}{[\textbf{todo}: #1]}}

\hypersetup{%
  pdftitle={\plaintitle},
% Use \plainauthor for final version.
%  pdfauthor={\plainauthor},
  pdfauthor={\emptyauthor},
  pdfkeywords={\plainkeywords},
  pdfdisplaydoctitle=true, % For Accessibility
  bookmarksnumbered,
  pdfstartview={FitH},
  colorlinks,
  citecolor=black,
  filecolor=black,
  linkcolor=black,
  urlcolor=linkColor,
  breaklinks=true,
  hypertexnames=false
}
\begin{document}

\title{\plaintitle}

\numberofauthors{1}
\author{%
  \alignauthor{Anonymous Authors\\
    \affaddr{Unseen University}\\
    \affaddr{Discworld}\\
    \email{e-mail address}}\\
}

\maketitle

\begin{abstract}
TBA
\end{abstract}

% \category{H.5.m.}{Information Interfaces and Presentation
%   (e.g. HCI)}{Miscellaneous} \category{See
%   \url{http://acm.org/about/class/1998/} for the full list of ACM
%   classifiers. This section is required.}{}{}
%
% \keywords{\plainkeywords}

\section{Introduction}

some background

\subsection{Example}

put some screenshot here?

\subsection{Our Approach and Contributions}
We present blah blah.
Our contributions are:

\begin{itemize}
\item Novel language design based on design principles
\item Open source implementation on the web with cool use cases
\item Critical evaluation
\end{itemize}

Long-term goal of making journalism better

\section{Related work}

\textbf{Visual tools.}

\textbf{Programming tools.}
Notebooks

\textbf{Journalism.}
Idyll

\section{Overview}

some walkthrough illustrating thegamma with code and screenshots

\newpage

\section{Design principles}
The design of TheGamma is based on two categories of design goals that we identify in this
section. The first group of goals follow the observation that majority of users will not be
expert programmers and can only invest limited amount of time into learning a new tool. The
second group of goals are derived from our focus on journalistic applications and the specific
challenges of this domain.

\subsection{Principles for designing simple language}
Our approach originates in a careful consideration of the challenges faced by users learning
how to use main-stream programming languages for data science scripting. \todo{There must be
some research on learning programming we can cite}

\todo{To me, some of the things below sound kind of trivial. Is that just me, or do we
need more fundamental principles?}

\textbf{A1. Allow learning from examples.} Users of tools such as spreadsheets often learn by
looking at existing problem solutions \todo{Advait's PPIG}. Our design should allow this by
making it possible to inspect and retrace steps used while solving a problem in an existing
application.

\textbf{A2. Allow learning by experimentation.} Another principle of spreadsheets that we want
to keep is the ability to experiment and see results immediately. Our design should allow users
to try invoking an operation or modifying a parameter and quickly see if this leads to the
desired results.

\textbf{A3. Choice over construction.} To minimize the amount of information that users have to
learn and remember, our system should work in a way that allows constructing programs by
choosing from options that can reasonably appear in a current context, rather than requiring
users to recall particular syntax or exact identifier name.
\todo{recognition over recall?}

\textbf{A4. Power for power users.} Some users of the system may, over time, become advanced users
and the system should support those. In other words, the upper bound on what can be achieved
should be well above the most common use cases. \todo{I think I got this idea of "boundaries" on what
is possible from some paper, but cannot recall which...}

\textbf{A5. Minimal viable scenario.} At the same time, the complex features that power users might
need should not affect the most elementary uses of the system and should remain completely hidden
until needed. In other words, the lower bound on what one needs to know to use the system for basic
tasks should be as low as possible.

\todo{I think the above would probably be better if it was just 3 principles. A1+A2 = ``learnability'',
A3 as is, A4+A5 = ``boundaries''. The length of UIST papers also tends to be quite short, so that
might be better for space reasons too.}

\subsection{Principles addressing challenges in journalism}
In addition to consideration based on our focus on non-expert users, we also identify a number
of principles that are specific to our target domain. Following a careful consideration of
current debates in journalism, we identify the following principles. \todo{we should find some
references on trends in journalism to motivate these}

\textbf{B1. Trust through transparency.} The system should allow fact checking of the analyses to
build trust. This means that viewers should be able to determine what is the source of analysed
data and how has the data been transformed.

\textbf{B2. Opening the process to readers.} Journalists are increasingly opening the way they work
to readers in order to build trust. The ability to view how an analysis has been constructed should
not be limited just to experts (say, by running a Jupyter Notebook from GitHub), but should be
available to all interested readers.

\textbf{B3. Providing meaningful engagement mechanism.} Finally, the system should provide a
mechanism through which readers can engage in a meaningful discussion. For example, it should allow
them modify parameters of a data visualization in order to show how, e.g. different choice of
countries affects the result.

\section{Systematic description}

\subsection{Language}

Code should be visible to allow (A1)

auto-complete friendly language allows (A2)

language design addresses (A3, A4) by making easy things easy and hard
things possible (if you know the right syntax)

\subsection{Type providers}

\section{Design principles evaluation}

\section{System evaluation}

\subsection{Something measurable}

\subsection{Case studies}

\subsection{Scalability}

\section{Discussion}

\subsection{Study limitations}
exploratory in nature so we do not make any quantitative claims about effects

not comparing against other systems

\subsection{Design principles}

\subsection{Design issues}
modifying code

\newpage
TODO: Evaluating systems paper

\balance{}
\bibliographystyle{SIGCHI-Reference-Format}
\bibliography{sample}
\end{document}

### 2019-02-22 design/paper.tex
Restructuring design principles

@@
 \newpage
+~
+\newpage
 
 \section{Design principles}
-The design of TheGamma is based on two categories of design goals that we identify in this
-section. The first group of goals follow the observation that majority of users will not be
-expert programmers and can only invest limited amount of time into learning a new tool. The
-second group of goals are derived from our focus on journalistic applications and the specific
-challenges of this domain.
+TheGamma is based on design principles that we identify in this section. We start by careful
+consideration of our target application domain, i.e.~data analyses of open data produced by
+journalists and published by online media. By considering the users, challenges and typical
+scenarios in this domain, we derive more technical design principles for our system.
 
-\subsection{Principles for designing simple language}
-Our approach originates in a careful consideration of the challenges faced by users learning
-how to use main-stream programming languages for data science scripting. \todo{There must be
-some research on learning programming we can cite}
+\subsection{Principles addressing challenges in journalism}
+Modern journalism faces many challenges \todo{cite something}. There are many responses to those
+challenges. To develop a new kind of trust, journalists are increasingly presenting not just
+outcomes of their analysis, but also the process they used. To develop relationship with readers,
+journalists are increasingly looking for meaningful ways of engagement. Those developments
+inform the following design principles of TheGamma system. \todo{rewrite this with
+more references}
 
-\todo{To me, some of the things below sound kind of trivial. Is that just me, or do we
-need more fundamental principles?}
+\textbf{A1. Trust through transparency.} The system should allow fact checking of the analyses to
+build trust. This means that viewers should be able to determine what is the source of analysed
+data and how has the data been transformed.
 
-\textbf{A1. Allow learning from examples.} Users of tools such as spreadsheets often learn by
-looking at existing problem solutions \todo{Advait's PPIG}. Our design should allow this by
-making it possible to inspect and retrace steps used while solving a problem in an existing
-application.
+\textbf{A2. Opening the process.} Journalists are increasingly opening the way they work
+to readers in order to build trust. The ability to view how an analysis has been constructed should
+not be limited just to experts (say, by running a Jupyter Notebook), but should be
+available to all interested readers.
 
-\textbf{A2. Allow learning by experimentation.} Another principle of spreadsheets that we want
-to keep is the ability to experiment and see results immediately. Our design should allow users
-to try invoking an operation or modifying a parameter and quickly see if this leads to the
-desired results.
+\textbf{A3. Providing meaningful engagement mechanism.} The system should provide a
+mechanism through which readers can engage in a meaningful discussion. For example, it should allow
+modifying parameters of a data visualization in order to show how, e.g. different choice of
+countries affects the result.
 
-\textbf{A3. Choice over construction.} To minimize the amount of information that users have to
+\subsection{Principles for open data journalism tool design}
+One important observation from the above list is that our tool should be accessible to non-expert
+users such as readers and non-technical journalists, while providing extra capabilities for more
+technical users. It should be easy to use if one just wants to modify existing code and should
+encourage experimentation. Considering these challenges, we identify the following technical
+design principles. \todo{There must be papers on learning programming that can be referenced
+here.}
+
+\textbf{B1. Learning from examples and by experimentation.} We should support two ways of learning.
+Users of tools such as spreadsheets
+often learn by looking at existing problem solutions \todo{Advait's PPIG}. Our design should allow
+this by making it possible to inspect and retrace steps used while solving a problem in an existing
+application. Another principle of spreadsheets that we want to keep is the ability to experiment
+and see results immediately. Our design should allow users to try invoking an operation or modifying
+a parameter and quickly see if this leads to the desired results.
+
+\textbf{B2. Choice over construction.} To minimize the amount of information that users have to
 learn and remember, our system should work in a way that allows constructing programs by
@@
 
-\textbf{A4. Power for power users.} Some users of the system may, over time, become advanced users
+\textbf{B3. Make simple things easy and complex things possible.} Some users of the system may, over time, become advanced users
 and the system should support those. In other words, the upper bound on what can be achieved
-should be well above the most common use cases. \todo{I think I got this idea of "boundaries" on what
+should be well above the most common use cases. At the same time, the complex features that power users might
+need should not affect the most elementary uses of the system and should remain completely hidden
+until needed. In other words, the lower bound on what one needs to know to use the system for basic
+tasks should be as low as possible. \todo{I think I got this idea of "boundaries" on what
 is possible from some paper, but cannot recall which...}
 
-\textbf{A5. Minimal viable scenario.} At the same time, the complex features that power users might
-need should not affect the most elementary uses of the system and should remain completely hidden
-until needed. In other words, the lower bound on what one needs to know to use the system for basic
-tasks should be as low as possible.
-
-\todo{I think the above would probably be better if it was just 3 principles. A1+A2 = ``learnability'',
-A3 as is, A4+A5 = ``boundaries''. The length of UIST papers also tends to be quite short, so that
-might be better for space reasons too.}
-
-\subsection{Principles addressing challenges in journalism}
-In addition to consideration based on our focus on non-expert users, we also identify a number
-of principles that are specific to our target domain. Following a careful consideration of
-current debates in journalism, we identify the following principles. \todo{we should find some
-references on trends in journalism to motivate these}
-
-\textbf{B1. Trust through transparency.} The system should allow fact checking of the analyses to
-build trust. This means that viewers should be able to determine what is the source of analysed
-data and how has the data been transformed.
-
-\textbf{B2. Opening the process to readers.} Journalists are increasingly opening the way they work
-to readers in order to build trust. The ability to view how an analysis has been constructed should
-not be limited just to experts (say, by running a Jupyter Notebook from GitHub), but should be
-available to all interested readers.
-
-\textbf{B3. Providing meaningful engagement mechanism.} Finally, the system should provide a
-mechanism through which readers can engage in a meaningful discussion. For example, it should allow
-them modify parameters of a data visualization in order to show how, e.g. different choice of
-countries affects the result.
+\textbf{B4. Visibility of state.} To support transparency, the system should make its entire state
+transparent -- when reviewing a data analysis, all parameters should be immediately visible and the
+user should not need to, e.g.~navigate through complex user interface to find them.
 
 \section{Systematic description}
+Describe in some more systematic detail how things work
 
@@
 
-Code should be visible to allow (A1)
-
-auto-complete friendly language allows (A2)
-
-language design addresses (A3, A4) by making easy things easy and hard
-things possible (if you know the right syntax)
-
 \subsection{Type providers}
 
+~
+
 \section{Design principles evaluation}
+Evaluate the system with respect to the design principles
+
+~
 
 \section{System evaluation}
+More discussion and evaluation, featuring some numbers and case studies
 
@@
 \subsection{Scalability}
+
+~
 
@@
 \subsection{Design principles}
+How well did we do wrt design principles?
 
 \subsection{Design issues}
-modifying code
+future challenges and limitations of the model - such as issues when modifying code
+in the middle of the call chain
 
-\newpage
-TODO: Evaluating systems paper
+~
+
+~
+
+TODO: Add evaluation according to the evaluating systems paper list
 

### 2019-04-01 wip-compost/paper.tex
Adding some rough notes on Compost

\documentclass[acmsmall,anonymous,fleqn]{acmart}\settopmatter{printfolios=false,printccs=false,printacmref=false}

%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{OOPSLA}
\acmArticle{1}
\acmYear{2019}
\acmMonth{1}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

\settopmatter{printfolios=true}
%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2018}           %% If different from \acmYear

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
\citestyle{acmauthoryear}   %% For author/year citations

\usepackage{booktabs}
\usepackage{subcaption}
\usepackage{lineno,hyperref,xcolor}
\usepackage{flushend}
\usepackage{stmaryrd}
\usepackage{amssymb}
\usepackage{xypic}
\usepackage{semantic}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{enumerate}
\usepackage{amsmath}
\usepackage[T1]{fontenc}

\setlist{leftmargin=6mm}
\newcounter{thc}
\newcounter{dfc}

\theoremstyle{plain}
\newtheorem{lem}[thc]{Lemma}
\newtheorem{theorem}[thc]{Theorem}

\theoremstyle{definition}
\newtheorem{axiom}[dfc]{Axiom}
\newtheorem{definition}[dfc]{Definition}

\title{Compost: Library for composable data visualization}
%\author{Tomas Petricek}
%\affiliation{
%  \institution{University of Kent}
%  \country{United Kingdom}
%}
%\email{tomas@tomasp.net}

%\begin{CCSXML}
%<ccs2012>
%<concept>
%<concept_id>10011007.10011006.10011008</concept_id>
%<concept_desc>Software and its engineering~General programming languages</concept_desc>
%<concept_significance>500</concept_significance>
%</concept>
%<concept>
%<concept_id>10003456.10003457.10003521.10003525</concept_id>
%<concept_desc>Social and professional topics~History of programming languages</concept_desc>
%<concept_significance>300</concept_significance>
%</concept>
%</ccs2012>
%\end{CCSXML}
%
%\ccsdesc[500]{Software and its engineering~General programming languages}
%\ccsdesc[300]{Social and professional topics~History of programming languages}

\definecolor{cmtclr}{rgb}{0.0,0.6,0.0}
\definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
\definecolor{idclr}{rgb}{0.0,0.0,0.0}
\definecolor{numclr}{rgb}{0.0,0.4,0.0}
\definecolor{strclr}{rgb}{0.4,0.4,0.0}
\definecolor{rstrclr}{rgb}{0.5,0.1,0.0}
\definecolor{prepclr}{rgb}{0.6,0.0,0.2}
\newcommand{\vect}[1]{\langl #1 \rangl}
\newcommand{\langl}{\begin{picture}(4.5,7)
\put(1.1,2.5){\rotatebox{60}{\line(1,0){5.5}}}
\put(1.1,2.5){\rotatebox{300}{\line(1,0){5.5}}}
\end{picture}}
\newcommand{\rangl}{\begin{picture}(4.5,7)
\put(.9,2.5){\rotatebox{120}{\line(1,0){5.5}}}
\put(.9,2.5){\rotatebox{240}{\line(1,0){5.5}}}
\end{picture}}
\newcommand{\ball}[1]{\FPeval{\result}{clip(201+#1)}\textnormal{\ding{\result}}}
\newcommand{\lsep}{\;\;|\;\;}
\newcommand{\num}[1]{\textcolor{numclr}{#1}}
\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
\newcommand{\strf}[1]{\textnormal{\textcolor{strclr}{\sffamily #1}}}
\newcommand{\rstr}[1]{\textnormal{\textcolor{rstrclr}{\sffamily "#1"}}}
\newcommand{\ident}[1]{\textnormal{\textcolor{idclr}{\sffamily #1}}}
\newcommand{\qident}[1]{\textnormal{\sffamily \textquotesingle #1\textquotesingle}}
\newcommand{\dom}{\ident{dom}}
\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}

\newcommand{\bndclr}[1]{\textcolor{kvdclr}{#1}}
\newcommand{\bkndclr}[1]{\textcolor{prepclr}{#1}}
\newcommand{\blblclr}[1]{\textcolor{numclr}{#1}}
\newcommand{\bnd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
\newcommand{\bknd}[1]{\textnormal{\textcolor{prepclr}{\sffamily #1}}}
\newcommand{\blbl}[1]{\textnormal{\textcolor{numclr}{\sffamily #1}}}
\newcommand{\narrow}[1]{\hspace{-0.6em}#1\hspace{-0.6em}}
\newcommand{\rname}[1]{{\sffamily\small(#1)}}
\newcommand{\ername}[1]{\vspace{0.75em}\rname{#1}\hspace{0.5em}}
\newcommand{\preview}[1]{\,\textnormal{\guillemotleft} #1\textnormal{\guillemotright}\,}

\begin{document}
%\begin{abstract}
%\end{abstract}
\maketitle

% ==================================================================================================

\section{Domain specific language}
\label{sec:intro}

\subsection*{Domain-specific values}
All values that are provided when constructing charts are from the chart domain. For example,
when visualizing the population of different countries, X values will be categorical values
such as \strf{United Kingdom} or \strf{Czech Republic} and Y values will be continuous values
such as $\num{66.04}$ million or $\num{10.58}$ million. Compost will automatically map the value
from a domain value to a pixel value. The user does not need to do any rescaling to obtain
value in pixels. This is akin to most high-level charting libraries, but in contrast with some
more flexible libraries like, say, D3.

Compost supports only two-dimensional charts and values can be either categorical (such as
different countries) or continuous (such as population size). A value needs to determine an
exact position in the space available for the chart. For continuous values, this simply means
applying a linear transformation. For categorical values, the situation is more difficult. For example,
say we are drawing a bar chart with population of three different countries with countries
on the X axis and population on the Y axis. The scale of the X axis is categorical and
Compost divides the available space into three ranges of equal width. However, then \strf{United Kingdom}
or \strf{Czech Republic} does not determine an exact position but instead a range. To determine
a unique position, we need to attach a value between $\num{0}$ and $\num{1}$ to the
categorical value which identifies a relative position in the available range. This is
illustrated in Figure~\ref{fig:scales}.

\begin{figure}[!h]
\includegraphics[scale=1,trim={0cm 6.5cm 6cm 0cm},clip]{values.pdf}
\caption{Values on a continous scale and categorical scale.}
\label{fig:scales}
\end{figure}

More formally, we write $c$ for categorical values such as countries (represented as country
names using strings), $r$ for ratio between $0$ and $1$ and we write $n$ for numerical continuous
values with any range. A value $v$ used in a Compost chart is defined as:
%
\begin{equation*}
\begin{array}{rcl}
v & = & \kvd{cat}~c, r \lsep \kvd{cont}~n
\end{array}
\end{equation*}
%
A value $v$ that is used to create Compost charts can be categorical, written as \kvd{cat} and
consisting of the category and a ratio that determines point in the space allocated for the
cateogry. A continous value, written as \kvd{cont}, consists of a number and will be mapped to
a location based on the implicitly obtained re-scaling.

\subsection*{Language of composable charts}
Now that we defined our representation of values, we can define a basic composable language of
data visualizations. A shape $s$ represents a chart. In the most basic form, it can be text label,
line connecting a list of points and a filled polygon determined by a list of points. For line,
text and polygon, we also include a parameter $\gamma$ indicating the color to be used. In addition,
a shape can be constructed by overlaying several other shapes:
%
\begin{equation*}
\begin{array}{rcl}
s & = & \kvd{overlay}~s_1, \ldots, s_n\\
  & | & \kvd{text}~\gamma, v_x, v_y, t\\
  & | & \kvd{line}~\gamma, v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\\
  & | & \kvd{fill}~\gamma, v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}
\end{array}
\end{equation*}
%
As an example, consider bar chart with \strf{UK} and \strf{CZ} as two categorical values on the X
axis and their population in millions (continuous value) on the Y axis. This can be constructed by
creating two filled rectangles using \kvd{fill} and overlaying them using \kvd{overlay}:
%
\begin{equation*}
\begin{array}{l}
\kvd{overlay}\\
\quad \kvd{fill}~\strf{\#0000ff},
  ~(\kvd{cat}~\strf{UK}, \num{0}), (\kvd{cont}~\num{00.00}), (\kvd{cat}~\strf{UK}, \num{0}), (\kvd{cont}~\num{66.04}),\\
\hspace{5.9em}(\kvd{cat}~\strf{UK}, \num{1}), (\kvd{cont}~\num{66.04}), (\kvd{cat}~\strf{UK}, \num{1}), (\kvd{cont}~\num{00.00}),\\
\quad \kvd{fill}~\strf{\#ff0000},
  ~(\kvd{cat}~\strf{CZ}, \num{0}), (\kvd{cont}~\num{00.00}), (\kvd{cat}~\strf{CZ}, \num{0}), (\kvd{cont}~\num{10.58}),\\
\hspace{5.9em}(\kvd{cat}~\strf{CZ}, \num{1}), (\kvd{cont}~\num{10.58}), (\kvd{cat}~\strf{CZ}, \num{1}), (\kvd{cont}~\num{00.00})\\
\end{array}
\end{equation*}
%
The shape specification overlays two bars of different colours. To construct a rectangle, we
define four points. For the UK, two of the points have Y value set to 0 (bottom of the bar) and
two have it set to 66.04 (top of the bar indiciating the population value). The values on the X
axis are categorical \strf{UK} values -- two are on the left as indicated by the ratio $\num{0}$
and two are on the right as indicated by ratio $\num{1}$. The ratio is always relative to the
category, so the ratios for the second bar are also $\num{0}$ and $\num{1}$.

\subsection*{Calculating scales and projections}
When processing shape specification such as the one given in the previous section, the Compost
library infers the scales for the X and Y axes and computes a projection function that can be
used for mapping domain-specific values to pixels. While doing this, it also checks for consistency --
all values on one axis have to be either categorical or continuous, but not a mix of the two.

The first step is to determine the scales for each of the axes. A scale can be continuous,
defined by a minimal and maximal value, or a categorical, defined by a list of categorical values.
Note that we do not need to track the ratios used as each category
will be allocated equal amount of space, regardless of where in that space shapes need to
appear. A scale $l$ is thus defined as:
%
\begin{equation*}
\begin{array}{rcl}
l & = & \kvd{categorical}~c_1, \ldots, c_k \lsep \kvd{continuous}~n_{min}, \ldots, n_{max}
\end{array}
\end{equation*}
%
A scale is obtained by recursively walking over the shape and gradually constructing two scales,
for X and Y axis, from the X and Y coordinates that appear in the shape. The process is fairly
straightforward and we won't discuss it in detail (for now). The first value encountered determines
what kind of a scale will be constructed. If the type of a later value does not match the type of a
scale, the process fails.

Once we obtain scales for both X and Y axes, we calculate a projection function for each axis.
The function takes a value $v$ together with total space available (say, in pixels) and produces
a position in pixels corresponding to $v$. We also omit the details here (for now), but briefly --
for continuous values, we produce a simple linear transformation; for categorical values, we split
the available space into $k$ equally sized regions (where $k$ is the number of categorical values
in the scale) and then map a categorical value $\kvd{cat}~c, r$ to the region corresponding to $c$
according to the ratio $r$.

\subsection*{Controlling and nesting scales}
The process of computing scales can be controlled by additional primitives in the Compost
domain-specific language for describing shapes that were not discussed earlier. The following
definition lists additional primitives that are also a part of the definition of $s$. Note that
the primitives can be applied either to the X scale or to the Y scale - we write $x/y$ to indicate
that a sub-script on those primitives can be set either to $x$ or to $y$ (both scales can be
modified by nesting two primitives):
%
\begin{equation*}
\begin{array}{rcl}
s & = & (\ldots)\\
  & | & \kvd{axis}_{x/y} s\\
  & | & \kvd{roundScale}_{x/y} s\\
  & | & \kvd{explicitScale}_{x/y}~l, s\\
  & | & \kvd{nest}_{x/y}~v_{min}, v_{max}, s
\end{array}
\end{equation*}
%
We first focus on the first three primitives, while the \kvd{nest} primitive will be discussed in
the next section. The \kvd{axis} and \kvd{roundScale} constructs could be defined as derived constructs,
but it is easier to consider them as primitives for now. The \kvd{axis} primitive simply takes the shape
$s$ and draws an axis around the contents of $s$, using the inferred scale of $s$ to determine values
displayed on the axis. This could be a derived construct, because axes can be rendered using lines
and text labels.

The \kvd{roundScale} operation takes the inferred X or Y scale of the shape $s$ and, if it is a
continuous scale, rounds its minimal and maximal values to ``nice'' numbers. For example, if a
continuous scale has minimum $\num{0}$ and maximum $\num{66.04}$, the resulting scale would have
maximum $\num{70}$. For categorical scale, the operation does not have any effect.
The \kvd{explicitScale} operation is similar, but it replaces the inferred scale with an explicitly
provided scale (the type of the inferred scale has to match with the type of the explicitly given
scale). For example, assuming \ident{populationBar} is the example given earlier, we can write:
%
\begin{equation*}
\begin{array}{l}
\kvd{axis}_x~(\kvd{axis}_y~(\kvd{roundScale}_y\\
\quad(\kvd{explicitScale}_x~(\kvd{categorical \strf{CZ},\strf{MN},\strf{UK}})~\ident{populationBar})))
\end{array}
\end{equation*}
%
This replaces the X axis with an explicitly given one that includes extra country code and also
overrides the implicitly inferred order of the categorical values. We then ask Compost to
automatically round the Y scale (showing population) so that we get a ``nice'' number as the
maximum. Finally, we add axes around the shape, producing a usual labelled chart.

As noted earlier, both \kvd{axis} and \kvd{roundScale} could be defined as derived -- \kvd{axis}
would have to infer the scales of the nested shape, calculate appropriate labels and insert
suitable lines and labels; the \kvd{roundScale} operation would need to infer the scale too and
then add an explicit \kvd{explicitScale} with a rounded minimal and maximal value of a continuous
scale.

\begin{figure}[!h]
\includegraphics[scale=1,trim={0cm 7.5cm 6cm 0cm},clip]{nest.pdf}
\caption{A continuous scale with values from $0$ to $6$, nested in another scale.}
\label{fig:nesting}
\end{figure}

\subsection*{Nesting of charts and scales}
Finally, one more primitive that we added in the above definition and that we have not explained
yet is $\kvd{nest}_{x/y}$. The parameters of the primitive are two values, $v_{min}$ and $v_{max}$
together with a shape $s$. The primitive makes it possible to nest one scale inside another
one.

When inferring the scales, Compost infers the scale of the nested shape $s$. This will then take
the area determined by the two points $v_{min}$ and $v_{max}$ in the scale determined by other
shapes that are overlaid with the $\kvd{nest}$ shape. The two points are also used to determine
the scales of the overall shape, but the scales of $s$ are nested and do not have any effect on
the outer scales. In fact, the scale of $s$ does not even have to be of the same type as the
scale of the outer shape.

Figure~\ref{fig:nesting} illustrates this with an example. Here, we are nesting a continuous scale
with values ranging from $\num{0}$ to $\num{6}$ into another scale. The values $v_{max}$ and
$v_{min}$ do not even have to be continuous. We can, for example, nest a line chart inside a
bar of a bar chart and then the two values could be $\kvd{cat}~\strf{One}, 0$ and $\kvd{cat}~\strf{One}, 1$
(which are two values that define a region of a categorical scale).

One area where nesting of scales is particularly useful is when combining multiple charts in
a single view. For example, let's say that we have two line charts showing prices of two different
stocks over the same period of time (for simplicity, we can just use UNIX timestamp as a time).
The prices of the two stocks are orders of magnitude different, so they cannot fit easily into the
same chart. We want to show two line charts side-by-side (one above each other). They should each
have their own Y axis (price), but the X axis (time) should be shared.

Assuming we have \ident{googPrices} and \ident{fbPrices} as two line charts without any axes,
covering the same date range, we can write:
%
\begin{equation*}
\begin{array}{l}
\kvd{axis}_x~(\kvd{overlay}\\
\quad (\kvd{nest}_y~(\kvd{cat}~\strf{top-chart}, \num{0}), (\kvd{cat}~\strf{top-chart}, \num{1}),
  (\kvd{axis}_y~\ident{fbPrices})),\\
\quad (\kvd{nest}_y~(\kvd{cat}~\strf{bottom-chart}, \num{0}), (\kvd{cat}~\strf{bottom-chart}, \num{1}),
  (\kvd{axis}_y~\ident{googPrices}))~)\\
\end{array}
\end{equation*}
%
Reading the code from the inner-most part to the outer-most, we first add separate Y axes to both
of the line charts. Given the difference in the prices, the axes will have quite different values.
We then nest the (continuous) Y axes using the \kvd{nest} primitive and, at the same time, implicitly
define a new outer Y axis. The outer Y axis is categorical with just two values, \strf{top-chart}
and \strf{bottom-chart}. This means that the two nested charts will take equal amount of space,
one above the other (each of the charts takes the full space allocated for the category -- the
minimal value has ratio $0$ while the maximal value has ratio $1$).

~

~

\noindent
[TODO: We really need illustrations for those example charts, but that would be more work!]

~

\section{Random ideas}
\begin{itemize}
\item We could define a type system to catch categorical/continuous value mismatch in a single scale.
\item It would be worth thinking about ordinal values, which are categorical but can be sorted.
\item There might be other kinds of scales - for example, color scale (can have meaning in scatter plot)
  or a scale for secondary markers (like sizes of bubbles in a bubble chart). We could really say that
  every value (including bar chart colors) should be coming from some scale...
\item Implementing something like \kvd{axis} or \kvd{roundScale} as an actual derived primitive is
  a bit tricky, because it needs to invoke a part of the normal rendering workflow (to run the
  automatic inference of scales on the nested shape) -- this might be just implementation issue,
  but it could be some more basic problem.
\end{itemize}
\end{document}

### 2019-09-23 wip-live-programming/paper-v2.tex
Updating to programming template, tweaks

--- 2019-02-11 wip-live-programming/paper.tex
@@
 
-\title{Live coding environment for a data exploration language}
+\title{A live coding environment for a data exploration language}
 \author{Tomas Petricek}
@@
 scientists write code that introduces few or no abstractions. Second, data exploration is an
-interactive process and it blurs the traditional distinction between development-time and run-time.
-This poses an interesting challenge for programming language research, because it allows us to
-revisit a number of traditional assumptions about programming languages and focus our attention
-on a new kind of programming tools that data analysts need.
-
-In this paper, we present \emph{data exploration calculus}, a formally tractable
-language that captures the structure of simple data analyses as done, for example, by journalists
-exploring open government data. We implement a programming environment for data exploration that
-evaluates code instantaneously during editing and shows previews of the results,
-while allowing the user to modify code in an unrestricted way in a text
-editor. Supporting a text editor is tricky as any edit can change the structure
-of code and fully recomputing program after every change would be too expensive.
-We present a new technique that allows evaluating and type-checking code while it is
-being written while reusing results of previous runs. We formalize the technique using the data
+interactive process that blurs the traditional distinction between development-time and run-time.
+This poses an interesting challenge for programming language research, inviting us to
+revisit a number of traditional assumptions about programming languages and design a new kind of
+programming tools that are suited for data analyst needs.
+
+We present the \emph{data exploration calculus}, a formal language that captures the structure of
+scripts written to explore data. We focus on the simplest kind of data analyses such as those
+written by journalists to analyse open government data. We implement a programming environment,
+based on this calculus, that evaluates code instantly during editing and shows previews of the results,
+while allowing the user to modify code in an unrestricted way in a text editor. Supporting interactive
+editing is tricky. Any edit can change the structure of code and fully recomputing the output would
+be too expensive. We present a technique that allows evaluating and type-checking code while it is
+being written and that reuses results of previous runs. We formalize the technique using the data
 exploration calculus, prove that it is correct and specify when previous results are reused.
-Finally, we also illustrate the practicality of our approach using empirical evaluation
+Finally, we illustrate the practicality of our approach using empirical evaluation
 and a case study.
 
-As data analysis becomes ever more important use of programming, research on programming languages
-and tools needs to consider new kinds of code that people need to write and new kinds of tools
-that can support them. The present paper is one step in this important direction.
+As data analysis becomes an ever more important use of programming, research on programming languages
+and tools needs to consider new kinds of programming workflows appropriate for those domains and
+conceive new kinds of tools that can support them. The present paper is one step in this important direction.
 \end{abstract}
@@
 \label{sec:intro}
-One of the aspects that make spreadsheets easier to use than programming tools is their
+One of the aspects that make spreadsheets easier to use than other programming tools is their
 liveness. When you change a value in a cell in Excel, the whole spreadsheet updates instantly
@@
 
-Increasing number of programming environments aim to provide the same live development experience
+Increasing numbers of programming environments aim to provide a live development experience
 for standard programming languages, but doing this is not easy. Fully recomputing the whole program
-after every keystroke is inefficient and calculating how a change in the source code changes the
+after every keystroke is inefficient. Calculating how a change in the source code changes the
 result is extremely hard when the text editor allows arbitrary changes.
-Consider the following snippet that gets the release years of 10 most expensive movies from a
+Consider the following snippet that gets the release years of the 10 most expensive movies from a
 data set \ident{movies}:
@@
 %
-A live coding environment computes and shows the list of years, but then the programmer modifies the code by
-making the constant $\num{10}$ a variable and changing the date format to see the full date:
+A live coding environment computes and shows the list of years. Suppose that the programmer then modifies the code by
+making $\num{10}$ a variable and changing the date format to see the full date:
 %
@@
 \ident{map} operation to differently format the release dates of already computed top 10 movies.
-We present such a live coding environment, describe it formally and empirically evaluate it.
-We discuss related work in Section~\ref{sec:future}, but we review the most important directions
-next, in order to situate our contributions.
+This can already be done in a number of systems based on structured code editors, but our
+approach allows this behaviour using a standard text editor. We discuss related work in more detail in
+Section~\ref{sec:future}, but we review the most important directions here, in order to
+situate our contributions.
 
 \subsection{Related work}
-We focus on simple data exploration as done, for example, by journalists \cite{ddj}
-and we aim to bring the experience of using transparent and reproducible programming tools
-closer to that of using spreadsheets. When a user edits a cell in a spreadsheet, the results
-automatically update. Recomputation in spreadsheets is well-studied,~e.g. by \citet{spreadsheet},
-but it is simpler than in our case, because editing is limited to individual cells.
-
-Many visual data exploration tools support interactivity \cite{control,tableau,vizdom} and some,
-can even export the workflow as a script \cite{wrangler}, but data analysts who prefer working with
+Simple data exploration performed, for example, by journalists \cite{ddj} is done either
+programmatically or using spreadsheets. The former is reproducible and transparent, but requires
+expert programming skills. The latter is easy, but it limits reproducibility and is error-prone.
+We aim to bring some aspects of spreadsheets, especially liveness, to programmatic data exploration.
+Recomputation in spreadsheets is well studied,~e.g. by \citet{spreadsheet}, but extending it to
+code written and edited in an ordinary text editor poses interesting challenges.
+
+Many visual data exploration tools support interactivity \cite{control,tableau,vizdom} and some
+can even export the workflow as a script \cite{wrangler}. Data analysts who prefer working with
 code typically resort to notebook systems such as Jupyter or R Markdown \cite{jupyter,rmarkdown}.
@@
 cells and manually reevaluate cells. Many notebook systems do not track dependencies
-between cells, which can lead to inconsistencies; an issue that has been addressed
-by e.g.~\citet{dataflow,noworkflow} or \citet{wrattler}.
-
-We aim to keep the transparency and reproducibility of notebooks, but add a correct and
-efficient instantaneous evaluation mechanism. Existing text-based editors that provide instant,
+between cells, which can lead to inconsistencies, an issue that has been addressed
+by e.g.~\citet{dataflow,noworkflow} and \citet{wrattler}.
+
+Existing text-based editors that provide instant feedback,
 such as Lighttable \cite{lighttable} and Chrome DevTools, often work only in certain
-situations or require full recomputation. A more principled approach has been used by
+situations or require full recomputation. A more principled approach can be used by
 systems based on structured editing \cite{livenut,lamdu}, which benefit from the fact that code
-can only be modified via known operations with known effect on the computation graph
-(``extract variable'' has no effect on the result; ``change constant value''
+can is modified via known operations with known effect on the computation graph
+(e.g.~``extract variable'' has no effect on the result; ``change constant value''
 forces recomputation of subsequent code). However, we aim to cater for the many users who
 prefer to edit programs as free-form text.
-
-The technical aspects of our technique for providing instantaneous feedback
+%
+The technical aspects of our technique for providing instant feedback
 are related to the work on incremental computation \cite{selfadjusting,incremental}. This focuses
-on cases where \emph{data} changes but program stays the same, while our focus is on cases where
-\emph{program} changes but data stays. Yet, the technical aspects are related and,
-as noted in Section~\ref{sec:future}, language that supports incremental computation could be used
+on cases where \emph{data changes} but program stays the same, while our focus is on cases where
+\emph{program changes} but data stays. Yet, the technical aspects are related and,
+as noted in Section~\ref{sec:future}, a language that supports incremental computation could be used
 to implement the kind of system presented in this paper.
@@
 We present the design and implementation of a live coding environment for a data exploration
-language that provides a correct and efficient instantaneous feedback, yet, is integrated into an
+language that provides correct and efficient instant feedback, yet is integrated into an
 ordinary text editor. Our most important contributions are:
@@
 \begin{itemize}[itemsep=3pt]
-\item We introduce \emph{data exploration calculus} (Section~\ref{sec:calculus}), a small formally
-  tractable programming language for data exploration. The language is motivated by our review
-  of how data analysts work (Section~\ref{sec:background}) and captures the idea that scripts are
-  often lists of commands with few or no abstractions with most logic provided by external libraries.
+\item We introduce the \emph{data exploration calculus} (Section~\ref{sec:calculus}), a small formally
+  tractable language for data exploration. The calculus is motivated by our review
+  of how data analysts work (Section~\ref{sec:background}) and makes it possible to write simple,
+  transparent and reproducible data analyses.
 
 \item Implementing a live coding environment requires a different way of thinking than the one
-  presented in classic compiler literature. We capture the essence of the new perspective
-  (Section~\ref{sec:formal}) and use it to build our \emph{live preview} mechanism
-  (Section~\ref{sec:previews}) evaluates code instantaneously during editing.
+  presented in classic compiler literature, because it does not operate in a batch mode. We capture
+  the essence of the new perspective (Section~\ref{sec:formal}) and use it to build our \emph{live
+  preview} mechanism (Section~\ref{sec:previews}) that evaluates code instantly during editing.
 
@@
 
-\item We extend our core environment in two ways, First, we add instantaneous type checker
+\item We describe two extensions of our core environment. First, we add instant type checker
   that relies on the same core infrastructure as live previews and enables richer user experience
-  through \emph{type providers} (Section~\ref{sec:extra-types}). Second, we add an abstraction mechanism
-  that preserves the live coding experience (Section~\ref{sec:extra-abstraction}).
+  through \emph{type providers} (Section~\ref{sec:extra-types}). Second, we add a code reuse mechanism
+  that supports the live coding experience (Section~\ref{sec:extra-abstraction}).
 \end{itemize}
@@
 \begin{figure}[b]
+\vspace{-0.5em}
 \includegraphics[scale=0.5]{notebook.png}
+\vspace{-0.5em}
 \caption{An excerpt from a Jupyter notebook by Financial Times reporters, which loads trade
@@
 \label{fig:ft-uncomtrade}
+\vspace{-0.5em}
 \end{figure}
@@
 code they write and the way they interact with the system is very different from how software
-engineers work \cite{workflow}. This paper focuses on simple data wrangling and data exploration as done, for
+engineers work. The different nature of data science workflow has been documented
+by \citet{workflow}. This paper focuses on simple data wrangling and data exploration as done, for
 example, by journalists analysing government datasets. This section illustrates how such data
-analyses look and provides a justification for the design of our data exploration calculus.
+analyses typically look and provides a justification for the design of our data exploration calculus.
 
@@
 Data analysts increasingly use notebook systems such as Jupyter, which
-make it possible to combine text, equations and code with results of running the code such as tables
+make it possible to combine text, equations and code with results of running the code, such as tables
 or visualizations. Notebooks blur the distinction between development and execution.
@@
 models of physical systems to journalists who load data, perform simple aggregations and create visualizations. Our
-initial focus is on simple use cases, such as those of journalists. Tooling that makes notebook
-systems more interactive and spreadsheet-like can encourage users to choose programming tools
-over spreadsheets and produce reproducible and transparent data analyses.
+initial focus is on the simplest use cases. We believe that making programmatic
+data exploration more interactive and spreadsheet-like can encourage users to choose programming tools
+over spreadsheets, allowing them to produce reproducible and transparent data analyses.
 
@@
 builds a visualization comparing waste flows in 2017 and 2018. Figure~\ref{fig:ft-uncomtrade} shows
-an extract from one notebook from the data analysis. The code has a number of remarkable properties:
+an extract from one notebook of the data analysis. The code has a number of important properties:
 %
@@
   needs to be set to handle Namibia correctly. These are discovered interactively by writing and
-  running a code snippet and examining the output, so providing a rapid feedback is essential.
+  running a code snippet and examining the output, so providing an instant feedback is essential.
 \end{itemize}
@@
     \raggedright\footnotesize The Gamma script to aggregate Olympic medals. The analyst selects
-      Rio 2016 games and counts the number of distinct events per athlete. After typing `.',
+      Rio 2016 games and counts the number of distinct events per athlete. After typing `.'
       a type provider offers further aggregation operations.
@@
   \subcaptionbox{
-    \raggedright\footnotesize Live preview in The Gamma programming environment developed based on
-      the theory presented in this paper. The table is produced as
+    \raggedright\footnotesize
+      Our live coding environment for The Gamma language based on the theory presented in this paper.
+      The table is produced as
       the data analyst types and updates on-the-fly to show result at the current cursor position.
@@
   }{\includegraphics[scale=0.26]{gamma2.png}}
-  \caption{\small Data exploration in The Gamma -- using a type provider (left) and with a live
-    preview (right).}
+  \caption{\small Previous work on The Gamma by \citet{gamma} (left) and our new editor with live preview (right).}
   \label{fig:gamma-intro}
@@
 
-Data exploration of the kind discussed in the previous section has been the motivation for a
-recent work by \citet{gamma} that develops a small scripting language called The Gamma. The language
-follows the structure discussed above. A script is a sequence of commands that can either define a
-variable or produce an output. The language does not support top-level function declarations
+Data exploration of the kind discussed in the previous section has been the motivation for
+recent work by \citet{gamma} that develops a small scripting language called The Gamma.
+Scripts in The Gamma are sequences of commands that can either define a variable or produce an
+output. The language does not support top-level function declarations
 and lambda functions can be used only as method arguments.
@@
 Given the limited expressiveness of The Gamma, libraries are implemented in other languages,
-such as JavaScript. The Gamma also supports type providers \cite{providers-fsharp},
-which can be used to generate types, on the fly, for accessing and exploring external data sources.
+such as JavaScript. The Gamma supports type providers \cite{providers-fsharp},
+which generate types on the fly, for accessing and exploring external data sources.
 Type providers provide object types with members and The Gamma makes using those convenient by
@@
 As illustrated by \citet{gamma}, the combination of type providers and auto-complete on~`.' makes it
-possible to complete a large number of data exploration tasks through a very simple interaction of
+possible to complete a large number of data exploration tasks through the very simple interaction of
 selecting operations from a list. An example in Figure~\ref{fig:gamma-dot} summarizes data on
-Olympic medals using a type provider. The names of identifiers such as \qident{sum Bronze} do not have
-any special meaning. They are merely members generated by the type provider, based on information
+Olympic medals using a type provider. Identifiers such as \qident{sum Bronze}
+are merely names of members generated by the type provider, based on information
 about the data source. The type provider used in this example generates an object with members for
@@
 produced by journalists, economists and other users with other than programming background.
-When we refer to The Gamma in this paper, readers not familiar with it can assume that we focus
-on a subset of Python.
+When we refer to The Gamma in this paper, readers not familiar with it can consider a subset of Python.
 
@@
 
-The practical implementation work that complements the theory presented in this paper implements a live
-coding environment for The Gamma and is discussed in Section~\ref{sec:evaluation-case} in detail. Earlier work by
-\citet{gamma} discussed The Gamma language and the type provider mechanism, but the implementation
-provided only a standard text editor with auto-completion. Our work brings the experience of using
-The Gamma closer to that of spreadsheets by developing an editor that evaluates scripts
-instantaneously in background and displays the results as \emph{live previews}.
+The implementation that accompanies this paper builds a live coding environment for The Gamma
+language. The implementation is discussed in Section~\ref{sec:evaluation-case} and it replaces
+the text editor with auto-complete, originally used by The Gamma, with a live coding environment
+that provides \emph{live previews}. We choose to extend The Gamma over using a subset of Python
+for its simplicity as it lets us focus on the principles of live coding systems.
 
 An example of a live preview is shown in Figure~\ref{fig:gamma-preview}. As noted earlier, The
-Gamma program consist of a list of commands which are either expressions or let bindings. Our
+Gamma programs consist of lists of commands which are either expressions or let bindings. Our
 editor displays a live preview below the command that the user is currently editing. The preview
@@
 
-There is a number of guiding principles that our work follows. First, we allow the analyst
+There are a number of guiding principles that inform our design. First, we allow the analyst
 to edit code in an unrestricted form in a text editor. Although structured editors provide an
-appealing alternative and make recomputation easier, we prefer the transparency
-and openness of plain text. Second, we focus on the scenario when code changes, but input
+appealing alternative and make recomputation easier, we prefer the flexibility of plain text.
+Second, we focus on the scenario when code changes, but input
 does not. Rapid feedback allows the analyst to quickly adapt code to correctly handle
@@
 
-The \emph{data exploration calculus} is a small, formally tractable language for data exploration.
-The calculus is not Turing-complete and it can only be used together with external libraries that
+The \emph{data exploration calculus} is a small formal language for data exploration.
+The calculus is not, in itself, Turing-complete and it can only be used together with external libraries that
 define what objects are available and what the behaviour of their members is. This is
@@
 We initially present the data exploration calculus without a static type system as the live preview
-mechanism does not rely on types. We consider type checking later in Section~\ref{sec:extra-types}
-as the mechanism that is used for live previews can also be used for instantaneous type checking.
+mechanism does not rely on types. We consider type checking later in Section~\ref{sec:extra-types}.
+Interestingly, the mechanism that is used for live previews can also be used for instant type checking.
 
@@
 \begin{array}{lcl}
-p &\narrow{=}& c_1; \ldots; c_n\\
-c &\narrow{=}& \kvd{let}~x = t \lsep t\\
+p &\narrow{::=}& c_1; \ldots; c_n\\
+c &\narrow{::=}& \kvd{let}~x = t \lsep t\\
 \end{array}
@@
 \begin{array}{lcl}
-t &\narrow{=}& o \lsep x \\
+t &\narrow{::=}& o \lsep x \\
   &\narrow{|}& t.m(e, \ldots, e)
@@
 \begin{array}{lcl}
-e &\narrow{=}& t \lsep \lambda x\rightarrow e\\
-v &\narrow{=}& o \lsep \lambda x\rightarrow e\\
+e &\narrow{::=}& t \lsep \lambda x\rightarrow e\\
+v &\narrow{::=}& o \lsep \lambda x\rightarrow e\\
 \end{array}
@@
 
-The data exploration calculus is a call-by-value language. We model the evaluation as a small-step
+The data exploration calculus is a call-by-value language. We model evaluation as a small-step
 reduction $\rightsquigarrow$. Fully evaluating a program results in an irreducible sequence of
@@
 by a relation $\rightsquigarrow_\epsilon$ that models the functionality of the external libraries
-used with the calculus and defines reduction rules for member accesses. The relation has the
+used with the calculus and defines the reduction behaviour for member accesses. The relation has the
 following form:
@@
 %
-The operation is invoked on an object and takes values (objects or function values) as arguments.
-It always results in an opaque object.
+Here, the operation $m$ is invoked on an object and takes values (objects or function values) as arguments.
+The reduction always results in an object.
 Figure~\ref{fig:dec-calculus} defines the reduction rules in terms of $\rightsquigarrow_\epsilon$
 and evaluation contexts; $C_e$ specifies left-to-right evaluation of arguments of a method call,
-$C_c$ specifies evaluation of a command and $C_p$ defines top-to-bottom evaluation of a program.
-The rule \rname{external} applies reductions provided by an external library in a
-call-by-value order and \rname{let} substitutes a value of an evaluated variable in
-all subsequent commands. Note that \rname{let} leaves the value of the variable
-in the resulting list of commands.
+$C_c$ specifies evaluation of a command and $C_p$ defines left-to-right evaluation of a program.
+The rule \rname{external} calls a method provided by an external library in a
+call-by-value fashion and \rname{let} substitutes a value of an evaluated variable in
+all subsequent commands and leaves the result in the list of commands.
+Note that our semantics does not define how $\lambda$ functions are reduced. This is done by
+external libraries, which will typically supply functions with arguments using standard $\beta$-reduction.
+The behaviour is subject to constraints discussed next.
 
 \subsection{Properties}
-The data exploration calculus is a very simple language and it has a number of desirable
-properties. However, some of those require that the relation $\rightsquigarrow_\epsilon$, which
-defines evaluation for external libraries, satisfies a number of conditions. Our main motivation
-is to capture properties that allow us to prove that our method of evaluating live previews,
-discussed in Section~\ref{sec:evaluation-correctness}, is correct.
+The data exploration calculus has a number of desirable properties. Some of those require that the
+relation $\rightsquigarrow_\epsilon$, which defines evaluation for external libraries, satisfies a
+number of conditions. We discuss \emph{normalization} and \emph{let elimination} in this section.
+Those two are particularly important as they will allow us to prove correctness of our method of
+evaluating live previews in Section~\ref{sec:evaluation-correctness}.
 
@@
 We say that two expressions
-$e$ and $e'$ are observationally equivalent if, for any context $C$, the expressions $C[e]$ and
+$e$ and $e'$ are \emph{observationally equivalent} if, for any context $C$, the expressions $C[e]$ and
 $C[e']$ reduce to the same value. Lambda functions $\lambda x\!\rightarrow\!2$ and
@@
 \begin{itemize}
-\item[-] \emph{Completeness.} For all $o, m, i$ and all $v_1, \ldots, v_i$, there exists $o'$ such
+\item[-] \emph{Totality.} For all $o, m, i$ and all $v_1, \ldots, v_i$, there exists $o'$ such
   that $o.m(v_1, \ldots, v_i) \rightsquigarrow_\epsilon o'$.
@@
 The compositionality condition is essential for proving the correctness of our live preview mechanism.
-Completeness allows us to prove normalization, i.e.~all programs reduce to
+Totality allows us to prove normalization, i.e.~all programs reduce to
 a value -- although the resulting value may be an error value provided by the external library.
@@
 
-\section{Formalising live coding environment}
+\section{Formalising a live coding environment}
 \label{sec:formal}
@@
 A naive way of providing live previews during code editing would be to re-evaluate the code
-after each change. This would be wasteful -- for example, when writing
-code, most changes are additive and the preview can be updated by evaluating just the newly added
+after each change. This would be wasteful -- when writing code to explore data, most changes are
+additive and the preview can be updated by evaluating just the newly added
 code. In this section, we develop foundations for an efficient way of displaying live previews
@@
 The key idea behind our method is to maintain a dependency graph \cite{dependencies} with
-nodes representing individual operations of the computation that can be (partially) evaluated
+nodes representing individual operations of the computation that can be evaluated
 to obtain a preview. Each time the program text is modified, we parse it afresh (using an
@@
 
-The nodes of the graph serve as unique keys into a lookup table with previously
+The nodes of the graph serve as unique keys into a lookup table containing previously
 evaluated parts of the program. When a preview is requested for an expression, we use the graph
@@
 \begin{array}{ll}
-\bkndclr{k}~=~\bknd{fun}(x)~|~\bknd{mem}(m)&(\textit{Node kinds})\\[0.2em]
+\bkndclr{k}~::=~\bknd{fun}(x)~|~\bknd{mem}(m)&(\textit{Node kinds})\\[0.2em]
 \Delta(\bkndclr{k}, [(\bndclr{v_1},\blblclr{l_1}), \ldots, (\bndclr{v_n},\blblclr{l_n})])  & (\textit{Lookup for a node})
@@
 $\bndclr{v_1}, \ldots, \bndclr{v_n}$ labelled with labels $\blblclr{l_1}, \ldots, \blblclr{l_n}$.
-We write $\Delta(\bndclr{k}, l)\!\downarrow$ when a value for a given key is not defined.
+We write $\Delta(\bkndclr{k}, l)\!\downarrow$ when a value for a given key is not defined.
 For example, when creating the graph in Figure~\ref{fig:dep-graph-b},
@@
 this lookup fails. We then construct a new node $\bnd{mem}(\ident{take}, s_2)$ using a fresh
-symbol $s_2$.
+symbol $s_2$ and later insert it into the cache.
 
@@
 The binding process formalised in Section~\ref{sec:formal-bind} specifies how to update the
-dependency graph after updated program text is parsed. During live coding, this is done
+dependency graph after program is updated and parsed. During live coding, this is done
 repeatedly as the programmer edits code. Throughout the process, we maintain a series of
@@
 nodes of the graph. Since the binding process reuses nodes when their dependencies do not change,
-previews for expressions (or sub-expressions) of a program can be reused when updating a preview.
+previews for expressions (or sub-expressions) of a program are reused when updating a preview.
 
 In this section, we describe how previews are evaluated. The evaluation is done over the dependency
-graph, rather than over the structure of the program expressions as in the operational semantics
+graph, rather than over the structure of program expressions as in the operational semantics
 given in Section~\ref{sec:calculus-semantics}. We analyse the preview evaluation formally in
 Section~\ref{sec:evaluation} and show that the resulting previews are the same as the result we
-would get by directly evaluating the code and, also, that no recomputation occurs when code is
-edited in certain ways.
+would get by directly evaluating the code and, equally importantly, we show that no recomputation
+occurs when code is edited in certain ways.
 
@@
 
-Programs in the data exploration calculus consist of sequence of commands. Those can always be
+Programs in the data exploration calculus consist of sequence of commands. Those are
 evaluated to a value with a preview that can be displayed to the user. However, we also support
-previews for all sub-expressions. This can be problematic if the current sub-expression is inside
+previews for sub-expressions. This can be problematic if the current sub-expression is inside
 the body of a function. For example:
@@
 force the evaluation of the surrounding expression, capture (a number of) values passed as inputs
-to the lambda function and display a preview based on those. Finally, in Section~\ref{sec:extra-abstraction},
-we also consider a more speculative design for an abstraction mechanism that supports live previews
-that could, in some cases, replace lambda function.
+to the lambda function and display a preview based on those. In Section~\ref{sec:extra-abstraction}
+we also consider a more speculative design for an abstraction mechanism that supports live previews,
+which could, in some cases, replace lambda function.
 
@@
 that would allow us to express the splicing operation used in \rname{mem-expr}. In categorical
-terms, the context-dependent proposition can be modelled as an indexed comonad \cite{effectrev,graded}.
+terms, the context-dependent proposition can be modelled as a graded comonad \cite{effectrev,graded}.
 The evaluation of a preview with no context dependencies (built implicitly into our evaluation rules)
@@
 
-The key performance claim about our method of providing instantaneous feedback is that it is more
+The key performance claim about our method of providing instant feedback is that it is more
 efficient than recomputing values for the whole program (or the current command) after every
@@
 we also implemented an online data exploration environment, equipped with live editor for The
-Gamma language that provides instantaneous feedback during coding. The environment uses the
+Gamma language that provides instant feedback during coding. The environment uses the
 principles presented in this paper to build a more comprehensive system that allows users,
@@
 live preview mechanism, but it lacks many important features. We outline two extensions.
-First, we discuss how to add instantaneous type checking by using the dependency graph construction
+First, we discuss how to add instant type checking by using the dependency graph construction
 described in Section~\ref{sec:formal}. Second, we consider an abstraction mechanism that preserves
@@
 
-Instantaneous live previews give analysts quick feedback when they write incorrect code, but
+Instant live previews give analysts quick feedback when they write incorrect code, but
 having type information is still valuable. First, it can help give better error messages. Second,
@@
 as an edge from the node of the function. Figure~\ref{fig:graph-func} shows the
-revised dependency graph for $o.m(\lambda x\rightarrow x)$. A revised version of the algorithm
-that constructs the dependency graph is included in Appendix~\ref{sec:app-types}.
+revised dependency graph for $o.m(\lambda x\rightarrow x)$. %A revised version of the algorithm
+%that constructs the dependency graph is included in Appendix~\ref{sec:app-types}.
 
@@
 The typing rules for variables, functions and member access are shown in Figure~\ref{fig:tc}.
-Remaining rules are available in Appendix~\ref{sec:app-types}.
+%Remaining rules are available in Appendix~\ref{sec:app-types}.
 When type checking a member access \rname{mem}, we find its dependencies $\bndclr{v_i}$ and
@@
 delayed previews inside body of a function. We believe that extending the data exploration calculus
-with an abstraction mechanism that would support development with instantaneous feedback is an
+with an abstraction mechanism that would support development with instant feedback is an
 interesting design problem and we briefly outline a possible solution here.
@@
 We aim to make creating transparent data analyses, such as those done by journalists, easier and we
-present a tool that provides instantaneous feedback when writing code in a text editor. Many our
+present a tool that provides instant feedback when writing code in a text editor. Many our
 design decisions have alternatives that have been explored in a rich body of literature.
@@
 One of the reasons that make spreadsheets easier to use than programming tools is that they
-provide instant feedback. We aim to make programming tools as instantaneous as spreadsheets.
+provide instant feedback. We aim to make programming tools as instant as spreadsheets.
 We formalized a number of key aspects of simple data analyses such as those don by journalists
@@
 
-\section{Appendix - Type checking}
-\label{sec:app-types}
-
-tbd
+% \section{Appendix - Type checking}
+% \label{sec:app-types}
+%
+% tbd
 

### 2019-09-28 wip-live-programming/paper.tex
vspace vspace vspace...

@@
-\documentclass[acmsmall,anonymous,fleqn]{acmart}\settopmatter{printfolios=false,printccs=false,printacmref=false}
-
-%% For double-blind review submission, w/ CCS and ACM Reference
-%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
-%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
-%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
-%% For single-blind review submission, w/ CCS and ACM Reference
-%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
-%% For final camera-ready submission, w/ required CCS and ACM Reference
-%\documentclass[acmsmall]{acmart}\settopmatter{}
-
-\acmJournal{PACMPL}
-\acmVolume{1}
-\acmNumber{OOPSLA}
-\acmArticle{1}
-\acmYear{2019}
-\acmMonth{1}
-\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
-\startPage{1}
-
-\settopmatter{printfolios=true}
-%% Copyright information
-%% Supplied to authors (based on authors' rights management selection;
-%% see authors.acm.org) by publisher for camera-ready submission;
-%% use 'none' for review submission.
-\setcopyright{none}
-%\setcopyright{acmcopyright}
-%\setcopyright{acmlicensed}
-%\setcopyright{rightsretained}
-%\copyrightyear{2018}           %% If different from \acmYear
-
-%% Bibliography style
-\bibliographystyle{ACM-Reference-Format}
-%% Citation style
-%% Note: author/year citations are required for papers published as an
-%% issue of PACMPL.
-\citestyle{acmauthoryear}   %% For author/year citations
-
-\usepackage{booktabs}
-\usepackage{subcaption}
-\usepackage{lineno,hyperref,xcolor}
-\usepackage{flushend}
-\usepackage{stmaryrd}
+\documentclass[english,submission]{programming}
+\usepackage[backend=biber]{biblatex}
+\addbibresource{paper.bib}
+\usepackage{multicol}
+\usepackage{microtype}
 \usepackage{amssymb}
@@
 \usepackage{semantic}
-\usepackage{booktabs}
-\usepackage{enumitem}
-\usepackage{enumerate}
 \usepackage{amsmath}
-\usepackage[T1]{fontenc}
-
-\setlist{leftmargin=6mm}
+\usepackage{amsthm}
+
 \newcounter{thc}
@@
 
-\title{Live coding environment for a data exploration language}
-\author{Tomas Petricek}
-
-\affiliation{
-  \institution{University of Kent}
-  \country{United Kingdom}
-}
-\email{tomas@tomasp.net}
-
-\begin{CCSXML}
-<ccs2012>
-<concept>
-<concept_id>10011007.10011006.10011008</concept_id>
-<concept_desc>Software and its engineering~General programming languages</concept_desc>
-<concept_significance>500</concept_significance>
-</concept>
-<concept>
-<concept_id>10003456.10003457.10003521.10003525</concept_id>
-<concept_desc>Social and professional topics~History of programming languages</concept_desc>
-<concept_significance>300</concept_significance>
-</concept>
-</ccs2012>
-\end{CCSXML}
-
-\ccsdesc[500]{Software and its engineering~General programming languages}
-\ccsdesc[300]{Social and professional topics~History of programming languages}
-
 \definecolor{cmtclr}{rgb}{0.0,0.6,0.0}
 \definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
-\definecolor{idclr}{rgb}{0.3,0.3,0.3}
+\definecolor{idclr}{rgb}{0.0,0.0,0.0}
 \definecolor{numclr}{rgb}{0.0,0.4,0.0}
@@
 \newcommand{\narrow}[1]{\hspace{-0.6em}#1\hspace{-0.6em}}
-\newcommand{\rname}[1]{{\sffamily\small(#1)}}
+\newcommand{\rname}[1]{{\sffamily(#1)}}
 \newcommand{\ername}[1]{\vspace{0.75em}\rname{#1}\hspace{0.5em}}
@@
 \begin{document}
+\title{A live coding environment for a data exploration language}
+%\titlerunning{Learning to live with errors}
+
+\author[a]{Tomas Petricek}
+\authorinfo{Tomas is xxx. Contact him at \email{tomas@tomasp.net}.}
+\affiliation[a]{University of Kent, UK}
+\keywords{xxx}
+\paperdetails{
+  perspective=engineering,
+  area={History and philosophy of programming},
+}
+
+% https://dl.acm.org/ccs/ccs.cfm and generate your Classification
+\begin{CCSXML}
+<ccs2012>
+<concept>
+<concept_id>10002944.10011122.10002945</concept_id>
+<concept_desc>General and reference~Surveys and overviews</concept_desc>
+<concept_significance>300</concept_significance>
+</concept>
+<concept>
+<concept_id>10003456.10003457.10003521</concept_id>
+<concept_desc>Social and professional topics~History of computing</concept_desc>
+<concept_significance>300</concept_significance>
+</concept>
+</ccs2012>
+\end{CCSXML}
+\ccsdesc[300]{General and reference~Surveys and overviews}
+\ccsdesc[300]{Social and professional topics~History of computing}
+
+\maketitle
+
 \begin{abstract}
+TODO!
+
+Rewrite according to guidelines.
+
+TODO!
 
@@
 scientists write code that introduces few or no abstractions. Second, data exploration is an
-interactive process and it blurs the traditional distinction between development-time and run-time.
-This poses an interesting challenge for programming language research, because it allows us to
-revisit a number of traditional assumptions about programming languages and focus our attention
-on a new kind of programming tools that data analysts need.
-
-In this paper, we present \emph{data exploration calculus}, a formally tractable
-language that captures the structure of simple data analyses as done, for example, by journalists
-exploring open government data. We implement a programming environment for data exploration that
-evaluates code instantaneously during editing and shows previews of the results,
-while allowing the user to modify code in an unrestricted way in a text
-editor. Supporting a text editor is tricky as any edit can change the structure
-of code and fully recomputing program after every change would be too expensive.
-We present a new technique that allows evaluating and type-checking code while it is
-being written while reusing results of previous runs. We formalize the technique using the data
+interactive process that blurs the traditional distinction between development-time and run-time.
+This poses an interesting challenge for programming language research, inviting us to
+revisit a number of traditional assumptions about programming languages and design a new kind of
+programming tools that are suited for data analyst needs.
+
+We present the \emph{data exploration calculus}, a formal language that captures the structure of
+scripts written to explore data. We focus on the simplest kind of data analyses such as those
+written by journalists to analyse open government data. We implement a programming environment,
+based on this calculus, that evaluates code instantly during editing and shows previews of the results,
+while allowing the user to modify code in an unrestricted way in a text editor. Supporting interactive
+editing is tricky. Any edit can change the structure of code and fully recomputing the output would
+be too expensive. We present a technique that allows evaluating and type-checking code while it is
+being written and that reuses results of previous runs. We formalize the technique using the data
 exploration calculus, prove that it is correct and specify when previous results are reused.
-Finally, we also illustrate the practicality of our approach using empirical evaluation
+Finally, we illustrate the practicality of our approach using empirical evaluation
 and a case study.
 
-As data analysis becomes ever more important use of programming, research on programming languages
-and tools needs to consider new kinds of code that people need to write and new kinds of tools
-that can support them. The present paper is one step in this important direction.
+As data analysis becomes an ever more important use of programming, research on programming languages
+and tools needs to consider new kinds of programming workflows appropriate for those domains and
+conceive new kinds of tools that can support them. The present paper is one step in this important direction.
 \end{abstract}
-\maketitle
 
@@
 \label{sec:intro}
-One of the aspects that make spreadsheets easier to use than programming tools is their
-liveness. When you change a value in a cell in Excel, the whole spreadsheet updates instantly
+One of the aspects that make spreadsheets easier to use than other programming tools is their
+liveness. When you change a cell in Excel, the whole spreadsheet updates instantly
 and you immediately see new results, without having to explicitly trigger re-computation
@@
 
-Increasing number of programming environments aim to provide the same live development experience
+An increasing number of programming environments aim to provide a live development experience
 for standard programming languages, but doing this is not easy. Fully recomputing the whole program
@@
 result is extremely hard when the text editor allows arbitrary changes.
-Consider the following snippet that gets the release years of 10 most expensive movies from a
+Consider the following snippet that gets the release years of the $10$ most expensive movies from a
 data set \ident{movies}:
@@
 %
-A live coding environment computes and shows the list of years, but then the programmer modifies the code by
-making the constant $\num{10}$ a variable and changing the date format to see the full date:
+A live coding environment computes and displays the list of years. Suppose that the programmer then
+makes $\num{10}$ a variable and changes the date format:
 %
@@
 Ideally, the live coding environment should understand the change, reuse a cached result of the
-first two transformations (sorting and taking first 10 elements) and only evaluate the
-\ident{map} operation to differently format the release dates of already computed top 10 movies.
-We present such a live coding environment, describe it formally and empirically evaluate it.
-We discuss related work in Section~\ref{sec:future}, but we review the most important directions
-next, in order to situate our contributions.
+first two transformations (sorting and taking the first 10 elements) and only evaluate the
+\ident{map} operation to differently format the release dates of top 10 movies. Our environment
+does this for a simple data exploration language in an unrestricted text editor. We discuss related
+work in Section~\ref{sec:future}, but we briefly review the most important directions here, in
+order to situate our contributions.
 
 \subsection{Related work}
-We focus on simple data exploration as done, for example, by journalists \cite{ddj}
-and we aim to bring the experience of using transparent and reproducible programming tools
-closer to that of using spreadsheets. When a user edits a cell in a spreadsheet, the results
-automatically update. Recomputation in spreadsheets is well-studied,~e.g. by \citet{spreadsheet},
-but it is simpler than in our case, because editing is limited to individual cells.
-
-Many visual data exploration tools support interactivity \cite{control,tableau,vizdom} and some,
-can even export the workflow as a script \cite{wrangler}, but data analysts who prefer working with
+Simple data exploration performed, for example, by journalists \cite{ddj} is done either
+programmatically or using spreadsheets. The latter is easy but error-prone while the former
+requires expert programming skills. We aim to bring liveness of spreadsheets to programmatic
+data exploration. This requires extending recomputation as done in spreadsheets \cite{spreadsheet}
+to code written in an ordinary text editor.
+
+Many visual data exploration tools support interactivity \cite{control,tableau,vizdom} and some
+can even export the workflow as a script \cite{wrangler}. Data analysts who prefer working with
 code typically resort to notebook systems such as Jupyter or R Markdown \cite{jupyter,rmarkdown}.
@@
 cells and manually reevaluate cells. Many notebook systems do not track dependencies
-between cells, which can lead to inconsistencies; an issue that has been addressed
-by e.g.~\citet{dataflow,noworkflow} or \citet{wrattler}.
-
-We aim to keep the transparency and reproducibility of notebooks, but add a correct and
-efficient instantaneous evaluation mechanism. Existing text-based editors that provide instant,
+between cells which can lead to well-documented inconsistencies \cite{dataflow,noworkflow,wrattler}.
+
+Existing text-based editors that provide instant feedback,
 such as Lighttable \cite{lighttable} and Chrome DevTools, often work only in certain
-situations or require full recomputation. A more principled approach has been used by
-systems based on structured editing \cite{livenut,lamdu}, which benefit from the fact that code
-can only be modified via known operations with known effect on the computation graph
-(``extract variable'' has no effect on the result; ``change constant value''
-forces recomputation of subsequent code). However, we aim to cater for the many users who
+situations or require full recomputation. A more principled approach can be used by
+systems based on structured editing \cite{livenut,lamdu} where code is only
+modified via known operations with known effect on the computation graph
+(e.g.~``extract variable'' has no effect on the result; ``change constant value''
+forces recomputation of subsequent code). We aim to cater for the many users who
 prefer to edit programs as free-form text.
-
-The technical aspects of our technique for providing instantaneous feedback
-are related to the work on incremental computation \cite{selfadjusting,incremental}. This focuses
-on cases where \emph{data} changes but program stays the same, while our focus is on cases where
-\emph{program} changes but data stays. Yet, the technical aspects are related and,
-as noted in Section~\ref{sec:future}, language that supports incremental computation could be used
-to implement the kind of system presented in this paper.
+%
+The technical aspects of our implementation are related to incremental computation \cite{selfadjusting,incremental}.
+This focuses on cases where \emph{data changes} but program stays the same, while our focus is on cases where
+\emph{program changes} but data stays the same.
 
 \subsection{Contributions}
-We present the design and implementation of a live coding environment for a data exploration
-language that provides a correct and efficient instantaneous feedback, yet, is integrated into an
-ordinary text editor. Our most important contributions are:
-
-\begin{itemize}[itemsep=3pt]
-\item We introduce \emph{data exploration calculus} (Section~\ref{sec:calculus}), a small formally
-  tractable programming language for data exploration. The language is motivated by our review
-  of how data analysts work (Section~\ref{sec:background}) and captures the idea that scripts are
-  often lists of commands with few or no abstractions with most logic provided by external libraries.
-
-\item Implementing a live coding environment requires a different way of thinking than the one
-  presented in classic compiler literature. We capture the essence of the new perspective
-  (Section~\ref{sec:formal}) and use it to build our \emph{live preview} mechanism
-  (Section~\ref{sec:previews}) evaluates code instantaneously during editing.
+We present the design and implementation of a live coding environment for a simple data exploration
+language that provides correct and efficient instant feedback, yet is integrated into an
+ordinary text editor. Our key contributions are:
+
+\begin{itemize}
+\item We introduce the \emph{data exploration calculus} (Section~\ref{sec:calculus}), a small formally
+  tractable language for data exploration. The calculus is motivated by our review
+  of how data analysts work (Section~\ref{sec:background}) and makes it possible to write simple,
+  transparent and reproducible data analyses.
+
+\item A live coding environment does not operate in batch mode and so we cannot follow classic
+  compiler literature. We capture the essence of such new perspective (Section~\ref{sec:formal})
+  and use it to build our \emph{live preview} mechanism (Section~\ref{sec:previews}) that evaluates
+  code instantly during editing.
 
@@
   (Section~\ref{sec:evaluation-empirical}) and a case study (Section~\ref{sec:evaluation-case}).
-
-\item We extend our core environment in two ways, First, we add instantaneous type checker
-  that relies on the same core infrastructure as live previews and enables richer user experience
-  through \emph{type providers} (Section~\ref{sec:extra-types}). Second, we add an abstraction mechanism
-  that preserves the live coding experience (Section~\ref{sec:extra-abstraction}).
 \end{itemize}
@@
 
-\begin{figure}[b]
-\includegraphics[scale=0.5]{notebook.png}
-\caption{An excerpt from a Jupyter notebook by Financial Times reporters, which loads trade
-  data from the UN trade database (stored in a local file) and adds ISO country codes (from
-  an Excel spreadsheet).}
+\begin{figure}
+\vspace{-1em}
+\hspace{1em}\includegraphics[scale=0.52]{figures/notebook.png}
+\vspace{-0.5em}
+\caption{Financial Times analysis that joins UN trade database with ISO country codes.}
 \label{fig:ft-uncomtrade}
+\vspace{-0.5em}
 \end{figure}
@@
 code they write and the way they interact with the system is very different from how software
-engineers work \cite{workflow}. This paper focuses on simple data wrangling and data exploration as done, for
-example, by journalists analysing government datasets. This section illustrates how such data
-analyses look and provides a justification for the design of our data exploration calculus.
+engineers work \cite{workflow}. This paper focuses on simple data wrangling and data exploration as
+done, for example, by journalists analysing government datasets. This new kind of non-expert
+programmers is worth our attention as they often work on informing the public. They need easy-to-use
+tools, but not necessarily a full programming langugae. In this section, we ilustrate how such
+data analyses look and we provide a justification for the design of our data exploration calculus.
 
@@
 
-Data analysts increasingly use notebook systems such as Jupyter, which
-make it possible to combine text, equations and code with results of running the code such as tables
-or visualizations. Notebooks blur the distinction between development and execution.
-Data analysts write small snippets of code, run them to see results immediately and then revise
-them.
-
-Jupyter notebooks are used by a variety of users ranging from scientists who implement complex
-models of physical systems to journalists who load data, perform simple aggregations and create visualizations. Our
-initial focus is on simple use cases, such as those of journalists. Tooling that makes notebook
-systems more interactive and spreadsheet-like can encourage users to choose programming tools
-over spreadsheets and produce reproducible and transparent data analyses.
-
-As an example, the analysis done by the Financial Times for an article on plastic waste
-\cite{ftnotebooks,ftarticle} joins datasets from Eurostat, UN Comtrade and more, aggregates the data and
+Data analysts increasingly use notebook systems such as Jupyter, which make it possible to
+combine text, equations and code with results of running the code, such as tables or visualizations.
+Notebooks blur the conventional distinction between development and execution. Data analysts write
+small snippets of code, run them to see results immediately and then revise them.
+
+Notebooks are used by users ranging from scientists who implement complex models of physical
+systems to journalists who perform simple data aggregations and create visualizations. Our
+focus is on the simplest use cases. Making programmatic data exploration more
+spreadsheet-like should encourage users to choose programming tools over spreadsheets, resulting
+in more reproducible and transparent data analyses.
+
+Consider the Financial Times analysis of plastic waste \cite{ftnotebooks,ftarticle}. It joins
+datasets from Eurostat, UN Comtrade and more, aggregates the data and
 builds a visualization comparing waste flows in 2017 and 2018. Figure~\ref{fig:ft-uncomtrade} shows
-an extract from one notebook from the data analysis. The code has a number of remarkable properties:
-%
-\begin{itemize}[itemsep=3pt]
+an excerpt from one notebook of the data analysis. The code has a number of important properties:
+%
+\begin{itemize}
 \item There is no abstraction. The analysis uses lambda functions as arguments to library calls,
-  but it does not define any top-level functions. Parameterization is done by defining a list of
-  inputs and then having a for loop, or by defining a top-level variable and setting its value.
-  In our example, \ident{material} is set to \str{plastics} and a comment lists other options.
-  This lets the analyst easily see and check the results of intermediate steps of the analysis.
+  but it does not define any custom functions. Code is parameterized by having a global variable
+  \ident{material} set to \str{plastics} and keeping other possible values in a comment.
+  This lets the analyst run and check results of intermediate steps.
 
@@
   which provides operations for data wrangling such as \ident{merge} to join datasets
-  or \ident{drop\_duplicates} to delete rows with duplicate column values. Those libraries are
-  external to the data analysis and may even be implemented in other languages -- many Python
-  libraries rely on C++.
+  or \ident{drop\_duplicates} to delete rows with duplicate column values. Such standard libraries
+  are external to the data analysis and are often implemented in another language like C++.
 
@@
 
-\item There are many corner cases, such as the fact that the \ident{keep\_default\_na} parameter
-  needs to be set to handle Namibia correctly. These are discovered interactively by writing and
-  running a code snippet and examining the output, so providing a rapid feedback is essential.
+\item There are many corner cases, such as the fact that \ident{keep\_default\_na}
+  needs to be set to handle Namibia correctly. These are discovered interactively by
+  running the code and examining the output, so providing an instant feedback is essential.
 \end{itemize}
@@
 \noindent
-Many Jupyter notebooks are more complex than the example discussed here. They might define
-helper functions or even structure code using object-oriented programming. However, this paper
-is motivated by our conviction that simple data analyses such as the one discussed here are frequent
-enough to pose an interesting and important domain for programming tools research.
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure}[b]
+Many Jupyter notebooks are more complex than the above example and might use helper
+functions or object-oriented code. However, simple data analyses such as the one discussed
+here are frequent enough and pose interesting problems for programming tools.
+This paper aims to bring such analyses to the attention of programming research community
+by capturing their essential properties as a formal calculus.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}
+  \vspace{-1.5em}
   \centering
   \subcaptionbox{
-    \raggedright\footnotesize The Gamma script to aggregate Olympic medals. The analyst selects
-      Rio 2016 games and counts the number of distinct events per athlete. After typing `.',
-      a type provider offers further aggregation operations.
+    \raggedright\footnotesize The analysis counts the number of distinct events per
+      athlete. After typing `.' the editor offers further aggregation operations.
     \label{fig:gamma-dot}
-  }{\includegraphics[scale=0.26]{gamma1.png}}
+  }{\includegraphics[scale=0.26]{figures/gamma1.png}}
   \subcaptionbox{
-    \raggedright\footnotesize Live preview in The Gamma programming environment developed based on
-      the theory presented in this paper. The table is produced as
-      the data analyst types and updates on-the-fly to show result at the current cursor position.
+    \raggedright\footnotesize
+      Our live coding environment for The Gamma. The table is updated on-the-fly a
+      and shows the result at the current cursor position.
     \label{fig:gamma-preview}
-  }{\includegraphics[scale=0.26]{gamma2.png}}
-  \caption{\small Data exploration in The Gamma -- using a type provider (left) and with a live
-    preview (right).}
+  }{\includegraphics[scale=0.26]{figures/gamma2.png}}
+  \caption{\small Previous work on The Gamma (left) and our new editor with live preview (right).}
   \label{fig:gamma-intro}
+  \vspace{-0.5em}
 \end{figure}
+
+% --------------------------------------------------------------------------------------------------
 
@@
 
-Data exploration of the kind discussed in the previous section has been the motivation for a
-recent work by \citet{gamma} that develops a small scripting language called The Gamma. The language
-follows the structure discussed above. A script is a sequence of commands that can either define a
-variable or produce an output. The language does not support top-level function declarations
-and lambda functions can be used only as method arguments.
+Simple data exploration as discussed in the previous section has been the motivation for
+a scripting language The Gamma \cite{gamma}. Scripts in The Gamma are sequences of commands
+that either define a variable or produce an output. It does not support top-level
+functions and lambda functions can be used only as method arguments.
 
 Given the limited expressiveness of The Gamma, libraries are implemented in other languages,
-such as JavaScript. The Gamma also supports type providers \cite{providers-fsharp},
-which can be used to generate types, on the fly, for accessing and exploring external data sources.
-Type providers provide object types with members and The Gamma makes using those convenient by
+such as JavaScript. The Gamma uses type providers \cite{providers-fsharp}
+for accessing and exploring external data sources.
+Type providers generate object types with members and The Gamma makes using those convenient by
 providing auto-complete when the user types the dot symbol (`.') to access a member.
-
-As illustrated by \citet{gamma}, the combination of type providers and auto-complete on~`.' makes it
-possible to complete a large number of data exploration tasks through a very simple interaction of
-selecting operations from a list. An example in Figure~\ref{fig:gamma-dot} summarizes data on
-Olympic medals using a type provider. The names of identifiers such as \qident{sum Bronze} do not have
-any special meaning. They are merely members generated by the type provider, based on information
-about the data source. The type provider used in this example generates an object with members for
-different data transformations, such as \qident{group data}, which return further objects with
-members for specifying transformation parameters, such as selecting the grouping key using
-\qident{by Athlete}.
+The combination of type providers and auto-complete makes it
+possible to solve a large number of data exploration tasks through the very simple interaction of
+selecting operations from a list.
+
+The example in Figure~\ref{fig:gamma-dot} summarizes data on Olympic medals. Identifiers such
+as \qident{sum Bronze} are names of members generated by the type provider. The type provider
+used in this example generates an object with members for data transformations such as
+\qident{group data}, which return further objects with members for specifying transformation
+parameters, such as selecting the grouping key using \qident{by Athlete}.
 
@@
 produced by journalists, economists and other users with other than programming background.
-When we refer to The Gamma in this paper, readers not familiar with it can assume that we focus
-on a subset of Python.
+When we refer to The Gamma in this paper, readers not familiar with it can consider a small
+subset of Python.
 
@@
 
-The practical implementation work that complements the theory presented in this paper implements a live
-coding environment for The Gamma and is discussed in Section~\ref{sec:evaluation-case} in detail. Earlier work by
-\citet{gamma} discussed The Gamma language and the type provider mechanism, but the implementation
-provided only a standard text editor with auto-completion. Our work brings the experience of using
-The Gamma closer to that of spreadsheets by developing an editor that evaluates scripts
-instantaneously in background and displays the results as \emph{live previews}.
+The implementation that accompanies this paper builds a live coding environment for The Gamma.
+The implementation is discussed in Section~\ref{sec:evaluation-case} and it replaces
+the original text editor with just auto-complete with a live coding environment
+that provides \emph{live previews}. We choose to extend The Gamma over using a subset of Python
+as the simplicity of TheGamma lets us focus on the principles of live coding systems.
 
 An example of a live preview is shown in Figure~\ref{fig:gamma-preview}. As noted earlier, The
-Gamma program consist of a list of commands which are either expressions or let bindings. Our
+Gamma programs consist of lists of commands which are either expressions or let bindings. Our
 editor displays a live preview below the command that the user is currently editing. The preview
@@
 
-The \emph{data exploration calculus} is a small, formally tractable language for data exploration.
-The calculus is not Turing-complete and it can only be used together with external libraries that
+The \emph{data exploration calculus} is a small formal language for data exploration. The calculus
+is not, in itself, Turing-complete and it can only be used together with external libraries that
 define what objects are available and what the behaviour of their members is. This is
 sufficient to capture the simple data analyses discussed in Section~\ref{sec:background}. We
-define the calculus in this section and then use it as the basis for discussing our live preview
+define the calculus in this section and then use it to formalise our live preview
 mechanism in Section~\ref{sec:formal}.
-
-We initially present the data exploration calculus without a static type system as the live preview
-mechanism does not rely on types. We consider type checking later in Section~\ref{sec:extra-types}
-as the mechanism that is used for live previews can also be used for instantaneous type checking.
+%
+The live preview mechanism does not rely on types and so we postpone the discussion of static
+typing to Section~\ref{sec:extra-types}. Interestingly, it reuses the mechanism used for live
+previews.
 
@@
 \begin{figure}
+\vspace{-0.5em}
 \raggedright
-\hspace{1.5em}{\small\sffamily Programs, commands, terms, expressions and values}
+\hspace{0.5em}{\small\sffamily Programs, commands, terms, expressions and values}
+\vspace{-0.5em}
 \begin{equation*}
 \begin{array}{lcl}
-p &\narrow{=}& c_1; \ldots; c_n\\
-c &\narrow{=}& \kvd{let}~x = t \lsep t\\
+p &\narrow{::=}& c_1; \ldots; c_n\\
+c &\narrow{::=}& \kvd{let}~x = t \lsep t\\
 \end{array}
@@
 \vspace{0.5em}
-\hspace{1.5em}{\small\sffamily Evaluation contexts of expressions}
+\hspace{0.5em}{\small\sffamily Evaluation contexts of expressions}
+\vspace{-0.5em}
 \begin{equation*}
@@
 \vspace{0.5em}
-\hspace{1.5em}{\small\sffamily Let elimination and member reduction}
+\hspace{0.5em}{\small\sffamily Let elimination and member reduction}
+\vspace{-0.5em}
 \begin{equation*}
@@
 \label{fig:dec-calculus}
+\vspace{-0.5em}
 \end{figure}
@@
 
-A program $p$ in the data exploration calculus consists of a sequence of commands $c$. A command
+A program $p$ in the data exploration calculus consists of a sequence of commands~$c$. A command
 can be either a let binding or a term. Let bindings define variables $x$ that can be used in
-subsequent commands. The calculus supports lambda functions, but they
+subsequent commands. Lambda functions
 can only appear as arguments in method calls. A term $t$ can be a value, variable or a
 member access, while an expression $e$, which can appear as an argument in member access,
-can be a lambda function or a term.\footnote{Similar but weaker restrictions on the use of lambda
-functions exist in other languages. For example, in order to guide type inference, lambda functions
-in C\# can appear as either method arguments or assigned to an explicitly typed variable, but they
-cannot be assigned to an implicitly typed variable.}
+can be a lambda function or a term.
+
+%\footnote{Similar but weaker restrictions
+%exist in other languages. To guide type inference, lambda functions
+%in C\# can appear as either method arguments or assigned to an explicitly typed variable, but they
+%cannot be assigned to an implicitly typed variable.}
 
@@
 \item[-] We write $\rightsquigarrow^{*}$ for the reflexive, transitive closure of $\rightsquigarrow$
-\item[-] We write $\rightsquigarrow_\ident{let}$ for a call-by-name let binding elimination\\
-  $c_1;\; \ldots;\; c_{k-1};\; \kvd{let}~x=t;\; c_{k+1};\; \ldots; c_n ~\rightsquigarrow_\ident{let}~
+\item[-] We write $\rightsquigarrow_\ident{let}$ for a call-by-name let binding elimination
+  $c_1;\; \ldots;\; c_{k-1};\;$\\ $\kvd{let}~x=t;\; c_{k+1};\; \ldots; c_n ~\rightsquigarrow_\ident{let}~
   c_1;\; \ldots;\; c_{k-1};\; t; c_{k+1}[x\leftarrow t];\; \ldots; c_n[x\leftarrow t]$
@@
 observationally equivalent values as arguments, it should return the same value. Second, the
-evaluation of $o.m(v_1,\ldots,v_n)$ should be defined for all $o, i$ and $v_i$. The external
+evaluation of $o.m(v_1,\ldots,v_n)$ should be defined for all $o, n$ and $v_i$. The external
 library will typically satisfy this by defining an error object and reducing all invalid calls
-to the error object, following the standard method of \citet{gowrong}.
+to the error object \cite{gowrong}.
 
@@
 \begin{itemize}
-\item[-] \emph{Completeness.} For all $o, m, i$ and all $v_1, \ldots, v_i$, there exists $o'$ such
+\item[-] \emph{Totality.} For all $o, m, i$ and all $v_1, \ldots, v_i$, there exists $o'$ such
   that $o.m(v_1, \ldots, v_i) \rightsquigarrow_\epsilon o'$.
-\vspace{0.5em}
-\item[-] \emph{Compositionality.} For observationally equivalent arguments, the reduction returns
+\item[-] \emph{Compositionality.} For observationally equivalent arguments, the reduction should always return
   the same object, i.e.~given $e_0, e_1, \ldots, e_n$ and $e'_0, e'_1, \ldots, e'_n$ and $m$ such that
@@
 \noindent
-The compositionality condition is essential for proving the correctness of our live preview mechanism.
-Completeness allows us to prove normalization, i.e.~all programs reduce to
+Compositionality is essential for proving the correctness of our live preview mechanism.
+Totality allows us to prove normalization, i.e.~all programs reduce to
 a value -- although the resulting value may be an error value provided by the external library.
@@
 \label{thm:normalization}
-For all $p$, there exists $n$ and $o_1, \ldots, o_n$ such that $p\rightsquigarrow^{*} o_1;\ldots;o_n$.
+For all $p$, there exists $n, o_1, \ldots, o_n$ such that $p\rightsquigarrow^{*} o_1;\ldots;o_n$.
 \end{theorem}
@@
 \begin{proof}
-The elimination of let binding transforms a program $c_1;\; \ldots;\; c_{k-1};\; \kvd{let}~x=t;\; c_{k+1};\; \ldots; c_n$
-to a program $c_1;\; \ldots;\; c_{k-1};\; t; c_{k+1}[x\leftarrow t];\; \ldots; c_n[x\leftarrow t]$.
-The reduction steps for the new program can be constructed using the steps of $p \rightsquigarrow^{*} o_1;\ldots;o_n$.
-The new command $t$ reduces to an object $o$ using the same steps as the original term $t$
-in $\kvd{let}~x=t$ but with context $C_c = -$ rather than $C_c = \kvd{let}~x=-$; the terms $t$
-introduced by substitution also reduce using the same steps as before, but using
-contexts in which the variable $x$ originally appeared.
+By constructing $p' \rightsquigarrow^{*} o_1;\ldots;o_n$ from
+$p \rightsquigarrow^{*} o_1;\ldots;o_n$. See Appendix~\ref{sec:let-lang-elimination}.
 \end{proof}
 
-\newpage
-
 % ==================================================================================================
 
-\section{Formalising live coding environment}
+\section{Formalising a live coding environment}
 \label{sec:formal}
@@
 A naive way of providing live previews during code editing would be to re-evaluate the code
-after each change. This would be wasteful -- for example, when writing
-code, most changes are additive and the preview can be updated by evaluating just the newly added
-code. In this section, we develop foundations for an efficient way of displaying live previews
-for the data exploration calculus.
+after each change. This would be wasteful -- when writing code to explore data, most changes are
+additive. To update a preview, we only need to evaluate newly added code. We describe an
+efficient mechanism in this section.
 
@@
 When binding a new expression to the graph, we reuse previously created nodes as long as
-they have the same dependencies. For expressions that have a new structure, we create new nodes.
-
-The nodes of the graph serve as unique keys into a lookup table with previously
+they have the same structure and the same dependencies. For expressions that have a new
+structure, we create new nodes.
+
+The nodes of the graph serve as unique keys into a lookup table containing previously
 evaluated parts of the program. When a preview is requested for an expression, we use the graph
@@
 
-\begin{figure}[b]
+\begin{figure}
 \centering
@@
   \raggedright\small
-  Graph constructed from the following initial expression:
+  Graph constructed from initial expression:
   $\kvd{let}~x = \num{15}~\kvd{in}~\ident{data.skip}(\num{10}).\ident{take}(x)$
@@
 
-Furthermore, the graph edges are labelled with labels indicating the kind of dependency. For
+The graph edges are labelled with labels indicating the kind of dependency. For
 a method call, the labels are ``first argument'', ``second argument'' and so on. Writing
@@
 \begin{array}{rcll}
-\bndclr{v}&=&\bnd{val}(o) \lsep \bnd{var}(x)\lsep \bnd{mem}(m, s)\lsep \bnd{fun}(x, s)&(\textit{Vertices})\\
-\blblclr{l}&=&\blbl{body} \lsep \blbl{arg}(i)&(\textit{Edge labels})\\
+\bndclr{v}&=&\bnd{val}(o) \lsep \bnd{var}(x)\lsep \bnd{mem}(m, s)\lsep \bnd{fun}(x, s)&\qquad(\textit{Vertices})\\
+\blblclr{l}&=&\blbl{body} \lsep \blbl{arg}(i)&\qquad(\textit{Edge labels})\\
 \end{array}
@@
 different dependencies will contain a different symbol. Dependencies of member access are labelled
-with \blbl{arg} indicating the index of the argument (the instance has index $0$ and arguments
-start with $1$). Finally, nodes \bnd{fun} and \bnd{var} represent function values and variables
-bound by $\lambda$ abstraction. For simplicity, we use variable names rather than de Bruijn
-indices and so renaming a bound variable forces recomputation.
+with \blbl{arg} indicating the index of the argument ($0$ for the instance and $1,\ldots$ for the
+arguments). Finally, nodes \bnd{fun} and \bnd{var} represent function values and variables
+bound by $\lambda$ abstraction.
 
@@
 \subsubsection{Example graph}
-Figure~\ref{fig:dep-graph} illustrates how we construct and update the
+Figure~\ref{fig:dep-graph} illustrates how we build the
 dependency graph. Node representing $\ident{take}(x)$ depends on the argument -- the
 number $\num{15}$ -- and the instance, which is a node representing $\ident{skip}(\num{10})$.
-This, in turn, depends on the instance \ident{data} and the number $\num{10}$. Note that variables
+This depends on the instance \ident{data} and the number $\num{10}$. Note that variables
 bound via \kvd{let} binding such as $x$ do not appear as $\bnd{var}$ nodes. The node using it
-depends directly on the node representing the expression that is assigned to $x$.
+depends directly on the node representing the expression assigned to $x$.
 
@@
 \subsubsection{Reusing graph nodes}
-The binding process takes an expression and constructs a dependency graph reusing existing nodes
-when possible. For this, we use a lookup table of member access and function value nodes. The key
-in the lookup table is formed by a node kind (for disambiguation) together with a list of
-dependencies. A node kind is a member access containing the member name or a function containing
-the bound variable name; a lookup table $\Delta$ then maps a node kind with a list of dependencies
-to a cached node:
+The binding process takes an expression and constructs a dependency graph.
+It uses a lookup table to reuse previously created member access and function value nodes. The key
+in the lookup table is formed by a node kind together with a list of dependencies. A node kind
+includes the member or variable name; a lookup table $\Delta$ then maps a node kind with a list
+of dependencies to a cached node:
 %
@@
 $\bndclr{v_1}, \ldots, \bndclr{v_n}$ labelled with labels $\blblclr{l_1}, \ldots, \blblclr{l_n}$.
-We write $\Delta(\bndclr{k}, l)\!\downarrow$ when a value for a given key is not defined.
-For example, when creating the graph in Figure~\ref{fig:dep-graph-b},
-we perform the following lookup for the \ident{skip} member access:
-%
-\[ \Delta(\bknd{mem}(\ident{skip}), [(\bnd{var}(\ident{data}),\blbl{arg}(0)), (\bnd{val}(\num{10}), \blbl{arg}(1))]) \]
-%
-The lookup returns the node $\bnd{mem}(\ident{skip}, s_0)$ known from the previous step. We then perform
-the following lookup for the \ident{take} member access:
-%
-\[ \Delta(\bknd{mem}(\ident{take}), [(\bnd{mem}(\ident{skip}, s_0),\blbl{arg}(0)), (\bnd{val}(\num{10}), \blbl{arg}(1))]) \]
-%
-In the previous graph, the argument of \ident{take} was $\num{15}$ rather than $\num{10}$ and so
-this lookup fails. We then construct a new node $\bnd{mem}(\ident{take}, s_2)$ using a fresh
-symbol $s_2$.
+We write $\Delta(\bkndclr{k}, l)\!\downarrow$ when a value for a key is not defined.
+When creating the graph in Figure~\ref{fig:dep-graph-b},
+we perform the following lookups:
+%
+\begin{equation*}
+\begin{array}{ll}
+\Delta(\bknd{mem}(\ident{skip}), [(\bnd{var}(\ident{data}),\blbl{arg}(0)), (\bnd{val}(\num{10}), \blbl{arg}(1))]) &\qquad(1)\\
+\Delta(\bknd{mem}(\ident{take}), [(\bnd{mem}(\ident{skip}, s_0),\blbl{arg}(0)), (\bnd{val}(\num{10}), \blbl{arg}(1))]) &\qquad(2)
+\end{array}
+\end{equation*}
+%
+First, we look for the \ident{skip} member access. The result is the $\bnd{mem}(\ident{skip}, s_0)$
+known from the previous step. We then look for the \ident{take} member access. In the earlier
+step, the argument of \ident{take} was $\num{15}$ rather than $\num{10}$ and so this lookup fails.
+We then construct a new node $\bnd{mem}(\ident{take}, s_2)$ and later add it to the cache.
 
@@
 \begin{figure*}[t]
-\begin{equation*}
-\hspace{-3.5em}
+\vspace{-1em}
+\begin{equation*}
+\hspace{-1.5em}
 \begin{array}{ll}
-\ident{bind-expr}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n)) = \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n) &(1)~~\\
+\ident{bind-expr}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n)) = \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n) &\hspace{-2em}(1)~~\\
 \quad \textnormal{when}~\bndclr{v_i}, (V_i, E_i) = \ident{bind-expr}_{\Gamma, \Delta}(e_i)
@@
 \quad \textnormal{let}~E = \{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v}, \bndclr{v_n}, \blbl{arg}(n))\}
-\\[1em]
-\ident{bind-expr}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n)) = \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n)&(2)\\
+\\[0.75em]
+\ident{bind-expr}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n)) = \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n)&\hspace{-2em}(2)\\
 \quad \textnormal{when}~\bndclr{v_i}, (V_i, E_i) = \ident{bind-expr}_{\Gamma, \Delta}(e_i)
@@
 ~\textnormal{and}~E = \{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v},\bndclr{v_n}, \blbl{arg}(n)) \}
-\\[1em]
-\ident{bind-expr}_{\Gamma, \Delta}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V, \{e\} \cup E)&(3)\\
+\\[0.75em]
+\ident{bind-expr}_{\Gamma, \Delta}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V, \{e\} \cup E)&\hspace{-2em}(3)\\
 \quad \textnormal{when}~\Gamma_1 = \Gamma \cup \{ x, \bnd{var}(x) \}
@@
 \quad \textnormal{let}~e = (\bndclr{v}, \bndclr{v_0}, \blbl{body})
-\\[1em]
-\ident{bind-expr}_{\Gamma, \Delta}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V, \{e\} \cup E)&(4)\\
+\\[0.75em]
+\ident{bind-expr}_{\Gamma, \Delta}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V, \{e\} \cup E)&\hspace{-2em}(4)\\
 \quad \textnormal{when}~\Gamma_1 = \Gamma \cup \{ x, \bnd{var}(x) \}
@@
 ~\textnormal{and}~e = (\bndclr{v},\bndclr{v_0},\blbl{body})
-\\[1em]
-\ident{bind-expr}_{\Gamma, \Delta}(o) = \bnd{val}(o), (\{ \bnd{val}(o) \}, \emptyset) &(5)
-\\[1em]
-\ident{bind-expr}_{\Gamma, \Delta}(x) = \bndclr{v}, (\{ \bndclr{v} \}, \emptyset)~\textnormal{when}~\bndclr{v} = \Gamma(x)&(6)
-\end{array}
-\end{equation*}
+\\[0.75em]
+\ident{bind-expr}_{\Gamma, \Delta}(o) = \bnd{val}(o), (\{ \bnd{val}(o) \}, \emptyset) &\hspace{-2em}(5)
+\\[0.75em]
+\ident{bind-expr}_{\Gamma, \Delta}(x) = \bndclr{v}, (\{ \bndclr{v} \}, \emptyset)~\textnormal{when}~\bndclr{v} = \Gamma(x)&\hspace{-2em}(6)
+\end{array}
+\end{equation*}
+\vspace{-1em}
 \caption{Binding rules that define a construction of a dependency graph for an expression.}
@@
 \begin{equation*}
-\hspace{-3.5em}
+\vspace{-0.5em}
+\hspace{-1.5em}
 \begin{array}{ll}
 \ident{bind-prog}_{\Gamma, \Delta}(\kvd{let}~x=e; c_2; \ldots; c_n) = \bndclr{v_1};\ldots;\bndclr{v_n}, (\{\bndclr{v_1}\} \cup V \cup V_1, E \cup E_1)
-  \hspace{9.35em}&(7)\\
+  \hspace{9.35em}&\hspace{2.5em}(7)\\
 \quad \textnormal{let}~\bndclr{v_1}, (V_1,E_1) = \ident{bind-expr}_{\Gamma,\Delta}(e_1)~\textnormal{and}~\Gamma_1 = \Gamma \cup \{(x,\bndclr{v_1})\} \\
 \quad \textnormal{and}~\bndclr{v_2}; \ldots; \bndclr{v_n}, (V, E) = \ident{bind-prog}_{\Gamma_1, \Delta}(c_2; \ldots; c_n)
-\\[1em]
-\ident{bind-prog}_{\Gamma, \Delta}(e; c_2; \ldots; c_n) = \bndclr{v_1};\ldots;\bndclr{v_n}, (\{\bndclr{v_1}\} \cup V \cup V_1, E \cup E_1)&(8)\\
+\\[0.75em]
+\ident{bind-prog}_{\Gamma, \Delta}(e; c_2; \ldots; c_n) = \bndclr{v_1};\ldots;\bndclr{v_n}, (\{\bndclr{v_1}\} \cup V \cup V_1, E \cup E_1)&\hspace{2.5em}(8)\\
 \quad \textnormal{let}~\bndclr{v_1}, (V_1,E_1) = \ident{bind-expr}_{\Gamma,\Delta}(e)
 ~\textnormal{and}~\bndclr{v_2}; \ldots; \bndclr{v_n}, (V, E) = \ident{bind-prog}_{\Gamma_1, \Delta}(c_2; \ldots; c_n)
-\\[1em]
-\ident{bind-prog}_{\Gamma, \Delta}([]) = [], (\emptyset, \emptyset)&(9)
-\end{array}
-\end{equation*}
+\\[0.75em]
+\ident{bind-prog}_{\Gamma, \Delta}([]) = [], (\emptyset, \emptyset)&\hspace{2.5em}(9)
+\end{array}
+\end{equation*}
+\vspace{-1em}
 \caption{Binding rules that define a construction of a dependency graph for a program.}
 \label{fig:binding-rules-prog}
+\vspace{-0.5em}
 \end{figure*}
@@
 
-After parsing updated code, we update the dependency graph and link each node of the abstract
+After parsing modified code, we update the dependency graph and link each node of the abstract
 syntax tree to a node of the dependency graph. This process is called
@@
 and \ident{bind-prog} function (Figure~\ref{fig:binding-rules-prog}). Both functions are
-annotated with a lookup table $\Delta$ discussed in Section~\ref{sec:formal-deps} and a variable
-context $\Gamma$. The variable context is a map from variable names to dependency graph nodes and
+annotated with a lookup table $\Delta$ and a variable
+context~$\Gamma$. The variable context is a map from variable names to dependency graph nodes and
 is used for variables bound using \kvd{let} binding.
 
-When applied on an expression $e$, binding $\ident{bind-expr}_{\Gamma,\Delta}(e)$ returns
-a node $\bndclr{v}$ corresponding to the expression $e$ paired with a dependency graph $(V, E)$.
-In the graph, $V$ is a set of nodes $\bndclr{v}$ and $E$ is a set of labelled edges
-$(\bndclr{v_1}, \bndclr{v_2}, \blblclr{l})$. We attach the label directly to the edge rather than
-keeping a separate colouring function as this makes the formalisation simpler.
-The $\ident{bind-prog}_{\Gamma,\Delta}$ function works similarly, but takes a sequence of
-commands and returns a sequence of nodes.
-
-\subsubsection{Binding expressions.} When binding member access, we use \ident{bind-expr} recursively
+When invoked, $\ident{bind-expr}_{\Gamma,\Delta}(e)$ returns a node $\bndclr{v}$ that corresponds
+to the expression~$e$, paired with a dependency graph $(V, E)$ formed by nodes $V$ and labelled
+edges $E$. That edges are written as $(\bndclr{v_1}, \bndclr{v_2}, \blblclr{l})$ and include a
+label $\blblclr{l}$. The $\ident{bind-prog}_{\Gamma,\Delta}$ function works similarly, but turns
+a sequence of commands into a sequence of nodes.
+
+When binding a member access, we use \ident{bind-expr} recursively
 to get a node and a dependency graph for each sub-expression. The nodes representing sub-expressions are
-then used as dependencies for lookup into $\Delta$, together with their labels. If a node already
+then used for lookup into $\Delta$, together with their labels. If a node already
 exists in $\Delta$ it is reused (1). Alternatively, we create a new node containing a fresh symbol~(2).
-
-If a lambda function uses its argument, we will not be able to evaluate its body. In this case, the
-graph node bound to a function will depend on a synthetic node $\bnd{var}(x)$ that represents the
-variable with unknown value. When binding a function, we create the synthetic variable and add it
+%
+The graph node bound to a function depends on a synthetic node $\bnd{var}(x)$ that represents a
+variable of unknown value. When binding a function, we create a variable node and add it
 to the variable context $\Gamma_1$ before binding the body. As with member access, the node
-representing a function may (3) or may not (4) be already present in the lookup table.
-
-\subsubsection{Binding programs.} When binding a program, we bind the first command and then
-recursively process remaining commands until we reach an empty list of commands (9).
-For \kvd{let} binding (7), we bind the expression $e$ assigned to the variable to obtain a
-graph node $\bndclr{v_1}$. We then bind the remaining commands using a variable context $\Gamma_1$
-that maps the value of the variable to the graph node $\bndclr{v_1}$. The variable context is used
-when binding a variable (6) and so all variables declared using \kvd{let} binding will be bound to
-a graph node representing the value assigned to the variable. When the command is just an
-expression (8), we bind the expression using \ident{bind-expr}.
+representing a function may (3) or may not (4) already exist in the lookup table.
+
+When binding a program, we bind the first command and
+recursively process remaining commands (9). For \kvd{let} binding (7), we bind the expression $e$
+assigned to the variable to obtain a graph node $\bndclr{v_1}$. We then store the node in the
+variable context $\Gamma_1$ and bind the remaining commands. The variable context is used
+when binding a variable in \ident{bind-expr} (6) and so all variables declared using \kvd{let}
+will be bound to a graph node representing the value assigned to the variable. When the command
+is just an expression (8), we bind the expression using \ident{bind-expr}.
 
@@
 
-The binding process formalised in Section~\ref{sec:formal-bind} specifies how to update the
-dependency graph after updated program text is parsed. During live coding, this is done
-repeatedly as the programmer edits code. Throughout the process, we maintain a series of
-lookup table states $\Delta_0, \Delta_1, \Delta_2, \ldots$ The initial lookup table is
-empty, i.e.~$\Delta_0 = \emptyset$. At a step $i$, we parse a program $p_i$ (consisting of
-several commands) and obtain a new dependency graph using the previous $\Delta$. The result is
+During editing, the dependency graph is repeatedly updated according to the binding rules.
+We maintain a series of lookup table states $\Delta_0, \Delta_1, \Delta_2, \ldots$ The initial
+lookup table is empty, i.e.~$\Delta_0 = \emptyset$. At a step $i$, we parse a program $p_i$
+and obtain a new dependency graph using the previous $\Delta$. The result is
 a sequence of nodes corresponding to commands of the program and a graph $(V, E)$:
@@
 %
-The new state of the cache is computed by calling the $\ident{update}_{V, E}(\Delta_{i-1})$ function
+The new state of the cache is computed using $\ident{update}_{V, E}(\Delta_{i-1})$
 defined in Figure~\ref{fig:loop}. The function adds newly created nodes from the graph
-$(V, E)$ to the previous cache $\Delta_{i-1}$ and returns a new cache $\Delta_{i}$. We only cache
-nodes for function and member accesses -- nodes for variables and primitive values will remain
-the same thanks to the way they are constructed.
+$(V, E)$ to the previous cache $\Delta_{i-1}$ and returns a new cache $\Delta_{i}$.
 
@@
 
-The binding process described in the previous section constructs a dependency graph after code
-changes. The nodes in the dependency graph correspond to individual operations that will be performed
-when running the program. When evaluating a preview, we attach (partial) results to
-nodes of the graph. Since the binding process reuses nodes when their dependencies do not change,
-previews for expressions (or sub-expressions) of a program can be reused when updating a preview.
+The binding process constructs a dependency graph after code changes. The nodes in the dependency
+graph correspond to individual operations that will be performed when running the program. When
+evaluating a preview, we attach partial results to nodes of the graph. Since the binding process
+reuses nodes, previews for sub-expressions attached to graph nodes will also be reused.
 
 In this section, we describe how previews are evaluated. The evaluation is done over the dependency
-graph, rather than over the structure of the program expressions as in the operational semantics
-given in Section~\ref{sec:calculus-semantics}. We analyse the preview evaluation formally in
-Section~\ref{sec:evaluation} and show that the resulting previews are the same as the result we
-would get by directly evaluating the code and, also, that no recomputation occurs when code is
-edited in certain ways.
+graph, rather than over the structure of program expressions as in the operational semantics
+given in Section~\ref{sec:calculus-semantics}. In Section~\ref{sec:evaluation}, we prove that
+resulting previews are the same as the result we would get by directly evaluating code and we
+also show that no recomputation occurs when code is edited in certain ways.
 
@@
 a \emph{delayed preview} is a partially evaluated expression with free variables:
-
+%
 \begin{equation*}
@@
 variables. A possibly delayed preview $d$ can be either an evaluated preview $p$ or an expression
-$e$ that requires variables $\Gamma$. For simplicity, we use an untyped language and so $\Gamma$
-is just a list of variables $x_1, \ldots, x_n$.
-
-A delayed preview is not necessarily the body of a lambda function as it appears in the source
-code. We partially evaluate sub-expressions of the body that do not have free variables or that
-have free variables bound by an earlier let binding.
-
-Our implementation does not currently display
-delayed previews to the user, but there is a number of possible approaches for doing that.
-Most interestingly, since lambda functions always appear as arguments of a member access, we could
-force the evaluation of the surrounding expression, capture (a number of) values passed as inputs
-to the lambda function and display a preview based on those. Finally, in Section~\ref{sec:extra-abstraction},
-we also consider a more speculative design for an abstraction mechanism that supports live previews
-that could, in some cases, replace lambda function.
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure*}[b]
-\begin{equation*}
-\hspace{-2.5em}
+$e$ that requires variables $\Gamma$. We use an untyped language and so $\Gamma$ is just a list of
+variables $x_1, \ldots, x_n$. As discussed in Appendix~\ref{sec:app-theories},
+delayed previews have an interesting theoretical link with graded comonads.
+The body of a lambda function may have a fully evaluated preview if it uses only variables that
+are bound by earlier let bindings, but it will typically be delayed. We consider a speculative
+design for an abstraction mechanism that better supports live previews in
+Appendix~\ref{sec:extra-abstraction}.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure*}
+\begin{equation*}
+\vspace{-2em}
+\hspace{-0.5em}
 \begin{array}{l}
@@
 \begin{equation*}
-\hspace{-2.5em}
+\hspace{-1em}
 \begin{array}{l}
@@
 \end{equation*}
-
+\vspace{-0.5em}
 \caption{Rules that define evaluation of previews over a dependency graph for a program}
 \label{fig:eval}
+\vspace{-0.5em}
 \end{figure*}
@@
 For simplicity, the relation $\Downarrow$ in Figure~\ref{fig:eval} does not specify how previews
-are cached and linked to graph nodes. In practice, this is done by maintaining a lookup table
-from graph nodes $\bndclr{v}$ to (possibly delayed) previews $p$.
-Whenever $\Downarrow$ is used to obtain a preview for a graph node, we first
-attempt to find an already evaluated preview using the lookup table. If the preview has not
-been previously evaluated, we evaluate it and add it to the lookup table.
-
-The cached evaluated previews can be reused in two ways. First, multiple nodes can
-depend on one sub-graph in a single dependency graph (if the same sub-expression appears
-twice in the program). Second, the keys of the lookup table are graph nodes and nodes are
-reused when a new dependency graph is constructed after the user edits the source code.
-
-\subsection{Theories of delayed previews}
+are cached. In practice, this is done by maintaining a lookup table from graph nodes $\bndclr{v}$
+to previews $p$. Whenever $\Downarrow$ is used to obtain a preview for a graph node, we first
+check the lookup table. If the preview has not been previously evaluated, we evaluate it and add
+it to the lookup. Cached previews can be reused in two ways. First, if the same sub-expression
+appears multiple times in the program, it will share a graph node and the preview will be resued.
+Second, when binding modified source code, the process reuses graph nodes and so previews are
+also reused during code editing.
+
+% ==================================================================================================
+
+\section{Evaluating live coding environment}
+\label{sec:evaluation}
+
+Computing previews using a dependency graph implements a correct and efficient optimization.
+In this section we show that this is the case, first theoretically in Section~\ref{sec:evaluation-reuse},
+and then empirically in Section~\ref{sec:evaluation-empirical}. We also describe a case study where
+we developed an online service for data exploration based on the methods discussed in this paper
+(Section~\ref{sec:evaluation-case}).
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Correctness of previews}
+\label{sec:evaluation-correctness}
+
+To show that the previews are correct, we prove two properties. Correctness (Theorem~\ref{thm:correcntess})
+guarantees that, the previews we calculate using a dependency graph are the same as the values we
+would obtain by evaluating the program directly. Determinacy (Theorem~\ref{thm:determinacy})
+guarantees that previews assigned to a graph node based on an earlier graph are the same as
+previews that we would obtain afres using an updated graph.
+
+To simplify the proofs, we consider programs without let bindings. Eliminating let bindings does
+not change the result of evaluation, as shown in Lemma~\ref{thm:let-lang-elimination}, and it also
+does not change the constructed dependency graph as shown below in Lemma~\ref{thm:let-grp-elimination}.
+
+\begin{lem}[Let elimintion for a dependency graph]
+\label{thm:let-grp-elimination}
+Given programs $p_1, p_2$ such that $p_1 \rightsquigarrow_\ident{let} p_2$ and a lookup table
+$\Delta_0$ then if $\bndclr{v_1}; \ldots; \bndclr{v_n}, (V, E) = \ident{bind-prog}_{\emptyset, \Delta_0}(p_1)$ and
+$\bndclr{v'_1};\ldots;\bndclr{v'_n}, (V', E') = \ident{bind-prog}_{\emptyset, \Delta_1}(p_2)$ such that $\Delta_1 = \ident{update}_{V, E}(\Delta_0)$
+then for all $i$, $\bndclr{v_i} = \bndclr{v'_i}$ and also $(V, E) = (V', E')$.
+\end{lem}
+\begin{proof}
+By analysis of the binding process. See Appendix~\ref{sec:app-let-grp-elimination}.
+\end{proof}
+
+\noindent
+The Lemma~\ref{thm:let-grp-elimination} provides a way of removing let bindings from a program,
+such that the resulting dependency graph remains the same. Here, we bind the original program
+first, which adds the node for $e$ to $\Delta$. In our implementation, this is not needed
+because $\Delta$ is updated while the graph is being constructed using $\ident{bind-expr}$.
+To keep the formalisation simpler, we separate the process of building the dependency graph
+and updating $\Delta$ and thus Lemma~\ref{thm:let-grp-elimination} requires an extra binding step.
+
+Now, we can show that, given a let-free expression, the preview obtained using a correctly
+constructed dependency graph is the same as the one we would obtain by directly evaluating the
+expression. This requires a simple auxiliary lemma.
+
+\begin{lem}[Lookup inversion]
+\label{thm:lemma-lookup}
+Given $\Delta$ obtained using \ident{update} in Figure~\ref{fig:loop} then:
+\begin{itemize}
+\raggedright
+\item[--] If $\bndclr{v}=\Delta(\bknd{fun}(x),[(\bndclr{v_0}, \blblclr{l_0})])$
+then $\bndclr{v}=\bnd{fun}(x, s)$ for some $s$.
+\item[--] If $\bndclr{v}=\Delta(\bknd{mem}(m),[(\bndclr{v_0}, \blblclr{l_0}), \ldots, (\bndclr{v_n}, \blblclr{l_n})])$
+then $\bndclr{v}=\bnd{mem}(m, s)$ for some $s$.
+\end{itemize}
+\end{lem}
+\begin{proof}
+By construction of $\Delta$ in Figure~\ref{fig:loop}.
+\end{proof}
+
+\begin{theorem}[Term preview correctness]
+\label{thm:let-free-correct}
+Given a term $t$ that has no free variables, together with a lookup table $\Delta$ obtained
+from any sequence of programs using $\ident{bind-prog}$ (Figure~\ref{fig:binding-rules-prog}) and
+$\ident{update}$ (Figure~\ref{fig:loop}), then
+let $\bndclr{v}, (V, E) = \ident{bind-expr}_{\emptyset, \Delta}(t)$.
+
+\vspace{0.25em}
+\noindent
+If $\bndclr{v}\Downarrow p$
+over a graph $(V, E)$ then $p = o$ for some value $o$ and $t \rightsquigarrow^{*} o$.
+\end{theorem}
+
+\begin{proof}
+By induction over the binding process. See Appendix~\ref{sec:app-let-free-correct}.
+\end{proof}
+
+\begin{theorem}[Program preview correctness]
+\label{thm:correcntess}
+Consider a program $p=c_1;\ldots;c_n$ that has no free variables, together with a lookup table
+$\Delta_0$ obtained from any sequence of programs using $\ident{bind-prog}$
+(Figure~\ref{fig:binding-rules-prog}) and $\ident{update}$ (Figure~\ref{fig:loop}). Assume a
+let-free program $p'=t_1;\ldots;t_n$ such that $p\rightsquigarrow^{*}_\ident{let}p'$.
+
+\vspace{0.25em}
+\noindent
+Let $\bndclr{v_1};\ldots;\bndclr{v_n}, (V, E) = \ident{bind-prog}_{\emptyset, \Delta_0}(p)$
+and define updated lookup table $\Delta_1=\ident{update}_{V, E}(\Delta_0)$ and
+let $\bndclr{v'_1};\ldots;\bndclr{v'_n}, (V', E') = \ident{bind-prog}_{\emptyset, \Delta_1}(p)$.
+
+\vspace{0.25em}
+\noindent
+If $\bndclr{v'_i}\Downarrow p_i$ over a graph $(V', E')$ then $p_i=o_i$ for some value $o_i$ and
+$t_i \rightsquigarrow o_i$.
+\end{theorem}
+\begin{proof}
+Direct consequence of Lemma~\ref{thm:let-grp-elimination} and Theorem~\ref{thm:let-free-correct}.
+\end{proof}
+
+\noindent
+Our implementation updates $\Delta$ during the recursive binding process and so a stronger version
+of the property holds: previews calculated over a graph obtained directly for the original program
+$p$ are the same as the values of the fully evaluated program. Our formalisation omits this for
+simplicity.
+
+The second important property is determinacy, which makes it possible to cache the previews
+evaluated via~$\Downarrow$ using the corresponding graph node as a lookup key.
+
+\begin{theorem}[Preview determinacy]
+\label{thm:determinacy}
+For some $\Delta$ and for any programs $p, p'$, assume that the first program is bound,
+i.e.~$\bndclr{v_1};\ldots;\bndclr{v_n}, (V, E) = \ident{bind-prog}_{\emptyset, \Delta}(p)$,
+the graph node cache is updated $\Delta' = \ident{update}_{V, E}(\Delta)$ and the second program is
+bound, i.e.~$\bndclr{v'_1};\ldots;\bndclr{v'_m}, (V', E') = \ident{bind-prog}_{\emptyset, \Delta'}(p')$.
+Now, for any $\bndclr{v}$, if $\bndclr{v} \Downarrow p$ over $(V, E)$ then also
+ $\bndclr{v} \Downarrow p$ over $(V', E')$.
+\end{theorem}
+\begin{proof}
+By induction over $\Downarrow$, show that the same evaluation rules also
+apply over $(V', E')$. This is the case, because graph nodes added to $\Delta'$ by $\ident{update}_{V, E}$
+are added as new nodes in $\ident{bind-prog}_{\emptyset, \Delta'}$ and
+nodes and edges of $(V, E)$ are unaffected.
+\end{proof}
+
+\noindent
+The cache of previews (Section~\ref{sec:previews-cache}) associates a preview $d$ with a node
+$\bndclr{v}$ as the key. Theorem~\ref{thm:determinacy} guarantees that this is valid. As we update dependency
+graph during code editing, previous nodes will continue representing the same sub-expressions.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\raggedright
+\vspace{-0.5em}
+{\small\sffamily Edit contexts of expressions}
+\begin{equation*}
+\begin{array}{lcl}
+K_e[-] &=& K_e[-].m(e_1, \ldots, e_n) ~\lsep~ e.m(e_1, \ldots, e_{l-1}, K_e[-], e_{l+1},\ldots , e_n) ~\lsep~ -\\
+K_c[-] &=& \kvd{let}~x = K_e[-] ~\lsep~ K_e[-]\\
+\end{array}
+\end{equation*}
+
+\vspace{0.5em}
+{\small\sffamily Code edit operations preserving preview for a sub-expression}
+
+\hspace{2.5em}\begin{minipage}[c]{0.868\textwidth}
+  \raggedright\setlength{\parindent}{-1em}
+
+  \ername{let-intro-var}
+  $\overline{c_1}; \preview{e}; \overline{c_2}$ changes to
+  $\overline{c_1}; \kvd{let}\,x = e; \preview{x}; \overline{c_2}$ where $x$ is fresh.
+
+  \ername{let-intro-ins}
+  $\overline{c_1}; \overline{c_2}; \preview{K_c[e]}; \overline{c_3}$ is changed to
+  $\overline{c_1}; \kvd{let}\,x=e; \overline{c_2}; \preview{K_c[x]}; \overline{c_3}$ via
+  a semantically non-equivalent expression
+  $\overline{c_1}; \overline{c_2}; K_c[x]; \overline{c_3}$ where $x$ is free.
+
+  \ername{let-intro-del}
+  $\overline{c_1}; \overline{c_2}; \preview{K_c[e]}; \overline{c_3}$ is changed to
+  $\overline{c_1}; \kvd{let}\,x=e; \overline{c_2}; \preview{K_c[x]}; \overline{c_3}$ via
+  an expression $\overline{c_1}; \kvd{let}\,x=e; \overline{c_2}; K_c[e]; \overline{c_3}$
+  with unused variable $x$.
+
+  \ername{let-elim-del}
+  $\overline{c_1}; \kvd{let}\,x=e; \overline{c_2}; \preview{K_c[x]}; \overline{c_3}$ is changed to
+  $\overline{c_1}; \overline{c_2}; \preview{K_c[e]}; \overline{c_3}$ via
+  a semantically non-equivalent expression
+  $\overline{c_1}; \overline{c_2}; K_c[x]; \overline{c_3}$ where $x$ is free.
+
+  \ername{let-elim-ins}
+  $\overline{c_1}; \kvd{let}\,x=e; \overline{c_2}; \preview{K_c[x]}; \overline{c_3}$ is changed to
+  $\overline{c_1}; \overline{c_2}; \preview{K_c[e]}; \overline{c_3}$ via
+  an expression $\overline{c_1}; \kvd{let}\,x=e; \overline{c_2}; K_c[e]; \overline{c_3}$
+  with unused variable $x$.
+
+  \ername{edit-mem}
+  $\overline{c_1}; K_c[\preview{e_0}.m(\overline{e})]; \overline{c_2}$ is changed to
+  $\overline{c_1}; K_c[\preview{e_0}.m'(\overline{e'})]; \overline{c_2}$
+
+  \ername{edit-let}
+  $\overline{c_1}; \kvd{let}\,x=e_1; \overline{c_2}; K_c[\preview{e_2}]; \overline{c_3}$ is changed to
+  $\overline{c_1}; \kvd{let}\,x=e'_1; \overline{c_2}; K_c[\preview{e_2}]; \overline{c_3}$ \hspace{2em} when
+  $x\notin FV(e_2)$.
+\end{minipage}
+
+\caption{Code edit operations that preserve previously evaluated preview}
+\label{fig:operations}
+\vspace{-0.5em}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Reuse of previews}
+\label{sec:evaluation-reuse}
+
+In this section, we identify a number of code edit operations where the previously evaluated
+values for a sub-expression can be reused. This includes the motivating example from
+Section~\ref{sec:intro} where the data analyst extracted a constant into a let binding and
+modified a parameter of the last method call in a call chain.
+
+The list of preview-preserving edits is shown in Figure~\ref{fig:operations}.
+It includes several ways of introducing and eliminating let bindings and edits where the analyst
+modifies an unrelated part of the program. The list is not exhaustive. Rather, it illustrates
+typical edits that the data analyst might perform when writing code. To express the operations we
+define an editing context $K$ which is similar to evaluation context $C$ from Figure~\ref{fig:dec-calculus},
+but allows sub-expressions appearing anywhere in the program.
+
+We use the notation $\preview{e}$ to mark parts of expressions that are not recomputed
+during the edit; we write $\overline{c}$ and $\overline{e}$ for a list of commands and expressions,
+respectively. In some of the edit operations, we also specify an intermediate program that may be
+semantically different and only has a partial preview. This illustrates a typical way of
+working with code in a text editor using cut and paste. For example, in
+\rname{let-intro-ins}, the analyst cuts a sub-expression $e$, replaces it with
+a variable $x$ and then adds a let binding for a variable $x$ and inserts the expression $e$
+from the clipboard. The \rname{let-intro-del} operation captures the same edit, but
+done in a different order.
+
+Theorem~\ref{thm:preview-reuse} proves that the operations given in Figure~\ref{fig:operations}
+preserve the preview for a marked sub-expression. It relies on the following lemma that
+generally characterizes one common kind of edits. Given two versions of a program
+that both contain the same sub-expression $e$, if the let bindings that define the values of
+variables used in $e$ do not change, then the graph node assigned to $e$ will be the same
+when binding the original and the updated program.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{lem}[Binding sub-expressions]
+\label{thm:sub-expr}
+Assume we have programs $p_1 = c_1; \ldots; c_k; K_c[e]; c_{k+1};\ldots; c_n$ and
+$p_2 = c'_1; \ldots; c'_k; K'_c[e]; c'_{k+1};\ldots;c'_n$ and $I\subseteq \{1\ldots k\}$ such
+that $\forall i\!\in\!I.\,c_i\!=\!c'_i$ and for each $x \in \bigcup_{i\in I}FV(c_i) \cup FV(e)$
+there exists $j\in I$ such that $c_j = \kvd{let}\,x=e$ for some $e$.
+%
+Given any $\Delta$, assume that the the first program is bound,
+i.e.~$\bndclr{v_1};\ldots;\bndclr{v_n}, (V, E) = \ident{bind-prog}_{\emptyset, \Delta}(p_1)$,
+the cache is updated $\Delta' = \ident{update}_{V, E}(\Delta)$ and the second
+program is bound,
+i.e.~$\bndclr{v'_1};\ldots;\bndclr{v'_n}, (V', E') = \ident{bind-prog}_{\emptyset, \Delta'}(p_2)$.
+
+Now, assume $\bndclr{v}, G = \ident{bind-expr}_{\Gamma, \Delta}(e)$ and
+$\bndclr{v'}, G' = \ident{bind-expr}_{\Gamma', \Delta'}(e)$ are the recursive calls to bind
+$e$ during the first and the second binding, respectively. Then, the graph nodes assigned to the
+sub-expression $e$ are the same, i.e.~$\bndclr{v} = \bndclr{v'}$.
+\end{lem}
+\begin{proof}
+First, assuming that $\forall x\in FV(e). \Gamma(x) = \Gamma'(x)$, we show by induction over the binding process of $e$
+for the first program that the result is the same. In cases (1) and (3), the updated $\Delta'$
+contains the required key and so the second binding proceeds using the same case. In cases
+(2) and (4), the second binding reuses the node created by the first binding using case (1) and
+(3), respectively. Cases (5) and (6) are the same.
+
+Second, when binding let bindings in $c_1; \ldots; c_k$, the initial $\Gamma = \emptyset$ during
+both bindings. Nodes added to $\Gamma$ and $\Gamma'$ for commands $c_j$ such that $j\in I$ are
+the same and nodes added for remaining commands do not add any new nodes referenced from $e$ and
+so $\bndclr{v} = \bndclr{v'}$ using the above.
+\end{proof}
+
+\begin{theorem}[Preview reuse]
+\label{thm:preview-reuse}
+Given the sequence of expressions as specified in Figure~\ref{fig:operations}, if the expressions
+are bound in sequence and graph node cache updated as specified in Figure~\ref{fig:loop}, then
+the graph nodes assigned to the specified sub-expressions are the same.
+\end{theorem}
+\begin{proof}
+Cases \rname{edit-let} and \rname{edit-mem} are direct consequences of Lemma~\ref{thm:sub-expr};
+for \rname{let-intro-var}, the node assigned to $x$ is the node assigned to $e$ which is the
+same as before the edit from Lemma~\ret{thm:sub-expr}.
+Cases \rname{let-intro-ins} and \rname{let-intro-del} are similar to \rname{let-intro-var}, but
+also require using induction over the binding of $K_c[e]$. Finally, cases \rname{let-elim-ins}
+and \rname{let-elim-del} are similar and also use Lemma~\ref{thm:sub-expr}
+together with induction over the binding of $K_c[x]$.
+\end{proof}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Empirical evaluation of efficiency}
+\label{sec:evaluation-empirical}
+
+The key performance claim about our method of providing instant feedback is that it is more
+efficient than recomputing values for the whole program (or the current command) after every
+keystroke. In the previous section, we formally proved that this is true and gave examples of
+code edit operations that do not cause recomputation. In this section, we further support this
+claim with an empirical evaluation. The purpose of this section is not to precisely evaluate
+overheads of our implementation, but to compare how much recomputation different evaluation
+strategies perform.
+
+For the purpose of the evaluation, we use a simple image manipulation library that provides
+operations for loading, greyscaling, blurring and combining images. We compare delays in
+updating the preview for three different evaluation strategies, while performing the same
+sequence of code edit operations. Using image processing as an example gives us a way to
+visualize the reuse of previously computed values. As in a typical data exploration scenario,
+the individual operations are relatively expensive compared to the overheads of building the
+dependency graph.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[b]
+\raggedright
+
+\vspace{0.5em}\hspace{1.5em}{\small\sffamily (1) Enter the following code and then change parameter of blur from 4 to 8:}
+\begin{equation*}
+\begin{array}{l}
+\ident{image}.\ident{load}(\str{shadow.png}).\ident{greyScale}().\ident{blur}(\num{4})
+\end{array}
+\end{equation*}
+
+\vspace{0.5em}\hspace{1.5em}{\small\sffamily (2) Assign the result to a variable and start writing code for further operations:}
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}\,\ident{shadow}=\ident{image}.\ident{load}(\str{shadow.png}).\ident{greyScale}().\ident{blur}(\num{8})\\
+\ident{shadow}.\ident{combine}
+\end{array}
+\end{equation*}
+
+\vspace{0.5em}\hspace{1.5em}{\small\sffamily (3) Finish code to combine two images and then change parameter of combine from 20 to 80:}
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}\,\ident{shadow}=\ident{image}.\ident{load}(\str{shadow.png}).\ident{greyScale}().\ident{blur}(\num{8})\\
+\ident{shadow}.\ident{combine}(\ident{image}.\ident{load}(\str{pope.png}), \num{20})
+\end{array}
+\end{equation*}
+
+\vspace{0.5em}\hspace{1.5em}{\small\sffamily (4) Extract the parameter of combine to a let bound variable:}
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}\,\ident{ratio} = \num{80} \\
+\kvd{let}\,\ident{shadow}=\ident{image}.\ident{load}(\str{shadow.png}).\ident{greyScale}().\ident{blur}(\num{8})\\
+\ident{shadow}.\ident{combine}(\ident{image}.\ident{load}(\str{pope.png}), \ident{ratio})
+\end{array}
+\end{equation*}
+
+\caption{Code edit operations that are used in the experimental evaluation}
+\label{fig:image-edits}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\vspace{-0.5em}
+\includegraphics[scale=0.425]{figures/drawing.png}
+\caption{Time required to recompute the results of a sample program after individual tokens of
+  the program are added or modified for three different evaluations strategies.}
+\label{fig:drawing}
+\vspace{-0.5em}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsubsection{Code edit operations and methodology}
+
+To avoid clutter when visualizing the performance, we update the preview after complete tokens are
+added (such as identifier, number, dot or left parenthesis) rather than after individual keystrokes.
+Figure~\ref{fig:image-edits} shows the sequence of code edit operations that we use to measure
+the delays in updating a live preview. We first enter an expression to load, greyscale and blur
+an image (1) then introduce let binding (2) and add more operations (3). Finally, we extract
+one of the parameters into a variable (4). Most of the operations are simply adding code, but
+there are two cases where we modify existing code and change value of a parameter for \ident{blur}
+and \ident{combine} immediately after (1) and (3), respectively.
+
+We implement the algorithm described in Sections~\ref{sec:formal} and \ref{sec:previews} in a
+simple web-based environment that allows the user to modify code and explicitly trigger
+recomputation. It then measures time needed to recompute values for the whole program and
+displays the resulting image. If the parsing fails, e.g.~because of an unclosed parenthesis,
+previews are not updated and we record only the time taken by parsing attempt. We compare the
+delays of three different evaluation strategies:
+
+\begin{itemize}
+\item \emph{Call-by-value.} Following the semantics in Section~\ref{sec:calculus-semantics},
+  all sub-expressions are fully evaluated before an expression is evaluated. In this case, work may
+  be done even if the evaluation results in an error. For example, we parse
+  $\ident{image}.\ident{load}(\str{shadow.png}).\ident{blur}$ as member access with no arguments.
+  The evaluation loads the image, but then fails because blur requires one argument.
+
+\item \emph{Lazy.} To address the wastefulness of call-by-value strategy discussed above, we
+  simulate lazy evaluation by implementing a version of the image processing library where
+  operations build a delayed computation and only evaluate it when rendering an image. Using
+  this strategy, failing computations do not perform unnecessary work.
+
+\item \emph{Live.} Finally, the live strategy uses the algorithm described in
+  Sections~\ref{sec:formal} and \ref{sec:previews}. The cache is empty at the beginning of the
+  experiment and we update it after each token is added. This is the only strategy that does
+  not start afresh after reparsing code.
+\end{itemize}
+
+\subsubsection{Efficiency evaluation}
+The experimental environment is implemented in F\# and compiled to JavaScript using the Fable
+compiler. We run the experiments in Firefox (version 64.0.2, 32bit) on Windows 10
+(build 1809, 64bit) on a computer with Intel Core i7-7660U CPU and 16Gb RAM.
+
+Figure~\ref{fig:drawing} shows times needed to recompute previews after individual tokens
+of the program are added, deleted or modified, according to the script in
+Figure~\ref{fig:image-edits}, resulting in 38 measurements for each of the 3
+evaluation strategies. We mark a number of notable points in the chart:
+
+\begin{enumerate}
+\item Loading image for the first time incurs small extra overhead in the live strategy.
+\item Greyscaling using the live strategy does not need to re-load the image.
+\item Accessing the blur member without arguments causes delay for call-by-value strategy.
+\item When varying the parameter of blur, the live strategy reuses previously greyscaled image.
+\item Introducing let binding does not cause recomputation when using live strategy.
+\item As in (c), accessing combine member without valid argument only affects call-by-value.
+\item Varying parameter of combine in live strategy does not recompute blur and is much faster.
+\item Introducing let binding, again, causes full recomputation for lazy can call-by-value.
+\end{enumerate}
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}
+\vspace{-0.5em}
+\noindent
+\includegraphics[scale=0.48]{figures/hist-all.png}
+\includegraphics[scale=0.48]{figures/hist-slow.png}
+\vspace{-0.25em}
+\caption{Histogram showing the distribution of delays incurred when updating previews.
+  We show a histogram computed from all delays (left) and only from delays larger
+  than 15ms (right).}
+\label{fig:drawing-hist}
+\vspace{-0.75em}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\noindent
+A summarized view of the delays is provided in Figure~\ref{fig:drawing-hist}, which shows
+histograms illustrating the distribution of delays for each of the three evaluation methods.
+A large proportion of delays is very small (less than 15ms) because the parser used in our
+experimental environment often fails (e.g.~for unclosed parentheses). The histogram
+on the right summarizes only delays for edit operations where the delay for at least one of
+the strategies was over 15ms. A more robust parser would be able to recover and the delays
+for call-by-value and lazy strategies would be even more significant.
+
+\subsubsection{Summary of results}
+The purpose of our experiment is not to exactly assess the overhead of our implementation.
+Our goal is to illustrate how often can previously evaluated results be reused and the impact
+this has when writing code. The experiment presented in this section is small-scale, but it is
+sufficient for this purpose. When recomputing results after every edit
+using the \emph{call-by-value} strategy, the time needed to update results grows
+continually. The \emph{lazy} strategy removes the overhead for programs that fail, but
+keeps the same trend. Our \emph{live} strategy reuses values computed
+previously. Consequently, expensive operations (such as (d) and (g) in Figure~\ref{fig:drawing})
+are significantly faster, because they do not need to recompute operations done previously when
+writing the code. As shown in Figure~\ref{fig:drawing-hist}, there are almost no very expensive
+operations (taking over 1 second) in the \emph{live} strategy in contrast to several in the
+other two strategies.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\noindent
+\hspace{-1em}
+\includegraphics[scale=0.245]{figures/gallery1.png}
+\includegraphics[scale=0.245]{figures/gallery2.png}
+\caption{\small Data analysis counting the number of libraries per county in the UK. We load
+  and aggregate data (left) and then create a chart with a label (right).
+  The example is adapted from \url{http://gallery.thegamma.net/73/}.}
+\label{fig:gallery}
+\vspace{-1em}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Transparent tools for data journalism}
+\label{sec:evaluation-case}
+
+In Section~\ref{sec:background-jupyter}, we motivated our work by considering how journalists
+explore open data. In addition to the theoretical and experimental work presented in this paper,
+we also implemented an online data exploration environment, equipped with live editor for The
+Gamma language that provides instant feedback during coding. The environment uses the
+principles presented in this paper to build a more comprehensive system that allows users,
+such as journalists, to analyse, summarize and visualize open data. In this section, we briefly
+report on our experience with the system.
+
+The environment is available at \url{http://gallery.thegamma.net}. Two screenshots shown
+in Figure~\ref{fig:gallery} illustrate a number of interesting features:
+
+\begin{itemize}
+\item The left screenshot uses a type provider for data aggregation \cite{gamma}
+  (Section~\ref{sec:background-gamma}). Our system views the type provider as an external library
+  (with objects, members and reduction relation). Type providers rely on type information to
+  provide editor auto-complete, which we support by implementing type checking over a
+  dependency graph as discussed in Section~\ref{sec:extra-types}.
+
+\item When displaying live preview for code written using the data aggregation type provider (left),
+  our environment also generates tabes that show individual steps of the data transformation.
+  A tab is selected based on the cursor position and shows preview for the current sub-expression
+  (e.g.~raw data before grouping in the screenshot).
+
+\item Another external library provides support for charting (right). The environment does not
+  handle charts in a special way and displays the result of evaluating the whole command. The
+  screenshot shows a case where the user modifies parameters of the chart. Thanks to our live evaluation
+  strategy, this is done efficiently without reevaluating the data transformation.
+\end{itemize}
+
+\noindent
+A user study to evaluate the usability of the system from a human-computer interaction perspective
+is left for future work. As an anecdotal evidence of the usability, the code in
+Figure~\ref{fig:gallery} was developed by an attendee of a tutorial at Mozilla Festival 2017
+who had no prior programming experience.
+
+% ==================================================================================================
+
+\section{Related and future work}
+\label{sec:future}
+
+We aim to make creating transparent data analyses, such as those done by journalists, easier and we
+present a tool that provides instant feedback when writing code in a text editor. Many our
+design decisions have alternatives that have been explored in a rich body of literature.
+
+\subsubsection{Notebooks and data science tools}
+Basic feedback mechanism, the REPL (read-eval-print-loop), originated in Lisp
+\cite{lisp} and is now common in many languages \cite{drscheme}. Notebook systems
+such as Jupyter and R markdown \cite{rmarkdown,jupyter,ipython} structure code in cells,
+but still rely on explicitly executing individual cells using a stateful REPL environment.
+Integrating provenance tracking with notebooks \cite{dataflow,noworkflow,wrattler} addresses
+the reproducibility issues of this model, but keeps cells as units of work. Ideas such as
+dependency tracking and efficient recomputation exist in many visual data exploration tools
+\cite{control,tableau,vizdom} and scientific workflow systems \cite{taverna,kepler}. Some of the
+implementation techniques are related to our work, but we focus on text-based scripts.
+Tempe \cite{tempe} focuses on streaming data, but provides a text-based scripting environment
+with automatic code update and \cite{ripple} empirically evaluate its usability in contrast to
+REPLs. The challenges faced by Tempe are more akin to dynamic software update \cite{dsu}.
+
+\subsubsection{Live coding and editing}
+Live coding based on textual programs has been popularised by \cite{principle,learnable} and
+is actively developed in domains such as live coded music \cite{beyond,sonic}. \cite{liveroad}
+reviews the use of live coding in a Smalltalk derived environment. Lighttable by \cite{lighttable}
+and Chrome DevTools provide limited live previews akin to those presented in this paper,
+but without well specified recomputation model. Finally, work on keeping state during
+code edits \cite{alive,livingit} would be relevant for supporting streaming data. In the future,
+we also plan to explore how direct manipulation \cite{direct} could be integrated with our preview
+mechanism. \cite{liveroad}
+
+Using structured editors \cite{structure-based} rather than text editors is a promising alternative
+direction as editors can recompute program based on understanding of edit operations
+\cite{interactive,livenut,lamdu}. This can be elegantly implemented using bi-directional
+lambda calculus \cite{hazelnut}, but it also makes us consider more human-centric abstractions
+\cite{subtext,directprog} along the lines of the one outlined in Section~\ref{sec:extra-abstraction}.
+
+\subsubsection{Incremental computation and dependency analysis}
+Work on self-adjusting and incremental computation \cite{selfadjusting,adapton} handles
+recomputation when the program stays the same, but input changes. It relies on dependency tracking
+similar to ours. We rely on static code analysis whereas most incremental systems,
+e.g.~\cite{incremental,adaptivefp,deltaml} evaluate the program and use programmer-supplied
+information to build a dependency graph. \cite{adapton} implement a small adaptive interpreter
+that treats code as changing input data, suggesting an interesting implementation technique for
+live coding systems. Our use of dependency graphs \cite{dependencies} is static and first-order
+and can be seen as a form of program slicing \cite{slicing}, although our binding process is
+more directly inspired by Roslyn \cite{roslyn}, which uses it for efficient background type-checking.
+
+\subsubsection{Semantics and partial evaluation}
+The evaluation of previews is a form of partial evaluation \cite{partial}, done in a
+way that allows caching. This can be done implicitly or explicitly in the form of
+multi-stage programming \cite{metaml}. Semantically, the evaluation of previews can be seen as a
+modality \cite{modal} and delayed previews are linked to contextual modal type theory \cite{cmtt},
+formally modelled using comonads \cite{cmtt-denotation}. This suggests a direction for rigorous
+analysis of the presented system.
+
+% ==================================================================================================
+%
+% \section{TODO}
+%
+% TBD: Define live preview somewhere
+%
+% TBD: In section about dependency graphs, include a subsection discussing how this relates
+% to incremental computation (just like "Semantics of delayed previews" links that to CMTT etc.)
+%
+%
+% \cite{selfadjusting}
+% \cite{incremental}
+%
+% Incremental computation
+%
+%   Self-adjusting computation
+%   http://www.cs.cmu.edu/~rwh/theses/acar.pdf
+%
+%   Incremental Computation with Names
+%   https://www.cs.tufts.edu/~jfoster/papers/oopsla15.pdf
+%   -> dsl interpreter
+%
+%   (and other adapton stuff http://adapton.org/)
+%
+% Type checking
+%
+%   co-contextual
+%   http://www.informatik.uni-marburg.de/~seba/publications/cocontextual-type-checking.pdf
+%   http://drops.dagstuhl.de/opus/volltexte/2017/7262/pdf/LIPIcs-ECOOP-2017-18.pdf
+%
+% https://dl.acm.org/citation.cfm?id=664109
+%
+\section{Summary}
+One of the reasons that make spreadsheets easier to use than programming tools is that they
+provide instant feedback. We aim to make programming tools as instant as spreadsheets.
+We formalized a number of key aspects of simple data analyses such as those don by journalists
+and then used our observations to build both theory and simple practical data analytics tools.
+
+Our \emph{data exploration calculus} is a simple formally tractable language
+for data exploration. The calculus captures key observations about simple data analyses. They
+rely on logic from external libraries, implement few abstractions and are written as lists of
+commands. We later revised the calculus to add type checking and simple abstraction mechanism.
+
+Our main technical contribution is a \emph{live preview} mechanism that efficiently evaluates
+code during editing and instantly provides a preview of the result. We allow users to edit code
+in unconstrained way in an text editor, which makes this particularly challenging. The key trick
+is to separate the process into fast \emph{binding phase}, which constructs a dependency graph
+and slower \emph{evaluation phase} that can cache results. This makes it possible to quickly parse
+updated code, reconstruct dependency graph and compute preview using previous, partially
+evaluated, results.
+
+We evaluated our approach in three ways. First, we proved that our mechanism is correct and
+that it reuses evaluated values for many common code edit operations. Second, we conducted an
+experimental study that illustrates how often are previously evaluated results reused during
+typical programming scenario. Thirdly, we used our research as a basis for online data exploration
+environment, which shows the practical usability of our work.
+
+% ==================================================================================================
+
+\printbibliography
+
+% ==================================================================================================
+
+\newpage
+\appendix
+
+\section{Appendix}
+
+\subsection{Normalization}
+\label{sec:app-normalization}
+
+\begin{theorem}[Normalization]
+For all $p$, there exists $n$ and $o_1, \ldots, o_n$ such that $p\rightsquigarrow^{*} o_1;\ldots;o_n$.
+\end{theorem}
+\begin{proof}
+We define $\ident{size}$ of a program in data exploration calculus as follows:
+\begin{equation}
+\begin{array}{rcl}
+\ident{size}(c_1; \ldots; c_n) &=& 1 + \Sigma^{n}_{i=1} \ident{size}(c_i)\\
+\ident{size}(\kvd{let}~x=t) &=& 1 + \ident{size}(t)\\
+\ident{size}(e_0.m(e_1, \ldots, e_n)) &=& 1 + \Sigma^{n}_{i=0} \ident{size}(e_i)\\
+\ident{size}(\lambda x\rightarrow e) &=& 1 + \ident{size}(e)\\
+\ident{size}(o) = \ident{size}(x) &=& 1\\
+\end{array}
+\end{equation}
+
+\noindent
+The property holds because, first, both \rname{let} and \rname{external} decrease the \ident{size}
+of the program and, second, a program is either fully evaluated, i.e.~$o_1;\ldots;o_n$ for some $n$
+or, it can be reduced using one of the reduction rules.
+\end{proof}
+
+\subsection{Term preview correctness}
+\label{sec:app-correctness}
+
+\begin{theorem}[Term preview correctness]
+Given a term $t$ that has no free variables, together with a lookup table $\Delta$ obtained
+from any sequence of programs using $\ident{bind-prog}$ (Figure~\ref{fig:binding-rules-prog}) and
+$\ident{update}$ (Figure~\ref{fig:loop}), then
+let $\bndclr{v}, (V, E) = \ident{bind-expr}_{\emptyset, \Delta}(t)$. If $\bndclr{v}\Downarrow p$
+over a graph $(V, E)$ then $p = o$ for some value $o$ and $t \rightsquigarrow^{*} o$.
+\end{theorem}
+
+\begin{proof}
+First note that, when combining recursively constructed sub-graphs, the \ident{bind-expr} function
+adds new nodes and edges leading from those new nodes. Therefore, an evaluation using~$\Downarrow$
+over a sub-graph will also be valid over the new graph -- the newly added nodes and edges do not introduce
+non-determinism to the rules given in Figure~\ref{fig:eval}.
+
+We prove a more general property showing that for any $e$, its binding
+$\bndclr{v}, (V, E) = \ident{bind-expr}_{\emptyset, \Delta}(e)$ and any evaluation context $C$
+such that $C[e]\rightsquigarrow o$ for some $o$, one of the following holds:
+%
+\begin{enumerate}
+\item[a.] If $FV(e)\,=\,\emptyset$ then $\bndclr{v} \Downarrow p$ for some $p$ and $C[p] \rightsquigarrow o$
+\item[b.] If $FV(e)\neq\emptyset$ then $\bndclr{v} \Downarrow \llbracket e_p \rrbracket_{FV(e)}$ for some $e_p$ and $C[e_p] \rightsquigarrow o$
+\end{enumerate}
+%
+In the first case, $p$ is a value, but it is not always the case that $e \rightsquigarrow^{*} p$,
+because $p$ may be lambda function and preview evaluation may reduce sub-expression in the body of
+the function. Using a context $C$ in which the value reduces to an object avoids this problem.
+
+The proof of the theorem follows from the more general property. Using a context $C[-]=-$, the
+term $t$ reduces $t \rightsquigarrow^{*}t' \rightsquigarrow_\epsilon o$ for some $o$ and the
+preview $p$ is a value $o$ because $C[p] = p = o$.
+The proof is by induction over the binding process, which follows the structure of the expression:
+
+\vspace{0.75em}\noindent(1) $\ident{bind-expr}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n))$ --
+  Here $e = e_0.m(e_1, \ldots, e_n)$, $\bndclr{v_i}$ are graph nodes obtained by induction for
+  expressions $e_i$ and $\{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v}, \bndclr{v_n}, \blbl{arg}(n))\} \subseteq E$.
+  From lookup inversion Lemma~\ref{thm:lemma-lookup}, $\bndclr{v} = \bnd{mem}(m, s)$ for some $s$.
+
+  If $FV(e)=\emptyset$, then $\bndclr{v_i} \Downarrow p_i$ for $i\in 0\ldots n$ and
+  $\bndclr{v}\Downarrow p$ using \rname{mem-val} such that $p_0.m(p_1, \ldots, p_n) \rightsquigarrow p$.
+  From induction hypothesis and \emph{compositionality} of external libraries (Definition~\ref{def:external}),
+  it holds that for any $C$ such that $C[e_0.m(e_1, \ldots, e_n)] \rightsquigarrow o$ for some $o$
+  then also $C[p_0.m(p_1, \ldots, p_n)] \rightsquigarrow C[p] \rightsquigarrow o$.
+
+  If $FV(e)\neq\emptyset$, then $\bndclr{v_i} \Downarrow_\ident{lift} \llbracket e'_i \rrbracket$ for $i\in 0\ldots n$ and
+  $\bndclr{v}\Downarrow \llbracket e'_0.m(e'_1, \ldots, e'_n) \rrbracket_{FV(e)}$ using \rname{mem-expr}.
+  From induction hypothesis and \emph{compositionality} of external libraries (Definition~\ref{def:external}),
+  it holds that for any $C$ such that $C[e_0.m(e_1, \ldots, e_n)] \rightsquigarrow o$ for some $o$
+  then also $C[e'_0.m(e'_1, \ldots, e'_n)] \rightsquigarrow o$.
+
+\vspace{0.75em}\noindent(2) $\ident{bind-expr}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n))$ --
+  This case is similar to (1), except that the fact that $\bndclr{v} = \bnd{mem}(m, s)$
+  holds by construction, rather than using Lemma~\ref{thm:lemma-lookup}.
+
+\vspace{0.75em}\noindent(3) $\ident{bind-expr}_{\Gamma, \Delta}(\lambda x\rightarrow e_b)$ --
+  Here $e = \lambda x\rightarrow e_b$, $\bndclr{v_b}$ is the graph node obtained by induction
+  for the expression $e_b$ and $(\bndclr{v}, \bndclr{v_b}, \blbl{body}) \in E$.
+  From lookup inversion Lemma~\ref{thm:lemma-lookup}, $\bndclr{v} = \bnd{fun}(x, s)$ for some $s$.
+  The evaluation can use one of three rules:
+
+  \begin{enumerate}
+  \item[i.]
+  If $FV(e)=\emptyset$ then $\bndclr{v_b}\Downarrow p_b$ for some $p_b$ and $\bndclr{v}\Downarrow \lambda x\rightarrow p_b$
+  using \rname{fun-val}. Let $e'_b=p_b$.
+
+  \item[ii.]
+  If $FV(e_b)=\{x\}$ then $\bndclr{v_b}\Downarrow \llbracket e'_b \rrbracket_x$ for some $e'_b$ and
+  $\bndclr{v}\Downarrow \lambda x\rightarrow e'_b$ using \rname{fun-bind}.
+
+  \item[iii.]
+  Otherwise, $\bndclr{v_b}\Downarrow \llbracket e'_b \rrbracket_{x,\Gamma}$ for some $e'_b$ and
+  $\bndclr{v}\Downarrow \llbracket\lambda x\rightarrow e'_b\rrbracket_\Gamma$ using \rname{fun-expr}.
+  \end{enumerate}
+
+\noindent
+For i.) and ii.) we show that a.) is the case; for iii.) we show that b.) is the case; that is
+for any $C$, if $C[\lambda x\rightarrow e_b] \rightsquigarrow o$ then also
+$C[\lambda x\rightarrow e'_b] \rightsquigarrow o$. For a given $C$, let $C'[-] = C[\lambda x\rightarrow -]$
+and use the induction hypothesis, i.e.~if $C'[e_b] \rightsquigarrow o$ for some $o$ then also $C'[e'_b] \rightsquigarrow o$.
+
+\vspace{0.75em}\noindent(4) $\ident{bind-expr}_{\Gamma, \Delta}(\lambda x\rightarrow e)$  --
+  This case is similar to (3), except that the fact that $\bndclr{v} = \bnd{fun}(x, s)$
+  holds by construction, rather than using Lemma~\ref{thm:lemma-lookup}.
+
+\vspace{0.75em}\noindent(5) $\ident{bind-expr}_{\Gamma, \Delta}(o)$ -- In this case $e=o$ and $\bndclr{v} = \bnd{val}(o)$
+  and $\bnd{val}(o)\Downarrow o$ using \rname{val} and so the case a.) trivially holds.
+
+\vspace{0.75em}\noindent(6) $\ident{bind-expr}_{\Gamma, \Delta}(x)$ -- The initial $\Gamma$ is empty,
+  so $x$ must have been added to $\Gamma$ by case (3) or (4). Hence,
+  $\bndclr{v}=\bnd{var}(x)$, $\bndclr{v}\Downarrow \llbracket x \rrbracket_{x}$ using \rname{var}
+  and so $e_p = e = x$ and the case b.) trivially holds.
+
+\end{proof}
+
+% ==================================================================================================
+
+\section{Theories of delayed previews}
+\label{sec:app-theories}
+
 The operational semantics presented in this paper serves two purposes. It gives a simple guide
@@
 
-\section{Evaluating live coding environment}
-\label{sec:evaluation}
-
-Evaluating previews using a dependency graph makes it possible to cache partial results.
-The mechanism satisfies two properties. First, if we evaluate a preview using dependency
-graph with caching, it is the same as the value we would obtain by evaluating the expression
-directly. Second, the evaluation of previews using dependency graphs reuses -- in some cases --
-previously evaluated partial results. In other words, we show that the mechanism is correct
-and implements an effective optimization. We discuss correctness first and then evaluate
-effectiveness, both theoretically (Section~\ref{sec:evaluation-reuse}) and empirically
-(Section~\ref{sec:evaluation-empirical}). Finally, we describe a case study where we used the
-methods outlined in the paper to develop a larger online service for data exploration
-(Section~\ref{sec:evaluation-case}).
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Correctness of previews}
-\label{sec:evaluation-correctness}
-
-To show that the previews are correct, we prove two properties. Correctness
-(Theorem~\ref{thm:correcntess}) guarantees that, no matter how a dependency graph is constructed,
-when we use it to evaluate previews for a program, the previews are the same as the values we
-would obtain by evaluating the program commands directly. Determinacy (Theorem~\ref{thm:determinacy})
-guarantees that if we cache a preview for a graph node and update the graph, the preview we would
-evaluate using the updated graph would be the same as the cached preview.
-
-To simplify the proofs, we consider programs without let bindings. This is possible, because
-eliminating let bindings does not change the result of evaluation, as shown earlier in
-Lemma~\ref{thm:let-lang-elimination}, and it also does not change the constructed dependency graph
-as shown below in Lemma~\ref{thm:let-grp-elimination}.
+\section{Proofs for section 3}
+\label{sec:let-lang-elimination}
+
+\begin{lem}[Let elimination for a program]
+\label{thm:let-lang-elimination-proof}
+Given any program $p$ such that $p \rightsquigarrow^{*} o_1;\ldots;o_n$ for some $n$ and $o_1, \ldots, o_n$
+then if $p \rightsquigarrow_\ident{let} p'$ for some $p'$ then also $p' \rightsquigarrow^{*} o_1;\ldots;o_n$.
+\end{lem}
+\begin{proof}
+The elimination of let binding transforms a program $c_1;\; \ldots;\; c_{k-1};\; \kvd{let}~x=t;\; c_{k+1};\; \ldots; c_n$
+to a program $c_1;\; \ldots;\; c_{k-1};\; t; c_{k+1}[x\leftarrow t];\; \ldots; c_n[x\leftarrow t]$.
+The reduction steps for the new program can be constructed using the steps of $p \rightsquigarrow^{*} o_1;\ldots;o_n$.
+The new command $t$ reduces to an object $o$ using the same steps as the original term $t$
+in $\kvd{let}~x=t$ but with context $C_c = -$ rather than $C_c = \kvd{let}~x=-$; the terms $t$
+introduced by substitution also reduce using the same steps as before, but using
+contexts in which the variable $x$ originally appeared.
+\end{proof}
+
+\section{Correctness of previews}
+\label{sec:app-correctness}
+
+\subsection{Let elimination for a dependency graph}
+\label{sec:app-let-grp-elimination}
 
 \begin{lem}[Let elimintion for a dependency graph]
-\label{thm:let-grp-elimination}
+\label{thm:app-let-grp-elimination}
 Given programs $p_1, p_2$ such that $p_1 \rightsquigarrow_\ident{let} p_2$ and a lookup table
@@
 
-The Lemma~\ref{thm:let-grp-elimination} provides a way of removing let bindings from a program,
-such that the resulting dependency graph remains the same. Here, we bind the original program
-first, which adds the node for $e$ to $\Delta$. In our implementation, this is not needed
-because $\Delta$ is updated while the graph is being constructed using $\ident{bind-expr}$.
-To keep the formalisation simpler, we separate the process of building the dependency graph
-and updating $\Delta$ and thus Lemma~\ref{thm:let-grp-elimination} requires an extra binding step.
-
-Now, we can show that, given a let-free expression, the preview obtained using a correctly
-constructed dependency graph is the same as the one we would obtain by directly evaluating the
-expression. This requires a simple auxiliary lemma.
-
-\begin{lem}[Lookup inversion]
-\label{thm:lemma-lookup}
-Given $\Delta$ obtained using \ident{update} as defined in Figure~\ref{fig:loop} then:
-\begin{itemize}
-\raggedright
-\item[--] If $\bndclr{v}=\Delta(\bknd{fun}(x),[(\bndclr{v_0}, \blblclr{l_0})])$
-then $\bndclr{v}=\bnd{fun}(x, s)$ for some $s$.
-\item[--] If $\bndclr{v}=\Delta(\bknd{mem}(m),[(\bndclr{v_0}, \blblclr{l_0}), \ldots, (\bndclr{v_n}, \blblclr{l_n})])$
-then $\bndclr{v}=\bnd{mem}(m, s)$ for some $s$.
-\end{itemize}
-\end{lem}
-\begin{proof}
-By construction of $\Delta$ in Figure~\ref{fig:loop}.
-\end{proof}
+\subsection{Term preview correctness}
+\label{sec:app-let-free-correct}
 
 \begin{theorem}[Term preview correctness]
-\label{thm:let-free-correct}
+\label{thm:app-let-free-correct}
 Given a term $t$ that has no free variables, together with a lookup table $\Delta$ obtained
@@
 
-\noindent
-The correctness theorem combines the previous two results.
-
-\begin{theorem}[Program preview correctness]
-\label{thm:correcntess}
-Consider a program $p=c_1;\ldots;c_n$ that has no free variables, together with a lookup table
-$\Delta_0$ obtained from any sequence of programs using $\ident{bind-prog}$
-(Figure~\ref{fig:binding-rules-prog}) and $\ident{update}$ (Figure~\ref{fig:loop}). Assume a
-let-free program $p'=t_1;\ldots;t_n$ such that $p\rightsquigarrow^{*}_\ident{let}p'$.
-
-\vspace{0.25em}
-\noindent
-Let $\bndclr{v_1};\ldots;\bndclr{v_n}, (V, E) = \ident{bind-prog}_{\emptyset, \Delta_0}(p)$
-and define updated lookup table $\Delta_1=\ident{update}_{V, E}(\Delta_0)$ and
-let $\bndclr{v'_1};\ldots;\bndclr{v'_n}, (V', E') = \ident{bind-prog}_{\emptyset, \Delta_1}(p)$.
-
-\vspace{0.25em}
-\noindent
-If $\bndclr{v'_i}\Downarrow p_i$ over a graph $(V', E')$ then $p_i=o_i$ for some value $o_i$ and
-$t_i \rightsquigarrow o_i$.
-\end{theorem}
-\begin{proof}
-Direct consequence of Lemma~\ref{thm:let-grp-elimination} and Theorem~\ref{thm:let-free-correct}.
-\end{proof}
-
-\noindent
-Our implementation updates $\Delta$ during the recursive binding process and so a stronger version
-of the property holds: previews calculated over a graph obtained directly for the original program
-$p$ are the same as the values of the fully evaluated program. Our formalisation omits this for
-simplicity.
-
-The second important property is determinacy, which makes it possible to cache the previews
-evaluated using~$\Downarrow$ using the corresponding graph node as a lookup key.
-
-\begin{theorem}[Program preview determinacy]
-\label{thm:determinacy}
-For some $\Delta$ and for any programs $p, p'$, assume that the first program is bound,
-i.e.~$\bndclr{v_1};\ldots;\bndclr{v_n}, (V, E) = \ident{bind-prog}_{\emptyset, \Delta}(p)$,
-the graph node cache is updated $\Delta' = \ident{update}_{V, E}(\Delta)$ and the second program is
-bound, i.e.~$\bndclr{v'_1};\ldots;\bndclr{v'_m}, (V', E') = \ident{bind-prog}_{\emptyset, \Delta'}(p')$.
-Now, for any $\bndclr{v}$, if $\bndclr{v} \Downarrow p$ over $(V, E)$ then also
- $\bndclr{v} \Downarrow p$ over $(V', E')$.
-\end{theorem}
-\begin{proof}
-By induction over $\Downarrow$ over $(V, E)$, we show that the same evaluation rules also
-apply over $(V', E')$. This is the case, because graph nodes added to $\Delta'$ by $\ident{update}_{V, E}$
-are added as new nodes in $\ident{bind-prog}_{\emptyset, \Delta'}$ and
-nodes and edges of $(V, E)$ used during the evaluation are unaffected.
-\end{proof}
-
-\noindent
-The mechanism used for caching previews (Section~\ref{sec:previews-cache}),
-keeps a preview preview $d$ in a lookup table indexed by nodes $\bndclr{v}$.
-Theorem~\ref{thm:determinacy} guarantees that this is a valid -- as we update dependency
-graph during code editing, previous nodes will continue representing the same sub-expressions.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Reuse of previews}
-\label{sec:evaluation-reuse}
-
-In the motivating example given in Section~\ref{sec:intro}, the data analyst first extracted a
-constant value into a let binding and then modified a parameter of the last method call in a call
-chain. We argued that a live coding environment should reuse partially evaluated previews for these
-two cases. We now show that this is, indeed, the case in our system.
-
-In this section, we identify a number of code edit operations where the previously evaluated
-values for a sub-expression can be reused. The list of operations is shown in Figure~\ref{fig:operations}.
-To express the operations we define an editing context $K$ which is defined
-similarly to evaluation context $C$ from Figure~\ref{fig:dec-calculus}, but captures sub-expressions
-appearing anywhere in the program.
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
-\raggedright
-\vspace{-0.5em}
-\hspace{1.5em}{\small\sffamily Edit contexts of expressions}
-\begin{equation*}
-\begin{array}{lcl}
-K_e[-] &=& K_e[-].m(e_1, \ldots, e_n) ~\lsep~ e.m(e_1, \ldots, e_{l-1}, K_e[-], e_{l+1},\ldots , e_n) ~\lsep~ -\\
-K_c[-] &=& \kvd{let}~x = K_e[-] ~\lsep~ K_e[-]\\
-\end{array}
-\end{equation*}
-
-\vspace{0.5em}
-\hspace{1.5em}{\small\sffamily Code edit operations preserving preview for a sub-expression}
-
-\hspace{4em}\begin{minipage}[c]{0.82\textwidth}
-  \raggedright\setlength{\parindent}{-1em}
-
-  \ername{let-intro-var}
-  $\overline{c_1}; \preview{e}; \overline{c_2}$ changes to
-  $\overline{c_1}; \kvd{let}\,x = e; \preview{x}; \overline{c_2}$ where $x$ is a fresh name.
-
-  \ername{let-intro-ins}
-  $\overline{c_1}; \overline{c_2}; \preview{K_c[e]}; \overline{c_3}$ is changed to
-  $\overline{c_1}; \kvd{let}\,x=e; \overline{c_2}; \preview{K_c[x]}; \overline{c_3}$ via
-  a semantically non-equivalent expression
-  $\overline{c_1}; \overline{c_2}; K_c[x]; \overline{c_3}$ where $x$ is free.
-
-  \ername{let-intro-del}
-  $\overline{c_1}; \overline{c_2}; \preview{K_c[e]}; \overline{c_3}$ is changed to
-  $\overline{c_1}; \kvd{let}\,x=e; \overline{c_2}; \preview{K_c[x]}; \overline{c_3}$ via
-  an expression $\overline{c_1}; \kvd{let}\,x=e; \overline{c_2}; K_c[e]; \overline{c_3}$
-  with unused variable $x$.
-
-  \ername{let-elim-del}
-  $\overline{c_1}; \kvd{let}\,x=e; \overline{c_2}; \preview{K_c[x]}; \overline{c_3}$ is changed to
-  $\overline{c_1}; \overline{c_2}; \preview{K_c[e]}; \overline{c_3}$ via
-  a semantically non-equivalent expression
-  $\overline{c_1}; \overline{c_2}; K_c[x]; \overline{c_3}$ where $x$ is free.
-
-  \ername{let-elim-ins}
-  $\overline{c_1}; \kvd{let}\,x=e; \overline{c_2}; \preview{K_c[x]}; \overline{c_3}$ is changed to
-  $\overline{c_1}; \overline{c_2}; \preview{K_c[e]}; \overline{c_3}$ via
-  an expression $\overline{c_1}; \kvd{let}\,x=e; \overline{c_2}; K_c[e]; \overline{c_3}$
-  with unused variable $x$.
-
-  \ername{edit-mem}
-  $\overline{c_1}; K_c[\preview{e_0}.m(\overline{e})]; \overline{c_2}$ is changed to
-  $\overline{c_1}; K_c[\preview{e_0}.m'(\overline{e'})]; \overline{c_2}$
-
-  \ername{edit-let}
-  $\overline{c_1}; \kvd{let}\,x=e_1; \overline{c_2}; K_c[\preview{e_2}]; \overline{c_3}$ is changed to
-  $\overline{c_1}; \kvd{let}\,x=e'_1; \overline{c_2}; K_c[\preview{e_2}]; \overline{c_3}$ \hspace{2em} when
-  $x\notin FV(e_2)$.
-\end{minipage}
-
-\caption{Code edit operations that preserve previously evaluated preview}
-\label{fig:operations}
-\vspace{-0.5em}
-\end{figure}
-
-% --------------------------------------------------------------------------------------------------
-
-We use the notation $\preview{e}$ to mark the part of the expressions before and after the edit
-that have the same value and will not need to be recomputed after the change; we write
-$\overline{c}$ and $\overline{e}$ for a list of commands and expressions, respectively.
-In some of the edit operations, we also specify an intermediate program that may be
-semantically different and only has a partial preview. This illustrates a typical way of
-working with code in a text editor using cut and paste. For example, in
-\rname{let-intro-ins}, we the analyst cuts a sub-expression $e$, replaces it with
-a variable $x$ and then adds a let binding for a variable $x$ and inserts the expression $e$
-from the clipboard. The \rname{let-intro-del} operation captures the same edit, but
-done in a different order.
-
-The of preview-preserving code edit operations includes several ways of introducing a let
-binding, several ways of eliminating a let binding and two edit operations where the data analyst
-modifies an unrelated part of the program. It is worth noting that the list is not exhaustive.
-Rather, it aims to illustrate some of the typical operations that the data analyst might perform
-when writing code.
-
-Theorem~\ref{thm:preview-reuse} proves that the operations given in Figure~\ref{fig:operations}
-preserve the preview for a marked sub-expression. It relies on the following lemma that
-characterizes one of the common kinds of edits more generally. Given two versions of a program
-that both contain the same sub-expression $e$, if the let bindings that define the values of
-variables used in $e$ do not change, then the graph node assigned to $e$ will be the same
-when binding the original and the updated program.
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{lem}[Binding sub-expressions]
-\label{thm:sub-expr}
-Assume we have programs $p_1 = c_1; \ldots; c_k; K_c[e]; c_{k+1};\ldots; c_n$ and
-$p_2 = c'_1; \ldots; c'_k; K'_c[e]; c'_{k+1};\ldots;c'_n$ and $I\subseteq \{1\ldots k\}$ such
-that $\forall i\!\in\!I.\,c_i\!=\!c'_i$ and for each $x \in \bigcup_{i\in I}FV(c_i) \cup FV(e)$
-there exists $j\in I$ such that $c_j = \kvd{let}\,x=e$ for some $e$.
-%
-Given any $\Delta$, assume that the the first program is bound,
-i.e.~$\bndclr{v_1};\ldots;\bndclr{v_n}, (V, E) = \ident{bind-prog}_{\emptyset, \Delta}(p_1)$,
-the cache is updated $\Delta' = \ident{update}_{V, E}(\Delta)$ and the second
-program is bound,
-i.e.~$\bndclr{v'_1};\ldots;\bndclr{v'_n}, (V', E') = \ident{bind-prog}_{\emptyset, \Delta'}(p_2)$.
-
-Now, assume $\bndclr{v}, G = \ident{bind-expr}_{\Gamma, \Delta}(e)$ and
-$\bndclr{v'}, G' = \ident{bind-expr}_{\Gamma', \Delta'}(e)$ are the recursive calls to bind
-$e$ during the first and the second binding, respectively. Then, the graph nodes assigned to the
-sub-expression $e$ are the same, i.e.~$\bndclr{v} = \bndclr{v'}$.
-\end{lem}
-\begin{proof}
-First, assuming that $\forall x\in FV(e). \Gamma(x) = \Gamma'(x)$, we show by induction over the binding process of $e$
-for the first program that the result is the same. In cases (1) and (3), the updated $\Delta'$
-contains the required key and so the second binding proceeds using the same case. In cases
-(2) and (4), the second binding reuses the node created by the first binding using case (1) and
-(3), respectively. Cases (5) and (6) are the same.
-
-Second, when binding let bindings in $c_1; \ldots; c_k$, the initial $\Gamma = \emptyset$ during
-both bindings. Nodes added to $\Gamma$ and $\Gamma'$ for commands $c_j$ such that $j\in I$ are
-the same and nodes added for remaining commands do not add any new nodes referenced from $e$ and
-so $\bndclr{v} = \bndclr{v'}$ using the above.
-\end{proof}
-
-\begin{theorem}[Preview reuse]
-\label{thm:preview-reuse}
-Given the sequence of expressions as specified in Figure~\ref{fig:operations}, if the expressions
-are bound in sequence and graph node cache updated as specified in Figure~\ref{fig:loop}, then
-the graph nodes assigned to the specified sub-expressions are the same.
-\end{theorem}
-\begin{proof}
-Cases \rname{edit-let} and \rname{edit-mem} are direct consequences of Lemma~\ref{thm:sub-expr};
-for \rname{let-intro-var}, the node assigned to $x$ is the node assigned to $e$ which is the
-same as before the edit from Lemma~\ret{thm:sub-expr}.
-Cases \rname{let-intro-ins} and \rname{let-intro-del} are similar to \rname{let-intro-var}, but
-also require using induction over the binding of $K_c[e]$. Finally, cases \rname{let-elim-ins}
-and \rname{let-elim-del} are similar and also use Lemma~\ref{thm:sub-expr}
-together with induction over the binding of $K_c[x]$.
-\end{proof}
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Empirical evaluation of efficiency}
-\label{sec:evaluation-empirical}
-
-The key performance claim about our method of providing instantaneous feedback is that it is more
-efficient than recomputing values for the whole program (or the current command) after every
-keystroke. In the previous section, we formally proved that this is true and gave examples of
-code edit operations that do not cause recomputation. In this section, we further support this
-claim with an empirical evaluation. The purpose of this section is not to precisely evaluate
-overheads of our implementation, but to compare how much recomputation different evaluation
-strategies perform.
-
-For the purpose of the evaluation, we use a simple image manipulation library that provides
-operations for loading, greyscaling, blurring and combining images. We compare delays in
-updating the preview for three different evaluation strategies, while performing the same
-sequence of code edit operations. Using image processing as an example gives us a way to
-visualize the reuse of previously computed values. As in a typical data exploration scenario,
-the individual operations are relatively expensive compared to the overheads of building the
-dependency graph.
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure}[b]
-\raggedright
-
-\vspace{0.5em}\hspace{1.5em}{\small\sffamily (1) Enter the following code and then change parameter of blur from 4 to 8:}
-\begin{equation*}
-\begin{array}{l}
-\ident{image}.\ident{load}(\str{shadow.png}).\ident{greyScale}().\ident{blur}(\num{4})
-\end{array}
-\end{equation*}
-
-\vspace{0.5em}\hspace{1.5em}{\small\sffamily (2) Assign the result to a variable and start writing code for further operations:}
-\begin{equation*}
-\begin{array}{l}
-\kvd{let}\,\ident{shadow}=\ident{image}.\ident{load}(\str{shadow.png}).\ident{greyScale}().\ident{blur}(\num{8})\\
-\ident{shadow}.\ident{combine}
-\end{array}
-\end{equation*}
-
-\vspace{0.5em}\hspace{1.5em}{\small\sffamily (3) Finish code to combine two images and then change parameter of combine from 20 to 80:}
-\begin{equation*}
-\begin{array}{l}
-\kvd{let}\,\ident{shadow}=\ident{image}.\ident{load}(\str{shadow.png}).\ident{greyScale}().\ident{blur}(\num{8})\\
-\ident{shadow}.\ident{combine}(\ident{image}.\ident{load}(\str{pope.png}), \num{20})
-\end{array}
-\end{equation*}
-
-\vspace{0.5em}\hspace{1.5em}{\small\sffamily (4) Extract the parameter of combine to a let bound variable:}
-\begin{equation*}
-\begin{array}{l}
-\kvd{let}\,\ident{ratio} = \num{80} \\
-\kvd{let}\,\ident{shadow}=\ident{image}.\ident{load}(\str{shadow.png}).\ident{greyScale}().\ident{blur}(\num{8})\\
-\ident{shadow}.\ident{combine}(\ident{image}.\ident{load}(\str{pope.png}), \ident{ratio})
-\end{array}
-\end{equation*}
-
-\caption{Code edit operations that are used in the experimental evaluation}
-\label{fig:image-edits}
-\end{figure}
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
-\vspace{-0.5em}
-\includegraphics[scale=0.425]{drawing.png}
-\caption{Time required to recompute the results of a sample program after individual tokens of
-  the program are added or modified for three different evaluations strategies.}
-\label{fig:drawing}
-\vspace{-0.5em}
-\end{figure}
-
-% --------------------------------------------------------------------------------------------------
-
-\subsubsection{Code edit operations and methodology}
-
-To avoid clutter when visualizing the performance, we update the preview after complete tokens are
-added (such as identifier, number, dot or left parenthesis) rather than after individual keystrokes.
-Figure~\ref{fig:image-edits} shows the sequence of code edit operations that we use to measure
-the delays in updating a live preview. We first enter an expression to load, greyscale and blur
-an image (1) then introduce let binding (2) and add more operations (3). Finally, we extract
-one of the parameters into a variable (4). Most of the operations are simply adding code, but
-there are two cases where we modify existing code and change value of a parameter for \ident{blur}
-and \ident{combine} immediately after (1) and (3), respectively.
-
-We implement the algorithm described in Sections~\ref{sec:formal} and \ref{sec:previews} in a
-simple web-based environment that allows the user to modify code and explicitly trigger
-recomputation. It then measures time needed to recompute values for the whole program and
-displays the resulting image. If the parsing fails, e.g.~because of an unclosed parenthesis,
-previews are not updated and we record only the time taken by parsing attempt. We compare the
-delays of three different evaluation strategies:
-
-\begin{itemize}[itemsep=3pt]
-\item \emph{Call-by-value.} Following the semantics in Section~\ref{sec:calculus-semantics},
-  all sub-expressions are fully evaluated before an expression is evaluated. In this case, work may
-  be done even if the evaluation results in an error. For example, we parse
-  $\ident{image}.\ident{load}(\str{shadow.png}).\ident{blur}$ as member access with no arguments.
-  The evaluation loads the image, but then fails because blur requires one argument.
-
-\item \emph{Lazy.} To address the wastefulness of call-by-value strategy discussed above, we
-  simulate lazy evaluation by implementing a version of the image processing library where
-  operations build a delayed computation and only evaluate it when rendering an image. Using
-  this strategy, failing computations do not perform unnecessary work.
-
-\item \emph{Live.} Finally, the live strategy uses the algorithm described in
-  Sections~\ref{sec:formal} and \ref{sec:previews}. The cache is empty at the beginning of the
-  experiment and we update it after each token is added. This is the only strategy that does
-  not start afresh after reparsing code.
-\end{itemize}
-
-\subsubsection{Efficiency evaluation}
-The experimental environment is implemented in F\# and compiled to JavaScript using the Fable
-compiler. We run the experiments in Firefox (version 64.0.2, 32bit) on Windows 10
-(build 1809, 64bit) on a computer with Intel Core i7-7660U CPU and 16Gb RAM.
-
-Figure~\ref{fig:drawing} shows times needed to recompute previews after individual tokens
-of the program are added, deleted or modified, according to the script in
-Figure~\ref{fig:image-edits}, resulting in 38 measurements for each of the 3
-evaluation strategies. We mark a number of notable points in the chart:
-
-\begin{enumerate}[(a)]
-\item Loading image for the first time incurs small extra overhead in the live strategy.
-\item Greyscaling using the live strategy does not need to re-load the image.
-\item Accessing the blur member without arguments causes delay for call-by-value strategy.
-\item When varying the parameter of blur, the live strategy reuses previously greyscaled image.
-\item Introducing let binding does not cause recomputation when using live strategy.
-\item As in (c), accessing combine member without valid argument only affects call-by-value.
-\item Varying parameter of combine in live strategy does not recompute blur and is much faster.
-\item Introducing let binding, again, causes full recomputation for lazy can call-by-value.
-\end{enumerate}
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure}
-\vspace{-0.5em}
-\noindent
-\includegraphics[scale=0.48]{hist-all.png}
-\includegraphics[scale=0.48]{hist-slow.png}
-\vspace{-0.25em}
-\caption{Histogram showing the distribution of delays incurred when updating previews.
-  We show a histogram computed from all delays (left) and only from delays larger
-  than 15ms (right).}
-\label{fig:drawing-hist}
-\vspace{-0.75em}
-\end{figure}
-
-% --------------------------------------------------------------------------------------------------
-
-\noindent
-A summarized view of the delays is provided in Figure~\ref{fig:drawing-hist}, which shows
-histograms illustrating the distribution of delays for each of the three evaluation methods.
-A large proportion of delays is very small (less than 15ms) because the parser used in our
-experimental environment often fails (e.g.~for unclosed parentheses). The histogram
-on the right summarizes only delays for edit operations where the delay for at least one of
-the strategies was over 15ms. A more robust parser would be able to recover and the delays
-for call-by-value and lazy strategies would be even more significant.
-
-\subsubsection{Summary of results}
-The purpose of our experiment is not to exactly assess the overhead of our implementation.
-Our goal is to illustrate how often can previously evaluated results be reused and the impact
-this has when writing code. The experiment presented in this section is small-scale, but it is
-sufficient for this purpose. When recomputing results after every edit
-using the \emph{call-by-value} strategy, the time needed to update results grows
-continually. The \emph{lazy} strategy removes the overhead for programs that fail, but
-keeps the same trend. Our \emph{live} strategy reuses values computed
-previously. Consequently, expensive operations (such as (d) and (g) in Figure~\ref{fig:drawing})
-are significantly faster, because they do not need to recompute operations done previously when
-writing the code. As shown in Figure~\ref{fig:drawing-hist}, there are almost no very expensive
-operations (taking over 1 second) in the \emph{live} strategy in contrast to several in the
-other two strategies.
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
-\noindent
-\hspace{-1em}
-\includegraphics[scale=0.245]{gallery1.png}
-\includegraphics[scale=0.245]{gallery2.png}
-\caption{\small Data analysis counting the number of libraries per county in the UK. We load
-  and aggregate data (left) and then create a chart with a label (right).
-  The example is adapted from \url{http://gallery.thegamma.net/73/}.}
-\label{fig:gallery}
-\vspace{-1em}
-\end{figure}
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Transparent tools for data journalism}
-\label{sec:evaluation-case}
-
-In Section~\ref{sec:background-jupyter}, we motivated our work by considering how journalists
-explore open data. In addition to the theoretical and experimental work presented in this paper,
-we also implemented an online data exploration environment, equipped with live editor for The
-Gamma language that provides instantaneous feedback during coding. The environment uses the
-principles presented in this paper to build a more comprehensive system that allows users,
-such as journalists, to analyse, summarize and visualize open data. In this section, we briefly
-report on our experience with the system.
-
-The environment is available at \url{http://gallery.thegamma.net}. Two screenshots shown
-in Figure~\ref{fig:gallery} illustrate a number of interesting features:
-
-\begin{itemize}[itemsep=3pt]
-\item The left screenshot uses a type provider for data aggregation \cite{gamma}
-  (Section~\ref{sec:background-gamma}). Our system views the type provider as an external library
-  (with objects, members and reduction relation). Type providers rely on type information to
-  provide editor auto-complete, which we support by implementing type checking over a
-  dependency graph as discussed in Section~\ref{sec:extra-types}.
-
-\item When displaying live preview for code written using the data aggregation type provider (left),
-  our environment also generates tabes that show individual steps of the data transformation.
-  A tab is selected based on the cursor position and shows preview for the current sub-expression
-  (e.g.~raw data before grouping in the screenshot).
-
-\item Another external library provides support for charting (right). The environment does not
-  handle charts in a special way and displays the result of evaluating the whole command. The
-  screenshot shows a case where the user modifies parameters of the chart. Thanks to our live evaluation
-  strategy, this is done efficiently without reevaluating the data transformation.
-\end{itemize}
-
-\noindent
-A user study to evaluate the usability of the system from a human-computer interaction perspective
-is left for future work. As an anecdotal evidence of the usability, the code in
-Figure~\ref{fig:gallery} was developed by an attendee of a tutorial at Mozilla Festival 2017
-who had no prior programming experience.
-
 % ==================================================================================================
@@
 
-\section{Related and future work}
-\label{sec:future}
-
-We aim to make creating transparent data analyses, such as those done by journalists, easier and we
-present a tool that provides instantaneous feedback when writing code in a text editor. Many our
-design decisions have alternatives that have been explored in a rich body of literature.
-
-\subsubsection*{Notebooks and data science tools.}
-Basic feedback mechanism, the REPL (read-eval-print-loop), originated in Lisp
-\cite{lisp} and is now common in many languages \cite{drscheme}. Notebook systems
-such as Jupyter and R markdown \cite{rmarkdown,jupyter,ipython} structure code in cells,
-but still rely on explicitly executing individual cells using a stateful REPL environment.
-Integrating provenance tracking with notebooks \cite{dataflow,noworkflow,wrattler} addresses
-the reproducibility issues of this model, but keeps cells as units of work. Ideas such as
-dependency tracking and efficient recomputation exist in many visual data exploration tools
-\cite{control,tableau,vizdom} and scientific workflow systems \cite{taverna,kepler}. Some of the
-implementation techniques are related to our work, but we focus on text-based scripts.
-Tempe \cite{tempe} focuses on streaming data, but provides a text-based scripting environment
-with automatic code update and \citet{ripple} empirically evaluate its usability in contrast to
-REPLs. The challenges faced by Tempe are more akin to dynamic software update \cite{dsu}.
-
-\subsubsection*{Live coding and editing.}
-Live coding based on textual programs has been popularised by \citet{principle,learnable} and
-is actively developed in domains such as live coded music \cite{beyond,sonic}. \citet{liveroad}
-reviews the use of live coding in a Smalltalk derived environment. Lighttable by \citet{lighttable}
-and Chrome DevTools provide limited live previews akin to those presented in this paper,
-but without well specified recomputation model. Finally, work on keeping state during
-code edits \cite{alive,livingit} would be relevant for supporting streaming data. In the future,
-we also plan to explore how direct manipulation \cite{direct} could be integrated with our preview
-mechanism. \citet{liveroad}
-
-Using structured editors \cite{structure-based} rather than text editors is a promising alternative
-direction as editors can recompute program based on understanding of edit operations
-\cite{interactive,livenut,lamdu}. This can be elegantly implemented using bi-directional
-lambda calculus \cite{hazelnut}, but it also makes us consider more human-centric abstractions
-\cite{subtext,directprog} along the lines of the one outlined in Section~\ref{sec:extra-abstraction}.
-
-\subsubsection*{Incremental computation and dependency analysis.}
-Work on self-adjusting and incremental computation \cite{selfadjusting,adapton} handles
-recomputation when the program stays the same, but input changes. It relies on dependency tracking
-similar to ours. We rely on static code analysis whereas most incremental systems,
-e.g.~\cite{incremental,adaptivefp,deltaml} evaluate the program and use programmer-supplied
-information to build a dependency graph. \citet{adapton} implement a small adaptive interpreter
-that treats code as changing input data, suggesting an interesting implementation technique for
-live coding systems. Our use of dependency graphs \cite{dependencies} is static and first-order
-and can be seen as a form of program slicing \cite{slicing}, although our binding process is
-more directly inspired by Roslyn \cite{roslyn}, which uses it for efficient background type-checking.
-
-\subsubsection*{Semantics and partial evaluation.}
-The evaluation of previews is a form of partial evaluation \cite{partial}, done in a
-way that allows caching. This can be done implicitly or explicitly in the form of
-multi-stage programming \cite{metaml}. Semantically, the evaluation of previews can be seen as a
-modality \cite{modal} and delayed previews are linked to contextual modal type theory \cite{cmtt},
-formally modelled using comonads \cite{cmtt-denotation}. This suggests a direction for rigorous
-analysis of the presented system.
-
-% ==================================================================================================
-%
-% \section{TODO}
-%
-% TBD: Define live preview somewhere
-%
-% TBD: In section about dependency graphs, include a subsection discussing how this relates
-% to incremental computation (just like "Semantics of delayed previews" links that to CMTT etc.)
-%
-%
-% \cite{selfadjusting}
-% \cite{incremental}
-%
-% Incremental computation
-%
-%   Self-adjusting computation
-%   http://www.cs.cmu.edu/~rwh/theses/acar.pdf
-%
-%   Incremental Computation with Names
-%   https://www.cs.tufts.edu/~jfoster/papers/oopsla15.pdf
-%   -> dsl interpreter
-%
-%   (and other adapton stuff http://adapton.org/)
-%
-% Type checking
-%
-%   co-contextual
-%   http://www.informatik.uni-marburg.de/~seba/publications/cocontextual-type-checking.pdf
-%   http://drops.dagstuhl.de/opus/volltexte/2017/7262/pdf/LIPIcs-ECOOP-2017-18.pdf
-%
-% https://dl.acm.org/citation.cfm?id=664109
-%
-\section{Summary}
-One of the reasons that make spreadsheets easier to use than programming tools is that they
-provide instant feedback. We aim to make programming tools as instantaneous as spreadsheets.
-We formalized a number of key aspects of simple data analyses such as those don by journalists
-and then used our observations to build both theory and simple practical data analytics tools.
-
-Our \emph{data exploration calculus} is a simple formally tractable language
-for data exploration. The calculus captures key observations about simple data analyses. They
-rely on logic from external libraries, implement few abstractions and are written as lists of
-commands. We later revised the calculus to add type checking and simple abstraction mechanism.
-
-Our main technical contribution is a \emph{live preview} mechanism that efficiently evaluates
-code during editing and instantly provides a preview of the result. We allow users to edit code
-in unconstrained way in an text editor, which makes this particularly challenging. The key trick
-is to separate the process into fast \emph{binding phase}, which constructs a dependency graph
-and slower \emph{evaluation phase} that can cache results. This makes it possible to quickly parse
-updated code, reconstruct dependency graph and compute preview using previous, partially
-evaluated, results.
-
-We evaluated our approach in three ways. First, we proved that our mechanism is correct and
-that it reuses evaluated values for many common code edit operations. Second, we conducted an
-experimental study that illustrates how often are previously evaluated results reused during
-typical programming scenario. Thirdly, we used our research as a basis for online data exploration
-environment, which shows the practical usability of our work.
-
-\bibliography{paper}
-
-\newpage
-\appendix
-
-\section{Appendix}
-
-\subsection{Normalization}
-\label{sec:app-normalization}
-
-\begin{theorem}[Normalization]
-For all $p$, there exists $n$ and $o_1, \ldots, o_n$ such that $p\rightsquigarrow^{*} o_1;\ldots;o_n$.
-\end{theorem}
-\begin{proof}
-We define $\ident{size}$ of a program in data exploration calculus as follows:
-\begin{equation}
-\begin{array}{rcl}
-\ident{size}(c_1; \ldots; c_n) &=& 1 + \Sigma^{n}_{i=1} \ident{size}(c_i)\\
-\ident{size}(\kvd{let}~x=t) &=& 1 + \ident{size}(t)\\
-\ident{size}(e_0.m(e_1, \ldots, e_n)) &=& 1 + \Sigma^{n}_{i=0} \ident{size}(e_i)\\
-\ident{size}(\lambda x\rightarrow e) &=& 1 + \ident{size}(e)\\
-\ident{size}(o) = \ident{size}(x) &=& 1\\
-\end{array}
-\end{equation}
-
-\noindent
-The property holds because, first, both \rname{let} and \rname{external} decrease the \ident{size}
-of the program and, second, a program is either fully evaluated, i.e.~$o_1;\ldots;o_n$ for some $n$
-or, it can be reduced using one of the reduction rules.
-\end{proof}
-
-\subsection{Term preview correctness}
-\label{sec:app-correctness}
-
-\begin{theorem}[Term preview correctness]
-Given a term $t$ that has no free variables, together with a lookup table $\Delta$ obtained
-from any sequence of programs using $\ident{bind-prog}$ (Figure~\ref{fig:binding-rules-prog}) and
-$\ident{update}$ (Figure~\ref{fig:loop}), then
-let $\bndclr{v}, (V, E) = \ident{bind-expr}_{\emptyset, \Delta}(t)$. If $\bndclr{v}\Downarrow p$
-over a graph $(V, E)$ then $p = o$ for some value $o$ and $t \rightsquigarrow^{*} o$.
-\end{theorem}
-
-\begin{proof}
-First note that, when combining recursively constructed sub-graphs, the \ident{bind-expr} function
-adds new nodes and edges leading from those new nodes. Therefore, an evaluation using~$\Downarrow$
-over a sub-graph will also be valid over the new graph -- the newly added nodes and edges do not introduce
-non-determinism to the rules given in Figure~\ref{fig:eval}.
-
-We prove a more general property showing that for any $e$, its binding
-$\bndclr{v}, (V, E) = \ident{bind-expr}_{\emptyset, \Delta}(e)$ and any evaluation context $C$
-such that $C[e]\rightsquigarrow o$ for some $o$, one of the following holds:
-%
-\begin{enumerate}
-\item[a.] If $FV(e)\,=\,\emptyset$ then $\bndclr{v} \Downarrow p$ for some $p$ and $C[p] \rightsquigarrow o$
-\item[b.] If $FV(e)\neq\emptyset$ then $\bndclr{v} \Downarrow \llbracket e_p \rrbracket_{FV(e)}$ for some $e_p$ and $C[e_p] \rightsquigarrow o$
-\end{enumerate}
-%
-In the first case, $p$ is a value, but it is not always the case that $e \rightsquigarrow^{*} p$,
-because $p$ may be lambda function and preview evaluation may reduce sub-expression in the body of
-the function. Using a context $C$ in which the value reduces to an object avoids this problem.
-
-The proof of the theorem follows from the more general property. Using a context $C[-]=-$, the
-term $t$ reduces $t \rightsquigarrow^{*}t' \rightsquigarrow_\epsilon o$ for some $o$ and the
-preview $p$ is a value $o$ because $C[p] = p = o$.
-The proof is by induction over the binding process, which follows the structure of the expression:
-
-\vspace{0.75em}\noindent(1) $\ident{bind-expr}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n))$ --
-  Here $e = e_0.m(e_1, \ldots, e_n)$, $\bndclr{v_i}$ are graph nodes obtained by induction for
-  expressions $e_i$ and $\{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v}, \bndclr{v_n}, \blbl{arg}(n))\} \subseteq E$.
-  From lookup inversion Lemma~\ref{thm:lemma-lookup}, $\bndclr{v} = \bnd{mem}(m, s)$ for some $s$.
-
-  If $FV(e)=\emptyset$, then $\bndclr{v_i} \Downarrow p_i$ for $i\in 0\ldots n$ and
-  $\bndclr{v}\Downarrow p$ using \rname{mem-val} such that $p_0.m(p_1, \ldots, p_n) \rightsquigarrow p$.
-  From induction hypothesis and \emph{compositionality} of external libraries (Definition~\ref{def:external}),
-  it holds that for any $C$ such that $C[e_0.m(e_1, \ldots, e_n)] \rightsquigarrow o$ for some $o$
-  then also $C[p_0.m(p_1, \ldots, p_n)] \rightsquigarrow C[p] \rightsquigarrow o$.
-
-  If $FV(e)\neq\emptyset$, then $\bndclr{v_i} \Downarrow_\ident{lift} \llbracket e'_i \rrbracket$ for $i\in 0\ldots n$ and
-  $\bndclr{v}\Downarrow \llbracket e'_0.m(e'_1, \ldots, e'_n) \rrbracket_{FV(e)}$ using \rname{mem-expr}.
-  From induction hypothesis and \emph{compositionality} of external libraries (Definition~\ref{def:external}),
-  it holds that for any $C$ such that $C[e_0.m(e_1, \ldots, e_n)] \rightsquigarrow o$ for some $o$
-  then also $C[e'_0.m(e'_1, \ldots, e'_n)] \rightsquigarrow o$.
-
-\vspace{0.75em}\noindent(2) $\ident{bind-expr}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n))$ --
-  This case is similar to (1), except that the fact that $\bndclr{v} = \bnd{mem}(m, s)$
-  holds by construction, rather than using Lemma~\ref{thm:lemma-lookup}.
-
-\vspace{0.75em}\noindent(3) $\ident{bind-expr}_{\Gamma, \Delta}(\lambda x\rightarrow e_b)$ --
-  Here $e = \lambda x\rightarrow e_b$, $\bndclr{v_b}$ is the graph node obtained by induction
-  for the expression $e_b$ and $(\bndclr{v}, \bndclr{v_b}, \blbl{body}) \in E$.
-  From lookup inversion Lemma~\ref{thm:lemma-lookup}, $\bndclr{v} = \bnd{fun}(x, s)$ for some $s$.
-  The evaluation can use one of three rules:
-
-  \begin{enumerate}
-  \item[i.]
-  If $FV(e)=\emptyset$ then $\bndclr{v_b}\Downarrow p_b$ for some $p_b$ and $\bndclr{v}\Downarrow \lambda x\rightarrow p_b$
-  using \rname{fun-val}. Let $e'_b=p_b$.
-
-  \item[ii.]
-  If $FV(e_b)=\{x\}$ then $\bndclr{v_b}\Downarrow \llbracket e'_b \rrbracket_x$ for some $e'_b$ and
-  $\bndclr{v}\Downarrow \lambda x\rightarrow e'_b$ using \rname{fun-bind}.
-
-  \item[iii.]
-  Otherwise, $\bndclr{v_b}\Downarrow \llbracket e'_b \rrbracket_{x,\Gamma}$ for some $e'_b$ and
-  $\bndclr{v}\Downarrow \llbracket\lambda x\rightarrow e'_b\rrbracket_\Gamma$ using \rname{fun-expr}.
-  \end{enumerate}
-
-\noindent
-For i.) and ii.) we show that a.) is the case; for iii.) we show that b.) is the case; that is
-for any $C$, if $C[\lambda x\rightarrow e_b] \rightsquigarrow o$ then also
-$C[\lambda x\rightarrow e'_b] \rightsquigarrow o$. For a given $C$, let $C'[-] = C[\lambda x\rightarrow -]$
-and use the induction hypothesis, i.e.~if $C'[e_b] \rightsquigarrow o$ for some $o$ then also $C'[e'_b] \rightsquigarrow o$.
-
-\vspace{0.75em}\noindent(4) $\ident{bind-expr}_{\Gamma, \Delta}(\lambda x\rightarrow e)$  --
-  This case is similar to (3), except that the fact that $\bndclr{v} = \bnd{fun}(x, s)$
-  holds by construction, rather than using Lemma~\ref{thm:lemma-lookup}.
-
-\vspace{0.75em}\noindent(5) $\ident{bind-expr}_{\Gamma, \Delta}(o)$ -- In this case $e=o$ and $\bndclr{v} = \bnd{val}(o)$
-  and $\bnd{val}(o)\Downarrow o$ using \rname{val} and so the case a.) trivially holds.
-
-\vspace{0.75em}\noindent(6) $\ident{bind-expr}_{\Gamma, \Delta}(x)$ -- The initial $\Gamma$ is empty,
-  so $x$ must have been added to $\Gamma$ by case (3) or (4). Hence,
-  $\bndclr{v}=\bnd{var}(x)$, $\bndclr{v}\Downarrow \llbracket x \rrbracket_{x}$ using \rname{var}
-  and so $e_p = e = x$ and the case b.) trivially holds.
-
-\end{proof}
-
-\section{Appendix - Type checking}
-\label{sec:app-types}
-
-tbd
-
-% Evaluating live previews can be an expensive operations, so being able to cache partial previews
-% is a must for a live coding environment. Type checking is typically fast, so the main focus of
-% this paper is on live previews. However, asynchronous type providers in The Gamma (Section~\ref{sec:types-providers})
-% can make type checking time consuming, and so we use the dependency graph also for type checking
-% (Section~\ref{sec:types-graph}). Type checking lambda functions (Section~\ref{sec:types-funs})
-% requires a slight extension of the model discussed in Section~\ref{sec:formal}.
-%
-% % --------------------------------------------------------------------------------------------------
-%
-% \subsection{Asynchronously provided types}
-% \label{sec:types-providers}
-%
-% Data available in The Gamma can be defined using several kinds of type providers. The type provider
-% used in Figure~\ref{fig:thegamma} as asynchronous \cite{async}. It downloads the sample URL and generates
-% types based on the contents of the web page. The parameter to \ident{web.scrape} is a static
-% parameter and is evaluated during type-checking. We omit details in this paper, but we note this
-% works similarly to F\# type providers \cite{providers-fsharp}.
-%
-% Type providers can also be implemented as REST services \cite{rest} to allow anyone implement a data source
-% in the language of their choice. In this case, each member of a call-chain returns a type that is
-% generated based on the result of an HTTP request. For example, when the user types \ident{worldbank}
-% (to access information about countries), the type provider makes a request to
-% \url{http://thegamma-services.azurewebsites.net/worldbank}, which returns two members:
-% %
-% \begin{equation*}
-% \begin{array}{l}
-% [~\{\rstr{name}\!:\str{byYear},\rstr{returns}\!:\\
-% ~\quad \{\rstr{kind}\!:\str{nested},\rstr{endpoint}\!:\str{/pickYear}\}\}\\
-% \hspace{0.55em}\{\rstr{name}\!:\str{byCountry},\rstr{returns}\!:\\
-% ~\quad \{\rstr{kind}\!:\str{nested},\rstr{endpoint}\!:\str{/pickCountry}\}\}~]
-% \end{array}
-% \end{equation*}
-% %
-% This indicates that \ident{worldbank} has members \ident{byYear} and \ident{byCountry}.
-% If the user types \ident{worldbank.byCountry}, a request is made to the specified URL
-% \url{http://thegamma-services.azurewebsites.net/worldbank/pickCountry}:
-% %
-% \begin{equation*}
-% \begin{array}{l}
-% [~\{\rstr{name}\!:\str{Andorra},\rstr{trace}\!:[\str{"country=AR"}], \\
-% \quad \rstr{returns}\!:\{\rstr{kind}\!:\str{nested},\rstr{endpoint}\!:\str{/pickTopic}\}\}\\
-% \hspace{0.55em}\{\rstr{name}\!:\str{Afghanistan},\rstr{trace}\!:[\str{"country=AF"}],\\
-% \quad \rstr{returns}\!:\{\rstr{kind}\!:\str{nested},\rstr{endpoint}\!:\str{/pickTopic}\}\},\,..]
-% \end{array}
-% \end{equation*}
-% %
-% This returns a list of countries which can then be accessed as members via
-% \ident{worldbank.byCountry.Andorra}, etc.
-%
-% This is one reason for why type checking in The Gamma can be time consuming. Other
-% type providers may perform other more computationally intensive work to provide types and so
-% it is desirable to reuse type-checking results during live coding. The rest of this section
-% shows how this is done using the dependency graph discussed in Section~\ref{sec:formal}.
-%
-% \begin{figure}[!b]
-% \vspace{-0.5em}
-% \begin{equation*}
-% \xymatrix{
-% &\bnd{val}(o)\\
-% &\bnd{mem}(m, s_0) \ar[u]_{\blbl{arg}(0)} \ar[dl]^{\blbl{arg}(1)} \\
-% \bnd{fun}(x, s_2) \ar@/^/[uur]^{\blbl{callsite}(m, 1)} \ar[rr]^{\blbl{body}} && \bnd{var}(x, s_1) \ar@/_/[uul]_{\blbl{callsite}(m, 1)}
-% }
-% \end{equation*}
-% \caption{Dependency graph for $o.m(\lambda x\rightarrow x)$}
-% \label{fig:graph-func}
-% \end{figure}
-%
-% % --------------------------------------------------------------------------------------------------
-%
-% \begin{figure*}[t]
-% \vspace{-0.5em}
-% \begin{equation*}
-% \begin{array}{ll}
-% \ident{bind-expr}_{\Gamma, \Delta, c}(e_0.m(e_1, \ldots, e_n)) = &(2b)\\
-% \qquad \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n)\\
-% \quad \textnormal{when}~\bndclr{v_0}, (V_0, E_0) = \ident{bind-expr}_{\Gamma, \Delta, \bot}(e_0)\\
-% \quad \textnormal{and}~c_i = (\bndclr{v_0}, \blbl{callsite}(m,i))\hspace{4.8em} (i \in 1\ldots n)\\
-% \quad \textnormal{and}~\bndclr{v_i}, (V_i, E_i) = \ident{bind-expr}_{\Gamma, \Delta, c_i}(e_i)\qquad (i\in 1\ldots n)\\
-% \quad \textnormal{and}~(\bknd{mem}(m),[(\bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v_n}, \blbl{arg}(n))]) \notin \ident{dom}(\Delta)\hspace{-200em}.\\
-% \quad \textnormal{let}~\bndclr{v} = \bnd{mem}(m, s), s~\textnormal{fresh}\\
-% \quad \textnormal{let}~E = \{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v},\bndclr{v_n}, \blbl{arg}(n)) \}
-% \end{array}
-% \qquad\qquad
-% \begin{array}{ll}
-% \ident{bind-expr}_{\Gamma, \Delta, (\bndclr{v_c}, \blblclr{l_c} )}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V_0, E \cup E_0)&(7b)\\
-% \quad \textnormal{when}~(\bknd{var}(x),[(\bndclr{v_c}, \blblclr{l_c} )])\notin \ident{dom}(\Delta)\\
-% \quad \textnormal{let}~\bndclr{v_x} = \bnd{var}(x, s_x), s_x~\textnormal{fresh}\\
-% \quad \textnormal{and}~\Gamma_1 = \Gamma \cup \{ x, \bndclr{v_x} \}\\
-% \quad \textnormal{and}~\bndclr{v_0}, (V_0, E_0) = \ident{bind-expr}_{\Gamma_1, \Delta, \bot}(e)\\
-% \quad \textnormal{when}~(\bknd{fun}(x),[(\bndclr{v_0}, \blbl{body}), (\bndclr{v_c}, \blblclr{l_c} )])\notin \ident{dom}(\Delta)\\
-% \quad \textnormal{let}~\bndclr{v} = \bnd{fun}(x, s_f), s_f~\textnormal{fresh}\\
-% \quad \textnormal{let}~E = \{ (\bndclr{v},\bndclr{v_0},\blbl{body}), (\bndclr{v},\bndclr{v_c},\blblclr{l_c}), (\bndclr{v_x},\bndclr{v_c},\blblclr{l_c}) \}
-% \\[0.75em]
-% \end{array}
-% \end{equation*}
-% \vspace{-0.5em}
-% \caption{Revised binding rules, tracking call sites of function values.}
-% \label{fig:binding-rules-callsite}
-% \vspace{-0.5em}
-% \end{figure*}
-%
-% % --------------------------------------------------------------------------------------------------
-%
-% \subsection{Revised binding of functions}
-% \label{sec:types-funs}
-%
-% The Gamma script supports lambda functions, but only in a limited way. A function can be passed
-% as a parameter to a method, which makes type checking of functions easier. For example, consider:
-% %
-% \begin{equation*}
-% \begin{array}{l}
-% \ident{movies}.\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())
-% \end{array}
-% \end{equation*}
-% %
-% If \ident{movies} is a collection of \ident{Movie} objects, the type of the lambda function must be
-% $\ident{Movie}\rightarrow\ident{bool}$ and so the type of $x$ is \ident{Movie}. This is similar to
-% type checking of lambda functions in C\# \cite{csharp}, where type is also inferred from the context (or has
-% to be specified explicitly).
-%
-% We do not currently allow lambda functions as stand-alone let-bound
-% values. This could be done by requiring explicit types, or introducing polymorphism, but it was not
-% necessary for the limited domain of non-expert data exploration.
-%
-% \paragraph{Dependency graph for functions.}
-% In the binding process specified in Section~\ref{sec:formal}, a variable is a leaf of the dependency
-% graph. In the revised model, it depends on the context in which it appears. A new edge labeled
-% $\blbl{callsite}(m, i)$ indicates that the source node is the input variable of a function
-% passed as the $i^\textnormal{th}$ argument to the $m$ member of the expression represented by the
-% target node. A node representing function is linked to the call site using the same edge.
-%
-% Figure~\ref{fig:graph-func} shows the result of binding $o.m(\lambda x\rightarrow x)$. Both
-% $\bnd{fun}(x, s_2)$ and $\bnd{var}(x, s_1)$ now depend on the node representing $o$. The new
-% $\blbl{callsite}$ edge makes it possible to type-check function and variable nodes just using their
-% dependencies. As before, the member invocation $\bnd{mem}(m, s_0)$
-% depends on the instance using $\blbl{arg}(0)$ and on its argument using $\blbl{arg}(1)$.
-%
-% % --------------------------------------------------------------------------------------------------
-%
-% \begin{figure*}
-% \vspace{0.5em}
-% \begin{equation*}
-% \begin{array}{l}
-% \qquad\inference[\rname{val}~]
-%   {\Sigma(n) = \alpha}
-%   {\bnd{val}(n) \vdash \alpha }
-% \quad\qquad
-% \inference[\rname{var}~]
-%   { (\bnd{var}(x,s), \bndclr{v}, \blbl{callsite}(m, i)) \in E &
-%     \bndclr{v} \vdash (.., m:(\tau_1, \ldots, \tau_k)\rightarrow \tau,..) & \tau_i=\tau'\rightarrow\tau'' }
-%   { \bnd{var}(x, s) \vdash \tau' }
-% \\[2.5em]
-% \inference[\rname{fun}~]
-%   { \{(\bnd{fun}(x,s), \bndclr{v_b}, \blbl{body}), (\bnd{var}(x,s), \bndclr{v_c}, \blbl{callsite}(m, i))\} \subseteq E &
-%     \bndclr{v_c} \vdash (.., m:(\tau_1, \ldots, \tau_k)\rightarrow \tau,..) &
-%     \tau_i=\tau'\rightarrow\tau'' & \bndclr{v_b} \vdash \tau'' }
-%   { \bnd{fun}(x, s) \vdash \tau' \rightarrow \tau'' }
-% \\[2.5em]
-% \qquad\qquad\quad\inference[\rname{mem}~]
-%   { \forall i\in\{0\ldots k\}.(\bnd{mem}(m, s), \bndclr{v_i}, \blbl{arg}(i)) \in E &
-%     \bndclr{v_0} \vdash (.., m:(\tau_1, \ldots, \tau_k) \rightarrow \tau, ..)& \bndclr{v_i} \vdash \tau_i }
-%   { \bnd{mem}(m, s) \vdash \tau }
-% \\[1em]
-% \end{array}
-% \end{equation*}
-% \caption{Rules that define evaluation of previews over a dependency graph for an expression}
-% \vspace{0.5em}
-% \label{fig:tc}
-% \end{figure*}
-%
-% % --------------------------------------------------------------------------------------------------
-%
-% \paragraph{Revised binding process.}
-% For the revised binding process, we introduce a new edge label $\blbl{callsite}$. Variable nodes
-% now have dependencies and so we cache them and attach a symbol $s$ to \bnd{var}, so we also
-% introduce a node kind $\bknd{var}$ as part of a lookup key for $\Delta$.
-%
-% The \ident{bind-expr} function now has a parameter $c$, in addition to $\Gamma$ and $\Delta$, which
-% represents the context in which the binding happens. This is either a member invocation (labeled
-% with instance node $\bndclr{v}$ and \blbl{callsite} label $\blblclr{l}$), or not a member
-% invocation written as $\bot$. The updated definitions are:
-% %
-% \begin{equation*}
-% \begin{array}{lcll}
-% \bndclr{v} &\hspace{-0.5em}\in\hspace{-0.5em}&\bnd{val}(n)~|~\bnd{var}(x, s)~|~\bnd{mem}(m, s)~|~\bnd{fun}(x, s)&(\textit{Vertices})\\
-% \blblclr{l}&\hspace{-0.5em}\in\hspace{-0.5em}&\blbl{body}~|~\blbl{arg}(i)~|~\blbl{callsite}(m,i)&(\textit{Edge labels})\\
-% \bkndclr{k}&\hspace{-0.5em}\in\hspace{-0.5em}&\bknd{fun}(x)~|~\bknd{mem}(m)~|~\bknd{var}(x)&(\textit{Node kinds})\\
-%           c&\hspace{-0.5em}\in\hspace{-0.5em}&\bot ~|~(\bndclr{v}, \blblclr{l})&(\textit{Call sites})\\
-% \end{array}
-% \end{equation*}
-% %
-% The key parts of the revised definition of the \ident{bind-expr} function are shown in Figure~\ref{fig:binding-rules-callsite}.
-% We now write $\ident{bind-expr}_{\Gamma,\Delta,c}$ where $c$ represents the context in which the
-% binding occurs. This is set to $\bot$ in all cases, except when binding arguments of a member call.
-% In (2b), we first recursively bind the instance node using $\bot$ as the context and then
-% bind all the arguments using $(\bndclr{v_0}, \blbl{callsite}(m, i))$ as the context for $i^{\textnormal{th}}$
-% argument. The rest is as in case (2) before. The case (1) is updated similarly and is not shown
-% here for brevity.
-%
-% When binding a function (7b), we now also store variable nodes in $\Delta$ and so we check if a
-% variable with the given call site exists. If no, we create a fresh node $\bnd{var}(x, s_x)$.
-% The node is added to $\Gamma$ as before. At the end, we now also include a call site edge from the
-% variable node and from the function node in $E$. We omit a similar variant of the case (6).
-% The remaining cases (3)-(5) are the same, except that \ident{bind-expr} has the additional $c$ parameter
-% and recursive calls always set it to $\bot$.
-%
-% Finally, the \ident{update} function in Figure~\ref{fig:loop} also needs to be updated to store the
-% newly created \bnd{var} nodes. This is done by adding the following single case:
-%
-% \begin{equation*}
-% \begin{array}{l}
-% \Delta_{i}(\bknd{var}(x), [(\bndclr{v},\blbl{callsite}(m,i))]) = \bnd{var}(x, s)\\
-% \quad\quad \textnormal{for all}~\bnd{var}(x, s) \in V\\
-% \quad\quad \textnormal{such that}~(\bnd{var}(x, s), \bndclr{v}, \blbl{callsite}(m,i)) \in E
-% \end{array}
-% \end{equation*}
-%
-% % --------------------------------------------------------------------------------------------------
-%
-% \subsection{Type checking over dependency graphs}
-% \label{sec:types-graph}
-%
-% The type system for The Gamma supports a number of primitive types (such as integers and strings)
-% written as $\alpha$. Composed types include functions and objects with members. Objects are be
-% provided by type providers, but we omit the details here. Types of object members are written as
-% $\sigma$ and can have multiple arguments:
-% %
-% \begin{equation*}
-% \begin{array}{lcll}
-% \tau&\in&\alpha ~|~ \tau \rightarrow \tau ~|~ \{m_1\!:\!\sigma_1, \ldots, m_n\!:\!\sigma_n\} &\quad(\textit{Types})\\
-% \sigma &\in& (\tau_1, \ldots, \tau_n) \rightarrow \tau &\quad(\textit{Members})\\
-% \end{array}
-% \end{equation*}
-% %
-% The typing judgements are written in the form $\bnd{v}\vdash \tau$. They are parameterised by the
-% dependency graph $(V, E)$, but this is not modified during type checking so we keep it implicit
-% rather than writing e.g.~$\bnd{v}\vdash_{(V,E)} \tau$.
-%
-% \paragraph{Type checking.}
-% The typing rules are shown in Figure~\ref{fig:tc}. Types of primitive values $n$ are obtained
-% using a global lookup table $\Sigma$ \rname{val}. When type checking a member call \rname{mem},
-% we find its dependencies $\bndclr{v_i}$ and check that the first one (instance) is an object with
-% the required member $m$. The types of input arguments of the member then need to match the types
-% of the remaining (non-instance) nodes.
-%
-% Type checking a function \rname{fun} and a variable \rname{var} is similar. In both cases,
-% we follow the \blbl{callsite} edge to find the member that accepts the function as an argument.
-% We obtain the type of the function from the type of the $i^\textnormal{th}$ argument of the
-% member. We use the input type as the type of variable \rname{var}. For functions, we also check
-% that the resulting type matches the type of the body \rname{fun}.
-%
-% \paragraph{Caching results.}
-% Performing type checking over the dependency graph, rather than over the abstract syntax tree,
-% enables us to reuse the results of previously type checked parts of a program. As when
-% caching evaluated previews (Section~\ref{sec:previews}), we build a lookup table mapping graph
-% nodes to types. When type checking a node, we first check the cache and, only if it is new,
-% follow the $\vdash$ relation to obtain the type.
-%
-% As a result, code can be type checked on-the-fly during editing, even when asynchronous type
-% providers are used, and the programmer gets instant feedback without delays.
-%
-% % --------------------------------------------------------------------------------------------------
-%
-% \subsection{Properties of type checking}
-%
-% As noted in Section~\ref{sec:types}, the focus of this paper is on live previews, but we also
-% use the method based on reusing nodes in a dependency graph for type checking. We do not discuss
-% properties of type checking in detail, but we briefly note how the different properties of
-% live previews extend to corresponding properties of type checking.
-%
-% \paragraph{Type checking result reuse.} In Section~\ref{sec:evaluation-reuse}, we show that
-% certain source code edits do not cause the recomputation of previews for the whole expression
-% or a sub-expression. The edits are given in Figure~\ref{fig:operations}. The proof uses the fact
-% that the newly bound graph (after code edit) reuses nodes of the previous graph. This implies that
-% type checking results can be reused in exactly the same way as live previews -- they are also
-% stored in a lookup table with graph nodes as keys.
-%
-% \paragraph{Correctness.} The correctness property (Theorem~\ref{thm:correctness}) shows that
-% graph-based preview evaluation matches direct evaluation of expressions. To show a corresponding
-% property for type checking, we would need to provide ordinary type system based on the structure
-% of the expression and prove that the two are equivalent. In our implementation, we only use the
-% presented graph-based type checking method, so we do not provide an alternate account in this paper.
-%
-% \paragraph{Determinacy.} The determinacy property (Theorem~\ref{thm:determinacy}) guarantees that
-% previews can be cached, because evaluating them again, using $\Downarrow$ over an updated graph,
-% would yield the same result. The same property holds for $\vdash$, meaning that type checking
-% results can be cached. Although the Theorem~\ref{thm:determinacy} talks explicitly about
-% $\Downarrow$, it can be easily extended for $\vdash$, because the proof depends on how the graph
-% is updated using $\ident{update}_{V,E}$ and the binding process.
-%
 \end{document}

### 2019-09-30 wip-live-programming/paper.tex
Adding submitted version

@@
 \usepackage{amsthm}
+\usepackage{enumitem}
 
@@
 \author[a]{Tomas Petricek}
-\authorinfo{Tomas is xxx. Contact him at \email{tomas@tomasp.net}.}
+\authorinfo{
+\noindent
+Tomas is a lecturer at University of Kent. He is working on making programming and data science
+easier and more accessible. His research interests span a range of areas including theory of
+programming languages, history and philosophy of programming, as well as open data and data
+journalism.
+%
+Previously, he contributed to functional programming and the development of the F\# language and
+type providers at Microsoft Research and obtained PhD from University of Cambridge for his work
+on coeffects, a theory of context-aware programming languages. Along the way, he became interested
+in understanding programming through the perspective of history and philosophy of science and wrote
+papers about the evolution of programming concepts such as types and errors.
+%
+Contact him at \email{tomas@tomasp.net}.}
 \affiliation[a]{University of Kent, UK}
-\keywords{xxx}
+\keywords{Data exploration, live coding, data journalism, instant previews}
 \paperdetails{
-  perspective=engineering,
-  area={History and philosophy of programming},
+  perspective=sciencetheoretical,
+  area={Programming environments},
 }
@@
 <concept>
-<concept_id>10002944.10011122.10002945</concept_id>
-<concept_desc>General and reference~Surveys and overviews</concept_desc>
+<concept_id>10002951.10003227.10003351</concept_id>
+<concept_desc>Information systems~Data mining</concept_desc>
 <concept_significance>300</concept_significance>
@@
 <concept>
-<concept_id>10003456.10003457.10003521</concept_id>
-<concept_desc>Social and professional topics~History of computing</concept_desc>
+<concept_id>10011007.10011006.10011041</concept_id>
+<concept_desc>Software and its engineering~Compilers</concept_desc>
+<concept_significance>300</concept_significance>
+</concept>
+<concept>
+<concept_id>10003120.10003121.10003129</concept_id>
+<concept_desc>Human-centered computing~Interactive systems and tools</concept_desc>
 <concept_significance>300</concept_significance>
@@
 \end{CCSXML}
-\ccsdesc[300]{General and reference~Surveys and overviews}
-\ccsdesc[300]{Social and professional topics~History of computing}
-
+
+\ccsdesc[300]{Human-centered computing~Interactive systems and tools}
+\ccsdesc[300]{Information systems~Data mining}
+\ccsdesc[300]{Software and its engineering~Compilers}
 \maketitle
 
+\vspace{1em}
+\noindent
+{\scriptsize
+~\\
+\textbf{Note to reviewers:} This paper has been previously submitted to a different
+venue. The revisions made in response to past reviews are summarized in Appendix~\ref{sec:revisions}
+(to be removed for publication).}
+\vspace{2em}
+
 \begin{abstract}
-TODO!
-
-Rewrite according to guidelines.
-
-TODO!
-
-A growing amount of code is written to explore and analyze data, but the way data analysts
-work with code differs from the way software engineers do. First, many data
-scientists write code that introduces few or no abstractions. Second, data exploration is an
-interactive process that blurs the traditional distinction between development-time and run-time.
-This poses an interesting challenge for programming language research, inviting us to
-revisit a number of traditional assumptions about programming languages and design a new kind of
-programming tools that are suited for data analyst needs.
-
-We present the \emph{data exploration calculus}, a formal language that captures the structure of
-scripts written to explore data. We focus on the simplest kind of data analyses such as those
-written by journalists to analyse open government data. We implement a programming environment,
-based on this calculus, that evaluates code instantly during editing and shows previews of the results,
-while allowing the user to modify code in an unrestricted way in a text editor. Supporting interactive
-editing is tricky. Any edit can change the structure of code and fully recomputing the output would
-be too expensive. We present a technique that allows evaluating and type-checking code while it is
-being written and that reuses results of previous runs. We formalize the technique using the data
-exploration calculus, prove that it is correct and specify when previous results are reused.
-Finally, we illustrate the practicality of our approach using empirical evaluation
-and a case study.
-
+%
+% Context: What is the broad context of the work? What is the importance of the general research area?
+A growing amount of code is written to explore and analyze data, often by data analysts
+who do not have a traditional background in programming, for example by journalists.
+%
+% Inquiry: What problem or question does the paper address? How has this problem or question been addressed by others (if at all)?
+The way such data anlysts write code is different from the way software engineers do so.
+They use few abstractions, work interactively and rely heavily on external libraries.
+We aim to capture this way of working and build a programming environment that makes data exploration
+easier by providing instant live feedback.
+
+% Approach: What was done that unveiled new knowledge?
+We combine theoretical and applied approach. We present the \emph{data exploration calculus},
+a formal language that captures the structure of code written by data analysts. We then
+implement a data exploration environment that evaluates code instantly during editing and shows
+previews of the results.
+%
+% Knowledge: What new facts were uncovered? If the research was not results oriented, what new capabilities are enabled by the work?
+We formally describe an algorithm for providing instant previews for the data exploration calculus
+that allows the user to modify code in an unrestricted way in a text editor. Supporting interactive
+editing is tricky as any edit can change the structure of code and fully recomputing the output
+would be too expensive.
+%
+% Grounding: What argument, feasibility proof, artifacts, or results and evaluation support this work?
+We prove that our algorithm is correct and that it reuses previous results when updating previews
+after a number of common code edit operations. We also illustrate the practicality of our approach
+with an empirical evaluation and a case study.
+
+% Importance: Why does this work matter?
 As data analysis becomes an ever more important use of programming, research on programming languages
@@
 conceive new kinds of tools that can support them. The present paper is one step in this important direction.
+
 \end{abstract}
@@
 
-\subsection{Data wrangling in Jupyter}
+\subsection{Simple data exploration in Jupyter}
 \label{sec:background-jupyter}
@@
 Here, we can directly evaluate sub-expressions \ident{movies} and $\ident{movies}.\ident{take}(\num{10})$,
-but not $x.\ident{getReleased}()$ and $x.\ident{getReleased}().\ident{format}(\str{dd-mm-yyyy})$
-because they contain a free variable $x$. Our preview evaluation algorithm addresses this by
+but not $x.\ident{getReleased}()$ because it contains a free variable $x$. Our preview evaluation algorithm addresses this by
 producing two kinds of previews. A \emph{fully evaluated preview} is just a value, while
@@
 \begin{equation*}
-\vspace{-2em}
+\vspace{-2.5em}
 \hspace{-0.5em}
@@
 \end{equation*}
+\vspace{-0.5em}
 \begin{equation*}
@@
 \end{equation*}
-\vspace{-0.5em}
+\vspace{-0.75em}
 \caption{Rules that define evaluation of previews over a dependency graph for a program}
 \label{fig:eval}
-\vspace{-0.5em}
+\vspace{-0.75em}
 \end{figure*}
@@
 The evaluation of previews is defined in Figure~\ref{fig:eval}. Given a dependency graph $(V, E)$,
-we define a relation $\bndclr{v}\Downarrow d$ that evaluates a sub-expression corresponding to
+the relation $\bndclr{v}\Downarrow d$ evaluates a sub-expression corresponding to
 the node $\bndclr{v}$ to a possibly delayed preview $d$. The nodes $V$ and edges $E$ of the
-dependency graph are parameters of the $\Downarrow$ relation, but they do not change during the
+graph are parameters of $\Downarrow$, but they do not change during the
 evaluation and so we do not explicitly write them.
@@
 \begin{proof}
-By induction over the binding process. See Appendix~\ref{sec:app-let-free-correct}.
+By induction over the binding process. See Appendix~\ref{sec:app-correctness}.
 \end{proof}
@@
 Theorem~\ref{thm:preview-reuse} proves that the operations given in Figure~\ref{fig:operations}
-preserve the preview for a marked sub-expression. It relies on the following lemma that
-generally characterizes one common kind of edits. Given two versions of a program
-that both contain the same sub-expression $e$, if the let bindings that define the values of
+preserve the preview for a marked sub-expression. It relies on a Lemma~\ref{thm:sub-expr} given in
+Section~\ref{sec:sub-expr} that generally characterizes one common kind of edits. Given two versions
+of a program that both contain the same sub-expression $e$, if the let bindings that define the values of
 variables used in $e$ do not change, then the graph node assigned to $e$ will be the same
@@
 % --------------------------------------------------------------------------------------------------
-
-\begin{lem}[Binding sub-expressions]
-\label{thm:sub-expr}
-Assume we have programs $p_1 = c_1; \ldots; c_k; K_c[e]; c_{k+1};\ldots; c_n$ and
-$p_2 = c'_1; \ldots; c'_k; K'_c[e]; c'_{k+1};\ldots;c'_n$ and $I\subseteq \{1\ldots k\}$ such
-that $\forall i\!\in\!I.\,c_i\!=\!c'_i$ and for each $x \in \bigcup_{i\in I}FV(c_i) \cup FV(e)$
-there exists $j\in I$ such that $c_j = \kvd{let}\,x=e$ for some $e$.
-%
-Given any $\Delta$, assume that the the first program is bound,
-i.e.~$\bndclr{v_1};\ldots;\bndclr{v_n}, (V, E) = \ident{bind-prog}_{\emptyset, \Delta}(p_1)$,
-the cache is updated $\Delta' = \ident{update}_{V, E}(\Delta)$ and the second
-program is bound,
-i.e.~$\bndclr{v'_1};\ldots;\bndclr{v'_n}, (V', E') = \ident{bind-prog}_{\emptyset, \Delta'}(p_2)$.
-
-Now, assume $\bndclr{v}, G = \ident{bind-expr}_{\Gamma, \Delta}(e)$ and
-$\bndclr{v'}, G' = \ident{bind-expr}_{\Gamma', \Delta'}(e)$ are the recursive calls to bind
-$e$ during the first and the second binding, respectively. Then, the graph nodes assigned to the
-sub-expression $e$ are the same, i.e.~$\bndclr{v} = \bndclr{v'}$.
-\end{lem}
-\begin{proof}
-First, assuming that $\forall x\in FV(e). \Gamma(x) = \Gamma'(x)$, we show by induction over the binding process of $e$
-for the first program that the result is the same. In cases (1) and (3), the updated $\Delta'$
-contains the required key and so the second binding proceeds using the same case. In cases
-(2) and (4), the second binding reuses the node created by the first binding using case (1) and
-(3), respectively. Cases (5) and (6) are the same.
-
-Second, when binding let bindings in $c_1; \ldots; c_k$, the initial $\Gamma = \emptyset$ during
-both bindings. Nodes added to $\Gamma$ and $\Gamma'$ for commands $c_j$ such that $j\in I$ are
-the same and nodes added for remaining commands do not add any new nodes referenced from $e$ and
-so $\bndclr{v} = \bndclr{v'}$ using the above.
-\end{proof}
 
@@
 for \rname{let-intro-var}, the node assigned to $x$ is the node assigned to $e$ which is the
-same as before the edit from Lemma~\ret{thm:sub-expr}.
+same as before the edit from Lemma~\ref{thm:sub-expr}.
 Cases \rname{let-intro-ins} and \rname{let-intro-del} are similar to \rname{let-intro-var}, but
@@
 
-\begin{figure}[b]
+\begin{figure}
 \raggedright
-
-\vspace{0.5em}\hspace{1.5em}{\small\sffamily (1) Enter the following code and then change parameter of blur from 4 to 8:}
+\vspace{-0.5em}
+{\small\sffamily (1) Enter the following code and then change parameter of blur from 4 to 8:}
+\vspace{-0.5em}
 \begin{equation*}
@@
 
-\vspace{0.5em}\hspace{1.5em}{\small\sffamily (2) Assign the result to a variable and start writing code for further operations:}
+{\small\sffamily (2) Assign the result to a variable and start writing code for further operations:}
+\vspace{-0.5em}
 \begin{equation*}
@@
 
-\vspace{0.5em}\hspace{1.5em}{\small\sffamily (3) Finish code to combine two images and then change parameter of combine from 20 to 80:}
+{\small\sffamily (3) Finish code to combine two images and then change parameter of combine from 20 to 80:}
+\vspace{-0.5em}
 \begin{equation*}
@@
 
-\vspace{0.5em}\hspace{1.5em}{\small\sffamily (4) Extract the parameter of combine to a let bound variable:}
+{\small\sffamily (4) Extract the parameter of combine to a let bound variable:}
+\vspace{-0.5em}
 \begin{equation*}
@@
 
+\vspace{-1em}
 \caption{Code edit operations that are used in the experimental evaluation}
 \label{fig:image-edits}
+\vspace{-0.5em}
 \end{figure}
@@
 
-\begin{figure}[t]
-\vspace{-0.5em}
-\includegraphics[scale=0.425]{figures/drawing.png}
-\caption{Time required to recompute the results of a sample program after individual tokens of
-  the program are added or modified for three different evaluations strategies.}
-\label{fig:drawing}
-\vspace{-0.5em}
-\end{figure}
-
-% --------------------------------------------------------------------------------------------------
-
-\subsubsection{Code edit operations and methodology}
-
 To avoid clutter when visualizing the performance, we update the preview after complete tokens are
-added (such as identifier, number, dot or left parenthesis) rather than after individual keystrokes.
-Figure~\ref{fig:image-edits} shows the sequence of code edit operations that we use to measure
+added rather than after individual keystrokes.
+Figure~\ref{fig:image-edits} shows the sequence of edits that we use to measure
 the delays in updating a live preview. We first enter an expression to load, greyscale and blur
@@
 recomputation. It then measures time needed to recompute values for the whole program and
-displays the resulting image. If the parsing fails, e.g.~because of an unclosed parenthesis,
-previews are not updated and we record only the time taken by parsing attempt. We compare the
+displays the resulting image. If the parsing fails, we record only the time taken by parsing attempt. We compare the
 delays of three different evaluation strategies:
@@
 \item \emph{Call-by-value.} Following the semantics in Section~\ref{sec:calculus-semantics},
-  all sub-expressions are fully evaluated before an expression is evaluated. In this case, work may
-  be done even if the evaluation results in an error. For example, we parse
-  $\ident{image}.\ident{load}(\str{shadow.png}).\ident{blur}$ as member access with no arguments.
-  The evaluation loads the image, but then fails because blur requires one argument.
-
-\item \emph{Lazy.} To address the wastefulness of call-by-value strategy discussed above, we
+  all sub-expressions are evaluated before an expression. This is often wasteful. For example,
+  we parse the expression $\ident{image}.\ident{load}(\str{shadow.png}).\ident{blur}$ as a member access with no
+  arguments. The evaluation loads the image, but then fails because blur requires one argument.
+
+\item \emph{Lazy.} To address the wastefulness of call-by-value strategy, we
   simulate lazy evaluation by implementing a version of the image processing library where
@@
 
-\item \emph{Live.} Finally, the live strategy uses the algorithm described in
+\item \emph{Live.} Finally, we use the algorithm described in
   Sections~\ref{sec:formal} and \ref{sec:previews}. The cache is empty at the beginning of the
-  experiment and we update it after each token is added. This is the only strategy that does
-  not start afresh after reparsing code.
+  experiment and we update it after each token is added. This is the only strategy where
+  evaluation does not start afresh after reparsing code.
 \end{itemize}
 
-\subsubsection{Efficiency evaluation}
-The experimental environment is implemented in F\# and compiled to JavaScript using the Fable
-compiler. We run the experiments in Firefox (version 64.0.2, 32bit) on Windows 10
-(build 1809, 64bit) on a computer with Intel Core i7-7660U CPU and 16Gb RAM.
-
-Figure~\ref{fig:drawing} shows times needed to recompute previews after individual tokens
-of the program are added, deleted or modified, according to the script in
-Figure~\ref{fig:image-edits}, resulting in 38 measurements for each of the 3
-evaluation strategies. We mark a number of notable points in the chart:
-
-\begin{enumerate}
-\item Loading image for the first time incurs small extra overhead in the live strategy.
-\item Greyscaling using the live strategy does not need to re-load the image.
-\item Accessing the blur member without arguments causes delay for call-by-value strategy.
-\item When varying the parameter of blur, the live strategy reuses previously greyscaled image.
-\item Introducing let binding does not cause recomputation when using live strategy.
-\item As in (c), accessing combine member without valid argument only affects call-by-value.
-\item Varying parameter of combine in live strategy does not recompute blur and is much faster.
-\item Introducing let binding, again, causes full recomputation for lazy can call-by-value.
-\end{enumerate}
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure}
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\vspace{-0.5em}
+\includegraphics[scale=0.425]{figures/drawing.png}
+\caption{Time required to recompute the results of a sample program after individual tokens
+  are added or modified for three different evaluations strategies.}
+\label{fig:drawing}
+\vspace{-0.5em}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[b]
 \vspace{-0.5em}
@@
 \vspace{-0.25em}
-\caption{Histogram showing the distribution of delays incurred when updating previews.
+\caption{Distribution of delays incurred when updating previews.
   We show a histogram computed from all delays (left) and only from delays larger
@@
 % --------------------------------------------------------------------------------------------------
+
+The experimental environment is implemented in F\# and compiled to JavaScript using the Fable
+compiler. We run the experiments in Firefox (version 64.0.2, 32bit) on Windows 10
+(build 1809, 64bit) on Intel Core i7-7660U CPU with 16Gb RAM.
+
+Figure~\ref{fig:drawing} shows times needed to recompute previews after individual tokens
+are added, deleted or modified, according to the script in
+Figure~\ref{fig:image-edits}, resulting in 38 measurements. We mark a number of notable points in the chart:
+
+\begin{enumerate}[itemsep=0em,label=\alph*.]
+\item Loading image for the first time incurs small extra overhead in the live strategy.
+\item Greyscaling using the live strategy does not need to re-load the image.
+\item Accessing the blur member without arguments causes delay for call-by-value.
+\item When varying the parameter of blur, the live strategy reuses the greyscaled image.
+\item Introducing let binding does not cause recomputation when using live strategy.
+\item As in (c), accessing a member without an argument only affects call-by-value.
+\item The live strategy is much faster when varying the parameter of combine.
+\item Introducing let binding, again, causes full recomputation for lazy and call-by-value.
+\end{enumerate}
 
@@
 on the right summarizes only delays for edit operations where the delay for at least one of
-the strategies was over 15ms. A more robust parser would be able to recover and the delays
-for call-by-value and lazy strategies would be even more significant.
-
-\subsubsection{Summary of results}
+the strategies was over 15ms. The histogram shows that the live strategy eliminates the longest
+delays (by caching partial results), with the exception of a few where the underlying operation
+takes a long time (such as blurring the image). The results would be even more significant with
+an error-recovering parser.
+
 The purpose of our experiment is not to exactly assess the overhead of our implementation.
@@
 keeps the same trend. Our \emph{live} strategy reuses values computed
-previously. Consequently, expensive operations (such as (d) and (g) in Figure~\ref{fig:drawing})
+previously. Consequently, expensive operations such as (d) and (g) in Figure~\ref{fig:drawing}
 are significantly faster, because they do not need to recompute operations done previously when
@@
 such as journalists, to analyse, summarize and visualize open data. In this section, we briefly
-report on our experience with the system.
-
-The environment is available at \url{http://gallery.thegamma.net}. Two screenshots shown
+report on our experience with the system. Two screenshots shown
 in Figure~\ref{fig:gallery} illustrate a number of interesting features:
@@
 \begin{itemize}
-\item The left screenshot uses a type provider for data aggregation \cite{gamma}
-  (Section~\ref{sec:background-gamma}). Our system views the type provider as an external library
-  (with objects, members and reduction relation). Type providers rely on type information to
+\item The left screenshot uses a type provider for data aggregation \cite{gamma}.
+  Type provider are treated as an external library
+  (with objects, members and reduction relation). Type providers also rely on type information to
   provide editor auto-complete, which we support by implementing type checking over a
-  dependency graph as discussed in Section~\ref{sec:extra-types}.
+  dependency graph (Section~\ref{sec:extra-types}).
 
 \item When displaying live preview for code written using the data aggregation type provider (left),
-  our environment also generates tabes that show individual steps of the data transformation.
-  A tab is selected based on the cursor position and shows preview for the current sub-expression
-  (e.g.~raw data before grouping in the screenshot).
-
-\item Another external library provides support for charting (right). The environment does not
-  handle charts in a special way and displays the result of evaluating the whole command. The
-  screenshot shows a case where the user modifies parameters of the chart. Thanks to our live evaluation
+  our environment generates tabes that show individual steps of the data transformation.
+  A tab is selected based on the cursor position and shows a preview for the current sub-expression
+  (e.g.~raw data before grouping).
+
+\item Another external library provides support for charting (right). Here, the environment
+  displays the result of evaluating the whole command. The screenshot shows a case where the user
+  modifies parameters of the chart. Thanks to our live evaluation
   strategy, this is done efficiently without reevaluating the data transformation.
@@
 \noindent
+The environment is available at \url{http://gallery.thegamma.net}.
 A user study to evaluate the usability of the system from a human-computer interaction perspective
-is left for future work. As an anecdotal evidence of the usability, the code in
-Figure~\ref{fig:gallery} was developed by an attendee of a tutorial at Mozilla Festival 2017
+is left for future work. As an anecdotal evidence, the code in
+Figure~\ref{fig:gallery} was developed by an attendee of a Mozilla Festival 2017
 who had no prior programming experience.
@@
 We aim to make creating transparent data analyses, such as those done by journalists, easier and we
-present a tool that provides instant feedback when writing code in a text editor. Many our
-design decisions have alternatives that have been explored in a rich body of literature.
-
-\subsubsection{Notebooks and data science tools}
+present a tool that provides instant feedback when writing code in a text editor.
+There is a rich body of work exploring complementary designs.
+
+\paragraph{Notebooks and data science tools.}
 Basic feedback mechanism, the REPL (read-eval-print-loop), originated in Lisp
-\cite{lisp} and is now common in many languages \cite{drscheme}. Notebook systems
-such as Jupyter and R markdown \cite{rmarkdown,jupyter,ipython} structure code in cells,
-but still rely on explicitly executing individual cells using a stateful REPL environment.
-Integrating provenance tracking with notebooks \cite{dataflow,noworkflow,wrattler} addresses
-the reproducibility issues of this model, but keeps cells as units of work. Ideas such as
-dependency tracking and efficient recomputation exist in many visual data exploration tools
-\cite{control,tableau,vizdom} and scientific workflow systems \cite{taverna,kepler}. Some of the
+\cite{lisp} and has become standard \cite{drscheme}. Notebook systems
+such as Jupyter \cite{rmarkdown,jupyter,ipython} structure code in cells,
+but still rely on executing individual cells using a stateful REPL.
+Adding provenance tracking \cite{dataflow,noworkflow,wrattler} addresses
+the reproducibility issues of notebooks, but keeps cells as units of work. Ideas such as
+dependency tracking and efficient recomputation also exist in visual data exploration tools
+\cite{control,tableau,vizdom} and scientific workflow systems \cite{taverna,kepler}. The
 implementation techniques are related to our work, but we focus on text-based scripts.
 Tempe \cite{tempe} focuses on streaming data, but provides a text-based scripting environment
-with automatic code update and \cite{ripple} empirically evaluate its usability in contrast to
-REPLs. The challenges faced by Tempe are more akin to dynamic software update \cite{dsu}.
-
-\subsubsection{Live coding and editing}
-Live coding based on textual programs has been popularised by \cite{principle,learnable} and
-is actively developed in domains such as live coded music \cite{beyond,sonic}. \cite{liveroad}
-reviews the use of live coding in a Smalltalk derived environment. Lighttable by \cite{lighttable}
+with automatic code update; its usability in contrast to REPLs has been empirically
+evaluated \cite{ripple}. The challenges faced by Tempe are more akin to dynamic software update \cite{dsu}.
+
+\paragraph{Live coding and editing.}
+Live coding based on textual programs has been popularised by Victor \cite{principle,learnable} and
+is actively developed in domains such as live coded music \cite{beyond,sonic}. Kubelka et al.~\cite{liveroad}
+review the use of live coding in a Smalltalk derived environment. Lighttable \cite{lighttable}
 and Chrome DevTools provide limited live previews akin to those presented in this paper,
@@
 we also plan to explore how direct manipulation \cite{direct} could be integrated with our preview
-mechanism. \cite{liveroad}
-
+mechanism.
+%
 Using structured editors \cite{structure-based} rather than text editors is a promising alternative
@@
 lambda calculus \cite{hazelnut}, but it also makes us consider more human-centric abstractions
-\cite{subtext,directprog} along the lines of the one outlined in Section~\ref{sec:extra-abstraction}.
-
-\subsubsection{Incremental computation and dependency analysis}
+\cite{subtext,directprog} further discussed in Section~\ref{sec:extra-abstraction}.
+
+\paragraph{Incremental computation and dependency analysis.}
 Work on self-adjusting and incremental computation \cite{selfadjusting,adapton} handles
@@
 
-\subsubsection{Semantics and partial evaluation}
+\paragraph{Semantics and partial evaluation.}
 The evaluation of previews is a form of partial evaluation \cite{partial}, done in a
@@
 %
-% \section{TODO}
-%
-% TBD: Define live preview somewhere
-%
-% TBD: In section about dependency graphs, include a subsection discussing how this relates
-% to incremental computation (just like "Semantics of delayed previews" links that to CMTT etc.)
-%
-%
-% \cite{selfadjusting}
-% \cite{incremental}
-%
-% Incremental computation
-%
-%   Self-adjusting computation
-%   http://www.cs.cmu.edu/~rwh/theses/acar.pdf
-%
-%   Incremental Computation with Names
-%   https://www.cs.tufts.edu/~jfoster/papers/oopsla15.pdf
-%   -> dsl interpreter
-%
-%   (and other adapton stuff http://adapton.org/)
-%
-% Type checking
-%
-%   co-contextual
-%   http://www.informatik.uni-marburg.de/~seba/publications/cocontextual-type-checking.pdf
-%   http://drops.dagstuhl.de/opus/volltexte/2017/7262/pdf/LIPIcs-ECOOP-2017-18.pdf
-%
-% https://dl.acm.org/citation.cfm?id=664109
-%
 \section{Summary}
-One of the reasons that make spreadsheets easier to use than programming tools is that they
+One of the aspects that make spreadsheets easier to use than programming tools is that they
 provide instant feedback. We aim to make programming tools as instant as spreadsheets.
-We formalized a number of key aspects of simple data analyses such as those don by journalists
+We described a number of key aspects of simple data analyses such as those done by journalists
 and then used our observations to build both theory and simple practical data analytics tools.
@@
 for data exploration. The calculus captures key observations about simple data analyses. They
-rely on logic from external libraries, implement few abstractions and are written as lists of
-commands. We later revised the calculus to add type checking and simple abstraction mechanism.
+rely on logic defined by external libraries, implement few abstractions and are written as lists of
+commands.
 
@@
 in unconstrained way in an text editor, which makes this particularly challenging. The key trick
-is to separate the process into fast \emph{binding phase}, which constructs a dependency graph
-and slower \emph{evaluation phase} that can cache results. This makes it possible to quickly parse
+is to separate the process into a fast \emph{binding phase}, which constructs a dependency graph
+and a slower \emph{evaluation phase} that can cache results. This makes it possible to quickly parse
 updated code, reconstruct dependency graph and compute preview using previous, partially
@@
 
-\section{Appendix}
-
-\subsection{Normalization}
+\section{Details of proofs}
+
+\subsection{Normalization for data exploration calculus}
 \label{sec:app-normalization}
@@
 \begin{theorem}[Normalization]
-For all $p$, there exists $n$ and $o_1, \ldots, o_n$ such that $p\rightsquigarrow^{*} o_1;\ldots;o_n$.
+\raggedrigth
+For all $p$, there exists $n$ and $o_1, \ldots, o_n$ \\
+such that $p\rightsquigarrow^{*} o_1;\ldots;o_n$.
 \end{theorem}
@@
 
-\subsection{Term preview correctness}
-\label{sec:app-correctness}
-
-\begin{theorem}[Term preview correctness]
-Given a term $t$ that has no free variables, together with a lookup table $\Delta$ obtained
-from any sequence of programs using $\ident{bind-prog}$ (Figure~\ref{fig:binding-rules-prog}) and
-$\ident{update}$ (Figure~\ref{fig:loop}), then
-let $\bndclr{v}, (V, E) = \ident{bind-expr}_{\emptyset, \Delta}(t)$. If $\bndclr{v}\Downarrow p$
-over a graph $(V, E)$ then $p = o$ for some value $o$ and $t \rightsquigarrow^{*} o$.
-\end{theorem}
-
-\begin{proof}
-First note that, when combining recursively constructed sub-graphs, the \ident{bind-expr} function
-adds new nodes and edges leading from those new nodes. Therefore, an evaluation using~$\Downarrow$
-over a sub-graph will also be valid over the new graph -- the newly added nodes and edges do not introduce
-non-determinism to the rules given in Figure~\ref{fig:eval}.
-
-We prove a more general property showing that for any $e$, its binding
-$\bndclr{v}, (V, E) = \ident{bind-expr}_{\emptyset, \Delta}(e)$ and any evaluation context $C$
-such that $C[e]\rightsquigarrow o$ for some $o$, one of the following holds:
-%
-\begin{enumerate}
-\item[a.] If $FV(e)\,=\,\emptyset$ then $\bndclr{v} \Downarrow p$ for some $p$ and $C[p] \rightsquigarrow o$
-\item[b.] If $FV(e)\neq\emptyset$ then $\bndclr{v} \Downarrow \llbracket e_p \rrbracket_{FV(e)}$ for some $e_p$ and $C[e_p] \rightsquigarrow o$
-\end{enumerate}
-%
-In the first case, $p$ is a value, but it is not always the case that $e \rightsquigarrow^{*} p$,
-because $p$ may be lambda function and preview evaluation may reduce sub-expression in the body of
-the function. Using a context $C$ in which the value reduces to an object avoids this problem.
-
-The proof of the theorem follows from the more general property. Using a context $C[-]=-$, the
-term $t$ reduces $t \rightsquigarrow^{*}t' \rightsquigarrow_\epsilon o$ for some $o$ and the
-preview $p$ is a value $o$ because $C[p] = p = o$.
-The proof is by induction over the binding process, which follows the structure of the expression:
-
-\vspace{0.75em}\noindent(1) $\ident{bind-expr}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n))$ --
-  Here $e = e_0.m(e_1, \ldots, e_n)$, $\bndclr{v_i}$ are graph nodes obtained by induction for
-  expressions $e_i$ and $\{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v}, \bndclr{v_n}, \blbl{arg}(n))\} \subseteq E$.
-  From lookup inversion Lemma~\ref{thm:lemma-lookup}, $\bndclr{v} = \bnd{mem}(m, s)$ for some $s$.
-
-  If $FV(e)=\emptyset$, then $\bndclr{v_i} \Downarrow p_i$ for $i\in 0\ldots n$ and
-  $\bndclr{v}\Downarrow p$ using \rname{mem-val} such that $p_0.m(p_1, \ldots, p_n) \rightsquigarrow p$.
-  From induction hypothesis and \emph{compositionality} of external libraries (Definition~\ref{def:external}),
-  it holds that for any $C$ such that $C[e_0.m(e_1, \ldots, e_n)] \rightsquigarrow o$ for some $o$
-  then also $C[p_0.m(p_1, \ldots, p_n)] \rightsquigarrow C[p] \rightsquigarrow o$.
-
-  If $FV(e)\neq\emptyset$, then $\bndclr{v_i} \Downarrow_\ident{lift} \llbracket e'_i \rrbracket$ for $i\in 0\ldots n$ and
-  $\bndclr{v}\Downarrow \llbracket e'_0.m(e'_1, \ldots, e'_n) \rrbracket_{FV(e)}$ using \rname{mem-expr}.
-  From induction hypothesis and \emph{compositionality} of external libraries (Definition~\ref{def:external}),
-  it holds that for any $C$ such that $C[e_0.m(e_1, \ldots, e_n)] \rightsquigarrow o$ for some $o$
-  then also $C[e'_0.m(e'_1, \ldots, e'_n)] \rightsquigarrow o$.
-
-\vspace{0.75em}\noindent(2) $\ident{bind-expr}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n))$ --
-  This case is similar to (1), except that the fact that $\bndclr{v} = \bnd{mem}(m, s)$
-  holds by construction, rather than using Lemma~\ref{thm:lemma-lookup}.
-
-\vspace{0.75em}\noindent(3) $\ident{bind-expr}_{\Gamma, \Delta}(\lambda x\rightarrow e_b)$ --
-  Here $e = \lambda x\rightarrow e_b$, $\bndclr{v_b}$ is the graph node obtained by induction
-  for the expression $e_b$ and $(\bndclr{v}, \bndclr{v_b}, \blbl{body}) \in E$.
-  From lookup inversion Lemma~\ref{thm:lemma-lookup}, $\bndclr{v} = \bnd{fun}(x, s)$ for some $s$.
-  The evaluation can use one of three rules:
-
-  \begin{enumerate}
-  \item[i.]
-  If $FV(e)=\emptyset$ then $\bndclr{v_b}\Downarrow p_b$ for some $p_b$ and $\bndclr{v}\Downarrow \lambda x\rightarrow p_b$
-  using \rname{fun-val}. Let $e'_b=p_b$.
-
-  \item[ii.]
-  If $FV(e_b)=\{x\}$ then $\bndclr{v_b}\Downarrow \llbracket e'_b \rrbracket_x$ for some $e'_b$ and
-  $\bndclr{v}\Downarrow \lambda x\rightarrow e'_b$ using \rname{fun-bind}.
-
-  \item[iii.]
-  Otherwise, $\bndclr{v_b}\Downarrow \llbracket e'_b \rrbracket_{x,\Gamma}$ for some $e'_b$ and
-  $\bndclr{v}\Downarrow \llbracket\lambda x\rightarrow e'_b\rrbracket_\Gamma$ using \rname{fun-expr}.
-  \end{enumerate}
-
-\noindent
-For i.) and ii.) we show that a.) is the case; for iii.) we show that b.) is the case; that is
-for any $C$, if $C[\lambda x\rightarrow e_b] \rightsquigarrow o$ then also
-$C[\lambda x\rightarrow e'_b] \rightsquigarrow o$. For a given $C$, let $C'[-] = C[\lambda x\rightarrow -]$
-and use the induction hypothesis, i.e.~if $C'[e_b] \rightsquigarrow o$ for some $o$ then also $C'[e'_b] \rightsquigarrow o$.
-
-\vspace{0.75em}\noindent(4) $\ident{bind-expr}_{\Gamma, \Delta}(\lambda x\rightarrow e)$  --
-  This case is similar to (3), except that the fact that $\bndclr{v} = \bnd{fun}(x, s)$
-  holds by construction, rather than using Lemma~\ref{thm:lemma-lookup}.
-
-\vspace{0.75em}\noindent(5) $\ident{bind-expr}_{\Gamma, \Delta}(o)$ -- In this case $e=o$ and $\bndclr{v} = \bnd{val}(o)$
-  and $\bnd{val}(o)\Downarrow o$ using \rname{val} and so the case a.) trivially holds.
-
-\vspace{0.75em}\noindent(6) $\ident{bind-expr}_{\Gamma, \Delta}(x)$ -- The initial $\Gamma$ is empty,
-  so $x$ must have been added to $\Gamma$ by case (3) or (4). Hence,
-  $\bndclr{v}=\bnd{var}(x)$, $\bndclr{v}\Downarrow \llbracket x \rrbracket_{x}$ using \rname{var}
-  and so $e_p = e = x$ and the case b.) trivially holds.
-
-\end{proof}
-
-% ==================================================================================================
-
-\section{Theories of delayed previews}
-\label{sec:app-theories}
-
-The operational semantics presented in this paper serves two purposes. It gives a simple guide
-for implementing text-based live coding environments for data science and we use it to prove that
-our optimized way of producing live previews is correct. However, some aspects of our mechanism
-are related to important work in semantics of programming languages and deserve to be mentioned.
-
-The construction of delayed previews is related to meta-programming. Assuming
-we have delayed previews $\llbracket e_0 \rrbracket_x$ and $\llbracket e_1 \rrbracket_y$ and
-we invoke a member $m$ on $e_0$ using $e_1$ as an argument. To do this, we construct a new
-delayed preview $\llbracket e_0.m(e_1) \rrbracket_{x, y}$. This operation is akin to expression
-splicing from meta-programming \cite{metaml,quotations}.
-
-The semantics of delayed previews can be more formally captured by Contextual Modal Type Theory
-(CMTT) \cite{cmtt} and comonads \cite{cmtt-denotation}. In CMTT, $\lbrack \Psi \rbrack A$ denotes
-that a proposition $A$ is valid in context $\Psi$, which is similar to our delayed previews written
-as $\llbracket A \rrbracket_\Psi$. CMTT defines rules for composing context-dependent propositions
-that would allow us to express the splicing operation used in \rname{mem-expr}. In categorical
-terms, the context-dependent proposition can be modelled as a graded comonad \cite{effectrev,graded}.
-The evaluation of a preview with no context dependencies (built implicitly into our evaluation rules)
-corresponds to the counit operation of a comonad and would be explicitly written as
-$\llbracket A \rrbracket_\emptyset \rightarrow A$.
-
-% ==================================================================================================
-
-\section{Proofs for section 3}
+\subsection{Let elimmination for a program}
 \label{sec:let-lang-elimination}
@@
 \end{proof}
-
-\section{Correctness of previews}
-\label{sec:app-correctness}
 
@@
 \subsection{Term preview correctness}
-\label{sec:app-let-free-correct}
+\label{sec:app-correctness}
 
 \begin{theorem}[Term preview correctness]
-\label{thm:app-let-free-correct}
 Given a term $t$ that has no free variables, together with a lookup table $\Delta$ obtained
@@
 $\ident{update}$ (Figure~\ref{fig:loop}), then
-let $\bndclr{v}, (V, E) = \ident{bind-expr}_{\emptyset, \Delta}(t)$.
-
-\vspace{0.25em}
-\noindent
-If $\bndclr{v}\Downarrow p$
+let $\bndclr{v}, (V, E) = \ident{bind-expr}_{\emptyset, \Delta}(t)$. If $\bndclr{v}\Downarrow p$
 over a graph $(V, E)$ then $p = o$ for some value $o$ and $t \rightsquigarrow^{*} o$.
@@
 \begin{proof}
-First note that, when combining recursively constructed sub-graphs, the \ident{bind-expr} function
+When combining recursively constructed sub-graphs, the \ident{bind-expr} function
 adds new nodes and edges leading from those new nodes. Therefore, an evaluation using~$\Downarrow$
@@
 \item[a.] If $FV(e)\,=\,\emptyset$ then $\bndclr{v} \Downarrow p$ for some $p$ and $C[p] \rightsquigarrow o$
-\item[b.] If $FV(e)\,\neq\,\emptyset$ then $\bndclr{v} \Downarrow \llbracket e_p \rrbracket_{FV(e)}$ for some $e_p$ and $C[e_p] \rightsquigarrow o$
+\item[b.] If $FV(e)\neq\emptyset$ then $\bndclr{v} \Downarrow \llbracket e_p \rrbracket_{FV(e)}$ for some $e_p$ and $C[e_p] \rightsquigarrow o$
 \end{enumerate}
@@
 preview $p$ is a value $o$ because $C[p] = p = o$.
-The proof is by induction over the binding process, which follows the structure of the expression.
-The full proof is shown in Appendix~\ref{sec:app-correctness}.
+The proof is by induction over the binding process, which follows the structure of the expression:
+
+\vspace{0.75em}\noindent(1) $\ident{bind-expr}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n))$ --
+  Here $e = e_0.m(e_1, \ldots, e_n)$, $\bndclr{v_i}$ are graph nodes obtained by induction for
+  expressions $e_i$ and $\{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v}, \bndclr{v_n}, \blbl{arg}(n))\} \subseteq E$.
+  From lookup inversion Lemma~\ref{thm:lemma-lookup}, $\bndclr{v} = \bnd{mem}(m, s)$ for some $s$.
+
+  If $FV(e)=\emptyset$, then $\bndclr{v_i} \Downarrow p_i$ for $i\in 0\ldots n$ and
+  $\bndclr{v}\Downarrow p$ using \rname{mem-val} such that $p_0.m(p_1, \ldots, p_n) \rightsquigarrow p$.
+  From induction hypothesis and \emph{compositionality} of external libraries (Definition~\ref{def:external}),
+  it holds that for any $C$ such that $C[e_0.m(e_1, \ldots, e_n)] \rightsquigarrow o$ for some $o$
+  then also $C[p_0.m(p_1, \ldots, p_n)] \rightsquigarrow C[p] \rightsquigarrow o$.
+
+  If $FV(e)\neq\emptyset$, then $\bndclr{v_i} \Downarrow_\ident{lift} \llbracket e'_i \rrbracket$ for $i\in 0\ldots n$ and
+  $\bndclr{v}\Downarrow \llbracket e'_0.m(e'_1, \ldots, e'_n) \rrbracket_{FV(e)}$ using \rname{mem-expr}.
+  From induction hypothesis and \emph{compositionality} of external libraries (Definition~\ref{def:external}),
+  it holds that for any $C$ such that $C[e_0.m(e_1, \ldots, e_n)] \rightsquigarrow o$ for some $o$
+  then also $C[e'_0.m(e'_1, \ldots, e'_n)] \rightsquigarrow o$.
+
+\vspace{0.75em}\noindent(2) $\ident{bind-expr}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n))$ --
+  This case is similar to (1), except that the fact that $\bndclr{v} = \bnd{mem}(m, s)$
+  holds by construction, rather than using Lemma~\ref{thm:lemma-lookup}.
+
+\vspace{0.75em}\noindent(3) $\ident{bind-expr}_{\Gamma, \Delta}(\lambda x\rightarrow e_b)$ --
+  Here $e = \lambda x\rightarrow e_b$, $\bndclr{v_b}$ is the graph node obtained by induction
+  for the expression $e_b$ and $(\bndclr{v}, \bndclr{v_b}, \blbl{body}) \in E$.
+  From lookup inversion Lemma~\ref{thm:lemma-lookup}, $\bndclr{v} = \bnd{fun}(x, s)$ for some $s$.
+  The evaluation can use one of three rules:
+
+  \begin{enumerate}
+  \item[i.]
+  If $FV(e)=\emptyset$ then $\bndclr{v_b}\Downarrow p_b$ for some $p_b$ and $\bndclr{v}\Downarrow \lambda x\rightarrow p_b$
+  using \rname{fun-val}. Let $e'_b=p_b$.
+
+  \item[ii.]
+  If $FV(e_b)=\{x\}$ then $\bndclr{v_b}\Downarrow \llbracket e'_b \rrbracket_x$ for some $e'_b$ and
+  $\bndclr{v}\Downarrow \lambda x\rightarrow e'_b$ using \rname{fun-bind}.
+
+  \item[iii.]
+  Otherwise, $\bndclr{v_b}\Downarrow \llbracket e'_b \rrbracket_{x,\Gamma}$ for some $e'_b$ and
+  $\bndclr{v}\Downarrow \llbracket\lambda x\rightarrow e'_b\rrbracket_\Gamma$ using \rname{fun-expr}.
+  \end{enumerate}
+
+\noindent
+For i.) and ii.) we show that a.) is the case; for iii.) we show that b.) is the case; that is
+for any $C$, if $C[\lambda x\rightarrow e_b] \rightsquigarrow o$ then also
+$C[\lambda x\rightarrow e'_b] \rightsquigarrow o$. For a given $C$, let $C'[-] = C[\lambda x\rightarrow -]$
+and use the induction hypothesis, i.e.~if $C'[e_b] \rightsquigarrow o$ for some $o$ then also $C'[e'_b] \rightsquigarrow o$.
+
+\vspace{0.75em}\noindent(4) $\ident{bind-expr}_{\Gamma, \Delta}(\lambda x\rightarrow e)$  --
+  This case is similar to (3), except that the fact that $\bndclr{v} = \bnd{fun}(x, s)$
+  holds by construction, rather than using Lemma~\ref{thm:lemma-lookup}.
+
+\vspace{0.75em}\noindent(5) $\ident{bind-expr}_{\Gamma, \Delta}(o)$ -- In this case $e=o$ and $\bndclr{v} = \bnd{val}(o)$
+  and $\bnd{val}(o)\Downarrow o$ using \rname{val} and so the case a.) trivially holds.
+
+\vspace{0.75em}\noindent(6) $\ident{bind-expr}_{\Gamma, \Delta}(x)$ -- The initial $\Gamma$ is empty,
+  so $x$ must have been added to $\Gamma$ by case (3) or (4). Hence,
+  $\bndclr{v}=\bnd{var}(x)$, $\bndclr{v}\Downarrow \llbracket x \rrbracket_{x}$ using \rname{var}
+  and so $e_p = e = x$ and the case b.) trivially holds.
+
 \end{proof}
 
+\subsection{Binding sub-expressions}
+\label{sec:sub-expr}
+
+\begin{lem}[Binding sub-expressions]
+\label{thm:sub-expr}
+Assume we have programs $p_1 = c_1; \ldots; c_k; K_c[e]; c_{k+1};\ldots; c_n$ and
+$p_2 = c'_1; \ldots; c'_k; K'_c[e]; c'_{k+1};\ldots;c'_n$ and $I\subseteq \{1\ldots k\}$ such
+that $\forall i\!\in\!I.\,c_i\!=\!c'_i$ and for each $x \in \bigcup_{i\in I}FV(c_i) \cup FV(e)$
+there exists $j\in I$ such that $c_j = \kvd{let}\,x=e$ for some $e$.
+%
+Given any $\Delta$, assume that the the first program is bound,
+i.e.~$\bndclr{v_1};\ldots;\bndclr{v_n}, (V, E) = \ident{bind-prog}_{\emptyset, \Delta}(p_1)$,
+the cache is updated $\Delta' = \ident{update}_{V, E}(\Delta)$ and the second
+program is bound,
+i.e.~$\bndclr{v'_1};\ldots;\bndclr{v'_n}, (V', E') = \ident{bind-prog}_{\emptyset, \Delta'}(p_2)$.
+
+Now, assume $\bndclr{v}, G = \ident{bind-expr}_{\Gamma, \Delta}(e)$ and
+$\bndclr{v'}, G' = \ident{bind-expr}_{\Gamma', \Delta'}(e)$ are the recursive calls to bind
+$e$ during the first and the second binding, respectively. Then, the graph nodes assigned to the
+sub-expression $e$ are the same, i.e.~$\bndclr{v} = \bndclr{v'}$.
+\end{lem}
+\begin{proof}
+First, assuming that $\forall x\in FV(e). \Gamma(x) = \Gamma'(x)$, we show by induction over the binding process of $e$
+for the first program that the result is the same. In cases (1) and (3), the updated $\Delta'$
+contains the required key and so the second binding proceeds using the same case. In cases
+(2) and (4), the second binding reuses the node created by the first binding using case (1) and
+(3), respectively. Cases (5) and (6) are the same.
+
+Second, when binding let bindings in $c_1; \ldots; c_k$, the initial $\Gamma = \emptyset$ during
+both bindings. Nodes added to $\Gamma$ and $\Gamma'$ for commands $c_j$ such that $j\in I$ are
+the same and nodes added for remaining commands do not add any new nodes referenced from $e$ and
+so $\bndclr{v} = \bndclr{v'}$ using the above.
+\end{proof}
+
 % ==================================================================================================
 
-\section{Extending data exploration calculus}
-\label{sec:extra}
-
-The data exploration calculus presented in Section~\ref{sec:calculus} allowed us to formalise the
-live preview mechanism, but it lacks many important features. We outline two extensions.
-First, we discuss how to add instant type checking by using the dependency graph construction
-described in Section~\ref{sec:formal}. Second, we consider an abstraction mechanism that preserves
-the live coding experience.
-
-% --------------------------------------------------------------------------------------------------
+\section{Theories of delayed previews}
+\label{sec:app-theories}
+
+The operational semantics presented in this paper serves two purposes. It gives a simple guide
+for implementing text-based live coding environments for data science and we use it to prove that
+our optimized way of producing live previews is correct. However, some aspects of our mechanism
+are related to important work in semantics of programming languages and deserve to be mentioned.
+
+The construction of delayed previews is related to meta-programming. Assuming
+we have delayed previews $\llbracket e_0 \rrbracket_x$ and $\llbracket e_1 \rrbracket_y$ and
+we invoke a member $m$ on $e_0$ using $e_1$ as an argument. To do this, we construct a new
+delayed preview $\llbracket e_0.m(e_1) \rrbracket_{x, y}$. This operation is akin to expression
+splicing from meta-programming \cite{metaml,quotations}.
+
+The semantics of delayed previews can be more formally captured by Contextual Modal Type Theory
+(CMTT) \cite{cmtt} and comonads \cite{cmtt-denotation}. In CMTT, $\lbrack \Psi \rbrack A$ denotes
+that a proposition $A$ is valid in context $\Psi$, which is similar to our delayed previews written
+as $\llbracket A \rrbracket_\Psi$. CMTT defines rules for composing context-dependent propositions
+that would allow us to express the splicing operation used in \rname{mem-expr}. In categorical
+terms, the context-dependent proposition can be modelled as a graded comonad \cite{effectrev,graded}.
+The evaluation of a preview with no context dependencies (built implicitly into our evaluation rules)
+corresponds to the counit operation of a comonad and would be explicitly written as
+$\llbracket A \rrbracket_\emptyset \rightarrow A$.
+
+% ==================================================================================================
 
@@
 
-\subsection{Type checking}
+\section{Type checking}
 \label{sec:extra-types}
 
-Instant live previews give analysts quick feedback when they write incorrect code, but
+Live previews give analysts quick feedback when they write incorrect code, but
 having type information is still valuable. First, it can help give better error messages. Second,
 types can be used to provide auto-complete -- when the user types `.' we can offer available
-members without having to evaluate the wait until the value of the object is available.
-
-\subsubsection{Revised dependency graph}
+members without having to wait until the value of the object is available.
+
+\paragraph{Revised dependency graph.}
 Type checking of small programs is typically fast enough that no caching is necessary. However,
@@
 
-\subsubsection{Type checking}
+\paragraph{Type checking.}
 The structure of typing rules is similar to the evaluation relation $\bndclr{v}\Downarrow d$
@@
 
-\begin{figure}[b]
-\vspace{-0.5em}
-\begin{equation*}
-\hspace{-3.5em}
+\begin{figure}
+\begin{equation*}
+\hspace{-1em}
 \begin{array}{l}
-\inference[\rname{mem}~]
-  { \forall i\in\{0\ldots k\}.(\bnd{mem}(m, s), \bndclr{v_i}, \blbl{arg}(i)) \in E &
+\hspace{7em}\inference[\rname{var}~]
+  { (\bnd{var}(x,s), \bndclr{v}, \blbl{callsite}(m, i)) \in E \\
+    \bndclr{v} \vdash \{.., m:(\tau_1, \ldots, \tau_k)\rightarrow \tau,..\} & \tau_i=\tau'\rightarrow\tau'' }
+  { \bnd{var}(x, s) \vdash \tau' }
+\\[1.5em]
+\hspace{7em}\inference[\rname{mem}~]
+  { \forall i\in\{0\ldots k\}.(\bnd{mem}(m, s), \bndclr{v_i}, \blbl{arg}(i)) \in E \\
   \bndclr{v_0} \vdash \{.., m:(\tau_1, \ldots, \tau_k) \rightarrow \tau, ..\} & \bndclr{v_i} \vdash \tau_i }
@@
 \\[1.5em]
-\inference[\rname{var}~]
-  { (\bnd{var}(x,s), \bndclr{v}, \blbl{callsite}(m, i)) \in E &
-    \bndclr{v} \vdash \{.., m:(\tau_1, \ldots, \tau_k)\rightarrow \tau,..\} & \tau_i=\tau'\rightarrow\tau'' }
-  { \bnd{var}(x, s) \vdash \tau' }
-\\[1.5em]
-\inference[\rname{fun}~]
+\hspace{4em}\inference[\rname{fun}~]
   { \{(\bnd{fun}(x,s), \bndclr{v_b}, \blbl{body}), (\bnd{var}(x,s), \bndclr{v_c}, \blbl{callsite}(m, i))\} \subseteq E \\
@@
 
-\subsection{Feedback-friendly abstraction}
+\section{Feedback-friendly abstraction}
 \label{sec:extra-abstraction}
@@
 
+\newpage
+\section{Manuscript version history (to be removed for publication)}
+\label{sec:revisions}
+
+This note briefly reviews the revisions made to this manuscript in response to reviews for an
+earlier version of the paper.
+
+\paragraph{Unclear motivation.}
+An earlier review pointed out that it was not clear who is the target user for the described
+system. The newly added Section~\ref{sec:background-jupyter} provides the answer. Our focus
+is on non-expert programmers conducting simple data exploration such as journalists working
+with government data. This is perhaps an unusual focus for a programming language research, but
+we believe it is increasingly important.
+
+\paragraph{Unclear scope of the work.}
+An earlier version of the paper did not make it clear what is prior work and
+what is new in this paper. The paper has been restructured to make this clear --
+we introduce the data exploration calculus (Section~\ref{sec:calculus}) and a
+theory for building a live coding environment for the calculus (Section~\ref{sec:formal})
+with live preview support (Section~\ref{sec:previews}). This is accompanied by an implementation
+that extends an editor for The Gamma, described earlier \cite{gamma} -- we add live preview
+support and auto-complete using type checking based on the dependency graph (Section~\ref{sec:extra-types}).
+
+\paragraph{Simplicity of the language.}
+An earlier review argued that the language used in the paper is overly simplistic.
+The newly added Section~\ref{sec:background-jupyter} explains the motivation behind the
+data exploration calculus. The simplicity of the language reflects the use cases that
+motivated this paper. The newly added case study (Section~\ref{sec:evaluation-case}) shows that
+the language is sufficient for building a simple, but realistic and practically useful
+data exploration environment. We also discuss extensions to the langauge in
+Section~\ref{sec:extra-abstraction}.
+
+\paragraph{Missing empirical evaluation.}
+An earlier version of the paper lacked empirical evaluation. It is not our aim to measure
+the exact overhead of our implementation (which is typically negligible with respect to the
+cost of evaluating the data transformations), but the newly added empirical evaluation
+(Section~\ref{sec:evaluation-empirical}) shows the effect of our evaluation strategy during
+a reprsentative coding task.
+
+\paragraph{Missing related work.}
+An earlier version of the paper lacked discussion of important related work, in particular that
+on incremental computation. This has now been addressed -- incremental computation focuses on
+the case where data change, while our focus is on cases where code changes, but the implementation
+techniques are closely related.
+
 \end{document}

### 2020-02-13 wip-live-programming/paper.tex
Final version edits

@@
-\documentclass[english,submission]{programming}
+\documentclass[english,crc]{programming}
 \usepackage[backend=biber]{biblatex}
@@
 \newcounter{dfc}
-
 \theoremstyle{plain}
@@
 \begin{document}
-\title{A live coding environment for a data exploration language}
+\title{Foundations of a live data exploration environment}
 %\titlerunning{Learning to live with errors}
-
-\author[a]{Tomas Petricek}
+\author[a,b]{Tomas Petricek}
 \authorinfo{
@@
 easier and more accessible. His research interests span a range of areas including theory of
-programming languages, history and philosophy of programming, as well as open data and data
-journalism.
+programming languages, tools for data science as well as history and philosophy of programming.
 %
@@
 \affiliation[a]{University of Kent, UK}
-\keywords{Data exploration, live coding, data journalism, instant previews}
+\affiliation[b]{and The Alan Turing Institute, UK}
+\keywords{Data exploration, live programming, data journalism, instant previews}
 \paperdetails{
@@
 }
-
 % https://dl.acm.org/ccs/ccs.cfm and generate your Classification
@@
 \end{CCSXML}
-
 \ccsdesc[300]{Human-centered computing~Interactive systems and tools}
@@
 \maketitle
-
-\vspace{1em}
-\noindent
-{\scriptsize
-~\\
-\textbf{Note to reviewers:} This paper has been previously submitted to a different
-venue. The revisions made in response to past reviews are summarized in Appendix~\ref{sec:revisions}
-(to be removed for publication).}
-\vspace{2em}
 
@@
 %
-A live coding environment computes and displays the list of years. Suppose that the programmer then
+A live programming environment computes and displays the list of years. Suppose that the programmer then
 makes $\num{10}$ a variable and changes the date format:
@@
 %
-Ideally, the live coding environment should understand the change, reuse a cached result of the
+Ideally, the live programming environment should understand the change, reuse a cached result of the
 first two transformations (sorting and taking the first 10 elements) and only evaluate the
@@
 
-\subsection{Related work}
-Simple data exploration performed, for example, by journalists \cite{ddj} is done either
-programmatically or using spreadsheets. The latter is easy but error-prone while the former
-requires expert programming skills. We aim to bring liveness of spreadsheets to programmatic
-data exploration. This requires extending recomputation as done in spreadsheets \cite{spreadsheet}
-to code written in an ordinary text editor.
-
-Many visual data exploration tools support interactivity \cite{control,tableau,vizdom} and some
-can even export the workflow as a script \cite{wrangler}. Data analysts who prefer working with
-code typically resort to notebook systems such as Jupyter or R Markdown \cite{jupyter,rmarkdown}.
-Those are text-based, but have a limited model of recomputation. Users structure code in
-cells and manually reevaluate cells. Many notebook systems do not track dependencies
-between cells which can lead to well-documented inconsistencies \cite{dataflow,noworkflow,wrattler}.
-
-Existing text-based editors that provide instant feedback,
-such as Lighttable \cite{lighttable} and Chrome DevTools, often work only in certain
-situations or require full recomputation. A more principled approach can be used by
-systems based on structured editing \cite{livenut,lamdu} where code is only
-modified via known operations with known effect on the computation graph
-(e.g.~``extract variable'' has no effect on the result; ``change constant value''
-forces recomputation of subsequent code). We aim to cater for the many users who
-prefer to edit programs as free-form text.
-%
-The technical aspects of our implementation are related to incremental computation \cite{selfadjusting,incremental}.
-This focuses on cases where \emph{data changes} but program stays the same, while our focus is on cases where
-\emph{program changes} but data stays the same.
-
-\subsection{Contributions}
-We present the design and implementation of a live coding environment for a simple data exploration
+\subsection*{Contributions}
+We present the design and implementation of a live programming environment for a simple data exploration
 language that provides correct and efficient instant feedback, yet is integrated into an
@@
 
-\item A live coding environment does not operate in batch mode and so we cannot follow classic
+\item A live programming environment does not operate in batch mode and so we cannot follow classic
   compiler literature. We capture the essence of such new perspective (Section~\ref{sec:formal})
-  and use it to build our \emph{live preview} mechanism (Section~\ref{sec:previews}) that evaluates
+  and use it to build our \emph{instant preview} mechanism (Section~\ref{sec:previews}) that evaluates
   code instantly during editing.
 
-\item We prove that our live preview mechanism is correct (Section~\ref{sec:evaluation-correctness}) and that
+\item We prove that our instant preview mechanism is correct (Section~\ref{sec:evaluation-correctness}) and that
   it reuses previously evaluated values when possible (Section~\ref{sec:evaluation-reuse}). We
@@
 
-\begin{figure}
-\vspace{-1em}
+\begin{figure}[b]
 \hspace{1em}\includegraphics[scale=0.52]{figures/notebook.png}
-\vspace{-0.5em}
 \caption{Financial Times analysis that joins UN trade database with ISO country codes.}
 \label{fig:ft-uncomtrade}
-\vspace{-0.5em}
 \end{figure}
@@
 \begin{itemize}
+\vspace{0.5em}
 \item There is no abstraction. The analysis uses lambda functions as arguments to library calls,
@@
 
+\vspace{0.5em}
 \item The code relies on external libraries. Our example uses Pandas \cite{pandas},
@@
 
+\vspace{0.5em}
 \item There are many corner cases, such as the fact that \ident{keep\_default\_na}
@@
 
+\vspace{0.5em}
 \noindent
@@
 \begin{figure}
-  \vspace{-1.5em}
 \centering
@@
   \raggedright\footnotesize
-      Our live coding environment for The Gamma. The table is updated on-the-fly a
+    Our live programming environment for The Gamma. The table is updated on-the-fly a
     and shows the result at the current cursor position.
@@
 }{\includegraphics[scale=0.26]{figures/gamma2.png}}
-  \caption{\small Previous work on The Gamma (left) and our new editor with live preview (right).}
+\vspace{0.5em}
+\caption{\small Previous work on The Gamma with auto-complete based on type information (left)
+  and our new editor with instant preview (right).}
 \label{fig:gamma-intro}
-  \vspace{-0.5em}
 \end{figure}
@@
 
-Simple data exploration as discussed in the previous section has been the motivation for
+Simple data exploration has been the motivation for
 a scripting language The Gamma \cite{gamma}. Scripts in The Gamma are sequences of commands
@@
 functions and lambda functions can be used only as method arguments.
-
 Given the limited expressiveness of The Gamma, libraries are implemented in other languages,
 such as JavaScript. The Gamma uses type providers \cite{providers-fsharp}
-for accessing and exploring external data sources.
-Type providers generate object types with members and The Gamma makes using those convenient by
-providing auto-complete when the user types the dot symbol (`.') to access a member.
+for accessing external data sources.
+Type providers generate object types with members and The Gamma offers those in an
+auto-complete list when the user types dot (`.') to access a member.
 The combination of type providers and auto-complete makes it
@@
 
-\subsection{Live coding for data exploration}
-
-The implementation that accompanies this paper builds a live coding environment for The Gamma.
-The implementation is discussed in Section~\ref{sec:evaluation-case} and it replaces
-the original text editor with just auto-complete with a live coding environment
-that provides \emph{live previews}. We choose to extend The Gamma over using a subset of Python
-as the simplicity of TheGamma lets us focus on the principles of live coding systems.
-
-An example of a live preview is shown in Figure~\ref{fig:gamma-preview}. As noted earlier, The
+\subsection{Live programming for data exploration}
+
+The implementation that accompanies this paper builds a live programming environment for The Gamma.
+It is discussed in Section~\ref{sec:evaluation-case} and it replaces
+the original text editor with just auto-complete with a live programming environment
+that provides \emph{instant previews}.
+
+An example of a instant preview is shown in Figure~\ref{fig:gamma-preview}. As noted earlier, The
 Gamma programs consist of lists of commands which are either expressions or let bindings. Our
-editor displays a live preview below the command that the user is currently editing. The preview
+editor displays a instant preview below the command that the user is currently editing. The preview
 shows the result of evaluating the expression or the value assigned to a bound variable. When the
@@
 sufficient to capture the simple data analyses discussed in Section~\ref{sec:background}. We
-define the calculus in this section and then use it to formalise our live preview
+define the calculus in this section and then use it to formalise our instant preview
 mechanism in Section~\ref{sec:formal}.
 %
-The live preview mechanism does not rely on types and so we postpone the discussion of static
-typing to Section~\ref{sec:extra-types}. Interestingly, it reuses the mechanism used for live
+The instant preview mechanism does not rely on types and so we postpone the discussion of static
+typing to Appendix~\ref{sec:extra-types}. Interestingly, it reuses the mechanism used for live
 previews.
@@
 \begin{figure}
-\vspace{-0.5em}
 \raggedright
@@
 \quad\text{\rname{let}}\\[1.2em]
-{o.m(v_1, \ldots, v_n) \rightsquigarrow_\epsilon o'}\implies
+\inference
+  {o.m(v_1, \ldots, v_n) \rightsquigarrow_\epsilon o'}
   {C_p[o.m(v_1, \ldots, v_n)] \rightsquigarrow C_p[o']} &
@@
 
-\vspace{-0.5em}
 \caption{Syntax, contexts and reduction rules of the data exploration calculus}
 \label{fig:dec-calculus}
-\vspace{-0.5em}
 \end{figure}
@@
 The rule \rname{external} calls a method provided by an external library in a
-call-by-value fashion and \rname{let} substitutes a value of an evaluated variable in
+call-by-value fashion; \rname{let} substitutes a value of an evaluated variable in
 all subsequent commands and leaves the result in the list of commands.
-Note that our semantics does not define how $\lambda$ functions are reduced. This is done by
+Note that our semantics does not define how $\lambda$ applications are reduced. This is done by
 external libraries, which will typically supply functions with arguments using standard $\beta$-reduction.
 The behaviour is subject to constraints discussed next.
+
+\subsection{Example external library}
+\label{sec:calculus-example}
+
+The data exploration calculus is not limited to the data exploration domain.
+It can be used with external libraries for a wide range of other simple programming tasks, such as image
+manipulation, as done in Section~\ref{sec:evaluation-empirical}. However, we choose a name that reflects the
+domain that motivated this paper. To illustrate how a definition of an external library looks,
+consider the following simple data manipulation script:
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~l = \ident{list}.\ident{range}(\num{0}, \num{10})\\
+l.\ident{map}(\lambda x \rightarrow \ident{math}.\ident{mul}(x, \num{10}))
+\end{array}
+\end{equation*}
+%
+An external library provides the \ident{list} and \ident{math} values with members
+\ident{range}, \ident{map} and \ident{mul}. The objects of the external library consist of
+numbers $n$, lists of objects $[o_1, \ldots, o_k]$ and failed computations $\bot$ \cite{gowrong}.
+Next, the external library needs to define the $\rightsquigarrow_\epsilon$ relation that defines
+the evaluation of member accesses. The following shows the rules for members of lists, assuming
+the only supported member is \ident{map}:
+%
+\begin{equation*}
+\begin{array}{l}
+\\[-0.75em]
+\inference
+  {e[x \leftarrow n_i] \rightsquigarrow o_i \quad (\textit{for all $i\in 1\ldots k$})}
+  {[ n_1, \ldots, n_k ].\ident{map}(\lambda x\rightarrow e) \rightsquigarrow_\epsilon [ o_1, \ldots, o_k ]}
+\qquad
+\inference
+  {(\textit{otherwise})}
+  {[ n_1, \ldots, n_k ].m(v_1, \ldots, v_n) \rightsquigarrow_\epsilon \bot}
+\\[-0.75em]
+~
+\end{array}
+\end{equation*}
+%
+When evaluating \ident{map}, we apply the provided function to all elements of the list using
+standard $\beta$-reduction, defined recursively using $\rightsquigarrow$, and return a list with
+resulting objects. The $\rightsquigarrow_\epsilon$ relation is defined on all member accesses,
+but non-existent members reduce to the failed computation $\bot$.
 
@@
 Those two are particularly important as they will allow us to prove correctness of our method of
-evaluating live previews in Section~\ref{sec:evaluation-correctness}.
+evaluating instant previews in Section~\ref{sec:evaluation-correctness}.
 
@@
 observationally equivalent values as arguments, it should return the same value. Second, the
-evaluation of $o.m(v_1,\ldots,v_n)$ should be defined for all $o, n$ and $v_i$. The external
-library will typically satisfy this by defining an error object and reducing all invalid calls
-to the error object \cite{gowrong}.
+evaluation of $o.m(v_1,\ldots,v_n)$ should be defined for all $o, n$ and $v_i$.
+The definition in Section~\ref{sec:calculus-example} satisfies those by using standard
+$\beta$-reduction for lambda functions and by reducing all invalid calls to the $\bot$ object.
 
@@
 \noindent
-Compositionality is essential for proving the correctness of our live preview mechanism.
+Compositionality is essential for proving the correctness of our instant preview mechanism and
+implies determinism of external libraries.
 Totality allows us to prove normalization, i.e.~all programs reduce to
@@
 using the $\rightsquigarrow_\ident{let}$ reduction does not affect the result. This
-simplifies our later proof of live preview correctness in Section~\ref{sec:evaluation-correctness}.
+simplifies our later proof of instant preview correctness in Section~\ref{sec:evaluation-correctness}.
 
@@
 
-\section{Formalising a live coding environment}
+\section{Formalising a live programming environment}
 \label{sec:formal}
 
-A naive way of providing live previews during code editing would be to re-evaluate the code
+A naive way of providing instant previews during code editing would be to re-evaluate the code
 after each change. This would be wasteful -- when writing code to explore data, most changes are
@@
 
-\subsection{Maintaining dependency graph}
+\subsection{Maintaining the dependency graph}
 \label{sec:formal-deps}
@@
 the current node.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsubsection{Elements of the graph}
+The nodes of the graph represent individual operations of the computation. In our design, the nodes
+are used as cache keys, so we attach a unique symbol $s$ to some of the nodes. That way, we can
+create two unique nodes representing, for example, access to a member named \ident{take} which
+differ in their dependencies.
+
+The graph edges are labelled with labels indicating the kind of dependency. For
+a method call, the labels are ``first argument'', ``second argument'' and so on. Writing
+$s$ for symbols and $i$ for integers, nodes (vertices) $\bndclr{v}$ and edge labels $\blblclr{l}$
+are defined as:
+%
+\begin{equation*}
+\begin{array}{rcll}
+\bndclr{v}&=&\bnd{val}(o) \lsep \bnd{var}(x)\lsep \bnd{mem}(m, s)\lsep \bnd{fun}(x, s)&\qquad(\textit{Vertices})\\
+\blblclr{l}&=&\blbl{body} \lsep \blbl{arg}(i)&\qquad(\textit{Edge labels})\\
+\end{array}
+\end{equation*}
+%
+The \bnd{val} node represents a primitive value and contains the object itself. Two occurrences
+of $\num{10}$ in the source code will be represented by the same node. Member access \bnd{mem}
+contains the member name, together with a unique symbol $s$ -- two member access nodes with
+different dependencies will contain a different symbol. Dependencies of member access are labelled
+with \blbl{arg} indicating the index of the argument ($0$ for the instance and $1,\ldots$ for the
+arguments). Finally, nodes \bnd{fun} and \bnd{var} represent function values and variables
+bound by $\lambda$ abstraction.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsubsection{Example graph}
+Figure~\ref{fig:dep-graph} illustrates how we build the
+dependency graph. Node representing $\ident{take}(x)$ depends on the argument -- the
+number $\num{15}$ -- and the instance, which is a node representing $\ident{skip}(\num{10})$.
+This depends on the instance \ident{data} and the number $\num{10}$. Note that variables
+bound via \kvd{let} binding such as $x$ do not appear as $\bnd{var}$ nodes. The node using it
+depends directly on the node representing the expression assigned to $x$.
+
+After changing the value of $x$, we create a new graph. The dependencies of the node
+$\bnd{mem}(\ident{skip}, s_0)$ are unchanged and so the symbol $s_0$ attached to the node remains
+the same and previously computed previews can be reused. This part of the program is not recomputed.
+The $\blbl{arg}(1)$ dependency of the \ident{take} call
+changed and so we create a new node $\bnd{mem}(\ident{skip}, s_2)$ with a fresh symbol $s_2$.
+The preview for this node is then computed as needed using the already known values of its
+dependencies.
 
@@
   }\qquad$
+  \vspace{0.5em}
 }
@@
 }\qquad$
+\vspace{0.5em}
 }
+\vspace{0.5em}
 \caption{Dependency graphs formed by two steps of the live programming process. }
 \label{fig:dep-graph}
+\vspace{0.5em}
 \end{figure}
-
-% --------------------------------------------------------------------------------------------------
-
-\subsubsection{Elements of the graph}
-The nodes of the graph represent individual operations of the computation. In our design, the nodes
-are used as cache keys, so we attach a unique symbol $s$ to some of the nodes. That way, we can
-create two unique nodes representing, for example, access to a member named \ident{take} which
-differ in their dependencies.
-
-The graph edges are labelled with labels indicating the kind of dependency. For
-a method call, the labels are ``first argument'', ``second argument'' and so on. Writing
-$s$ for symbols and $i$ for integers, nodes (vertices) $\bndclr{v}$ and edge labels $\blblclr{l}$
-are defined as:
-%
-\begin{equation*}
-\begin{array}{rcll}
-\bndclr{v}&=&\bnd{val}(o) \lsep \bnd{var}(x)\lsep \bnd{mem}(m, s)\lsep \bnd{fun}(x, s)&\qquad(\textit{Vertices})\\
-\blblclr{l}&=&\blbl{body} \lsep \blbl{arg}(i)&\qquad(\textit{Edge labels})\\
-\end{array}
-\end{equation*}
-%
-The \bnd{val} node represents a primitive value and contains the object itself. Two occurrences
-of $\num{10}$ in the source code will be represented by the same node. Member access \bnd{mem}
-contains the member name, together with a unique symbol $s$ -- two member access nodes with
-different dependencies will contain a different symbol. Dependencies of member access are labelled
-with \blbl{arg} indicating the index of the argument ($0$ for the instance and $1,\ldots$ for the
-arguments). Finally, nodes \bnd{fun} and \bnd{var} represent function values and variables
-bound by $\lambda$ abstraction.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsubsection{Example graph}
-Figure~\ref{fig:dep-graph} illustrates how we build the
-dependency graph. Node representing $\ident{take}(x)$ depends on the argument -- the
-number $\num{15}$ -- and the instance, which is a node representing $\ident{skip}(\num{10})$.
-This depends on the instance \ident{data} and the number $\num{10}$. Note that variables
-bound via \kvd{let} binding such as $x$ do not appear as $\bnd{var}$ nodes. The node using it
-depends directly on the node representing the expression assigned to $x$.
-
-After changing the value of $x$, we create a new graph. The dependencies of the node
-$\bnd{mem}(\ident{skip}, s_0)$ are unchanged and so the symbol $s_0$ attached to the node remains
-the same and previously computed previews can be reused. This part of the program is not recomputed.
-The $\blbl{arg}(1)$ dependency of the \ident{take} call
-changed and so we create a new node $\bnd{mem}(\ident{skip}, s_2)$ with a fresh symbol $s_2$.
-The preview for this node is then computed as needed using the already known values of its
-dependencies.
 
@@
 \end{equation*}
-\vspace{-1em}
+\vspace{-0.5em}
 \caption{Updating the node cache after binding a new graph}
@@
 
-\section{Computing live previews}
+\section{Computing instant previews}
 \label{sec:previews}
@@
 are bound by earlier let bindings, but it will typically be delayed. We consider a speculative
-design for an abstraction mechanism that better supports live previews in
+design for an abstraction mechanism that better supports instant previews in
 Appendix~\ref{sec:extra-abstraction}.
@@
 \begin{equation*}
-\vspace{-2.5em}
+\vspace{-1.5em}
 \hspace{-0.5em}
@@
 \end{equation*}
-\vspace{-0.75em}
 \caption{Rules that define evaluation of previews over a dependency graph for a program}
 \label{fig:eval}
-\vspace{-0.75em}
 \end{figure*}
@@
 
-\section{Evaluating live coding environment}
+\section{Evaluating live programming environment}
 \label{sec:evaluation}
@@
 
-To avoid clutter when visualizing the performance, we update the preview after complete tokens are
+To avoid distractions when visualizing the performance, we update the preview after complete tokens are
 added rather than after individual keystrokes.
 Figure~\ref{fig:image-edits} shows the sequence of edits that we use to measure
-the delays in updating a live preview. We first enter an expression to load, greyscale and blur
+the delays in updating a instant preview. We first enter an expression to load, greyscale and blur
 an image (1) then introduce let binding (2) and add more operations (3). Finally, we extract
@@
 
-\begin{enumerate}[itemsep=0em,label=\alph*.]
+\begin{enumerate}[itemsep=0.1em,label=\alph*.]
 \item Loading image for the first time incurs small extra overhead in the live strategy.
@@
 
-\begin{figure}[t]
+\begin{figure}[b]
 \noindent
@@
   and aggregate data (left) and then create a chart with a label (right).
-  The example is adapted from \url{http://gallery.thegamma.net/73/}.}
+  The example has been created using our tool discussed in Section~\ref{sec:evaluation-case}
+  by a non-programmer and is available at \url{http://gallery.thegamma.net/73/}.}
 \label{fig:gallery}
@@
 
-\item When displaying live preview for code written using the data aggregation type provider (left),
+\item When displaying instant preview for code written using the data aggregation type provider (left),
   our environment generates tabes that show individual steps of the data transformation.
@@
 
-We aim to make creating transparent data analyses, such as those done by journalists, easier and we
-present a tool that provides instant feedback when writing code in a text editor.
-There is a rich body of work exploring complementary designs.
+Simple data exploration performed, for example, by journalists \cite{ddj} is done either
+programmatically or using spreadsheets. The latter is easy but error-prone while the former
+requires expert programming skills. We aim to bring liveness of spreadsheets to programmatic
+data exploration. This requires extending recomputation as done in spreadsheets \cite{spreadsheet}
+to code written in an ordinary text editor.
 
 \paragraph{Notebooks and data science tools.}
-Basic feedback mechanism, the REPL (read-eval-print-loop), originated in Lisp
-\cite{lisp} and has become standard \cite{drscheme}. Notebook systems
-such as Jupyter \cite{rmarkdown,jupyter,ipython} structure code in cells,
-but still rely on executing individual cells using a stateful REPL.
-Adding provenance tracking \cite{dataflow,noworkflow,wrattler} addresses
-the reproducibility issues of notebooks, but keeps cells as units of work. Ideas such as
-dependency tracking and efficient recomputation also exist in visual data exploration tools
+Visual data exploration tools are interactive \cite{control,tableau,vizdom} and some
+can export the workflow as a script \cite{wrangler}, but data analysts who prefer code
+typically resort to notebook systems such as Jupyter or R Markdown \cite{jupyter,rmarkdown}.
+Those are text-based, but have a limited model of recomputation. Users structure code in
+cells and manually reevaluate cells. Many notebook systems are based on the REPL (read-eval-print-loop)
+\cite{lisp,drscheme} and do not track dependencies between cells, which can lead to well-documented
+inconsistencies \cite{dataflow,noworkflow,wrattler}.
+
+Ideas such as dependency tracking and efficient recomputation exist in visual data exploration tools
 \cite{control,tableau,vizdom} and scientific workflow systems \cite{taverna,kepler}. The
@@
 with automatic code update; its usability in contrast to REPLs has been empirically
-evaluated \cite{ripple}. The challenges faced by Tempe are more akin to dynamic software update \cite{dsu}.
-
-\paragraph{Live coding and editing.}
-Live coding based on textual programs has been popularised by Victor \cite{principle,learnable} and
-is actively developed in domains such as live coded music \cite{beyond,sonic}. Kubelka et al.~\cite{liveroad}
-review the use of live coding in a Smalltalk derived environment. Lighttable \cite{lighttable}
-and Chrome DevTools provide limited live previews akin to those presented in this paper,
+evaluated \cite{ripple}. The challenges faced by Tempe are more akin to dynamic software update~\cite{dsu}.
+
+\paragraph{Live and exploratory programming.}
+Live programming based on textual programs has been popularised by Victor \cite{principle,learnable} and
+is actively developed in domains such as live coded music \cite{beyond,sonic}.
+The notion of exploratory programming has recently been analysed from the perspective of
+human-computer interaction \cite{exploratory}, which led to new tools \cite{variolite},
+complementary to our instant previews. Kubelka et al.~\cite{liveroad}
+review the use of live programming in a Smalltalk derived environment. Lighttable \cite{lighttable}
+and Chrome DevTools provide limited instant previews akin to those presented in this paper,
 but without well specified recomputation model. Finally, work on keeping state during
-code edits \cite{alive,livingit} would be relevant for supporting streaming data. In the future,
-we also plan to explore how direct manipulation \cite{direct} could be integrated with our preview
-mechanism.
-%
-Using structured editors \cite{structure-based} rather than text editors is a promising alternative
-direction as editors can recompute program based on understanding of edit operations
-\cite{interactive,livenut,lamdu}. This can be elegantly implemented using bi-directional
+code edits \cite{alive,livingit} would be relevant for supporting streaming data.
+
+A more principled approach can be used by systems based on structured editing \cite{structure-based,interactive,livenut,lamdu}
+where code is only modified via known operations with known effect on the computation graph
+(e.g.~``extract variable'' has no effect on the result; ``change constant value''
+forces recomputation of subsequent code). This can be elegantly implemented using bi-directional
 lambda calculus \cite{hazelnut}, but it also makes us consider more human-centric abstractions
@@
 Work on self-adjusting and incremental computation \cite{selfadjusting,adapton} handles
-recomputation when the program stays the same, but input changes. It relies on dependency tracking
-similar to ours. We rely on static code analysis whereas most incremental systems,
-e.g.~\cite{incremental,adaptivefp,deltaml} evaluate the program and use programmer-supplied
-information to build a dependency graph. \cite{adapton} implement a small adaptive interpreter
-that treats code as changing input data, suggesting an interesting implementation technique for
-live coding systems. Our use of dependency graphs \cite{dependencies} is static and first-order
+recomputation when the program stays the same, but input changes.
+Most incremental systems, e.g.~\cite{incremental,adaptivefp,deltaml} evaluate the program and use
+programmer-supplied information to build a dependency graph, whereas our system uses static code analysis;
+\cite{adapton} implement a small adaptive interpreter
+that treats code as changing input data, suggesting an implementation technique for
+live programming systems. Our use of dependency graphs \cite{dependencies} is static and first-order
 and can be seen as a form of program slicing \cite{slicing}, although our binding process is
@@
 modality \cite{modal} and delayed previews are linked to contextual modal type theory \cite{cmtt},
-formally modelled using comonads \cite{cmtt-denotation}. This suggests a direction for rigorous
-analysis of the presented system.
+formally modelled using comonads \cite{cmtt-denotation}.
 
@@
 
-Our main technical contribution is a \emph{live preview} mechanism that efficiently evaluates
+Our main technical contribution is a \emph{instant preview} mechanism that efficiently evaluates
 code during editing and instantly provides a preview of the result. We allow users to edit code
@@
 
+\acks
+We thank Dominic Orchard, Stephen Kell, Roly Perera and Jonathan Edwards for many discussions
+about the work presented in the paper. Dominic Orchard provided invaluable feedback on earlier
+version of the paper. The paper also benefited from suggestions made by anonymous reviewers
+of The Programming Journal as well as reviwers of earlier versions of the paper.
+This work was partly supported by The Alan Turing Institute under the EPSRC grant EP/N510129/1.
+
 % ==================================================================================================
@@
 \begin{theorem}[Normalization]
-\raggedrigth
 For all $p$, there exists $n$ and $o_1, \ldots, o_n$ \\
@@
 The operational semantics presented in this paper serves two purposes. It gives a simple guide
-for implementing text-based live coding environments for data science and we use it to prove that
-our optimized way of producing live previews is correct. However, some aspects of our mechanism
+for implementing text-based live programming environments for data science and we use it to prove that
+our optimized way of producing instant previews is correct. However, some aspects of our mechanism
 are related to important work in semantics of programming languages and deserve to be mentioned.
@@
 
-Live previews give analysts quick feedback when they write incorrect code, but
+Instant previews give analysts quick feedback when they write incorrect code, but
 having type information is still valuable. First, it can help give better error messages. Second,
@@
 reason, type checking can be relatively time consuming and can benefit from the same caching
-facilities as those available for live previews.
+facilities as those available for instant previews.
 
@@
 
-The results of type checking can be cached and reused in the same way as live previews, although
+The results of type checking can be cached and reused in the same way as instant previews, although
 we leave out the details. A property akin to correctness (Theorem~\ref{thm:correcntess})
@@
 
-Providing live previews inside ordinary functions is problematic, because we do not have readily
+Providing instant previews inside ordinary functions is problematic, because we do not have readily
 available values for input parameters and our mechanism for lambda functions only provides
@@
 
-This minimalistic abstraction mechanism supports code reuse without affecting how live previews
+This minimalistic abstraction mechanism supports code reuse without affecting how instant previews
 are computed. Commands in a labelled block require variables to be defined before the block.
@@
 
-\newpage
-\section{Manuscript version history (to be removed for publication)}
-\label{sec:revisions}
-
-This note briefly reviews the revisions made to this manuscript in response to reviews for an
-earlier version of the paper.
-
-\paragraph{Unclear motivation.}
-An earlier review pointed out that it was not clear who is the target user for the described
-system. The newly added Section~\ref{sec:background-jupyter} provides the answer. Our focus
-is on non-expert programmers conducting simple data exploration such as journalists working
-with government data. This is perhaps an unusual focus for a programming language research, but
-we believe it is increasingly important.
-
-\paragraph{Unclear scope of the work.}
-An earlier version of the paper did not make it clear what is prior work and
-what is new in this paper. The paper has been restructured to make this clear --
-we introduce the data exploration calculus (Section~\ref{sec:calculus}) and a
-theory for building a live coding environment for the calculus (Section~\ref{sec:formal})
-with live preview support (Section~\ref{sec:previews}). This is accompanied by an implementation
-that extends an editor for The Gamma, described earlier \cite{gamma} -- we add live preview
-support and auto-complete using type checking based on the dependency graph (Section~\ref{sec:extra-types}).
-
-\paragraph{Simplicity of the language.}
-An earlier review argued that the language used in the paper is overly simplistic.
-The newly added Section~\ref{sec:background-jupyter} explains the motivation behind the
-data exploration calculus. The simplicity of the language reflects the use cases that
-motivated this paper. The newly added case study (Section~\ref{sec:evaluation-case}) shows that
-the language is sufficient for building a simple, but realistic and practically useful
-data exploration environment. We also discuss extensions to the langauge in
-Section~\ref{sec:extra-abstraction}.
-
-\paragraph{Missing empirical evaluation.}
-An earlier version of the paper lacked empirical evaluation. It is not our aim to measure
-the exact overhead of our implementation (which is typically negligible with respect to the
-cost of evaluating the data transformations), but the newly added empirical evaluation
-(Section~\ref{sec:evaluation-empirical}) shows the effect of our evaluation strategy during
-a reprsentative coding task.
-
-\paragraph{Missing related work.}
-An earlier version of the paper lacked discussion of important related work, in particular that
-on incremental computation. This has now been addressed -- incremental computation focuses on
-the case where data change, while our focus is on cases where code changes, but the implementation
-techniques are closely related.
-
 \end{document}

### 2020-03-09 wip-design/paper.tex
Final version of programming, wip on design

\documentclass{sigchi}
\usepackage{xspace}
\usepackage{fancyvrb}
\usepackage{inconsolata}
\usepackage{xcolor}

\definecolor{kvdclr}{rgb}{0.3,0.0,0.8}
\DefineVerbatimEnvironment{thegamma}{Verbatim}{fontfamily=zi4,numbers=left,xleftmargin=6mm,fontsize=\small,commandchars=\\\{\}}
%\newcommand{\kvd}[1]{\textcolor{kvdclr}{#1}}
\newcommand{\kvd}[1]{\textbf{#1}}
\newcommand{\ikvd}[1]{{\fontfamily{zi4}\selectfont #1}}
\newcommand{\tg}{The Gamma\xspace}

% Use this section to set the ACM copyright statement (e.g. for
% preprints).  Consult the conference website for the camera-ready
% copyright statement.

% Copyright
\CopyrightYear{2020}
%\setcopyright{acmcopyright}
\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}
% DOI
\doi{https://doi.org/10.1145/3313831.XXXXXXX}
% ISBN
\isbn{978-1-4503-6708-0/20/04}
%Conference
\conferenceinfo{UIST'20,}{October  20--23, 2020, Minneapolis, MN, USA}
%Price
\acmPrice{\$15.00}

% Use this command to override the default ACM copyright statement
% (e.g. for preprints).  Consult the conference website for the
% camera-ready copyright statement.

%% HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP --
%% Please note you need to make sure the copy for your specific
%% license is used here!
% \toappear{
% Permission to make digital or hard copies of all or part of this work
% for personal or classroom use is granted without fee provided that
% copies are not made or distributed for profit or commercial advantage
% and that copies bear this notice and the full citation on the first
% page. Copyrights for components of this work owned by others than ACM
% must be honored. Abstracting with credit is permitted. To copy
% otherwise, or republish, to post on servers or to redistribute to
% lists, requires prior specific permission and/or a fee. Request
% permissions from \href{mailto:Permissions@acm.org}{Permissions@acm.org}. \\
% \emph{UIST '20},  October 20--23, 2020, Minneapolis, MN, USA \\
% ACM xxx-x-xxxx-xxxx-x/xx/xx\ldots \$15.00 \\
% DOI: \url{http://dx.doi.org/xx.xxxx/xxxxxxx.xxxxxxx}
% }

% Arabic page numbers for submission.  Remove this line to eliminate
% page numbers for the camera ready copy
% \pagenumbering{arabic}

% Load basic packages
\usepackage{balance}       % to better equalize the last page
\usepackage{graphics}      % for EPS, load graphicx instead
\usepackage[T1]{fontenc}   % for umlauts and other diaeresis
\usepackage{txfonts}
\usepackage{mathptmx}
\usepackage[pdflang={en-US},pdftex]{hyperref}
\usepackage{color}
\usepackage{booktabs}
\usepackage{textcomp}

% Some optional stuff you might like/need.
\usepackage{microtype}        % Improved Tracking and Kerning
% \usepackage[all]{hypcap}    % Fixes bug in hyperref caption linking
\usepackage{ccicons}          % Cite your images correctly!
% \usepackage[utf8]{inputenc} % for a UTF8 editor only

% If you want to use todo notes, marginpars etc. during creation of
% your draft document, you have to enable the "chi_draft" option for
% the document class. To do this, change the very first line to:
% "\documentclass[chi_draft]{sigchi}". You can then place todo notes
% by using the "\todo{...}"  command. Make sure to disable the draft
% option again before submitting your final document.
\usepackage{todonotes}

% Paper metadata (use plain text, for PDF inclusion and later
% re-using, if desired).  Use \emtpyauthor when submitting for review
% so you remain anonymous.
\def\plaintitle{The Gamma: Data Exploration through Iterative Prompting}
\def\plainauthor{First Author, Second Author, Third Author,
  Fourth Author, Fifth Author, Sixth Author}
\def\emptyauthor{}
\def\plainkeywords{Data exploration; End-user programming; Data journalism; Programming languages; Type providers}
\def\plaingeneralterms{Documentation, Standardization}

% llt: Define a global style for URLs, rather that the default one
\makeatletter
\def\url@leostyle{%
  \@ifundefined{selectfont}{
    \def\UrlFont{\sf}
  }{
    \def\UrlFont{\small\bf\ttfamily}
  }}
\makeatother
\urlstyle{leo}

% To make various LaTeX processors do the right thing with page size.
\def\pprw{8.5in}
\def\pprh{11in}
\special{papersize=\pprw,\pprh}
\setlength{\paperwidth}{\pprw}
\setlength{\paperheight}{\pprh}
\setlength{\pdfpagewidth}{\pprw}
\setlength{\pdfpageheight}{\pprh}

% Make sure hyperref comes last of your loaded packages, to give it a
% fighting chance of not being over-written, since its job is to
% redefine many LaTeX commands.
\definecolor{linkColor}{RGB}{6,125,233}
\hypersetup{%
  pdftitle={\plaintitle},
% Use \plainauthor for final version.
%  pdfauthor={\plainauthor},
  pdfauthor={\emptyauthor},
  pdfkeywords={\plainkeywords},
  pdfdisplaydoctitle=true, % For Accessibility
  bookmarksnumbered,
  pdfstartview={FitH},
  colorlinks,
  citecolor=black,
  filecolor=black,
  linkcolor=black,
  urlcolor=linkColor,
  breaklinks=true,
  hypertexnames=false
}

% create a shortcut to typeset table headings
% \newcommand\tabhead[1]{\small\textbf{#1}}

% End of preamble. Here it comes the document.
\begin{document}

\title{\plaintitle}

\numberofauthors{3}
\author{%
  \alignauthor{Leave Authors Anonymous\\
    \affaddr{for Submission}\\
    \affaddr{City, Country}\\
    \email{e-mail address}}\\
  \alignauthor{Leave Authors Anonymous\\
    \affaddr{for Submission}\\
    \affaddr{City, Country}\\
    \email{e-mail address}}\\
  \alignauthor{Leave Authors Anonymous\\
    \affaddr{for Submission}\\
    \affaddr{City, Country}\\
    \email{e-mail address}}\\
}

\maketitle

\begin{abstract}
Governments, non-profit organizations and citizen initiatives publish increasing amounts of
data, but extracting insights from such data and presenting them to the public is hard.
First, data comes in a variety of formats that each requires a different tool. Second, many
data exploration tools do not reveal how a result was obtained, making it difficult to reproduce
the results and check how they were obtained.
%
We contribute The Gamma, a novel data exploration environment for non-experts. The Gamma is based
on a single interaction principle and using it results in transparent and reproducible scripts.
This allows transfer of knowledge from one data source to another and
learning from previously created data analyses. We evaluate the usability and learnability of
The Gamma through a user study on non-technical employees of a research institute.
%
We argue that the our approach allows journalists and the public to benefit from the rise
of open data, by making data exploration easier, more transparent and more reproducible.
\end{abstract}

% ACM Classfication
%
% \begin{CCSXML}
% <ccs2012>
% <concept>
% <concept_id>10003120.10003121</concept_id>
% <concept_desc>Human-centered computing~Human computer interaction (HCI)</concept_desc>
% <concept_significance>500</concept_significance>
% </concept>
% <concept>
% <concept_id>10003120.10003121.10003125.10011752</concept_id>
% <concept_desc>Human-centered computing~Haptic devices</concept_desc>
% <concept_significance>300</concept_significance>
% </concept>
% <concept>
% <concept_id>10003120.10003121.10003122.10003334</concept_id>
% <concept_desc>Human-centered computing~User studies</concept_desc>
% <concept_significance>100</concept_significance>
% </concept>
% </ccs2012>
% \end{CCSXML}
%
% \ccsdesc[500]{Human-centered computing~Human computer interaction (HCI)}
% \ccsdesc[300]{Human-centered computing~Haptic devices}
% \ccsdesc[100]{Human-centered computing~User studies}

% Author Keywords
\keywords{\plainkeywords}

% Print the classficiation codes
% \printccsdesc
% Please use the 2012 Classifiers and see this link to embed them in the text: \url{https://dl.acm.org/ccs/ccs_flat.cfm}

\begin{figure}
\includegraphics[width=1\columnwidth]{figures/thegamma-annot}
\caption{Teams with the greatest number of gold medals from Rio 2016
Olympics with a reproducible The Gamma script (1), an auto-complete prompt offering ways
of sorting the data (2) and instant preview (3).}
\label{fig:thegamma}
\vspace{-0.5em}
\end{figure}

\section{Introduction}
Data science has more capabilities to help us understand the world than ever before, yet at the
same time post-truth politics and increasing public distrust in statistics makes data-driven insights
increasingly less relevant in public discourse~\cite{howstatslost}. This should perhaps not be a surpirse.
Journalists can access increasing amounts of data, but producing engaging and transparent data-driven
reports that are easy to interpret is expensive and requires expert programming skills \cite{ddj}.

The design of a data exploration tool for journalists poses a unique mix of challenges. First, the
tool needs to be easy to learn for end-users working under tight deadlines. Second, it needs to
support a wide range of data sources in a way where the expertise gained when working with one data
source is relevant for other data sources. Third, the resulting data-driven insights need to be
transparent, allowing the readers to verify the claims and learn how to reproduce the work.

We present The Gamma, a text-based data exploration environment for non-experts. The Gamma
is based on a single interaction principle, which provides uniform
access to a range of data sources including data tables, graph databases and data cubes.
An anlysis created in The Gamma is a transparent script that can be followed to reproduce the
result from scratch. This allows learning from existing analyses and encourages readers
to engage with the results. Our key contributions are:

\begin{itemize}
\item We identify the design requirements for a data exploration tool for journalists
  (Section~\ref{sec:motivation}) and follow those to build a novel programming environment
  The Gamma (Section~\ref{sec:overview}).

\item We introduce \emph{iterative prompting} (Section~\ref{sec:design}),
  an interaction design principle that can be used to complete a variety of programming
  tasks in a uniform way that allows transfer of knowledge between different tasks.

\item We show how to use the iterative prompting principle for querying of distinct data
  sources including data tables, graph databases and data cubes (Section~\ref{sec:implementation}).

\item We discuss a number of case studies (Section~\ref{sec:cases}) and
  conduct a user study to evaluate the usability of The Gamma and the extent to which users can,
  (i) learn from examples and (ii)~transfer knowledge between tasks (Section~\ref{sec:study}).
\end{itemize}

The Gamma is available as open-source at \href{http://thegamma.net}{\small\bf\ttfamily thegamma.net}.

\section{Related work}

\textbf{Visual tools.}

\textbf{Programming tools.}
Notebooks

\textbf{Journalism.}
Idyll \cite{idyll}

\textbf{Type providers.}
PL work

\newpage
~
\newpage

\begin{figure}
\includegraphics[width=1\columnwidth]{figures/nyt}
\caption{New York Times article on Obama's legacy \cite{youdraw}. The article asks the reader to make a guess
(engagement), but only lists ``Immigration and Customs Enforcement'' as a source of data.}
\label{fig:nyt}
\end{figure}

\section{Background and design goals}
\label{sec:motivation}

The Gamma aims to adapt the recent innovations in programming language research, especially the
work on type providers, into a form where it could be used in practice by journalists and other
non-expert interested in data exploration. We start with a careful consideration of our target
application domain, i.e.~data analyses produced by journalists and citizen data scientists that
are published online. We look at both practical requirements for such programming environment
and requirements arising from our focus on journalism. This analysis is based on the author's
experience of collaborating with journalists\footnote{Citations removed to preserve anonymity.},
review of literature on data journalism, e.g.~\cite{ddj,edcj17,edcj18} and more general trends in
journalism.

\subsection{Open Journalism}
Journalism continually develops and responds to the many challenges it faces \cite{future}.
Two recent challenges are relevant to our work. The first is building trust in media.
One way of establishing trust in the age of fake news is to be more transparent about editorial
decisions, process and original sources. Many journalists believe that opening up the process
shows the quality and trustworthiness of their work~\cite{transparency}.
The second challenge is reader engagement. To develop a relationship with readers, journalists are
increasingly looking for meaningful ways of engagement. This includes reader comments, involvement
of citizen journalists \cite{comments,citizen} and the development of new interactive formats
\cite{youdraw}. To address the above challenges, a tool for data exploration should satisfy the
following three requirements.

\paragraph{Trust Through Transparency}
To support trustworthiness, data analyses should be transparent. The reader should be able to
determine what is the source of analysed data and how has the data been transformed. As much as
possible, these capabilities should also be accessible to non-expert readers.

\paragraph{Reproducibility for Fact Checking}
It should be possible to re-run the analysis to verify that it produces the presented results.
However, running an opaque script is not enough. A reader should be able to recreate the analysis
by following the necessary steps from the original data source to the end result.

\paragraph{Encouraging Meaningful Engagement}
The tool should support a mechanism through which readers can engage in a meaningful discussion.
For example, it should allow modifying of parameters of a data visualization in order to show
how different choices affect the final result.

\begin{figure}
\includegraphics[width=1\columnwidth]{figures/ft}
\caption{Financial Times analysis of recyclable waste. Full source is provided as Jupyter Notebooks
on GitHub \cite{ftnotebooks}, but re-running the analysis is difficult, even for an expert.}
\label{fig:ft}
\end{figure}

\subsection{End-user Data Exploration}
Our aim is to make programmatic data exploration accessible to journalists, but we want to keep
the desirable properties of text-based programming. In particular, source code of a data
exploration should provide a full reproducible record of how the data analysis has been done.
As end-users, journalists have a number of interesting characteristics. They work under tight
deadline and data exploration is only a complementary skill. They also need to work with a wide
range of data sources, including big data tables (e.g.~Iraq War documents leak) or graph
databases (e.g.~Panama Papers). This leads to a number of practical requirements on the programming
environment.

\paragraph{Conceptual Simplicity}
We target end-users who cannot dedicate much time to learning about a tool prior
to using it. Consequently, using the tool should require understanding of only a small number
of concepts. Once the user understand a small number of concepts, they should be able to complete
basic data exploration tasks.

\paragraph{Uniformity across Data Sources}
The users should be able to navigate through large databases, query relational databases and
query graph databases through the same mechanism. Ideally, expertise gained with one data source
should also be transferable to working with another data source.

\paragraph{Learning without Experts}
Sarkar \cite{learning} reports that users learn how to use Excel either by talking to experts,
or by seeing a feature in a spreadsheet received from a colleague. In our circumstances, experts
are unlikely to be available, so the tool should support learning from examples. When looking at
a work done and published by another person, the user needs to see (and be able to understand)
how a task was completed.

\newpage
~
\newpage

\section{Overview}
\label{sec:overview}

\tg is a text-based programming environment that allows non-experts create simple data exploration
scripts using a single interaction principle -- choosing an item from an auto-complete list.
It supports a range of data sources including tabular data, graph data and data cubes.

\subsection{Querying Travel Expenses}
To introduce \tg, we walk through a simple problem that a local journalist might want to solve.
The UK government publishes travel expense claims by members of the House of Lords. A journalist
wants to find out which of the members representing the Kent county spend the most on travel.
The following shows a subset of the data:\footnote{ \url{https://www.parliament.uk/mps-lords-and-offices/members-allowances/house-of-lords/holallowances/} }

\begin{thegamma}
\textbf{Name, County, Days Attended, Days Away, Travel Costs}
Lord Adonis, London, 8, 0, 504
Baroness Afshar, Yorkshire, 2, 0, 0
Lord Alderdice, Oxfordshire, 3, 0, 114
Lord Alli, London, 5, 0, 0
Baroness Amos, London, 3, 0, 0
\end{thegamma}

After the analyst imports the CSV file (through a web interface), the environment is initialized
with code that refers to the imported variable \ikvd{expenses}. The analyst then types `.' (dot):

\begin{thegamma}
expenses.
\end{thegamma}
\vspace{-0.4em}\qquad\includegraphics[width=0.5\columnwidth]{figures/lords1}

The type provider for tabular data allows analysts to construct simple queries. It first offers
a list of operations that the analyst might want to perform such as grouping, filtering and sorting.
To find members of the House of Lords from Kent, the analyst chooses \ikvd{filter data},
types `.' and then chooses \ikvd{County is} from the offered list, types `.' and starts
typing Kent:

\begin{thegamma}
expenses
  .'filter data'.'County is'.Ke
\end{thegamma}
\vspace{-0.4em}\qquad\includegraphics[width=0.5\columnwidth]{figures/lords2}

The completion list is generated from the values in the \ikvd{County} column of the dataset.
After selecting \ikvd{Kent}, the live preview is updated to only show records according to the
specified filter:

\includegraphics[width=1\columnwidth]{figures/lords3}

To finish specifying filtering conditions, the analyst chooses \ikvd{then} and is offered the
same list of querying operations as in the first step. To sort House of Lords members by their
travel costs, she now chooses \ikvd{sort data} and types `.' (dot):

\begin{thegamma}
expenses
  .'filter data'.'County is'.Kent.then
  .'sort data'.
\end{thegamma}
\qquad\includegraphics[width=0.5\columnwidth]{figures/lords4}

The auto-complete offers a list of columns that can be used for sorting, each with ascending
(default) and descending order option. After choosing one or more sort keys, the analyst selects
the \ikvd{then} member and is, again, offered the list of querying operations. They use
\ikvd{paging} to get top 5 records and \ikvd{get series} to obtain a data series with just
the House of Lords member name and their travel expenses.

\begin{thegamma}
expenses
  .'filter data'.'County is'.Kent.then
  .'sort data'.'by Travel Costs descending'.then
  .paging.take(5)
  .'get series'.'with key Name'.'and value Travel Costs'
\end{thegamma}

When the code evaluates to a data series with a categorical (textual) key and a numerical value,
\tg switches from displaying the result as a table to a column chart:

\includegraphics[width=1\columnwidth]{figures/lords5}

\subsection{Querying via Iterative Prompting}
The

\newpage

\section{Design}
\label{sec:design}

interaction = iterative prompting
how is this different from just auto-complete?

trace = for transparency and learning

One important observation from the above list is that our tool should be accessible to non-expert
users such as readers and non-technical journalists, while providing extra capabilities for more
technical users. It should be easy to use if one just wants to modify existing code and should
encourage experimentation. Considering these challenges, we identify the following technical
design principles. todo{There must be papers on learning programming that can be referenced
here.}

\paragraph{B1. Learning from examples and by experimentation.} We should support two ways of learning.
Users of tools such as spreadsheets
often learn by looking at existing problem solutions todo{Advait's PPIG}. Our design should allow
this by making it possible to inspect and retrace steps used while solving a problem in an existing
application. Another principle of spreadsheets that we want to keep is the ability to experiment
and see results immediately. Our design should allow users to try invoking an operation or modifying
a parameter and quickly see if this leads to the desired results.

\paragraph{B2. Choice over construction.} To minimize the amount of information that users have to
learn and remember, our system should work in a way that allows constructing programs by
choosing from options that can reasonably appear in a current context, rather than requiring
users to recall particular syntax or exact identifier name.
todo{recognition over recall?}

\paragraph{B3. Make simple things easy and complex things possible.} Some users of the system may, over time, become advanced users
and the system should support those. In other words, the upper bound on what can be achieved
should be well above the most common use cases. At the same time, the complex features that power users might
need should not affect the most elementary uses of the system and should remain completely hidden
until needed. In other words, the lower bound on what one needs to know to use the system for basic
tasks should be as low as possible. todo{I think I got this idea of "boundaries" on what
is possible from some paper, but cannot recall which...}

\paragraph{B4. Visibility of state.} To support transparency, the system should make its entire state
transparent -- when reviewing a data analysis, all parameters should be immediately visible and the
user should not need to, e.g.~navigate through complex user interface to find them.

\section{Implementation}
\label{sec:implementation}

\subsection{Language}

\subsection{Type providers}

\section{Use cases}
\label{sec:cases}

\section{User study}
\label{sec:study}

this is between-user user study

FIRST: How one area transfers to another?

DEMO: WorldBank
explain everything

DISPLAY: Public order and safety, Defence
\begin{verbatim}
expenditure.byService.'Public order and safety'.inTermsOf.GDP
expenditure.byService.Defence.inTermsOf.GDP
\end{verbatim}

DEMO: Graph DB
explain everything

\begin{verbatim}
drWho.Character.Doctor.'ENEMY_OF'.'[any]'
  .'APPEARED_IN'.'[any]'.'explore_properties'.explore
  .'group data'.'by 1-name'.'count distinct 2-title'
\end{verbatim}

DISPLAY: Who has larges travel expenses?
(and in London only?)

\begin{verbatim}
lords.'sort data'.'by Travel Costs descending'

lords
  .'filter data'.'County is'.London.then
  .'sort data'.'by Travel Costs descending'
\end{verbatim}

SECOND: No experts are needed

DEMO: explain how live preview works, explain how '.' works,
explain how newlines and indentation work

DISPLAY 'CO2 emissions (metric tons per capita)'

\begin{verbatim}
worldbank.byCountry.'United Kingdom'
  .'Economy & Growth'.'GDP per capita (current US$)'

worldbank.byCountry.Germany
  .'Economy & Growth'.'GDP per capita (current US$)'

worldbank.byCountry.'Czech Republic'
  .'Economy & Growth'.'GDP per capita (current US$)'
\end{verbatim}

DISPLAY: Top athletes from London

\begin{verbatim}
olympics
  .'group data'.'by Team'.'sum Gold'.then
  .'sort data'.'by Gold descending'.then
  .paging.take(5)
  .'get series'.'with key Team'.'and value Gold'
\end{verbatim}

THIRD: 'then'

DEMO: Show worldbank
GIVE: Commented source code using 'olympics'
DISPLAY: Top athletes from London (think-aloud)

QUESTIONS

1) Did I tell you enough in the introduction to get started?

Say we want to provide educational materials for journalists
(with limited budgets), what would be the most important?

2) Video or just code samples?

We'll have more data sources than we can write tutorials for

3) Do we just teach them how the environment works?

4) What do we need to teach about a data source?

\section{Discussion}

\subsection{Study limitations}
exploratory in nature so we do not make any quantitative claims about effects

not comparing against other systems

\subsection{Design principles}
How well did we do wrt design principles?

\subsection{Design issues}
future challenges and limitations of the model - such as issues when modifying code
in the middle of the call chain

\section{Conclusions}

\newpage
~
\newpage

\section{Acknowledgments}
Yo

% BALANCE COLUMNS
\balance{}

% REFERENCES FORMAT
% References must be the same font size as other body text.
\bibliographystyle{SIGCHI-Reference-Format}
\bibliography{paper}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2020-03-09 programming-2020/paper.tex
Move published paper

--- 2020-02-13 wip-live-programming/paper.tex
@@
-\documentclass[english,crc]{programming}
+\AtBeginDocument{\paperdetails{
+    submitted=2019-09-30,
+    published=2020-02-17,
+    year=2020,
+    volume=4,
+    issue=3,
+    articlenumber=8,
+  }}
+\documentclass[english,crc,references=cleveref]{programming}
 \usepackage[backend=biber]{biblatex}
 \addbibresource{paper.bib}
-\usepackage{multicol}
-\usepackage{microtype}
-\usepackage{amssymb}
+% \usepackage{multicol}
+% \usepackage{microtype}
+% \usepackage{amssymb}
 \usepackage{xypic}
 \usepackage{semantic}
-\usepackage{amsmath}
+% \usepackage{amsmath}
 \usepackage{amsthm}
-\usepackage{enumitem}
-
+% \usepackage{enumitem}
+
+\AtBeginDocument{
   \newcounter{thc}
@@
 \newtheorem{definition}[dfc]{Definition}
+}
 
@@
 \newcommand{\lsep}{\;\;|\;\;}
-\newcommand{\num}[1]{\textcolor{numclr}{#1}}
+% \newcommand{\Num}[1]{\textcolor{numclr}{#1}}
+\newcommand{\Num}[1]{\num[color=numclr]{#1}}
 \newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
@@
 \author[a,b]{Tomas Petricek}
-\authorinfo{
-\noindent
-Tomas is a lecturer at University of Kent. He is working on making programming and data science
+\authorinfo{%
+%
+ % Tomas
+  is a lecturer at University of Kent. He is working on making programming and data science
 easier and more accessible. His research interests span a range of areas including theory of
@@
 \kvd{let}~\ident{top} =\ident{movies}.\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\\
-\qquad .\ident{take}(\num{10}).\ident{map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{yyyy}))\\
+\qquad .\ident{take}(\Num{10}).\ident{map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{yyyy}))\\
 \end{array}
@@
 A live programming environment computes and displays the list of years. Suppose that the programmer then
-makes $\num{10}$ a variable and changes the date format:
+makes $\Num{10}$ a variable and changes the date format:
 %
@@
 \begin{array}{l}
-\kvd{let}~\ident{count} = \num{10}\\
+\kvd{let}~\ident{count} = \Num{10}\\
 \kvd{let}~\ident{top} = \ident{movies}.\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\\
@@
 does this for a simple data exploration language in an unrestricted text editor. We discuss related
-work in Section~\ref{sec:future}, but we briefly review the most important directions here, in
+work in \cref{sec:future}, but we briefly review the most important directions here, in
 order to situate our contributions.
@@
 \begin{itemize}
-\item We introduce the \emph{data exploration calculus} (Section~\ref{sec:calculus}), a small formally
+\item We introduce the \emph{data exploration calculus} (\cref{sec:calculus}), a small formally
   tractable language for data exploration. The calculus is motivated by our review
-  of how data analysts work (Section~\ref{sec:background}) and makes it possible to write simple,
+  of how data analysts work (\cref{sec:background}) and makes it possible to write simple,
   transparent and reproducible data analyses.
@@
 \item A live programming environment does not operate in batch mode and so we cannot follow classic
-  compiler literature. We capture the essence of such new perspective (Section~\ref{sec:formal})
-  and use it to build our \emph{instant preview} mechanism (Section~\ref{sec:previews}) that evaluates
+  compiler literature. We capture the essence of such new perspective (\cref{sec:formal})
+  and use it to build our \emph{instant preview} mechanism (\cref{sec:previews}) that evaluates
   code instantly during editing.
 
-\item We prove that our instant preview mechanism is correct (Section~\ref{sec:evaluation-correctness}) and that
-  it reuses previously evaluated values when possible (Section~\ref{sec:evaluation-reuse}). We
+\item We prove that our instant preview mechanism is correct (\cref{sec:evaluation-correctness}) and that
+  it reuses previously evaluated values when possible (\cref{sec:evaluation-reuse}). We
   illustrate the practicality of the mechanism using an empirical evaluation
-  (Section~\ref{sec:evaluation-empirical}) and a case study (Section~\ref{sec:evaluation-case}).
+  (\cref{sec:evaluation-empirical}) and a case study (\cref{sec:evaluation-case}).
 \end{itemize}
@@
 
-\begin{figure}[b]
-\hspace{1em}\includegraphics[scale=0.52]{figures/notebook.png}
+\begin{figure}
+  \begin{wide}
+    \includegraphics[width=\linewidth]{figures/notebook}
+  \end{wide}
   \caption{Financial Times analysis that joins UN trade database with ISO country codes.}
@@
 code they write and the way they interact with the system is very different from how software
-engineers work \cite{workflow}. This paper focuses on simple data wrangling and data exploration as
+engineers work~\cite{workflow}. This paper focuses on simple data wrangling and data exploration as
 done, for example, by journalists analysing government datasets. This new kind of non-expert
@@
 
-Consider the Financial Times analysis of plastic waste \cite{ftnotebooks,ftarticle}. It joins
+Consider the Financial Times analysis of plastic waste~\cite{ftnotebooks,ftarticle}. It joins
 datasets from Eurostat, UN Comtrade and more, aggregates the data and
-builds a visualization comparing waste flows in 2017 and 2018. Figure~\ref{fig:ft-uncomtrade} shows
+builds a visualization comparing waste flows in 2017 and 2018. \Cref{fig:ft-uncomtrade} shows
 an excerpt from one notebook of the data analysis. The code has a number of important properties:
@@
 \begin{itemize}
-\vspace{0.5em}
+%\vspace{0.5em}
 \item There is no abstraction. The analysis uses lambda functions as arguments to library calls,
@@
 
-\vspace{0.5em}
-\item The code relies on external libraries. Our example uses Pandas \cite{pandas},
+%\vspace{0.5em}
+\item The code relies on external libraries. Our example uses Pandas~\cite{pandas},
   which provides operations for data wrangling such as \ident{merge} to join datasets
@@
 
-\vspace{0.5em}
+%\vspace{0.5em}
 \item The code is structured as a sequence of commands. Some commands define a variable, either by
@@
 
-\vspace{0.5em}
+%\vspace{0.5em}
 \item There are many corner cases, such as the fact that \ident{keep\_default\_na}
@@
 \end{itemize}
-
-\vspace{0.5em}
-\noindent
+%
+%\vspace{0.5em}
+%
 Many Jupyter notebooks are more complex than the above example and might use helper
@@
 \begin{figure}
-\centering
-\subcaptionbox{
-  \raggedright\footnotesize The analysis counts the number of distinct events per
-    athlete. After typing `.' the editor offers further aggregation operations.
+  \begin{wide}\centering
+    \begin{subfigure}{.49\linewidth}
+      \includegraphics[width=\linewidth]{figures/gamma1.png}
+      \caption{The analysis counts the number of distinct events per athlete.
+        After typing `.' the editor offers further aggregation operations.}
   \label{fig:gamma-dot}
-}{\includegraphics[scale=0.26]{figures/gamma1.png}}
-\subcaptionbox{
-  \raggedright\footnotesize
-    Our live programming environment for The Gamma. The table is updated on-the-fly a
-    and shows the result at the current cursor position.
+    \end{subfigure}\hfill
+    \begin{subfigure}{.49\linewidth}
+      \includegraphics[width=\linewidth]{figures/gamma2.png}
+      \caption{Our live programming environment for The Gamma. The table is
+        updated on-the-fly and shows the result at the current cursor
+        position.}
   \label{fig:gamma-preview}
-}{\includegraphics[scale=0.26]{figures/gamma2.png}}
-\vspace{0.5em}
-\caption{\small Previous work on The Gamma with auto-complete based on type information (left)
+    \end{subfigure}
+  \end{wide}
+% \vspace{0.5em}
+\caption{Previous work on The Gamma with auto-complete based on type information (left)
   and our new editor with instant preview (right).}
@@
 
+\begin{sloppypar}
 Simple data exploration has been the motivation for
-a scripting language The Gamma \cite{gamma}. Scripts in The Gamma are sequences of commands
+a scripting language The Gamma~\cite{gamma}. Scripts in The Gamma are sequences of commands
 that either define a variable or produce an output. It does not support top-level
@@
 Given the limited expressiveness of The Gamma, libraries are implemented in other languages,
-such as JavaScript. The Gamma uses type providers \cite{providers-fsharp}
+such as JavaScript. The Gamma uses type providers~\cite{providers-fsharp}
 for accessing external data sources.
@@
 selecting operations from a list.
-
-The example in Figure~\ref{fig:gamma-dot} summarizes data on Olympic medals. Identifiers such
+\end{sloppypar}
+
+The example in \cref{fig:gamma-dot} summarizes data on Olympic medals. Identifiers such
 as \qident{sum Bronze} are names of members generated by the type provider. The type provider
@@
 
-The Gamma language is richer, but the example in Figure~\ref{fig:gamma-dot} shows that non-trivial
+The Gamma language is richer, but the example in \cref{fig:gamma-dot} shows that non-trivial
 data exploration can be done using a very simple language. The assumptions about structure of
@@
 The implementation that accompanies this paper builds a live programming environment for The Gamma.
-It is discussed in Section~\ref{sec:evaluation-case} and it replaces
+It is discussed in \cref{sec:evaluation-case} and it replaces
 the original text editor with just auto-complete with a live programming environment
@@
 
-An example of a instant preview is shown in Figure~\ref{fig:gamma-preview}. As noted earlier, The
+An example of a instant preview is shown in \cref{fig:gamma-preview}. As noted earlier, The
 Gamma programs consist of lists of commands which are either expressions or let bindings. Our
@@
 define what objects are available and what the behaviour of their members is. This is
-sufficient to capture the simple data analyses discussed in Section~\ref{sec:background}. We
+sufficient to capture the simple data analyses discussed in \cref{sec:background}. We
 define the calculus in this section and then use it to formalise our instant preview
-mechanism in Section~\ref{sec:formal}.
+mechanism in \cref{sec:formal}.
 %
 The instant preview mechanism does not rely on types and so we postpone the discussion of static
-typing to Appendix~\ref{sec:extra-types}. Interestingly, it reuses the mechanism used for live
+typing to \cref{sec:extra-types}. Interestingly, it reuses the mechanism used for live
 previews.
@@
 \raggedright
-\hspace{0.5em}{\small\sffamily Programs, commands, terms, expressions and values}
-\vspace{-0.5em}
+\hspace{0.5em}{\sffamily Programs, commands, terms, expressions and values}
+%
 \begin{equation*}
@@
 
-\vspace{0.5em}
-\hspace{0.5em}{\small\sffamily Evaluation contexts of expressions}
-\vspace{-0.5em}
+%
+\hspace{0.5em}{\sffamily Evaluation contexts of expressions}
+%
 \begin{equation*}
@@
 
-\vspace{0.5em}
-\hspace{0.5em}{\small\sffamily Let elimination and member reduction}
-\vspace{-0.5em}
+%
+\hspace{0.5em}{\sffamily Let elimination and member reduction}
+%
 \begin{equation*}
@@
 The calculus combines object-oriented features such as member access with functional features
-including lambda functions. The syntax is defined in Figure~\ref{fig:dec-calculus}. Object values
+including lambda functions. The syntax is defined in \cref{fig:dec-calculus}. Object values
 $o$ are defined by external libraries that are used in conjunction with the core calculus.
@@
 The reduction always results in an object.
-Figure~\ref{fig:dec-calculus} defines the reduction rules in terms of $\rightsquigarrow_\epsilon$
+\Cref{fig:dec-calculus} defines the reduction rules in terms of $\rightsquigarrow_\epsilon$
 and evaluation contexts; $C_e$ specifies left-to-right evaluation of arguments of a method call,
@@
 It can be used with external libraries for a wide range of other simple programming tasks, such as image
-manipulation, as done in Section~\ref{sec:evaluation-empirical}. However, we choose a name that reflects the
+manipulation, as done in \cref{sec:evaluation-empirical}. However, we choose a name that reflects the
 domain that motivated this paper. To illustrate how a definition of an external library looks,
@@
 \begin{array}{l}
-\kvd{let}~l = \ident{list}.\ident{range}(\num{0}, \num{10})\\
-l.\ident{map}(\lambda x \rightarrow \ident{math}.\ident{mul}(x, \num{10}))
+\kvd{let}~l = \ident{list}.\ident{range}(\Num{0}, \Num{10})\\
+l.\ident{map}(\lambda x \rightarrow \ident{math}.\ident{mul}(x, \Num{10}))
 \end{array}
@@
 \ident{range}, \ident{map} and \ident{mul}. The objects of the external library consist of
-numbers $n$, lists of objects $[o_1, \ldots, o_k]$ and failed computations $\bot$ \cite{gowrong}.
+numbers $n$, lists of objects $[o_1, \ldots, o_k]$ and failed computations $\bot$~\cite{gowrong}.
 Next, the external library needs to define the $\rightsquigarrow_\epsilon$ relation that defines
@@
 Those two are particularly important as they will allow us to prove correctness of our method of
-evaluating instant previews in Section~\ref{sec:evaluation-correctness}.
+evaluating instant previews in \cref{sec:evaluation-correctness}.
 
@@
 \begin{itemize}
-\item[-] We write $\rightsquigarrow^{*}$ for the reflexive, transitive closure of $\rightsquigarrow$
-\item[-] We write $\rightsquigarrow_\ident{let}$ for a call-by-name let binding elimination
+\item We write $\rightsquigarrow^{*}$ for the reflexive, transitive closure of $\rightsquigarrow$
+\item We write $\rightsquigarrow_\ident{let}$ for a call-by-name let binding elimination
   $c_1;\; \ldots;\; c_{k-1};\;$\\ $\kvd{let}~x=t;\; c_{k+1};\; \ldots; c_n ~\rightsquigarrow_\ident{let}~
@@
 \end{definition}
-
-\noindent
+%
+%
 We say that two expressions
-$e$ and $e'$ are \emph{observationally equivalent} if, for any context $C$, the expressions $C[e]$ and
+$e$ and $e'$ are \emph{observationally equivalent} if, for any context~$C$, the expressions $C[e]$ and
 $C[e']$ reduce to the same value. Lambda functions $\lambda x\!\rightarrow\!2$ and
@@
 evaluation of $o.m(v_1,\ldots,v_n)$ should be defined for all $o, n$ and $v_i$.
-The definition in Section~\ref{sec:calculus-example} satisfies those by using standard
+The definition in \cref{sec:calculus-example} satisfies those by using standard
 $\beta$-reduction for lambda functions and by reducing all invalid calls to the $\bot$ object.
@@
 $\rightsquigarrow_\epsilon$ that satisfies the following two properties:
-\begin{itemize}
-\item[-] \emph{Totality.} For all $o, m, i$ and all $v_1, \ldots, v_i$, there exists $o'$ such
+\begin{description}
+\item [Totality] For all $o, m, i$ and all $v_1, \ldots, v_i$, there exists $o'$ such
   that $o.m(v_1, \ldots, v_i) \rightsquigarrow_\epsilon o'$.
-\item[-] \emph{Compositionality.} For observationally equivalent arguments, the reduction should always return
-  the same object, i.e.~given $e_0, e_1, \ldots, e_n$ and $e'_0, e'_1, \ldots, e'_n$ and $m$ such that
+\item[Compositionality] For observationally equivalent arguments, the reduction should always return
+  the same object, i.\hairspace e.~given $e_0, e_1, \ldots, e_n$ and $e'_0, e'_1, \ldots, e'_n$ and $m$ such that
   $e_0.m(e_1, \ldots, e_n) \rightsquigarrow^{*} o$ and $e'_0.m(e'_1, \ldots, e'_n) \rightsquigarrow^{*} o'$ then
@@
   $C_i[e'_i] \rightsquigarrow^{*} o_i$ for some $o_i$ then $o = o'$.
-\end{itemize}
+\end{description}
 \end{definition}
-
-\noindent
+%
+%
 Compositionality is essential for proving the correctness of our instant preview mechanism and
 implies determinism of external libraries.
-Totality allows us to prove normalization, i.e.~all programs reduce to
+Totality allows us to prove normalization, i.\hairspace e.~all programs reduce to
 a value -- although the resulting value may be an error value provided by the external library.
@@
 A program that is not a sequence of values can be reduced and reduction decreases the size of the
-program. See Appendix~\ref{sec:app-normalization} for more detail.
+program. See \cref{sec:app-normalization} for more detail.
 \end{proof}
-
-\noindent
+%
+%
 Although the reduction rules \rname{let} and \rname{external} of the data exploration calculus
@@
 using the $\rightsquigarrow_\ident{let}$ reduction does not affect the result. This
-simplifies our later proof of instant preview correctness in Section~\ref{sec:evaluation-correctness}.
+simplifies our later proof of instant preview correctness in \cref{sec:evaluation-correctness}.
 
@@
 By constructing $p' \rightsquigarrow^{*} o_1;\ldots;o_n$ from
-$p \rightsquigarrow^{*} o_1;\ldots;o_n$. See Appendix~\ref{sec:let-lang-elimination}.
+$p \rightsquigarrow^{*} o_1;\ldots;o_n$. See \cref{sec:let-lang-elimination}.
 \end{proof}
@@
 
-The key idea behind our method is to maintain a dependency graph \cite{dependencies} with
+The key idea behind our method is to maintain a dependency graph~\cite{dependencies} with
 nodes representing individual operations of the computation that can be evaluated
@@
 The \bnd{val} node represents a primitive value and contains the object itself. Two occurrences
-of $\num{10}$ in the source code will be represented by the same node. Member access \bnd{mem}
+of $\Num{10}$ in the source code will be represented by the same node. Member access \bnd{mem}
 contains the member name, together with a unique symbol $s$ -- two member access nodes with
 different dependencies will contain a different symbol. Dependencies of member access are labelled
-with \blbl{arg} indicating the index of the argument ($0$ for the instance and $1,\ldots$ for the
+with \blbl{arg} indicating the index of the argument ($0$~for the instance and $1,\ldots$ for the
 arguments). Finally, nodes \bnd{fun} and \bnd{var} represent function values and variables
@@
 \subsubsection{Example graph}
-Figure~\ref{fig:dep-graph} illustrates how we build the
+\cref{fig:dep-graph} illustrates how we build the
 dependency graph. Node representing $\ident{take}(x)$ depends on the argument -- the
-number $\num{15}$ -- and the instance, which is a node representing $\ident{skip}(\num{10})$.
-This depends on the instance \ident{data} and the number $\num{10}$. Note that variables
+number $\Num{15}$ -- and the instance, which is a node representing $\ident{skip}(\Num{10})$.
+This depends on the instance \ident{data} and the number $\Num{10}$. Note that variables
 bound via \kvd{let} binding such as $x$ do not appear as $\bnd{var}$ nodes. The node using it
@@
 \subcaptionbox{
-  \raggedright\small
+  \raggedright
   Graph constructed from initial expression:
-  $\kvd{let}~x = \num{15}~\kvd{in}~\ident{data.skip}(\num{10}).\ident{take}(x)$
+  $\kvd{let}~x = \Num{15}~\kvd{in}~\ident{data.skip}(\Num{10}).\ident{take}(x)$
   \label{fig:dep-graph-a}
@@
   $\xymatrix{
-  \bnd{val}(\num{10}) & \bnd{val}(\num{15})\\
+  \bnd{val}(\Num{10}) & \bnd{val}(\Num{15})\\
   \bnd{mem}(\ident{skip},s_0)\ar[d]_{\blbl{arg}(0)} \ar[u]^{\blbl{arg}(1)} &
@@
 \subcaptionbox{
-  \raggedright\small
+  \raggedright
   Updated graph after changing $x$ to $10$:
-  $\kvd{let}~x = \num{10}~\kvd{in}~\ident{data.skip}(\num{10}).\ident{take}(x)$
+  $\kvd{let}~x = \Num{10}~\kvd{in}~\ident{data.skip}(\Num{10}).\ident{take}(x)$
   \label{fig:dep-graph-b}
@@
   $\xymatrix{
-  \bnd{val}(\num{10})&\\
+  \bnd{val}(\Num{10})&\\
   \bnd{mem}(\ident{skip}, s_0)\ar[d]_{\blbl{arg}(0)} \ar[u]^{\blbl{arg}(1)} &
@@
 We write $\Delta(\bkndclr{k}, l)\!\downarrow$ when a value for a key is not defined.
-When creating the graph in Figure~\ref{fig:dep-graph-b},
+When creating the graph in \cref{fig:dep-graph-b},
 we perform the following lookups:
@@
 \begin{array}{ll}
-\Delta(\bknd{mem}(\ident{skip}), [(\bnd{var}(\ident{data}),\blbl{arg}(0)), (\bnd{val}(\num{10}), \blbl{arg}(1))]) &\qquad(1)\\
-\Delta(\bknd{mem}(\ident{take}), [(\bnd{mem}(\ident{skip}, s_0),\blbl{arg}(0)), (\bnd{val}(\num{10}), \blbl{arg}(1))]) &\qquad(2)
+\Delta(\bknd{mem}(\ident{skip}), [(\bnd{var}(\ident{data}),\blbl{arg}(0)), (\bnd{val}(\Num{10}), \blbl{arg}(1))]) &\qquad(1)\\
+\Delta(\bknd{mem}(\ident{take}), [(\bnd{mem}(\ident{skip}, s_0),\blbl{arg}(0)), (\bnd{val}(\Num{10}), \blbl{arg}(1))]) &\qquad(2)
 \end{array}
@@
 known from the previous step. We then look for the \ident{take} member access. In the earlier
-step, the argument of \ident{take} was $\num{15}$ rather than $\num{10}$ and so this lookup fails.
+step, the argument of \ident{take} was $\Num{15}$ rather than $\Num{10}$ and so this lookup fails.
 We then construct a new node $\bnd{mem}(\ident{take}, s_2)$ and later add it to the cache.
@@
 
-\begin{figure*}[t]
-\vspace{-1em}
-\begin{equation*}
-\hspace{-1.5em}
-\begin{array}{ll}
-\ident{bind-expr}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n)) = \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n) &\hspace{-2em}(1)~~\\
-\quad \textnormal{when}~\bndclr{v_i}, (V_i, E_i) = \ident{bind-expr}_{\Gamma, \Delta}(e_i)
-~\textnormal{and}~\bndclr{v} = \Delta(\bknd{mem}(m),[(\bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v_n}, \blbl{arg}(n))])\\
-\quad \textnormal{let}~E = \{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v}, \bndclr{v_n}, \blbl{arg}(n))\}
-\\[0.75em]
-\ident{bind-expr}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n)) = \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n)&\hspace{-2em}(2)\\
-\quad \textnormal{when}~\bndclr{v_i}, (V_i, E_i) = \ident{bind-expr}_{\Gamma, \Delta}(e_i)
-~\textnormal{and}~\Delta(\bknd{mem}(m),[(\bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v_n}, \blbl{arg}(n))])\!\downarrow\\
-\quad \textnormal{let}~\bndclr{v} = \bnd{mem}(m, s), s~\textnormal{fresh}
-~\textnormal{and}~E = \{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v},\bndclr{v_n}, \blbl{arg}(n)) \}
-\\[0.75em]
-\ident{bind-expr}_{\Gamma, \Delta}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V, \{e\} \cup E)&\hspace{-2em}(3)\\
-\quad \textnormal{when}~\Gamma_1 = \Gamma \cup \{ x, \bnd{var}(x) \}
+\begin{figure*}
+  \begin{wide}
+    \begin{align}
+&\ident{bind-expr}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n)) = \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n) \\
+&\quad \textnormal{when}~\bndclr{v_i}, (V_i, E_i) = \ident{bind-expr}_{\Gamma, \Delta}(e_i)
+~\textnormal{and}~\bndclr{v} = \Delta(\bknd{mem}(m),[(\bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v_n}, \blbl{arg}(n))])\notag\\
+&\quad \textnormal{let}~E = \{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v}, \bndclr{v_n}, \blbl{arg}(n))\}\notag\\
+%
+&\ident{bind-expr}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n)) = \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n)\\
+&\quad \textnormal{when}~\bndclr{v_i}, (V_i, E_i) = \ident{bind-expr}_{\Gamma, \Delta}(e_i)
+~\textnormal{and}~\Delta(\bknd{mem}(m),[(\bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v_n}, \blbl{arg}(n))])\!\downarrow\notag\\
+&\quad \textnormal{let}~\bndclr{v} = \bnd{mem}(m, s), s~\textnormal{fresh}
+~\textnormal{and}~E = \{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v},\bndclr{v_n}, \blbl{arg}(n)) \}\notag\\
+%
+&\ident{bind-expr}_{\Gamma, \Delta}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V, \{e\} \cup E)\\
+&\quad \textnormal{when}~\Gamma_1 = \Gamma \cup \{ x, \bnd{var}(x) \}
 ~\textnormal{and}~\bndclr{v_0}, (V, E) = \ident{bind-expr}_{\Gamma_1, \Delta}(e)
-~\textnormal{and}~\bndclr{v} = \Delta(\bknd{fun}(x),[(\bndclr{v_0}, \blbl{body})])\\
-\quad \textnormal{let}~e = (\bndclr{v}, \bndclr{v_0}, \blbl{body})
-\\[0.75em]
-\ident{bind-expr}_{\Gamma, \Delta}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V, \{e\} \cup E)&\hspace{-2em}(4)\\
-\quad \textnormal{when}~\Gamma_1 = \Gamma \cup \{ x, \bnd{var}(x) \}
+~\textnormal{and}~\bndclr{v} = \Delta(\bknd{fun}(x),[(\bndclr{v_0}, \blbl{body})])\notag\\
+&\quad \textnormal{let}~e = (\bndclr{v}, \bndclr{v_0}, \blbl{body})\notag\\
+%
+&\ident{bind-expr}_{\Gamma, \Delta}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V, \{e\} \cup E)\\
+&\quad \textnormal{when}~\Gamma_1 = \Gamma \cup \{ x, \bnd{var}(x) \}
 ~\textnormal{and}~\bndclr{v_0}, (V, E) = \ident{bind-expr}_{\Gamma_1, \Delta}(e)
-~\textnormal{and}~\Delta(\bknd{fun}(x),[(\bndclr{v_0}, \blbl{body})])\!\downarrow\\
-\quad \textnormal{let}~\bndclr{v} = \bnd{fun}(x, s), s~\textnormal{fresh}
-~\textnormal{and}~e = (\bndclr{v},\bndclr{v_0},\blbl{body})
-\\[0.75em]
-\ident{bind-expr}_{\Gamma, \Delta}(o) = \bnd{val}(o), (\{ \bnd{val}(o) \}, \emptyset) &\hspace{-2em}(5)
-\\[0.75em]
-\ident{bind-expr}_{\Gamma, \Delta}(x) = \bndclr{v}, (\{ \bndclr{v} \}, \emptyset)~\textnormal{when}~\bndclr{v} = \Gamma(x)&\hspace{-2em}(6)
-\end{array}
-\end{equation*}
-\vspace{-1em}
+~\textnormal{and}~\Delta(\bknd{fun}(x),[(\bndclr{v_0}, \blbl{body})])\!\downarrow\notag\\
+&\quad \textnormal{let}~\bndclr{v} = \bnd{fun}(x, s), s~\textnormal{fresh}
+~\textnormal{and}~e = (\bndclr{v},\bndclr{v_0},\blbl{body})\notag\\
+%
+&\ident{bind-expr}_{\Gamma, \Delta}(o) = \bnd{val}(o), (\{\bnd{val}(o) \}, \emptyset) \\
+&\ident{bind-expr}_{\Gamma, \Delta}(x) = \bndclr{v}, (\{ \bndclr{v} \}, \emptyset)~\textnormal{when}~\bndclr{v} = \Gamma(x)
+\end{align}
+  \end{wide}
 \caption{Binding rules that define a construction of a dependency graph for an expression.}
@@
 
-\begin{figure*}[t]
-\begin{equation*}
-\vspace{-0.5em}
-\hspace{-1.5em}
-\begin{array}{ll}
-\ident{bind-prog}_{\Gamma, \Delta}(\kvd{let}~x=e; c_2; \ldots; c_n) = \bndclr{v_1};\ldots;\bndclr{v_n}, (\{\bndclr{v_1}\} \cup V \cup V_1, E \cup E_1)
-  \hspace{9.35em}&\hspace{2.5em}(7)\\
-\quad \textnormal{let}~\bndclr{v_1}, (V_1,E_1) = \ident{bind-expr}_{\Gamma,\Delta}(e_1)~\textnormal{and}~\Gamma_1 = \Gamma \cup \{(x,\bndclr{v_1})\} \\
-\quad \textnormal{and}~\bndclr{v_2}; \ldots; \bndclr{v_n}, (V, E) = \ident{bind-prog}_{\Gamma_1, \Delta}(c_2; \ldots; c_n)
-\\[0.75em]
-\ident{bind-prog}_{\Gamma, \Delta}(e; c_2; \ldots; c_n) = \bndclr{v_1};\ldots;\bndclr{v_n}, (\{\bndclr{v_1}\} \cup V \cup V_1, E \cup E_1)&\hspace{2.5em}(8)\\
-\quad \textnormal{let}~\bndclr{v_1}, (V_1,E_1) = \ident{bind-expr}_{\Gamma,\Delta}(e)
-~\textnormal{and}~\bndclr{v_2}; \ldots; \bndclr{v_n}, (V, E) = \ident{bind-prog}_{\Gamma_1, \Delta}(c_2; \ldots; c_n)
-\\[0.75em]
-\ident{bind-prog}_{\Gamma, \Delta}([]) = [], (\emptyset, \emptyset)&\hspace{2.5em}(9)
-\end{array}
-\end{equation*}
-\vspace{-1em}
+\begin{figure*}
+\begin{wide}
+\begin{align}
+&\ident{bind-prog}_{\Gamma, \Delta}(\kvd{let}~x=e; c_2; \ldots; c_n) = \bndclr{v_1};\ldots;\bndclr{v_n}, (\{\bndclr{v_1}\} \cup V \cup V_1, E \cup E_1)\\
+&\quad \textnormal{let}~\bndclr{v_1}, (V_1,E_1) = \ident{bind-expr}_{\Gamma,\Delta}(e_1)~\textnormal{and}~\Gamma_1 = \Gamma \cup \{(x,\bndclr{v_1})\} \notag\\
+&\quad \textnormal{and}~\bndclr{v_2}; \ldots; \bndclr{v_n}, (V, E) = \ident{bind-prog}_{\Gamma_1, \Delta}(c_2; \ldots; c_n)\notag\\
+%
+&\ident{bind-prog}_{\Gamma, \Delta}(e; c_2; \ldots; c_n) = \bndclr{v_1};\ldots;\bndclr{v_n}, (\{\bndclr{v_1}\} \cup V \cup V_1, E \cup E_1)\\
+&\quad \textnormal{let}~\bndclr{v_1}, (V_1,E_1) = \ident{bind-expr}_{\Gamma,\Delta}(e)
+~\textnormal{and}~\bndclr{v_2}; \ldots; \bndclr{v_n}, (V, E) = \ident{bind-prog}_{\Gamma_1, \Delta}(c_2; \ldots; c_n)\notag\\
+%
+&\ident{bind-prog}_{\Gamma, \Delta}([]) = [], (\emptyset, \emptyset)
+\end{align}
+\end{wide}
 \caption{Binding rules that define a construction of a dependency graph for a program.}
 \label{fig:binding-rules-prog}
-\vspace{-0.5em}
+%
 \end{figure*}
@@
 syntax tree to a node of the dependency graph. This process is called
-binding and is defined by the \ident{bind-expr} function (Figure~\ref{fig:binding-rules-expr})
-and \ident{bind-prog} function (Figure~\ref{fig:binding-rules-prog}). Both functions are
+binding and is defined by the \ident{bind-expr} function (\cref{fig:binding-rules-expr})
+and \ident{bind-prog} function (\cref{fig:binding-rules-prog}). Both functions are
 annotated with a lookup table $\Delta$ and a variable
@@
 \begin{figure}
-\vspace{-0.5em}
+%
 \begin{equation*}
@@
 \ident{update}_{V,E}(\Delta_{i-1}) = \Delta_i~\textnormal{such that:}
-\\[0.75em]
+\\
 \quad\Delta_{i}(\bknd{mem}(m), [(\bndclr{v_0},\blbl{arg}(0)),\ldots, (\bndclr{v_n},\blbl{arg}(n))]) = \bnd{mem}(m, s)\\
@@
 ~\textnormal{and}~(\bnd{mem}(m, s), \bndclr{v_i}, \blbl{arg}(i)) \in E~\textnormal{for}~i\in 0,..,n
-\\[0.75em]
+\\
 \quad\Delta_{i}(\bknd{fun}(x), [(\bndclr{v_1},\blbl{body})]) = \bnd{fun}(x, s)\\
@@
 ~\textnormal{and}~(\bnd{fun}(x, s), \bndclr{v_1}, \blbl{body}) \in E
-\\[0.75em]
+\\
 \quad\Delta_{i}(\bkndclr{v}) = \Delta_{i-1}(\bkndclr{v})\quad \textnormal{(otherwise)}
@@
 \end{equation*}
-\vspace{-0.5em}
+%
 \caption{Updating the node cache after binding a new graph}
 \label{fig:loop}
-\vspace{-0.5em}
+%
 \end{figure}
@@
 We maintain a series of lookup table states $\Delta_0, \Delta_1, \Delta_2, \ldots$ The initial
-lookup table is empty, i.e.~$\Delta_0 = \emptyset$. At a step $i$, we parse a program $p_i$
+lookup table is empty, i.\hairspace e.~$\Delta_0 = \emptyset$. At a step $i$, we parse a program $p_i$
 and obtain a new dependency graph using the previous $\Delta$. The result is
@@
 The new state of the cache is computed using $\ident{update}_{V, E}(\Delta_{i-1})$
-defined in Figure~\ref{fig:loop}. The function adds newly created nodes from the graph
+defined in \cref{fig:loop}. The function adds newly created nodes from the graph
 $(V, E)$ to the previous cache $\Delta_{i-1}$ and returns a new cache $\Delta_{i}$.
@@
 graph, rather than over the structure of program expressions as in the operational semantics
-given in Section~\ref{sec:calculus-semantics}. In Section~\ref{sec:evaluation}, we prove that
+given in \cref{sec:calculus-semantics}. In \cref{sec:evaluation}, we prove that
 resulting previews are the same as the result we would get by directly evaluating code and we
@@
 \begin{array}{l}
-\kvd{let}~\ident{top} = \ident{movies}.\ident{take}(\num{10}).\ident{map}(\lambda x \rightarrow x.\ident{getReleased}().\ident{format}(\str{dd-mm-yyyy}))
-\end{array}
-\end{equation*}
-%
-Here, we can directly evaluate sub-expressions \ident{movies} and $\ident{movies}.\ident{take}(\num{10})$,
+\kvd{let}~\ident{top} = \ident{movies}.\ident{take}(\Num{10}).\ident{map}(\lambda x \rightarrow x.\ident{getReleased}().\ident{format}(\str{dd-mm-yyyy}))
+\end{array}
+\end{equation*}
+%
+Here, we can directly evaluate sub-expressions \ident{movies} and $\ident{movies}.\ident{take}(\Num{10})$,
 but not $x.\ident{getReleased}()$ because it contains a free variable $x$. Our preview evaluation algorithm addresses this by
@@
 $e$ that requires variables $\Gamma$. We use an untyped language and so $\Gamma$ is just a list of
-variables $x_1, \ldots, x_n$. As discussed in Appendix~\ref{sec:app-theories},
+variables $x_1, \ldots, x_n$. As discussed in \cref{sec:app-theories},
 delayed previews have an interesting theoretical link with graded comonads.
@@
 design for an abstraction mechanism that better supports instant previews in
-Appendix~\ref{sec:extra-abstraction}.
+\cref{sec:extra-abstraction}.
 
@@
 \begin{equation*}
-\vspace{-1.5em}
+%
 \hspace{-0.5em}
@@
 \end{equation*}
-\vspace{-0.5em}
+%
 \begin{equation*}
@@
 \subsection{Evaluation of previews}
-The evaluation of previews is defined in Figure~\ref{fig:eval}. Given a dependency graph $(V, E)$,
+The evaluation of previews is defined in \cref{fig:eval}. Given a dependency graph $(V, E)$,
 the relation $\bndclr{v}\Downarrow d$ evaluates a sub-expression corresponding to
@@
 
-For simplicity, the relation $\Downarrow$ in Figure~\ref{fig:eval} does not specify how previews
+For simplicity, the relation $\Downarrow$ in \cref{fig:eval} does not specify how previews
 are cached. In practice, this is done by maintaining a lookup table from graph nodes $\bndclr{v}$
@@
 Computing previews using a dependency graph implements a correct and efficient optimization.
-In this section we show that this is the case, first theoretically in Section~\ref{sec:evaluation-reuse},
-and then empirically in Section~\ref{sec:evaluation-empirical}. We also describe a case study where
+In this section we show that this is the case, first theoretically in \cref{sec:evaluation-reuse},
+and then empirically in \cref{sec:evaluation-empirical}. We also describe a case study where
 we developed an online service for data exploration based on the methods discussed in this paper
-(Section~\ref{sec:evaluation-case}).
+(\cref{sec:evaluation-case}).
 
@@
 
-To show that the previews are correct, we prove two properties. Correctness (Theorem~\ref{thm:correcntess})
+To show that the previews are correct, we prove two properties. Correctness (\cref{thm:correcntess})
 guarantees that, the previews we calculate using a dependency graph are the same as the values we
-would obtain by evaluating the program directly. Determinacy (Theorem~\ref{thm:determinacy})
+would obtain by evaluating the program directly. Determinacy (\cref{thm:determinacy})
 guarantees that previews assigned to a graph node based on an earlier graph are the same as
@@
 To simplify the proofs, we consider programs without let bindings. Eliminating let bindings does
-not change the result of evaluation, as shown in Lemma~\ref{thm:let-lang-elimination}, and it also
-does not change the constructed dependency graph as shown below in Lemma~\ref{thm:let-grp-elimination}.
+not change the result of evaluation, as shown in \cref{thm:let-lang-elimination}, and it also
+does not change the constructed dependency graph as shown below in \cref{thm:let-grp-elimination}.
 
@@
 \begin{proof}
-By analysis of the binding process. See Appendix~\ref{sec:app-let-grp-elimination}.
+By analysis of the binding process. See \cref{sec:app-let-grp-elimination}.
 \end{proof}
-
-\noindent
-The Lemma~\ref{thm:let-grp-elimination} provides a way of removing let bindings from a program,
+%
+%
+The \cref{thm:let-grp-elimination} provides a way of removing let bindings from a program,
 such that the resulting dependency graph remains the same. Here, we bind the original program
@@
 To keep the formalisation simpler, we separate the process of building the dependency graph
-and updating $\Delta$ and thus Lemma~\ref{thm:let-grp-elimination} requires an extra binding step.
+and updating $\Delta$ and thus \cref{thm:let-grp-elimination} requires an extra binding step.
 
@@
 \label{thm:lemma-lookup}
-Given $\Delta$ obtained using \ident{update} in Figure~\ref{fig:loop} then:
+Given $\Delta$ obtained using \ident{update} in \cref{fig:loop} then:
 \begin{itemize}
 \raggedright
-\item[--] If $\bndclr{v}=\Delta(\bknd{fun}(x),[(\bndclr{v_0}, \blblclr{l_0})])$
+\item If $\bndclr{v}=\Delta(\bknd{fun}(x),[(\bndclr{v_0}, \blblclr{l_0})])$
 then $\bndclr{v}=\bnd{fun}(x, s)$ for some $s$.
-\item[--] If $\bndclr{v}=\Delta(\bknd{mem}(m),[(\bndclr{v_0}, \blblclr{l_0}), \ldots, (\bndclr{v_n}, \blblclr{l_n})])$
+\item If $\bndclr{v}=\Delta(\bknd{mem}(m),[(\bndclr{v_0}, \blblclr{l_0}), \ldots, (\bndclr{v_n}, \blblclr{l_n})])$
 then $\bndclr{v}=\bnd{mem}(m, s)$ for some $s$.
@@
 \begin{proof}
-By construction of $\Delta$ in Figure~\ref{fig:loop}.
+By construction of $\Delta$ in \cref{fig:loop}.
 \end{proof}
@@
 Given a term $t$ that has no free variables, together with a lookup table $\Delta$ obtained
-from any sequence of programs using $\ident{bind-prog}$ (Figure~\ref{fig:binding-rules-prog}) and
-$\ident{update}$ (Figure~\ref{fig:loop}), then
+from any sequence of programs using $\ident{bind-prog}$ (\cref{fig:binding-rules-prog}) and
+$\ident{update}$ (\cref{fig:loop}), then
 let $\bndclr{v}, (V, E) = \ident{bind-expr}_{\emptyset, \Delta}(t)$.
 
-\vspace{0.25em}
-\noindent
 If $\bndclr{v}\Downarrow p$
@@
 \begin{proof}
-By induction over the binding process. See Appendix~\ref{sec:app-correctness}.
+By induction over the binding process. See \cref{sec:app-correctness}.
 \end{proof}
@@
 $\Delta_0$ obtained from any sequence of programs using $\ident{bind-prog}$
-(Figure~\ref{fig:binding-rules-prog}) and $\ident{update}$ (Figure~\ref{fig:loop}). Assume a
+(\cref{fig:binding-rules-prog}) and $\ident{update}$ (\cref{fig:loop}). Assume a
 let-free program $p'=t_1;\ldots;t_n$ such that $p\rightsquigarrow^{*}_\ident{let}p'$.
 
-\vspace{0.25em}
-\noindent
+\begin{sloppypar}
 Let $\bndclr{v_1};\ldots;\bndclr{v_n}, (V, E) = \ident{bind-prog}_{\emptyset, \Delta_0}(p)$
@@
 let $\bndclr{v'_1};\ldots;\bndclr{v'_n}, (V', E') = \ident{bind-prog}_{\emptyset, \Delta_1}(p)$.
-
-\vspace{0.25em}
-\noindent
+\end{sloppypar}
+
 If $\bndclr{v'_i}\Downarrow p_i$ over a graph $(V', E')$ then $p_i=o_i$ for some value $o_i$ and
@@
 \begin{proof}
-Direct consequence of Lemma~\ref{thm:let-grp-elimination} and Theorem~\ref{thm:let-free-correct}.
+Direct consequence of \cref{thm:let-grp-elimination} and \cref{thm:let-free-correct}.
 \end{proof}
-
-\noindent
+%
+%
 Our implementation updates $\Delta$ during the recursive binding process and so a stronger version
@@
 For some $\Delta$ and for any programs $p, p'$, assume that the first program is bound,
-i.e.~$\bndclr{v_1};\ldots;\bndclr{v_n}, (V, E) = \ident{bind-prog}_{\emptyset, \Delta}(p)$,
+i.\hairspace e.~$\bndclr{v_1};\ldots;\bndclr{v_n}, (V, E) = \ident{bind-prog}_{\emptyset, \Delta}(p)$,
 the graph node cache is updated $\Delta' = \ident{update}_{V, E}(\Delta)$ and the second program is
-bound, i.e.~$\bndclr{v'_1};\ldots;\bndclr{v'_m}, (V', E') = \ident{bind-prog}_{\emptyset, \Delta'}(p')$.
+bound, i.\hairspace e.~$\bndclr{v'_1};\ldots;\bndclr{v'_m}, (V', E') = \ident{bind-prog}_{\emptyset, \Delta'}(p')$.
 Now, for any $\bndclr{v}$, if $\bndclr{v} \Downarrow p$ over $(V, E)$ then also
@@
 \end{proof}
-
-\noindent
-The cache of previews (Section~\ref{sec:previews-cache}) associates a preview $d$ with a node
-$\bndclr{v}$ as the key. Theorem~\ref{thm:determinacy} guarantees that this is valid. As we update dependency
+%
+%
+The cache of previews (\cref{sec:previews-cache}) associates a preview $d$ with a node
+$\bndclr{v}$ as the key. \Cref{thm:determinacy} guarantees that this is valid. As we update dependency
 graph during code editing, previous nodes will continue representing the same sub-expressions.
@@
 \raggedright
-\vspace{-0.5em}
-{\small\sffamily Edit contexts of expressions}
+%
+{\sffamily Edit contexts of expressions}
 \begin{equation*}
@@
 
-\vspace{0.5em}
-{\small\sffamily Code edit operations preserving preview for a sub-expression}
+%
+{\sffamily Code edit operations preserving preview for a sub-expression}
 
@@
 \label{fig:operations}
-\vspace{-0.5em}
+%
 \end{figure}
@@
 values for a sub-expression can be reused. This includes the motivating example from
-Section~\ref{sec:intro} where the data analyst extracted a constant into a let binding and
+\cref{sec:intro} where the data analyst extracted a constant into a let binding and
 modified a parameter of the last method call in a call chain.
 
-The list of preview-preserving edits is shown in Figure~\ref{fig:operations}.
+The list of preview-preserving edits is shown in \cref{fig:operations}.
 It includes several ways of introducing and eliminating let bindings and edits where the analyst
@@
 typical edits that the data analyst might perform when writing code. To express the operations we
-define an editing context $K$ which is similar to evaluation context $C$ from Figure~\ref{fig:dec-calculus},
+define an editing context $K$ which is similar to evaluation context $C$ from \cref{fig:dec-calculus},
 but allows sub-expressions appearing anywhere in the program.
@@
 
-Theorem~\ref{thm:preview-reuse} proves that the operations given in Figure~\ref{fig:operations}
-preserve the preview for a marked sub-expression. It relies on a Lemma~\ref{thm:sub-expr} given in
-Section~\ref{sec:sub-expr} that generally characterizes one common kind of edits. Given two versions
+\Cref{thm:preview-reuse} proves that the operations given in \cref{fig:operations}
+preserve the preview for a marked sub-expression. It relies on a \cref{thm:sub-expr} given in
+\cref{sec:sub-expr} that generally characterizes one common kind of edits. Given two versions
 of a program that both contain the same sub-expression $e$, if the let bindings that define the values of
@@
 \label{thm:preview-reuse}
-Given the sequence of expressions as specified in Figure~\ref{fig:operations}, if the expressions
-are bound in sequence and graph node cache updated as specified in Figure~\ref{fig:loop}, then
+Given the sequence of expressions as specified in \cref{fig:operations}, if the expressions
+are bound in sequence and graph node cache updated as specified in \cref{fig:loop}, then
 the graph nodes assigned to the specified sub-expressions are the same.
@@
 \begin{proof}
-Cases \rname{edit-let} and \rname{edit-mem} are direct consequences of Lemma~\ref{thm:sub-expr};
+Cases \rname{edit-let} and \rname{edit-mem} are direct consequences of \cref{thm:sub-expr};
 for \rname{let-intro-var}, the node assigned to $x$ is the node assigned to $e$ which is the
-same as before the edit from Lemma~\ref{thm:sub-expr}.
+same as before the edit from \cref{thm:sub-expr}.
 Cases \rname{let-intro-ins} and \rname{let-intro-del} are similar to \rname{let-intro-var}, but
 also require using induction over the binding of $K_c[e]$. Finally, cases \rname{let-elim-ins}
-and \rname{let-elim-del} are similar and also use Lemma~\ref{thm:sub-expr}
+and \rname{let-elim-del} are similar and also use \cref{thm:sub-expr}
 together with induction over the binding of $K_c[x]$.
@@
 \raggedright
-\vspace{-0.5em}
-{\small\sffamily (1) Enter the following code and then change parameter of blur from 4 to 8:}
-\vspace{-0.5em}
+%
+{\sffamily (1) Enter the following code and then change parameter of blur from 4 to 8:}
+%
 \begin{equation*}
 \begin{array}{l}
-\ident{image}.\ident{load}(\str{shadow.png}).\ident{greyScale}().\ident{blur}(\num{4})
-\end{array}
-\end{equation*}
-
-{\small\sffamily (2) Assign the result to a variable and start writing code for further operations:}
-\vspace{-0.5em}
+\ident{image}.\ident{load}(\str{shadow.png}).\ident{greyScale}().\ident{blur}(\Num{4})
+\end{array}
+\end{equation*}
+
+{\sffamily (2) Assign the result to a variable and start writing code for further operations:}
+%
 \begin{equation*}
 \begin{array}{l}
-\kvd{let}\,\ident{shadow}=\ident{image}.\ident{load}(\str{shadow.png}).\ident{greyScale}().\ident{blur}(\num{8})\\
+\kvd{let}\,\ident{shadow}=\ident{image}.\ident{load}(\str{shadow.png}).\ident{greyScale}().\ident{blur}(\Num{8})\\
 \ident{shadow}.\ident{combine}
@@
 
-{\small\sffamily (3) Finish code to combine two images and then change parameter of combine from 20 to 80:}
-\vspace{-0.5em}
+{\sffamily (3) Finish code to combine two images and change parameter of combine from 20 to 80:}
+%
 \begin{equation*}
 \begin{array}{l}
-\kvd{let}\,\ident{shadow}=\ident{image}.\ident{load}(\str{shadow.png}).\ident{greyScale}().\ident{blur}(\num{8})\\
-\ident{shadow}.\ident{combine}(\ident{image}.\ident{load}(\str{pope.png}), \num{20})
-\end{array}
-\end{equation*}
-
-{\small\sffamily (4) Extract the parameter of combine to a let bound variable:}
-\vspace{-0.5em}
+\kvd{let}\,\ident{shadow}=\ident{image}.\ident{load}(\str{shadow.png}).\ident{greyScale}().\ident{blur}(\Num{8})\\
+\ident{shadow}.\ident{combine}(\ident{image}.\ident{load}(\str{pope.png}), \Num{20})
+\end{array}
+\end{equation*}
+
+{\sffamily (4) Extract the parameter of combine to a let bound variable:}
+%
 \begin{equation*}
 \begin{array}{l}
-\kvd{let}\,\ident{ratio} = \num{80} \\
-\kvd{let}\,\ident{shadow}=\ident{image}.\ident{load}(\str{shadow.png}).\ident{greyScale}().\ident{blur}(\num{8})\\
+\kvd{let}\,\ident{ratio} = \Num{80} \\
+\kvd{let}\,\ident{shadow}=\ident{image}.\ident{load}(\str{shadow.png}).\ident{greyScale}().\ident{blur}(\Num{8})\\
 \ident{shadow}.\ident{combine}(\ident{image}.\ident{load}(\str{pope.png}), \ident{ratio})
@@
 \end{equation*}
-
-\vspace{-1em}
+%
 \caption{Code edit operations that are used in the experimental evaluation}
 \label{fig:image-edits}
-\vspace{-0.5em}
+%
 \end{figure}
@@
 added rather than after individual keystrokes.
-Figure~\ref{fig:image-edits} shows the sequence of edits that we use to measure
+\Cref{fig:image-edits} shows the sequence of edits that we use to measure
 the delays in updating a instant preview. We first enter an expression to load, greyscale and blur
@@
 
-We implement the algorithm described in Sections~\ref{sec:formal} and \ref{sec:previews} in a
+We implement the algorithm described in \cref{sec:formal} \cref{sec:previews} in a
 simple web-based environment that allows the user to modify code and explicitly trigger
@@
 
-\begin{itemize}
-\item \emph{Call-by-value.} Following the semantics in Section~\ref{sec:calculus-semantics},
+\begin{description}
+\item[Call-by-value] Following the semantics in \cref{sec:calculus-semantics},
   all sub-expressions are evaluated before an expression. This is often wasteful. For example,
@@
 
-\item \emph{Lazy.} To address the wastefulness of call-by-value strategy, we
+\item[Lazy] To address the wastefulness of call-by-value strategy, we
   simulate lazy evaluation by implementing a version of the image processing library where
@@
 
-\item \emph{Live.} Finally, we use the algorithm described in
-  Sections~\ref{sec:formal} and \ref{sec:previews}. The cache is empty at the beginning of the
+\item[Live] Finally, we use the algorithm described in
+  \cref{sec:formal} \cref{sec:previews}. The cache is empty at the beginning of the
   experiment and we update it after each token is added. This is the only strategy where
   evaluation does not start afresh after reparsing code.
-\end{itemize}
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
-\vspace{-0.5em}
-\includegraphics[scale=0.425]{figures/drawing.png}
+\end{description}
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}
+%
+\begin{wide}\includegraphics[width=\linewidth]{figures/drawing.png} \end{wide}
 \caption{Time required to recompute the results of a sample program after individual tokens
@@
 \label{fig:drawing}
-\vspace{-0.5em}
+%
 \end{figure}
@@
 
-\begin{figure}[b]
-\vspace{-0.5em}
-\noindent
-\includegraphics[scale=0.48]{figures/hist-all.png}
-\includegraphics[scale=0.48]{figures/hist-slow.png}
-\vspace{-0.25em}
+\begin{figure}
+%
+\begin{wide}
+\includegraphics[width=.5\linewidth]{figures/hist-all.png}%
+\includegraphics[width=.5\linewidth]{figures/hist-slow.png}
+\end{wide}
 \caption{Distribution of delays incurred when updating previews.
   We show a histogram computed from all delays (left) and only from delays larger
-  than 15ms (right).}
+  than \SI{15}{\ms} (right).}
 \label{fig:drawing-hist}
-\vspace{-0.75em}
+%
 \end{figure}
@@
 compiler. We run the experiments in Firefox (version 64.0.2, 32bit) on Windows 10
-(build 1809, 64bit) on Intel Core i7-7660U CPU with 16Gb RAM.
-
-Figure~\ref{fig:drawing} shows times needed to recompute previews after individual tokens
+(build 1809, \SI{64}{\bit}) on Intel Core i7-7660U CPU with \SI{16}{\giga\byte} RAM.
+
+\Cref{fig:drawing} shows times needed to recompute previews after individual tokens
 are added, deleted or modified, according to the script in
-Figure~\ref{fig:image-edits}, resulting in 38 measurements. We mark a number of notable points in the chart:
-
-\begin{enumerate}[itemsep=0.1em,label=\alph*.]
+\cref{fig:image-edits}, resulting in 38 measurements. We mark a number of notable points in the chart:
+
+~
+
+\begin{enumerate}
+  \renewcommand{\theenumi}{\alph{enumi}}%
 \item Loading image for the first time incurs small extra overhead in the live strategy.
@@
 \end{enumerate}
-
-\noindent
-A summarized view of the delays is provided in Figure~\ref{fig:drawing-hist}, which shows
+%
+%
+A summarized view of the delays is provided in \cref{fig:drawing-hist}, which shows
 histograms illustrating the distribution of delays for each of the three evaluation methods.
-A large proportion of delays is very small (less than 15ms) because the parser used in our
-experimental environment often fails (e.g.~for unclosed parentheses). The histogram
+A large proportion of delays is very small (less than \SI{15}{\ms}) because the parser used in our
+experimental environment often fails (e.\hairspace g.~for unclosed parentheses). The histogram
 on the right summarizes only delays for edit operations where the delay for at least one of
-the strategies was over 15ms. The histogram shows that the live strategy eliminates the longest
+the strategies was over \SI{15}{\ms}. The histogram shows that the live strategy eliminates the longest
 delays (by caching partial results), with the exception of a few where the underlying operation
@@
 keeps the same trend. Our \emph{live} strategy reuses values computed
-previously. Consequently, expensive operations such as (d) and (g) in Figure~\ref{fig:drawing}
+previously. Consequently, expensive operations such as (d) and (g) in \cref{fig:drawing}
 are significantly faster, because they do not need to recompute operations done previously when
-writing the code. As shown in Figure~\ref{fig:drawing-hist}, there are almost no very expensive
+writing the code. As shown in \cref{fig:drawing-hist}, there are almost no very expensive
 operations (taking over 1 second) in the \emph{live} strategy in contrast to several in the
@@
 
-\begin{figure}[b]
-\noindent
-\hspace{-1em}
-\includegraphics[scale=0.245]{figures/gallery1.png}
-\includegraphics[scale=0.245]{figures/gallery2.png}
-\caption{\small Data analysis counting the number of libraries per county in the UK. We load
+\begin{figure}
+  \begin{wide}
+\includegraphics[width=.5\linewidth]{figures/gallery1.png}
+\includegraphics[width=.5\linewidth]{figures/gallery2.png}
+\end{wide}
+\caption{Data analysis counting the number of libraries per county in the UK. We load
   and aggregate data (left) and then create a chart with a label (right).
-  The example has been created using our tool discussed in Section~\ref{sec:evaluation-case}
+  The example has been created using our tool discussed in \cref{sec:evaluation-case}
   by a non-programmer and is available at \url{http://gallery.thegamma.net/73/}.}
 \label{fig:gallery}
-\vspace{-1em}
+%
 \end{figure}
@@
 
-In Section~\ref{sec:background-jupyter}, we motivated our work by considering how journalists
+In \cref{sec:background-jupyter}, we motivated our work by considering how journalists
 explore open data. In addition to the theoretical and experimental work presented in this paper,
@@
 report on our experience with the system. Two screenshots shown
-in Figure~\ref{fig:gallery} illustrate a number of interesting features:
+in \cref{fig:gallery} illustrate a number of interesting features:
 
 \begin{itemize}
-\item The left screenshot uses a type provider for data aggregation \cite{gamma}.
+\item The left screenshot uses a type provider for data aggregation~\cite{gamma}.
   Type provider are treated as an external library
@@
   provide editor auto-complete, which we support by implementing type checking over a
-  dependency graph (Section~\ref{sec:extra-types}).
+  dependency graph (\cref{sec:extra-types}).
 
@@
   A tab is selected based on the cursor position and shows a preview for the current sub-expression
-  (e.g.~raw data before grouping).
+  (e.\hairspace g.~raw data before grouping).
 
@@
 \end{itemize}
-
-\noindent
+%
+%
 The environment is available at \url{http://gallery.thegamma.net}.
@@
 is left for future work. As an anecdotal evidence, the code in
-Figure~\ref{fig:gallery} was developed by an attendee of a Mozilla Festival 2017
+\cref{fig:gallery} was developed by an attendee of a Mozilla Festival 2017
 who had no prior programming experience.
@@
 
-Simple data exploration performed, for example, by journalists \cite{ddj} is done either
+Simple data exploration performed, for example, by journalists~\cite{ddj} is done either
 programmatically or using spreadsheets. The latter is easy but error-prone while the former
 requires expert programming skills. We aim to bring liveness of spreadsheets to programmatic
-data exploration. This requires extending recomputation as done in spreadsheets \cite{spreadsheet}
+data exploration. This requires extending recomputation as done in spreadsheets~\cite{spreadsheet}
 to code written in an ordinary text editor.
@@
 \paragraph{Notebooks and data science tools.}
-Visual data exploration tools are interactive \cite{control,tableau,vizdom} and some
-can export the workflow as a script \cite{wrangler}, but data analysts who prefer code
-typically resort to notebook systems such as Jupyter or R Markdown \cite{jupyter,rmarkdown}.
+Visual data exploration tools are interactive~\cite{control,tableau,vizdom} and some
+can export the workflow as a script~\cite{wrangler}, but data analysts who prefer code
+typically resort to notebook systems such as Jupyter or R Markdown~\cite{jupyter,rmarkdown}.
 Those are text-based, but have a limited model of recomputation. Users structure code in
-cells and manually reevaluate cells. Many notebook systems are based on the REPL (read-eval-print-loop)
-\cite{lisp,drscheme} and do not track dependencies between cells, which can lead to well-documented
-inconsistencies \cite{dataflow,noworkflow,wrattler}.
-
-Ideas such as dependency tracking and efficient recomputation exist in visual data exploration tools
-\cite{control,tableau,vizdom} and scientific workflow systems \cite{taverna,kepler}. The
+cells and manually reevaluate cells. Many notebook systems are based on the REPL (read-eval-print-loop)%
+~\cite{lisp,drscheme} and do not track dependencies between cells, which can lead to well-documented
+inconsistencies~\cite{dataflow,noworkflow,wrattler}.
+
+Ideas such as dependency tracking and efficient recomputation exist in visual data exploration tools%
+~\cite{control,tableau,vizdom} and scientific workflow systems~\cite{taverna,kepler}. The
 implementation techniques are related to our work, but we focus on text-based scripts.
-Tempe \cite{tempe} focuses on streaming data, but provides a text-based scripting environment
+Tempe~\cite{tempe} focuses on streaming data, but provides a text-based scripting environment
 with automatic code update; its usability in contrast to REPLs has been empirically
-evaluated \cite{ripple}. The challenges faced by Tempe are more akin to dynamic software update~\cite{dsu}.
+evaluated~\cite{ripple}.
 
 \paragraph{Live and exploratory programming.}
-Live programming based on textual programs has been popularised by Victor \cite{principle,learnable} and
-is actively developed in domains such as live coded music \cite{beyond,sonic}.
+Live programming based on textual programs has been popularised by \citeauthor{learnable}~\cite{learnable,principle} and
+is actively developed in domains such as live coded music~\cite{beyond,sonic}.
+The notions of exploratory and live programming have been extensively studied~\cite{review}.
 The notion of exploratory programming has recently been analysed from the perspective of
-human-computer interaction \cite{exploratory}, which led to new tools \cite{variolite},
-complementary to our instant previews. Kubelka et al.~\cite{liveroad}
-review the use of live programming in a Smalltalk derived environment. Lighttable \cite{lighttable}
+human-computer interaction \cite{exploratory}, which led to new tools~\cite{variolite},
+complementary to our instant previews. \Citeauthor{liveroad}~\cite{liveroad}
+review the use of live programming in a Smalltalk derived environment. Lighttable~\cite{lighttable}
 and Chrome DevTools provide limited instant previews akin to those presented in this paper,
 but without well specified recomputation model. Finally, work on keeping state during
-code edits \cite{alive,livingit} would be relevant for supporting streaming data.
-
-A more principled approach can be used by systems based on structured editing \cite{structure-based,interactive,livenut,lamdu}
+code edits~\cite{alive,livingit} would be relevant for supporting streaming data.
+
+A more principled approach can be used by systems based on structured editing~\cite{structure-based,interactive,livenut,lamdu}
 where code is only modified via known operations with known effect on the computation graph
-(e.g.~``extract variable'' has no effect on the result; ``change constant value''
+(e.\hairspace g.~``extract variable'' has no effect on the result; ``change constant value''
 forces recomputation of subsequent code). This can be elegantly implemented using bi-directional
-lambda calculus \cite{hazelnut}, but it also makes us consider more human-centric abstractions
-\cite{subtext,directprog} further discussed in Section~\ref{sec:extra-abstraction}.
+lambda calculus~\cite{hazelnut}, but it also makes us consider more human-centric abstractions%
+~\cite{subtext,directprog} further discussed in \cref{sec:extra-abstraction}.
 
 \paragraph{Incremental computation and dependency analysis.}
-Work on self-adjusting and incremental computation \cite{selfadjusting,adapton} handles
+Work on self-adjusting and incremental computation~\cite{selfadjusting,adapton} handles
 recomputation when the program stays the same, but input changes.
-Most incremental systems, e.g.~\cite{incremental,adaptivefp,deltaml} evaluate the program and use
-programmer-supplied information to build a dependency graph, whereas our system uses static code analysis;
+Most incremental systems, e.\hairspace g.~\cite{incremental,adaptivefp,deltaml} evaluate the program and use
+programmer-supplied information to build a dependency graph, whereas our system uses static code analysis;%
 \cite{adapton} implement a small adaptive interpreter
 that treats code as changing input data, suggesting an implementation technique for
-live programming systems. Our use of dependency graphs \cite{dependencies} is static and first-order
-and can be seen as a form of program slicing \cite{slicing}, although our binding process is
-more directly inspired by Roslyn \cite{roslyn}, which uses it for efficient background type-checking.
+live programming systems. Our use of dependency graphs~\cite{dependencies} is static and first-order
+and can be seen as a form of program slicing~\cite{slicing}, although our binding process is
+more directly inspired by Roslyn~\cite{roslyn}, which uses it for efficient background type-checking.
 
 \paragraph{Semantics and partial evaluation.}
-The evaluation of previews is a form of partial evaluation \cite{partial}, done in a
+The evaluation of previews is a form of partial evaluation~\cite{partial}, done in a
 way that allows caching. This can be done implicitly or explicitly in the form of
-multi-stage programming \cite{metaml}. Semantically, the evaluation of previews can be seen as a
-modality \cite{modal} and delayed previews are linked to contextual modal type theory \cite{cmtt},
-formally modelled using comonads \cite{cmtt-denotation}.
+multi-stage programming~\cite{metaml}. Semantically, the evaluation of previews can be seen as a
+modality and delayed previews are linked to contextual modal type theory~\cite{cmtt},
+formally modelled using comonads~\cite{cmtt-denotation}.
 
@@
 \acks
-We thank Dominic Orchard, Stephen Kell, Roly Perera and Jonathan Edwards for many discussions
-about the work presented in the paper. Dominic Orchard provided invaluable feedback on earlier
-version of the paper. The paper also benefited from suggestions made by anonymous reviewers
-of The Programming Journal as well as reviwers of earlier versions of the paper.
-This work was partly supported by The Alan Turing Institute under the EPSRC grant EP/N510129/1.
-
+\begin{sloppypar}
+ We thank Dominic Orchard, Stephen Kell, Roly Perera and Jonathan
+  Edwards for many discussions about the work presented in the paper. Dominic
+  Orchard provided invaluable feedback on earlier version of the paper. The
+  paper also benefited from suggestions made by anonymous reviewers of The
+  Programming Journal as well as reviwers of earlier versions of the paper.
+  This work was partly supported by The Alan Turing Institute under the EPSRC
+  grant EP/N510129/1.
+\end{sloppypar}
 % ==================================================================================================
@@
 
-\newpage
 \appendix
@@
 \end{equation}
-
-\noindent
+%
+%
 The property holds because, first, both \rname{let} and \rname{external} decrease the \ident{size}
-of the program and, second, a program is either fully evaluated, i.e.~$o_1;\ldots;o_n$ for some $n$
+of the program and, second, a program is either fully evaluated, i.\hairspace e.~$o_1;\ldots;o_n$ for some $n$
 or, it can be reduced using one of the reduction rules.
@@
 Given a term $t$ that has no free variables, together with a lookup table $\Delta$ obtained
-from any sequence of programs using $\ident{bind-prog}$ (Figure~\ref{fig:binding-rules-prog}) and
-$\ident{update}$ (Figure~\ref{fig:loop}), then
+from any sequence of programs using $\ident{bind-prog}$ (\cref{fig:binding-rules-prog}) and
+$\ident{update}$ (\cref{fig:loop}), then
 let $\bndclr{v}, (V, E) = \ident{bind-expr}_{\emptyset, \Delta}(t)$. If $\bndclr{v}\Downarrow p$
@@
 over a sub-graph will also be valid over the new graph -- the newly added nodes and edges do not introduce
-non-determinism to the rules given in Figure~\ref{fig:eval}.
+non-determinism to the rules given in \cref{fig:eval}.
 
@@
 \begin{enumerate}
-\item[a.] If $FV(e)\,=\,\emptyset$ then $\bndclr{v} \Downarrow p$ for some $p$ and $C[p] \rightsquigarrow o$
-\item[b.] If $FV(e)\neq\emptyset$ then $\bndclr{v} \Downarrow \llbracket e_p \rrbracket_{FV(e)}$ for some $e_p$ and $C[e_p] \rightsquigarrow o$
+  \renewcommand{\theenumi}{\alph{enumi}}%
+\item If $FV(e)\,=\,\emptyset$ then $\bndclr{v} \Downarrow p$ for some $p$ and $C[p] \rightsquigarrow o$
+\item If $FV(e)\neq\emptyset$ then $\bndclr{v} \Downarrow \llbracket e_p \rrbracket_{FV(e)}$ for some $e_p$ and $C[e_p] \rightsquigarrow o$
 \end{enumerate}
@@
 
-\vspace{0.75em}\noindent(1) $\ident{bind-expr}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n))$ --
+\begin{enumerate}
+  \renewcommand*\labelenumi{\usekomafont{listlabel}\theenumi}
+  \renewcommand{\theenumi}{(\arabic{enumi})}%
+
+\item \label{itm:1} $\ident{bind-expr}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n))$ --
   Here $e = e_0.m(e_1, \ldots, e_n)$, $\bndclr{v_i}$ are graph nodes obtained by induction for
   expressions $e_i$ and $\{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v}, \bndclr{v_n}, \blbl{arg}(n))\} \subseteq E$.
-  From lookup inversion Lemma~\ref{thm:lemma-lookup}, $\bndclr{v} = \bnd{mem}(m, s)$ for some $s$.
+  From lookup inversion \cref{thm:lemma-lookup}, $\bndclr{v} = \bnd{mem}(m, s)$ for some $s$.
 
@@
   $\bndclr{v}\Downarrow p$ using \rname{mem-val} such that $p_0.m(p_1, \ldots, p_n) \rightsquigarrow p$.
-  From induction hypothesis and \emph{compositionality} of external libraries (Definition~\ref{def:external}),
+  From induction hypothesis and \emph{compositionality} of external libraries (\cref{def:external}),
   it holds that for any $C$ such that $C[e_0.m(e_1, \ldots, e_n)] \rightsquigarrow o$ for some $o$
@@
   $\bndclr{v}\Downarrow \llbracket e'_0.m(e'_1, \ldots, e'_n) \rrbracket_{FV(e)}$ using \rname{mem-expr}.
-  From induction hypothesis and \emph{compositionality} of external libraries (Definition~\ref{def:external}),
+  From induction hypothesis and \emph{compositionality} of external libraries (\cref{def:external}),
   it holds that for any $C$ such that $C[e_0.m(e_1, \ldots, e_n)] \rightsquigarrow o$ for some $o$
@@
 
-\vspace{0.75em}\noindent(2) $\ident{bind-expr}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n))$ --
-  This case is similar to (1), except that the fact that $\bndclr{v} = \bnd{mem}(m, s)$
-  holds by construction, rather than using Lemma~\ref{thm:lemma-lookup}.
-
-\vspace{0.75em}\noindent(3) $\ident{bind-expr}_{\Gamma, \Delta}(\lambda x\rightarrow e_b)$ --
+\item  $\ident{bind-expr}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n))$ --
+  This case is similar to \labelcref{itm:1}, except that the fact that $\bndclr{v} = \bnd{mem}(m, s)$
+  holds by construction, rather than using \cref{thm:lemma-lookup}.
+
+\item \label{itm:3} $\ident{bind-expr}_{\Gamma, \Delta}(\lambda x\rightarrow e_b)$ --
   Here $e = \lambda x\rightarrow e_b$, $\bndclr{v_b}$ is the graph node obtained by induction
   for the expression $e_b$ and $(\bndclr{v}, \bndclr{v_b}, \blbl{body}) \in E$.
-  From lookup inversion Lemma~\ref{thm:lemma-lookup}, $\bndclr{v} = \bnd{fun}(x, s)$ for some $s$.
+  From lookup inversion \cref{thm:lemma-lookup}, $\bndclr{v} = \bnd{fun}(x, s)$ for some $s$.
   The evaluation can use one of three rules:
@@
   \begin{enumerate}
-  \item[i.]
+    \renewcommand{\theenumii}{\roman{enumii}}%
+  \item
   If $FV(e)=\emptyset$ then $\bndclr{v_b}\Downarrow p_b$ for some $p_b$ and $\bndclr{v}\Downarrow \lambda x\rightarrow p_b$
@@
 
-  \item[ii.]
+  \item
   If $FV(e_b)=\{x\}$ then $\bndclr{v_b}\Downarrow \llbracket e'_b \rrbracket_x$ for some $e'_b$ and
@@
 
-  \item[iii.]
+  \item
   Otherwise, $\bndclr{v_b}\Downarrow \llbracket e'_b \rrbracket_{x,\Gamma}$ for some $e'_b$ and
@@
   \end{enumerate}
-
-\noindent
+%
+%
 For i.) and ii.) we show that a.) is the case; for iii.) we show that b.) is the case; that is
@@
 $C[\lambda x\rightarrow e'_b] \rightsquigarrow o$. For a given $C$, let $C'[-] = C[\lambda x\rightarrow -]$
-and use the induction hypothesis, i.e.~if $C'[e_b] \rightsquigarrow o$ for some $o$ then also $C'[e'_b] \rightsquigarrow o$.
-
-\vspace{0.75em}\noindent(4) $\ident{bind-expr}_{\Gamma, \Delta}(\lambda x\rightarrow e)$  --
-  This case is similar to (3), except that the fact that $\bndclr{v} = \bnd{fun}(x, s)$
-  holds by construction, rather than using Lemma~\ref{thm:lemma-lookup}.
-
-\vspace{0.75em}\noindent(5) $\ident{bind-expr}_{\Gamma, \Delta}(o)$ -- In this case $e=o$ and $\bndclr{v} = \bnd{val}(o)$
+and use the induction hypothesis, i.\hairspace e.~if $C'[e_b] \rightsquigarrow o$ for some $o$ then also $C'[e'_b] \rightsquigarrow o$.
+
+\item \label{itm:4} $\ident{bind-expr}_{\Gamma, \Delta}(\lambda x\rightarrow e)$  --
+  This case is similar to \labelcref{itm:3}, except that the fact that $\bndclr{v} = \bnd{fun}(x, s)$
+  holds by construction, rather than using \cref{thm:lemma-lookup}.
+
+\item $\ident{bind-expr}_{\Gamma, \Delta}(o)$ -- In this case $e=o$ and $\bndclr{v} = \bnd{val}(o)$
   and $\bnd{val}(o)\Downarrow o$ using \rname{val} and so the case a.) trivially holds.
 
-\vspace{0.75em}\noindent(6) $\ident{bind-expr}_{\Gamma, \Delta}(x)$ -- The initial $\Gamma$ is empty,
-  so $x$ must have been added to $\Gamma$ by case (3) or (4). Hence,
+\item $\ident{bind-expr}_{\Gamma, \Delta}(x)$ -- The initial $\Gamma$ is empty,
+  so $x$ must have been added to $\Gamma$ by case \labelcref{itm:3} or \labelcref{itm:4}. Hence,
   $\bndclr{v}=\bnd{var}(x)$, $\bndclr{v}\Downarrow \llbracket x \rrbracket_{x}$ using \rname{var}
   and so $e_p = e = x$ and the case b.) trivially holds.
-
+\end{enumerate}
 \end{proof}
@@
 \label{thm:sub-expr}
-Assume we have programs $p_1 = c_1; \ldots; c_k; K_c[e]; c_{k+1};\ldots; c_n$ and
+Assume we have programs $p_1, p_2$ such that $p_1 = c_1; \ldots; c_k; K_c[e]; c_{k+1};\ldots; c_n$ and
 $p_2 = c'_1; \ldots; c'_k; K'_c[e]; c'_{k+1};\ldots;c'_n$ and $I\subseteq \{1\ldots k\}$ such
@@
 Given any $\Delta$, assume that the the first program is bound,
-i.e.~$\bndclr{v_1};\ldots;\bndclr{v_n}, (V, E) = \ident{bind-prog}_{\emptyset, \Delta}(p_1)$,
+i.\hairspace e.~$\bndclr{v_1};\ldots;\bndclr{v_n}, (V, E) = \ident{bind-prog}_{\emptyset, \Delta}(p_1)$,
 the cache is updated $\Delta' = \ident{update}_{V, E}(\Delta)$ and the second
 program is bound,
-i.e.~$\bndclr{v'_1};\ldots;\bndclr{v'_n}, (V', E') = \ident{bind-prog}_{\emptyset, \Delta'}(p_2)$.
+i.\hairspace e.~$\bndclr{v'_1};\ldots;\bndclr{v'_n}, (V', E') = \ident{bind-prog}_{\emptyset, \Delta'}(p_2)$.
 
@@
 $e$ during the first and the second binding, respectively. Then, the graph nodes assigned to the
-sub-expression $e$ are the same, i.e.~$\bndclr{v} = \bndclr{v'}$.
+sub-expression $e$ are the same, i.\hairspace e.~$\bndclr{v} = \bndclr{v'}$.
 \end{lem}
@@
 delayed preview $\llbracket e_0.m(e_1) \rrbracket_{x, y}$. This operation is akin to expression
-splicing from meta-programming \cite{metaml,quotations}.
+splicing from meta-programming~\cite{metaml,quotations}.
 
 The semantics of delayed previews can be more formally captured by Contextual Modal Type Theory
-(CMTT) \cite{cmtt} and comonads \cite{cmtt-denotation}. In CMTT, $\lbrack \Psi \rbrack A$ denotes
+(CMTT)~\cite{cmtt} and comonads~\cite{cmtt-denotation}. In CMTT, $\lbrack \Psi \rbrack A$ denotes
 that a proposition $A$ is valid in context $\Psi$, which is similar to our delayed previews written
@@
 that would allow us to express the splicing operation used in \rname{mem-expr}. In categorical
-terms, the context-dependent proposition can be modelled as a graded comonad \cite{effectrev,graded}.
+terms, the context-dependent proposition can be modelled as a graded comonad~\cite{effectrev,graded}.
 The evaluation of a preview with no context dependencies (built implicitly into our evaluation rules)
@@
 Type checking of small programs is typically fast enough that no caching is necessary. However,
-The Gamma supports \emph{type providers} \cite{providers-fsharp,providers-idris}, which can
-generate types based on an external file or a REST service call, e.g.~\cite{fsdata}. For this
+The Gamma supports \emph{type providers}~\cite{providers-fsharp,providers-idris}, which can
+generate types based on an external file or a REST service call, e.\hairspace g.~\cite{fsdata}. For this
 reason, type checking can be relatively time consuming and can benefit from the same caching
@@
 Adding type checking requires revising the way we construct the dependency graph introduced in
-Section~\ref{sec:formal}. Previously, a variable bound by a lambda function had no dependencies.
+\cref{sec:formal}. Previously, a variable bound by a lambda function had no dependencies.
 However, the type of the variable depends on the context in which it appears. Given an expression
@@
 member of the expression represented by the target node. We also add $\blbl{callsite}(m, i)$
-as an edge from the node of the function. Figure~\ref{fig:graph-func} shows the
+as an edge from the node of the function. \Cref{fig:graph-func} shows the
 revised dependency graph for $o.m(\lambda x\rightarrow x)$. %A revised version of the algorithm
-%that constructs the dependency graph is included in Appendix~\ref{sec:app-types}.
+%that constructs the dependency graph is included in \cref{sec:app-types}.
 
@@
 
-The typing rules for variables, functions and member access are shown in Figure~\ref{fig:tc}.
-%Remaining rules are available in Appendix~\ref{sec:app-types}.
+The typing rules for variables, functions and member access are shown in \cref{fig:tc}.
+%Remaining rules are available in \cref{sec:app-types}.
 When type checking a member access \rname{mem}, we find its dependencies $\bndclr{v_i}$ and
@@
 The results of type checking can be cached and reused in the same way as instant previews, although
-we leave out the details. A property akin to correctness (Theorem~\ref{thm:correcntess})
+we leave out the details. A property akin to correctness (\cref{thm:correcntess})
 requires defining standard type checking over the structure of expressions, which we also
@@
 
-The data analysis by Financial Times in Section~\ref{sec:background-jupyter} illustrates why
+The data analysis by Financial Times in \cref{sec:background-jupyter} illustrates why
 notebook users often avoid abstraction. Wrapping code in a function makes it impossible to
@@
 \end{equation*}
-
-\noindent
+%
+%
 We introduce two new kinds of commands: a labelled sequence of commands and a reference to a label.
@@
 the block repeatedly. The correct use of dynamic scoping can be checked using
-coeffects \cite{coeffects}.
+coeffects~\cite{coeffects}.
 
@@
 intend to implement this mechanism in a future version of our data
-exploration environment (Section~\ref{sec:evaluation-case}).
+exploration environment (\cref{sec:evaluation-case}).
 

### 2020-03-12 wip-design/paper.tex
Design section

@@
 \newcommand{\kvd}[1]{\textbf{#1}}
-\newcommand{\ikvd}[1]{{\fontfamily{zi4}\selectfont #1}}
+\newcommand{\ikvd}[1]{{\fontfamily{zi4}\selectfont\small #1}}
 \newcommand{\tg}{The Gamma\xspace}
@@
 
+Both follow from assistant tools for software development \cite{assistants}. Our implementation,
+outlined in the previous section, implements iterative prompting through a code editor mechanism
+designed for code completion. Finally, the work on type providers \cite{inforich,fsdata,dotdriven},
+which we extend, also utilizes code completion.
+
 \textbf{Journalism.}
@@
 
-\section{Background and design goals}
+\section{Motivation}
 \label{sec:motivation}
@@
 \includegraphics[width=1\columnwidth]{figures/lords5}
+\vspace{-1.5em}
 
 \subsection{Querying via Iterative Prompting}
-The
+In the example, the journalists constructs a query that filters and sorts a data table. This is
+not unlike writing a query in the SQL language. However, the whole query is constructed through
+a single interaction mechanism that we refer to as \emph{iterative prompting}. The user triggers
+the interaction by typing `.' at the end of the query constructed so far. She is then offered a
+list of available operations in the current context. This may include a choice of top-level
+operations (e.g.~at the start and then again after choosing \ikvd{then}) or possible
+parameters for an operation (such as possible sorting keys and directions after choosing
+\ikvd{sort data}). The user then merely needs to choose one of the available options, rather than
+learning the syntax and semantics of SQL in order to know what constructs can be used in the
+given context.
+We discuss the important aspects of The Gamma design in the next section and then discuss the
+integration of specific data sources in detail, revisiting the type provider for querying
+data tables.
 
-\newpage
+% ==================================================================================================
 
@@
 
-interaction = iterative prompting
-how is this different from just auto-complete?
+We identified a number of requirements for a data exploration tool for journalists earlier.
+First, using the tool should result in transparent and reproducible data analyses that can be
+understood and modified by non-experts. Second, the tool should minimze the number of concepts
+that the user needs to understand and, subsequently, allow them to learn through exploration
+and from examples. In this section, we discuss how our design aims to satisfy those requirements.
 
-trace = for transparency and learning
+\subsection{Lowering the Barrier to Entry}
+Data exploration tasks, such as querying tables, have a certain irreducible complexity.
+Regardless of the interface, the user will be exposed to concepts such as filtering,
+sorting and grouping. Our design aims to lower the barrier to entry by stratifying the concepts
+into first-level \emph{iterative prompting} principle and second-level \emph{domain specific
+languages} for each kind of data source.
+The user needs to master the iterative prompting principle in order to start working
+with the system. However, using individual data sources can be mastered gradually.
 
-One important observation from the above list is that our tool should be accessible to non-expert
-users such as readers and non-technical journalists, while providing extra capabilities for more
-technical users. It should be easy to use if one just wants to modify existing code and should
-encourage experimentation. Considering these challenges, we identify the following technical
-design principles. todo{There must be papers on learning programming that can be referenced
-here.}
+\paragraph{Iterative Prompting}
+Iterative prompting is an interaction principle where the user repeatedly invokes an
+auto-complete prompt and makes a selection from the offered options. The principle is closely
+related to both code completion and the use of command line. Iterative prompting is novel as an
+overarching interaction principle for program construction.
 
-\paragraph{B1. Learning from examples and by experimentation.} We should support two ways of learning.
-Users of tools such as spreadsheets
-often learn by looking at existing problem solutions todo{Advait's PPIG}. Our design should allow
-this by making it possible to inspect and retrace steps used while solving a problem in an existing
-application. Another principle of spreadsheets that we want to keep is the ability to experiment
-and see results immediately. Our design should allow users to try invoking an operation or modifying
-a parameter and quickly see if this leads to the desired results.
+In code completion, the auto-complete
+prompt is invoked only in certain contexts, e.g.~when accessing a member of an object through a
+variable defined earlier in code. It requires the user to be sufficiently familiar with the
+programming language in order to get to a point where they are offered a list of members. In
+contrast, our system only requires choosing the initial data source. The rest of the programming
+is done via iterative prompting. The iterative nature of iterative prompting makes it similar to
+using the command line or REPL (read-eval-print-loop) tools. However, those repeatedly ask the
+user to type code or commands.
 
-\paragraph{B2. Choice over construction.} To minimize the amount of information that users have to
-learn and remember, our system should work in a way that allows constructing programs by
-choosing from options that can reasonably appear in a current context, rather than requiring
-users to recall particular syntax or exact identifier name.
-todo{recognition over recall?}
+We argue that iterative prompting is easier than other forms of interaction, because it
+follows the \emph{recognition over recall} usability heuristic. The users are only required to
+choose from an offered list of options, rather than having to recall a possible command
+(to type in a command line) or a syntax in a text-based programming environment.
 
-\paragraph{B3. Make simple things easy and complex things possible.} Some users of the system may, over time, become advanced users
-and the system should support those. In other words, the upper bound on what can be achieved
-should be well above the most common use cases. At the same time, the complex features that power users might
-need should not affect the most elementary uses of the system and should remain completely hidden
-until needed. In other words, the lower bound on what one needs to know to use the system for basic
-tasks should be as low as possible. todo{I think I got this idea of "boundaries" on what
-is possible from some paper, but cannot recall which...}
+\paragraph{Domain Specific Languages}
+The access to individual data sources in The Gamma is facilitated through a domain specific
+language, which defines the primitives that are offered to the user in the auto-complete prompts
+invoked through iterative prompting. The domain specific languages are embedded in The Gamma --
+they define merely the available members (and specify how to evaluate a chain of members), but
+they cannot define any custom syntax.
 
-\paragraph{B4. Visibility of state.} To support transparency, the system should make its entire state
-transparent -- when reviewing a data analysis, all parameters should be immediately visible and the
-user should not need to, e.g.~navigate through complex user interface to find them.
+We discuss the individual domain specific languages for querying data tables, graph databases and
+data cubes in a later section. The complexity of those languages differs. The previously discussed
+language for querying tabular data is the most complex one. However, The Gamma makes it easy for
+the user to start exploring and learning new languages, because they are all accessible via
+iterative prompting.
+
+\subsection{Building the Magic Escalator of Knowledge}
+As discussed earlier, The Gamma is designed for users who work under tight deadlines and only
+analyse data as their secondary task. The Gamma aims to support such users by having a low
+barrier for entry and making it easy to learn independently.
+
+\paragraph{Design for Percolation}
+When analysing how Excel users learn Sarkar \cite{learning} points out that users learn new
+features opportunistically when the usage of a feature is apparent in a spreadsheet. For
+example, users can learn different functions to use in formulas, because those are visible in
+the cell. Learning how to use a wizard for creating charts in this way is not possible because
+it leaves no trace in the spreadsheet -- only the final result. Sarkar's recommendation is to
+\emph{design for percolation}, i.e.~in a way where looking at the final result makes it apparent
+what feature has been used and how.
+
+\paragraph{Text-based Source Code}
+In The Gamma, each step in the iterative prompting process results in an identifier that is
+added to the source code. This means that a program constructed solely through iterative prompting
+keeps a full trace of how it was created. Seeing the resulting source code provides the user all
+information that they need to recreate the program, not just by copying it, but also by using the
+iterative prompting mechanism. The Gamma represents code as text and allows the user to edit it
+freely, so not all interactions leave a trace. For example, deleting the most recently selected
+option and choosing a different one is not apparent from the result.
+
+\subsection{Making Complex Things Possible}
+Although the focus on The Gamma is simple data exploration and almost all the use cases we
+envision can be achieved using the iterative prompting interaction principle, there remain a
+few cases where more flexibility is needed. We aim to \emph{make simple things easy and
+complex things possible}. The language thus supports a small number of additional constructs that
+can only be used through text editing. The following example illustrates three additional
+constructs:
+
+\begin{thegamma}
+\kvd{let} topTravel =
+  expenses
+    .'sort data'.'by Travel Costs descending'.then
+    .paging.take(5).'get series'
+    .'with key Name'.'and value Travel Costs'
+
+charts.column(topTravel)
+  .setTitle("House of Lords members by travel expenses")
+  .setColors(["red","blue","green"])
+\end{thegamma}
+
+First, The Gamma allows operations with parameters such as \ikvd{take(3)} or \ikvd{setTitle("...")}.
+This is currently needed when writing a query that skips or takes first N elements from a table.
+The remaining features are not needed for basic data exploration, but allow other scenarios, such
+as manual creation of charts. The \ikvd{let} construct can be used to define (immutable) variables
+and The Gamma also supports lists written as \ikvd{[1,2,3]}.
+
+% ==================================================================================================
 
@@
 
-\subsection{Type providers}
+\subsection{Data sources}
+
+% ==================================================================================================
 
@@
 
+"i haven't done 'then' "
+
+USABILITY ISSUES
+- need to delete the rest of the line
+- when you leave ".c" there, things break
+- typing . without indentation does not work
+- purple box vs. black wrench icons in the completion
+- how do you know top level data sources
+- 'games is' bit (with the year) - methods suck!
+
 \section{Discussion}

### 2020-03-14 wip-design/paper.tex
Case study evaluation

@@
 \usepackage{xcolor}
+\usepackage{tikz}
+\newcommand*{\priority}[1]{\begin{tikzpicture}[scale=0.12]%
+    \draw (0,0) circle (1);
+    \fill[fill opacity=1,fill=black] (0,0) -- (90:1) arc (90:90-#1*3.6:1) -- cycle;
+    \end{tikzpicture}}
 
@@
 
+%https://www.nngroup.com/articles/recognition-and-recall/
 We argue that iterative prompting is easier than other forms of interaction, because it
@@
 
+\newpage
+~
+\newpage
+
+% ==================================================================================================
+
 \section{User study}
@@
 
+business team of a non-profit research organization, including a former journalist
+
+\subsection{Research questions}
+
+\paragraph{RQ1: Can end users explore data with The Gamma?}
+
+\paragraph{RQ2: Can knowledge transfer between data sources?}
+
+\paragraph{RQ3: Can users learn from just code samples?}
+
+\subsection{Study design}
 this is between-user user study
 
-FIRST: How one area transfers to another?
+\subsection{Results}
+it is good
 
-DEMO: WorldBank
-explain everything
+\paragraph{Can end users explore data with The Gamma?}
+Yes, the table shows the results.
+People who spent between 10 and 25 minutes (average 18 minutes) working with The Gamma
+Out of 13 people, all completed some work, 1 person completed partly and 6 were able
+  to complete the task with small assistance (related to methods and syntax, as discussed below)
 
-DISPLAY: Public order and safety, Defence
-\begin{verbatim}
-expenditure.byService.'Public order and safety'.inTermsOf.GDP
-expenditure.byService.Defence.inTermsOf.GDP
-\end{verbatim}
+\begin{quote}
+\emph{``This is actually pretty simple to use. You think about the logic of what you're actually asking and
+    then you try to get it into the format you can. But knowing where it comes from would tell you how
+    to trust it.''}
+\end{quote}
 
-DEMO: Graph DB
-explain everything
+we should use it as a training resource for students - would students get this?
 
-\begin{verbatim}
-drWho.Character.Doctor.'ENEMY_OF'.'[any]'
-  .'APPEARED_IN'.'[any]'.'explore_properties'.explore
-  .'group data'.'by 1-name'.'count distinct 2-title'
-\end{verbatim}
+\begin{quote}
+\emph{``I don't think they'd need more than 5 minute video. This is a data
+  source, this is what's in there.''}
+\end{quote}
 
-DISPLAY: Who has larges travel expenses?
-(and in London only?)
+\begin{quote}
+\emph{``for somebody who does not do coding or programming, this does not feel that daunting. it's not like you're giving me large screen full of code, which is reassurring" (~36)''}
+\end{quote}
 
-\begin{verbatim}
-lords.'sort data'.'by Travel Costs descending'
+\paragraph{RQ2: Can knowledge transfer between data sources?}
+expenditure task (#1-#5) and lords task (#9-#11) - is transfer from worldbank to expenditure and
+from graph data to lords; the former transfers more directly, the latter less so (mainly because
+of then and methods, which we discuss below)
 
-lords
-  .'filter data'.'County is'.London.then
-  .'sort data'.'by Travel Costs descending'
-\end{verbatim}
+\begin{quote}
+  \emph{``I found it quite easy to translate what you showed us in the demo to the new dataset, i though it was quite easy to just get how that works''}
+\end{quote}
 
-SECOND: No experts are needed
+\paragraph{RQ3: Can users learn from just code samples?}
+in the worldbank task (#6-#9), users were given just a code sample and demo using no real
+data source. They were able to complete based on just following and modifying code sample (to find other data).
+(no problems they had were related to data source)
 
-DEMO: explain how live preview works, explain how '.' works,
-explain how newlines and indentation work
+- "a video would just be this [code samples] anyway" (~19)
+- "keystroke by keystroke video would be good"
 
-DISPLAY 'CO2 emissions (metric tons per capita)'
+\begin{quote}
+  \emph{``I think providing one video of one example is good and maybe a couple of little examples of code wher people can see the kind of things you can do" (28:02)''}
+\end{quote}
 
-\begin{verbatim}
-worldbank.byCountry.'United Kingdom'
-  .'Economy & Growth'.'GDP per capita (current US$)'
+"more than one example of code would be useful because I'd be less confused about options"
 
-worldbank.byCountry.Germany
-  .'Economy & Growth'.'GDP per capita (current US$)'
+\subsection{Further observations}
 
-worldbank.byCountry.'Czech Republic'
-  .'Economy & Growth'.'GDP per capita (current US$)'
-\end{verbatim}
+\paragraph{Making complex things possible hurts}
+#11, #12, #13 all struggled with methods (they show in the completion lists)
 
-DISPLAY: Top athletes from London
+\paragraph{Benefits and drawbacks of text}
 
-\begin{verbatim}
-olympics
-  .'group data'.'by Team'.'sum Gold'.then
-  .'sort data'.'by Gold descending'.then
-  .paging.take(5)
-  .'get series'.'with key Team'.'and value Gold'
-\end{verbatim}
+AG - confused by indentation (Monaco inserts it automatically - duh!)
+it does not work if you have it and add a new to-level thing;
 
-THIRD: 'then'
+BM uses copy & paste to copy thing and replace one identifier in the middle
+(which would be difficult to do with non-textual representation)
 
-DEMO: Show worldbank
-GIVE: Commented source code using 'olympics'
-DISPLAY: Top athletes from London (think-aloud)
+people occasionally miss '.' when typing and try to use a space
 
-QUESTIONS
+\paragraph{How users understand the then keyword}
 
-1) Did I tell you enough in the introduction to get started?
+what is then: "allows us to chain together the operations"
+is it special? "maybe I need it just to split over multiple lines" (experiments..) "no!"
 
-Say we want to provide educational materials for journalists
-(with limited budgets), what would be the most important?
+"when you explained about the "dot then" that was a really useful thing to know.
+When I found that, I was like 'this is fine, this is doable'.
+If I knew this from the start, it would [have been easier]" (~27)
 
-2) Video or just code samples?
+"I thought 'then' starts a new line. (...) I realised there was no 'then' after the 'paging' command. "
+"It starts a new line or a new query? I assumed there would be one after 'paging' too."
 
-We'll have more data sources than we can write tutorials for
+\subsection{Conclusions}
+(former journalist)
+Do you think this is something journalists could figure out how to use?
+"Yeah, I think so. There's a lot of effort going into data journalism that programming could make much quicker,
+but I was always nervous about code." ... "Something like this would really simplify things."
 
-3) Do we just teach them how the environment works?
+\begin{table}
+  \centering
+  \begin{tabular}{l l l c l}
+    \toprule
+      & {\small \textit{Task}}
+      & {\small \textit{Kind}} & {\small \textit{Completed}}
+      & {\small \textit{Notes}} \\
+    \midrule
+    \small \#1 & \small expenditure & \small cube & \priority{50} & {\small Obtained one of two data series}\\
+    \small \#2 & \small expenditure & \small cube & \priority{100} & {\small Explored furhter data series }\\
+    \small \#3 & \small expenditure & \small cube & \priority{100}& {\small Explored further data series }\\
+    \small \#4 & \small expenditure & \small cube & \priority{75}& {\small With hint to use another member }\\
+    \small \#5 & \small expenditure & \small cube & \priority{100}& {\small Explored further data series }\\
+    \small \#6 & \small worldbank & \small cube & \priority{75} & {\small With general syntax hint }\\
+    \small \#7 & \small worldbank & \small cube & \priority{100} & {\small Completed very quickly }\\
+    \small \#8 & \small worldbank & \small cube & \priority{100} & {\small Extra time to find data }\\
+    \small \#9 & \small lords & \small table & \priority{75} & {\small  Struggled with composition}\\
+    \small \#10 & \small lords & \small table & \priority{100} & {\small Completed very quickly }\\
+    \small \#11 & \small lords & \small table & \priority{75} & {\small With a hint to avoid methods}\\
+    \small \#12 & \small olympics & \small table & \priority{75}  & {\small With a hint to avoid methods}\\
+    \small \#13 & \small olympics & \small table & \priority{75}  & {\small With a hint to avoid methods}\\
+    \bottomrule
+  \end{tabular}
+  \caption{Table captions should be placed below the table. We
+    recommend table lines be 1 point, 25\% black. Minimize use of
+    table grid lines.}~\label{tab:table1}
+\end{table}
 
-4) What do we need to teach about a data source?
+\newpage
+~
+\newpage
 
-"i haven't done 'then' "
-
-USABILITY ISSUES
-- need to delete the rest of the line
-- when you leave ".c" there, things break
-- typing . without indentation does not work
-- purple box vs. black wrench icons in the completion
-- how do you know top level data sources
-- 'games is' bit (with the year) - methods suck!
+% ==================================================================================================
 

### 2020-03-21 wip-design/paper.tex
Evaluation

@@
     \end{tikzpicture}}
+\newcommand*{\priorityc}[1]{
+\resizebox{0.8em}{0.8em}{
+  {\protect\tikz{ \protect
+    \draw[line width=1mm, black] (0,0) circle (1); \fill[fill opacity=1,fill=black] (0,0) -- (90:1) arc (90:90-#1*3.6:1) -- cycle;
+    } }}}
 
@@
 
-\subsection{Data sources}
+\subsection{Data cube}
+
+\subsection{Tabular data}
+
+\subsection{Graph database}
 
@@
 
-business team of a non-profit research organization, including a former journalist
+Our goal is to develop an easy-to-learn tool that journalists and other non-programmers can use
+for producing transparent data analyses. To evaluate the extent to which The Gamma achieves this,
+we conduct a user study. We recruit participants among the operations and business team of a
+non-profit research organization and investigate three research questions.
 
-\subsection{Research questions}
-
-\paragraph{RQ1: Can end users explore data with The Gamma?}
+\paragraph{RQ1: Can non-programmers explore data with The Gamma?}
+We give participants one of four simeple data exploration tasks and see if they can complete the
+task using The Gamma and, possibly, how much assistance they need.
 
 \paragraph{RQ2: Can knowledge transfer between data sources?}
+Our first hypothesis is that users familiar with the iterative prompting principle will be able to
+use an unfamiliar data source. In two of the tasks, participants are shown one data source and
+then asked to work with a different one.
 
 \paragraph{RQ3: Can users learn from just code samples?}
+Our second hypothesis is that users can learn through precolation, i.e.~by looking at the source
+code of published analyses. In one of our tasks, participants are given a code sample, but only a
+minimal explanation of The Gamma environment.
 
 \subsection{Study design}
-this is between-user user study
+We perform between-subjects study to evaluate the first experience of using The Gamma.
+We recruit 13 participants (5 male, 8 female) from a business team
+of a research lab working in various non-technical roles (project management,
+partnerships, communications) including one former journalist.
+
+We split participants into 4 groups. We first gave a brief 5 minute demonstration of The
+Gamma and then asked participants to complete a task. Finally, we conducted a short semi-structured
+group interview and later sent participants a follow-up questionnaire. The four tasks were:
+
+\begin{itemize}
+\item \emph{Expenditure.} Participants were given a demo using \emph{worldbank}.
+  They were asked to use the \emph{expenditure} data source to compare the UK government spending
+  on ``Public order and safety'' and ``Defence'' in terms of GDP.
+\item \emph{Lodrs.} Participants were given a demo using \emph{worldbank}.
+  They were asked to use the \emph{lords} data source (a table with House of Lords
+  members expenses) to find a members representing London with the highest travel costs.
+\item \emph{Worldbank.} Participants were given a minimal explanation of The Gamma environment and
+  a code sample using \emph{worldbank}. They were asked to solve a different task using
+  the \emph{worldbank} data source.
+\item \emph{Olympics.} Participnts were given a demo using \emph{olympics}.
+  They were asked to solve a more complex problem using the same data source.
+\end{itemize}
+
+We let participants work independently, but offered guidance if they got stuck for longer period of
+time. Tasks \emph{expenditure} and \emph{lords} aim to answer the questions RQ1 and RQ2; the task
+\emph{worldbank} aims to answer RQ1 and RQ3. In \emph{olympics}, we test RQ1 using a more complex
+data source and we ask further questions to explore their understanding of the data source.
+
+\begin{table}
+  \centering
+  \begin{tabular}{l l l c l}
+    \toprule
+      & {\small \textit{Task}}
+      & {\small \textit{Kind}} & {\small \textit{Done}}
+      & {\small \textit{Notes}} \\
+    \midrule
+    \small \#1 & \small expenditure & \small cube & \priority{50} & {\small Obtained one of two data series}\\
+    \small \#2 & \small expenditure & \small cube & \priority{100} & {\small Explored furhter data series }\\
+    \small \#3 & \small expenditure & \small cube & \priority{100}& {\small Explored further data series }\\
+    \small \#4 & \small expenditure & \small cube & \priority{75}& {\small With hint to use another member }\\
+    \small \#5 & \small expenditure & \small cube & \priority{100}& {\small Explored further data series }\\
+    \small \#6 & \small worldbank & \small cube & \priority{75} & {\small With general syntax hint }\\
+    \small \#7 & \small worldbank & \small cube & \priority{100} & {\small Completed very quickly }\\
+    \small \#8 & \small worldbank & \small cube & \priority{100} & {\small Extra time to find data }\\
+    \small \#9 & \small lords & \small table & \priority{75} & {\small  Struggled with composition}\\
+    \small \#10 & \small lords & \small table & \priority{100} & {\small Completed very quickly }\\
+    \small \#11 & \small lords & \small table & \priority{75} & {\small With a hint to avoid operations}\\
+    \small \#12 & \small olympics & \small table & \priority{75}  & {\small With a hint to avoid operations}\\
+    \small \#13 & \small olympics & \small table & \priority{75}  & {\small With a hint to avoid operations}\\
+    \bottomrule
+  \end{tabular}
+  \caption{Overview of the work completed by individual participants. The marks denote:
+    \priorityc{75} = required some guidance, \priorityc{50} = partially completed.}
+  \label{tab:tasks}
+  \vspace{-1em}
+\end{table}
 
 \subsection{Results}
-it is good
+Table~\ref{tab:tasks} summarizes the work done by the study participants.
+For each participant, we record the task, the kind of data source used in the task and the
+level of completion. For participants who needed assistance, the notes section details the help
+given. We analyse the hints in the next section. Table~\ref{tab:quest} shows the results of the
+follow-up questionnaire.
 
-\paragraph{Can end users explore data with The Gamma?}
-Yes, the table shows the results.
-People who spent between 10 and 25 minutes (average 18 minutes) working with The Gamma
-Out of 13 people, all completed some work, 1 person completed partly and 6 were able
-  to complete the task with small assistance (related to methods and syntax, as discussed below)
+\paragraph{RQ1: Can non-programmers explore data with The Gamma?}
+Three results allow us to answer RQ1 in the affirmative. First, 7 out of 9
+participants agree or strongly agree that they ``found the system easy to use''. Second,
+participants spent between 10 and 25 minutes (average 17 minutes) working with The Gamma and
+all 12 out of 13 participants completed the task; 6 participants required some assistance,
+but 3 of those faced one issue (discussed later) that could be covered in the introductory
+presentation. Third, a number of participants shared positive comments in the semi-structured
+interview.
+
+Participant \#3 found the system simple, but also points out an issue about data provenance,
+which we discuss later:
 
 \begin{quote}
-\emph{``This is actually pretty simple to use. You think about the logic of what you're actually asking and
-    then you try to get it into the format you can. But knowing where it comes from would tell you how
-    to trust it.''}
+\emph{``This is actually pretty simple to use. You think about the logic of what you're actually
+  asking and then you try to get it into the format you can. But knowing where it comes from
+  would tell you how to trust it.''}
 \end{quote}
 
-we should use it as a training resource for students - would students get this?
+Similarly, participant \#2, notes that The Gamma alleviated their unease about code:
 
 \begin{quote}
-\emph{``I don't think they'd need more than 5 minute video. This is a data
+\emph{``For somebody who does not do coding or programming, this does not feel that daunting.
+  It's not like you're giving me large screen full of code, which is reassurring.''}
+\end{quote}
+
+Finally, participant \#5 suggested the system could be used as an educational tool for teaching
+critical thinking with data. They answer a follow-up question about what training materials would
+the students need as follows:
+
+\begin{quote}
+\emph{``I don't think they'd need more than 5 minute video (..) this is the data
   source, this is what's in there.''}
@@
 
+\paragraph{RQ2: Can knowledge transfer between data sources?}
+Our study does not conclusively answer RQ2. There is some evidence in favor of a positive answer.
+In the practical part, two of the tasks (\emph{expenditure} and \emph{lords}) used a different data
+source in the introductory presentation than the one that the participants were asked to use.
+Participants were able to complete those tasks, although \emph{lords} has been more challenging
+as it involves a more complex data source. In the interview, participant \#2 also gives a positive
+answer:
+
 \begin{quote}
-\emph{``for somebody who does not do coding or programming, this does not feel that daunting. it's not like you're giving me large screen full of code, which is reassurring" (~36)''}
+\emph{``I found it quite easy to translate what you showed us in the demo to the new dataset.
+   I though it was quite easy to just get how that works.''}
 \end{quote}
 
-\paragraph{RQ2: Can knowledge transfer between data sources?}
-expenditure task (#1-#5) and lords task (#9-#11) - is transfer from worldbank to expenditure and
-from graph data to lords; the former transfers more directly, the latter less so (mainly because
-of then and methods, which we discuss below)
+Negative evidence is offered by the follow-up questionnaire. When asked
+whether they would know how to approach a task using a new data source, 5 out of 9 participants
+diagree or strongly disagree that they would know how to approach it without any further guidance.
+Participants did not believe that knowledge can easily transfer to another
+data source.
+
+\paragraph{RQ3: Can users learn from just code samples?}
+A positive answer to RQ3 is supported by the \emph{worldbank} task results, the follow-up questionnaire
+and interview comments. In the task, participants were given only a
+minimal demo of the iterative prompting principle together with print-out of 2 code samples.
+All three were able to complete a related task using the same data source. In the follow-up
+questionnaire, only 2 out of 9 participants disagree or strongly disagree that
+they would know how to approach a task using an unfamiliar data source when given ``a number
+of code samples''.
+
+In the semi-structured interview, participants were asked what would be the most useful format
+of educational material about The Gamma (code samples, video tutorials, etc.).
+Participant \#7 noted that \emph{``a video would just be this [i.e.~a code sample] anyway''}, while
+participant \#13 (former journalist) answered:
 
 \begin{quote}
-  \emph{``I found it quite easy to translate what you showed us in the demo to the new dataset, i though it was quite easy to just get how that works''}
+  \emph{``I think providing one video of one example is good and maybe a couple of little examples
+  of code wher people can see the kind of things you can do.''}
 \end{quote}
 
-\paragraph{RQ3: Can users learn from just code samples?}
-in the worldbank task (#6-#9), users were given just a code sample and demo using no real
-data source. They were able to complete based on just following and modifying code sample (to find other data).
-(no problems they had were related to data source)
-
-- "a video would just be this [code samples] anyway" (~19)
-- "keystroke by keystroke video would be good"
-
-\begin{quote}
-  \emph{``I think providing one video of one example is good and maybe a couple of little examples of code wher people can see the kind of things you can do" (28:02)''}
-\end{quote}
-
-"more than one example of code would be useful because I'd be less confused about options"
+This is aligned with our design goal. Once the user understands the iterative prompting principle
+(which can be done through a video tutorial), they can learn how to use any specific data source
+just from code samples.
 
 \subsection{Further observations}
+In this section, we briefly discuss a number of observations about The Gamma design that
+emerged from the study, some of which suggest ways of improving the system.
 
-\paragraph{Making complex things possible hurts}
-#11, #12, #13 all struggled with methods (they show in the completion lists)
+\paragraph{Making complex things possible may hurt}
+As discussed earlier, The Gamma language supports operations such as \ikvd{take(5)}. Most type
+providers never generate those, but the provider for working with tabular data is an exception.
+When filtering data, the provider allows specifying a condition on numerical attributes such as
+\ikvd{olympics.\textquotesingle filter data\textquotesingle.\textquotesingle Year is greater than\textquotesingle(2004)}.
+
+Three participants (\#11, \#12, \#13) struggled to complete a task, because they
+initially attempted to use those operations. However, those violate the iterative prompting
+principle as one cannot type `.' after \ikvd{\textquotesingle Year is greater than\textquotesingle}.
+This suggests that we should avoid operations in type providers for data access, even if it
+limits the expressive power of the type provider.
 
@@
 
+\begin{table}
+  \centering
+  \begin{tabular}{p{19em} c c}
+    \toprule
+      {\small \textit{Questionnaire item}} & {\small \textit{Avg}} & {\small \textit{Sdv}} \\
+    \midrule
+    \small Considering your experience \& possible use of The Gamma:\\
+    \small \quad I found the system easy to use. & \small 2.11 & \small 0.87\\
+    \small \quad Journalists will be able to use it to analyse data. & \small 2.67 & \small 1.05\\
+    \small \quad Readers will be able to critically examine the analyses. & \small 2.44 & \small 0.68\\
+    \small Thinking about the usability of the system:\\
+    \small \quad The resulting programs are easy to understand. & \small 2.33 & \small 0.82\\
+    \small \quad The text editor for creating them was easy to use. & \small 2.33 & \small 0.94\\
+    \small \quad Finding the right item in the auto-complete was easy. & \small 2.56 & \small 1.42\\
+    \small I would know how to approach a task using a new data source\\
+    \small \quad If given a comprehensive video tutorial. & \small 2.00 & \small 1.24\\
+    \small \quad If given a number of code samples. & \small 2.78 & \small 1.13\\
+    \small \quad Without any further guidance. & \small 3.56 & \small 1.17\\
+    \bottomrule
+  \end{tabular}
+  \caption{Summary of follow-up questionnaire responses using
+    a 5-point Likert scale (1 = strongly agree, 5 = strongly disagree) over 9 subjects.}
+  \label{tab:quest}
+\end{table}
+
 \subsection{Conclusions}
@@
 but I was always nervous about code." ... "Something like this would really simplify things."
-
-\begin{table}
-  \centering
-  \begin{tabular}{l l l c l}
-    \toprule
-      & {\small \textit{Task}}
-      & {\small \textit{Kind}} & {\small \textit{Completed}}
-      & {\small \textit{Notes}} \\
-    \midrule
-    \small \#1 & \small expenditure & \small cube & \priority{50} & {\small Obtained one of two data series}\\
-    \small \#2 & \small expenditure & \small cube & \priority{100} & {\small Explored furhter data series }\\
-    \small \#3 & \small expenditure & \small cube & \priority{100}& {\small Explored further data series }\\
-    \small \#4 & \small expenditure & \small cube & \priority{75}& {\small With hint to use another member }\\
-    \small \#5 & \small expenditure & \small cube & \priority{100}& {\small Explored further data series }\\
-    \small \#6 & \small worldbank & \small cube & \priority{75} & {\small With general syntax hint }\\
-    \small \#7 & \small worldbank & \small cube & \priority{100} & {\small Completed very quickly }\\
-    \small \#8 & \small worldbank & \small cube & \priority{100} & {\small Extra time to find data }\\
-    \small \#9 & \small lords & \small table & \priority{75} & {\small  Struggled with composition}\\
-    \small \#10 & \small lords & \small table & \priority{100} & {\small Completed very quickly }\\
-    \small \#11 & \small lords & \small table & \priority{75} & {\small With a hint to avoid methods}\\
-    \small \#12 & \small olympics & \small table & \priority{75}  & {\small With a hint to avoid methods}\\
-    \small \#13 & \small olympics & \small table & \priority{75}  & {\small With a hint to avoid methods}\\
-    \bottomrule
-  \end{tabular}
-  \caption{Table captions should be placed below the table. We
-    recommend table lines be 1 point, 25\% black. Minimize use of
-    table grid lines.}~\label{tab:table1}
-\end{table}
 
@@
 \section{Discussion}
+
+Data provenance (where data comes from)
+What are the data sources availbale (how to start)?
 

### 2020-03-22 wip-design/paper.tex
Working on implementation section

@@
 \label{sec:implementation}
+The Gamma consists of a programming language with a web-based coding environment and a number
+of type providers that provide access to various kinds of data sources. The system is implemented
+in F\# and compiled to JavaScript using Fable.
 
-\subsection{Language}
+\subsection{The Gamma Language}
+In The Gamma, a program is a sequence of commands. A command can be either a variable declaration
+or an expression that evaluates to a value such as a data table or a chart.
+An expression is a reference to a data source followed by a chain of member accesses.
+A member can be either an ordinary member or an operation which takes a list of parameters
+enclosed in parentheses. Our implementation supports a few other features, but those
+are not used in this paper.
 
-\subsection{Data cube}
+The language uses a type system to infer what members are available at a given point in a chain
+of members. The type is an object with a list of members that, in turn, have their own types.
+The types are not built-in, but are generated by a type provider for a particular data source.
+The types are also generated lazily. A type is only computed when the corresponding member
+is actually used in code.
 
-\subsection{Tabular data}
+The web-based coding environment uses the Monaco\footnote{\url{https://microsoft.github.io/monaco-editor/}}
+editor and implemts auto-completion feature that is triggered when the user types `.' and relies
+on information from the type system. The can display the value of a current command inline
+as in Figure~\ref{fig:thegamma}, or values of all commands on the side as in Figure~\ref{fig:sidebyside}.
+In the latter case, it can also combine one or more data series into a single chart.
 
-\subsection{Graph database}
+\subsection{Data Cube Type Provider}
+
+\subsection{Tabular Data Type Provider}
+
+\subsection{Graph Database Type Provider}
+
+\newpage
+
+\begin{figure}
+\includegraphics[width=1\columnwidth]{figures/sidebyside}
+\caption{The Gamma environment with an automatically generated chart comparing CO2 emissions
+of China and United States.}
+\label{fig:sidebyside}
+\vspace{-0.5em}
+\end{figure}
 
 % ==================================================================================================
+
+\newpage
+~
+\newpage
 
@@
     \small \#12 & \small olympics & \small table & \priority{75}  & {\small With a hint to avoid operations}\\
-    \small \#13 & \small olympics & \small table & \priority{75}  & {\small With a hint to avoid operations}\\
+    \small \#13 & \small olympics & \small table & \priority{75}  & {\small Hints about `then' and operations}\\
     \bottomrule
@@
 \paragraph{Benefits and drawbacks of text}
+The Gamma is based on text to aid transparency. Text hints that there is no hidden state and
+the reader sees the full code. The study suggests that using a text editor has both
+benefits and drawbacks compared to alternatives such as structured editors~\cite{structure-based,livenut,lamdu}.
+Most participants had no difficulty navigating around source code, making edits or deleting code
+fragments, which is arguably harder in an unfamiliar structured editor.
 
-AG - confused by indentation (Monaco inserts it automatically - duh!)
-it does not work if you have it and add a new to-level thing;
-
-BM uses copy & paste to copy thing and replace one identifier in the middle
-(which would be difficult to do with non-textual representation)
-
-people occasionally miss '.' when typing and try to use a space
-
-\paragraph{How users understand the then keyword}
-
-what is then: "allows us to chain together the operations"
-is it special? "maybe I need it just to split over multiple lines" (experiments..) "no!"
-
-"when you explained about the "dot then" that was a really useful thing to know.
-When I found that, I was like 'this is fine, this is doable'.
-If I knew this from the start, it would [have been easier]" (~27)
-
-"I thought 'then' starts a new line. (...) I realised there was no 'then' after the 'paging' command. "
-"It starts a new line or a new query? I assumed there would be one after 'paging' too."
+We observed two issues in the study. Participant \#2 struggled with correct indentation, starting a second command
+with more indentation than needed and participant \#6 had a syntax error in an unrelated command,
+which prevents charts from rendering. Some participants used the text editor effectively,
+e.g.~participant \#5, who used copy-and-paste to fetch the same data series for multiple countries.
 
@@
 
+\paragraph{How users understand the `then' member}
+When interviewing participants who worked with a tabular data source, we
+asked about their understanding of the \ikvd{then} member. This is a regular member generated by the
+type provider (i.e.~not a keyword), but it has a special meaning. Consider
+the following example which calculates the average travel expenses and total number of
+representatives per county using the UK House of Lords expenses data:
+
+\begin{thegamma}
+expneses.'group data'.'by County'
+  .'average Travel Costs'.'count distinct Name'.then
+\end{thegamma}
+
+The \ikvd{then} member is used to complete a part of a query where the user can continue
+adding items to build a list. Here, we select two aggregations to be calculated for each
+county before choosing \ikvd{then} and applying other operations such as sorting.
+
+Two participants (\#12 and \#13) initially thought that \ikvd{then} is used to split a command
+over multiple lines, but rejected the idea after experimenting and noting that they can insert
+line breaks elsewhere. One correctly concluded that it ``allows us to chain together the
+operations'' of the query. Participant \#13 needed a hint about using the
+\ikvd{then} member and later reflected:
+
+\begin{quote}
+  \emph{``When you explained about the `dot then' that was a really useful thing to know.
+  When I found that, I was like this is fine, this is doable. If I knew this from the start,
+  it would [have been easier].''}
+\end{quote}
+
+This comment summarizes an important fact. Although iterative prompting allows the users to start
+exploring new data sources, the domain specific languages used by more complex data sources have
+their own design principles that users need to understand to use the data source effectively.
+
 \subsection{Conclusions}
-(former journalist)
-Do you think this is something journalists could figure out how to use?
-"Yeah, I think so. There's a lot of effort going into data journalism that programming could make much quicker,
-but I was always nervous about code." ... "Something like this would really simplify things."
+The motivation behind The Gamma is to make simple programmatic data exploration accessible to
+journalists and other non-experts. Main-stream programming languages have the benefit of
+reproducibility and transparency, but are difficult to use. Our study shows that non-experts
+are able to solve basic data exploration tasks using The Gamma. When asked whether The Gamma
+is something that journalists could learn how to use, participant \#13 (a former journalist) answered:
 
+\begin{quote}
+\emph{``Yeah, I think so. There's a lot of effort going into data journalism that
+  programming could make much quicker, but I was always nervous about code. (...)
+  Something like this would really simplify things.''}
+\end{quote}
+
+The Gamma is based on the iterative prompting principle which lowers the barrier to entry, but
+it is worth noting that it does not fully eliminate complexity involved in data querying.
+Instead, it provides a way of structuring the complexity. Iterative prompting makes it easy to
+get started, but complex data sources will inevitably require additional learning. The Gamma
+makes this easier by allowing users to learn from published code samples.
 \newpage

### 2020-03-22 wip-design/paper.tex
Adding implementation section

@@
 
+\begin{figure}
+\includegraphics[width=1\columnwidth]{figures/sidebyside}
+\caption{The Gamma environment with an automatically generated chart comparing CO2 emissions
+of China and United States.}
+\label{fig:sidebyside}
+\end{figure}
+
 \section{Implementation}
@@
 A member can be either an ordinary member or an operation which takes a list of parameters
-enclosed in parentheses. Our implementation supports a few other features, but those
-are not used in this paper.
+enclosed in parentheses. When the member name contains non-alphanumerical characters such as
+whitespcae, it needs to be written in quotes. Our implementation supports a few other features,
+but those are not used in this paper.
 
-The language uses a type system to infer what members are available at a given point in a chain
-of members. The type is an object with a list of members that, in turn, have their own types.
-The types are not built-in, but are generated by a type provider for a particular data source.
-The types are also generated lazily. A type is only computed when the corresponding member
-is actually used in code.
+The Gamma uses a type system to infer what members are available at a given point in a chain.
+A type is an object with a list of members that, in turn, have their own types.
+The types are not built-in, but are generated by a type providers for individual data sources.
+The types are generated lazily and are only computed when a member appears in code.
 
@@
 
+\begin{figure}
+\centering
+\includegraphics[scale=0.28]{figures/cubetp}
+\caption{Exploring data cubes. The World Bank provider allows users to first choose
+  a country or a year and then an indicator.}
+\label{fig:cubetp}
+\end{figure}
+
 \subsection{Data Cube Type Provider}
+Our first type provider allows users explore data form a data cube. For example, the World
+Bank\footnote{\url{https://data.worldbank.org/}} collects a range of indicators about many
+countries in the world each year. The data set is a three-dimensional cube with dimensions
+corresponding to countries, indicators and years. The following example, also shown in
+Figure~\ref{fig:sidebyside}, uses the provider to access CO2 emission data for United States:
+
+\begin{thegamma}
+worldbank.byCountry.'United States'
+  .'Climate Change'.'CO2 emissions (kt)'
+\end{thegamma}
+
+As illustrated in Figure~\ref{fig:cubetp}, the provider allows uses to select a data series
+from the data cube. By choosing \ikvd{byCountry.\textquotesingle United States\textquotesingle},
+we restrict the cube to a two-dimensional plane. We then choose an indicator category
+\ikvd{\textquotesingle Climate Change\textquotesingle} and a specific indicator
+\ikvd{\textquotesingle CO2 emissions (kt)\textquotesingle}, obtaining a time series with
+years as keys and emission data as values. The World Bank provider also supports filtering
+by year and indicator.
+
+Another type provider available in The Gamma that is based on the data cube structure allows
+the user to explore UK government expenditure:
+
+\begin{thegamma}
+expenditure.byService.Defence.inTermsOf.GDP
+\end{thegamma}
+
+The dimensions of the cube are government services, years and value type (adjusted, nominal,
+per GDP). Here, we select the \ikvd{Defence} service and \ikvd{GDP} value type. Our implementation
+does not yet use a standardized data cube storage and so adding another data cube source
+currently requires implementing a new type provider.
 
@@
 
-\subsection{Graph Database Type Provider}
+Our second type provider allows users to construct queries to explore data in tabular formats
+such as CSV files. It is based on the theory developed Petricek~\cite{dotdriven}. Unlike the
+data cube provider, the provider for tabular data does not just allow selecting a subset of the
+data, but it can be used to construct SQL-like query. Consider the example from Figure~\ref{fig:thegamma}:
 
-\newpage
+\begin{thegamma}
+olympics.'filter data'.'Games is'.'Rio (2016)'.then
+  .'group data'.'by Team'.'sum Gold'.'sum Silver'.then
+  .'sort data'.'by Gold descending'
+\end{thegamma}
+
+The example works with a CSV file that records individual medals awarded in Olympic games.
+The chain constructs a query that selects rows corresponding to the Rio 2016 Olympics and then
+calculates total number of gold and silver medals for each team (country) before sorting the data.
+
+When using the provider, the user specifies a sequence of operations. Members such as
+\ikvd{\textquotesingle filter data\textquotesingle} or \ikvd{\textquotesingle group data\textquotesingle}
+determine the operation type. Those are followed by operation parameters. For example, when grouping
+data, we first select the key and then add a number of aggregations to calculate over the group.
+Unlike SQL, the provider only allows users to choose from pre-defined aggregations such as
+calculating the sum, average or the number of distinct values. As illustrated in
+Section~\ref{sec:cases}, this still allows us to construct a wide range of practical queries.
 
 \begin{figure}
-\includegraphics[width=1\columnwidth]{figures/sidebyside}
-\caption{The Gamma environment with an automatically generated chart comparing CO2 emissions
-of China and United States.}
-\label{fig:sidebyside}
-\vspace{-0.5em}
+\centering
+\includegraphics[scale=0.28]{figures/graphtp}
+\vspace{0.5em}
+\caption{Exploring graph databases. The user specifies a path through the data, possibly with
+  placeholders to select multiple nodes.}
+\label{fig:graphtp}
 \end{figure}
 
+\subsection{Graph Database Type Provider}
+Our third type provider allows users explore data from graph databases. A graph database
+consists of nodes representing entities and relationships between them. In the context of
+data journalism, graph databases have been used for example in The Panama Papers reporting \cite{panama}.
+
+The following example explores a database of Doctor Who characters and episodes. It retrieves
+all enemies of the Doctor that appear in the Day of the Moon episode:
+
+\begin{thegamma}
+drwho.Character.Doctor
+  .'ENEMY OF'.'[any]'
+  .'APPEARED IN'.'Day of the Moon'
+\end{thegamma}
+
+We start from the \ikvd{Doctor} node and then follow two relationships. We use
+\ikvd{\textquotesingle ENEMY OF\textquotesingle.\textquotesingle [any]\textquotesingle}
+to follow links to all enemies of the Doctor and then specify
+\ikvd{\textquotesingle APPEARED IN\textquotesingle.\textquotesingle Day of the Moon\textquotesingle}
+to only select only such enemies that appear in a specific episode. The resulting query
+is illustrated in Figure~\ref{fig:graphtp}.
+
+The provider works with any graph database and generates members automatically, based on the
+data in the database. In the above example, \ikvd{ENEMY OF} and \ikvd{APPEARED IN} are labels
+of relations and \ikvd{Doctor} and \ikvd{Day of the Moon} are labels of nodes. The
+\ikvd{[any]} member defines a placeholder that can be filled with any node with the specified
+relationships. The results returned by the provider is a table of properties of all nodes
+along the specified path. As discussed in the next section, such table can be further queried
+using the tabular data type provider.
+
 % ==================================================================================================
-
-\newpage
-~
-\newpage
 

### 2020-03-26 wip-design/paper.tex
References, case studies

@@
 
+Evaluating ui systems research
+\cite{evaluating}
+
+Type providers
+\cite{inforich,fsdata,dotdriven}
+
+Assistants for SW / auto-complete
+\cite{assistants}
+
+Journalism tools
+\cite{idyll}
+
+Notebook systems
+\cite{jupyter} or more transparency \cite{wrattler}
+managing messes \cite{messes}
+
+Visual data exploration tools
+\cite{control,tableau,vizdom}
+
+Wrangler
+\cite{wrangler}
+
+Victor
+\cite{learnable,principle}
+
+Live and exploratory programming
+\cite{review} - review of the two
+\cite{exploratory}, which led to new tools~\cite{variolite} (supports informal versioning),
+
+\cite{liveroad}
+review the use of live programming in a Smalltalk derived environment. Lighttable~\cite{lighttable}
+
+A more principled approach can be used by systems based on structured editing~\cite{structure-based,livenut,lamdu}
+
+Subtext
+\cite{subtext,directprog}
+
+sketch and sketch
+\cite{sketchnsketch}
+
+lower this barrier to valid, replicablestatistical analysis, we introduce Tea1, a high-level declar-ative language and runtime system
+(still JSON, but good aims)
+\cite{tea}
+
+DS.js: Turn Any Webpage into an Example-Centric Live Programming Environment for Learning Data Science
+(cool but just JS)
+\cite{dsjs}
+
+Eviza: A Natural Language Interface for Visual Analysis
+(from Tableau; uses auto-complete to suggest natural language phrases)
+\cite{eviza}
+
+Declarative Interaction Design for Data Visualization
+(interactive construction of vega-like grammar of graphics)
+\cite{interactionviz}
+
+programming by demonstration
+\cite{watch}
+
+"Dontcheva et al. [4] use PBD techniques to enable
+extraction, integration, and templated search of web data.
+Gulwani [8] introduces an algorithm for learning expres-
+sive string processing programs from input-output examples"
+
+programming by example as an alternative
+\cite{spreadsheetpbe,flashextract}
+
+Lyra: An interactive visualization design environment
+\cite{lyra}
+
+end-user programming
+\cite{enduser} - overview, \cite{smallmatter} book
+
+touch based data science??? must inevitably have fewer intraections
+NA
+
+autocomplete / predictive text work
+\cite{predictive} - predicting data transformation scripts in a somewhat more complex language
+\cite{proactive} -
+
+\cite{potter}
+Potter’s Wheel: An Interactive Data Cleaning System
+n interactive data clean-ing system that tightly integrates transformation anddiscrepancy detection.
+Users  gradually build  trans-formations  to  clean  the  data  by  adding  or  undoingtransforms on a spreadsheet-like interface; the effectof a transform is shown at once on records visible onscreen.
+
+\newpage
+
 \textbf{Visual tools.}
@@
 
+The Gamma aims to simplify programmatic data exploration while keeping enough expressive power
+to allow users to create interesting data visualizations. We consider the expressive power in
+this section and assess simplicity in the next one. We used The Gamma to analyse
+the UK government expenditure, activities of a research institute, Olympic medals and
+information about the Doctor Who series\footnote{Links removed to preserve anonymity.}.
+In this section, we consider two interesting larger examples.
+
+\subsection{If Michael Phelps were a Country}
+Michael Phelps has won so many medals that media compared him to
+countries.\footnote{\href{https://www.npr.org/sections/thetorch/2016/08/14/489832779/}{\small\bf\ttfamily npr.org/sections/thetorch/2016/08/14/489832779/}}
+Using the tabular data provider, we construct a chart shown in Figure~\ref{fig:cases} that
+produces a country league table including Michael Phelps:\footnote{Available at: \url{http://gallery.thegamma.net/86/}}
+
+\begin{thegamma}
+\kvd{let} data = olympics
+    .'group data'.'by Team'.'sum Gold'.then
+    .'sort data'.'by Gold descending'.then
+    .paging.skip(43).take(4)
+    .'get series'.'with key Team'.'and value Gold'
+
+\kvd{let} phelps = olympics
+  .'filter data'.'Athlete is'.'Michael Phelps'.then
+  .'group data'.'by Athlete'.'sum Gold'.then
+  .'get series'.'with key Athlete'.'and value Gold'
+
+charts.bar(data.append(phelps))
+  .setColors(["#aec7e8","#aec7e8","#1f77b4"])
+\end{thegamma}
+
+The core of the data analysis is done in two commands. The first counts gold medals by countries
+and uses paging to fetch 4 countries with suitable number of medals. In the second, we abuse the
+grouping operation to aggregate data for just a single group. The two data series are then assigned
+to local variables (for readability) and passed to the \ikvd{chart.columns} function.
+In this case study, the core data exploration can be constructed via iterative prompting, but
+building a chart requires invoking the \ikvd{append} operation and specifying colors using a list.
+
+\subsection{The Most Frequent Doctor Who Villains}
+In the second case study, we list Doctor Who villains by the number of episodes in which they
+appear.\footnote{Available at: \url{http://gallery.thegamma.net/87/}} This is interesting as it
+combines the graph database provider for fetching the data with the tabular data provider for
+summarization:
+
+\begin{thegamma}
+\kvd{let} topEnemies = drWho.Character.Doctor
+  .'ENEMY OF'.'[any]'.'APPEARED IN'.'[any]'.explore
+  .'group data'.'by Character name'
+    .'count distinct Episode name'.then
+  .'sort data'.'by Episode name descending'.then
+  .paging.take(8).'get series'
+    .'with key Character name'.'and value Episode name'
+
+chart.bar(topEnemies)
+\end{thegamma}
+
+Lines 1-2 use the graph database provider to find all paths linking the Doctor node with any
+character linked via \ikvd{ENEMY OF}, followed by any episode linked by \ikvd{APPEARED IN}.
+This produces a table that can be analysed using the tabular data provider by choosing the
+\ikvd{explore} member. For each character (the villain) we count the total number of
+distinct episodes in the table. The result is shown in Figure~\ref{fig:cases}.
+
+The logic is not trivial, but it performs a fairly sophisticated data analysis that involves a
+graph database query followed by an SQL-like data aggregation. The code can be constructed
+using iterative prompting (with the exception of the numbers in paging). It can also be constructed
+gradually and instantaneous prviews support the user while doing so.
+
 \newpage
+
+\begin{figure}
+\centering
+\includegraphics[width=1\columnwidth]{figures/cases}
+\caption{Charts produced by two The Gamma case studies.}
+\label{fig:cases}
+\end{figure}
+
 ~

### 2020-03-27 wip-design/paper.tex
Related work

@@
 
-\numberofauthors{3}
+\numberofauthors{1}
 \author{%
@@
     \affaddr{City, Country}\\
-    \email{e-mail address}}\\
-  \alignauthor{Leave Authors Anonymous\\
-    \affaddr{for Submission}\\
-    \affaddr{City, Country}\\
-    \email{e-mail address}}\\
-  \alignauthor{Leave Authors Anonymous\\
-    \affaddr{for Submission}\\
-    \affaddr{City, Country}\\
-    \email{e-mail address}}\\
+    \email{e-mail address}}
 }
@@
 
-Evaluating ui systems research
-\cite{evaluating}
-
-Type providers
-\cite{inforich,fsdata,dotdriven}
-
-Assistants for SW / auto-complete
-\cite{assistants}
-
-Journalism tools
-\cite{idyll}
-
-Notebook systems
-\cite{jupyter} or more transparency \cite{wrattler}
-managing messes \cite{messes}
-
-Visual data exploration tools
-\cite{control,tableau,vizdom}
-
-Wrangler
-\cite{wrangler}
-
-Victor
-\cite{learnable,principle}
-
-Live and exploratory programming
-\cite{review} - review of the two
-\cite{exploratory}, which led to new tools~\cite{variolite} (supports informal versioning),
-
-\cite{liveroad}
-review the use of live programming in a Smalltalk derived environment. Lighttable~\cite{lighttable}
-
-A more principled approach can be used by systems based on structured editing~\cite{structure-based,livenut,lamdu}
-
-Subtext
-\cite{subtext,directprog}
-
-sketch and sketch
-\cite{sketchnsketch}
-
-lower this barrier to valid, replicablestatistical analysis, we introduce Tea1, a high-level declar-ative language and runtime system
-(still JSON, but good aims)
-\cite{tea}
-
-DS.js: Turn Any Webpage into an Example-Centric Live Programming Environment for Learning Data Science
-(cool but just JS)
-\cite{dsjs}
-
-Eviza: A Natural Language Interface for Visual Analysis
-(from Tableau; uses auto-complete to suggest natural language phrases)
-\cite{eviza}
-
-Declarative Interaction Design for Data Visualization
-(interactive construction of vega-like grammar of graphics)
-\cite{interactionviz}
-
-programming by demonstration
-\cite{watch}
-
-"Dontcheva et al. [4] use PBD techniques to enable
-extraction, integration, and templated search of web data.
-Gulwani [8] introduces an algorithm for learning expres-
-sive string processing programs from input-output examples"
-
-programming by example as an alternative
-\cite{spreadsheetpbe,flashextract}
-
-Lyra: An interactive visualization design environment
-\cite{lyra}
-
-end-user programming
-\cite{enduser} - overview, \cite{smallmatter} book
-
-touch based data science??? must inevitably have fewer intraections
-NA
-
-autocomplete / predictive text work
-\cite{predictive} - predicting data transformation scripts in a somewhat more complex language
-\cite{proactive} -
-
-\cite{potter}
-Potter’s Wheel: An Interactive Data Cleaning System
-n interactive data clean-ing system that tightly integrates transformation anddiscrepancy detection.
-Users  gradually build  trans-formations  to  clean  the  data  by  adding  or  undoingtransforms on a spreadsheet-like interface; the effectof a transform is shown at once on records visible onscreen.
-
-\newpage
-
-\textbf{Visual tools.}
-
-\textbf{Programming tools.}
-Notebooks
-
-Both follow from assistant tools for software development \cite{assistants}. Our implementation,
-outlined in the previous section, implements iterative prompting through a code editor mechanism
-designed for code completion. Finally, the work on type providers \cite{inforich,fsdata,dotdriven},
-which we extend, also utilizes code completion.
-
-\textbf{Journalism.}
-Idyll \cite{idyll}
-
-\textbf{Type providers.}
-PL work
-
-\newpage
-~
+The Gamma brings together recent programming language innovations for working with data
+\cite{inforich}, work on making programming accessible to non-programmers \cite{enduser,smallmatter}
+and work on tools for journalists such as Idyll \cite{idyll}. Related work includes a range of
+visual and programmatic tools used by data scientists, work that aims to simplify programming
+or explore alternative paradigms and various predictive tools that assist users in completing
+programming or data exploration tasks.
+
+% Evaluating ui systems research
+% \cite{evaluating}
+
+\subsection{Data science tools}
+Data exploration is typically done using spreadsheets, comprehensive visual tools or
+via code using notebooks. Notebooks such as Jupyter \cite{jupyter} are also popular with technically
+skilled journalists as they allow combining code with commentary and visual outputs. However,
+notebooks can lead to messy hard to reproduce code \cite{messes,wrattler}.
+
+Visual data analytics tools \cite{control,vizdom} such as Tableau \cite{tableau} do not require
+programming skills, but require mastering a range of different interactions. They also make it
+hard to see what data transformations have been done. Several systems for data exploration,
+e.g.~Potter's Wheel and Wrangler \cite{potter,wrangler}, and visualization, e.g.~Lyra \cite{lyra},
+provide a GUI, but record a transparent script that can be seen and modified by the user.
+
+\subsection{Easier programming tools}
+We aim to design a programming environment that is easy to use and learn. Many approaches to this
+goal have been tried. Victor \cite{learnable,principle} introduced design principles that
+inspired many to build live programming systems \cite{review,liveroad,lighttable} that give
+immediate feedaback to help programmers understand how code relates to output and
+exploratory systems \cite{variolite,exploratory} that assist with completing open-ended tasks.
+
+To avoid difficulties with editing code as text, some systems use structured editors~\cite{structure-based,livenut,lamdu}.
+In Subtext \cite{subtext,directprog} the language itself is co-designed with the editor to make
+the interactions with code more natural. Finally, programming can be made easier by designing
+high-level declarative abstractions. Tea does so for statistical analyses \cite{tea} and
+Satyanarayan et al. \cite{interactionviz,vegalite} does so for interactive data visualizations.
+
+The Gamma is live in that it gives an instant preview of the results. It is text-based,
+although it is syntactically simple. The code written in The Gamma is high-level and so work
+on declarative abstractions is a valuable inspiration.
+
+\subsection{Code completion and assistants}
+A key component in The Gamma design is the use of auto-complete for offering possible
+operations. Code completion has been pioneerd by Kaiser~\cite{assistants}, but our work is more
+directly inspired by the work on type providers \cite{inforich,fsdata,dotdriven}, which utilize
+code completion to integrate external data into a statically typed programming language.
+
+When using type providers, completions are offered as the user types code. However, completion
+tools can also recmmend scripts when users begin interacting with data \cite{predictive,proactive}.
+Alternatively, they can also be guided by natural language as for example Eviza~\cite{eviza},
+which uses natural language for data exploration. CodeMend~\cite{codemend} uses natural language
+for more general code edits, but uses data visualization as a case study.
+Finally, data entry interfaces based on completion such as Dasher \cite{dasher} utilize an
+interaction principle similar to our iterative prompting.
+
+\subsection{Programming without writing code}
+
+A number of tools for data exploration use alternative approaches to programming where
+the user does not write code. In programming by example \cite{byexample}, the user gives
+examples of desired results. In data science, this approach has been used for example for
+specifying data transformations in spreadsheets and data extraction \cite{spreadsheetpbe,flashextract}.
+
+In direct manipulation \cite{direct}, a program is specified by directly interacting with the
+output. This has been used in the visual domain \cite{sketchnsketch}, but also for data querying
+\cite{dynamicq,vlang}. The VQE language~\cite{visage} also considers how to allow code reuse and
+modification in this context. Direct manipulation can also support data exploration by letting
+users partially edit queries,~e.g. by changing quantifiers as in DataPlay~\cite{dataplay}.
+
 \newpage

### 2020-03-28 wip-design/paper.tex
Discussion and conclusions

@@
 
-% Evaluating ui systems research
-% \cite{evaluating}
-
 \subsection{Data science tools}
@@
 
-\newpage
-
-\begin{figure}
-\includegraphics[width=1\columnwidth]{figures/nyt}
-\caption{New York Times article on Obama's legacy \cite{youdraw}. The article asks the reader to make a guess
-(engagement), but only lists ``Immigration and Customs Enforcement'' as a source of data.}
-\label{fig:nyt}
-\end{figure}
+% ==================================================================================================
 
@@
 One way of establishing trust in the age of fake news is to be more transparent about editorial
-decisions, process and original sources. Many journalists believe that opening up the process
+decisions, process and original sources.
+
+Many journalists believe that opening up the process
 shows the quality and trustworthiness of their work~\cite{transparency}.
@@
 \begin{figure}
-\includegraphics[width=1\columnwidth]{figures/ft}
-\caption{Financial Times analysis of recyclable waste. Full source is provided as Jupyter Notebooks
-on GitHub \cite{ftnotebooks}, but re-running the analysis is difficult, even for an expert.}
-\label{fig:ft}
+\includegraphics[width=1\columnwidth]{figures/nyt}
+\caption{New York Times article on Obama's legacy \cite{youdraw}. The article asks the reader to make a guess
+(engagement), but only lists ``Immigration and Customs Enforcement'' as a source of data.}
+\label{fig:nyt}
 \end{figure}
@@
 
-\newpage
-~
-\newpage
+\begin{figure}
+\includegraphics[width=1\columnwidth]{figures/ft}
+\caption{Financial Times analysis of recyclable waste. Full source is provided as Jupyter Notebooks
+on GitHub \cite{ftnotebooks}, but re-running the analysis is difficult, even for an expert.}
+\label{fig:ft}
+\end{figure}
+
+% ==================================================================================================
 
@@
 
-\newpage
-
 \begin{figure}
@@
 \end{figure}
-
-~
-\newpage
 
@@
 
-\subsection{Conclusions}
+\subsection{Study conclusions}
 The motivation behind The Gamma is to make simple programmatic data exploration accessible to
@@
 makes this easier by allowing users to learn from published code samples.
-\newpage
-~
-\newpage
 
@@
 \section{Discussion}
-
-Data provenance (where data comes from)
-What are the data sources availbale (how to start)?
-
-\subsection{Study limitations}
-exploratory in nature so we do not make any quantitative claims about effects
-
-not comparing against other systems
-
-\subsection{Design principles}
-How well did we do wrt design principles?
-
-\subsection{Design issues}
-future challenges and limitations of the model - such as issues when modifying code
-in the middle of the call chain
+We design a programming environment that can make programmatic data exploration accessible
+to journalists. This is aligned with recent developments in journalism, which aims to build
+trust through transparency and greater reproducibility as well as provide meaningful ways of
+engagement.
+
+\subsection{Evaluating The Gamma}
+Data exploration environments are complex systems that do not yield to simple controlled
+experimentation. We evaluate The Gamma using criteria inspired by Olsen~\cite{evaluating}.
+
+\begin{itemize}
+\item \emph{Importance.} Open journalism has the potential to make factual claims backed by
+data more commonplace and enable wider audience engage with such claims. As such, we contribute
+towards solving an important societal issue.
+
+\item \emph{Expressive leverage.} The Gamma reduces the number of concepts that the user needs
+to master to just the \emph{iterative prompting} principle, although complex data sources may
+add more concepts through their domain specific languages.
+
+\item \emph{Empowering new participants.} As demonstrated by our user study, The Gamma allows
+non-experts, including those not comfortable with code, to perform basic programmatic
+data exploration tasks.
+
+\item \emph{Generality.} Finally, our case studies show that The Gamma can be used to solve
+a wide range of data manipulation tasks involving many data sources including tabular data,
+data cubes and graph databases.
+\end{itemize}
+
+\subsection{Remaining design issues}
+There remain a number of aspects of data exploration in the context of journalism that
+The Gamma does not address. Two of those, data provenance and data availability were also
+observed by the participants in our study.
+
+\paragraph{Data provenance}
+Data sources such as \ikvd{olympics} or \ikvd{worldbank} are defined when initializing The Gamma
+programming environment, but the system does not currently show where such data comes from.
+For some data sources, the source is e.g.~a CSV file published by the government. In this case,
+we can easily show the source. However, other type providers may pre-process data. Displaying
+data source in such cases would require more sophisticated provenance tracking \cite{provenance}.
+
+\paragraph{Data availability}
+In the current version, The Gamma does not have a way of informing the user what data sources
+are available. In other words, the user needs to know the first identifier, such as \ikvd{olympics},
+to get started. We could address this by choosing a data source as the first step of iterative
+prompting and perhaps typing \ikvd{.olympics}. However, a more fundamental issue is finding
+the data source in the first place. This could be partly addressed by a type provider for a
+currated online database such as Enigma Public\footnote{\url{https://docs.enigma.com/public}}.
+Providing access to open government data repositories such as \href{http://data.gov.uk}{\small\bf\ttfamily data.gov.uk}
+is more appealing, but very challenging due to their unstructured nature.
+
+\subsection{Evaluation limitations}
+The evaluation conducted in this paper is qualitative and exploratory in nature. In particular,
+we do not make any quantitative claims about the usability of The Gamma and its learning curve.
+Our study also focused on the core iterative prompting interaction, but our case studies often
+required using other features such as operations with parameters.
+
+Although The Gamma is open-source, it has not been deployed in production in a newsroom so far.
+This would lead to valuable insights, but it requires finding a suitable fortuitous opportunity.
+Finally, we also do not compare the usability of The Gamma with the usability of other popular
+systems such as Tableau \cite{tableau}. It would be possible to set tasks that can be completed
+in both systems, but the systems have very different aims making such comparison problematic.
 
 \section{Conclusions}
-
-\newpage
-~
-\newpage
-
-\section{Acknowledgments}
-Yo
+We present The Gamma, a simple data exploration environment. The Gamma is motivated by the needs
+of journalists working with data, who need to produce transparent and easy-to-understand reports
+backed by data that are transparent and engaging, all the while facing tight deadlines.
+
+The Gamma is based on a single interaction principle, \emph{iterative prompting}, can be used to
+complete a wide range of data exploration tasks using a variety of data sources including tabular
+data, data cubes and graph databases. The design lowers the barrier to entry for programmatic
+data exploration and it makes it easy to learn the system independently from examples and by
+experimentation.
+
+% \section{Acknowledgments}
+% Yo
 

### 2020-04-21 wip-compost/paper.tex
Updating to programming template, tweaks
Introduction for compost paper

@@
-\documentclass[acmsmall,anonymous,fleqn]{acmart}\settopmatter{printfolios=false,printccs=false,printacmref=false}
+\NeedsTeXFormat{LaTeX2e}
+\documentclass{jfp}
 
-%% For double-blind review submission, w/ CCS and ACM Reference
-%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
-%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
-%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
-%% For single-blind review submission, w/ CCS and ACM Reference
-%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
-%% For final camera-ready submission, w/ required CCS and ACM Reference
-%\documentclass[acmsmall]{acmart}\settopmatter{}
-
-\acmJournal{PACMPL}
-\acmVolume{1}
-\acmNumber{OOPSLA}
-\acmArticle{1}
-\acmYear{2019}
-\acmMonth{1}
-\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
-\startPage{1}
-
-\settopmatter{printfolios=true}
-%% Copyright information
-%% Supplied to authors (based on authors' rights management selection;
-%% see authors.acm.org) by publisher for camera-ready submission;
-%% use 'none' for review submission.
-\setcopyright{none}
-%\setcopyright{acmcopyright}
-%\setcopyright{acmlicensed}
-%\setcopyright{rightsretained}
-%\copyrightyear{2018}           %% If different from \acmYear
-
-%% Bibliography style
-\bibliographystyle{ACM-Reference-Format}
-%% Citation style
-%% Note: author/year citations are required for papers published as an
-%% issue of PACMPL.
-\citestyle{acmauthoryear}   %% For author/year citations
-
-\usepackage{booktabs}
-\usepackage{subcaption}
-\usepackage{lineno,hyperref,xcolor}
-\usepackage{flushend}
-\usepackage{stmaryrd}
-\usepackage{amssymb}
-\usepackage{xypic}
-\usepackage{semantic}
-\usepackage{booktabs}
-\usepackage{enumitem}
-\usepackage{enumerate}
-\usepackage{amsmath}
-\usepackage[T1]{fontenc}
-
-\setlist{leftmargin=6mm}
-\newcounter{thc}
-\newcounter{dfc}
-
-\theoremstyle{plain}
-\newtheorem{lem}[thc]{Lemma}
-\newtheorem{theorem}[thc]{Theorem}
-
-\theoremstyle{definition}
-\newtheorem{axiom}[dfc]{Axiom}
-\newtheorem{definition}[dfc]{Definition}
-
-\title{Compost: Library for composable data visualization}
-%\author{Tomas Petricek}
-%\affiliation{
-%  \institution{University of Kent}
-%  \country{United Kingdom}
-%}
-%\email{tomas@tomasp.net}
-
-%\begin{CCSXML}
-%<ccs2012>
-%<concept>
-%<concept_id>10011007.10011006.10011008</concept_id>
-%<concept_desc>Software and its engineering~General programming languages</concept_desc>
-%<concept_significance>500</concept_significance>
-%</concept>
-%<concept>
-%<concept_id>10003456.10003457.10003521.10003525</concept_id>
-%<concept_desc>Social and professional topics~History of programming languages</concept_desc>
-%<concept_significance>300</concept_significance>
-%</concept>
-%</ccs2012>
-%\end{CCSXML}
-%
-%\ccsdesc[500]{Software and its engineering~General programming languages}
-%\ccsdesc[300]{Social and professional topics~History of programming languages}
-
-\definecolor{cmtclr}{rgb}{0.0,0.6,0.0}
-\definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
-\definecolor{idclr}{rgb}{0.0,0.0,0.0}
-\definecolor{numclr}{rgb}{0.0,0.4,0.0}
-\definecolor{strclr}{rgb}{0.4,0.4,0.0}
-\definecolor{rstrclr}{rgb}{0.5,0.1,0.0}
-\definecolor{prepclr}{rgb}{0.6,0.0,0.2}
-\newcommand{\vect}[1]{\langl #1 \rangl}
-\newcommand{\langl}{\begin{picture}(4.5,7)
-\put(1.1,2.5){\rotatebox{60}{\line(1,0){5.5}}}
-\put(1.1,2.5){\rotatebox{300}{\line(1,0){5.5}}}
-\end{picture}}
-\newcommand{\rangl}{\begin{picture}(4.5,7)
-\put(.9,2.5){\rotatebox{120}{\line(1,0){5.5}}}
-\put(.9,2.5){\rotatebox{240}{\line(1,0){5.5}}}
-\end{picture}}
-\newcommand{\ball}[1]{\FPeval{\result}{clip(201+#1)}\textnormal{\ding{\result}}}
-\newcommand{\lsep}{\;\;|\;\;}
-\newcommand{\num}[1]{\textcolor{numclr}{#1}}
-\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
-\newcommand{\strf}[1]{\textnormal{\textcolor{strclr}{\sffamily #1}}}
-\newcommand{\rstr}[1]{\textnormal{\textcolor{rstrclr}{\sffamily "#1"}}}
-\newcommand{\ident}[1]{\textnormal{\textcolor{idclr}{\sffamily #1}}}
-\newcommand{\qident}[1]{\textnormal{\sffamily \textquotesingle #1\textquotesingle}}
-\newcommand{\dom}{\ident{dom}}
-\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
-
-\newcommand{\bndclr}[1]{\textcolor{kvdclr}{#1}}
-\newcommand{\bkndclr}[1]{\textcolor{prepclr}{#1}}
-\newcommand{\blblclr}[1]{\textcolor{numclr}{#1}}
-\newcommand{\bnd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
-\newcommand{\bknd}[1]{\textnormal{\textcolor{prepclr}{\sffamily #1}}}
-\newcommand{\blbl}[1]{\textnormal{\textcolor{numclr}{\sffamily #1}}}
-\newcommand{\narrow}[1]{\hspace{-0.6em}#1\hspace{-0.6em}}
-\newcommand{\rname}[1]{{\sffamily\small(#1)}}
-\newcommand{\ername}[1]{\vspace{0.75em}\rname{#1}\hspace{0.5em}}
-\newcommand{\preview}[1]{\,\textnormal{\guillemotleft} #1\textnormal{\guillemotright}\,}
+\title{Composing data visualizations}
+\author[Tomas Petricek]{TOMAS PETRICEK\\
+       University of Kent, UK\\
+       \email{t.petricek@kent.ac.uk}}
 
 \begin{document}
-%\begin{abstract}
-%\end{abstract}
-\maketitle
+\maketitle[f]
 
-% ==================================================================================================
+% \begin{abstract}
+% This guide is for authors who are preparing papers for the \emph{Journal of
+% Functional Programming} using the \LaTeXe\ document-preparation system
+% and the Functional Programming class file (\texttt{jfp1.cls}).
+% \end{abstract}
+
+\bibliographystyle{jfp}
+
+\section{Introduction}
+Let's say we want to create the two charts in Figure~\ref{fig:charts}. The chart on the left is
+a bar chart, except that it shows two different values for each bar. The chart on the
+right is a line chart, except that it also highlights two parts of the timeline with two different
+colors.
+
+There is a plenty of libraries that can draw bar charts and line charts, but adding those extra
+features will only be possible if the author already thought about your exact scenario. For
+example, Google Charts supports the left chart (it is called Dual-X Bar Chart) but there is no
+way for adding a background, let alone based on the X value. The alternative is to use a more
+low-level library such as D3. In D3 you construct the chart piece by piece, but then you have to
+tediously transform your values to coordinates in pixels. For scientific data visualizations,
+you could use an implementation of the Grammar of Graphics such as ggplot2. This lets you
+map data to geometric objects (such as points, bars, and lines) and their visual properties
+(X and Y coordinate, shape and color). However, the number of options is still somewhat limited.
+
+To a functional programmer, this situation is disappointing. We should be able to
+compose a variety of charts from a small number of primitives. The shapes should be defined
+not in terms of pixels, but numbers of seats or exchange rates. We should also be able to easily
+derive high-level abstractions, such as a bar chart, from those low-level primitives.
+
+\begin{figure}
+  \includegraphics[scale=0.57]{figures/charts}
+  \vspace{0.25em}
+  \caption{Two charts about the UK politics: Comparison of election results from 2017 and 2019 (left)
+    and GBP-USD exchange rate with highlighted areas before and after the 23 June 2016 Brexit vote.}
+  \label{fig:charts}
+\end{figure}
+
+In this paper, we...
+
+\newpage
 
@@
 \label{sec:intro}
+\subsection*{Domain-specific values}
 
-\subsection*{Domain-specific values}
 All values that are provided when constructing charts are from the chart domain. For example,
@@
 more flexible libraries like, say, D3.
-
 Compost supports only two-dimensional charts and values can be either categorical (such as
@@
 exact position in the space available for the chart. For continuous values, this simply means
-applying a linear transformation. For categorical values, the situation is more difficult. For example,
+applying a linear transformation. For categorical values, the situation is more difficult. For % example,
 say we are drawing a bar chart with population of three different countries with countries
 on the X axis and population on the Y axis. The scale of the X axis is categorical and
-Compost divides the available space into three ranges of equal width. However, then \strf{United Kingdom}
+Compost divides the available space into three ranges of equal width. However, then \strf{United % Kingdom}
 or \strf{Czech Republic} does not determine an exact position but instead a range. To determine
@@
 illustrated in Figure~\ref{fig:scales}.
-
 \begin{figure}[!h]
@@
 \end{figure}
-
 More formally, we write $c$ for categorical values such as countries (represented as country
@@
 a location based on the implicitly obtained re-scaling.
-
 \subsection*{Language of composable charts}
@@
 \quad \kvd{fill}~\strf{\#0000ff},
-  ~(\kvd{cat}~\strf{UK}, \num{0}), (\kvd{cont}~\num{00.00}), (\kvd{cat}~\strf{UK}, \num{0}), (\kvd{cont}~\num{66.04}),\\
-\hspace{5.9em}(\kvd{cat}~\strf{UK}, \num{1}), (\kvd{cont}~\num{66.04}), (\kvd{cat}~\strf{UK}, \num{1}), (\kvd{cont}~\num{00.00}),\\
+ ~(\kvd{cat}~\strf{UK}, \num{0}), (\kvd{cont}~\num{00.00}), (\kvd{cat}~\strf{UK}, \num{0}), % (\kvd{cont}~\num{66.04}),\\
+\hspace{5.9em}(\kvd{cat}~\strf{UK}, \num{1}), (\kvd{cont}~\num{66.04}), (\kvd{cat}~\strf{UK}, % \num{1}), (\kvd{cont}~\num{00.00}),\\
 \quad \kvd{fill}~\strf{\#ff0000},
-  ~(\kvd{cat}~\strf{CZ}, \num{0}), (\kvd{cont}~\num{00.00}), (\kvd{cat}~\strf{CZ}, \num{0}), (\kvd{cont}~\num{10.58}),\\
-\hspace{5.9em}(\kvd{cat}~\strf{CZ}, \num{1}), (\kvd{cont}~\num{10.58}), (\kvd{cat}~\strf{CZ}, \num{1}), (\kvd{cont}~\num{00.00})\\
+ ~(\kvd{cat}~\strf{CZ}, \num{0}), (\kvd{cont}~\num{00.00}), (\kvd{cat}~\strf{CZ}, \num{0}), % (\kvd{cont}~\num{10.58}),\\
+\hspace{5.9em}(\kvd{cat}~\strf{CZ}, \num{1}), (\kvd{cont}~\num{10.58}), (\kvd{cat}~\strf{CZ}, % \num{1}), (\kvd{cont}~\num{00.00})\\
 \end{array}
@@
 category, so the ratios for the second bar are also $\num{0}$ and $\num{1}$.
-
 \subsection*{Calculating scales and projections}
@@
 library infers the scales for the X and Y axes and computes a projection function that can be
-used for mapping domain-specific values to pixels. While doing this, it also checks for consistency --
+used for mapping domain-specific values to pixels. While doing this, it also checks for consistency % --
 all values on one axis have to be either categorical or continuous, but not a mix of the two.
-
 The first step is to determine the scales for each of the axes. A scale can be continuous,
@@
 scale, the process fails.
-
 Once we obtain scales for both X and Y axes, we calculate a projection function for each axis.
@@
 according to the ratio $r$.
-
 \subsection*{Controlling and nesting scales}
@@
 We first focus on the first three primitives, while the \kvd{nest} primitive will be discussed in
-the next section. The \kvd{axis} and \kvd{roundScale} constructs could be defined as derived constructs,
-but it is easier to consider them as primitives for now. The \kvd{axis} primitive simply takes the shape
-$s$ and draws an axis around the contents of $s$, using the inferred scale of $s$ to determine values
+the next section. The \kvd{axis} and \kvd{roundScale} constructs could be defined as derived % constructs,
+but it is easier to consider them as primitives for now. The \kvd{axis} primitive simply takes the % shape
+$s$ and draws an axis around the contents of $s$, using the inferred scale of $s$ to determine % values
 displayed on the axis. This could be a derived construct, because axes can be rendered using lines
 and text labels.
-
 The \kvd{roundScale} operation takes the inferred X or Y scale of the shape $s$ and, if it is a
@@
 \kvd{axis}_x~(\kvd{axis}_y~(\kvd{roundScale}_y\\
-\quad(\kvd{explicitScale}_x~(\kvd{categorical \strf{CZ},\strf{MN},\strf{UK}})~\ident{populationBar})))
+\quad(\kvd{explicitScale}_x~(\kvd{categorical % \strf{CZ},\strf{MN},\strf{UK}})~\ident{populationBar})))
 \end{array}
@@
 maximum. Finally, we add axes around the shape, producing a usual labelled chart.
-
 As noted earlier, both \kvd{axis} and \kvd{roundScale} could be defined as derived -- \kvd{axis}
@@
 scale.
-
 \begin{figure}[!h]
@@
 \end{figure}
-
 \subsection*{Nesting of charts and scales}
@@
 one.
-
 When inferring the scales, Compost infers the scale of the nested shape $s$. This will then take
@@
 scale of the outer shape.
-
 Figure~\ref{fig:nesting} illustrates this with an example. Here, we are nesting a continuous scale
@@
 $v_{min}$ do not even have to be continuous. We can, for example, nest a line chart inside a
-bar of a bar chart and then the two values could be $\kvd{cat}~\strf{One}, 0$ and $\kvd{cat}~\strf{One}, 1$
+bar of a bar chart and then the two values could be $\kvd{cat}~\strf{One}, 0$ and % $\kvd{cat}~\strf{One}, 1$
 (which are two values that define a region of a categorical scale).
-
 One area where nesting of scales is particularly useful is when combining multiple charts in
@@
 have their own Y axis (price), but the X axis (time) should be shared.
-
 Assuming we have \ident{googPrices} and \ident{fbPrices} as two line charts without any axes,
@@
  (\kvd{axis}_y~\ident{fbPrices})),\\
-\quad (\kvd{nest}_y~(\kvd{cat}~\strf{bottom-chart}, \num{0}), (\kvd{cat}~\strf{bottom-chart}, \num{1}),
+\quad (\kvd{nest}_y~(\kvd{cat}~\strf{bottom-chart}, \num{0}), (\kvd{cat}~\strf{bottom-chart}, % \num{1}),
  (\kvd{axis}_y~\ident{googPrices}))~)\\
@@
 of the line charts. Given the difference in the prices, the axes will have quite different values.
-We then nest the (continuous) Y axes using the \kvd{nest} primitive and, at the same time, implicitly
+We then nest the (continuous) Y axes using the \kvd{nest} primitive and, at the same time, % implicitly
 define a new outer Y axis. The outer Y axis is categorical with just two values, \strf{top-chart}
@@
 minimal value has ratio $0$ while the maximal value has ratio $1$).
-
 ~
-
 ~
-
 \noindent
 [TODO: We really need illustrations for those example charts, but that would be more work!]
-
 ~
-
+[TODO: Maybe use this chart as a motivation: https://wellcomeopenresearch.org/articles/4-63/v1]
+~
 \section{Random ideas}
 \begin{itemize}
-\item We could define a type system to catch categorical/continuous value mismatch in a single scale.
+\item We could define a type system to catch categorical/continuous value mismatch in a single % scale.
 \item It would be worth thinking about ordinal values, which are categorical but can be sorted.
-\item There might be other kinds of scales - for example, color scale (can have meaning in scatter plot)
-  or a scale for secondary markers (like sizes of bubbles in a bubble chart). We could really say that
+\item There might be other kinds of scales - for example, color scale (can have meaning in scatter % plot)
+ or a scale for secondary markers (like sizes of bubbles in a bubble chart). We could really say % that
  every value (including bar chart colors) should be coming from some scale...
@@
 \end{document}
+
+%
+
+% \usepackage{booktabs}
+% \usepackage{subcaption}
+% \usepackage{lineno,hyperref,xcolor}
+% \usepackage{flushend}
+% \usepackage{stmaryrd}
+% \usepackage{amssymb}
+% \usepackage{xypic}
+% \usepackage{semantic}
+% \usepackage{booktabs}
+% \usepackage{enumitem}
+% \usepackage{enumerate}
+% \usepackage{amsmath}
+% \usepackage[T1]{fontenc}
+%
+% \setlist{leftmargin=6mm}
+% \newcounter{thc}
+% \newcounter{dfc}
+%
+% \theoremstyle{plain}
+% \newtheorem{lem}[thc]{Lemma}
+% \newtheorem{theorem}[thc]{Theorem}
+%
+% \theoremstyle{definition}
+% \newtheorem{axiom}[dfc]{Axiom}
+% \newtheorem{definition}[dfc]{Definition}
+%
+% \title{Compost: Library for composable data visualization}
+% %\author{Tomas Petricek}
+% %\affiliation{
+% %  \institution{University of Kent}
+% %  \country{United Kingdom}
+% %}
+% %\email{tomas@tomasp.net}
+%
+% %\begin{CCSXML}
+% %<ccs2012>
+% %<concept>
+% %<concept_id>10011007.10011006.10011008</concept_id>
+% %<concept_desc>Software and its engineering~General programming languages</concept_desc>
+% %<concept_significance>500</concept_significance>
+% %</concept>
+% %<concept>
+% %<concept_id>10003456.10003457.10003521.10003525</concept_id>
+% %<concept_desc>Social and professional topics~History of programming languages</concept_desc>
+% %<concept_significance>300</concept_significance>
+% %</concept>
+% %</ccs2012>
+% %\end{CCSXML}
+% %
+% %\ccsdesc[500]{Software and its engineering~General programming languages}
+% %\ccsdesc[300]{Social and professional topics~History of programming languages}
+%
+% \definecolor{cmtclr}{rgb}{0.0,0.6,0.0}
+% \definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
+% \definecolor{idclr}{rgb}{0.0,0.0,0.0}
+% \definecolor{numclr}{rgb}{0.0,0.4,0.0}
+% \definecolor{strclr}{rgb}{0.4,0.4,0.0}
+% \definecolor{rstrclr}{rgb}{0.5,0.1,0.0}
+% \definecolor{prepclr}{rgb}{0.6,0.0,0.2}
+% \newcommand{\vect}[1]{\langl #1 \rangl}
+% \newcommand{\langl}{\begin{picture}(4.5,7)
+% \put(1.1,2.5){\rotatebox{60}{\line(1,0){5.5}}}
+% \put(1.1,2.5){\rotatebox{300}{\line(1,0){5.5}}}
+% \end{picture}}
+% \newcommand{\rangl}{\begin{picture}(4.5,7)
+% \put(.9,2.5){\rotatebox{120}{\line(1,0){5.5}}}
+% \put(.9,2.5){\rotatebox{240}{\line(1,0){5.5}}}
+% \end{picture}}
+% \newcommand{\ball}[1]{\FPeval{\result}{clip(201+#1)}\textnormal{\ding{\result}}}
+% \newcommand{\lsep}{\;\;|\;\;}
+% \newcommand{\num}[1]{\textcolor{numclr}{#1}}
+% \newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
+% \newcommand{\strf}[1]{\textnormal{\textcolor{strclr}{\sffamily #1}}}
+% \newcommand{\rstr}[1]{\textnormal{\textcolor{rstrclr}{\sffamily "#1"}}}
+% \newcommand{\ident}[1]{\textnormal{\textcolor{idclr}{\sffamily #1}}}
+% \newcommand{\qident}[1]{\textnormal{\sffamily \textquotesingle #1\textquotesingle}}
+% \newcommand{\dom}{\ident{dom}}
+% \newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
+%
+% \newcommand{\bndclr}[1]{\textcolor{kvdclr}{#1}}
+% \newcommand{\bkndclr}[1]{\textcolor{prepclr}{#1}}
+% \newcommand{\blblclr}[1]{\textcolor{numclr}{#1}}
+% \newcommand{\bnd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
+% \newcommand{\bknd}[1]{\textnormal{\textcolor{prepclr}{\sffamily #1}}}
+% \newcommand{\blbl}[1]{\textnormal{\textcolor{numclr}{\sffamily #1}}}
+% \newcommand{\narrow}[1]{\hspace{-0.6em}#1\hspace{-0.6em}}
+% \newcommand{\rname}[1]{{\sffamily\small(#1)}}
+% \newcommand{\ername}[1]{\vspace{0.75em}\rname{#1}\hspace{0.5em}}
+% \newcommand{\preview}[1]{\,\textnormal{\guillemotleft} #1\textnormal{\guillemotright}\,}
+%
+% \begin{document}
+% %\begin{abstract}
+% %\end{abstract}
+% \maketitle
+%
+% % ==================================================================================================
+%

### 2020-04-25 wip-compost/paper.tex
Finish intro, start DSL section

@@
 \documentclass{jfp}
+\usepackage{xcolor}
 
@@
 
+\definecolor{kvdclr}{rgb}{0.5,0.0,0.5}
+\definecolor{fkvdclr}{rgb}{0.0,0.0,0.6}
+\definecolor{urlclr}{rgb}{0.0,0.0,0.8}
+\definecolor{idclr}{rgb}{0.0,0.0,0.0}
+\definecolor{numclr}{rgb}{0.0,0.4,0.0}
+\definecolor{strclr}{rgb}{0.6,0.2,0.0}
+
+\newcommand{\lsep}{\;\;|\;\;}
+\newcommand{\num}[1]{\textcolor{numclr}{#1}}
+\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
+\newcommand{\strf}[1]{\textnormal{\textcolor{strclr}{\sffamily #1}}}
+\newcommand{\ident}[1]{\textnormal{\textcolor{idclr}{\sffamily #1}}}
+\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
+\newcommand{\fkvd}[1]{\textnormal{\textcolor{fkvdclr}{\sffamily #1}}}
+\newcommand{\narrow}[1]{\hspace{-0.6em}#1\hspace{-0.6em}}
+\newcommand{\urrl}[1]{\textnormal{\textcolor{urlclr}{\url{#1}}}}
 \begin{document}
@@
 Let's say we want to create the two charts in Figure~\ref{fig:charts}. The chart on the left is
-a bar chart, except that it shows two different values for each bar. The chart on the
-right is a line chart, except that it also highlights two parts of the timeline with two different
-colors.
+a bar chart that shows two different values for each bar. The chart on the right consists of two
+line charts that share the X axis and highlight two parts of the timeline with two different colors.
 
@@
 example, Google Charts supports the left chart (it is called Dual-X Bar Chart) but there is no
-way for adding a background, let alone based on the X value. The alternative is to use a more
+way for adding a background, or sharing an axis between charts. The alternative is to use a more
 low-level library such as D3. In D3 you construct the chart piece by piece, but then you have to
-tediously transform your values to coordinates in pixels. For scientific data visualizations,
-you could use an implementation of the Grammar of Graphics such as ggplot2. This lets you
-map data to geometric objects (such as points, bars, and lines) and their visual properties
-(X and Y coordinate, shape and color). However, the number of options is still somewhat limited.
+tediously transform your values to coordinates in pixels yourself. For scientific plots,
+you could use an implementation of Grammar of Graphics such as ggplot2, where a chart is
+a mapping from data to geometric objects (such as points, bars, and lines) and their visual
+properties (X and Y coordinate, shape and color). However, the range of charts that can be
+created using this systematic approach is still somewhat limited.
 
-To a functional programmer, this situation is disappointing. We should be able to
-compose a variety of charts from a small number of primitives. The shapes should be defined
-not in terms of pixels, but numbers of seats or exchange rates. We should also be able to easily
-derive high-level abstractions, such as a bar chart, from those low-level primitives.
+What would an elegant functional approach to data visualization look like? A functional programmer
+would want a domain-specific language that has a small number of primitives; allow us to define
+high level abstractions such as a bar chart and its basic building blocks are expressed in terms
+of domain values such as the exchange rate, rather than pixels.
 
-\begin{figure}
+\begin{figure}[h]
   \includegraphics[scale=0.57]{figures/charts}
@@
 
-In this paper, we...
+As is often the case with domain-specific languages, finding the right primitives is more of an art
+than science. For this reason, I present my answer -- a library named Compost -- as a functional pearl.
+I hope to convince the reader that Compost is elegant and I illustrate this with a wide range
+of examples. Compost has a number of specific desirable properties:
 
-\newpage
+\begin{itemize}
+\item Charts are composed from a small number of primitive building blocks using a small number of
+  combinators. In particular, concepts such as bar charts, line charts or charts with aligned
+  axes are expressed in terms of more basic concepts.
+\item The primitives are specified in domain terms. When drawing a line, the value of an Y coordinate
+  is an exchange rate of 1.36 USD/GBP, not 137 pixels from the bottom.
+\item Most common chart types can be easily captured as high level abstractions, but there is an
+  elengant way of creating a majority of more interesting custom charts.
+\item The approach can easily be extended to creating web-based charts that involve animations
+  or interaction with the user.
+\end{itemize}
+%
+The presentation in this paper focuses on explaining the primitives and combinators of the
+domain-specific language. I outline the structure of an implementation, but omit the details. Filling
+those in requires careful thinking about geometry and projections, but there are no unexpected
+surprises. A complete F\# implementation, including the examples used in this paper, is available
+at: \urrl{http://github.com/compostjs}.
 
-\section{Domain specific language}
-\label{sec:intro}
-\subsection*{Domain-specific values}
+\section{Composing visualization primitives}
+I will introduce individual features of the Compost library gradually. The first important aspect of
+Compost is that properties of shapes are defined in terms of domain-specific values. I first
+explain what this means and then use domain-specific values to specify the core part of the
+UK election results bar chart.
 
-All values that are provided when constructing charts are from the chart domain. For example,
-when visualizing the population of different countries, X values will be categorical values
-such as \strf{United Kingdom} or \strf{Czech Republic} and Y values will be continuous values
-such as $\num{66.04}$ million or $\num{10.58}$ million. Compost will automatically map the value
-from a domain value to a pixel value. The user does not need to do any rescaling to obtain
-value in pixels. This is akin to most high-level charting libraries, but in contrast with some
-more flexible libraries like, say, D3.
-Compost supports only two-dimensional charts and values can be either categorical (such as
-different countries) or continuous (such as population size). A value needs to determine an
-exact position in the space available for the chart. For continuous values, this simply means
-applying a linear transformation. For categorical values, the situation is more difficult. For % example,
-say we are drawing a bar chart with population of three different countries with countries
-on the X axis and population on the Y axis. The scale of the X axis is categorical and
-Compost divides the available space into three ranges of equal width. However, then \strf{United % Kingdom}
-or \strf{Czech Republic} does not determine an exact position but instead a range. To determine
-a unique position, we need to attach a value between $\num{0}$ and $\num{1}$ to the
-categorical value which identifies a relative position in the available range. This is
-illustrated in Figure~\ref{fig:scales}.
-\begin{figure}[!h]
-\includegraphics[scale=1,trim={0cm 6.5cm 6cm 0cm},clip]{values.pdf}
-\caption{Values on a continous scale and categorical scale.}
+\subsection{Domain-specific values}
+
+In the election results chart in Figure~\ref{fig:charts} (left), the X values are categorical
+values representing the political parties such as \strf{Conservative} or \strf{Labour}. The
+Y values are numerical values representing the number of seats won such as $\num{365}$ MPs.
+When creating data visualizations, those are the values that the user needs to specify. This is
+akin to most high-level charting libraries such as Google Charts, but in contrast with more
+flexible libraries like D3.
+
+Our design focuses on two-dimensional charts with X and Y axes. Values mapped to those axes
+can be either categorical (such as different political parties or countries) or continuous
+(such as number of votes or exchange rates). The mapping from categorical and continuous values
+to exact positions on the chart is done automatically. For continuous values, this simply means
+applying a linear transformation. For categorical values, the mapping is more difficult.
+
+For example, in the UK election results chart, the X axis is categorical. The library automatically
+divides the available space between the six categorical values (political parties). The value
+\strf{Green} does not determine an exact position on the axis, but rather a range. To determine
+an exact position, we also need to attach a value between $\num{0}$ and $\num{1}$ to the
+categorical value. This identifies a relative position in the available range.
+
+\begin{figure}
+\includegraphics[scale=1,trim={0cm 6.5cm 6cm 0cm},clip]{figures/values.pdf}
+\caption{On a continuous scale (above), an exact position is determined by a number.
+  On a categorical scale (below), an exact position is determined by the category and a
+  numerical ratio from 0 to 1.}
 \label{fig:scales}
 \end{figure}
-More formally, we write $c$ for categorical values such as countries (represented as country
-names using strings), $r$ for ratio between $0$ and $1$ and we write $n$ for numerical continuous
-values with any range. A value $v$ used in a Compost chart is defined as:
+
+Figure~\ref{fig:scales} illustrates the two kinds of values using the axes from the UK
+election results chart. More formally, we define a value $v$ as either a continuous value
+$\kvd{cont}~n$ containing any number $n$ or a categorical value $\kvd{cat}~c, r$, consisting
+of a categorical value $c$ (implemented as a string) and a ratio $r$ between $0$ and $1$:
 %
@@
 \end{equation*}
-%
-A value $v$ that is used to create Compost charts can be categorical, written as \kvd{cat} and
-consisting of the category and a ratio that determines point in the space allocated for the
-cateogry. A continous value, written as \kvd{cont}, consists of a number and will be mapped to
-a location based on the implicitly obtained re-scaling.
+
+\vspace{-2.5em}
+\subsection{Visualization primitives}
+
+\begin{equation*}
+\begin{array}{rcl}
+s & = & \kvd{overlay}~[\,s_1, \ldots, s_n\,]\\
+ & | & \kvd{text}~\gamma, v_x, v_y, t\\
+ & | & \kvd{line}~\gamma, v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\\
+ & | & \kvd{fill}~\gamma, v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\\
+ & | & \kvd{axis}_{x/y} s
+\end{array}
+\end{equation*}
+
+~
+
+xx
+
+~
+
+\begin{equation*}
+\begin{array}{l}
+  \kvd{axis}_x~(\kvd{axis}_y~(\kvd{overlay}~[\\
+  \quad \fkvd{for}~\ident{party}, \ident{clr17}, \ident{clr19}, \ident{mp17}, \ident{mp19}~\fkvd{in}~\ident{elections}~\rightarrow\\
+  \quad\quad \kvd{padding}~0,10,0,10,\,\kvd{overlay}~[\\
+  \quad\quad\quad \kvd{fill}~\ident{clr},~(\kvd{cat}~\ident{party}, \num{0}), (\kvd{cont}~\num{0}), (\kvd{cat}~\ident{party}, \num{0}), (\kvd{cont}~\ident{mp17}),\\
+  \quad\quad\quad \hspace{2.65em}           ~(\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\ident{mp17}), (\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\num{0})\\
+  \quad\quad\quad \kvd{fill}~\ident{clr},~(\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\num{0}), (\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\ident{mp19}),\\
+  \quad\quad\quad \hspace{2.65em}           ~(\kvd{cat}~\ident{party}, \num{1}), (\kvd{cont}~\ident{mp19}), (\kvd{cat}~\ident{party}, \num{1}), (\kvd{cont}~\num{0})~~]\\
+  ]\,)\,)
+\end{array}
+\end{equation*}
+
+\subsection{Inferring scales and projections}
+
+\section{Fancy scale features}
+
+\section{Definining abstractions}
+
+\section{Interactive charts}
+
+\section{Implementation structure}
+
+\newpage
+
+\section{Domain-specific language}
+\label{sec:intro}
+\subsection*{Domain-specific values}
+
 \subsection*{Language of composable charts}
@@
 \quad \kvd{fill}~\strf{\#0000ff},
- ~(\kvd{cat}~\strf{UK}, \num{0}), (\kvd{cont}~\num{00.00}), (\kvd{cat}~\strf{UK}, \num{0}), % (\kvd{cont}~\num{66.04}),\\
-\hspace{5.9em}(\kvd{cat}~\strf{UK}, \num{1}), (\kvd{cont}~\num{66.04}), (\kvd{cat}~\strf{UK}, % \num{1}), (\kvd{cont}~\num{00.00}),\\
+ ~(\kvd{cat}~\strf{UK}, \num{0}), (\kvd{cont}~\num{00.00}), (\kvd{cat}~\strf{UK}, \num{0}), (\kvd{cont}~\num{66.04}),\\
+\hspace{5.9em}(\kvd{cat}~\strf{UK}, \num{1}), (\kvd{cont}~\num{66.04}), (\kvd{cat}~\strf{UK}, \num{1}), (\kvd{cont}~\num{00.00}),\\
 \quad \kvd{fill}~\strf{\#ff0000},
- ~(\kvd{cat}~\strf{CZ}, \num{0}), (\kvd{cont}~\num{00.00}), (\kvd{cat}~\strf{CZ}, \num{0}), % (\kvd{cont}~\num{10.58}),\\
-\hspace{5.9em}(\kvd{cat}~\strf{CZ}, \num{1}), (\kvd{cont}~\num{10.58}), (\kvd{cat}~\strf{CZ}, % \num{1}), (\kvd{cont}~\num{00.00})\\
+ ~(\kvd{cat}~\strf{CZ}, \num{0}), (\kvd{cont}~\num{00.00}), (\kvd{cat}~\strf{CZ}, \num{0}), (\kvd{cont}~\num{10.58}),\\
+\hspace{5.9em}(\kvd{cat}~\strf{CZ}, \num{1}), (\kvd{cont}~\num{10.58}), (\kvd{cat}~\strf{CZ}, \num{1}), (\kvd{cont}~\num{00.00})\\
 \end{array}
@@
 library infers the scales for the X and Y axes and computes a projection function that can be
-used for mapping domain-specific values to pixels. While doing this, it also checks for consistency % --
+used for mapping domain-specific values to pixels. While doing this, it also checks for consistency --
 all values on one axis have to be either categorical or continuous, but not a mix of the two.
@@
 We first focus on the first three primitives, while the \kvd{nest} primitive will be discussed in
-the next section. The \kvd{axis} and \kvd{roundScale} constructs could be defined as derived % constructs,
-but it is easier to consider them as primitives for now. The \kvd{axis} primitive simply takes the % shape
-$s$ and draws an axis around the contents of $s$, using the inferred scale of $s$ to determine % values
+the next section. The \kvd{axis} and \kvd{roundScale} constructs could be defined as derived constructs,
+but it is easier to consider them as primitives for now. The \kvd{axis} primitive simply takes the shape
+$s$ and draws an axis around the contents of $s$, using the inferred scale of $s$ to determine values
 displayed on the axis. This could be a derived construct, because axes can be rendered using lines
@@
 \kvd{axis}_x~(\kvd{axis}_y~(\kvd{roundScale}_y\\
-\quad(\kvd{explicitScale}_x~(\kvd{categorical % \strf{CZ},\strf{MN},\strf{UK}})~\ident{populationBar})))
+\quad(\kvd{explicitScale}_x~(\kvd{categorical \strf{CZ},\strf{MN},\strf{UK}})~\ident{populationBar})))
 \end{array}
@@
 $v_{min}$ do not even have to be continuous. We can, for example, nest a line chart inside a
-bar of a bar chart and then the two values could be $\kvd{cat}~\strf{One}, 0$ and % $\kvd{cat}~\strf{One}, 1$
+bar of a bar chart and then the two values could be $\kvd{cat}~\strf{One}, 0$ and $\kvd{cat}~\strf{One}, 1$
 (which are two values that define a region of a categorical scale).
@@
  (\kvd{axis}_y~\ident{fbPrices})),\\
-\quad (\kvd{nest}_y~(\kvd{cat}~\strf{bottom-chart}, \num{0}), (\kvd{cat}~\strf{bottom-chart}, % \num{1}),
+\quad (\kvd{nest}_y~(\kvd{cat}~\strf{bottom-chart}, \num{0}), (\kvd{cat}~\strf{bottom-chart}, \num{1}),
  (\kvd{axis}_y~\ident{googPrices}))~)\\
@@
 of the line charts. Given the difference in the prices, the axes will have quite different values.
-We then nest the (continuous) Y axes using the \kvd{nest} primitive and, at the same time, % implicitly
+We then nest the (continuous) Y axes using the \kvd{nest} primitive and, at the same time, implicitly
 define a new outer Y axis. The outer Y axis is categorical with just two values, \strf{top-chart}
@@
 \begin{itemize}
-\item We could define a type system to catch categorical/continuous value mismatch in a single % scale.
+\item We could define a type system to catch categorical/continuous value mismatch in a single %scale.
 \item It would be worth thinking about ordinal values, which are categorical but can be sorted.
-\item There might be other kinds of scales - for example, color scale (can have meaning in scatter % plot)
- or a scale for secondary markers (like sizes of bubbles in a bubble chart). We could really say % that
+\item There might be other kinds of scales - for example, color scale (can have meaning in scatter plot)
+ or a scale for secondary markers (like sizes of bubbles in a bubble chart). We could really say that
  every value (including bar chart colors) should be coming from some scale...

### 2020-04-26 wip-compost/paper.tex
Scale inference and example

@@
 
-\section{Composing visualization primitives}
+\section{Composing simple visualizations}
 I will introduce individual features of the Compost library gradually. The first important aspect of
@@
 
-\subsection{Domain-specific values}
+\begin{figure}[t]
+\includegraphics[scale=1,trim={0cm 6.5cm 6cm 0cm},clip]{figures/values.pdf}
+\caption{On a continuous scale (above), an exact position is determined by a number.
+  On a categorical scale (below), an exact position is determined by the category and a
+  numerical ratio from 0 to 1.}
+\label{fig:scales}
+\end{figure}
+
+\subsection{Working with domain-specific values}
 
@@
 
+Figure~\ref{fig:scales} illustrates the two kinds of values using the axes from the UK
+election results chart. In Figure~\ref{fig:shape}, we define a value $v$ as either a continuous value
+$\kvd{cont}~n$ containing any number $n$ or a categorical value $\kvd{cat}~c, r$, consisting
+of a categorical value $c$ (implemented as a string) and a ratio $r$ between $0$ and $1$.
+%
 \begin{figure}
-\includegraphics[scale=1,trim={0cm 6.5cm 6cm 0cm},clip]{figures/values.pdf}
-\caption{On a continuous scale (above), an exact position is determined by a number.
-  On a categorical scale (below), an exact position is determined by the category and a
-  numerical ratio from 0 to 1.}
-\label{fig:scales}
+\begin{equation*}
+\begin{array}{rcl}
+v & = & \kvd{cat}~c, r \\
+  & | & \kvd{cont}~n\\
+  ~
+\end{array}
+\qquad
+\begin{array}{rclcl}
+s & = &\kvd{line}~\gamma, [\,v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\,] & | & \kvd{overlay}~[\,s_1, \ldots, s_n\,]\\
+ & | & \kvd{fill}~\gamma, [\,v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\,] & | &\kvd{axis}_{l/r/t/b}~s\\
+ & | & \kvd{text}~\gamma, v_x, v_y, t & | &\kvd{padding}~n_t,n_r,n_b,n_l,s
+\end{array}
+\end{equation*}
+\caption{Core primitives of the Compost domain-specific language. Values $v$ are either categorical
+  or continuous; a shape $s$ is then defined as a simple recursive algebraic data type.}
+\label{fig:shape}
 \end{figure}
 
-Figure~\ref{fig:scales} illustrates the two kinds of values using the axes from the UK
-election results chart. More formally, we define a value $v$ as either a continuous value
-$\kvd{cont}~n$ containing any number $n$ or a categorical value $\kvd{cat}~c, r$, consisting
-of a categorical value $c$ (implemented as a string) and a ratio $r$ between $0$ and $1$:
+\subsection{Introducing basic primitives and combinators}
+
+Now that we know how Compost represents values, we can define the basic elements of its
+domain-specific langauge. A chart is represeted by the shape $s$ defined in Figure~\ref{fig:shape}.
+A primitive shape can be a text label, a line connecting a list of points or a filled polygon
+defined by a list of points. The position of points is specified by X and Y coordinates,
+which can be either categorical or continuous values. For text, line and polygon, we also include
+a parameter $\gamma$ that specifies the element color.
+
+Figure~\ref{fig:shape} also defines three combinators. The most important is $\kvd{overlay}$,
+which overlays all shapes from a given list. When doing this, Compost automatically infers the
+scales of X and Y axes and calculates suitable projections using a method discussed in the next
+section. Finally, $\kvd{padding}$ adds padding around a specified shape and $\kvd{axis}$ adds
+an axis showing the inferred scale on the left, right, top or bottom of a given shape.
+Using those primitives, we can construct the simple UK election results bar chart in Figure~\ref{fig:simple} as follows:
 %
 \begin{equation*}
-\begin{array}{rcl}
-v & = & \kvd{cat}~c, r \lsep \kvd{cont}~n
+\begin{array}{l}
+\kvd{axis}_l~(\kvd{axis}_b~(\kvd{overlay}~[\\
+\quad \kvd{fill}~\strf{\#0000ff},
+ ~[\;\,(\kvd{cat}~\strf{Conservative}, \num{0}), (\kvd{cont}~\num{0}), (\kvd{cat}~\strf{Conservative}, \num{0}), (\kvd{cont}~\num{365}),\\
+\hspace{6.7em}\;(\kvd{cat}~\strf{Conservative}, \num{1}), (\kvd{cont}~\num{365}), (\kvd{cat}~\strf{Conservative}, \num{1}), (\kvd{cont}~\num{0})\;\,],\\
+\quad \kvd{fill}~\strf{\#ff0000},
+ ~[\;\,(\kvd{cat}~\strf{Labour}, \num{0}), (\kvd{cont}~\num{0}), (\kvd{cat}~\strf{Labour}, \num{0}), (\kvd{cont}~\num{202}),\\
+\hspace{6.7em}\;(\kvd{cat}~\strf{Labour}, \num{1}), (\kvd{cont}~\num{202}), (\kvd{cat}~\strf{Labour}, \num{1}), (\kvd{cont}~\num{0})\,\;]~]))\\
 \end{array}
@@
 
-\vspace{-2.5em}
-\subsection{Visualization primitives}
+\noindent
+The chart specification overlays two bars of different colors and then adds axes to the bottom and
+left of the chart. The two bars are filled rectangles defined using four corner points. The Y
+coordinates are specified as continuous values, while the X coordinates are categorical. For
+the Conservative party, two of the points have the Y coordinate set to $\kvd{cont}~\num{0}$ (bottom of the bar)
+and two have the Y coordinate set to $\kvd{cont}~\num{365}$ (top of the bar). The two X coordinates
+are the start and the end of the range allocated for the \strf{Conservative} category,
+i.e.~$\kvd{cat}~\strf{Conservative}, \num{0}$ on the left and $\kvd{cat}~\strf{Conservative}, \num{1}$
+on the right.
+
+\begin{figure}
+  \includegraphics[scale=0.57]{figures/simple}
+  \vspace{0.25em}
+  \caption{Two charts about the UK politics: Comparison of election results from 2017 and 2019 (left)
+    and GBP-USD exchange rate with highlighted areas before and after the 23 June 2016 Brexit vote.}
+  \label{fig:simple}
+\end{figure}
+
+Extending the snippet to generate a grouped bar chart that shows two results
+for each party as in Figure~\ref{fig:charts} is easy. Given a party $p$, we need to
+generate two rectangles, one with X coordinates $\kvd{cat}~p, 0$ and $\kvd{cat}~p, 0.5$
+and the other with X coordinates $\kvd{cat}~p, 0.5$ and $\kvd{cat}~p, 1$.
+In the following snippet, we use a \fkvd{for} comprehension to generate the list. All remaining
+constructs are primitives of the Compost domain-specific language. Assuming \ident{elections} is a
+list of election results containing a five-element tuple consisting of a party name, colors for
+2017 and 2019 and results for 2017 and 2019 we create the chart using:
+%
+\begin{equation*}
+\begin{array}{l}
+  \kvd{axis}_l~(\kvd{axis}_b~(\kvd{overlay}~[\\
+  \quad \fkvd{for}~\ident{party}, \ident{clr17}, \ident{clr19}, \ident{mp17}, \ident{mp19}~\fkvd{in}~\ident{elections}~\rightarrow\\
+  \quad\quad \kvd{padding}~0,10,0,10,\,\kvd{overlay}~[\\
+  \quad\quad\quad \kvd{fill}~\ident{clr17},~[\,(\kvd{cat}~\ident{party}, \num{0}), (\kvd{cont}~\num{0}), (\kvd{cat}~\ident{party}, \num{0}), (\kvd{cont}~\ident{mp17}),\\
+  \quad\quad\quad \hspace{4.08em}           ~(\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\ident{mp17}), (\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\num{0}) \,], \\
+  \quad\quad\quad \kvd{fill}~\ident{clr19},~[\,(\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\num{0}), (\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\ident{mp19}),\\
+  \quad\quad\quad \hspace{4.08em}           ~(\kvd{cat}~\ident{party}, \num{1}), (\kvd{cont}~\ident{mp19}), (\kvd{cat}~\ident{party}, \num{1}), (\kvd{cont}~\num{0})\,]~~]~~]\,)\,)\\
+\end{array}
+\end{equation*}
+
+\noindent
+Aside from iterating over all available parties and splitting the bar, the example also adds padding
+around the bars. A padding is specified in pixels rather than in terms of domain values. This is
+sometimes preferrable over, for example, drawing a bar using a range from $\num{0.05}$ to $\num{0.5}$.
+One remaining feature of the chart in Figure~\ref{fig:charts} that is still missing is the caption.
+We will add this in Section~\ref{sec:abstractions}.
+
+\subsection{Inferring scales and projections automatically}
+
+When composing shapes using the \kvd{overlay} primitive, the user does not need to specify how
+to position the child elements relatively to each other. The Compost library positions the elements
+automatically. This is done in two steps. First, Compost infers the \emph{scales} for X and Y
+axes. A scale represents the range of values that needs to fit in the space available for the chart.
+Second, Compost calculates a \emph{projection}, a mapping from domain-specific values of the scale
+to the available screen space. A scale $l$ can be either categorical or continuous:
 
@@
 \begin{array}{rcl}
-s & = & \kvd{overlay}~[\,s_1, \ldots, s_n\,]\\
- & | & \kvd{text}~\gamma, v_x, v_y, t\\
- & | & \kvd{line}~\gamma, v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\\
- & | & \kvd{fill}~\gamma, v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\\
- & | & \kvd{axis}_{x/y} s
+l & = & \kvd{continuous}~n_{min}, n_{max} \lsep \kvd{categorical}~[\,c_1, \ldots, c_k\,]
 \end{array}
@@
 
-~
+\vspace{-0.5em}
+\noindent
+A continuous scale is defined by a minimal and maximal value that need to be mapped to the
+available chart space. A categorical scale is defined by a list of individual categorical values.
+Note that we do not need a minimal and maximal ratios of the used categorical values as Compost
+will use an equal space for each category, regardless of where in this space a shape needs to
+appear.
 
-xx
-
-~
-
+Inferring scales is done by a simple recursive function that walks over the given shape and
+constructs two scales for the X and Y axis, using the X and Y coordinates that appear in the shape.
+Most of the work is done by a simple helper function that takes two scales, $l_1$ and $l_2$,
+and produces a new scale that represents the union of the two:
+%
 \begin{equation*}
 \begin{array}{l}
-  \kvd{axis}_x~(\kvd{axis}_y~(\kvd{overlay}~[\\
-  \quad \fkvd{for}~\ident{party}, \ident{clr17}, \ident{clr19}, \ident{mp17}, \ident{mp19}~\fkvd{in}~\ident{elections}~\rightarrow\\
-  \quad\quad \kvd{padding}~0,10,0,10,\,\kvd{overlay}~[\\
-  \quad\quad\quad \kvd{fill}~\ident{clr},~(\kvd{cat}~\ident{party}, \num{0}), (\kvd{cont}~\num{0}), (\kvd{cat}~\ident{party}, \num{0}), (\kvd{cont}~\ident{mp17}),\\
-  \quad\quad\quad \hspace{2.65em}           ~(\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\ident{mp17}), (\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\num{0})\\
-  \quad\quad\quad \kvd{fill}~\ident{clr},~(\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\num{0}), (\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\ident{mp19}),\\
-  \quad\quad\quad \hspace{2.65em}           ~(\kvd{cat}~\ident{party}, \num{1}), (\kvd{cont}~\ident{mp19}), (\kvd{cat}~\ident{party}, \num{1}), (\kvd{cont}~\num{0})~~]\\
-  ]\,)\,)
+\ident{union}~(\kvd{continuous}~n_l, n_h)~(\kvd{continuous}~n'_l, n'_h) =\\
+\qquad\kvd{continuous}~\min(n_l, n'_l), \max(n_h, n'_h)\\[0.5em]
+\ident{union}~(\kvd{categorical}~[\,c_1, \ldots, c_p\,])~(\kvd{categorical}~[\,c'_1, \ldots, c'_q\,]) =\\
+\qquad \kvd{categorical}~[\,c_1, \ldots, c_p\,] \;@\; [\,c'_i~|~ \forall i\in 1\ldots q, \nexists j. c_j = c'_i \;]
 \end{array}
@@
 
-\subsection{Inferring scales and projections}
+\vspace{-0.5em}
+\noindent
+When unioning two continuous scales, the minimum and maximum of the resulting scale is the smallest
+and largest of the two minimums and maximums, respectively. When unioning two categorical scales,
+we take all values of the first scale and append all values of the second scale that do not appear
+in the first one. Note that this means that the order of categorical values in a scale depends on
+the order in which they appear in the shape. (A possible improvement to Compost would be to support
+ordinal values, which are categorical values with a well-defined ordering.) It is also worth noting
+that a categorical scale cannot be combined with a continuous scale. In other words, mixing
+categorical and continuous values in a single scale results in an error.
 
-\section{Fancy scale features}
+Once Compost computes scales for the given shape and all its sub-shapes, it constructs a projection
+function that maps domain-specific values to the available chart space. I say more about this in
+Section~\ref{sec:impl}. For a continuous scale, the projection is a linear transformation. For
+categorical scale with $k$ values, we split the available chart space into $k$ equally sized
+regions and then map a categorical value $\kvd{cat}~c, r$ to the region corresponding to $c$
+according to the ratio $r$.
+
+\newpage
+
+\section{Composing advanced visualizations}
+\label{sec:fancy}
+
+\subsection{Creating scales with nice ranges}
+
+\subsection{Controlling the composition of scales}
+
+\newpage
 
 \section{Definining abstractions}
+\label{sec:abstractions}
 
@@
 \section{Implementation structure}
+\label{sec:impl}
 
@@
 
-\section{Domain-specific language}
-\label{sec:intro}
-\subsection*{Domain-specific values}
-
-\subsection*{Language of composable charts}
-Now that we defined our representation of values, we can define a basic composable language of
-data visualizations. A shape $s$ represents a chart. In the most basic form, it can be text label,
-line connecting a list of points and a filled polygon determined by a list of points. For line,
-text and polygon, we also include a parameter $\gamma$ indicating the color to be used. In addition,
-a shape can be constructed by overlaying several other shapes:
-%
-\begin{equation*}
-\begin{array}{rcl}
-s & = & \kvd{overlay}~s_1, \ldots, s_n\\
- & | & \kvd{text}~\gamma, v_x, v_y, t\\
- & | & \kvd{line}~\gamma, v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\\
- & | & \kvd{fill}~\gamma, v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}
-\end{array}
-\end{equation*}
-%
-As an example, consider bar chart with \strf{UK} and \strf{CZ} as two categorical values on the X
-axis and their population in millions (continuous value) on the Y axis. This can be constructed by
-creating two filled rectangles using \kvd{fill} and overlaying them using \kvd{overlay}:
-%
-\begin{equation*}
-\begin{array}{l}
-\kvd{overlay}\\
-\quad \kvd{fill}~\strf{\#0000ff},
- ~(\kvd{cat}~\strf{UK}, \num{0}), (\kvd{cont}~\num{00.00}), (\kvd{cat}~\strf{UK}, \num{0}), (\kvd{cont}~\num{66.04}),\\
-\hspace{5.9em}(\kvd{cat}~\strf{UK}, \num{1}), (\kvd{cont}~\num{66.04}), (\kvd{cat}~\strf{UK}, \num{1}), (\kvd{cont}~\num{00.00}),\\
-\quad \kvd{fill}~\strf{\#ff0000},
- ~(\kvd{cat}~\strf{CZ}, \num{0}), (\kvd{cont}~\num{00.00}), (\kvd{cat}~\strf{CZ}, \num{0}), (\kvd{cont}~\num{10.58}),\\
-\hspace{5.9em}(\kvd{cat}~\strf{CZ}, \num{1}), (\kvd{cont}~\num{10.58}), (\kvd{cat}~\strf{CZ}, \num{1}), (\kvd{cont}~\num{00.00})\\
-\end{array}
-\end{equation*}
-%
-The shape specification overlays two bars of different colours. To construct a rectangle, we
-define four points. For the UK, two of the points have Y value set to 0 (bottom of the bar) and
-two have it set to 66.04 (top of the bar indiciating the population value). The values on the X
-axis are categorical \strf{UK} values -- two are on the left as indicated by the ratio $\num{0}$
-and two are on the right as indicated by ratio $\num{1}$. The ratio is always relative to the
-category, so the ratios for the second bar are also $\num{0}$ and $\num{1}$.
-\subsection*{Calculating scales and projections}
-When processing shape specification such as the one given in the previous section, the Compost
-library infers the scales for the X and Y axes and computes a projection function that can be
-used for mapping domain-specific values to pixels. While doing this, it also checks for consistency --
-all values on one axis have to be either categorical or continuous, but not a mix of the two.
-The first step is to determine the scales for each of the axes. A scale can be continuous,
-defined by a minimal and maximal value, or a categorical, defined by a list of categorical values.
-Note that we do not need to track the ratios used as each category
-will be allocated equal amount of space, regardless of where in that space shapes need to
-appear. A scale $l$ is thus defined as:
-%
-\begin{equation*}
-\begin{array}{rcl}
-l & = & \kvd{categorical}~c_1, \ldots, c_k \lsep \kvd{continuous}~n_{min}, \ldots, n_{max}
-\end{array}
-\end{equation*}
-%
-A scale is obtained by recursively walking over the shape and gradually constructing two scales,
-for X and Y axis, from the X and Y coordinates that appear in the shape. The process is fairly
-straightforward and we won't discuss it in detail (for now). The first value encountered determines
-what kind of a scale will be constructed. If the type of a later value does not match the type of a
-scale, the process fails.
-Once we obtain scales for both X and Y axes, we calculate a projection function for each axis.
-The function takes a value $v$ together with total space available (say, in pixels) and produces
-a position in pixels corresponding to $v$. We also omit the details here (for now), but briefly --
-for continuous values, we produce a simple linear transformation; for categorical values, we split
-the available space into $k$ equally sized regions (where $k$ is the number of categorical values
-in the scale) and then map a categorical value $\kvd{cat}~c, r$ to the region corresponding to $c$
-according to the ratio $r$.
 \subsection*{Controlling and nesting scales}

### 2020-05-01 wip-compost/paper.tex
Nice ranges, code for nesting scales

@@
 
-\section{Composing simple visualizations}
+\section{Basic charts: Overlaying chart primitives}
 I will introduce individual features of the Compost library gradually. The first important aspect of
@@
 
-\subsection{Working with domain-specific values}
+\subsection{Domain-specific values}
 
@@
 
-\subsection{Introducing basic primitives and combinators}
+\subsection{Basic primitives and combinators}
 
@@
 an axis showing the inferred scale on the left, right, top or bottom of a given shape.
-Using those primitives, we can construct the simple UK election results bar chart in Figure~\ref{fig:simple} as follows:
+Using those primitives, we can construct the simple UK election results bar chart in Figure~\ref{fig:simple} (left).
+We use the \fkvd{let} construct of the host functional language to structure the code:
 %
@@
 \begin{array}{l}
-\kvd{axis}_l~(\kvd{axis}_b~(\kvd{overlay}~[\\
+\fkvd{let}~\ident{conservative},~\ident{labour}~=\\
 \quad \kvd{fill}~\strf{\#0000ff},
@@
  ~[\;\,(\kvd{cat}~\strf{Labour}, \num{0}), (\kvd{cont}~\num{0}), (\kvd{cat}~\strf{Labour}, \num{0}), (\kvd{cont}~\num{202}),\\
-\hspace{6.7em}\;(\kvd{cat}~\strf{Labour}, \num{1}), (\kvd{cont}~\num{202}), (\kvd{cat}~\strf{Labour}, \num{1}), (\kvd{cont}~\num{0})\,\;]~]))\\
+\hspace{6.7em}\;(\kvd{cat}~\strf{Labour}, \num{1}), (\kvd{cont}~\num{202}), (\kvd{cat}~\strf{Labour}, \num{1}), (\kvd{cont}~\num{0})\,\;]\\[0.5em]
+\kvd{axis}_l~(\kvd{axis}_b~(\kvd{overlay}~[~\ident{labour}, \ident{conservative}~]))\\
 \end{array}
@@
   \vspace{0.25em}
-  \caption{Two charts about the UK politics: Comparison of election results from 2017 and 2019 (left)
-    and GBP-USD exchange rate with highlighted areas before and after the 23 June 2016 Brexit vote.}
+  \caption{Simple chart showing the UK election results; using automatically inferred scales (left)
+    and using rounded Y scale and explicitly defined (reordered) X scale (right).}
   \label{fig:simple}
@@
 
-\subsection{Inferring scales and projections automatically}
+\subsection{Inferring scales and projections}
 
@@
 Second, Compost calculates a \emph{projection}, a mapping from domain-specific values of the scale
-to the available screen space. A scale $l$ can be either categorical or continuous:
+to the available screen space. A scale $l$ is defined in Figure~\ref{fig:scale}.
 
+\begin{figure}
 \begin{equation*}
@@
 \end{equation*}
+\caption{A scale $l$ can be continuous, defined by a range, or categorical, defined by a list of values.}
+\label{fig:scale}
+\end{figure}
 
-\vspace{-0.5em}
-\noindent
 A continuous scale is defined by a minimal and maximal value that need to be mapped to the
@@
 
-Once Compost computes scales for the given shape and all its sub-shapes, it constructs a projection
-function that maps domain-specific values to the available chart space. I say more about this in
+Once Compost computes scales for a given shape and its sub-shapes, it constructs a projection
+that maps domain-specific values to the available chart space. We discuss this in
 Section~\ref{sec:impl}. For a continuous scale, the projection is a linear transformation. For
@@
 
+\section{Advanced charts: Controlling scale composition}
+\label{sec:fancy}
+
+Most charts have one X and one Y axis that are determined by the values the chart shows,
+but there are interesting exceptions. The chart in Figure~\ref{fig:charts} (right)
+has two different Y axes, one for GBP-USD and one for GBP-EUR. In the next two sections, we look at
+three combinators that control the scale inferrence process and what flexibility this enables.
+
+\subsection{Defining nice scale ranges}
+
+The automatic scale inference often results in scales where the maximum is a non-round number.
+This leads to charts that fully utilize the available space, but may not be easy to read.
+The first two primitives, shown in Figure~\ref{fig:control} (left) allow the chart designer
+to adjust the automatically inferred range of scales.
+
+\begin{figure}
+  \begin{equation*}
+  \begin{array}{rclcl}
+  s & = & \kvd{roundScale}_{x/y} s      &|& \kvd{nest}_{x/y}~v_{min}, v_{max}, s \\
+    & | & \kvd{explicitScale}_{x/y}~l, s &|&  (\ldots)
+  \end{array}
+  \end{equation*}
+  \caption{Additional combinators for controlling scales; \kvd{roundScale} and \kvd{explicitScale}
+    can be used for creating nice scale ranges; \kvd{nest} makes it possible to create charts with
+    multiple scales.}
+\label{fig:control}
+\end{figure}
+
+The two combinators for controlling the range of scales are \kvd{roundScale} and \kvd{explicitScale}.
+The oprations can be applied to either the X scale or the Y scale, which is indicated by the
+$x/y$ sub-script. The \kvd{roundScale} primitve takes the inferred X or Y scale of the shape $s$
+and, if it is a continuous scale, rounds its minimal and maximal values to a ``nice'' number.
+For example, if a continuous scale has minimum $\num{0}$ and maximum $\num{365}$, the resulting
+scale would have a maximum $\num{400}$. For categorical scale, the operation does not have any effect.
+The \kvd{explicitScale} operation is similar, but it replaces the inferred scale with an explicitly
+provided scale (the type of the inferred scale has to match with the type of the explicitly given
+scale). For example, the chart in Figure~\ref{fig:simple} (right) is constructed using the
+following code (reusing the \ident{labour} and \ident{conservative} variables defined earlier):
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{axis}_l~(\kvd{axis}_b~(\kvd{roundScale}_y~(\kvd{explicitScale}_x~(\kvd{categorical}~[\,\strf{Labour},\strf{Conservative}\,]),\\
+\qquad \kvd{overlay}~[~\ident{labour}, \ident{conservative}~]~)))\\
+\end{array}
+\end{equation*}
+
+\noindent
+Reading the code from the inside out, the snippet first overlays the two coloured bars defined
+earlier; it then replaces the X axis with an explicitly given one that changes the order of the
+values. As a result, the bar for \strf{Labour} will appear on the left, even though the value
+comes later in the list of overlaid chart elements.
+
+The code next uses \kvd{roundScale} to automatically round the minimum and maximum of the
+continuous Y scale (showing the total number of seats). Finally, we add axes around the shape,
+producing a usual labelled chart.  It is worth noting that \kvd{axis} and \kvd{roundScale}
+could be implemented as derived operations; \kvd{roundScale} would need to infer the scale of
+the nested shape and then insert \kvd{explicitScale} with a rounded number; \kvd{axis}
+would also need to infer the scales and then generates labels and lines in suitable locations.
+
+\subsection{Nested scales}
+
 \newpage
 
-\section{Composing advanced visualizations}
-\label{sec:fancy}
+codezzz
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{overlay}~[~
+\kvd{nest}_y~(\kvd{continuous}~\num{0},\num{50}), (\kvd{axis}_l~(\kvd{axis}_r~(\kvd{axis}_b~(\kvd{line}~\strf{\#202020}~\ident{gbpusd}))))\\
+\hspace{3.68em} \kvd{nest}_y~(\kvd{continuous}~\num{50},\num{100}), (\kvd{axis}_l~(\kvd{axis}_r~(\kvd{axis}_b~(\kvd{line}~\strf{\#202020}~\ident{gbpeur}))))~]\\
+\end{array}
+\end{equation*}
 
-\subsection{Creating scales with nice ranges}
+moar codezzzz
 
-\subsection{Controlling the composition of scales}
+\begin{equation*}
+\begin{array}{l}
+\fkvd{let}~\ident{brexitRate}~(\ident{lo}, \ident{hi})~\ident{rates}~=\\
+\quad \kvd{overlay}~[\\
+\qquad \kvd{fill}~\strf{\#1F77B460}, [\;\,\kvd{cont}~\num{0}, \kvd{cont}~\ident{lo}, \kvd{cont}~\num{16}, \kvd{cont}~\ident{lo}, \kvd{cont}~\num{16}, \kvd{cont}~\ident{hi}, \kvd{cont}~\num{0}, \kvd{cont}~\ident{hi}\;],\\
+\qquad \kvd{fill}~\strf{\#D6272860}, [\;\,\kvd{cont}~\num{16}, \kvd{cont}~\ident{lo}, \kvd{cont}~\num{44}, \kvd{cont}~\ident{lo}, \kvd{cont}~\num{44}, \kvd{cont}~\ident{hi}, \kvd{cont}~\num{16}, \kvd{cont}~\ident{hi}\;],\\
+\qquad \kvd{line}~\strf{\#202020}~\ident{rates}~]\\[0.5em]
+\kvd{overlay}~[~
+\kvd{nest}_y~(\kvd{continuous}~\num{0},\num{50}), (\kvd{axis}_l~(\kvd{axis}_r~(\kvd{axis}_b~(\ident{brexitRate}~(\num{1.25}, \num{1.50})~\ident{gbpusd}))))\\
+\hspace{3.68em} \kvd{nest}_y~(\kvd{continuous}~\num{50},\num{100}), (\kvd{axis}_l~(\kvd{axis}_r~(\kvd{axis}_b~(\ident{brexitRate}~(\num{1.15}, \num{1.30})~\ident{gbpeur}))))~]\\
+\end{array}
+\end{equation*}
+
+\begin{figure}
+  \includegraphics[scale=0.57]{figures/lines}
+  \vspace{0.25em}
+  \caption{TBA.}
+  \label{fig:lines}
+\end{figure}
 
@@
 \subsection*{Controlling and nesting scales}
-The process of computing scales can be controlled by additional primitives in the Compost
-domain-specific language for describing shapes that were not discussed earlier. The following
-definition lists additional primitives that are also a part of the definition of $s$. Note that
-the primitives can be applied either to the X scale or to the Y scale - we write $x/y$ to indicate
-that a sub-script on those primitives can be set either to $x$ or to $y$ (both scales can be
-modified by nesting two primitives):
-%
-\begin{equation*}
-\begin{array}{rcl}
-s & = & (\ldots)\\
- & | & \kvd{axis}_{x/y} s\\
- & | & \kvd{roundScale}_{x/y} s\\
- & | & \kvd{explicitScale}_{x/y}~l, s\\
- & | & \kvd{nest}_{x/y}~v_{min}, v_{max}, s
-\end{array}
-\end{equation*}
-%
-We first focus on the first three primitives, while the \kvd{nest} primitive will be discussed in
-the next section. The \kvd{axis} and \kvd{roundScale} constructs could be defined as derived constructs,
-but it is easier to consider them as primitives for now. The \kvd{axis} primitive simply takes the shape
-$s$ and draws an axis around the contents of $s$, using the inferred scale of $s$ to determine values
-displayed on the axis. This could be a derived construct, because axes can be rendered using lines
-and text labels.
-The \kvd{roundScale} operation takes the inferred X or Y scale of the shape $s$ and, if it is a
-continuous scale, rounds its minimal and maximal values to ``nice'' numbers. For example, if a
-continuous scale has minimum $\num{0}$ and maximum $\num{66.04}$, the resulting scale would have
-maximum $\num{70}$. For categorical scale, the operation does not have any effect.
-The \kvd{explicitScale} operation is similar, but it replaces the inferred scale with an explicitly
-provided scale (the type of the inferred scale has to match with the type of the explicitly given
-scale). For example, assuming \ident{populationBar} is the example given earlier, we can write:
-%
-\begin{equation*}
-\begin{array}{l}
-\kvd{axis}_x~(\kvd{axis}_y~(\kvd{roundScale}_y\\
-\quad(\kvd{explicitScale}_x~(\kvd{categorical \strf{CZ},\strf{MN},\strf{UK}})~\ident{populationBar})))
-\end{array}
-\end{equation*}
-%
-This replaces the X axis with an explicitly given one that includes extra country code and also
-overrides the implicitly inferred order of the categorical values. We then ask Compost to
-automatically round the Y scale (showing population) so that we get a ``nice'' number as the
-maximum. Finally, we add axes around the shape, producing a usual labelled chart.
-As noted earlier, both \kvd{axis} and \kvd{roundScale} could be defined as derived -- \kvd{axis}
-would have to infer the scales of the nested shape, calculate appropriate labels and insert
-suitable lines and labels; the \kvd{roundScale} operation would need to infer the scale too and
-then add an explicit \kvd{explicitScale} with a rounded minimal and maximal value of a continuous
-scale.
+
 \begin{figure}[!h]
@@
 \end{figure}
+
 \subsection*{Nesting of charts and scales}

### 2020-05-02 wip-compost/paper.tex
Nesting of scales, abstractions code

@@
 
-\title{Composing data visualizations}
+\title{Composable data visualizations}
 \author[Tomas Petricek]{TOMAS PETRICEK\\
@@
 \end{equation*}
+\vspace{-1em}
 \caption{A scale $l$ can be continuous, defined by a range, or categorical, defined by a list of values.}
 \label{fig:scale}
+\vspace{-1em}
+\end{figure}
+\begin{figure}
+\begin{equation*}
+\begin{array}{rclcl}
+s & = & \kvd{roundScale}_{x/y} s      &|& \kvd{nest}_{x/y}~v_{min}, v_{max}, s \\
+  & | & \kvd{explicitScale}_{x/y}~l, s &|&  (\ldots)
+\end{array}
+\end{equation*}
+\vspace{-0.5em}
+\caption{Additional combinators for controlling and nesting scales, extending earlier definition of $s$.}
+\label{fig:control}
 \end{figure}
@@
 
-\begin{figure}
-  \begin{equation*}
-  \begin{array}{rclcl}
-  s & = & \kvd{roundScale}_{x/y} s      &|& \kvd{nest}_{x/y}~v_{min}, v_{max}, s \\
-    & | & \kvd{explicitScale}_{x/y}~l, s &|&  (\ldots)
-  \end{array}
-  \end{equation*}
-  \caption{Additional combinators for controlling scales; \kvd{roundScale} and \kvd{explicitScale}
-    can be used for creating nice scale ranges; \kvd{nest} makes it possible to create charts with
-    multiple scales.}
-\label{fig:control}
-\end{figure}
-
 The two combinators for controlling the range of scales are \kvd{roundScale} and \kvd{explicitScale}.
@@
 \end{equation*}
+\vspace{-0.5em}
 
@@
 
+\begin{figure}
+\includegraphics[scale=0.57]{figures/lines}
+\vspace{0.25em}
+\caption{Two charts showing currency exchange rates with a shared X scale and separate Y scales.}
+\label{fig:lines}
+\end{figure}
+
 \subsection{Nested scales}
 
-\newpage
+The most interesting primitive for controlling scale composition defined in Figure~\ref{fig:control}
+is $\kvd{nest}_{x/y}$. The combinator takes two values, $v_{min}, v_{max}$ and a shape $s$ as arguments
+and it nests the scale of the shape $s$ inside the region defined by $v_{min}, v_{max}$. When inferring
+scales of shapes, the scale of $\kvd{nest}_{x/y}~l, s$ will be a categorical or continuous scale
+inferred using the values $v_{min}$ and $v_{max}$, regardless of the values that are used inside
+the shape $s$. The chart space between $v_{max}$ and $v_{min}$ will then be used to render the
+nested shape $s$ using its inferred scale. An example of nesting is shown in Figure~\ref{fig:nesting}.
+Here, a chart with a continous scale from $\num{1.1}$ to $\num{1.4}$ (e.g.~GBP-EUR exchange rates)
+is nested in the left half of another chart, which has a continuous scale from $\num{0}$ to $\num{100}$.
 
-codezzz
+\begin{figure}
+\includegraphics[scale=1,trim={0cm 7.5cm 6cm 0cm},clip]{figures/nest}
+\caption{A continuous scale with values from $0$ to $6$, nested in another scale.}
+\label{fig:nesting}
+\end{figure}
+
+The nesting of scales can be used in a variety of ways. We can, for example, nest a line chart
+inside a bar of a bar chart. In that case, the values for $v_{min}$ and $v_{max}$ would be
+$\kvd{cat}~\strf{ABC}, 0$ and $\kvd{cat}~\strf{ABC}, 1$, which define the start and the end of
+the region allocated to the $\strf{ABC}$ category on a categorical scale. A simpler use case for
+the combinator is showing multiple charts in a single view. For example, the motivating example
+in Figure~\ref{fig:charts} (right) comapres aligned line charts of exchange rates for two
+different currencies. Assuming \ident{gbpusd} and \ident{gbpeur} are lists containing days as X
+values and exchange rates as Y values, we can construct a simple chart with two line charts,
+shown in Figure~\ref{fig:lines} (left), using:
 %
@@
 \kvd{overlay}~[~
-\kvd{nest}_y~(\kvd{continuous}~\num{0},\num{50}), (\kvd{axis}_l~(\kvd{axis}_r~(\kvd{axis}_b~(\kvd{line}~\strf{\#202020}~\ident{gbpusd}))))\\
-\hspace{3.68em} \kvd{nest}_y~(\kvd{continuous}~\num{50},\num{100}), (\kvd{axis}_l~(\kvd{axis}_r~(\kvd{axis}_b~(\kvd{line}~\strf{\#202020}~\ident{gbpeur}))))~]\\
+\kvd{nest}_y~(\kvd{cont}~\num{0}),(\kvd{cont}~\num{50}), (\kvd{axis}_l~(\kvd{axis}_r~(\kvd{axis}_b~(\kvd{line}~\strf{\#202020}~\ident{gbpusd}))))\\
+\hspace{3.68em} \kvd{nest}_y~(\kvd{cont}~\num{50}),(\kvd{cont}~\num{100}), (\kvd{axis}_l~(\kvd{axis}_r~(\kvd{axis}_b~(\kvd{line}~\strf{\#202020}~\ident{gbpeur}))))~]\\
 \end{array}
@@
 
-moar codezzzz
+\vspace{-0.5em}
+\noindent
+In this example, the X scale shows the days of the year. This scale is shared by both of the charts.
+Indeed, if data was only available for the second half of the month for one of the charts,
+we would want the line to start in the middle of the chart. However, the Y scale needs to be
+separate for each of the charts. To achieve this, we use $\kvd{nest}_y$. The scale of the inner
+shapes is continuous, from the minimal to the maximal exchange rate for a given period. The
+outer scale is determined by the explicitly defined points. For the upper chart, these are
+$\kvd{cont}~\num{0}$ and $\kvd{cont}~\num{50}$; for the lower chart, these are
+$\kvd{cont}~\num{50}$ and $\kvd{cont}~\num{100}$. The continuous values define a scale that only
+contain two shapes -- one in the upper half, one in the lower half -- and so the three numbers could
+have equally been, for example, $\num{0},\num{1}, \num{2}$. The outer scale used here is
+synthetic and it is not aligned with other chart elements. An example of a more complex chart
+that follows a similar style, but does not have synthetic outer scale would be pairplot from
+the seaborn Python library. % \cite
+
+For completeness, the following code snippet shows how to construct the full currency exchange
+rate chart shown in Figure~\ref{fig:lines} (right), including the blue and red background:
+%
+\begin{equation*}
+\begin{array}{l}
+\fkvd{let}~\ident{brexitRate}~(\ident{lo}, \ident{hi})~\ident{rates}~=\kvd{overlay}~[\\
+\quad \kvd{fill}~\strf{\#1F77B460}, [\;\,\kvd{cont}~\num{0}, \kvd{cont}~\ident{lo}, \kvd{cont}~\num{16}, \kvd{cont}~\ident{lo}, \kvd{cont}~\num{16}, \kvd{cont}~\ident{hi}, \kvd{cont}~\num{0}, \kvd{cont}~\ident{hi}\;],\\
+\quad \kvd{fill}~\strf{\#D6272860}, [\;\,\kvd{cont}~\num{16}, \kvd{cont}~\ident{lo}, \kvd{cont}~\num{44}, \kvd{cont}~\ident{lo}, \kvd{cont}~\num{44}, \kvd{cont}~\ident{hi}, \kvd{cont}~\num{16}, \kvd{cont}~\ident{hi}\;],\\
+\quad \kvd{line}~\strf{\#202020}~\ident{rates}~]\\[0.5em]
+\kvd{overlay}~[~
+\kvd{nest}_y~(\kvd{cont}~\num{0}),(\kvd{cont}~\num{50}), (\kvd{axis}_l~(\kvd{axis}_r~(\kvd{axis}_b~(\ident{brexitRate}~(\num{1.25}, \num{1.50})~\ident{gbpusd}))))\\
+\hspace{3.68em} \kvd{nest}_y~(\kvd{cont}~\num{50}),(\kvd{cont}~\num{100}), (\kvd{axis}_l~(\kvd{axis}_r~(\kvd{axis}_b~(\ident{brexitRate}~(\num{1.15}, \num{1.30})~\ident{gbpeur}))))~]\\
+\end{array}
+\end{equation*}
+
+\noindent
+Here, we use the \fkvd{let} binding of the host language to define a function that takes
+a the data \ident{rates} together with the minimum and maximum. This is used for drawing
+two filled rectangles, covering the first 16 days of the view in blue and the rest in red.
+The shapes combined using \kvd{overlay} are rendered in the order in which they appear and so
+the line shape is last, so that it appears above the background.
+
+\section{Definining abstractions}
+\label{sec:abstractions}
 
@@
 \begin{array}{l}
-\fkvd{let}~\ident{brexitRate}~(\ident{lo}, \ident{hi})~\ident{rates}~=\\
-\quad \kvd{overlay}~[\\
-\qquad \kvd{fill}~\strf{\#1F77B460}, [\;\,\kvd{cont}~\num{0}, \kvd{cont}~\ident{lo}, \kvd{cont}~\num{16}, \kvd{cont}~\ident{lo}, \kvd{cont}~\num{16}, \kvd{cont}~\ident{hi}, \kvd{cont}~\num{0}, \kvd{cont}~\ident{hi}\;],\\
-\qquad \kvd{fill}~\strf{\#D6272860}, [\;\,\kvd{cont}~\num{16}, \kvd{cont}~\ident{lo}, \kvd{cont}~\num{44}, \kvd{cont}~\ident{lo}, \kvd{cont}~\num{44}, \kvd{cont}~\ident{hi}, \kvd{cont}~\num{16}, \kvd{cont}~\ident{hi}\;],\\
-\qquad \kvd{line}~\strf{\#202020}~\ident{rates}~]\\[0.5em]
-\kvd{overlay}~[~
-\kvd{nest}_y~(\kvd{continuous}~\num{0},\num{50}), (\kvd{axis}_l~(\kvd{axis}_r~(\kvd{axis}_b~(\ident{brexitRate}~(\num{1.25}, \num{1.50})~\ident{gbpusd}))))\\
-\hspace{3.68em} \kvd{nest}_y~(\kvd{continuous}~\num{50},\num{100}), (\kvd{axis}_l~(\kvd{axis}_r~(\kvd{axis}_b~(\ident{brexitRate}~(\num{1.15}, \num{1.30})~\ident{gbpeur}))))~]\\
+\fkvd{let}~\ident{title}~t~s~=\ident{overlay}~[\\
+\quad \kvd{nest}_x~(\kvd{cont}~\num{0}), (\kvd{cont}~\num{100}),
+  (\kvd{nest}_y~(\kvd{cont}~\num{0}), (\kvd{cont}~\num{15}), \\
+\quad\quad \kvd{explicitScale}_x~(\kvd{continuous}~\num{0},\num{100}),~(\kvd{explicitScale}_y~(\kvd{continuous}~\num{0},\num{100}),\\
+\quad\qquad \kvd{text}~\strf{\#000000}, (\kvd{cont}~\num{50}), (\kvd{cont}~\num{50}),~t)~)\\
+\quad \kvd{nest}_x~(\kvd{cont}~\num{0}), (\kvd{cont}~\num{100}),
+  (\kvd{nest}_y~(\kvd{cont}~\num{15}), (\kvd{cont}~\num{100}),~s)~]
 \end{array}
@@
 
-\begin{figure}
-  \includegraphics[scale=0.57]{figures/lines}
-  \vspace{0.25em}
-  \caption{TBA.}
-  \label{fig:lines}
-\end{figure}
+\begin{equation*}
+\begin{array}{l}
+\fkvd{let}~\ident{pairplot}~\ident{attrs}~\ident{data}~=~\kvd{overlay}~[\\
+\quad \kvd{for}~x~\kvd{in}~\ident{attrs}~\rightarrow~\kvd{for}~y~\kvd{in}~\ident{attrs}~\rightarrow\\
+\qquad \kvd{nest}_x~(\kvd{cat}~x, 0), (\kvd{cat}~x, 1), (\kvd{nest}_y~(\kvd{cat}~y, 0), (\kvd{cat}~y, 1), \\
+\quad\qquad (\kvd{axis}_l~(\kvd{axis}_b~(\kvd{overlay}~[~\fkvd{for}~v~\kvd{in}~\ident{data}~\rightarrow~\kvd{bubble}~(\ident{get}~x~v), (\ident{get}~y~v), \num{1}, \num{1} ])))~)]\\
+\end{array}
+\end{equation*}
+
+let
+  Shape.Layered [
+    for x in ["sepal.length"; "sepal.width"; "petal.length"; "petal.width" ] do
+      for y in ["sepal.length"; "sepal.width"; "petal.length"; "petal.width" ] do
+        yield Shape.OuterScale(Some(Categorical [| ca x |]), Some(Categorical [| ca y |]),
+          Shape.Axes(false, false, true, true, Shape.Layered [
+            for v, k in iris ->
+              Derived.StrokeColor(colors.[k], Shape.Bubble(numv v.[x], numv v.[y], 1., 1.))
+          ]))
+  ] |> render "out3"
 
 \newpage
-
-\section{Definining abstractions}
-\label{sec:abstractions}
 
@@
 
-\begin{figure}[!h]
-\includegraphics[scale=1,trim={0cm 7.5cm 6cm 0cm},clip]{nest.pdf}
-\caption{A continuous scale with values from $0$ to $6$, nested in another scale.}
-\label{fig:nesting}
-\end{figure}
+\subsection*{Nesting of charts and scales}
 
-\subsection*{Nesting of charts and scales}
-Finally, one more primitive that we added in the above definition and that we have not explained
-yet is $\kvd{nest}_{x/y}$. The parameters of the primitive are two values, $v_{min}$ and $v_{max}$
-together with a shape $s$. The primitive makes it possible to nest one scale inside another
-one.
-When inferring the scales, Compost infers the scale of the nested shape $s$. This will then take
-the area determined by the two points $v_{min}$ and $v_{max}$ in the scale determined by other
-shapes that are overlaid with the $\kvd{nest}$ shape. The two points are also used to determine
-the scales of the overall shape, but the scales of $s$ are nested and do not have any effect on
-the outer scales. In fact, the scale of $s$ does not even have to be of the same type as the
-scale of the outer shape.
-Figure~\ref{fig:nesting} illustrates this with an example. Here, we are nesting a continuous scale
-with values ranging from $\num{0}$ to $\num{6}$ into another scale. The values $v_{max}$ and
-$v_{min}$ do not even have to be continuous. We can, for example, nest a line chart inside a
-bar of a bar chart and then the two values could be $\kvd{cat}~\strf{One}, 0$ and $\kvd{cat}~\strf{One}, 1$
-(which are two values that define a region of a categorical scale).
-One area where nesting of scales is particularly useful is when combining multiple charts in
-a single view. For example, let's say that we have two line charts showing prices of two different
-stocks over the same period of time (for simplicity, we can just use UNIX timestamp as a time).
-The prices of the two stocks are orders of magnitude different, so they cannot fit easily into the
-same chart. We want to show two line charts side-by-side (one above each other). They should each
-have their own Y axis (price), but the X axis (time) should be shared.
-Assuming we have \ident{googPrices} and \ident{fbPrices} as two line charts without any axes,
-covering the same date range, we can write:
-%
-\begin{equation*}
-\begin{array}{l}
-\kvd{axis}_x~(\kvd{overlay}\\
-\quad (\kvd{nest}_y~(\kvd{cat}~\strf{top-chart}, \num{0}), (\kvd{cat}~\strf{top-chart}, \num{1}),
- (\kvd{axis}_y~\ident{fbPrices})),\\
-\quad (\kvd{nest}_y~(\kvd{cat}~\strf{bottom-chart}, \num{0}), (\kvd{cat}~\strf{bottom-chart}, \num{1}),
- (\kvd{axis}_y~\ident{googPrices}))~)\\
-\end{array}
-\end{equation*}
-%
-Reading the code from the inner-most part to the outer-most, we first add separate Y axes to both
-of the line charts. Given the difference in the prices, the axes will have quite different values.
-We then nest the (continuous) Y axes using the \kvd{nest} primitive and, at the same time, implicitly
-define a new outer Y axis. The outer Y axis is categorical with just two values, \strf{top-chart}
-and \strf{bottom-chart}. This means that the two nested charts will take equal amount of space,
-one above the other (each of the charts takes the full space allocated for the category -- the
-minimal value has ratio $0$ while the maximal value has ratio $1$).
 ~

### 2020-05-09 wip-compost/paper.tex
New implementation approach, abstractions

@@
 Our design focuses on two-dimensional charts with X and Y axes. Values mapped to those axes
-can be either categorical (such as different political parties or countries) or continuous
-(such as number of votes or exchange rates). The mapping from categorical and continuous values
-to exact positions on the chart is done automatically. For continuous values, this simply means
+can be either categorical (e.g.~political parties, countries) or continuous
+(e.g.~number of votes, exchange rates). The mapping from categorical and continuous values
+to positions on the chart is done automatically. For continuous values, this involves
 applying a linear transformation. For categorical values, the mapping is more difficult.
@@
  & | & \kvd{fill}~\gamma, [\,v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\,] & | &\kvd{axis}_{l/r/t/b}~s\\
- & | & \kvd{text}~\gamma, v_x, v_y, t & | &\kvd{padding}~n_t,n_r,n_b,n_l,s
+ & | & \kvd{text}~\gamma, v_x, v_y, t & | &\kvd{padding}~n_t,n_r,n_b,n_l,s\\
+ & | & \kvd{bubble}~\gamma, v_x, v_y, w, h
 \end{array}
@@
 domain-specific langauge. A chart is represeted by the shape $s$ defined in Figure~\ref{fig:shape}.
-A primitive shape can be a text label, a line connecting a list of points or a filled polygon
-defined by a list of points. The position of points is specified by X and Y coordinates,
-which can be either categorical or continuous values. For text, line and polygon, we also include
-a parameter $\gamma$ that specifies the element color.
+A primitive shape can be a text label, a line connecting a list of points, a filled polygon
+defined by a list of points or a bubble at a given point with a given width and height.
+The position of points is specified by X and Y coordinates, which can be either categorical or
+continuous values. For text, line, polygon and bubble, we also include a parameter $\gamma$
+that specifies the element color. The width and height of a bubble is given in pixels.
 
@@
 sometimes preferrable over, for example, drawing a bar using a range from $\num{0.05}$ to $\num{0.5}$.
-One remaining feature of the chart in Figure~\ref{fig:charts} that is still missing is the caption.
-We will add this in Section~\ref{sec:abstractions}.
+The chart is still missing a title, which we add in Section~\ref{sec:abstractions}.
 
@@
 \begin{array}{l}
-\fkvd{let}~\ident{brexitRate}~(\ident{lo}, \ident{hi})~\ident{rates}~=\kvd{overlay}~[\\
+\fkvd{let}~\ident{xrate}~(\ident{lo}, \ident{hi})~\ident{rates}~=\kvd{overlay}~[\\
 \quad \kvd{fill}~\strf{\#1F77B460}, [\;\,\kvd{cont}~\num{0}, \kvd{cont}~\ident{lo}, \kvd{cont}~\num{16}, \kvd{cont}~\ident{lo}, \kvd{cont}~\num{16}, \kvd{cont}~\ident{hi}, \kvd{cont}~\num{0}, \kvd{cont}~\ident{hi}\;],\\
@@
 \kvd{overlay}~[~
-\kvd{nest}_y~(\kvd{cont}~\num{0}),(\kvd{cont}~\num{50}), (\kvd{axis}_l~(\kvd{axis}_r~(\kvd{axis}_b~(\ident{brexitRate}~(\num{1.25}, \num{1.50})~\ident{gbpusd}))))\\
-\hspace{3.68em} \kvd{nest}_y~(\kvd{cont}~\num{50}),(\kvd{cont}~\num{100}), (\kvd{axis}_l~(\kvd{axis}_r~(\kvd{axis}_b~(\ident{brexitRate}~(\num{1.15}, \num{1.30})~\ident{gbpeur}))))~]\\
+\kvd{nest}_y~(\kvd{cont}~\num{0}),(\kvd{cont}~\num{50}), (\kvd{axis}_l~(\kvd{axis}_r~(\kvd{axis}_b~(\ident{xrate}~(\num{1.25}, \num{1.50})~\ident{gbpusd}))))\\
+\hspace{3.68em} \kvd{nest}_y~(\kvd{cont}~\num{50}),(\kvd{cont}~\num{100}), (\kvd{axis}_l~(\kvd{axis}_r~(\kvd{axis}_b~(\ident{xrate}~(\num{1.15}, \num{1.30})~\ident{gbpeur}))))~]\\
 \end{array}
@@
 
-\section{Definining abstractions}
+\begin{figure}
+  \includegraphics[scale=0.57]{figures/standard}
+  \vspace{0.25em}
+  \caption{Sample charts built using derived abstractions; a scatter plot visualizing the Iris\\
+    dataset with a title (left) and a pairplot comparing two Iris features (right).}
+  \label{fig:standard}
+\end{figure}
+
+\section{Standard charts: Definining new abstractions}
 \label{sec:abstractions}
 
+The Compost library does not introduce a rich collection of standard charts and chart features
+as most charting libraries. However, the functional domain-specific language design makes
+it easy to define high-level chart features based on the low-level primitives of the core language.
+To illustrate this, we give two examples.
+
+First, one last remaining feature of the two charts in Figure~\ref{fig:charts} is a chart title.
+This can be added to any chart using the following derived combinator:
+%
 \begin{equation*}
 \begin{array}{l}
-\fkvd{let}~\ident{title}~t~s~=\ident{overlay}~[\\
+\fkvd{let}~\ident{title}~t~s~=\kvd{overlay}~[\\
 \quad \kvd{nest}_x~(\kvd{cont}~\num{0}), (\kvd{cont}~\num{100}),
@@
 
+\vspace{-0.5em}
+\noindent
+The \ident{title} combinator is a function defined using \fkvd{let} in the host language. It takes
+a title $t$ and a shape $s$. It overlays two shapes. To position the title above the chart, the first
+shape has an outer Y scale $\kvd{continuous}~\num{0},\num{15}$ while the second has an outer Y scale
+$\kvd{continuous}~\num{15},\num{100}$. These are defined using the $\kvd{nest}_y$ primitive. Similarly,
+the outer X scale of both is $\kvd{continuous}~\num{0},\num{100}$, defined using $\kvd{nest}_x$.
+
+The second shape simply wraps the specified chart $s$ to which we are attaching the title. The first
+positions the text title in the middle of the available space. To do so, we explicitly set the X and
+Y scales inside the upper shape to continuous scales from $\num{0}$ to $\num{100}$ and then position
+the text label in the middle, at a point $(\kvd{cont}~\num{50}), (\kvd{cont}~\num{50})$.
+Figure~\ref{fig:standard} (left) shows a sample scatter plot chart with a title created using the
+\ident{title} combinator.
+
+A more complex type of chart that can be easily composed using the Compost primitives is the
+pairplot chart from the seaborn Python library. % \cite
+Pairplot visualizes pairwise relationships between features of a dataset.
+An example using three features (sepal width, petal width, petal length) from the Iris dataset is shown
+in Figure~\ref{fig:standard} (right). % \cite
+A pairplot draws a grid of charts, each visualizing the relationship between two numerical features. For distinct
+features, pairplot shows a scatter plot using one fature for X values and the other for Y values.
+At the diagonal (when the fatures are the same), it draws a histogram of the values of the feature.
+Additionally, a categorical feature can be used to determine the color of dots in the scatter plots.
+
+To generate a pairplot, we use \kvd{nest} to overlay and align a grid of plots. Each of those
+overlays a number of bubbles or filled shapes and adds left and bottom axis. As before, we use
+\fkvd{let} to define a function and list comprehensions to generate individual chart elements.
+We assume that \ident{data} is a list of rows, \ident{attrs} is a list of available attributes and
+$\ident{get}~a~r$ obtains the attribute $a$ of a row $r$. We also assume the dataset contains the \str{color} attribute.
+%
 \begin{equation*}
@@
 \fkvd{let}~\ident{pairplot}~\ident{attrs}~\ident{data}~=~\kvd{overlay}~[\\
-\quad \kvd{for}~x~\kvd{in}~\ident{attrs}~\rightarrow~\kvd{for}~y~\kvd{in}~\ident{attrs}~\rightarrow\\
-\qquad \kvd{nest}_x~(\kvd{cat}~x, 0), (\kvd{cat}~x, 1), (\kvd{nest}_y~(\kvd{cat}~y, 0), (\kvd{cat}~y, 1), \\
-\quad\qquad (\kvd{axis}_l~(\kvd{axis}_b~(\kvd{overlay}~[~\fkvd{for}~v~\kvd{in}~\ident{data}~\rightarrow~\kvd{bubble}~(\ident{get}~x~v), (\ident{get}~y~v), \num{1}, \num{1} ])))~)]\\
+\quad \fkvd{for}~x~\fkvd{in}~\ident{attrs}~\rightarrow~\fkvd{for}~y~\fkvd{in}~\ident{attrs}~\rightarrow\\
+\qquad \kvd{nest}_x~(\kvd{cat}~x, 0), (\kvd{cat}~x, 1), (\kvd{nest}_y~(\kvd{cat}~y, 0), (\kvd{cat}~y, 1), \kvd{axis}_l~(\kvd{axis}_b\\
+\quad\qquad (~\fkvd{if}~x \neq y~\fkvd{then}~\kvd{overlay}~[\;\fkvd{for}~r~\fkvd{in}~\ident{data}~\rightarrow~\\
+\qquad\qquad \hspace{0.7em} \kvd{bubble}~(\ident{get}~\str{color}~r),~(\ident{get}~x~r), (\ident{get}~y~r), \num{1}, \num{1}~]\\
+ \quad\qquad \hspace{0.7em} \fkvd{else}~\kvd{overlay}~[\;\fkvd{for}~x_1, x_2, y~\fkvd{in}~\ident{bins}~x~\ident{data}~\rightarrow~\\
+\qquad\qquad \hspace{0.7em}  \kvd{fill}~\strf{\#808080}~[\; x_1, y, x_2, y, x_2, \num{0}, x_1, \num {0}\;]\;])))] \\
 \end{array}
@@
 
-let
-  Shape.Layered [
-    for x in ["sepal.length"; "sepal.width"; "petal.length"; "petal.width" ] do
-      for y in ["sepal.length"; "sepal.width"; "petal.length"; "petal.width" ] do
-        yield Shape.OuterScale(Some(Categorical [| ca x |]), Some(Categorical [| ca y |]),
-          Shape.Axes(false, false, true, true, Shape.Layered [
-            for v, k in iris ->
-              Derived.StrokeColor(colors.[k], Shape.Bubble(numv v.[x], numv v.[y], 1., 1.))
-          ]))
-  ] |> render "out3"
+\vspace{-0.5em}
+\noindent
+As before, \kvd{nest} is essential for composing individual charts. Here, the points that determine
+the locations of individual charts are categorical values defined by the attributes of the dataset.
+The choice between two possible nested charts is made using the host language \fkvd{if} construct.
+Scatter plots are generated by overlaying bubbles with X and Y coordinates obtained using $\ident{get}~x~r$
+and $\ident{get}~y~r$. Histograms are composed from filled shapes. To obtain their locations, we use
+a helper function $\ident{bins}~x~\ident{data}$, which returns a list of bins specified by a tripple
+consisting of a lower and an upper range $x_1, x_2$ and the count $y$.
+
+The example shows that the Compost is simple yet expressive. With just
+a few lines of code, we are able to construct charts that, in other systems, require dedicated
+libraries. The essential aspect of the language making this possible is the automatic inference
+of scales and their mapping to the available space as well as the \kvd{nest} operation.
+
+\section{Interactive charts}
 
 \newpage
-
-\section{Interactive charts}
 

### 2020-05-11 wip-compost/paper.tex
Interactivity

@@
 
-\section{Interactive charts}
+\section{Interactive charts: Domain-specific event handling}
+Data visualizations published on the web often also feature interactivity. Standard forms of
+interactivity include animations, hover labels or zooming. More interesting custom forms
+include the ``You draw It'' visualization introduced by the New York Times. % cite
+The chart shows only the first half of the data, such as a timeline, and the reader has to
+guess the second half before clicking a button and seeing the actual data. Standard forms of
+interactivity are supported by high-level libraries. Google Charts supports panning using
+drag \& drop, zooming to a selected chart range and animations. Custom interactivity can be
+implemented using low-level libraries such as D3, but doing so requires directly handling
+JavaScript events.
+
+\end{figure}
+\begin{figure}
+\begin{equation*}
+\begin{array}{rclcl}
+s & = & \kvd{mouseUp}~(\lambda \,x~y \rightarrow e), s  &|& \kvd{mouseDown}~(\lambda \,x~y\rightarrow e), s  \\
+    &|& \kvd{mouseMove}~(\lambda \,x~y \rightarrow e), s &|& (\ldots)
+\end{array}
+\end{equation*}
+\vspace{-0.5em}
+\caption{Additional combinators for mouse-based interaction, extending earlier definition of $s$.}
+\label{fig:mouse}
+\end{figure}
+
+Compost can be used in conjunction with a virtual-dom library to create interactive data
+visualizations based on the Elm architecture. An interactive visualization is described using
+a pair of functions -- the \emph{view} function creates a chart based on the current state and
+the \emph{update} function specifies how the state changes when an event occurs.
+
+To support handling of mouse-based events, Compost adds three additional primitives to the definition
+of shape, shown in Figure~\ref{fig:mouse}. The three new primitives make it possible to handle
+three common mouse events using custom functions $\lambda \;x~y\rightarrow e$, specified in the
+host language. The most interesting aspect is that the functions are given X and Y coordinates
+of the event specified in the domain of the chart. This means that if the user clicks on the
+bar representing the Conservative party in a bar chart, the values might be, for example,
+$\kvd{cat}~\strf{Conservative},\num{0.75}$ for X and $\kvd{cont}~\num{120.5}$ for Y.
+
+To illustrate building of interactive data visualizations using Compost, we look at one aspect of
+the ``You draw It'' visualizations. We want to create a bar chart where the user can use drag \& drop
+to move individual bars. The resulting chart can be seen in Figure~\ref{fig:youdraw}.
+The first step is to define types representing the state and events that can occur:
+%
+\begin{equation*}
+\begin{array}{lll}
+\fkvd{type}~\ident{State} &\narrow{=}& \ident{bool}~*~(\ident{string}~*~\ident{int})~\ident{list}\\[0.5em]
+\fkvd{type}~\ident{Event} &\narrow{=}& \ident{Update}~\fkvd{of}~(\ident{string}~*~\ident{int})\\
+                          &\;\narrow{|}& \ident{Moving}~\fkvd{of}~\ident{bool}
+\end{array}
+\end{equation*}
+
+yadda
+
+\begin{equation*}
+\begin{array}{l}
+\fkvd{let}~\ident{update}~(\fkvd{true},s)~(\ident{Update}(p, v)) =
+  \fkvd{true}, \ident{map}~(\lambda(k, o) \rightarrow k, \fkvd{if}~k = p~\fkvd{then}~v~\fkvd{else}~o)~s\\
+\hspace{1.05em}~\ident{update}~(m,s)~(\ident{Update}(\_, \_)) = m,s\\
+\hspace{1.05em}~\ident{update}~(\_,s)~(\ident{Moving}(m)) = m,s\\
+\end{array}
+\end{equation*}
+
+yadda
+
+\begin{equation*}
+\begin{array}{l}
+\fkvd{let}~\ident{render}~t~(m, s)~=\\
+\quad \kvd{axis}_l~(\kvd{axis}_b~(\kvd{explicitScale}_y~(\kvd{continuous}~\num{0},\num{400}),\\
+\qquad (~\kvd{mouseMove}~(\lambda \,(\kvd{cat}~p,\_)~(\kvd{cont}~v)\rightarrow \ident{trigger}(\ident{Update}(p, v))),\\
+\qquad (~\kvd{mouseUp}~(\lambda \,\_~\_\rightarrow \ident{trigger}(\ident{Moving}(\fkvd{true}))),\\
+\qquad (~\kvd{mouseDown}~(\lambda \,\_~\_\rightarrow \ident{trigger}(\ident{Moving}(\fkvd{false}))),~\kvd{overlay}~[\\
+\qquad \qquad \fkvd{for}~\ident{party}, \ident{mps}~\fkvd{in}~s~\rightarrow~\kvd{padding}~0,10,0,10,~\kvd{fill}~(\ident{color}~\ident{party}),\\
+\qquad \qquad\quad\quad [\,(\kvd{cat}~\ident{party}, \num{0}), (\kvd{cont}~\num{0}), (\kvd{cat}~\ident{party}, \num{0}), (\kvd{cont}~\ident{mp}),\\
+\qquad \qquad\quad\quad \hspace{0.38em}(\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\ident{mps}), (\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\num{0}) \,])~~])))))\\
+\end{array}
+\end{equation*}
 
 \newpage
+
+\begin{figure}[h]
+  \includegraphics[scale=0.57]{figures/youdraw}
+  \vspace{0.25em}
+  \caption{You draw.}
+  \label{fig:youdraw}
+\end{figure}
 

### 2020-05-17 wip-compost/paper.tex
Interactive vis and implementation

@@
 \subsection{Basic primitives and combinators}
+\label{sec:basic-primitives}
 
@@
 \subsection{Inferring scales and projections}
+\label{sec:basic-scales}
 
@@
 guess the second half before clicking a button and seeing the actual data. Standard forms of
-interactivity are supported by high-level libraries. Google Charts supports panning using
-drag \& drop, zooming to a selected chart range and animations. Custom interactivity can be
+interactivity are supported by high-level libraries. For example, Google Charts supports panning
+using drag \& drop, zooming to a selected chart range and animations. Custom interactivity can be
 implemented using low-level libraries such as D3, but doing so requires directly handling
 JavaScript events.
+
+\subsection{Domain-specific events}
+Compost can be used in conjunction with a virtual-dom library to create interactive data
+visualizations based on the Elm architecture. An interactive visualization is described using
+a pair of functions -- the \emph{view} function creates a chart based on the current state and
+the \emph{update} function specifies how the state changes when an event occurs.
+
+To support handling of mouse-based events, Compost adds three additional primitives to the definition
+of shape, shown in Figure~\ref{fig:mouse}. The three new primitives make it possible to handle
+three common mouse events using custom functions $\lambda \;x~y\rightarrow e$, specified in the
+host language. The most interesting aspect is that the functions are given X and Y coordinates
+of the event specified in the domain of the chart. This means that if the user clicks on the
+bar representing the Conservative party in a bar chart, the values might be, for example,
+$\kvd{cat}~\strf{Conservative},\num{0.75}$ for X and $\kvd{cont}~\num{120.5}$ for Y.
 
@@
 
-Compost can be used in conjunction with a virtual-dom library to create interactive data
-visualizations based on the Elm architecture. An interactive visualization is described using
-a pair of functions -- the \emph{view} function creates a chart based on the current state and
-the \emph{update} function specifies how the state changes when an event occurs.
-
-To support handling of mouse-based events, Compost adds three additional primitives to the definition
-of shape, shown in Figure~\ref{fig:mouse}. The three new primitives make it possible to handle
-three common mouse events using custom functions $\lambda \;x~y\rightarrow e$, specified in the
-host language. The most interesting aspect is that the functions are given X and Y coordinates
-of the event specified in the domain of the chart. This means that if the user clicks on the
-bar representing the Conservative party in a bar chart, the values might be, for example,
-$\kvd{cat}~\strf{Conservative},\num{0.75}$ for X and $\kvd{cont}~\num{120.5}$ for Y.
-
+\subsection{You draw it data visualizations}
 To illustrate building of interactive data visualizations using Compost, we look at one aspect of
@@
 
-yadda
+\vspace{-0.5em}
+\noindent
+The state is a pair of boolean, indicating whether the user is currently dragging, together with
+a list of key value pairs, storing the number of seats for each political party. Two types of events
+can occur in the visualization. First, the user may start or stop dragging, which is indicated using
+$\ident{Moving}(\fkvd{true})$ and $\ident{Moving}(\fkvd{false})$, respectively. Second, the user
+may change a value for a party, which is captured by the $\ident{Update}$ event.
 
+The next part of the visualization is the \ident{update} function which takes an old state together
+with an event and produces a new state:
+%
 \begin{equation*}
@@
 
-yadda
+\vspace{-0.5em}
+\noindent
+The first two cases handle the \ident{Update} event. The event carries two values, $p, v$, which
+represent the party (which bar the user is dragging) and the new value (new number of seats).
+If the user is currently dragging, we replace the value associated with the party $p$ in the
+list $s$ using the \ident{map} function. If the user is not currently dragging, the event is ingored.
+The last case handles the \ident{Moving} event, which simply replaces the first component of the
+state tuple, i.e.~a flag indicating whether a mouse button is down.
 
+\begin{figure}[t]
+  \includegraphics[scale=0.57]{figures/youdraw}
+  \vspace{0.25em}
+  \caption{Interactive ``You draw it'' data visualization. The user moves cursor to a bar (left),
+    pushes a mouse button and drags the bar to the position that they think is the correct one (right).}
+  \label{fig:youdraw}
+\end{figure}
+
+Finally, the \ident{render} function takes the current state and builds the data visualization
+using the Compost domain-specific language. In addition, it also takes a parameter \ident{trigger},
+which is a function of type $\ident{Event}\rightarrow\ident{unit}$ that can be used to trigger
+events in handlers, registered using primitives such as \kvd{mouseMove}.
+To build the bar chart in Figure~\ref{fig:youdraw}, we use the same approach as in
+Section~\ref{sec:basic-primitives}. The only addition are the event handlers registered using
+\kvd{mouseMove}, \kvd{mouseUp} and \kvd{mouseDown}:
+%
 \begin{equation*}
 \begin{array}{l}
-\fkvd{let}~\ident{render}~t~(m, s)~=\\
+\fkvd{let}~\ident{render}~\ident{trigger}~(\_, \ident{state})~=\\
 \quad \kvd{axis}_l~(\kvd{axis}_b~(\kvd{explicitScale}_y~(\kvd{continuous}~\num{0},\num{400}),\\
@@
 \qquad (~\kvd{mouseDown}~(\lambda \,\_~\_\rightarrow \ident{trigger}(\ident{Moving}(\fkvd{false}))),~\kvd{overlay}~[\\
-\qquad \qquad \fkvd{for}~\ident{party}, \ident{mps}~\fkvd{in}~s~\rightarrow~\kvd{padding}~0,10,0,10,~\kvd{fill}~(\ident{color}~\ident{party}),\\
+\qquad \qquad \fkvd{for}~\ident{party}, \ident{mps}~\fkvd{in}~\ident{state}~\rightarrow~\kvd{padding}~0,10,0,10,~(\kvd{fill}~(\ident{color}~\ident{party}),\\
 \qquad \qquad\quad\quad [\,(\kvd{cat}~\ident{party}, \num{0}), (\kvd{cont}~\num{0}), (\kvd{cat}~\ident{party}, \num{0}), (\kvd{cont}~\ident{mp}),\\
@@
 
-\newpage
+\vspace{-0.5em}
+\noindent
+When the user interacts with the visualization created using Compost, the library
+translates the coordinates associated with events from pixels to domain-specific
+values. In case of the above bar chart, when the user moves a mouse, the function registed
+using \kvd{mouseMove} is given a categorical value $\kvd{cat}~p, r$ as the X coordinate and
+a continuous value $\kvd{cont}~v$ as the Y coordinate. It then takes $p$, which is the name
+of the party corresponding to the bar and the value $v$ corresponding to the number of seats
+and triggers the $\ident{Update}(p, v)$ event to update the state. The handlers for
+\kvd{mouseUp} and \kvd{mouseDown} do not use the coordinates. They simply switch the flag
+indicating whether the user is currently dragging or not.
 
-\begin{figure}[h]
-  \includegraphics[scale=0.57]{figures/youdraw}
-  \vspace{0.25em}
-  \caption{You draw.}
-  \label{fig:youdraw}
-\end{figure}
+The pair of functions, \ident{update} and \ident{render}, together with an initial state is
+all that is needed to create an interactive data visualization. Compost calls
+\ident{render} each time the state changes and uses virtual-dom to update the chart
+displayed in the browser. Although creating an interactive visualization is more work than
+creating a static one, the domain-specific nature of Compost is invaluable. We can
+simply take the values $p$ and $v$ produced by a mouse event, use those to update the state
+and then, again, render an updated chart.
 
@@
 
+The Compost library is open-source and the full source code, implmented in the functional
+language F\#, can be found at \urrl{http://github.com/compostjs}.
+As is often the case with functional domain-specific languages, the implementation is easy once
+we find the right collection of basic primitives. This largely applies to the Compost library
+and so we will not go into the implementation details. However, it is worth giving an
+outline of the implmentation structure.
+
+As mentioned in Section~\ref{sec:basic-scales}, the rendering of shapes proceeds in two stages.
+First, the library infers the scales of a shape. When doing so, it also annotates some shapes
+with additional information that are needed later for rendering. Second, the library projects
+the shape onto an available space and produces the chart, represented as an SVG object.
+
+\subsection{Inferring the scales of a shape}
+
+\begin{equation*}
+\begin{array}{rclcl}
+s &=& \kvd{scaledNest}_x~v_{min},\;v_{max},\;s_x,\;s &|& (\ldots)
+\end{array}
+\end{equation*}
+
+\begin{equation*}
+\ident{calculateScales} ~:~ \ident{Shape} \rightarrow \ident{Scale}*\ident{Scale}\rightarrow\ident{Shape}
+\end{equation*}
+
+\begin{equation*}
+\begin{array}{l}
+\ident{calculateScales}~(\kvd{nest}_x~v_{min},\;v_{max},\;s) = \\
+\quad\fkvd{let}~(s_x, s_y),s' = \ident{calculateScales}~s\\
+\quad(\ident{union}~(\ident{singleton}~v_{min})~(\ident{singleton}~v_{max}), s_y),~\kvd{scaledNest}_x~v_{min},\;v_{max},\;s_x,\;s'\\
+\\
+\ident{calculateScales}~(\kvd{overlay}~l) = \\
+\quad\fkvd{let}~\ident{scales},\;l'~=~\ident{unzip}~(\ident{map}~\ident{calculateScales}~l)\\
+\quad\fkvd{let}~s_x,s_y~=~\ident{unzip}~\ident{scales}\\
+\quad(\ident{reduce}~\ident{union}~s_x,~\ident{reduce}~\ident{union}~s_y), \kvd{overlay}~l'
+\end{array}
+\end{equation*}
+
+\subsection{Drawing shapes}
+
+\begin{equation*}
+\ident{project}~:~\ident{float}*\ident{float}\rightarrow \ident{Scale}\rightarrow\ident{Value} \rightarrow \ident{float}
+\end{equation*}
+
+\begin{equation*}
+\ident{drawShape}~:~\ident{Area} \rightarrow \ident{Scale}*\ident{Scale}\rightarrow\ident{Shape}\rightarrow\ident{Svg}
+\end{equation*}
+
+\begin{equation*}
+\begin{array}{l}
+\ident{drawShape}~(x_1, y_1, x_2, y_2)~(s_x, s_y)~(\kvd{scaledNest}_x~v_{min},\;v_{max},\;{is}_x,\;\ident{shape}) = \\
+\quad\fkvd{let}~x_1' = \ident{project}~(x_1, x_2)~s_x~v_{min}\\
+\quad\fkvd{let}~x_2' = \ident{project}~(x_1, x_2)~s_x~v_{max}\\
+\quad\ident{drawShape}~(x_1', y_1, x_2', y_2)~({is}_x, s_y)~\ident{shape}\\
+\\
+\ident{drawShape}~a~s~(\kvd{overlay}~l) = \\
+\quad\ident{append}~(\ident{map}~(\ident{drawShape}~a~s)~l)
+
+\end{array}
+\end{equation*}
+
+a
+
+b
+
+c
+
+d
+
+e
+
+f
+
+g
+
+h
+
+i
 \newpage

### 2020-05-23 wip-compost/paper.tex
Implementation, working on references

@@
 \documentclass{jfp}
+\usepackage{amssymb}
 \usepackage{xcolor}
@@
 % \end{abstract}
-
-\bibliographystyle{jfp}
 
@@
 \end{equation*}
-\vspace{-0.5em}
 \caption{Additional combinators for controlling and nesting scales, extending earlier definition of $s$.}
@@
 \end{equation*}
-\vspace{-0.5em}
 \caption{Additional combinators for mouse-based interaction, extending earlier definition of $s$.}
+\vspace{0.5em}
 \label{fig:mouse}
@@
 \begin{array}{l}
-\fkvd{let}~\ident{update}~(\fkvd{true},s)~(\ident{Update}(p, v)) =
-  \fkvd{true}, \ident{map}~(\lambda(k, o) \rightarrow k, \fkvd{if}~k = p~\fkvd{then}~v~\fkvd{else}~o)~s\\
+\fkvd{let}~\ident{update}~(\_,s)~(\ident{Moving}(m)) = m,s\\
+\hspace{1.05em}~\ident{update}~(\fkvd{true},s)~(\ident{Update}(p, v)) =\\
+\hspace{1.05em}\qquad  \fkvd{true}, \ident{map}~(\lambda(k, o) \rightarrow k, \fkvd{if}~k = p~\fkvd{then}~v~\fkvd{else}~o)~s\\
 \hspace{1.05em}~\ident{update}~(m,s)~(\ident{Update}(\_, \_)) = m,s\\
-\hspace{1.05em}~\ident{update}~(\_,s)~(\ident{Moving}(m)) = m,s\\
 \end{array}
@@
 represent the party (which bar the user is dragging) and the new value (new number of seats).
-If the user is currently dragging, we replace the value associated with the party $p$ in the
+If the user is currently dragging, we replace the value associated with the party $p$ in \newpage \noindent the
 list $s$ using the \ident{map} function. If the user is not currently dragging, the event is ingored.
@@
 \begin{figure}[t]
+  \vspace{0.5em}
   \includegraphics[scale=0.57]{figures/youdraw}
-  \vspace{0.25em}
+  \vspace{1em}
   \caption{Interactive ``You draw it'' data visualization. The user moves cursor to a bar (left),
@@
 
-\section{Implementation structure}
+\section{Implementation structure: Scale inference and projection}
 \label{sec:impl}
 
-The Compost library is open-source and the full source code, implmented in the functional
-language F\#, can be found at \urrl{http://github.com/compostjs}.
+Compost is an open-source library, implmented in the functional language F\#. The full source
+code can be found at \urrl{http://github.com/compostjs}.
 As is often the case with functional domain-specific languages, the implementation is easy once
-we find the right collection of basic primitives. This largely applies to the Compost library
-and so we will not go into the implementation details. However, it is worth giving an
-outline of the implmentation structure.
+we find the right collection of basic primitives and the right structure for the implementation.
+This largely applies to the Compost library and so we will not go into the implementation details.
+It is, however, worth giving an outline of the implmentation structure.
 
@@
 
+In order to render a shape, we need to know the range of values that should appear on the X and
+Y axes. This is done by inferring a scale for each of the axes from the individual X and Y
+coordinates that specify shape locations. As discussed earlier, a scale can be either categorical
+(displaying only categorical values) or contiuous (displaying only continuous values). When
+inerring scales, we use two helper operations; \ident{union}, discussed earlier, combines two
+scales and \ident{singleton} creates a scale from a single coordinate.
+
+The operation that infers the scales of a shape is \ident{calculateScales}. It takes a shape
+and produces a pair of X and Y scales, together with a transformed shape:
+%
+\begin{equation*}
+  \ident{calculateScales} ~:~ \ident{Shape} \rightarrow (\ident{Scale}*\ident{Scale}) * \ident{Shape}
+\end{equation*}
+
+\vspace{-1em}
+\noindent
+The operation does not need to transform the shape in most cases. The exception is the
+shape $\kvd{nest}_{x/y}~v_{min}, v_{max}, s$. In this case, the returned scale is based solely on the
+values of $v_{min}$ and $v_{max}$. For rendering we need to keep the inferred scales of the nested
+shape $s$. To do so, the operation replaces the $\kvd{nest}_{x/y}$ shape with an auxiliary shape
+$\kvd{scaledNest}_{x/y}$:
+%
 \begin{equation*}
 \begin{array}{rclcl}
-s &=& \kvd{scaledNest}_x~v_{min},\;v_{max},\;s_x,\;s &|& (\ldots)
+s &=& \kvd{scaledNest}_{x/y}~v_{min},\;v_{max},\;s_{x/y},\;s &|& (\ldots)
 \end{array}
@@
 
-\begin{equation*}
-\ident{calculateScales} ~:~ \ident{Shape} \rightarrow \ident{Scale}*\ident{Scale}\rightarrow\ident{Shape}
-\end{equation*}
-
+\vspace{-1em}
+\noindent
+There are two kinds of cases handled by \ident{calculateScales}. For primitives, it constructs
+a pair of scales from individual coordinates using \ident{union} and \ident{singleton}. For
+shapes containing a sub-shape, the operation calculates the scales of a sub-shape recursively
+and then adapts those somehow. To illustrate, we consider two interesting cases:
+%
 \begin{equation*}
@@
 
-\subsection{Drawing shapes}
+\vspace{-0.5em}
+\noindent
+When calculating the scales of the $\kvd{nest}_x$, the function first calculate scales of the
+sub-shape $s$ recursively. The resulting Y scale $s_y$ is returned as the result, while the X
+scale is obtained from the two coordinates $v_{min}$ and $v_{max}$. This is also the case where
+the shape is transformed and the returned $\kvd{scaledNest}_x$ shape stores the inferred X scale
+$s_x$ of the sub-shape $s$. The second example is the $\kvd{overlay}$ case which recursively
+infers scales of all sub-shapes and combines those using the list folding function
+\ident{reduce} with $\ident{union}$ as an argument.
 
+\subsection{Projecting coordinates and drawing}
+The key operation that needs to be performed when drawing a shape is projecting coordinates from
+domain-specific values to the screen coordinates. As we draw a shape, we keep the X and Y scale
+and the space in pixels that it should be drawn on. Initially, the X and Y scales are those
+inferred for the entire shape and the space in pixels is $\num{0}\ldots\ident{width}$ and
+$\num{0}\ldots\ident{height}$ where $\ident{width}\times\ident{height}$ is the size of the target
+SVG element.
+
+The key calculation is done by the \ident{project} function, which takes the space in pixels (as a
+pair of floating-point numbers representing the range), the current scale and a domain-specific value
+and produces a coordinate in pixels:
+%
 \begin{equation*}
@@
 
+\vspace{-1.25em}
+\noindent
+The function is only defined if the value and scale are compatible. If both are continuous, the
+function performs a simple linear transformation. If both are categorical, the available pixel
+space is divided into a equally-sized bins, one for each categorical value on the scale, and the
+value is then projected into the appropriate bin.
+
+The drawing of shapes is done by a function that takes the available area as a quadruple $(x_1, y_1), (x_2, y_2)$
+together with the X and Y scale corresponding to the area and a shape to be drawn. The result is
+a data structure representing a SVG document:
+%
 \begin{equation*}
-\ident{drawShape}~:~\ident{Area} \rightarrow \ident{Scale}*\ident{Scale}\rightarrow\ident{Shape}\rightarrow\ident{Svg}
+\ident{drawShape}~:~(\ident{float}*\ident{float})*(\ident{float}*\ident{float}) \rightarrow \ident{Scale}*\ident{Scale}\rightarrow\ident{Shape}\rightarrow\ident{Svg}
 \end{equation*}
 
+\vspace{-1.25em}
+\noindent
+For primitive shapes, the operation projects the coordinates using \ident{project} and construct
+a corresponding SVG document. For shapes with sub-shapes it calls itself recursively, possibly
+with an adjusted scale or area. The two cases discussed earlier illustrate this:
+%
 \begin{equation*}
 \begin{array}{l}
-\ident{drawShape}~(x_1, y_1, x_2, y_2)~(s_x, s_y)~(\kvd{scaledNest}_x~v_{min},\;v_{max},\;{is}_x,\;\ident{shape}) = \\
+\ident{drawShape}~a~s~(\kvd{overlay}~l) = \\
+\quad\ident{concat}~(\ident{map}~(\ident{drawShape}~a~s)~l)\\
+\\[-0.5em]
+\ident{drawShape}~((x_1, y_1), (x_2, y_2))~(s_x, s_y)~(\kvd{scaledNest}_x~v_{min},\;v_{max},\;{ns}_x,\;\ident{shape}) = \\
 \quad\fkvd{let}~x_1' = \ident{project}~(x_1, x_2)~s_x~v_{min}\\
 \quad\fkvd{let}~x_2' = \ident{project}~(x_1, x_2)~s_x~v_{max}\\
-\quad\ident{drawShape}~(x_1', y_1, x_2', y_2)~({is}_x, s_y)~\ident{shape}\\
-\\
-\ident{drawShape}~a~s~(\kvd{overlay}~l) = \\
-\quad\ident{append}~(\ident{map}~(\ident{drawShape}~a~s)~l)
-
+\quad\ident{drawShape}~((x_1', y_1), (x_2', y_2))~({ns}_x, s_y)~\ident{shape}\\
 \end{array}
@@
 
-a
+\vspace{-0.5em}
+\noindent
+When drawing \kvd{overlay}, the function draws all sub-shapes onto the same
+area using the same scales and then concatenates the returned SVG components the \ident{concat} helper.
+The $\kvd{scaledNest}_x$ case is more illuminating. Here, we first
+use \ident{project} to find the range $x_1', x_2'$ corresponding to the domain-values
+$v_{min},v_{max}$. This defines the area corresponding to the nested scale ${ns}_x$, onto
+which the X coordinates in the sub-shape $\ident{shape}$ should be projected.
+To do this, we recursively call \ident{drawShape}, but use $x_1'$ and $x_2'$ as the X coordinates
+of the target area and ${ns}_x$ as the X scale. The Y area and scales are propagated unchanged.
 
-b
-
-c
-
-d
-
-e
-
-f
-
-g
-
-h
-
-i
 \newpage
 
-\subsection*{Controlling and nesting scales}
-
-\subsection*{Nesting of charts and scales}
-
-~
-~
-\noindent
-[TODO: We really need illustrations for those example charts, but that would be more work!]
-~
-[TODO: Maybe use this chart as a motivation: https://wellcomeopenresearch.org/articles/4-63/v1]
-~
-\section{Random ideas}
+\section{Conclusions}
 \begin{itemize}
@@
 \end{itemize}
+
+\cite{gcharts}
+\cite{seaborn}
+\cite{youdraw}
+\cite{grammar}
+\cite{layered}
+\cite{vega-lite}
+\cite{vega}
+\cite{d3}
+\cite{polaris}
+
+\bibliographystyle{jfp}
+\bibliography{paper}
+
 \end{document}
-
 %
-
-% \usepackage{booktabs}
-% \usepackage{subcaption}
-% \usepackage{lineno,hyperref,xcolor}
-% \usepackage{flushend}
-% \usepackage{stmaryrd}
-% \usepackage{amssymb}
-% \usepackage{xypic}
-% \usepackage{semantic}
-% \usepackage{booktabs}
-% \usepackage{enumitem}
-% \usepackage{enumerate}
-% \usepackage{amsmath}
-% \usepackage[T1]{fontenc}
-%
-% \setlist{leftmargin=6mm}
-% \newcounter{thc}
-% \newcounter{dfc}
-%
-% \theoremstyle{plain}
-% \newtheorem{lem}[thc]{Lemma}
-% \newtheorem{theorem}[thc]{Theorem}
-%
-% \theoremstyle{definition}
-% \newtheorem{axiom}[dfc]{Axiom}
-% \newtheorem{definition}[dfc]{Definition}
-%
-% \title{Compost: Library for composable data visualization}
-% %\author{Tomas Petricek}
-% %\affiliation{
-% %  \institution{University of Kent}
-% %  \country{United Kingdom}
-% %}
-% %\email{tomas@tomasp.net}
-%
-% %\begin{CCSXML}
-% %<ccs2012>
-% %<concept>
-% %<concept_id>10011007.10011006.10011008</concept_id>
-% %<concept_desc>Software and its engineering~General programming languages</concept_desc>
-% %<concept_significance>500</concept_significance>
-% %</concept>
-% %<concept>
-% %<concept_id>10003456.10003457.10003521.10003525</concept_id>
-% %<concept_desc>Social and professional topics~History of programming languages</concept_desc>
-% %<concept_significance>300</concept_significance>
-% %</concept>
-% %</ccs2012>
-% %\end{CCSXML}
-% %
-% %\ccsdesc[500]{Software and its engineering~General programming languages}
-% %\ccsdesc[300]{Social and professional topics~History of programming languages}
-%
-% \definecolor{cmtclr}{rgb}{0.0,0.6,0.0}
-% \definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
-% \definecolor{idclr}{rgb}{0.0,0.0,0.0}
-% \definecolor{numclr}{rgb}{0.0,0.4,0.0}
-% \definecolor{strclr}{rgb}{0.4,0.4,0.0}
-% \definecolor{rstrclr}{rgb}{0.5,0.1,0.0}
-% \definecolor{prepclr}{rgb}{0.6,0.0,0.2}
-% \newcommand{\vect}[1]{\langl #1 \rangl}
-% \newcommand{\langl}{\begin{picture}(4.5,7)
-% \put(1.1,2.5){\rotatebox{60}{\line(1,0){5.5}}}
-% \put(1.1,2.5){\rotatebox{300}{\line(1,0){5.5}}}
-% \end{picture}}
-% \newcommand{\rangl}{\begin{picture}(4.5,7)
-% \put(.9,2.5){\rotatebox{120}{\line(1,0){5.5}}}
-% \put(.9,2.5){\rotatebox{240}{\line(1,0){5.5}}}
-% \end{picture}}
-% \newcommand{\ball}[1]{\FPeval{\result}{clip(201+#1)}\textnormal{\ding{\result}}}
-% \newcommand{\lsep}{\;\;|\;\;}
-% \newcommand{\num}[1]{\textcolor{numclr}{#1}}
-% \newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
-% \newcommand{\strf}[1]{\textnormal{\textcolor{strclr}{\sffamily #1}}}
-% \newcommand{\rstr}[1]{\textnormal{\textcolor{rstrclr}{\sffamily "#1"}}}
-% \newcommand{\ident}[1]{\textnormal{\textcolor{idclr}{\sffamily #1}}}
-% \newcommand{\qident}[1]{\textnormal{\sffamily \textquotesingle #1\textquotesingle}}
-% \newcommand{\dom}{\ident{dom}}
-% \newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
-%
-% \newcommand{\bndclr}[1]{\textcolor{kvdclr}{#1}}
-% \newcommand{\bkndclr}[1]{\textcolor{prepclr}{#1}}
-% \newcommand{\blblclr}[1]{\textcolor{numclr}{#1}}
-% \newcommand{\bnd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
-% \newcommand{\bknd}[1]{\textnormal{\textcolor{prepclr}{\sffamily #1}}}
-% \newcommand{\blbl}[1]{\textnormal{\textcolor{numclr}{\sffamily #1}}}
-% \newcommand{\narrow}[1]{\hspace{-0.6em}#1\hspace{-0.6em}}
-% \newcommand{\rname}[1]{{\sffamily\small(#1)}}
-% \newcommand{\ername}[1]{\vspace{0.75em}\rname{#1}\hspace{0.5em}}
-% \newcommand{\preview}[1]{\,\textnormal{\guillemotleft} #1\textnormal{\guillemotright}\,}
-%
-% \begin{document}
-% %\begin{abstract}
-% %\end{abstract}
-% \maketitle
-%
-% % ==================================================================================================
-%

### 2020-05-27 wip-compost/paper.tex
Conclusions, references

@@
 a bar chart that shows two different values for each bar. The chart on the right consists of two
-line charts that share the X axis and highlight two parts of the timeline with two different colors.
+line charts that share the X axis with parts of the timeline highlighted using two different colors.
 
 There is a plenty of libraries that can draw bar charts and line charts, but adding those extra
-features will only be possible if the author already thought about your exact scenario. For
-example, Google Charts supports the left chart (it is called Dual-X Bar Chart) but there is no
+features will only be possible if the author already thought about your exact scenario.
+Google Charts \cite{gcharts} supports the left chart (it is called Dual-X Bar Chart) but there is no
 way for adding a background, or sharing an axis between charts. The alternative is to use a more
-low-level library such as D3. In D3 you construct the chart piece by piece, but then you have to
+low-level library. In D3 \cite{d3} you construct the chart piece by piece, but you have to
 tediously transform your values to coordinates in pixels yourself. For scientific plots,
-you could use an implementation of Grammar of Graphics such as ggplot2, where a chart is
-a mapping from data to geometric objects (such as points, bars, and lines) and their visual
-properties (X and Y coordinate, shape and color). However, the range of charts that can be
+you could use ggplot2 \cite{ggplot2}, based on the Grammar of Graphics \cite{grammar}.
+A chart is a mapping from data to geometric objects (points, bars, lines) and their visual
+properties (X and Y coordinate, shape, color). However, the range of charts that can be
 created using this systematic approach is still somewhat limited.
@@
 As is often the case with domain-specific languages, finding the right primitives is more of an art
-than science. For this reason, I present my answer -- a library named Compost -- as a functional pearl.
-I hope to convince the reader that Compost is elegant and I illustrate this with a wide range
+than science. For this reason, we present an answer, a library named Compost, as a functional pearl.
+We hope to convince the reader that Compost is elegant and we illustrate this with a wide range
 of examples. Compost has a number of specific desirable properties:
@@
   combinators. In particular, concepts such as bar charts, line charts or charts with aligned
-  axes are expressed in terms of more basic concepts.
+  axes are all expressed in terms of more basic concepts.
 \item The primitives are specified in domain terms. When drawing a line, the value of an Y coordinate
-  is an exchange rate of 1.36 USD/GBP, not 137 pixels from the bottom.
+  is an exchange rate of 1.36 USD/GBP, not 67 pixels from the bottom.
 \item Most common chart types can be easily captured as high level abstractions, but there is an
   elengant way of creating a majority of more interesting custom charts.
-\item The approach can easily be extended to creating web-based charts that involve animations
-  or interaction with the user.
+\item The approach can easily be integrated with the Elm architecture \cite{elm}
+	to creating web-based charts that involve animations or interaction with the user.
 \end{itemize}
@@
 The presentation in this paper focuses on explaining the primitives and combinators of the
-domain-specific language. I outline the structure of an implementation, but omit the details. Filling
+domain-specific language. We outline the structure of an implementation, but omit the details. Filling
 those in requires careful thinking about geometry and projections, but there are no unexpected
@@
 \section{Basic charts: Overlaying chart primitives}
-I will introduce individual features of the Compost library gradually. The first important aspect of
-Compost is that properties of shapes are defined in terms of domain-specific values. I first
-explain what this means and then use domain-specific values to specify the core part of the
+We introduce individual features of the Compost library gradually. The first important aspect of
+Compost is that properties of shapes are defined in terms of domain-specific values. In this
+section, we explain what this means and then use domain-specific values to specify the core part of the
 UK election results bar chart.
@@
 
-In the election results chart in Figure~\ref{fig:charts} (left), the X values are categorical
+In the election results chart in Figure~\ref{fig:charts} (left), the X axis shows categorical
 values representing the political parties such as \strf{Conservative} or \strf{Labour}. The
-Y values are numerical values representing the number of seats won such as $\num{365}$ MPs.
+Y axis shows numerical values representing the number of seats won such as $\num{365}$ MPs.
 When creating data visualizations, those are the values that the user needs to specify. This is
@@
 that follows a similar style, but does not have synthetic outer scale would be pairplot from
-the seaborn Python library. % \cite
+the seaborn Python library \cite{seaborn}.
 
@@
 A more complex type of chart that can be easily composed using the Compost primitives is the
-pairplot chart from the seaborn Python library. % \cite
+pairplot chart from the seaborn library \cite{seaborn}.
 Pairplot visualizes pairwise relationships between features of a dataset.
 An example using three features (sepal width, petal width, petal length) from the Iris dataset is shown
-in Figure~\ref{fig:standard} (right). % \cite
+in Figure~\ref{fig:standard} (right).
 A pairplot draws a grid of charts, each visualizing the relationship between two numerical features. For distinct
@@
 interactivity include animations, hover labels or zooming. More interesting custom forms
-include the ``You draw It'' visualization introduced by the New York Times. % cite
+include the ``You draw It'' visualization introduced by the New York Times \cite{youdraw}.
 The chart shows only the first half of the data, such as a timeline, and the reader has to
@@
 implemented using low-level libraries such as D3, but doing so requires directly handling
-JavaScript events.
+JavaScript events and modifying the browser DOM.
 
@@
 Compost can be used in conjunction with a virtual-dom library to create interactive data
-visualizations based on the Elm architecture. An interactive visualization is described using
+visualizations based on the Elm architecture \cite{mvu}. An interactive visualization is described using
 a pair of functions -- the \emph{view} function creates a chart based on the current state and
@@
 
-\newpage
+\section{Conclusions}
+This paper presents a functional perspective on the problem of finding easy to use, but flexible
+abstractions for composing data visualizations. We hope to find a sweet between high-level, but
+inflexible approaches and low-level, but hard to use approaches.
 
-\section{Conclusions}
-\begin{itemize}
-\item We could define a type system to catch categorical/continuous value mismatch in a single %scale.
-\item It would be worth thinking about ordinal values, which are categorical but can be sorted.
-\item There might be other kinds of scales - for example, color scale (can have meaning in scatter plot)
- or a scale for secondary markers (like sizes of bubbles in a bubble chart). We could really say that
- every value (including bar chart colors) should be coming from some scale...
-\item Implementing something like \kvd{axis} or \kvd{roundScale} as an actual derived primitive is
- a bit tricky, because it needs to invoke a part of the normal rendering workflow (to run the
- automatic inference of scales on the nested shape) -- this might be just implementation issue,
- but it could be some more basic problem.
-\end{itemize}
+Most work in this space is based on Grammar of Graphics \cite{grammar}, designing more or less complex and
+powerful variants \cite{layered,vega-lite,vega,polaris}. In Grammar of Graphics, a chart is a mapping from
+data to a chart elements and their visual attributes. In contrast, in Compost, the mapping is
+specified in the host functional programming language and a chart is merely a resulting data type
+describing the visual elements using domain-specific primitives.
 
-\cite{gcharts}
-\cite{seaborn}
-\cite{youdraw}
-\cite{grammar}
-\cite{layered}
-\cite{vega-lite}
-\cite{vega}
-\cite{d3}
-\cite{polaris}
+Our approach is very flexible as it lets the user compose primitive visual elements in any
+way they want; it lets them define their own high-level abstractions and it also integrates well
+with reactive programming architectures to support interactive data visualizations.
+
+In this paper, we focus on presenting the core ideas behind Compost. However, much
+remains to be explored, both in terms of finding the best set of primitives and in terms of
+their language integration. First, we only support categorical and continuous values, but
+there are also ordinal values (which cannot be compared, but can be sorted). Second, some of our
+primitives, namely \kvd{axis} and \kvd{roundScale} could be implemented as derived operations, but
+we treat those as built-in for simplicity. Third, we only treat X and Y as scales, but we could
+similarly treat other visual features (colors of bars, size of bubbles) as scales, which would
+allow a more high-level specification of certain charts. Finally, there are also some obvious
+ways in which a type system could prevent errors, for example by checking that categorical and
+continuous values are not mixed on a single scale.
+
+Even in the basic form, the Compost library makes it possible to
+create a wide range of standard, atypical and even interactive charts. Moreover, we hope that
+other functional programmers agree that it does so in a simple, logical and easy-to-understand way.
+In other words, we hope the presented functional library is, indeed, a functional pearl.
 

### 2020-06-08 wip-compost/paper.tex
Add types and units section

@@
 \usepackage{xcolor}
+\usepackage{semantic}
 
@@
 \newcommand{\urrl}[1]{\textnormal{\textcolor{urlclr}{\url{#1}}}}
+
+\newcommand{\langl}{\begin{picture}(4.5,7)
+\put(1.1,2.5){\rotatebox{60}{\line(1,0){5.5}}}
+\put(1.1,2.5){\rotatebox{300}{\line(1,0){5.5}}}
+\end{picture}}
+\newcommand{\rangl}{\begin{picture}(4.5,7)
+\put(.9,2.5){\rotatebox{120}{\line(1,0){5.5}}}
+\put(.9,2.5){\rotatebox{240}{\line(1,0){5.5}}}
+\end{picture}}
+
 \begin{document}
@@
   \caption{Two charts about the UK politics: Comparison of election results from 2017 and 2019 (left)
-    and GBP-USD exchange rate with highlighted areas before and after the 23 June 2016 Brexit vote.}
+    and GBP/USD exchange rate with highlighted areas before and after the 23 June 2016 Brexit vote.}
   \label{fig:charts}
@@
 
+\subsection{Types and units of measure}
+We introduce the Compost domain-specific language as untyped, but there are some obvious ways in
+which types could make composing charts in Compost safer. First, a type representing a shape could
+specify whether the X and Y axes represent categorical or continous values. This would rule out
+mixing of different values on a single scale and guarantee that the \ident{union} operation,
+sketched in the previous section, is never called in a way leading to an undefined result.
+Second, the type of values mapped to an axis could be further annotated with units of measure
+\cite{units}. Using the F\# notation where $n\langl u\rangl$ is a number $n$ with unit $u$,
+an axis containing a value $\kvd{cont}~\num{317}\langl \ident{mp}\rangl$ would then be incompatible
+with an axis containing a value $\kvd{cont}~\num{1.32}\langl \ident{gbp}/\ident{usd}\rangl$.
+
+The design of the type system for Compost is straightforward, so we only present a brief sketch.
+There are two kinds of types; $\sigma$ is a type of values and $\tau$ is a type of shapes.
+Assuming $u$ denotes a unit of measure, the types are defined as:
+%
+\begin{equation*}
+\begin{array}{rcl}
+\sigma & = & \kvd{Cat}~u \lsep \kvd{Cont}~u
+\end{array}
+\qquad\qquad
+\begin{array}{rcl}
+\tau & = & \kvd{Shape}~\sigma_x, \sigma_y \\
+\end{array}
+\end{equation*}
+
+\vspace{-1em}
+\noindent
+Correspondingly, there are two kinds of judgements; $v\vdash \sigma$ indicates the type of a
+value, while $s\vdash \tau$ indicates the type of a shape. The typing rules for two of the
+basic chart primitives, \kvd{line} and \kvd{overlay} look as follows:
+%
+\begin{equation*}
+\inference
+  {v_{xi} \vdash \sigma_x & v_{yi} \vdash \sigma_y}
+  {\kvd{line}~\gamma, [\,v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\,] \vdash \kvd{Shape}~\sigma_x, \sigma_y}
+\qquad
+\inference
+  {s_i \vdash \kvd{Shape}~\sigma_x, \sigma_y}
+  {\kvd{overlay}~[\,s_1, \ldots, s_n\,] \vdash \kvd{Shape}~\sigma_x, \sigma_y}
+\end{equation*}
+
+\vspace{-0.5em}
+\noindent
+The rule for \kvd{line} ensures that all X and Y values have the same types, $\sigma_x$ and $\sigma_y$,
+respectively and infers $\kvd{Shape}~\sigma_x, \sigma_y$ as the type of the shape. The rule for
+\kvd{overlay} ensures that all composed shapes have the same type, including the type of X and Y scales.
+
 \section{Advanced charts: Controlling scale composition}
@@
 but there are interesting exceptions. The chart in Figure~\ref{fig:charts} (right)
-has two different Y axes, one for GBP-USD and one for GBP-EUR. In the next two sections, we look at
+has two different Y axes, one for GBP/USD and one for GBP/EUR. In the next two sections, we look at
 three combinators that control the scale inferrence process and what flexibility this enables.
@@
 nested shape $s$ using its inferred scale. An example of nesting is shown in Figure~\ref{fig:nesting}.
-Here, a chart with a continous scale from $\num{1.1}$ to $\num{1.4}$ (e.g.~GBP-EUR exchange rates)
+Here, a chart with a continous scale from $\num{1.1}$ to $\num{1.4}$ (e.g.~GBP/EUR exchange rates)
 is nested in the left half of another chart, which has a continuous scale from $\num{0}$ to $\num{100}$.
@@
 similarly treat other visual features (colors of bars, size of bubbles) as scales, which would
-allow a more high-level specification of certain charts. Finally, there are also some obvious
-ways in which a type system could prevent errors, for example by checking that categorical and
-continuous values are not mixed on a single scale.
+allow a more high-level specification of certain charts.
 

### 2020-06-12 wip-compost/paper.tex
Final edits

@@
 
+\newcommand{\X}{\emph{x}\ }
+\newcommand{\Y}{\emph{y}\ }
 \newcommand{\lsep}{\;\;|\;\;}
@@
 a bar chart that shows two different values for each bar. The chart on the right consists of two
-line charts that share the X axis with parts of the timeline highlighted using two different colors.
+line charts that share the \X axis with parts of the timeline highlighted using two different colors.
 
-There is a plenty of libraries that can draw bar charts and line charts, but adding those extra
+Many libraries that can draw bar charts and line charts, but adding those extra
 features will only be possible if the author already thought about your exact scenario.
 Google Charts \cite{gcharts} supports the left chart (it is called Dual-X Bar Chart) but there is no
-way for adding a background, or sharing an axis between charts. The alternative is to use a more
+way to add a background, or sharing an axis between charts. The alternative is to use a more
 low-level library. In D3 \cite{d3} you construct the chart piece by piece, but you have to
@@
 A chart is a mapping from data to geometric objects (points, bars, lines) and their visual
-properties (X and Y coordinate, shape, color). However, the range of charts that can be
+properties (\X and \Y coordinate, shape, color). However, the range of charts that can be
 created using this systematic approach is still somewhat limited.
@@
 What would an elegant functional approach to data visualization look like? A functional programmer
-would want a domain-specific language that has a small number of primitives; allow us to define
-high level abstractions such as a bar chart and its basic building blocks are expressed in terms
-of domain values such as the exchange rate, rather than pixels.
+would want a domain-specific language that has a small number of primitives that allow us to define
+high level abstractions such as a bar chart and that uses domain values such as the exchange
+rate, rather than pixels, in its basic building blocks.
 
@@
   \vspace{0.25em}
-  \caption{Two charts about the UK politics: Comparison of election results from 2017 and 2019 (left)
+  \caption{Two charts about UK politics: Comparison of election results from 2017 and 2019 (left)
     and GBP/USD exchange rate with highlighted areas before and after the 23 June 2016 Brexit vote.}
@@
 As is often the case with domain-specific languages, finding the right primitives is more of an art
-than science. For this reason, we present an answer, a library named Compost, as a functional pearl.
+than science. For this reason, we present our solution, a library named Compost, as a functional pearl.
 We hope to convince the reader that Compost is elegant and we illustrate this with a wide range
@@
 \begin{itemize}
-\item Charts are composed from a small number of primitive building blocks using a small number of
-  combinators. In particular, concepts such as bar charts, line charts or charts with aligned
-  axes are all expressed in terms of more basic concepts.
-\item The primitives are specified in domain terms. When drawing a line, the value of an Y coordinate
+\item Concepts such as bar charts, line charts or charts with aligned
+  axes are all expressed in terms of more primitive building blocks using a small number of combinators.
+\item The primitives are specified in domain terms. When drawing a line, the value of a \\ \Y coordinate
   is an exchange rate of 1.36 USD/GBP, not 67 pixels from the bottom.
-\item Most common chart types can be easily captured as high level abstractions, but there is an
-  elengant way of creating a majority of more interesting custom charts.
+\item Most common chart types can be easily captured as high level abstractions, but there is a
+  way of creating a majority of more interesting custom charts.
 \item The approach can easily be integrated with the Elm architecture \cite{elm}
@@
 The presentation in this paper focuses on explaining the primitives and combinators of the
-domain-specific language. We outline the structure of an implementation, but omit the details. Filling
-those in requires careful thinking about geometry and projections, but there are no unexpected
-surprises. A complete F\# implementation, including the examples used in this paper, is available
+domain-specific language. We outline the structure of an implementation, but omit the details; filling
+those in merely requires careful thinking about geometry and projections. An F\#
+implementation, including all our examples, is available
 at: \urrl{http://github.com/compostjs}.
@@
 \begin{figure}[t]
-\includegraphics[scale=1,trim={0cm 6.5cm 6cm 0cm},clip]{figures/values.pdf}
+\includegraphics[scale=1,trim={0cm 6.9cm 6cm 0cm},clip]{figures/values.pdf}
 \caption{On a continuous scale (above), an exact position is determined by a number.
   On a categorical scale (below), an exact position is determined by the category and a
-  numerical ratio from 0 to 1.}
+  number between 0 and 1.}
 \label{fig:scales}
@@
 
-In the election results chart in Figure~\ref{fig:charts} (left), the X axis shows categorical
+In the election results chart in Figure~\ref{fig:charts} (left), the \X axis shows categorical
 values representing the political parties such as \strf{Conservative} or \strf{Labour}. The
-Y axis shows numerical values representing the number of seats won such as $\num{365}$ MPs.
+\Y axis shows numerical values representing the number of seats won such as $\num{365}$ MPs.
 When creating data visualizations, those are the values that the user needs to specify. This is
@@
 
-Our design focuses on two-dimensional charts with X and Y axes. Values mapped to those axes
+Our design focuses on two-dimensional charts with \X and \Y axes. Values mapped to those axes
 can be either categorical (e.g.~political parties, countries) or continuous
 (e.g.~number of votes, exchange rates). The mapping from categorical and continuous values
-to positions on the chart is done automatically. For continuous values, this involves
-applying a linear transformation. For categorical values, the mapping is more difficult.
+to positions on the chart is done automatically. We discuss this in Section~\ref{sec:basic-scales}.
 
-For example, in the UK election results chart, the X axis is categorical. The library automatically
+For example, in the UK election results chart, the \X axis is categorical. The library automatically
 divides the available space between the six categorical values (political parties). The value
@@
 election results chart. In Figure~\ref{fig:shape}, we define a value $v$ as either a continuous value
-$\kvd{cont}~n$ containing any number $n$ or a categorical value $\kvd{cat}~c, r$, consisting
-of a categorical value $c$ (implemented as a string) and a ratio $r$ between $0$ and $1$.
+$\kvd{cont}~n$ containing any number $n$ or a categorical value $\kvd{cat}~c, r$ consisting
+of a categorical value $c$ and a number $r$ between $0$ and $1$. As discussed in
+Section~\ref{sec:basic-types}, continuous and categorical values can also be annotated with
+units of measure to make the values more descriptive.
 %
@@
 v & = & \kvd{cat}~c, r \\
-  & | & \kvd{cont}~n\\
+  & | & \kvd{cont}~n\\[1.25em]
   ~
@@
 
-Now that we know how Compost represents values, we can define the basic elements of its
-domain-specific langauge. A chart is represeted by the shape $s$ defined in Figure~\ref{fig:shape}.
+Now we can define the besic elements of the Compost domain-specific langauge.
+A chart element is represeted by the shape $s$ defined in Figure~\ref{fig:shape}.
 A primitive shape can be a text label, a line connecting a list of points, a filled polygon
 defined by a list of points or a bubble at a given point with a given width and height.
-The position of points is specified by X and Y coordinates, which can be either categorical or
+The position of points is specified by \X and \Y coordinates, which can be either categorical or
 continuous values. For text, line, polygon and bubble, we also include a parameter $\gamma$
-that specifies the element color. The width and height of a bubble is given in pixels.
+that specifies the element color. The width and height of a bubble is given in pixels rather than
+in domain units.
 
 Figure~\ref{fig:shape} also defines three combinators. The most important is $\kvd{overlay}$,
-which overlays all shapes from a given list. When doing this, Compost automatically infers the
-scales of X and Y axes and calculates suitable projections using a method discussed in the next
-section. Finally, $\kvd{padding}$ adds padding around a specified shape and $\kvd{axis}$ adds
+which overlays given shapes. When doing this, Compost infers the range of values on the \X and \Y
+axes and calculates suitable projections using a method discussed in the next
+section. The $\kvd{padding}$ combinator adds padding around a specified shape and $\kvd{axis}$ adds
 an axis showing the inferred scale on the left, right, top or bottom of a given shape.
-Using those primitives, we can construct the simple UK election results bar chart in Figure~\ref{fig:simple} (left).
-We use the \fkvd{let} construct of the host functional language to structure the code:
+Using those primitives, we can construct the simple UK election results bar chart in Figure~\ref{fig:simple} (left):
 %
@@
 
+\vspace{-0.25em}
 \noindent
-The chart specification overlays two bars of different colors and then adds axes to the bottom and
-left of the chart. The two bars are filled rectangles defined using four corner points. The Y
-coordinates are specified as continuous values, while the X coordinates are categorical. For
-the Conservative party, two of the points have the Y coordinate set to $\kvd{cont}~\num{0}$ (bottom of the bar)
-and two have the Y coordinate set to $\kvd{cont}~\num{365}$ (top of the bar). The two X coordinates
+We use the \fkvd{let} construct of the host functional language to structure the code. The chart
+specification overlays two bars of different colors and then adds axes to the bottom and left of the chart.
+The two bars are filled rectangles defined using four corner points. The \Y
+coordinates are specified as continuous values, while the \X coordinates are categorical. For
+the Conservative party, two of the points have the \Y coordinate set to $\kvd{cont}~\num{0}$ (bottom of the bar)
+and two have the \Y coordinate set to $\kvd{cont}~\num{365}$ (top of the bar). The two \X coordinates
 are the start and the end of the range allocated for the \strf{Conservative} category,
@@
 Extending the snippet to generate a grouped bar chart that shows two results
-for each party as in Figure~\ref{fig:charts} is easy. Given a party $p$, we need to
-generate two rectangles, one with X coordinates $\kvd{cat}~p, 0$ and $\kvd{cat}~p, 0.5$
-and the other with X coordinates $\kvd{cat}~p, 0.5$ and $\kvd{cat}~p, 1$.
+for each party as in Figure~\ref{fig:charts} is not much harder. Given a party $p$, we need to
+generate two rectangles, one with \X coordinates $\kvd{cat}~p, 0$ and $\kvd{cat}~p, 0.5$
+and the other with \X coordinates $\kvd{cat}~p, 0.5$ and $\kvd{cat}~p, 1$.
 In the following snippet, we use a \fkvd{for} comprehension to generate the list. All remaining
@@
 
+\vspace{-0.5em}
 \noindent
 Aside from iterating over all available parties and splitting the bar, the example also adds padding
-around the bars. A padding is specified in pixels rather than in terms of domain values. This is
-sometimes preferrable over, for example, drawing a bar using a range from $\num{0.05}$ to $\num{0.5}$.
+around the bars. This is specified in pixels rather than in terms of domain values. A similar
+result could be be achieved by drawing a bar using a range from $\num{0.05}$ to $\num{0.5}$, but
+specifying padding precisely in pixels is sometimes preferable.
 The chart is still missing a title, which we add in Section~\ref{sec:abstractions}.
@@
 
+We follow the terminology of Vega \cite{vega} and use the term \emph{scale} to refer to the mapping of
+values to positions on a screen; a \emph{coordinate} is a value representing a position on a
+scale and the term \emph{axis} is used to refer to the visual representation of a scale.
+
+Scales are an important concept in Compost.
 When composing shapes using the \kvd{overlay} primitive, the user does not need to specify how
 to position the child elements relatively to each other. The Compost library positions the elements
-automatically. This is done in two steps. First, Compost infers the \emph{scales} for X and Y
-axes. A scale represents the range of values that needs to fit in the space available for the chart.
-Second, Compost calculates a \emph{projection}, a mapping from domain-specific values of the scale
-to the available screen space. A scale $l$ is defined in Figure~\ref{fig:scale}.
+automatically. This is done in two steps. During pre-processing, Compost infers the scales for
+\X and \Y axes. A scale represents the range of values that needs to fit in the space available for
+the chart. When rendering a shape, Compost projects domain-specific values to the available screen
+space based on the inferred scale. A scale $l$ is defined in Figure~\ref{fig:scale}.
+%
+A continuous scale is defined by a minimal and maximal value that need to be mapped to the
+available chart space. A categorical scale is defined by a list of individual categorical values.
+Note that we do not need a minimal and maximal ratios of the used categorical values as Compost
+will use an equal space for each category, regardless of where in this space a shape needs to
+appear.
 
@@
 
-A continuous scale is defined by a minimal and maximal value that need to be mapped to the
-available chart space. A categorical scale is defined by a list of individual categorical values.
-Note that we do not need a minimal and maximal ratios of the used categorical values as Compost
-will use an equal space for each category, regardless of where in this space a shape needs to
-appear.
-
-Inferring scales is done by a simple recursive function that walks over the given shape and
-constructs two scales for the X and Y axis, using the X and Y coordinates that appear in the shape.
-Most of the work is done by a simple helper function that takes two scales, $l_1$ and $l_2$,
+Scale inference is done by a simple recursive function that walks over the given shape and
+constructs two scales for the \X and \Y axis, using the \X and \Y coordinates that appear in the shape.
+Most of the work is done by a simple helper function that takes two scales, $l_1$~and $l_2$,
 and produces a new scale that represents the union of the two:
@@
 
-Once Compost computes scales for a given shape and its sub-shapes, it constructs a projection
-that maps domain-specific values to the available chart space. We discuss this in
-Section~\ref{sec:impl}. For a continuous scale, the projection is a linear transformation. For
+The scales inferred during pre-processing are later used when rendering a shape.
+We discuss the implementation in Section~\ref{sec:impl}. The key operation is projection
+which, given a coordinate, a scale and an area on the screen, produces a position on
+the screen. For a continuous scale, the projection is a linear transformation. For
 categorical scale with $k$ values, we split the available chart space into $k$ equally sized
@@
 \subsection{Types and units of measure}
+\label{sec:basic-types}
+
 We introduce the Compost domain-specific language as untyped, but there are some obvious ways in
-which types could make composing charts in Compost safer. First, a type representing a shape could
-specify whether the X and Y axes represent categorical or continous values. This would rule out
+which types can make composing charts in Compost safer. First, a type representing a shape could
+specify whether the \X and \Y axes represent categorical or continous values. This would rule out
 mixing of different values on a single scale and guarantee that the \ident{union} operation,
@@
 
-The design of the type system for Compost is straightforward, so we only present a brief sketch.
-There are two kinds of types; $\sigma$ is a type of values and $\tau$ is a type of shapes.
-Assuming $u$ denotes a unit of measure, the types are defined as:
+We only outline the type system here. There are two kinds of types; $\sigma$ is a type of values
+and $\tau$ is a type of shapes. Assuming $u$ denotes a unit of measure, the types are defined as:
 %
@@
 
-\vspace{-1em}
+\vspace{-1.25em}
 \noindent
-Correspondingly, there are two kinds of judgements; $v\vdash \sigma$ indicates the type of a
+Correspondingly, there are two kinds of typing judgements; $v\vdash \sigma$ indicates the type of a
 value, while $s\vdash \tau$ indicates the type of a shape. The typing rules for two of the
@@
 respectively and infers $\kvd{Shape}~\sigma_x, \sigma_y$ as the type of the shape. The rule for
-\kvd{overlay} ensures that all composed shapes have the same type, including the type of X and Y scales.
+\kvd{overlay} ensures that all composed shapes have the same type, including the type of \X and \Y scales.
 
@@
 
-Most charts have one X and one Y axis that are determined by the values the chart shows,
+Most charts have one \X and one \Y scale that are determined by the values the chart shows,
 but there are interesting exceptions. The chart in Figure~\ref{fig:charts} (right)
-has two different Y axes, one for GBP/USD and one for GBP/EUR. In the next two sections, we look at
-three combinators that control the scale inferrence process and what flexibility this enables.
+has two different \Y axes, one for GBP/USD and one for GBP/EUR. In the next two sections, we look at
+three combinators that control the scale inference process and what flexibility this enables.
 
@@
 to adjust the automatically inferred range of scales.
-
-The two combinators for controlling the range of scales are \kvd{roundScale} and \kvd{explicitScale}.
-The oprations can be applied to either the X scale or the Y scale, which is indicated by the
-$x/y$ sub-script. The \kvd{roundScale} primitve takes the inferred X or Y scale of the shape $s$
+The oprations can be applied to either the \X scale or the \Y scale, which is indicated by the
+$x/y$ subscript. The \kvd{roundScale} primitive takes the inferred \X or \Y scale of the shape $s$
 and, if it is a continuous scale, rounds its minimal and maximal values to a ``nice'' number.
@@
 scale would have a maximum $\num{400}$. For categorical scale, the operation does not have any effect.
-The \kvd{explicitScale} operation is similar, but it replaces the inferred scale with an explicitly
+The \kvd{explicitScale} operation replaces the inferred scale with an explicitly
 provided scale (the type of the inferred scale has to match with the type of the explicitly given
@@
 \end{equation*}
-\vspace{-0.5em}
+\vspace{-0.75em}
 
@@
 
-\begin{figure}
-\includegraphics[scale=0.57]{figures/lines}
-\vspace{0.25em}
-\caption{Two charts showing currency exchange rates with a shared X scale and separate Y scales.}
-\label{fig:lines}
-\end{figure}
-
 \subsection{Nested scales}
 
-The most interesting primitive for controlling scale composition defined in Figure~\ref{fig:control}
-is $\kvd{nest}_{x/y}$. The combinator takes two values, $v_{min}, v_{max}$ and a shape $s$ as arguments
-and it nests the scale of the shape $s$ inside the region defined by $v_{min}, v_{max}$. When inferring
-scales of shapes, the scale of $\kvd{nest}_{x/y}~l, s$ will be a categorical or continuous scale
-inferred using the values $v_{min}$ and $v_{max}$, regardless of the values that are used inside
+The most interesting primitive for controlling scale composition
+is $\kvd{nest}_{x/y}$. The behaviour is similar to that of the SVG viewport \cite{svg}.
+It takes two values, $v_{min}, v_{max}$ and a shape $s$ as arguments and
+nests the scale of the shape $s$ inside the region defined by $v_{min}, v_{max}$.
+When inferring scales of shapes, the scale of $\kvd{nest}_{x/y}~l, s$ will be a categorical or continuous scale
+constructed from the values $v_{min}$ and $v_{max}$, regardless of the values that are used inside
 the shape $s$. The chart space between $v_{max}$ and $v_{min}$ will then be used to render the
@@
 
-The nesting of scales can be used in a variety of ways. We can, for example, nest a line chart
-inside a bar of a bar chart. In that case, the values for $v_{min}$ and $v_{max}$ would be
-$\kvd{cat}~\strf{ABC}, 0$ and $\kvd{cat}~\strf{ABC}, 1$, which define the start and the end of
-the region allocated to the $\strf{ABC}$ category on a categorical scale. A simpler use case for
-the combinator is showing multiple charts in a single view. For example, the motivating example
+The nesting of scales can be used in a variety of ways. For example, to nest a scatter plot
+showing individual data points inside a bar of a histogram, we would use $\kvd{cat}~\strf{ABC}, 0$
+and $\kvd{cat}~\strf{ABC}, 1$ as the points that define the start and the end of the region. A simpler
+use case for the combinator is showing multiple charts in a single view. For example, the motivating example
 in Figure~\ref{fig:charts} (right) comapres aligned line charts of exchange rates for two
-different currencies. Assuming \ident{gbpusd} and \ident{gbpeur} are lists containing days as X
-values and exchange rates as Y values, we can construct a simple chart with two line charts,
+different currencies. Assuming \ident{gbpusd} and \ident{gbpeur} are lists containing days as \X
+values and exchange rates as \Y values, we can construct a simple chart with two line charts,
 shown in Figure~\ref{fig:lines} (left), using:
@@
 \noindent
-In this example, the X scale shows the days of the year. This scale is shared by both of the charts.
+In this example, the \X scale shows the days of the year. This scale is shared by both of the charts.
 Indeed, if data was only available for the second half of the month for one of the charts,
-we would want the line to start in the middle of the chart. However, the Y scale needs to be
+we would want the line to start in the middle of the chart. However, the \Y scale needs to be
 separate for each of the charts. To achieve this, we use $\kvd{nest}_y$. The scale of the inner
@@
 have equally been, for example, $\num{0},\num{1}, \num{2}$. The outer scale used here is
-synthetic and it is not aligned with other chart elements. An example of a more complex chart
-that follows a similar style, but does not have synthetic outer scale would be pairplot from
-the seaborn Python library \cite{seaborn}.
+synthetic and it is not aligned with other chart elements. A chart that does not have synthetic
+outer scale is pairplot, discussed in the next section.
+
+\begin{figure}
+\includegraphics[scale=0.57]{figures/lines}
+\vspace{0.25em}
+\caption{Two charts showing currency exchange rates with a shared X scale and separate Y scales.}
+\label{fig:lines}
+\end{figure}
 
@@
 Here, we use the \fkvd{let} binding of the host language to define a function that takes
-a the data \ident{rates} together with the minimum and maximum. This is used for drawing
+the data \ident{rates} together with the minimum and maximum. This is used for drawing
 two filled rectangles, covering the first 16 days of the view in blue and the rest in red.
@@
 
-\begin{figure}
-  \includegraphics[scale=0.57]{figures/standard}
-  \vspace{0.25em}
-  \caption{Sample charts built using derived abstractions; a scatter plot visualizing the Iris\\
-    dataset with a title (left) and a pairplot comparing two Iris features (right).}
-  \label{fig:standard}
-\end{figure}
-
 \section{Standard charts: Definining new abstractions}
@@
 
-The Compost library does not introduce a rich collection of standard charts and chart features
-as most charting libraries. However, the functional domain-specific language design makes
-it easy to define high-level chart features based on the low-level primitives of the core language.
-To illustrate this, we give two examples.
+The functional domain-specific language design makes it easy to define high-level chart features
+and chart types, known from standard charting libraries, using the low-level primitives of the
+core language. To illustrate this, we give two examples.
 
@@
 \noindent
-The \ident{title} combinator is a function defined using \fkvd{let} in the host language. It takes
-a title $t$ and a shape $s$. It overlays two shapes. To position the title above the chart, the first
-shape has an outer Y scale $\kvd{continuous}~\num{0},\num{15}$ while the second has an outer Y scale
-$\kvd{continuous}~\num{15},\num{100}$. These are defined using the $\kvd{nest}_y$ primitive. Similarly,
-the outer X scale of both is $\kvd{continuous}~\num{0},\num{100}$, defined using $\kvd{nest}_x$.
+We use \fkvd{let} in the host language to define \ident{title} as a function taking a title $t$ and
+a shape $s$. It overlays two shapes. To position the title above the chart, the~first
+shape has an outer \Y scale $\kvd{continuous}~\num{0},\num{15}$ while the second has an outer \Y scale
+$\kvd{continuous}~\num{15},\num{100}$. Similarly, the outer \X scale of both is $\kvd{continuous}~\num{0},\num{100}$.
+These are defined using $\kvd{nest}_{x/y}$.
+
+\begin{figure}
+  \includegraphics[scale=0.57]{figures/standard}
+  \vspace{0.25em}
+  \caption{Sample charts built using derived abstractions; a scatter plot visualizing the Iris\\
+    dataset with a title (left) and a pairplot comparing two Iris features (right).}
+  \label{fig:standard}
+\end{figure}
 
 The second shape simply wraps the specified chart $s$ to which we are attaching the title. The first
-positions the text title in the middle of the available space. To do so, we explicitly set the X and
-Y scales inside the upper shape to continuous scales from $\num{0}$ to $\num{100}$ and then position
+positions the text title in the middle of the available space. To do so, we explicitly set the \X and
+\Y scales inside the upper shape to continuous scales from $\num{0}$ to $\num{100}$ and then position
 the text label in the middle, at a point $(\kvd{cont}~\num{50}), (\kvd{cont}~\num{50})$.
@@
 
-A more complex type of chart that can be easily composed using the Compost primitives is the
-pairplot chart from the seaborn library \cite{seaborn}.
-Pairplot visualizes pairwise relationships between features of a dataset.
+A more complex chart that can be composed using the Compost primitives is pairplot from the seaborn
+library \cite{seaborn}. Pairplot visualizes pairwise relationships between features of a dataset.
 An example using three features (sepal width, petal width, petal length) from the Iris dataset is shown
@@
 A pairplot draws a grid of charts, each visualizing the relationship between two numerical features. For distinct
-features, pairplot shows a scatter plot using one fature for X values and the other for Y values.
-At the diagonal (when the fatures are the same), it draws a histogram of the values of the feature.
-Additionally, a categorical feature can be used to determine the color of dots in the scatter plots.
+features, pairplot shows a scatter plot using one fature for \X values and the other for \Y values.
+When the fatures are the same (the diagonal), it draws a histogram of the feature values.
+A categorical feature can be used to determine the color of dots in the scatter plots.
 
@@
 The choice between two possible nested charts is made using the host language \fkvd{if} construct.
-Scatter plots are generated by overlaying bubbles with X and Y coordinates obtained using $\ident{get}~x~r$
+Scatter plots are generated by overlaying bubbles with \X and \Y coordinates obtained using $\ident{get}~x~r$
 and $\ident{get}~y~r$. Histograms are composed from filled shapes. To obtain their locations, we use
@@
 \section{Interactive charts: Domain-specific event handling}
-Data visualizations published on the web often also feature interactivity. Standard forms of
-interactivity include animations, hover labels or zooming. More interesting custom forms
-include the ``You draw It'' visualization introduced by the New York Times \cite{youdraw}.
+Many data visualizations published on the web feature interactivity. Standard forms of
+interactivity include animations, hover labels or zooming. More interesting custom
+visualizations include ``You Draw It'' introduced by the New York Times \cite{youdraw}.
 The chart shows only the first half of the data, such as a timeline, and the reader has to
 guess the second half before clicking a button and seeing the actual data. Standard forms of
-interactivity are supported by high-level libraries. For example, Google Charts supports panning
-using drag \& drop, zooming to a selected chart range and animations. Custom interactivity can be
-implemented using low-level libraries such as D3, but doing so requires directly handling
+interactivity are often supported by high-level libraries; Google Charts supports panning
+using drag \& drop, zooming to a selected chart range and animations. Custom interactivity is
+typically implemented using low-level libraries such as D3, but doing so requires directly handling
 JavaScript events and modifying the browser DOM.
@@
 of shape, shown in Figure~\ref{fig:mouse}. The three new primitives make it possible to handle
-three common mouse events using custom functions $\lambda \;x~y\rightarrow e$, specified in the
-host language. The most interesting aspect is that the functions are given X and Y coordinates
-of the event specified in the domain of the chart. This means that if the user clicks on the
+three common mouse events using custom functions $\lambda x~y\rightarrow e$, specified in the
+host language. The most interesting aspect is that the functions are given \X and \Y coordinates
+of the event specified in the domain units of the chart. This means that if the user clicks on the
 bar representing the Conservative party in a bar chart, the values might be, for example,
-$\kvd{cat}~\strf{Conservative},\num{0.75}$ for X and $\kvd{cont}~\num{120.5}$ for Y.
+$\kvd{cat}~\strf{Conservative},\num{0.75}$ for \X and $\kvd{cont}~\num{120.5}$ for \Y.
 
@@
 
-\subsection{You draw it data visualizations}
+\subsection{You Draw It data visualizations}
 To illustrate building of interactive data visualizations using Compost, we look at one aspect of
-the ``You draw It'' visualizations. We want to create a bar chart where the user can use drag \& drop
-to move individual bars. The resulting chart can be seen in Figure~\ref{fig:youdraw}.
-The first step is to define types representing the state and events that can occur:
-%
+``You Draw It''. We want to create a bar chart where the user can use drag \& drop
+to move individual bars. Figure~\ref{fig:youdraw} shows the interactive chart before and after an
+interaction. The first step is to define types representing the state and events that can occur:
+
+\noindent
 \begin{equation*}
 \begin{array}{lll}
-\fkvd{type}~\ident{State} &\narrow{=}& \ident{bool}~*~(\ident{string}~*~\ident{int})~\ident{list}\\[0.5em]
-\fkvd{type}~\ident{Event} &\narrow{=}& \ident{Update}~\fkvd{of}~(\ident{string}~*~\ident{int})\\
-                          &\;\narrow{|}& \ident{Moving}~\fkvd{of}~\ident{bool}
+\fkvd{type}~\ident{State} &\narrow{=}& \ident{bool}~*~(\ident{string}~*~\ident{int})~\ident{list}\\
+\fkvd{type}~\ident{Event} &\narrow{=}& \ident{Update}~\fkvd{of}~(\ident{string}~*~\ident{int}) ~~|~~ \ident{Moving}~\fkvd{of}~\ident{bool}
 \end{array}
@@
 \noindent
-The state is a pair of boolean, indicating whether the user is currently dragging, together with
+The state is a pair of boolean, indicating whether the user is currently dragging, and
 a list of key value pairs, storing the number of seats for each political party. Two types of events
@@
 $\ident{Moving}(\fkvd{true})$ and $\ident{Moving}(\fkvd{false})$, respectively. Second, the user
-may change a value for a party, which is captured by the $\ident{Update}$ event.
+may change a value for a party, which is represented by the $\ident{Update}$ event.
 
-The next part of the visualization is the \ident{update} function which takes an old state together
+The next part of the implementation is the \ident{update} function which takes an old state together
 with an event and produces a new state:
@@
 \fkvd{let}~\ident{update}~(\_,s)~(\ident{Moving}(m)) = m,s\\
-\hspace{1.05em}~\ident{update}~(\fkvd{true},s)~(\ident{Update}(p, v)) =\\
-\hspace{1.05em}\qquad  \fkvd{true}, \ident{map}~(\lambda(k, o) \rightarrow k, \fkvd{if}~k = p~\fkvd{then}~v~\fkvd{else}~o)~s\\
+\hspace{1.05em}~\ident{update}~(\fkvd{true},s)~(\ident{Update}(p, v)) =
+%\hspace{1.05em}\qquad
+  \fkvd{true}, \ident{map}~(\lambda(k, o) \rightarrow k, \fkvd{if}~k = p~\fkvd{then}~v~\fkvd{else}~o)~s\\
 \hspace{1.05em}~\ident{update}~(m,s)~(\ident{Update}(\_, \_)) = m,s\\
@@
 \noindent
-The first two cases handle the \ident{Update} event. The event carries two values, $p, v$, which
+The first case handles the \ident{Moving} event, which replaces the first component of the
+state tuple, i.e.~a flag indicating whether a mouse button is down.
+The next two cases handle the \ident{Update} event. The event carries two values, $p, v$, which
 represent the party (which bar the user is dragging) and the new value (new number of seats).
-If the user is currently dragging, we replace the value associated with the party $p$ in \newpage \noindent the
+If the user is currently dragging, we replace the value associated with the party $p$ in \noindent the
 list $s$ using the \ident{map} function. If the user is not currently dragging, the event is ingored.
-The last case handles the \ident{Moving} event, which simply replaces the first component of the
-state tuple, i.e.~a flag indicating whether a mouse button is down.
 
 \begin{figure}[t]
-  \vspace{0.5em}
   \includegraphics[scale=0.57]{figures/youdraw}
   \vspace{1em}
-  \caption{Interactive ``You draw it'' data visualization. The user moves cursor to a bar (left),
+  \caption{Interactive ``You Draw it'' data visualization. The user moves cursor to a bar (left),
     pushes a mouse button and drags the bar to the position that they think is the correct one (right).}
@@
 
-Finally, the \ident{render} function takes the current state and builds the data visualization
+Finally, the \ident{view} function takes the current state and builds the data visualization
 using the Compost domain-specific language. In addition, it also takes a parameter \ident{trigger},
@@
 \begin{array}{l}
-\fkvd{let}~\ident{render}~\ident{trigger}~(\_, \ident{state})~=\\
+\fkvd{let}~\ident{view}~\ident{trigger}~(\_, \ident{state})~=\\
 \quad \kvd{axis}_l~(\kvd{axis}_b~(\kvd{explicitScale}_y~(\kvd{continuous}~\num{0},\num{400}),\\
@@
 
-\vspace{-0.5em}
-\noindent
 When the user interacts with the visualization created using Compost, the library
@@
 values. In case of the above bar chart, when the user moves a mouse, the function registed
-using \kvd{mouseMove} is given a categorical value $\kvd{cat}~p, r$ as the X coordinate and
-a continuous value $\kvd{cont}~v$ as the Y coordinate. It then takes $p$, which is the name
+using \kvd{mouseMove} is given a categorical value $\kvd{cat}~p, r$ as the \X coordinate and
+a continuous value $\kvd{cont}~v$ as the \Y coordinate. It then takes $p$, which is the name
 of the party corresponding to the bar and the value $v$ corresponding to the number of seats
@@
 
-The pair of functions, \ident{update} and \ident{render}, together with an initial state is
+The pair of functions, \ident{update} and \ident{view}, together with an initial state is
 all that is needed to create an interactive data visualization. Compost calls
-\ident{render} each time the state changes and uses virtual-dom to update the chart
+\ident{view} each time the state changes and uses virtual-dom to update the chart
 displayed in the browser. Although creating an interactive visualization is more work than
@@
 
-\section{Implementation structure: Scale inference and projection}
+\section{Implementation structure: Scale inference and rendering}
 \label{sec:impl}
@@
 code can be found at \urrl{http://github.com/compostjs}.
-As is often the case with functional domain-specific languages, the implementation is easy once
-we find the right collection of basic primitives and the right structure for the implementation.
+As is often the case with functional domain-specific languages, the implementation is not difficult
+once we find the right collection of basic primitives and the right structure for the implementation.
 This largely applies to the Compost library and so we will not go into the implementation details.
@@
 First, the library infers the scales of a shape. When doing so, it also annotates some shapes
-with additional information that are needed later for rendering. Second, the library projects
+with additional information that is needed later for rendering. Second, the library projects
 the shape onto an available space and produces the chart, represented as an SVG object.
@@
 
-In order to render a shape, we need to know the range of values that should appear on the X and
-Y axes. This is done by inferring a scale for each of the axes from the individual X and Y
+In order to render a shape, we need to know the range of values that should appear on the \X and
+\Y axes. This is done by inferring a scale for each of the axes from the individual \X and \Y
 coordinates that specify shape locations. As discussed earlier, a scale can be either categorical
@@
 The operation that infers the scales of a shape is \ident{calculateScales}. It takes a shape
-and produces a pair of X and Y scales, together with a transformed shape:
+and produces a pair of \X and \Y scales, together with a transformed shape:
 %
@@
 
-\vspace{-1em}
+\vspace{-1.25em}
 \noindent
@@
 $\kvd{scaledNest}_{x/y}$:
-%
+
 \begin{equation*}
@@
 When calculating the scales of the $\kvd{nest}_x$, the function first calculate scales of the
-sub-shape $s$ recursively. The resulting Y scale $s_y$ is returned as the result, while the X
+sub-shape $s$ recursively. The resulting \Y scale $s_y$ is returned as the result, while the \X
 scale is obtained from the two coordinates $v_{min}$ and $v_{max}$. This is also the case where
-the shape is transformed and the returned $\kvd{scaledNest}_x$ shape stores the inferred X scale
+the shape is transformed and the returned $\kvd{scaledNest}_x$ shape stores the inferred \X scale
 $s_x$ of the sub-shape $s$. The second example is the $\kvd{overlay}$ case which recursively
-infers scales of all sub-shapes and combines those using the list folding function
+computes scales of all sub-shapes and combines those using the list folding function
 \ident{reduce} with $\ident{union}$ as an argument.
@@
 The key operation that needs to be performed when drawing a shape is projecting coordinates from
-domain-specific values to the screen coordinates. As we draw a shape, we keep the X and Y scale
-and the space in pixels that it should be drawn on. Initially, the X and Y scales are those
+domain-specific values to the screen coordinates. As we draw a shape, we keep the \X and \Y scale
+and the space in pixels that it should be drawn on. Initially, the \X and \Y scales are those
 inferred for the entire shape and the space in pixels is $\num{0}\ldots\ident{width}$ and
@@
 \noindent
-The function is only defined if the value and scale are compatible. If both are continuous, the
+The function is only defined if the value and scale are compatible. As discussed in Section~\ref{sec:basic-types},
+this could be guaranteed using a simple type system. If both are continuous, the
 function performs a simple linear transformation. If both are categorical, the available pixel
@@
 The drawing of shapes is done by a function that takes the available area as a quadruple $(x_1, y_1), (x_2, y_2)$
-together with the X and Y scale corresponding to the area and a shape to be drawn. The result is
-a data structure representing a SVG document:
+together with the \X and \Y scale mapped onto the area and a shape to be drawn. The result is
+a data structure representing an SVG document:
 %
@@
 
-\vspace{-1.25em}
-\noindent
-For primitive shapes, the operation projects the coordinates using \ident{project} and construct
+For primitive shapes, the operation projects the coordinates using \ident{project} and constructs
 a corresponding SVG document. For shapes with sub-shapes it calls itself recursively, possibly
@@
 When drawing \kvd{overlay}, the function draws all sub-shapes onto the same
-area using the same scales and then concatenates the returned SVG components the \ident{concat} helper.
+area using the same scales and then concatenates the returned SVG components using the \ident{concat} helper.
 The $\kvd{scaledNest}_x$ case is more illuminating. Here, we first
@@
 $v_{min},v_{max}$. This defines the area corresponding to the nested scale ${ns}_x$, onto
-which the X coordinates in the sub-shape $\ident{shape}$ should be projected.
-To do this, we recursively call \ident{drawShape}, but use $x_1'$ and $x_2'$ as the X coordinates
-of the target area and ${ns}_x$ as the X scale. The Y area and scales are propagated unchanged.
+which the \X coordinates in the sub-shape $\ident{shape}$ should be projected.
+To do this, we recursively call \ident{drawShape}, but use $x_1'$ and $x_2'$ as the \X coordinates
+of the target area and ${ns}_x$ as the \X scale. The \Y area and scales are propagated unchanged.
 
 \section{Conclusions}
-This paper presents a functional perspective on the problem of finding easy to use, but flexible
+This paper presents a functional take on the problem of designing easy to use, but flexible
 abstractions for composing data visualizations. We hope to find a sweet between high-level, but
@@
 data to a chart elements and their visual attributes. In contrast, in Compost, the mapping is
-specified in the host functional programming language and a chart is merely a resulting data type
+specified in the host programming language and a chart is merely a resulting data type
 describing the visual elements using domain-specific primitives.
@@
 primitives, namely \kvd{axis} and \kvd{roundScale} could be implemented as derived operations, but
-we treat those as built-in for simplicity. Third, we only treat X and Y as scales, but we could
+we treat those as built-in for simplicity. Third, we only treat \X and \Y as scales, but we could
 similarly treat other visual features (colors of bars, size of bubbles) as scales, which would

### 2020-06-12 wip-compost/paper.tex
Fix typos

@@
 
-Now we can define the besic elements of the Compost domain-specific langauge.
-A chart element is represeted by the shape $s$ defined in Figure~\ref{fig:shape}.
+Now we can define the basic elements of the Compost domain-specific language.
+A chart element is represented by the shape $s$ defined in Figure~\ref{fig:shape}.
 A primitive shape can be a text label, a line connecting a list of points, a filled polygon
@@
 which types can make composing charts in Compost safer. First, a type representing a shape could
-specify whether the \X and \Y axes represent categorical or continous values. This would rule out
+specify whether the \X and \Y axes represent categorical or continuous values. This would rule out
 mixing of different values on a single scale and guarantee that the \ident{union} operation,
@@
 The automatic scale inference often results in scales where the maximum is a non-round number.
-This leads to charts that fully utilize the available space, but may not be easy to read.
+This leads to charts that fully utilise the available space, but may not be easy to read.
 The first two primitives, shown in Figure~\ref{fig:control} (left) allow the chart designer
 to adjust the automatically inferred range of scales.
-The oprations can be applied to either the \X scale or the \Y scale, which is indicated by the
+The operations can be applied to either the \X scale or the \Y scale, which is indicated by the
 $x/y$ subscript. The \kvd{roundScale} primitive takes the inferred \X or \Y scale of the shape $s$
@@
 nested shape $s$ using its inferred scale. An example of nesting is shown in Figure~\ref{fig:nesting}.
-Here, a chart with a continous scale from $\num{1.1}$ to $\num{1.4}$ (e.g.~GBP/EUR exchange rates)
+Here, a chart with a continuous scale from $\num{1.1}$ to $\num{1.4}$ (e.g.~GBP/EUR exchange rates)
 is nested in the left half of another chart, which has a continuous scale from $\num{0}$ to $\num{100}$.
@@
 use case for the combinator is showing multiple charts in a single view. For example, the motivating example
-in Figure~\ref{fig:charts} (right) comapres aligned line charts of exchange rates for two
+in Figure~\ref{fig:charts} (right) compares aligned line charts of exchange rates for two
 different currencies. Assuming \ident{gbpusd} and \ident{gbpeur} are lists containing days as \X
@@
 
-\section{Standard charts: Definining new abstractions}
+\section{Standard charts: Defining new abstractions}
 \label{sec:abstractions}
@@
 A pairplot draws a grid of charts, each visualizing the relationship between two numerical features. For distinct
-features, pairplot shows a scatter plot using one fature for \X values and the other for \Y values.
-When the fatures are the same (the diagonal), it draws a histogram of the feature values.
+features, pairplot shows a scatter plot using one feature for \X values and the other for \Y values.
+When the features are the same (the diagonal), it draws a histogram of the feature values.
 A categorical feature can be used to determine the color of dots in the scatter plots.
@@
 and $\ident{get}~y~r$. Histograms are composed from filled shapes. To obtain their locations, we use
-a helper function $\ident{bins}~x~\ident{data}$, which returns a list of bins specified by a tripple
+a helper function $\ident{bins}~x~\ident{data}$, which returns a list of bins specified by a triple
 consisting of a lower and an upper range $x_1, x_2$ and the count $y$.
@@
 If the user is currently dragging, we replace the value associated with the party $p$ in \noindent the
-list $s$ using the \ident{map} function. If the user is not currently dragging, the event is ingored.
+list $s$ using the \ident{map} function. If the user is not currently dragging, the event is ignored.
 
@@
 translates the coordinates associated with events from pixels to domain-specific
-values. In case of the above bar chart, when the user moves a mouse, the function registed
+values. In case of the above bar chart, when the user moves a mouse, the function registered
 using \kvd{mouseMove} is given a categorical value $\kvd{cat}~p, r$ as the \X coordinate and
@@
 
-Compost is an open-source library, implmented in the functional language F\#. The full source
+Compost is an open-source library, implemented in the functional language F\#. The full source
 code can be found at \urrl{http://github.com/compostjs}.
@@
 This largely applies to the Compost library and so we will not go into the implementation details.
-It is, however, worth giving an outline of the implmentation structure.
+It is, however, worth giving an outline of the implementation structure.
 
@@
 coordinates that specify shape locations. As discussed earlier, a scale can be either categorical
-(displaying only categorical values) or contiuous (displaying only continuous values). When
-inerring scales, we use two helper operations; \ident{union}, discussed earlier, combines two
+(displaying only categorical values) or continuous (displaying only continuous values). When
+inferring scales, we use two helper operations; \ident{union}, discussed earlier, combines two
 scales and \ident{singleton} creates a scale from a single coordinate.

### 2020-09-11 wip-design/paper.tex
Design paper - related work and introduction

@@
-\documentclass{sigchi}
-\usepackage{xspace}
-\usepackage{fancyvrb}
-\usepackage{inconsolata}
-\usepackage{xcolor}
-\usepackage{tikz}
-\newcommand*{\priority}[1]{\begin{tikzpicture}[scale=0.12]%
-    \draw (0,0) circle (1);
-    \fill[fill opacity=1,fill=black] (0,0) -- (90:1) arc (90:90-#1*3.6:1) -- cycle;
-    \end{tikzpicture}}
-\newcommand*{\priorityc}[1]{
-\resizebox{0.8em}{0.8em}{
-  {\protect\tikz{ \protect
-    \draw[line width=1mm, black] (0,0) circle (1); \fill[fill opacity=1,fill=black] (0,0) -- (90:1) arc (90:90-#1*3.6:1) -- cycle;
-    } }}}
-
-\definecolor{kvdclr}{rgb}{0.3,0.0,0.8}
-\DefineVerbatimEnvironment{thegamma}{Verbatim}{fontfamily=zi4,numbers=left,xleftmargin=6mm,fontsize=\small,commandchars=\\\{\}}
-%\newcommand{\kvd}[1]{\textcolor{kvdclr}{#1}}
-\newcommand{\kvd}[1]{\textbf{#1}}
-\newcommand{\ikvd}[1]{{\fontfamily{zi4}\selectfont\small #1}}
-\newcommand{\tg}{The Gamma\xspace}
-
-% Use this section to set the ACM copyright statement (e.g. for
-% preprints).  Consult the conference website for the camera-ready
-% copyright statement.
-
-% Copyright
-\CopyrightYear{2020}
-%\setcopyright{acmcopyright}
-\setcopyright{acmlicensed}
-%\setcopyright{rightsretained}
-%\setcopyright{usgov}
-%\setcopyright{usgovmixed}
-%\setcopyright{cagov}
-%\setcopyright{cagovmixed}
-% DOI
-\doi{https://doi.org/10.1145/3313831.XXXXXXX}
-% ISBN
-\isbn{978-1-4503-6708-0/20/04}
-%Conference
-\conferenceinfo{UIST'20,}{October  20--23, 2020, Minneapolis, MN, USA}
-%Price
-\acmPrice{\$15.00}
-
-% Use this command to override the default ACM copyright statement
-% (e.g. for preprints).  Consult the conference website for the
-% camera-ready copyright statement.
-
-%% HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP --
-%% Please note you need to make sure the copy for your specific
-%% license is used here!
-% \toappear{
-% Permission to make digital or hard copies of all or part of this work
-% for personal or classroom use is granted without fee provided that
-% copies are not made or distributed for profit or commercial advantage
-% and that copies bear this notice and the full citation on the first
-% page. Copyrights for components of this work owned by others than ACM
-% must be honored. Abstracting with credit is permitted. To copy
-% otherwise, or republish, to post on servers or to redistribute to
-% lists, requires prior specific permission and/or a fee. Request
-% permissions from \href{mailto:Permissions@acm.org}{Permissions@acm.org}. \\
-% \emph{UIST '20},  October 20--23, 2020, Minneapolis, MN, USA \\
-% ACM xxx-x-xxxx-xxxx-x/xx/xx\ldots \$15.00 \\
-% DOI: \url{http://dx.doi.org/xx.xxxx/xxxxxxx.xxxxxxx}
-% }
-
-% Arabic page numbers for submission.  Remove this line to eliminate
-% page numbers for the camera ready copy
-% \pagenumbering{arabic}
-
-% Load basic packages
-\usepackage{balance}       % to better equalize the last page
-\usepackage{graphics}      % for EPS, load graphicx instead
-\usepackage[T1]{fontenc}   % for umlauts and other diaeresis
-\usepackage{txfonts}
-\usepackage{mathptmx}
-\usepackage[pdflang={en-US},pdftex]{hyperref}
-\usepackage{color}
-\usepackage{booktabs}
-\usepackage{textcomp}
-
-% Some optional stuff you might like/need.
-\usepackage{microtype}        % Improved Tracking and Kerning
-% \usepackage[all]{hypcap}    % Fixes bug in hyperref caption linking
-\usepackage{ccicons}          % Cite your images correctly!
-% \usepackage[utf8]{inputenc} % for a UTF8 editor only
-
-% If you want to use todo notes, marginpars etc. during creation of
-% your draft document, you have to enable the "chi_draft" option for
-% the document class. To do this, change the very first line to:
-% "\documentclass[chi_draft]{sigchi}". You can then place todo notes
-% by using the "\todo{...}"  command. Make sure to disable the draft
-% option again before submitting your final document.
-\usepackage{todonotes}
-
-% Paper metadata (use plain text, for PDF inclusion and later
-% re-using, if desired).  Use \emtpyauthor when submitting for review
-% so you remain anonymous.
-\def\plaintitle{The Gamma: Data Exploration through Iterative Prompting}
-\def\plainauthor{First Author, Second Author, Third Author,
-  Fourth Author, Fifth Author, Sixth Author}
-\def\emptyauthor{}
-\def\plainkeywords{Data exploration; End-user programming; Data journalism; Programming languages; Type providers}
-\def\plaingeneralterms{Documentation, Standardization}
-
-% llt: Define a global style for URLs, rather that the default one
-\makeatletter
-\def\url@leostyle{%
-  \@ifundefined{selectfont}{
-    \def\UrlFont{\sf}
-  }{
-    \def\UrlFont{\small\bf\ttfamily}
-  }}
-\makeatother
-\urlstyle{leo}
-
-% To make various LaTeX processors do the right thing with page size.
-\def\pprw{8.5in}
-\def\pprh{11in}
-\special{papersize=\pprw,\pprh}
-\setlength{\paperwidth}{\pprw}
-\setlength{\paperheight}{\pprh}
-\setlength{\pdfpagewidth}{\pprw}
-\setlength{\pdfpageheight}{\pprh}
-
-% Make sure hyperref comes last of your loaded packages, to give it a
-% fighting chance of not being over-written, since its job is to
-% redefine many LaTeX commands.
-\definecolor{linkColor}{RGB}{6,125,233}
-\hypersetup{%
-  pdftitle={\plaintitle},
-% Use \plainauthor for final version.
-%  pdfauthor={\plainauthor},
-  pdfauthor={\emptyauthor},
-  pdfkeywords={\plainkeywords},
-  pdfdisplaydoctitle=true, % For Accessibility
-  bookmarksnumbered,
-  pdfstartview={FitH},
-  colorlinks,
-  citecolor=black,
-  filecolor=black,
-  linkcolor=black,
-  urlcolor=linkColor,
-  breaklinks=true,
-  hypertexnames=false
+\documentclass[manuscript,review,anonymous]{acmart}
+\usepackage{hyperref}
+
+% NOTE: Good things
+%  - what is iterative prompting? "Like auto-complete but less limited" - hmm, maybe
+%  - sell more main idea: can do much with very limited set of commands / via auto-complete
+%  - paper is convincing in that this works for SQL-like queries (maybe not general purpose, but this is good)
+%  - journalism is a good focus / but needs to deemphasize this a bit because of the study
+%  - #1 Limited language for data exploration (what is included / excluded / essential / could be added)
+%  - #2 Iterative prompting - what's new? people can get to all programs via '.' and up/down/enter.
+
+% NOTE: Random
+%  - How choices were made?
+%  - Terminology: data analysis / data exploration / data visualization - what is the goal of the system? (fair enough)
+%    (it's not about writing charts - just about getting the right data - we have limited charting functionality,
+%     one could integrate with Vega, or do future work based on iterative prompting)
+%  - Technical details - you can only create valid programs!! (it's contextual)
+%  - Minor: what was "demo" at the start of a session
+%  - Table 1 vs. Table 2 numbers of participants
+%  - What data quality is needed to make this work?
+%  - "Magic Escalator of Knowledge" - remove bs?
+%  - Clarify how we support 3D cubes (no unified API); mention "done by intern in X weeks"
+
+% TODO Study improvements
+% - The questionaire needs a bit more introduction when you present  the study.
+% - When looking at the data for the questionaire, it seems somewhat inconclusive
+%    as all the values lie around 2.5.  I think this warrants a bit more discussion.
+% - Connected to this, I would recommend to highlight the successes of the study a bit more in the
+%     discussion and  presentation of the results. You actually got non-programmers to do
+%     sophisticated queries! That you didn't get support for all RQs is perhaps more a result of being quite  ambitious!
+
+% TODO
+%   - SAY SOMEWHERE: choose function *and* specify arguments via auto-complete
+%   - ADD: table of all available operations? (shorten introduction)
+
+% NOTE: Contribution
+%  - What is the exact goal? (so that it can be compared with R, Python, Vega-Lite, Tableau)
+%  - How the design goals were generated? (not by talking to journalists)
+%  - "notebooks are messy" - not arguing we can fix that
+
+% NOTE: Motivation
+% - what if you could write the whole program using auto-complete?
+
+% extend autocomplete into a full-blown interactive programming mechanism
+% and show that it is viable for small programming tasks like data exploration
+%
+% --- remove some of the motivation (makes ppl ask bad questions)
+
+%
+% Iterative prompting - should we call it expression completion
+% (is it more general result, or is it a new thing?)
+% -> emphasize the why this is called differently paragraph
+%
+% Explain type providers a bit more (we say what it does, but not much about what
+% it is - give dynamic typing in a static language)
+%
+% The 'then' operation was mysterious - can we add one line explanation to the operations?
+% (yes we can and we should totally do that)
+
+% \usepackage{xspace}
+% \usepackage{fancyvrb}
+% \usepackage{inconsolata}
+% \usepackage{xcolor}
+% \usepackage{tikz}
+% \newcommand*{\priority}[1]{\begin{tikzpicture}[scale=0.12]%
+%     \draw (0,0) circle (1);
+%     \fill[fill opacity=1,fill=black] (0,0) -- (90:1) arc (90:90-#1*3.6:1) -- cycle;
+%     \end{tikzpicture}}
+% \newcommand*{\priorityc}[1]{
+% \resizebox{0.8em}{0.8em}{
+%   {\protect\tikz{ \protect
+%     \draw[line width=1mm, black] (0,0) circle (1); \fill[fill opacity=1,fill=black] (0,0) -- (90:1) arc (90:90-#1*3.6:1) -- cycle;
+%     } }}}
+%
+% \definecolor{kvdclr}{rgb}{0.3,0.0,0.8}
+% \DefineVerbatimEnvironment{thegamma}{Verbatim}{fontfamily=zi4,numbers=left,xleftmargin=6mm,fontsize=\small,commandchars=\\\{\}}
+% %\newcommand{\kvd}[1]{\textcolor{kvdclr}{#1}}
+% \newcommand{\kvd}[1]{\textbf{#1}}
+% \newcommand{\ikvd}[1]{{\fontfamily{zi4}\selectfont\small #1}}
+% \newcommand{\tg}{The Gamma\xspace}
+% \usepackage{balance}       % to better equalize the last page
+% \usepackage{graphics}      % for EPS, load graphicx instead
+% \usepackage[T1]{fontenc}   % for umlauts and other diaeresis
+% \usepackage{txfonts}
+% \usepackage{mathptmx}
+% \usepackage[pdflang={en-US},pdftex]{hyperref}
+% \usepackage{color}
+% \usepackage{booktabs}
+% \usepackage{textcomp}
+
+\setcopyright{acmwhatever}
+\copyrightyear{2020}
+\acmYear{2020}
+%\acmDOI{10.1145/1122445.1122456}
+\acmConference[Woodstock '18]{Woodstock '18: ACM Symposium on Neural
+  Gaze Detection}{June 03--05, 2018}{Woodstock, NY}
+\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
+  June 03--05, 2018, Woodstock, NY}
+\acmPrice{15.00}
+%\acmISBN{978-1-4503-XXXX-X/18/06}
+
+\begin{document}
+\title{The Gamma: Data Exploration through Iterative Prompting}
+
+\author{Tomas Petricek}
+\email{t.petricek@kent.ac.uk}
+\affiliation{%
+  \institution{University of Kent}
+  \city{Canterbury, UK}
 }
-
-% create a shortcut to typeset table headings
-% \newcommand\tabhead[1]{\small\textbf{#1}}
-
-% End of preamble. Here it comes the document.
-\begin{document}
-
-\title{\plaintitle}
-
-\numberofauthors{1}
-\author{%
-  \alignauthor{Leave Authors Anonymous\\
-    \affaddr{for Submission}\\
-    \affaddr{City, Country}\\
-    \email{e-mail address}}
-}
-
-\maketitle
+%\renewcommand{\shortauthors}{Trovato and Tobin, et al.}
 
@@
   %
-We argue that the our approach allows journalists and the public to benefit from the rise
+  We argue that our approach allows journalists and the public to benefit from the rise
   of open data, by making data exploration easier, more transparent and more reproducible.
+
 \end{abstract}
 
-% ACM Classfication
-%
-% \begin{CCSXML}
-% <ccs2012>
-% <concept>
-% <concept_id>10003120.10003121</concept_id>
-% <concept_desc>Human-centered computing~Human computer interaction (HCI)</concept_desc>
-% <concept_significance>500</concept_significance>
-% </concept>
-% <concept>
-% <concept_id>10003120.10003121.10003125.10011752</concept_id>
-% <concept_desc>Human-centered computing~Haptic devices</concept_desc>
-% <concept_significance>300</concept_significance>
-% </concept>
-% <concept>
-% <concept_id>10003120.10003121.10003122.10003334</concept_id>
-% <concept_desc>Human-centered computing~User studies</concept_desc>
-% <concept_significance>100</concept_significance>
-% </concept>
-% </ccs2012>
-% \end{CCSXML}
-%
-% \ccsdesc[500]{Human-centered computing~Human computer interaction (HCI)}
-% \ccsdesc[300]{Human-centered computing~Haptic devices}
-% \ccsdesc[100]{Human-centered computing~User studies}
-
-% Author Keywords
-\keywords{\plainkeywords}
-
-% Print the classficiation codes
-% \printccsdesc
-% Please use the 2012 Classifiers and see this link to embed them in the text: \url{https://dl.acm.org/ccs/ccs_flat.cfm}
+\begin{CCSXML}
+<ccs2012>
+   <concept>
+       <concept_id>10003120.10003121.10003124</concept_id>
+       <concept_desc>Human-centered computing~Interaction paradigms</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10011007.10011006.10011050.10011017</concept_id>
+       <concept_desc>Software and its engineering~Domain specific languages</concept_desc>
+       <concept_significance>300</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10011007.10011006.10011066.10011069</concept_id>
+       <concept_desc>Software and its engineering~Integrated and visual development environments</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+ </ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Human-centered computing~Interaction paradigms}
+\ccsdesc[500]{Software and its engineering~Integrated and visual development environments}
+\ccsdesc[300]{Software and its engineering~Domain specific languages}
+\keywords{data exploration; end-user programming; data journalism; programming languages; type providers}
+
+\maketitle
 
 \begin{figure}
-\includegraphics[width=1\columnwidth]{figures/thegamma-annot}
-\caption{Teams with the greatest number of gold medals from Rio 2016
-Olympics with a reproducible The Gamma script (1), an auto-complete prompt offering ways
-of sorting the data (2) and instant preview (3).}
+\includegraphics[width=.54\columnwidth]{figures/thegamma-annot}
+\caption{Obtaining teams with the greatest number of gold medals from Rio 2016
+Olympics with a reproducible The Gamma script (1), contextual iterative prompting mechainsm
+offering ways of sorting the data (2) and an instant preview of results (3).}
 \label{fig:thegamma}
-\vspace{-0.5em}
 \end{figure}
@@
 same time post-truth politics and increasing public distrust in statistics makes data-driven insights
-increasingly less relevant in public discourse~\cite{howstatslost}. This should perhaps not be a surpirse.
-Journalists can access increasing amounts of data, but producing engaging and transparent data-driven
-reports that are easy to interpret is expensive and requires expert programming skills \cite{ddj}.
-
-The design of a data exploration tool for journalists poses a unique mix of challenges. First, the
-tool needs to be easy to learn for end-users working under tight deadlines. Second, it needs to
-support a wide range of data sources in a way where the expertise gained when working with one data
-source is relevant for other data sources. Third, the resulting data-driven insights need to be
-transparent, allowing the readers to verify the claims and learn how to reproduce the work.
-
-We present The Gamma, a text-based data exploration environment for non-experts. The Gamma
-is based on a single interaction principle, which provides uniform
+increasingly less relevant in public discourse~\cite{howstatslost}. To reverse this trend, we
+need tools that let non-experts, including journalists and other information-literate citizens,
+produce transparent, engaging data analyses that are easy to interpret without requiring expert
+programming skill~\cite{ddj}. The design of such data exploration tool poses a unique mix of chellenges.
+First, the tool needs to have a very low barrier to entry. Second, it needs to support a wide
+range of data sources in a uniform way. Third, the resulting data analyses need to assist
+readers in learning how to reproduce the work and verify the claims it makes.
+
+We contribut The Gamma, a text-based data exploration environment for non-experts. The Gamma
+is based on a single, easy to understand interaction principle and provides a uniform
 access to a range of data sources including data tables, graph databases and data cubes.
-An anlysis created in The Gamma is a transparent script that can be followed to reproduce the
+The resulting analysis is a transparent script that can be followed to reproduce the
 result from scratch. This allows learning from existing analyses and encourages readers
-to engage with the results. Our key contributions are:
+to engage with data.
+
+\subsubsection*{Iterative prompting}
+The key idea in The Gamma, which we term the \emph{iterative prompting} principle, is that all
+valid data exploration scripts can be constructed by repeatedly choosing from a list
+of offered options. This way, non-programmers can write entire scripts through auto-complete,
+without learning a programming language, but they still produce transparent and reproducible
+source code. In other words, iterative prompting turns auto-complete from a programmer assistance
+tool into a non-expert programming mechanism. A crucial feature is that iterative prompting
+only offer operations that are valid in a given context and that it offer all such operations;
+it is both correct and complete.
+
+\subsubsection*{Date exploration}
+The Gamma focuses on data exploration of the kind illustrated in Figure~\ref{fig:thegamma}.
+The user accesses data available in a structured format. They make several experiments to find an
+interesting way of looking at the data, e.g.~by applying different aggregations or filters. They
+may choose to view the results as a table or a basic chart before publishing their analysis. The
+Gamma makes such simple programming with data simple enough for non-experts. Scraping and cleaning
+of messy data or building rich data visualizations is outside of the scope of this paper, but
+exposing those using the iterative prompting approach is an interesting future challenge.
+
+\subsubsection*{Contributions}
+The Gamma is available (non-anonymously) at \href{http://thegamma.net}{\textcolor{blue}{thegamma.net}},
+both as a JavaScript library and a hosted data exploration service. In this paper, we describe and
+evaluate the design principles behind the project:
 
 \begin{itemize}
-\item We identify the design requirements for a data exploration tool for journalists
-  (Section~\ref{sec:motivation}) and follow those to build a novel programming environment
-  The Gamma (Section~\ref{sec:overview}).
-
-\item We introduce \emph{iterative prompting} (Section~\ref{sec:design}),
-  an interaction design principle that can be used to complete a variety of programming
-  tasks in a uniform way that allows transfer of knowledge between different tasks.
-
-\item We show how to use the iterative prompting principle for querying of distinct data
-  sources including data tables, graph databases and data cubes (Section~\ref{sec:implementation}).
-
-\item We discuss a number of case studies (Section~\ref{sec:cases}) and
-  conduct a user study to evaluate the usability of The Gamma and the extent to which users can,
-  (i) learn from examples and (ii)~transfer knowledge between tasks (Section~\ref{sec:study}).
+\item We introduce the iterative prompting principle in The Gamma (Section~\ref{sec:overview})
+  and show that it can be used for querying of distinct data sources including data tables, graph
+  databases and data cubes (Section~\ref{sec:implementation}).
+
+\item We show that iterative prompting is a complete and correct program construction method and
+  consider how it lowers the barrier to entry and allows learning through examples, without
+  expert guidance (Section~\ref{sec:design}).
+
+\item We evaluate the system through a number of case studies (Section~\ref{sec:cases})
+  and a user study (Section~\ref{sec:study}). Our study shows that non-programmers can use The Gamma
+  to construct non-trivial data queries and ascertains the extent to which users can,
+  (i) learn from examples and (ii)~transfer knowledge between tasks.
 \end{itemize}
 
-The Gamma is available as open-source at \href{http://thegamma.net}{\small\bf\ttfamily thegamma.net}.
-
 \section{Related work}
 
-The Gamma brings together recent programming language innovations for working with data
-\cite{inforich}, work on making programming accessible to non-programmers \cite{enduser,smallmatter}
-and work on tools for journalists such as Idyll \cite{idyll}. Related work includes a range of
-visual and programmatic tools used by data scientists, work that aims to simplify programming
-or explore alternative paradigms and various predictive tools that assist users in completing
-programming or data exploration tasks.
-
-\subsection{Data science tools}
-Data exploration is typically done using spreadsheets, comprehensive visual tools or
-via code using notebooks. Notebooks such as Jupyter \cite{jupyter} are also popular with technically
-skilled journalists as they allow combining code with commentary and visual outputs. However,
-notebooks can lead to messy hard to reproduce code \cite{messes,wrattler}.
-
-Visual data analytics tools \cite{control,vizdom} such as Tableau \cite{tableau} do not require
-programming skills, but require mastering a range of different interactions. They also make it
-hard to see what data transformations have been done. Several systems for data exploration,
-e.g.~Potter's Wheel and Wrangler \cite{potter,wrangler}, and visualization, e.g.~Lyra \cite{lyra},
-provide a GUI, but record a transparent script that can be seen and modified by the user.
-
-\subsection{Easier programming tools}
-We aim to design a programming environment that is easy to use and learn. Many approaches to this
+The key contribution of our work is that it develops a new, fundamentally different, way of using the
+established auto-completion mechanism. Unlike most past work dating back to Kaiser~\cite{assistants},
+we do not view it as a programmer assistance tool, but as an interaction mechanism through which
+non-experts can create entire programs. We build on recent research on information-rich
+programming \cite{inforich} and aim to make those advances available to non-programmers
+\cite{enduser,smallmatter}, in the context of data exploration as done, most notably, by
+journalists \cite{ddj}. Our work features a novel combination of characteristics in that
+the iterative prompting we develop (i) is centered around editing and understanding of program code,
+(ii) its conceptual complexity is reduced to a single basic kind of interaction, yet (iii) it is
+correct and complete in that it can be used to construct all meaningful programs.
+
+\subsubsection*{Code completion for data science}
+
+A key component in The Gamma is the use of auto-complete for offering possible operations.
+Our work follows type providers \cite{inforich,fsdata}, which integrate external data into a
+static type system of F\#, allowing the use of auto-completion; for querying data tables, we utilize
+the theory developed by Petricek \cite{dotdriven}. The key difference in our work is that The Gamma
+can be used without a programming language expertise.
+
+Most similar to our approach are tools that recommend scripts when users begin interacting
+with data. Those based on machine learning code completion for domain-specific languages \cite{predictive,proactive}
+differ in that they do not guarantee completeness, i.e.~the user cannot create all possible
+scripts. Approaches based on natural language can effectively support data exploration
+or visualization \cite{eviza,codemend}, but hide the structure of the underlying language and
+require more than just selecting options. Conversational agents \cite{iris} improve on such work
+in that they can provide more guidance. Code completion based on machine learning or statistical
+methods \cite{mlcomplete,statcomplete} also exists for general-purpose programming languages used
+by data scientists such as Python \cite{pythia}, providing assistance to expert programmers.
+Finally, DS.js \cite{dsjs} is interesting in that it enables querying of data on the
+web; it uses JavaScript with rich contextual code completion.
+
+\subsubsection*{Notebooks and business intelligence tools}
+
+Notebooks such as Jupyter \cite{jupyter}, which allow combining source code with commentary and
+visual outputs, are widely used by data scientists, but require expert programming skills.
+The Gamma targets non-experts, but could be easily integrated with a multi-language notebook system
+such as Wrattler \cite{wrattler}.
+
+Spreadsheets and business intelligence tools \cite{tableau,powerbi} do
+not involve programming, but require mastering a complex GUI. This is also the case
+for other visual data analytics tools \cite{control,vizdom}. In contrast, The Gamma is
+based on a single kind of interaction. Several visual systems \cite{potter,wrangler,lyra} record
+interactions with the GUI as a script that can be seen and modified by the user.
+Unlike in The Gamma, the source code does not guide the user in learning how to use the system.
+
+\subsubsection*{Easier programming tools}
+We aim to build an easy to use and learn programming system. Many approaches to this
 goal have been tried. Victor \cite{learnable,principle} introduced design principles that
 inspired many to build live programming systems \cite{review,liveroad,lighttable} that give
-immediate feedaback to help programmers understand how code relates to output and
+immediate feedback to help programmers understand how code relates to output and
 exploratory systems \cite{variolite,exploratory} that assist with completing open-ended tasks.
-
+A system combining textual language with visualization also exists for graph querying \cite{guess}.
 To avoid difficulties with editing code as text, some systems use structured editors~\cite{structure-based,livenut,lamdu}.
 In Subtext \cite{subtext,directprog} the language itself is co-designed with the editor to make
-the interactions with code more natural. Finally, programming can be made easier by designing
-high-level declarative abstractions. Tea does so for statistical analyses \cite{tea} and
-Satyanarayan et al. \cite{interactionviz,vegalite} does so for interactive data visualizations.
-
-The Gamma is live in that it gives an instant preview of the results. It is text-based,
-although it is syntactically simple. The code written in The Gamma is high-level and so work
-on declarative abstractions is a valuable inspiration.
-
-\subsection{Code completion and assistants}
-A key component in The Gamma design is the use of auto-complete for offering possible
-operations. Code completion has been pioneerd by Kaiser~\cite{assistants}, but our work is more
-directly inspired by the work on type providers \cite{inforich,fsdata,dotdriven}, which utilize
-code completion to integrate external data into a statically typed programming language.
-
-When using type providers, completions are offered as the user types code. However, completion
-tools can also recmmend scripts when users begin interacting with data \cite{predictive,proactive}.
-Alternatively, they can also be guided by natural language as for example Eviza~\cite{eviza},
-which uses natural language for data exploration. CodeMend~\cite{codemend} uses natural language
-for more general code edits, but uses data visualization as a case study.
-Finally, data entry interfaces based on completion such as Dasher \cite{dasher} utilize an
-interaction principle similar to our iterative prompting.
-
-\subsection{Programming without writing code}
-
-A number of tools for data exploration use alternative approaches to programming where
+the interactions with code more natural. The Gamma is live in that our editor gives an instant
+preview of the results.
+
+Many systems simplify programming by designing high-level declarative abstractions,
+e.g.~for interactive news articles \cite{idyll}, statistical analyses \cite{tea}
+or interactive data visualization \cite{interactionviz,vegalite}. The Gamma uses
+high-level abstractions for data querying, but designing high-level iterative prompting
+abstractions for other tasks remains future work.
+
+\subsubsection*{Programming without writing code}
+There are two main approaches to programming where
 the user does not write code. In programming by example \cite{byexample}, the user gives
-examples of desired results. In data science, this approach has been used for example for
-specifying data transformations in spreadsheets and data extraction \cite{spreadsheetpbe,flashextract}.
-
+examples of desired results. This has been used, e.g.~for specifying data transformations
+in spreadsheets and data extraction \cite{spreadsheetpbe,flashextract}.
 In direct manipulation \cite{direct}, a program is specified by directly interacting with the
@@
 users partially edit queries,~e.g. by changing quantifiers as in DataPlay~\cite{dataplay}.
+
+\subsubsection*{Guestures and data entry}
+Although our focus is on program construction, our work can be positioned in the
+broader context of input methods. Akin to Dasher \cite{dasher}, our system provides a way of
+navigating through a complete space of options, while on-screen feedforward \cite{octopocus} allows
+efficient selection in guesture-based interfaces. Those provide compelling alternatives to
+auto-completion menus, although the efficiency of input methods is not an issue in programming.
+
+\section{Overview}
+
+design motivated by exploratory question - what if we could program with autocomplete
+no talking to journalists hence shorter paper
+
+maybe talk about "design space"
+
+\newpage
 
@@
 
-\tg is a text-based programming environment that allows non-experts create simple data exploration
+\tg is a text-based programming environment that allows non-experts to create simple data exploration
 scripts using a single interaction principle -- choosing an item from an auto-complete list.
@@
 
+\subsection{Correct and complete}
+any program can be created by autocomplete, all programs created by autocomplete are correct
+
 \subsection{Lowering the Barrier to Entry}
@@
 
-As illustrated in Figure~\ref{fig:cubetp}, the provider allows uses to select a data series
+As illustrated in Figure~\ref{fig:cubetp}, the provider allows users to select a data series
 from the data cube. By choosing \ikvd{byCountry.\textquotesingle United States\textquotesingle},
@@
 Our second type provider allows users to construct queries to explore data in tabular formats
-such as CSV files. It is based on the theory developed Petricek~\cite{dotdriven}. Unlike the
+such as CSV files. It is based on the theory developed Petricek~\cite{dotdriven}. Unlike the  % grammar!
 data cube provider, the provider for tabular data does not just allow selecting a subset of the
@@
 \paragraph{RQ1: Can non-programmers explore data with The Gamma?}
-We give participants one of four simeple data exploration tasks and see if they can complete the
+We give participants one of four simple data exploration tasks and see if they can complete the
 task using The Gamma and, possibly, how much assistance they need.
@@
 \paragraph{RQ3: Can users learn from just code samples?}
-Our second hypothesis is that users can learn through precolation, i.e.~by looking at the source
+Our second hypothesis is that users can learn through percolation, i.e.~by looking at the source
 code of published analyses. In one of our tasks, participants are given a code sample, but only a
@@
 % References must be the same font size as other body text.
-\bibliographystyle{SIGCHI-Reference-Format}
+\bibliographystyle{ACM-Reference-Format}
 \bibliography{paper}

### 2020-09-12 wip-design/paper.tex
Edit overview, move implementation

@@
 \documentclass[manuscript,review,anonymous]{acmart}
+\usepackage{caption}
+\usepackage{subcaption}
 \usepackage{hyperref}
@@
 % \usepackage{xspace}
-% \usepackage{fancyvrb}
+\usepackage{fancyvrb}
 % \usepackage{inconsolata}
-% \usepackage{xcolor}
+\usepackage{xcolor}
 % \usepackage{tikz}
@@
 %
-% \definecolor{kvdclr}{rgb}{0.3,0.0,0.8}
-% \DefineVerbatimEnvironment{thegamma}{Verbatim}{fontfamily=zi4,numbers=left,xleftmargin=6mm,fontsize=\small,commandchars=\\\{\}}
+\definecolor{kvdclr}{rgb}{0.3,0.0,0.8}
+\DefineVerbatimEnvironment{thegamma}{Verbatim}{fontfamily=zi4,numbers=left,xleftmargin=6mm,fontsize=\small,commandchars=\\\{\}}
 % %\newcommand{\kvd}[1]{\textcolor{kvdclr}{#1}}
-% \newcommand{\kvd}[1]{\textbf{#1}}
-% \newcommand{\ikvd}[1]{{\fontfamily{zi4}\selectfont\small #1}}
-% \newcommand{\tg}{The Gamma\xspace}
+\newcommand{\kvd}[1]{\textbf{#1}}
+\newcommand{\ikvd}[1]{{\fontfamily{zi4}\selectfont\small #1}}
 % \usepackage{balance}       % to better equalize the last page
@@
 
-\subsubsection*{Iterative prompting}
+\subsubsection*{Iterative Prompting}
 The key idea in The Gamma, which we term the \emph{iterative prompting} principle, is that all
@@
 
-\subsubsection*{Date exploration}
+\subsubsection*{Data Exploration}
 The Gamma focuses on data exploration of the kind illustrated in Figure~\ref{fig:thegamma}.
@@
 \subsubsection*{Contributions}
-The Gamma is available (non-anonymously) at \href{http://thegamma.net}{\textcolor{blue}{thegamma.net}},
+The Gamma is available (non-anonymously) at \url{http://thegamma.net},
 both as a JavaScript library and a hosted data exploration service. In this paper, we describe and
@@
 
-\section{Related work}
+\section{Related Work}
 
@@
 
-\subsubsection*{Code completion for data science}
+\subsubsection*{Code Completion for Data Science}
 
@@
 
-\subsubsection*{Notebooks and business intelligence tools}
+\subsubsection*{Notebooks and Business Intelligence Tools}
 
@@
 
-\subsubsection*{Easier programming tools}
+\subsubsection*{Easier Programming Tools}
 We aim to build an easy to use and learn programming system. Many approaches to this
@@
 
-\subsubsection*{Programming without writing code}
+\subsubsection*{Programming without Writing Code}
 There are two main approaches to programming where
@@
 
-\subsubsection*{Guestures and data entry}
+\subsubsection*{Guestures and Data Entry}
 Although our focus is on program construction, our work can be positioned in the
@@
 
+\begin{figure}[b]
+\includegraphics[width=1\columnwidth]{figures/thegamma-walk}
+\caption{Constructing a script that charts the top 5 members of the House of Lords for Kent, based
+on their travel costs.}
+\label{fig:walkthrough}
+\end{figure}
+
 \section{Overview}
-
-design motivated by exploratory question - what if we could program with autocomplete
-no talking to journalists hence shorter paper
-
-maybe talk about "design space"
+\label{sec:overview}
+
+The review of related work suggests that there is an unexplored point in the design space of tools
+for data exploration. Although various efforts make text-based programming easier, e.g. by providing
+high-level declarative abstractions, most systems that target non-experts shy away from code, by using
+either programming by example, natural language or graphical user interface. As spelled out in
+Section~\ref{sec:design}, text-based programming has notable benefits for public-facing data analyses
+such as learnability and transparency. Our work is thus motivated primarily by the question whether we
+can make text-based programming easy enough for non-experts? The study presented in Section~\ref{sec:study}
+shows that this is, indeed, possible at least for typical data querying tasks.
+The Gamma consists of a programming language, a web-based coding environment and a number of type providers
+that enable access to various kinds of data sources. It allows non-experts to create programs using the
+\emph{iterative prompting} interaction principle -- by repeately selecting an item from an auto-complete list.
+We start with a walkthrough of the system (Section~\ref{sec:overview-walk}), before looking at details of
+The Gamma language (Section~\ref{sec:overview-lang}) and individual type providers (Section~\ref{sec:overview-tps}).
+
+\subsection{Querying Travel Expenses}
+\label{sec:overview-walk}
+
+To introduce The Gamma, we walk through a simple problem that a local journalist might want to solve.
+The UK government publishes travel expense claims by members of the House of Lords. We want to find
+out which of the members representing the Kent county spend the most on travel.
+The following shows a subset of the data:\footnote{Full data set has been obtained from
+\url{https://www.parliament.uk/mps-lords-and-offices/members-allowances/house-of-lords/holallowances/} }
+
+\begin{thegamma}
+\textbf{Name, County, Days Attended, Days Away, Travel Costs}
+Lord Adonis, London, 8, 0, 504
+Baroness Afshar, Yorkshire, 2, 0, 0
+Lord Alderdice, Oxfordshire, 3, 0, 114
+Lord Alli, London, 5, 0, 0
+\end{thegamma}
+% Baroness Amos, London, 3, 0, 0
+
+\noindent
+The data is available as a CSV file. After the analyst imports the file through a web interface,
+the environment is initialized with code that refers to the imported data as \ikvd{expenses}
+and she starts exploring the data using the type provider for tabular data (Section~\ref{sec:overview-tps}),
+following the steps illustrated in Figure~\ref{fig:walkthrough}:
+
+\begin{enumerate}
+\item The analyst types `.' (dot) to trigger auto-completion on \ikvd{expenses}. The type provider
+  offers a list of operations that the analyst might want to perform such as grouping,
+  filtering and sorting.
+
+\item To find the House of Lords with the largest spending, the analyst chooses the
+  \ikvd{sort data} operation. Next, she is offered a list of possible arguments based on the
+  schema of the tabular data and chooses the desired one.
+
+\item The Gamma evaluates the source code on-the-fly and shows a preview
+  of results. After choosing the sorting key, the analyst sees a table with House of Lords members
+  from more remote counties of the UK.
+
+\item To finish specifying the (possibly compound) sorting key, the analyst chooses \ikvd{then}
+  and is offered the same list of querying operations as in the first step. To obtain House
+  members from Kent, she chooses \ikvd{filter data}. To navigate through the offered list more
+  efficiently, she types first two characters of the name.
+
+\item After selecting \ikvd{County is} to specify the desired type of condition, the analyst types
+  `.' and is offered a list of options based on the values of the \ikvd{County} column in the
+  source data set. She types \ikvd{Ke} and selects \ikvd{Kent}.
+
+\item The analyst chooses \ikvd{then} and is,
+  again, offered the list of querying operation. She uses \ikvd{paging} to get top 5 records,
+  which requires typing \ikvd{5} as the argument. She then uses the \ikvd{get series} operation
+  to obtain a data series associating travel expenses with a name, which is automatically
+  visualized using a bar chart.
+\end{enumerate}
+
+\noindent
+The constructed code is not unlike an SQL query, except that the whole script is constructed using
+iterative prompting, by repeatedly selecting one of the offered members. Those represent both
+operations, such as \ikvd{sort by} and arguments, such as \ikvd{Kent}. The only exception
+is when the analyst needs to type the number \ikvd{5} to specify the number of items to take.
+
+\subsection{The Gamma Programming Environment}
+\label{sec:overview-lang}
+
+The Gamma consists of a text-based programming language with a web-based coding environment,
+based on the Monaco editor\footnote{For more information, see \url{https://microsoft.github.io/monaco-editor/}},
+and type providers that provide access to data tables, graph databases and data cubes.
+
+\subsubsection*{The Gamma Language}
+A program in The Gamma is a sequence of commands. A command can be either a variable declaration
+or an expression that evaluates to a value such as a data table or a chart.
+An expression is a reference to a data source followed by a chain of member accesses.
+A member can be either an ordinary member such as \ikvd{paging} or an operation which takes a
+list of parameters enclosed in parentheses as in \ikvd{take(5)}. When the member name contains
+non-alphanumerical characters it is  written in quotes such as \ikvd{\textquotesingle sort by\textquotesingle}.
+
+The Gamma uses a type system to infer what members are available at a given point in a chain.
+Each expression has a type with a list of members that, in turn, have their own types.
+The types are not built-in, but are generated by type providers for individual data sources.
+The programming environment for The Gamma is based on a text editor. When the user types `.'
+the editor triggers auto-completion and retrives a list of available members based on the type
+information. The programming environment evaluates scripts on-the-fly and shows a preview as
+illustrated in Figure~\ref{fig:thegamma}.
+
+\subsubsection*{Making Complex Things Possible}
+As illustrated by the \ikvd{take(5)} operation, there is a handful of situations where The
+Gamma does not yet fully support the iterative prompting principle. The language supports
+a small number of other features that can be used by more advanced users through text editing:
+
+\vspace{0.3em}
+\begin{thegamma}
+\kvd{let} topTravel = expenses.'sort data'.'by Travel Costs descending'.then
+  .paging.take(5).'get series'.'with key Name'.'and value Travel Costs'
+charts.column(topTravel).setColors(["red","blue","green"])
+  .setTitle("House of Lords members by travel expenses")
+\end{thegamma}
+\vspace{0.3em}
+
+\noindent
+First, The Gamma allows operations with parameters such as \ikvd{take(5)} or \ikvd{setTitle("...")}.
+This is currently needed when writing a query that skips or takes first N elements from a table.
+The remaining features are not needed for basic data exploration. The \ikvd{let} construct can be
+used to define (immutable) variables and The Gamma also supports lists written as \ikvd{[1,2,3]}.
+Advanced language features are currently used when building custom charts, but we expect that a
+charting library compatible with iterative prompting would alleviate the need for most of those.
+
+\subsection{Providers}
+\label{sec:overview-tps}
+
+all the important logic
+
+define DSLs
+
+\subsubsection*{Data Cube Type Provider}
+Our first type provider allows users explore data form a data cube. For example, the World
+Bank\footnote{\url{https://data.worldbank.org/}} collects a range of indicators about many
+countries in the world each year. The data set is a three-dimensional cube with dimensions
+corresponding to countries, indicators and years. The following example, also shown in
+Figure~\ref{fig:sidebyside}, uses the provider to access CO2 emission data for United States:
+
+\begin{thegamma}
+worldbank.byCountry.'United States'
+  .'Climate Change'.'CO2 emissions (kt)'
+\end{thegamma}
+
+As illustrated in Figure~\ref{fig:cubetp}, the provider allows users to select a data series
+from the data cube. By choosing \ikvd{byCountry.\textquotesingle United States\textquotesingle},
+we restrict the cube to a two-dimensional plane. We then choose an indicator category
+\ikvd{\textquotesingle Climate Change\textquotesingle} and a specific indicator
+\ikvd{\textquotesingle CO2 emissions (kt)\textquotesingle}, obtaining a time series with
+years as keys and emission data as values. The World Bank provider also supports filtering
+by year and indicator.
+
+Another type provider available in The Gamma that is based on the data cube structure allows
+the user to explore UK government expenditure:
+
+\begin{thegamma}
+expenditure.byService.Defence.inTermsOf.GDP
+\end{thegamma}
+
+The dimensions of the cube are government services, years and value type (adjusted, nominal,
+per GDP). Here, we select the \ikvd{Defence} service and \ikvd{GDP} value type. Our implementation
+does not yet use a standardized data cube storage and so adding another data cube source
+currently requires implementing a new type provider.
+
+\subsubsection*{Tabular Data Type Provider}
+
+Our second type provider allows users to construct queries to explore data in tabular formats
+such as CSV files. It is based on the theory developed Petricek~\cite{dotdriven}. Unlike the  % grammar!
+data cube provider, the provider for tabular data does not just allow selecting a subset of the
+data, but it can be used to construct SQL-like query. Consider the example from Figure~\ref{fig:thegamma}:
+
+\begin{thegamma}
+olympics.'filter data'.'Games is'.'Rio (2016)'.then
+  .'group data'.'by Team'.'sum Gold'.'sum Silver'.then
+  .'sort data'.'by Gold descending'
+\end{thegamma}
+
+The example works with a CSV file that records individual medals awarded in Olympic games.
+The chain constructs a query that selects rows corresponding to the Rio 2016 Olympics and then
+calculates total number of gold and silver medals for each team (country) before sorting the data.
+
+When using the provider, the user specifies a sequence of operations. Members such as
+\ikvd{\textquotesingle filter data\textquotesingle} or \ikvd{\textquotesingle group data\textquotesingle}
+determine the operation type. Those are followed by operation parameters. For example, when grouping
+data, we first select the key and then add a number of aggregations to calculate over the group.
+Unlike SQL, the provider only allows users to choose from pre-defined aggregations such as
+calculating the sum, average or the number of distinct values. As illustrated in
+Section~\ref{sec:cases}, this still allows us to construct a wide range of practical queries.
+
+\begin{figure}
+\centering
+\begin{subfigure}[b]{0.5\textwidth}
+  \centering
+  \includegraphics[scale=0.25]{figures/cubetp}
+  \vspace{0.5em}
+  \caption{The World Bank provider allows users to first choose
+    a country or a year and then an indicator.}
+  \label{fig:cubetp}
+\end{subfigure}
+\hfill
+\begin{subfigure}[b]{0.45\textwidth}
+  \centering
+  \includegraphics[scale=0.28]{figures/graphtp}
+  \caption{The user specifies a path through the data, possibly with
+    placeholders to select multiple nodes.}
+  \label{fig:graphtp}
+\end{subfigure}
+\vspace{-0.5em}
+\caption{Design of type providers for exploring data cubes and graph databases.}
+\label{fig:three graphs}
+\end{figure}
+
+\subsubsection*{Graph Database Type Provider}
+Our third type provider allows users explore data from graph databases. A graph database
+consists of nodes representing entities and relationships between them. In the context of
+data journalism, graph databases have been used for example in The Panama Papers reporting \cite{panama}.
+
+The following example explores a database of Doctor Who characters and episodes. It retrieves
+all enemies of the Doctor that appear in the Day of the Moon episode:
+
+\begin{thegamma}
+drwho.Character.Doctor
+  .'ENEMY OF'.'[any]'
+  .'APPEARED IN'.'Day of the Moon'
+\end{thegamma}
+
+We start from the \ikvd{Doctor} node and then follow two relationships. We use
+\ikvd{\textquotesingle ENEMY OF\textquotesingle.\textquotesingle [any]\textquotesingle}
+to follow links to all enemies of the Doctor and then specify
+\ikvd{\textquotesingle APPEARED IN\textquotesingle.\textquotesingle Day of the Moon\textquotesingle}
+to only select only such enemies that appear in a specific episode. The resulting query
+is illustrated in Figure~\ref{fig:graphtp}.
+
+The provider works with any graph database and generates members automatically, based on the
+data in the database. In the above example, \ikvd{ENEMY OF} and \ikvd{APPEARED IN} are labels
+of relations and \ikvd{Doctor} and \ikvd{Day of the Moon} are labels of nodes. The
+\ikvd{[any]} member defines a placeholder that can be filled with any node with the specified
+relationships. The results returned by the provider is a table of properties of all nodes
+along the specified path. As discussed in the next section, such table can be further queried
+using the tabular data type provider.
 
 \newpage
-
 % ==================================================================================================
@@
 
-\section{Overview}
-\label{sec:overview}
-
-\tg is a text-based programming environment that allows non-experts to create simple data exploration
-scripts using a single interaction principle -- choosing an item from an auto-complete list.
-It supports a range of data sources including tabular data, graph data and data cubes.
-
-\subsection{Querying Travel Expenses}
-To introduce \tg, we walk through a simple problem that a local journalist might want to solve.
-The UK government publishes travel expense claims by members of the House of Lords. A journalist
-wants to find out which of the members representing the Kent county spend the most on travel.
-The following shows a subset of the data:\footnote{ \url{https://www.parliament.uk/mps-lords-and-offices/members-allowances/house-of-lords/holallowances/} }
-
-\begin{thegamma}
-\textbf{Name, County, Days Attended, Days Away, Travel Costs}
-Lord Adonis, London, 8, 0, 504
-Baroness Afshar, Yorkshire, 2, 0, 0
-Lord Alderdice, Oxfordshire, 3, 0, 114
-Lord Alli, London, 5, 0, 0
-Baroness Amos, London, 3, 0, 0
-\end{thegamma}
-
-After the analyst imports the CSV file (through a web interface), the environment is initialized
-with code that refers to the imported variable \ikvd{expenses}. The analyst then types `.' (dot):
-
-\begin{thegamma}
-expenses.
-\end{thegamma}
-\vspace{-0.4em}\qquad\includegraphics[width=0.5\columnwidth]{figures/lords1}
-
-The type provider for tabular data allows analysts to construct simple queries. It first offers
-a list of operations that the analyst might want to perform such as grouping, filtering and sorting.
-To find members of the House of Lords from Kent, the analyst chooses \ikvd{filter data},
-types `.' and then chooses \ikvd{County is} from the offered list, types `.' and starts
-typing Kent:
-
-\begin{thegamma}
-expenses
-  .'filter data'.'County is'.Ke
-\end{thegamma}
-\vspace{-0.4em}\qquad\includegraphics[width=0.5\columnwidth]{figures/lords2}
-
-The completion list is generated from the values in the \ikvd{County} column of the dataset.
-After selecting \ikvd{Kent}, the live preview is updated to only show records according to the
-specified filter:
-
-\includegraphics[width=1\columnwidth]{figures/lords3}
-
-To finish specifying filtering conditions, the analyst chooses \ikvd{then} and is offered the
-same list of querying operations as in the first step. To sort House of Lords members by their
-travel costs, she now chooses \ikvd{sort data} and types `.' (dot):
-
-\begin{thegamma}
-expenses
-  .'filter data'.'County is'.Kent.then
-  .'sort data'.
-\end{thegamma}
-\qquad\includegraphics[width=0.5\columnwidth]{figures/lords4}
-
-The auto-complete offers a list of columns that can be used for sorting, each with ascending
-(default) and descending order option. After choosing one or more sort keys, the analyst selects
-the \ikvd{then} member and is, again, offered the list of querying operations. They use
-\ikvd{paging} to get top 5 records and \ikvd{get series} to obtain a data series with just
-the House of Lords member name and their travel expenses.
-
-\begin{thegamma}
-expenses
-  .'filter data'.'County is'.Kent.then
-  .'sort data'.'by Travel Costs descending'.then
-  .paging.take(5)
-  .'get series'.'with key Name'.'and value Travel Costs'
-\end{thegamma}
-
-When the code evaluates to a data series with a categorical (textual) key and a numerical value,
-\tg switches from displaying the result as a table to a column chart:
-
-\includegraphics[width=1\columnwidth]{figures/lords5}
-\vspace{-1.5em}
-
-\subsection{Querying via Iterative Prompting}
-In the example, the journalists constructs a query that filters and sorts a data table. This is
-not unlike writing a query in the SQL language. However, the whole query is constructed through
-a single interaction mechanism that we refer to as \emph{iterative prompting}. The user triggers
-the interaction by typing `.' at the end of the query constructed so far. She is then offered a
-list of available operations in the current context. This may include a choice of top-level
-operations (e.g.~at the start and then again after choosing \ikvd{then}) or possible
-parameters for an operation (such as possible sorting keys and directions after choosing
-\ikvd{sort data}). The user then merely needs to choose one of the available options, rather than
-learning the syntax and semantics of SQL in order to know what constructs can be used in the
-given context.
-We discuss the important aspects of The Gamma design in the next section and then discuss the
-integration of specific data sources in detail, revisiting the type provider for querying
-data tables.
-
 % ==================================================================================================
@@
 
-\subsection{Making Complex Things Possible}
-Although the focus on The Gamma is simple data exploration and almost all the use cases we
-envision can be achieved using the iterative prompting interaction principle, there remain a
-few cases where more flexibility is needed. We aim to \emph{make simple things easy and
-complex things possible}. The language thus supports a small number of additional constructs that
-can only be used through text editing. The following example illustrates three additional
-constructs:
-
-\begin{thegamma}
-\kvd{let} topTravel =
-  expenses
-    .'sort data'.'by Travel Costs descending'.then
-    .paging.take(5).'get series'
-    .'with key Name'.'and value Travel Costs'
-
-charts.column(topTravel)
-  .setTitle("House of Lords members by travel expenses")
-  .setColors(["red","blue","green"])
-\end{thegamma}
-
-First, The Gamma allows operations with parameters such as \ikvd{take(3)} or \ikvd{setTitle("...")}.
-This is currently needed when writing a query that skips or takes first N elements from a table.
-The remaining features are not needed for basic data exploration, but allow other scenarios, such
-as manual creation of charts. The \ikvd{let} construct can be used to define (immutable) variables
-and The Gamma also supports lists written as \ikvd{[1,2,3]}.
-
 % ==================================================================================================
@@
 \end{figure}
-
-\section{Implementation}
-\label{sec:implementation}
-The Gamma consists of a programming language with a web-based coding environment and a number
-of type providers that provide access to various kinds of data sources. The system is implemented
-in F\# and compiled to JavaScript using Fable.
-
-\subsection{The Gamma Language}
-In The Gamma, a program is a sequence of commands. A command can be either a variable declaration
-or an expression that evaluates to a value such as a data table or a chart.
-An expression is a reference to a data source followed by a chain of member accesses.
-A member can be either an ordinary member or an operation which takes a list of parameters
-enclosed in parentheses. When the member name contains non-alphanumerical characters such as
-whitespcae, it needs to be written in quotes. Our implementation supports a few other features,
-but those are not used in this paper.
-
-The Gamma uses a type system to infer what members are available at a given point in a chain.
-A type is an object with a list of members that, in turn, have their own types.
-The types are not built-in, but are generated by a type providers for individual data sources.
-The types are generated lazily and are only computed when a member appears in code.
-
-The web-based coding environment uses the Monaco\footnote{\url{https://microsoft.github.io/monaco-editor/}}
-editor and implemts auto-completion feature that is triggered when the user types `.' and relies
-on information from the type system. The can display the value of a current command inline
-as in Figure~\ref{fig:thegamma}, or values of all commands on the side as in Figure~\ref{fig:sidebyside}.
-In the latter case, it can also combine one or more data series into a single chart.
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.28]{figures/cubetp}
-\caption{Exploring data cubes. The World Bank provider allows users to first choose
-  a country or a year and then an indicator.}
-\label{fig:cubetp}
-\end{figure}
-
-\subsection{Data Cube Type Provider}
-Our first type provider allows users explore data form a data cube. For example, the World
-Bank\footnote{\url{https://data.worldbank.org/}} collects a range of indicators about many
-countries in the world each year. The data set is a three-dimensional cube with dimensions
-corresponding to countries, indicators and years. The following example, also shown in
-Figure~\ref{fig:sidebyside}, uses the provider to access CO2 emission data for United States:
-
-\begin{thegamma}
-worldbank.byCountry.'United States'
-  .'Climate Change'.'CO2 emissions (kt)'
-\end{thegamma}
-
-As illustrated in Figure~\ref{fig:cubetp}, the provider allows users to select a data series
-from the data cube. By choosing \ikvd{byCountry.\textquotesingle United States\textquotesingle},
-we restrict the cube to a two-dimensional plane. We then choose an indicator category
-\ikvd{\textquotesingle Climate Change\textquotesingle} and a specific indicator
-\ikvd{\textquotesingle CO2 emissions (kt)\textquotesingle}, obtaining a time series with
-years as keys and emission data as values. The World Bank provider also supports filtering
-by year and indicator.
-
-Another type provider available in The Gamma that is based on the data cube structure allows
-the user to explore UK government expenditure:
-
-\begin{thegamma}
-expenditure.byService.Defence.inTermsOf.GDP
-\end{thegamma}
-
-The dimensions of the cube are government services, years and value type (adjusted, nominal,
-per GDP). Here, we select the \ikvd{Defence} service and \ikvd{GDP} value type. Our implementation
-does not yet use a standardized data cube storage and so adding another data cube source
-currently requires implementing a new type provider.
-
-\subsection{Tabular Data Type Provider}
-
-Our second type provider allows users to construct queries to explore data in tabular formats
-such as CSV files. It is based on the theory developed Petricek~\cite{dotdriven}. Unlike the  % grammar!
-data cube provider, the provider for tabular data does not just allow selecting a subset of the
-data, but it can be used to construct SQL-like query. Consider the example from Figure~\ref{fig:thegamma}:
-
-\begin{thegamma}
-olympics.'filter data'.'Games is'.'Rio (2016)'.then
-  .'group data'.'by Team'.'sum Gold'.'sum Silver'.then
-  .'sort data'.'by Gold descending'
-\end{thegamma}
-
-The example works with a CSV file that records individual medals awarded in Olympic games.
-The chain constructs a query that selects rows corresponding to the Rio 2016 Olympics and then
-calculates total number of gold and silver medals for each team (country) before sorting the data.
-
-When using the provider, the user specifies a sequence of operations. Members such as
-\ikvd{\textquotesingle filter data\textquotesingle} or \ikvd{\textquotesingle group data\textquotesingle}
-determine the operation type. Those are followed by operation parameters. For example, when grouping
-data, we first select the key and then add a number of aggregations to calculate over the group.
-Unlike SQL, the provider only allows users to choose from pre-defined aggregations such as
-calculating the sum, average or the number of distinct values. As illustrated in
-Section~\ref{sec:cases}, this still allows us to construct a wide range of practical queries.
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.28]{figures/graphtp}
-\vspace{0.5em}
-\caption{Exploring graph databases. The user specifies a path through the data, possibly with
-  placeholders to select multiple nodes.}
-\label{fig:graphtp}
-\end{figure}
-
-\subsection{Graph Database Type Provider}
-Our third type provider allows users explore data from graph databases. A graph database
-consists of nodes representing entities and relationships between them. In the context of
-data journalism, graph databases have been used for example in The Panama Papers reporting \cite{panama}.
-
-The following example explores a database of Doctor Who characters and episodes. It retrieves
-all enemies of the Doctor that appear in the Day of the Moon episode:
-
-\begin{thegamma}
-drwho.Character.Doctor
-  .'ENEMY OF'.'[any]'
-  .'APPEARED IN'.'Day of the Moon'
-\end{thegamma}
-
-We start from the \ikvd{Doctor} node and then follow two relationships. We use
-\ikvd{\textquotesingle ENEMY OF\textquotesingle.\textquotesingle [any]\textquotesingle}
-to follow links to all enemies of the Doctor and then specify
-\ikvd{\textquotesingle APPEARED IN\textquotesingle.\textquotesingle Day of the Moon\textquotesingle}
-to only select only such enemies that appear in a specific episode. The resulting query
-is illustrated in Figure~\ref{fig:graphtp}.
-
-The provider works with any graph database and generates members automatically, based on the
-data in the database. In the above example, \ikvd{ENEMY OF} and \ikvd{APPEARED IN} are labels
-of relations and \ikvd{Doctor} and \ikvd{Day of the Moon} are labels of nodes. The
-\ikvd{[any]} member defines a placeholder that can be filled with any node with the specified
-relationships. The results returned by the provider is a table of properties of all nodes
-along the specified path. As discussed in the next section, such table can be further queried
-using the tabular data type provider.
 

### 2020-09-12 wip-design/paper.tex
Providers, merge design and motivation

@@
 \begin{figure}
-\includegraphics[width=.54\columnwidth]{figures/thegamma-annot}
+\includegraphics[width=.6\columnwidth]{figures/thegamma-annot}
 \caption{Obtaining teams with the greatest number of gold medals from Rio 2016
@@
 \end{figure}
+
+% ==================================================================================================
 
@@
 \begin{itemize}
-\item We introduce the iterative prompting principle in The Gamma (Section~\ref{sec:overview})
+\item We introduce the iterative prompting principle in The Gamma
   and show that it can be used for querying of distinct data sources including data tables, graph
-  databases and data cubes (Section~\ref{sec:implementation}).
+  databases and data cubes (Section~\ref{sec:overview}).
+
+% TODO REWRITE
 
@@
 \end{itemize}
+
+% ==================================================================================================
 
@@
 
-\subsection{Providers}
+\subsection{Type Providers for Data Querying}
 \label{sec:overview-tps}
 
-all the important logic
-
-define DSLs
+The Gamma can be extended to support any kind of data source by implementing a \emph{type provider}.
+Conceptually, a type provider defines a domain-specific language for exploring data of a particular
+kind. Techincally, it generates object types with members (such as \ikvd{sort by} or \ikvd{Count is})
+that are accessed via iterative prompting. We describe type provider for exploring data cubes
+(inspired by an example from Syme et al. \cite{inforich}), tabular data (based on theory developed
+by Petricek \cite{dotdriven}), and a novel type provider for exploring graph databases.
 
 \subsubsection*{Data Cube Type Provider}
-Our first type provider allows users explore data form a data cube. For example, the World
-Bank\footnote{\url{https://data.worldbank.org/}} collects a range of indicators about many
-countries in the world each year. The data set is a three-dimensional cube with dimensions
-corresponding to countries, indicators and years. The following example, also shown in
-Figure~\ref{fig:sidebyside}, uses the provider to access CO2 emission data for United States:
+Our first type provider allows users to explore data form a data cube. For example, the World
+Bank\footnote{The data is retrived using an API available at: \url{https://data.worldbank.org/}}
+collects a range of indicators about many countries in the world each year. The data set is a
+three-dimensional cube with dimensions corresponding to countries, indicators and years. The
+following example uses the provider to access CO$_{2}$ emission data for United States:
 
 \begin{thegamma}
-worldbank.byCountry.'United States'
-  .'Climate Change'.'CO2 emissions (kt)'
+worldbank.byCountry.'United States'.'Climate Change'.'CO2 emissions (kt)'
 \end{thegamma}
 
+\noindent
 As illustrated in Figure~\ref{fig:cubetp}, the provider allows users to select a data series
-from the data cube. By choosing \ikvd{byCountry.\textquotesingle United States\textquotesingle},
-we restrict the cube to a two-dimensional plane. We then choose an indicator category
-\ikvd{\textquotesingle Climate Change\textquotesingle} and a specific indicator
-\ikvd{\textquotesingle CO2 emissions (kt)\textquotesingle}, obtaining a time series with
-years as keys and emission data as values. The World Bank provider also supports filtering
-by year and indicator.
-
-Another type provider available in The Gamma that is based on the data cube structure allows
-the user to explore UK government expenditure:
+from the data cube. Choosing \ikvd{byCountry.\textquotesingle United States\textquotesingle},
+restricts the cube to a two-dimensional plane. We then choose an indicator category
+and a specific indicator \ikvd{\textquotesingle CO2 emissions (kt)\textquotesingle}, obtaining
+a time series with years as keys and emission data as values. Similarly, we could first filter the
+data by a year or an indicator. The same mechanism can be used for exploring the UK government
+expenditure:
 
@@
 
+\noindent
 The dimensions of the cube are government services, years and value type (adjusted, nominal,
-per GDP). Here, we select the \ikvd{Defence} service and \ikvd{GDP} value type. Our implementation
-does not yet use a standardized data cube storage and so adding another data cube source
+per GDP). Here, we select the \ikvd{Defence} service and \ikvd{GDP} value type.
+As there is no widely used standard format for data cubes, adding another data cube to The Gamma
 currently requires implementing a new type provider.
@@
 Our second type provider allows users to construct queries to explore data in tabular formats
-such as CSV files. It is based on the theory developed Petricek~\cite{dotdriven}. Unlike the  % grammar!
-data cube provider, the provider for tabular data does not just allow selecting a subset of the
-data, but it can be used to construct SQL-like query. Consider the example from Figure~\ref{fig:thegamma}:
+such as CSV files. Unlike the data cube provider, the provider for tabular data does not just
+allow selecting a subset of the data, but it can be used to construct SQL-like query. Consider
+the example from Figure~\ref{fig:thegamma}:
 
@@
 
+\noindent
 The example works with a CSV file that records individual medals awarded in Olympic games.
@@
 determine the operation type. Those are followed by operation parameters. For example, when grouping
-data, we first select the key and then add a number of aggregations to calculate over the group.
+data, we first select the key and then choose a number of aggregations to calculate over the group.
 Unlike SQL, the provider only allows users to choose from pre-defined aggregations such as
 calculating the sum, average or the number of distinct values. As illustrated in
-Section~\ref{sec:cases}, this still allows us to construct a wide range of practical queries.
+Section~\ref{sec:cases}, this is sufficient to construct a wide range of practical queries.
 
@@
   \vspace{0.5em}
-  \caption{The World Bank provider allows users to first choose
-    a country or a year and then an indicator.}
+  \caption{Exploring World Bank data using the data cube type provider, users
+    choose values from two dimensions to obtain a data series.}
   \label{fig:cubetp}
@@
   \includegraphics[scale=0.28]{figures/graphtp}
-  \caption{The user specifies a path through the data, possibly with
+  \caption{To query graph data, the user specifies a path through the data, possibly with
     placeholders to select multiple nodes.}
@@
 \subsubsection*{Graph Database Type Provider}
-Our third type provider allows users explore data from graph databases. A graph database
-consists of nodes representing entities and relationships between them. In the context of
-data journalism, graph databases have been used for example in The Panama Papers reporting \cite{panama}.
-
+Our third type provider allows users to explore data from graph databases, which store
+nodes representing entities and relationships between them. In the context of data journalism,
+graph databases have been used for example in The Panama Papers reporting \cite{panama}.
 The following example explores a database of Doctor Who characters and episodes. It retrieves
@@
 
+\noindent
 We start from the \ikvd{Doctor} node and then follow two relationships. We use
@@
 \ikvd{\textquotesingle APPEARED IN\textquotesingle.\textquotesingle Day of the Moon\textquotesingle}
-to only select only such enemies that appear in a specific episode. The resulting query
+to select only enemies that appear in a specific episode. The resulting query
 is illustrated in Figure~\ref{fig:graphtp}.
@@
 relationships. The results returned by the provider is a table of properties of all nodes
-along the specified path. As discussed in the next section, such table can be further queried
-using the tabular data type provider.
-
-\newpage
+along the specified path. As illustrated by an example discussed in Section~\ref{sec:cases},
+the returned table can be further queried using the tabular data type provider.
+
 % ==================================================================================================
 
-\section{Motivation}
-\label{sec:motivation}
-
-The Gamma aims to adapt the recent innovations in programming language research, especially the
-work on type providers, into a form where it could be used in practice by journalists and other
-non-expert interested in data exploration. We start with a careful consideration of our target
+\section{Design}
+\label{sec:design}
+
+\subsection{Journalism}
+
+We start with a careful consideration of our target
 application domain, i.e.~data analyses produced by journalists and citizen data scientists that
@@
 
-\subsection{Open Journalism}
+% Open Journalism
+
 Journalism continually develops and responds to the many challenges it faces \cite{future}.
@@
 
-\begin{figure}
-\includegraphics[width=1\columnwidth]{figures/nyt}
-\caption{New York Times article on Obama's legacy \cite{youdraw}. The article asks the reader to make a guess
-(engagement), but only lists ``Immigration and Customs Enforcement'' as a source of data.}
-\label{fig:nyt}
-\end{figure}
-
-\subsection{End-user Data Exploration}
+\subsection{Lowering barriers}
 Our aim is to make programmatic data exploration accessible to journalists, but we want to keep
@@
 
-\paragraph{Learning without Experts}
-Sarkar \cite{learning} reports that users learn how to use Excel either by talking to experts,
-or by seeing a feature in a spreadsheet received from a colleague. In our circumstances, experts
-are unlikely to be available, so the tool should support learning from examples. When looking at
-a work done and published by another person, the user needs to see (and be able to understand)
-how a task was completed.
-
-\begin{figure}
-\includegraphics[width=1\columnwidth]{figures/ft}
-\caption{Financial Times analysis of recyclable waste. Full source is provided as Jupyter Notebooks
-on GitHub \cite{ftnotebooks}, but re-running the analysis is difficult, even for an expert.}
-\label{fig:ft}
-\end{figure}
-
-% ==================================================================================================
-
-% ==================================================================================================
-
-\section{Design}
-\label{sec:design}
-
-We identified a number of requirements for a data exploration tool for journalists earlier.
-First, using the tool should result in transparent and reproducible data analyses that can be
-understood and modified by non-experts. Second, the tool should minimze the number of concepts
-that the user needs to understand and, subsequently, allow them to learn through exploration
-and from examples. In this section, we discuss how our design aims to satisfy those requirements.
-
-\subsection{Correct and complete}
-any program can be created by autocomplete, all programs created by autocomplete are correct
-
-\subsection{Lowering the Barrier to Entry}
 Data exploration tasks, such as querying tables, have a certain irreducible complexity.
@@
 \paragraph{Iterative Prompting}
-Iterative prompting is an interaction principle where the user repeatedly invokes an
-auto-complete prompt and makes a selection from the offered options. The principle is closely
-related to both code completion and the use of command line. Iterative prompting is novel as an
-overarching interaction principle for program construction.
+%Iterative prompting is an interaction principle where the user repeatedly invokes an
+%auto-complete prompt and makes a selection from the offered options. The principle is closely
+%related to both code completion and the use of command line. Iterative prompting is novel as an
+%overarching interaction principle for program construction.
 
@@
 
-\subsection{Building the Magic Escalator of Knowledge}
+\subsection{Learnign without experts}
+
+Sarkar \cite{learning} reports that users learn how to use Excel either by talking to experts,
+or by seeing a feature in a spreadsheet received from a colleague. In our circumstances, experts
+are unlikely to be available, so the tool should support learning from examples. When looking at
+a work done and published by another person, the user needs to see (and be able to understand)
+how a task was completed.
+
 As discussed earlier, The Gamma is designed for users who work under tight deadlines and only
@@
 option and choosing a different one is not apparent from the result.
+
+\subsection{Correct and complete}
+any program can be created by autocomplete, all programs created by autocomplete are correct
+
+% ==================================================================================================
+
+\newpage
+
+% \begin{figure}
+% \includegraphics[width=1\columnwidth]{figures/nyt}
+% \caption{New York Times article on Obama's legacy \cite{youdraw}. The article asks the reader to make a guess
+% (engagement), but only lists ``Immigration and Customs Enforcement'' as a source of data.}
+% \label{fig:nyt}
+% \end{figure}
+
+% \begin{figure}
+% \includegraphics[width=1\columnwidth]{figures/ft}
+% \caption{Financial Times analysis of recyclable waste. Full source is provided as Jupyter Notebooks
+% on GitHub \cite{ftnotebooks}, but re-running the analysis is difficult, even for an expert.}
+% \label{fig:ft}
+% \end{figure}
+
+% ==================================================================================================
+~
+\newpage
 

### 2020-09-15 wip-design/paper.tex
Finish design, tweaks in case studies

@@
 
-% TODO Study improvements
-% - The questionaire needs a bit more introduction when you present  the study.
-% - When looking at the data for the questionaire, it seems somewhat inconclusive
-%    as all the values lie around 2.5.  I think this warrants a bit more discussion.
-% - Connected to this, I would recommend to highlight the successes of the study a bit more in the
-%     discussion and  presentation of the results. You actually got non-programmers to do
-%     sophisticated queries! That you didn't get support for all RQs is perhaps more a result of being quite  ambitious!
-
 % TODO
@@
 
-% extend autocomplete into a full-blown interactive programming mechanism
-% and show that it is viable for small programming tasks like data exploration
-%
-% --- remove some of the motivation (makes ppl ask bad questions)
-
-%
-% Iterative prompting - should we call it expression completion
-% (is it more general result, or is it a new thing?)
-% -> emphasize the why this is called differently paragraph
-%
 % Explain type providers a bit more (we say what it does, but not much about what
@@
 \usepackage{xcolor}
-% \usepackage{tikz}
-% \newcommand*{\priority}[1]{\begin{tikzpicture}[scale=0.12]%
-%     \draw (0,0) circle (1);
-%     \fill[fill opacity=1,fill=black] (0,0) -- (90:1) arc (90:90-#1*3.6:1) -- cycle;
-%     \end{tikzpicture}}
-% \newcommand*{\priorityc}[1]{
-% \resizebox{0.8em}{0.8em}{
-%   {\protect\tikz{ \protect
-%     \draw[line width=1mm, black] (0,0) circle (1); \fill[fill opacity=1,fill=black] (0,0) -- (90:1) arc (90:90-#1*3.6:1) -- cycle;
-%     } }}}
+\usepackage{tikz}
+\newcommand*{\priority}[1]{\begin{tikzpicture}[scale=0.12]%
+   \draw (0,0) circle (1);
+   \fill[fill opacity=1,fill=black] (0,0) -- (90:1) arc (90:90-#1*3.6:1) -- cycle;
+   \end{tikzpicture}}
+\newcommand*{\priorityc}[1]{
+\resizebox{0.8em}{0.8em}{
+ {\protect\tikz{ \protect
+   \draw[line width=1mm, black] (0,0) circle (1); \fill[fill opacity=1,fill=black] (0,0) -- (90:1) arc (90:90-#1*3.6:1) -- cycle;
+   } }}}
 %
@@
 \begin{figure}
-\includegraphics[width=.6\columnwidth]{figures/thegamma-annot}
+\includegraphics[width=.58\columnwidth]{figures/thegamma-annot}
+\vspace{-0.5em}
 \caption{Obtaining teams with the greatest number of gold medals from Rio 2016
@@
 \label{fig:thegamma}
+\vspace{-0.5em}
 \end{figure}
@@
 
-% TODO REWRITE
-
-\item We show that iterative prompting is a complete and correct program construction method and
-  consider how it lowers the barrier to entry and allows learning through examples, without
-  expert guidance (Section~\ref{sec:design}).
+\item We reflect how our design addresses challenges faced by journalists (Section~\ref{sec:design-ddj})
+  and analyse the design theoretically. We argue that our design lowers barriers to entry (Section~\ref{sec:design-bar}),
+  supports learning without experts (Section~\ref{sec:design-expert}) and offers a
+  complete and correct program construction method (Section~\ref{sec:design-cc}).
 
@@
 Our second type provider allows users to construct queries to explore data in tabular formats
-such as CSV files. Unlike the data cube provider, the provider for tabular data does not just
+such as CSV files. A prominent example of tabular data used by journalists is, e.g.~the
+Iraq War documents leak~\cite{iraq}. Unlike the data cube provider, the provider for tabular data does not just
 allow selecting a subset of the data, but it can be used to construct SQL-like query. Consider
@@
 \caption{Design of type providers for exploring data cubes and graph databases.}
-\label{fig:three graphs}
+\label{fig:tps}
 \end{figure}
@@
 drwho.Character.Doctor
-  .'ENEMY OF'.'[any]'
-  .'APPEARED IN'.'Day of the Moon'
+  .'ENEMY OF'.'[any]'.'APPEARED IN'.'Day of the Moon'
 \end{thegamma}
@@
 
-\section{Design}
+\section{Reflections on Design}
 \label{sec:design}
 
-\subsection{Journalism}
-
-We start with a careful consideration of our target
-application domain, i.e.~data analyses produced by journalists and citizen data scientists that
-are published online. We look at both practical requirements for such programming environment
-and requirements arising from our focus on journalism. This analysis is based on the author's
-experience of collaborating with journalists\footnote{Citations removed to preserve anonymity.},
-review of literature on data journalism, e.g.~\cite{ddj,edcj17,edcj18} and more general trends in
-journalism.
-
-% Open Journalism
-
-Journalism continually develops and responds to the many challenges it faces \cite{future}.
-Two recent challenges are relevant to our work. The first is building trust in media.
-One way of establishing trust in the age of fake news is to be more transparent about editorial
-decisions, process and original sources.
-
-Many journalists believe that opening up the process
-shows the quality and trustworthiness of their work~\cite{transparency}.
-The second challenge is reader engagement. To develop a relationship with readers, journalists are
-increasingly looking for meaningful ways of engagement. This includes reader comments, involvement
-of citizen journalists \cite{comments,citizen} and the development of new interactive formats
-\cite{youdraw}. To address the above challenges, a tool for data exploration should satisfy the
-following three requirements.
-
-\paragraph{Trust Through Transparency}
-To support trustworthiness, data analyses should be transparent. The reader should be able to
-determine what is the source of analysed data and how has the data been transformed. As much as
-possible, these capabilities should also be accessible to non-expert readers.
-
-\paragraph{Reproducibility for Fact Checking}
-It should be possible to re-run the analysis to verify that it produces the presented results.
-However, running an opaque script is not enough. A reader should be able to recreate the analysis
-by following the necessary steps from the original data source to the end result.
-
-\paragraph{Encouraging Meaningful Engagement}
-The tool should support a mechanism through which readers can engage in a meaningful discussion.
-For example, it should allow modifying of parameters of a data visualization in order to show
-how different choices affect the final result.
-
-\subsection{Lowering barriers}
-Our aim is to make programmatic data exploration accessible to journalists, but we want to keep
-the desirable properties of text-based programming. In particular, source code of a data
-exploration should provide a full reproducible record of how the data analysis has been done.
-As end-users, journalists have a number of interesting characteristics. They work under tight
-deadline and data exploration is only a complementary skill. They also need to work with a wide
-range of data sources, including big data tables (e.g.~Iraq War documents leak) or graph
-databases (e.g.~Panama Papers). This leads to a number of practical requirements on the programming
-environment.
-
-\paragraph{Conceptual Simplicity}
-We target end-users who cannot dedicate much time to learning about a tool prior
-to using it. Consequently, using the tool should require understanding of only a small number
-of concepts. Once the user understand a small number of concepts, they should be able to complete
-basic data exploration tasks.
-
-\paragraph{Uniformity across Data Sources}
-The users should be able to navigate through large databases, query relational databases and
-query graph databases through the same mechanism. Ideally, expertise gained with one data source
-should also be transferable to working with another data source.
-
-Data exploration tasks, such as querying tables, have a certain irreducible complexity.
-Regardless of the interface, the user will be exposed to concepts such as filtering,
-sorting and grouping. Our design aims to lower the barrier to entry by stratifying the concepts
-into first-level \emph{iterative prompting} principle and second-level \emph{domain specific
-languages} for each kind of data source.
-The user needs to master the iterative prompting principle in order to start working
-with the system. However, using individual data sources can be mastered gradually.
-
-\paragraph{Iterative Prompting}
-%Iterative prompting is an interaction principle where the user repeatedly invokes an
-%auto-complete prompt and makes a selection from the offered options. The principle is closely
-%related to both code completion and the use of command line. Iterative prompting is novel as an
-%overarching interaction principle for program construction.
-
-In code completion, the auto-complete
-prompt is invoked only in certain contexts, e.g.~when accessing a member of an object through a
-variable defined earlier in code. It requires the user to be sufficiently familiar with the
-programming language in order to get to a point where they are offered a list of members. In
-contrast, our system only requires choosing the initial data source. The rest of the programming
-is done via iterative prompting. The iterative nature of iterative prompting makes it similar to
-using the command line or REPL (read-eval-print-loop) tools. However, those repeatedly ask the
-user to type code or commands.
-
-%https://www.nngroup.com/articles/recognition-and-recall/
-We argue that iterative prompting is easier than other forms of interaction, because it
-follows the \emph{recognition over recall} usability heuristic. The users are only required to
-choose from an offered list of options, rather than having to recall a possible command
-(to type in a command line) or a syntax in a text-based programming environment.
-
-\paragraph{Domain Specific Languages}
-The access to individual data sources in The Gamma is facilitated through a domain specific
-language, which defines the primitives that are offered to the user in the auto-complete prompts
-invoked through iterative prompting. The domain specific languages are embedded in The Gamma --
-they define merely the available members (and specify how to evaluate a chain of members), but
-they cannot define any custom syntax.
-
-We discuss the individual domain specific languages for querying data tables, graph databases and
-data cubes in a later section. The complexity of those languages differs. The previously discussed
-language for querying tabular data is the most complex one. However, The Gamma makes it easy for
-the user to start exploring and learning new languages, because they are all accessible via
-iterative prompting.
-
-\subsection{Learnign without experts}
-
-Sarkar \cite{learning} reports that users learn how to use Excel either by talking to experts,
-or by seeing a feature in a spreadsheet received from a colleague. In our circumstances, experts
-are unlikely to be available, so the tool should support learning from examples. When looking at
-a work done and published by another person, the user needs to see (and be able to understand)
-how a task was completed.
-
-As discussed earlier, The Gamma is designed for users who work under tight deadlines and only
-analyse data as their secondary task. The Gamma aims to support such users by having a low
-barrier for entry and making it easy to learn independently.
-
-\paragraph{Design for Percolation}
-When analysing how Excel users learn Sarkar \cite{learning} points out that users learn new
-features opportunistically when the usage of a feature is apparent in a spreadsheet. For
-example, users can learn different functions to use in formulas, because those are visible in
-the cell. Learning how to use a wizard for creating charts in this way is not possible because
-it leaves no trace in the spreadsheet -- only the final result. Sarkar's recommendation is to
-\emph{design for percolation}, i.e.~in a way where looking at the final result makes it apparent
-what feature has been used and how.
-
-\paragraph{Text-based Source Code}
-In The Gamma, each step in the iterative prompting process results in an identifier that is
+The design of The Gamma has been motivated by a curiosity as to whether iterative
+prompting can make text-based programming with data accessible to non-experts. In this section,
+we theoretically assess the resulting design and position it in the context of a possible application
+in the context of data journalism.
+
+\subsection{Data Journalism Perspective}
+\label{sec:design-ddj}
+
+Although our work targets non-expert data exploration in general, data journalism is an important
+specific domain with interesting design requirements. We reflect those on the basis of relevant
+literature, e.g.~\cite{ddj,edcj17,edcj18} and past experience of collaborating with
+journalists. Two challenges faced by journalists \cite{future} are particularly relevant to The Gamma.
+
+\subsubsection*{Trust Through Transparency}
+The first issue is trust in media. Many practitioners believe that transparency
+about the editorial process and information sources can serve as a proof of quality and
+trustworthiness~\cite{transparency}. This applies to data analyses too. For example, the Financial Times
+increasingly share source code of their analyses on GitHub, e.g.~\cite{ftnotebooks},
+but reproducing such analyses is difficult even for an expert. The Gamma has a potential to
+improve on the status quo in that a part of the analysis can be directly embedded in an article.
+The source code provides full account of what the data sources are and how are they used.
+The Gamma makes this code, to a greater extent, accessible to non-expert readers.
+
+\subsubsection*{Encouraging Meaningful Engagement}
+
+The second challenge is developing relationship with readers. Journalists are looking
+for meaningful ways of engagement through reader comments, involvement of citizen journalists
+\cite{comments,citizen} and the development of new interactive formats \cite{youdraw}.
+The Gamma can support this aim by providing a data exploration environment that can engage
+non-expert readers in a meaningful discussion. As argued below, non-experts can learn to use the
+system and can, for example, modify parameters or change filtering criteria to get a
+new perspective on a story.
+
+\subsection{Lowering Barriers to Entry}
+\label{sec:design-bar}
+
+Data exploration has a certain irreducible essential complexity. To make it accessible to users who
+cannot dedicate much time to learning a tool prior to using it, this complexity needs to be
+carefully stratified. The Gamma uses a two-level structure. At the first level, the user needs
+to learn iterative prompting to be able to start exploring data. At the second level, users will
+need to learn domain-specific languages defined by individual type providers.
+
+\subsubsection*{Iterative Prompting Principle}
+Iterative prompting is a suitable first level principle, because it is easy to trigger.
+In conventional programming languages, auto-complete assistance is only available once basic
+code structure is written. In contrast, user of In The Gamma only needs to choose the initial
+data source. Iterative prompting is also easier to use than e.g.~a command line or a REPL
+(read-eval-print-loop), because it follows the \emph{recognition over recall} usability heuristic.
+The users are not required to recall and type a command. They merely need to select one from a
+list of options.
+
+\subsubsection*{Stratifying Data Exploration Complexity}
+At the first level, any data source is accessed through the same mechanism. At the second level,
+each data source defines its own domain specific language, consisting of the primitives that are
+offered to the user in auto-complete prompts. The domain specific languages are embedded in The Gamma
+language -- they define merely the available members and cannot extend the core language syntax.
+Although the complexity of individual languages differs, the users can always start exploring
+and learning new languages using the familiar first level iterative prompting principle.
+An important question, which we study in Section~\ref{sec:study}, is whether the expertise gained
+with one data source be transferred to working with another data source.
+
+\subsection{Learning without Experts}
+\label{sec:design-expert}
+
+A typical user of The Gamma will not dedicate significant time to learning it in advance and
+they will not have access to experts. Most learning thus needs to happen from examples.
+When analysing how Excel users learn, Sarkar \cite{learning} points out that users learn new
+features when the usage of a feature is apparent in a spreadsheet. For example, users can learn
+different functions in formulas, because those are visible in the cell. Learning how to use a
+wizard for creating charts is not possible because the operation leaves no full trace in the
+spreadsheet. Sarkar's recommendation is to \emph{design for percolation},
+i.e.~in a way where looking at the final result makes it apparent what feature has been used and how.
+
+In The Gamma, each step of iterative prompting results in an identifier that is
 added to the source code. This means that a program constructed solely through iterative prompting
-keeps a full trace of how it was created. Seeing the resulting source code provides the user all
+keeps a full trace of how it was created. The resulting source code provides the user all
 information that they need to recreate the program, not just by copying it, but also by using the
-iterative prompting mechanism. The Gamma represents code as text and allows the user to edit it
-freely, so not all interactions leave a trace. For example, deleting the most recently selected
-option and choosing a different one is not apparent from the result.
-
-\subsection{Correct and complete}
-any program can be created by autocomplete, all programs created by autocomplete are correct
+iterative prompting mechanism. We assess the viability of this way of learning in the study in
+Section~\ref{sec:study}.
+
+\subsection{Correctness and Completeness}
+\label{sec:design-cc}
+
+An important characteristic of our design is that, barring a few exceptions discussed below, the iterative
+prompting mechanism is both correct and complete with respect to possible data exploration scripts.
+This means that (i) auto-complete lists offered by iterative prompting contain
+only options that are valid in a given context and that (ii) any script that can be written
+in The Gamma can be contructed via iterative prompting.
+
+\subsubsection*{Correctness} The Gamma uses an object-based type system for error-checking and for
+generating auto-complete lists. A well-typed script can always be executed. When using iterative
+prompting, a selected option, which is a valid object member, is added to the end of a script,
+leading to a well-typed script. This distinguishes our system from auto-completion based on machine
+learning or dictionary-based methods, which may offer members not valid in a given context. Advanced
+users of The Gamma, modifying scripts as text, can still violate correctness -- for example, if the
+user uses auto-complete to replace a member in a middle of a member access chain with a member of another type.
+
+\subsubsection*{Completeness} Auto-completion lists offered via iterative prompting contain
+all available members and so the user can construct all possible scripts. The range of such scripts
+is determined by type providers, which support limited set of options (e.g. group aggregations in
+the tabular data type provider). Two exceptions to completeness in our current design is specifying
+numerical parameters as in \ikvd{take(5)} and advanced language features such as let binding.
+We believe that alleviating the need for those poses an interesting further research challenge.
+
+\begin{figure}[b]
+\centering
+\begin{subfigure}[b]{0.49\textwidth}
+  \centering
+  \includegraphics[width=1\columnwidth]{figures/cases-mp}
+  \caption{Exploring Olympic medalists using tabular data type provider}
+  \label{fig:cases-mp}
+\end{subfigure}
+\hfill
+\begin{subfigure}[b]{0.49\textwidth}
+  \centering
+  \includegraphics[width=1\columnwidth]{figures/cases-dr}
+  \caption{Exploring Dr Who graph database by composing type providers}
+  \label{fig:cases-dr}
+\end{subfigure}
+\vspace{-0.5em}
+\caption{Charts produced by two case studies of using The Gamma.}
+\label{fig:cases}
+\end{figure}
 
@@
 
-\newpage
-
-% \begin{figure}
-% \includegraphics[width=1\columnwidth]{figures/nyt}
-% \caption{New York Times article on Obama's legacy \cite{youdraw}. The article asks the reader to make a guess
-% (engagement), but only lists ``Immigration and Customs Enforcement'' as a source of data.}
-% \label{fig:nyt}
-% \end{figure}
-
-% \begin{figure}
-% \includegraphics[width=1\columnwidth]{figures/ft}
-% \caption{Financial Times analysis of recyclable waste. Full source is provided as Jupyter Notebooks
-% on GitHub \cite{ftnotebooks}, but re-running the analysis is difficult, even for an expert.}
-% \label{fig:ft}
-% \end{figure}
-
-% ==================================================================================================
-~
-\newpage
-
-% ==================================================================================================
-
-\begin{figure}
-\includegraphics[width=1\columnwidth]{figures/sidebyside}
-\caption{The Gamma environment with an automatically generated chart comparing CO2 emissions
-of China and United States.}
-\label{fig:sidebyside}
-\end{figure}
-
-% ==================================================================================================
-
-\section{Use cases}
+\section{Case Studies}
 \label{sec:cases}
@@
 the UK government expenditure, activities of a research institute, Olympic medals and
-information about the Doctor Who series\footnote{Links removed to preserve anonymity.}.
-In this section, we consider two interesting larger examples.
-
-\subsection{If Michael Phelps were a Country}
-Michael Phelps has won so many medals that media compared him to
-countries.\footnote{\href{https://www.npr.org/sections/thetorch/2016/08/14/489832779/}{\small\bf\ttfamily npr.org/sections/thetorch/2016/08/14/489832779/}}
-Using the tabular data provider, we construct a chart shown in Figure~\ref{fig:cases} that
-produces a country league table including Michael Phelps:\footnote{Available at: \url{http://gallery.thegamma.net/86/}}
+information about the Doctor Who series\footnote{The analyses are available (non-anonymously)
+at \url{http://gallery.thegamma.net}, \url{http://turing.thegamma.net} and \url{http://rio2016.thegamma.net}}.
+In this section, we consider two interesting larger examples\footnote{Available
+(non-anonymously) at: \url{http://gallery.thegamma.net/86/} and  \url{http://gallery.thegamma.net/87/}, respectively.}.
+
+\subsubsection*{If Michael Phelps were a Country}
+Michael Phelps has won so many medals that media compared the number to
+countries.\footnote{e.g.~``If Michael Phelps Were A Country, Where Would His Gold Medal Tally Rank?''
+ \url{https://www.npr.org/sections/thetorch/2016/08/14/489832779/}}
+Media illustrated the story with a chart showing a country league table including Michael Phelps.
+We reproduce the chart, shown in Figure~\ref{fig:cases-mp}, using the tabular data type provider:
 
 \begin{thegamma}
-\kvd{let} data = olympics
-    .'group data'.'by Team'.'sum Gold'.then
+\kvd{let} data = olympics.'group data'.'by Team'.'sum Gold'.then
     .'sort data'.'by Gold descending'.then
-    .paging.skip(43).take(4)
-    .'get series'.'with key Team'.'and value Gold'
-
-\kvd{let} phelps = olympics
-  .'filter data'.'Athlete is'.'Michael Phelps'.then
+    .paging.skip(43).take(4).'get series'.'with key Team'.'and value Gold'
+
+\kvd{let} phelps = olympics.'filter data'.'Athlete is'.'Michael Phelps'.then
   .'group data'.'by Athlete'.'sum Gold'.then
@@
 
-charts.bar(data.append(phelps))
-  .setColors(["#aec7e8","#aec7e8","#1f77b4"])
+charts.bar(data.append(phelps)).setColors(["#aec7e8","#aec7e8","#1f77b4"])
 \end{thegamma}
 
-The core of the data analysis is done in two commands. The first counts gold medals by countries
+\noindent
+The data analysis is done in two commands. The first counts gold medals by countries
 and uses paging to fetch 4 countries with suitable number of medals. In the second, we abuse the
@@
 to local variables (for readability) and passed to the \ikvd{chart.columns} function.
-In this case study, the core data exploration can be constructed via iterative prompting, but
-building a chart requires invoking the \ikvd{append} operation and specifying colors using a list.
-
-\subsection{The Most Frequent Doctor Who Villains}
-In the second case study, we list Doctor Who villains by the number of episodes in which they
-appear.\footnote{Available at: \url{http://gallery.thegamma.net/87/}} This is interesting as it
+The case study illustrates when more advanced language features are necessary. The data exploration
+itself has been completed via iterative prompting, but producing the final chart currently requires
+some manual programming. In practice, this would likely be done with the help of an expert
+or by copying code from an example.
+
+\subsubsection*{The Most Frequent Doctor Who Villains}
+In the second case study, use a graph database on the Dr Who series and list Dr Who villains
+by the number of episodes in which they appear. This case study is interesting as it
 combines the graph database provider for fetching the data with the tabular data provider for
@@
 \begin{thegamma}
-\kvd{let} topEnemies = drWho.Character.Doctor
-  .'ENEMY OF'.'[any]'.'APPEARED IN'.'[any]'.explore
-  .'group data'.'by Character name'
-    .'count distinct Episode name'.then
+drWho.Character.Doctor.'ENEMY OF'.'[any]'.'APPEARED IN'.'[any]'.explore
+  .'group data'.'by Character name'.'count distinct Episode name'.then
   .'sort data'.'by Episode name descending'.then
-  .paging.take(8).'get series'
-    .'with key Character name'.'and value Episode name'
-
-chart.bar(topEnemies)
+  .paging.take(8).'get series'.'with key Character name'.'and value Episode name'
 \end{thegamma}
 
-Lines 1-2 use the graph database provider to find all paths linking the Doctor node with any
+\noindent
+Line 1 use the graph database provider to find all paths linking the Doctor node with any
 character linked via \ikvd{ENEMY OF}, followed by any episode linked by \ikvd{APPEARED IN}.
@@
 \ikvd{explore} member. For each character (the villain) we count the total number of
-distinct episodes in the table. The result is shown in Figure~\ref{fig:cases}.
-
-The logic is not trivial, but it performs a fairly sophisticated data analysis that involves a
-graph database query followed by an SQL-like data aggregation. The code can be constructed
-using iterative prompting (with the exception of the numbers in paging). It can also be constructed
-gradually and instantaneous prviews support the user while doing so.
-
-\begin{figure}
-\centering
-\includegraphics[width=1\columnwidth]{figures/cases}
-\caption{Charts produced by two The Gamma case studies.}
-\label{fig:cases}
-\end{figure}
+distinct episodes in the table. The result is shown in Figure~\ref{fig:cases-dr}.
+The example performs a fairly sophisticated data analysis that involves a graph database query
+followed by an SQL-like data aggregation. The code can be fully constructed using iterative
+prompting (with the exception of the numbers in paging). Another important feature is that
+instantaneous previews support the user during the construction, showing the intermediate result
+at each step of the construction.
 
@@
 
-\section{User study}
+% TODO Study improvements
+% - The questionaire needs a bit more introduction when you present  the study.
+% - When looking at the data for the questionaire, it seems somewhat inconclusive
+%    as all the values lie around 2.5.  I think this warrants a bit more discussion.
+% - Connected to this, I would recommend to highlight the successes of the study a bit more in the
+%     discussion and  presentation of the results. You actually got non-programmers to do
+%     sophisticated queries! That you didn't get support for all RQs is perhaps more a result of being quite  ambitious!
+
+\section{User Study}
 \label{sec:study}
 
-Our goal is to develop an easy-to-learn tool that journalists and other non-programmers can use
+Our goal was to develop an easy-to-learn tool that non-programmers can use
 for producing transparent data analyses. To evaluate the extent to which The Gamma achieves this,
@@
 
-\paragraph{RQ1: Can non-programmers explore data with The Gamma?}
+\vspace{0.5em}\noindent\emph{RQ1: Can non-programmers explore data with The Gamma?}
 We give participants one of four simple data exploration tasks and see if they can complete the
@@
 
-\paragraph{RQ2: Can knowledge transfer between data sources?}
+\vspace{0.5em}\noindent\emph{RQ2: Can knowledge transfer between data sources?}
 Our first hypothesis is that users familiar with the iterative prompting principle will be able to
@@
 
-\paragraph{RQ3: Can users learn from just code samples?}
+\vspace{0.5em}\noindent\emph{RQ3: Can users learn from just code samples?}
 Our second hypothesis is that users can learn through percolation, i.e.~by looking at the source
@@
 We recruit 13 participants (5 male, 8 female) from a business team
-of a research lab working in various non-technical roles (project management,
-partnerships, communications) including one former journalist.
-
-We split participants into 4 groups. We first gave a brief 5 minute demonstration of The
-Gamma and then asked participants to complete a task. Finally, we conducted a short semi-structured
-group interview and later sent participants a follow-up questionnaire. The four tasks were:
+of a research lab working in non-technical roles (project management,
+partnerships, communications) including one former journalist. Only one participant (\#12)
+had prior programming experience.
+
+We split participants into 4 groups. We first gave a brief 5 minute overview of The
+Gamma and then asked participants to complete a task.
+ The four tasks were:
 
@@
   on ``Public order and safety'' and ``Defence'' in terms of GDP.
-\item \emph{Lodrs.} Participants were given a demo using \emph{worldbank}.
+\item \emph{Lords.} Participants were given a demo using \emph{worldbank}.
   They were asked to use the \emph{lords} data source (a table with House of Lords
@@
 
+Finally, we conducted a short semi-structured
+group interview and later sent participants a follow-up questionnaire.
+
 We let participants work independently, but offered guidance if they got stuck for longer period of
@@
 
-\begin{table}
+\begin{figure}
+  \begin{subfigure}[b]{0.49\textwidth}
     \centering
@@
       \midrule
-    \small \#1 & \small expenditure & \small cube & \priority{50} & {\small Obtained one of two data series}\\
-    \small \#2 & \small expenditure & \small cube & \priority{100} & {\small Explored furhter data series }\\
-    \small \#3 & \small expenditure & \small cube & \priority{100}& {\small Explored further data series }\\
-    \small \#4 & \small expenditure & \small cube & \priority{75}& {\small With hint to use another member }\\
-    \small \#5 & \small expenditure & \small cube & \priority{100}& {\small Explored further data series }\\
+      \small \#1 & \small expenditure & \small cube & \priority{50} & {\small Obtained one data series}\\
+      \small \#2 & \small expenditure & \small cube & \priority{100} & {\small Explored furhter data }\\
+      \small \#3 & \small expenditure & \small cube & \priority{100}& {\small Explored further data  }\\
+      \small \#4 & \small expenditure & \small cube & \priority{75}& {\small Correct member hint }\\
+      \small \#5 & \small expenditure & \small cube & \priority{100}& {\small Explored further data }\\
       \small \#6 & \small worldbank & \small cube & \priority{75} & {\small With general syntax hint }\\
@@
       \small \#8 & \small worldbank & \small cube & \priority{100} & {\small Extra time to find data }\\
-    \small \#9 & \small lords & \small table & \priority{75} & {\small  Struggled with composition}\\
+      \small \#9 & \small lords & \small table & \priority{75} & {\small  Composition issues}\\
       \small \#10 & \small lords & \small table & \priority{100} & {\small Completed very quickly }\\
-    \small \#11 & \small lords & \small table & \priority{75} & {\small With a hint to avoid operations}\\
-    \small \#12 & \small olympics & \small table & \priority{75}  & {\small With a hint to avoid operations}\\
-    \small \#13 & \small olympics & \small table & \priority{75}  & {\small Hints about `then' and operations}\\
+      \small \#11 & \small lords & \small table & \priority{75} & {\small Hint to avoid operations}\\
+      \small \#12 & \small olympics & \small table & \priority{75}  & {\small Hint to avoid operations}\\
+      \small \#13 & \small olympics & \small table & \priority{75}  & {\small Two operations hints}\\
       \bottomrule
     \end{tabular}
-  \caption{Overview of the work completed by individual participants. The marks denote:
+    \caption{Work completed by individual participants. The marks denote:
       \priorityc{75} = required some guidance, \priorityc{50} = partially completed.}
     \label{tab:tasks}
-  \vspace{-1em}
-\end{table}
+  \end{subfigure}
+  \hfill
+  \begin{subfigure}[b]{0.49\textwidth}
+    \centering
+    \begin{tabular}{p{17.25em} c c}
+      \toprule
+        {\small \textit{Questionnaire item}} & {\small \textit{Avg}} & {\small \textit{Sdv}} \\
+      \midrule
+      \small Considering your experience and possible use of The Gamma in the context of data journalism:\\
+      \small \quad I found the system easy to use. & \small 3.89 & \small 0.87\\
+      \small \quad Journalists will be able to use it to analyse data. & \small 3.33 & \small 1.05\\
+      \small \quad Readers will be able to examine the analyses. & \small 3.56 & \small 0.68\\
+      \small Thinking about the usability of the system:\\
+      \small \quad Resulting programs are easy to understand. & \small 3.67 & \small 0.82\\
+      \small \quad Text editor for creating them was easy to use. & \small 3.67 & \small 0.94\\
+      \small \quad Finding items in the auto-complete was easy. & \small 3.44 & \small 1.42\\
+      \small I would know how to explore a new data source\\
+      \small \quad If given a comprehensive video tutorial. & \small 4.00 & \small 1.24\\
+      \small \quad If given a number of code samples. & \small 3.22 & \small 1.13\\
+      \small \quad Without any further guidance. & \small 2.44 & \small 1.17 \vspace{0.1em}\\
+      \bottomrule
+    \end{tabular}
+    \caption{Follow-up questionnaire responses using
+      a 5-point Likert scale (1 = strongly disagree, 5 = strongly agree) over 9 subjects.}
+    \label{tab:quest}
+  \end{subfigure}
+  \caption{Summary of the work completed by individual participants and responses in a follow-up questionnaire}
+  \label{tab:tabs}
+\end{figure}
 
@@
 follow-up questionnaire.
+
+x
+
+y
+
+z
+
+a
+
+b
+
+c
+
+d
 
@@
 e.g.~participant \#5, who used copy-and-paste to fetch the same data series for multiple countries.
-
-\begin{table}
-  \centering
-  \begin{tabular}{p{19em} c c}
-    \toprule
-      {\small \textit{Questionnaire item}} & {\small \textit{Avg}} & {\small \textit{Sdv}} \\
-    \midrule
-    \small Considering your experience \& possible use of The Gamma:\\
-    \small \quad I found the system easy to use. & \small 2.11 & \small 0.87\\
-    \small \quad Journalists will be able to use it to analyse data. & \small 2.67 & \small 1.05\\
-    \small \quad Readers will be able to critically examine the analyses. & \small 2.44 & \small 0.68\\
-    \small Thinking about the usability of the system:\\
-    \small \quad The resulting programs are easy to understand. & \small 2.33 & \small 0.82\\
-    \small \quad The text editor for creating them was easy to use. & \small 2.33 & \small 0.94\\
-    \small \quad Finding the right item in the auto-complete was easy. & \small 2.56 & \small 1.42\\
-    \small I would know how to approach a task using a new data source\\
-    \small \quad If given a comprehensive video tutorial. & \small 2.00 & \small 1.24\\
-    \small \quad If given a number of code samples. & \small 2.78 & \small 1.13\\
-    \small \quad Without any further guidance. & \small 3.56 & \small 1.17\\
-    \bottomrule
-  \end{tabular}
-  \caption{Summary of follow-up questionnaire responses using
-    a 5-point Likert scale (1 = strongly agree, 5 = strongly disagree) over 9 subjects.}
-  \label{tab:quest}
-\end{table}
 
@@
 \section{Conclusions}
+[TODO: THIS IS A SHORT PAPER.... no design probe]
+
 We present The Gamma, a simple data exploration environment. The Gamma is motivated by the needs

### 2020-09-15 wip-design/paper.tex
Study and discussion improvements

@@
 
-We contribut The Gamma, a text-based data exploration environment for non-experts. The Gamma
+We contribute The Gamma, a text-based data exploration environment for non-experts. The Gamma
 is based on a single, easy to understand interaction principle and provides a uniform
@@
 Most similar to our approach are tools that recommend scripts when users begin interacting
-with data. Those based on machine learning code completion for domain-specific languages \cite{predictive,proactive}
+with data. Those based on machine learning code completion for domain specific languages \cite{predictive,proactive}
 differ in that they do not guarantee completeness, i.e.~the user cannot create all possible
@@
 The Gamma can be extended to support any kind of data source by implementing a \emph{type provider}.
-Conceptually, a type provider defines a domain-specific language for exploring data of a particular
+Conceptually, a type provider defines a domain specific language for exploring data of a particular
 kind. Techincally, it generates object types with members (such as \ikvd{sort by} or \ikvd{Count is})
@@
 to learn iterative prompting to be able to start exploring data. At the second level, users will
-need to learn domain-specific languages defined by individual type providers.
+need to learn domain specific languages defined by individual type providers.
 
@@
 for producing transparent data analyses. To evaluate the extent to which The Gamma achieves this,
-we conduct a user study. We recruit participants among the operations and business team of a
+we conducted a user study. We recruited participants among the operations and business team of a
 non-profit research organization and investigate three research questions.
 
-\vspace{0.5em}\noindent\emph{RQ1: Can non-programmers explore data with The Gamma?}
-We give participants one of four simple data exploration tasks and see if they can complete the
-task using The Gamma and, possibly, how much assistance they need.
-
-\vspace{0.5em}\noindent\emph{RQ2: Can knowledge transfer between data sources?}
-Our first hypothesis is that users familiar with the iterative prompting principle will be able to
-use an unfamiliar data source. In two of the tasks, participants are shown one data source and
-then asked to work with a different one.
-
-\vspace{0.5em}\noindent\emph{RQ3: Can users learn from just code samples?}
-Our second hypothesis is that users can learn through percolation, i.e.~by looking at the source
-code of published analyses. In one of our tasks, participants are given a code sample, but only a
-minimal explanation of The Gamma environment.
-
-\subsection{Study design}
-We perform between-subjects study to evaluate the first experience of using The Gamma.
-We recruit 13 participants (5 male, 8 female) from a business team
+\vspace{0.5em}\noindent\emph{RQ1: Can non-programmers use The Gamma?}\hspace{0.3em}
+Our first hypothesis is that non-programmers will be able to use The Gamma to explore data. To test
+this, we gave participants one of four simple data exploration tasks and assess whether they were
+able to complete the task, as well as how much assistance, if any, they needed.
+
+\vspace{0.5em}\noindent\emph{RQ2: Can knowledge transfer between data sources?}\hspace{0.3em}
+Our second hypothesis is that users familiar with the iterative prompting principle will be able to
+use an unfamiliar data source. We designed two of the four tasks to test this. Participants were
+shown a demo using one data source and then asked to complete a task using a different one.
+
+\vspace{0.5em}\noindent\emph{RQ3: Can users learn from just code samples?}\hspace{0.3em}
+Our third hypothesis is that users can learn through percolation, i.e.~by looking at the source
+code of published analyses. In one of our tasks, participants were given a minimal explanation
+of The Gamma (showing how to initiate the iterative prompting process) together with
+an extensive code sample.
+
+\subsection{Study Design}
+We performed a between-subjects study to evaluate the first experience of using The Gamma.
+We recruited 13 participants (5 male, 8 female) from a business team
 of a research lab working in non-technical roles (project management,
 partnerships, communications) including one former journalist. Only one participant (\#12)
-had prior programming experience.
-
-We split participants into 4 groups. We first gave a brief 5 minute overview of The
-Gamma and then asked participants to complete a task.
- The four tasks were:
+had prior programming experience. We split participants into 4 groups. We gave participants a
+brief overview of The Gamma (with content depending on the task) and then asked participants
+to complete a task. The four tasks were:
 
@@
   members expenses) to find a members representing London with the highest travel costs.
-\item \emph{Worldbank.} Participants were given a minimal explanation of The Gamma environment and
+\item \emph{Worldbank.} Participants were given a minimal demo of iterative prompting and
   a code sample using \emph{worldbank}. They were asked to solve a different task using
@@
 \item \emph{Olympics.} Participnts were given a demo using \emph{olympics}.
-  They were asked to solve a more complex problem using the same data source.
+  They were asked to solve a more complex problem, involving grouping and aggregation,
+  using the same data source.
 \end{itemize}
 
-Finally, we conducted a short semi-structured
-group interview and later sent participants a follow-up questionnaire.
-
-We let participants work independently, but offered guidance if they got stuck for longer period of
-time. Tasks \emph{expenditure} and \emph{lords} aim to answer the questions RQ1 and RQ2; the task
+\noindent
+We let participants work independently, but offered guidance if they got stuck.
+Tasks \emph{expenditure} and \emph{lords} aim to answer the questions RQ1 and RQ2; the task
 \emph{worldbank} aims to answer RQ1 and RQ3. In \emph{olympics}, we test RQ1 using a more complex
-data source and we ask further questions to explore their understanding of the data source.
+data source and we ask further questions to explore the understanding of the data source. Following the
+experiment, we conducted a short semi-structured group interview and later sent participants a follow-up
+questionnaire.
 
@@
     \end{tabular}
-    \caption{Work completed by individual participants. The marks denote:
-      \priorityc{75} = required some guidance, \priorityc{50} = partially completed.}
+    \caption{Work completed by individual participants; \priorityc{100} = completed,
+      \priorityc{75} = required some guidance, \priorityc{50} = partially completed}
     \label{tab:tasks}
@@
       \small Considering your experience and possible use of The Gamma in the context of data journalism:\\
-      \small \quad I found the system easy to use. & \small 3.89 & \small 0.87\\
-      \small \quad Journalists will be able to use it to analyse data. & \small 3.33 & \small 1.05\\
-      \small \quad Readers will be able to examine the analyses. & \small 3.56 & \small 0.68\\
+      \small \quad I found the system easy to use. & \small 4.00 & \small 0.89\\
+      \small \quad Journalists will be able to use it to analyse data. & \small 3.45 & \small 1.04\\
+      \small \quad Readers will be able to examine the analyses. & \small 3.73 & \small 0.79\\
       \small Thinking about the usability of the system:\\
-      \small \quad Resulting programs are easy to understand. & \small 3.67 & \small 0.82\\
-      \small \quad Text editor for creating them was easy to use. & \small 3.67 & \small 0.94\\
-      \small \quad Finding items in the auto-complete was easy. & \small 3.44 & \small 1.42\\
+      \small \quad Resulting programs are easy to understand. & \small 3.73 & \small 0.79\\
+      \small \quad Text editor for creating them was easy to use. & \small 3.73 & \small 0.90\\
+      \small \quad Finding items in the auto-complete was easy. & \small 3.45 & \small 1.37\\
       \small I would know how to explore a new data source\\
-      \small \quad If given a comprehensive video tutorial. & \small 4.00 & \small 1.24\\
-      \small \quad If given a number of code samples. & \small 3.22 & \small 1.13\\
-      \small \quad Without any further guidance. & \small 2.44 & \small 1.17 \vspace{0.1em}\\
+      \small \quad If given a comprehensive video tutorial. & \small 4.00 & \small 1.26\\
+      \small \quad If given a number of code samples. & \small 3.36 & \small 1.12\\
+      \small \quad Without any further guidance. & \small 2.45 & \small 1.13 \vspace{0.1em}\\
       \bottomrule
@@
     \caption{Follow-up questionnaire responses using
-      a 5-point Likert scale (1 = strongly disagree, 5 = strongly agree) over 9 subjects.}
+      a 5-point Likert scale (1 = strongly disagree, 5 = strongly agree) over 11 subjects.}
     \label{tab:quest}
   \end{subfigure}
-  \caption{Summary of the work completed by individual participants and responses in a follow-up questionnaire}
+  \vspace{-0.5em}
+  \caption{Summary of the work completed by individual participants and responses to the follow-up questionnaire.}
   \label{tab:tabs}
+  \vspace{-0.25em}
 \end{figure}
 
-\subsection{Results}
-Table~\ref{tab:tasks} summarizes the work done by the study participants.
+\subsection{Study Results}
+Figure~\ref{tab:tabs} shows the results of the study. Most notably, all participants were able to
+complete, at least partially, a non-trivial data exploration task and only half of them required
+further guidance. Table~\ref{tab:tasks} summarizes the work done by the study participants.
 For each participant, we record the task, the kind of data source used in the task and the
 level of completion. For participants who needed assistance, the notes section details the help
-given. We analyse the hints in the next section. Table~\ref{tab:quest} shows the results of the
-follow-up questionnaire.
-
-x
-
-y
-
-z
-
-a
-
-b
-
-c
-
-d
-
-\paragraph{RQ1: Can non-programmers explore data with The Gamma?}
-Three results allow us to answer RQ1 in the affirmative. First, 7 out of 9
-participants agree or strongly agree that they ``found the system easy to use''. Second,
-participants spent between 10 and 25 minutes (average 17 minutes) working with The Gamma and
-all 12 out of 13 participants completed the task; 6 participants required some assistance,
-but 3 of those faced one issue (discussed later) that could be covered in the introductory
-presentation. Third, a number of participants shared positive comments in the semi-structured
-interview.
-
+given. We discuss possible design improvements based on the experience in Section~\ref{sec:study-obs}.
+
+The follow-up questionnaire was sent to all participants a week after the experiment. It was
+completed all but two participants (\#8 and \#12). We asked general questions about the usability
+of the system together with questions about learning designed to answer RQ2 and RQ3. A summary
+of the answers is shown in Table~\ref{tab:quest}.
+
+\subsection*{RQ1: Can non-programmers explore data with The Gamma?}
+Three facts allow us to answer RQ1 in the affirmative. First, 9 out of 11 participants agree or
+strongly agree that they ``found the system easy to use''. Second, participants spent 10--25 minutes
+(average 17 minutes) working with The Gamma and 12 out of 13 completed the task; 6 required assistance,
+but 3 of those faced one issue (discussed later) that could be addressed in the introduction.
+Third, a number of participants shared positive comments in the interviews.
 Participant \#3 found the system simple, but also points out an issue about data provenance,
-which we discuss later:
+which we revisit later:
 
@@
 
+\noindent
 Similarly, participant \#2, notes that The Gamma alleviated their unease about code:
@@
 
+\noindent
 Finally, participant \#5 suggested the system could be used as an educational tool for teaching
@@
 
-\paragraph{RQ2: Can knowledge transfer between data sources?}
+\subsection*{RQ2: Can knowledge transfer between data sources?}
 Our study does not conclusively answer RQ2. There is some evidence in favor of a positive answer.
 In the practical part, two of the tasks (\emph{expenditure} and \emph{lords}) used a different data
-source in the introductory presentation than the one that the participants were asked to use.
+source in the introductory presentation than the one that the participants were asked to explore.
 Participants were able to complete those tasks, although \emph{lords} has been more challenging
@@
 \emph{``I found it quite easy to translate what you showed us in the demo to the new dataset.
-   I though it was quite easy to just get how that works.''}
+   I thought it was quite easy to just get how that works.''}
 \end{quote}
 
+\noindent
 Negative evidence is offered by the follow-up questionnaire. When asked
-whether they would know how to approach a task using a new data source, 5 out of 9 participants
-diagree or strongly disagree that they would know how to approach it without any further guidance.
-Participants did not believe that knowledge can easily transfer to another
-data source.
-
-\paragraph{RQ3: Can users learn from just code samples?}
+whether they would know how to approach a task using a new data source, 6 out of 11 participants
+disagree or strongly disagree that they would know how to approach it without any further guidance.
+The nature of RQ2 makes it a more challenging question to study and so finding a conclusive answer
+arguably requires further research.
+
+\subsection*{RQ3: Can users learn from just code samples?}
 A positive answer to RQ3 is supported by the \emph{worldbank} task results, the follow-up questionnaire
@@
 All three were able to complete a related task using the same data source. In the follow-up
-questionnaire, only 2 out of 9 participants disagree or strongly disagree that
+questionnaire, only 2 out of 11 participants disagree or strongly disagree that
 they would know how to approach a task using an unfamiliar data source when given ``a number
@@
   \emph{``I think providing one video of one example is good and maybe a couple of little examples
-  of code wher people can see the kind of things you can do.''}
+  of code where people can see the kind of things you can do.''}
 \end{quote}
 
+\noindent
 This is aligned with our design goal. Once the user understands the iterative prompting principle
@@
 
-\subsection{Further observations}
+\subsection{Further Observations}
+\label{sec:study-obs}
 In this section, we briefly discuss a number of observations about The Gamma design that
-emerged from the study, some of which suggest ways of improving the system.
-
-\paragraph{Making complex things possible may hurt}
-As discussed earlier, The Gamma language supports operations such as \ikvd{take(5)}. Most type
+emerged from the experiments and follow-up interviews, some of which suggest ways of improving the system.
+
+\subsubsection*{Making Complex Things Possible May Hurt}
+The Gamma supports operations such as \ikvd{take(5)}. Most type
 providers never generate those, but the provider for working with tabular data is an exception.
@@
 Three participants (\#11, \#12, \#13) struggled to complete a task, because they
-initially attempted to use those operations. However, those violate the iterative prompting
+initially attempted to use those operations. They violate the iterative prompting
 principle as one cannot type `.' after \ikvd{\textquotesingle Year is greater than\textquotesingle}.
-This suggests that we should avoid operations in type providers for data access, even if it
-limits the expressive power of the type provider.
-
-\paragraph{Benefits and drawbacks of text}
-The Gamma is based on text to aid transparency. Text hints that there is no hidden state and
+This suggests that we should either avoid such operations, or hide them under an ``advanced operations''
+tab as a caution to novice users.
+
+\subsubsection*{Benefits and Drawbacks of Text}
+The Gamma is based on text to aid transparency. Text implies that there is no hidden state and
 the reader sees the full code. The study suggests that using a text editor has both
@@
 
-We observed two issues in the study. Participant \#2 struggled with correct indentation, starting a second command
-with more indentation than needed and participant \#6 had a syntax error in an unrelated command,
-which prevents charts from rendering. Some participants used the text editor effectively,
+On the one hand, we observed two issues in the study. Participant \#2 struggled with correct indentation, starting
+a second command with more indentation than needed and participant \#6 had a syntax error in an unrelated command,
+which prevents charts from rendering. On the other hand, some participants used the text editor effectively,
 e.g.~participant \#5, who used copy-and-paste to fetch the same data series for multiple countries.
 
-\paragraph{How users understand the `then' member}
-When interviewing participants who worked with a tabular data source, we
-asked about their understanding of the \ikvd{then} member. This is a regular member generated by the
-type provider (i.e.~not a keyword), but it has a special meaning. Consider
-the following example which calculates the average travel expenses and total number of
-representatives per county using the UK House of Lords expenses data:
+\subsubsection*{Understanding the `then' Member}
+We asked participants who worked with tabular data about their understanding of the \ikvd{then} member.
+This is a regular member (not a keyword), but it has a special meaning in the domain specific language.
+Consider the calculation of average travel costs and number of representatives per county in the House of Lords:
 
 \begin{thegamma}
-expneses.'group data'.'by County'
-  .'average Travel Costs'.'count distinct Name'.then
+expneses.'group data'.'by County'.'average Travel Costs'.'count distinct Name'.then
 \end{thegamma}
 
-The \ikvd{then} member is used to complete a part of a query where the user can continue
-adding items to build a list. Here, we select two aggregations to be calculated for each
+\noindent
+The \ikvd{then} member is used to complete a part of a query where the user repeatedly
+add items to build a list. Here, we select two aggregations to be calculated for each
 county before choosing \ikvd{then} and applying other operations such as sorting.
-
-Two participants (\#12 and \#13) initially thought that \ikvd{then} is used to split a command
+Two participants (\#12 and \#13) thought that \ikvd{then} is used to split a command
 over multiple lines, but rejected the idea after experimenting and noting that they can insert
 line breaks elsewhere. One correctly concluded that it ``allows us to chain together the
-operations'' of the query. Participant \#13 needed a hint about using the
-\ikvd{then} member and later reflected:
+operations'' of the query. Following a hint, participant \#13 reflected:
 
@@
 
+\noindent
 This comment summarizes an important fact. Although iterative prompting allows the users to start
@@
 
-\subsection{Study conclusions}
-The motivation behind The Gamma is to make simple programmatic data exploration accessible to
-journalists and other non-experts. Main-stream programming languages have the benefit of
-reproducibility and transparency, but are difficult to use. Our study shows that non-experts
-are able to solve basic data exploration tasks using The Gamma. When asked whether The Gamma
-is something that journalists could learn how to use, participant \#13 (a former journalist) answered:
+% ==================================================================================================
+
+\section{Discussion}
+We examine an unexplored point in the design space of tools for data exploration. The Gamma is a
+text-based programming environment, but targets non-programmers such as data journalists. We
+conclude with a more general discussion of evaluation of the system and the possible applicability
+of The Gamma in data journalism.
+
+\subsection{Evaluating Exploratory Research}
+The research presented in this paper is qualitative and exploratory in nature. In particular,
+we do not make any quantitative claims about the usability of The Gamma and its learning curve.
+Our investigation focused on the core iterative prompting principle, but some our case studies
+also required using features such as operations with parameters.
+
+Although The Gamma is open-source, it has not been deployed in production in a newsroom so far.
+This would lead to valuable insights, but it requires finding a suitable fortuitous opportunity.
+Finally, we also do not compare the usability of The Gamma with other popular systems
+such as Tableau \cite{tableau}. It would be possible to set tasks that can be completed
+in both systems, but the systems have very different aims making such comparison problematic.
+
+\subsection{Evaluating Complex Systems}
+Data exploration environments are complex systems that do not yield to simple controlled
+experimentation. Olsen~\cite{evaluating} proposes criteria for judging whether a system advances
+the state of the art. A number of those apply to The Gamma:
+
+\begin{itemize}
+\item \emph{Importance.} Data journalism can make factual claims backed by data more commonplace
+and enable wider audience to engage with such claims. As such, we contribute towards
+solving an important societal issue.
+
+\item \emph{Expressive leverage.} Iterative prompting stratifies the complexity of data exploration
+such that different data sources are accessed through the same unified iterative prompting
+interaction principle.
+
+\item \emph{Empowering new participants.} As demonstrated by our user study, The Gamma allows
+non-experts, including those not comfortable with code, to perform basic programmatic
+data exploration tasks.
+
+\item \emph{Generality.} The Gamma can be used to query data from a wide range of data
+sources including tabular data, data cubes and graph databases. The range of possible tasks is
+illustrated by case studies presented earlier.
+\end{itemize}
+
+\subsection{Applications to Data Journalism}
+Although The Gamma targets a broad audience of non-programmers, some of our work has been
+particularly motivated by the use of data in journalism. The Gamma is aligned with recent challenges
+faced by journalists in that it can help build trust through transparency as well as provide
+meaningful ways of reader engagement.
+
+Our study shows that non-experts with background similar
+to journalists are able to solve basic data exploration tasks using The Gamma. When asked whether
+The Gamma is something that journalists could learn how to use, a former journalist who
+participated in our study (\#13) answered:
 
@@
 
-The Gamma is based on the iterative prompting principle which lowers the barrier to entry, but
-it is worth noting that it does not fully eliminate complexity involved in data querying.
-Instead, it provides a way of structuring the complexity. Iterative prompting makes it easy to
-get started, but complex data sources will inevitably require additional learning. The Gamma
-makes this easier by allowing users to learn from published code samples.
-
-% ==================================================================================================
-
-\section{Discussion}
-We design a programming environment that can make programmatic data exploration accessible
-to journalists. This is aligned with recent developments in journalism, which aims to build
-trust through transparency and greater reproducibility as well as provide meaningful ways of
-engagement.
-
-\subsection{Evaluating The Gamma}
-Data exploration environments are complex systems that do not yield to simple controlled
-experimentation. We evaluate The Gamma using criteria inspired by Olsen~\cite{evaluating}.
-
-\begin{itemize}
-\item \emph{Importance.} Open journalism has the potential to make factual claims backed by
-data more commonplace and enable wider audience engage with such claims. As such, we contribute
-towards solving an important societal issue.
-
-\item \emph{Expressive leverage.} The Gamma reduces the number of concepts that the user needs
-to master to just the \emph{iterative prompting} principle, although complex data sources may
-add more concepts through their domain specific languages.
-
-\item \emph{Empowering new participants.} As demonstrated by our user study, The Gamma allows
-non-experts, including those not comfortable with code, to perform basic programmatic
-data exploration tasks.
-
-\item \emph{Generality.} Finally, our case studies show that The Gamma can be used to solve
-a wide range of data manipulation tasks involving many data sources including tabular data,
-data cubes and graph databases.
-\end{itemize}
-
-\subsection{Remaining design issues}
+\noindent
+The answer suggests that iterative prompting, does indeed, lower the barrier to entry. Although it
+does not fully eliminate complexity involved in data querying, it provides a way of stratifying
+it. Iterative prompting makes it easy to get started with data exploration, addressing the initial
+``nervousness about code''. By making the source code of data anlyses visible, The Gamma then
+enables further learning through percolation.
+
+\subsection{Further Design Issues}
 There remain a number of aspects of data exploration in the context of journalism that
@@
 
-\paragraph{Data provenance}
-Data sources such as \ikvd{olympics} or \ikvd{worldbank} are defined when initializing The Gamma
-programming environment, but the system does not currently show where such data comes from.
-For some data sources, the source is e.g.~a CSV file published by the government. In this case,
-we can easily show the source. However, other type providers may pre-process data. Displaying
+\subsubsection*{Data provenance}
+Data sources such as \ikvd{olympics} or \ikvd{worldbank} are defined when initializing The Gamma,
+but the system does not currently show where such data comes from.
+For some tabular data sources, the source is a CSV file published, e.g.~by the government. In this case,
+we can easily show the source URL. However, other type providers may pre-process data. Displaying
 data source in such cases would require more sophisticated provenance tracking \cite{provenance}.
 
-\paragraph{Data availability}
+\subsubsection*{Data availability}
 In the current version, The Gamma does not have a way of informing the user what data sources
@@
 currated online database such as Enigma Public\footnote{\url{https://docs.enigma.com/public}}.
-Providing access to open government data repositories such as \href{http://data.gov.uk}{\small\bf\ttfamily data.gov.uk}
-is more appealing, but very challenging due to their unstructured nature.
-
-\subsection{Evaluation limitations}
-The evaluation conducted in this paper is qualitative and exploratory in nature. In particular,
-we do not make any quantitative claims about the usability of The Gamma and its learning curve.
-Our study also focused on the core iterative prompting interaction, but our case studies often
-required using other features such as operations with parameters.
-
-Although The Gamma is open-source, it has not been deployed in production in a newsroom so far.
-This would lead to valuable insights, but it requires finding a suitable fortuitous opportunity.
-Finally, we also do not compare the usability of The Gamma with the usability of other popular
-systems such as Tableau \cite{tableau}. It would be possible to set tasks that can be completed
-in both systems, but the systems have very different aims making such comparison problematic.
+Providing access to open government data repositories such as \url{http://data.gov} and
+\url{http://data.gov.uk} is more appealing, but challenging due to their unstructured nature.
 
 \section{Conclusions}
-[TODO: THIS IS A SHORT PAPER.... no design probe]
-
-We present The Gamma, a simple data exploration environment. The Gamma is motivated by the needs
-of journalists working with data, who need to produce transparent and easy-to-understand reports
-backed by data that are transparent and engaging, all the while facing tight deadlines.
-
-The Gamma is based on a single interaction principle, \emph{iterative prompting}, can be used to
-complete a wide range of data exploration tasks using a variety of data sources including tabular
-data, data cubes and graph databases. The design lowers the barrier to entry for programmatic
-data exploration and it makes it easy to learn the system independently from examples and by
-experimentation.
-
-% \section{Acknowledgments}
-% Yo
-
-% BALANCE COLUMNS
-\balance{}
-
-% REFERENCES FORMAT
-% References must be the same font size as other body text.
+Exploring data in a programming environment that makes the full source code available increases
+transparency, reproducibility and empowers users to ask critical questions about the data anlysis.
+But can we make those features accessible to non-programmers? In this paper, we presented The Gamma,
+a simple data exploration environment for non-programmers that answers this question in the
+affirmative.
+
+The Gamma is based on a single interaction principle, \emph{iterative prompting}. It can be used to
+complete a range of data exploration tasks using tabular data, data cubes and graph databases.
+The design lowers the barrier to entry for programmatic data exploration and makes it easy to learn
+the system independently through examples and by experimentation. We implemented The Gamma, make it
+available as open source and conducted a user study, which lets us conclude that
+The Gamma can be used by non-programmers to construct non-trivial data exploration scripts.
+
 \bibliographystyle{ACM-Reference-Format}
@@
 \end{document}
-
-%%% Local Variables:
-%%% mode: latex
-%%% TeX-master: t
-%%% End:

### 2020-09-16 wip-design/paper.tex
Finalizing submission

@@
 We aim to build an easy to use and learn programming system. Many approaches to this
-goal have been tried. Victor \cite{learnable,principle} introduced design principles that
+goal have been tried. Victor \cite{principle} introduced design principles that
 inspired many to build live programming systems \cite{review,liveroad,lighttable} that give

### 2020-09-16 wip-design/paper.tex
Typos

@@
 \caption{Obtaining teams with the greatest number of gold medals from Rio 2016
-Olympics with a reproducible The Gamma script (1), contextual iterative prompting mechainsm
+Olympics with a reproducible The Gamma script (1), contextual iterative prompting mechanism
 offering ways of sorting the data (2) and an instant preview of results (3).}
@@
 produce transparent, engaging data analyses that are easy to interpret without requiring expert
-programming skill~\cite{ddj}. The design of such data exploration tool poses a unique mix of chellenges.
+programming skill~\cite{ddj}. The design of such data exploration tool poses a unique mix of challenges.
 First, the tool needs to have a very low barrier to entry. Second, it needs to support a wide
@@
 
-\subsubsection*{Guestures and Data Entry}
+\subsubsection*{Gestures and Data Entry}
 Although our focus is on program construction, our work can be positioned in the
@@
 navigating through a complete space of options, while on-screen feedforward \cite{octopocus} allows
-efficient selection in guesture-based interfaces. Those provide compelling alternatives to
+efficient selection in gesture-based interfaces. Those provide compelling alternatives to
 auto-completion menus, although the efficiency of input methods is not an issue in programming.
@@
 that enable access to various kinds of data sources. It allows non-experts to create programs using the
-\emph{iterative prompting} interaction principle -- by repeately selecting an item from an auto-complete list.
+\emph{iterative prompting} interaction principle -- by repeatedly selecting an item from an auto-complete list.
 We start with a walkthrough of the system (Section~\ref{sec:overview-walk}), before looking at details of
@@
 The data is available as a CSV file. After the analyst imports the file through a web interface,
-the environment is initialized with code that refers to the imported data as \ikvd{expenses}
+the environment is initialised with code that refers to the imported data as \ikvd{expenses}
 and she starts exploring the data using the type provider for tabular data (Section~\ref{sec:overview-tps}),
@@
 The programming environment for The Gamma is based on a text editor. When the user types `.'
-the editor triggers auto-completion and retrives a list of available members based on the type
+the editor triggers auto-completion and retrieves a list of available members based on the type
 information. The programming environment evaluates scripts on-the-fly and shows a preview as
@@
 Conceptually, a type provider defines a domain specific language for exploring data of a particular
-kind. Techincally, it generates object types with members (such as \ikvd{sort by} or \ikvd{Count is})
+kind. Technically, it generates object types with members (such as \ikvd{sort by} or \ikvd{Count is})
 that are accessed via iterative prompting. We describe type provider for exploring data cubes
@@
 Our first type provider allows users to explore data form a data cube. For example, the World
-Bank\footnote{The data is retrived using an API available at: \url{https://data.worldbank.org/}}
+Bank\footnote{The data is retrieved using an API available at: \url{https://data.worldbank.org/}}
 collects a range of indicators about many countries in the world each year. The data set is a
@@
 only options that are valid in a given context and that (ii) any script that can be written
-in The Gamma can be contructed via iterative prompting.
+in The Gamma can be constructed via iterative prompting.
 
@@
   \includegraphics[width=1\columnwidth]{figures/cases-mp}
-  \caption{Exploring Olympic medalists using tabular data type provider}
+  \caption{Exploring Olympic medallists using tabular data type provider}
   \label{fig:cases-mp}
@@
   the \emph{worldbank} data source.
-\item \emph{Olympics.} Participnts were given a demo using \emph{olympics}.
+\item \emph{Olympics.} Participants were given a demo using \emph{olympics}.
   They were asked to solve a more complex problem, involving grouping and aggregation,
@@
       \small \#1 & \small expenditure & \small cube & \priority{50} & {\small Obtained one data series}\\
-      \small \#2 & \small expenditure & \small cube & \priority{100} & {\small Explored furhter data }\\
+      \small \#2 & \small expenditure & \small cube & \priority{100} & {\small Explored further data }\\
       \small \#3 & \small expenditure & \small cube & \priority{100}& {\small Explored further data  }\\
@@
 \emph{``For somebody who does not do coding or programming, this does not feel that daunting.
-  It's not like you're giving me large screen full of code, which is reassurring.''}
+  It's not like you're giving me large screen full of code, which is reassuring.''}
 \end{quote}
@@
 \begin{thegamma}
-expneses.'group data'.'by County'.'average Travel Costs'.'count distinct Name'.then
+expenses.'group data'.'by County'.'average Travel Costs'.'count distinct Name'.then
 \end{thegamma}
@@
 it. Iterative prompting makes it easy to get started with data exploration, addressing the initial
-``nervousness about code''. By making the source code of data anlyses visible, The Gamma then
+``nervousness about code''. By making the source code of data analyses visible, The Gamma then
 enables further learning through percolation.
@@
 the data source in the first place. This could be partly addressed by a type provider for a
-currated online database such as Enigma Public\footnote{\url{https://docs.enigma.com/public}}.
+curated online database such as Enigma Public\footnote{\url{https://docs.enigma.com/public}}.
 Providing access to open government data repositories such as \url{http://data.gov} and
@@
 Exploring data in a programming environment that makes the full source code available increases
-transparency, reproducibility and empowers users to ask critical questions about the data anlysis.
+transparency, reproducibility and empowers users to ask critical questions about the data analysis.
 But can we make those features accessible to non-programmers? In this paper, we presented The Gamma,

### 2021-03-22 wip-compost/paper.tex
Design paper - related work and introduction
Tweaks in pages 1-6

@@
 \usepackage{semantic}
+\usepackage{comment}
+
+\begin{comment}
+Dear editors,
+I'm grateful to the reviewers for their thorough assessment of the paper.
+I was very happy to read that the reviewer #2 perfectly captured my hope for the paper ("I must confess that at first Compost seemed too simplistic---but upon reflection I realize that is rather the point"), but the reviewers 1 and 3 made a number of excellent points that deserved clarification.
+I made a number of changes in response to the reviews.
+
+MAJOR CHANGES
+The following two major additions should clarify the paper w.r.t.
+many of the points made by reviewer 1 about the language design and its expressive power and limitations, as well as to answer points made by other reviewers (Haskell Chart, pie charts):
+
+* I added a new section 2.3 (Choosing the level of language abstraction) which provides more background on the design of the DSL and contrasts it with Haskell Chart library and the Pan language.
+This should address the concerns of reviewer 1 and explain that, e.g.
+supporting combinators like "above s1 s2" was not the intention of the design.
+
+* I added a new section 7 (Limitations and further work) that documents some of the things that are currently not supported.
+This discusses issues such as Pie charts (mentioned by reviewers 1 and 3), logarithmic and contracted axes (reviewer 3) and also visual aspects (reviewer 3).
+The section explains what is "not yet implemented" vs.
+what is "not supported by design".
+
+MINOR CHANGES
+The following are the notable minor changes that directly address the more specific reviewer comments:
+
+* Reviewer 1 asked for clarification re "nest" - the explanation has been extended to be more clear
+* Reviewer 2 pointed out a number of typos - all have been fixed (Thank you!)
+* Reviewer 2 asked about figure placement - I tried to improve this (but sometimes unsuccessfuly as I also want to keep long code snippets on one page)
+* Reviewer 3 asked about tight links with F# and Elm - I extended the discussion of implementation at the end of Section 1
+* Reviewer 3 asked for clarification of syntactic categories in Figure 3 - this is now done.
+* Reviewer 3 asked about syntax of host language vs. DSL - this is now explained at the start of Section 2.2
+* Reviewer 3 asked about units of measure - I believe this is covered in Section 2.5 (in the context of types)
+* Reviewer 3 asked for clarification re interaction - I added new paragraph in Section 5 and also made changes throughout to answer other specific questions raised
+
+\end{comment}
 
@@
 \newcommand{\narrow}[1]{\hspace{-0.6em}#1\hspace{-0.6em}}
-\newcommand{\urrl}[1]{\textnormal{\textcolor{urlclr}{\url{#1}}}}
+\newcommand{\urrl}[1]{\textnormal{\textcolor{urlclr}{#1}}}
 
@@
 
-Many libraries that can draw bar charts and line charts, but adding those extra
-features will only be possible if the author already thought about your exact scenario.
+Many libraries can draw bar charts and line charts, but extra features like multiple bars
+for each label, alignment of multiple charts or custom color coding can only be used if the
+library author already thought about your exact scenario.
 Google Charts \cite{gcharts} supports the left chart (it is called Dual-X Bar Chart) but there is no
-way to add a background, or sharing an axis between charts. The alternative is to use a more
+way to add a background or share an axis between charts. The alternative is to use a more
 low-level library. In D3 \cite{d3} you construct the chart piece by piece, but you have to
@@
 
-What would an elegant functional approach to data visualization look like? A functional programmer
-would want a domain-specific language that has a small number of primitives that allow us to define
-high level abstractions such as a bar chart and that uses domain values such as the exchange
-rate, rather than pixels, in its basic building blocks.
-
-\begin{figure}[h]
-  \includegraphics[scale=0.57]{figures/charts}
+\begin{figure}[b]
+  \vspace{0.5em}
+  \includegraphics[scale=0.55]{figures/charts}
   \vspace{0.25em}
@@
 \end{figure}
+
+What would an elegant functional approach to data visualization look like? A functional programmer
+would want a domain-specific language that has a small number of primitives that allow us to define
+high level abstractions such as a bar chart and that uses domain values such as the exchange
+rate, rather than pixels, in its basic building blocks.
 
@@
   is an exchange rate of 1.36 USD/GBP, not 67 pixels from the bottom.
-\item Most common chart types can be easily captured as high level abstractions, but there is a
-  way of creating a majority of more interesting custom charts.
+\item Common chart types such as bar charts or line charts can be easily captured as high level
+  abstractions, but many interesting custom charts can be created as well.
 \item The approach can easily be integrated with the Elm architecture \cite{elm}
-	to creating web-based charts that involve animations or interaction with the user.
+	to create web-based charts that involve animations or interaction with the user.
 \end{itemize}
@@
 domain-specific language. We outline the structure of an implementation, but omit the details; filling
-those in merely requires careful thinking about geometry and projections. An F\#
-implementation, including all our examples, is available
-at: \urrl{http://github.com/compostjs}.
+those in merely requires careful thinking about geometry and projections.
+
+Compost is available as open-source at \urrl{http://compostjs.github.io}. It is implemented in F\#,
+but is available as a plain JavaScript library thanks to the Fable F\# to JavaScript compiler.
+The core logic consists of 800 lines of code and depends on the virtual-dom
+library (\urrl{http://npmjs.com/package/virtual-dom}) for the
+implementation of the interactive features, meaking it easily portable to other functional
+programming languages.
 
@@
 \begin{equation*}
-\begin{array}{rcl}
-v & = & \kvd{cat}~c, r \\
-  & | & \kvd{cont}~n\\[1.25em]
-  ~
+\begin{array}{llcl}
+\textit{Category} & c\\
+\textit{Ratio} & r\\
+\textit{Number} & n\\
+\textit{Text} & t \\
+\textit{Color} & \gamma \\
+\textit{Value} & v & = & \kvd{cat}~c, r \\
+&   & | & \kvd{cont}~n
 \end{array}
-\qquad
-\begin{array}{rclcl}
-s & = &\kvd{line}~\gamma, [\,v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\,] & | & \kvd{overlay}~[\,s_1, \ldots, s_n\,]\\
- & | & \kvd{fill}~\gamma, [\,v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\,] & | &\kvd{axis}_{l/r/t/b}~s\\
- & | & \kvd{text}~\gamma, v_x, v_y, t & | &\kvd{padding}~n_t,n_r,n_b,n_l,s\\
- & | & \kvd{bubble}~\gamma, v_x, v_y, w, h
+\begin{array}{lrcl}
+\textit{Shape} & s & = &\kvd{line}~\gamma, [\,v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\,]\\
+&  & | & \kvd{fill}~\gamma, [\,v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\,] \\
+&  & | & \kvd{text}~\gamma, v_x, v_y, t \\
+&  & | & \kvd{bubble}~\gamma, v_x, v_y, n_w, n_h \\
+&  & | & \kvd{overlay}~[\,s_1, \ldots, s_n\,] \\
+&  & | &\kvd{axis}_{l/r/t/b}~s\\
+&  & | &\kvd{padding}~n_t,n_r,n_b,n_l,s\\
 \end{array}
@@
 
-Now we can define the basic elements of the Compost domain-specific language.
-A chart element is represented by the shape $s$ defined in Figure~\ref{fig:shape}.
+Compost is an embedded domain-specific language, implemented as a set of functions. In the
+subsequent code samples, we will use color to distinguish primitives of the Compost language,
+such as \kvd{overlay} or \kvd{cat} from primitives of the host language such as \fkvd{let} or
+\fkvd{for}.
+
+A chart element is represented by a shape $s$, as defined in Figure~\ref{fig:shape}.
 A primitive shape can be a text label, a line connecting a list of points, a filled polygon
@@
 Using those primitives, we can construct the simple UK election results bar chart in Figure~\ref{fig:simple} (left):
-%
+
 \begin{equation*}
@@
 \hspace{6.7em}\;(\kvd{cat}~\strf{Labour}, \num{1}), (\kvd{cont}~\num{202}), (\kvd{cat}~\strf{Labour}, \num{1}), (\kvd{cont}~\num{0})\,\;]\\[0.5em]
-\kvd{axis}_l~(\kvd{axis}_b~(\kvd{overlay}~[~\ident{labour}, \ident{conservative}~]))\\
+\kvd{axis}_l~(\kvd{axis}_b~(\kvd{overlay}~[~\ident{conservative}, \ident{labour}~]))\\
 \end{array}
@@
 \begin{figure}
-  \includegraphics[scale=0.57]{figures/simple}
+  \includegraphics[scale=0.55]{figures/simple}
   \vspace{0.25em}
@@
   \label{fig:simple}
+  \vspace{-1.5em}
 \end{figure}
@@
 Aside from iterating over all available parties and splitting the bar, the example also adds padding
-around the bars. This is specified in pixels rather than in terms of domain values. A similar
+around the bars, which is specified in pixels. A similar
 result could be be achieved by drawing a bar using a range from $\num{0.05}$ to $\num{0.5}$, but
@@
 The chart is still missing a title, which we add in Section~\ref{sec:abstractions}.
+
+\subsection{Choosing the level of language abstraction}
+\label{sec:basic-level}
+
+Perhaps the most important aspect of the design of any domain-specific language is the level
+of abstraction it uses. The bar chart example discussed in the previous section illustrates
+the choice made in Compost. On the one hand, Compost gives us flexibility by letting us
+compose charts from shapes. On the other hand, Compost limits what we can do by using
+two-dimensional space with positions determined by categorical or continuous values.
+In other words, the Compost design lies in the middle of a broader spectrum.
+
+An example of a more general domain-specific language is the Pan language \cite{fun} for producing
+images. Pan represents images as functions from a 2D point to a color. This makes it possible to
+create powerful combinators, e.g.~polar image transformation, but it makes it harder
+to express logic important for charting, e.g.~automatic alignment of shapes defined in terms
+of categorical values. Compost makes it easy to put two bars side-by-side in a
+bar chart, but harder to define generic combinators for aligning images.
+
+An example of a less general domain-specific language is the Haskell Chart library \cite{hcharts}.
+Haskell Chart provides a wide range of plots (such as lines, candles, areas, points, error bars, pies, etc.),
+but those can only be composed in limited ways by overlaying them or arranging them in a grid.
+The language is closer to the domain of the most common applications, but it places more
+restrictions on what can be expressed.
+
+The design of the Compost domain-specific language aims to capture the key principles shared
+by most charts, but avoid using a long list of different plot types. Different types of charts
+are all produced by composing shapes, but the ways in which shapes can be composed and transformed
+are limited to those that are needed for typical charts. We discuss the limitations of this
+approach in more detail in Section~\ref{sec:limits}.
 
@@
 available chart space. A categorical scale is defined by a list of individual categorical values.
-Note that we do not need a minimal and maximal ratios of the used categorical values as Compost
-will use an equal space for each category, regardless of where in this space a shape needs to
+Note that we do not need minimal and maximal ratios of the used categorical values as Compost
+will use equal space for each category, regardless of where in this space a shape needs to
 appear.
-
-\begin{figure}
-\vspace{-0.5em}
-\begin{equation*}
-\begin{array}{rcl}
-l & = & \kvd{continuous}~n_{min}, n_{max} \lsep \kvd{categorical}~[\,c_1, \ldots, c_k\,]
-\end{array}
-\end{equation*}
-\vspace{-1em}
-\caption{A scale $l$ can be continuous, defined by a range, or categorical, defined by a list of values.}
-\label{fig:scale}
-\vspace{-1em}
-\end{figure}
-\begin{figure}
-\begin{equation*}
-\begin{array}{rclcl}
-s & = & \kvd{roundScale}_{x/y} s      &|& \kvd{nest}_{x/y}~v_{min}, v_{max}, s \\
-  & | & \kvd{explicitScale}_{x/y}~l, s &|&  (\ldots)
-\end{array}
-\end{equation*}
-\caption{Additional combinators for controlling and nesting scales, extending earlier definition of $s$.}
-\label{fig:control}
-\end{figure}
 
@@
 and produces a new scale that represents the union of the two:
-%
+
 \begin{equation*}
@@
 according to the ratio $r$.
+
+\begin{figure}
+\vspace{-0.5em}
+\begin{equation*}
+\begin{array}{rcl}
+\textit{Scale}~~~l & = & \kvd{continuous}~n_{min}, n_{max} \lsep \kvd{categorical}~[\,c_1, \ldots, c_k\,]
+\end{array}
+\end{equation*}
+\vspace{-1em}
+\caption{A scale $l$ can be continuous, defined by a range, or categorical, defined by a list of values.}
+\label{fig:scale}
+\vspace{-1em}
+\end{figure}
 
@@
 
+\begin{figure}
+\begin{equation*}
+\begin{array}{rclcl}
+\textit{Shape}~~~s & = & \kvd{roundScale}_{x/y} s      &|& \kvd{nest}_{x/y}~v_{min}, v_{max}, s \\
+   & | & \kvd{explicitScale}_{x/y}~l, s &|&  (\ldots)
+\end{array}
+\end{equation*}
+\caption{Additional combinators for controlling and nesting scales, extending earlier definition of $s$.}
+\label{fig:control}
+\end{figure}
+
 \section{Advanced charts: Controlling scale composition}
@@
 scale). For example, the chart in Figure~\ref{fig:simple} (right) is constructed using the
-following code (reusing the \ident{labour} and \ident{conservative} variables defined earlier):
+following code (reusing the \ident{conservative} and \ident{labour} variables defined earlier):
 %
@@
 \kvd{axis}_l~(\kvd{axis}_b~(\kvd{roundScale}_y~(\kvd{explicitScale}_x~(\kvd{categorical}~[\,\strf{Labour},\strf{Conservative}\,]),\\
-\qquad \kvd{overlay}~[~\ident{labour}, \ident{conservative}~]~)))\\
+\qquad \kvd{overlay}~[~\ident{conservative}, \ident{labour}~]~)))\\
 \end{array}
@@
 The most interesting primitive for controlling scale composition
-is $\kvd{nest}_{x/y}$. The behaviour is similar to that of the SVG viewport \cite{svg}.
-It takes two values, $v_{min}, v_{max}$ and a shape $s$ as arguments and
+is $\kvd{nest}_{x/y}$. As with other primitives like \kvd{padding}, the primitive takes a shape
+with some additional parameters and defines a new shape.
+Its behaviour is similar to that of the SVG viewport \cite{svg}.
+The \kvd{nest} primitive takes two values, $v_{min}, v_{max}$ and a shape $s$ as arguments and
 nests the scale of the shape $s$ inside the region defined by $v_{min}, v_{max}$.
-When inferring scales of shapes, the scale of $\kvd{nest}_{x/y}~l, s$ will be a categorical or continuous scale
-constructed from the values $v_{min}$ and $v_{max}$, regardless of the values that are used inside
-the shape $s$. The chart space between $v_{max}$ and $v_{min}$ will then be used to render the
-nested shape $s$ using its inferred scale. An example of nesting is shown in Figure~\ref{fig:nesting}.
+When inferring scales of shapes, the scale of $\kvd{nest}_{x/y}~v_{min}, v_{max}, s$ will be a
+categorical or continuous scale constructed from the values $v_{min}$ and $v_{max}$, regardless
+of the values that are used inside the shape $s$. The chart space between $v_{max}$ and $v_{min}$
+will then be used to render the nested shape $s$ using its inferred scale. In other words, the
+operation defines a virtual coordinate system that exists only inside the newly created shape,
+but is invisible to anything outside of the shape.
+An example of nesting is shown in Figure~\ref{fig:nesting}.
 Here, a chart with a continuous scale from $\num{1.1}$ to $\num{1.4}$ (e.g.~GBP/EUR exchange rates)
@@
   \vspace{0.25em}
-  \caption{Sample charts built using derived abstractions; a scatter plot visualizing the Iris\\
-    dataset with a title (left) and a pairplot comparing two Iris features (right).}
+  \caption{Sample charts built using derived abstractions; a scatter plot visualizing the Iris dataset with a title (left) and a pairplot comparing two Iris features (right).}
   \label{fig:standard}
@@
 \Y scales inside the upper shape to continuous scales from $\num{0}$ to $\num{100}$ and then position
-the text label in the middle, at a point $(\kvd{cont}~\num{50}), (\kvd{cont}~\num{50})$.
+the text label in the middle, at a point $(\kvd{cont}~\num{50}), (\kvd{cont}~\num{50})$. We assume
+that the \kvd{text} primitive centers the text, although the actual implementation also allows the
+user to specify horizontal and vertical alignment.
 Figure~\ref{fig:standard} (left) shows a sample scatter plot chart with a title created using the
@@
 
-The example shows that the Compost is simple yet expressive. With just
+The example shows that Compost is simple yet expressive. With just
 a few lines of code, we are able to construct charts that, in other systems, require dedicated
@@
 
+Compost uses the Elm architecture \cite{mvu} to support interactive data visualizations.
+In this model, an interactive visualization is described using a pair of user-defined types
+and a pair of user-defined functions. The \emph{state} type represents the current state of what
+is displayed (e.g.~animation step or selection) and the \emph{event} type represents actions that
+the user can perform (e.g.~start an animation or draw a selection range). The two functions
+use the state and event types. The \emph{view} function creates a chart based on the current
+state and the \emph{update} function specifies how the state changes when an event occurs.
+
 \subsection{Domain-specific events}
-Compost can be used in conjunction with a virtual-dom library to create interactive data
-visualizations based on the Elm architecture \cite{mvu}. An interactive visualization is described using
-a pair of functions -- the \emph{view} function creates a chart based on the current state and
-the \emph{update} function specifies how the state changes when an event occurs.
 
@@
 
-\end{figure}
 \begin{figure}
@@
 \subsection{You Draw It data visualizations}
-To illustrate building of interactive data visualizations using Compost, we look at one aspect of
+To illustrate building interactive data visualizations using Compost, we look at one aspect of
 ``You Draw It''. We want to create a bar chart where the user can use drag \& drop
@@
 \noindent
-The state is a pair of boolean, indicating whether the user is currently dragging, and
-a list of key value pairs, storing the number of seats for each political party. Two types of events
+The state is a pair of a boolean, indicating whether the user is currently dragging, and
+a list of key/value pairs, storing the number of seats for each political party. Two types of events
 can occur in the visualization. First, the user may start or stop dragging, which is indicated using
@@
 
-\begin{figure}[t]
-  \includegraphics[scale=0.57]{figures/youdraw}
-  \vspace{1em}
-  \caption{Interactive ``You Draw it'' data visualization. The user moves cursor to a bar (left),
-    pushes a mouse button and drags the bar to the position that they think is the correct one (right).}
-  \label{fig:youdraw}
-\end{figure}
-
 Finally, the \ident{view} function takes the current state and builds the data visualization
 using the Compost domain-specific language. In addition, it also takes a parameter \ident{trigger},
-which is a function of type $\ident{Event}\rightarrow\ident{unit}$ that can be used to trigger
-events in handlers, registered using primitives such as \kvd{mouseMove}.
+which is an effectful function of type $\ident{Event}\rightarrow\ident{unit}$ that can be used to
+trigger events in handlers, registered using primitives such as \kvd{mouseMove}. The \ident{trigger}
+function is provided by the Compost runtime. When it is invoked from an event handler, it takes
+the current state, transforms it using the \emph{update} function, sets the new state as the
+current state and invokes the \emph{view} function to display the new state.
+
 To build the bar chart in Figure~\ref{fig:youdraw}, we use the same approach as in
@@
 
+The primitives for specifying mouse event handlers can be nested or appear in multiple sub-shapes
+of the composed shape. This makes it possible to attach different event handlers to different
+parts of a chart and get event coordinates in local units. In case of nesting, the nested handler
+will capture events that occur in the space occupied by the shape it wraps, but it will ignore
+events occuring outside of this area.
+
 The pair of functions, \ident{update} and \ident{view}, together with an initial state is
@@
 and then, again, render an updated chart.
+
+\begin{figure}[t]
+  \includegraphics[scale=0.57]{figures/youdraw}
+  \vspace{1em}
+  \caption{Interactive ``You Draw it'' data visualization. The user moves cursor to a bar (left),
+    pushes a mouse button and drags the bar to the position that they think is the correct one (right).}
+  \label{fig:youdraw}
+\end{figure}
 
@@
 \subsection{Inferring the scales of a shape}
+\label{sec:impl-infer}
 
@@
 
+\vspace{-1.25em}
+\noindent
 For primitive shapes, the operation projects the coordinates using \ident{project} and constructs
@@
 
+\section{Limitations and future work}
+\label{sec:limits}
+
+As discussed in Section~\ref{sec:basic-level}, the Compost library chooses a level of abstraction
+that makes it possible to express a wide range of charts, but does not allow arbitrary image
+manipulation. The examples discussed so far provide a good review of what can be expressed
+using Compost. It is also worth considering what cannot currently be expressed. For many of the
+current limitations, we also consider what additional primitive would address the problem.
+
+\subsection{Radial charts and image transformations}
+Compost cannot currently produce pie charts and other radial charts. This could be supported by
+defining a primitive \kvd{polar} that renders a shape $s$ specified as a parameter using
+a polar coordinate system instead of the default Cartesian system. Like the \kvd{nest} primitive,
+this would create a new shape that occupies a newly defined chart region. The \kvd{polar}
+primitive would make it possible to create pie charts, but also more elaborate Circos charts
+used to visualize genomic data \cite{circos}.
+
+Radial charts provide a clear motivation for supporting polar geometries, but we do not currently
+expect the need for more general image transformations such as those supported by Pan \cite{fun}.
+Those are useful for producing visually appealing images, but may not be necessary for data
+visualization. Arguably, we also do not expect the need for more general layout combinators such
+as \ident{above} or \ident{besides}  \cite{monoids}. Those can be expressed elegantly using image transformations.
+In Compost, we can achieve similar effect using \kvd{nest} and \kvd{explicitScale}, as shown when
+definining \ident{title} in Section~\ref{sec:abstractions}.
+
+\subsection{Combinations and transformations of scales}
+
+Another area in which Compost could be extended is to allow more flexible handling of scales.
+Currently, categorical scales are mapped to bins of equal size and continuous scales are mapped
+using a linear transformation. The current design does not make it possible to use logarithmic
+scale or, for example, contracted axis where a sub-range of values in the middle is omitted.
+Both of these could be supported if Compost allowed the user to specify a custom value
+transformation function.
+
+Another interesting challenge is to allow overlaying of charts with multiple scales. This can
+currently be done by using \kvd{overlay} together with \kvd{nest}. However, a more principled
+approach would be to allow the user to specify multiple, possibly named, scales for each shape.
+The \ident{calculateScales} operation discussed in Section~\ref{sec:impl-infer} would then need
+to return a list of scales rather than just a pair.
+
+\subsection{Controlling visual elements of a chart}
+
+There is also a number of occasions where the user might require more control over various
+visual elements of the chart such as fonts, text alignment or visual aspects of the automatically
+generated axes. The current implementation of Compost already allows control over fonts, font
+sizes and text alignment, but we omit the details for brevity.
+
+Controlling the visual aspects of axes is a more interesting problem. In fact, the
+\kvd{axis} primitive described in this paper is not a primitive operation, but rather a derived
+one. It is implemented by calculating the scales of the shape specified as an argument and
+overlaying it with lines (for axes and grid), text elements (for labels) and adding a padding.
+The current implementation does not allow much customization, but the user can look at the
+implementation and easily create their own version, much like they can create their own version
+of the \ident{title} operation described in Section~\ref{sec:abstractions}.
+
 \section{Conclusions}
 This paper presents a functional take on the problem of designing easy to use, but flexible
-abstractions for composing data visualizations. We hope to find a sweet between high-level, but
+abstractions for composing data visualizations. We hope to find a sweet spot between high-level, but
 inflexible approaches and low-level, but hard to use approaches.
@@
 powerful variants \cite{layered,vega-lite,vega,polaris}. In Grammar of Graphics, a chart is a mapping from
-data to a chart elements and their visual attributes. In contrast, in Compost, the mapping is
+data to chart elements and their visual attributes. In contrast, in Compost, the mapping is
 specified in the host programming language and a chart is merely a resulting data type
@@
 
-Even in the basic form, the Compost library makes it possible to
-create a wide range of standard, atypical and even interactive charts. Moreover, we hope that
-other functional programmers agree that it does so in a simple, logical and easy-to-understand way.
-In other words, we hope the presented functional library is, indeed, a functional pearl.
+\paragraph{Acknowledgements}
+The Compost library is the result of my prolonged effort to create an elegant charting API for
+F\#, which was supported, at various stages, by Don Syme at Microsoft Research and Howard Mansell
+at BlueMountain Capital. The idea of Compost first came together in discussion with Mathias
+Brandewinder and was (much much later) implemented thanks to the support of Google Digital News
+Initiative and The Alan Turing Institute. The final motivation for this paper was an invitation
+to talk at the LambdaDays conference in Krak\'ow and the positive comments from the attendees.
+Finally, the anonymous referees provided valuable feedback that made this a better paper.
 

### 2021-03-22 wip-design/paper.tex
Tweaks in pages 1-6

@@
 \begin{abstract}
-  Governments, non-profit organizations and citizen initiatives publish increasing amounts of
-  data, but extracting insights from such data and presenting them to the public is hard.
-  First, data comes in a variety of formats that each requires a different tool. Second, many
-  data exploration tools do not reveal how a result was obtained, making it difficult to reproduce
-  the results and check how they were obtained.
+  Data exploration tools based on code have many desirable characteristics. They can easily access
+  a wide range of different data sources, result in reproducible scripts and encourage users to
+  reuse and modify existing data analyses. Unfortunately, most programming tools require expert
+  coding skills. Can we make data exploration based on code accessible to non-experts?
   %
-  We contribute The Gamma, a novel data exploration environment for non-experts. The Gamma is based
-  on a single interaction principle and using it results in transparent and reproducible scripts.
-  This allows transfer of knowledge from one data source to another and
-  learning from previously created data analyses. We evaluate the usability and learnability of
-  The Gamma through a user study on non-technical employees of a research institute.
+  We present The Gamma, a novel programmatic data exploration environment.
+  The Gamma is based on a single interaction principle and using it results in transparent and
+  reproducible scripts. This lowers the barrier to entry and allows learning from previously
+  created data analyses. We evaluate the usability and learnability of The Gamma through a user
+  study on non-technical employees of a research institute.
   %
-  We argue that our approach allows journalists and the public to benefit from the rise
-  of open data, by making data exploration easier, more transparent and more reproducible.
-
+  Our work shows that we may not need to shy away from code in order to build accessible,
+  reproducible and transparent tools that will allow a broad audience to benefit from the
+  rise of open data.
 \end{abstract}
@@
 
-We contribute The Gamma, a text-based data exploration environment for non-experts. The Gamma
+We contribute The Gamma, a programmatic data exploration environment for non-experts. The Gamma
 is based on a single, easy to understand interaction principle and provides a uniform
@@
 \subsubsection*{Iterative Prompting}
-The key idea in The Gamma, which we term the \emph{iterative prompting} principle, is that all
-valid data exploration scripts can be constructed by repeatedly choosing from a list
-of offered options. This way, non-programmers can write entire scripts through auto-complete,
-without learning a programming language, but they still produce transparent and reproducible
-source code. In other words, iterative prompting turns auto-complete from a programmer assistance
-tool into a non-expert programming mechanism. A crucial feature is that iterative prompting
-only offer operations that are valid in a given context and that it offer all such operations;
-it is both correct and complete.
+The main contribution of our work is the \emph{iterative prompting} interaction principle,
+which makes it possible to construct all valid data exploration scripts by repeatedly choosing
+from a list of offered options. This way, non-programmers can write entire scripts through
+auto-complete without learning a programming language first, but still produce transparent
+and reproducible code. In other words, iterative prompting turns auto-complete from a programmer
+assistance tool into a non-expert programming mechanism. A crucial feature is that iterative
+prompting only offer operations that are valid in a given context and that it offer all such
+operations; it is both correct and complete.
 
@@
 may choose to view the results as a table or a basic chart before publishing their analysis. The
-Gamma makes such simple programming with data simple enough for non-experts. Scraping and cleaning
-of messy data or building rich data visualizations is outside of the scope of this paper, but
-exposing those using the iterative prompting approach is an interesting future challenge.
-
-\subsubsection*{Contributions}
+Gamma makes such programmatic data exploration simple enough for non-experts, but scraping and
+cleaning of messy data or building custom data visualizations is currently outside of the scope of
+our work. Exposing those using iterative prompting remains an interesting future challenge.
+
+\subsubsection*{Paper Overview}
 The Gamma is available (non-anonymously) at \url{http://thegamma.net},
 both as a JavaScript library and a hosted data exploration service. In this paper, we describe and
-evaluate the design principles behind the project:
+evaluate the design principles behind the project\footnote{This paper is X pages, without citations,
+to be commensurate with the size of the contribution.}:
 
@@
 \item We introduce the iterative prompting principle in The Gamma
-  and show that it can be used for querying of distinct data sources including data tables, graph
+  and show how it can be used for querying of distinct data sources including data tables, graph
   databases and data cubes (Section~\ref{sec:overview}).
 
-\item We reflect how our design addresses challenges faced by journalists (Section~\ref{sec:design-ddj})
-  and analyse the design theoretically. We argue that our design lowers barriers to entry (Section~\ref{sec:design-bar}),
+\item We reflect how our design lowers barriers to entry (Section~\ref{sec:design-bar}),
   supports learning without experts (Section~\ref{sec:design-expert}) and offers a
@@
 \item We evaluate the system through a number of case studies (Section~\ref{sec:cases})
-  and a user study (Section~\ref{sec:study}). Our study shows that non-programmers can use The Gamma
-  to construct non-trivial data queries and ascertains the extent to which users can,
-  (i) learn from examples and (ii)~transfer knowledge between tasks.
+  and a user study (Section~\ref{sec:study}), which confirms  that non-programmers can use The Gamma
+  to construct non-trivial data queries.
 \end{itemize}
@@
 established auto-completion mechanism. Unlike most past work dating back to Kaiser~\cite{assistants},
-we do not view it as a programmer assistance tool, but as an interaction mechanism through which
-non-experts can create entire programs. We build on recent research on information-rich
+we do not view it as a programmer assistance tool. Instead, we turn it into an interaction mechanism
+through which non-experts can create entire programs. We build on recent research on information-rich
 programming \cite{inforich} and aim to make those advances available to non-programmers
-\cite{enduser,smallmatter}, in the context of data exploration as done, most notably, by
-journalists \cite{ddj}. Our work features a novel combination of characteristics in that
-the iterative prompting we develop (i) is centered around editing and understanding of program code,
-(ii) its conceptual complexity is reduced to a single basic kind of interaction, yet (iii) it is
-correct and complete in that it can be used to construct all meaningful programs.
+\cite{enduser,smallmatter}, in the context of data exploration as done, for example, by journalists
+\cite{ddj}. Our work features a novel combination of characteristics in that our iterative prompting
+interaction principle (i) is centered around editing and understanding of code,
+(ii) reduces conceptual complexity to a single basic kind of interaction, yet (iii) it is
+correct and complete in that it can be used to construct all meaningful queries for a variety
+of data sources.
 
@@
 
-Most similar to our approach are tools that recommend scripts when users begin interacting
-with data. Those based on machine learning code completion for domain specific languages \cite{predictive,proactive}
-differ in that they do not guarantee completeness, i.e.~the user cannot create all possible
-scripts. Approaches based on natural language can effectively support data exploration
-or visualization \cite{eviza,codemend}, but hide the structure of the underlying language and
-require more than just selecting options. Conversational agents \cite{iris} improve on such work
-in that they can provide more guidance. Code completion based on machine learning or statistical
+Most similar to our approach are tools that make recommendations when users begin interacting
+with data. Those based on machine learning-based code completion for domain specific languages \cite{predictive,proactive}
+differ in that they do not guarantee completeness, i.e.~it is unclear whether the user can create all possible
+scripts. Approaches based on natural language are effective \cite{eviza,codemend}, but hide the
+underlying structure and do not help the user understand it. Conversational agents \cite{iris}
+share similar characteristics, except that the construction process is iterative.
+
+Code completion based on machine learning or statistical
 methods \cite{mlcomplete,statcomplete} also exists for general-purpose programming languages used
@@
 Finally, DS.js \cite{dsjs} is interesting in that it enables querying of data on the
-web; it uses JavaScript with rich contextual code completion.
+web. It uses JavaScript, but with rich contextual code completion.
 
@@
 visual outputs, are widely used by data scientists, but require expert programming skills.
-The Gamma targets non-experts, but could be easily integrated with a multi-language notebook system
-such as Wrattler \cite{wrattler}.
-
-Spreadsheets and business intelligence tools \cite{tableau,powerbi} do
-not involve programming, but require mastering a complex GUI. This is also the case
-for other visual data analytics tools \cite{control,vizdom}. In contrast, The Gamma is
-based on a single kind of interaction. Several visual systems \cite{potter,wrangler,lyra} record
-interactions with the GUI as a script that can be seen and modified by the user.
-Unlike in The Gamma, the source code does not guide the user in learning how to use the system.
+The Gamma targets non-experts, but could also be integrated with a multi-language notebook system
+\cite{wrattler}.
+%
+Spreadsheets, business intelligence tools \cite{tableau,powerbi} and other visual data analytics
+tools \cite{control,vizdom} do not involve programming, but require mastering a complex GUI.
+In contrast, The Gamma is based on a single kind of interaction, through which all available operations
+can be completed. Several systems \cite{potter,wrangler,lyra} record interactions with the GUI
+as a script that can be modified by the user. Unlike in The Gamma, the source code does not
+guide the user in learning how to use the system.
 
@@
 preview of the results.
-
-Many systems simplify programming by designing high-level declarative abstractions,
+%
+Many systems simplify programming by offering high-level abstractions,
 e.g.~for interactive news articles \cite{idyll}, statistical analyses \cite{tea}
-or interactive data visualization \cite{interactionviz,vegalite}. The Gamma uses
-high-level abstractions for data querying, but designing high-level iterative prompting
-abstractions for other tasks remains future work.
+or interactive data visualization \cite{interactionviz,vegalite}. The Gamma exposes a number
+of data sources through high-level abstractions that support iterative prompting, but support
+for tasks other than querying remains furture work.
 
@@
 efficient selection in gesture-based interfaces. Those provide compelling alternatives to
-auto-completion menus, although the efficiency of input methods is not an issue in programming.
+auto-completion menus, although the efficiency of input methods is typically not an issue in programming.
 
@@
 high-level declarative abstractions, most systems that target non-experts shy away from code, by using
-either programming by example, natural language or graphical user interface. As spelled out in
-Section~\ref{sec:design}, text-based programming has notable benefits for public-facing data analyses
-such as learnability and transparency. Our work is thus motivated primarily by the question whether we
-can make text-based programming easy enough for non-experts? The study presented in Section~\ref{sec:study}
-shows that this is, indeed, possible at least for typical data querying tasks.
+either programming by example, natural language or graphical user interface. However, text-based
+programming has notable benefits such as learnability and transparency. Our work is primarily
+motivated by the question whether we can make text-based programming easy enough for non-experts?
+The study presented in Section~\ref{sec:study} shows that this is, indeed, possible at least for typical data querying tasks.
 The Gamma consists of a programming language, a web-based coding environment and a number of type providers
@@
 \emph{iterative prompting} interaction principle -- by repeatedly selecting an item from an auto-complete list.
+%
 We start with a walkthrough of the system (Section~\ref{sec:overview-walk}), before looking at details of
-The Gamma language (Section~\ref{sec:overview-lang}) and individual type providers (Section~\ref{sec:overview-tps}).
+The Gamma language (Section~\ref{sec:overview-lang}) and type providers (Section~\ref{sec:overview-tps}).
 
@@
 
-To introduce The Gamma, we walk through a simple problem that a local journalist might want to solve.
-The UK government publishes travel expense claims by members of the House of Lords. We want to find
-out which of the members representing the Kent county spend the most on travel.
-The following shows a subset of the data:\footnote{Full data set has been obtained from
-\url{https://www.parliament.uk/mps-lords-and-offices/members-allowances/house-of-lords/holallowances/} }
+To introduce The Gamma, we walk through a simple scenario.
+A journalist from Kent is exploring travel expense claims by members of the House of Lords published
+by the UK government. The data is available as a CSV file\footnote{Full data set has been obtained from
+\url{https://www.parliament.uk/mps-lords-and-offices/members-allowances/house-of-lords/holallowances/} }:
 
@@
 \noindent
-The data is available as a CSV file. After the analyst imports the file through a web interface,
+After the analyst imports the file through a web interface,
 the environment is initialised with code that refers to the imported data as \ikvd{expenses}
@@
 \item The analyst types `.' (dot) to trigger auto-completion on \ikvd{expenses}. The type provider
-  offers a list of operations that the analyst might want to perform such as grouping,
-  filtering and sorting.
-
-\item To find the House of Lords with the largest spending, the analyst chooses the
-  \ikvd{sort data} operation. Next, she is offered a list of possible arguments based on the
-  schema of the tabular data and chooses the desired one.
-
-\item The Gamma evaluates the source code on-the-fly and shows a preview
-  of results. After choosing the sorting key, the analyst sees a table with House of Lords members
-  from more remote counties of the UK.
-
-\item To finish specifying the (possibly compound) sorting key, the analyst chooses \ikvd{then}
-  and is offered the same list of querying operations as in the first step. To obtain House
-  members from Kent, she chooses \ikvd{filter data}. To navigate through the offered list more
-  efficiently, she types first two characters of the name.
-
-\item After selecting \ikvd{County is} to specify the desired type of condition, the analyst types
-  `.' and is offered a list of options based on the values of the \ikvd{County} column in the
-  source data set. She types \ikvd{Ke} and selects \ikvd{Kent}.
+  offers a list of operations that the analyst can perform. To find House of Lords
+  members from Kent, the analyst chooses \ikvd{filter data}.
+
+\item The analyst is offered a list of columns based on the schema of the tabular data and choses
+  \ikvd{County is}.  She is then offered a list of counties in the data set and types \ikvd{ke} to
+  search for Kent and she selects \ikvd{Kent}.
+
+\item The Gamma evaluates the code on-the-fly and shows a
+  preview of results. The analyst now sees a table with House of Lords members from Kent.
+  She wants to see if there are any members who missed any House sessions.
+
+\item The analyst finishes specifying the (possibly compound) sorting key by choosing \ikvd{then}
+  and is offered the same list of querying operations as in the first step. She selects
+  \ikvd{sort data} followed by \ikvd{by Days Away descending}.
+
+\item The analyst sees that there are no reported ``days away'' and decides to compare
+  travel costs. She hits the backspace key a number of times, is offered the list
+  of keys again and selects \ikvd{by Travel Costs descending}.
 
@@
 
-The Gamma consists of a text-based programming language with a web-based coding environment,
-based on the Monaco editor\footnote{For more information, see \url{https://microsoft.github.io/monaco-editor/}},
-and type providers that provide access to data tables, graph databases and data cubes.
-
-\subsubsection*{The Gamma Language}
 A program in The Gamma is a sequence of commands. A command can be either a variable declaration
@@
 A member can be either an ordinary member such as \ikvd{paging} or an operation which takes a
-list of parameters enclosed in parentheses as in \ikvd{take(5)}. When the member name contains
-non-alphanumerical characters it is  written in quotes such as \ikvd{\textquotesingle sort by\textquotesingle}.
+list of parameters enclosed in parentheses as in \ikvd{take(5)}.
+Names with non-alphanumerical characters are escaped using quotes.
 
@@
 The types are not built-in, but are generated by type providers for individual data sources.
-The programming environment for The Gamma is based on a text editor. When the user types `.'
+The programming environment for The Gamma is based on the Monaco editor\footnote{For
+more information, see \url{https://microsoft.github.io/monaco-editor/}}. When the user types `.'
 the editor triggers auto-completion and retrieves a list of available members based on the type
@@
 
-\subsubsection*{Making Complex Things Possible}
-As illustrated by the \ikvd{take(5)} operation, there is a handful of situations where The
-Gamma does not yet fully support the iterative prompting principle. The language supports
-a small number of other features that can be used by more advanced users through text editing:
-
-\vspace{0.3em}
-\begin{thegamma}
-\kvd{let} topTravel = expenses.'sort data'.'by Travel Costs descending'.then
-  .paging.take(5).'get series'.'with key Name'.'and value Travel Costs'
-charts.column(topTravel).setColors(["red","blue","green"])
-  .setTitle("House of Lords members by travel expenses")
-\end{thegamma}
-\vspace{0.3em}
-
-\noindent
-First, The Gamma allows operations with parameters such as \ikvd{take(5)} or \ikvd{setTitle("...")}.
-This is currently needed when writing a query that skips or takes first N elements from a table.
-The remaining features are not needed for basic data exploration. The \ikvd{let} construct can be
-used to define (immutable) variables and The Gamma also supports lists written as \ikvd{[1,2,3]}.
-Advanced language features are currently used when building custom charts, but we expect that a
-charting library compatible with iterative prompting would alleviate the need for most of those.
-
-\subsection{Type Providers for Data Querying}
-\label{sec:overview-tps}
-
-The Gamma can be extended to support any kind of data source by implementing a \emph{type provider}.
-Conceptually, a type provider defines a domain specific language for exploring data of a particular
-kind. Technically, it generates object types with members (such as \ikvd{sort by} or \ikvd{Count is})
-that are accessed via iterative prompting. We describe type provider for exploring data cubes
-(inspired by an example from Syme et al. \cite{inforich}), tabular data (based on theory developed
-by Petricek \cite{dotdriven}), and a novel type provider for exploring graph databases.
-
-\subsubsection*{Data Cube Type Provider}
-Our first type provider allows users to explore data form a data cube. For example, the World
-Bank\footnote{The data is retrieved using an API available at: \url{https://data.worldbank.org/}}
-collects a range of indicators about many countries in the world each year. The data set is a
-three-dimensional cube with dimensions corresponding to countries, indicators and years. The
-following example uses the provider to access CO$_{2}$ emission data for United States:
-
-\begin{thegamma}
-worldbank.byCountry.'United States'.'Climate Change'.'CO2 emissions (kt)'
-\end{thegamma}
-
-\noindent
-As illustrated in Figure~\ref{fig:cubetp}, the provider allows users to select a data series
-from the data cube. Choosing \ikvd{byCountry.\textquotesingle United States\textquotesingle},
-restricts the cube to a two-dimensional plane. We then choose an indicator category
-and a specific indicator \ikvd{\textquotesingle CO2 emissions (kt)\textquotesingle}, obtaining
-a time series with years as keys and emission data as values. Similarly, we could first filter the
-data by a year or an indicator. The same mechanism can be used for exploring the UK government
-expenditure:
-
-\begin{thegamma}
-expenditure.byService.Defence.inTermsOf.GDP
-\end{thegamma}
-
-\noindent
-The dimensions of the cube are government services, years and value type (adjusted, nominal,
-per GDP). Here, we select the \ikvd{Defence} service and \ikvd{GDP} value type.
-As there is no widely used standard format for data cubes, adding another data cube to The Gamma
-currently requires implementing a new type provider.
-
-\subsubsection*{Tabular Data Type Provider}
-
-Our second type provider allows users to construct queries to explore data in tabular formats
-such as CSV files. A prominent example of tabular data used by journalists is, e.g.~the
-Iraq War documents leak~\cite{iraq}. Unlike the data cube provider, the provider for tabular data does not just
-allow selecting a subset of the data, but it can be used to construct SQL-like query. Consider
-the example from Figure~\ref{fig:thegamma}:
-
-\begin{thegamma}
-olympics.'filter data'.'Games is'.'Rio (2016)'.then
-  .'group data'.'by Team'.'sum Gold'.'sum Silver'.then
-  .'sort data'.'by Gold descending'
-\end{thegamma}
-
-\noindent
-The example works with a CSV file that records individual medals awarded in Olympic games.
-The chain constructs a query that selects rows corresponding to the Rio 2016 Olympics and then
-calculates total number of gold and silver medals for each team (country) before sorting the data.
-
-When using the provider, the user specifies a sequence of operations. Members such as
-\ikvd{\textquotesingle filter data\textquotesingle} or \ikvd{\textquotesingle group data\textquotesingle}
-determine the operation type. Those are followed by operation parameters. For example, when grouping
-data, we first select the key and then choose a number of aggregations to calculate over the group.
-Unlike SQL, the provider only allows users to choose from pre-defined aggregations such as
-calculating the sum, average or the number of distinct values. As illustrated in
-Section~\ref{sec:cases}, this is sufficient to construct a wide range of practical queries.
+There is a handful of situations where The Gamma does not yet fully support the iterative prompting
+principle. First, it allows operations with parameters such as \ikvd{take(5)}. This is currently
+needed when writing a query that skips or takes the first N elements from a table. Second,
+The Gamma allows the user to declare (immutable) variables using \ikvd{let}. This is not needed
+for basic data exploration, but allows advanced users to better structure more complex code.
 
@@
   \centering
-  \includegraphics[scale=0.25]{figures/cubetp}
+  \includegraphics[scale=0.24]{figures/cubetp}
   \vspace{0.5em}
@@
   \centering
-  \includegraphics[scale=0.28]{figures/graphtp}
+  \includegraphics[scale=0.27]{figures/graphtp}
   \caption{To query graph data, the user specifies a path through the data, possibly with
@@
 
+\subsection{Type Providers for Data Querying}
+\label{sec:overview-tps}
+
+The Gamma can be extended to support any kind of data source by implementing a \emph{type provider},
+which defines a domain specific language for exploring data of a particular kind.
+Technically, a type provider generates object types with members (such as \ikvd{paging} or \ikvd{Kent})
+that are accessed via iterative prompting. We describe type provider for exploring data cubes
+(inspired by an example from Syme et al. \cite{inforich}), tabular data (based on theory developed
+by Petricek \cite{dotdriven}), and a novel type provider for exploring graph databases.
+
+\subsubsection*{Data Cube Type Provider}
+Our first type provider allows users to explore data cubes, which are multi-dimensional
+arrays of values. For example, the World Bank collects a range of indicators about many countries
+each year while the UK government expenditure records spending for different government services,
+over time, with different adjustments:
+
+\begin{thegamma}
+worldbank.byCountry.'United States'.'Climate Change'.'CO2 emissions (kt)'
+expenditure.byService.Defence.inTermsOf.GDP
+\end{thegamma}
+
+\noindent
+The dimensions of the \ikvd{worldbank} cube are countries, years and indicators, whereas the dimensions
+of \ikvd{expenditure} are government services, years and value type (adjusted, nominal,
+per GDP).  As illustrated in Figure~\ref{fig:cubetp}, the provider allows users to select a data series
+from the data cube. Choosing \ikvd{byCountry.\textquotesingle United States\textquotesingle},
+restricts the cube to a two-dimensional plane; selecting \ikvd{\textquotesingle CO2 emissions (kt)\textquotesingle}
+then gives a series with years as keys and emission data as values. Similarly, we could first filter the
+data by a year or an indicator. The same mechanism is used to select UK government spending on
+defence in terms of GDP over all available years.
+As there is no widely used standard format for data cubes, adding another data cube to The Gamma
+currently requires implementing a new type provider.
+
+\subsubsection*{Tabular Data Type Provider}
+
+Our second type provider allows users to construct queries to explore data in tabular formats
+such as CSV files. Unlike the data cube provider, the provider for tabular data does not just
+allow selecting a subset of the data, but it can be used to construct SQL-like query. Consider
+the example from Figure~\ref{fig:thegamma}:
+
+\begin{thegamma}
+olympics.'filter data'.'Games is'.'Rio (2016)'.then
+  .'group data'.'by Team'.'sum Gold'.'sum Silver'.then
+  .'sort data'.'by Gold descending'
+\end{thegamma}
+
+\noindent
+The example works with a CSV file that records individual medals awarded in Olympic games.
+The chain constructs a query that selects rows corresponding to the Rio 2016 Olympics and then
+calculates total number of gold and silver medals for each team (country) before sorting the data.
+
+When using the provider, the user specifies a sequence of operations. Members such as
+\ikvd{\textquotesingle filter data\textquotesingle} or \ikvd{\textquotesingle group data\textquotesingle}
+determine the operation type. Those are followed by operation parameters. For example, when grouping
+data, we first select the key and then choose a number of aggregations to calculate over the group.
+Unlike SQL, the provider only allows users to choose from pre-defined aggregations such as
+calculating the sum, average or the number of distinct values. As illustrated in
+Section~\ref{sec:cases}, this is sufficient to construct a wide range of practical queries.
+
 \subsubsection*{Graph Database Type Provider}
@@
 the returned table can be further queried using the tabular data type provider.
-
-% ==================================================================================================
-
-\section{Reflections on Design}
-\label{sec:design}
-
-The design of The Gamma has been motivated by a curiosity as to whether iterative
-prompting can make text-based programming with data accessible to non-experts. In this section,
-we theoretically assess the resulting design and position it in the context of a possible application
-in the context of data journalism.
-
-\subsection{Data Journalism Perspective}
-\label{sec:design-ddj}
-
-Although our work targets non-expert data exploration in general, data journalism is an important
-specific domain with interesting design requirements. We reflect those on the basis of relevant
-literature, e.g.~\cite{ddj,edcj17,edcj18} and past experience of collaborating with
-journalists. Two challenges faced by journalists \cite{future} are particularly relevant to The Gamma.
-
-\subsubsection*{Trust Through Transparency}
-The first issue is trust in media. Many practitioners believe that transparency
-about the editorial process and information sources can serve as a proof of quality and
-trustworthiness~\cite{transparency}. This applies to data analyses too. For example, the Financial Times
-increasingly share source code of their analyses on GitHub, e.g.~\cite{ftnotebooks},
-but reproducing such analyses is difficult even for an expert. The Gamma has a potential to
-improve on the status quo in that a part of the analysis can be directly embedded in an article.
-The source code provides full account of what the data sources are and how are they used.
-The Gamma makes this code, to a greater extent, accessible to non-expert readers.
-
-\subsubsection*{Encouraging Meaningful Engagement}
-
-The second challenge is developing relationship with readers. Journalists are looking
-for meaningful ways of engagement through reader comments, involvement of citizen journalists
-\cite{comments,citizen} and the development of new interactive formats \cite{youdraw}.
-The Gamma can support this aim by providing a data exploration environment that can engage
-non-expert readers in a meaningful discussion. As argued below, non-experts can learn to use the
-system and can, for example, modify parameters or change filtering criteria to get a
-new perspective on a story.
-
-\subsection{Lowering Barriers to Entry}
-\label{sec:design-bar}
-
-Data exploration has a certain irreducible essential complexity. To make it accessible to users who
-cannot dedicate much time to learning a tool prior to using it, this complexity needs to be
-carefully stratified. The Gamma uses a two-level structure. At the first level, the user needs
-to learn iterative prompting to be able to start exploring data. At the second level, users will
-need to learn domain specific languages defined by individual type providers.
-
-\subsubsection*{Iterative Prompting Principle}
-Iterative prompting is a suitable first level principle, because it is easy to trigger.
-In conventional programming languages, auto-complete assistance is only available once basic
-code structure is written. In contrast, user of In The Gamma only needs to choose the initial
-data source. Iterative prompting is also easier to use than e.g.~a command line or a REPL
-(read-eval-print-loop), because it follows the \emph{recognition over recall} usability heuristic.
-The users are not required to recall and type a command. They merely need to select one from a
-list of options.
-
-\subsubsection*{Stratifying Data Exploration Complexity}
-At the first level, any data source is accessed through the same mechanism. At the second level,
-each data source defines its own domain specific language, consisting of the primitives that are
-offered to the user in auto-complete prompts. The domain specific languages are embedded in The Gamma
-language -- they define merely the available members and cannot extend the core language syntax.
-Although the complexity of individual languages differs, the users can always start exploring
-and learning new languages using the familiar first level iterative prompting principle.
-An important question, which we study in Section~\ref{sec:study}, is whether the expertise gained
-with one data source be transferred to working with another data source.
-
-\subsection{Learning without Experts}
-\label{sec:design-expert}
-
-A typical user of The Gamma will not dedicate significant time to learning it in advance and
-they will not have access to experts. Most learning thus needs to happen from examples.
-When analysing how Excel users learn, Sarkar \cite{learning} points out that users learn new
-features when the usage of a feature is apparent in a spreadsheet. For example, users can learn
-different functions in formulas, because those are visible in the cell. Learning how to use a
-wizard for creating charts is not possible because the operation leaves no full trace in the
-spreadsheet. Sarkar's recommendation is to \emph{design for percolation},
-i.e.~in a way where looking at the final result makes it apparent what feature has been used and how.
-
-In The Gamma, each step of iterative prompting results in an identifier that is
-added to the source code. This means that a program constructed solely through iterative prompting
-keeps a full trace of how it was created. The resulting source code provides the user all
-information that they need to recreate the program, not just by copying it, but also by using the
-iterative prompting mechanism. We assess the viability of this way of learning in the study in
-Section~\ref{sec:study}.
-
-\subsection{Correctness and Completeness}
-\label{sec:design-cc}
-
-An important characteristic of our design is that, barring a few exceptions discussed below, the iterative
-prompting mechanism is both correct and complete with respect to possible data exploration scripts.
-This means that (i) auto-complete lists offered by iterative prompting contain
-only options that are valid in a given context and that (ii) any script that can be written
-in The Gamma can be constructed via iterative prompting.
-
-\subsubsection*{Correctness} The Gamma uses an object-based type system for error-checking and for
-generating auto-complete lists. A well-typed script can always be executed. When using iterative
-prompting, a selected option, which is a valid object member, is added to the end of a script,
-leading to a well-typed script. This distinguishes our system from auto-completion based on machine
-learning or dictionary-based methods, which may offer members not valid in a given context. Advanced
-users of The Gamma, modifying scripts as text, can still violate correctness -- for example, if the
-user uses auto-complete to replace a member in a middle of a member access chain with a member of another type.
-
-\subsubsection*{Completeness} Auto-completion lists offered via iterative prompting contain
-all available members and so the user can construct all possible scripts. The range of such scripts
-is determined by type providers, which support limited set of options (e.g. group aggregations in
-the tabular data type provider). Two exceptions to completeness in our current design is specifying
-numerical parameters as in \ikvd{take(5)} and advanced language features such as let binding.
-We believe that alleviating the need for those poses an interesting further research challenge.
 
@@
 
+% ==================================================================================================
+
+\section{Reflections on Design}
+\label{sec:design}
+
+The design of The Gamma has been motivated by a curiosity as to whether iterative
+prompting can make text-based programming with data accessible to non-experts. In this section,
+we theoretically assess the resulting design and position it in the context of a possible application
+in the context of data journalism.
+
+\subsection{Lowering Barriers to Entry}
+\label{sec:design-bar}
+
+Data exploration has a certain irreducible essential complexity. To make it accessible to users who
+cannot dedicate much time to learning a tool prior to using it, this complexity needs to be
+carefully stratified. The Gamma uses a two-level structure. At the first level, the user needs
+to learn iterative prompting to be able to start exploring data. At the second level, users will
+need to learn domain specific languages defined by individual type providers.
+
+\subsubsection*{Iterative Prompting Principle}
+Iterative prompting is a suitable first level principle, because it is easy to trigger.
+In conventional programming languages, auto-complete assistance is only available once basic
+code structure is written. In contrast, user of In The Gamma only needs to choose the initial
+data source. Iterative prompting is also easier to use than e.g.~a command line or a REPL
+(read-eval-print-loop), because it follows the \emph{recognition over recall} usability heuristic.
+The users are not required to recall and type a command. They merely need to select one from a
+list of options.
+
+\subsubsection*{Stratifying Data Exploration Complexity}
+At the first level, any data source is accessed through the same mechanism. At the second level,
+each data source defines its own domain specific language, consisting of the primitives that are
+offered to the user in auto-complete prompts. The domain specific languages are embedded in The Gamma
+language -- they define merely the available members and cannot extend the core language syntax.
+Although the complexity of individual languages differs, the users can always start exploring
+and learning new languages using the familiar first level iterative prompting principle.
+An important question, which we study in Section~\ref{sec:study}, is whether the expertise gained
+with one data source be transferred to working with another data source.
+
+\subsection{Learning without Experts}
+\label{sec:design-expert}
+
+A typical user of The Gamma will not dedicate significant time to learning it in advance and
+they will not have access to experts. Most learning thus needs to happen from examples.
+When analysing how Excel users learn, Sarkar \cite{learning} points out that users learn new
+features when the usage of a feature is apparent in a spreadsheet. For example, users can learn
+different functions in formulas, because those are visible in the cell. Learning how to use a
+wizard for creating charts is not possible because the operation leaves no full trace in the
+spreadsheet. Sarkar's recommendation is to \emph{design for percolation},
+i.e.~in a way where looking at the final result makes it apparent what feature has been used and how.
+
+In The Gamma, each step of iterative prompting results in an identifier that is
+added to the source code. This means that a program constructed solely through iterative prompting
+keeps a full trace of how it was created. The resulting source code provides the user all
+information that they need to recreate the program, not just by copying it, but also by using the
+iterative prompting mechanism. We assess the viability of this way of learning in the study in
+Section~\ref{sec:study}.
+
+\subsection{Correctness and Completeness}
+\label{sec:design-cc}
+
+An important characteristic of our design is that, barring a few exceptions discussed below, the iterative
+prompting mechanism is both correct and complete with respect to possible data exploration scripts.
+This means that (i) auto-complete lists offered by iterative prompting contain
+only options that are valid in a given context and that (ii) any script that can be written
+in The Gamma can be constructed via iterative prompting.
+
+\subsubsection*{Correctness} The Gamma uses an object-based type system for error-checking and for
+generating auto-complete lists. A well-typed script can always be executed. When using iterative
+prompting, a selected option, which is a valid object member, is added to the end of a script,
+leading to a well-typed script. This distinguishes our system from auto-completion based on machine
+learning or dictionary-based methods, which may offer members not valid in a given context. Advanced
+users of The Gamma, modifying scripts as text, can still violate correctness -- for example, if the
+user uses auto-complete to replace a member in a middle of a member access chain with a member of another type.
+
+\subsubsection*{Completeness} Auto-completion lists offered via iterative prompting contain
+all available members and so the user can construct all possible scripts. The range of such scripts
+is determined by type providers, which support limited set of options (e.g. group aggregations in
+the tabular data type provider). Two exceptions to completeness in our current design is specifying
+numerical parameters as in \ikvd{take(5)} and advanced language features such as let binding.
+We believe that alleviating the need for those poses an interesting further research challenge.
+
 \section{Discussion}

### 2021-03-23 wip-design/paper.tex
Simplifying case study discussion

@@
   %
-  We present The Gamma, a novel programmatic data exploration environment.
+  We present The Gamma, a novel text-based data exploration environment.
   The Gamma is based on a single interaction principle and using it results in transparent and
@@
 
-We contribute The Gamma, a programmatic data exploration environment for non-experts. The Gamma
+We contribute The Gamma, a text-based data exploration environment for non-experts. The Gamma
 is based on a single, easy to understand interaction principle and provides a uniform
@@
 \begin{figure}[b]
-\includegraphics[width=1\columnwidth]{figures/thegamma-walk}
+\includegraphics[width=0.94\columnwidth]{figures/thegamma-walk}
 \caption{Constructing a script that charts the top 5 members of the House of Lords for Kent, based
@@
 
-The review of related work suggests that there is an unexplored point in the design space of tools
-for data exploration. Although various efforts make text-based programming easier, e.g. by providing
-high-level declarative abstractions, most systems that target non-experts shy away from code, by using
-either programming by example, natural language or graphical user interface. However, text-based
-programming has notable benefits such as learnability and transparency. Our work is primarily
-motivated by the question whether we can make text-based programming easy enough for non-experts?
-The study presented in Section~\ref{sec:study} shows that this is, indeed, possible at least for typical data querying tasks.
-The Gamma consists of a programming language, a web-based coding environment and a number of type providers
-that enable access to various kinds of data sources. It allows non-experts to create programs using the
-\emph{iterative prompting} interaction principle -- by repeatedly selecting an item from an auto-complete list.
-%
-We start with a walkthrough of the system (Section~\ref{sec:overview-walk}), before looking at details of
-The Gamma language (Section~\ref{sec:overview-lang}) and type providers (Section~\ref{sec:overview-tps}).
-
+The Gamma aims to make text-based data exploration easy enough for non-experts. The aim is
+motivated, in part, by the desirable properties of text-based data exploration tools such
+as transparency, reproducibility and learnability and, in part, by an aim to explore an
+unexplored point in the design space of data exploration tools. Although various efforts
+make text-based programming easier, most systems that target non-experts shy away from code.
+
+The Gamma is a text-based data exploration environment that allows non-experts explore data using
+iterative prompting -- by repeatedly selecting an item from an auto-complete list. The study
+presented in Section~\ref{sec:study} confirms that the kind of data exploration shown in the next
+section can, indeed, be successfully done by non-experts.
+%
 \subsection{Querying Travel Expenses}
@@
 
-To introduce The Gamma, we walk through a simple scenario.
-A journalist from Kent is exploring travel expense claims by members of the House of Lords published
-by the UK government. The data is available as a CSV file\footnote{Full data set has been obtained from
-\url{https://www.parliament.uk/mps-lords-and-offices/members-allowances/house-of-lords/holallowances/} }:
+The walkthough in Figure~\ref{fig:walkthrough} shows a typical task completed using The Gamma.
+A data analyst from Kent is exploring travel expense claims by members of the House of Lords published
+by the UK government \cite{lords}. The following shows a subset of the data in the CSV format:
 
@@
 \noindent
-After the analyst imports the file through a web interface,
-the environment is initialised with code that refers to the imported data as \ikvd{expenses}
-and she starts exploring the data using the type provider for tabular data (Section~\ref{sec:overview-tps}),
-following the steps illustrated in Figure~\ref{fig:walkthrough}:
+The analyst imports the file through a web interface, the environment is initialised with code
+that refers to the imported data as \ikvd{expenses} and she starts exploring the data using the type provider for tabular data (Section~\ref{sec:overview-tps}):
 
@@
 The types are not built-in, but are generated by type providers for individual data sources.
-The programming environment for The Gamma is based on the Monaco editor\footnote{For
-more information, see \url{https://microsoft.github.io/monaco-editor/}}. When the user types `.'
+The programming environment for The Gamma is based on the Monaco editor \cite{monaco}. When the user types `.'
 the editor triggers auto-completion and retrieves a list of available members based on the type
-information. The programming environment evaluates scripts on-the-fly and shows a preview as
+information. The Gamma evaluates scripts on-the-fly and shows a preview as
 illustrated in Figure~\ref{fig:thegamma}.
@@
 
-The Gamma can be extended to support any kind of data source by implementing a \emph{type provider},
+The Gamma can be extended to support any data source by implementing a \emph{type provider},
 which defines a domain specific language for exploring data of a particular kind.
-Technically, a type provider generates object types with members (such as \ikvd{paging} or \ikvd{Kent})
+A type provider generates object types with members (such as \ikvd{paging} or \ikvd{Kent})
 that are accessed via iterative prompting. We describe type provider for exploring data cubes
-(inspired by an example from Syme et al. \cite{inforich}), tabular data (based on theory developed
-by Petricek \cite{dotdriven}), and a novel type provider for exploring graph databases.
+(inspired by Syme et al. \cite{inforich}), tabular data (based on theory developed
+by Petricek \cite{dotdriven}), and graph databases.
 
@@
 of \ikvd{expenditure} are government services, years and value type (adjusted, nominal,
-per GDP).  As illustrated in Figure~\ref{fig:cubetp}, the provider allows users to select a data series
-from the data cube. Choosing \ikvd{byCountry.\textquotesingle United States\textquotesingle},
-restricts the cube to a two-dimensional plane; selecting \ikvd{\textquotesingle CO2 emissions (kt)\textquotesingle}
+per GDP).  Figure~\ref{fig:cubetp} how the provider allows users to slice the data cube.
+Choosing \ikvd{byCountry.\textquotesingle United States\textquotesingle},
+restricts the cube to a plane and selecting \ikvd{\textquotesingle CO2 emissions (kt)\textquotesingle}
 then gives a series with years as keys and emission data as values. Similarly, we could first filter the
 data by a year or an indicator. The same mechanism is used to select UK government spending on
-defence in terms of GDP over all available years.
-As there is no widely used standard format for data cubes, adding another data cube to The Gamma
-currently requires implementing a new type provider.
+defence in terms of GDP.
+%  As there is no widely used standard format for data cubes, adding another data cube to The Gamma
+% currently requires implementing a new type provider.
 
@@
 
-Our second type provider allows users to construct queries to explore data in tabular formats
-such as CSV files. Unlike the data cube provider, the provider for tabular data does not just
+Our second type provider allows users to construct queries to explore data in tabular formats.
+Unlike the data cube provider, the provider for tabular data does not just
 allow selecting a subset of the data, but it can be used to construct SQL-like query. Consider
@@
 \noindent
-The example works with a CSV file that records individual medals awarded in Olympic games.
+The example queries a table that records individual medals awarded in Olympic games.
 The chain constructs a query that selects rows corresponding to the Rio 2016 Olympics and then
 calculates total number of gold and silver medals for each team (country) before sorting the data.
-
+%
 When using the provider, the user specifies a sequence of operations. Members such as
@@
 Unlike SQL, the provider only allows users to choose from pre-defined aggregations such as
-calculating the sum, average or the number of distinct values. As illustrated in
-Section~\ref{sec:cases}, this is sufficient to construct a wide range of practical queries.
+calculating the sum, average or the number of distinct values.
+Section~\ref{sec:cases} shows that this is sufficient to construct a range of practical queries.
 
@@
 Our third type provider allows users to explore data from graph databases, which store
-nodes representing entities and relationships between them. In the context of data journalism,
-graph databases have been used for example in The Panama Papers reporting \cite{panama}.
+nodes representing entities and relationships between them.
 The following example explores a database of Doctor Who characters and episodes. It retrieves
@@
 \begin{thegamma}
-drwho.Character.Doctor
-  .'ENEMY OF'.'[any]'.'APPEARED IN'.'Day of the Moon'
+drwho.Character.Doctor.'ENEMY OF'.'[any]'.'APPEARED IN'.'Day of the Moon'
 \end{thegamma}
@@
 
-\begin{figure}[b]
+\begin{figure}[t]
 \centering
+\begin{subfigure}[b]{0.49\textwidth}
+  \centering
+  \includegraphics[width=1\columnwidth]{figures/cases-dr}
+  \caption{Exploring Dr Who graph database by composing type providers}
+  \label{fig:cases-dr}
+\end{subfigure}
+\hfill
 \begin{subfigure}[b]{0.49\textwidth}
@@
 \end{subfigure}
-\hfill
-\begin{subfigure}[b]{0.49\textwidth}
-  \centering
-  \includegraphics[width=1\columnwidth]{figures/cases-dr}
-  \caption{Exploring Dr Who graph database by composing type providers}
-  \label{fig:cases-dr}
-\end{subfigure}
 \vspace{-0.5em}
@@
 The Gamma aims to simplify programmatic data exploration while keeping enough expressive power
-to allow users to create interesting data visualizations. We consider the expressive power in
-this section and assess simplicity in the next one. We used The Gamma to analyse
-the UK government expenditure, activities of a research institute, Olympic medals and
-information about the Doctor Who series\footnote{The analyses are available (non-anonymously)
+to allow users to create interesting data explorations. We assess the simplicity of The Gamma
+in Section~\ref{sec:study}. Here, we consider two case studies that evaluate expressivity and show what
+can be achieved using the simple iterative prompting principle\footnote{Available (non-anonymously) at:
+\url{http://gallery.thegamma.net/86/} and  \url{http://gallery.thegamma.net/87/}, respectively.}.
+We used The Gamma for larger projects that analyse the UK government expenditure, present
+activities of a research institute, analyse Olympic medals and explore information about the Doctor
+Who series\footnote{The analyses are available (non-anonymously)
 at \url{http://gallery.thegamma.net}, \url{http://turing.thegamma.net} and \url{http://rio2016.thegamma.net}}.
-In this section, we consider two interesting larger examples\footnote{Available
-(non-anonymously) at: \url{http://gallery.thegamma.net/86/} and  \url{http://gallery.thegamma.net/87/}, respectively.}.
-
-\subsubsection*{If Michael Phelps were a Country}
-Michael Phelps has won so many medals that media compared the number to
-countries.\footnote{e.g.~``If Michael Phelps Were A Country, Where Would His Gold Medal Tally Rank?''
- \url{https://www.npr.org/sections/thetorch/2016/08/14/489832779/}}
-Media illustrated the story with a chart showing a country league table including Michael Phelps.
-We reproduce the chart, shown in Figure~\ref{fig:cases-mp}, using the tabular data type provider:
-
-\begin{thegamma}
-\kvd{let} data = olympics.'group data'.'by Team'.'sum Gold'.then
-    .'sort data'.'by Gold descending'.then
-    .paging.skip(43).take(4).'get series'.'with key Team'.'and value Gold'
-
-\kvd{let} phelps = olympics.'filter data'.'Athlete is'.'Michael Phelps'.then
-  .'group data'.'by Athlete'.'sum Gold'.then
-  .'get series'.'with key Athlete'.'and value Gold'
-
-charts.bar(data.append(phelps)).setColors(["#aec7e8","#aec7e8","#1f77b4"])
-\end{thegamma}
-
-\noindent
-The data analysis is done in two commands. The first counts gold medals by countries
-and uses paging to fetch 4 countries with suitable number of medals. In the second, we abuse the
-grouping operation to aggregate data for just a single group. The two data series are then assigned
-to local variables (for readability) and passed to the \ikvd{chart.columns} function.
-The case study illustrates when more advanced language features are necessary. The data exploration
-itself has been completed via iterative prompting, but producing the final chart currently requires
-some manual programming. In practice, this would likely be done with the help of an expert
-or by copying code from an example.
 
 \subsubsection*{The Most Frequent Doctor Who Villains}
-In the second case study, use a graph database on the Dr Who series and list Dr Who villains
+Our first case study uses a graph database with data from the Dr Who series. It lists Dr Who villains
 by the number of episodes in which they appear. This case study is interesting as it
@@
 \noindent
-Line 1 use the graph database provider to find all paths linking the Doctor node with any
+Line 1 use the graph provider to find all paths linking the Doctor with any
 character linked via \ikvd{ENEMY OF}, followed by any episode linked by \ikvd{APPEARED IN}.
-This produces a table that can be analysed using the tabular data provider by choosing the
-\ikvd{explore} member. For each character (the villain) we count the total number of
-distinct episodes in the table. The result is shown in Figure~\ref{fig:cases-dr}.
-The example performs a fairly sophisticated data analysis that involves a graph database query
-followed by an SQL-like data aggregation. The code can be fully constructed using iterative
-prompting (with the exception of the numbers in paging). Another important feature is that
-instantaneous previews support the user during the construction, showing the intermediate result
-at each step of the construction.
+This produces a table that can be analysed using the tabular data provider by selecting
+\ikvd{explore}. For each character (the villain) we count the number of
+distinct episodes. The result is shown in Figure~\ref{fig:cases-dr}.
+Despite performing a sophisticated data analysis that involves a graph database query,
+followed by an SQL-like data aggregation, the code can be constructed using iterative
+prompting, with the exception of the numbers in paging.
+
+\subsubsection*{If Michael Phelps were a Country}
+Michael Phelps has won so many medals that media compared the number to
+countries~\cite{phelps}, often using a chart that shows a country league table including Michael Phelps
+as an additional data point.
+We reproduce the chart, shown in Figure~\ref{fig:cases-mp}, using the tabular data type provider:
+%
+\begin{thegamma}
+\kvd{let} data = olympics.'group data'.'by Team'.'sum Gold'.then
+    .'sort data'.'by Gold descending'.then
+    .paging.skip(43).take(4).'get series'.'with key Team'.'and value Gold'
+\vspace{5em}
+\kvd{let} phelps = olympics.'filter data'.'Athlete is'.'Michael Phelps'.then
+  .'group data'.'by Athlete'.'sum Gold'.then
+  .'get series'.'with key Athlete'.'and value Gold'
+
+charts.bar(data.append(phelps)).setColors(["#aec7e8","#aec7e8","#1f77b4"])
+\end{thegamma}
+%
+The data analysis is done in three commands. The first counts gold medals by countries
+and uses paging to fetch 4 countries with suitable number of medals. In the second, we use the
+grouping operation to aggregate data for just a single group. The two data series are then assigned
+to local variables (for readability) and passed to the \ikvd{chart.columns} function.
+The example illustrates a case when more advanced language features are necessary. The data exploration
+itself has been completed via iterative prompting, but producing the final chart currently requires
+some manual programming. %In practice, this would likely be done with the help of an expert
+%or by copying code from an example.
 
@@
 
-Our goal was to develop an easy-to-learn tool that non-programmers can use
-for producing transparent data analyses. To evaluate the extent to which The Gamma achieves this,
-we conducted a user study. We recruited participants among the operations and business team of a
-non-profit research organization and investigate three research questions.
-
-\vspace{0.5em}\noindent\emph{RQ1: Can non-programmers use The Gamma?}\hspace{0.3em}
-Our first hypothesis is that non-programmers will be able to use The Gamma to explore data. To test
-this, we gave participants one of four simple data exploration tasks and assess whether they were
-able to complete the task, as well as how much assistance, if any, they needed.
-
-\vspace{0.5em}\noindent\emph{RQ2: Can knowledge transfer between data sources?}\hspace{0.3em}
-Our second hypothesis is that users familiar with the iterative prompting principle will be able to
-use an unfamiliar data source. We designed two of the four tasks to test this. Participants were
-shown a demo using one data source and then asked to complete a task using a different one.
-
-\vspace{0.5em}\noindent\emph{RQ3: Can users learn from just code samples?}\hspace{0.3em}
-Our third hypothesis is that users can learn through percolation, i.e.~by looking at the source
-code of published analyses. In one of our tasks, participants were given a minimal explanation
-of The Gamma (showing how to initiate the iterative prompting process) together with
-an extensive code sample.
+Our objective is to develop a text-based data exploration tool that can be used by non-programmers.
+Using the characterization by Olsen~\cite{evaluating}, data exploration environments are complex
+systems that do not yield to simple controlled experimentation. We discuss the limits of evaluating
+The Gamma in Section~\ref{sec:discussion-eval} and do not attempt to make make quantitative comparison
+e.g.~with SQL or Tabelau~\cite{tableau}. Consequently, our goals are more modest.
 
 \subsection{Study Design}
+We evaluate the extent to which non-programmers can use The Gamma to conduct
+text-based data exploration (\emph{RQ1}). To test this, we gave volunteers one of four data
+exploration tasks and assessed whether they were able to complete the task, as well as how much
+assistance, if any, they needed.
+%
+Some aspects of the study aim to shed light on three questions concerning learnability
+of The Gamma, in particular whether knowledge can be trasferred between different data sources
+(\emph{RQ2a}), whether users can learn from code samples without a detailed guidance
+(\emph{RQ2b}) and to what extent they form a correct mental model of a more complex query
+language used in the tabular data source (\emph{RQ2c}).
+
 We performed a between-subjects study to evaluate the first experience of using The Gamma.
@@
 brief overview of The Gamma (with content depending on the task) and then asked participants
-to complete a task. The four tasks were:
+to complete a task. The participants worked on the tasks for 30 minutes, after which we conducted
+a 30 minute semi-structured group interview. We let participants work independently, but offered
+guidance if they got stuck and asked for help. The four tasks were:
+% The participants were volunteers who responded to an invitation sent to an internal mailing list.
 
@@
 
+%\vspace{0.5em}\noindent\emph{RQ2: Can knowledge transfer between data sources?}\hspace{0.3em}
+%Our second hypothesis is that users familiar with the iterative prompting principle will be able to
+%use an unfamiliar data source. We designed two of the four tasks to test this. Participants were
+%shown a demo using one data source and then asked to complete a task using a different one.
+%
+%\vspace{0.5em}\noindent\emph{RQ3: Can users learn from just code samples?}\hspace{0.3em}
+%Our third hypothesis is that users can learn through percolation, i.e.~by looking at the source
+%code of published analyses. In one of our tasks, participants were given a minimal explanation
+%of The Gamma (showing how to initiate the iterative prompting process) together with
+%an extensive code sample.
+%
+
 \noindent
-We let participants work independently, but offered guidance if they got stuck.
-Tasks \emph{expenditure} and \emph{lords} aim to answer the questions RQ1 and RQ2; the task
-\emph{worldbank} aims to answer RQ1 and RQ3. In \emph{olympics}, we test RQ1 using a more complex
-data source and we ask further questions to explore the understanding of the data source. Following the
-experiment, we conducted a short semi-structured group interview and later sent participants a follow-up
-questionnaire.
-
-\begin{figure}
-  \begin{subfigure}[b]{0.49\textwidth}
+All of the tasks aim to answer the primary question RQ1 for one of the supported data sources.
+In addition, the tasks \emph{expenditure} and \emph{lords} study RQ2a because they use different
+data source in introduction and taks; \emph{worldbank} study RQ2b in that it comes with only brief
+introduction and \emph{olumpics} studies RQ2c using our most complex data source.
+
+\begin{table}
 \centering
@@
   \midrule
-      \small \#1 & \small expenditure & \small cube & \priority{50} & {\small Obtained one data series}\\
-      \small \#2 & \small expenditure & \small cube & \priority{100} & {\small Explored further data }\\
-      \small \#3 & \small expenditure & \small cube & \priority{100}& {\small Explored further data  }\\
-      \small \#4 & \small expenditure & \small cube & \priority{75}& {\small Correct member hint }\\
-      \small \#5 & \small expenditure & \small cube & \priority{100}& {\small Explored further data }\\
+  \small \#1\qquad\qquad & \small expenditure \qquad\qquad & \small cube \qquad\qquad & \qquad\priority{50}\qquad\qquad & {\small Obtained one of two data series}\\
+  \small \#2 & \small expenditure & \small cube & \priority{100} & {\small Explored furhter data series }\\
+  \small \#3 & \small expenditure & \small cube & \priority{100}& {\small Explored further data series }\\
+  \small \#4 & \small expenditure & \small cube & \priority{75}& {\small With hint to use another member }\\
+  \small \#5 & \small expenditure & \small cube & \priority{100}& {\small Explored further data series }\\
   \small \#6 & \small worldbank & \small cube & \priority{75} & {\small With general syntax hint }\\
@@
   \small \#8 & \small worldbank & \small cube & \priority{100} & {\small Extra time to find data }\\
-      \small \#9 & \small lords & \small table & \priority{75} & {\small  Composition issues}\\
+  \small \#9 & \small lords & \small table & \priority{75} & {\small  Struggled with composition}\\
   \small \#10 & \small lords & \small table & \priority{100} & {\small Completed very quickly }\\
-      \small \#11 & \small lords & \small table & \priority{75} & {\small Hint to avoid operations}\\
-      \small \#12 & \small olympics & \small table & \priority{75}  & {\small Hint to avoid operations}\\
-      \small \#13 & \small olympics & \small table & \priority{75}  & {\small Two operations hints}\\
+  \small \#11 & \small lords & \small table & \priority{75} & {\small With a hint to avoid operations}\\
+  \small \#12 & \small olympics & \small table & \priority{75}  & {\small With a hint to avoid operations}\\
+  \small \#13 & \small olympics & \small table & \priority{75}  & {\small Hints about `then' and operations}\\
   \bottomrule
 \end{tabular}
-    \caption{Work completed by individual participants; \priorityc{100} = completed,
-      \priorityc{75} = required some guidance, \priorityc{50} = partially completed}
+
+\caption{\centering Overview of work completed by individual participants in the study. \\The marks denote:
+ \priorityc{100} = completed, \priorityc{75} = required some guidance, \priorityc{50} = partially completed}
 \label{tab:tasks}
-  \end{subfigure}
-  \hfill
-  \begin{subfigure}[b]{0.49\textwidth}
-    \centering
-    \begin{tabular}{p{17.25em} c c}
-      \toprule
-        {\small \textit{Questionnaire item}} & {\small \textit{Avg}} & {\small \textit{Sdv}} \\
-      \midrule
-      \small Considering your experience and possible use of The Gamma in the context of data journalism:\\
-      \small \quad I found the system easy to use. & \small 4.00 & \small 0.89\\
-      \small \quad Journalists will be able to use it to analyse data. & \small 3.45 & \small 1.04\\
-      \small \quad Readers will be able to examine the analyses. & \small 3.73 & \small 0.79\\
-      \small Thinking about the usability of the system:\\
-      \small \quad Resulting programs are easy to understand. & \small 3.73 & \small 0.79\\
-      \small \quad Text editor for creating them was easy to use. & \small 3.73 & \small 0.90\\
-      \small \quad Finding items in the auto-complete was easy. & \small 3.45 & \small 1.37\\
-      \small I would know how to explore a new data source\\
-      \small \quad If given a comprehensive video tutorial. & \small 4.00 & \small 1.26\\
-      \small \quad If given a number of code samples. & \small 3.36 & \small 1.12\\
-      \small \quad Without any further guidance. & \small 2.45 & \small 1.13 \vspace{0.1em}\\
-      \bottomrule
-    \end{tabular}
-    \caption{Follow-up questionnaire responses using
-      a 5-point Likert scale (1 = strongly disagree, 5 = strongly agree) over 11 subjects.}
-    \label{tab:quest}
-  \end{subfigure}
-  \vspace{-0.5em}
-  \caption{Summary of the work completed by individual participants and responses to the follow-up questionnaire.}
-  \label{tab:tabs}
-  \vspace{-0.25em}
-\end{figure}
+\vspace{-2em}
+\end{table}
 
 \subsection{Study Results}
-Figure~\ref{tab:tabs} shows the results of the study. Most notably, all participants were able to
-complete, at least partially, a non-trivial data exploration task and only half of them required
-further guidance. Table~\ref{tab:tasks} summarizes the work done by the study participants.
-For each participant, we record the task, the kind of data source used in the task and the
-level of completion. For participants who needed assistance, the notes section details the help
-given. We discuss possible design improvements based on the experience in Section~\ref{sec:study-obs}.
-
-The follow-up questionnaire was sent to all participants a week after the experiment. It was
-completed all but two participants (\#8 and \#12). We asked general questions about the usability
-of the system together with questions about learning designed to answer RQ2 and RQ3. A summary
-of the answers is shown in Table~\ref{tab:quest}.
-
-\subsection*{RQ1: Can non-programmers explore data with The Gamma?}
+Table~\ref{tab:tasks} summarizes the work done by the study participants. Most notably, all
+participants were able to complete, at least partially, a non-trivial data exploration task and
+only half of them required further guidance. For each participant, we record the task, the kind
+of data source used in the task and the level of completion. For participants who needed assistance,
+the notes section details the help given. The experience suggests a number of possible design
+improvements for The Gamma, which are discussed below.
+
+\subsubsection*{Can Non-programmers Explore Data with The Gamma?}
 Three facts allow us to answer RQ1 in the affirmative. First, 9 out of 11 participants agree or
@@
 Third, a number of participants shared positive comments in the interviews.
-Participant \#3 found the system simple, but also points out an issue about data provenance,
-which we revisit later:
-
-\begin{quote}
-\emph{``This is actually pretty simple to use. You think about the logic of what you're actually
-  asking and then you try to get it into the format you can. But knowing where it comes from
-  would tell you how to trust it.''}
-\end{quote}
-
-\noindent
-Similarly, participant \#2, notes that The Gamma alleviated their unease about code:
-
-\begin{quote}
+For example, participant \#3 noted taht \emph{``this is actually pretty simple to use.
+You think about the logic of what you're actually asking and then you try to get it into the format you can.''}
+Participant \#2 noted that The Gamma alleviated their unease about code:
 \emph{``For somebody who does not do coding or programming, this does not feel that daunting.
   It's not like you're giving me large screen full of code, which is reassuring.''}
-\end{quote}
-
-\noindent
-Finally, participant \#5 suggested the system could be used as an educational tool for teaching
-critical thinking with data. They answer a follow-up question about what training materials would
-the students need as follows:
-
-\begin{quote}
-\emph{``I don't think they'd need more than 5 minute video (..) this is the data
-  source, this is what's in there.''}
-\end{quote}
-
-\subsection*{RQ2: Can knowledge transfer between data sources?}
-Our study does not conclusively answer RQ2. There is some evidence in favor of a positive answer.
-In the practical part, two of the tasks (\emph{expenditure} and \emph{lords}) used a different data
-source in the introductory presentation than the one that the participants were asked to explore.
-Participants were able to complete those tasks, although \emph{lords} has been more challenging
-as it involves a more complex data source. In the interview, participant \#2 also gives a positive
-answer:
-
-\begin{quote}
-\emph{``I found it quite easy to translate what you showed us in the demo to the new dataset.
-   I thought it was quite easy to just get how that works.''}
-\end{quote}
-
-\noindent
-Negative evidence is offered by the follow-up questionnaire. When asked
-whether they would know how to approach a task using a new data source, 6 out of 11 participants
-disagree or strongly disagree that they would know how to approach it without any further guidance.
-The nature of RQ2 makes it a more challenging question to study and so finding a conclusive answer
-arguably requires further research.
-
-\subsection*{RQ3: Can users learn from just code samples?}
-A positive answer to RQ3 is supported by the \emph{worldbank} task results, the follow-up questionnaire
-and interview comments. In the task, participants were given only a
-minimal demo of the iterative prompting principle together with print-out of 2 code samples.
-All three were able to complete a related task using the same data source. In the follow-up
-questionnaire, only 2 out of 11 participants disagree or strongly disagree that
-they would know how to approach a task using an unfamiliar data source when given ``a number
-of code samples''.
-
-In the semi-structured interview, participants were asked what would be the most useful format
-of educational material about The Gamma (code samples, video tutorials, etc.).
-Participant \#7 noted that \emph{``a video would just be this [i.e.~a code sample] anyway''}, while
-participant \#13 (former journalist) answered:
-
-\begin{quote}
-  \emph{``I think providing one video of one example is good and maybe a couple of little examples
-  of code where people can see the kind of things you can do.''}
-\end{quote}
-
-\noindent
-This is aligned with our design goal. Once the user understands the iterative prompting principle
-(which can be done through a video tutorial), they can learn how to use any specific data source
-just from code samples.
-
-\subsection{Further Observations}
-\label{sec:study-obs}
-In this section, we briefly discuss a number of observations about The Gamma design that
-emerged from the experiments and follow-up interviews, some of which suggest ways of improving the system.
+
+% \noindent
+% Finally, participant \#5 suggested the system could be used as an educational tool for teaching
+% critical thinking with data. They answer a follow-up question about what training materials would
+% the students need as follows:
+% \begin{quote}
+% \emph{``I don't think they'd need more than 5 minute video (..) this is the data
+%   source, this is what's in there.''}
+% \end{quote}
+
+\subsubsection*{How Users Learn The Gamma?}
+The nature of RQ2 makes it a challenging question to study and finding conclusive answers
+may require further research. There is some evidence that knowledge can be transferred between
+different data sources (RQ2a). Two of the tasks (\emph{expenditure} and \emph{lords}) used
+one data source in the introduction and another for the task. Participants were able to complete
+those, although \emph{lords} has been more challenging as it involves a complex data source.
+Participant \#2 also gives a positive answer \emph{``I found it quite easy to translate what you
+showed us in the demo to the new dataset.''}.
+
+There is also some evidence that users can learn just from code samples (RQ2b). In the
+\emph{worldbank} task, participants were given only a minimal demo of how to invoke iterative
+prompting together with print-out of 2 code samples. All three participants were able to complete
+a related task using the same data source. When discussing suitable educational materials for
+The Gamma, participant \#7 also confirmed that having code is sufficient when they noted that
+\emph{``a video would just be this [i.e.~a code sample] anyway''}. This supports our hypothesis
+that, once a user understands the iterative prompting principle, they can learn how to use any
+specific data source just from code samples.
+
+\subsubsection*{How Users Understand Complex Query Languages?}
+Although iterative prompting itself is simple, specific data sources may
+expose richer structure. To ascertain their understanding of the tabular type provider (RQ2c), we
+asked participants who worked with tabular data (\emph{lords} or \emph{olympics}) about their
+understanding of the  \ikvd{then} member. This is a regular member, but it has a
+special meaning in the domain specific language. For example, when specifying a list of
+aggregations, the user can repeatedly select further aggregations. Selecting \ikvd{then} completes
+the operation and allows the user to choose other operations such as sorting.
+
+Two participants (\#12 and \#13) initially thought that \ikvd{then} is used to split a command
+over multiple lines, but rejected the idea after experimenting.
+Participant \#12 then correctly concluded that it \emph{``allows us
+to chain together the operations''} of the query; after a hint, participant \#13 reflected
+that \emph{``if I knew this from the start, it would [have been easier].''} This summarizes an
+important fact. Iterative prompting allows the users to start exploring new data sources, but
+the structures exposed by more complex data sources have their own further design principles that
+the users need to understand.
 
@@
 \ikvd{olympics.\textquotesingle filter data\textquotesingle.\textquotesingle Year is greater than\textquotesingle(2004)}.
-
+%
 Three participants (\#11, \#12, \#13) struggled to complete a task, because they
-initially attempted to use those operations. They violate the iterative prompting
+initially attempted to use those operations. However, those operations violate the iterative prompting
 principle as one cannot type `.' after \ikvd{\textquotesingle Year is greater than\textquotesingle}.
@@
 
-\subsubsection*{Understanding the `then' Member}
-We asked participants who worked with tabular data about their understanding of the \ikvd{then} member.
-This is a regular member (not a keyword), but it has a special meaning in the domain specific language.
-Consider the calculation of average travel costs and number of representatives per county in the House of Lords:
-
-\begin{thegamma}
-expenses.'group data'.'by County'.'average Travel Costs'.'count distinct Name'.then
-\end{thegamma}
-
-\noindent
-The \ikvd{then} member is used to complete a part of a query where the user repeatedly
-add items to build a list. Here, we select two aggregations to be calculated for each
-county before choosing \ikvd{then} and applying other operations such as sorting.
-Two participants (\#12 and \#13) thought that \ikvd{then} is used to split a command
-over multiple lines, but rejected the idea after experimenting and noting that they can insert
-line breaks elsewhere. One correctly concluded that it ``allows us to chain together the
-operations'' of the query. Following a hint, participant \#13 reflected:
-
-\begin{quote}
-  \emph{``When you explained about the `dot then' that was a really useful thing to know.
-  When I found that, I was like this is fine, this is doable. If I knew this from the start,
-  it would [have been easier].''}
-\end{quote}
-
-\noindent
-This comment summarizes an important fact. Although iterative prompting allows the users to start
-exploring new data sources, the domain specific languages used by more complex data sources have
-their own design principles that users need to understand to use the data source effectively.
-
 % ==================================================================================================
@@
 
-\section{Reflections on Design}
+\section{Design Reflections}
 \label{sec:design}
@@
 \subsection{Evaluating Complex Systems}
+\label{sec:discussion-eval}
 Data exploration environments are complex systems that do not yield to simple controlled

### 2021-04-06 wip-design/paper.tex
Finish cleanup for resubmission

@@
 \begin{document}
-\title{The Gamma: Data Exploration through Iterative Prompting}
+\title{Iterative Prompting: Programmatic Data Exploration for Non-programmers}
 
@@
   a wide range of different data sources, result in reproducible scripts and encourage users to
-  reuse and modify existing data analyses. Unfortunately, most programming tools require expert
+  reuse and modify existing code. Unfortunately, most programming tools require expert
   coding skills. Can we make data exploration based on code accessible to non-experts?
   %
-  We present The Gamma, a novel text-based data exploration environment.
-  The Gamma is based on a single interaction principle and using it results in transparent and
-  reproducible scripts. This lowers the barrier to entry and allows learning from previously
-  created data analyses. We evaluate the usability and learnability of The Gamma through a user
-  study on non-technical employees of a research institute.
+  We present The Gamma, a novel text-based data exploration environment that answers the question in
+  the affirmative. The Gamma is based on a novel interaction principle, \emph{iterative prompting},
+  which lets users create transparent and reproducible scripts without writing code. The Gamma lowers
+  the barrier to entry and learning from previously created data analyses. We evaluate the usability
+  of The Gamma through a user study on non-technical employees of a research institute.
   %
@@
 \begin{figure}
-\includegraphics[width=.58\columnwidth]{figures/thegamma-annot}
+\includegraphics[width=.63\columnwidth]{figures/thegamma-annot}
 \vspace{-0.5em}
@@
 \section{Introduction}
-Data science has more capabilities to help us understand the world than ever before, yet at the
-same time post-truth politics and increasing public distrust in statistics makes data-driven insights
-increasingly less relevant in public discourse~\cite{howstatslost}. To reverse this trend, we
-need tools that let non-experts, including journalists and other information-literate citizens,
-produce transparent, engaging data analyses that are easy to interpret without requiring expert
-programming skill~\cite{ddj}. The design of such data exploration tool poses a unique mix of challenges.
-First, the tool needs to have a very low barrier to entry. Second, it needs to support a wide
-range of data sources in a uniform way. Third, the resulting data analyses need to assist
-readers in learning how to reproduce the work and verify the claims it makes.
+Despite the advances on visual tooling, programmatic data exploration remains the choice of
+expert analysts. It is flexibile, offers greater reusability and leads to transparent
+analyses. We aim to make programmatic data exploration accessible to a wider range of users.
+The design of a data exploration tool that would make this possible poses a number of challenges.
+First, the tool needs to have a low barrier to entry. Second, it needs to support a wide
+range of data sources in a uniform way. Third, the users should be able to learn how to use the
+tool by looking at existing data analyses.
 
@@
 which makes it possible to construct all valid data exploration scripts by repeatedly choosing
-from a list of offered options. This way, non-programmers can write entire scripts through
-auto-complete without learning a programming language first, but still produce transparent
-and reproducible code. In other words, iterative prompting turns auto-complete from a programmer
+an item from a list of options offered through auto-complete. The design favors \emph{recognition
+over recall} and allows non-programmers to write entire scripts without typing code and without
+learning a programming language first. Yet, it still results in transparent and reproducible code.
+In other words, iterative prompting turns auto-complete from a programmer
 assistance tool into a non-expert programming mechanism. A crucial feature is that iterative
-prompting only offer operations that are valid in a given context and that it offer all such
+prompting only offers operations that are valid in a given context and that it offers all such
 operations; it is both correct and complete.
@@
 cleaning of messy data or building custom data visualizations is currently outside of the scope of
-our work. Exposing those using iterative prompting remains an interesting future challenge.
+our work. Exposing those using iterative prompting remains an interesting and worthwhile future
+challenge.
 
@@
 both as a JavaScript library and a hosted data exploration service. In this paper, we describe and
-evaluate the design principles behind the project\footnote{This paper is X pages, without citations,
-to be commensurate with the size of the contribution.}:
+evaluate the design principles behind the project\footnote{This paper is 11 pages (5500 words),
+without citations, to be commensurate with the size of the contribution.}:
 
@@
 
-\item We reflect how our design lowers barriers to entry (Section~\ref{sec:design-bar}),
-  supports learning without experts (Section~\ref{sec:design-expert}) and offers a
-  complete and correct program construction method (Section~\ref{sec:design-cc}).
-
 \item We evaluate the system through a number of case studies (Section~\ref{sec:cases})
@@
   to construct non-trivial data queries.
+
+\item We reflect how our design lowers barriers to entry, supports learning without experts and offers a
+  complete and correct program construction method (Section~\ref{sec:discuss}).
 \end{itemize}
@@
 to follow links to all enemies of the Doctor and then specify
-\ikvd{\textquotesingle APPEARED IN\textquotesingle.\textquotesingle Day of the Moon\textquotesingle}
-to select only enemies that appear in a specific episode. The resulting query
-is illustrated in Figure~\ref{fig:graphtp}.
-
+\ikvd{\textquotesingle APPEARED IN\textquotesingle}
+to select only enemies that appear in a specific episode. The result appears in
+in Figure~\ref{fig:graphtp}.
+%
 The provider works with any graph database and generates members automatically, based on the
-data in the database. In the above example, \ikvd{ENEMY OF} and \ikvd{APPEARED IN} are labels
+data. In the above, \ikvd{ENEMY OF} and \ikvd{APPEARED IN} are labels
 of relations and \ikvd{Doctor} and \ikvd{Day of the Moon} are labels of nodes. The
@@
 \begin{figure}[t]
+\vspace{-0.5em}
 \centering
@@
   \centering
-  \includegraphics[width=1\columnwidth]{figures/cases-dr}
+  \includegraphics[width=0.96\columnwidth]{figures/cases-dr}
   \caption{Exploring Dr Who graph database by composing type providers}
@@
   \centering
-  \includegraphics[width=1\columnwidth]{figures/cases-mp}
+  \includegraphics[width=0.96\columnwidth]{figures/cases-mp}
   \caption{Exploring Olympic medallists using tabular data type provider}
@@
 \label{fig:cases}
+\vspace{-0.5em}
 \end{figure}
@@
 The Gamma aims to simplify programmatic data exploration while keeping enough expressive power
-to allow users to create interesting data explorations. We assess the simplicity of The Gamma
-in Section~\ref{sec:study}. Here, we consider two case studies that evaluate expressivity and show what
-can be achieved using the simple iterative prompting principle\footnote{Available (non-anonymously) at:
+to allow users to create interesting data explorations. In this section, we consider two case
+studies that evaluate expressivity and show what can be achieved using the simple
+iterative prompting principle\footnote{Available (non-anonymously) at:
 \url{http://gallery.thegamma.net/86/} and  \url{http://gallery.thegamma.net/87/}, respectively.}.
-We used The Gamma for larger projects that analyse the UK government expenditure, present
-activities of a research institute, analyse Olympic medals and explore information about the Doctor
-Who series\footnote{The analyses are available (non-anonymously)
-at \url{http://gallery.thegamma.net}, \url{http://turing.thegamma.net} and \url{http://rio2016.thegamma.net}}.
+We used The Gamma for larger projects exploring the UK government expenditure, activities of a
+research institute adn Olympic medal winners\footnote{Available (non-anonymously)
+at \url{http://turing.thegamma.net} and \url{http://rio2016.thegamma.net}}.
 
@@
 
-Our objective is to develop a text-based data exploration tool that can be used by non-programmers.
 Using the characterization by Olsen~\cite{evaluating}, data exploration environments are complex
-systems that do not yield to simple controlled experimentation. We discuss the limits of evaluating
-The Gamma in Section~\ref{sec:discussion-eval} and do not attempt to make make quantitative comparison
-e.g.~with SQL or Tabelau~\cite{tableau}. Consequently, our goals are more modest.
+systems that do not yield to simple controlled experimentation. Consequently, our goals are modest and
+we do not attempt to quantitatively compare our work with other tools. We merely aim to study whether
+The Gamma can be successfully used by non-programmers.
+
+We conducted a study in which we gave volunteers one of four data exploration tasks and assessed
+whether they were able to complete the task and how much assistance, if any, they needed.
+Some aspects of the study offer insights into how users learn and understand The Gamma. We summarize
+those points, but do not claim conclusive results.
 
 \subsection{Study Design}
-We evaluate the extent to which non-programmers can use The Gamma to conduct
-text-based data exploration (\emph{RQ1}). To test this, we gave volunteers one of four data
-exploration tasks and assessed whether they were able to complete the task, as well as how much
-assistance, if any, they needed.
-%
-Some aspects of the study aim to shed light on three questions concerning learnability
-of The Gamma, in particular whether knowledge can be trasferred between different data sources
-(\emph{RQ2a}), whether users can learn from code samples without a detailed guidance
-(\emph{RQ2b}) and to what extent they form a correct mental model of a more complex query
-language used in the tabular data source (\emph{RQ2c}).
 
@@
 We recruited 13 participants (5 male, 8 female) from a business team
-of a research lab working in non-technical roles (project management,
-partnerships, communications) including one former journalist.  Only one participant (\#12)
-had prior programming experience. We split participants into 4 groups. We gave participants a
-brief overview of The Gamma (with content depending on the task) and then asked participants
-to complete a task. The participants worked on the tasks for 30 minutes, after which we conducted
+of a research institute working in non-technical roles (project management,
+partnerships, communications). Only one participant (\#12) had prior programming experience.
+We split participants into 4 groups and asked each group to complete a different task.
+We gave participants a brief overview of The Gamma (content depended on the task).
+The participants worked for 30 minutes, after which we conducted
 a 30 minute semi-structured group interview. We let participants work independently, but offered
-guidance if they got stuck and asked for help. The four tasks were:
+guidance if they got stuck. The four tasks were:
 % The participants were volunteers who responded to an invitation sent to an internal mailing list.
@@
   the \emph{worldbank} data source.
-\item \emph{Olympics.} Participants were given a demo using \emph{olympics}.
+\item \emph{Olympics.} Participants were given a basic demo using \emph{olympics}.
   They were asked to solve a more complex problem, involving grouping and aggregation,
@@
 \end{itemize}
+%
+%Some aspects of the study aim to shed light on three questions concerning learnability
+%of The Gamma, in particular whether knowledge can be trasferred between different data sources
+%(\emph{RQ2a}), whether users can learn from code samples without a detailed guidance
+%(\emph{RQ2b}) and to what extent they form a correct mental model of a more complex query
+%language used in the tabular data source (\emph{RQ2c}).
+%
 
@@
 \noindent
-All of the tasks aim to answer the primary question RQ1 for one of the supported data sources.
-In addition, the tasks \emph{expenditure} and \emph{lords} study RQ2a because they use different
-data source in introduction and taks; \emph{worldbank} study RQ2b in that it comes with only brief
-introduction and \emph{olumpics} studies RQ2c using our most complex data source.
+Our primary hypothesis was that non-programmers will be able to use The Gamma to explore
+data. This was tested by all four tasks for one of the supported data sources. Some aspects of
+the study shed light on questions concerning the learnability of The Gamma. The tasks
+\emph{expenditure} and \emph{lords} test if knowledge can be transferred between different
+data sources by using one sources in the introduction and another in the task. The task
+\emph{worldbank} explores whether users can learn from just code samples by providing only minimal
+upfront explanation. The task \emph{lords} lets us study to what extent participants form a
+correct mental model of the more complex query language used in the tabular data source.
 
@@
   \small \#1\qquad\qquad & \small expenditure \qquad\qquad & \small cube \qquad\qquad & \qquad\priority{50}\qquad\qquad & {\small Obtained one of two data series}\\
-  \small \#2 & \small expenditure & \small cube & \priority{100} & {\small Explored furhter data series }\\
-  \small \#3 & \small expenditure & \small cube & \priority{100}& {\small Explored further data series }\\
-  \small \#4 & \small expenditure & \small cube & \priority{75}& {\small With hint to use another member }\\
-  \small \#5 & \small expenditure & \small cube & \priority{100}& {\small Explored further data series }\\
-  \small \#6 & \small worldbank & \small cube & \priority{75} & {\small With general syntax hint }\\
+  \small \#2 & \small expenditure & \small cube & \priority{100} & {\small Explored furhter data series independently}\\
+  \small \#3 & \small expenditure & \small cube & \priority{100}& {\small Explored further data series independently}\\
+  \small \#4 & \small expenditure & \small cube & \priority{75}& {\small Completed following a hint to use another member }\\
+  \small \#5 & \small expenditure & \small cube & \priority{100}& {\small Explored further data series independently}\\
+  \small \#6 & \small worldbank & \small cube & \priority{75} & {\small Completed after a syntax hint about whitespace}\\
   \small \#7 & \small worldbank & \small cube & \priority{100} & {\small Completed very quickly }\\
-  \small \#8 & \small worldbank & \small cube & \priority{100} & {\small Extra time to find data }\\
-  \small \#9 & \small lords & \small table & \priority{75} & {\small  Struggled with composition}\\
+  \small \#8 & \small worldbank & \small cube & \priority{100} & {\small Completed, but needed longer to find correct data }\\
+  \small \#9 & \small lords & \small table & \priority{75} & {\small Struggled with composition of operations}\\
   \small \#10 & \small lords & \small table & \priority{100} & {\small Completed very quickly }\\
-  \small \#11 & \small lords & \small table & \priority{75} & {\small With a hint to avoid operations}\\
-  \small \#12 & \small olympics & \small table & \priority{75}  & {\small With a hint to avoid operations}\\
-  \small \#13 & \small olympics & \small table & \priority{75}  & {\small Hints about `then' and operations}\\
+  \small \#11 & \small lords & \small table & \priority{75} & {\small With a hint to avoid operations taking  arguments}\\
+  \small \#12 & \small olympics & \small table & \priority{75}  & {\small With a hint to avoid operations taking arguments}\\
+  \small \#13 & \small olympics & \small table & \priority{75}  & {\small With hints about `then' and operations taking arguments}\\
   \bottomrule
@@
 \subsection{Study Results}
-Table~\ref{tab:tasks} summarizes the work done by the study participants. Most notably, all
-participants were able to complete, at least partially, a non-trivial data exploration task and
-only half of them required further guidance. For each participant, we record the task, the kind
-of data source used in the task and the level of completion. For participants who needed assistance,
-the notes section details the help given. The experience suggests a number of possible design
-improvements for The Gamma, which are discussed below.
+Table~\ref{tab:tasks} summarizes the work done by the study participants. For each participant, we
+record the task, the kind of data source used and the level of completion. For
+participants who needed assistance, the notes section details the help given. The experience
+suggests a number of possible design improvements for The Gamma, which are discussed below.
 
-\subsubsection*{Can Non-programmers Explore Data with The Gamma?}
-Three facts allow us to answer RQ1 in the affirmative. First, 9 out of 11 participants agree or
-strongly agree that they ``found the system easy to use''. Second, participants spent 10--25 minutes
-(average 17 minutes) working with The Gamma and 12 out of 13 completed the task; 6 required assistance,
-but 3 of those faced one issue (discussed later) that could be addressed in the introduction.
-Third, a number of participants shared positive comments in the interviews.
+\subsubsection*{Can non-programmers explore data with The Gamma?}
+All participants were able to complete, at least partially, a non-trivial data exploration task and
+only half of them required further guidance. Participants spent 10--25 minutes (average 17 minutes)
+working with The Gamma and 12 out of 13 completed the task; 6 required assistance, but 3 of those
+faced an issue related to operations taking arguments (discussed later), which could be addressed
+in the introduction. A number of participants also shared positive comments in the interviews.
 For example, participant \#3 noted taht \emph{``this is actually pretty simple to use.
@@
 
-\subsubsection*{How Users Learn The Gamma?}
-The nature of RQ2 makes it a challenging question to study and finding conclusive answers
-may require further research. There is some evidence that knowledge can be transferred between
-different data sources (RQ2a). Two of the tasks (\emph{expenditure} and \emph{lords}) used
-one data source in the introduction and another for the task. Participants were able to complete
-those, although \emph{lords} has been more challenging as it involves a complex data source.
-Participant \#2 also gives a positive answer \emph{``I found it quite easy to translate what you
-showed us in the demo to the new dataset.''}.
+\subsubsection*{How users learn The Gamma?}
+There is some evidence that knowledge can be transferred between different data sources. Two of the
+tasks (\emph{expenditure} and \emph{lords}) used different data sources in the introduction and the
+task. Participants were able to complete those, although \emph{lords} has been challenging as it
+involves a complex data source. Participant \#2 also shared a positive comment:
+\emph{``I found it quite easy to translate what you showed us in the demo to the new dataset.''}.
 
-There is also some evidence that users can learn just from code samples (RQ2b). In the
+There is also some evidence that users can learn just from code samples. In the
 \emph{worldbank} task, participants were given only a minimal demo of how to invoke iterative
@@
 
-\subsubsection*{How Users Understand Complex Query Languages?}
-Although iterative prompting itself is simple, specific data sources may
-expose richer structure. To ascertain their understanding of the tabular type provider (RQ2c), we
-asked participants who worked with tabular data (\emph{lords} or \emph{olympics}) about their
-understanding of the  \ikvd{then} member. This is a regular member, but it has a
-special meaning in the domain specific language. For example, when specifying a list of
-aggregations, the user can repeatedly select further aggregations. Selecting \ikvd{then} completes
-the operation and allows the user to choose other operations such as sorting.
-
-Two participants (\#12 and \#13) initially thought that \ikvd{then} is used to split a command
-over multiple lines, but rejected the idea after experimenting.
-Participant \#12 then correctly concluded that it \emph{``allows us
+\subsubsection*{How users understand complex query languages?}
+The tabular type provider has a rich structure and uses a member named \ikvd{then} to complete
+the specification of a current operation, for example when specifying a list of aggregation
+operations. We asked participants who worked with tabular data (\emph{lords} or \emph{olympics})
+about their understanding of the \ikvd{then} member. Two participants (\#12 and \#13) initially
+thought that \ikvd{then} is used to split a command over multiple lines, but rejected the idea after
+experimenting. Participant \#12 then correctly concluded that it \emph{``allows us
 to chain together the operations''} of the query; after a hint, participant \#13 reflected
-that \emph{``if I knew this from the start, it would [have been easier].''} This summarizes an
-important fact. Iterative prompting allows the users to start exploring new data sources, but
+that \emph{``if I knew this from the start, it would [have been easier].''}
+In summary, iterative prompting allows users to start exploring new data sources, but
 the structures exposed by more complex data sources have their own further design principles that
@@
 
-\subsubsection*{Making Complex Things Possible May Hurt}
-The Gamma supports operations such as \ikvd{take(5)}. Most type
-providers never generate those, but the provider for working with tabular data is an exception.
-When filtering data, the provider allows specifying a condition on numerical attributes such as
+\subsubsection*{What would make The Gamma easier to use?}
+The type provider for tabular data generates operations that take arguments such as \ikvd{take(5)}
+in a number of places. When filtering data, it allows writing
 \ikvd{olympics.\textquotesingle filter data\textquotesingle.\textquotesingle Year is greater than\textquotesingle(2004)}.
-%
-Three participants (\#11, \#12, \#13) struggled to complete a task, because they
-initially attempted to use those operations. However, those operations violate the iterative prompting
-principle as one cannot type `.' after \ikvd{\textquotesingle Year is greater than\textquotesingle}.
+Those operations violate the iterative prompting principle as one cannot type `.' after
+\ikvd{\textquotesingle Year is greater than\textquotesingle}. Three participants (\#11, \#12, \#13)
+struggled to complete a task, because they initially attempted to use those operations.
 This suggests that we should either avoid such operations, or hide them under an ``advanced operations''
-tab as a caution to novice users.
+tab.
 
-\subsubsection*{Benefits and Drawbacks of Text}
-The Gamma is based on text to aid transparency. Text implies that there is no hidden state and
-the reader sees the full code. The study suggests that using a text editor has both
-benefits and drawbacks compared to alternatives such as structured editors~\cite{structure-based,livenut,lamdu}.
-Most participants had no difficulty navigating around source code, making edits or deleting code
-fragments, which is arguably harder in an unfamiliar structured editor.
-
-On the one hand, we observed two issues in the study. Participant \#2 struggled with correct indentation, starting
-a second command with more indentation than needed and participant \#6 had a syntax error in an unrelated command,
-which prevents charts from rendering. On the other hand, some participants used the text editor effectively,
-e.g.~participant \#5, who used copy-and-paste to fetch the same data series for multiple countries.
+The Gamma uses an ordinary text editor. This has both benefits
+and drawbacks compared to structured editors~\cite{structure-based,livenut,lamdu}.
+Most participants had no difficulty navigating around code, making edits or deleting
+fragments, which is arguably harder in a structured editor. Some participants used
+text editor effectively, e.g.~participant \#5, who used copy-and-paste to fetch the same data
+for multiple countries. However, we also observed two issues. Participant \#2 struggled
+with indentation and participant \#6 had a syntax error in an unrelated command, which
+prevents charts from rendering.
 
@@
 
-\section{Design Reflections}
-\label{sec:design}
+\section{Discussion}
+\label{sec:discuss}
 
-The design of The Gamma has been motivated by a curiosity as to whether iterative
-prompting can make text-based programming with data accessible to non-experts. In this section,
-we theoretically assess the resulting design and position it in the context of a possible application
-in the context of data journalism.
+The Gamma examines an unexplored point in the design space of tools for data exploration. It is
+a text-based programming environment for non-programmers. Its design has been motivated
+by a curiosity as to whether iterative prompting can make text-based programming with data
+accessible to non-experts. In this section, we theoretically assess the resulting design,
+making use of criteria for judging whether a system advances the state of the art proposed by
+Olsen~\cite{evaluating}.
 
-\subsection{Lowering Barriers to Entry}
-\label{sec:design-bar}
+\subsubsection*{Learning without experts}
+The Gamma allows non-programmers to produce transparent and reproducible scripts that explore
+data from a wide range of sources. It allows new participants to benefit from the capabilities
+offered by programmatic data exploration, satisfying the \emph{exmpowering new participants}
+criteria proposed by Olsen~\cite{evaluating}. To empower new participants, our design aims to
+make The Gamma suitable for users who cannot dedicate significant amount of time to learning
+it in advance and may not have access to experts. This is supported in two ways.
 
-Data exploration has a certain irreducible essential complexity. To make it accessible to users who
-cannot dedicate much time to learning a tool prior to using it, this complexity needs to be
-carefully stratified. The Gamma uses a two-level structure. At the first level, the user needs
-to learn iterative prompting to be able to start exploring data. At the second level, users will
-need to learn domain specific languages defined by individual type providers.
-
-\subsubsection*{Iterative Prompting Principle}
-Iterative prompting is a suitable first level principle, because it is easy to trigger.
-In conventional programming languages, auto-complete assistance is only available once basic
-code structure is written. In contrast, user of In The Gamma only needs to choose the initial
-data source. Iterative prompting is also easier to use than e.g.~a command line or a REPL
+First, the iterative prompting principle makes it easy for users to start experimenting.
+The user needs to select an initial data source and then repeatedly choose an
+item from a list of choices. Iterative prompting is easier to use than e.g.~a command line or a REPL
 (read-eval-print-loop), because it follows the \emph{recognition over recall} usability heuristic.
 The users are not required to recall and type a command. They merely need to select one from a
-list of options.
+list of options. Second, the resulting source code serves as a trace of how the analysis was
+created. It provides the user with all information that they need to recreate the
+program, not just by copying it, but also by using iterative prompting. Such design
+has been termed \emph{design for percolation} by Sarkar \cite{learning}, who studies how Excel users
+learn. He points out that users learn new features when the usage of a feature is apparent in a spreadsheet.
+For example, users can learn different functions in formulas, because those are visible in the cell.
+Learning how to use a wizard for creating charts is hard because the operation leaves no
+full trace in the spreadsheet. Design for percolation thus supports learnability.
 
-\subsubsection*{Stratifying Data Exploration Complexity}
-At the first level, any data source is accessed through the same mechanism. At the second level,
-each data source defines its own domain specific language, consisting of the primitives that are
-offered to the user in auto-complete prompts. The domain specific languages are embedded in The Gamma
-language -- they define merely the available members and cannot extend the core language syntax.
-Although the complexity of individual languages differs, the users can always start exploring
-and learning new languages using the familiar first level iterative prompting principle.
-An important question, which we study in Section~\ref{sec:study}, is whether the expertise gained
-with one data source be transferred to working with another data source.
+\subsubsection*{Lowering barriers to entry}
 
-\subsection{Learning without Experts}
-\label{sec:design-expert}
+Data exploration has a certain irreducible essential complexity. To make a system usable, this
+complexity needs to be carefully stratified. The Gamma uses a two level structure. The first level
+consists of the language itself with the iterative prompting mechanism. The second level consists of
+the individual members generated by a type provider. This can be seen as a \emph{domain specific
+language}, embedded in The Gamma language. Although the complexity of individual domain specific
+languages differs, the user can always start exploring through iterative prompting, even when
+faced with an unfamiliar data source.
+%
+In tackling complexity, The Gamma satisfies two criteria proposed by Olsen \cite{evaluating}.
+The Gamma satisfies the \emph{generality} criteria in that it can be used uniformly with a wide
+range of data sources. It also satisfies the \emph{expressive leverage} criteria in that it factors
+out common aspects of different data queries into the core language (first level) and leaves the
+specifics of each data source to the second level.
 
-A typical user of The Gamma will not dedicate significant time to learning it in advance and
-they will not have access to experts. Most learning thus needs to happen from examples.
-When analysing how Excel users learn, Sarkar \cite{learning} points out that users learn new
-features when the usage of a feature is apparent in a spreadsheet. For example, users can learn
-different functions in formulas, because those are visible in the cell. Learning how to use a
-wizard for creating charts is not possible because the operation leaves no full trace in the
-spreadsheet. Sarkar's recommendation is to \emph{design for percolation},
-i.e.~in a way where looking at the final result makes it apparent what feature has been used and how.
+\subsubsection*{Correctness and completeness}
 
-In The Gamma, each step of iterative prompting results in an identifier that is
-added to the source code. This means that a program constructed solely through iterative prompting
-keeps a full trace of how it was created. The resulting source code provides the user all
-information that they need to recreate the program, not just by copying it, but also by using the
-iterative prompting mechanism. We assess the viability of this way of learning in the study in
-Section~\ref{sec:study}.
+An important characteristic of our design is that the iterative prompting mechanism is both \emph{correct}
+and \emph{complete} with respect to possible data exploration scripts. The two properties are
+a consequence of the fact that a program is a formed by a chain of operations and
+that the auto-completion leverages a static type system. When invoking iterative prompting
+at the end of a well-typed script, a selected option, which is a valid object member, is added to
+the end of the script, resulting in another well-typed script. This distinguishes our system from
+auto-completion based on machine learning, which may offer members not
+valid in a given context. Auto-completion lists offered via iterative prompting contain
+all available members and so the user can construct all possible scripts. Two exceptions to
+completeness in our current design are the let binding and specifying numerical parameters as in
+\ikvd{take(5)}.
 
-\subsection{Correctness and Completeness}
-\label{sec:design-cc}
+\subsubsection*{Importance of applications}
+Although The Gamma targets a broad audience of non-programmers, some of our work has been
+particularly motivated by the use of data in journalism. It has the potential to enable journalists
+to make factual claims backed by data more commonplace and enable wider audience to engage with
+such claims. As such The Gamma satisfies the \emph{importance} criteria proposed by Olsen~\cite{evaluating}.
+Although The Gamma is open-source, it has not been deployed in a newsroom so far.
+This would lead to valuable insights, but it requires finding a suitable fortuitous opportunity.
+The potential is indicated by a comment from a former journalist who participated in our study (\#13):
+\emph{``There's a lot of effort going into data journalism that  programming could make much quicker,
+  (...) like this would really simplify things.''}
 
-An important characteristic of our design is that, barring a few exceptions discussed below, the iterative
-prompting mechanism is both correct and complete with respect to possible data exploration scripts.
-This means that (i) auto-complete lists offered by iterative prompting contain
-only options that are valid in a given context and that (ii) any script that can be written
-in The Gamma can be constructed via iterative prompting.
-
-\subsubsection*{Correctness} The Gamma uses an object-based type system for error-checking and for
-generating auto-complete lists. A well-typed script can always be executed. When using iterative
-prompting, a selected option, which is a valid object member, is added to the end of a script,
-leading to a well-typed script. This distinguishes our system from auto-completion based on machine
-learning or dictionary-based methods, which may offer members not valid in a given context. Advanced
-users of The Gamma, modifying scripts as text, can still violate correctness -- for example, if the
-user uses auto-complete to replace a member in a middle of a member access chain with a member of another type.
-
-\subsubsection*{Completeness} Auto-completion lists offered via iterative prompting contain
-all available members and so the user can construct all possible scripts. The range of such scripts
-is determined by type providers, which support limited set of options (e.g. group aggregations in
-the tabular data type provider). Two exceptions to completeness in our current design is specifying
-numerical parameters as in \ikvd{take(5)} and advanced language features such as let binding.
-We believe that alleviating the need for those poses an interesting further research challenge.
-
-\section{Discussion}
-We examine an unexplored point in the design space of tools for data exploration. The Gamma is a
-text-based programming environment, but targets non-programmers such as data journalists. We
-conclude with a more general discussion of evaluation of the system and the possible applicability
-of The Gamma in data journalism.
-
-\subsection{Evaluating Exploratory Research}
-The research presented in this paper is qualitative and exploratory in nature. In particular,
-we do not make any quantitative claims about the usability of The Gamma and its learning curve.
-Our investigation focused on the core iterative prompting principle, but some our case studies
-also required using features such as operations with parameters.
-
-Although The Gamma is open-source, it has not been deployed in production in a newsroom so far.
-This would lead to valuable insights, but it requires finding a suitable fortuitous opportunity.
-Finally, we also do not compare the usability of The Gamma with other popular systems
-such as Tableau \cite{tableau}. It would be possible to set tasks that can be completed
-in both systems, but the systems have very different aims making such comparison problematic.
-
-\subsection{Evaluating Complex Systems}
-\label{sec:discussion-eval}
-Data exploration environments are complex systems that do not yield to simple controlled
-experimentation. Olsen~\cite{evaluating} proposes criteria for judging whether a system advances
-the state of the art. A number of those apply to The Gamma:
-
-\begin{itemize}
-\item \emph{Importance.} Data journalism can make factual claims backed by data more commonplace
-and enable wider audience to engage with such claims. As such, we contribute towards
-solving an important societal issue.
-
-\item \emph{Expressive leverage.} Iterative prompting stratifies the complexity of data exploration
-such that different data sources are accessed through the same unified iterative prompting
-interaction principle.
-
-\item \emph{Empowering new participants.} As demonstrated by our user study, The Gamma allows
-non-experts, including those not comfortable with code, to perform basic programmatic
-data exploration tasks.
-
-\item \emph{Generality.} The Gamma can be used to query data from a wide range of data
-sources including tabular data, data cubes and graph databases. The range of possible tasks is
-illustrated by case studies presented earlier.
-\end{itemize}
-
-\subsection{Applications to Data Journalism}
-Although The Gamma targets a broad audience of non-programmers, some of our work has been
-particularly motivated by the use of data in journalism. The Gamma is aligned with recent challenges
-faced by journalists in that it can help build trust through transparency as well as provide
-meaningful ways of reader engagement.
-
-Our study shows that non-experts with background similar
-to journalists are able to solve basic data exploration tasks using The Gamma. When asked whether
-The Gamma is something that journalists could learn how to use, a former journalist who
-participated in our study (\#13) answered:
-
-\begin{quote}
-\emph{``Yeah, I think so. There's a lot of effort going into data journalism that
-  programming could make much quicker, but I was always nervous about code. (...)
-  Something like this would really simplify things.''}
-\end{quote}
-
-\noindent
-The answer suggests that iterative prompting, does indeed, lower the barrier to entry. Although it
-does not fully eliminate complexity involved in data querying, it provides a way of stratifying
-it. Iterative prompting makes it easy to get started with data exploration, addressing the initial
-``nervousness about code''. By making the source code of data analyses visible, The Gamma then
-enables further learning through percolation.
-
-\subsection{Further Design Issues}
-There remain a number of aspects of data exploration in the context of journalism that
-The Gamma does not address. Two of those, data provenance and data availability were also
-observed by the participants in our study.
-
-\subsubsection*{Data provenance}
-Data sources such as \ikvd{olympics} or \ikvd{worldbank} are defined when initializing The Gamma,
-but the system does not currently show where such data comes from.
-For some tabular data sources, the source is a CSV file published, e.g.~by the government. In this case,
-we can easily show the source URL. However, other type providers may pre-process data. Displaying
-data source in such cases would require more sophisticated provenance tracking \cite{provenance}.
-
-\subsubsection*{Data availability}
-In the current version, The Gamma does not have a way of informing the user what data sources
-are available. In other words, the user needs to know the first identifier, such as \ikvd{olympics},
-to get started. We could address this by choosing a data source as the first step of iterative
-prompting and perhaps typing \ikvd{.olympics}. However, a more fundamental issue is finding
-the data source in the first place. This could be partly addressed by a type provider for a
-curated online database such as Enigma Public\footnote{\url{https://docs.enigma.com/public}}.
-Providing access to open government data repositories such as \url{http://data.gov} and
-\url{http://data.gov.uk} is more appealing, but challenging due to their unstructured nature.
+% The answer suggests that iterative prompting, does indeed, lower the barrier to entry. Although it
+% does not fully eliminate complexity involved in data querying, it provides a way of stratifying
+% it. Iterative prompting makes it easy to get started with data exploration, addressing the initial
+% ``nervousness about code''. By making the source code of data analyses visible, The Gamma then
+% enables further learning through percolation.
+%
+% \section{Future work}
+% There remain a number of aspects of data exploration in the context of journalism that
+% The Gamma does not address. Two of those, data provenance and data availability were also
+% observed by the participants in our study.
+%
+% \subsubsection*{Data provenance}
+% Data sources such as \ikvd{olympics} or \ikvd{worldbank} are defined when initializing The Gamma,
+% but the system does not currently show where such data comes from.
+% For some tabular data sources, the source is a CSV file published, e.g.~by the government. In this case,
+% we can easily show the source URL. However, other type providers may pre-process data. Displaying
+% data source in such cases would require more sophisticated provenance tracking \cite{provenance}.
+%
+% \subsubsection*{Data availability}
+% In the current version, The Gamma does not have a way of informing the user what data sources
+% are available. In other words, the user needs to know the first identifier, such as \ikvd{olympics},
+% to get started. We could address this by choosing a data source as the first step of iterative
+% prompting and perhaps typing \ikvd{.olympics}. However, a more fundamental issue is finding
+% the data source in the first place. This could be partly addressed by a type provider for a
+% curated online database such as Enigma Public\footnote{\url{https://docs.enigma.com/public}}.
+% Providing access to open government data repositories such as \url{http://data.gov} and
+% \url{http://data.gov.uk} is more appealing, but challenging due to their unstructured nature.
+%
+% xxx
+%
+% The research presented in this paper is qualitative and exploratory in nature. In particular,
+% we do not make any quantitative claims about the usability of The Gamma and its learning curve.
+% Our investigation focused on the core iterative prompting principle, but some our case studies
+% also required using features such as operations with parameters.
 

### 2022-06-10 wip-design-v2/paper.tex
Edits, VL/HCC version

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
\usepackage{caption}
\usepackage{subcaption}
\usepackage{xcolor}
\usepackage[colorlinks]{hyperref}
\usepackage{fancyvrb}
\usepackage{tikz}
\usepackage{booktabs}

\newcommand*{\priority}[1]{\begin{tikzpicture}[scale=0.12]%
   \draw (0,0) circle (1);
   \fill[fill opacity=1,fill=black] (0,0) -- (90:1) arc (90:90-#1*3.6:1) -- cycle;
   \end{tikzpicture}}
\newcommand*{\priorityc}[1]{
\resizebox{0.8em}{0.8em}{
 {\protect\tikz{ \protect
   \draw[line width=1mm, black] (0,0) circle (1); \fill[fill opacity=1,fill=black] (0,0) -- (90:1) arc (90:90-#1*3.6:1) -- cycle;
   } }}}
%
\definecolor{kvdclr}{rgb}{0.3,0.0,0.8}
\DefineVerbatimEnvironment{thegamma}{Verbatim}{fontfamily=zi4,numbers=left,xleftmargin=6mm,fontsize=\small,commandchars=\\\{\}}
\newcommand{\kvd}[1]{\textbf{#1}}
\newcommand{\ikvd}[1]{{\fontfamily{zi4}\selectfont\small #1}}

\begin{document}
\title{\textbf{The Gamma}: Programmatic Data\\ Exploration for Non-programmers}
\author{\IEEEauthorblockN{Tomas Petricek}
\IEEEauthorblockA{University of Kent, Canterbury, UK \&\\ Charles University, Prague, Czech Republic}}

\maketitle

\begin{abstract}
Data exploration tools based on code can access any data source, result in reproducible scripts and
encourage users to verify, reuse and modify existing code. Unfortunately, they are hard to use and
require expert coding skills. Can we make data exploration tools based on code accessible to non-experts?

We present The Gamma, a novel text-based data exploration environment that answers the question in
the affirmative. The Gamma takes the idea of code completion to the limit. Users create transparent
and reproducible scripts without writing code, by repeatedly choosing from offered code completions.

The Gamma is motivated by the needs of data journalists and shows that we may not need to shy away
from code for building accessible, reproducible and transparent tools that allow a broad public to
benefit from the rise of open data.
\end{abstract}

\begin{IEEEkeywords}
data exploration, data journalism
\end{IEEEkeywords}
\IEEEpubid{978-1-6654-4214-5/22/\$31.00 \copyright 2022 IEEE}
\IEEEpubidadjcol

% ==================================================================================================

\section{Introduction}
\noindent
Despite the advances on visual tooling, programmatic data exploration remains the choice of
expert analysts. It is flexibile, offers greater reusability and leads to transparent
analyses. The design of a programmatic data exploration tool that would be accessible to data
journalists poses a number of design challenges. First, the tool needs to have a low barrier to
entry to support first-time users without training. Second, it needs to support multiple
data sources in a uniform way to allow transfer of knowledge across domains. Finally, users need
to be able to learn by looking at existing data analyses.

We present The Gamma, a text-based data exploration tool for non-experts that is based on a single,
easy to understand interaction principle. It provides a uniform access to data tables, graph
databases and data cubes and leads to transparent analyses that can be easily reproduced,
encouraging learning and critical engagement with data.

The Gamma is based on \emph{iterative prompting}, which turns code completion from a programmer
assistance tool into a non-expert programming mechanism that allows users to construct all valid
data exploration scripts just by repeatedly choosing an item from a list of offered options.
The design favors \emph{recognition over recall} and allows non-programmers to construct entire
scripts without first learning to code. Yet, the result remains a transparent and reproducible script.
A crucial feature is that iterative prompting only offers operations that are valid in a given
context and that it offers all such operations; it is both correct and complete.

The Gamma focuses on tasks that a data journalist may want to complete (Figure~\ref{fig:thegamma}).
The user accesses data available in a structured format. They make several experiments to find an
interesting insight, e.g.~by applying different aggregations or filters. They
visualize the results using a table or a chart before publishing their analysis. The
Gamma makes such programmatic data exploration simple enough for non-programmers. Scraping and
cleaning of messy data or building custom data visualizations is outside of the scope of
our work, but exposing such functionality using iterative prompting is an interesting and
worthwhile future challenge.

\begin{figure}
\includegraphics[width=1\columnwidth]{figures/thegamma-annot}
\caption{Obtaining teams with the greatest number of gold medals from Rio 2016
Olympics: (1) Reproducible The Gamma script; (2) contextual iterative prompting
offering ways of sorting the data; (3) an instant preview of results.}
\label{fig:thegamma}
\vspace{-1em}
\end{figure}

\IEEEpubidadjcol
In this paper, we describe and evaluate the design principles behind The Gamma project:

\begin{itemize}
\setlength\itemsep{0.25em}
\item We introduce the iterative prompting principle in The Gamma (Section~\ref{sec:overview})
  and show how it can be used for querying of distinct data sources including data tables, graph
  databases and data cubes (Section~\ref{sec:system}).

\item We illustrate the expressiveness of the system through a case study (Section~\ref{sec:cases})
  and evaluate it through a user study (Section~\ref{sec:study}), confirming that non-programmers
  can use The Gamma to construct non-trivial data queries.

\item We reflect how our design lowers barriers to entry, supports learning without experts and offers a
  complete and correct program construction method (Section~\ref{sec:discuss}).
\end{itemize}

\noindent
The Gamma is available at \url{http://thegamma.net},
both as an open-source library and a hosted data exploration service.

% ==================================================================================================

\section{Related Work}
\noindent
We aim to make recent advances on information-rich programming~\cite{inforich} available to
non-programmers~\cite{enduser,smallmatter}, in the context of data journalism~\cite{ddj}.
Our work features a novel combination of characteristics in that our iterative prompting
interaction principle is centered around code, but reduces the conceptual complexity of coding to
a single basic kind of interaction.

\vspace{0.5em}
\subsubsection*{Code Completion for Data Science}

The Gamma utilizes type providers~\cite{inforich,fsdata}, which integrate external data into a
static type system. This enables auto-completion~\cite{assistants}, which we turn into a tool for
non-programmers. Similar systems based on machine learning and domain specific
languages~\cite{predictive,proactive} do not guarantee completeness, i.e.~it is unclear whether the user
can create all possible scripts. Approaches based on natural language are
effective~\cite{eviza,codemend,iris}, but hide the underlying structure and do not help users understand
the exact operations performed. Code completion based on machine learning~\cite{mlcomplete,statcomplete}
also exists for general-purpose programming languages used by data scientists such as
Python~\cite{pythia}, but this focuses on providing assistance to expert programmers.

\vspace{0.5em}
\subsubsection*{Notebooks and Business Intelligence Tools}

Notebooks such as Jupyter~\cite{jupyter} are widely used data exploration environments for
programmers. The Gamma targets non-experts, but could be integrated with a multi-language
notebook system~\cite{wrattler}.
%
Spreadsheets, business intelligence~\cite{tableau,powerbi} and other visual data analytics
tools~\cite{control,vizdom} do not involve programming, but require mastering a complex GUI.
In contrast, in The Gamma, all operations can be completed through a single kind of interaction.
Several systems~\cite{potter,wrangler,lyra} record interactions with the GUI as a script that can
be modified by the user. Unlike in The Gamma, the generated code does not guide the user in
learning how to use the system.

\vspace{0.5em}
\subsubsection*{Easier Programming Tools}
Many systems aim to make programming easier. Victor~\cite{principle} inspired work on live
environments environments~\cite{review,liveroad,lighttable} that help programmers understand how
code relates to output; exploratory systems~\cite{variolite,exploratory} assist with completing
open-ended tasks; and system combining code with visualization also exists for graph
querying~\cite{guess}. The Gamma is live in that our editor gives an instant preview of the results.
To avoid difficulties with editing code as text, some systems use structured
editors~\cite{structure-based,livenut,lamdu,subtext,directprog}. Many systems simplify programming by
offering high-level abstractions, e.g.~for interactive news articles~\cite{idyll}, statistical
analyses~\cite{tea}, data visualization~\cite{interactionviz,vegalite}. The Gamma
provides high-level abstractions for data querying, but supporting other tasks remains future work.

\vspace{0.5em}
\subsubsection*{Programming without Writing Code}
In programming by example~\cite{byexample}, used for example in spreadsheets~\cite{spreadsheetpbe,flashextract},
the user gives examples of desired results. In direct manipulation~\cite{direct}, a program is
specified by interacting with the output. This has been used in the visual domain~\cite{sketchnsketch}, but also for data
querying~\cite{dynamicq,vlang,visage}. Direct manipulation can also support data exploration by letting
users partially edit queries,~e.g. by changing quantifiers as in DataPlay~\cite{dataplay}.

% ==================================================================================================

\section{Overview}
\label{sec:overview}

\noindent
The Gamma is a text-based system that allows non-experts to explore data using
iterative prompting -- by repeatedly selecting an item from an auto-complete list. The study
presented in Section~\ref{sec:study} confirms that the kind of data exploration shown in the next
section can be successfully done by non-experts.

We introduce The Gamma by walking through a
typical data exploration task. A data journalist from Kent is exploring travel expense claims
by members of the House of Lords published by the UK government \cite{lords}. After importing
the CSV file through a web interface, the environment is initialized with code that refers
to the imported data as \ikvd{expenses} using the type provider for tabular data (Section~\ref{sec:system}).
The journalist types `.' (dot) to start exploring the data:

\vspace{0.5em}
\begin{thegamma}
expenses.
\end{thegamma}
\qquad\includegraphics[width=0.7\columnwidth]{figures/lords0}
\vspace{0.5em}

\noindent
The type provider offers a list of operations that the journalist can perform. To find House of Lords
members from Kent, the journalist chooses \ikvd{filter data}. She is then offered a list of columns
based on the schema of the CSV file and chooses \ikvd{County is}. The completion lists
counties in the data set:

\vspace{0.5em}
\begin{thegamma}
expenses.'filter data'.'County is'.
\end{thegamma}
\qquad\includegraphics[width=0.7\columnwidth]{figures/lords1}
\vspace{0.5em}

\noindent
The journalist chooses Kent. The Gamma evaluates the code on-the-fly and shows a
preview of results.

\vspace{0.5em}
\begin{thegamma}
expenses.'filter data'.'County is'.Kent
\end{thegamma}
\vspace{0.4em}\qquad\includegraphics[width=0.95\columnwidth]{figures/lords2}
\vspace{0.5em}

\noindent
The journalist decides to compare travel costs. She finishes specifying the
filtering condition by choosing \ikvd{then} and is offered the same list of querying operations as
in the first step. She selects  \ikvd{sort data} and is offered a list of sorting options:

\vspace{0.5em}
\begin{thegamma}
expenses.'filter data'.'County is'.Kent.then.
  'sort data'.
\end{thegamma}
\qquad\includegraphics[width=0.7\columnwidth]{figures/lords3}
\vspace{0.5em}

\noindent
The journalist chooses \ikvd{then} and is, again, offered the list of querying operation. She
uses \ikvd{paging} to get the top 4 records, which requires typing \ikvd{4} as the argument.
She then uses the \ikvd{get series} operation to obtain a data series associating travel
expenses with a name, which is automatically visualized:

\vspace{0.5em}
\begin{thegamma}
expenses.'filter data'.'County is'.Kent.then
  .'sort data'.'by Travel Costs descending'.then
  .paging.take(4).'get series'
    .'with key Name'.'and value Travel Costs'
\end{thegamma}
\vspace{0.2em}\quad\includegraphics[width=0.95\columnwidth]{figures/lords4}

The code is not unlike an SQL query, except that the whole script is constructed using
iterative prompting, by repeatedly selecting one of the offered members. Those represent both
operations, such as \ikvd{sort by} and arguments, such as \ikvd{Kent}. The only exception
is when the analyst needs to type the number \ikvd{4} to specify the number of items to take.

\section{System description}
\label{sec:system}

\noindent
A program in The Gamma is a sequence of commands that can be either a variable declarations or
an expression that evaluates to a value. An expression is a reference to a data source followed
by a chain of member accesses. Each expression has a type that is used to generate options in
auto-completion. A type defines a list of members that, in turn, have their own types.
The types are not built-in, but are generated by type providers for individual data sources.
The syntax and semantics of the language has been described elsewhere \cite{tglive}.

A new data source can be supported by implementing a \emph{type provider},
which defines a domain specific language for exploring data of a particular kind.
A type provider generates object types with members (such as \ikvd{paging} or \ikvd{Kent})
that are accessed via iterative prompting. We outline type providers for exploring data cubes
(inspired by Syme et al. \cite{inforich}), tabular data (formalized elsewhere \cite{dotdriven}), and graph databases.

\vspace{0.5em}
\subsubsection*{Data Cube Provider}
Data cubes are multi-dimensional arrays of values. For example, the World Bank collects a range of
indicators about many countries each year while the UK government expenditure records spending for
different government services, over time, with different adjustments:

\vspace{0.25em}
\begin{thegamma}
worldbank.byCountry.'United States'.
  'Climate Change'.'CO2 emissions (kt)'

expenditure.byService.Defence.inTermsOf.GDP
\end{thegamma}
\vspace{0.25em}

\noindent
The dimensions of the \ikvd{worldbank} cube are countries, years and indicators, whereas the dimensions
of \ikvd{expenditure} are government services, years and value type (adjusted, nominal,
per GDP).  Figure~\ref{fig:cubetp} illustrates how the provider allows users to slice the data cube.
Choosing \ikvd{byCountry.\textquotesingle United States\textquotesingle},
restricts the cube to a plane and \ikvd{\textquotesingle CO2 emissions (kt)\textquotesingle}
then gives a series with years as keys and emission data as values. Similarly, we could first filter the
data by a year or an indicator. The same mechanism is used to select UK government spending on
defence in terms of GDP.
%  As there is no widely used standard format for data cubes, adding another data cube to The Gamma
% currently requires implementing a new type provider.

\vspace{0.5em}
\subsubsection*{Graph Database Type Provider}
Graph databases store nodes representing entities and relationships between them.
The following example explores a database of Doctor Who characters and episodes. It retrieves
all enemies of the Doctor that appear in the Day of the Moon episode:

\vspace{0.25em}
\begin{thegamma}
drwho.Character.Doctor.'ENEMY OF'.'[any]'
  .'APPEARED IN'.'Day of the Moon'
\end{thegamma}
\vspace{0.25em}

\noindent
We start from the \ikvd{Doctor} node and then follow two relationships. We use
\ikvd{\textquotesingle ENEMY OF\textquotesingle.\textquotesingle [any]\textquotesingle}
to follow links to all enemies of the Doctor and then specify
\ikvd{\textquotesingle APPEARED IN\textquotesingle}
to select only enemies that appear in a specific episode. The query is illustrated in
in Figure~\ref{fig:graphtp}. The members are generated from the data;
\ikvd{ENEMY OF} and \ikvd{APPEARED IN} are labels
of relations and \ikvd{Doctor} and \ikvd{Day of the Moon} are labels of nodes. The
\ikvd{[any]} member defines a placeholder that can be filled with any node with the specified
relationships. The results returned by the provider is a table of properties of all nodes
along the specified path, which can be further queried and visualized.

\vspace{0.5em}
\subsubsection*{Tabular Data Provider}

Unlike the graph and data cube providers, the type provider for tabular data does not just
allow selecting a subset of the data, but it can be used to construct SQL-like query. Consider
the example of querying expense claims from Section~\ref{sec:overview}, which filters and then
sorts the data.

When using the provider, the user specifies a sequence of operations. Members such as
\ikvd{\textquotesingle filter data\textquotesingle} or \ikvd{\textquotesingle sort data\textquotesingle}
determine the operation type. Those are followed by members that specify operation parameters.
For example, when filtering data, we first select the column and then choose a desired value.
Unlike SQL, the provider only allows users to choose from pre-defined filtering conditions,
but this is sufficient for constructing a range of practical queries.

\begin{figure}[h]
\centering
\begin{subfigure}[b]{0.5\textwidth}
  \centering
  \includegraphics[scale=0.27]{figures/cubetp}
  \vspace{0.5em}
  \caption{Exploring World Bank data using the data cube type provider, users
    choose values from two dimensions to obtain a data series.}
  \label{fig:cubetp}
  \vspace{1em}
\end{subfigure}
\begin{subfigure}[b]{0.45\textwidth}
  \centering
  \includegraphics[scale=0.31]{figures/graphtp}
  \caption{To query graph data, the user specifies a path through the data, possibly with
    placeholders to select multiple nodes.}
  \label{fig:graphtp}
  \vspace{0.5em}
\end{subfigure}
\caption{Type providers for exploring cube and graph data.}
\label{fig:tps}
\vspace{-0.5em}
\end{figure}

% ==================================================================================================

\section{Case Study}
\label{sec:cases}

\noindent
The Gamma aims to simplify programmatic data exploration while keeping enough expressive power
to allow users to create interesting data explorations. To show what can be achieved by
interactive prompting, we present a case study that explores a graph database with Dr Who
series data.\footnote{See: \url{http://gallery.thegamma.net/87/}. We also used The Gamma for
projects exploring the UK government expenditure, activities of a research institute and Olympic
medal winners, available at \url{http://turing.thegamma.net} and \url{http://rio2016.thegamma.net}}.

The following constructs a chart (Figure~\ref{fig:cases-dr}) of top Dr Who villains by the number of
episodes in which they appear. This case is interesting as it combines the graph database provider
for fetching the data with the tabular data provider:

\vspace{0.25em}
\begin{thegamma}
drWho.Character.Doctor.'ENEMY OF'.'[any]'
     .'APPEARED IN'.'[any]'.explore
  .'group data'.'by Character name'
     .'count distinct Episode name'.then
  .'sort data'.'by Episode name descending'.then
  .paging.take(8).'get series'
     .'with key Character name'
     .'and value Episode name'
\end{thegamma}
\vspace{0.25em}

\noindent
Line 1 use the graph provider to find all paths linking the Doctor with any
character linked via \ikvd{ENEMY OF}, followed by any episode linked by \ikvd{APPEARED IN}.
This produces a table that can be analysed using the tabular data provider by selecting
\ikvd{explore}. For each character (the villain) we count the number of
distinct episodes. The result is shown in Figure~\ref{fig:cases-dr}.
Despite performing a sophisticated data analysis that involves a graph database query,
followed by an SQL-like data aggregation, the code can be constructed using iterative
prompting, with the exception of the numbers in paging.

% ==================================================================================================

% TODO Study improvements
% - The questionaire needs a bit more introduction when you present  the study.
% - When looking at the data for the questionaire, it seems somewhat inconclusive
%    as all the values lie around 2.5.  I think this warrants a bit more discussion.
% - Connected to this, I would recommend to highlight the successes of the study a bit more in the
%     discussion and  presentation of the results. You actually got non-programmers to do
%     sophisticated queries! That you didn't get support for all RQs is perhaps more a result of being quite  ambitious!

% \begin{table*}
% \vspace{-0.5em}
% \centering
% \renewcommand{\arraystretch}{1.25}
% \begin{tabular}{l l l c l}
%   \toprule
%     & {\small \textit{Task}}
%     & {\small \textit{Kind}} & {\small \textit{Done}}
%     & {\small \textit{Notes}} \\
%   \midrule
%   \small \#1\qquad\qquad & \small expenditure \qquad\qquad & \small cube \qquad\qquad & \qquad\priority{50}\qquad\qquad & {\small Obtained one of two data series}\\
%   \small \#2 & \small expenditure & \small cube & \priority{100} & {\small Explored furhter data series independently}\\
%   \small \#3 & \small expenditure & \small cube & \priority{100}& {\small Explored further data series independently}\\
%   \small \#4 & \small expenditure & \small cube & \priority{75}& {\small Completed following a hint to use another member }\\
%   \small \#5 & \small expenditure & \small cube & \priority{100}& {\small Explored further data series independently}\\
%   \small \#6 & \small worldbank & \small cube & \priority{75} & {\small Completed after a syntax hint about whitespace}\\
%   \small \#7 & \small worldbank & \small cube & \priority{100} & {\small Completed very quickly }\\
%   \small \#8 & \small worldbank & \small cube & \priority{100} & {\small Completed, but needed longer to find correct data }\\
%   \small \#9 & \small lords & \small table & \priority{75} & {\small Struggled with composition of operations}\\
%   \small \#10 & \small lords & \small table & \priority{100} & {\small Completed very quickly }\\
%   \small \#11 & \small lords & \small table & \priority{75} & {\small With a hint to avoid operations taking  arguments}\\
%   \small \#12 & \small olympics & \small table & \priority{75}  & {\small With a hint to avoid operations taking arguments}\\
%   \small \#13 & \small olympics & \small table & \priority{75}  & {\small With hints about `then' and operations taking arguments}\\
%   \bottomrule
% \end{tabular}
%
% \caption{\centering Overview of work completed by individual participants in the study. \\The marks denote:
%  \priorityc{100} = completed, \priorityc{75} = required some guidance, \priorityc{50} = partially completed}
% \label{tab:tasks}
% \vspace{-0.5em}
% \end{table*}

\section{User Study}
\label{sec:study}

\noindent
Data exploration environments are complex systems that do not yield to simple controlled
experimentation~\cite{evaluating}. Rather than comparing our work with other tools, we
evaluate whether The Gamma can be successfully used by non-programmers.

\begin{figure}[t]
\centering
\vspace{-0.25em}
\includegraphics[width=0.95\columnwidth]{figures/cases-dr}
\vspace{-0.5em}
\caption{Who does the Dr Who fight most frequently?}
\label{fig:cases-dr}
\vspace{-1em}
\end{figure}

We performed a between-subjects study to assess whether non-programmers are able to complete a
simple data exploration task using The Gamma. We recruited 13 participants (5 male, 8 female)
from a business team of a research institute working in non-technical roles (project management,
communications). Only one participant (\#12) had prior programming experience.
%
We split participants into 4 groups and asked each group to complete a different task.
We gave participants a brief overview of The Gamma. The participants then worked for 30
minutes, after which we conducted a semi-structured group interview. We offered guidance
if participants were unable to progress for more than 5 minutes. The four tasks were:

\vspace{0.5em}
\begin{itemize}
\setlength\itemsep{0.5em}
\item \emph{Expenditure.} Participants were shown the \emph{worldbank} data cube
  and were asked to compare UK spending on `Public order and safety'' and ``Defence''
  using another data cube.
\item \emph{Lords.} Participants were shown \emph{worldbank} and were asked to use the
  \emph{expenses} data table provider to sort London House of Lords members by their
  travel costs.
\item \emph{Worldbank.} Participants were given a minimal iterative prompting demo and
  a code sample using \emph{worldbank}. They were asked to solve another \emph{worldbank} task.
\item \emph{Olympics.} Participants were given a demo using \emph{olympics} that did not involve grouping.
  They were asked to solve a problem involving grouping and aggregation.
\end{itemize}
\vspace{0.5em}

%
%Some aspects of the study aim to shed light on three questions concerning learnability
%of The Gamma, in particular whether knowledge can be trasferred between different data sources
%(\emph{RQ2a}), whether users can learn from code samples without a detailed guidance
%(\emph{RQ2b}) and to what extent they form a correct mental model of a more complex query
%language used in the tabular data source (\emph{RQ2c}).
%

%\vspace{0.5em}\noindent\emph{RQ2: Can knowledge transfer between data sources?}\hspace{0.3em}
%Our second hypothesis is that users familiar with the iterative prompting principle will be able to
%use an unfamiliar data source. We designed two of the four tasks to test this. Participants were
%shown a demo using one data source and then asked to complete a task using a different one.
%
%\vspace{0.5em}\noindent\emph{RQ3: Can users learn from just code samples?}\hspace{0.3em}
%Our third hypothesis is that users can learn through percolation, i.e.~by looking at the source
%code of published analyses. In one of our tasks, participants were given a minimal explanation
%of The Gamma (showing how to initiate the iterative prompting process) together with
%an extensive code sample.
%

\noindent
Our primary hypothesis was that non-programmers will be able to use The Gamma to explore
data. This was tested by all four tasks for one of the supported data sources.

The tasks \emph{expenditure} and \emph{lords} further test if knowledge can be transferred between
different data sources by using one sources in the introduction and another in the task;
\emph{worldbank} explores whether users can learn how to use a data source from just code samples;
and \emph{lords} lets us study to what extent participants form a correct mental model of the
more complex query language used in the tabular data source.

\vspace{0.5em}
\emph{Can non-programmers explore data with The Gamma?}~
All participants were able to complete, at least partially, a non-trivial data exploration task and
only half of them required further guidance. Participants spent 10--25 minutes (average 17)
working with The Gamma and 12 out of 13 completed the task; 6 required assistance, but 3 of those
faced the same issue related to operations taking arguments (discussed later).

A number of participants shared positive comments in the group interviews.
Participant \#3 noted that \emph{``this is actually pretty simple to use,''}
participant \#2 said that The Gamma alleviated their unease about code:
\emph{``for somebody who does not do coding or programming, this does not feel that daunting.''}
and participant \#5 suggested that the system could be used as an educational tool for teaching
critical thinking with data.

\vspace{0.5em}
\emph{How users learn The Gamma?}
There is some evidence that knowledge can be transferred between different data sources. In
\emph{expenditure} and \emph{lords}, participants were able to complete tasks after seeing a demo
using another data source. Participant \#2 \emph{``found it quite easy to translate what you showed
us in the demo to the new dataset.''}. However, the \emph{lords} task has been more challenging as
it involves a more complex data source.

There is also some evidence that, once a user understands iterative prompting, they can learn
from just code samples. All three participants were able to complete the \emph{worldbank} task,
where they were given printed code samples, but no demo using any data source. When discussing
suitable educational materials for The Gamma, participant \#7 also confirmed that
\emph{``a video would just be this [i.e.~a code sample] anyway''}.

\vspace{0.5em}
\emph{How users understand complex query languages?}
The tabular type provider uses a member \ikvd{then} to complete the specification of a
current operation, for example when specifying a list of aggregation operations. Two participants
(\#12 and \#13) initially thought that \ikvd{then} is used to split a command over multiple lines,
but rejected the idea after experimenting. Participant \#12 then correctly concluded that it
\emph{``allows us to chain together the operations''} of the query. While iterative prompting
allows users to start exploring new data sources, the structures exposed by more complex data
sources have their own further design principles that the users need to understand.

\vspace{0.5em}
\emph{What would make The Gamma easier to use?}
Three participants (\#11, \#12, \#13) struggled to complete a task using the tabular data source,
because they attempted to use operation that takes a numerical parameter and thus
violates the iterative prompting principle. This could be avoided by removing such operations
or by hiding them under an ``advanced'' tab.

The Gamma uses an ordinary text editor and most participants had no difficulty navigating around code,
making edits or deleting fragments, which is harder in a structure editor. Some
participants used the text editor effectively, e.g.~leveraging copy-and-paste. However, two
participants struggled with indentation and a syntax error in an unrelated command. This could
likely be alleviated through better error reporting.

% ==================================================================================================

\section{Discussion}
\label{sec:discuss}

\noindent
As a text-based programming environment for non-prog\-ram\-mers, The Gamma examines an unexplored point
in the design space of tools for data exploration. It has been particularly motivated by
the use of data in journalism. The Gamma has the potential to enable journalists
to make factual claims backed by data more commonplace and enable wider audience to engage with
such claims, satisfying the \emph{importance} criteria~\cite{evaluating} for advancing the
state of the art. It also satisfies a number of design goals important in the data journalism
context.

\vspace{0.5em}
\subsubsection*{Learning without experts}
Our design aims to make The Gamma suitable for users who cannot dedicate significant amount of time
to learning it in advance and may not have access to experts, satisfying the \emph{empowering new
participants} criteria~\cite{evaluating}. This is supported in two ways.

First, the iterative prompting principle makes it easy for users to start experimenting.
The user needs to select an initial data source and then repeatedly choose an
item from a list of choices. This is easier to use than a command line or a REPL
(read-eval-print-loop) interface, because it follows the \emph{recognition over recall} usability heuristic.
The users are not required to recall and type a command. They merely need to select one from a
list of options.

Second, the resulting code serves as a trace of how the analysis was
created. It provides the user with all information that they need to recreate the
program, not just by copying it, but also by using iterative prompting. Such design
has been called \emph{design for percolation}~\cite{learning} and it supports learnability.
In Excel, studied by Sarkar \cite{learning}, users learn new features when their usage is apparent
in a spreadsheet, e.g.~different functions in formulas, but learning how to use a wizard for
creating charts is hard because the operation does not leave a full trace in the spreadsheet.

\vspace{0.5em}
\subsubsection*{Lowering barriers to entry}

Data exploration has a certain irreducible essential complexity~\cite{silverbullet}. To make a system usable, this
complexity needs to be carefully stratified. The Gamma uses a two level structure. The first level
consists of the language itself with the iterative prompting mechanism. The second level consists of
the individual members generated by a type provider. This can be seen as a \emph{domain specific
language}, embedded in The Gamma language. Although the complexity of individual domain specific
languages differs, the user can always start exploring through iterative prompting, even when
faced with an unfamiliar data source.

In tackling complexity, The Gamma satisfies two criteria proposed by Olsen \cite{evaluating}:
\emph{generality} in that it can be used uniformly with a wide
range of data sources, and \emph{expressive leverage} in that it factors
out common aspects of different data queries into the core language (first level) and leaves the
specifics of each data source to the second level.

\vspace{0.5em}
\subsubsection*{Correctness and completeness}

An important characteristic of our design is that the iterative prompting mechanism is both \emph{correct}
and \emph{complete} with respect to possible data exploration scripts. The two properties are
a consequence of the fact that a program is a formed by a chain of operations and
that the auto-completion leverages a static type system. When invoking iterative prompting
at the end of a well-typed script, a selected option, which is a valid object member, is added to
the end of the script, resulting in another well-typed script. This distinguishes our system from
auto-completion based on machine learning, which may offer members not
valid in a given context. Auto-completion lists offered via iterative prompting contain
all available members and so the user can construct all possible scripts. Two exceptions to
completeness in our current design are the let binding and specifying numerical parameters as in
\ikvd{take(5)}.

\section{Conclusions}

\noindent
Exploring data in a programming environment that makes the full source code available increases
transparency, reproducibility and empowers users to ask critical questions about the data analysis.
But can we make those features accessible to non-programmers? In this paper, we presented The Gamma,
a simple data exploration environment for non-programmers that answers this question in the
affirmative.

The Gamma is based on a single interaction principle, \emph{iterative prompting}. It can be used to
complete a range of data exploration tasks using tabular data, data cubes and graph databases.
The design lowers the barrier to entry for programmatic data exploration and makes it easy to learn
the system independently through examples and by experimentation. We implemented The Gamma, make it
available as open source and conducted a user study, which lets us conclude that
The Gamma can be used by non-programmers to construct non-trivial data exploration scripts.

\section*{Acknowledgments}
\addcontentsline{toc}{section}{Acknowledgments}

We thank to May Yong and Nour Boulahcen for their contributions to The Gamma type providers.
The author is also grateful to Don Syme, James Geddes, Jonathan Edwards and Roly Perera for
numerous discussions about data science tooling and type providers, as well as Luke Church for
discussions about human-computer interaction and Clemens Klokmose for numerous suggestions on
framing of this paper. Anonymous reviewers of this and earlier versions of the paper also provided
valuable feedback. This work was partly supported by The Alan Turing Institute under the EPSRC
grant EP/N510129/1 and by a Google Digital News Initiative grant.

\IEEEtriggeratref{47}
\bibliographystyle{IEEEtran}
\bibliography{paper}

\end{document}

### 2022-06-10 wip-design/paper.tex
Edits, VL/HCC version

@@
 \begin{document}
-\title{Iterative Prompting: Programmatic Data Exploration for Non-programmers}
+\title{The Gamma: Programmatic Data Exploration for Non-programmers}
 
@@
 coding skills. Can we make data exploration based on code accessible to non-experts?
-  %
+
 We present The Gamma, a novel text-based data exploration environment that answers the question in
-  the affirmative. The Gamma is based on a novel interaction principle, \emph{iterative prompting},
-  which lets users create transparent and reproducible scripts without writing code. The Gamma lowers
-  the barrier to entry and learning from previously created data analyses. We evaluate the usability
-  of The Gamma through a user study on non-technical employees of a research institute.
-  %
+the affirmative. The Gamma is takes the code completion mechanism to the limit.
+By repeatedly choosing from offered completions, users can create transparent and reproducible
+scripts without writing code. We evaluate the usability of The Gamma through a user
+study on non-technical employees of a research institute.
+
 Our work shows that we may not need to shy away from code in order to build accessible,

### 2022-06-10 wip-design/studies/supplementary/source/graph-service/app.js

@@ -1,0 +1,109 @@
+var app = require('express')();
+var http = require('http').createServer(app);
+var api = require('./neo4j');
+const fs = require('fs');
+
+function ConvertToCSV(objArray,keys) {
+  var array = typeof objArray != 'object' ? JSON.parse(objArray) : objArray;
+  var str = '';
+  var lines = '';
+  for (var key in keys) {
+    lines += keys[key] +','
+  }
+  lines = lines.substring(0, lines.length - 1);
+  str += lines + '\r\n';
+  for (var i = 0; i < array.length; i++) {
+      var line = '';
+      for (var index in array[i]) {
+          if (line != '') line += ','
+          if (typeof(array[i][index]) == typeof("") && array[i][index].includes(",") ) {
+            line += array[i][index].replace(/,/g, ";"); // to fix the issue with comma in string, I replace all the comma by ;
+          }else {
+            line += array[i][index];
+          }
+
+      }
+
+        str += line + '\r\n';
+      }
+  return str;
+}
+
+
+app.use(function(req, res, next) {
+  res.header("Access-Control-Allow-Origin", "*");
+  res.header("Access-Control-Allow-Headers", "Origin, X-Requested-With, Content-Type, Accept");
+  next();
+});
+
+app.get('/drWho', function(req, res) {
+    api.All_nodes().then(resultJson => {
+      res.setHeader('Content-Type', 'text/plain');
+      res.end(resultJson);
+    });
+});
+
+app.get('/drWho/:trace/nodes_of_type/:type', function(req, res) {
+    api.nodes_of_type(req.params.type, req.params.trace).then(resultJson => {
+      res.setHeader('Content-Type', 'text/plain');
+      res.end(resultJson);
+    });
+});
+
+app.get('/drWho/:trace/links_from_node/:node_id', function(req, res) {
+  api.links_from_node(req.params.node_id,  req.params.trace).then(resultJson => {
+    res.setHeader('Content-Type', 'text/plain');
+    res.end(resultJson);
+  });
+});
+
+app.get('/drWho/:trace/links_from_any_node/:label_type', function(req, res) {
+    api.links_from_any_node(req.params.label_type, req.params.trace).then(resultJson => {
+      res.setHeader('Content-Type', 'text/plain');
+      res.end(resultJson);
+    });
+
+});
+
+app.get('/drWho/:trace/linked_from_node/:node_id/:relation', function(req, res) {
[…]

### 2022-06-10 wip-design/studies/supplementary/source/graph-service/neo4j.js

@@ -1,0 +1,462 @@
+const neo4j = require('neo4j-driver').v1;
+const fs = require('fs');
+const _ = require('underscore');
+
+var config = null;
+try {
+  config = JSON.parse(fs.readFileSync("config.json"))
+} catch {
+  config = {
+    NEO4J_PASSWORD: process.env.NEO4J_PASSWORD,
+    NEO4J_URL: process.env.NEO4J_URL,
+    NEO4J_USER: process.env.NEO4J_USER,
+    GRAPH_SERVICE: process.env.GRAPH_SERVICE,
+    GALLERY_SERVICE: process.env.GALLERY_SERVICE }
+}
+const driver = neo4j.driver(config.NEO4J_URL, neo4j.auth.basic(config.NEO4J_USER, config.NEO4J_PASSWORD));
+const session = driver.session();
+
+function constructDataQuery(traceString) {
+  const trace = traceString.split('&')
+  var conditions = "n1.label = $label";
+  var returns = "n1";
+  var query = "(n1)";
+  var args =
+    { label:trace[0],
+    }
+
+  var numNodes = trace.length/2;
+  for(var n = 1; n <= numNodes; n++)
+  {
+    var nodeName = trace[(n-1)*2+1];
+    if (n > 1)
+    {
+      returns += ", n" + n;
+      query += "-[r" + (n-1) + "]->"
+      query += "(n" + n + ")";
+
+      conditions += " AND type(r" + (n-1) + ") = $r" + (n-1) + "type"
+      args["r" + (n-1) + "type"] = trace[(n-1)*2];
+
+    }
+    if (nodeName != "[any]") {
+      conditions += " AND n" + n + ".name = $n" + n + "name"
+      args["n" + n + "name"] = nodeName;
+    }
+  }
+  return ['MATCH ' + query + ' WHERE ' + conditions + ' RETURN ' + returns, args];
+}
+
+
+
+function list_of_key_and_type(traceString){
+  var [query, args] = constructDataQuery(traceString)
+  var key = [];
+  var type = [];
+  const resultPromise = session.run(query, args);
+  return resultPromise.then(result => {
+    session.close();
+
+    for(var res of result.records) {
+      for(var i = 0; i < res._fields.length; i++) {
+        for(var prop of Object.keys(res._fields[i].properties)) {
+          key.push("node" + (i+1) + "." + prop);
+          type.push(typeof(res._fields[i].properties[prop]));
+
+        }
+      }
+    }
[…]

### 2022-06-10 wip-design/studies/supplementary/source/sample-web/server.js

@@ -1,0 +1,19 @@
+var express = require('express');
+var app = express();
+var port = process.env.port || 8089;
+
+app.use('/', express.static(__dirname + '/web'));
+var exposedNodeModules = [ 
+  '/node_modules/requirejs',
+  '/node_modules/monaco-editor',
+  '/node_modules/thegamma-script',
+  '/node_modules/babel-standalone' ];
+  
+for(var i=0; i<exposedNodeModules.length; i++) {
+  var subdir = exposedNodeModules[i];
+  app.use(subdir, express.static(__dirname + subdir));
+}  
+  
+app.listen(port);
+
+console.log('Listening on port: ', port);
